### 2015-09-30 HSF-TN/templates/template.tex
Add templates directory

\documentclass[12pt,a4paper]{article}

% Variables that controls behaviour
\usepackage{ifthen} % for conditional statements
\newboolean{pdflatex}
\setboolean{pdflatex}{true} % False for eps figures

\newboolean{inbibliography}
\setboolean{inbibliography}{false} %True once you enter the bibliography

\textheight=230mm
\textwidth=160mm
\oddsidemargin=7mm
\evensidemargin=-10mm
\topmargin=-10mm
\headsep=20mm
\columnsep=5mm
\addtolength{\belowcaptionskip}{0.5em}

\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}

\setlength{\hoffset}{-2cm}
\setlength{\voffset}{-2cm}

% Page defaults ...
\topmargin=0.5cm
\oddsidemargin=2.5cm
\textwidth=16cm
\textheight=22cm

% Don't chase after perfection
\raggedbottom
\sloppy

\usepackage{microtype}
\usepackage{lineno}    % Line numbering during drafting
\usepackage{xspace}    % Avoid problems with missing or double spaces after predefined symbold
\usepackage{caption}   % These three command get the figure and table captions automatically small
\renewcommand{\captionfont}{\small}
\renewcommand{\captionlabelfont}{\small}

%% Graphics
\usepackage{graphicx}  % to include figures (can also use other packages)
\usepackage{color}
\usepackage{colortbl}

%% Math
\usepackage{amsmath} % Adds a large collection of math symbols
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{upgreek} % Adds in support for greek letters in roman typeset

%% fix to allow peaceful coexistence of line numbering and
%% mathematical objects
%% http://www.latex-community.org/forum/viewtopic.php?f=5&t=163
%%
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname
end#1\endcsname
 \renewenvironment{#1}%
   {\linenomath\csname old#1\endcsname}%
   {\csname oldend#1\endcsname\endlinenomath}%
}
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}%
}
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
\patchBothAmsMathEnvironmentsForLineno{eqnarray}%
}

% Get hyperlinks to captions and in references.
% These do not work with revtex. Use "hypertext" as class option instead.
\usepackage{hyperref}    % Hyperlinks in references
\usepackage[all]{hypcap} % Internal hyperlinks to floats.

% Make this the last packages you include before the \begin{document}
\usepackage{cite} % Allows for ranges in citations
\usepackage{mciteplus}

\usepackage{longtable} % only for template

\begin{document}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\begin{titlepage}
\pagenumbering{roman}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2015-01 \\  % ID 
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{4.0cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
  HSF Technical Notes policy
\end{center}
}

\vspace*{2.0cm}

% Authors -------------------------------------------------
\begin{center}
A.~McNab$^1$
\bigskip\\
{\it\footnotesize
$ ^1$University of Manchester
}
\end{center}

\vspace{\fill}

% Abstract -----------------------------------------------
\begin{abstract}
  \noindent
  The note describes the HSF Technical Notes policy, the rationale behind
  the notes series, and further recommendations. 
  This is the first version of the policy for the HEP
  Software Foundation
  (HSF) Technical Notes series, and itself serves as an example technical note.

\end{abstract}

\vspace*{2.0cm}

\vspace{\fill}

{\footnotesize 
\centerline{\copyright~Named authors on behalf of the HSF, licence \href{http://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0}.}}
\vspace*{2mm}

\end{titlepage}

% empty page may follow the title page in long documents
%\newpage
%\setcounter{page}{2}
%\mbox{~}

\cleardoublepage

\renewcommand{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

%%%% Uncomment next 2 lines if desired
%\tableofcontents
%\cleardoublepage

\pagestyle{plain} % restore page numbers for the main text
\setcounter{page}{1}
\pagenumbering{arabic}

%% Uncomment during drafting and review.
%% Comment before a final submission.
\linenumbers

\section{Introduction}
\label{sec:Introduction}

This is the first version of the policy for the HEP Software Foundation
(HSF) Technical Notes series, and itself serves as an example technical note.

The technical notes series provides an early example of community level co-operation and 
interoperation that can be fostered by HSF.
It is hoped that groups of projects will co-operate on writing notes for
interfaces they provide or use, and HSF will offer support and publicity for
any meetings and workshops that result from these collaborative efforts.

The notes series provides archived, numbered documents that can be referenced in
software documentation and papers, and the notes themselves may also become the 
basis for papers associated with HSF member projects.

\section{Policies}
\label{sec:Policies}

\begin{enumerate}

\item HSF technical notes are informational, and they do not
constitute standards or recommendations.

\item If a note needs to describe the required and optional elements of an interface between systems,
then use of the conventional key words 
``MUST'', 
``MUST NOT'', 
``REQUIRED'', 
``SHALL'', 
``SHALL NOT'', 
``SHOULD'', 
``SHOULD NOT'', 
``RECOMMENDED'', 
``MAY'', and      
``OPTIONAL''
as described in RFC 2119\cite{SHOULD-MUST} is preferred.

\item HSF limits its editorial input to checking that the document is within the
scope of HSF, and has the required format, grammar, and spelling.

\item HSF avoids attempting to judge whether the subject matter of a proposed technical 
note is sufficiently significant or not. Acceptance of a technical note
does not constitute an endorsement of its contents or importance.

\item A simple format for notes will be provided as a LaTeX template, but technical
notes must be supplied by authors as PDF files. Any method for creating a PDF file in
a similar format is acceptable, such as Word or Pages as a subsitute for 
LaTeX.

\item The copyright of a technical note remains with its authors, but the note
must be made available with a Creative Commons Attribution license, 
CC-BY-4.0 or later.

\item Notes which have been produced as reports for other purposes may be 
accepted in PDF form as HSF technical notes provided the title page of the
submitted PDF file conforms
to the HSF format including the assigned HSF identifier, and the entire document
is made available under the above Creative Commons license.

\item An archive of technical notes will be made available on a website
managed by HSF.

\item Each technical note will be assigned an identifier of the format
HSF-TN-YYYY-NN when it is finally accepted and archived by HSF. Notes should
be referenced as 
{\it Author(s), HSF-TN-YYYY-NN ``The Title'' (HEP Software
Foundation) } rather than with a URL.

\item Once published, substantive revisions to the technical note will not 
be accepted. This is similar to the IETF RFC model\cite{IETF-RFC}.

\item Subsequent notes may be accepted
% from the same authors 
covering updates to the same subject matter 
% after one year, 
and a new note identifier will be issued. Projects are encouraged
to give version numbers to their interfaces to clarify the process of
documenting their interfaces as they evolve over time. HSF may prefix copies 
of older notes presented on the archive website with 
forward references to newer versions.

\end{enumerate}

\section{Procedure}
\label{sec:Procedure}

Draft technical notes should be submitted following the process published in
the Technical Notes section of the HSF website. 
The acceptance process will be managed by editors for the HSF Technical 
Notes series appointed by the HSF governing body (initially, by the HSF
interim Foundation Board). The series editors may co-opt
other people to act as additional editors for a given technical note.
The editorial process will include at least one week of consultation
on the proposed final draft within the HSF community.

\section{Recommendations}
\label{sec:Recommendations}

HSF working groups are asked to work towards producing technical notes which
record their outcomes.

Interested projects or groups of projects are encouraged to suggest 
technical notes documenting their interfaces and methods. It is likely that
many projects have documents like this already (e.g. as wiki pages, sections
of published papers or PhD theses) that can readily be reworked and
perhaps expanded into technical notes.

\section{Standards}
\label{sec:Standards}

If there is a need for HSF
to manage standards, then a procedure will be developed for endorsing
suitable technical notes as standards documents.

\section{Summary}
\label{sec:Summary}

This technical note sets out the rationale, policies, and recommendations
for further HSF technical notes.

%\section*{Acknowledgments}

%\section*{References}

\begin{thebibliography}{9}
% Use references in the format expected by JPCS (as used for CHEP proceedings)

\bibitem{SHOULD-MUST} S. Bradner, RFC2119 ``Key words for use in RFCs to Indicate Requirement Levels'' (Internet Engineering Task Force)

\bibitem{IETF-RFC} S. Bradner, RFC2026 ``The Internet Standards Process -- Revision 3'' (Internet Engineering Task Force)

\end{thebibliography}

%\addcontentsline{toc}{section}{References}

\end{document}

### 2015-09-30 HSF-TN/draft-2015-MJF/draft-HSF-TN-2015-MJF.tex
Import

\documentclass[12pt,a4paper]{article}

% Variables that controls behaviour
\usepackage{ifthen} % for conditional statements
\newboolean{pdflatex}
\setboolean{pdflatex}{true} % False for eps figures

\newboolean{inbibliography}
\setboolean{inbibliography}{false} %True once you enter the bibliography

\textheight=230mm
\textwidth=160mm
\oddsidemargin=7mm
\evensidemargin=-10mm
\topmargin=-10mm
\headsep=20mm
\columnsep=5mm
\addtolength{\belowcaptionskip}{0.5em}

\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}

\setlength{\hoffset}{-2cm}
\setlength{\voffset}{-2cm}

% Page defaults ...
\topmargin=0.5cm
\oddsidemargin=2.5cm
\textwidth=16cm
\textheight=22cm

% Don't chase after perfection
\raggedbottom
\sloppy

\usepackage{microtype}
\usepackage{lineno}    % Line numbering during drafting
\usepackage{xspace}    % Avoid problems with missing or double spaces after predefined symbold
\usepackage{caption}   % These three command get the figure and table captions automatically small
\renewcommand{\captionfont}{\small}
\renewcommand{\captionlabelfont}{\small}

%% Graphics
\usepackage{graphicx}  % to include figures (can also use other packages)
\usepackage{color}
\usepackage{colortbl}

%% Math
\usepackage{amsmath} % Adds a large collection of math symbols
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{upgreek} % Adds in support for greek letters in roman typeset

%% fix to allow peaceful coexistence of line numbering and
%% mathematical objects
%% http://www.latex-community.org/forum/viewtopic.php?f=5&t=163
%%
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname
end#1\endcsname
 \renewenvironment{#1}%
   {\linenomath\csname old#1\endcsname}%
   {\csname oldend#1\endcsname\endlinenomath}%
}
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}%
}
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
\patchBothAmsMathEnvironmentsForLineno{eqnarray}%
}

% Get hyperlinks to captions and in references.
% These do not work with revtex. Use "hypertext" as class option instead.
\usepackage{hyperref}    % Hyperlinks in references
\usepackage[all]{hypcap} % Internal hyperlinks to floats.

% Make this the last packages you include before the \begin{document}
\usepackage{cite} % Allows for ranges in citations
\usepackage{mciteplus}

\usepackage{longtable} % only for template

\begin{document}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\begin{titlepage}
\pagenumbering{roman}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2015-MJF \\  % ID 
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{4.0cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
  Machine/Job Features
\end{center}
}

\vspace*{2.0cm}

% Authors -------------------------------------------------
\begin{center}
A.~McNab$^1$, 
Stefan Roiser$^2$, Tony Cass$^2$, Ulrich Schwickerath$^2$, ...
\bigskip\\
{\it\footnotesize
$ ^1$University of Manchester \\
$ ^2$CERN
}
\end{center}

\vspace{\fill}

% Abstract -----------------------------------------------
\begin{abstract}
  \noindent

Within the HEPiX virtualization group and the WLCG MJF Task Force, a mechanism 
has been developed which gives 
access to detailed information about the current host and the current job.
This allows user
payloads to access meta information, independent of the current batch system,
to access information like the performance of the node or calculate the
remaining run time available to them.

\end{abstract}

\vspace*{2.0cm}

\vspace{\fill}

{\footnotesize 
\centerline{\copyright~Named authors on behalf of the HSF, licence \href{http://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0}.}}
\vspace*{2mm}

\end{titlepage}

% empty page may follow the title page in long documents
%\newpage
%\setcounter{page}{2}
%\mbox{~}

\cleardoublepage

\renewcommand{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

%%%% Uncomment next 2 lines if desired
%\tableofcontents
%\cleardoublepage

\pagestyle{plain} % restore page numbers for the main text
\setcounter{page}{1}
\pagenumbering{arabic}

%% Uncomment during drafting and review.
%% Comment before a final submission.
\linenumbers

\section{Introduction}
\label{sec:Introduction}

The proposed schema is made to be extensible so that it can be used to add
additional information. The purpose of this document is to define the
specifications and use case of this schema. It should be seen as the source
of information for the actual implementation of the required scripts by the
sites.

\section{Definitions}
\label{sec:Definitions}

The resource provider for batch systems is the owner of the worker nodes. 
When jobs are running within virtual machines, the entity that performs
system level configuration within the VM (typically with root access) acts
as the resource provider referred to in the rest of this document.

\subsection{Environment variables}
\label{sec:EnvironmentVariables}

For each job, two environment variables may be set, with the following
names:

\begin{tabular}{l l l}
Variable	& Contents		& Comments \\
\hline
MACHINEFEATURES	& Path to a directory or URL & Information about the execution host\\
JOBFEATURES	& Path to a directory or URL & Job specific information \\
\end{tabular}

These environment variables are the base interface for the user payload.
They must be set for the job environment by the resource provider.

In the
case of virtual machines on IaaS cloud platforms, the resource provider may
discover the values to set for the environment variables from the
machinefeatures and jobfeatures metadata keys provided by the cloud
infrastructure. These metadata keys should only be accessed once in the
lifetime of each virtual machine.

\subsection{Directories}
\label{sec:Directories}

The environment variables point to directories or URLs created by the resource
provider. Inside, the file name is the key, the contents are the values, so
that files can be referred to with expressions like
\$MACHINEFEATURES/shutdowntime . The directory name should not include the
trailing slash. These directories are either local directories in the
filesystem or sections of the URL space on an HTTP(S) server. The user
positively determines whether the files are to be opened locally or over
HTTP(S) by checking for a leading slash or the prefix http:// or https://
respectively. Typically this can achieved using library functions which can
transparently handle local files and remote URLs when opening files. The
files may be accessed multiple times to check for changes in value or in the
absence of caching by the user. The HTTP(S) server may provide HTTP cache
control and expiration information which the user may use to reduce the
number of queries.
The \$MACHINEFEATURES and \$JOBFEATURES directories contain files created by
the resource provider.
All files in the directories must be readable by both the user and the
resource provider services.

\section{Use cases}
\label{sec:Use cases}

Use cases to be covered are

\begin{tabular}{l l l p{5cm}}
Identifier & Actors & Pre-conditions & Scenario \\
\hline
1.	& user	& job starter script	& The job needs to calculate the remaining time it is allowed to run	 	  \\
2.	& user	& job starter scripts	& The job needs to know how long it was already running	 	  \\
3.	& user	& host setup		& The job wants to know the performance of the host in order to calculate the remaining time it will need to complete (for CPU intensive jobs)	 	  \\
4.	& site	& host setup		& A host needs to be drained. The payload needs to be informed of the planned shutdown time	 	  \\
5.	& site	& job starter script	& A multi-core user job on a non-exclusive node needs to know how many threads it is allowed to start. This is specifically of interest in a late-binding scenario where the pilot reserved the cores and the user payload needs to know about this.	  \\
6.	& site	& job starter script	& A user job wants to know how many job slots are allocated to the current job	  \\
7.	& site	& job starter script	& A user jobs wants to know the maximum amount of disk space it is allowed to use	  \\
8.	& site	& job starter script	& A user job wants to setup memory limits to protect itself from being killed by the batch system automatically	  \\
\end{tabular}

\section{Requirements}
\label{sec:Requirements}

\begin{itemize}
\item The proposed schema must be unique and leave no room for interpretation of the values provided.
\item For this reason, basic information is used which is well defined across sites.
\item Host and Job information can be both static (like the HS06 rating) and dynamic (eg shutdown time may be set at any time by the site).
% \item Job specific files will be owned by the user and possibly reside on a /tmp like area
\end{itemize}

\section{General explications}
\label{sec:GeneralExplications}

The implementation, that is the creation of the files and their contents, can
be highly site specific. A sample implementation can be done per batch
system in use, but it is understood that sites are allowed to change the
implementation, provided that the created numbers match the definitions
given in this document.

\section{Normalization and CPU factors}
\label{sec:Normalization}

At many sites batch resources consist of a mixture of different hardware
types which have different performance. When a user submits a job to a queue,
this queue typically sets limits on the CPU time and the wall clock time of
the job. If such a job ends up on a faster node, it will terminate quicker.
To avoid that jobs which run on slower nodes are terminated prematurely, CPU
and Wall clock times are usually scaled with a factor which depends on the
performance of the machine. This factor is called the CPU factor. A
reference machine has a CPU factor of 1. For such a machine, normalized and
real time values for CPU are the same. A CPU factor below 1 means that the
worker node is slower than a given reference. In this case normalized times
are larger than the real time values, and jobs are allowed to run longer in
order to terminate. A CPU factor above 1 means that the worker node is
faster than a given reference. In this case normalized times are smaller
than the real time values.

\section{List of requirements}
\label{sec:ListOfRequirements}

Job specific information which are:
\begin{itemize}
\item found in the directory pointed to by \$JOBFEATURES
\item owned by the user who is executing the original job. e.g. the batch
system job start script. In the case of pilots this would be the pilot user at the site.
\item created before the user job starts, eg during a job starter script.
\end{itemize}

\begin{tabular}{l l l p{2.5cm} p{5cm}}
Identifier & File Name (key) & Use cases & Value & (Optional) Comments \\
\hline
1.1	& cpufactor\_lrms		& 1,3	& Normalization factor as used by the batch system. & Can be site specific \\
1.2.1	& cpu\_limit\_secs\_lrms	& 1	& CPU limit in seconds, normalized & Divide by cpufactor\_lrms to retrieve the real time seconds. For multi-core jobs it's the total. \\
1.2.2	& cpu\_limit\_secs		& 1	& CPU limit in seconds, real time (not normalized) & For multi-core jobs it's the total. \\
1.3.1	& wall\_limit\_secs\_lrms	& 1	& Run time limit in seconds, normalized & Divide by cpufactor\_lrms to retrieve the real time seconds \\
1.3.2	& wall\_limit\_secs		& 1	& Run time limit in seconds, real time (not normalized) &  \\
1.4	& disk\_limit\_GB		& 7	& Scratch space limit in GB (if any) & If no quotas are used on a shared system, this corresponds to the full scratch space available to all jobs which run on the host. Counting is 1GB = 1000MB = 1000x1000kB \\
1.5	& jobstart\_secs		& 2	& Unix time stamp (in seconds) of the time when the job started in the batch farm. & This is what the batch system sees, not when the user payload started to work. \\
1.6	& mem\_limit\_MB		& 8	& Memory limit (if any) in MB. & Total memory. Count with 1000 not 1024, that is 4GB corresponds to 4000MB \\
1.7	& allocated\_CPU		& 5	& number of allocated cores to the current job & Allocated cores can be physical or logical \\
1.8	& shutdowntime\_job		& 1	& dynamic value, shutdown time as a UNIX time stamp (in seconds) & optional, if the file is missing no job shutdown is foreseen. The job needs to have finished all its processing when the shutdowntime has arrived \\
\end{tabular}

Host specific information which are:
\begin{itemize}
\item found in the directory pointed to by \$MACHINEFEATURES
\item readable by the user who is executing the original job. In the case of pilots this would be the pilot user at the site.
\item created before the user job starts
\end{itemize}

\begin{tabular}{l l l p{2.5cm} p{5cm}}
Identifier & File Name (key) & Use cases & Value & (Optional) Comments \\
\hline
2.1	& hs06			& 3	& HS06 rating of the full machine in it's current setup & Static value. HS06 is measured following the HEPiX recommendations. If Hyperthreading is enabled, the additional cores are treats as if they were full cores \\
2.2	& shutdowntime		& 4	& dynamic value, shutdowntime as a UNIX time stamp (in seconds) & Dynamic. If the file is missing, no shutdown is foreseen. The value is in real time, and must be in the future. Must be removed if the shutdowntime has arrived \\
2.3	& jobslots		& 6	& Number of job slots for the host & dynamic value, can change with batch reconfigurations \\
2.4	& phys\_cores		& 3	& number of physical cores &  \\
2.5	& log\_cores		& 3	& number of logical cores & can be zero if hyperthreading is off \\
2.6	& shutdown\_command	& 4	& path to a command on the machine & optional, only relevant for virtual machines. A command provided by the site which provides a hook for the user to properly destroy the virtual machine and unregister it \\
\end{tabular}

\clearpage

\section{Summary}
\label{sec:Summary}

MJF

%\section*{Acknowledgments}

%\section*{References}

%\begin{thebibliography}{9}
%% Use references in the format expected by JPCS (as used for CHEP proceedings)
%
%\bibitem{SHOULD-MUST} S. Bradner, RFC2119 ``Key words for use in RFCs to Indicate Requirement Levels'' (Internet Engineering Task Force)
%
%\bibitem{IETF-RFC} S. Bradner, RFC2026 ``The Internet Standards Process -- Revision 3'' (Internet Engineering Task Force)
%
%\end{thebibliography}

%\addcontentsline{toc}{section}{References}

\end{document}

### 2015-11-01 HSF-TN/draft-2015-NAM/draft-HSF-TN-2015-NAM.tex
add draft of naming conventions

\documentclass[12pt,a4paper]{article}

% Variables that controls behaviour
\usepackage{ifthen} % for conditional statements
\newboolean{pdflatex}
\setboolean{pdflatex}{true} % False for eps figures

\newboolean{inbibliography}
\setboolean{inbibliography}{false} %True once you enter the bibliography

\textheight=230mm
\textwidth=160mm
\oddsidemargin=7mm
\evensidemargin=-10mm
\topmargin=-10mm
\headsep=20mm
\columnsep=5mm
\addtolength{\belowcaptionskip}{0.5em}

\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}

\setlength{\hoffset}{-2cm}
\setlength{\voffset}{-2cm}

% Page defaults ...
\topmargin=0.5cm
\oddsidemargin=2.5cm
\textwidth=16cm
\textheight=22cm

% Don't chase after perfection
\raggedbottom
\sloppy

\usepackage{microtype}
\usepackage{lineno}    % Line numbering during drafting
\usepackage{xspace}    % Avoid problems with missing or double spaces after predefined symbold
\usepackage{caption}   % These three command get the figure and table captions automatically small
\renewcommand{\captionfont}{\small}
\renewcommand{\captionlabelfont}{\small}

%% Graphics
\usepackage{graphicx}  % to include figures (can also use other packages)
\usepackage{color}
\usepackage{colortbl}
\usepackage[table]{xcolor}

%% Math
\usepackage{amsmath} % Adds a large collection of math symbols
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{upgreek} % Adds in support for greek letters in roman typeset

%% fix to allow peaceful coexistence of line numbering and
%% mathematical objects
%% http://www.latex-community.org/forum/viewtopic.php?f=5&t=163
%%
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname
end#1\endcsname
 \renewenvironment{#1}%
   {\linenomath\csname old#1\endcsname}%
   {\csname oldend#1\endcsname\endlinenomath}%
}
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}%
}
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
\patchBothAmsMathEnvironmentsForLineno{eqnarray}%
}

% Get hyperlinks to captions and in references.
% These do not work with revtex. Use "hypertext" as class option instead.
\usepackage{hyperref}    % Hyperlinks in references
\usepackage[all]{hypcap} % Internal hyperlinks to floats.

% Make this the last packages you include before the \begin{document}
\usepackage{cite} % Allows for ranges in citations
\usepackage{mciteplus}

\usepackage{longtable} % only for template

\usepackage[colorinlistoftodos]{todonotes}

\date{\today}

\begin{document}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\begin{titlepage}
\pagenumbering{roman}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2015-XXX \\  % ID 
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{4.0cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
  HSF Platform Naming Conventions - A Proposal
\end{center}
}

\vspace*{2.0cm}

% Authors -------------------------------------------------
\begin{center}
B.~Hegner$^1$
\bigskip\\
{\it\footnotesize
$ ^1$CERN 
}
\end{center}

\vspace{\fill}

% Abstract -----------------------------------------------
\begin{abstract}
  \noindent
  The note describes a proposal for a common platform naming scheme for HEP and tools to automatize the platform identification. 
\end{abstract}

\vspace*{2.0cm}

\vspace{\fill}

{\footnotesize 
\centerline{\copyright~Named authors on behalf of the HSF, licence \href{http://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0}.}}
\vspace*{2mm}

\end{titlepage}

\pagestyle{plain} % restore page numbers for the main text
\setcounter{page}{1}
\pagenumbering{arabic}

\section{Rationale}

Multiple ways of denoting hardware platform vs. compiler vs. operating-system vs. build type combinations exist throughout the field of high energy physics. Quite often packages with different naming conventions have to be combined and a significant amount of time is spent to transform between these different conventions. Furthermore, this transformation is error prone and has to be updated from time to time when new hardware platforms or compiler implementations arrive.

\section{Proposal}

In the view of the fact, that naming conventions are just agreements we propose to adopt existing naming conventions wherever possible and useful.

We consider the following pieces as part of the convention. A build configuration is denoted by the scheme
\[architecture-OS-compiler-buildtype\]
where \emph{architecture}, \emph{OS}, \emph{compiler}, and \emph{buildtype} are described in the following. Furthermore, we propose a common tool that helps identifying and building these strings.

\subsection{Architecture}

Different OSes may call the same \emph{architecture} with different names. We propose to rely on the abstraction of the Python package \emph{platform}, namely \texttt{platform.machine()}. This translates to e.g. x86\_64 for current Intel and AMD CPUs. In cases where optimized builds for a certain processor generation are required and software is not runable on the generic architecture, the generic architecture should be replaced by \emph{architecture+instructionset}. 
There exist no stable conventions for the instruction set, thus we propose to create ad-hoc conventions.

\subsection{Operating System}
The operating system (\emph{OS}) is a combination of the name of the operating system itself and its major version, e.g. \emph{ubuntu15} or \emph{slc6} in case a canonical abbreviation exists.

\subsubsection{Linux}
The name and version are as given by Python's \texttt{platform.linux\_distribution()}, using the short name of the distribution, transformed to all lower-case. This leads to names like \emph{ubuntu, centos, fedora}.

\emph{nota bene: Due to mistakes in the distribution of Scientific Linux CERN 6, this convention identifies SLC as redhat. This is fixed in the tool mentioned further below.}

\subsubsection{MacOS X}
The name is set to \emph{macos} and the version as given by the first two parts of \texttt{platform.mac\_ver()}, e.g. yielding \emph{macos1010}.

\subsubsection{Windows}
The name is set to \emph{win} and the version as given by the first two parts of \texttt{platform.win32\_ver()}.

\subsection{Compiler}
The \emph{compiler} is a combination of compiler name and compiler version. We propose to use the self-given names of the compilers like \emph{gcc, clang, msvc, icc} and adding major, minor version, and patch version numbers to it, e.g. \emph{clang350} or \emph{gcc511}. In case the system compiler is being used, the compiler should be denoted as \emph{native}.

\subsection{Buildtype}
The \emph{buildtype} denotes whether there is a debug build, an optimized build, or any other special setting. We propose to use \emph{opt} for optimized and \emph{dbg} for debug builds. The buildtype can be used to specify further custom build options, e.g. the used C++ standard, leading to opt11 or dbg14.
In general, any special compile time setting can be tracked in this name component.

\subsection{Generic Cases}
In some cases the mentioned components do not influence the build-artifacts created or are not relevant. In those cases, the corresponding name component is replaced by the string \emph{all}. Examples are pure-Python packages that do not require compilation at install time.

\section{Examples}
\texttt{x86\_64-centos7-gcc511-opt}
\\
\texttt{x86\_64-slc6-clang350-dbg}

\section{Identification tool -- hsf\_get\_platform}
As to simplify the adoption of these conventions, we provide a minimal command line tool \emph{hsf\_get\_platform}, that is able to identify architecture, operating system, and compiler in a transparent way. It can be used to e.g. auto-derive everything in a given environment or to extract one part of the platform string
\\
\texttt{hsf\_get\_platform.py --buildtype opt $\rightarrow$ "x86\_64-ubuntu15-gcc49-opt"}
\\
\texttt{hsf\_get\_platform.py --get compiler $\rightarrow$ "gcc49"}.

\section{Possible Extensions to denote Platform Compatibility}
The proposed naming convention identifies a given platform. In many cases, different platforms, however, may behave the same or yield compatible binaries.

For example, \emph{slc6} is compatible with \emph{sl6} and \emph{redhat6}. Similarly, patch versions of compilers are \emph{usually} compatible with each other.

We provide an additional tool \texttt{hsf\_platform\_compatibility} to tell the compatibility of platforms. The logic and mapping of these compatibilities cannot be derived from first principles and thus has to be maintained actively. In the long view, the maintenance burden should be shared across the community. It can be used like follows
\\
\newline
\texttt{hsf\_platform\_compatibility.py x86\_64-slc5-gcc481-opt x86\_64-slc6-gcc481-opt}
\newline 
\texttt{$\rightarrow$ "OS'es slc5 and slc6 are incompatible"} + \emph{return code 1}

\section{Resources}
Both of the tools mentioned are available at \url{https://github.com/HEP-SF/tools}.

\section*{Acknowledgements}
We would like to thank Ben Couturier and Brett Viren for fruitful discussions and feedback on earlier drafts.

\end{document}

### 2015-11-03 HSF-TN/draft-2015-PKG/draft-HSF-TN-2015-PKG.tex
add packaging WG report draft

\documentclass[12pt,a4paper]{article}

% Variables that controls behaviour
\usepackage{ifthen} % for conditional statements
\newboolean{pdflatex}
\setboolean{pdflatex}{true} % False for eps figures

\newboolean{inbibliography}
\setboolean{inbibliography}{false} %True once you enter the bibliography
%https://4e49633d0951e732a4b4-e055a20cd140c2a2b51d085d456340d1.ssl.cf5.rackcdn.com/kyczzsrnjkkf/page/f3ecf59ad94a62d438bfcb00fcc1d00be6852995.jpeg

\textheight=230mm
\textwidth=160mm
\oddsidemargin=7mm
\evensidemargin=-10mm
\topmargin=-10mm
\headsep=20mm
\columnsep=5mm
\addtolength{\belowcaptionskip}{0.5em}

\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}

\setlength{\hoffset}{-2cm}
\setlength{\voffset}{-2cm}

% Page defaults ...
\topmargin=0.5cm
\oddsidemargin=2.5cm
\textwidth=16cm
\textheight=22cm

% Don't chase after perfection
\raggedbottom
\sloppy

\usepackage{microtype}
\usepackage{lineno}    % Line numbering during drafting
\usepackage{xspace}    % Avoid problems with missing or double spaces after predefined symbold
\usepackage{caption}   % These three command get the figure and table captions automatically small
\renewcommand{\captionfont}{\small}
\renewcommand{\captionlabelfont}{\small}

%% Graphics
\usepackage{graphicx}  % to include figures (can also use other packages)
\usepackage{color}
\usepackage{colortbl}
\usepackage[table]{xcolor}

%% Math
\usepackage{amsmath} % Adds a large collection of math symbols
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{upgreek} % Adds in support for greek letters in roman typeset

%% fix to allow peaceful coexistence of line numbering and
%% mathematical objects
%% http://www.latex-community.org/forum/viewtopic.php?f=5&t=163
%%
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname
end#1\endcsname
 \renewenvironment{#1}%
   {\linenomath\csname old#1\endcsname}%
   {\csname oldend#1\endcsname\endlinenomath}%
}
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}%
}
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
\patchBothAmsMathEnvironmentsForLineno{eqnarray}%
}

% Get hyperlinks to captions and in references.
% These do not work with revtex. Use "hypertext" as class option instead.
\usepackage{hyperref}    % Hyperlinks in references
\usepackage[all]{hypcap} % Internal hyperlinks to floats.

% Make this the last packages you include before the \begin{document}
\usepackage{cite} % Allows for ranges in citations
\usepackage{mciteplus}

\usepackage{longtable} % only for template

\usepackage[colorinlistoftodos]{todonotes}

\date{\today}

\begin{document}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\begin{titlepage}
\pagenumbering{roman}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2015-XXX \\  % ID 
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{4.0cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
  HSF Packaging Working Group Report
\end{center}
}

\vspace*{2.0cm}

% Authors -------------------------------------------------
% LSK: Note that people should add themselves when they start editing
\begin{center}
L.~Sexton-Kennedy$^1$, B.~Hegner$^2$
\bigskip\\
{\it\footnotesize
$ ^1$FNAL,$^2$CERN 
}
\end{center}

\vspace{\fill}

% Abstract -----------------------------------------------
\begin{abstract}
  \noindent
  The note describes the outcome of the discussions in the HSF Packaging Working Group. 
  It summarizes the discussion on existing configuration and build tools and the possibility to 
  converge on more common solutions.
\end{abstract}

\vspace*{2.0cm}

\vspace{\fill}

{\footnotesize 
\centerline{\copyright~Named authors on behalf of the HSF, licence \href{http://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0}.}}
\vspace*{2mm}

\end{titlepage}
\pagestyle{plain} % restore page numbers for the main text
\setcounter{page}{1}
\pagenumbering{arabic}

\section{Introduction}

Software development in high energy physics follows the paradigm of open-source software (OSS). Experiments as well as the theory community heavily rely on software being developed outside of the field. The number of such third party software (so-called \emph{externals}) used within a given context can easily reach over 100 interdependent software packages.\footnote{ A package in this context is a revision controlled collection of source files including files to provide an implementation of how to build the source into binary products. These can either be object libraries or executables.}
Creating a consistent and working stack out of 100s of packages, on a variety of platforms is a non-trivial task. Within the field multiple technical solutions exist to configure and build those stacks, in the following these will be referred to as \emph{build tools}.

Furthermore, quite often software has to be ported to new platforms and operating systems and subsequently patches to the individual externals need to be created. This is a manual and time consuming task, requiring a very special kind of expert knowledge.

None of this work is experiment specific and our working group agrees that this effort is being duplicated. The aim of the HSF packaging working group is to see whether a better synergy within the field on porting/patching and build tools is possible.

This document describes the discussions and findings of the working group.
\subsection{The Working Group}
The working group consisted of members of the HEP community from the energy and intensity frontiers as well as accelerator modeling. The group met in a series of meetings over the months of May and June~\cite{meetings}.
Additional discussions took place via the public \href{https://groups.google.com/forum/#!forum/hep-sf-packaging-wg}{packaging working group mailing list}.

\section{The HEP build and packaging tool landscape}

Software stacks are needed by many organizations most notably HEP experiments.  If a package is authored by the organization it is not considered an external package however it still will need build recipes. Any system that builds external packages into a coherent whole must also be capable of building the internal packages as well.  In the area of external packages it will be useful to distinguish between "foreign" and "domain" external packages. Domain packages are authored by groups within the field of HEP and therefor the HSF could influence the behavior of the domain packages. The foreign packages have to be dealt with as they exist. Examples of foreign packages include: gcc, fftw3, eigen and lapack. Examples of domain packages include clhep and ROOT. In the following a few of the solutions used within the community are summarized. For more details we refer to the project websites.

\subsection{Solution 1 -- Worch}
\emph{Worch}\cite{worch} is the build tool being proposed for the DUNE collaboration.  It has been used to build the software suite common to many liquid argon experiments, called larsoft, as well as the experiment specific code that builds on top of larsoft.

\subsection{Solution 2 -- LCGCMake}

\emph{LCGCMake}\cite{lcgcmake} is the build tool used for creating the LCG releases used by the LHC experiments ATLAS and LHCb. It is based on a set of CMake macros and each external package is described using one of these macros in a semi-declarative style.

\subsection{Solution 3 -- cmsBuild}

\emph{cmsBuild}\cite{cmsbuild} is the build tool used for making releases of CMSSW for CMS.  It orchestrates the building of the whole stack from the compiler on up on an as needed basis for any particular release/architecture combination.  It is based on rpm build, python scripts, and text files that specify dependencies, library names and other optional information.

\subsection{Solution 4 -- contractor} 
\emph{Contractor} is the build tool used by the accelerator modeling community and was developed for the Synergia project.  The project must support a wide diversity of platforms from laptops to super-computers.  Looking at the requirements of this community and the solutions they chose to meet them should inform the HSF community as we look forward to a time when many more diverse platforms will be need in HEP computing.  Contractor has no external dependencies and is implemented in Python.

\subsection{Solution 5 -- SciSoft (mrb/ups/cetbuildtools)}
MRB is an acronym for Multi-Repository-Build and is used by some of the intensity frontier experiments at FNAL. Unlike other tools in this list it requires ups (an acronym for Unix-Product-Support) to create an environment for it to operate in properly.  cetbuildtools is a set of scripts that drive cmake to build an interdependent set of packages.  The resulting packages are uploaded to a distribution server called SciSoft.

\section{Taxonomy of non-HEP tools}
\subsection{Solution 6 - Homebrew }
\emph{Homebrew}\cite{homebrew} is a MacOS specific build tool. Packages build instructions are declared by writing Ruby classes with a given set of attributes. The feature set of these declarations is similar to those of cmsBuild and lcgcmake. A project fork focusses on the addition of Linux as a supported platform.

\subsection{Solution 7 - Nix }
\emph{Nix}\cite{nix} is a cross-platform build tool. Package build instructions are written in the \emph{Nix expression language}.

\section{Tool Features and Requirements}
We compared the feature set of the tools described. In evaluating them we defined a set of criteria in which to measure the tools against each other.  One experiment's strict requirement may just be considered a feature for another experiment. This is why we only list the findings, but do not make an overall assessment at the end of the report. In the following we explain the comparison criteria and the situation for the individual tools.

\subsection{Supported Platforms and Environments}
The most basic criterion is the support for the various platforms and environments used in high-energy physics. There are three categories of environments:
\begin{enumerate}
\item \textbf{Linux} on physical or virtual machines
\item \textbf{MacOS X} on mainly physical machines
\item \textbf{Windows} on desktop machines
\end{enumerate}
The first category is split into several flavours of distributions. The main distribution are the RedHat derived Scientific Linux\cite{sl} and CentOS\cite{centos}.

In addition, multiple hardware architectures are in use or will be in use in the foreseeable future - x86, ARM, PowerPC, MIC, and various dedicated super-computers. Some of them impose the requirement of cross compilation onto the build and packaging solution, which is why the keyword \textbf{Xcompiler} is added to the platform table.
The table~\ref{tab:taxonomyPlatforms} summarizes this information for the tools considered by the working group so far. Both Linux and MacOS X are supported by all of the tools, while Windows is not supported at all. X-compilation does not seem a priority of the projects yet.

\begin{table}
\centering
\begin{tabular}{lcccc}
& Linux & MacOS X & Windows & Xcompiler \\ \hline
cmsBuild & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} +  \\
Contractor & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & ?\\
Homebrew & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{red} -\\
LCGCMake & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} o   & \cellcolor{yellow} o \\
Nix & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} o  & \cellcolor{yellow} o \\
SciSoft &  \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & ? \\
Worch &  \cellcolor{green} + &  \cellcolor{green}  + & \cellcolor{red} -  & ?
\end{tabular}
\caption{\label{tab:taxonomyPlatforms}Supported platforms of the different build tools.}
\end{table}

\subsection{Build and Installation Variants}
One important feature of build tools in HEP is the support for the installation of multiple stacks or package versions in parallel. 
\begin{enumerate}
\item \textbf{Multi-Rel}: The support for building and installing multiple stacks in parallel. This is important if multiple versions of experiment software need to coexist on a given installation.
\item \textbf{Multi-BuildVar}: The support for multiple versions of the same package within the same stack. There are many possible reasons why this maybe needed, what they have in common is the need to specify compiler switches that have to be applied consistently across the build, like a debugging option or a compiler dialect.
\item \textbf{MutiShell-RTE}: The support for setting up the runtime environment for a given stack version, supporting multiple shell flavours.
\item \textbf{Relocation}: Whether the build tool supports the relocation of packages or creates fully relocatable packages.
\end{enumerate}

The assessment of the tools according to these criteria are listed in Table~\ref{tab:taxonomyVariants}.

\begin{table}
\centering
\begin{tabular}{lcccc}
&Multi-Rel & Multi-BuildVar & MultiShell-RTE & Relocation \\ \hline
cmsBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} & \cellcolor{green} + \\
Contractor & \cellcolor{green} + & \cellcolor{green} + & ? & ? \\
Homebrew &  \cellcolor{red} - & \cellcolor{red} -   & \cellcolor{yellow} NA   & \cellcolor{red} -  \\
LCGCMake & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} +\\
Nix & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} +  & \cellcolor{red} -\\
SciSoft &  \cellcolor{green} + & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{green} + \\
Worch &  \cellcolor{green} + &  \cellcolor{green}  + & ? & ?
\end{tabular}
\caption{\label{tab:taxonomyVariants}Support for multiple releases and build variants. \emph{NA} denotes that the criteria are not-applicable for homebrew. As it only provides one variant it can rely on default system paths.}
\end{table}

\subsection{Ease of Install and Use}
The criteria in this category are:
\begin{enumerate}
\item \textbf{Depends-On}: the dependencies of the build tool itself.
\item \textbf{Ease-Add-Pkg}: the ease of adding another package to a stack.
\item \textbf{Ease-Bootstrap}: the ease of bootstrapping the build system itself. A prerequisite is the possibility to use it w/o root privileges. 
\item \textbf{Documentation}: existence and quality of documentation
\end{enumerate}

The findings are summarized in Table~\ref{tab:taxonomyUsability}. The dependencies of most of the tools seem well under control. The addition of new packages is of similar complexity in all cases. Only the ease of bootstrapping differs. For most of them it is a simple checkout. However, others like Nix an entire (fake-)root environment is necessary. The documentation of the domain specific packages is in general very poor, the documentation of the non-HEP tools surprisingly good and complete.

\begin{table}
\centering
\begin{tabular}{lcccc}
&Depends-On & Ease-Add-Pkg & Ease-Bootstrap & Documentation \\ \hline
cmsBuild & \cellcolor{yellow} Python,rpm,apt & \cellcolor{green} spec-file & \cellcolor{yellow} o  & \cellcolor{red} -\\
Contractor &  \cellcolor{green} Python & ? & ?  & \cellcolor{red} -\\
Homebrew & \cellcolor{green} Ruby & \cellcolor{green} Formula & \cellcolor{green} + & \cellcolor{green} + \\
LCGCMake & \cellcolor{green} Python, Cmake & \cellcolor{green} Cmake-macro & \cellcolor{green} + & \cellcolor{yellow} o\\
Nix & \cellcolor{green} Perl & \cellcolor{green} expression & \cellcolor{red} - & \cellcolor{green} +\\
SciSoft &  \cellcolor{green} Cmake & \cellcolor{green} Cmake-macro & \cellcolor{yellow} o & \cellcolor{red} -\\
Worch &   \cellcolor{green} Waf,Python &  \cellcolor{green}  Text-files & \cellcolor{green} + & \cellcolor{yellow} o\\
\end{tabular}
\caption{\label{tab:taxonomyUsability}Ease of use of the various build tools.}
\end{table}

\subsection{Other Criteria}
During the meetings proponents brought up other creiteria that are not so easy to group together.  Some might consider these requirements, other might disagree. For now we've just listed them as "other". The criteria in this category are:
\begin{enumerate}
\item \textbf{Performance}: The relevant metrics are a. length of time to build the entire stack, this includes support for parallel builds. b. length of time to incrementally build a developer defined subset of the stack. 
%BH: I merged Parallel + Performance again
\item \textbf{Sys-Reuse}: the ability of the build system to reuse parts of the system software that is being built if desired.
\item \textbf{Community}: whether the tool is used by a wider community or just by one collaboration 
\item \textbf{Unique-IDs}: A method of uniquely identifying a build product such that if it already exists, it does not have to be built again. 
  \item \textbf{VCS-Support} Integrated support for check-out of given tags or branches from the projects’ repositories directly.
\end{enumerate}
The findings are summarized in Table~\ref{tab:taxonomyOthers}. 
\begin{table}
\centering
\begin{tabular}{lcccccc}

&Performance & Sys-Reuse & Community & Unique-IDs & VCS-Support\\ \hline
cmsBuild & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{red} - & \cellcolor{green} +& \cellcolor{green} +  \\
Contractor & ? & ?  & ? & \cellcolor{red} - & ?\\
Homebrew & \cellcolor{green} +  & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} + \\
LCGCMake & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + \\
Nix & \cellcolor{green} + & \cellcolor{red} -  & ? & \cellcolor{green} + & ? \\
SciSoft & \cellcolor{yellow} o & ? & \cellcolor{green} + & ? & \cellcolor{green}+  \\
Worch &  ? &  \cellcolor{green} + & \cellcolor{red} -  & ? & \cellcolor{green} + 
\end{tabular}
\caption{\label{tab:taxonomyOthers}Further considered features of the build tools. Details in the text.}
\end{table}

\section{Sharing of porting/patching -- "Build recipes"}
At the core of all packaging systems sits the execution of a given set of build instructions for a particular package, so-called \emph{build recipes}. Framed by collaboration-specific pre- and post-processing steps. The behaviour of these build instructions may be platform dependent or they may be influenced by parameters like build type or paths to their dependencies.

A lot, probably even most of the work in the packaging infrastructures goes into porting software and preparing these \emph{recipes} for the various platform-dependency-compiler-version combinations. When comparing the (implicit) interfaces of the packaging systems towards their individual recipes, they are strikingly similar. Thus starting with sharing these recipes looks like a viable solution. Various options have been discussed on the HSF packaging github tracker~\cite{github:discussion}. A first implementation of this idea was done in the context of the ALICE experiment~\cite{alidist}.

\section{Summary}
Together with most of the software librarians of the HEP community we assessed the existing build and packaging tools. For this we identified 17 criteria, which are however of different importance to different users. The goal was to see whether there is a base for working on a future common build tool.

While in general of high quality, the non-HEP tools we looked at seem very weak on the support for multi-stack, multi-configuration setups. There are however more tools in the open-source community we did not investigate on yet.
Among the HEP specific tools examined so far, \emph{LCGCMake} currently seems the most suited candidate for generalization. If decided, it could serve as the tool for the \emph{HSF Reference Builds} of the projects participating in the HSF. A more ambitious goal would be finding a common tool for building experiment stacks.

In parallel to the assessment of existing build tools, we discussed the concept of shared \emph{build recipes}. As one o the next steps in the working group we will try to lay this out in more concrete.

\section*{Acknowledgements}
We would like to thank all people participating in the constructive and lively discussions. In particular Jim Amundson, Marco Clemencic, Giulio Eulisse, Ben Morgan, Pere Mato, Shahzad Muzaffar, and Brett Viren.

\begin{thebibliography}{9}
\bibitem{meetings} Indico agendas of \href{https://indico.cern.ch/event/373973/}{25.2.2015}, \href{https://indico.cern.ch/event/398344/}{2.6.2015}, \href{https://indico.cern.ch/event/400272/}{9.6.2015}, \href{https://indico.cern.ch/event/402229/}{16.6.2015}, \href{https://indico.cern.ch/event/403790/}{23.6.2015}

\bibitem{worch} Worch: \url{https://github.com/brettviren/worch}
\bibitem{lcgcmake} LCGCMake: \url{http://ph-dep-sft.web.cern.ch/document/using-lcgcmake}
\bibitem{cmsbuild} cmsBuild: \url{https://github.com/cmsbuild/cmsdist}
\bibitem{homebrew} Homebrew: \url{http://brew.sh/}
\bibitem{nix} Nix: \url{https://nixos.org/nix/}

\bibitem{sl} Scientific Linux: \url{https://www.scientificlinux.org/}
\bibitem{centos} CentOS: \url{https://www.centos.org/}
\bibitem{github:discussion} Packaging Protocol Discussion:  \url{https://github.com/HEP-SF/packaging/issues/1}
\bibitem{alidist} The alidist package: \url{https://github.com/alisw/alidist}.
\end{thebibliography}

\end{document}

### 2015-11-03 HSF-TN/draft-2015-PKG/draft-HSF-TN-2015-PKG.tex
including B. Viren's comments; see https://groups.google.com/d/msg/hep-sf-packaging-wg/MFUKV0TSfsI/uBlVkoEzBwAJ

@@
   The note describes the outcome of the discussions in the HSF Packaging Working Group. 
-  It summarizes the discussion on existing configuration and build tools and the possibility to 
-  converge on more common solutions.
+  It summarizes the discussion on existing configuration and build tools and the possibility to converge on more common solutions.
 \end{abstract}
@@
 Any system that builds external packages into a coherent whole must also be capable of building the internal packages as well.
-In the area of external packages it will be useful to distinguish between "foreign" and "domain" external packages.
+In the area of external packages it will be useful to distinguish between ``foreign'' and ``domain'' external packages.
 Domain packages are authored by groups within the field of HEP and therefor the HSF could influence the behavior of the domain packages.
@@
 Unlike other tools in this list it requires ups (an acronym for Unix-Product-Support) to create an environment for it to operate in properly.
-cetbuildtools is a set of scripts that drive cmake to build an interdependent set of packages.
+cetbuildtools is a set of scripts that drive CMake to build an interdependent set of packages.
 The resulting packages are uploaded to a distribution server called SciSoft.
@@
 The feature set of these declarations is similar to those of cmsBuild and lcgcmake.
-A project fork focusses on the addition of Linux as a supported platform.
+A project fork focuses on the addition of Linux as a supported platform.
 
@@
 In evaluating them we defined a set of criteria in which to measure the tools against each other.
-One experiment's strict requirement may just be considered a feature for another experiment.
-This is why we only list the findings, but do not make an overall assessment at the end of the report.
+However, one experiment's strict requirement may just be considered a feature for another experiment.
 In the following we explain the comparison criteria and the situation for the individual tools.
@@
 \begin{enumerate}
-\item \textbf{Linux} on physical or virtual machines
-\item \textbf{MacOS X} on mainly physical machines
-\item \textbf{Windows} on desktop machines
+\item \textbf{Linux}
+\item \textbf{MacOS X}
+\item \textbf{Windows}
 \end{enumerate}
-The first category is split into several flavours of distributions. The main distribution are the RedHat derived Scientific Linux\cite{sl} and CentOS\cite{centos}.
+The first category is split into several flavours of distributions.
+The main distribution are the RedHat derived Scientific Linux~\cite{sl} and CentOS~\cite{centos}.
+Compared to these, Debian~\cite{debian} and Debian based based distributions play a smaller role in computing centres.
+On desktops, the Debian based Ubuntu~\cite{ubuntu} seems rather popular.
 
@@
 SciSoft &  \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & ? \\
-Worch &  \cellcolor{green} + &  \cellcolor{green}  + & \cellcolor{red} -  & ?
+Worch &  \cellcolor{green} + &  \cellcolor{green}  + & \cellcolor{red} -  &  \cellcolor{yellow} o
 \end{tabular}
-\caption{\label{tab:taxonomyPlatforms}Supported platforms of the different build tools.}
+\caption{\label{tab:taxonomyPlatforms}Supported platforms of the different build tools. A yellow ``o'' denotes that the support is untested or not of production quality.}
 \end{table}
@@
 SciSoft &  \cellcolor{green} + & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{green} + \\
-Worch &  \cellcolor{green} + &  \cellcolor{green}  + & ? & ?
+Worch &  \cellcolor{green} + &  \cellcolor{green}  + &  \cellcolor{green} + &  \cellcolor{green} +
 \end{tabular}
@@
 Some might consider these requirements, other might disagree.
-For now we've just listed them as "other".
+For now we've just listed them as ``other''.
 The criteria in this category are:
@@
 SciSoft & \cellcolor{yellow} o & ? & \cellcolor{green} + & ? & \cellcolor{green}+  \\
-Worch &  ? &  \cellcolor{green} + & \cellcolor{red} -  & ? & \cellcolor{green} + 
+Worch & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{red} -  & \cellcolor{yellow} o & \cellcolor{green} + 
 \end{tabular}
@@
 
-\section{Sharing of porting/patching -- "Build recipes"}
+\section{Sharing of porting/patching -- ``Build recipes''}
 At the core of all packaging systems sits the execution of a given set of build instructions for a particular package, so-called \emph{build recipes}.
@@
 \bibitem{centos} CentOS: \url{https://www.centos.org/}
+\bibitem{debian} Debian: \url{https://www.debian.org/}
+\bibitem{ubuntu} Ubuntu: \url{https://www.ubuntu.com/}
 \bibitem{github:discussion} Packaging Protocol Discussion:  \url{https://github.com/HEP-SF/packaging/issues/1}

### 2015-11-09 HSF-TN/draft-2015-LIC/draft-HSF-TN-2015-LIC.tex
Initial version of licensing TN converted to LaTeX

\documentclass[12pt,a4paper]{article}

% Variables that controls behaviour
\usepackage{ifthen} % for conditional statements
\newboolean{pdflatex}
\setboolean{pdflatex}{true} % False for eps figures

\newboolean{inbibliography}
\setboolean{inbibliography}{false} %True once you enter the bibliography

\usepackage[utf8]{inputenc} % Support accentuated characters, e.g. for author names

\textheight=230mm
\textwidth=160mm
\oddsidemargin=7mm
\evensidemargin=-10mm
\topmargin=-10mm
\headsep=20mm
\columnsep=5mm
\addtolength{\belowcaptionskip}{0.5em}

\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}

\setlength{\hoffset}{-2cm}
\setlength{\voffset}{-2cm}

% Page defaults ...
\topmargin=0.5cm
\oddsidemargin=2.5cm
\textwidth=16cm
\textheight=22cm

% Don't chase after perfection
\raggedbottom
\sloppy

\usepackage{microtype}
\usepackage{lineno}    % Line numbering during drafting
\usepackage{xspace}    % Avoid problems with missing or double spaces after predefined symbold
\usepackage{caption}   % These three command get the figure and table captions automatically small
\renewcommand{\captionfont}{\small}
\renewcommand{\captionlabelfont}{\small}

%% Graphics
\usepackage{graphicx}  % to include figures (can also use other packages)
\usepackage{color}
\usepackage{colortbl}

%% Math
\usepackage{amsmath} % Adds a large collection of math symbols
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{upgreek} % Adds in support for greek letters in roman typeset

%% fix to allow peaceful coexistence of line numbering and
%% mathematical objects
%% http://www.latex-community.org/forum/viewtopic.php?f=5&t=163
%%
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname
end#1\endcsname
 \renewenvironment{#1}%
   {\linenomath\csname old#1\endcsname}%
   {\csname oldend#1\endcsname\endlinenomath}%
}
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}%
}
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
\patchBothAmsMathEnvironmentsForLineno{eqnarray}%
}

% Get hyperlinks to captions and in references.
% These do not work with revtex. Use "hypertext" as class option instead.
\usepackage{hyperref}    % Hyperlinks in references
\usepackage[all]{hypcap} % Internal hyperlinks to floats.

% Make this the last packages you include before the \begin{document}
\usepackage{cite} % Allows for ranges in citations
\usepackage{mciteplus}

\usepackage{longtable} % only for template

\usepackage{verbatim} % support for comments

\usepackage{enumerate} % enumerated list

\begin{document}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\begin{titlepage}
\pagenumbering{roman}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2015-01 \\  % ID 
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{4.0cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
 Software Licence Agreements\newline
 HSF Policy Guidelines
\end{center}
}

\vspace*{2.0cm}

% Authors -------------------------------------------------
\begin{center}
J.~Harvey$^1$
M.~Jouvin$^2$
A.~McNab$^3$
E.~Sexton-Kennedy$^4$
T.~Wenaus$^5$
\bigskip\\
{\it\footnotesize
$ ^1$CERN
$ ^2$Laboratoire de l’Accélérateur Linéaire (CNRS)
$ ^3$University of Manchester
$ ^4$Fermi National Accelerator Laboratory
$ ^5$Brookhaven National Laboratory  
}
\end{center}

\vspace{\fill}

% Abstract -----------------------------------------------
\begin{abstract}
  \noindent
 These guidelines were prepared by the HSF Startup Team in order to serve as an aid in establishing a Software Licence Agreement for software projects
 hosted by the HSF. The report contains background information on open source licences approved by the Open Source Initiative and concludes with a set of
 recommendations for choosing a licence and instructions for drafting text to include with the source code.

\end{abstract}

\vspace*{2.0cm}

\vspace{\fill}

{\footnotesize 
\centerline{\copyright~Named authors on behalf of the HSF, licence \href{http://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0}.}}
\vspace*{2mm}

\end{titlepage}

% empty page may follow the title page in long documents
%\newpage
%\setcounter{page}{2}
%\mbox{~}

\cleardoublepage

\renewcommand{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

%%%% Uncomment next 2 lines if desired
%\tableofcontents
%\cleardoublepage

\pagestyle{plain} % restore page numbers for the main text
\setcounter{page}{1}
\pagenumbering{arabic}

%% Uncomment during drafting and review.
%% Comment before a final submission.
\linenumbers

\section{Introduction}
\label{sec:Introduction}

A simple survey of software packages in common usage in HEP reveals a de facto widespread adoption of different software licences. Moreover policy documents on software licensing in HEP are difficult to find. In 2011 CERN setup a Task Force to provide recommendations for the licensing of software developed at CERN. The final report \cite{[1]} is worth reading as it contains much useful background information and reference material. The main recommendation states that, "Whenever possible, software owned in whole or in part by CERN should be made available as open-source software and that the open-source licences used for CERN-owned software should be widely used licences approved by the Open Source Initiative (OSI)". This approach is in broad agreement with the policies adopted at other laboratories and has therefore been taken as the starting point for the recommendations contained in this report.

The philosophy of openness is enshrined throughout our field as exemplified by making the results of our experimental and theoretical work generally available. The same approach is followed here in order to help achieve our goal of providing reliable and long-lived software products through collaborative open-source software development.  Open Source Software (OSS) is computer software with its source code made available with a licence in which the copyright holder provides the rights to study, change, and distribute the software to anyone and for any purpose. The goal of the guidelines described in this note is to allow HSF and its projects to distribute and build upon their respective work. In this regard, HSF follows the example of other leading open-source software endeavours, such as the Apache Software Foundation \cite{[2]}. This does not preclude the full rights of contributors (copyright owners) to use their original contributions for any other purpose outside HSF.

\section{Basic Terminology}
\label{sec:BasicTerminology}

We begin by defining some of the key terms as described in \cite{[3]}.
\begin{itemize}
\item Copyright is a legal right created by the law of a country that grants the creator of original work exclusive rights to its use and distribution, usually for a limited time. Copyright is a form of intellectual property, applicable to any expressed representation of a creative work. It is often shared among multiple authors, each of whom holds a set of rights to use or licence the work. These rights frequently include reproduction, control over derivative works, distribution, and "moral rights" such as attribution.

\item Public domain software is software that has been placed in the public domain. In other words there is absolutely no ownership such as copyright, trademark, or patent. Unlike other classes of licences, there are no restrictions as to what can be done with the software. The software can be modified, distributed, or sold even without any attribution. This is the simplest way to make open source software and allows people to share the program and their improvements, if they are so minded, but it also allows the program to be converted into proprietary software. They can make changes, many or few, and distribute the result as a proprietary product. People who receive the program in that modified form do not have the freedom that the original author gave them. 
A software licence is a legal instrument governing the use or redistribution of software. A typical software licence grants an end-user permission to use one or more copies of software in ways where such a use would otherwise potentially constitute copyright infringement of the software owner's exclusive rights under copyright law.

\item An open source software licence is a notice that grants the recipient of a piece of software extensive rights to modify and redistribute that software. Copyright law usually prohibits these actions, but the rights-holder (usually the author) of a piece of software can remove these restrictions by accompanying the software with a software licence that grants the recipient these rights. Software using such a licence can meet the conditions to be classed as open source software as conferred by the copyright holder. Open source licenses broadly divide into free software licenses and permissive software licenses.
Free software licences include "copyleft" provisions, which require all future versions to also be distributed with these freedoms. These are termed "restrictive" licences.

\item Permissive software licences do not impose these additional conditions and are usually just a grant of rights and a disclaimer of warranty, thus also allowing distributors to add restrictions for further recipients, or to produce an extended proprietary version of the software
All major open source software licences require that acknowledgement is given to authors of the software in documentation and/or at runtime. In an academic context these provisions can be useful in establishing the impact of a software project, and even when software released under a permissive licence is reused in a closed-source commercial product.

\end{itemize}

\section{Main License Types}
\label{sec:MainLicenseTypes}

\subsection{Copyleft License}
\label{sec:CopyleftLicense}

Copyleft is a general method for making a program free, and requiring all modified and extended versions of the program to be free as well. Copyleft says that anyone who redistributes the software, with or without changes, must pass along the freedom to further copy and change it. Copyleft guarantees that every user has freedom and provides an incentive for other programmers to add to free software. A good example is the GNU C++ compiler.

The spirit behind a Copyleft licence is the creation of an open community of users or developers where the licencees are encouraged not only to improve, correct, complement and integrate the software they receive but also to make available these enhancements to the entire community. The difference between copyleft and non-copyleft licences is that users cannot take the free software and turn it into proprietary software, thus preventing any member of this open community to depart from the principles of reciprocal contribution.

The Copyleft principles were laid down by Richard Stallman of the Free Software Foundation in 1985, which was at the inception of the OSS movement through the creation of the GNU project. Copyleft is a general concept, and therefore cannot be used directly; you can only use a specific implementation of the concept. In the GNU Project, for example, the specific distribution terms used for most software are contained in the GNU General Public Licence (GPL) \cite{[4]}. GPL version 3 was published in 2007 but many copyleft projects (eg Linux) have chosen to continue with GPL version 2.
GPL 2 and 3 software amount to 24% and 10% of open source software respectively [6].

\subsection{Weak Copyleft License}
\label{sec:WeakCopyleft}

These typically follow the same rules as the GPL except that the user may use, unmodified, the free software component in a larger program which is released under a licence different from the free licence. The chief consequence is that the user is not obliged to provide the full source code of its larger work under a copyleft licence.

The most widely used example of this type of licence is the GNU Lesser General Public Licence (LGPL). Licences such as LGPL target libraries of software, which are designed to be incorporated unchanged into larger programs. For example, the ROOT software project \cite{[5]} has adopted an LGPL licence.

LGPL is also frequently used for non-library software when there is a particular concern from the licensor that the obligation to release the source of a work incorporating unchanged the GPL-licensed software would seriously hamper its wide adoption. The most common case is when a free library's features are readily available for proprietary software through other alternative libraries. In that case, the library cannot give free software any particular advantage, so it is better to use the LGPL for that library. The LGPL licence is used for the GNU C library, for example, since using the GPL would have driven proprietary software developers to use one of the many other C libraries.

As with full copyleft licences which prevent modified versions from being distributed under a proprietary licence, weak copyleft licenses are intended to ensure the non-appropriation by third parties of the Open Source software. 
As of July 2013, the LGPL was used by 7% of all open source licenced projects \cite{[6]}.

\subsection{Permissive License}
\label{sec:PermissiveLicense}

These licences allow redistribution of the original or modified software and source code, including under a different licence.  Depending on the terms of the permissive licence,  the different licences may be proprietary licences or copyleft licences or other permissive licences.

The Apache Software License (ASL), initially from 1999 and currently at version 2.0, is one of the most widely used examples of a permissive licence. Like other open source software licences, the licence allows the user of the software the freedom to use the software for any purpose, to distribute it, to modify it, and to distribute modified versions of the software, under the terms of the licence, without concern for royalties.  The Apache Licence does not require a derivative work of the software, or modifications to the original, to be distributed using the same licence (unlike copyleft licences). The Apache Software Foundation and the Free Software Foundation agreed that the Apache Licence 2.0 is a free software licence, compatible with version 3 of the GPL licence, meaning that code under GPL version 3 and Apache Licence 2.0 can be combined, as long as the resulting software is licensed under GPL version 3.

Other well-known examples of widely used free software licences approved by the OSI include the MIT and BSD licences. The MIT licences from 1988 onwards permit reuse within proprietary software provided all copies of the licensed software include a copy of the MIT Licence terms and the copyright notice. Such proprietary software retains its proprietary nature even though it incorporates software under the MIT Licence. The licence is also GPL-compatible, meaning that the GPL permits combination and redistribution with software that uses the MIT Licence.

BSD licences from 1988 onwards are another family of permissive free software licences, imposing minimal restrictions on the redistribution of covered software. Two variants of the licence, the New BSD Licence/Modified BSD Licence (3-clause), and the Simplified BSD Licence/FreeBSD Licence (2-clause) have been verified as GPL-compatible free software licences by the Free Software Foundation, and have been vetted as open source licences by the Open Source Initiative.
As of July 2013, the ASL, BSD and MIT permissive licences accounted for 42% of all open source licensed projects [6].

\section{Specific Constraints}
\label{sec:SpecificConstraints}

\subsection{Changing the License}
\label{sec:LicenseChange}

The ability to change the license term of a project, including the right to dual-license it, is an exclusive right of copyright holders. Except when explicitly stated otherwise, copyright holders are all the people who contributed to the project. In large projects, after some time, it may make impossible to change the license used by a project.

Although this rule applies to any license, it is more a concern for copyleft licences as permissive licenses give anybody the right to fork the project with a new license. For this reason, some projects, when there is no risk (or a low risk) of appropriation of the work by a third party, prefer to use a permissive license in order to keep a greater flexibility to evolve (including restrict) the licence in the future. A well known example is Apache where a development community exists and where most people (including commercial vendors) contribute their modification back to the community even though this is not a legal requirement of the permissive Apache licence.

To avoid problems in changing licence, some projects or software foundations (like the Apache Software Foundation) have an explicit transfer of copyright to one single legal entity by each project contributor. This is the main alternative for project with copyleft licenses. As with any change related to licensing, it has to be decided early in the life of the project as it requires the agreement of all copyright holders.

Where a non-permissive licence is required to distribute software binaries or packages, one option is maintain the source code repository under a permissive licence but re-licence the software at distribution time under the required licence. This maintains flexibility about what licence to use in the future, but allows linking or repackaging with more restrictively-licensed open source software in the present.

\subsection{Collaboration Agreements}
\label{sec:CollaborationAgreements}

For software developed in collaboration between partners from different institutes consideration may be given to establishing a Collaboration Agreement. This should define the licence to be used for the jointly developed software and typically also describes other rules for governing the way decisions are taken e.g. rules for accepting new members and rules for managing the development life cycle of the product. Typically, it also identifies a 'prime distributor' that takes the role of managing and deploying new releases of the software. Transferring copyright to the prime distributor may also help ensure the software can be maintained over the full life-time of the project in situations where the original developer (i.e. owner) can no longer be contacted.

\subsection{Commercial Exploitation}
\label{sec:Commercial Exploitation}

Any software distributed under a given licence may also be distributed under one or more different licence(s). This is often referred to as dual or multiple licensing.  A frequent case of dual licensing is the public release of a programme under a Copyleft licence (such as GPL) and, contemporaneously, a bilateral agreement between the programme owner and a third party company for the commercial exploitation of the software.

In the case of permissive free software licences, as all permissions for appropriation have been given to any third party, and so commercial exploitation by dual licensing becomes unnecessary. Therefore, permissive licenses, such as the ASL, MIT and BSD licenses, are preferred by many companies because such licenses make it possible to use open-source software code without having to turn proprietary enhancements back over to the open source software community. These licenses encourage commercial adoption of open-source software because they make it possible for companies to profit from investing in enhancements made to existing open-source software solutions.

\section{Recommendations}
\label{sec:Recommendations}

HSF encourages all its members and partners to make available the software they develop as Open Source, unless forbidden due to external constraints such as collaborative agreement. Only open-source software can become HSF projects. The open-source licence(s) adopted should be widely used licences approved by the Open Source Initiative (OSI). It should not be necessary to create a new licence and using a unusual licence may hinder the redistribution of the software by third parties.

The exact licence chosen may depend on several factors but they should enable the following key points:
\begin{itemize}	
\item Make the software distributable by other projects through their natural software distribution channels. This should anticipate their need to distribute modified versions of the software to fix bugs downstream or address compatibility requirements. 
\item Make the software and its source code reusable by other HSF or open-source projects using the most widely used open-source licences, whether copyleft or permissive.
\item Build a community around the software project and maximize the contributions by the users back to the project.
\end{itemize}

The GNU and Apache projects have demonstrated that these goals can be achieved either with copyleft or permissive licence approaches. Both approaches have vocal supporters and no consensus has emerged in the last 30 years of open source software development.

For projects producing libraries and taking the copyleft route, LGPL should be preferred for program libraries when the goal is to allow wide and rapid adoptions by applications with different licenses.

Permissive licences are good candidates when adoption by commercial partners must be possible and that there is a risk that at a later stage it will be difficult to contact all the copyright holders to discuss dual licensing. This is sometimes a requirement in projects funded by governmental bodies. In the copyleft case, it may be necessary to require that the copyright of contributions are assigned to the project to achieve this.

Whatever the licence chosen, software must contain in the notice a statement acknowledging the copyright owner(s) and the licence chosen. See next section for examples.

In addition, the following points must be taken into consideration:
\begin{enumerate}
\item	When contributing to an existing project, release your modified versions under the same licence as the original work.  
\item	A licence should be assigned to tutorials, reference manuals and other large works of documentation. The GNU Free Documentation Licence (GFDL) \cite{[7]} is a strong copyleft licence for educational works, initially written for software manuals, and includes terms that specifically address common issues arising when those works are distributed or modified. Licences from the Creative Commons family are also gaining ground in this area and provide a viable alternative.\cite{[8]}
\end{enumerate}

\section{Examples}
\label{sec:Examples}

This section contains examples for specifying licence terms, based on real licenses from different HEP laboratories. You can use them as a source of inspiration but you need to customize them to your specific needs and local context.

The licence should contain a statement in the header of each source file acknowledging (i) the copyright of the owner(s) and (ii) the applicable licence.
\newline\verb?(i)	Copyright?

In the following we give some examples of Copyright statements that are used by CERN depending on whether the software is owned solely by CERN or by CERN and external partners:
\begin{itemize}
\item “© Copyright [year] CERN”
for software owned solely by a single institute, in this case CERN
\item “© Copyright [year] CERN [for the benefit of the [Name of appropriate group] Collaboration]”
for software developed by a collaboration but where ownership has been transferred to a single institute, in this case CERN
\item  “© Copyright [year] [names of all copyright holders]”
for software owned by partners in small collaborations
\item “© Copyright [year] Copyright Holders of [name of the collaboration or joint project]. See [https://link] for details of the Copyright Holders”
for software owned by partners in large collaborations
\end{itemize}

(ii)	Applicable licence

One of the following licence statements must be included (immediately following the copyright statement):

“This software is distributed under the terms of the GNU General Public Licence version 3 (GPL Version 3), copied verbatim in the file “COPYING” /copied verbatim below”.
For software distributed under the default GPLv3 licence
“This software is distributed under the terms of the GNU Lesser General Public Licence version 3 (LGPL Version 3), copied verbatim in the file “COPYING” /copied verbatim below”.
For software distributed under the LGPLv3 licence
“This software is distributed under the terms of the Apache version 2.0 licence, copied verbatim in the file “COPYING” /copied verbatim below”.
For software distributed under the Apache licence v2
“This software is distributed under the terms of the BSD-2-Clause licence, copied verbatim in the file “COPYING” /copied verbatim below”.
For software distributed under the BSD-2-Clause licence
“This software is distributed under the terms of the BSD-3-Clause licence, copied verbatim in the file “COPYING” /copied verbatim below”.
For software distributed under the BSD-3-Clause licence
“This software is distributed under the terms of the MIT licence, copied verbatim in the file “COPYING” /copied verbatim below”.
For software distributed under the MIT licence

The verbatim text of the licence should be copied either in a dedicated file which is part of the distribution (in this case the filename is COPYING) or directly below the licence statement.

The text of each licence to be copied verbatim for each of these licences can be found here [9,10,11,12,13,14].

\begin{thebibliography}{9}
\bibitem{[1]} Final Report from the task force on Open Source Software Licence at CERN: \url{http://indico.cern.ch/category/4252/}
\bibitem{[2]} The Apache Software Foundation: \url{http://www.apache.org}
\bibitem{[3]} \url{http://en.wikipedia.org/wiki/Software_licence}
\bibitem{[4]} GNU GENERAL PUBLIC LICENCE Version 3, 29 June 2007 \url{http://www.gnu.org/copyleft/gpl.html}
\bibitem{[5]} ROOT software terms and conditions: \url{https://root.cern.ch/root/License.html}
\bibitem{[6]} Top Open Source Licences, source BLACKDUCK, July 2015: \url{https://www.blackducksoftware.com/resources/data/top-20-open-source-licenses}
\bibitem{[7]} GNU Free Documentation Licence (GFDL): \url{http://www.gnu.org/licences/fdl.html}
\bibitem{[8]} Creative Commons licences: \url{http://creativecommons.org}
\bibitem{[9]} Text of GPL v3 licence, June 2007: \url{http://opensource.org/licenses/GPL-3.0} 
\bibitem{[10]} Text of LGPL v3 licence, June 2007: \url{http://opensource.org/licenses/LGPL-3.0}
\bibitem{[11]} Text of Apache 2.0 licence Jan 2004: \url{http://opensource.org/licenses/Apache-2.0}
\bibitem{[12]} Text of BSD-2-Clause licence: \url{http://opensource.org/licenses/BSD-2-Clause}
\bibitem{[13]} Text of BSD-3-Clause licence: \url{http://opensource.org/licenses/BSD-3-Clause}
\bibitem{[14]} Text of MIT licence: \url{http://opensource.org/licenses/MIT}
\end{thebibliography}

%\addcontentsline{toc}{section}{References}

\end{document}

### 2015-11-09 HSF-TN/draft-2015-LIC/draft-HSF-TN-2015-LIC.tex

@@
 \begin{itemize}
-\item Copyright is a legal right created by the law of a country that grants the creator of original work exclusive rights to its use and distribution, usually for a limited time.
+\item \textit{Copyright} is a legal right created by the law of a country that grants the creator of original work exclusive rights to its use and distribution, usually for a limited time.
 Copyright is a form of intellectual property, applicable to any expressed representation of a creative work.
@@
 
-\item Public domain software is software that has been placed in the public domain.
+\item \textit{Public domain software} is software that has been placed in the public domain.
 In other words there is absolutely no ownership such as copyright, trademark, or patent.
@@
 
-\item An open source software licence is a notice that grants the recipient of a piece of software extensive rights to modify and redistribute that software.
+\item An \textit{open source software licence} is a notice that grants the recipient of a piece of software extensive rights to modify and redistribute that software.
 Copyright law usually prohibits these actions, but the rights-holder (usually the author) of a piece of software can remove these restrictions by accompanying the software with a software licence that grants the recipient these rights.
@@
 
-\item Permissive software licences do not impose these additional conditions and are usually just a grant of rights and a disclaimer of warranty, thus also allowing distributors to add restrictions for further recipients, or to produce an extended proprietary version of the software
+\item \textit{Permissive software licences} do not impose these additional conditions and are usually just a grant of rights and a disclaimer of warranty, thus also allowing distributors to add restrictions for further recipients, or to produce an extended proprietary version of the software
 All major open source software licences require that acknowledgement is given to authors of the software in documentation and/or at runtime.
@@
 GPL version 3 was published in 2007 but many copyleft projects (eg Linux) have chosen to continue with GPL version 2.
-GPL 2 and 3 software amount to 24% and 10% of open source software respectively [6].
+GPL 2 and 3 software amount to 24\% and 10\% of open source software respectively [6].
 
@@
 As with full copyleft licences which prevent modified versions from being distributed under a proprietary licence, weak copyleft licenses are intended to ensure the non-appropriation by third parties of the Open Source software.
-As of July 2013, the LGPL was used by 7% of all open source licenced projects \cite{[6]}.
+As of July 2013, the LGPL was used by 7\% of all open source licenced projects \cite{[6]}.
 
@@
 Two variants of the licence, the New BSD Licence/Modified BSD Licence (3-clause), and the Simplified BSD Licence/FreeBSD Licence (2-clause) have been verified as GPL-compatible free software licences by the Free Software Foundation, and have been vetted as open source licences by the Open Source Initiative.
-As of July 2013, the ASL, BSD and MIT permissive licences accounted for 42% of all open source licensed projects [6].
+
+As of July 2013, the ASL, BSD and MIT permissive licences accounted for 42\% of all open source licensed projects [6].
 
@@
 
-The licence should contain a statement in the header of each source file acknowledging (i) the copyright of the owner(s) and (ii) the applicable licence.
+The licence should contain a statement in the header of each source file acknowledging the copyright of the owner(s) and the applicable licence.
 \newline\verb?(i)	Copyright?
+
+\subsection{Copyright}
+\label{sec:CopyrightExamples}
 
@@
 \begin{itemize}
-\item “© Copyright [year] CERN”
-for software owned solely by a single institute, in this case CERN
-\item “© Copyright [year] CERN [for the benefit of the [Name of appropriate group] Collaboration]”
-for software developed by a collaboration but where ownership has been transferred to a single institute, in this case CERN
-\item  “© Copyright [year] [names of all copyright holders]”
-for software owned by partners in small collaborations
-\item “© Copyright [year] Copyright Holders of [name of the collaboration or joint project]. See [https://link] for details of the Copyright Holders”
-for software owned by partners in large collaborations
+\item for software owned solely by a single institute, in this case CERN:
+\begin{quote}
+© Copyright [year] CERN
+\end{quote}
+\item for software developed by a collaboration but where ownership has been transferred to a single institute, in this case CERN:
+\begin{quote}
+© Copyright [year] CERN [for the benefit of the [Name of appropriate group] Collaboration]
+\end{quote}
+\item for software owned by partners in small collaborations:
+\begin{quote}
+© Copyright [year] [names of all copyright holders]
+\end{quote}
+\item for software owned by partners in large collaborations:
+\begin{quote}
+© Copyright [year] Copyright Holders of [name of the collaboration or joint project]. See [https://link] for details of the Copyright Holders
+\end{quote}
 \end{itemize}
 
-(ii)	Applicable licence
+\subsection{Applicable licence}
+\label{sec:ApplicableLicenceExamples}
 
-One of the following licence statements must be included (immediately following the copyright statement):
+One of the following licence statements must be included, immediately following the copyright statement, \textbf{and followed by the text of the relevant license as shown in the references}:
 
-“This software is distributed under the terms of the GNU General Public Licence version 3 (GPL Version 3), copied verbatim in the file “COPYING” /copied verbatim below”.
-For software distributed under the default GPLv3 licence
-“This software is distributed under the terms of the GNU Lesser General Public Licence version 3 (LGPL Version 3), copied verbatim in the file “COPYING” /copied verbatim below”.
-For software distributed under the LGPLv3 licence
-“This software is distributed under the terms of the Apache version 2.0 licence, copied verbatim in the file “COPYING” /copied verbatim below”.
-For software distributed under the Apache licence v2
-“This software is distributed under the terms of the BSD-2-Clause licence, copied verbatim in the file “COPYING” /copied verbatim below”.
-For software distributed under the BSD-2-Clause licence
-“This software is distributed under the terms of the BSD-3-Clause licence, copied verbatim in the file “COPYING” /copied verbatim below”.
-For software distributed under the BSD-3-Clause licence
-“This software is distributed under the terms of the MIT licence, copied verbatim in the file “COPYING” /copied verbatim below”.
-For software distributed under the MIT licence
+\begin{itemize}
+\item For software distributed under the default GPLv3 licence\cite{[9]}:
+\begin{quote}
+This software is distributed under the terms of the GNU General Public Licence version 3 (GPL Version 3).
+\end{quote}
+\item For software distributed under the LGPLv3 licence\cite{[10]}:
+\begin{quote}
+This software is distributed under the terms of the GNU Lesser General Public Licence version 3 (LGPL Version 3).
+\end{quote}
+\item For software distributed under the Apache licence v2\cite{[11]}:
+\begin{quote}
+This software is distributed under the terms of the Apache version 2.0 licence.
+\end{quote}
+\item For software distributed under the BSD-2-Clause licence\cite{[12]}:
+\begin{quote}
+This software is distributed under the terms of the BSD-2-Clause licence.
+\end{quote}
+\item For software distributed under the BSD-3-Clause licence\cite{[13]}:
+\begin{quote}
+This software is distributed under the terms of the BSD-3-Clause licence.
+\end{quote}
+\item For software distributed under the MIT licence\cite{[14]}:
+\begin{quote}
+This software is distributed under the terms of the MIT licence.
+\end{quote}
+\end{itemize}
 

### 2016-02-04 HSF-TN/draft-2016-MJF/draft-HSF-TN-2016-MJF.tex
Import final version from task force

\documentclass[12pt,a4paper]{article}

% Variables that controls behaviour
\usepackage{ifthen} % for conditional statements
\newboolean{pdflatex}
\setboolean{pdflatex}{true} % False for eps figures

\newboolean{inbibliography}
\setboolean{inbibliography}{false} %True once you enter the bibliography

\textheight=230mm
\textwidth=160mm
\oddsidemargin=7mm
\evensidemargin=-10mm
\topmargin=-10mm
\headsep=20mm
\columnsep=5mm
\addtolength{\belowcaptionskip}{0.5em}

\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}

\setlength{\hoffset}{-2cm}
\setlength{\voffset}{-2cm}

% Page defaults ...
\topmargin=0.5cm
\oddsidemargin=2.5cm
\textwidth=16cm
\textheight=22cm

% Don't chase after perfection
\raggedbottom
\sloppy

\usepackage{microtype}
\usepackage{lineno}    % Line numbering during drafting
\usepackage{xspace}    % Avoid problems with missing or double spaces after predefined symbold
\usepackage{caption}   % These three command get the figure and table captions automatically small
\renewcommand{\captionfont}{\small}
\renewcommand{\captionlabelfont}{\small}

%% Graphics
\usepackage{graphicx}  % to include figures (can also use other packages)
\usepackage{color}
\usepackage{colortbl}

%% Math
\usepackage{amsmath} % Adds a large collection of math symbols
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{upgreek} % Adds in support for greek letters in roman typeset

%% fix to allow peaceful coexistence of line numbering and
%% mathematical objects
%% http://www.latex-community.org/forum/viewtopic.php?f=5&t=163
%%
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname
end#1\endcsname
 \renewenvironment{#1}%
   {\linenomath\csname old#1\endcsname}%
   {\csname oldend#1\endcsname\endlinenomath}%
}
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}%
}
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
\patchBothAmsMathEnvironmentsForLineno{eqnarray}%
}

% Get hyperlinks to captions and in references.
% These do not work with revtex. Use "hypertext" as class option instead.
\usepackage{hyperref}    % Hyperlinks in references
\usepackage[all]{hypcap} % Internal hyperlinks to floats.

% Make this the last packages you include before the \begin{document}
\usepackage{cite} % Allows for ranges in citations
\usepackage{mciteplus}

\usepackage{longtable} % only for template

\begin{document}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\begin{titlepage}
\pagenumbering{roman}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2016-MJF \\  % ID 
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{4.0cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
  Machine/Job Features
\end{center}
}

\vspace*{2.0cm}

% Authors -------------------------------------------------
\begin{center}
M.~Alef$^1$,
T.~Cass$^2$,
J.J.~Keijser$^3$,
A.~McNab$^4$,
S.~Roiser$^2$,
U.~Schwickerath$^2$,
I.~Sfiligoi$^5$
\bigskip\\
{\it\footnotesize
$ ^1$Karlsruhe Institute of Technology \\
$ ^2$CERN \\
$ ^3$NIKHEF \\
$ ^4$University of Manchester \\
$ ^5$Fermi National Accelerator Laboratory \\
}
\end{center}

\vspace{\fill}

% Abstract -----------------------------------------------
\begin{abstract}
  \noindent

Within the HEPiX virtualization group and the 
WLCG MJF Task Force, a mechanism 
has been developed which provides
access to detailed information about the current host and the current job to
the job itself. This allows user
payloads to access meta information, independent of the current batch
system or virtual machine model. This
information includes the performance of the node and the
remaining run time for the current job.

\end{abstract}

\vspace*{2.0cm}

\vspace{\fill}

{\footnotesize 
\centerline{\copyright~Named authors on behalf of the HSF, licence \href{http://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0}.}}
\vspace*{2mm}

\end{titlepage}

% empty page may follow the title page in long documents
%\newpage
%\setcounter{page}{2}
%\mbox{~}

\cleardoublepage

\renewcommand{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

%%%% Uncomment next 2 lines if desired
%\tableofcontents
%\cleardoublepage

\pagestyle{plain} % restore page numbers for the main text
\setcounter{page}{1}
\pagenumbering{arabic}

%% Uncomment during drafting and review.
%% Comment before a final submission.
%\linenumbers

\section{Introduction}
\label{sec:Introduction}

Within the HEPiX virtualization group\cite{HEPIXMJF} and the 
WLCG MJF Task Force\cite{WLCGMJF}, a mechanism 
has been developed which provides
access to detailed information about the current host and the current job to
the job itself. This allows user
payloads to access meta information, independent of the current batch
system or virtual machine model.

The proposed schema is made to be extensible so that 
additional information can be added. The purpose of this note is to define the
specifications and use case of this schema. It should be seen as the source
of information for the actual implementation of the scripts required by the
sites to provide it.

\section{Aims}
\label{sec:Aims}

\begin{itemize}
\item The proposed schema must be unique and leave no room for interpretation of the values provided.
      For this reason, basic information is used which is well defined across sites.
\item Host and job information can be both static (like the HS06\cite{HEPIXHS06} rating of the 
      hardware) and dynamic (eg shutdown time may be set at any time by the site.)
\item Job specific files will be readable and possibly owned by the user and
      residing on a /tmp like area
\item The implementation, that is the creation of the files and their contents, can
      be highly site specific. A sample implementation can be done per batch
      system in use, but it is understood that sites are allowed to change the
      implementation, provided that the created numbers match the definitions
      given in this note.
\end{itemize}

\section{Use cases}
\label{sec:Use cases}

The use cases considered in developing the protocols included:

\begin{enumerate}
\item The job needs to calculate the remaining time it is allowed to run.
\item The job needs to know how long it was already running.
\item The job wants to know the performance of the processors allocated to it in order to calculate the remaining time it will need to complete (for CPU intensive jobs).
\item A host needs to be drained, and the payload needs to be informed of the planned shutdown time.
\item A multiprocessor user job on a non-exclusive node needs to know how many threads or processes it is allowed to start. This
      especially useful in a late-binding scenario where the pilot reserved the processors and the user payload needs to discover this.
\item A user job wants to know how many processors are allocated to the current job.
\item A user job wants to know the maximum amount of scratch disk it is allowed to use.
\item A user job wants to set up memory limits to protect itself from being killed by the batch system automatically.
\end{enumerate}

\section{Definitions}
\label{sec:Definitions}

On VM-based systems, references to ``jobs'' are to be interpreted
as ``virtual machines'' and ``machines'' as ``hypervisors''.

When jobs are running within virtual machines, the entity that provides 
the system level configuration or contextualization of the VM acts
as the resource provider referred to in the rest of this note.

\section{Environment variables}
\label{sec:EnvironmentVariables}

For each job, two environment variables may be set, with the names
\$MACHINEFEATURES and \$JOBFEATURES.

These environment variables are the base interface for the user payload.
Their values must be provided for the job by the resource provider.

In the
case of virtual machines on IaaS cloud platforms, the virtual machine
may discover the values to set for the environment variables from 
``machinefeatures'' and ``jobfeatures'' metadata keys provided by resource
provider via the cloud
infrastructure. These metadata keys should only be accessed once in the
lifetime of each virtual machine. Alternatively, the values to set may
be supplied as part of the contextualization of the virtual machines.

\section{Directories}
\label{sec:Directories}

The environment variables point to directories created by the resource
provider. Inside, the file name is the key, the contents are the values, so
that files can be referred to with expressions like
\$MACHINEFEATURES/shutdowntime . The directory name should not include the
trailing slash. These directories are either local directories in the
filesystem or sections of the URL space on an HTTP(S) server. The user
positively determines whether the files are to be opened locally or over
HTTP(S) by checking for a leading slash or the prefix http:// or https://
respectively. Typically this can achieved using library functions which can
transparently handle local files and remote URLs when opening files.

Unlike metadata keys, the key/value files may be
accessed multiple times to check for changes in value or in the
absence of caching by the user. An HTTP(S) server may provide HTTP cache
control and expiration information which the user may use to reduce the
number of queries.
All files in the directories must be readable by both the user and the
resource provider services, and have file names which only consist of
lowercase letters, numbers, and underscores.

\section{\$MACHINEFEATURES}
\label{sec:MACHINEFEATURES}

Host-specific key/value pairs which are all:

\begin{itemize}
\item Found in the directory pointed to by \$MACHINEFEATURES
\item Readable by the user who is executing the original job. In the case
      of pilots this would be the pilot user at the site.
\item Required unless the resource provider cannot determine their value.
\item Static unless otherwise stated.
\end{itemize}

\begin{description}

\item[total\_cpu] Number of processors which may be allocated to
jobs. Typically the number of processors seen by the operating system on one
worker node (that is the number of ``processor :'' lines in /proc/cpuinfo on
Linux), but potentially set to more or less than this for performance
reasons. (Use case 3.)

\item[hs06] Total HS06 rating of the full machine in its current
setup. HS06 is measured following the HEPiX recommendations\cite{HEPIXHS06}, with 
HS06 benchmarks run in parallel, one for each processor which may be 
allocated to jobs. (Use case 3.)

\item[shutdowntime] Shutdown time for the machine as a UNIX time stamp 
in seconds. The value is dynamic and optional. If the file is missing, 
no shutdown is foreseen.  (Use case 4.)

\item[grace\_secs] If the resource provider announces a
shutdown time to the jobs on this host, that time will not be less than 
grace\_secs seconds after the moment the shutdown time is set. This 
allows jobs to begin packages of work knowing that there will be sufficient 
time for them to be completed even if a shutdown time is announced. 
This value is required if a shutdown time will be set or changed
which will affect any jobs which have already started on this host.

\end{description}

\section{\$JOBFEATURES}
\label{sec:JOBFEATURES}

Job specific key/value pairs which are all:

\begin{itemize}
\item Found in the directory pointed to by \$JOBFEATURES
\item Readable and possibly owned by the user who is executing the original
      job. In the case of pilots this would be the pilot user at the site.
\item Required unless the resource provider cannot determine their value.
\item Created before the job starts and static unless otherwise stated, or 
      unless the batch system has a recognised way of changing the
      parameters of the job in a way the job is guaranteed to be aware
      of. For example, if there is a mechanism for a job to
      release processors, then the resource provider may update
      allocated\_cpu when this happens.
\end{itemize}

\begin{description}

\item[allocated\_cpu] Number of processors allocated to the current job. 
(Use case~5.)

\item[hs06\_job] Total HS06 rating for the processors allocated
to this job. The job's share is calculated by the resource provider from 
per-processor HS06 measurements made for the machine. (Use case~3.)

\item[shutdowntime\_job] Dynamic value. Shutdown time as a UNIX time 
stamp in seconds. If the file is missing no job shutdown
is foreseen. The job needs to have finished all of its processing when
the shutdown time has arrived. (Use case~1.)

\item[grace\_secs\_job] If the resource provider announces a
shutdowntime\_job to the job, it will not be less than grace\_secs\_job
seconds after the moment the shutdown time is set. This allows jobs to
begin packages of work knowing that there will be sufficient time for
them to be completed even if a shutdown time is announced.
This value is static and required if a shutdown time will be set or changed
after the job has started.

\item[jobstart\_secs] UNIX time stamp in seconds of the time when the job 
started on the worker node. For a pilot job scenario, this is when the 
batch system started the pilot job, not when the user payload started to 
run. (Use case~2.)

\item[job\_id] A string of printable non-whitespace ASCII characters
used by the resource provider to
identify the job at the site. In batch environments, this should simply be
the job ID. In virtualized environments, job\_id will typically contain the
UUID of the VM.

\item[wall\_limit\_secs] Elapsed time limit in seconds, starting from
jobstart\_secs. This is not scaled up for multiprocessor jobs.
(Use case~1.)

\item[cpu\_limit\_secs] CPU time limit in seconds. For multiprocessor jobs
this is the total for all processes started by the job. (Use case~1.)

\item[max\_rss\_bytes] Resident memory usage limit, if any, in bytes
for all processes started by this job. (Use case~8.)

\item[max\_swap\_bytes] Swap limit, if any, in bytes for all processes
started by this job. (Use case~8)

\item[scratch\_limit\_bytes] Scratch space limit if any. If no quotas
are used on a shared system, this corresponds to the full scratch 
space available to all jobs which run on the host. User jobs from
EGI-registered VOs expect the ``max size of scratch space used by jobs''
value on their VO ID Card\cite{VOIDCARD} to be available to each job in the worst case.
If there is a recognised procedure for informing the job of the location
of the scratch space (eg EGI's \$TMPDIR policy\cite{EGITMPDIR}), then this value refers to that
space. (Use case~7.)

\end{description}

\clearpage

\section{Summary}
\label{sec:Summary}

This note describes how the \$MACHINEFEATURES and \$JOBFEATURES variables
may be set and used by jobs to obtain meta information from resource
providers in a uniform way across different batch and virtual machine
systems.
\vspace{0.5cm}

The following key/value pairs have been defined:

\begin{center}
\begin{tabular}{|l|l|p{5cm}}
\hline
\$MACHINEFEATURES & \$JOBFEATURES \\
\hline
total\_cpu & allocated\_cpu \\
hs06 & hs06\_job \\
shutdowntime & shutdowntime\_job \\
grace\_secs & grace\_secs\_job \\
 & jobstart\_secs \\
 & job\_id \\
 & wall\_limit\_secs \\
 & cpu\_limit\_secs \\
 & max\_rss\_bytes \\
 & max\_swap\_bytes \\
 & scratch\_limit\_bytes \\
\hline
\end{tabular}
\end{center}

%\section*{Acknowledgments}

%\section*{References}

\begin{thebibliography}{9}
% Use references in the format expected by JPCS (as used for CHEP proceedings)

\bibitem{HEPIXMJF} T. Cass, ``Environmental Information on WN'', Grid
Deployment Board, CERN, 13 June 2012, retrieved from
https://indico.cern.ch/event/155069/

\bibitem{WLCGMJF} ``Machine / Job Features Task Force'', \\
 https://twiki.cern.ch/twiki/bin/view/LCG/MachineJobFeatures

\bibitem{HEPIXHS06} ``HEP-SPEC06 Benchmark'', https://w3.hepix.org/benchmarks/

\bibitem{VOIDCARD} ``The VO ID Card system'', http://operations-portal.egi.eu/vo/help

\bibitem{EGITMPDIR} P. Solagna, ``EGI policy for the TMPDIR environment variable usage'',
 EGI Document 1119, retrieved from https://documents.egi.eu/public/ShowDocument?docid=1119

%\bibitem{SHOULD-MUST} S. Bradner, RFC2119 ``Key words for use in RFCs to Indicate Requirement Levels'' (Internet Engineering Task Force)

%\bibitem{IETF-RFC} S. Bradner, RFC2026 ``The Internet Standards Process -- Revision 3'' (Internet Engineering Task Force)

\end{thebibliography}

%\addcontentsline{toc}{section}{References}

\end{document}

### 2016-02-11 HSF-TN/2016-01/HSF-TN-2016-01.tex
"Trivial changes" described in https://github.com/HEP-SF/documents/issues/4#issuecomment-183078367

--- 2015-11-09 HSF-TN/draft-2015-LIC/draft-HSF-TN-2015-LIC.tex
@@
 \\
- & & HSF-TN-2015-01 \\  % ID 
- & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
+ & & HSF-TN-2016-01 \\  % ID 
+ & & 11 February 2016 \\ % Date - Can also hardwire e.g.: 23 March 2010
  & & \\
@@
 %% Comment before a final submission.
-\linenumbers
+%% \linenumbers
 
@@
 Moreover policy documents on software licensing in HEP are difficult to find.
-In 2011 CERN setup a Task Force to provide recommendations for the licensing of software developed at CERN.
+In 2011 CERN set up a Task Force to provide recommendations for the licensing of software developed at CERN.
 The final report \cite{[1]} is worth reading as it contains much useful background information and reference material.
@@
 People who receive the program in that modified form do not have the freedom that the original author gave them.
-A software licence is a legal instrument governing the use or redistribution of software.
+
+\item A \textit{software licence} is a legal instrument governing the use or redistribution of software.
 A typical software licence grants an end-user permission to use one or more copies of software in ways where such a use would otherwise potentially constitute copyright infringement of the software owner's exclusive rights under copyright law.
@@
 Software using such a licence can meet the conditions to be classed as open source software as conferred by the copyright holder.
-Open source licenses broadly divide into free software licenses and permissive software licenses.
-Free software licences include "copyleft" provisions, which require all future versions to also be distributed with these freedoms. These are termed "restrictive" licences.
 
-\item \textit{Permissive software licences} do not impose these additional conditions and are usually just a grant of rights and a disclaimer of warranty, thus also allowing distributors to add restrictions for further recipients, or to produce an extended proprietary version of the software
-All major open source software licences require that acknowledgement is given to authors of the software in documentation and/or at runtime.
+\item Open source licences broadly divide into free software licences (also known as restrictive software licences) and permissive software licences.
+\textit{Restrictive software licences} include "copyleft" provisions, which require all future versions to also be distributed with these freedoms.
+\textit{Permissive software licences} do not impose these additional conditions and are usually just a grant of rights and a disclaimer of warranty, thus also allowing distributors to add restrictions for further recipients, or to produce an extended proprietary version of the software
+
+\item All major open source software licences require that \textit{acknowledgement} is given to authors of the software in documentation and/or at runtime.
 In an academic context these provisions can be useful in establishing the impact of a software project, and even when software released under a permissive licence is reused in a closed-source commercial product.
@@
 
-\section{Main License Types}
-\label{sec:MainLicenseTypes}
+\section{Main Licence Types}
+\label{sec:MainLicenceTypes}
 
-\subsection{Copyleft License}
-\label{sec:CopyleftLicense}
+\subsection{Copyleft Licence}
+\label{sec:CopyleftLicence}
 
@@
 GPL version 3 was published in 2007 but many copyleft projects (eg Linux) have chosen to continue with GPL version 2.
-GPL 2 and 3 software amount to 24\% and 10\% of open source software respectively [6].
+GPL 2 and 3 software amount to 24\% and 10\% of open source software respectively [5].
 
-\subsection{Weak Copyleft License}
+\subsection{Weak Copyleft Licence}
 \label{sec:WeakCopyleft}
@@
 Licences such as LGPL target libraries of software, which are designed to be incorporated unchanged into larger programs.
-For example, the ROOT software project \cite{[5]} has adopted an LGPL licence.
+For example, the ROOT software project \cite{[6]} has adopted an LGPL licence.
 
@@
 
-As with full copyleft licences which prevent modified versions from being distributed under a proprietary licence, weak copyleft licenses are intended to ensure the non-appropriation by third parties of the Open Source software.
-As of July 2013, the LGPL was used by 7\% of all open source licenced projects \cite{[6]}.
+As with full copyleft licences which prevent modified versions from being distributed under a proprietary licence, weak copyleft licences are intended to ensure the non-appropriation by third parties of the Open Source software.
+As of July 2013, the LGPL was used by 7\% of all open source licensed projects \cite{[5]}.
 
-\subsection{Permissive License}
-\label{sec:PermissiveLicense}
+\subsection{Permissive Licence}
+\label{sec:PermissiveLicence}
 
@@
 
-The Apache Software License (ASL), initially from 1999 and currently at version 2.0, is one of the most widely used examples of a permissive licence.
+The Apache Software Licence (ASL), initially from 1999 and currently at version 2.0, is one of the most widely used examples of a permissive licence.
 Like other open source software licences, the licence allows the user of the software the freedom to use the software for any purpose, to distribute it, to modify it, and to distribute modified versions of the software, under the terms of the licence, without concern for royalties.
@@
 
-As of July 2013, the ASL, BSD and MIT permissive licences accounted for 42\% of all open source licensed projects [6].
+As of July 2013, the ASL, BSD and MIT permissive licences accounted for 42\% of all open source licensed projects [5].
 
@@
 
-\subsection{Changing the License}
-\label{sec:LicenseChange}
+\subsection{Changing the Licence}
+\label{sec:LicenceChange}
 
-The ability to change the license term of a project, including the right to dual-license it, is an exclusive right of copyright holders.
+The ability to change the licence term of a project, including the right to dual-license it, is an exclusive right of copyright holders.
 Except when explicitly stated otherwise, copyright holders are all the people who contributed to the project.
-In large projects, after some time, it may make impossible to change the license used by a project.
+In large projects, after some time, it may make impossible to change the licence used by a project.
 
-Although this rule applies to any license, it is more a concern for copyleft licences as permissive licenses give anybody the right to fork the project with a new license.
-For this reason, some projects, when there is no risk (or a low risk) of appropriation of the work by a third party, prefer to use a permissive license in order to keep a greater flexibility to evolve (including restrict) the licence in the future.
+Although this rule applies to any licence, it is more a concern for copyleft licences as permissive licences give anybody the right to fork the project with a new licence.
+For this reason, some projects, when there is no risk (or a low risk) of appropriation of the work by a third party, prefer to use a permissive licence in order to keep a greater flexibility to evolve (including restrict) the licence in the future.
 A well known example is Apache where a development community exists and where most people (including commercial vendors) contribute their modification back to the community even though this is not a legal requirement of the permissive Apache licence.
@@
 To avoid problems in changing licence, some projects or software foundations (like the Apache Software Foundation) have an explicit transfer of copyright to one single legal entity by each project contributor.
-This is the main alternative for project with copyleft licenses.
+This is the main alternative for project with copyleft licences.
 As with any change related to licensing, it has to be decided early in the life of the project as it requires the agreement of all copyright holders.
 
-Where a non-permissive licence is required to distribute software binaries or packages, one option is maintain the source code repository under a permissive licence but re-licence the software at distribution time under the required licence.
+Where a non-permissive licence is required to distribute software binaries or packages, one option is maintain the source code repository under a permissive licence but re-license the software at distribution time under the required licence.
 This maintains flexibility about what licence to use in the future, but allows linking or repackaging with more restrictively-licensed open source software in the present.
@@
 
-In the case of permissive free software licences, as all permissions for appropriation have been given to any third party, and so commercial exploitation by dual licensing becomes unnecessary.
-Therefore, permissive licenses, such as the ASL, MIT and BSD licenses, are preferred by many companies because such licenses make it possible to use open-source software code without having to turn proprietary enhancements back over to the open source software community.
-These licenses encourage commercial adoption of open-source software because they make it possible for companies to profit from investing in enhancements made to existing open-source software solutions.
+In the case of permissive free software licences, as all permissions for appropriation have been given to any third party, commercial exploitation by dual licensing becomes unnecessary.
+Therefore, permissive licences, such as the ASL, MIT and BSD licences, are preferred by many companies because such licences make it possible to use open-source software code without having to turn proprietary enhancements back over to the open source software community.
+These licences encourage commercial adoption of open-source software because they make it possible for companies to profit from investing in enhancements made to existing open-source software solutions.
 
@@
 
-For projects producing libraries and taking the copyleft route, LGPL should be preferred for program libraries when the goal is to allow wide and rapid adoptions by applications with different licenses.
+For projects producing libraries and taking the copyleft route, LGPL should be preferred for program libraries when the goal is to allow wide and rapid adoptions by applications with different licences.
 
@@
 
-This section contains examples for specifying licence terms, based on real licenses from different HEP laboratories.
+This section contains examples for specifying licence terms, based on real licences from different HEP laboratories.
 You can use them as a source of inspiration but you need to customize them to your specific needs and local context.
@@
 
-One of the following licence statements must be included, immediately following the copyright statement, \textbf{and followed by the text of the relevant license as shown in the references}:
+One of the following licence statements must be included, immediately following the copyright statement, \textbf{and followed by the text of the relevant licence as shown in the references}:
 
@@
 \bibitem{[4]} GNU GENERAL PUBLIC LICENCE Version 3, 29 June 2007 \url{http://www.gnu.org/copyleft/gpl.html}
-\bibitem{[5]} ROOT software terms and conditions: \url{https://root.cern.ch/root/License.html}
-\bibitem{[6]} Top Open Source Licences, source BLACKDUCK, July 2015: \url{https://www.blackducksoftware.com/resources/data/top-20-open-source-licenses}
+\bibitem{[5]} Top Open Source Licences, source BLACKDUCK, July 2015: \url{https://www.blackducksoftware.com/resources/data/top-20-open-source-licenses}
+\bibitem{[6]} ROOT software terms and conditions: \url{https://root.cern.ch/root/License.html}
 \bibitem{[7]} GNU Free Documentation Licence (GFDL): \url{http://www.gnu.org/licences/fdl.html}

### 2016-02-12 HSF-TN/2016-01/HSF-TN-2016-01.tex

@@
 
+\subsection{Licence Compatibility}
+\label{sec:LicenceCompatibility}
+
+The software stacks used in HEP experiments are generally composed of large numbers of software packages, with complex inter-dependencies on one another.
+The term licence compatibility refers to the problem of combining software programmes (such as the software packages in typical HEP stacks) when they are subject to different licences that may contain contradictory requirements.
+This issue and the related terminology are well described in the Annexes to the final report of the CERN Task Force on OSS Licences~\cite{[1]}.
+
+As compatibility is not necessarily reciprocal, compatibility is generally described as a directional property between two licences.
+One refers to upstream or downstream compatibility to specify the direction of the licence compatibility; when the compatibility direction is not specified, this is usually understood as equivalent to the upstream compatibility.
+For example, as mentioned above, the Apache v2.0 licence is (upstream-)compatible with GPL v3, and GPL v3 is downstream-compatible with Apache v2.0: it is possible for programmes licensed under GPL v3 to incorporate programmes licensed under Apache v2.0, with the resulting work still licensed under GPL v3, but the opposite is not true.
+
+Licence compatibility is an important factor that the owners of a specific package should keep in mind when choosing the licence for their software.
+Downstream-compatibility, with the licences of any other programs upon which a given package depends, may effectively restrict the choice of the licence that can be adopted for that package, as some licences might be prohibited by compatibility issues.
+Amongst the licences allowed by downstream-compatibility, the most appropriate licence for a given package should be chosen taking into account its potential upstream-compatibility with the licences chosen by the prospective customers of that package: the choice of a particular licence for a package may in fact be an encouragement or a deterrent for other people to use that package within their software.
+These considerations are reflected in the HSF recommendations described in section~\ref{sec:Recommendations}.
+
 \subsection{Changing the Licence}

### 2016-02-12 HSF-TN/2016-01/HSF-TN-2016-01.tex

--- 2015-11-09 HSF-TN/draft-2015-LIC/draft-HSF-TN-2015-LIC.tex
@@
 \\
- & & HSF-TN-2015-01 \\  % ID 
- & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
+ & & HSF-TN-2016-01 \\  % ID 
+ & & 1 February 2016 \\ % Date - Can also hardwire e.g.: 23 March 2010
  & & \\
@@
 \label{sec:SpecificConstraints}
+
+\subsection{Licence Compatibility}
+\label{sec:LicenceCompatibility}
+
+The software stacks used in HEP experiments are generally composed of large numbers of software packages, with complex inter-dependencies on one another.
+The term licence compatibility refers to the problem of combining software programmes (such as the software packages in typical HEP stacks) when they are subject to different licences that may contain contradictory requirements.
+This issue and the related terminology are well described in the Annexes to the final report of the CERN Task Force on OSS Licences~\cite{[1]}.
+
+As compatibility is not necessarily reciprocal, compatibility is generally described as a directional property between two licences.
+One refers to upstream or downstream compatibility to specify the direction of the licence compatibility; when the compatibility direction is not specified, this is usually understood as equivalent to the upstream compatibility.
+For example, as mentioned above, the Apache v2.0 licence is (upstream-)compatible with GPL v3, and GPL v3 is downstream-compatible with Apache v2.0: it is possible for programmes licensed under GPL v3 to incorporate programmes licensed under Apache v2.0, with the resulting work still licensed under GPL v3, but the opposite is not true.
+
+Licence compatibility is an important factor that the owners of a specific package should keep in mind when choosing the licence for their software.
+Downstream-compatibility, with the licences of any other programs upon which a given package depends, may effectively restrict the choice of the licence that can be adopted for that package, as some licences might be prohibited by compatibility issues.
+Amongst the licences allowed by downstream-compatibility, the most appropriate licence for a given package should be chosen taking into account its potential upstream-compatibility with the licences chosen by the prospective customers of that package: the choice of a particular licence for a package may in fact be an encouragement or a deterrent for other people to use that package within their software.
+These considerations are reflected in the HSF recommendations described in section~\ref{sec:Recommendations}.
 

### 2016-02-22 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
Import

\documentclass[12pt,a4paper]{article}

% Variables that controls behaviour
\usepackage{ifthen} % for conditional statements
\newboolean{pdflatex}
\setboolean{pdflatex}{true} % False for eps figures

\newboolean{inbibliography}
\setboolean{inbibliography}{false} %True once you enter the bibliography

\textheight=230mm
\textwidth=160mm
\oddsidemargin=7mm
\evensidemargin=-10mm
\topmargin=-10mm
\headsep=20mm
\columnsep=5mm
\addtolength{\belowcaptionskip}{0.5em}

\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}

\setlength{\hoffset}{-2cm}
\setlength{\voffset}{-2cm}

% Page defaults ...
\topmargin=0.5cm
\oddsidemargin=2.5cm
\textwidth=16cm
\textheight=22cm

% Don't chase after perfection
\raggedbottom
\sloppy

\usepackage{microtype}
\usepackage{lineno}    % Line numbering during drafting
\usepackage{xspace}    % Avoid problems with missing or double spaces after predefined symbold
\usepackage{caption}   % These three command get the figure and table captions automatically small
\renewcommand{\captionfont}{\small}
\renewcommand{\captionlabelfont}{\small}

%% Graphics
\usepackage{graphicx}  % to include figures (can also use other packages)
\usepackage{color}
\usepackage{colortbl}

%% Math
\usepackage{amsmath} % Adds a large collection of math symbols
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{upgreek} % Adds in support for greek letters in roman typeset

%% fix to allow peaceful coexistence of line numbering and
%% mathematical objects
%% http://www.latex-community.org/forum/viewtopic.php?f=5&t=163
%%
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname
end#1\endcsname
 \renewenvironment{#1}%
   {\linenomath\csname old#1\endcsname}%
   {\csname oldend#1\endcsname\endlinenomath}%
}
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}%
}
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
\patchBothAmsMathEnvironmentsForLineno{eqnarray}%
}

% Get hyperlinks to captions and in references.
% These do not work with revtex. Use "hypertext" as class option instead.
\usepackage{hyperref}    % Hyperlinks in references
\usepackage[all]{hypcap} % Internal hyperlinks to floats.

% Make this the last packages you include before the \begin{document}
\usepackage{cite} % Allows for ranges in citations
\usepackage{mciteplus}

\usepackage{longtable} % only for template

\begin{document}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\begin{titlepage}
\pagenumbering{roman}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2016-VACPLAT \\  % ID 
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{4.0cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
  Vacuum Platform
\end{center}
}

\vspace*{2.0cm}

% Authors -------------------------------------------------
\begin{center}
A.~McNab$^1$
\bigskip\\
{\it\footnotesize
$ ^1$University of Manchester
}
\end{center}

\vspace{\fill}

% Abstract -----------------------------------------------
\begin{abstract}
  \noindent

This technical note describes components of GridPP's Vacuum Platform for 
managing VMs, including the \texttt{\$JOBOUTPUTS}, VacQuery, and VacUserData 
interfaces.

\end{abstract}

\vspace*{2.0cm}

\vspace{\fill}

{\footnotesize 
\centerline{\copyright~Named authors on behalf of the HSF, licence \href{http://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0}.}}
\vspace*{2mm}

\end{titlepage}

\cleardoublepage

\renewcommand{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

%%%% Uncomment next 2 lines if desired
%\tableofcontents
%\cleardoublepage

\pagestyle{plain} % restore page numbers for the main text
\setcounter{page}{1}
\pagenumbering{arabic}

%% Uncomment during drafting and review.
%% Comment before a final submission.
\linenumbers

\section{Introduction}
\label{sec:Introduction}

This technical note describes components of GridPP's Vacuum Platform for 
managing virtual machines (VMs) to run jobs for WLCG and other HEP 
experiments.

The \texttt{\$JOBOUTPUTS}, VacQuery, and VacUserData interfaces are described, 
which have been developed for managing the VM lifecycle. These are used 
by two GridPP software systems, Vac and Vcycle, which can be described 
as VM lifecycle managers (VMLMs).

This note is written in terms of VMs, but the interfaces have been 
designed to be generalised to other forms of logical machine in the
future, such as Docker containers and unikernels.

The term ``resource provider'' is used to refer to the entity which is 
able to take the decision about creating each VM. That is, the decision 
about whether resources will be provided or not. Typically, this is owner 
of an OpenStack tenancy managed by Vcycle or the manager of Vac VM 
factories. Beyond this may be many layers of resource provision
in terms of legal owners of services and hardware and even funding agencies.

A location at which VMs can be created managed by one or more VMLMs
which are cooperating to achieve a set of target shares is referred to
as a ``space''. This is equivalent to a Compute Element at a Grid site,
and spaces must be given DNS names in DNS space available to the
resource provider. However, it is not necessary to register the 
space name in the corresponding DNS zone. For Vac, a space is a set
of VM factory machines which are communicating via VacQuery and may
be said to be neighbours. For Vcycle, a space corresponds to an 
OpenStack tenancy or project, with a specified endpoint to contact
and identity tokens to use.

\section{Environment}
\label{sec:env}

Where possible, VMLMs should provide an approximation of OpenStack's 
environment for VMs, which is derived from EC2. Any contexualization 
user\_data file required and a metadata service should be provided via a 
``Magic IP'' 
HTTP service at 169.254.169.254 from the point of view of the VM. 
Monolithic VM images which do not use a user\_data file require a 
metadata service to be able to discover the 
URLs of the \texttt{\$MACHINEFEATURES}, \texttt{\$JOBFEATURES}, and 
\texttt{\$JOBOUTPUTS} locations.

As not all IaaS cloud systems provide metadata services, VMs and VMLMs 
should also implement the standard VacUserData substitutions described 
in section~\ref{sec:vacuserdata}. These include substitutions giving the 
URLs of the \texttt{\$MACHINEFEATURES}, \texttt{\$JOBFEATURES}, and 
\texttt{\$JOBOUTPUTS} locations.

VMLMs must also support VMs which use Cloud Init contextualization.

\section{Machine/Job Features}
\label{sec:mjf}

The Machine/Job Features (MJF) mechanism described in \cite{HSF-MJF} allows
resource providers to communicate information to batch jobs and virtual
machines, including the number of processors they are allocated and
how long they may run for. The MJF terminology is derived from batch
job environments, and job equates to virtual machine when applied to
virtualized environments such as the Vacuum Platform.

Resource providers in the Vacuum Platform must make the MJF
\texttt{\$MACHINEFEATURES} and \texttt{\$JOBFEATURES} locations
available over HTTP(S) rather than through a shared filesystem, and
should publish the URLs of these locations in OpenStack/EC2 metadata 
tags and using the VacUserData substitutions in the user\_data files
supplied to VMs.

The value of \texttt{\$JOBFEATURES/job\_id} should be set to the VM UUID by
the VMLM as soon as it is known. For example, with Vac the UUID is
chosen by VMLM and its value can be set when the first
\texttt{\$JOBFEATURES}
key/values are created. However with Vcycle managing OpenStack, the
VM UUID is only available after the VM has been created, and is then
recorded by Vcycle in the \texttt{\$JOBFEATURES} directory it provides.

\section{\$JOBOUTPUTS}
\label{sec:joboutputs}

The \texttt{\$JOBOUTPUTS} mechanism is an extension to Machine/Job Features by
which the URL of a location to which VMs can write status and log files
is communicated to the VMs. This value of the \texttt{\$JOBOUTPUTS} URL should
be given in the same way as the \texttt{\$MACHINEFEATURES} and 
\texttt{\$JOBFEATURES} URLs.

Any log file which the VMs wish to make available to resource providers
may be written to the \texttt{\$JOBOUTPUTS} location, for later examiniation
in case of problems. All of these files must have unique names, and
they are all written to the same level (``directory'') of the URL
space on the \texttt{\$JOBOUTPUTS} HTTP(S) server. This mechanism is also used 
to provide the \texttt{shutdown\_message} file described in the next section.

\subsection{Shutdown Messages}
\label{sec:shutdownmessages}

When VMs finish, they should write a \texttt{shutdown\_message} file to
\texttt{\$JOBOUTPUTS/shutdown\_message} containing one line of text without
a trailing newline character. This text consists of a
three digit shutdown message code in the range 100-999, a space, 
and then a human-readable description of the message code.

The message code (and not the human-readable description) will
be used by the resource provider's software to determine why
the VM finished and whether to create more VMs of this type in
the immediate future as slots become available.

\begin{center}
\begin{table}[h]
\label{table:shutdowncodes}
\begin{center}
\begin{tabular}{l}
\\
100 Shutdown as requested by the VM's host/hypervisor \\
200 Intended work completed ok \\
300 No more work available from task queue \\
400 Site/host/VM is currently banned/disabled from receiving more work \\
500 Problem detected with environment/VM provided by the site \\
600 Grid-wide problem with job agent or application within VM \\
700 Transient problem with job agent or application within VM \\
\\
\end{tabular}
\caption{\label{opt}Shutdown codes and messages}
\end{center}
\end{table}
\end{center}

The shutdown codes are designed to be extensible by the insertion of 
intermediate numbers for finer-grained reporting. This is similar to the
three digit response codes of internet protcols such as SMTP and HTTP.

\section{Image URLs}
\label{sec:imageurls}

Experiments should provide the HTTPS URL of the image file required to boot
their VMs, which VMLMs should use. VMLMs should support both standard CAs
and IGTF endorsed-CAs when verifying the X.509 certificates used by the
relevant HTTPS webserver.

To avoid overloading these webservers, VMLMs must cache images by 
Last-Modified time, and should use the HTTP If-Modified-Since mechanism
when fetching images. If this header is used, then it is acceptable to
check the URL for updates each time a VM is created.

Where the VMLM is unable to update the image used to boot the VMs itself, 
it should attempt to verify that the image being used is current and refuse
to create new VMs with an old image. Typically this applies to IaaS cloud
systems where users are unable to upload new images, or a manual upload step
is required. VMLM authors should consider how resource providers will be
made aware of this situation when it arises, but for scalability reasons,
the VMLM should not rely on the experiment suffering from VMs starting and
then failing 
due to an out of date VM image and then notifying resource providers.

\section{VacUserData templates}
\label{sec:vacuserdata}

In most cases, a generic image such as CernVM is used which then
requires further contextualization as the VM starts using a user\_data
file supplied by VMLM. The VMLM must be able to retrieve a template for
the user\_data file
from an HTTPS URL nominated by the experiment each time a VM is to 
be created. That is, without any caching. The VMLM must include an
appropriate HTTP User-Agent header indicating the VMLM implementation
and version when making this request to allow experiments to monitor
which VMLMs are in use. The VMLM should support both standard CAs
and IGTF endorsed-CAs when verifying the X.509 certificates used by the
relevant HTTPS webserver.

The VMLM must apply the following pattern
based substitutions to the user\_data template supplied by the experiment.
These patterns are all in the form \texttt{\#\#user\_data\_XXX\#\#}.

The following substitutions are performed automatically
using data the VMLM holds internally:

\begin{quote}
\begin{description}
\setlength{\parskip}{0pt}
\item{\texttt{\#\#user\_data\_space\#\#}} Space name
\item{\texttt{\#\#user\_data\_machinetype\#\#}} Name of the machinetype of this VM
\item{\texttt{\#\#user\_data\_machine\_hostname\#\#}} Hostname assigned to the VM by the VMLM
\item{\texttt{\#\#user\_data\_manager\_version\#\#}} A string giving the VMLM version
\item{\texttt{\#\#user\_data\_manager\_hostname\#\#}} Hostname of the VMLM
\item{\texttt{\#\#user\_data\_manager\_machinefeatures\_url\#\#}} \texttt{\$MACHINEFEATURES} URL (section~\ref{sec:mjf})
\item{\texttt{\#\#user\_data\_manager\_jobfeatures\_url\#\#}} \texttt{\$JOBFEATURES} URL (section~\ref{sec:mjf})
\item{\texttt{\#\#user\_data\_manager\_joboutputs\_url\#\#}} \texttt{\$JOBOUTPUTS} URL (section~\ref{sec:joboutputs})
\end{description}
\end{quote}

The VMLM must also provide a mechanism for the resource provider which uses
the VMLM to specify strings or files which will used in substitutions
required by the VM. These patterns take the form
\texttt{\#\#user\_data\_option\_XXX\#\#} where XXX is an arbitrary string
consisting of letters, numbers, and underscores.

If the VM requires an X.509 proxy, it must expect that the special pattern
\texttt{\#\#user\_data\_option\_x509\_proxy\#\#} will be replaced by the 
PEM encoded X.509 certificates and RSA private key which compromise the
proxy. VMLM's should provide a mechanism for creating X.509 proxies
dynamically for each VM instance from a host or robot certificate owned
by the resource provider, with an X.509 proxy lifetime reflecting
the maximum VM lifetime.

\section{VacQuery}
\label{sec:vacquery}

The principle use of the VacQuery protocol is to allow Vac factories to
gather information from their neighbours about what VMs are running for
what machinetypes. This done using the machinetypes\_query and
machinetype\_status UDP messages. Factory and machine message pairs
are also supported which can be used for automated or manual 
monitoring of Vac-based sites.

VacQuery queries sent to Vac daemons take the form of JSON documents
in packets directed to the unused UDP port 
995\footnote{In Roman numerals, V=5 and M=1000. 995 could be written
as VM = 1000 - 5, although this violates conventions invented in 
modern times.} The protocol has been designed to keep JSON message and
IP header below the ethernet MTU of 1500 bytes to avoid fragmentation
on local networks. Responses are sent to the UDP port from which the
query was sent.

\subsection{Factory messages}
\label{sec:factorymessages}

The factory messages factory\_query and factory\_status are intended
for monitoring the state of the factories themselves, including
generic Unix health metrics such as free disk and CPU load. As well
as manual queries by administrators, these messages may also be
used for automated Nagios-style monitoring and alarms.

\subsubsection{factory\_query}
\label{sec:factoryquery}

The factory\_query message is sent to a factory to request a 
factory\_status message in response.

\begin{quote}
\begin{description}
\setlength{\parskip}{0pt}
\item[message\_type] ``factory\_query''
\item[vac\_version] Software version of Vac
\item[vacquery\_version] Version of the VacQuery protocol
\item[space] Vac space name
\item[cookie] Freely chosen by the sender
\end{description}
\end{quote}

\subsubsection{factory\_status}
\label{sec:factorystatus}

factory\_status messages are returned in response to 
factory\_query messages directed to a factory. They may
also be generated spontaneously and sent to a VacMon
service as described in section~\ref{sec:vacmon}.

\begin{quote}
\begin{description}
\setlength{\parskip}{0pt}
\item[message\_type] ``factory\_status''
\item[vac\_version] Software version of Vac
\item[vacquery\_version] Version of the VacQuery protocol
\item[cookie] Matching the value supplied by the recipient
\item[space] Vac space name
\item[factory] FQDN of the factory
\item[time\_sent] Time in Unix seconds (since 00:00:00 1st Jan 1970)
\item[total\_cpus] Number of (logical) processors available to VMs
\item[running\_cpus] Number of processors assigned to running VMs
\item[running\_machines] Number of running VMs
\item[total\_machines] Maximum possible number of VMs
\item[total\_hs06] Total HS06 available to VMs
\item[vac\_disk\_avail\_kb] Free space available in Vac's workspace, in units of 1024 bytes
\item[root\_disk\_avail\_kb] Free space available on the root partition, in units of 1024 bytes
\item[vac\_disk\_avail\_inodes] Free inodes available in Vac's workspace
\item[root\_disk\_avail\_inodes] Free inodes available on the root partition
\item[load\_average] The 15 minute Unix load average on the factory
\item[kernel\_version] The running kernel version of the factory
\item[os\_issue] A string identifying the operating system (typically the first line of /etc/issue)
\item[boot\_time] The time when the factory booted up in Unix seconds
\item[factory\_heartbeat\_time] Time of the last heartbeat created by the VM factory agent in Unix secconds
\item[responder\_heartbeat\_time] Time of the last heartbeat created by the VacQuery responder service in Unix secconds
\item[mjf\_heartbeat\_time] Time of the last heartbeat created by the HTTP Machine/Job Features service in Unix secconds
\item[metadata\_heartbeat\_time] Time of the last heartbeat created by the HTTP Metadata service in Unix secconds
\item[swap\_used\_kb] Swap space in use on the factory, in units of 1024 bytes
\item[swap\_free\_kb] Free swap space, in units of 1024 bytes
\item[mem\_used\_kb] Physical memory in use on the factory, in units of 1024 bytes
\item[mem\_total\_kb] Free physical memory, in units of 1024 bytes
\item[site] Name of the site registered in the GOCDB, or the Vac space name if the site is not registered
\end{description}
\end{quote}

\subsection{Machine messages}
\label{sec:machinemessages}

The machines\_query (plural) and machine\_status (singular) messages can
be used to create views of the VMs running within a Vac space, similar to
the views from the top command of running processes on a single host.

\subsubsection{machines\_query}
\label{sec:machinesquery}

The machines\_query message is sent to a factory to request a 
machine\_status message for each of its VM slots.

\begin{quote}
\begin{description}
\setlength{\parskip}{0pt}
\item[message\_type] ``machines\_query''
\item[vac\_version] Software version of Vac
\item[vacquery\_version] Version of the VacQuery protocol
\item[space] Vac space name
\item[cookie] Freely chosen by the sender
\end{description}
\end{quote}

\subsubsection{machine\_status}
\label{sec:machinestatus}

machine\_status messages are returned in response to machines\_query messages directed to a factory.

\begin{quote}
\begin{description}
\setlength{\parskip}{0pt}
\item[message\_type] ``factory\_status''
\item[vac\_version] Software version of Vac
\item[vacquery\_version] Version of the VacQuery protocol
\item[cookie] Matching the value supplied by the recipient
\item[space] Vac space name
\item[factory] FQDN of the factory
\item[num\_machines] Number of machine\_status messages to expect from this factory
\item[time\_sent] Time in Unix seconds (since 00:00:00 1st Jan 1970)
\item[machine] Hostname of the VM slot
\item[state] State of the current or most recent VM in this slot, as a string
\item[uuid] Lowlevel UUID, as used by libvirtd
\item[created\_time] Unix time of the VM's creation
\item[started\_time] Unix time the VM entered the running state
\item[heartbeat\_time] Unix time when the VM was last observed to be running (this is not the same as any heartbeat generated within the VM)
\item[cpu\_seconds] CPU seconds used by the VM
\item[cpu\_percentage] Recent CPU percentage use. May be over 100\% for mulitprocessor VMs
\item[hs06] Total HEPSPEC06 for the number of processors assigned to this VM
\item[machinetype] Name of the machinetype
\item[shutdown\_message] Any shutdown message given by the last VM to run in this slot
\item[shutdown\_time] Unix time of the shutdown\_message
\end{description}
\end{quote}

\subsection{Machinetype messages}
\label{sec:machinetypemessages}

The machinetypes\_query (plural) and machinetype\_status (singular) messages 
are used by factories to gather information from neighbours within the
same Vac space about what they are running, and outcomes of recently started
VMs which have finished.

\subsubsection{machinetypes\_query}
\label{sec:machinetypesquery}

The machinetypes\_query message is sent to a factory to request a 
machinetype\_status message for each of the machinetypes it supports.

\begin{quote}
\begin{description}
\setlength{\parskip}{0pt}
\item[message\_type] ``machinetypes\_query''
\item[vac\_version] Software version of Vac
\item[vacquery\_version] Version of the VacQuery protocol
\item[space] Vac space name
\item[cookie] Freely chosen by the sender
\end{description}
\end{quote}

\subsubsection{machinetype\_status}
\label{sec:machinetypestatus}

machinetype\_status messages are returned in response to machinetypes\_query 
messages directed to a factory. They may
also be generated spontaneously and sent to a VacMon
service as described in section~\ref{sec:vacmon}.

\begin{quote}
\begin{description}
\setlength{\parskip}{0pt}
\item[message\_type] ``factory\_status''
\item[vac\_version] Software version of Vac
\item[vacquery\_version] Version of the VacQuery protocol
\item[cookie] Matching the value supplied by the recipient
\item[space] Vac space name
\item[factory] FQDN of the factory
\item[num\_machinetypes] Number of machinetype\_status messages to expect from this factory
\item[time\_sent] Time in Unix seconds (since 00:00:00 1st Jan 1970)
\item[machinetype] Name of the machinetype
\item[total\_hs06] Total HEPSPEC06 of all the processors allocated to running VMs for this machinetype on this factory
\item[total\_machines] Number of running VMs for this machinetype on this factory
\item[num\_before\_fizzle] Number of running VMs which have not yet reached fizzle\_seconds
\item[shutdown\_message] Shutdown message given by the most recently created VM for this machinetype on this factory which has finished
\item[shutdown\_time] Unix time of the shutdown\_message
\item[shutdown\_machine] Name of the VM slot associated with the shutdown\_message
\end{description}
\end{quote}

\subsection{VacMon services}
\label{sec:vacmon}

VacMon services receive factory\_status and machinetype\_status
messages from Vac daemons on UDP port 8884. These may be used for
Ganglia-style monitoring of individual sites or groups of sites.
As VacQuery messages as JSON documents, they may be conveniently 
recorded in data stores such as ElasticSearch.

\section{APEL}
\label{sec:apel}

VMLMs should support reporting of usage to the central APEL service
with messages of the type ``APEL-individual-job-message''.

In this case, VLMs must include the following in the messages:
\begin{quote}
\begin{description}
\setlength{\parskip}{0pt}
\item[FQAN] VOMS FQAN specified by the experiment when configuring the space
\item[SubmitHost] Must be of the form 
  \texttt{[space name] + "/" + [vmlm] + "-" + [VMLM host name]}, where
  ``vmlm'' is a lowercase name for the VMLM software such as ``vac''
\item[LocalJobId] VM UUID
\item[LocalUserId] VMLM hostname
\item[Queue] Name of the machinetype
\item[GlobalUserName] The space name converted to an X.500 DN with DC
  components. For example, vac01.example.com would become /DC=vac01/DC=example/DC=com
\item[InfrastructureDescription] Such as APEL-VAC or APEL-VCYCLE, with APEL and then
  an uppercase name for the VMLM software.
\item[Processors] The number of virtual CPUs assigned to the VM.
\end{description}
\end{quote}

APEL Sync records must also be sent, and these can conveniently be
generated by each VMLM instance from an archive of the individual
job messages, as the SubmitHost is unique to the VMLM instance
in both cases.

In addition to APEL records generated by the VMLM, an underlying
cloud infrastracture may also be instrumented to submit APEL usage records
to the cental APEL service.
This is especially likely at sites participating in the EGI Federated Cloud.
In this case, resource providers must ensure that double counting is avoided
by disabling reporting from the VMLM to the central APEL service.

\section{GLUE2}
\label{sec:glue2}

VMLMs should publish status information as JSON documents using the GLUE 2.0 
schema, at an HTTPS URL advertised in the GOCDB service entry for the space,
in the Grid Information / URL field.

\clearpage

\section{Summary}
\label{sec:Summary}

Vacuum platform interfaces have been described ...

%\section*{Acknowledgments}

\section*{References}

%\begin{thebibliography}{9}
%% Use references in the format expected by JPCS (as used for CHEP proceedings)
%
%\bibitem{SHOULD-MUST} S. Bradner, RFC2119 ``Key words for use in RFCs to Indicate Requirement Levels'' (Internet Engineering Task Force)
%
%\bibitem{IETF-RFC} S. Bradner, RFC2026 ``The Internet Standards Process -- Revision 3'' (Internet Engineering Task Force)
%
%\end{thebibliography}

%\addcontentsline{toc}{section}{References}

\end{document}

### 2016-02-22 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
Spell out Unix seconds

@@
 
-The principle use of the VacQuery protocol is to allow Vac factories to
+The principal use of the VacQuery protocol is to allow Vac factories to
 gather information from their neighbours about what VMs are running for
@@
 query was sent.
+
+All dates/times in VacQuery messages are expressed as Unix seconds. That
+is, the number of seconds since 00:00:00 1st Jan 1970.
 
@@
 \item[factory] FQDN of the factory
-\item[time\_sent] Time in Unix seconds (since 00:00:00 1st Jan 1970)
+\item[time\_sent] Time in Unix seconds
 \item[total\_cpus] Number of (logical) processors available to VMs
@@
 \item[num\_machines] Number of machine\_status messages to expect from this factory
-\item[time\_sent] Time in Unix seconds (since 00:00:00 1st Jan 1970)
+\item[time\_sent] Time in Unix seconds
 \item[machine] Hostname of the VM slot
@@
 \item[num\_machinetypes] Number of machinetype\_status messages to expect from this factory
-\item[time\_sent] Time in Unix seconds (since 00:00:00 1st Jan 1970)
+\item[time\_sent] Time in Unix seconds
 \item[machinetype] Name of the machinetype

### 2016-02-22 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
Clarify use of /proc values etc

@@
 for monitoring the state of the factories themselves, including
-generic Unix health metrics such as free disk and CPU load. As well
+generic Linux health metrics such as free disk and CPU load. As well
 as manual queries by administrators, these messages may also be
@@
 
+The format and units of the disk and memory values are
+aligned with the values returned by the relevant system
+calls and the /proc interface.
+
 \begin{quote}
@@
 \item[time\_sent] Time in Unix seconds
+\item[site] Name of the site registered in the GOCDB, or the Vac space name if the site is not registered
 \item[total\_cpus] Number of (logical) processors available to VMs
@@
 \item[total\_hs06] Total HS06 available to VMs
-\item[vac\_disk\_avail\_kb] Free space available in Vac's workspace, in units of 1024 bytes
-\item[root\_disk\_avail\_kb] Free space available on the root partition, in units of 1024 bytes
-\item[vac\_disk\_avail\_inodes] Free inodes available in Vac's workspace
-\item[root\_disk\_avail\_inodes] Free inodes available on the root partition
-\item[load\_average] The 15 minute Unix load average on the factory
-\item[kernel\_version] The running kernel version of the factory
-\item[os\_issue] A string identifying the operating system (typically the first line of /etc/issue)
 \item[boot\_time] The time when the factory booted up in Unix seconds
@@
 \item[metadata\_heartbeat\_time] Time of the last heartbeat created by the HTTP Metadata service in Unix secconds
+\item[vac\_disk\_avail\_kb] Free space available in Vac's workspace, in units of 1024 bytes
+\item[root\_disk\_avail\_kb] Free space available on the root partition, in units of 1024 bytes
+\item[vac\_disk\_avail\_inodes] Free inodes available in Vac's workspace
+\item[root\_disk\_avail\_inodes] Free inodes available on the root partition
+\item[load\_average] The 15 minute load average on the factory
+\item[kernel\_version] The kernel version of the factory
+\item[os\_issue] A string identifying the operating system (typically the first line of /etc/issue)
 \item[swap\_used\_kb] Swap space in use on the factory, in units of 1024 bytes
@@
 \item[mem\_total\_kb] Free physical memory, in units of 1024 bytes
-\item[site] Name of the site registered in the GOCDB, or the Vac space name if the site is not registered
 \end{description}

### 2016-02-22 HSF-TN/2016-01/HSF-TN-2016-01.tex

@@
 \begin{center}
-J.~Harvey$^1$
-M.~Jouvin$^2$
+M.~Jouvin$^1$
+J.~Harvey$^2$
 A.~McNab$^3$
@@
 {\it\footnotesize
-$ ^1$CERN
-$ ^2$Laboratoire de l’Accélérateur Linéaire (CNRS)
+$ ^1$Laboratoire de l’Accélérateur Linéaire (CNRS)
+$ ^2$CERN
 $ ^3$University of Manchester
@@
 \label{sec:SpecificConstraints}
-
-\subsection{Licence Compatibility}
-\label{sec:LicenceCompatibility}
-
-The software stacks used in HEP experiments are generally composed of large numbers of software packages, with complex inter-dependencies on one another.
-The term licence compatibility refers to the problem of combining software programmes (such as the software packages in typical HEP stacks) when they are subject to different licences that may contain contradictory requirements.
-This issue and the related terminology are well described in the Annexes to the final report of the CERN Task Force on OSS Licences~\cite{[1]}.
-
-As compatibility is not necessarily reciprocal, compatibility is generally described as a directional property between two licences.
-One refers to upstream or downstream compatibility to specify the direction of the licence compatibility; when the compatibility direction is not specified, this is usually understood as equivalent to the upstream compatibility.
-For example, as mentioned above, the Apache v2.0 licence is (upstream-)compatible with GPL v3, and GPL v3 is downstream-compatible with Apache v2.0: it is possible for programmes licensed under GPL v3 to incorporate programmes licensed under Apache v2.0, with the resulting work still licensed under GPL v3, but the opposite is not true.
-
-Licence compatibility is an important factor that the owners of a specific package should keep in mind when choosing the licence for their software.
-Downstream-compatibility, with the licences of any other programs upon which a given package depends, may effectively restrict the choice of the licence that can be adopted for that package, as some licences might be prohibited by compatibility issues.
-Amongst the licences allowed by downstream-compatibility, the most appropriate licence for a given package should be chosen taking into account its potential upstream-compatibility with the licences chosen by the prospective customers of that package: the choice of a particular licence for a package may in fact be an encouragement or a deterrent for other people to use that package within their software.
-These considerations are reflected in the HSF recommendations described in section~\ref{sec:Recommendations}.
 

### 2016-02-26 HSF-TN/2016-01/HSF-TN-2016-02.tex
Create HSF-TN-2016-02.tex
First version of the HSF Packaging working group report. The group is now sending it out for public comment.

--- 2015-11-03 HSF-TN/draft-2015-PKG/draft-HSF-TN-2015-PKG.tex
@@
 \begin{center}
-L.~Sexton-Kennedy$^1$, B.~Hegner$^2$
+L.~Sexton-Kennedy$^1$, B.~Hegner$^2$, B.~Viren$^3$
 \bigskip\\
 {\it\footnotesize
-$ ^1$FNAL,$^2$CERN 
+$ ^1$FNAL,$^2$CERN,$^3$BNL
 }
@@
 Additional discussions took place via the public \href{https://groups.google.com/forum/#!forum/hep-sf-packaging-wg}{packaging working group mailing list}.
+\subsection{The Packaging Problem Space}
+Collectively across HEP and even within some large institutions there is
+a need to maintain software installations which span a broad
+configuration space.  It is helpful to consider this as an actual
+multi-dimensional space and decompose it along the following orthogonal
+and discrete dimensions:
+
+\begin{description}
+
+\item[project] enumerates a set of software units each of which provides
+some cohesive functionality, often in source code form and which can be
+used to produce a number of library or executable files.  Examples
+include ROOT, Geant4 or BOOST.
+
+\item[version] labels the state of a project's software either as a
+formal release or as a snapshot in time.  Of course the exact label
+convention will cosmetically differ between projects but this coordinate
+can nonetheless be considered ordered in some manner and as a union of
+all possible conventions.  Examples include 6.04/10 (ROOT), 10.01p02
+(Geant4), 2015118 (some snapshot) or \texttt{0400b69} (abbreviated git
+commit hash).
+
+\item[variant] describes the set of options applied during the process
+of building a particular (project,version) pair.  The variant of a
+package is in general itself multi-dimensional.  Between any given set
+of (package,version) points there is some amount of degeneracy in the
+variant dimension.  For example, it is common now that all C++ packages
+that must work together must be built with a consistent compiler and C++
+standard.  However, some packages may be built with optimization and no
+debug symbols and others, vice versa.
+
+\end{description}
+
+A single (\textit{project,version,variant}) point in this space
+represents one built \textit{package} (for now, ignoring details of file
+system layout, bundling, distribution, run-time environment).  The
+software that is provided by the operating system may be thought of as a
+single point in this space.
+
+In this picture, an end-user \textit{software suite} is represented as a
+tree of connected points in this space.  Connections are made along
+lines of dependency and such that desired build consistency is secured.
+It is the act of building one package on top of another that a line is
+drawn.
+
+This picture allows the expression of some required and desired featues
+of a packaging system:
+% these remain to be discussed, for sure!
+
+\begin{itemize}
+
+\item require ability to precisely describe, record and instantiate a
+package-space tree.
+
+\item require the ability to produce different package-space trees.
+
+\item require the ability for end-users to selectively activate a
+specific package-space tree.
+
+\item require the ability to prune installed package-space trees but
+keep shared branches needed by any retained trees.
+
+\item desire ability to avoid rebuilding common branches when two trees
+overlap.
+
+\item require ability to precisely assemble package-space trees from
+individual points (built packages) exactly as they were built
+
+\item desire ability to loosely assemble package-space trees from
+individual points close to where they were built (eg, build against
+Python 2.7.9 but run against Python 2.7.10).
+
+\item require ability to derive a novel package-space tree from an
+existing one (eg, upgrade ROOT to new version, make debug/opt versions
+of some packages, use native compiler or build one as part of the tree).
+
+\end{itemize}
+Many HEP experiments define their own naming conventions for the \textit{variant} dimension in this space, and they are all equally valid but inhibit sharing due to trivial language differences.
+A sub-goal of this working group is to agree on a field wide naming convention.
+The HSF Packaging group has a draft proposal, available here: https://github.com/HEP-SF/documents/blob/master/HSF-TN/draft-2015-NAM/draft-HSF-TN-2015-NAM.pdf
+% OK there is probably a better way to reference this but I don't know it... -LSK
 
@@
 
-\subsection{Solution 4 -- contractor} 
-\emph{Contractor} is the build tool used by the accelerator modeling community and was developed for the Synergia project.
+\subsection{Solution 4 -- Contractor} 
+\emph{Contractor}\cite{contractor} is a build tool used in the accelerator modeling community and was developed for the Synergia project.
 The project must support a wide diversity of platforms from laptops to super-computers.
@@
 
+\subsection{Solution 6 -- aliBuild}
+\emph{aliBuild}\cite{aliBuild} is an evolution of cmsBuild that fixes many of the weakness of cmsBuild.
+Relative to the above it is not dependent on rpm and builds tar-files that can be currated into a number of diffent package managers such as yum, and apt.
+% I need to get Giulio to sign off on my short description. -LSK
+
 \section{Taxonomy of non-HEP tools}
-\subsection{Solution 6 - Homebrew }
+\subsection{Solution 7 - Homebrew }
 \emph{Homebrew}\cite{homebrew} is a MacOS specific build tool.
@@
 
-\subsection{Solution 7 - Nix }
+\subsection{Solution 8 - Nix }
 \emph{Nix}\cite{nix} is a cross-platform build tool. Package build instructions are written in the \emph{Nix expression language}.
+
+\subsection{Solution 9 - Spack }
+\emph{Spack}\cite{Spack} is a packaging system specifically developed for the HPC use case which has a lot in common with the HEP requirements.
+``Spack provides a novel, re-cursive specification syntax to invoke parametric builds of packages and dependencies.
+It allows any number of builds to coexist on the same system, and it ensures that installed packages can find their dependencies, regardless of the environment." On creating Spack the authors did a review of existing packaging systems, including a few of those discussed in this document.
+
+\subsection{Commonalities}
+All these non-HEP tools rely on properly set RPATHs and do not provide any relocatability options. The targeted deployment scenarios are  different ones compared to the current approach in HEP.
 
@@
 Some of them impose the requirement of cross compilation onto the build and packaging solution, which is why the keyword \textbf{Xcompiler} is added to the platform table.
+However this also refers to the ability to build linux releases on mac, or ability to build mac releases on linux.
 The table~\ref{tab:taxonomyPlatforms} summarizes this information for the tools considered by the working group so far.
@@
 & Linux & MacOS X & Windows & Xcompiler \\ \hline
+aliBuild & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} +  \\
 cmsBuild & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} +  \\
-Contractor & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & ?\\
-Homebrew & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{red} -\\
+Contractor & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} + \\
+Homebrew & \cellcolor{yellow} o & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{red} -\\
 LCGCMake & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} o   & \cellcolor{yellow} o \\
 Nix & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} o  & \cellcolor{yellow} o \\
-SciSoft &  \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & ? \\
+SciSoft &  \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{red} - \\
+Spack & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} +\\
 Worch &  \cellcolor{green} + &  \cellcolor{green}  + & \cellcolor{red} -  &  \cellcolor{yellow} o
@@
 \item \textbf{Multi-Rel}: The support for building and installing multiple stacks in parallel. This is important if multiple versions of experiment software need to coexist on a given installation.
-\item \textbf{Multi-BuildVar}: The support for multiple versions of the same package within the same stack.
+\item \textbf{Multi-BuildVar}: The support for multiple varients of the same package, release pair within the same stack.
 There are many possible reasons why this maybe needed, what they have in common is the need to specify compiler switches that have to be applied consistently across the build, like a debugging option or a compiler dialect.
-\item \textbf{MutiShell-RTE}: The support for setting up the runtime environment for a given stack version, supporting multiple shell flavours.
+\item \textbf{MultiShell-RTE}: The support for setting up the runtime environment for a given stack triplet (project,version,variant), supporting multiple shell flavours.
 \item \textbf{Relocation}: Whether the build tool supports the relocation of packages or creates fully relocatable packages.
@@
 &Multi-Rel & Multi-BuildVar & MultiShell-RTE & Relocation \\ \hline
+aliBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} & \cellcolor{green} + \\
 cmsBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} & \cellcolor{green} + \\
-Contractor & \cellcolor{green} + & \cellcolor{green} + & ? & ? \\
+Contractor & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA  &  \cellcolor{red} - \\
 Homebrew &  \cellcolor{red} - & \cellcolor{red} -   & \cellcolor{yellow} NA   & \cellcolor{red} -  \\
@@
 SciSoft &  \cellcolor{green} + & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{green} + \\
+Spack & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA & \cellcolor{red} -\\
 Worch &  \cellcolor{green} + &  \cellcolor{green}  + &  \cellcolor{green} + &  \cellcolor{green} +
@@
 \caption{\label{tab:taxonomyVariants}Support for multiple releases and build variants.
-\emph{NA} denotes that the criteria are not-applicable for homebrew.
+\emph{NA} denotes that the criteria are not-applicable for homebrew and Spack.
 As it only provides one variant it can rely on default system paths.}
@@
 &Depends-On & Ease-Add-Pkg & Ease-Bootstrap & Documentation \\ \hline
+aliBuild & \cellcolor{green} Python & \cellcolor{green} spec-file & \cellcolor{green} +  & \cellcolor{yellow} o\\
 cmsBuild & \cellcolor{yellow} Python,rpm,apt & \cellcolor{green} spec-file & \cellcolor{yellow} o  & \cellcolor{red} -\\
-Contractor &  \cellcolor{green} Python & ? & ?  & \cellcolor{red} -\\
+Contractor &  \cellcolor{green} Python & \cellcolor{green} python file & \cellcolor{green} git checkout  & \cellcolor{red} -\\
 Homebrew & \cellcolor{green} Ruby & \cellcolor{green} Formula & \cellcolor{green} + & \cellcolor{green} + \\
 LCGCMake & \cellcolor{green} Python, Cmake & \cellcolor{green} Cmake-macro & \cellcolor{green} + & \cellcolor{yellow} o\\
-Nix & \cellcolor{green} Perl & \cellcolor{green} expression & \cellcolor{red} - & \cellcolor{green} +\\
+Nix & \cellcolor{green} Perl & \cellcolor{yellow} expression & \cellcolor{red} - & \cellcolor{green} +\\
 SciSoft &  \cellcolor{green} Cmake & \cellcolor{green} Cmake-macro & \cellcolor{yellow} o & \cellcolor{red} -\\
+Spack & \cellcolor{green} Python & \cellcolor{green} auto template & \cellcolor{green} git checkout& \cellcolor{green} +\\
 Worch &   \cellcolor{green} Waf,Python &  \cellcolor{green}  Text-files & \cellcolor{green} + & \cellcolor{yellow} o\\
@@
 length of time to incrementally build a developer defined subset of the stack.
+c.
+Ability to reuse binary packages to speedup builds on different machines.
 %BH: I merged Parallel + Performance again
 \item \textbf{Sys-Reuse}: the ability of the build system to reuse parts of the system software that is being built if desired.
-\item \textbf{Community}: whether the tool is used by a wider community or just by one collaboration 
+\item \textbf{Community}: whether the tool is used by a wider community or not.
+Green means it is a standard opensource project (e.g.
+homebrew, conda, nix), yellow means it is HEP / science only community, red means it is a single experiment or person effort.
 \item \textbf{Unique-IDs}: A method of uniquely identifying a build product such that if it already exists, it does not have to be built again.
+This feature insures that the software suite will always be built consistently throughout the whole stack.
+Further discussion with the group reveals that this point is complicated.
+We will expand on this in the next section.
+%It is also desirable to have the ability to develop a local copy of the package and still build using the tool (including noticing changes in the local area and rebuilding a consistent build).
   \item \textbf{VCS-Support} Integrated support for check-out of given tags or branches from the projects’ repositories directly.
@@
 &Performance & Sys-Reuse & Community & Unique-IDs & VCS-Support\\ \hline
-cmsBuild & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{red} - & \cellcolor{green} +& \cellcolor{green} +  \\
-Contractor & ? & ?  & ? & \cellcolor{red} - & ?\\
+aliBuild & \cellcolor{green} + & \cellcolor{yellow} o & \cellcolor{red} - & \cellcolor{green} +& \cellcolor{green} +  \\
+cmsBuild & \cellcolor{green} + & \cellcolor{yellow} o & \cellcolor{red} - & \cellcolor{green} +& \cellcolor{green} +  \\
+Contractor & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{red} - & \cellcolor{green} + \\
 Homebrew & \cellcolor{green} +  & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} + \\
-LCGCMake & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + \\
-Nix & \cellcolor{green} + & \cellcolor{red} -  & ? & \cellcolor{green} + & ? \\
-SciSoft & \cellcolor{yellow} o & ? & \cellcolor{green} + & ? & \cellcolor{green}+  \\
+LCGCMake & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} + & \cellcolor{green} + & \cellcolor{green} + \\
+Nix & \cellcolor{green} + & \cellcolor{red} -  & \cellcolor{green} + & \cellcolor{green} + & ? \\
+SciSoft & \cellcolor{yellow} o & \cellcolor{yellow} o & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{green}+  \\
+Spack & \cellcolor{yellow} o & \cellcolor{yellow} o  & \cellcolor{green} + & \cellcolor{green} +& \cellcolor{green} +\\
 Worch & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{red} -  & \cellcolor{yellow} o & \cellcolor{green} + 
@@
 While in general of high quality, the non-HEP tools we looked at seem very weak on the support for multi-stack, multi-configuration setups.
-There are however more tools in the open-source community we did not investigate on yet.
-Among the HEP specific tools examined so far, \emph{LCGCMake} currently seems the most suited candidate for generalization.
-If decided, it could serve as the tool for the \emph{HSF Reference Builds} of the projects participating in the HSF.
-A more ambitious goal would be finding a common tool for building experiment stacks.
+On the other hand, the tool Spack, originating from the HPC community, covers all these points excellently.
+The problematic points of Spack are the lack of a possiblity to install pre-compiled binaries, and the non-relocatability of build artifacts.
+The former is a blocker for the HEP use-case, as the number of installations is much larger than in the very centralized HPC environment -- even though the usage of CVMFS for software distribution changed the deployment scenarios significantly for HEP.
+Preliminary investigations indicate that enhancing Spack and/or combining it with other tool(s) would be straightforward.
+Therefore Spack currently seems to be the most suitable candidate for a common packaging tool.
+Among the HEP-specific tools examined so far, \emph{LCGCMake} and \emph{aliBuild} currently seem to be the primary candidates for generalization.
+
+The next step would be to investigate whether the weak points of Spack can be addressed.
+Successful first steps were done in an HSF-specific fork of Spack.
+If successful, it could serve as the tool for the \emph{HSF Reference Builds} of the projects participating in the HSF.
+A more ambitious goal would be enabling it as a common tool for building experiment stacks.
 
 In parallel to the assessment of existing build tools, we discussed the concept of shared \emph{build recipes}.
-As one o the next steps in the working group we will try to lay this out in more concrete.
+As one of the next steps in the working group we will try to lay this out in a more concrete way.
 
 \section*{Acknowledgements}
-We would like to thank all people participating in the constructive and lively discussions.
-In particular Jim Amundson, Marco Clemencic, Giulio Eulisse, Ben Morgan, Pere Mato, Shahzad Muzaffar, and Brett Viren.
+We would like to thank all people participating in the constructive and lively discussions. In particular James Amundson, Marco Clemencic, Giulio Eulisse, Ben Morgan, Pere Mato, and Shahzad Muzaffar.
 
@@
 \bibitem{cmsbuild} cmsBuild: \url{https://github.com/cmsbuild/cmsdist}
+\bibitem{aliBuild} aliBuild: \url{http://alisw.github.io/alibuild/}
+\bibitem{Spack} Spack: \url{http://www.computer.org/csdl/proceedings/sc/2015/3723/00/2807623.pdf}
+
+\bibitem{contractor} Contractor: \url{https://cdcvs.fnal.gov/redmine/projects/contractor/wiki}
 \bibitem{homebrew} Homebrew: \url{http://brew.sh/}

### 2016-02-29 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
Updates and GOCDB section

@@
 
-This technical note describes components of GridPP's Vacuum Platform for 
-managing VMs, including the \texttt{\$JOBOUTPUTS}, VacQuery, and VacUserData 
-interfaces.
+This technical note describes components of the Vacuum Platform developed by
+GridPP for managing VMs, including the \texttt{\$JOBOUTPUTS}, VacQuery, and 
+VacUserData interfaces.
 
@@
 about whether resources will be provided or not. Typically, this is owner 
-of an OpenStack tenancy managed by Vcycle or the manager of Vac VM 
+of an OpenStack or other cloud tenancy managed by Vcycle or the manager of Vac VM 
 factories. Beyond this may be many layers of resource provision
@@
 As not all IaaS cloud systems provide metadata services, VMs and VMLMs 
-should also implement the standard VacUserData substitutions described 
+should also implement the VacUserData substitutions described 
 in section~\ref{sec:vacuserdata}. These include substitutions giving the 
@@
 
-The Machine/Job Features (MJF) mechanism described in \cite{HSF-MJF} allows
+The Machine/Job Features (MJF) mechanism described in \cite{HSFMJF} allows
 resource providers to communicate information to batch jobs and virtual
@@
 
-Resource providers in the Vacuum Platform must make the MJF
+Resource providers using the Vacuum Platform must make the MJF
 \texttt{\$MACHINEFEATURES} and \texttt{\$JOBFEATURES} locations
 available over HTTP(S) rather than through a shared filesystem, and
-should publish the URLs of these locations in OpenStack/EC2 metadata 
+should publish the URLs of these locations in OpenStack/EC2 
+\texttt{machinefeatures} and \texttt{jobfeatures} metadata 
 tags and using the VacUserData substitutions in the user\_data files
@@
 be given in the same way as the \texttt{\$MACHINEFEATURES} and 
-\texttt{\$JOBFEATURES} URLs.
+\texttt{\$JOBFEATURES} URLs, using a VacUserData substitution and an
+OpenStack/EC2 \texttt{joboutputs} metadata key.
 
@@
 in case of problems. All of these files must have unique names, and
-they are all written to the same level (``directory'') of the URL
+are all written to the same level (``directory'') of the URL
 space on the \texttt{\$JOBOUTPUTS} HTTP(S) server. This mechanism is also used 
@@
 their VMs, which VMLMs should use. VMLMs should support both standard CAs
-and IGTF endorsed-CAs when verifying the X.509 certificates used by the
-relevant HTTPS webserver.
+and Internationl Grid Trust Federation (IGTF) endorsed CAs when verifying
+the X.509 certificates used by the relevant HTTPS webserver.
 
@@
 made aware of this situation when it arises, but for scalability reasons,
-the VMLM should not rely on the experiment suffering from VMs starting and
-then failing 
+the VMLM should not rely on the experiment suffering from VMs failing 
 due to an out of date VM image and then notifying resource providers.
@@
 and version when making this request to allow experiments to monitor
-which VMLMs are in use. The VMLM should support both standard CAs
-and IGTF endorsed-CAs when verifying the X.509 certificates used by the
+which VMLM versions are in use. The VMLM should support both standard CAs
+and IGTF-endorsed CAs when verifying the X.509 certificates used by the
 relevant HTTPS webserver.
@@
 
-The VMLM must also provide a mechanism for the resource provider which uses
-the VMLM to specify strings or files which will used in substitutions
-required by the VM. These patterns take the form
+The VMLM must also provide a mechanism for the resource provider
+to specify strings or files whose static values will be used in pattern 
+substitutions required by the VM. These patterns take the form
 \texttt{\#\#user\_data\_option\_XXX\#\#} where XXX is an arbitrary string
@@
 
+The VM must not assume that any other grid, HEP middleware, or scripts
+are running as part of the VMLM and able to provide dynamic values for
+pattern substitutions. For example, it
+must not require that resource providers provide proxies with VOMS
+attributes to the VM. If this is needed, the VM should use
+the proxy provided to obtain the VOMS credentials itself, using software 
+managed by the experiment within the VM.
+
 \section{VacQuery}
@@
 
+The VacQuery protocol specifies queries and status messages
+which can be sent over UDP as short JSON documents.
+
 The principal use of the VacQuery protocol is to allow Vac factories to
 gather information from their neighbours about what VMs are running for
-what machinetypes. This done using the machinetypes\_query and
+what machinetypes. This is done using the machinetypes\_query and
 machinetype\_status UDP messages. Factory and machine message pairs
@@
 in packets directed to the unused UDP port 
-995\footnote{In Roman numerals, V=5 and M=1000. 995 could be written
+995.\footnote{In Roman numerals, V=5 and M=1000. 995 could be written
 as VM = 1000 - 5, although this violates conventions invented in 
-modern times.} The protocol has been designed to keep JSON message and
-IP header below the ethernet MTU of 1500 bytes to avoid fragmentation
-on local networks. Responses are sent to the UDP port from which the
-query was sent.
+modern times.} Responses are sent to the UDP port from which the
+query was sent. The protocol has been designed to keep JSON messages and
+IP headers below the ethernet MTU of 1500 bytes to avoid fragmentation
+on local networks.
 
@@
 \item[message\_type] ``factory\_query''
-\item[vac\_version] Software version of Vac
-\item[vacquery\_version] Version of the VacQuery protocol
+\item[vac\_version] Name and software version of Vac
+\item[vacquery\_version] Name and version of the VacQuery protocol
 \item[space] Vac space name
@@
 \item[message\_type] ``factory\_status''
-\item[vac\_version] Software version of Vac
-\item[vacquery\_version] Version of the VacQuery protocol
+\item[vac\_version] Name and software version of Vac
+\item[vacquery\_version] Name and version of the VacQuery protocol
 \item[cookie] Matching the value supplied by the recipient
@@
 \item[message\_type] ``machines\_query''
-\item[vac\_version] Software version of Vac
-\item[vacquery\_version] Version of the VacQuery protocol
+\item[vac\_version] Name and software version of Vac
+\item[vacquery\_version] Name and version of the VacQuery protocol
 \item[space] Vac space name
@@
 \item[cpu\_percentage] Recent CPU percentage use. May be over 100\% for mulitprocessor VMs
-\item[hs06] Total HEPSPEC06 for the number of processors assigned to this VM
+\item[hs06] Total HEPSPEC06 for the processors assigned to this VM
 \item[machinetype] Name of the machinetype
@@
 \item[message\_type] ``machinetypes\_query''
-\item[vac\_version] Software version of Vac
-\item[vacquery\_version] Version of the VacQuery protocol
+\item[vac\_version] Name and software version of Vac
+\item[vacquery\_version] Name and version of the VacQuery protocol
 \item[space] Vac space name
@@
 Ganglia-style monitoring of individual sites or groups of sites.
-As VacQuery messages as JSON documents, they may be conveniently 
+As VacQuery messages are sent as JSON documents, they may be conveniently 
 recorded in data stores such as ElasticSearch.
@@
 VMLMs should support reporting of usage to the central APEL service
-with messages of the type ``APEL-individual-job-message''.
+with messages of the type ``APEL-individual-job-message''. These are
+the records used for conventional grid sites, rather than those developed
+for cloud resources.
 
-In this case, VLMs must include the following in the messages:
+VLMs must include the following in the messages:
 \begin{quote}
@@
 
-In addition to APEL records generated by the VMLM, an underlying
-cloud infrastracture may also be instrumented to submit APEL usage records
-to the cental APEL service.
+In addition to grid-style APEL records generated by the VMLM, an underlying
+cloud infrastracture may also be instrumented to submit cloud-style APEL 
+usage records to the cental APEL service.
 This is especially likely at sites participating in the EGI Federated Cloud.
@@
 by disabling reporting from the VMLM to the central APEL service.
+
+\section{GOCDB}
+\label{sec:gocdb}
+
+Spaces should be registered in the GOCDB entries for the site, using 
+appropriate service types. The service types \texttt{uk.ac.gridpp.vac} and 
+\texttt{uk.ac.gridpp.vcycle} have been created for Vac and Vcycle
+spaces.
+
+Registration allows new Vacuum Platform resources to be discovered more easily
+by experiments, and permits the declaration of downtimes for these services.
 
@@
 
-\section*{References}
+%\section*{References}
 
-%\begin{thebibliography}{9}
+\begin{thebibliography}{9}
 %% Use references in the format expected by JPCS (as used for CHEP proceedings)
-%
+
 %\bibitem{SHOULD-MUST} S. Bradner, RFC2119 ``Key words for use in RFCs to Indicate Requirement Levels'' (Internet Engineering Task Force)
-%
-%\bibitem{IETF-RFC} S. Bradner, RFC2026 ``The Internet Standards Process -- Revision 3'' (Internet Engineering Task Force)
-%
-%\end{thebibliography}
+
+\bibitem{HSFMJF} M.~Alef et al, HSF-TN-2016-02 ``Machine/Job Features Specification'' (HEP Software Foundation)
+
+\end{thebibliography}
 

### 2016-03-05 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
Spell out cvmfs proxy pattern

@@
 
+If the VM requires the address(es) of HTTP proxies to use with CernVM-FS,
+it must expect this value as the special pattern 
+\texttt{\#\#user\_data\_option\_cvmfs\_proxy\#\#} The VM must be able to
+accept compound CernVM-FS proxy expressions containing semicolon and pipe 
+characters. Typically this will involve placing the substitution pattern
+in appropriate quotation marks within the user\_data template.
+
 If the VM requires an X.509 proxy, it must expect that the special pattern

### 2016-03-14 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
Improved wording

@@
 of an OpenStack or other cloud tenancy managed by Vcycle or the manager of Vac VM 
-factories. Beyond this may be many layers of resource provision
-in terms of legal owners of services and hardware and even funding agencies.
+factories. The term is not used here to refer to higher or lower layers of resource
+provision in terms of legal owners of services and hardware, funding
+agencies, operators of the site infrastructure etc.
 
@@
 which are cooperating to achieve a set of target shares is referred to
-as a ``space''. This is equivalent to a Compute Element at a Grid site,
-and spaces must be given DNS names in DNS space available to the
+as a ``space''. This is equivalent to a Compute Element (CE) at a Grid site,
+and spaces must be given CE-style DNS names in DNS space available to the
 resource provider. However, it is not necessary to register the 
@@
 be said to be neighbours. For Vcycle, a space corresponds to an 
-OpenStack tenancy or project, with a specified endpoint to contact
-and identity tokens to use.
+OpenStack or similar tenancy or project, with a specified endpoint to 
+contact and identity tokens to use.
 

### 2016-03-14 HSF-TN/draft-2016-PKG/draft-HSF-TN-2016-PKG.tex
PKG 2015->2016

--- 2016-02-26 HSF-TN/2016-01/HSF-TN-2016-02.tex
@@
 \\
- & & HSF-TN-2015-XXX \\  % ID 
+ & & HSF-TN-2016-XXX \\  % ID
  & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
@@
 The working group consisted of members of the HEP community from the energy and intensity frontiers as well as accelerator modeling.
-The group met in a series of meetings over the months of May and June~\cite{meetings}.
+The group met in a series of meetings over the months of May and June 2015~\cite{meetings}.
 Additional discussions took place via the public \href{https://groups.google.com/forum/#!forum/hep-sf-packaging-wg}{packaging working group mailing list}.
@@
 standard.  However, some packages may be built with optimization and no
-debug symbols and others, vice versa.
+debug symbols and others, vice versa. Debug symbols and the choice of a
+particular C++ standard are the most used variants in HEP.
+
+\item[platform] denotes the combination of operating system, hardware
+architecture and chosen compiler and compiler version. Many HEP experiments
+define their own naming conventions for the \textit{platform} dimension in this
+space, and they are all equally valid but inhibit sharing due to trivial
+language differences.  A sub-goal of this working group is to agree on a field
+wide naming convention, in form of another technical
+note~\cite{naming-conventions}.
+
+\item[dependencies] of a particular build of a project can be of different
+version and variant. The chosen setup of the dependencies are thus
+another dimension in the configuration space.
 
@@
 
-A single (\textit{project,version,variant}) point in this space
-represents one built \textit{package} (for now, ignoring details of file
+For the purpose of this note a single (\textit{project,version,variant,platform,dependencies})
+point in this space is called a built \textit{package} (for now, ignoring details of file
 system layout, bundling, distribution, run-time environment).  The
@@
 
-This picture allows the expression of some required and desired featues
+This picture allows the expression of some required and desired features
 of a packaging system:
@@
 existing one (eg, upgrade ROOT to new version, make debug/opt versions
-of some packages, use native compiler or build one as part of the tree).
+of some projects, use native compiler or build one as part of the tree).
 
 \end{itemize}
-Many HEP experiments define their own naming conventions for the \textit{variant} dimension in this space, and they are all equally valid but inhibit sharing due to trivial language differences.
-A sub-goal of this working group is to agree on a field wide naming convention.
-The HSF Packaging group has a draft proposal, available here: https://github.com/HEP-SF/documents/blob/master/HSF-TN/draft-2015-NAM/draft-HSF-TN-2015-NAM.pdf
-% OK there is probably a better way to reference this but I don't know it... -LSK
 
@@
 Software stacks are needed by many organizations most notably HEP experiments.
-If a package is authored by the organization it is not considered an external package however it still will need build recipes.
-Any system that builds external packages into a coherent whole must also be capable of building the internal packages as well.
-In the area of external packages it will be useful to distinguish between ``foreign'' and ``domain'' external packages.
-Domain packages are authored by groups within the field of HEP and therefor the HSF could influence the behavior of the domain packages.
-The foreign packages have to be dealt with as they exist.
-Examples of foreign packages include: gcc, fftw3, eigen and lapack.
-Examples of domain packages include clhep and ROOT.
-In the following a few of the solutions used within the community are summarized.
-For more details we refer to the project websites.
+If a project is authored by the organization it is not considered an external
+project however it still will need build recipes. Any system that builds
+external projects into a coherent whole must also be capable of building the
+internal projects as well.  In the area of external projects it will be useful
+to distinguish between ``foreign'' and ``domain'' external projects. Domain
+projects are authored by groups within the field of HEP and therefor the HSF
+could influence the behavior of the domain projects. The foreign projects have
+to be dealt with as they exist. Examples of foreign projects include:
+gcc, fftw3, eigen and lapack. Examples of domain projects include clhep and
+ROOT. In the following a few of the solutions used within the community are
+summarized. For more details we refer to the project websites.
 
 \subsection{Solution 1 -- Worch}
-\emph{Worch}\cite{worch} is the build tool being proposed for the DUNE collaboration.
-It has been used to build the software suite common to many liquid argon experiments, called larsoft, as well as the experiment specific code that builds on top of larsoft.
+\emph{Worch}\cite{worch} is the build tool being proposed for the DUNE
+collaboration.  It has been used to build the software suite common to many
+liquid argon experiments, called larsoft, as well as the experiment specific
+code that builds on top of larsoft.
 
@@
 
-\emph{LCGCMake}\cite{lcgcmake} is the build tool used for creating the LCG releases used by the LHC experiments ATLAS and LHCb.
-It is based on a set of CMake macros and each external package is described using one of these macros in a semi-declarative style.
+\emph{LCGCMake}\cite{lcgcmake} is the build tool used for creating the LCG
+releases used by the LHC experiments ATLAS and LHCb. It is based on a set of
+CMake macros and each external project is described using one of these macros
+in a semi-declarative style.
 
@@
 
-\emph{cmsBuild}\cite{cmsbuild} is the build tool used for making releases of CMSSW for CMS.
-It orchestrates the building of the whole stack from the compiler on up on an as needed basis for any particular release/architecture combination.
-It is based on rpm build, python scripts, and text files that specify dependencies, library names and other optional information.
+\emph{cmsBuild}\cite{cmsbuild} is the build tool used for making releases of
+CMSSW for CMS. It orchestrates the building of the whole stack from the
+compiler on up on an as needed basis for any particular release/architecture
+combination.  It is based on rpm build, python scripts, and text files that
+specify dependencies, library names and other optional information.
 
@@
 \emph{aliBuild}\cite{aliBuild} is an evolution of cmsBuild that fixes many of the weakness of cmsBuild.
-Relative to the above it is not dependent on rpm and builds tar-files that can be currated into a number of diffent package managers such as yum, and apt.
+Relative to the above it is not dependent on rpm and builds tar-files that can be curated into a number of diffent package managers such as yum, and apt.
 % I need to get Giulio to sign off on my short description. -LSK
@@
 \emph{Spack}\cite{Spack} is a packaging system specifically developed for the HPC use case which has a lot in common with the HEP requirements.
-``Spack provides a novel, re-cursive specification syntax to invoke parametric builds of packages and dependencies.
+``Spack provides a novel, recursive specification syntax to invoke parametric builds of packages and dependencies.
 It allows any number of builds to coexist on the same system, and it ensures that installed packages can find their dependencies, regardless of the environment." On creating Spack the authors did a review of existing packaging systems, including a few of those discussed in this document.
@@
 \section{Tool Features and Requirements}
-We compared the feature set of the tools described.
-In evaluating them we defined a set of criteria in which to measure the tools against each other.
-However, one experiment's strict requirement may just be considered a feature for another experiment.
-In the following we explain the comparison criteria and the situation for the individual tools.
+We compared the feature set of the tools described. In evaluating them we
+defined a set of criteria in which to measure the tools against each other.
+However, one experiment's strict requirement may just be considered a feature
+for another experiment. In the following we explain the comparison criteria and
+the situation for the individual tools.
 
 \subsection{Supported Platforms and Environments}
-The most basic criterion is the support for the various platforms and environments used in high-energy physics. There are three categories of environments:
+The most basic criterion is the support for the various platforms and
+environments used in high-energy physics. There are three categories of
+environments:
 \begin{enumerate}
@@
 \end{enumerate}
-The first category is split into several flavours of distributions.
-The main distribution are the RedHat derived Scientific Linux~\cite{sl} and CentOS~\cite{centos}.
-Compared to these, Debian~\cite{debian} and Debian based based distributions play a smaller role in computing centres.
-On desktops, the Debian based Ubuntu~\cite{ubuntu} seems rather popular.
+The first category is split into several flavours of distributions. The main
+distribution are the RedHat derived Scientific Linux~\cite{sl} and
+CentOS~\cite{centos}. Compared to these, Debian~\cite{debian} and Debian based
+distributions play a smaller role in computing centres. On desktops, the Debian
+based Ubuntu~\cite{ubuntu} seems rather popular.
 
-In addition, multiple hardware architectures are in use or will be in use in the foreseeable future - x86, ARM, PowerPC, MIC, and various dedicated super-computers.
-Some of them impose the requirement of cross compilation onto the build and packaging solution, which is why the keyword \textbf{Xcompiler} is added to the platform table.
-However this also refers to the ability to build linux releases on mac, or ability to build mac releases on linux.
-The table~\ref{tab:taxonomyPlatforms} summarizes this information for the tools considered by the working group so far.
-Both Linux and MacOS X are supported by all of the tools, while Windows is not supported at all.
-X-compilation does not seem a priority of the projects yet.
+In addition, multiple hardware architectures are in use or will be in use in the
+foreseeable future - x86, ARM, PowerPC, MIC, and various dedicated
+super-computers. Some of them impose the requirement of cross compilation onto
+the build and packaging solution, which is why the keyword \textbf{Xcompiler}
+is added to the platform table.  However this also refers to the ability to
+build linux releases on mac, or ability to build mac releases on linux.
+The table~\ref{tab:taxonomyPlatforms} summarizes this information for the tools
+considered by the working group so far. Both Linux and MacOS X are supported by
+all of the tools, while Windows is not supported at all. X-compilation does not
+seem a priority of the projects yet.
 
@@
 \item \textbf{Multi-Rel}: The support for building and installing multiple stacks in parallel. This is important if multiple versions of experiment software need to coexist on a given installation.
-\item \textbf{Multi-BuildVar}: The support for multiple varients of the same package, release pair within the same stack.
-There are many possible reasons why this maybe needed, what they have in common is the need to specify compiler switches that have to be applied consistently across the build, like a debugging option or a compiler dialect.
+\item \textbf{Multi-BuildVar}: The support for multiple variants of the same project, release pair within the same stack.
+There are many possible reasons why this may be needed, what they have in common is the need to specify compiler switches that have to be applied consistently across the build, like a debugging option or a compiler dialect.
 \item \textbf{MultiShell-RTE}: The support for setting up the runtime environment for a given stack triplet (project,version,variant), supporting multiple shell flavours.
@@
 &Multi-Rel & Multi-BuildVar & MultiShell-RTE & Relocation \\ \hline
-aliBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} & \cellcolor{green} + \\
-cmsBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} & \cellcolor{green} + \\
+aliBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + \\
+cmsBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + \\
 Contractor & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA  &  \cellcolor{red} - \\
@@
 
-The findings are summarized in Table~\ref{tab:taxonomyUsability}.
-The dependencies of most of the tools seem well under control.
-The addition of new packages is of similar complexity in all cases.
-Only the ease of bootstrapping differs.
-For most of them it is a simple checkout.
-However, others like Nix an entire (fake-)root environment is necessary.
-The documentation of the domain specific packages is in general very poor, the documentation of the non-HEP tools surprisingly good and complete.
+The findings are summarized in Table~\ref{tab:taxonomyUsability}. The
+dependencies of most of the tools seem well under control. The addition of new
+packages is of similar complexity in all cases. Only the ease of bootstrapping
+differs. For most of them it is a simple checkout. However, for others like Nix
+an entire (fake-)root environment is necessary. The documentation of the domain
+specific projects is in general very poor, the documentation of the non-HEP
+tools surprisingly good and complete.
 
@@
 \subsection{Other Criteria}
-During the meetings proponents brought up other creiteria that are not so easy to group together.
-Some might consider these requirements, other might disagree.
+During the meetings proponents brought up other criteria that are not so easy to group together.
+Some might consider these requirements, others might disagree.
 For now we've just listed them as ``other''.
@@
 \begin{enumerate}
-\item \textbf{Performance}: The relevant metrics are a.
-length of time to build the entire stack, this includes support for parallel builds.
-b.
-length of time to incrementally build a developer defined subset of the stack.
-c.
-Ability to reuse binary packages to speedup builds on different machines.
-%BH: I merged Parallel + Performance again
+\item \textbf{Performance}: The relevant metrics are a. Length of time to build
+the entire stack, this includes support for parallel builds. b. Length of time
+to incrementally build a developer defined subset of the stack. c. Ability to
+reuse binary packages to speedup builds on different machines.
 \item \textbf{Sys-Reuse}: the ability of the build system to reuse parts of the system software that is being built if desired.
@@
 \begin{thebibliography}{9}
-\bibitem{meetings} Indico agendas of \href{https://indico.cern.ch/event/373973/}{25.2.2015}, \href{https://indico.cern.ch/event/398344/}{2.6.2015}, \href{https://indico.cern.ch/event/400272/}{9.6.2015}, \href{https://indico.cern.ch/event/402229/}{16.6.2015}, \href{https://indico.cern.ch/event/403790/}{23.6.2015}
-
+\bibitem{meetings} Indico agendas of
+\href{https://indico.cern.ch/event/373973/}{25.2.2015},
+\href{https://indico.cern.ch/event/398344/}{2.6.2015},
+\href{https://indico.cern.ch/event/400272/}{9.6.2015},
+\href{https://indico.cern.ch/event/402229/}{16.6.2015},
+\href{https://indico.cern.ch/event/403790/}{23.6.2015},
+\href{https://indico.cern.ch/event/457365/}{4.11.2015},
+\href{https://indico.cern.ch/event/462334/}{18.11.2015},
+\href{https://indico.cern.ch/event/484006/}{10.2.2016}
+\bibitem{naming-conventions}Naming convention TN listed on \url{http://hepsoftwarefoundation.org/technical_notes.html}
 \bibitem{worch} Worch: \url{https://github.com/brettviren/worch}
@@
 \bibitem{cmsbuild} cmsBuild: \url{https://github.com/cmsbuild/cmsdist}
+\bibitem{contractor} Contractor: \url{https://cdcvs.fnal.gov/redmine/projects/contractor/wiki}
 \bibitem{aliBuild} aliBuild: \url{http://alisw.github.io/alibuild/}
-\bibitem{Spack} Spack: \url{http://www.computer.org/csdl/proceedings/sc/2015/3723/00/2807623.pdf}
-
-\bibitem{contractor} Contractor: \url{https://cdcvs.fnal.gov/redmine/projects/contractor/wiki}
 \bibitem{homebrew} Homebrew: \url{http://brew.sh/}
 \bibitem{nix} Nix: \url{https://nixos.org/nix/}
+\bibitem{Spack} Spack: \url{http://www.computer.org/csdl/proceedings/sc/2015/3723/00/2807623.pdf}
 

### 2016-04-02 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
VacQuery 0.4 changes

@@
 All dates/times in VacQuery messages are expressed as Unix seconds. That
-is, the number of seconds since 00:00:00 1st Jan 1970.
+is, the integer number of seconds since 00:00:00 1st Jan 1970.
 
@@
 \item[site] Name of the site registered in the GOCDB, or the Vac space name if the site is not registered
-\item[total\_cpus] Number of (logical) processors available to VMs
 \item[running\_cpus] Number of processors assigned to running VMs
 \item[running\_machines] Number of running VMs
-\item[total\_machines] Maximum possible number of VMs
-\item[total\_hs06] Total HS06 available to VMs
+\item[running\_hs06] Total HS06 of running VMs
+\item[max\_cpus] Maximum number of (logical) processors available to VMs
+\item[max\_machines] Maximum possible number of VMs
+\item[max\_hs06] Maximum HS06 available to all VMs
 \item[boot\_time] The time when the factory booted up in Unix seconds
@@
 \item[machinetype] Name of the machinetype
-\item[total\_hs06] Total HEPSPEC06 of all the processors allocated to running VMs for this machinetype on this factory
-\item[total\_machines] Number of running VMs for this machinetype on this factory
+\item[running\_hs06] Total HEPSPEC06 of all the processors allocated to running VMs for this machinetype on this factory
+\item[running\_machines] Number of running VMs for this machinetype on this factory
+\item[running\_cpus] Number of CPUs allocated to running VMs for this machinetype on this factory
 \item[num\_before\_fizzle] Number of running VMs which have not yet reached fizzle\_seconds

### 2016-04-06 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
Do not block 169.254.169.253!

@@
 
-Where possible, VMLMs should provide an approximation of OpenStack's 
-environment for VMs, which is derived from EC2. Any contexualization 
+Where possible, an approximation of OpenStack's environment for VMs, which 
+is derived from EC2, should be provided. Any contexualization 
 user\_data file required and a metadata service should be provided via a 
@@
 
-VMLMs must also support VMs which use Cloud Init contextualization.
+VMLMs must ensure they support VMs which use Cloud Init contextualization.
+
+VMs must not block access to the address 169.254.169.253, which is used
+by Amazon for DNS and by Vac for the default route, the local DNS, and 
+for the Machine/Job Features HTTP service.
 

### 2016-04-06 HSF-TN/2016-03/HSF-TN-2016-03.tex
PKG -> 03

--- 2016-03-14 HSF-TN/draft-2016-PKG/draft-HSF-TN-2016-PKG.tex
@@
 \\
- & & HSF-TN-2016-XXX \\  % ID
- & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
+ & & HSF-TN-2016-03 \\  % ID
+ & & March 23, 2016 \\ % Date - Can also hardwire e.g.: 23 March 2010
  & & \\
@@
 package is in general itself multi-dimensional.  Between any given set
-of (package,version) points there is some amount of degeneracy in the
+of (project,version) points there is some amount of degeneracy in the
 variant dimension.  For example, it is common now that all C++ packages

### 2016-10-04 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
.

@@
 
+\usepackage{float}
 \usepackage{microtype}
@@
 %%%% Uncomment next 2 lines if desired
-%\tableofcontents
+\tableofcontents
 %\cleardoublepage
@@
 %% Comment before a final submission.
-\linenumbers
+%\linenumbers
 
@@
 This note is written in terms of VMs, but the interfaces have been 
-designed to be generalised to other forms of logical machine in the
+designed to be generalise to other forms of logical machine in the
 future, such as Docker containers and unikernels.
@@
 contact and identity tokens to use.
+
+Each space is occupied by VMs which are instances of one or more 
+``machinetype'' that the VMLM is able to create. Each machinetype
+corresponds to a specific combination of VM boot image and 
+contextualization.
 
@@
 \begin{center}
-\begin{table}[h]
+\begin{table}[H]
 \label{table:shutdowncodes}
@@
 PEM encoded X.509 certificates and RSA private key which compromise the
-proxy. VMLM's should provide a mechanism for creating X.509 proxies
+proxy. VMLMs should provide a mechanism for creating X.509 proxies
 dynamically for each VM instance from a host or robot certificate owned
@@
 \setlength{\parskip}{0pt}
-\item[message\_type] ``factory\_status''
+\item[message\_type] ``machine\_status''
 \item[vac\_version] Name and software version of Vac
@@
 \setlength{\parskip}{0pt}
-\item[message\_type] ``factory\_status''
+\item[message\_type] ``machinetype\_status''
 \item[vac\_version] Name and software version of Vac
@@
 
-\section{GLUE2}
-\label{sec:glue2}
+%\section{GLUE2}
+%\label{sec:glue2}
+%
+%VMLMs should publish status information as JSON documents using the GLUE 2.0 
+%schema, at an HTTPS URL advertised in the GOCDB service entry for the space,
+%in the Grid Information / URL field.
 
-VMLMs should publish status information as JSON documents using the GLUE 2.0 
-schema, at an HTTPS URL advertised in the GOCDB service entry for the space,
-in the Grid Information / URL field.
-
-\clearpage
+%\clearpage
 
@@
 
-Vacuum platform interfaces have been described ...
+This note has described the various interfaces between virtual machines and
+virtual machine lifecycle managers required by the Vacuum Platform. Three new
+interfaces (\$JOBOUTPUTS, VacQuery, and VacUserData) have been introduced,
+and requirements set out for the use of Machine/Job Features, APEL, and
+GOCDB with the platform. Procedures for how experiments providing VMs should
+present their VM boot images and contextualization are also explained.
 

### 2017-01-12 HSF-TN/2016-03/HSF-TN-2016-03.tex

@@
  & & HSF-TN-2016-03 \\  % ID
- & & March 23, 2016 \\ % Date - Can also hardwire e.g.: 23 March 2010
+ & & January 12, 2017 \\ % Date - Can also hardwire e.g.: 23 March 2010
  & & \\
@@
 \begin{center}
-L.~Sexton-Kennedy$^1$, B.~Hegner$^2$, B.~Viren$^3$
+L.~Sexton-Kennedy$^1$, B.~Hegner$^2$, B.~Viren$^3$, G.~Amadio$^4$
 \bigskip\\
 {\it\footnotesize
-$ ^1$FNAL,$^2$CERN,$^3$BNL
+$^1$FNAL,$^2$CERN,$^3$BNL,$^4$UNESP
 }
@@
 
+\subsection{Solution 10 - Portage }
+\emph{Portage}\cite{portage} is a GPLv2 package management system
+based on \href{https://en.wikipedia.org/wiki/Ports_collection}{ports collections} from BSD-based operating systems.
+It was originally created for Gentoo Linux, but is also used by ChromeOS
+and CoreOS, among others. Portage packages, called \emph{ebuilds},
+contain build instructions written as bash shell scripts with a special
+syntax. Portage itself is written in a combination of python and bash.
+Portage can be used to manage software on Linux and several other Unix
+variants, including macOS (see \cite{prefix} for more details).
+
 \subsection{Commonalities}
@@
 Spack & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} +\\
-Worch &  \cellcolor{green} + &  \cellcolor{green}  + & \cellcolor{red} -  &  \cellcolor{yellow} o
+Worch &  \cellcolor{green} + &  \cellcolor{green}  + & \cellcolor{red} -  &  \cellcolor{yellow} o \\
+Portage &  \cellcolor{green} + &  \cellcolor{green}  + & \cellcolor{red} - &  \cellcolor{green} +
 \end{tabular}
@@
 Spack & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA & \cellcolor{red} -\\
-Worch &  \cellcolor{green} + &  \cellcolor{green}  + &  \cellcolor{green} + &  \cellcolor{green} +
+Worch &  \cellcolor{green} + &  \cellcolor{green}  + &  \cellcolor{green} + &  \cellcolor{green} + \\
+Portage & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA & \cellcolor{red} -
 \end{tabular}
@@
 Worch &   \cellcolor{green} Waf,Python &  \cellcolor{green}  Text-files & \cellcolor{green} + & \cellcolor{yellow} o\\
+Portage & \cellcolor{green} Python & \cellcolor{green} shell script & \cellcolor{green} +& \cellcolor{green} +
 \end{tabular}
@@
 Spack & \cellcolor{yellow} o & \cellcolor{yellow} o  & \cellcolor{green} + & \cellcolor{green} +& \cellcolor{green} +\\
-Worch & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{red} -  & \cellcolor{yellow} o & \cellcolor{green} +
+Worch & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{red} -  & \cellcolor{yellow} o & \cellcolor{green} + \\
+Portage & \cellcolor{green} + & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} +& \cellcolor{green} +
 \end{tabular}
@@
 \href{https://indico.cern.ch/event/462334/}{18.11.2015},
-\href{https://indico.cern.ch/event/484006/}{10.2.2016}
+\href{https://indico.cern.ch/event/484006/}{10.2.2016},
+\href{https://indico.cern.ch/event/581338/}{2.11.2016}
 \bibitem{naming-conventions}Naming convention TN listed on \url{http://hepsoftwarefoundation.org/technical_notes.html}
@@
 \bibitem{Spack} Spack: \url{http://www.computer.org/csdl/proceedings/sc/2015/3723/00/2807623.pdf}
-
+\bibitem{portage} Portage: \url{https://wiki.gentoo.org/wiki/Project:Portage}
+\bibitem{prefix} Gentoo Prefix project \url{https://wiki.gentoo.org/wiki/Project:Prefix}
 \bibitem{sl} Scientific Linux: \url{https://www.scientificlinux.org/}
@@
 \bibitem{debian} Debian: \url{https://www.debian.org/}
+\bibitem{gentoo} Gentoo: \url{https://www.gentoo.org/}
 \bibitem{ubuntu} Ubuntu: \url{https://www.ubuntu.com/}

### 2017-12-14 CWP/papers/roadmap/authors/a2tex.py
First version of author processing script
TODO: personal footnotes, institute addresses

@@ -1,0 +1,68 @@
+#! /usr/bin/env python
+#
+# Process CWP signatory list into a LaTeX author list
+
+import os
+import re
+import sys
+
+def main():
+    author_file = "authors.txt"
+    footnote_file = "footnotes.txt"
+
+    # Open and process the authors file
+    with open(author_file) as author_fh, open(footnote_file) as footnote_fh:
+        author_list = []
+        for author_line in author_fh:
+            if author_line.startswith("#"):
+                continue
+            author_elements = [ el.strip() for el in author_line.split(" - ") ]
+            try:
+                surname, forename = [ el.strip() for el in author_elements[0].split(",", 1) ]
+                affiliation = author_elements[1]
+                if len(author_elements) == 3:
+                    role = author_elements[2]
+                else:
+                    role = ""
+                author_list.append( {"surname": surname,
+                     "forename": forename,
+                     "affiliation": affiliation.replace("&", "\&"),
+                     "role": role.replace("&", "\&"),}
+                    )
+            except:
+                print "Got an exception processing this line:", author_line
+        
+        print author_list
+
+    # Generate the map to people's institutes
+    affiliation_map = {}
+    for author in author_list:
+        if author["affiliation"] in affiliation_map:
+            continue
+        affiliation_map[author["affiliation"]] = {
+            "address": author["affiliation"],
+            }
+    # Alphabetise and assign footnote marks
+    sorted_affiliations = affiliation_map.keys()[:]
+    sorted_affiliations.sort()
+    for footnote_mark, affiliation in enumerate(sorted_affiliations, start=1):
+        affiliation_map[affiliation]["footnotemark"] = str(footnote_mark)
+    
+        
+    # Write out the latex author file
+    with open("authors.tex", "w") as output:
+        for author in author_list:
+            print >>output, author["surname"] +", "  + \
+            author["forename"] + "$^{" + str(affiliation_map[author["affiliation"]]["footnotemark"]) + "}$" +\
+            ("" if author == author_list[-1] else ";")
+            #if author["role"]:
+            #    print >>output, "(" + author["role"] + ")"
+        print >>output
+        for affiliation in sorted_affiliations:
+            print affiliation
+            print >>output, "\par {\\footnotesize $^{" + str(affiliation_map[affiliation]["footnotemark"]) + "}$", affiliation_map[affiliation]["address"], "}"
+
+
+if __name__ == '__main__':
+    main()
+

### 2017-12-14 CWP/papers/roadmap/authors/a2tex.py
Support multiple affiliations, separaeted by &
Some fixes to badly formatted author entries (also fixed in gdoc)

@@ -2,6 +2,10 @@
 #
 # Process CWP signatory list into a LaTeX author list
 
+# The data source is expected to be 
+#   Lastname, Firstname - Institution (1) [- Role ]
+# If there are multiple affiliations, separate with &
+
 import os
 import re
 import sys
@@ -9,13 +13,22 @@ import sys
 def main():
     author_file = "authors.txt"
     footnote_file = "footnotes.txt"
-
+    affiliation_address_file = "address.txt"
+    
     # Open and process the authors file
     with open(author_file) as author_fh, open(footnote_file) as footnote_fh:
         author_list = []
         for author_line in author_fh:
-            if author_line.startswith("#"):
+            author_line = author_line.strip()
+            if author_line.startswith("#") or author_line == "":
                 continue
+            # Find any footnotes that the author wanted to add
+            footnotes_search = re.search(r"\(([\d,]+)\)$", author_line)
+            if footnotes_search:
+                footnotes = footnotes_search.group(1).split(",")
+                author_line = author_line.replace(footnotes_search.group(), "")
+            else:
+                footnotes = []
             author_elements = [ el.strip() for el in author_line.split(" - ") ]
             try:
                 surname, forename = [ el.strip() for el in author_elements[0].split(",", 1) ]
@@ -26,22 +39,36 @@ def main():
                     role = ""
                 author_list.append( {"surname": surname,
                      "forename": forename,
-                     "affiliation": affiliation.replace("&", "\&"),
-                     "role": role.replace("&", "\&"),}
+                     "affiliation": [ el.strip() for el in affiliation.split("&") ],
+                     "role": role.replace("&", "\&"),
+                     "footnotes": footnotes,}
                     )
-            except:
-                print "Got an exception processing this line:", author_line
-        
-        print author_list
-
+            except Exception as e:
+                print "Got an exception(", e, ") processing this line:", author_line
+                raise
+    #print author_list
+    
     # Generate the map to people's institutes
+    institute_address = {}
+    with open(affiliation_address_file) as address_fh:
+        for affiliation_line in address_fh:
+            affiliation_line = affiliation_line.strip()
+            if affiliation_line.startswith("#") or affiliation_line == "":
+                continue
+            institute, address = affiliation_line.split(": ", 1)
+            institute_address[institute] = address
+    #print institute_address
+    
     affiliation_map = {}
     for author in author_list:
[…]

### 2017-12-14 CWP/papers/roadmap/authors/a2tex.py
Add "footnote" functionality
Start listing institute addresses as well

@@ -16,7 +16,7 @@ def main():
     affiliation_address_file = "address.txt"
     
     # Open and process the authors file
-    with open(author_file) as author_fh, open(footnote_file) as footnote_fh:
+    with open(author_file) as author_fh:
         author_list = []
         for author_line in author_fh:
             author_line = author_line.strip()
@@ -75,20 +75,45 @@ def main():
     for footnote_mark, affiliation in enumerate(sorted_affiliations, start=1):
         affiliation_map[affiliation]["footnotemark"] = str(footnote_mark)
     
+    # Parse and assign footnotes
+    with open(footnote_file) as footnote_fh:
+        footnote_map = {}
+        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
+        footnotemark_index = 0
+        for footnote_line in footnote_fh:
+            footnote_line = footnote_line.strip()
+            if footnote_line.startswith("#") or footnote_line == "":
+                continue
+            number, note = footnote_line.split(" ", 1)
+            number = number.rstrip(".")
+            footnote_map[number] = {
+                "note": note.replace("&", "\&").replace("_", "\_"),
+                "footnotemark": footnote_symbols[footnotemark_index],
+                }
+            footnotemark_index += 1
+    sorted_footnotes = footnote_map.keys()[:]
+    sorted_footnotes.sort(cmp=lambda x,y: cmp(int(x), int(y)))
+    #print footnote_map
+        
         
     # Write out the latex author file
     with open("authors.tex", "w") as output:
         for author in author_list:
             affiliation_list = ",".join([ str(affiliation_map[affiliation]["footnotemark"]) for affiliation in author["affiliation"] ])
+            if author["footnotes"]:
+                footnote_list = ",".join( [ footnote_map[footnote]["footnotemark"] for footnote in author["footnotes"] ])
+                affiliation_list = affiliation_list + "," + footnote_list
             print >>output, author["surname"] +", "  + \
             author["forename"] + "$^{" + affiliation_list + "}$" +\
             ("" if author == author_list[-1] else ";")
             #if author["role"]:
             #    print >>output, "(" + author["role"] + ")"
-        print >>output
+        print >>output, "\\bigskip"
         for affiliation in sorted_affiliations:
-            print affiliation
[…]

### 2017-12-15 CWP/papers/roadmap/authors/a2tex.py
a2text: fixes for Python v3 compliance

@@ -9,6 +9,15 @@
 import os
 import re
 import sys
+from operator import attrgetter
+
+class Footnote():
+    def __init__(self, number_str, text):
+        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
+        self.mark = footnote_symbols[int(number_str)-1]
+        self.text = text
+    def __repr__(self):
+        return repr((self.mark, self.text))
 
 def main():
     author_file = "authors.txt"
@@ -16,7 +25,7 @@ def main():
     affiliation_address_file = "address.txt"
     
     # Open and process the authors file
-    with open(author_file) as author_fh:
+    with open(author_file, encoding='utf-8') as author_fh:
         author_list = []
         for author_line in author_fh:
             author_line = author_line.strip()
@@ -44,13 +53,13 @@ def main():
                      "footnotes": footnotes,}
                     )
             except Exception as e:
-                print "Got an exception(", e, ") processing this line:", author_line
+                print ("Got an exception({}) processing this line: {}".format(e, author_line))
                 raise
     #print author_list
     
     # Generate the map to people's institutes
     institute_address = {}
-    with open(affiliation_address_file) as address_fh:
+    with open(affiliation_address_file, encoding='utf-8') as address_fh:
         for affiliation_line in address_fh:
             affiliation_line = affiliation_line.strip()
             if affiliation_line.startswith("#") or affiliation_line == "":
@@ -70,50 +79,52 @@ def main():
     #print affiliation_map        
     
     # Alphabetise and assign footnote marks
-    sorted_affiliations = affiliation_map.keys()[:]
-    sorted_affiliations.sort()
+    sorted_affiliations = sorted(affiliation_map)
     for footnote_mark, affiliation in enumerate(sorted_affiliations, start=1):
         affiliation_map[affiliation]["footnotemark"] = str(footnote_mark)
     
     # Parse and assign footnotes
-    with open(footnote_file) as footnote_fh:
-        footnote_map = {}
-        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
-        footnotemark_index = 0
+    with open(footnote_file, encoding='utf-8') as footnote_fh:
+        footnote_list = []
         for footnote_line in footnote_fh:
             footnote_line = footnote_line.strip()
             if footnote_line.startswith("#") or footnote_line == "":
                 continue
             number, note = footnote_line.split(" ", 1)
             number = number.rstrip(".")
-            footnote_map[number] = {
[…]

### 2017-12-16 CWP/papers/roadmap/latex/authors.tex
Document as submitted to arXiv 2017-12-15

Alves Jr, Antonio Augusto $^{74}$;
Amadio, Guilherme $^{5}$;
Anh-Ky, Nguyen $^{111}$;
Aphecetche, Laurent $^{60}$;
Apostolakis, John $^{5}$;
Asai, Makoto $^{58,p}$;
Atzori, Luca $^{5}$;
Babik, Marian $^{5}$;
Bagliesi, Giuseppe $^{29}$;
Bandieramonte, Marilena $^{5}$;
Barisits, Martin $^{5}$;
Bauerdick, Lothar A. T. $^{16,c}$;
Belforte, Stefano $^{32}$;
Benjamin, Douglas $^{75}$;
Bernius, Catrin $^{58}$;
Bhimji, Wahid $^{42}$;
Bianchi, Riccardo Maria $^{97}$;
Bird, Ian $^{5}$;
Biscarat, Catherine $^{48}$;
Blomer, Jakob $^{5}$;
Bloom, Kenneth $^{89}$;
Boccali, Tommaso $^{29}$;
Bockelman, Brian $^{89}$;
Bold, Tomasz $^{39}$;
Bonacorsi, Daniele $^{24}$;
Boveia, Antonio $^{94}$;
Bozzi, Concezio $^{26}$;
Bracko, Marko $^{86,37}$;
Britton, David $^{79}$;
Buckley, Andy $^{79}$;
Buncic, Predrag $^{5,a}$;
Calafiura, Paolo $^{42}$;
Campana, Simone $^{5,a}$;
Canal, Philippe $^{16,c}$;
Canali, Luca $^{5}$;
Carlino, Gianpaolo $^{28}$;
Castro, Nuno $^{43,88,d}$;
Cattaneo, Marco $^{5}$;
Cerminara, Gianluca $^{5}$;
Chang, Philip $^{69}$;
Chapman, John $^{70}$;
Childers, Taylor $^{1}$;
Clarke, Peter $^{76}$;
Clemencic, Marco $^{5}$;
Cogneras, Eric $^{46}$;
Collier, Ian $^{56}$;
Corti, Gloria $^{5}$;
Cosmo, Gabriele $^{5}$;
Costanzo, Davide $^{103}$;
Couturier, Ben $^{5}$;
Cranmer, Kyle $^{53}$;
Cranshaw, Jack $^{1}$;
Crépé-Renaudin, Sabine $^{48}$;
Cristella, Leonardo $^{25}$;
Crooks, David $^{79}$;
Dallmeier-Tiessen, Sünje $^{5}$;
De Cian, Michel $^{80}$;
De, Kaushik $^{105}$;
Di Girolamo, Alessandro $^{5}$;
Dimitrov, Gancho $^{5}$;
Doglioni, Caterina $^{84,h}$;
Dotti, Andrea $^{58,p}$;
Duellmann, Dirk $^{5}$;
Duflot, Laurent $^{41}$;
Dykstra, Dave $^{16,c}$;
Dziedziniewicz-Wojcik, Katarzyna $^{5}$;
Dziurda, Agnieszka $^{5}$;
Egede, Ulrik $^{34}$;
Elmer, Peter $^{98,a}$;
Elmsheuser, Johannes $^{2}$;
Elvira, V. Daniel $^{16,c}$;
Eulisse, Giulio $^{5}$;
Ferber, Torben $^{67}$;
Filipcic, Andrej $^{37}$;
Fisk, Ian $^{59}$;
Fitzpatrick, Conor $^{14}$;
Flix, José $^{55,7,g}$;
Formica, Andrea $^{35}$;
Forti, Alessandra $^{85}$;
Gaede, Frank $^{13}$;
Ganis, Gerardo $^{5}$;
Gardner, Robert $^{73}$;
Garonne, Vincent $^{95}$;
Gellrich, Andreas $^{13}$;
Genser, Krzysztof $^{16,c}$;
George, Simon $^{57}$;
Geurts, Frank $^{99}$;
Gheata, Andrei $^{5}$;
Gheata, Mihaela $^{5}$;
Giacomini, Francesco $^{9}$;
Giagu, Stefano $^{101,31}$;
Giffels, Manuel $^{38}$;
Gingrich, Douglas $^{63}$;
Girone, Maria $^{5}$;
Gligorov, Vladimir V. $^{47}$;
Glushkov, Ivan $^{105}$;
Gohn, Wesley $^{81}$;
González Caballero, Isidro $^{96}$;
González Fernández, Juan R. $^{96}$;
Gonzalez Lopez, Jose Benito $^{5}$;
Govi, Giacomo $^{16}$;
Grandi, Claudio $^{24}$;
Grasland, Hadrien $^{41}$;
Gray, Heather $^{42}$;
Grillo, Lucia $^{85}$;
Guan, Wen $^{109}$;
Gutsche, Oliver $^{16,c}$;
Gyurjyan, Vardan $^{36}$;
Hanushevsky, Andrew $^{58,p}$;
Hariri, Farah $^{5}$;
Hartmann, Thomas $^{13}$;
Harvey, John $^{5,a}$;
Hauth, Thomas $^{38}$;
Hegner, Benedikt $^{5,a}$;
Heinemann, Beate $^{13}$;
Heinrich, Lukas $^{53}$;
Hernández, José M. $^{7,g}$;
Hildreth, Michael $^{91,f}$;
Hodgkinson, Mark $^{103}$;
Hoeche, Stefan $^{58,p}$;
Hristov, Peter $^{5}$;
Huang, Xingtao $^{102}$;
Ivanchenko, Vladimir N. $^{5,106}$;
Ivanov, Todor $^{104}$;
Jashal, Brij $^{62}$;
Jayatilaka, Bodhitha $^{16,c}$;
Jones, Roger $^{82,a}$;
Jouvin, Michel $^{41,a}$;
Jun, Soon Yung $^{16,c}$;
Kagan, Michael $^{58,p}$;
Kalderon, Charles William $^{84}$;
Karavakis, Edward $^{5}$;
Katz, Daniel S. $^{72}$;
Kcira, Dorian $^{11}$;
Kersevan, Borut Paul $^{83}$;
Kirby, Michael $^{16,c}$;
Klimentov, Alexei $^{2}$;
Klute, Markus $^{49}$;
Komarov, Ilya $^{32,n}$;
Koppenburg, Patrick $^{52}$;
Kowalkowski, Jim $^{16,c}$;
Kreczko, Luke $^{66}$;
Kuhr, Thomas $^{45,a}$;
Kutschke, Robert $^{16,a,c}$;
Kuznetsov, Valentin $^{12}$;
Lampl, Walter $^{65}$;
Lancon, Eric $^{2}$;
Lange, David $^{98}$;
Lassnig, Mario $^{5}$;
Laycock, Paul $^{5}$;
Leggett, Charles $^{42}$;
Letts, James $^{69}$;
Lewendel, Birgit $^{13}$;
Li, Teng $^{76}$;
Lima, Guilherme $^{16}$;
Linacre, Jacob $^{56,m}$;
Linden, Tomas $^{18}$;
Lo Presti, Giuseppe $^{5}$;
Lopienski, Sebastian $^{5}$;
Love, Peter $^{82}$;
Marshall, Zachary L $^{42}$;
Martelli, Edoardo $^{5}$;
Martin-Haugh, Stewart $^{56}$;
Mato, Pere $^{5}$;
Mazumdar, Kajari $^{62}$;
McCauley, Thomas $^{91}$;
McFayden, Josh $^{5}$;
McKee, Shawn $^{87,l}$;
McNab, Andrew $^{85}$;
Meinhard, Helge $^{5}$;
Menasce, Dario $^{27,a}$;
Mendez Lorenzo, Patricia $^{5}$;
Mete, Alaettin Serhan $^{68}$;
Michelotto, Michele $^{30}$;
Mitrevski, Jovan $^{45}$;
Moneta, Lorenzo $^{5}$;
Morgan, Ben $^{108}$;
Mount, Richard $^{58,p}$;
Moyse, Edward $^{64}$;
Murray, Sean $^{71,10}$;
Neubauer, Mark S $^{72,a,k}$;
Novaes, Sérgio $^{100}$;
Novak, Mihaly $^{5}$;
Oyanguren, Arantza $^{21}$;
Ozturk, Nurcan $^{105}$;
Pacheco Pages, Andres $^{55,19,j}$;
Paganini, Michela $^{110}$;
Pansanel, Jerome $^{33}$;
Pascuzzi, Vincent R. $^{107}$;
Pearce, Alex $^{5}$;
Pearson, Ben $^{50}$;
Pedro, Kevin $^{16,c}$;
Perdue, Gabriel $^{16}$;
Perez-Calero Yzquierdo, Antonio $^{55,7,g}$;
Perrozzi, Luca $^{15}$;
Petersen, Troels $^{51}$;
Petric, Marko $^{5}$;
Piedra, Jónatan $^{20}$;
Piilonen, Leo $^{112,i}$;
Piparo, Danilo $^{5}$;
Pokorski, Witold $^{5}$;
Polci, Francesco $^{47}$;
Potamianos, Karolos $^{13}$;
Psihas, Fernanda $^{23}$;
Raven, Gerhard $^{52}$;
Reuter, Jürgen $^{13}$;
Ribon, Alberto $^{5}$;
Ritter, Martin $^{45}$;
Robinson, James $^{13}$;
Rodrigues, Eduardo $^{74,a,e}$;
Roiser, Stefan $^{5,a}$;
Rousseau, David $^{41}$;
Roy, Gareth $^{79}$;
Sailer, Andre $^{5}$;
Sakuma, Tai $^{66}$;
Sánchez-Hernández, Alberto $^{8,o}$;
Santana, Renato $^{3}$;
Sartirana, Andrea $^{44}$;
Schellman, Heidi $^{54}$;
Schovancová, Jaroslava $^{5}$;
Schramm, Steven $^{78}$;
Schulz, Markus $^{5}$;
Sciabà, Andrea $^{5}$;
Seidel, Sally $^{90}$;
Sekmen, Sezen $^{40}$;
Serfon, Cedric $^{93}$;
Severini, Horst $^{92}$;
Sexton-Kennedy, Elizabeth $^{16,a,c}$;
Seymour, Michael $^{85}$;
Shapoval, Illya $^{42}$;
Shiers, Jamie $^{5}$;
Shiu, Jing-Ge $^{61}$;
Short, Hannah $^{5}$;
Siroli, Gian Piero $^{24}$;
Skipsey, Sam $^{79}$;
Smith, Tim $^{5}$;
Snyder, Scott $^{2}$;
Sokoloff, Michael D $^{74,a}$;
Stadie, Hartmut $^{17}$;
Stark, Giordon $^{6}$;
Stewart, Gordon $^{79}$;
Stewart, Graeme $^{5,a}$;
Templon, Jeff $^{52}$;
Tenaglia, Giacomo $^{5}$;
Tsulaia, Vakhtang $^{42}$;
Tunnell, Christopher $^{6}$;
Vaandering, Eric $^{16,c}$;
Valassi, Andrea $^{5}$;
Vallecorsa, Sofia $^{77}$;
Valsan, Liviu $^{5}$;
Van Gemmeren, Peter $^{1}$;
Vázquez Sierra, Carlos $^{52}$;
Vernet, Renaud $^{4}$;
Viren, Brett $^{2}$;
Vlimant, Jean-Roch $^{11,a}$;
Voss, Christian $^{13}$;
Vuosalo, Carl $^{113}$;
Wartel, Romain $^{5}$;
Wenaus, Torre $^{2}$;
Wenzel, Sandro $^{5}$;
Winklmeier, Frank $^{54}$;
Wissing, Christoph $^{13}$;
Wuerthwein, Frank $^{69}$;
Wynne, Benjamin $^{76}$;
Xiaomei, Zhang $^{22}$;
Yang, Wei $^{58,p}$;
Yazgan, Efe $^{22}$
\bigskip
\par {\footnotesize $^{1}$ High Energy Physics Division, Argonne National Laboratory, Argonne IL, USA}
\par {\footnotesize $^{2}$ Physics Department, Brookhaven National Laboratory, Upton NY, USA}
\par {\footnotesize $^{3}$ Centro Brasileiro de Pesquisas Fisicas, Rio de Janeiro, Brazil}
\par {\footnotesize $^{4}$ Centre de Calcul de l’IN2P3, Villeurbanne, Lyon, France}
\par {\footnotesize $^{5}$ CERN, Geneva, Switzerland}
\par {\footnotesize $^{6}$ Enrico Fermi Institute, University of Chicago, Chicago IL, USA}
\par {\footnotesize $^{7}$ Centro de Investigaciones Energéticas Medioambientales y Tecnológicas (CIEMAT), Madrid, Spain}
\par {\footnotesize $^{8}$ Cinvestav, Mexico City, Mexico}
\par {\footnotesize $^{9}$ Centro Nazionale Analisi Fotogrammi (CNAF), INFN, Bologna, Italy}
\par {\footnotesize $^{10}$ Center for High Performance Computing, Cape Town, South Africa}
\par {\footnotesize $^{11}$ California Institute of Technology, Pasadena, California, USA}
\par {\footnotesize $^{12}$ Cornell University, Ithaca, USA}
\par {\footnotesize $^{13}$ Deutsches Elektronen-Synchrotron, Hamburg, Germany}
\par {\footnotesize $^{14}$ Institute of Physics, Ecole Polytechnique Federale de Lausanne (EPFL), Lausanne, Switzerland}
\par {\footnotesize $^{15}$ ETH Zurich - Institute for Particle Physics and Astrophysics (IPA), Zurich, Switzerland}
\par {\footnotesize $^{16}$ Fermi National Accelerator Laboratory, Batavia, USA}
\par {\footnotesize $^{17}$ University of Hamburg, Hamburg, Germany}
\par {\footnotesize $^{18}$ Helsinki Institute of Physics, Helsinki, Finland}
\par {\footnotesize $^{19}$ Institut de Física d’Altes Energies and Departament de Física de la Universitat Autònoma de Barcelona and ICREA, Barcelona, Spain}
\par {\footnotesize $^{20}$ Instituto de Física de Cantabria (IFCA), CSIC-Universidad de Cantabria, Santander, Spain}
\par {\footnotesize $^{21}$ Instituto de Fisica Corpuscular, Centro Mixto Universidad de Valencia - CSIC, Valencia, Spain}
\par {\footnotesize $^{22}$ Institute of High Energy Physics, Chinese Academy of Sciences, Beijing}
\par {\footnotesize $^{23}$ Department of Physics, Indiana University, Bloomington IN, USA}
\par {\footnotesize $^{24}$ INFN Sezione di Bologna, Università di Bologna, Bologna, Italy}
\par {\footnotesize $^{25}$ INFN Sezione di Bari, Università di Bari, Politecnico di Bari, Bari, Italy}
\par {\footnotesize $^{26}$ Universita e INFN, Ferrara, Ferrara, Italy}
\par {\footnotesize $^{27}$ INFN Sezione di Milano-Bicocca, Milano, Italy}
\par {\footnotesize $^{28}$ INFN Sezione di Napoli, Università di Napoli, Napoli, Italy}
\par {\footnotesize $^{29}$ INFN Sezione di Pisa, Università di Pisa, Scuola Normale Superiore di Pisa, Pisa, Italy}
\par {\footnotesize $^{30}$ INFN Sezione di Padova, Università di Padova b, Padova, Italy}
\par {\footnotesize $^{31}$ INFN Sezione di Roma I, Università La Sapienza, Roma, Italy}
\par {\footnotesize $^{32}$ INFN Sezione di Trieste, Università di Trieste, Trieste, Italy}
\par {\footnotesize $^{33}$ Université de Strasbourg, CNRS, IPHC UMR 7178, F-67000 Strasbourg, France}
\par {\footnotesize $^{34}$ Imperial College London, London, United Kingdom}
\par {\footnotesize $^{35}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat a l’Energie Atomique), Gif-sur-Yvette, France}
\par {\footnotesize $^{36}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
\par {\footnotesize $^{37}$ Jozef Stefan Institute, Ljubljana, Slovenia}
\par {\footnotesize $^{38}$ Karlsruhe Institute of Technology, Karlsruhe, Germany}
\par {\footnotesize $^{39}$ AGH University of Science and Technology, Faculty of Physics and Applied Computer Science, Krakow, Poland}
\par {\footnotesize $^{40}$ Kyungpook National University, Daegu, Republic of Korea}
\par {\footnotesize $^{41}$ LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\par {\footnotesize $^{42}$ Lawrence Berkeley National Laboratory and University of California, Berkeley CA, USA}
\par {\footnotesize $^{43}$ Laboratorio de Instrumentacao e Fisica Experimental de Particulas (LIP), Lisboa, Portugal}
\par {\footnotesize $^{44}$ Laboratoire Leprince-Ringuet, Ecole polytechnique, CNRS/IN2P3, Université Paris-Saclay, Palaiseau, France}
\par {\footnotesize $^{45}$ Fakultät für Physik, Ludwig-Maximilians-Universität München, München, Germany}
\par {\footnotesize $^{46}$ Laboratoire de Physique Corpusculaire, Clermont Université and Université Blaise Pascal and CNRS/IN2P3, Clermont-Ferrand, France}
\par {\footnotesize $^{47}$ LPNHE, Universite Pierre et Marie Curie, Universite Paris Diderot, CNRS/IN2P3, Paris, France}
\par {\footnotesize $^{48}$ Laboratoire de Physique Subatomique et de Cosmologie, Université Joseph Fourier and CNRS/IN2P3 and Institut National Polytechnique de Grenoble, Grenoble, France}
\par {\footnotesize $^{49}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
\par {\footnotesize $^{50}$ Max-Planck-Institut für Physik (Werner-Heisenberg-Institut), München, Germany}
\par {\footnotesize $^{51}$ Niels Bohr Institute, University of Copenhagen, Kobenhavn, Denmark}
\par {\footnotesize $^{52}$ Nikhef National Institute for Subatomic Physics and University of Amsterdam, Amsterdam, Netherlands}
\par {\footnotesize $^{53}$ Department of Physics, New York University, New York, NY, USA}
\par {\footnotesize $^{54}$ Center for High Energy Physics, University of Oregon, Eugene OR, USA}
\par {\footnotesize $^{55}$ Port d’Informació Científica (PIC), Universitat Autònoma de Barcelona (UAB), Barcelona, Spain}
\par {\footnotesize $^{56}$ STFC Rutherford Appleton Laboratory, Didcot, United Kingdom}
\par {\footnotesize $^{57}$ Department of Physics, Royal Holloway University of London, Surrey, United Kingdom}
\par {\footnotesize $^{58}$ SLAC National Accelerator Laboratory, Menlo Park CA, USA}
\par {\footnotesize $^{59}$ Simons Foundation, New York, USA}
\par {\footnotesize $^{60}$ SUBATECH, IMT Atlantique, Université de Nantes, CNRS-IN2P3, Nantes, France}
\par {\footnotesize $^{61}$ National Taiwan University, Taipei, Taiwan}
\par {\footnotesize $^{62}$ Tata Institute of Fundamental Research, Mumbai, India}
\par {\footnotesize $^{63}$ Department of Physics, University of Alberta, Edmonton, AB, Canada}
\par {\footnotesize $^{64}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
\par {\footnotesize $^{65}$ Department of Physics, University of Arizona, Tucson, AZ, USA}
\par {\footnotesize $^{66}$ H.H. Wills Physics Laboratory, University of Bristol, Bristol, United Kingdom}
\par {\footnotesize $^{67}$ Department of Physics, University of British Columbia, Vancouver, BC, Canada}
\par {\footnotesize $^{68}$ Department of Physics and Astronomy, University of California Irvine, Irvine, CA, USA}
\par {\footnotesize $^{69}$ University of California, San Diego, La Jolla, USA}
\par {\footnotesize $^{70}$ Cavendish Laboratory, University of Cambridge, Cambridge, United Kingdom}
\par {\footnotesize $^{71}$ Physics Department, University of Cape Town, Cape Town, South Africa}
\par {\footnotesize $^{72}$ University of Illinois Urbana-Champaign, Champaign, Illinois, USA}
\par {\footnotesize $^{73}$ Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
\par {\footnotesize $^{74}$ University of Cincinnati, Cincinnati, OH, USA}
\par {\footnotesize $^{75}$ Department of Physics, Duke University, Durham, NC, USA}
\par {\footnotesize $^{76}$ SUPA - School of Physics and Astronomy, University of Edinburgh, Edinburgh, United Kingdom}
\par {\footnotesize $^{77}$ Gangneung-Wonju National University, South Korea}
\par {\footnotesize $^{78}$ Section de Physique, Université de Genève, Geneva, Switzerland}
\par {\footnotesize $^{79}$ SUPA - School of Physics and Astronomy, University of Glasgow, Glasgow, United Kingdom}
\par {\footnotesize $^{80}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
\par {\footnotesize $^{81}$ Department of Physics \& Astronomy, University of Kentucky, Lexington, USA}
\par {\footnotesize $^{82}$ Physics Department, Lancaster University, Lancaster, United Kingdom}
\par {\footnotesize $^{83}$ Department of Physics, Jožef Stefan Institute and University of Ljubljana, Ljubljana, Slovenia}
\par {\footnotesize $^{84}$ Fysiska institutionen, Lunds universitet, Lund, Sweden}
\par {\footnotesize $^{85}$ School of Physics and Astronomy, University of Manchester, Manchester, United Kingdom}
\par {\footnotesize $^{86}$ University of Maribor, Ljubljana, Slovenia}
\par {\footnotesize $^{87}$ Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
\par {\footnotesize $^{88}$ Departamento de Fisica, Universidade de Minho, Braga, Portugal}
\par {\footnotesize $^{89}$ University of Nebraska-Lincoln, Lincoln, USA}
\par {\footnotesize $^{90}$ Department of Physics and Astronomy, University of New Mexico, Albuquerque NM, USA}
\par {\footnotesize $^{91}$ University of Notre Dame, Notre Dame, USA}
\par {\footnotesize $^{92}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman OK, USA}
\par {\footnotesize $^{93}$ Department of Physics, University of Oslo, Oslo, Norway}
\par {\footnotesize $^{94}$ The Ohio State University, Columbus, USA}
\par {\footnotesize $^{95}$ Department of Physics, University of Oslo, Oslo, Norway}
\par {\footnotesize $^{96}$ Universidad de Oviedo, Oviedo, Spain}
\par {\footnotesize $^{97}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
\par {\footnotesize $^{98}$ Princeton University, Princeton, USA}
\par {\footnotesize $^{99}$ Rice University, Houston, USA}
\par {\footnotesize $^{100}$ Universidade Estadual Paulista, São Paulo, Brazil}
\par {\footnotesize $^{101}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
\par {\footnotesize $^{102}$ School of Physics, Shandong University, Shandong, China}
\par {\footnotesize $^{103}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
\par {\footnotesize $^{104}$ University of Sofia, Sofia, Bulgaria}
\par {\footnotesize $^{105}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
\par {\footnotesize $^{106}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
\par {\footnotesize $^{107}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
\par {\footnotesize $^{108}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
\par {\footnotesize $^{109}$ University ofWisconsin - Madison, Madison, WI, USA}
\par {\footnotesize $^{110}$ Department of Physics, Yale University, New Haven, CT, USA}
\par {\footnotesize $^{111}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
\par {\footnotesize $^{112}$ Virginia Tech, Blacksburg, Virginia, USA}
\par {\footnotesize $^{113}$ Department of Physics, University of Wisconsin, Madison WI, USA}
\bigskip
\par {\footnotesize $^{a}$ Community White Paper Editorial Board Member}
\par {\footnotesize $^{b}$ Vladimir V. Gligorov acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 724777 “RECEPT”.}
\par {\footnotesize $^{c}$ Supported by the US-DOE, DE-AC02-07CH11359}
\par {\footnotesize $^{d}$ Supported by FCT-Portugal, IF/00050/2013/CP1172/CT0002}
\par {\footnotesize $^{e}$ Supported by the US-NSF, ACI-1450319}
\par {\footnotesize $^{f}$ Supported by the US-NSF, PHY-1607578}
\par {\footnotesize $^{g}$ Supported by ES-MINECO, FPA2016-80994-c2-1-R \& MDM-2015-0509}
\par {\footnotesize $^{h}$ Caterina Doglioni acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 679305 “DARKJETS”.}
\par {\footnotesize $^{i}$ Supported by the US-DOE, DE-SC0009973}
\par {\footnotesize $^{j}$ Supported by the ES-MINECO FPA2016-80994-C2-2-R}
\par {\footnotesize $^{k}$ Supported by the US-DOE, DE-SC0018098 and US-NSF ACI-1558233}
\par {\footnotesize $^{l}$ Supported by the US-DOE, DE-SC0007859 and US-NSF, 76749/1136652/2}
\par {\footnotesize $^{m}$ Supported by funding from the European Union’s Horizon 2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement No. 752730}
\par {\footnotesize $^{n}$ Supported by Swiss National Science Foundation, Early Postdoc Mobility Fellowship, project number P2ELP2\_168556}
\par {\footnotesize $^{o}$ Supported by CONACYT (Mexico)}
\par {\footnotesize $^{p}$ Supported by the US-DOE, DE-AC02-76SF0051}

### 2017-12-17 CWP/papers/roadmap/latex/cwp.tex

% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.

% JHEP preprint template
\documentclass[12pt,a4paper]{article}
\usepackage{jheppub}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[titletoc]{appendix}
\usepackage{xspace}

\usepackage[utf8]{inputenc}

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{cwp.bib}
\addbibresource{cwp-chapters.bib}

%% Comment out before final submission
%\usepackage{lineno}  % for line numbering during review
%\linenumbers

\abstract{Particle physics has an ambitious and broad experimental 
programme for the coming decades. This programme requires large 
investments in detector hardware, either to build new facilities 
and experiments, or to upgrade existing ones.
Similarly, it requires commensurate investment in the R\&D of software to acquire, manage, process, and analyse the shear amounts of data to be recorded.
In planning for the HL-LHC in particular, it is critical that all of the
collaborating stakeholders agree on the software goals and priorities,
and that the efforts complement each other. In this spirit, this white paper
describes the R\&D activities required to prepare for this software upgrade.}

\begin{document}

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-01 \\
 & & December 15, 2017 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s}

\author{HEP Software Foundation\footnote{Authors are listed at the end of this report.}}

\maketitle

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Particle physics has an ambitious experimental programme for the
coming decades. The programme supports the strategic goals of the
particle physics community that have been laid out by the European
Strategy for Particle Physics~\cite{ESPP2013} and by
the Particle Physics Project Prioritization Panel (P5)~\cite{DOE-P5} in the United
States~\cite{2014pwa}. Broadly speaking, the
scientific goals are:

\begin{itemize}
\item Exploit the discovery of the Higgs boson as a precision tool
  for investigating Standard Model (SM) and Beyond the Standard Model
  (BSM) physics.
\item
  Etudy the decays of $b$- and $c$-hadrons, and tau leptons, in the search
  for manifestations of BSM physics, and investigate
  matter-antimatter differences.
\item
  Search for signatures of dark matter.
\item
  Probe neutrino oscillations and masses.
\item
  Study the Quark Gluon Plasma state of matter in heavy-ion collisions.
\item
  Explore the unknown.
\end{itemize}

The High-Luminosity Large Hadron Collider
(HL-LHC)~\cite{HL-LHC,1742-6596-515-1-012012,Apollinari:2284929} 
will be a major
upgrade of the current LHC~\cite{LHC} supporting the aim of 
an in-depth investigation of the properties of the Higgs boson and its
couplings to other particles (Figure~\ref{fig:hl-lhc}). The ATLAS~\cite{ATLAS} and CMS~\cite{CMS}
collaborations will continue to make measurements in the Higgs sector,
while searching for new physics Beyond the Standard Model (BSM). Should
a BSM discovery be made, a full exploration of that physics will be
pursued. Such BSM physics may help shed light on the nature of dark
matter, which we know makes up the majority of gravitational matter in
the universe, but which does not interact via the electromagnetic or
strong nuclear forces~\cite{Mangano2016}.

The LHCb experiment at the LHC~\cite{LHCb} and the Belle II experiment at
KEK~\cite{BelleII} study various aspects of heavy
flavour physics (b- and c-quark, and tau-lepton physics), where quantum
influences of very high mass particles manifest themselves in lower
energy phenomena. Their primary goal is to look for BSM physics, either
by studying CP violation (that is, asymmetries in the behaviour of
particles and their corresponding antiparticles) or modifications in
rate or angular distributions in rare heavy-flavour decays. Current
manifestations of such asymmetries do not explain why our universe is so
matter dominated. These flavour physics programmes are related to BSM
searches through effective field theory, and powerful constraints on new
physics keep coming from such studies.

\begin{figure}[bthp]
    \centering
    \includegraphics[width=0.94\textwidth]{HL-LHC-plan-2017-Plan-2.pdf}
    \caption{The current schedule for the LHC and HL-LHC upgrade
and run~\cite{HL-LHC}. Currently, the start of the HL-LHC run is foreseen for mid
2026. The long shutdowns, LS2 and LS3, will be used to upgrade both the
accelerator and the detector hardware.}
    \label{fig:hl-lhc}
\end{figure}

The study of neutrinos, their mass and oscillations, can also shed light
on matter-antimatter asymmetry. The DUNE experiment will provide a huge
improvement in our ability to probe neutrino physics, detecting
neutrinos from the Long Baseline Neutrino Facility at Fermilab, as well
as linking to astro-particle physics programmes, in particular through
the potential detection of supernovas and relic neutrinos. An overview
of the experimental programme scheduled at the Fermilab facility is
given in Figure~\ref{fig:fnal-if}.

\begin{figure}
\vspace*{0.3cm}
    \centering
    \includegraphics[width=0.94\textwidth]{10yr-PLAN-9Feb-2017-V7a.pdf}
    \caption{Run schedule for the Fermilab facility until 2026.}
    \label{fig:fnal-if}
\end{figure}

In the study of the early universe immediately after the Big Bang, it is
critical to understand the phase transition between the highly
compressed quark-gluon plasma and the nuclear matter in the universe
today. The ALICE experiment at the LHC~\cite{ALICE} and
the CBM~\cite{CBM} and PANDA~\cite{PANDA} experiments at the Facility for Antiproton and Ion
Research (FAIR) are specifically designed to probe this aspect of
nuclear and particle physics. In addition ATLAS, CMS and LHCb all
contribute to the LHC heavy-ion programme.

These experimental programmes require large investments in detector
hardware, either to build new facilities and experiments (e.g., FAIR and DUNE) or to upgrade existing ones (HL-LHC, Belle II). Similarly, they
require commensurate investment in the research and development
necessary to deploy software to acquire, manage, process, and analyse
the data recorded.

For the HL-LHC, which is scheduled to begin taking data in 2026 (Figure~\ref{fig:hl-lhc}) and to run into the 2030s, some 30 times more data than
the LHC has currently produced will be collected by ATLAS and CMS. As
the total amount of LHC data already collected is close to an exabyte,
it is clear that the problems to be solved require approaches beyond
simply scaling current solutions, assuming Moore's Law and more or less
constant operational budgets. The nature of computing hardware
(processors, storage, networks) is evolving with radically new
paradigms, the quantity of data to be processed is increasing
dramatically, its complexity is increasing, and more sophisticated
analyses will be required to maximise physics yield. Developing and
deploying sustainable software for future and upgraded experiments,
given these constraints, is both a technical and a social challenge, as
detailed in this paper. An important message of this report is that a
``software upgrade'' is needed to run in parallel with the hardware
upgrades planned for the HL-LHC in order to take full advantage of these hardware upgrades and to complete the HL-LHC physics programme.

In planning for the HL-LHC in particular, it is critical that all of the
collaborating stakeholders agree on the software goals and priorities,
and that the efforts complement each other. In this spirit, the HEP
Software Foundation (HSF) began a planning exercise in late 2016 to
prepare a Community White Paper (CWP)~\cite{HSF2017} at the behest of the Worldwide LHC
Computing Grid (WLCG) project~\cite{WLCG2016}. The role of the HSF is to
facilitate coordination and common efforts in HEP software and computing
internationally and to provide a structure for the community to set
goals and priorities for future work. The objective of the CWP is to
provide a roadmap for software R\&D in preparation for the HL-LHC and
for other HEP experiments on a similar timescale, which would identify
and prioritise the software research and development investments
required:

\begin{itemize}
\item
  to achieve improvements in software efficiency, scalability and
  performance, and to make use of advances in CPU, storage and network
  technologies in order to cope with the challenges ahead;
\item
  to enable new approaches to computing and software that can radically
  extend the physics reach of the detectors;
\item
  to ensure the long-term sustainability of the software through the
  lifetime of the HL- LHC;
\item
  to ensure data and knowledge preservation beyond the lifetime of
  individual experiments;
\item
  to attract the required new expertise by offering appropriate career
  recognition to physicists specialising in software development, and by
  an effective training effort to target all contributors in the
  community.
\end{itemize}

The CWP process, organised by the HSF with the participation of the LHC
experiments and the wider HEP software and computing community, began
with a kick-off workshop at the San Diego Supercomputer Centre (SDSC),
USA, in January 2017 and concluded after a final workshop in June 2017 at
the Laboratoire d'Annecy de Physique des Particules (LAPP), France, with
a large number of intermediate topical workshops and meetings
(Appendix \ref{appendix-a---list-of-workshops}). The
entire CWP process involved an estimated 250 participants.

To reach more widely than the LHC experiments, specific contact was made
with individuals with software and computing responsibilities in the
Fermilab muon and neutrino experiments, Belle II, the Linear Collider
community, as well as various national computing organisations. The CWP
process was able to build on all the links established since the
inception of the HSF in 2014.

Working groups were established on various topics which were expected to
be important parts of the HL-LHC roadmap: \emph{Careers, Staffing and
Training; Conditions Database; Data Organisation, Management and Access;
Data Analysis and Interpretation; Data and Software Preservation;
Detector Simulation; Data-Flow Processing Frameworks; Facilities and
Distributed Computing; Machine Learning; Physics Generators; Security;
Software Development, Deployment and Validation/Verification; Software
Trigger and Event Reconstruction;} and \emph{Visualisation.} The work of
each working group is summarised in this document.

This document is the result of the CWP process. Investing in the roadmap
outlined here will be fruitful for the whole of the HEP programme and
may also benefit other projects with similar technical challenges,
particularly in astrophysics, e.g., the Square Kilometre Array (SKA)~\cite{SKA},
the Cherenkov Telescope Array (CTA)~\cite{CTA} and the Large Synoptic Survey
Telescope (LSST)~\cite{LSST}.

\hypertarget{software-and-computing-challenges}{%
\section{Software and Computing
Challenges}\label{software-and-computing-challenges}}

Run 2 for the LHC started in 2015 and delivered a proton-proton
collision energy of
\href{http://home.cern/about/engineering/restarting-lhc-why-13-tev}{13
TeV}. By the end of LHC Run 2 in 2018, it is expected that about 150
fb\textsuperscript{-1} of physics data will have been collected by both
ATLAS and CMS. Together with ALICE and LHCb, the total size of LHC data
storage pledged by sites for the year 2017 is around 1 exabyte, as shown
in Table~\ref{tab:crsg2017} from the LHC's Computing Resource Scrutiny Group
(CRSG)~\cite{Lucchesi:2284575}. The CPU allocation from the CRSG for 2017 to each
experiment is also shown.

\begin{table}
    \centering
\begin{tabular}{lp{0.15\textwidth}p{0.15\textwidth}p{0.15\textwidth}p{0.15\textwidth}}
\hline
\textbf{Experiment} & \textbf{2017 Disk Pledges (PB)} & \textbf{2017 Tape Pledges (PB)} & \textbf{Total Disk and Tape Pledges (PB)} & \textbf{2017 CPU Pledges (kHS06)}\\
\hline
ALICE & 67 & 68 & 138 & 807\\
ATLAS & 172 & 251 & 423 & 2194\\
CMS & 123 & 204 & 327 & 1729\\
LHCb & 35 & 67 & 102 & 413\\
\textbf{Total} & \textbf{400} & \textbf{591} & \textbf{990} &
\textbf{5143}\\
\hline
\end{tabular}
    \caption{Resources pledged by WLCG sites to the 4 LHC experiments for
the year 2017 as described at the September 2017 session of the
Computing Resources Scrutiny Group (CRSG).}
    \label{tab:crsg2017}
\end{table}

Using an approximate conversion from HS06~\cite{HS06} to CPU cores of 10
means that LHC computing in 2017 is supported by about 500k CPU cores.
These resources are deployed ubiquitously, from close to the experiments
themselves at CERN to a worldwide distributed computing infrastructure,
the WLCG~\cite{WLCG}. Each experiment has developed its own workflow management and
data management software to manage its share of WLCG resources.

In order to process the data, the 4 largest LHC experiments have written
more than 20 million lines of program code over the last 15 years. This
has involved contributions from thousands of physicists and many
computing professionals, encompassing a wide range of skills and
abilities. The majority of this code was written for a single
architecture (x86\_64) and with a serial processing model in mind. There
is considerable anxiety in the experiments that much of this software is
not sustainable, with the original authors no longer in the field and
much of the code itself in a poorly maintained state, ill-documented, and
lacking tests. This code, which is largely experiment-specific, manages
the entire experiment data flow, including data acquisition, high-level
triggering, calibration and alignment, simulation, reconstruction (of
both real and simulated data), visualisation, and final data analysis.

HEP experiments are typically served with a large set of integrated and
configured common software components, which have been developed either
in-house or externally. Well-known examples include ROOT~\cite{Brun1996}, which is a data analysis toolkit that
also plays a critical role in the implementation of experiments' data
storage systems, and Geant4~\cite{Agostinelli2003}, a
simulation framework through which most detector simulation is achieved.
Other packages provide tools for supporting the development process;
they include compilers and scripting languages, as well as tools for
integrating, building, testing, and generating documentation. Physics
simulation is supported by a wide range of event generators provided by
the theory community (PYTHIA~\cite{PYTHIA}, SHERPA~\cite{Gleisberg:2008ta}, ALPGEN~\cite{Mangano:565290}, MADGRAPH~\cite{MADGRAPH}, HERWIG~\cite{HERWIG},
amongst many others). There is also code developed to support the
computing infrastructure itself, such as the CVMFS distributed caching
filesystem~\cite{1742-6596-331-4-042003}, the Frontier database caching 
mechanism~\cite{Frontier}, the XRootD file access software~\cite{XRootD} and a
number of storage systems (dCache, DPM, EOS). This list of packages is % REF?
by no means exhaustive, but illustrates the range of software employed
and its critical role in almost every aspect of the programme.

Already in Run 3 LHCb will process more than 40 times the number of
collisions that it does today, and ALICE will read out Pb-Pb collisions
continuously at 50 kHz. The upgrade to the HL-LHC for Run 4 then
produces a step change for ATLAS and CMS. The beam intensity will rise
substantially, giving bunch crossings where the number of discrete
proton-proton interactions (pileup) will rise to about 200, from about 60
today. This has important consequences for the operation of the
detectors and for the performance of the reconstruction software. The
two experiments will upgrade their trigger systems to record 5-10 times
as many events as they do today. It is anticipated that HL-LHC will
deliver about 300 fb\textsuperscript{-1} of data each year.

The steep rise in resources that are then required to manage this data
can be estimated from an extrapolation of the Run 2 computing model and
is shown in Figures~\ref{fig:cms-resources} and~\ref{fig:atlas-resources}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.47\textwidth}
        \includegraphics[width=\textwidth]{cms-cpu.png}
        \caption{}
        \label{fig:cms-cpu}
    \end{subfigure}
    \quad
    \begin{subfigure}[b]{0.47\textwidth}
        \includegraphics[width=\textwidth]{cms-disk.png}
        \caption{}
        \label{fig:cms-disk}
    \end{subfigure}
    \caption{CMS estimated CPU (\ref{fig:cms-cpu}) and disk space (\ref{fig:cms-disk}) 
    resources required into the HL-LHC era, using the current computing model 
    with parameters projected out for the next 12 years.}
    \label{fig:cms-resources}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.9\textwidth}
    	    \centering
        \includegraphics[width=0.6\textwidth]{atlas-cpuHLLHC.pdf}
        \caption{Estimated CPU resources (in kHS06) needed for the years 2018 to 2028 for both data and simulation processing. The blue points are estimates based on the current software performance estimates and using the ATLAS computing model parameters from 2017. The solid line shows the amount of resources expected to be available if a flat funding scenario is assumed, which implies an increase of 20\% per year, based on the current technology trends.}
        \label{fig:atlas-cpu}
    \end{subfigure}
    \quad
    \begin{subfigure}[b]{0.9\textwidth}
    	    \centering
        \includegraphics[width=0.6\textwidth]{atlas-diskHLLHC.pdf}
        \caption{Estimated total disk resources (in PB) needed for the years 2018 to 2028 for both data and simulation processing. The blue points are estimates based on the current event sizes estimates and using the ATLAS computing model parameters from 2017. The solid line shows the amount of resources expected to be available if a flat funding scenario is assumed, which implies an increase of 15\% per year, based on the current technology trends.}
        \label{fig:atlas-disk}
    \end{subfigure}
    \caption{ATLAS resources required into the HL-LHC era, using the current computing model 
    and software performance.\cite{ATLAS-HL-LHC-Plots}}
    \label{fig:atlas-resources}
\end{figure}

In general, it can be said that the amount of data that experiments can
collect and process in the future will be limited by affordable software
and computing, and therefore the physics reach during HL-LHC will be
limited by how efficiently these resources can be used.

The ATLAS numbers, in Figure~\ref{fig:atlas-resources}, are particularly interesting as they
estimate the resources that will be available to the experiment if a
flat funding profile is maintained, taking into account the expected
technology improvements given current trends~\cite{Panzer2017}. As can be
seen, the shortfall between needs and bare technology gains is
considerable: a factor 4 in CPU and a factor 7 in disk in 2027.

While the density of transistors on silicon continues to increase
following Moore's Law (albeit more slowly than in the past), power
density constraints have limited the clock speed of processors for more
than a decade. This has effectively stalled any progress in the
processing capacity of a single CPU core. Instead, increases in
potential processing capacity come from increases in the core count of
CPUs and wide CPU registers. Alternative processing architectures have
become more commonplace. These range from the many-core architecture
based on standard x86\_64 cores to numerous alternatives such as GPUs.
For GPUs, the processing model is very different, allowing a much
greater fraction of the die to be dedicated to arithmetic calculations,
but at a price in programming difficulty and memory handling for the
developer that tends to be specific to each processor generation.
Further developments may even see the use of FPGAs for more
general-purpose tasks. Fully exploiting these evolutions requires a
shift in programming model to one based on \emph{concurrency}.

Even with the throttling of clock speed to limit power consumption,
power remains a major issue. Low power architectures are in huge demand.
At one level this might challenge the dominance of x86\_64 by simply
replacing it with, for example, AArch64 devices that may achieve lower power
costs for the scale of HEP computing needs than Intel has achieved with its
Xeon architecture. More extreme is an architecture that would see
specialised processing units dedicated to particular tasks, but with
possibly large parts of the device switched off most of the time,
so-called dark silicon.

Limitations in affordable storage also pose a major challenge, as does
the I/O rates of higher capacity hard disks. Network
bandwidth will probably continue to increase at the required level, but
the ability to use it efficiently will need a closer integration with
applications. This will require software developments to
support distributed computing (data and workload management, software
distribution and data access) and an increasing awareness of the
extremely hierarchical view of data, from long latency tape access and
medium-latency network access through to the CPU memory hierarchy.

Taking advantage of these new architectures and programming paradigms
will be critical for HEP to increase the ability of our code to deliver
physics results efficiently, and to meet the processing challenges of
the future. Some of this work will be focused on re-optimised
implementations of existing algorithms. This will be complicated by the
fact that much of our code is written for the much simpler model of
serial processing, and without the software engineering needed for
sustainability. Proper support for taking advantage of concurrent
programming techniques, such as vectorisation and thread-based
programming, through frameworks and libraries, will be essential, as the
majority of the code will still be written by physicists. Other
approaches should examine new algorithms and techniques, including
highly parallelised code that can run on GPUs or the use of machine
learning techniques to replace computationally expensive pieces of
simulation or pattern recognition. The ensemble of computing work that
is needed by the experiments must remain sufficiently flexible to take
advantage of different architectures that will provide computing to HEP
in the future. The use of high performance computing sites and
commercial cloud providers will very likely be a requirement for the
community and will bring particular constraints and demand flexibility.

These technical challenges are accompanied by significant human
challenges. Software is written by many people in the collaborations,
with varying levels of expertise, from a few experts with precious
skills to novice coders. This implies organising training in effective
coding techniques and providing excellent documentation, examples and
support. Although it is inevitable that some developments will remain
within the scope of a single experiment, tackling software problems
coherently as a community will be critical to achieving success in the
future. This will range from sharing knowledge of techniques and best
practice to establishing common libraries and projects that will provide
generic solutions to the community. Writing code that supports a wider
subset of the community than just a single experiment will almost
certainly be mandated upon HEP and presents a greater challenge, but the
potential benefits are huge. Attracting, and retaining, people with the
required skills who can provide leadership is another significant
challenge, since it impacts on the need to give adequate recognition to
physicists who specialise in software development. This is an important
issue that is treated in more detail later in the report.

Particle physics is no longer alone in facing these massive data
challenges. Experiments in other fields, from astronomy to genomics,
will produce huge amounts of data in the future, and will need to
overcome the same challenges that we face, i.e., massive data handling and
efficient scientific programming. Establishing links with these fields
has already started. Additionally, interest from the computing science
community in solving these data challenges exists, and mutually
beneficial relationships would be possible where there are genuine
research problems that are of academic interest to that community and
provide practical solutions to ours. The efficient processing of massive
data volumes is also a challenge faced by industry, in particular the
internet economy, which developed novel and major new technologies under
the banner of \emph{Big Data} that may be applicable to our use cases.

Establishing a programme of investment in software for the HEP
community, with a view to ensuring effective and sustainable software
for the coming decades, will be essential to allow us to reap the
physics benefits of the multi-exabyte data to come. It was in recognition of
this fact that the HSF itself was set up and already works to promote
these common projects and community developments~\cite{HSF2015}.

\hypertarget{programme-of-work}{%
\section{Programme of Work}\label{programme-of-work}}

In the following we describe the programme of work being proposed for
the range of topics covered by the CWP working groups. We summarise the
main specific challenges each topic will face, describe current
practices, and propose a number of R\&D tasks that should be undertaken
in order to meet the challenges. R\&D tasks are grouped in two different
timescales: short term (by 2020, in time for the HL-LHC Computing Technical Design Reports of
ATLAS and CMS) and longer-term actions (by 2022, to be ready for testing
or deployment during LHC Run 3).

\hypertarget{physics-generators}{%
\subsection{Physics Generators}\label{physics-generators}}

\subsubsection*{Scope and Challenges}

Monte-Carlo event generators are a vital part of modern particle
physics, providing a key component of the understanding and
interpretation of experiment data. Collider experiments have a need for
theoretical QCD predictions at very high precision. Already in LHC Run
2, experimental uncertainties for many analyses are at the same level as,
or lower than, those from theory. Many analyses have irreducible
QCD-induced backgrounds, where statistical extrapolation into the signal
region can only come from theory calculations. With future experiment
and machine upgrades, as well as reanalysis of current data, measured
uncertainties will shrink even further, and this will increase the need
to reduce the corresponding errors from theory.

Increasing accuracy will compel the use of higher-order perturbation
theory generators with challenging computational demands. Generating
Monte Carlo events using leading order (LO) generators is only a small
part of the overall computing requirements for HEP experiments.
Next-to-leading order (NLO) event generation, used more during LHC Run
2, is already using significant resources. Higher accuracy
theoretical cross sections calculated at next-to-next-to-leading (NNLO),
already important in some Run 2 analyses, are not widely used
because of computational cost. By HL-LHC the use of NNLO event
generation will be more widely required, so these obstacles to their
adoption must be overcome.
Increasing the order of the generators increases greatly the complexity
of the phase space integration required to calculate the appropriate QCD
matrix elements. The difficulty of this integration arises from the need
to have sufficient coverage in a high-dimensional space (10-15
dimensions, with numerous local maxima); the appearance of negative event
weights; and the fact that many terms in the integration cancel, so that
a very high degree of accuracy of each term is required. Memory demands
for generators have generally been low and initialisation times have
been fast, but an increase in order means that memory consumption
becomes important and initialisation times can become a significant 
fraction of the job's run time.

For HEP experiments, in many cases, meaningful predictions can only be
obtained by combining higher-order perturbative calculations with parton
showers. This procedure is also needed as high-multiplicity final states
become more interesting at higher luminosities and event rates. Matching
(N)NLO fixed-order calculations to parton shower algorithms can have a
very low efficiency, and increases further the computational load needed
to generate the necessary number of particle-level events. In addition,
many of the current models for the combination of parton-level event
generators and parton shower codes are incompatible with requirements
for concurrency on modern architectures. It is a major challenge to
ensure that this software can run efficiently on next generation
hardware and software systems.

Developments in generator software are mainly done by the HEP theory
community. Theorists typically derive career recognition and advancement from
making contributions to theory itself, rather than by making improvements
to the computational efficiency of generators \emph{per se}. So,
improving the computational efficiency of event generators, and allowing
them to run effectively on resources such as high performance computing
facilities (HPCs), will mean engaging with experts in computational
optimisation who can work with the theorists who develop generators.

The challenge in the next decade is to advance the theory and practical
implementation of event generators to support the needs of future
experiments, reaching a new level of theory precision and recognising
the demands for computation and computational efficiency that this will
bring.

\subsubsection*{Current Practice}

Extensive use of LO generators and parton shower algorithms are still
made by most HEP experiments. Each experiment has its own simulation
needs, but for the LHC experiments tens of billions of generated events
are now used each year for Monte Carlo simulations. During LHC Run 2 more and
more NLO generators were used, because of their increased theoretical
precision and stability. The raw computational complexity of NLO
amplitudes, combined with many-body phase-space evaluations and the
inefficiencies of the matching process, leads to a potentially much-increased CPU
budget for physics event simulation for ATLAS and CMS.

The use of NLO generators by the experiments today is also limited because of
the way the generators are implemented, producing significant numbers of
negative event weights. This means that the total number of events the
experiments need to generate, simulate, and reconstruct can be many 
times larger for NLO than for LO samples. At the same time, the
experiments budget only a similar number of Monte Carlo simulation events as from the real data. Having large NLO samples is thus not consistent with
existing computing budgets until a different scheme is developed that
does not depend on negative event weights or produces them only at a
significantly reduced rate.

While most event generation is run on ``standard'' grid resources,
effort is ongoing to run more demanding tasks on HPC resources, e.g.,
W-boson + 5-jet events at the Argonne Mira HPC). However, scaling for
efficient running on some of the existing HPC resources is not trivial
and requires effort.

Standard HEP libraries such as LHAPDF~\cite{LHAPDF}, HepMC\cite{HepMC}, 
and Rivet~\cite{Rivet} are used by the generators for integration into
the experiments' event generation workflows. These require extensions
and sustained maintenance that should be considered a shared
responsibility of the theoretical and experimental communities in the
context of large-scale experiments. In practice, however, it has been
difficult to achieve the level of support that is really needed as there
has been a lack of recognition for this work. To help improve the
capabilities and performance of generators as used by the experimental
HEP programme, and to foster interaction between the communities, the 
MCnet~\cite{MCnet} short-term studentship programme has been very useful.
Interested experimental PhD students can join a generator group for
several months to work on improving a physics aspect of the simulation
that is relevant to their work, or to improve the integration of the
generator into an experimental framework.

\subsubsection*{Research and Development Programme}

As the Monte Carlo projects are funded mainly to develop theoretical
improvements, and not mainly as ``suppliers'' to the experimental HEP
programme, any strong requests towards efficiency improvements from the
experimental community would need to be backed up by plausible avenues
of support that can fund contributions from software engineers with the
correct technical skills in software optimisation to work within the
generator author teams.

In a similar way to the MCnet studentships, a matchmaking scheme could
focus on the software engineering side, and transfer some of the
expertise available in the experiments and facilities teams to the
generator projects. Sustainable improvements are unlikely to be
delivered by graduate students ``learning on the job'' and then leaving
after a few months, so meeting the requirement of transferring technical
expertise and effort will likely require placements for experienced
optimisation specialists and a medium- to long-term connection to the
generator project.

HEP experiments, which are now managed by very large collaborations including many
technical experts, can also play a key role in sustaining a healthy
relationship between theory and experiment software. Effort to work on
common tools that benefit both the experiment itself and the wider
community would provide shared value that justifies direct investment
from the stakeholders. This model would also be
beneficial for core HEP tools like LHAPDF, HepMC and Rivet, where future
improvements have no theoretical physics interest anymore, putting them
in a similar situation to generator performance improvements. One
structural issue blocking such a mode of operation is that some
experiments do not currently recognise contributions to external
projects as experiment service work --- a situation deserving of review
in areas where external software tools are critical to experiment
success.

In the following we describe specific areas of R\&D for event generation 
up to 2022 and beyond.

\begin{itemize}
\item
  The development of new and improved theoretical algorithms provides
  the largest potential for improving event generators. While it is not
  guaranteed that simply increasing the effort dedicated to this task
  will bring about the desired result, the long-term support of event
  generator development, and the creation of career opportunities in
  this research area, are critical given the commitment to experiments
  on multi-decade scales.
\item
  Expand development in reweighting event samples, where new physics
  signatures can be explored by updating the partonic weights according
  to new matrix elements. It is necessary that the phase space for the
  updated model be a subset of the original one, which is an important
  limitation. The procedure is more complex at NLO and can require
  additional information to be stored in the event files to properly
  reweight in different cases. Overcoming the technical issues from
  utilising negative event weights is crucial. Nevertheless, the method
  can be powerful in many cases, and would hugely reduce the time needed
  for the generation of BSM samples.
\item
  At a more technical level, concurrency is an avenue that has yet to be
  explored in depth for event generation. As the calculation of matrix
  elements requires VEGAS-style integration, this work would be helped
  by the development of a new Monte-Carlo integrator. For multi-particle
  interactions, factorising the full phase space integration into lower
  dimensional integrals would be a powerful method of parallelising,
  while the interference between different Feynman graphs can be handled
  with known techniques.
\item
  For many widely used generators, basic problems of concurrency and
  thread hostility need to be tackled, to make these packages suitable
  for efficient large scale use on modern processors and within modern
  HEP software frameworks. Providing appropriate common tools for
  interfacing, benchmarking and optimising multithreaded code would
  allow expertise to be shared effectively~\cite{HSF-CWP-2017-13}.
\item
  In most generators, parallelism was added post-facto, which leads to
  scaling problems when the level of parallelism becomes very large,
  e.g., on HPC machines. These HPC machines will be part of the
  computing resource pool used by HEP, so solving scaling issues on
  these resources for event generation is important, particularly as the
  smaller generator code bases can make porting to non-x86\_64
  architectures more tractable. The problem of long and inefficient
  initialisation when a job utilises hundreds or thousands of cores on
  an HPC needs to be tackled. While the memory consumption of event
  generators is generally modest, the generation of tree-level
  contributions to high multiplicity final states can use significant
  memory, and gains would be expected from optimising here.
\item
  An underexplored avenue is the efficiency of event generation as
  used by the experiments. An increasingly common usage is to generate
  very large inclusive event samples, which are filtered on event
  final-state criteria to decide which events are to be retained and
  passed onto detector simulation and reconstruction. This naturally
  introduces a large waste of very CPU-expensive event generation, which
  could be reduced by developing filtering tools within the generators
  themselves, designed for compatibility with the experiments'
  requirements. A particularly wasteful example is where events are
  separated into orthogonal subsamples by filtering, in which case the
  same large inclusive sample is generated many times, with each stream
  filtering the events into a different group: allowing a single
  inclusive event generation to be filtered into several orthogonal
  output streams would improve efficiency.
\end{itemize}

\hypertarget{detector-simulation}{%
\subsection{Detector Simulation}\label{detector-simulation}}

\subsubsection*{Scope and Challenges}

For all its success so far, the challenges faced by the HEP field in the
simulation domain are daunting. During the first two runs, the LHC
experiments produced, reconstructed, stored, transferred, and analysed
tens of billions of simulated events. This effort required more than
half of the total computing resources allocated to the experiments. As
part of the HL-LHC physics programme, the upgraded
experiments expect to collect 150 times more data than in Run 1; demand
for larger simulation samples to satisfy analysis needs will grow
accordingly. In addition, simulation tools have to serve diverse
communities, including accelerator-based particle physics research
utilising proton-proton colliders, neutrino, dark matter, and muon
experiments, as well as the cosmic frontier. The complex detectors of
the future, with different module- or cell-level shapes, finer
segmentation, and novel materials and detection techniques, require
additional features in geometry tools and bring new demands on physics
coverage and accuracy within the constraints of the available computing
budget. The diversification of the physics programmes also requires new
and improved physics models. More extensive use of Fast Simulation is a
potential solution, under the assumption that it is possible to improve
time performance without an unacceptable loss of physics accuracy.

The gains that can be made by speeding up critical elements of the
Geant4 simulation toolkit can be leveraged for all applications that use
it, and it is therefore well worth the investment in effort needed to
achieve it. The main challenges to be addressed if the required physics
and software performance goals are to be achieved are:

\begin{itemize}
\item
  Reviewing the physics models' assumptions, approximations, and
  limitations in order to achieve higher precision, and to extend the
  validity of models up to energies of the order of 100 TeV foreseen
  with the Future Circular Collider (FCC) project~\cite{FCC}.
\item
  Redesigning, developing, and commissioning detector simulation
  toolkits to be more efficient when executed on current vector CPUs and
  emerging new architectures, including GPUs, where use of SIMD
  vectorisation is vital; this includes porting and optimising the
  experiments' simulation applications to allow exploitation of large
  HPC facilities.
\item
  Exploring different Fast Simulation options, where the full detector
  simulation is replaced, in whole or in part, by computationally
  efficient techniques. An area of investigation is common frameworks
  for fast tuning and validation.
\item
  Developing, improving and optimising geometry tools that can be shared
  am\-ong experiments to make the modeling of complex detectors
  computationally more efficient, modular, and transparent.
\item
  Developing techniques for background modeling, including contributions
  of multiple hard interactions overlapping the event of interest in
  collider experiments (pileup).
\item
  Revisiting digitisation algorithms to improve performance and
  exploring opportunities for code sharing among experiments.
\item
  Recruiting, training, retaining human resources in all areas of
  expertise pertaining to the simulation domain, including software and
  physics.
\end{itemize}

It is obviously of critical importance that the whole community of
scientists working in the simulation domain continue to work together in
as efficient a way as possible in order to deliver the required
improvements. Very specific expertise is required across all simulation
domains, such as physics modeling, tracking through complex geometries
and magnetic fields, and building realistic applications that accurately
simulate highly complex detectors. Continuous support is needed to
recruit, train, and retain people with a unique set of skills needed to
guarantee the development, maintenance, and support of simulation codes
over the long timeframes foreseen in the HEP experimental programme.

\subsubsection*{Current Practices}

The Geant4 detector simulation toolkit is at the core of simulation in
almost every HEP experiment. Its continuous development, maintenance,
and support for the experiments is of vital importance. New or refined
functionality in physics coverage and accuracy continues to be delivered
in the ongoing development programme and software
performance improvements are introduced whenever possible.

Physics models are a critical part of the detector simulation, and are
continuously being reviewed, and in some cases reimplemented, in order
to improve accuracy and software performance. Electromagnetic (EM)
transport simulation is challenging as it occupies a large part of the
computing resources used in full detector simulation. Significant
efforts have been made in the recent past to better describe the
simulation of electromagnetic shower shapes, in particular to model the
$H \to \gamma \gamma$ signal and background accurately at the LHC. This
effort is being continued with an emphasis on reviewing the models'
assumptions, approximations, and limitations, especially at very high
energy, with a view to improving their respective software
implementations. In addition, a new ``theory-based'' model
(Goudsmit-Saunderson), for describing the \emph{multiple scattering} of
electrons and positrons, has been developed that has been demonstrated
to outperform, in terms of physics accuracy and speed, the current
models in Geant4. The models used to describe the \emph{bremsstrahlung}
process have also been reviewed, and recently an improved theoretical
description of the Landau-Pomeranchuk-Migdal effect was introduced
that plays a significant role at high energies. Theoretical review of
all electromagnetic models, including those of hadrons and ions, is
therefore of high priority both for HL-LHC and for FCC studies.

Hadronic physics simulation covers purely hadronic interactions. It is
not possible for a single model to describe all the physics encountered
in a simulation due to the large energy range that needs to be covered
and the simplified approximations that are used to overcome the
difficulty of solving the full theory (QCD). Currently the most-used
reference physics list for high energy and space applications is
FTFP\_BERT. It uses the Geant4 Bertini cascade for hadron--nucleus
interactions from 0 to 12 GeV incident hadron energy and the FTF parton
string model for hadron--nucleus interactions from 3 GeV upwards.
QGSP\_BERT is a popular alternative which replaces the FTF model with
the QGS model over the high energy range. The existence of more than one
model (for each energy range) is very valuable in order to be able to
determine the systematics effects related to the approximations used.
The use of highly granular calorimeters, such as the ones being designed
by the CALICE collaboration for future linear colliders, allows a
detailed validation of the development of hadronic showers with
test-beam data. Preliminary results suggest that the lateral profiles of
Geant4 hadronic showers are too narrow. Comparisons with LHC test-beam
data have shown that a fundamental ingredient for improving the
description of the lateral development of showers is the use of
intermediate and low energy models that can describe the cascading of
hadrons in nuclear matter. Additional work is currently being invested
in the further improvement of the QGS model, which is a more
theory-based approach than the phenomenological FTF model, and therefore
offers better confidence at high energies, up to a few TeV. This again
is a large endeavour and requires continuous effort over a long time.

The Geant4 collaboration is working closely with user communities to
enrich the physics models' validation system with data acquired during
physics runs and test beam campaigns. In producing new models of physics
interactions and improving the fidelity of the models that exist, it is
absolutely imperative that high-quality data are available. Simulation
model tuning often relies on test beam data, and a program to improve
the library of available data could be invaluable to the community. Such
data would ideally include both thin-target test beams for improving
interaction models and calorimeter targets for improving shower models.
This data could potentially be used for directly tuning Fast Simulation
models as well.

There are specific challenges associated with the Intensity Frontier
experimental programme, in particular simulation of the beamline and the
neutrino flux. Neutrino experiments rely heavily on detector simulations
to reconstruct neutrino energy, which requires accurate modelling of
energy deposition by a variety of particles across a range of energies.
Muon experiments such as Muon g-2 and Mu2e also face large simulation
challenges; since they are searching for extremely rare effects, they
must grapple with very low signal to background ratios and the modeling
of low cross-section background processes. Additionally, the size of the
computational problem is a serious challenge, as large simulation runs
are required to adequately sample all relevant areas of experimental
phase space, even when techniques to minimise the required computations
are used. There is also a need to simulate the effects of low energy
neutrons, which requires large computational resources. Geant4 is the
primary simulation toolkit for all of these experiments.

Simulation toolkits do not include effects like charge drift in an
electric field or models of the readout electronics of the experiments.
Instead, these effects are normally taken into account in a separate
step called digitisation. Digitisation is inherently local to a given
sub-detector and often even to a given readout element, so that there
are many opportunities for parallelism in terms of vectorisation and
multiprocessing or multithreading, if the code and the data objects are
designed optimally. Recently, both hardware and software projects have
benefitted from an increased level of sharing among experiments. The
LArSoft Collaboration develops and supports a shared base of physics
software across Liquid Argon (LAr) Time Projection Chamber (TPC)
experiments, which includes providing common digitisation code.
Similarly, an effort exists among the LHC experiments to share code for
modeling radiation damage effects in silicon. As ATLAS and CMS expect
to use similar readout chips in their future trackers, further code
sharing might be possible.

The Geant4 simulation toolkit will also evolve over the next decade to
include contributions from various R\&D projects, as described in the
following section. This is required to ensure the support of experiments
through continuous maintenance and improvement of the Geant4 simulation
toolkit. This is necessary until production versions of potentially
alternative engines, such as those resulting from ongoing R\&D work,
become available, integrated, and validated by experiments. The agreed
ongoing strategy to make this adoption possible is to ensure that new
developments resulting from the R\&D programme can be tested with
realistic prototypes and then be integrated, validated, and deployed in
a timely fashion in Geant4.

\subsubsection*{Research and Development Programme}

To meet the challenge of improving the performance by a large factor, an
ambitious R\&D programme is underway to investigate each component of
the simulation software for the long term. In the following we describe
in detail some of the studies to be performed in the next 3-5 years.

\begin{itemize}
\item
  Particle Transport and Vectorisation: the study of an efficient
  transport of particles (tracks) in groups so as to maximise the
  benefit of using SIMD operations.
\item
  Modularisation: improvement of Geant4 design to allow for a tighter
  and easier integration of single sub-packages of the code into
  experimental frameworks.
\item
  Physics Models: extensions and refinements of the physics algorithms
  to provide new and more performant physics capabilities.
\item
  Other activities: integration of multi-threading capabilities in
  experiment applications; experiment-agnostic software products to cope
  with increased pile\-up, fast simulation, digitisation, and efficient
  production of high-quality ran\-dom numbers.
\end{itemize}

\paragraph{Particle Transport and Vectorisation} One of the most
ambitious elements of the simulation R\&D programme is a new approach to
managing particle transport, which has been introduced by the GeantV
project. The aim is to deliver a multithreaded vectorised transport
engine that has the potential to deliver large performance benefits. Its
main feature is track-level parallelisation, bundling particles with
similar properties from different events to process them in a single
thread. This approach, combined with SIMD vectorisation coding
techniques and improved data locality, is expected to yield significant
speed-ups, which are to be measured in a realistic prototype currently
under development. For the GeantV transport engine to display its best
computing performance, it is necessary to vectorise and optimise the
accompanying modules, including geometry, navigation, and the physics
models. These are developed as independent libraries so that they can
also be used together with the current Geant4 transport engine. Of
course, when used with the current Geant4 they will not expose their
full performance potential, since transport in Geant4 is currently
sequential, but this allows for a preliminary validation and comparison
with the existing implementations. The benefit of this approach is that
new developments can be delivered as soon as they are available. The new
vectorised geometry package (VecGeom), developed as part of GeantV R\&D
and successfully integrated into Geant4, is an example that demonstrated
the benefit of this approach. By the end of 2018 it is intended to have
a proof-of-concept for the new particle transport engine that includes
vectorised EM physics, vectorised magnetic field propagation and that
uses the new vectorised geometry package. This will form a sound basis
for making performance comparisons for simulating EM showers in a
realistic detector.

\begin{itemize}
\item
  2019: the \emph{beta} release of the GeantV transport engine will
  contain enough functionality to build the first real applications.
  This will allow performance to be measured and give sufficient time to
  prepare for HL-LHC running. It should include the use of vectorisation
  in most of the components, including physics modelling for electrons,
  gammas and positrons, whilst still maintaining simulation
  reproducibility, and I/O in a concurrent environment and multi-event
  user data management.
\end{itemize}

\paragraph{Modularisation} Starting from the next release, a
modularisation of Geant4 is being pursued that will allow an easier
integration in experimental frameworks, with the possibility to include
only the Geant4 modules that are actually used. A further use case is
the possibility to use one of the Geant4 components in isolation, e.g.,
to use hadronic interaction modeling without kernel components from a
fast simulation framework. As a first step a preliminary review of
libraries' granularity is being pursued, which will be followed by a
review of intra-library dependencies with the final goal of reducing
their dependencies.

\begin{itemize}
\item
  2019: Redesign of some Geant4 kernel components to improve the
  efficiency of the simulation on HPC systems, starting from improved
  handling of Geant4 \emph{database}s on large core-count systems. A
  review will be made of the multithreading design to be closer to
  task-based frameworks, such as Intel's Thread\-ed Building Blocks
  (TBB)~\cite{TBB}.
\end{itemize}

\paragraph{Physics Models} It is intended to develop
new and extended physics models to
cover extended energy and physics processing of present and future
colliders, Intensity Frontier experiments, and direct dark matter search
experiments. The goal is to extend the missing models (e.g., neutrino
interactions), improve models' physics accuracy and, at the same time,
improve CPU and memory efficiency. The deliverables of these R\&D
efforts include physics modules that produce equivalent quality physics,
and will therefore require extensive validation in realistic
applications.

\begin{itemize}
\item
  2020: Improved implementation of hadronic cascade models for LHC and,
  in particular, Liquid Argon detectors. Improved accuracy models of
  EM interactions of photons and electrons. To address the needs of
  cosmic frontier experiments, optical photon transport must be improved
  and made faster.
\item
  2022: Implementation of EPOS string model for multi-GeV to multi-TeV
  interactions, for FCC detector simulation and systematic studies of
  HL-LHC detectors.
\end{itemize}

\paragraph{Experiment Applications} The experiment applications are
essential for validating the software and physics performance of new
versions of the simulation toolkit. ATLAS and CMS have already started
to integrate Geant4 multithreading capability in their simulation
applications; in the case of CMS the first Full Simulation production in
multithreaded mode was delivered in the autumn of 2017. Specific
milestones are as follows:

\begin{itemize}
\item
  2020: LHC, Neutrino, Dark Matter, and Muon experiments to demonstrate
  the ability to run their detector simulation in multithreaded mode,
  using the improved navigation and electromagnetic physics packages.
  This should bring experiments more accurate physics and improved
  performance.
\item
  2020: Early integration of the beta release of the GeantV transport
  engine in the experiments' simulation, including the implementation of
  the new user interfaces, which will allow the first performance
  measurements and physics validation to be made.
\item
  2022: The availability of a production version of the new track-level
  parallelisation and fully vectorised geometry, navigation, and physics
  libraries will offer the experiments the option to finalise
  integration into their frameworks; intensive work will be needed in
  physics validation and computing performance tests. If successful, the
  new engine could be in production on the timescale of the start of the
  HL-LHC run in 2026.
\end{itemize}

\paragraph{Pileup} Backgrounds to hard-scatter events have many
components including in-time pileup, out-of-time pileup, cavern
background and beam-gas collisions. All of these components can be
simulated, but they present storage and I/O challenges related to the
handling of the large simulated minimum bias samples used to model the
extra interactions. An R\&D programme is needed to study different
approaches to managing these backgrounds within the next 3 years:

\begin{itemize}
\item
  Real zero-bias events can be collected, bypassing any zero
  suppression, and overlaid on the fully simulated hard scatters. This
  approach faces challenges related to the collection of
  non-zero-suppressed samples or the use of suppressed events,
  non-linear effects when adding electronic signals from different
  samples, and sub-detector misalignment consistency between the
  simulation and the real experiment. Collecting calibration and
  alignment data at the start of a new Run would necessarily incur
  delays such that this approach is mainly of use in the final analyses.
  The experiments are expected to invest in the development of the
  zero-bias overlay approach by 2020.
\item
  The baseline option is to ``pre-mix'' together the minimum bias
  collisions into individual events that have the full background
  expected for a single collision of interest. Experiments will invest
  effort on improving their pre-mixing techniques, which allow the
  mixing to be performed at the digitisation level, reducing the disk and
  network usage for a single event.
\end{itemize}

\paragraph{Fast Simulation} The work on Fast Simulation is also
accelerating with the objective of producing a flexible framework that
permits Full and Fast simulation to be combined for different particles
in the same event. Various approaches to Fast Simulation are being tried
all with the same goal of saving computing time, under the assumption
that it is possible to improve time performance without an unacceptable
loss of physics accuracy. There has recently been a great deal of
interest in the use of Machine Learning in Fast Simulation, most of
which has focused on the use of multi-objective regression and
generative adversarial networks (GANs). Since use of GANs allows for
non-parametric learning in cases such as calorimetric shower
fluctuations, it is a promising avenue for generating non-Gaussian and
highly correlated physical effects. This is an obvious area for future
expansion and development, as it is currently in its infancy.

\begin{itemize}
\item
  2018: Assessment of the benefit of machine learning approach for Fast
  Simulation.
\item
  2019: ML-based Fast Simulation for some physics observables.
\item
  2022: Demonstrate the potential of a common Fast Simulation
  infrastructure applicable to the variety of detector configurations.
\end{itemize}

\paragraph{Digitisation} It is expected that, within the next 3
years, common digitisation efforts are well-established among
experiments, and advanced high-performance gener\-ic digitisation
examples, which experiments could use as a basis to develop their own
code, become available. For example, the development of next generation
silicon detectors requires realistic simulation of the charge collection
and digitisation processes. Owing to the large variety of technologies,
common software frameworks need to be flexible and modular to cater for
the different needs.

\begin{itemize}
\item
  2020: Deliver advanced high-performance, SIMD-friendly generic
  digitisation examples that experiments can use as a basis to develop
  their own code.
\item
  2022: Fully tested and validated optimised digitisation code that can
  be used by the HL-LHC and DUNE experiments.
\end{itemize}

\paragraph{Pseudorandom Number Generation} The selection of
pseudorandom number generators (PRNGs) presents challenges when running
on infrastructures with a large degree of parallelism, as
reproducibility is a key requirement. HEP will collaborate with
researchers in the development of PRNGs, seeking to obtain generators
that address better our challenging requirements. Specific milestones
are:

\begin{itemize}
\item
  2020: Develop a single library containing sequential and vectorised
  implementations of the set of state-of-the-art PRNGs, to replace the
  existing ROOT and CLHEP implementations. Potential use of C++11 PRNG
  interfaces and implementations, and their extension for our further
  requirements (output of multiple values, vectorisation) will be
  investigated.
\item
  2022: Promote a transition to the use of this library to replace
  existing implementations in ROOT and Geant4.
\end{itemize}

\hypertarget{software-trigger-and-event-reconstruction}{%
\subsection{Software Trigger and Event
Reconstruction}\label{software-trigger-and-event-reconstruction}}

\subsubsection*{Scope and Challenges}

The reconstruction of raw detector data and simulated data, and its
processing in real time, represent a major component of today's
computing requirements in HEP. Advances in the capabilities of 
facilities and future
experiments bring the potential for a dramatic increase in physics
reach, at the price of increased event complexities and rates. It is
therefore essential that event reconstruction algorithms and software
triggers continue to evolve so that they are able to efficiently exploit
future computing architectures, and deal with the increase in data rates
without loss of physics. Projections into future, e.g., at HL-LHC conditions,
show that without significant changes in approach or algorithms
the increase in resources needed would be incompatible with the 
the expected budget.

At the HL-LHC, the central challenge for object reconstruction is to
maintain excellent efficiency and resolution in the face of high pileup
values, especially at low transverse momentum ($p_T$).
Detector upgrades, such as increases in channel density, high-precision
timing, and improved detector geometric layouts, are essential to
overcome these problems. In many cases these new technologies bring
novel requirements to software trigger and/or event reconstruction
algorithms, or require new algorithms to be developed. Ones of
particular importance at the HL-LHC include high-granularity
calorimetry, precision timing detectors, and hardware triggers based on
tracking information, which may seed later software trigger and
reconstruction algorithms.

At the same time, trigger systems for next-generation experiments are
evolving to be more capable, both in their ability to select a wider
range of events of interest for the physics programme, and their ability
to stream a larger rate of events for further processing. ATLAS and CMS
both target systems where the output of the hardware trigger system is
increased by an order of magnitude over the current capability, up to 1
MHz~\cite{CERN-LHCC-2015-020,CMSCollaboration:2015zni}. In LHCb~\cite{CERN-LHCC-2014-016}
and ALICE~\cite{Buncic:2011297}, the full collision
rate (between 30 to 40 MHz for typical LHC proton-proton operations)
will be streamed to real-time or quasi-real-time software trigger
systems. The increase in event complexity also brings a ``problem'' of
an overabundance of signals to the experiments, and specifically to the
software trigger algorithms. The evolution towards a genuine real-time
analysis of data has been driven by the need to analyse more signal than
can be written out for traditional processing, and technological
developments that enable this without reducing the analysis sensitivity
or introducing biases.

Evolutions in computing technologies are an opportunity to move
beyond commodity x86\_64 technologies, which HEP has used very
effectively over the past 20 years, but also represent a significant challenge 
if we are to
derive sufficient event processing throughput per cost to reasonably
enable our physics programmes~\cite{Bird:1695401}.
Among these challenges, important items identified include the increase
of SIMD capabilities, the evolution towards multi- or many-core
architectures, the slow increase in memory bandwidth relative to CPU
capabilities, the rise of heterogeneous hardware, and the possible
evolution in facilities available to HEP production systems.

The move towards open source software development and continuous
integration systems brings opportunities to assist developers of
software trigger and event reconstruction algorithms. Continuous
integration systems based on standard open-source tools have already
allowed automated code quality and performance checks, both for
algorithm developers and code integration teams. Scaling these up to
allow for sufficiently high-statistics checks is still an
outstanding challenge. Also, code quality demands increase as
traditional offline analysis components migrate into trigger systems,
where algorithms can only be run once, and any problem means losing data
permanently.

\subsubsection*{Current Practices}

Substantial computing facilities are in use for both online and offline
event processing across all experiments surveyed. In most experiments,
online facilities are dedicated to the operation of the software
trigger, but a recent trend has been to use them opportunistically
for offline processing too, when the software trigger does not make them
100\% busy. On the other hand, offline facilities are shared with
event reconstruction, simulation, and analysis. CPU in use by
experiments is typically measured at the scale of tens or hundreds of
thousands of x86\_64 processing cores.

The CPU needed for event reconstruction tends to be dominated by charged
particle reconstruction (tracking), especially when the number of
collisions per bunch crossing is high and an efficient
reconstruction low p\textsubscript{T} particles is required.
Calorimetric reconstruction, particle flow reconstruction, and particle
identification algorithms also make up significant parts of the CPU
budget in some experiments. Disk storage is typically 10s to 100s of PBs
per experiment. It is dominantly used to make the output of the event
reconstruction, both for real data and simulation, available for
analysis.

Current experiments have moved towards smaller, but still
flexible, tiered data formats. These tiers are typically based on the
ROOT~\cite{Brun1996} file format and constructed to
facilitate both skimming of interesting events and the selection of
interesting pieces of events by individual analysis groups or through
centralised analysis processing systems. Initial implementations of
real-time analysis systems are in use within several experiments. These
approaches remove the detector data that typically makes up the raw data
tier kept for offline reconstruction, and keep only final analysis
objects~\cite{Aaij2016,ATL-DAQ-PUB-2017-003,Khachatryan:2149625}.

Systems critical for reconstruction, calibration, and alignment
generally implement a high level of automation in all experiments.
They are an integral part of the data taking and data reconstruction processing chain, both in the online systems as well as the offline processing setup.

\subsubsection*{Research and Development Programme}

Seven key areas, itemised below, have been identified where research and
development is necessary to enable the community to exploit the full
power of the enormous datasets that we will be collecting. Three of
these areas concern the increasingly parallel and heterogeneous
computing architectures that we will have to write our code for. In
addition to a general effort to vectorise our codebases, we must
understand what kinds of algorithms are best suited to what kinds of
hardware architectures, develop benchmarks that allow us to compare the
physics-per-dollar-per-watt performance of different algorithms across a
range of potential architectures, and find ways to optimally utilise
heterogeneous processing centres. The consequent increase in the
complexity and diversity of our codebase will necessitate both a
determined push to educate physicists in modern coding
practices, and a development of more sophisticated and automated quality
assurance and control. The increasing granularity of
our detectors, and the addition of timing information, which seems
mandatory to cope with the extreme pileup conditions at the HL-LHC, will
require new kinds of reconstruction algorithms that are sufficiently
fast for use in real-time. Finally, the increased
signal rates will mandate a push towards real-time analysis in many
areas of HEP, in particular those with low-p\textbf{\textsubscript{T}}
signatures.

\begin{itemize}
\item
  HEP developed toolkits and algorithms typically make poor use of
  vector units on commodity computing systems. Improving this will bring
  speedups to applications running on both current computing systems and
  most future architectures. The goal for work in this area is to evolve
  current toolkit and algorithm implementations, and best programming
  techniques, to better use SIMD capabilities of current and future
  CPU architectures.
\item
  Computing platforms are generally evolving towards having more cores
  in order to increase processing capability. This evolution has
  resulted in multithreaded frameworks in use, or in development, across
  HEP. Algorithm developers can improve throughput by being thread-safe
  and enabling the use of fine-grained parallelism. The goal is to
  evolve current event models, toolkits and algorithm implementations,
  and best programming techniques, to improve the throughput of
  multithreaded software trigger and event reconstruction applications.
\item
  Computing architectures using technologies beyond CPUs offer an
  interesting alternative for increasing throughput of the most
  time-consuming trigger or reconstruction algorithms. Examples
  such as GPUs and FPGAs could be integrated into dedicated
  trigger or specialised reconstruction processing facilities,
  in particular
  online computing farms. The goal is to demonstrate how the throughput
  of toolkits or algorithms can be improved
  in a production environment and to understand
  how much these new architectures require rethinking the algorithms
  used today. In addition, it is necessary to assess and minimise
  possible additional costs coming from the maintenance of multiple
  implementations of the same algorithm on different architectures.
\item
  HEP experiments have extensive continuous integration systems,
  including varying code regression checks that have enhanced the
  quality assurance (QA) and quality control (QC) procedures for
  software development in recent years. These are typically maintained
  by individual experiments and have not yet reached the point where
  statistical regression, technical, and physics performance checks can
  be performed for each proposed software change. The goal is to enable
  the development, automation, and deployment of extended QA and QC
  tools and facilities for software trigger and event reconstruction
  algorithms.
\item
  Real-time analysis techniques are being adopted to enable a wider
  range of physics signals to be saved by the trigger for final
  analysis. As rates increase, these techniques can become more
  important and widespread by enabling only the parts of an event
  associated with the signal candidates to be saved, reducing the
  disk space requirement. The goal is to evaluate and demonstrate the tools
  needed to facilitate real-time analysis techniques. Research topics
  include the study of compression and custom data formats, toolkits for real-time
  detector calibration and validation that enable full offline analysis
  chains to be ported into real-time, and frameworks that allow
  non-expert offline analysts to design and deploy real-time analyses
  without compromising data taking quality.
\item
  The central challenge for object reconstruction at the HL-LHC is to
  maintain excellent efficiency and resolution in the face of high
  pileup, especially at low object p\textsubscript{T}.
  Trigger systems and reconstruction software need to exploit new techniques
  and higher granularity detectors to maintain or even improve physics
  measurements in the future. It is also becoming increasingly clear
  that reconstruction in very high pileup environments, such as the
  HL-LHC or FCC-hh, will not be possible without adding some timing
  information to our detectors, in order to exploit the finite time
  during which the beams cross and the interactions are produced. The
  goal is to develop and demonstrate efficient techniques for physics
  object reconstruction and identification in complex environments.
\item
  Future experimental facilities will bring a large increase in event
  complexity. The performance scaling of current-generation algorithms with this
  complexity must be improved to avoid a large increase in resource
  needs. In addition, it may become necessary to deploy
  new algorithms in order to solve these problems, including advanced
  machine learning techniques. The goal is to evolve or rewrite existing
  toolkits and algorithms focused on their physics and technical
  performance at high event complexity, e.g., high pileup at HL-LHC.
  Most important targets are those which limit expected throughput
  performance at future facilities, e.g., charged-particle tracking. A
  number of such efforts are already in progress.
\end{itemize}

\hypertarget{data-analysis-and-interpretation}{%
\subsection{Data Analysis and
Interpretation}\label{data-analysis-and-interpretation}}

\subsubsection*{Scope and Challenges}

Scientific questions are answered by analysing the data obtained from
suitably designed experiments and comparing measurements with
predictions from models and theories. Such comparisons are typically
performed long after data taking, but can sometimes also be executed in
quasi-real time on selected samples of reduced size.

The final stages of analysis are undertaken by small groups or even
individual researchers. The baseline analysis model utilises successive
stages of data reduction, finally reaching a compact dataset for quick
real-time iterations. This approach aims at exploiting the maximum
possible scientific potential of the data, whilst minimising the ``time
to insight'' for a large number of different analyses performed in
parallel. It is a complicated combination of diverse criteria, ranging
from the need to make efficient use of computing resources to the
management styles of the experiment collaborations. Any analysis system
has to be flexible enough to cope with deadlines imposed by conference
schedules. Future analysis models must adapt to the massive increases in
data taken by the experiments, while retaining this essential ``time to
insight'' optimisation.

Over the past 20 years the HEP community has developed and gravitated
around a single analysis ecosystem based on ROOT~\cite{Brun1996}. 
ROOT is a general-purpose object
oriented framework that addresses the selection, integration,
development, and support of a number of foundation and utility class
libraries that can be used as a basis for developing HEP application
codes. The added value to the HEP community is that it provides an
integrated and validated toolkit, and its use encompasses the full event
processing chain; it has a major impact on the way HEP analysis is
performed. This lowers the hurdle to start an analysis, enabling the
community to communicate using a common analysis language, as well as
making common improvements as additions to the toolkit quickly become
available. The ongoing ROOT programme of work addresses important new
requirements, in both functionality and performance, and this is given
a high priority by the HEP community.

An important new development in the analysis domain has been the
emergence of new analysis tools coming from industry and open source
projects, and this presents new opportunities for improving the HEP
analysis software ecosystem. The HEP community is very interested in
using these software tools, together with established components, in an
interchangeable way. The main challenge will be to enable new
open-source tools to be plugged in dynamically to the existing ecosystem
and to provide mechanisms that allow the existing and new components to
interact and exchange data efficiently. To improve our ability to
analyse much larger datasets, R\&D will be needed to investigate file
formats, compression algorithms, and new ways of storing and accessing
data for analysis and to adapt workflows to run on future computing
infrastructures.

Reproducibility is the cornerstone of scientific results. It is
currently difficult to repeat most HEP analyses in exactly the manner
they were originally performed. This difficulty mainly arises due to the
number of scientists involved, the large number of steps in a typical
HEP analysis workflow, and the complexity of the analyses themselves. A
challenge specific to data analysis and interpretation is tracking the
evolution of relationships between all the different components of an
analysis.

Robust methods for data reinterpretation are also critical.
Collaborations typically interpret results in the context of specific
models for new physics searches and sometimes reinterpret those same
searches in the context of alternative theories. However, understanding
the full implications of these searches requires the interpretation of
the experimental results in the context of many more theoretical models
than are currently explored at the time of publication. Analysis
reproducibility and reinterpretation strategies need to be considered in
all new approaches under investigation, so that they become a
fundamental component of the system as a whole.

Adapting to the rapidly evolving landscape of software tools, as well as
to methodological approaches to data analysis, requires effort in
continuous training, both for novices as well as for experienced
researchers, as detailed in the Section 
\ref{training-and-careers}. The
maintenance and sustainability of the current analysis ecosystem also
present a major challenge, as currently this effort is provided by just
a few institutions. Legacy and less-used parts of the ecosystem need to
be managed appropriately. New policies are needed to retire little used
or obsolete components and free up effort for the development of new
components. These new tools should be made attractive and useful to a
significant part of the community to attract new contributors.

\subsubsection*{Current Practices}

Methods for analysing HEP data have been developed over many years and
successfully applied to produce physics results, including more than
2000 publications, during LHC Runs 1 and 2. Analysis at the LHC
experiments typically starts with users running code over centrally
managed data that is of O(100kB/event) and contains all of the
information required to perform a typical analysis leading to
publication. The most common approach is through a campaign of
\emph{data reduction} and \emph{refinement}, ultimately producing
simplified data structures of arrays of simple data types (``flat
ntuples'') and histograms used to make plots and tables, from which
physics results can be derived.

The current centrally-managed data typically used by a Run 2 data
analysis at the LHC (hundreds of TB) is far too large to be delivered
locally to the user. An often-stated requirement of the data reduction
steps is to arrive at a dataset that ``can fit on a laptop'', in order
to facilitate low-latency, high-rate access to a manageable amount of
data during the final stages of an analysis. Creating and retaining
intermediate datasets produced by data reduction campaigns, bringing and
keeping them ``close'' to the analysers, is designed to minimise latency
and the risks related to resource contention. At the same time, disk
space requirements are usually a key constraint of the experiment
computing models. The
LHC experiments have made a continuous effort to produce optimised
analysis-oriented data formats with enough information to avoid the need
to use intermediate formats. Another effective strategy has been to
combine analyses from different users and execute them within the same
batch jobs (so-called ``analysis trains''), thereby reducing the number
of times data must be read from the storage systems. This has improved
performance and usability, and simplified the task of the bookkeeping.

There has been a huge investment in using C++ for performance-critical
code, in particular in event reconstruction and simulation, and this
will continue in the future. However, for analysis applications, Python
has emerged as the language of choice in the data science community, and
its use continues to grow within HEP. Python is highly
appreciated for its ability to support fast development cycles, for its
ease-of-use, and it offers an abundance of well-maintained and advanced
open source software packages. Experience shows that the simpler
interfaces and code constructs of Python could reduce the complexity of
analysis code, and therefore contribute to decreasing the ``time to
insight'' for HEP analyses, as well as increasing their sustainability.
Increased HEP investment is needed to allow Python to become a first
class supported language.

One new model of data analysis, developed outside of HEP, maintains the
concept of sequential reduction, but mixes interactivity with
batch processing. These exploit new cluster management systems, most
notably Apache Spark, which uses open-source tools contributed both by
industry and the data-science community. Other products implementing the
same analysis concepts and workflows are emerging, such as TensorFlow,
Dask, Pachyderm, Blaze, Parsl, and Thrill. This approach can complement
the present and widely adopted Grid processing of datasets. It may
potentially simplify the access to data and the expression of
parallelism, thereby improving the exploitation of cluster resources.

An alternative approach, which was pioneered in astronomy but has become
more widespread throughout the Big Data world, is to perform \emph{fast
querying} of centrally managed data and compute remotely on the queried
data to produce the analysis products of interest. The analysis workflow
is accomplished without focus on persistence of data traditionally
associated with data reduction, although transient data may be generated
in order to efficiently accomplish this workflow and optionally can be
retained to facilitate an analysis ``checkpoint'' for subsequent
execution. In this approach, the focus is on obtaining the analysis
end-products in a way that does not necessitate a data reduction
campaign. It is of interest to understand the role
that such an approach could have in the global analysis infrastructure,
and if it can bring an optimisation of the global storage and computing
resources required for the processing of raw data to analysis.

Another active area regarding analysis in the world outside HEP is the
switch to a functional or declarative programming model, as for example
provided by Scala in the Spark environment. This allows scientists to
express the intended data transformation as a query on data. Instead of
having to define and control the ``how'', the analyst declares the
``what'' of their analysis, essentially removing the need to define the
event loop in an analysis, and leave it to underlying services and
systems to optimally iterate over events. It appears that these
high-level approaches will allow abstraction from the underlying
implementations, allowing the computing systems more freedom in
optimising the utilisation of diverse forms of computing resources. R\&D
is already under way, e.g., TDataFrame~\cite{TDataFrame} in ROOT, and
this needs to be continued with the ultimate goal of establishing a
prototype functional or declarative programming paradigm.

\subsubsection*{Research and Development Programme}

Towards HL-LHC, we envisage dedicated data analysis facilities for
experimenters, offering an extendable environment that can provide fully
functional analysis capabilities, integrating all these technologies
relevant for HEP. Initial prototypes of such analysis facilities are
currently under development. On the time scale of HL-LHC, such dedicated
analysis facilities would provide a complete system engineered for
latency optimisation and stability.

The following R\&D programme lists the tasks that need to be
accomplished. By 2020:

\begin{itemize}
\item
  Enable new open-source software tools to be plugged in dynamically to
  the existing ecosystem, and provide mechanisms to dynamically exchange
  parts of the ecosystem with new components.
\item
  Complete an advanced prototype of a low-latency response,
  high-capacity analysis facility, incorporating fast caching
  technologies to explore a query-based analysis approach and
  open-source cluster-management tools. It should in particular include
  an evaluation of additional storage layers, such as SSD storage and
  NVRAM-like storage, and cloud and Big Data orchestration systems.
\item
  Expand support of Python in our ecosystem with a strategy for ensuring
  long-term maintenance and sustainability. In particular in ROOT, the
  current Python bindings should evolve to reach the ease of use of native Python modules.
\item
  Prototype a comprehensive set of mechanisms for interacting and
  exchanging data between new open-source tools and the existing
  analysis ecosystem.
\item
  Develop a prototype based on a functional or declarative programming
  model for data analysis.
\item
  Conceptualise and prototype an analysis ``Interpretation Gateway'',
  including data repositories, e.g., HEPData~\cite{HEPData,HEPDataRepo}, and
  analysis preservation and reinterpretation tools.
\end{itemize}

By 2022:

\begin{itemize}
\item
  Evaluate chosen architectures for analysis facilities, verify their
  design and provide input for corrective actions to test them on a
  larger scale during Run 3.
\item
  Develop a blueprint for remaining analysis facility developments,
  system design and support model.
\end{itemize}

\hypertarget{machine-learning}{%
\subsection{Machine Learning}\label{machine-learning}}

Machine Learning (ML) is a rapidly evolving approach to characterising
and describing data with the potential to radically change how data is
reduced and analysed. Some applications will qualitatively improve the
physics reach of datasets. Others will allow much more efficient use of
processing and storage resources, effectively extending the physics
reach of experiments. Many of the activities in this area will
explicitly overlap with those in the other focus areas, whereas others
will be more generic. As a first approximation, the HEP community will
build domain-specific applications on top of existing toolkits and ML
algorithms developed by computer scientists, data scientists, and
scientific software developers from outside the HEP world. Work will
also be done to understand where problems do not map well onto existing
paradigms and how these problems can be recast into abstract
formulations of more general interest.

\subsubsection*{Scope and Challenges}

The Machine Learning, Statistics, and Data Science communities have
developed a variety of powerful ML approaches for classification (using
pre-defined categories), clustering (where categories are discovered),
regression (to produce continuous outputs), density estimation,
dimensionality reduction, etc. Some of these have been used productively
in HEP for more than 20 years, others have been introduced relatively
recently. The portfolio of ML techniques and tools is in constant
evolution, and a benefit is that many have well-documented open source
software implementations. ML has already become ubiquitous in some
HEP applications, most notably in classifiers used to discriminate
between signals and backgrounds in final offline analyses. It is
also increasingly used in both online and offline reconstruction and
particle identification algorithms, as well as the classification of
reconstruction-level objects, such as jets.

The abundance of, and advancements in, ML algorithms and implementations
present both opportunities and challenges for HEP. 
The community
needs to understand which are most appropriate for our use,
tradeoffs for using one tool compared to another, and
the tradeoffs of using ML algorithms compared to using more traditional software.
These issues are not necessarily ``factorisable'', and a key goal will
be to ensure that, as HEP research teams investigate the numerous
approaches at hand, the expertise acquired and lessons learned, get
adequately disseminated to the wider community. In general, each
\emph{team}, typically a small group of scientists from a collaboration,
will serve as a source of expertise, helping others develop and
deploy experiment-specific ML-based algorithms in their software stacks.
It should provide training to those developing new ML-based algorithms,
as well as those planning to use established ML tools.

With the advent of more powerful hardware and more performant ML
algorithms, the ML toolset will be used to develop application software
that could potentially, amongst other things:

\begin{itemize}
\item
  Replace the most computationally expensive parts of pattern
  recognition algorithms and parameter extraction algorithms for
  characterising reconstructed objects. For example, investigating how
  ML algorithms could improve the physics performance or execution speed
  of charged track and vertex reconstruction, one of the most CPU
  intensive elements of our current software.
\item
  Extend the use of ML algorithms for real-time event classification and
  analysis, as discussed in more detail in Section \ref{software-trigger-and-event-reconstruction}.
\item
  Extend the physics reach of experiments by extending the role of ML at
  the analysis stage: handling data/MC or control/signal region
  differences, interpolating between mass points, training in a
  systematics-aware way, etc.
\item
  Compress data significantly with negligible loss of fidelity in terms
  of physics utility.
\end{itemize}

As already discussed, many particle physics detectors produce much more
data than can be moved to permanent storage. The process of reducing the
size of the datasets is managed by the trigger system. ML algorithms
have already been used very successfully for triggering, to rapidly
characterise which events should be selected for additional
consideration and eventually saved to long-term storage. In the era
of the HL-LHC, the challenges will increase both quantitatively and
qualitatively as the number of proton-proton collisions per bunch
crossing increases. The scope of ML applications in the trigger will
need to expand in order to tackle the challenges to come.

\subsubsection*{Current Practices}

The use of ML in HEP analyses has become commonplace over the past two
decades, and the most common use case has been in signal/background
classification. The vast majority of HEP analyses published in recent
years have used the HEP-specific software package TMVA~\cite{TMVA} included in ROOT. Recently, however, many HEP analysts have begun
migrating to non-HEP ML packages such as scikit-learn~\cite{scikit-learn}
and Keras~\cite{Keras}, although these efforts have yet to result
in physics publications from major collaborations. Data scientists at
Yandex created a Python package that provides a consistent API to most
ML packages used in HEP~\cite{REP}. Packages like Spearmint~\cite{Spearmint} and scikit-optimize~\cite{scikit-optimize} perform Bayesian optimisation and can improve HEP Monte
Carlo work.

This shift in the set of ML techniques and packages utilised is
especially strong in the neutrino physics community, where new
experiments such as DUNE place ML at the very heart of their
reconstruction algorithms and event selection. The shift is also
occurring among LHC collaborations, where ML is becoming more and more
commonplace in reconstruction and real-time applications. Examples where
ML has already been deployed in a limited way include charged and
neutral particle reconstruction and identification, jet reconstruction
and identification, and determining a particle's production properties
(flavour tagging), based on information from the rest of the event. In
addition, ML algorithms have been developed that are
insensitive to changing detector performance, for use in real-time
applications, and algorithms that are minimally biased with respect to
the physical observables of interest.

At present, much of this development has happened in specific
collaborations. While each experiment has, or is likely to have,
different specific use cases, we expect that many of these will be
sufficiently similar to each other that R\&D can be done in common. Even
when this is not possible, experience with one type of problem will
provide insights into how to approach other types of problem. This is
why the Inter-experiment Machine Learning forum (IML~\cite{IML}) was
created at CERN in 2016, as a compliment to experiment specific ML R\&D groups.
It has already fostered closer collaboration between LHC and non-LHC
collaborations in the ML field.

\subsubsection*{Research and Development Roadmap and Goals}

The R\&D roadmap presented here is based on the preliminary work done in
recent years, coordinated by the HSF IML, which will remain the main
forum to coordinate work in ML in HEP and ensure the proper links
with the data science communities. The following programme of work is
foreseen.

By 2020:
\begin{itemize}
\item
  \emph{Particle identification and particle properties:} in calorimeters or
  time projection chambers (TPCs), where the data can be represented as
  a 2D or 3D image (or even in 4D, including timing information), the
  problems can be cast as a computer vision task. Deep Learning (DL),
  one class of ML algorithm, in which neural networks are used to
  reconstruct images from pixel intensities, is a good candidate to
  identify particles and extract many parameters. Promising DL
  architectures for these tasks include convolutional, recurrent, and
  adversarial neural networks. A particularly important application is
  to Liquid Argon TPCs (LArTPCs), which is the chosen detection
  technology for DUNE, the new flagship experiment in the neutrino programme. 
  A proof of
  concept and comparison of DL architectures should be finalised by
  2020. Particle identification can also be explored to tag the flavour
  of jets in collider experiments (e.g., so-called b-tagging). The
  investigation of these concepts, which connect to Natural Language
  Processing, has started at the LHC and is to be pursued on the same
  timescale.
\item
  \emph{ML middleware and data formats for offline usage:} HEP relies 
  on the ROOT format for its data, wheras the ML community
  has developed several other formats, often associated with specific ML
  tools. A desirable data format for ML applications should have the
  following attributes: high read-write speed for efficient training,
  sparse readability without loading the entire dataset into RAM,
  compressibility, and widespread adoption by the ML community. The thorough
  evaluation of the different data formats and their impact on ML
  performance in the HEP context must be continued, and it is necessary
  to define a strategy for bridging or migrating HEP formats to the
  chosen ML format(s), or vice-versa.
\item
  \emph{Computing resource optimisations:} managing large
  volume data transfers is one
  of the challenges facing current computing facilities. Networks 
  play a crucial role in data exchange and so a
  network-aware application layer may significantly improve experiment
  operations. ML is a promising technology to identify anomalies in
  network traffic, to predict and prevent network congestion, to detect
  bugs via analysis of self-learning networks, and for WAN path
  optimisation based on user access patterns.
\item
  \emph{ML as a Service (MLaaS):} current cloud providers rely on a MLaaS model
  exploiting interactive
  machine learning tools in order to make efficient use of resources, however,
  this is not yet widely used in HEP. HEP
  services for interactive analysis, such as CERN's Service for
  Web-based Analysis, SWAN~\cite{SWAN}, may play an important role in adoption
  of machine learning tools in HEP workflows. In order to use these tools
  more efficiently, sufficient and appropriately tailored hardware and
  instances other than SWAN will be identified.
\end{itemize}

By 2022:
\begin{itemize}
\item
  \emph{Detector anomaly detection:} data taking is
  continuously monitored by physicists taking shifts to monitor and
  assess the quality of the incoming data, largely using reference
  histograms produced by experts. A whole class of ML algorithms called anomaly detection
  can be useful for automating this important task. Such unsupervised algorithms are able
  to learn from data and produce an alert when deviations are observed.
  By monitoring many variables at the same time, such algorithms are
  sensitive to subtle signs forewarning of imminent failure, so that
  pre-emptive maintenance can be scheduled. These techniques are already
  used in industry.
\item
  \emph{Simulation:} recent progress in high fidelity fast generative models,
  such as Generative Adversarial Networks (GANs) and Variational
  Autoencoders (VAEs), which are able to sample high dimensional feature
  distributions by learning from existing data samples, offer a
  promising alternative for Fast Simulation. A simplified first attempt at
  using such techniques in simulation saw orders of magnitude increase
  in speed over existing Fast Simulation techniques, but has not yet
  reached the required accuracy~\cite{CaloGAN}.
\item
  \emph{Triggering and real-time analysis:} one of the challenges is the
  trade-off in algorithm complexity and performance under strict
  inference time constraints. To deal with the increasing event
  complexity at HL-LHC, the use of sophisticated ML
  algorithms will be explored at all trigger levels, building on the pioneering work of
  the LHC collaborations. A critical part of this work will be to
  understand which ML techniques allow us to maximally exploit
  future computing architectures.
\item
  \emph{Sustainable Matrix Element Method (MEM):} MEM is a powerful
  technique that can be utilised for making measurements of physical model
  parameters and direct searches for new phenomena. As it is
  very computationally intensive its use in HEP is limited.
  Although the use of neural networks for numerical integration is not new, 
  it is a
  technical challenge to design a network sufficiently rich
  to encode the complexity of the ME calculation for a given process
  over the phase space relevant to the signal process. Deep Neural
  Networks (DNNs) are good candidates~\cite{Bendavid2017,BendavidIML2017}.
\item
  \emph{Tracking:} pattern recognition is always a computationally challenging
  step. It becomes a huge challenge in the HL-LHC environment.
  Adequate ML techniques may provide a solution that scales linearly
  with LHC intensity. Several efforts in the HEP community have started
  to investigate ML algorithms for track pattern recognition on
  many-core processors.
\end{itemize}

\hypertarget{data-organisation-management-and-access}{%
\subsection{Data Organisation, Management and
Access}\label{data-organisation-management-and-access}}

The scientific reach of data-intensive experiments is limited by how
fast data can be accessed and digested by computational resources. Changes in
computing technology and large increases in data volume require new
computational models~\cite{Butler2013}, compatible with
budget constraints. The integration of newly emerging data
analysis paradigms into our computational model has the potential to
enable new analysis methods and increase scientific output. The field,
as a whole, has a window in which to adapt our data access and data
management schemes to ones that are more suited and optimally matched to
advanced computing models and a wide range of analysis applications.

\subsubsection*{Scope and Challenges}

The LHC experiments currently provision and manage about an exabyte of
storage, approximately half of which is archival, and half is
traditional disk storage. Other experiments that will soon start data taking have
similar needs, e.g., Belle II has the same data volumes as ATLAS. The
HL-LHC storage requirements per year are expected to jump by a factor
close to 10, which is a growth rate faster than can be 
accommodated by projected technology gains. Storage will remain one of the major
cost drivers for HEP computing, at a level roughly equal to the cost
of the computational resources. The combination of storage and analysis
computing costs may restrict scientific output and the potential physics
reach of the experiments, so new techniques and algorithms are likely to
be required.

In devising experiment computing models for this era many factors have
to be taken into account. In particular, the increasing availability of
very high-speed networks may reduce the need for CPU and data
co-location. Such networks may allow for more
extensive use of data access over the wide-area network (WAN), which may
provide failover capabilities, global and federated data namespaces, and
will have an impact on data caching. Shifts in data presentation and
analysis models, such as the use of event-based data streaming along
with more traditional dataset-based or file-based data access, will be
particularly important for optimising the utilisation of opportunistic
computing cycles on HPC facilities, commercial cloud resources, and
campus clusters. This can potentially resolve currently limiting factors
such as job eviction.

The three main challenges for data management in the HL-LHC follow:

\begin{itemize}
\item
  The experiments will significantly increase both
  the data rate and the data volume. The computing systems will need to
  handle this with as small a cost increase as possible and within
  evolving storage technology limitations.
\item
  The significantly increased computational requirements for the HL-LHC
  era will also place new requirements on data access. Specifically, the use of
  new types of computing resources (cloud, HPC) that have different dynamic
  availability and characteristics will require more dynamic data
  management and access systems.
\item
  Applications employing new techniques, such as 
  training for machine learning or high rate data query systems, will likely be employed to
  meet the computational constraints and to extend physics reach.
  These new applications will place new requirements
  on how and where data is accessed and produced. Specific applications,
  such as training for machine learning, may require use of specialised
  processor resources, such as GPUs, placing further requirements on
  data.
\end{itemize}

The projected event complexity of data from future HL-LHC
runs with high pileup and from high resolution Liquid Argon detectors at
DUNE will require advanced reconstruction algorithms and analysis tools
to interpret the data. The precursors of these tools, in the form of
new pattern recognition and tracking algorithms, 
are already proving to be drivers for the
compute needs of the HEP community. The storage systems that are
developed, and the data management techniques that are employed,
will need to be matched to these changes in computational work, so
as not to hamper potential improvements.

As with computing resources, the landscape of storage solutions
is trending towards heterogeneity. The ability to
leverage new storage technologies as they become available into existing
data delivery models is a challenge that we must be prepared for. This
also implies the need to leverage
``tactical storage'', i.e., storage that becomes more cost-effective as
it becomes available (e.g., from a cloud provider), and have a data
management and provisioning system that can exploit such resources at
short notice. Volatile data sources would impact many aspects of the
system: catalogues, job brokering, monitoring and alerting, accounting,
the applications themselves.

On the hardware side, R\&D is needed in alternative approaches to data
archiving to determine the possible cost/performance tradeoffs.
Currently, tape is extensively used to hold data that cannot be
economically made available online. While the data is still accessible,
it comes with a high latency penalty, limiting effective data access. We
suggest investigating either separate direct access-based archives
(e.g., disk or optical) or new models that hierarchically overlay online
direct access volumes with archive space. This is especially relevant
when access latency is proportional to storage density. Either approach
would need to also evaluate reliability risks and the effort needed to
provide data stability. For this work, we should exchange experiences
with communities that rely on large tape archives for their primary
storage.

Cost reductions in the maintenance and operation of storage
infrastructure can be realised through convergence of the major
experiments and resource providers on shared solutions. This does not
necessarily mean promoting a monoculture, as different solutions will be
adapted to certain major classes of use cases, type of site, or funding
environment. There will always be a judgement to make on the
desirability of using a variety of specialised systems, or of
abstracting the commonalities through a more limited, but common,
interface. Reduced costs and improved sustainability will be further
promoted by extending these concepts of convergence beyond HEP and into
the other large-scale scientific endeavours that will share the
infrastructure in the coming decade (e.g., the SKA and CTA experiments).
Efforts must be made as early as possible, during the formative design
phases of such projects, to create the necessary links.

Finally, all changes undertaken must not make the ease of access to data
any worse than it is under current computing models. We must also be
prepared to accept the fact that the best possible solution may require
significant changes in the way data is handled and analysed. What is
clear is that current practices will not scale to the needs of HL-LHC
and other major HEP experiments of the coming era.

\subsubsection*{Current Practices}

The original LHC computing models were based on simpler models used
before distributed computing was a central part of HEP computing. This
allowed for a reasonably clean separation between four different aspects
of interacting with data, namely data organisation, data management,
data access, and data granularity. The meaning of these terms may be 
summarised in what follows.

\begin{itemize}
\item
  \emph{Data organisation} is essentially how data is structured as it
  is written. Most data is written in files, in ROOT format, typically
  with a column-wise organisation of the data. The records corresponding
  to these columns are compressed. The internal details of this
  organisation are visible only to individual software applications.
\item
  In the past, the key challenge for \emph{data management} was the transition to
  use distributed computing in the form of the grid. The experiments
  developed dedicated data transfer and placement systems, along with
  catalogues, to move data between computing centres. Originally,
  computing models were rather static: data was placed at sites, and the
  relevant compute jobs were sent to the right locations. Since LHC
  startup, this model has been made more flexible to limit non-optimal
  pre-placement and to take into account data popularity. In addition,
  applications might interact with catalogues or, at times, the workflow
  management system does this on behalf of the applications.
\item
  \emph{Data access}: historically, various protocols have been used for
  direct reads (rfio, dcap, xrootd, etc.) where jobs are reading data
  explicitly staged-in or cached by the compute resource used or the
  site it belongs to. A recent move has been the convergence towards
  xrootd as the main protocol for direct access. With direct access,
  applications may use alternative protocols to those used by data
  transfers between sites. In addition, LHC experiments have been
  increasingly using remote access to the data, without any stage-in
  operations, using the possibilities offered by protocols such as xrootd
  or http.
\item
  \emph{Data granularity}: the data is split into datasets, as defined
  by physics selections and use cases, consisting of a set of individual
  files. While individual files in datasets can be processed in
  parallel, the files themselves are usually processed as a whole.
\end{itemize}

Before LHC turn-on, and in the first years of the LHC, these four
areas were to first order optimised independently. As LHC computing
matured, interest has turned to optimisations spanning multiple areas.
For example, the recent use of ``Data Federations'' mixes up Data
Management and Access. As we will see below, some of the foreseen
opportunities towards HL-LHC may require global optimisations.

Thus, in this section we take a broader view than traditional data
management and consider the combination of ``Data Organisation,
Management and Access'' (DOMA) together. We believe that this fuller
picture will provide important opportunities for improving
efficiency and scaleability, as we enter the many-exabyte era.

\subsubsection*{Research and Development Programme}

In the following, we describe
tasks that will need to be carried out in order to demonstrate that the
increased volume and complexity of data expected over the coming decade
can be stored, accessed, and analysed at an affordable cost.

\begin{itemize}
\item
  Sub-file granularity, e.g.,
  event-based, will be studied to see whether it can be implemented
  efficiently, and in a scalable, cost-effective manner, for all
  applications making use of event selection, to see whether it offers
  an advantage over current file-based granularity. The following tasks
  should be completed by 2020:
  \begin{itemize}
  \item
    Quantify the impact on
    performance and resource utilisation  of the storage and network
    for the main
    access patterns, i.e., simulation, reconstruction, analysis.
  \item
    Assess the impact on
    catalogues and data distribution.
  \item
    Assess whether
    event-granularity makes sense in object stores that tend to require
    large chunks of data for efficiency.
  \item
    Test for improvement in
    recoverability from preemption, in particular when using cloud spot
    resources and/or dynamic HPC resources.
  \end{itemize}
\item
  We will seek to derive benefits
  from data organisation and analysis technologies adopted by other big
  data users. A proof-of-concept that involves the following tasks needs
  to be established by 2020 to allow full implementations to be made in
  the years that follow.
  \begin{itemize}
  \item
    Study the impact of
    column-wise, versus row-wise, organisation of data on the
    performance of each kind of access.
  \item
    Investigate efficient data
    storage and access solutions that support the use of map-reduce or
    Spark-like analysis services.
  \item
    Evaluate just-in-time
    decompression schemes and mappings onto hardware architectures
    considering the flow of data, from spinning disk to memory and
    application.
  \end{itemize}
\item
  Investigate the role data
  placement optimisations can play, such as caching, in order to use
  computing resources effectively, and the technologies that can be used
  for this. The following tasks should be completed by 2020:
  \begin{itemize}
  \item
   Quantify the benefit of
    placement optimisation for reconstruction,
    analysis, and simulation.
  \item
    Assess the benefit of caching
    for Machine Learning-based applications, in particular for the
    learning phase, and follow-up the evolution of technology outside
    HEP.
  \end{itemize}
  In the longer term the benefits
  that can be derived from using different approaches to the way HEP is
  currently managing its data delivery systems should be studied. Two
  different content delivery methods will be looked at, namely Content
  Delivery Networks (CDN) and Named Data Networking (NDN).
\item
  Study how to minimise HEP
  infrastructure costs by exploiting varied quality of service from
  different storage technologies. In particular, study the role that
  opportunistic/tactical storage can play, as well as different archival
  storage solutions. A proof-of-concept should be made by 2020, with a
  full implementation to follow in the following years.
\end{itemize}

\begin{itemize}
\item
  Establish how to globally optimise data access latency, with respect
  to the efficiency of using CPU, at a sustainable cost. This involves
  studying the impact of concentrating data in fewer, larger locations
  (the ``data-lake'' approach), and making increased use of opportunistic
  compute resources located further from the data. Again, a
  proof-of-concept should be made by 2020, with a full implementation in
  the following years, if successful. This R\&D will be done in common
  with the related actions planned as part of Facilities and Distributed
  Computing.
\end{itemize}

\hypertarget{facilities-and-distributed-computing}{%
\subsection{Facilities and Distributed
Computing}\label{facilities-and-distributed-computing}}

\subsubsection*{Scope and Challenges}

As outlined in Section \ref{software-and-computing-challenges}, 
huge resource requirements are anticipated for
HL-LHC running. These need to be deployed and managed across the WLCG
infrastructure, which has evolved from the original ideas on deployment
before LHC data-taking started~\cite{MONARC}, to be a mature and
effective infrastructure that is now exploited by LHC experiments.
Currently, hardware costs are dominated by disk storage, closely
followed by CPU, followed by tape and networking. Naive estimates of
scaling to meet HL-LHC needs indicate that the current system would need
almost an order of magnitude more resources than will be available from
technology evolution alone. In addition, other initiatives such as Belle
II and DUNE in particle physics, but also other science projects such as
SKA, will require a comparable amount of resources on the same
infrastructure. Even anticipating substantial software improvements, the
major challenge in this area is to find the best configuration for
facilities and computing sites that make HL-LHC computing feasible. This
challenge is further complicated by substantial regional differences in
funding models, meaning that any solution must be sensitive to these
local considerations to be effective.

There are a number of changes that can be anticipated on the timescale
of the next decade that must be taken into account. There is an
increasing need to use highly heterogeneous resources, including the
use of HPC infrastructures (which can often have very particular setups
and policies that make their exploitation challenging); volunteer
computing (which is restricted in scope and unreliable, but can be a
significant resource); and cloud computing,
both commercial and research. All of these offer different resource
provisioning interfaces and can be significantly more dynamic than
directly funded HEP computing sites. In addition, diversity of computing
architectures is expected to become the norm, with different CPU
architectures, as well as more specialised GPUs and FPGAs.

This increasingly dynamic environment for resources, particularly CPU,
must be coupled with a highly reliable system for data storage and a
suitable network infrastructure for delivering this data to where it
will be processed. While CPU and disk capacity is expected to increase
by respectively 15\% and 25\% per year for the same cost~\cite{CERN-COST-EST}, the trends of
research network capacity increases show a much steeper growth, such as
two orders of magnitude from now to HL-LHC times. Therefore, the
evolution of the computing models would need to be more network centric.

In the network domain, there are new technology developments, such as
Software Defined Networks (SDNs), which enable user-defined high
capacity network paths to be controlled via experiment software, and
which could help manage these data flows. These new technologies require
considerable R\&D to prove their utility and practicality. In addition,
the networks used by HEP are likely to see large increases in traffic
from other science domains.

Underlying storage system technology will continue to evolve, for
example towards object stores, and, as proposed in  
Data Organisation, Management and Access  
(Section \ref{data-organisation-management-and-access}),
R\&D is also necessary to understand their usability and their role in
the HEP infrastructures. There is also the continual challenge of
assembling inhomogeneous systems and sites into an effective widely
distributed worldwide data management infrastructure that is usable by
experiments. This is particularly compounded by the scale increases for
HL-LHC where multiple replicas of data (for redundancy and availability)
will become extremely expensive.

Evolutionary change towards HL-LHC is required, as the experiments will
continue to use the current system. Mapping out a path for migration
then requires a fuller understanding of the costs and benefits of the
proposed changes. A model is needed in which the benefits of such
changes can be evaluated, taking into account hardware and human costs,
as well as the impact on software and workload performance that in turn
leads to physics impact. Even if HL-LHC is the use case used to build
this cost and performance model, because the ten years of experience
running large-scale experiments helped to define the needs, it is
believed that this work, and the resulting model, will be valuable for
other upcoming data intensive scientific initiatives. This includes
future HEP projects, such as Belle II, DUNE and possibly ILC
experiments, but also non-HEP projects, such as SKA.

\subsubsection*{Current Practices}

While there are many particular exceptions, most resources incorporated
into the current WLCG are done so in independently managed sites,
usually with some regional organisation structure, and mostly offering
both CPU and storage. The sites are usually funded directly to provide
computing to WLCG, and are in some sense then ``owned'' by HEP, albeit
often shared with others. Frequently substantial cost contributions are
made indirectly, for example through funding of energy costs or
additional staff effort, particularly at smaller centres. Tape is found
only at CERN and at large national facilities, such as the WLCG Tier-1s~\cite{Bird:1695401}.

Interfaces to these computing resources are defined by technical
operations in WLCG. Frequently there are choices that sites can make
among some limited set of approved options for interfaces. These can
overlap in functionality. Some are very HEP specific and recognised as
over-complex: work is in progress to get rid of them. The acceptable
architectures and operating systems are also defined at the WLCG level
(currently x86\_64, running Scientific Linux 6 and compatible), and
sites can deploy these either directly onto ``bare metal'' or can use an
abstraction layer, such as virtual machines or containers.

There are different logical networks being used to connect sites: LHCOPN
connects CERN with the Tier-1 centres and a mixture of LHCONE and
generic academic networks connect other sites.

Almost every experiment layers its own customised workload and data
management system on top of the base WLCG provision, with several
concepts, and a few lower-level components, in common. The pilot job
model for workloads is ubiquitous, where a real workload is dispatched
only once a job slot is secured. Data management layers aggregate files
in the storage systems into datasets and manage experiment-specific
metadata. In contrast to the MONARC model, sites are generally used more
flexibly and homogeneously by experiments, both in workloads and in data
stored.

In total, WLCG currently provides experiments with resources distributed
at about 170 sites, in 42 countries, which pledge every year the amount
of CPU and disk resources they are committed to delivering. The pledge
process is overseen by the Computing Resource Scrutiny Group (CRSG), mandated by
the funding agencies to validate the experiment requests, and to
identify mismatches with site pledges. These sites are connected by
10-100 Gb links, and deliver approximately 500k CPU cores and 1 EB of
storage, of which 400 PB is disk. More than 200M jobs are executed each day~\cite{Bird2017}.

\subsubsection*{Research and Development programme}

The following areas of study are ongoing, and will involve technology
evaluations, prototyping, and scale tests. Several of the items below
require some coordination with other topical areas discussed in this
document, and some work is still needed to finalise the detailed action
plan. These actions will need to be structured to meet the common
milestones of informing the HL-LHC Computing Technical Design Reports (TDRs), and deploying
advanced prototypes during LHC Run 3.

\begin{itemize}
\item
  Understand better the relationship between the performance and costs
  of the WLCG system, and how it delivers the necessary functionality to
  support LHC physics. This will be an ongoing process, started by the
  recently formed System Performance and Cost Modeling Working Group,
  and aims to provide a quantitative assessment for any proposed
  changes.
\item
  Define the functionality needed to implement a federated data centre
  concept (``data lake'') that aims to reduce the operational cost of
  storage for HL-LHC, and at the same time better manage network
  capacity, whilst maintaining the overall CPU efficiency. This would
  include the necessary qualities of service, and options for regionally
  distributed implementations, including the ability to flexibly respond
  to model changes in the balance between disk and tape. This work
  should be done in conjunction with the existing Data Organisation,
  Management and Access Working Group to evaluate the impact of the
  different access patterns and data organisations envisaged.
\item
  Establish an agreement on the common data management functionality
  that is required by experiments, targeting a consolidation and a lower
  maintenance burden. The intimate relationship between the management
  of elements in storage systems and metadata must be recognised. This
  work requires coordination with the Data Processing Frameworks
  Working Group. It
  needs to address at least the following use cases:

  \begin{itemize}
  \item
    processing sites that may have some small disk cache, but do not
    manage primary data;
  \item
    fine grained processing strategies that may enable processing of
    small chunks of data, with appropriate bookkeeping support;
  \item
    integration of heterogeneous processing resources, such as HPCs and
    clou\-ds.
  \end{itemize}
\item
  Explore scalable and uniform means of workload scheduling, which
  incorporate dynamic heterogenous resources, and the capabilities of
  finer grained processing that increases overall efficiency. The
  optimal scheduling of special workloads that require particular
  resources is clearly required.
\item
  Contribute to the prototyping and evaluation of a quasi-interactive
  analysis facility that would offer a different model for physics
  analysis, but would also need to be integrated into the data and
  workload management of the experiments. This is work to be done in
  collaboration with the Data Analysis and Interpretation Working Group.
\end{itemize}

\hypertarget{data-flow-processing-framework}{%
\subsection{Data-Flow Processing
Framework}\label{data-flow-processing-framework}}

\subsubsection*{Scope and Challenges}

Frameworks in HEP are used for the collaboration-wide
data processing tasks of triggering, reconstruction, and simulation, as
well as other tasks that subgroups of the collaboration are responsible
for, such as detector alignment and calibration. Providing framework
services and libraries that will satisfy the computing and data needs
for future HEP experiments in the next decade, while maintaining our
efficient exploitation of increasingly heterogeneous resources, is a
huge challenge.

To fully exploit the potential of modern processors, HEP data processing
frameworks need to allow for the parallel execution of reconstruction or
simulation algorithms on multiple events simultaneously. Frameworks face
the challenge of handling the massive parallelism and heterogeneity that
will be present in future computing facilities, including multi-core and
many-core systems, GPUs, Tensor Processing Units (TPUs), and tiered
memory systems, each integrated with storage and high-speed network
interconnections. Efficient running on heterogeneous resources will
require a tighter integration with the computing models' higher-level
systems of workflow and data management. Experiment frameworks must also
successfully integrate and marshall other HEP software that may have its
own parallelisation model, such as physics generators and detector
simulation.

Common developments across experiments are desirable in this area, but
are hampered by many decades of legacy work. Evolving our frameworks
also has to be done recognising the needs of the different stakeholders
in the system. This includes physicists, who are writing processing
algorithms for triggering, reconstruction or analysis; production
managers, who need to define processing workflows over massive datasets;
and facility managers, who require their infrastructures to be used
effectively. These frameworks are also constrained by security
requirements, mandated by the groups and agencies in charge of it.

\subsubsection*{Current Practices}

Although most frameworks used in HEP share common concepts, there are,
for mainly historical reasons, a number of different implementations;
some of these are shared between experiments. The Gaudi framework~\cite{Barrand:2001ny} was
originally developed by LHCb, but is also used by ATLAS and various
non-LHC experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff}, which was forked
to provide the art framework for the Fermilab Intensity Frontier experiments~\cite{Green:2012gv}. 
Belle II uses basf2~\cite{1742-6596-331-3-032024}. The linear collider community developed and uses
Marlin~\cite{Gaede:2006pj}. The FAIR experiments use FairROOT, closely related to ALICE's
AliROOT. The FAIR experiments and ALICE are now developing a new
framework, which is called O2~\cite{O2}. At the time of writing, most 
major frameworks support basic parallelisation, both within and across
events, based on a task-based model~\cite{Jones:2015soc}\cite{Clemencic:2015paa}.

Each framework has a processing model, which provides the means to
execute and apportion work. Mechanisms for this are threads, tasks,
processes, and inter-process communication. The different strategies used
reflect different trade-offs between constraints in the programming
model, efficiency of execution, and ease of adapting to inhomogeneous
resources. These concerns also reflect two different behaviours:
firstly, maximising throughput, where it is most important to maximise
the number of events that are processed by a given resource; secondly,
minimising latency, where the primary constraint is on how long it takes
to calculate an answer for a particular datum.

Current practice for throughput maximising system architectures have
con\-strain\-ed the scope of framework designs. Framework applications have
largely been viewed by the system as a batch job with complex
configuration, consuming resources according to rules dictated by the
computing model: one process using one core on one node, operating
independently with a fixed size memory space on a fixed set of files
(streamed or read directly). Only recently has CMS broken this tradition
starting at the beginning of Run 2, by utilising all available cores in one process space using threading. ATLAS is currently
using a multi-process fork-and-copy-on-write solution to remove the
constraint of one core/process. Both experiments were driven to solve
this problem by the ever-growing need for more memory per process
brought on by the increasing complexity of LHC events. Current practice
manages systemwide (or facility-wide) scaling by dividing up datasets,
generating a framework application configuration, and scheduling jobs on
nodes/cores to consume all available resources. Given anticipated
changes in hardware (heterogeneity, connectivity, memory, storage)
available at computing facilities, the interplay between
workflow and workload management systems and framework applications need to
be carefully examined. It may be advantageous to permit framework
applications (or systems) to span multi-node resources, allowing them
to be first-class participants in the business of scaling within a
facility. In our community some aspects of this approach, which maps
features with microservices or function as a service, is being pioneered
by the O2 framework.

\subsubsection*{Research and Development programme}

By the end of 2018: review the existing technologies that are the
important building blocks for data processing frameworks and reach
agreement on the main architectural concepts for the next generation of
frameworks. Community meetings and workshops, along the lines of the
original Concurrency Forum, are envisaged in order to foster collaboration in this work~\cite{ConcurrencyForum}. This includes the following:

\begin{itemize}
\item
  Libraries used for concurrency, their likely evolution and the issues
  in integrating the models used by detector simulation and physics
  generators into the frameworks.
\item
  Functional programming, as well as domain specific languages, as a way
  to describe the physics data processing that has to be undertaken
  rather than how it has to be implemented. This approach is based on
  the same concepts as the idea for functional approaches for
  (statistical) analysis as described in Section \ref{data-analysis-and-interpretation}.
\item
  Analysis of the functional differences between the existing frameworks
  and the different experiment use cases.
\end{itemize}

By 2020: prototype and demonstrator projects for the agreed
architectural concepts and baseline to inform the HL-LHC Computing TDRs
and to demonstrate advances over what is currently deployed. The
following specific items will have to be taken into account:

\begin{itemize}
\item
  These prototypes should be as common as possible between existing
  frameworks, or at least several of them, as a proof-of-concept of
  effort and component sharing between frameworks for their future
  evolution. Possible migration paths to more common implementations
  will be part of this activity.
\item
  In addition to covering the items mentioned for the review phase, they
  should particularly demonstrate possible approaches for scheduling the
  work across heterogeneous resources and using them efficiently, with a
  particular focus on the efficient use of co-processors, such as GPUs.
\item
  They need to identify data model changes that are required for an
  efficient use of new processor architectures (e.g., vectorisation),
  and for scaling I/O performance in the context of concurrency.
\item
  Prototypes of a more advanced integration with workload management,
  taking advantage in particular of the advanced features available at
  facilities for a finer control of the interactions with storage and
  network, and dealing efficiently with the specificities of HPC
  resources.
\end{itemize}

By 2022: production-quality framework libraries usable by several
experiment frameworks, covering the main areas successfully demonstrated
in the previous phase. During these activities we expect at least one
major paradigm shift to take place on this 5-year time scale. It will be
important to continue discussing their impact within the community,
which will be ensured through appropriate cross-experiment workshops
dedicated to data processing frameworks.

\hypertarget{conditions-data}{%
\subsection{Conditions Data}\label{conditions-data}}

\subsubsection*{Scope and Challenges}

Conditions data is defined as the non-event data required by
data-processing software to correctly simulate, digitise or reconstruct
the raw detector event data. The non-event data discussed here consists
mainly of detector calibration and alignment information, with some
additional data describing the detector configuration, the machine
parameters, as well as information from the detector control system.

Conditions data is different from event data in many respects, but one
of the important differences is that its volume scales with time rather
than with the luminosity. As a consequence, its growth is limited, as
compared to event data: conditions data volume is expected to be at the
terabyte scale and the update rate is modest (typically O(1)Hz).
However, conditions data is used by event processing applications
running on a very large distributed computing infrastructure, resulting
in tens of thousands of jobs that may try to access the conditions data
at the same time, and leading to a very significant rate of reading
(typically O(10) kHz).

To successfully serve such rates, some form of caching is needed, either
by using services such as web proxies (CMS and ATLAS use Frontier) or by
delivering the conditions data as files distributed to the jobs. For the
latter approach, CVMFS is an attractive solution due to its embedded
caching, and its advanced snapshotting and branching features. ALICE
have made some promising tests, and started to use this approach in Run
2; Belle II already took the same approach~\cite{WoodACAT2017}, and NA62
has also decided to adopt this solution. However, one particular
challenge to be overcome with the filesystem approach is to design an
efficient mapping of conditions data and metadata to files in order to
use the CVMFS caching layers efficiently.

Efficient caching is especially important in order to support the
high-reading rates that will be necessary for ATLAS and CMS experiments
starting with Run 4. For these experiments, a subset of the conditions
data is linked to the luminosity, leading to an interval of granularity
down to the order of a minute. Insufficient or inefficient caching may
impact the efficiency of the reconstruction processing.

Another important challenge is ensuring the long-term maintainability of
the conditions data storage infrastructure. Shortcomings in the initial
approach used in LHC Run 1 and Run 2, leading to complex
implementations, helped to identify the key requirements for an
efficient and sustainable condition data handling infrastructure. There
is now a consensus among experiments on these requirements~\cite{Laycock2017}: 
ATLAS and CMS are working on a common next-generation
conditions database~\cite{1742-6596-898-4-042047}.
The Belle II experiment, which is about to start its data taking, has already
developed a solution based on the same concepts and architecture. One
key point in this new design is to have a server mostly agnostic to the
data content with most of the intelligence on the client side. This new
approach should make it easier to rely on well-established open-source
products (e.g., Boost) or software components developed for the
processing of event data (e.g., CVMFS). With such an approach, it should
be possible to leverage technologies such as REST interfaces to simplify
insertion and read operations, and make them very efficient to reach the
rate levels foreseen. Also, to provide a resilient service to jobs that
depend on it, the client will be able to use multiple proxies or servers
to access the data.

One conditions data challenge may be linked to the use of an event
service, as ATLAS is doing currently, to use efficiently HPC facilities
for event simulation or processing. The event service allows better use
of resources that may be volatile by allocating and bookkeeping the work
done, not at the job granularity, but at the event granularity. This
reduces the possibility for optimising access to the conditions data at
the job level, and may lead to an increased pressure on the conditions
data infrastructure. This approach is still at an early stage, and more
experience is needed to better appreciate the exact impact on the
conditions data.

\subsubsection*{Current Practices}

The data model for conditions data management is an area where the
experiments have converged on something like a best common practice. The
time information for the validity of the Payloads is specified with a
parameter called an \emph{Interval Of Validity} (IOV), which can be
represented by a Run number, the ID of a luminosity section or a
universal timestamp. A fully qualified set of conditions data consists
of a set of payloads and their associate IOVs covering the time span
required by the workload. A label called a \emph{Tag} identifies the
version of the set and the global tag is the top-level configuration of
all conditions data. For a given detector subsystem and a given IOV, a
global tag will resolve to one, and only one, conditions data payload.
The global tag resolves to a particular system tag via the global tag
map table. A system tag consists of many intervals of validity or
entries in the IOV table. Finally, each entry in the IOV table maps to a
payload via its unique hash key.

A relational database is a good choice for implementing this design. One
advantage of this approach is that a payload has a unique identifier,
its hash key, and this identifier is the only way to access it. All
other information, such as tags and IOV, is metadata used to select a
particular payload. This allows a clear separation of the payload data
from the metadata, and may allow use of a different backend technology
to store the data and the metadata. This has potentially several
advantages:

\begin{itemize}
\item
  Payload objects can be cached independently of their metadata, using
  the appropriate technology, without the constraints linked to metadata
  queries.
\item
  Conditions data metadata are typically small compared to the
  conditions data themselves, which makes it easy to export them as a
  single file using technologies such as SQLite. This may help for
  long-term data preservation.
\item
  IOVs, being independent of the payload, can also be cached on their
  own.
\end{itemize}

A recent trend is the move to full reconstruction online, where the
calibrations and alignment are computed and applied in the High Level
Trigger (HLT). This is currently being tested by ALICE and LHCb, who
will adopt it for use in Run 3. This will offer an opportunity to
separate the distribution of conditions data to reconstruction jobs and
analysis jobs, as they will not run on the same infrastructure. However,
running reconstruction in the context of the HLT will put an increased
pressure on the access efficiency to the conditions data, due to the HLT
time budget constraints.

\subsubsection*{Research and Development Programme}

R\&D actions related to Conditions databases are already in progress,
and all the activities described below should be completed by 2020. This
will provide valuable input for the future HL-LHC TDRs, and allow these
services to be deployed during Run 3 to overcome the limitations seen in
today's solutions.

\begin{itemize}
\item
  File-system view of conditions data for analysis jobs: study how to
  leverage advanced snapshotting/branching features of CVMFS for
  efficiently distributing conditions data as well as ways to optimise
  data/metadata layout in order to benefit from CVMFS caching. Prototype
  production of the file-system view from the conditions database.
\item
  Identify and evaluate industry technologies that could replace
  HEP-specific components.
\item
  ATLAS: migrate current implementations based on COOL to the proposed
  REST-based approach; study how to avoid moving too much complexity on
  the client side, in particular for easier adoption by subsystems, e.g.,
  possibility of common modules/libraries. ALICE is also planning to
  explore this approach for the future, as an alternative or to
  complement the current CVMFS-based implementation.
\end{itemize}

\hypertarget{visualisation}{%
\subsection{Visualisation}\label{visualisation}}

\subsubsection*{Scope and Challenges}

In modern High Energy Physics (HEP) experiments, visualisation of data
has a key role in many activities and tasks across the whole data
processing chain: detector development, monitoring, event generation,
reconstruction, detector simulation, data analysis, as well as outreach
and education.

\emph{Event displays} are the main tool to explore experimental data at
the event level and to visualise the detector itself. There are two main
types of application: firstly, those integrated in the experiments'
frameworks, which are able to access and visualise all the experiments'
data, but at a cost in terms of complexity; secondly,
those designed as cross-platform applications, lightweight and fast,
delivering only a simplified version or a subset of the event data. In
the first case, access to data is tied intimately to an experiment's
data model (for both event and geometry data) and this inhibits
portability; in the second, processing the experiment data into a
generic format usually loses some detail and is an extra processing
step. In addition, there are various graphical backends that can be used
to visualise the final product, either standalone or within a browser,
and these can have a substantial impact on the types of devices
supported.

Beyond event displays, HEP also uses \emph{visualisation of statistical
information}, typically histograms, which allow the analyst to quickly
characterise the data. Unlike event displays, these visualisations are
not strongly linked to the detector geometry, and often aggregate data
from multiple events. Other types of visualisation are used to
\emph{display non-spatial data}, such as graphs for describing the
logical structure of the detector or for illustrating dependencies
between the data products of different reconstruction algorithms.

The main challenges in this domain are in the sustainability of the many
experi\-ment specific visualisation tools when common projects could
reduce duplication and increase quality and long-term maintenance. The
ingestion of events and other data could be eased by common formats,
which would need to be defined and satisfy all users. Changes to support
a client-server architecture would help broaden the ability to support
new devices, such as mobile phones. Making a good choice for the
libraries used to render 3D shapes is also key, impacting on the range
of output devices that can be supported and the level of interaction
with the user. Reacting to a fast-changing technology landscape is very
important -- HEP's effort is limited and generic solutions can often be
used with modest effort. This applies strongly to non-event
visualisation, where many open source and industry standard tools can be
exploited.

\subsubsection*{Current Practices}

Three key features characterise almost all HEP event displays:

\begin{itemize}
\item
  \textbf{Event-based workflow}: applications access experimental data
  on an event-by-event basis, visualising the data collections belonging
  to a particular event. Data can be related to the actual physics
  events (e.g., physics objects such as jets or tracks) or to the
  experimental conditions (e.g., detector descriptions, calibrations).
\item
  \textbf{Geometry visualisation}: The application can display the
  geometry of the detector, as retrieved from the experiments' software
  frameworks, or a simplified description, usually for the sake of speed
  or portability.
\item
  \textbf{Interactivity}: applications offer different interfaces and
  tools to users, in order to interact with the visualisation itself,
  select event data, and set cuts on objects' properties.
\end{itemize}

Experiments have often developed multiple event displays that either
take the full integration approach explained above or are standalone and
rely on extracted and simplified data.

The visualisation of data can be achieved through the low level OpenGL
API, by the use of higher-level OpenGL-based libraries, or within a web
browser using WebGL. Using OpenGL directly is robust and avoids other dependencies, but
implies a significant effort. Instead of using the API directly, a
library layer on top of OpenGL (e.g., Coin3D) can more closely match the
underlying data, such as geometry, and offers a higher level API that
simplifies development. However, this carries the risk that if the
library itself becomes deprecated, as has happened with Coin3D, the
experiment needs to migrate to a different solution or to take on the
maintenance burden itself. Standalone applications often use WebGL
technology to render 3D objects inside a web browser. This is a very
convenient way of rendering 3D graphics, due to the cross-platform
nature of web technologies, and offers many portability advantages (e.g.,
easier support for mobile or virtual reality devices), but at some cost
of not supporting the most complex visualisations requiring heavy
interaction with the experiments' data.

In recent years, video game engines, such as Unity or the Unreal Engine,
have become particularly popular in the game and architectural
visualisation industry. They provide very sophisticated graphics engines
and offer a lot of tools for user interaction, such as menu systems or
native handling of VR devices. They are well supported by industry and
tend to have a long lifespan (Unreal Engine is now  20 years old and
is still very popular). However, such engines are meant to be used as
development frameworks and their usage in HEP code is not always
evident. Code should be developed within them, while in HEP
framework-based applications we often want to use graphics libraries
that can be integrated in existing code. A number of HEP collaborations
have started experimenting in building event display tools with such
engines, among them Belle II and ATLAS, but their use is currently
limited to the display of simplified data only.

The new client-server architecture proposed as one of the visualisation
R\&D activities will ease the usage of WebGL technologies and game
engines in HEP.

For statistical data, ROOT has been the tool of choice in HEP for many
years and satisfies most use cases. However, increasing use of generic
tools and data formats means Matplotlib (Python) or JavaScript based
solutions (used, for example, in Jupyter notebooks) have made the
landscape more diverse. For visualising trees or graphs interactively,
there are many generic offerings and experiments have
started to take advantage of them.

\subsubsection*{Research and Development Roadmap}

The main goal of R\&D projects in this area will be to develop
techniques and tools that let visualisation applications and event
displays be less dependent on specific experiments' software frameworks,
leveraging common packages and common data formats. Exporters and
interface packages will be designed as bridges between the experiments'
frameworks, needed to access data at a high level of detail, and the
common packages based on the community standards that this group will
develop.

As part of this development work, demonstrators will be designed to show
the usability of our community solutions and tools. The goal will be to
get a final design of those tools so that the experiments can depend on
them in their future developments.

The working group will also work towards a more convenient access to geometry and
event data, through a client-server interface. In collaboration with the
Data Access and Management Working Group, an API or a service to deliver streamed
event data would be designed.

The work above should be completed by 2020.

Beyond that point, the focus will be on developing the actual
community-driven tools, to be used by the experiments for their
visualisation needs in production, potentially taking advantage of new
data access services.

The workshop that was held as part of the CWP process was felt to be
extremely useful for exchanging knowledge between developers in
different experiments, fostering collaboration and in bringing in ideas
from outside the community. This will now be held as an annual event and
will facilitate work on the common R\&D plan.

\hypertarget{software-development-deployment-validation-and-verification}{%
\subsection{Software Development, Deployment, Validation and
Verification}\label{software-development-deployment-validation-and-verification}}

\subsubsection*{Scope and Challenges}

Modern HEP experiments are often large distributed collaborations
with several hundred people actively writing software. It is
therefore vital that the processes and tools used for development are
streamlined to ease the process of contributing code and to facilitate
collaboration between geographically separated peers. At the same time,
we must properly manage the whole project, ensuring code quality,
reproducibility, and maintainability with the least effort possible.
Making sure this happens is largely a continuous process and shares a
lot with non-HEP specific software industries.

Work is ongoing to track and promote solutions in the following areas:

\begin{itemize}
\item
  Distributed development of software components, including the tools
  and processes required to do so (code organisation, documentation,
  issue tracking, artefact building), and the best practices in terms of
  code and people management.
\item
  Software quality, including aspects such as modularity and reusability
  of the developed components, architectural and performance best
  practices.
\item
  Software sustainability, including both development and maintenance
  efforts, as well as best practices given long timescales of HEP
  experiments.
\item
  Deployment of software and interaction with operations teams.
\item
  Validation of the software both at small scales (e.g., best practices
  on how to write a unit test) and larger ones (large scale validation
  of data produced by an experiment).
\item
  Software licensing and distribution, including their impact on
  software interoperability.
\item
  Recognition of the significant contribution that software makes to HEP
  as a field (also see Section \ref{training-and-careers} regarding
  career recognition).
\end{itemize}

HEP-specific challenges derive from the fact that HEP is a large,
inhomogeneous community with multiple sources of funding, mostly formed
of people belonging to university groups and HEP-focused laboratories.
Software development effort within an experiment usually encompasses a
huge range of experience and skills, from a few more or less full-time
experts to many physicist programmers with little formal software
training. In addition, the community is split between different
experiments that often diverge in timescales, size, and resources.
Experiment software is usually divided in two separate use cases:
production (being it data acquisition, data reconstruction or
simulation) and user analysis, whose requirements and lifecycles are
completely different. The former is very carefully managed in a
centralised and slow-moving manner, following the schedule of the
experiment itself. The latter is much more dynamic and strongly coupled
with conferences or article publication timelines. Finding solutions
that adapt well to both cases is not always obvious or even possible.

\subsubsection*{Current Practices}

Due to significant variations between experiments at various stages of
their lifecycles, there is a huge variation in practice across the
community. Thus, here we describe \emph{best practice}, with the
understanding that this ideal may be far from the reality for some
developers.

It is important that developers can focus on the design and
implementation of the code and do not have to spend a lot of time on
technical issues. Clear procedures and policies must exist to perform
administrative tasks in an easy and quick way. This starts with the
setup of the development environment. Supporting different platforms not
only allows developers to use their machines directly for 
development, it also provides a check of code portability. Clear
guidance and support for good design must be available in advance of
actual coding.

To maximise productivity, it is very beneficial to use development tools
that are not HEP-specific. There are many open source projects that are
of similar scale to large experiment software stacks and standard tools
are usually well documented. For source control HEP has generally chosen
to move to \emph{git}~\cite{git},
which is very welcome, as it also brings an
alignment with many open source projects and commercial organisations.
Likewise, \emph{CMake}~\cite{CMake} 
is widely used for the builds of software
packages, both within HEP and outside. Packaging many build products
together into a software stack is an area that still requires close
attention with respect to active developments (the HSF has an active
working group here).

Proper testing of changes to code should always be done in advance of a
change request to be accepted. Continuous integration, where `merge' or
`pull' requests are built and tested in advance, is now standard practice
in the open source community and in industry. Continuous integration can
run unit and integration tests, and can also incorporate code quality
checks and policy checks that help improve the consistency and quality
of the code at low human cost. Further validation on different platforms
and at large scales must be as automated as possible, including the
deployment of build artefacts for production.

Training (Section \ref{training-and-careers}) and documentation 
are key to efficient use
of developer effort. Documentation must cover best practices and
conventions as well as technical issues. For documentation that has to
be specific, the best solutions have a low barrier of entry for new
contributors, but also allow and encourage review of material.
Consequently, it is very useful to host documentation sources in a
repository with a similar workflow to code, and to use an engine that
translates the sources into modern web pages.

Recognition of software work as a key part of science has resulted in a
number of journals where developers can publish their work~\cite{SSI2017}. Journal publication also disseminates information to the
wider community in a permanent way and is the most established mechanism
for academic recognition. Publication in such journals provides proper
peer review, beyond that provided in conference papers, so it is
valuable for recognition as well as dissemination. However, this
practice is not widespread enough in the community and needs further
encouragement.

\subsubsection*{Research and Development Programme}

HEP must endeavour to be as responsive as possible to developments
outside of our field. In terms of hardware and software tools, there
remains great uncertainty as to what the platforms offering the best
value for money will be on the timescale of a decade. It therefore
behoves us to be as generic as possible in our technology choices,
retaining the necessary agility to adapt to this uncertain future.

Our vision is characterised by HEP being current with technologies and
para\-digms that are dominant in the wider software development community,
especially for open-source software, which we believe to be the right
model for our community. In order to achieve that aim, we propose that
the community establishes a development forum that allows for technology
tracking and discussion of new opportunities. The HSF can play a key
role in marshalling this group and in ensuring its findings are widely
disseminated. In addition, having wider and more accessible training for
developers in the field, that will teach the core skills needed for
effective software development, would be of great benefit.

Given our agile focus, it is better to propose here projects and
objectives to be investigated in the short to medium term, alongside
establishing the means to continually review and refocus the community
on the most promising areas. The main idea is to investigate new tools
as demonstrator projects where clear metrics for success in a reasonable
time should be established to avoid wasting community effort on
initially promising products that fail to live up to expectations.

Ongoing activities and short-term projects, include the following:

\begin{itemize}
\item
  Establish a common forum for the discussion of HEP software problems.
  This should be modeled along the lines of the Concurrency Forum~\cite{ConcurrencyForum}, which was very successful in establishing
  demonstrators and prototypes that were used as experiments started to
  develop parallel data processing frameworks.
\item
  Continue the HSF working group on \emph{Packaging}, with more
  prototype implementations based on the strongest candidates identified
  so far.
\item
  Provide practical advice on how to best set up new software packages,
  developing on the current project template work, and working to
  advertise this within the community.
\item
  Work with HEP experiments and other training projects to provide
  accessible core skills training to the community (see Section \ref{training-and-careers}). This
  training should be experiment-neutral, but could be usefully combined
  with the current experiment specific training. Specifically, this work
  can build on, and collaborate with, recent highly successful
  initiatives such as the LHCb \emph{Starterkit}~\cite{LHCbStarterkit}
  and ALICE \emph{Juniors}~\cite{ALICEJuniors}, and with established
  generic training initiatives such as \emph{Software Carpentry}~\cite{SoftwareCarpentry}.
\item
  Strengthen links with software communities and conferences outside of
  the HEP domain, presenting papers on the HEP experience and problem
  domain. The Scientific Computing with Python (SciPy), the
  Supercomputing Conferences (SCxx), the Conference of
  Research Software Engineers \emph{(RSE)}, and the Workshops on
  Sustainable Software for Science: Practice and Experiences (WSSSPE)
  would all be useful meetings to consider.
\item
  Write a paper that looks at case studies of successful and
  unsuccessful HEP software developments and that draws specific
  conclusions and advice for future projects.
\item
  Strengthen the publication record for important HEP software packages.
  Both peer-reviewed journals~\cite{SSI2017} and citable software version
  records (such as DOIs obtained via Zenodo~\cite{Zenodo}).
\end{itemize}

Medium term projects include the following:

\begin{itemize}
\item
  Prototype C++ refactoring tools, with specific use cases in migrating
  HEP code.
\item
  Prototyping of portable solutions for exploiting modern vector
  hardware on heterogenous platforms.
\item
  Support the adoption of industry standards and solutions over
  HEP-specific implementations whenever possible.
\item
  Develop tooling and instrumentation to measure software performance
  where tools with sufficient capabilities are not available from
  industry, especially in the domain of concurrency. This should
  primarily aim to further developments of existing tools, such as
  \emph{igprof}~\cite{Eulisse:2005zz},
  rather than to develop new ones.
\item
  Develop a common infrastructure to gather and analyse data about
  experiments' software, including profiling information and code
  metrics, and to ease sharing across different user communities.
\item
  Undertake a feasibility study of a common toolkit for statistical
  analysis that would be of use in regression testing for experiment's
  simulation and reconstruction software.
\end{itemize}

\hypertarget{data-and-software-preservation}{%
\subsection{Data and Software
Preservation}\label{data-and-software-preservation}}

\subsubsection*{Scope and Challenges}

Given the very large investment in particle physics experiments, it is
incumbent upon physicists to preserve the data and the knowledge that
leads to scientific results in a manner such that this investment is not
lost to future generations of scientists. For preserving ``data'', at
whatever stage of production, many of the aspects of the low level
bit-wise preservation have been covered by the Data Preservation for HEP
group~\cite{DPHEP}. ``Knowledge'' preservation encompasses the more
challenging aspects of retaining processing and analysis software,
documentation, and other components necessary for reusing a given
dataset. Preservation of this type can enable new analyses on older
data, as well as a way to revisit the details of a result after
publication. The latter can be especially important in resolving
conflicts between published results, applying new theoretical
assumptions, evaluating different theoretical models, or tuning new
modeling techniques.

Preservation enabling reuse can offer tangible benefits within a given
experiment. The preservation of software and workflows such that they
can be shared enhances collaborative work between analysts and analysis
groups, providing a way of capturing the knowledge behind a given
analysis during the review process. It enables easy transfer of knowledge
to new students or analysis teams, and could establish a manner by which
results can be generated automatically for submission to central
repositories, such as HEPData~\cite{Maguire:2017ypu}.
Preservation within an experiment can
provide ways of reprocessing and reanalysing data that could have been
collected more than a decade earlier. Benefits from preservation are
derived internally whether or not analysis work is approved through the
publication approval process for an experiment. Providing such immediate
benefits makes the adoption of data preservation in experiment workflows
particularly desirable.

A final series of motivations comes from the potential re-use by others
outside of the HEP experimental community. Significant outreach efforts
to bring the excitement of analysis and discovery to younger students
have been enabled by the preservation of experimental data and software
in an accessible format. Many examples also exist of phenomenology
papers reinterpreting the results of a particular analysis in a new
context. This has been extended further with published results based on
the reanalysis of processed data by scientists outside of the
collaborations. Engagement of external communities, such as machine
learning specialists, can be enhanced by providing the capability to
process and understand low-level HEP data in portable and relatively
platform-independent way, as happened with the Kaggle ML
challenges~\cite{pmlr-v42-cowa14}. 
This allows external users direct access to the same tools
and data as the experimentalists working in the collaborations.
Connections with industrial partners, such as those fostered by CERN
OpenLab, can be facilitated in a similar manner.

Preserving the knowledge of analysis, given the extremely wide scope of
how analysts do their work and experiments manage their workflows, is
far from easy. The level of reuse that is applicable needs to be
identified, and so a variety of preservation systems will probably be
appropriate given the different preservation needs between large central
experiment workflows and the work of an individual analyst. The larger
question is to what extent common low-level tools can be provided that
address similar needs across a wide scale of preservation problems.
These would range from capture tools, that preserve the details of an
analysis and its requirements, to ensuring that software and services
needed for a workflow would continue to function as required.

The above-mentioned steps can be considered to be consistent with the
FAIR data principles that are increasingly being mandated by funding
agencies~\cite{FAIRdata}.

\subsubsection*{Current Practices}

Each of the LHC experiments has adopted a data access and/or data
preservation policy, all of which can be found on the CERN Open Data
Portal~\cite{ODP}. All of the LHC experiments support public access to
some subset of the data in a highly reduced data format for the purposes
of outreach and education. CMS has gone one step further, releasing
substantial datasets in an Analysis Object Data (AOD) format that can be
used for new analyses. The current data release includes simulated data,
virtual machines that can instantiate the added analysis examples, and
extensive documentation~\cite{CMS-OpenData}. ALICE has promised to
release 10\% of their processed data after a five-year embargo and has
released 2010 data at this time~\cite{ALICE-OpenData}.
LHCb is willing to make access to reconstructed data available, but is
unable to commit to a specific timescale due to resource limitations. A
release of ntuple-level data for one high profile analysis, aimed
primarily at educational activities, is currently in preparation. ATLAS
has chosen a different direction for data release: data associated with
journal publications is made available, and ATLAS also strives to make
available additional material that allows reuse and reinterpretations of
the data in the context of new theoretical models~\cite{ATL-CB-PUB-2015-001}.
ATLAS is  exploring how
to provide the capability for reinterpretation of searches in the future
via a service such as 
RECAST~\cite{Cranmer:1299950}, in which the original internal analysis
code (including full detector simulation and reconstruction) is
preserved, as opposed to the re-coding approach with object-efficiency
calibrations used by external reinterpretation toolkits. All experiments
frequently provide detailed supplemental data along with publications to
allow for more detailed comparisons between results, or even
reinterpretation.

The LHC experiments have not yet set a formal policy addressing the new
capabilities of the CERN Analysis Preservation Portal (CAP)~\cite{CAP}
and whether or not some use of it will be required or merely encouraged.
All of them support some mechanisms for internal preservation of the
knowledge surrounding a physics publication~\cite{Shiers2017}.

\subsubsection*{Research and Development Programme}

There is a significant programme of work already happening in the data
preservation area. The feasibility and cost of common base services have
been studied for bit preservation, the preservation of executable
software environments, and the structured capturing of analysis metadata~\cite{Berghaus2016}.

The goals presented here should be orchestrated in conjunction with
projects conducted by the R\&D programmes of other working groups, since
the questions addressed are common. Goals to address on the timescale of
2020 are:

\begin{itemize}
\item
  Include embedded elements for the capture of preservation information
  and metadata and tools for the archiving of this information in
  developing a prototype analysis ecosystem(s). This should include an
  early demonstration of the CAP analysis preservation portal with a
  working UI.
\item
  Demonstrate the capability to provision and execute production
  workflows for experiments that are composed of multiple independent
  containers.
\item
  Collection of analysis use cases and elements that are necessary to
  preserve in order to enable re-use and to ensure these analyses can be
  captured in developing systems. This should track analysis evolution
  towards possible \emph{Big Data} environments and determine any elements
  that are difficult to capture, spawning further R\&D.
\item
  Evaluate, in the preservation area, the full potential and limitations
  of sandbox and ``freezing'' technologies, possibly coupled with
  version and history control software distribution systems.
\item
  Develop prototypes for the preservation and validation of large-scale
  production executables and workflows.
\item
  Integrate preservation capabilities into newly developed computing
  tools and workflows.
\item
  Extension and standardisation of the \emph{final data and analysis}
  preservation sche\-me via HEPData, Rivet and/or other reinterpretation
  tools. This could be used to preserve a sufficiently detailed
  re-usable record of many LHC Run 2 research outputs.
\end{itemize}

This would then lead naturally to deployed solutions that support data
preservation in the 2020-2022 time frame for the HEP experimental
programmes, in particular an analysis ecosystem that enables reuse for
any analysis that can be conducted in the ecosystem, and a system for
the preservation and validation of large-scale production workflows.

\hypertarget{security}{%
\subsection{Security}\label{security}}

\subsubsection*{Scope and Challenges}

Security is a cross-cutting area that impacts our projects,
collaborative work, users, and software infrastructure fundamentally. It
crucially shapes our reputation, our collaboration, the trust between
participants, and the users' perception of the quality and ease of use
of our services.

There are three key areas:

\begin{itemize}
\item
  Trust and policies; this includes trust models, policies, compliance,
  data protection issues.
\item
  Operational security; this includes threat intelligence, security
  operations, incident response.
\item
  Authentication and Authorisation; this includes identity management,
  identity federation, access control.
\end{itemize}

\paragraph{Trust and Policies}

Data Protection defines the boundaries that enable HEP work to be
conducted, in particular regarding data sharing aspects, for example
between the EU and the US. It is essential to establish a trusted
personal data exchange framework, minimising the amount of personal data
to be processed and ensuring legal compliance.

Beyond legal compliance and best practice, offering open access to
scientific resources and achieving shared goals requires prioritising
the protection of people and science, including the mitigation of the
effects of surveillance programs on scientific collaborations.

On the technical side, it is necessary to adapt the current, aging trust
model and security architecture relying solely on X.509 (which is no longer
the direction industry is taking), in order to include modern data
exchange design, for example involving commercial providers or hybrid
clouds. The future of our infrastructure involves increasingly diverse
resource providers connected through cloud gateways. For example,
HEPCloud~\cite{HEPCloud} at FNAL aims to connect Amazon,
Google Clouds, and HPC centres with our traditional grid computing
resources. The HNSciCloud European Project~\cite{HNSciCloud} aims to
support the enhancement of commercial cloud providers in order to be
leveraged by the scientific community. These are just two out of a
number of endeavours. As part of this modernisation, a transition is
needed from a model in which all participating organisations are bound
by custom HEP security policies to a more flexible approach where some
partners are not in a position to adopt such policies.

\paragraph{Operational Security and Threat Intelligence}

As attacks have become extremely sophisticated and costly to defend
against, the only cost-effective strategy is to address security threats
together, as a community. This involves constantly striving to liaise
with external organisations, including security vendors and law
enforcement entities, to enable the sharing of indicators of compromise
and threat intelligence between all actors. For organisations from all
sectors, including private companies, governments, and academia, threat
intelligence has become the main means by which to detect and manage
security breaches.

In addition, a global forum for HEP and the larger Research and Education
(R\&E) community needs to be built, where security experts feel
confident enough to share threat intelligence and security expertise. A
key to success is to ensure a closer collaboration between HEP security
contacts and campus security. The current gap at many HEP organisations
is both undermining the community's security posture and reducing the
effectiveness of the HEP security strategy.

There are several very active trust groups in the HEP community where
HEP participants share threat intelligence and organise coordinated
incident response \cite{EGI-CSIRT,REN-ISAC,XSEDE}.
There is unfortunately still no global Research and Education forum for
incident response, operational security, and threat intelligence
sharing. With its mature security operations and dense, global network
of HEP organisations, both of which are quite unique in the research
sector, the HEP community is ideally positioned to contribute to such a
forum and to benefit from the resulting threat intelligence, as it has
exposure, sufficient expertise, and connections to lead such an
initiative. It may play a key role in protecting multiple scientific
domains at a very limited cost.

There will be many technology evolutions as we start to take a serious
look at the next generation internet. For example, IPv6 is one upcoming
change that has yet to be fully understood from the security
perspective. Another high impact area is the internet of things (IoT),
connected devices on our networks that create new vectors of attack.

It will become necessary to evaluate and maintain operational security
in connected environments spanning public, private, and hybrid clouds.
The trust relationship between our community and such providers has yet
to be determined, including the allocation of responsibility for
coordinating and performing vulnerability management and incident
response. Incompatibilities between the e-Infrastructure approach to
community-based incident response and the ``pay-for-what-you-break''
model of certain commercial companies may come to light and must be
resolved.

\paragraph{Authentication and Authorisation Infrastructure}

It is now largely acknowledged that end-user certificates are
challenging to manage and create a certain entrance barrier to our
infrastructure for early career researchers. Integrating our access
control management system with new, user-friendly technologies and
removing our dependency on X.509 certificates is a key area of interest
for the HEP Community.

An initial step is to identify other technologies that can satisfy
traceability, isolation, privilege management and other requirements
necessary for HEP workflows. The chosen solution should prioritise
limiting the amount of change required to our services and follow
accepted standards to ease integration with external entities, such as
commercial clouds and HPC centres.

Trust federations and inter-federations, such as the R\&E standard
eduGAIN \cite{eduGAIN}, provide a needed functionality for
Authentication. They can remove the burden of identity provisioning from
our community and allow users to leverage their home organisation
credentials to access distributed computing resources. Although certain
web-based services have enabled authentication via such federations,
uptake is not yet widespread. The challenge remains to have the
necessary attributes published by each federation to provide robust
authentication.

The existing technologies leveraged by identity federations, e.g., the
Security Assertion Markup Language (SAML), have not supported non-web
applications historically. There is momentum within the wider community
to develop next-generation identity federations that natively support a
wider range of clients. In the meantime there are several viable interim
solutions that are able to provision users with the token required to
access a service (such as X.509) transparently, translated from their
home organisation identity.

Although federated identity provides a potential solution for our
challenges in Authentication, Authorisation should continue to be
tightly controlled by the HEP community. Enabling Virtual Organisation
(VO) membership for federated credentials and integrating such a
workflow with existing identity vetting processes is a major topic
currently being worked on, in particular within the WLCG community.
Commercial clouds and HPC centres have fundamentally different access
control models and technologies from our grid environment. We shall need
to enhance our access control model to ensure compatibility and
translate our grid-based identity attributes into those consumable by
such services.

\subsubsection*{Current Activities}

Multiple groups are working on policies and establishing a common trust
framework, including the EGI Security Policy Group
\cite{EGISecurityPolicyGroup} and the Security for Collaboration among
Infrastructures working group~\cite{SCI_WG}.

Operational security for the HEP community is being followed up in the
WLCG Working Group on Security Operations Centres~\cite{WLCG-SOC-WG}. The
HEP Community is actively involved in multiple operational security
groups and trust groups, facilitating the exchange of threat
intelligence and incident response communication. WISE~\cite{WISE}
provides a forum for e-Infrastructures to share and develop security
best practices and offers the opportunity to build relationships between
security representatives at multiple e-infrastructures of interest to
the HEP community.

The evolution of Authentication and Authorisation is being evaluated in
the recently created WLCG Working Group on Authorisation. 
In parallel, HEP is contributing to a wider effort
to document requirements for multiple Research Communities through the
work of FIM4R~\cite{FIM4R}. CERN's participation in the European
Authentication and Authorisation for Research and Collaboration (AARC)
project~\cite{AARC} provides the opportunity to ensure
that any directions chosen are consistent with those taken by the wider
community of research collaborations. The flow of attributes between
federated entities continues to be problematic, disrupting the
authentication flow. Trust between service providers and identity
providers is still evolving, and efforts within the R\&E Federations
Group (REFEDS)~\cite{REFEDS} and the AARC project aim to address the
visibility of both the level of assurance of identities and the security
capability of federation participants (through Sirtfi~\cite{Sirtfi}).

\subsubsection*{Research and Development Programme}

Over the next decade, it is expected that considerable changes will be
made to address security in the domains highlighted above. The
individual groups, in particular those mentioned above, working in the
areas of trust and policies, operational security, authentication and
authorisation, and technology evolutions, are driving the R\&D
activities. The list below summarises the most important actions:

\paragraph{Trust and Policies}
\begin{itemize}
\item 
By 2020:
\begin{itemize}
    \item
      Define and adopt policies in line with new EU Data Protection
      requirements.
    \item
      Develop frameworks to ensure trustworthy interoperability of
      infrastructures and communities.
\end{itemize}

\item
By 2022:
\begin{itemize}
    \item
      Create and promote community driven incident response policies and
      procedures.
\end{itemize}
\end{itemize}

\paragraph{Operational Security and threat intelligence}
\begin{itemize}
\item
By 2020:
\begin{itemize}
\item
  Offer a reference implementation, or at least specific guidance, for a
  Security Operation Centre deployment at HEP sites, enabling them to
  take action based on threat intelligence shared within the HEP
  community.
\end{itemize}
\item
By 2022:
\begin{itemize}
\item
  Participate in the founding of a global Research and Education Forum
  for incident response, since responding as a global community is the only
  effective solution against global security threats.
\item
  Build the capabilities to accommodate more participating organisations
  and streamline communication workflows, within and outside HEP,
  including maintaining a list of security contacts, secure communications
  channels, and security incident response mechanisms.
\item
  Reinforce the integration of HEP security capabilities with their
  respective home organisation, to ensure adequate integration of HEP
  security teams and site security teams.
\end{itemize}
\item 
By 2025:
\begin{itemize}
\item
  Prepare adequately as a community, in order to enable HEP
  organisations to operate defendable services against more
  sophisticated threats, stemming both from global cyber-criminal gangs
  targeting HEP resources (finance systems, intellectual property,
  ransomware), as well as from state actors targeting the energy and
  research sectors with advanced malware.
\end{itemize}
\end{itemize}

\paragraph{Authentication and Authorisation}

\begin{itemize}
\item
By 2020:
\begin{itemize}
\item
  Ensure that ongoing efforts in trust frameworks are sufficient to
  raise the level of confidence in federated identities to the
  equivalent of X.509, at which stage they could be a viable alternative
  to both grid certificates and CERN accounts.
\item
  Participate in setting directions for the future of identity
  federations, through the FIM4R~\cite{FIM4R} community.
\end{itemize}
\item
By 2022:
\begin{itemize}
\item
  Overhaul the current Authentication and Authorisation infrastructure,
  including Token Translation, integration with Community IdP-SP
  Proxies, and Membership Management tools. Enhancements in this area
  are needed to support a wider range of user identities for WLCG
  services.
\end{itemize}
\end{itemize}

\hypertarget{training-and-careers}{%
\section{Training and Careers}\label{training-and-careers}}

For HEP computing to be as successful as possible, the careers and
skills of the individuals who participate must be considered. Ensuring
that software developers can acquire the necessary skills and obtain
successful careers is considered an essential goal of the HSF, which has
the following specific objectives in its mission:

\begin{itemize}
\item
  To provide training opportunities for developers; this should include
  the support to the software schools for young scientists and computer
  engineers, and of a permanent training infrastructure for accomplished
  developers;
\item
  To provide career support for developers, for instance by listing job
  opportunities and by helping to shape well-defined career paths that
  provide advancement opportunities on a par with those in, for example,
  detector construction;
\item
  To increase the visibility of the value of software developers in HEP,
  recognising that it has scientific research value on an equal footing
  with other activities, and acknowledging and promoting specific
  ``champions'' in the field.
\end{itemize}

\subsection{Training Challenges}

HEP is facing major challenges with its software and computing that
require innovative solutions based on the proper adoption of new
technologies. More and more technologies are emerging as scientific
communities and industry face similar challenges and produce solutions
relevant to us. Integrating such technologies in our software and
computing infrastructure requires specialists, but it is also important that a
large fraction of the community is able to use these new tools and
paradigms. Specific solutions and optimisations must be implemented by
the HEP community itself, since many advanced requirements are unique to
our field.

Unlike the situation that is traditional in some other fields, in
which users express their requirements and computer specialists
implement solutions, there is a close collaboration, even overlap,
in HEP between users and developers that is essential for our success.
Many details of experiment data
cannot be known before data taking has started, and each change in
detector technology or machine performance improvement can have
important consequences for the software and computing infrastructure. In
the case of detectors, engineers and physicists are required to have a
good understanding of each other's field of expertise. In the same way,
it is necessary that physicists understand some of the complexities of
writing software, and that software experts are able to fathom the
requirements of physics problems.

Training must address an audience with very diverse computing skills,
ranging from novice programmers to advanced developers and users.
It must be used to spread best software engineering practices and
software technologies to a very large number of people, including the
physicists involved across the whole spectrum of data processing tasks,
from triggering to analysis. It must be done by people who have a sound
knowledge of the scientific and technical details, who prepare training
material despite the many calls on their time. Training thus needs
proper recognition to ensure that it happens and is carried out well.

HEP is seen as an interesting, innovative, and challenging field. 
This is a great advantage in attracting talented young people looking
for experience in a challenging and diverse environment in which
they can acquire skills that will be valuable, even in other fields.
As discussed in Software Development (Section
\ref{software-development-deployment-validation-and-verification}),
using industry standard tools across different experiments, 
and training people in how to use them properly,
helps with people's later career prospects
and makes our field even more attractive.
At the same time, experiments have a scientific programme to accomplish
and also to focus on the specific training required to
accomplish their specific goals. The right balance must be found between
these two requirements. It is necessary to find the right incentives to
favour training activities that bring more benefits in the medium to
long term, for the experiment, the community, and the careers of the
trainees.

\subsection{Possible Directions for Training}

To increase training activities in the community, whilst taking into
account the constraints of both the attendees and the trainers, we should
explore new approaches to training. The current ``school''
model is well established, as exemplified by three well-known successful
schools, the CERN School of Computing~\cite{CSC}, 
the Bertinoro School of Computing~\cite{ESC2017}
and the GridKa School of Computing~\cite{GridKASchool}.
They require a significant amount of
dedicated time of all the participants, at the same time and location,
and therefore are difficult to scale to meet the needs of a large number
of students. In view of this, we should identify opportunities to work
with HEP experiments and other training projects to provide accessible
core skills training to the community by basing them at laboratories
where students can easily travel. A number of highly successful
experiment-specific examples exist, such as the LHCb StarterKit~\cite{LHCbStarterkit} 
and ALICE Juniors~\cite{ALICEJuniors}, as well as
established generic training initiatives, such as 
Software Carpentry~\cite{SoftwareCarpentry}. As with hands-on tutorials organised during
conferences and workshops, the resulting networking is an important and
distinctive benefit of these events, where people build relationships
with other colleagues and experts.

In recent years, several R\&D projects, such as DIANA-HEP~\cite{DIANA-HEP} and
AMVA4\-New\-Physics~\cite{AMVA4NewPhysics}, have had training as one 
of their core activities. This
has provided an incentive to organise training events and has resulted
in the spread of expertise on advanced topics. We believe that training
should become an integral part of future major R\&D projects.

New pedagogical methods, such as active training and peer training, that
are complementary to schools or topical tutorials, also deserve more
attention. Online material can be shared by a student and a teacher to
provide the exchange of real examples and practical exercises. For
example, notebook technologies, such as Jupyter, support embedding of
runnable code and comments into the same document. The initial material
can be easily enriched by allowing other students and experts to add
comments and more examples in a collaborative way. The HSF started to
experiment with this approach with WikiToLearn~\cite{WikiToLearn}, a
platform developed in Italy outside HEP that promotes this kind of
training and collaborative enrichment of the training material. Projects
such as ROOT~\cite{Brun1996} have also started to provide some
training material based
on notebooks.

A lot of initiatives have been undertaken by the software community that
HEP can benefit from, and materials have been made available in the form
of online tutorials, active training, and Massive Open Online Courses
(MOOCs). Some effort needs to be invested to evaluate existing courses
and build a repository of selected ones that are appropriate to HEP
needs. This is not a negligible task and would require some dedicated
effort to reach the appropriate level of support. It should help to
increase training efficiency by making it easier to identify appropriate
courses or initiatives.

A model that emerged in recent years as a very valuable means of sharing
expertise is to use Question and Answer (Q\&A) systems, such as Stack
Overflow. A few such systems are run by experiments for their own needs,
but this is not necessarily optimal, as the value of these services is
increased by a large number of contributors with diverse backgrounds.
Running a cross-experiment Q\&A system has been discussed, but it has
not yet been possible to converge on a viable approach, both technically
and because of the effort required to run and support such a service.

\subsection{Career Support and Recognition}

Computer specialists in HEP are often physicists who have chosen to
specialise in computing. This has always been the case and needs to
continue. Nevertheless, for young people in particular, this leads to a
career recognition problem, as software and computing activities are not
well-recognised roles in various institutions supporting HEP research
and recruiting people working in the field. The exact situation is
highly dependent on policies and boundary conditions of the organisation
or country, but recognition of physicists tends to be based generally on
participation in data analysis or hardware developments. This is even a
bigger problem if the person is spending time contributing to training
efforts. This negatively impacts the future of these people and reduces
the possibility of HEP engaging them in the training effort of the
community when the community actually needs more people to participate
in this activity. Recognition of training efforts, either by direct
participation in training activities or by providing materials, is an
important issue to address, complementary to the incentives mentioned
above.

There is no easy solution to this problem. Part of the difficulty is
that organisations, and in particular the people inside them in charge
of the candidate selections for new positions and promotions, need to
adapt their expectations to these needs and to the importance of having
computing experts with a strong physics background as permanent members
of the community. Experts writing properly engineered and optimised
software can significantly reduce resource consumption and increase
physics reach, which provides huge financial value to modern HEP
experiments. The actual path for improvements in career recognition, as
the possible incentives for participating in the training efforts,
depends on the local conditions.

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

Future challenges for High Energy Physics in the domain of software and
computing are not simply an extrapolation of the challenges faced today.
The needs of the HEP programme in the high luminosity era far exceed those
that can be met by simply making incremental changes to today's code and
scaling up computing facilities within the anticipated budget. At the
same time, the limitation in single core CPU performance is making the
landscape of computing hardware far more diverse and challenging to
exploit, whilst offering huge performance boosts for suitable code.
Exploiting parallelism and other new techniques, such as modern machine
learning, offer great promise, but will require substantial work from
the community to adapt to our problems. If there were any lingering
notion that software or computing could be done cheaply by a few junior
people for modern experimental programmes, it should now be thoroughly
dispelled.

We believe HEP Software and Computing requires a step change in its profile and
effort to match the challenges ahead. We need investment in people who
can understand the problems we face, the solutions employed today, and
have the correct skills to provide innovative solutions for the future.
There needs to be recognition from the whole community for the work done
in this area, with a recognised career path for these experts. In
addition, we will need to invest heavily in training for the whole
software community as the contributions of the bulk of non-expert
physicists are also vital for our success.

We know that in any future scenario development effort will be
constrained, so it is vital that successful R\&D projects provide
sustainable software for the future. In many areas it is recognised 
that different experiments could have adopted common solutions, 
reducing overall development effort and increasing robustness and 
functionality. That model of duplicated development is not sustainable. We must
endeavour to achieve better coherence within HEP for future developments
to build advanced, open-source projects that can be shared and supported
in common. The HSF has already established itself as a forum that can
facilitate this. Establishing links outside of HEP, to other academic
disciplines, to industry, and to the computer science community, can
strengthen both the research and production phases of new solutions. We
should ensure that the best products are chosen, from inside and outside
HEP, and that they receive support from all parties, aiming at technical
excellence and economy of scale.

We have presented programmes of work that the community has identified
as being part of the roadmap for the future. While there is always some
scope to reorient current effort in the field, we would highlight the
following work programmes as being of the highest priority for
investment to address the goals that were set in the introduction.

\paragraph{Improvements in software efficiency, scalability and performance}

\begin{quote}
The bulk of CPU cycles consumed by experiments relate to the fundamental
challenges of simulation and reconstruction. Thus, the work programmes
in these areas, together with the frameworks that support them, are of
critical importance. The sheer volumes of data involved make research
into appropriate data formats and event content to reduce storage
requirements vital. Optimisation of our distributed computing systems,
including data and workload management, is paramount.
\end{quote}

\paragraph{Enable new approaches that can radically extend physics reach}

\begin{quote}
New techniques in simulation and reconstruction will be vital here.
Phys\-ics analysis is an area where new ideas can be particularly
fruitful. Exploring the full potential of machine learning is one common
theme that underpins many new approaches and the community should
endeavour to share knowledge widely across subdomains. New data analysis
paradigms coming from the Big Data industry, based on innovative
parallelised data processing on large computing farms, could transform
data analysis.
\end{quote}

\paragraph{Ensure the long-term sustainability of the software}

\begin{quote}
Applying modern software development techniques to our codes has
increased, and will continue to increase, developer productivity and
code quality. There is ample scope for more common tools and common
training to equip the community with the correct skills. Data
Preservation makes sustainability an immediate goal of development and
analysis and helps to reap the benefits of our experiments for decades
to come. Support for common software used across the community needs to
be recognised and accepted as a common task, borne by labs, institutes,
experiments, and funding agencies.
\end{quote}

The R\&D actions proposed in this Roadmap have taken into account the
charges that were laid down. When considering a specific project
proposal addressing our computing challenges, that project's impact, measured
against the charges, should be evaluated. Over the next decade, there
will almost certainly be disruptive changes that cannot be planned for,
and we must remain agile enough to adapt to these.

The HEP community has many natural subdivisions, between different
regional funding agencies, between universities and laboratories, and
between different experiments. It was in an attempt to overcome these
obstacles, and to encourage the community to work together in an
efficient and effective way, that the HEP Software Foundation was
established in 2014. This Community White Paper process has been
possible only because of the success of that effort in bringing the
community together. The need for more common developments in the future,
as underlined here, reinforces the importance of the HSF as a common
point of contact between all the parties involved, strengthening our
community spirit and continuing to help share expertise and identify
priorities. Even though this evolution will also require projects and
experiments to define clear priorities about these common developments,
we believe that the HSF, as a community effort, must be strongly
supported as part of our roadmap to success.

\newpage
\begin{appendices}

\hypertarget{appendix-a---list-of-workshops}{%
\section{List of Workshops}\label{appendix-a---list-of-workshops}}

\textbf{HEP Software Foundation Workshop}\\
\emph{Date:} 23-26 Jan, 2017\\
\emph{Location:} UCSD/SDSC (La Jolla, CA, USA)\\
\emph{URL:}
\href{http://indico.cern.ch/event/570249/}{{http://indico.cern.ch/event/570249/}}\\
\emph{Description:} This HSF workshop at SDSC/UCSD was the first
workshop supporting the CWP process. There were plenary sessions
covering topics of general interest as well as parallel sessions for the
many topical working groups in progress for the CWP.\\

\noindent
\textbf{Software Triggers and Event Reconstruction WG meeting}\\
\emph{Date:} 9 Mar, 2017\\
\emph{Location:} LAL-Orsay (Orsay, France)\\
\emph{URL:}
\href{https://indico.cern.ch/event/614111/}{{https://indico.cern.ch/event/614111/}}\\
\emph{Description:} This was a meeting of the Software Triggers and Event
Reconstruction CWP working group. It was held as a parallel session at
the ``Connecting the Dots'' workshop, which focuses on forward-looking
pattern recognition and machine learning algorithms for use in HEP.\\

\noindent
\textbf{IML Topical Machine Learning Workshop}\\
\emph{Date:} 20-22 Mar, 2017\\
\emph{Location:} CERN (Geneva, Switzerland)\\
\emph{URL:}
\href{https://indico.cern.ch/event/595059}{{https://indico.cern.ch/event/595059}}\\
\emph{Description:} This was a meeting of the Machine Learning CWP
working group. It was held as a parallel session at the
``Inter-experimental Machine Learning (IML)'' workshop, an organisation
formed in 2016 to facilitate communication regarding R\&D on ML
applications in the LHC experiments.\\

\noindent
\textbf{Community White Paper Follow-up at FNAL}\\
\emph{Date:} 23 Mar, 2017\\
\emph{Location:} FNAL (Batavia, IL, USA)\\
\emph{URL:}
\href{https://indico.fnal.gov/conferenceDisplay.py?confId=14032}{{https://indico.fnal.gov/conferenceDisplay.py?confId=14032}}\\
\emph{Description:} This one-day workshop was organised to engage with
the experimental HEP community involved in computing and software for
Intensity Frontier experiments at FNAL. Plans for the CWP were
described, with discussion about commonalities between the HL-LHC
challenges and the challenges of the FNAL neutrino and muon
experiments\\

\noindent
\textbf{CWP Visualisation Workshop}\\
\emph{Date:} 28-30 Mar, 2017\\
\emph{Location:} CERN (Geneva, Switzerland)\\
\emph{URL:}
\href{https://indico.cern.ch/event/617054/}{{https://indico.cern.ch/event/617054/}}\\
\emph{Description:} This workshop was organised by the Visualisation CWP
working group. It explored the current landscape of HEP visualisation
tools as well as visions for how these could evolve. There was
participation both from HEP developers and industry.\\

\noindent
\textbf{DS@HEP 2017 (Data Science in High Energy Physics)}\\
\emph{Date:} 8-12 May, 2017\\
\emph{Location:} FNAL (Batava, IL, USA)\\
\emph{URL:}
\href{https://indico.fnal.gov/conferenceDisplay.py?confId=13497}{{https://indico.fnal.gov/conferenceDisplay.py?confId=13497}}\\
\emph{Description:} This was a meeting of the Machine Learning CWP
working group. It was held as a parallel session at the ``Data Science
in High Energy Physics (DS@HEP)'' workshop, a workshop series begun in
2015 to facilitate communication regarding R\&D on ML applications in
HEP.\\

\noindent
\textbf{HEP Analysis Ecosystem Retreat}\\
\emph{Date:} 22-24 May, 2017\\
\emph{Location:} Amsterdam, the Netherlands\\
\emph{URL:}
\href{http://indico.cern.ch/event/613842/}{{http://indico.cern.ch/event/613842/}}\\
\emph{Summary report:}
\href{http://hepsoftwarefoundation.org/assets/AnalysisEcosystemReport20170804.pdf}{{http://cern.ch/go/mT8w}}\\
\emph{Description:} This was a general workshop, organised about the
HSF, about the ecosystem of analysis tools used in HEP and the ROOT
software framework. The workshop focused both on the current status and
the 5-10 year time scale covered by the CWP.\\

\noindent
\textbf{CWP Event Processing Frameworks Workshop}\\
\emph{Date:} 5-6 Jun, 2017\\
\emph{Location:} FNAL (Batavia, IL, USA)\\
\emph{URL:}
\href{https://indico.fnal.gov/conferenceDisplay.py?confId=14186}{{https://indico.fnal.gov/conferenceDisplay.py?confId=14186}}\\
\emph{Description:} This was a workshop held by the Event Processing Frameworks
CWP working group focused on writing an initial draft of the framework
white paper. Representatives from most of the current practice
frameworks participated.\\

\noindent
\textbf{HEP Software Foundation Workshop}\\
\emph{Date:} 26-30 Jun, 2017\\
\emph{Location:} LAPP (Annecy, France)\\
\emph{URL:}
\href{https://indico.cern.ch/event/613093/}{{https://indico.cern.ch/event/613093/}}\\
\emph{Description:} This was the final general workshop for the CWP
process. The CWP working groups came together to present their status
and plans, and develop consensus on the organisation and context for the
community roadmap. Plans were also made for the CWP writing phase that
followed in the few months following this last workshop.

\newpage
\hypertarget{appendix-b-glossary}{%
\section{Glossary}\label{appendix-b-glossary}}

\begin{description}

\item[AOD] Analysis Object Data is a summary of the reconstructed event and
contains sufficient information for common physics analyses.

\item[ALPGEN] An event generator designed for the generation of Standard Model
processes in hadronic collisions, with emphasis on final states with
large jet multiplicities. It is based on the exact LO evaluation of
partonic matrix elements, as well as top quark and gauge boson decays
with helicity correlations.

\item[BSM] Physics beyond the Standard Model (BSM) refers to the theoretical
developments needed to explain the deficiencies of the Standard Model
(SM), such as the
\href{https://en.wikipedia.org/wiki/Origin_of_mass}{origin of mass}, the
\href{https://en.wikipedia.org/wiki/Strong_CP_problem}{strong CP
problem},
\href{https://en.wikipedia.org/wiki/Neutrino_oscillation}{neutrino
oscillations},
\href{https://en.wikipedia.org/wiki/Baryon_asymmetry}{matter--antimatter
asymmetry}, and the nature of
\href{https://en.wikipedia.org/wiki/Dark_matter}{dark matter} and
\href{https://en.wikipedia.org/wiki/Dark_energy}{dark energy}.

\item[Coin3D] A C++ object oriented retained mode 3D graphics API
used to provide a higher layer of programming for OpenGL.

\item[COOL] LHC Conditions Database Project, a subproject of the POOL persistency framework.

Concurrency Forum: Software engineering is moving towards a paradigm shift in order to accommodate new CPU architectures with many cores, in which concurrency will play a more fundamental role in programming languages and libraries. The forum on concurrent programming models and frameworks aims to share knowledge among interested parties that work together to develop 'demonstrators' and agree on technology so that they can share code and compare results.

\item[CRSG] Computing Resources Scrutiny Group, a WLCG committee in charge of scrutinizing and assessing LHC experiment yearly resource requests to prepare funding agency decisions.

\item[CSIRT] Computer Security Incident Response Team. A CSIRT provides a reliable and trusted single point of contact for reporting computer security incidents and taking the appropriate measures in response tothem.

\item[CVMFS] The CERN Virtual Machine File System is a network file system
based on HTTP and optimised to deliver experiment software in a fast,
scalable, and reliable way through sophisticated caching strategies.

\item[CWP] The Community White Paper (this document) is the result of an
organised effort to describe the community strategy and a roadmap for
software and computing R\&D in HEP for the 2020s. This activity is
organised under the umbrella of the HSF.

\item[Deep Learning (DL)] one class of Machine Learning algorithms, based on a
high number of neural network layers.

\item[DNN] Deep Neural Network, class of neural networks with typically a
large number of hidden layers through which data is processed.

\item[DPHEP] The Data Preservation in HEP project is a collaboration for data
preservation and long term analysis.

\item[EGI] European Grid Initiative. A European organisation in charge of
delivering advanced computing services to support scientists,
multinational projects and research infrastructures, partially funded by
the European Union. It is operating both a grid infrastructure (many
WLCG sites in Europe are also EGI sites) and a federated cloud
infrastructure. It is also responsible for security incident response
for these infrastructures (CSIRT).

\item[FAIR] The Facility for Antiproton and Ion Research (FAIR) is located at
GSI Darmstadt. It is an international accelerator facility for research
with antiprotons and ions.

\item[FAIR] An abbreviation for a set of desirable data properties: Findable,
Accessible, Interoperable, and Re-usable.

\item[FCC] Future Circular Collider, a proposed new accelerator complex for
CERN, presently under study.

\item[FCC-hh] A 100 TeV proton-proton collider version of the FCC (the ``h''
stands for ``hadron'').

\item[GAN] Generative Adversarial Networks are a class of
\href{https://en.wikipedia.org/wiki/Artificial_intelligence}{artificial
intelligence} algorithms used in
\href{https://en.wikipedia.org/wiki/Unsupervised_machine_learning}{unsupervised
machine learning}, implemented by a system of two
\href{https://en.wikipedia.org/wiki/Neural_network}{neural networks}
contesting with each other in a
\href{https://en.wikipedia.org/wiki/Zero-sum_game}{zero-sum game}
framework.

\item[Geant4] A toolkit for the simulation of the passage of
particles through matter.

\item[GeantV] An R\&D project that aims to fully exploit the
parallelism, which is increasingly offered by the new generations of CPUs, in the field of detector simulation.

\item[GPGPU] General-Purpose computing on Graphics Processing Units is the use of a Graphics Processing Unit (GPU), which typically handles computation only for computer graphics, to perform computation in applications traditionally handled by the Central Processing Unit (CPU). Programming for GPUs is typically more challenging, but can offer significant gains in arithmetic throughput.

\item[HEPData] The Durham High Energy Physics Database is an open access
repository for scattering data from experimental particle physics.

\item[HERWIG] This is an event generator containing a wide range of Standard
Model, Higgs and supersymmetric processes. It uses the parton-shower approach
for initial- and final-state QCD radiation, including colour coherence effects and 
azimuthal correlations both within and between jets.

\item[HL-LHC] The High Luminosity Large Hadron Collider is a proposed upgrade to
the Large Hadron Collider to be made in 2026. The upgrade aims at increasing the 
luminosity of the machine by a factor of 10, up to 
$10^{35}\mathrm{cm}^{-2}\mathrm{s}^{-1}$, 
providing a better chance to see rare processes and improving
statistically marginal measurements.

\item[HLT] High Level Trigger. The computing resources, generally a large farm, close to the detector which process the events in real-time and select those who must be stored for further analysis.

\item[HPC] High Performance Computing.

\item[HS06] HEP-wide benchmark for measuring CPU performance based on the SPEC2006 benchmark
(\href{https://www.spec.org}{{https://www.spec.org}}).

\item[HSF] The HEP Software Foundation facilitates coordination and common
efforts in high energy physics (HEP) software and computing
internationally.

\item[IML] The Inter-experimental LHC Machine Learning (IML) Working Group is
focused on the development of modern state-of-the art machine learning
methods, techniques and practices for high-energy physics problems.

\item[IOV] Interval Of Validity, the period of time for which a specific piece
of conditions data is valid.

\item[JavaScript] A high-level, dynamic, weakly typed,
prototype-based, multi-paradigm, and interpreted programming language.
Alongside HTML and CSS, JavaScript is one of the three core technologies
of World Wide Web content production.

\item[Jupyter Notebook] This is a server-client application that allows
editing and running notebook documents via a web browser. Notebooks are
documents produced by the Jupyter Notebook App, which contain both
computer code (e.g., python) and rich text elements (paragraph,
equations, figures, links, etc...). Notebook documents are both
human-readable documents containing the analysis description and the
results (figures, tables, etc..) as well as executable documents which
can be run to perform data analysis.

\item[LHC] Large Hadron Collider, the main particle accelerator at CERN.

\item[LHCONE] A set of network circuits, managed worldwide by the National
Research and Education Networks, to provide dedicated transfer paths for
LHC T1/T2/T3 sites on the standard academic and research physical
network infrastructure.

\item[LHCOPN] LHC Optical Private Network. It is the private physical and IP
network that connects the Tier0 and the Tier1 sites of the WLCG.

\item[MADEVENT] This is a multi-purpose tree-level event generator. It is
powered by the matrix element event generator MADGRAPH, which generates
the amplitudes for all relevant sub-processes and produces the mappings
for the integration over the phase space.

\item[Matplotlib] This is a Python 2D plotting library that provides
publication quality figures in a variety of hardcopy formats and
interactive environments across platforms.

\item[ML] Machine learning is a field of computer science that gives computers
the ability to learn without being explicitly programmed. It focuses on
prediction making through the use of computers and emcompasses a lot of
algorithm classes (boosted decision trees, neural networks\ldots{}).

\item[MONARC] A model of large scale distributed computing based on many
regional centers, with a focus on LHC experiments at CERN. As part of
the MONARC project, a simulation framework was developed that provides a
design and optimisation tool. The MONARC model has been the initial
reference for building the WLCG infrastructure and to organise the data
transfers around it.

\item[OpenGL] Open Graphics Library is a cross-language, cross-platform
application programming interface(API) for rendering 2D and 3D vector
graphics. The API is typically used to interact with a graphics
processing unit(GPU), to achieve hardware-accelerated rendering.

\item[Openlab] CERN openlab is a public-private partnership that accelerates
the development of cutting-edge solutions for the worldwide LHC
community and wider scientific research.

\item[P5] The Particle Physics Project Prioritization Panel is a scientific
advisory panel tasked with recommending plans for U.S. investment in
particle physics research over the next ten years.

\item[PRNG] A PseudoRandom Number Generator is an algorithm for generating a
sequence of numbers whose properties approximate the properties of
sequences of random numbers.

\item[PyROOT] A \href{http://www.python.org/}{Python} extension module that
allows the user to interact with any ROOT class from the Python
interpreter.

\item[PYTHIA] A program for the generation of high-energy physics events, i.e.,
for the description of collisions at high energies between elementary
particles such as e+, e-, p and pbar in various combinations. It
contains theory and models for a number of physics aspects, including
hard and soft interactions, parton distributions, initial- and
final-state parton showers, multiparton interactions, fragmentation and
decay.

\item[QCD] Quantum Chromodynamics, the theory describing the strong
interaction between quarks and gluons.

\item[REST] Representational State Transfer
\href{https://en.wikipedia.org/wiki/Web_service}{web services} are a way
of providing interoperability between computer systems on the Internet.
One of its main features is stateless interactions between clients and
servers (every interaction is totally independent of the others),
allowing for very efficient caching.

\item[ROOT] A modular scientific software framework widely used in HEP data
processing applications.

\item[SAML] Security Assertion Markup Language. It is an open, XML-based,
standard for exchanging authentication and authorisation data between
parties, in particular, between an identity provider and a service
provider.

\item[SDN] Software-defined networking is an umbrella term encompassing several kinds of network technology aimed at making the network as agile
and flexible as the virtualised server and storage infrastructure of the modern data center.

\item[SHERPA] Sherpa is a Monte Carlo event generator for the Simulation of
High-Energy Reactions of PArticles in lepton-lepton, lepton-photon,
photon-photon, lepton-hadron and hadron-hadron collisions.

\item[SIMD] Single instruction, multiple data (\textbf{SIMD}), describes
computers with multiple processing elements that perform the same
operation on multiple data points simultaneously.

\item[SM] The Standard Model is the name given in the 1970s to a theory of
fundamental particles and how they interact. It is the currently
dominant theory explaining the elementary particles and their dynamics.

\item[SWAN] Service for Web based ANalysis is a platform for interactive data
mining in the CERN cloud using the Jupyter notebook interface.

\item[TBB] Intel Threading Building Blocks is a widely used C++ template
library for task parallelism. It lets you easily write parallel C++
programs that take full advantage of multicore performance.

\item[TMVA] The Toolkit for Multivariate Data Analysis with ROOT is a
standalone project that provides a ROOT-integrated machine learning
environment for the processing and parallel evaluation of sophisticated
multivariate classification techniques.

\item[VecGeom] The vectorised geometry library for particle-detector simulation.

\item[VO] Virtual Organisation. A group of users sharing a common interest
(for example, each LHC experiment is a VO), centrally managed, and used
in particular as the basis for authorisations in the WLCG
infrastructure.

\item[WebGL] The Web Graphics Library is a JavaScript API for rendering
interactive 2D and 3D graphics within any compatible web browser without
the use of plug-ins.

\item[WLCG] The Worldwide LHC Computing Grid project is a global collaboration
of more than 170 computing centres in 42 countries, linking up national
and international grid infrastructures. The mission of the WLCG project
is to provide global computing resources to store, distribute and
analyse data generated by the Large Hadron Collider (LHC) at CERN.

\item[X.509] A cryptographic standard which defines how to implement service
security using electronic certificates, based on the use of a private
and public key combination. It is widely used on web servers accessed
using the https protocol and is the main authentication mechanism on the
WLCG infrastructure.

\item[x86\_64] 64-bit version of the x86 instruction set.

\item[XRootD] Software framework that is a fully generic suite for fast, low latency and scalable data access.

\end{description}

\end{appendices}

\sloppy
\raggedright
\clearpage
\printbibliography[title={References},heading=bibintoc]

\clearpage
\section*{The HEP Software Foundation}
\input{authors.tex}

\end{document}

### 2017-12-17 CWP/papers/roadmap/authors/a2tex.py
a2text: update/improve description comment

@@ -1,10 +1,15 @@
 #! /usr/bin/env python
 #
 # Process CWP signatory list into a LaTeX author list
-
-# The data source is expected to be 
-#   Lastname, Firstname - Institution (1) [- Role ]
-# If there are multiple affiliations, separate with &
+#
+# The data source is expected to be made of 3 text files:
+#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey (i) [- Role ]
+#      with i a footnote number. If there are multiple affiliations, separate AffiliationKeys with &.
+#    - footnotes:txt: the list of footnotes. There must be one footnote per line starting with 'i.'
+#    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
+#      AffiliationKey must not contain any space.
+#
+# These 3 files are typically built from the Google Docs https://docs.google.com/document/d/1tBXwlNnQsxxZA3gVS1_KSpa8wRXGyk250EIsJwJ2T34/edit#
 
 import os
 import re

### 2017-12-17 CWP/papers/roadmap/authors/a2tex.py
a2tex: add a parser to pass input/output files

@@ -15,6 +15,19 @@ import os
 import re
 import sys
 from operator import attrgetter
+import argparse
+
+# Define some constants related to application exit status
+EXIT_STATUS_SUCCESS = 0
+EXIT_STATUS_OPTION_ERROR = 3
+EXIT_STATUS_FAILURE = 5
+
+# Default input files
+AUTHOR_FILE_DEFAULT = "authors.txt"
+FOOTNOTE_FILE_DEFAULT = "footnotes.txt"
+AFFILIATION_ADDRESS_FILE_DEFAULT = "address.txt"
+LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
+
 
 class Footnote():
     def __init__(self, number_str, text):
@@ -25,12 +38,25 @@ class Footnote():
         return repr((self.mark, self.text))
 
 def main():
-    author_file = "authors.txt"
-    footnote_file = "footnotes.txt"
-    affiliation_address_file = "address.txt"
-    
+
+    try:
+        parser = argparse.ArgumentParser()
+        parser.add_argument("--authors", dest="authors", default=AUTHOR_FILE_DEFAULT,
+                            help="File containining author list (D: {})".format(AUTHOR_FILE_DEFAULT))
+        parser.add_argument("--footnotes", dest="footnotes", default=FOOTNOTE_FILE_DEFAULT,
+                            help="File containining footnote list (D: {})".format(FOOTNOTE_FILE_DEFAULT))
+        parser.add_argument("--affiliations", dest="affiliations", default=AFFILIATION_ADDRESS_FILE_DEFAULT,
+                            help="File containining affiliation list (D: {})".format(AFFILIATION_ADDRESS_FILE_DEFAULT))
+        parser.add_argument("--output", dest="output_file", default=LATEX_AUTHOR_FILE_DEFAULT,
+                            help="Output Latex file (D: {})".format(LATEX_AUTHOR_FILE_DEFAULT))
+        options = parser.parse_args()
+    except Exception as e:
+        parser.invalid_option_value('Parsing error: {}'.format(e.msg))
+        return EXIT_STATUS_OPTION_ERROR
+
+
     # Open and process the authors file
-    with open(author_file, encoding='utf-8') as author_fh:
+    with open(options.authors, encoding='utf-8') as author_fh:
         author_list = []
         for author_line in author_fh:
             author_line = author_line.strip()
@@ -64,7 +90,7 @@ def main():
     
     # Generate the map to people's institutes
[…]

### 2017-12-17 CWP/papers/roadmap/authors/a2tex.py
a2tex: use an Author class instead of a dict

@@ -37,6 +37,26 @@ class Footnote():
     def __repr__(self):
         return repr((self.mark, self.text))
 
+class Author():
+    def __init__(self, author_line):
+        footnotes_search = re.search(r"\((?P<footnotes>[\d,]+)\)$", author_line)
+        if footnotes_search:
+            self.footnotes = footnotes_search.group('footnotes').split(",")
+            author_line = author_line.replace(footnotes_search.group(), "")
+        else:
+            self.footnotes = []
+        author_elements = [el.strip() for el in author_line.split(" - ")]
+        try:
+            self.surname, self.forename = [el.strip() for el in author_elements[0].split(",", 1)]
+            self.affiliations = [el.strip() for el in author_elements[1].split("&")]
+            if len(author_elements) == 3:
+                self.role = author_elements[2]
+            else:
+                self.role = ""
+        except Exception as e:
+            raise Exeption("Got an exception({}) processing line: {}".format(e, author_line))
+
+
 def main():
 
     try:
@@ -62,32 +82,8 @@ def main():
             author_line = author_line.strip()
             if author_line.startswith("#") or author_line == "":
                 continue
-            # Find any footnotes that the author wanted to add
-            footnotes_search = re.search(r"\(([\d,]+)\)$", author_line)
-            if footnotes_search:
-                footnotes = footnotes_search.group(1).split(",")
-                author_line = author_line.replace(footnotes_search.group(), "")
-            else:
-                footnotes = []
-            author_elements = [ el.strip() for el in author_line.split(" - ") ]
-            try:
-                surname, forename = [ el.strip() for el in author_elements[0].split(",", 1) ]
-                affiliation = author_elements[1]
-                if len(author_elements) == 3:
-                    role = author_elements[2]
-                else:
-                    role = ""
-                author_list.append( {"surname": surname,
-                     "forename": forename,
-                     "affiliation": [ el.strip() for el in affiliation.split("&") ],
-                     "role": role.replace("&", "\&"),
-                     "footnotes": footnotes,}
-                    )
-            except Exception as e:
-                print ("Got an exception({}) processing this line: {}".format(e, author_line))
-                raise
-    #print author_list
-    
+            author_list.append(Author(author_line))
+
     # Generate the map to people's institutes
     institute_address = {}
     with open(options.affiliations, encoding='utf-8') as address_fh:
@@ -97,18 +93,15 @@ def main():
                 continue
             institute, address = affiliation_line.split(": ", 1)
[…]

### 2017-12-17 CWP/papers/roadmap/authors/a2tex.py
a2tex refactoring: use a class for affiliations

@@ -57,6 +57,14 @@ class Author():
             raise Exeption("Got an exception({}) processing line: {}".format(e, author_line))
 
 
+class Affiliation():
+    def __init__(self,address):
+        self.address = address
+        self.mark = ''
+
+    def set_mark(self, i):
+        self.mark = str(i)
+
 def main():
 
     try:
@@ -98,14 +106,15 @@ def main():
     for author in author_list:
         for affiliation in author.affiliations:
             if affiliation not in affiliation_map:
-                affiliation_map[affiliation] = {
-                    "address": (institute_address[affiliation] if affiliation in institute_address else affiliation),
-                }
+                if affiliation in institute_address:
+                    affiliation_map[affiliation] = Affiliation(institute_address[affiliation])
+                else:
+                    raise Exception('Affiliation {} not defined'.format(affiliation))
 
     # Alphabetise and assign footnote marks
     sorted_affiliations = sorted(affiliation_map)
     for footnote_mark, affiliation in enumerate(sorted_affiliations, start=1):
-        affiliation_map[affiliation]["footnotemark"] = str(footnote_mark)
+        affiliation_map[affiliation].set_mark(footnote_mark)
     
     # Parse and assign footnotes
     with open(options.footnotes, encoding='utf-8') as footnote_fh:
@@ -118,13 +127,12 @@ def main():
             number = number.rstrip(".")
             footnote_list.append(Footnote(number, note))
     sorted_footnotes = sorted(footnote_list, key=lambda footnote: footnote.mark)
-    #print footnote_map
-        
+
         
     # Write out the latex author file
     with open(options.output_file, "w", encoding="utf-8") as output:
         for author in author_list:
-            affiliation_list = ",".join([ str(affiliation_map[affiliation]["footnotemark"]) for affiliation in author.affiliations ])
+            affiliation_list = ",".join([ affiliation_map[affiliation].mark for affiliation in author.affiliations ])
             if author.footnotes:
                 footnote_str = ",".join( [ footnote_list[int(id)-1].mark for id in author.footnotes ])
                 affiliation_list = affiliation_list + "," + footnote_str
@@ -137,13 +145,13 @@ def main():
                                                affiliation_list,
                                                eol_str),
                    file=output)
-            #if author["role"]:
-            #    print >>output, "(" + author["role"] + ")"
+
         print ("\\bigskip", file=output)
         for affiliation in sorted_affiliations:
[…]

### 2017-12-17 CWP/papers/roadmap/authors/a2tex.py
a2tex: print a sorted list of author

@@ -131,7 +131,7 @@ def main():
         
     # Write out the latex author file
     with open(options.output_file, "w", encoding="utf-8") as output:
-        for author in author_list:
+        for author in sorted(author_list, key=lambda author: author.surname):
             affiliation_list = ",".join([ affiliation_map[affiliation].mark for affiliation in author.affiliations ])
             if author.footnotes:
                 footnote_str = ",".join( [ footnote_list[int(id)-1].mark for id in author.footnotes ])

### 2017-12-17 CWP/papers/roadmap/authors/a2tex.py
a2tex: escape Latex special characters in affiliations and footnotes

@@ -65,6 +65,27 @@ class Affiliation():
     def set_mark(self, i):
         self.mark = str(i)
 
+def latex_escape(string):
+    """
+    Function to escape latex reserved characters
+    :param string: the string containing the characters to escape
+    :return: the escaped string
+    """
+
+    CHARS_TO_ESCAPE_PATTERN = re.compile(r'(?<!\\)(?P<char>[&%\$#_{}])')
+    SPECIAL_LATEX_CHARS = {'~':'$\\\\textasciitilde$',
+                           '^':'$\\\\textasciicircum$'}
+    # Do not attempt to escape \ as it may have been added as an escape character
+    # and is unlikely to be present in an affiliation or footnote...
+    # If it is present, if will have to be escaped manually
+    #                       '\\':'$\\\\textbackslash$'}
+    # Escape all reserved characters that can be escaped
+    string = CHARS_TO_ESCAPE_PATTERN.sub('\\\\\g<char>', string)
+    # Then replace special chars
+    for special_char, repl in SPECIAL_LATEX_CHARS.items():
+       string = re.sub(re.escape(special_char), repl, string)
+    return string
+
 def main():
 
     try:
@@ -149,13 +170,13 @@ def main():
         print ("\\bigskip", file=output)
         for affiliation in sorted_affiliations:
             print ("\\par {{\\footnotesize $^{{{}}}$ {}}}".format(str(affiliation_map[affiliation].mark),
-                                                               affiliation_map[affiliation].address),
+                                                                  latex_escape(affiliation_map[affiliation].address)),
                    file=output)
 
         print ("\\bigskip", file=output)
         for footnote in sorted_footnotes:
             print ("\\par {{\\footnotesize $^{{{}}}$ {}}}".format(footnote.mark,
-                                                                 footnote.text),
+                                                                  latex_escape(footnote.text)),
                    file=output)
 
 

### 2017-12-18 CWP/papers/roadmap/authors/a2tex.py
a2tex: require Python >= 3.5

@@ -3,17 +3,29 @@
 # Process CWP signatory list into a LaTeX author list
 #
 # The data source is expected to be made of 3 text files:
-#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey (i) [- Role ]
+#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey [- Role ] (i)
 #      with i a footnote number. If there are multiple affiliations, separate AffiliationKeys with &.
 #    - footnotes:txt: the list of footnotes. There must be one footnote per line starting with 'i.'
 #    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
 #      AffiliationKey must not contain any space.
 #
-# These 3 files are typically built from the Google Docs https://docs.google.com/document/d/1tBXwlNnQsxxZA3gVS1_KSpa8wRXGyk250EIsJwJ2T34/edit#
+# These 3 files are typically built from the Google Docs https://docs.google.com/document/d/1tBXwlNnQsxxZA3gVS1_KSpa8wRXGyk250EIsJwJ2T34
+#
+# Note: this script requires Python >= 3.5
+#
+# Initial version written by Graeme Stewart (graeme.andrew.stewart@cern.ch) and Michel Jouvin (jouvin@lal.in2p3.fr)
+
+import sys
+PYTHON_MIN = (3, 5)
+major, minor, _, _, _ = sys.version_info
+if (major, minor) < PYTHON_MIN:
+    python_min_str = '.'.join(str(x) for x in PYTHON_MIN)
+    # Keep old formatter syntax to be sure it works with old version of Python
+    print ("This script requires Python %s or later" % python_min_str)
+    sys.exit(2)
 
 import os
 import re
-import sys
 from operator import attrgetter
 import argparse
 

### 2017-12-18 CWP/papers/roadmap/authors/a2tex.py
Script to generate author list from Google Docs (#22)
* First version of author processing script

TODO: personal footnotes, institute addresses

* Support multiple affiliations, separaeted by &

Some fixes to badly formatted author entries (also fixed in gdoc)

* Add "footnote" functionality

Start listing institute addresses as well

* Signers update from Google Docs at 7 pm

* a2text: fixes for Python v3 compliance

* a2text: update/improve description comment

* a2tex: add a parser to pass input/output files

* a2tex: use an Author class instead of a dict

* a2tex refactoring: use a class

@@ -1,0 +1,197 @@
+#! /usr/bin/env python
+#
+# Process CWP signatory list into a LaTeX author list
+#
+# The data source is expected to be made of 3 text files:
+#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey [- Role ] (i)
+#      with i a footnote number. If there are multiple affiliations, separate AffiliationKeys with &.
+#    - footnotes:txt: the list of footnotes. There must be one footnote per line starting with 'i.'
+#    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
+#      AffiliationKey must not contain any space.
+#
+# These 3 files are typically built from the Google Docs https://docs.google.com/document/d/1tBXwlNnQsxxZA3gVS1_KSpa8wRXGyk250EIsJwJ2T34
+#
+# Note: this script requires Python >= 3.5
+#
+# Initial version written by Graeme Stewart (graeme.andrew.stewart@cern.ch) and Michel Jouvin (jouvin@lal.in2p3.fr)
+
+import sys
+PYTHON_MIN = (3, 5)
+major, minor, _, _, _ = sys.version_info
+if (major, minor) < PYTHON_MIN:
+    python_min_str = '.'.join(str(x) for x in PYTHON_MIN)
+    # Keep old formatter syntax to be sure it works with old version of Python
+    print ("This script requires Python %s or later" % python_min_str)
+    sys.exit(2)
+
+import os
+import re
+from operator import attrgetter
+import argparse
+
+# Define some constants related to application exit status
+EXIT_STATUS_SUCCESS = 0
+EXIT_STATUS_OPTION_ERROR = 3
+EXIT_STATUS_FAILURE = 5
+
+# Default input files
+AUTHOR_FILE_DEFAULT = "authors.txt"
+FOOTNOTE_FILE_DEFAULT = "footnotes.txt"
+AFFILIATION_ADDRESS_FILE_DEFAULT = "address.txt"
+LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
+
+
+class Footnote():
+    def __init__(self, number_str, text):
+        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
+        self.mark = footnote_symbols[int(number_str)-1]
+        self.text = text
+    def __repr__(self):
+        return repr((self.mark, self.text))
+
+class Author():
+    def __init__(self, author_line):
+        footnotes_search = re.search(r"\((?P<footnotes>[\d,]+)\)$", author_line)
[…]

### 2017-12-18 CWP/papers/roadmap/authors/a2tex.py
Minor changes to the processing script
- Don't add a space between the surname and the institute/footnote marks
- Also generate the author list in the form needed for arXiv author metadata
- Add a .gitignore to prevent adding generated files in error

@@ -39,6 +39,7 @@ AUTHOR_FILE_DEFAULT = "authors.txt"
 FOOTNOTE_FILE_DEFAULT = "footnotes.txt"
 AFFILIATION_ADDRESS_FILE_DEFAULT = "address.txt"
 LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
+ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
 
 
 class Footnote():
@@ -110,6 +111,8 @@ def main():
                             help="File containining affiliation list (D: {})".format(AFFILIATION_ADDRESS_FILE_DEFAULT))
         parser.add_argument("--output", dest="output_file", default=LATEX_AUTHOR_FILE_DEFAULT,
                             help="Output Latex file (D: {})".format(LATEX_AUTHOR_FILE_DEFAULT))
+        parser.add_argument("--arxivoutput", dest="arxiv_output_file", default=ARXIV_AUTHOR_FILE_DEFAULT,
+                            help="Output arXiv author list (D: {})".format(ARXIV_AUTHOR_FILE_DEFAULT))
         options = parser.parse_args()
     except Exception as e:
         parser.invalid_option_value('Parsing error: {}'.format(e.msg))
@@ -161,6 +164,11 @@ def main():
             footnote_list.append(Footnote(number, note))
     sorted_footnotes = sorted(footnote_list, key=lambda footnote: footnote.mark)
 
+    # Write out the author list to copy into the arXiv author file
+    with open(options.arxiv_output_file, "w", encoding="utf-8") as arxivoutput:
+        print("HEP Software Foundation: ", ", ".join([ "{} {}".format(author.forename, author.surname) 
+                                                      for author in  sorted(author_list, key=lambda author: author.surname) ]),
+              file=arxivoutput)
         
     # Write out the latex author file
     with open(options.output_file, "w", encoding="utf-8") as output:
@@ -173,7 +181,7 @@ def main():
                 eol_str = ""
             else:
                 eol_str = ";"
-            print ("{}, {} $^{{{}}}${}".format(author.surname,
+            print ("{}, {}$^{{{}}}${}".format(author.surname,
                                                author.forename,
                                                affiliation_list,
                                                eol_str),

### 2017-12-18 CWP/papers/roadmap/latex/authors.tex
Updated author list for 2017-12-18
- Generated from Google Doc on 2107-12-18T1300
- Add also a saved copy of the arXiv author metadata
Fix encoding problem with text files not being in utf-8
Fix white space issues and remove duplicate Oslo entry
Update key for SaoPaulo (minor)
Resulting authors.tex file copied to LaTeX area for roadmap

@@
-Alves Jr, Antonio Augusto $^{74}$;
-Amadio, Guilherme $^{5}$;
-Anh-Ky, Nguyen $^{111}$;
-Aphecetche, Laurent $^{60}$;
-Apostolakis, John $^{5}$;
-Asai, Makoto $^{58,p}$;
-Atzori, Luca $^{5}$;
-Babik, Marian $^{5}$;
-Bagliesi, Giuseppe $^{29}$;
-Bandieramonte, Marilena $^{5}$;
-Barisits, Martin $^{5}$;
-Bauerdick, Lothar A. T. $^{16,c}$;
-Belforte, Stefano $^{32}$;
-Benjamin, Douglas $^{75}$;
-Bernius, Catrin $^{58}$;
-Bhimji, Wahid $^{42}$;
-Bianchi, Riccardo Maria $^{97}$;
-Bird, Ian $^{5}$;
-Biscarat, Catherine $^{48}$;
-Blomer, Jakob $^{5}$;
-Bloom, Kenneth $^{89}$;
-Boccali, Tommaso $^{29}$;
-Bockelman, Brian $^{89}$;
-Bold, Tomasz $^{39}$;
-Bonacorsi, Daniele $^{24}$;
-Boveia, Antonio $^{94}$;
-Bozzi, Concezio $^{26}$;
-Bracko, Marko $^{86,37}$;
-Britton, David $^{79}$;
-Buckley, Andy $^{79}$;
-Buncic, Predrag $^{5,a}$;
-Calafiura, Paolo $^{42}$;
-Campana, Simone $^{5,a}$;
-Canal, Philippe $^{16,c}$;
-Canali, Luca $^{5}$;
-Carlino, Gianpaolo $^{28}$;
-Castro, Nuno $^{43,88,d}$;
-Cattaneo, Marco $^{5}$;
-Cerminara, Gianluca $^{5}$;
-Chang, Philip $^{69}$;
-Chapman, John $^{70}$;
-Childers, Taylor $^{1}$;
-Clarke, Peter $^{76}$;
-Clemencic, Marco $^{5}$;
-Cogneras, Eric $^{46}$;
-Collier, Ian $^{56}$;
-Corti, Gloria $^{5}$;
-Cosmo, Gabriele $^{5}$;
-Costanzo, Davide $^{103}$;
-Couturier, Ben $^{5}$;
-Cranmer, Kyle $^{53}$;
-Cranshaw, Jack $^{1}$;
-Crépé-Renaudin, Sabine $^{48}$;
-Cristella, Leonardo $^{25}$;
-Crooks, David $^{79}$;
-Dallmeier-Tiessen, Sünje $^{5}$;
-De Cian, Michel $^{80}$;
-De, Kaushik $^{105}$;
-Di Girolamo, Alessandro $^{5}$;
-Dimitrov, Gancho $^{5}$;
-Doglioni, Caterina $^{84,h}$;
-Dotti, Andrea $^{58,p}$;
-Duellmann, Dirk $^{5}$;
-Duflot, Laurent $^{41}$;
-Dykstra, Dave $^{16,c}$;
-Dziedziniewicz-Wojcik, Katarzyna $^{5}$;
-Dziurda, Agnieszka $^{5}$;
-Egede, Ulrik $^{34}$;
-Elmer, Peter $^{98,a}$;
-Elmsheuser, Johannes $^{2}$;
-Elvira, V. Daniel $^{16,c}$;
-Eulisse, Giulio $^{5}$;
-Ferber, Torben $^{67}$;
-Filipcic, Andrej $^{37}$;
-Fisk, Ian $^{59}$;
-Fitzpatrick, Conor $^{14}$;
-Flix, José $^{55,7,g}$;
-Formica, Andrea $^{35}$;
-Forti, Alessandra $^{85}$;
-Gaede, Frank $^{13}$;
-Ganis, Gerardo $^{5}$;
-Gardner, Robert $^{73}$;
-Garonne, Vincent $^{95}$;
-Gellrich, Andreas $^{13}$;
-Genser, Krzysztof $^{16,c}$;
-George, Simon $^{57}$;
-Geurts, Frank $^{99}$;
-Gheata, Andrei $^{5}$;
-Gheata, Mihaela $^{5}$;
-Giacomini, Francesco $^{9}$;
-Giagu, Stefano $^{101,31}$;
-Giffels, Manuel $^{38}$;
-Gingrich, Douglas $^{63}$;
-Girone, Maria $^{5}$;
-Gligorov, Vladimir V. $^{47}$;
-Glushkov, Ivan $^{105}$;
-Gohn, Wesley $^{81}$;
-González Caballero, Isidro $^{96}$;
-González Fernández, Juan R. $^{96}$;
-Gonzalez Lopez, Jose Benito $^{5}$;
-Govi, Giacomo $^{16}$;
-Grandi, Claudio $^{24}$;
-Grasland, Hadrien $^{41}$;
-Gray, Heather $^{42}$;
-Grillo, Lucia $^{85}$;
-Guan, Wen $^{109}$;
-Gutsche, Oliver $^{16,c}$;
-Gyurjyan, Vardan $^{36}$;
-Hanushevsky, Andrew $^{58,p}$;
-Hariri, Farah $^{5}$;
-Hartmann, Thomas $^{13}$;
-Harvey, John $^{5,a}$;
-Hauth, Thomas $^{38}$;
-Hegner, Benedikt $^{5,a}$;
-Heinemann, Beate $^{13}$;
-Heinrich, Lukas $^{53}$;
-Hernández, José M. $^{7,g}$;
-Hildreth, Michael $^{91,f}$;
-Hodgkinson, Mark $^{103}$;
-Hoeche, Stefan $^{58,p}$;
-Hristov, Peter $^{5}$;
-Huang, Xingtao $^{102}$;
-Ivanchenko, Vladimir N. $^{5,106}$;
-Ivanov, Todor $^{104}$;
-Jashal, Brij $^{62}$;
-Jayatilaka, Bodhitha $^{16,c}$;
-Jones, Roger $^{82,a}$;
-Jouvin, Michel $^{41,a}$;
-Jun, Soon Yung $^{16,c}$;
-Kagan, Michael $^{58,p}$;
-Kalderon, Charles William $^{84}$;
-Karavakis, Edward $^{5}$;
-Katz, Daniel S. $^{72}$;
-Kcira, Dorian $^{11}$;
-Kersevan, Borut Paul $^{83}$;
-Kirby, Michael $^{16,c}$;
-Klimentov, Alexei $^{2}$;
-Klute, Markus $^{49}$;
-Komarov, Ilya $^{32,n}$;
-Koppenburg, Patrick $^{52}$;
-Kowalkowski, Jim $^{16,c}$;
-Kreczko, Luke $^{66}$;
-Kuhr, Thomas $^{45,a}$;
-Kutschke, Robert $^{16,a,c}$;
-Kuznetsov, Valentin $^{12}$;
-Lampl, Walter $^{65}$;
-Lancon, Eric $^{2}$;
-Lange, David $^{98}$;
-Lassnig, Mario $^{5}$;
-Laycock, Paul $^{5}$;
-Leggett, Charles $^{42}$;
-Letts, James $^{69}$;
-Lewendel, Birgit $^{13}$;
-Li, Teng $^{76}$;
-Lima, Guilherme $^{16}$;
-Linacre, Jacob $^{56,m}$;
-Linden, Tomas $^{18}$;
-Lo Presti, Giuseppe $^{5}$;
-Lopienski, Sebastian $^{5}$;
-Love, Peter $^{82}$;
-Marshall, Zachary L $^{42}$;
-Martelli, Edoardo $^{5}$;
-Martin-Haugh, Stewart $^{56}$;
-Mato, Pere $^{5}$;
-Mazumdar, Kajari $^{62}$;
-McCauley, Thomas $^{91}$;
-McFayden, Josh $^{5}$;
-McKee, Shawn $^{87,l}$;
-McNab, Andrew $^{85}$;
-Meinhard, Helge $^{5}$;
-Menasce, Dario $^{27,a}$;
-Mendez Lorenzo, Patricia $^{5}$;
-Mete, Alaettin Serhan $^{68}$;
-Michelotto, Michele $^{30}$;
-Mitrevski, Jovan $^{45}$;
-Moneta, Lorenzo $^{5}$;
-Morgan, Ben $^{108}$;
-Mount, Richard $^{58,p}$;
-Moyse, Edward $^{64}$;
-Murray, Sean $^{71,10}$;
-Neubauer, Mark S $^{72,a,k}$;
-Novaes, Sérgio $^{100}$;
-Novak, Mihaly $^{5}$;
-Oyanguren, Arantza $^{21}$;
-Ozturk, Nurcan $^{105}$;
-Pacheco Pages, Andres $^{55,19,j}$;
-Paganini, Michela $^{110}$;
-Pansanel, Jerome $^{33}$;
-Pascuzzi, Vincent R. $^{107}$;
-Pearce, Alex $^{5}$;
-Pearson, Ben $^{50}$;
-Pedro, Kevin $^{16,c}$;
-Perdue, Gabriel $^{16}$;
-Perez-Calero Yzquierdo, Antonio $^{55,7,g}$;
-Perrozzi, Luca $^{15}$;
-Petersen, Troels $^{51}$;
-Petric, Marko $^{5}$;
-Piedra, Jónatan $^{20}$;
-Piilonen, Leo $^{112,i}$;
-Piparo, Danilo $^{5}$;
-Pokorski, Witold $^{5}$;
-Polci, Francesco $^{47}$;
-Potamianos, Karolos $^{13}$;
-Psihas, Fernanda $^{23}$;
-Raven, Gerhard $^{52}$;
-Reuter, Jürgen $^{13}$;
-Ribon, Alberto $^{5}$;
-Ritter, Martin $^{45}$;
-Robinson, James $^{13}$;
-Rodrigues, Eduardo $^{74,a,e}$;
-Roiser, Stefan $^{5,a}$;
-Rousseau, David $^{41}$;
-Roy, Gareth $^{79}$;
-Sailer, Andre $^{5}$;
-Sakuma, Tai $^{66}$;
-Sánchez-Hernández, Alberto $^{8,o}$;
-Santana, Renato $^{3}$;
-Sartirana, Andrea $^{44}$;
-Schellman, Heidi $^{54}$;
-Schovancová, Jaroslava $^{5}$;
-Schramm, Steven $^{78}$;
-Schulz, Markus $^{5}$;
-Sciabà, Andrea $^{5}$;
-Seidel, Sally $^{90}$;
-Sekmen, Sezen $^{40}$;
-Serfon, Cedric $^{93}$;
-Severini, Horst $^{92}$;
-Sexton-Kennedy, Elizabeth $^{16,a,c}$;
-Seymour, Michael $^{85}$;
-Shapoval, Illya $^{42}$;
-Shiers, Jamie $^{5}$;
-Shiu, Jing-Ge $^{61}$;
-Short, Hannah $^{5}$;
-Siroli, Gian Piero $^{24}$;
-Skipsey, Sam $^{79}$;
-Smith, Tim $^{5}$;
-Snyder, Scott $^{2}$;
-Sokoloff, Michael D $^{74,a}$;
-Stadie, Hartmut $^{17}$;
-Stark, Giordon $^{6}$;
-Stewart, Gordon $^{79}$;
-Stewart, Graeme $^{5,a}$;
-Templon, Jeff $^{52}$;
-Tenaglia, Giacomo $^{5}$;
-Tsulaia, Vakhtang $^{42}$;
-Tunnell, Christopher $^{6}$;
-Vaandering, Eric $^{16,c}$;
-Valassi, Andrea $^{5}$;
-Vallecorsa, Sofia $^{77}$;
-Valsan, Liviu $^{5}$;
-Van Gemmeren, Peter $^{1}$;
-Vázquez Sierra, Carlos $^{52}$;
-Vernet, Renaud $^{4}$;
-Viren, Brett $^{2}$;
-Vlimant, Jean-Roch $^{11,a}$;
-Voss, Christian $^{13}$;
-Vuosalo, Carl $^{113}$;
-Wartel, Romain $^{5}$;
-Wenaus, Torre $^{2}$;
-Wenzel, Sandro $^{5}$;
-Winklmeier, Frank $^{54}$;
-Wissing, Christoph $^{13}$;
-Wuerthwein, Frank $^{69}$;
-Wynne, Benjamin $^{76}$;
-Xiaomei, Zhang $^{22}$;
-Yang, Wei $^{58,p}$;
-Yazgan, Efe $^{22}$
+Alves Jr, Antonio Augusto$^{74}$;
+Amadio, Guilherme$^{5}$;
+Anh-Ky, Nguyen$^{110}$;
+Aphecetche, Laurent$^{60}$;
+Apostolakis, John$^{5}$;
+Asai, Makoto$^{58,p}$;
+Atzori, Luca$^{5}$;
+Babik, Marian$^{5}$;
+Bagliesi, Giuseppe$^{29}$;
+Bandieramonte, Marilena$^{5}$;
+Barisits, Martin$^{5}$;
+Bauerdick, Lothar A. T.$^{16,c}$;
+Belforte, Stefano$^{32}$;
+Benjamin, Douglas$^{75}$;
+Bernius, Catrin$^{58}$;
+Bhimji, Wahid$^{42}$;
+Bianchi, Riccardo Maria$^{96}$;
+Bird, Ian$^{5}$;
+Biscarat, Catherine$^{48}$;
+Blomer, Jakob$^{5}$;
+Bloom, Kenneth$^{89}$;
+Boccali, Tommaso$^{29}$;
+Bockelman, Brian$^{89}$;
+Bold, Tomasz$^{39}$;
+Bonacorsi, Daniele$^{24}$;
+Boveia, Antonio$^{93}$;
+Bozzi, Concezio$^{26}$;
+Bracko, Marko$^{86,37}$;
+Britton, David$^{79}$;
+Buckley, Andy$^{79}$;
+Buncic, Predrag$^{5,a}$;
+Calafiura, Paolo$^{42}$;
+Campana, Simone$^{5,a}$;
+Canal, Philippe$^{16,c}$;
+Canali, Luca$^{5}$;
+Carlino, Gianpaolo$^{28}$;
+Castro, Nuno$^{43,88,d}$;
+Cattaneo, Marco$^{5}$;
+Cerminara, Gianluca$^{5}$;
+Chang, Philip$^{69}$;
+Chapman, John$^{70}$;
+Chen, Gang$^{22}$;
+Childers, Taylor$^{1}$;
+Clarke, Peter$^{76}$;
+Clemencic, Marco$^{5}$;
+Cogneras, Eric$^{46}$;
+Collier, Ian$^{56}$;
+Corti, Gloria$^{5}$;
+Cosmo, Gabriele$^{5}$;
+Costanzo, Davide$^{102}$;
+Couturier, Ben$^{5}$;
+Cranmer, Kyle$^{53}$;
+Cranshaw, Jack$^{1}$;
+Cristella, Leonardo$^{25}$;
+Crooks, David$^{79}$;
+Crépé-Renaudin, Sabine$^{48}$;
+Dallmeier-Tiessen, Sünje$^{5}$;
+De, Kaushik$^{104}$;
+De Cian, Michel$^{80}$;
+Di Girolamo, Alessandro$^{5}$;
+Dimitrov, Gancho$^{5}$;
+Doglioni, Caterina$^{84,h}$;
+Dotti, Andrea$^{58,p}$;
+Duellmann, Dirk$^{5}$;
+Duflot, Laurent$^{41}$;
+Dykstra, Dave$^{16,c}$;
+Dziedziniewicz-Wojcik, Katarzyna$^{5}$;
+Dziurda, Agnieszka$^{5}$;
+Egede, Ulrik$^{34}$;
+Elmer, Peter$^{97,a}$;
+Elmsheuser, Johannes$^{2}$;
+Elvira, V. Daniel$^{16,c}$;
+Eulisse, Giulio$^{5}$;
+Ferber, Torben$^{67}$;
+Filipcic, Andrej$^{37}$;
+Fisk, Ian$^{59}$;
+Fitzpatrick, Conor$^{14}$;
+Flix, José$^{55,7,g}$;
+Formica, Andrea$^{35}$;
+Forti, Alessandra$^{85}$;
+Gaede, Frank$^{13}$;
+Ganis, Gerardo$^{5}$;
+Gardner, Robert$^{73}$;
+Garonne, Vincent$^{94}$;
+Gellrich, Andreas$^{13}$;
+Genser, Krzysztof$^{16,c}$;
+George, Simon$^{57}$;
+Geurts, Frank$^{98}$;
+Gheata, Andrei$^{5}$;
+Gheata, Mihaela$^{5}$;
+Giacomini, Francesco$^{9}$;
+Giagu, Stefano$^{100,31}$;
+Giffels, Manuel$^{38}$;
+Gingrich, Douglas$^{63}$;
+Girone, Maria$^{5}$;
+Gligorov, Vladimir V.$^{47}$;
+Glushkov, Ivan$^{104}$;
+Gohn, Wesley$^{81}$;
+Gonzalez Lopez, Jose Benito$^{5}$;
+González Caballero, Isidro$^{95}$;
+González Fernández, Juan R.$^{95}$;
+Govi, Giacomo$^{16}$;
+Grandi, Claudio$^{24}$;
+Grasland, Hadrien$^{41}$;
+Gray, Heather$^{42}$;
+Grillo, Lucia$^{85}$;
+Guan, Wen$^{108}$;
+Gutsche, Oliver$^{16,c}$;
+Gyurjyan, Vardan$^{36}$;
+Hanushevsky, Andrew$^{58,p}$;
+Hariri, Farah$^{5}$;
+Hartmann, Thomas$^{13}$;
+Harvey, John$^{5,a}$;
+Hauth, Thomas$^{38}$;
+Hegner, Benedikt$^{5,a}$;
+Heinemann, Beate$^{13}$;
+Heinrich, Lukas$^{53}$;
+Hernández, José M.$^{7,g}$;
+Hildreth, Michael$^{91,f}$;
+Hodgkinson, Mark$^{102}$;
+Hoeche, Stefan$^{58,p}$;
+Hristov, Peter$^{5}$;
+Huang, Xingtao$^{101}$;
+Ivanchenko, Vladimir N.$^{5,105}$;
+Ivanov, Todor$^{103}$;
+Jashal, Brij$^{62}$;
+Jayatilaka, Bodhitha$^{16,c}$;
+Jones, Roger$^{82,a}$;
+Jouvin, Michel$^{41,a}$;
+Jun, Soon Yung$^{16,c}$;
+Kagan, Michael$^{58,p}$;
+Kalderon, Charles William$^{84}$;
+Karavakis, Edward$^{5}$;
+Katz, Daniel S.$^{72}$;
+Kcira, Dorian$^{11}$;
+Kersevan, Borut Paul$^{83}$;
+Kirby, Michael$^{16,c}$;
+Klimentov, Alexei$^{2}$;
+Klute, Markus$^{49}$;
+Komarov, Ilya$^{32,n}$;
+Koppenburg, Patrick$^{52}$;
+Kowalkowski, Jim$^{16,c}$;
+Kreczko, Luke$^{66}$;
+Kuhr, Thomas$^{45,a}$;
+Kutschke, Robert$^{16,a,c}$;
+Kuznetsov, Valentin$^{12}$;
+Lampl, Walter$^{65}$;
+Lancon, Eric$^{2}$;
+Lange, David$^{97}$;
+Lassnig, Mario$^{5}$;
+Laycock, Paul$^{5}$;
+Leggett, Charles$^{42}$;
+Letts, James$^{69}$;
+Lewendel, Birgit$^{13}$;
+Li, Teng$^{76}$;
+Lima, Guilherme$^{16}$;
+Linacre, Jacob$^{56,m}$;
+Linden, Tomas$^{18}$;
+Lo Presti, Giuseppe$^{5}$;
+Lopienski, Sebastian$^{5}$;
+Love, Peter$^{82}$;
+Marshall, Zachary L$^{42}$;
+Martelli, Edoardo$^{5}$;
+Martin-Haugh, Stewart$^{56}$;
+Mato, Pere$^{5}$;
+Mazumdar, Kajari$^{62}$;
+McCauley, Thomas$^{91}$;
+McFayden, Josh$^{5}$;
+McKee, Shawn$^{87,l}$;
+McNab, Andrew$^{85}$;
+Meinhard, Helge$^{5}$;
+Menasce, Dario$^{27,a}$;
+Mendez Lorenzo, Patricia$^{5}$;
+Mete, Alaettin Serhan$^{68}$;
+Michelotto, Michele$^{30}$;
+Mitrevski, Jovan$^{45}$;
+Moneta, Lorenzo$^{5}$;
+Morgan, Ben$^{107}$;
+Mount, Richard$^{58,p}$;
+Moyse, Edward$^{64}$;
+Murray, Sean$^{71,10}$;
+Neubauer, Mark S$^{72,a,k}$;
+Novaes, Sérgio$^{99}$;
+Novak, Mihaly$^{5}$;
+Oyanguren, Arantza$^{21}$;
+Ozturk, Nurcan$^{104}$;
+Pacheco Pages, Andres$^{55,19,j}$;
+Paganini, Michela$^{109}$;
+Pansanel, Jerome$^{33}$;
+Pascuzzi, Vincent R.$^{106}$;
+Pearce, Alex$^{5}$;
+Pearson, Ben$^{50}$;
+Pedro, Kevin$^{16,c}$;
+Perdue, Gabriel$^{16}$;
+Perez-Calero Yzquierdo, Antonio$^{55,7,g}$;
+Perrozzi, Luca$^{15}$;
+Petersen, Troels$^{51}$;
+Petric, Marko$^{5}$;
+Piedra, Jónatan$^{20}$;
+Piilonen, Leo$^{111,i}$;
+Piparo, Danilo$^{5}$;
+Pokorski, Witold$^{5}$;
+Polci, Francesco$^{47}$;
+Potamianos, Karolos$^{13}$;
+Psihas, Fernanda$^{23}$;
+Raven, Gerhard$^{52}$;
+Reuter, Jürgen$^{13}$;
+Ribon, Alberto$^{5}$;
+Ritter, Martin$^{45}$;
+Robinson, James$^{13}$;
+Rodrigues, Eduardo$^{74,a,e}$;
+Roiser, Stefan$^{5,a}$;
+Rousseau, David$^{41}$;
+Roy, Gareth$^{79}$;
+Sailer, Andre$^{5}$;
+Sakuma, Tai$^{66}$;
+Santana, Renato$^{3}$;
+Sartirana, Andrea$^{44}$;
+Schellman, Heidi$^{54}$;
+Schovancová, Jaroslava$^{5}$;
+Schramm, Steven$^{78}$;
+Schulz, Markus$^{5}$;
+Sciabà, Andrea$^{5}$;
+Seidel, Sally$^{90}$;
+Sekmen, Sezen$^{40}$;
+Serfon, Cedric$^{94}$;
+Severini, Horst$^{92}$;
+Sexton-Kennedy, Elizabeth$^{16,a,c}$;
+Seymour, Michael$^{85}$;
+Shapoval, Illya$^{42}$;
+Shiers, Jamie$^{5}$;
+Shiu, Jing-Ge$^{61}$;
+Short, Hannah$^{5}$;
+Siroli, Gian Piero$^{24}$;
+Skipsey, Sam$^{79}$;
+Smith, Tim$^{5}$;
+Snyder, Scott$^{2}$;
+Sokoloff, Michael D$^{74,a}$;
+Stadie, Hartmut$^{17}$;
+Stark, Giordon$^{6}$;
+Stewart, Gordon$^{79}$;
+Stewart, Graeme$^{5,a}$;
+Sánchez-Hernández, Alberto$^{8,o}$;
+Templon, Jeff$^{52}$;
+Tenaglia, Giacomo$^{5}$;
+Tsulaia, Vakhtang$^{42}$;
+Tunnell, Christopher$^{6}$;
+Vaandering, Eric$^{16,c}$;
+Valassi, Andrea$^{5}$;
+Vallecorsa, Sofia$^{77}$;
+Valsan, Liviu$^{5}$;
+Van Gemmeren, Peter$^{1}$;
+Vernet, Renaud$^{4}$;
+Viren, Brett$^{2}$;
+Vlimant, Jean-Roch$^{11,a}$;
+Voss, Christian$^{13}$;
+Vuosalo, Carl$^{108}$;
+Vázquez Sierra, Carlos$^{52}$;
+Wartel, Romain$^{5}$;
+Wenaus, Torre$^{2}$;
+Wenzel, Sandro$^{5}$;
+Winklmeier, Frank$^{54}$;
+Wissing, Christoph$^{13}$;
+Wuerthwein, Frank$^{69}$;
+Wynne, Benjamin$^{76}$;
+Xiaomei, Zhang$^{22}$;
+Yang, Wei$^{58,p}$;
+Yazgan, Efe$^{22}$
 \bigskip
@@
 \par {\footnotesize $^{2}$ Physics Department, Brookhaven National Laboratory, Upton NY, USA}
-\par {\footnotesize $^{3}$ Centro Brasileiro de Pesquisas Fisicas, Rio de Janeiro, Brazil}
+\par {\footnotesize $^{3}$ Centro Brasileiro de Pesquisas Físicas, Rio de Janeiro, Brazil}
 \par {\footnotesize $^{4}$ Centre de Calcul de l’IN2P3, Villeurbanne, Lyon, France}
@@
 \par {\footnotesize $^{13}$ Deutsches Elektronen-Synchrotron, Hamburg, Germany}
-\par {\footnotesize $^{14}$ Institute of Physics, Ecole Polytechnique Federale de Lausanne (EPFL), Lausanne, Switzerland}
+\par {\footnotesize $^{14}$ Institute of Physics, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland}
 \par {\footnotesize $^{15}$ ETH Zurich - Institute for Particle Physics and Astrophysics (IPA), Zurich, Switzerland}
@@
 \par {\footnotesize $^{20}$ Instituto de Física de Cantabria (IFCA), CSIC-Universidad de Cantabria, Santander, Spain}
-\par {\footnotesize $^{21}$ Instituto de Fisica Corpuscular, Centro Mixto Universidad de Valencia - CSIC, Valencia, Spain}
+\par {\footnotesize $^{21}$ Instituto de Física Corpuscular, Centro Mixto Universidad de Valencia - CSIC, Valencia, Spain}
 \par {\footnotesize $^{22}$ Institute of High Energy Physics, Chinese Academy of Sciences, Beijing}
@@
 \par {\footnotesize $^{25}$ INFN Sezione di Bari, Università di Bari, Politecnico di Bari, Bari, Italy}
-\par {\footnotesize $^{26}$ Universita e INFN, Ferrara, Ferrara, Italy}
+\par {\footnotesize $^{26}$ Università e INFN, Ferrara, Ferrara, Italy}
 \par {\footnotesize $^{27}$ INFN Sezione di Milano-Bicocca, Milano, Italy}
@@
 \par {\footnotesize $^{34}$ Imperial College London, London, United Kingdom}
-\par {\footnotesize $^{35}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat a l’Energie Atomique), Gif-sur-Yvette, France}
+\par {\footnotesize $^{35}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat à l’Énergie Atomique), Gif-sur-Yvette, France}
 \par {\footnotesize $^{36}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
-\par {\footnotesize $^{37}$ Jozef Stefan Institute, Ljubljana, Slovenia}
+\par {\footnotesize $^{37}$ Jožef Stefan Institute, Ljubljana, Slovenia}
 \par {\footnotesize $^{38}$ Karlsruhe Institute of Technology, Karlsruhe, Germany}
@@
 \par {\footnotesize $^{42}$ Lawrence Berkeley National Laboratory and University of California, Berkeley CA, USA}
-\par {\footnotesize $^{43}$ Laboratorio de Instrumentacao e Fisica Experimental de Particulas (LIP), Lisboa, Portugal}
-\par {\footnotesize $^{44}$ Laboratoire Leprince-Ringuet, Ecole polytechnique, CNRS/IN2P3, Université Paris-Saclay, Palaiseau, France}
+\par {\footnotesize $^{43}$ Laboratório de Instrumentação e Física Experimental de Partículas (LIP), Lisboa, Portugal}
+\par {\footnotesize $^{44}$ Laboratoire Leprince-Ringuet, École Polytechnique, CNRS/IN2P3, Université Paris-Saclay, Palaiseau, France}
 \par {\footnotesize $^{45}$ Fakultät für Physik, Ludwig-Maximilians-Universität München, München, Germany}
 \par {\footnotesize $^{46}$ Laboratoire de Physique Corpusculaire, Clermont Université and Université Blaise Pascal and CNRS/IN2P3, Clermont-Ferrand, France}
-\par {\footnotesize $^{47}$ LPNHE, Universite Pierre et Marie Curie, Universite Paris Diderot, CNRS/IN2P3, Paris, France}
+\par {\footnotesize $^{47}$ LPNHE, Université Pierre et Marie Curie, Université Paris Diderot, CNRS/IN2P3, Paris, France}
 \par {\footnotesize $^{48}$ Laboratoire de Physique Subatomique et de Cosmologie, Université Joseph Fourier and CNRS/IN2P3 and Institut National Polytechnique de Grenoble, Grenoble, France}
@@
 \par {\footnotesize $^{80}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
-\par {\footnotesize $^{81}$ Department of Physics \& Astronomy, University of Kentucky, Lexington, USA}
+\par {\footnotesize $^{81}$ Department of Physics and Astronomy, University of Kentucky, Lexington, USA}
 \par {\footnotesize $^{82}$ Physics Department, Lancaster University, Lancaster, United Kingdom}
 \par {\footnotesize $^{83}$ Department of Physics, Jožef Stefan Institute and University of Ljubljana, Ljubljana, Slovenia}
-\par {\footnotesize $^{84}$ Fysiska institutionen, Lunds universitet, Lund, Sweden}
+\par {\footnotesize $^{84}$ Fysiska institutionen, Lunds Universitet, Lund, Sweden}
 \par {\footnotesize $^{85}$ School of Physics and Astronomy, University of Manchester, Manchester, United Kingdom}
@@
 \par {\footnotesize $^{87}$ Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
-\par {\footnotesize $^{88}$ Departamento de Fisica, Universidade de Minho, Braga, Portugal}
+\par {\footnotesize $^{88}$ Departamento de Física, Universidade do Minho, Braga, Portugal}
 \par {\footnotesize $^{89}$ University of Nebraska-Lincoln, Lincoln, USA}
@@
 \par {\footnotesize $^{92}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman OK, USA}
-\par {\footnotesize $^{93}$ Department of Physics, University of Oslo, Oslo, Norway}
-\par {\footnotesize $^{94}$ The Ohio State University, Columbus, USA}
-\par {\footnotesize $^{95}$ Department of Physics, University of Oslo, Oslo, Norway}
-\par {\footnotesize $^{96}$ Universidad de Oviedo, Oviedo, Spain}
-\par {\footnotesize $^{97}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
-\par {\footnotesize $^{98}$ Princeton University, Princeton, USA}
-\par {\footnotesize $^{99}$ Rice University, Houston, USA}
-\par {\footnotesize $^{100}$ Universidade Estadual Paulista, São Paulo, Brazil}
-\par {\footnotesize $^{101}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
-\par {\footnotesize $^{102}$ School of Physics, Shandong University, Shandong, China}
-\par {\footnotesize $^{103}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
-\par {\footnotesize $^{104}$ University of Sofia, Sofia, Bulgaria}
-\par {\footnotesize $^{105}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
-\par {\footnotesize $^{106}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
-\par {\footnotesize $^{107}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
-\par {\footnotesize $^{108}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
-\par {\footnotesize $^{109}$ University ofWisconsin - Madison, Madison, WI, USA}
-\par {\footnotesize $^{110}$ Department of Physics, Yale University, New Haven, CT, USA}
-\par {\footnotesize $^{111}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
-\par {\footnotesize $^{112}$ Virginia Tech, Blacksburg, Virginia, USA}
-\par {\footnotesize $^{113}$ Department of Physics, University of Wisconsin, Madison WI, USA}
+\par {\footnotesize $^{93}$ The Ohio State University, Columbus, USA}
+\par {\footnotesize $^{94}$ Department of Physics, University of Oslo, Oslo, Norway}
+\par {\footnotesize $^{95}$ Universidad de Oviedo, Oviedo, Spain}
+\par {\footnotesize $^{96}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
+\par {\footnotesize $^{97}$ Princeton University, Princeton, USA}
+\par {\footnotesize $^{98}$ Rice University, Houston, TX, USA}
+\par {\footnotesize $^{99}$ Universidade Estadual Paulista, São Paulo, Brazil}
+\par {\footnotesize $^{100}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
+\par {\footnotesize $^{101}$ School of Physics, Shandong University, Shandong, China}
+\par {\footnotesize $^{102}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
+\par {\footnotesize $^{103}$ University of Sofia, Sofia, Bulgaria}
+\par {\footnotesize $^{104}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
+\par {\footnotesize $^{105}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
+\par {\footnotesize $^{106}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
+\par {\footnotesize $^{107}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
+\par {\footnotesize $^{108}$ University of Wisconsin - Madison, Madison, WI, USA}
+\par {\footnotesize $^{109}$ Department of Physics, Yale University, New Haven, CT, USA}
+\par {\footnotesize $^{110}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
+\par {\footnotesize $^{111}$ Virginia Tech, Blacksburg, Virginia, USA}
 \bigskip
@@
 \par {\footnotesize $^{b}$ Vladimir V.
-Gligorov acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 724777 “RECEPT”.}
+Gligorov acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 724777 “RECEPT”}
 \par {\footnotesize $^{c}$ Supported by the US-DOE, DE-AC02-07CH11359}
@@
 \par {\footnotesize $^{g}$ Supported by ES-MINECO, FPA2016-80994-c2-1-R \& MDM-2015-0509}
-\par {\footnotesize $^{h}$ Caterina Doglioni acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 679305 “DARKJETS”.}
+\par {\footnotesize $^{h}$ Caterina Doglioni acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 679305 “DARKJETS”}
 \par {\footnotesize $^{i}$ Supported by the US-DOE, DE-SC0009973}

### 2017-12-19 CWP/papers/HSF-CWP-2017-013_soft-dev/latex/cwp-wg-software-development.tex
Add Software Dev CWP WG paper in LaTeX format
- Imported document from Google Doc via docx export and pandoc conversion
  to LaTeX
- Adopted HSF Roadmap JHEP style and made many small corrections to import
- Remove 'verification' from the title
- Expand the roadmap bibtex file (allow for maximum sharing of
  this resource)
- Rewrite cwp-chapters.bib to use `techreport` entry (use of `article`
  without a journal results in a pretty broken citation)
- Add proper citations through the paper
- Through proof read with many small corrections and improvements to spelling,
  grammar and consistency
-

% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.

% JHEP preprint template
\documentclass[12pt,a4paper]{article}
\usepackage{jheppub}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xspace}

\usepackage[utf8]{inputenc}

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{../../roadmap/latex/cwp.bib}
\addbibresource{../../roadmap/latex/cwp-chapters.bib}

%% Comment out before final submission
%\usepackage{lineno}  % for line numbering during review
%\linenumbers

\abstract{The High Energy Phyiscs community has developed and needs to maintain
many tens of millions of lines of code and to integrate effectively the
work of thousands of developers across large collaborations. 
Software needs to be built, validated, and deployed across hundreds of
sites. Software also has a lifetime of many years, frequently beyond that
of the original developer, it must be developed with sustainability in
mind. Adequate recognition of software development as a critical task in the HEP
community needs to be fostered and an appropriate publication and citation
strategy needs to be developed. As part of the HEP Softare Foundation's
Community White Paper process a working group on Software Development,
Deployment and Validation was formed to examine all of these
issues, identify best practice and to formulare recommendations for the next
decade. Its report is presented here.}

\begin{document}

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-13 \\
 & & December 21, 2017 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{HEP Software Foundation Community White Paper Working Group -- Software
Development, Deployment and Validation}

\author{HEP Software Foundation:}
\author[a]{Benjamin Couturier}
\author[a,1]{Giulio Eulisse}
\author[b]{Hadrien Grasland}
\author[a]{Benedikt Hegner}
\author[b]{Michel Jouvin}
\author[c]{Meghan Kane}
\author[d]{Daniel S. Katz}
\author[e]{Thomas Kuhr}
\author[f]{David Lange}
\author[a,1]{Patricia Mendez Lorenzo}
\author[e]{Martin Ritter}
\author[a,1]{Graeme Andrew Stewart}
\author[a]{Andrea Valassi}

\affiliation[a]{CERN, Geneva, Switzerland}
\affiliation[b]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\affiliation[c]{Soundcloud, Berlin, Germany}
\affiliation[d]{University of Illinois Urbana-Champaign, Champaign, Illinois,
USA}
\affiliation[e]{Fakultät für Physik, Ludwig-Maximilians-Universität München,
München, Germany}
\affiliation[f]{Princeton University, Princeton, USA}
\affiliation[1]{Paper Editor}

\maketitle

\newpage

\hypertarget{scope}{%
\section{Scope}\label{scope}}

As part of the process to layout the HEP Software Foundation's Community White
Paper Roadmap~\cite{HSF-CWP-2017-01} we formed a group to study issues around
software development, deployment and validation. The starting point for the
group was to recognise that modern High Energy Physics experiments are large
distributed collaborations comprising up to a few hundred people actively writing
software. It is therefore central that the processes and tools used for
development are streamlined to ease the task of contributing code and
to facilitate collaboration between geographically separated peers. At
the same time we must properly manage the whole project, ensuring code
quality, reproducibility and maintainability with the least effort
possible. Making sure this happens is largely a continuous process, and
shares a lot with non-HEP specific software industries.

The group is therefore interested in tracking and promoting solutions
for the following areas:

\begin{itemize}
\item
    Distributed development of software components, including the tools
  and processes required to do so (code organisation, documentation,
  issue tracking, artifact building) and the best practices in terms of
  code and people management.
  \item
    Software quality, including aspects such as modularity and reusability
  of the developed components, sustainability of the development effort,
  architectural and performance best practices.
  \item
    Deployment of software and interaction with operations teams.
  \item
    Validation of the software both at small scales (e.g., best practices
  on how to write a unit test) and larger ones (large scale validation
  of data produced by an experiment).
  \item
    Software licensing and distribution, including their impact on
  software interoperability.
  \item
    Recognition of the significant contribution that software makes to
  High Energy Physics as a field.
\end{itemize}

The goal of this document is to highlight the challenges the field will
face in the next 5 to 10 years, and to provide a strategy on how address
them. Adopting best practices from non-HEP communities and adapting them
to HEP-specific needs will be key for the success of the effort.

\hypertarget{main-challenges-in-the-next-5-10-years}{%
\section{Main challenges in the next 5-10
years}\label{main-challenges-in-the-next-5-10-years}}

The challenges can be divided into two broad categories: those generic
to large-scale distributed software development and those specific to
the HEP environment.

The HEP-specific challenges are those derived from the fact that HEP is
a large, inhomogeneous community with multiple sources of funding,
mostly formed of people belonging to small university groups and some
larger laboratories. Software development effort within an experiment
usually encompasses a huge range of experience and skills, from a few
more or less full time experts to many physicist programmers with little
formal software training. In addition, the community is split between
different experiments that often diverge in timescales, size and
resources.

Another peculiarity of HEP is that the experiment software is
usually divided in two separate use cases, production (data acquisition,
data reconstruction or simulation) and user analysis, whose requirements
and life-cycles are completely different. The former is very carefully
managed in a central and slow moving manner, following the schedule of the
experiment itself. The latter is much more dynamic and strongly coupled
with conferences or article publication timelines. Finding solutions
which adapt well to both cases is not always obvious or even possible.

Principle areas of challenge are:

\paragraph{Effective use of developer effort}

\begin{itemize}
\item
    The large fraction of short term students and collaborators
  found in our field requires a process for evaluating software
  contributions to common projects and for assuring not just their
  quality, but also their maintainability.
  \item
    For the same reason, we need to minimise the time required for people
  to begin to contribute effectively.
\end{itemize}

\paragraph{Knowledge sharing}

\begin{itemize}
\item
    There is a strong need for knowledge sharing, including effective
  training. This is required by groups that are geographically
  separated, but also amongst different fields, with different skills,
  languages and cultures. E.g., some students might be able to track new
  trends better than older colleagues, but they might lack vision;
  software engineers might provide better implementations than
  postdocs, but may lack a complete understanding of the broader
  picture for the experiment needed to set priorities.
  \item
    There is an historical and stubborn compartmentalisation between
  different experiments, so propagating knowledge horizontally across
  experiments is particularly hard.
\end{itemize}

\paragraph{Reusing code effectively}

\begin{itemize}
\item
    While striving towards common solutions, the group recognises that a
  single platform or set of tools will not fit all use cases, but we
  emphasise that generic best practices and recommendations can be made.
  A top-down approach, where a given tool or solution is mandated has
  not proven to be viable in the past. However, if we can
  identify success stories and make sure they can be extended (if
  needed) and advertised as possible general solutions, 
  duplicate efforts or developments are more likely to be avoided.
\end{itemize}

\paragraph{Validation of complex output}

\begin{itemize}
\item
    Efficient validation of the complex outputs of HEP code, involving
  inputs from highly specialised experts and unique sources, is hard to
  do, especially taking into account stochastic processes and long
  tails.
\end{itemize}

\paragraph{Data, code and knowledge preservation}

\begin{itemize}
\item
    Long term maintainability of the code to handle data that was obtained
  decades before as required by the lifetime of a HEP experiment.
\end{itemize}

\paragraph{Code lifecycle and deprecation}

\begin{itemize}
\item
    In a highly demanding software environment such as the HEP
  communities, there is great pressure to adopt updated versions of
  tools and libraries for performance or design benefits. However, there
  is little corresponding pressure to deprecate old components that may
  have been used for years, causing a bloat of the active software stack
  and provoking crises when abandoned upstream libraries start to fail
  to build. Hence there is a need to engage in effective, smooth and
  continuous lifecycle management of the codebase.
\end{itemize}

\paragraph{Hardware and software evolution}

\begin{itemize}
\item
    Hardware trends are now towards increasing processing capacity through
  concurrency (many cores and vectorisation) rather than any increase in
  serial processing capacity. This is at odds with the traditional HEP
  software workflows, which have largely been serial. Memory latencies
  and a deep memory and storage hierarchy have become key drivers of
  performance. Increasingly specialised processors are also arriving in
  the marketplace, which are not naturally suited to HEP codes. Adapting
  to this new hardware environment and maintaining the efficiency of HEP
  code becomes increasingly hard.
  \item
    New innovations in software development usually come from outside of
  the HEP community. These can provide significant advantages in
  exploiting new hardware or aiding efficient distributed development
  and such technologies need to be tracked, investigated and deployed.
\end{itemize}

\paragraph{Software deployment}

\begin{itemize}
\item
    As HEP software today runs in widely heterogeneous environments, from
  traditional grid computing sites to High Performance Computing sites,
  including the gamut of academic and commercial cloud services, there
  is a challenge to uniformly provide the appropriate software in all of
  these environments. End user analysis code also needs to be managed
  and distributed into these environments. CVMFS~\cite{7310920} has been hugely
  successful here, however it does not cover all cases and a viable
  common solution may emerge and be adopted more widely.
\end{itemize}

\paragraph{Recognition of software development}

\begin{itemize}
\item
    Writing high quality software is a skilled activity that greatly
  contributes to the success of the field. Ensuring that software
  developers receive recognition for their contributions will help
  retain people in the field and help their career development.
\end{itemize}

\hypertarget{meeting-the-challenges}{%
\section{Meeting the Challenges}\label{meeting-the-challenges}}

The previous section introduced a brief summary of the main challenges
this working group has identified. In the next paragraphs we both
expand on these and identify viable existing solutions and areas where
more investigation is required.

\hypertarget{effective-use-of-developer-effort}{%
\subsection{Effective use of developer
effort}\label{effective-use-of-developer-effort}}

\subsubsection{Design documentation and training}

For developers to be able to contribute effectively to experiment
software it is first of all necessary for them to have a clear idea of
how to write code that will mesh effectively with the framework or the
rest of the software stack. Thus it is essential that developers can
access clear documentation, with good examples as to how to model their
code. Effective contact with experts should be available so that good
design choices are made from the outset.

Once a clear picture of the framework is available to new developers the
first task of the contributor should be to make their intentions clear
to others. A common technique is to use a request for comments document
which outlines the design and allows other developers to provide
feedback before substantial coding effort is expended. Such documents
can be accompanied by demonstrator or prototype code to better
illustrate the purpose of the of the changes. This approach allows
documentation and testing to be incorporated into the development itself
and not left as an afterthought. This model was used by the Concurrency
Forum~\cite{ConcurrencyForum} in the past
and was very successful in selecting the most promising approaches for
the community and disseminating knowledge amongst different experiments.

The task for developers who might contribute a general purpose piece of
code that is used across experiments is more challenging as active
solicitation of input will need to be sought. However, if experiment
documentation is open to the HEP community then different experimental
boundary conditions will be more easily understood and translated into
the design and interface of a software product.

In general, good software design principles should be followed and thus
generic training in software design, but with a focus on the HEP problem
space, is a valuable investment for the community. We note that several
computing schools exist already, but that overall only a small fraction of
HEP students attend them as they are considered quite specialist.
Universities, in both their undergraduate and graduate programmes, play an
important role too in core skills training (and in some institutes also
in relevant advanced topics), but this can be uneven. Effort should be
made to incorporate more generic training into the education programmes of
the experiments; centering these activities around some of the larger
labs will also allow sharing of the training effort across different
experiments.

\hypertarget{standard-development-tools}{%
\subsubsection{Standard development
tools}\label{standard-development-tools}}

To maximise the productivity of developers it is very beneficial to use
development tools that are not HEP-specific. There are many open-source
projects and tools that are of similar scale to large experiment
software stacks and standard tools are usually well documented. Many
simple questions can be answered using standard resources, such as Stack
Overflow.

Insofar as an experiment requires customisation of tools, documentation
and training are evidently important. However, the costs of such
customisation can be easily outweighed by the cost of maintenance and
training, as well as a possible loss of generic skills for users.

For source control in particular, the open-source world has moved
strongly towards distributed version control as superior to centralised
solutions. The advantages for the individual developer are numerous,
including independence from central services and greater freedom to
experiment. Finished code enhancements can be proposed using a
pull request workflow so that high quality contributions are made with
the central repository receiving pristine patches. We note that the
community has generally chosen to move to git. This standardisation
is very welcome as it also brings an alignment with many open-source
projects and commercial organisations.

The HEP current community is widely using CMake for the builds of
software packages. This is an open-source system dedicated to building,
testing, and installing software that captures the specific build
specifications for each particular package in a platform independent
way. It is highly regarded because of its modular and scalable
characteristics and is also very widely used outside HEP with a lot of
support available. CMake's use of plugin modules allows scope for
sharing between HEP experiments in the cases of building HEP specific
code, though at the moment probably the best use is not yet made of this
feature in our community. CMake can orchestrate the builds of large
numbers of packages (e.g., the LCG releases of over 300 packages), but
can become quite specific when doing so and thus more limited. The HSF
Packaging group found
Spack~\cite{HSF-TN-2016-03,Gamblin:2015:SPM:2807591.2807623}
to be a promising
more general solution for the task of building an HEP software stack, however, it
currently has a relatively small user base, centered on HPC, and might
just become a niche product with little traction in the wider
development community. More generic tools, such as Homebrew or
Conda are interesting but may not have all the features needed by
HEP experiments.
Thus this area still requires close attention to active developments in
the open-source world.

For reporting problems, tracking bug fixes and planning improvements to
the software issue trackers should be used. The best of these offer
multi-level grouping of issues and provide direct support for useful
aspects of agile development discussed below. Optimal choices will offer
good integration with source code management (e.g., GitHub Issue
Tracker, Jira, GitLab Issue
Tracker) and the choice should usually be made in parallel
with an source code management tool.

\hypertarget{development-environment}{%
\subsubsection{Development environment}\label{development-environment}}

It is important that developers can focus on the design and
implementation of the code and do not have to spend a lot of time on
technical issues. Clear procedures and policies must exist to perform
administrative tasks in an easy and quick way. This starts with the
setup of the development environment, for example by having a minimal
set of requirements on the operating system and the installed packages.
Supporting different platforms not only allows the developers to use
their machine directly for the development, it also provides a check of
portability basically for free.

Other typical workflows are the update of versions or the submission of
code for review. It decreases the turnaround time and reduces
frustration if code quality criteria and policies are checked as early
as possible, ideally on commits to the local repository or even by the
editor. Modern IDEs, which offer good integration with modern
development tools, can be particularly valuable for novice coders. They
provide support for developers in navigating the code base as well as
catching elementary errors. It would be of great benefit for the
community to ensure that standard project templates
(e.g., HSF Project Template~\cite{HSFProjectTemplate}) work
with minimal configuration with standard IDEs, such as Eclipse, XCode,
KDevelop, Qt Creator, etc. For large experiment code bases, the scaling
of these IDEs to many millions of lines requires investigation.

Finally, support is important. A system that enables developers to
get quick and competent advice from experts will improve their
productivity.

\hypertarget{social-coding-sites}{%
\subsubsection{Social coding sites}\label{social-coding-sites}}

The advantages of social coding sites are that they allow developers to
share and discuss code through very streamlined and highly functional
web interfaces. Distributed version control systems are key enablers for
these sites. There are commercial choices, such as GitHub or BitBucket,
that can serve a community well; alternatively a privately hosted
service, such as CERN's GitLab instance, may mesh more easily with other
related services. We note that ``bridging'' code between different service
platforms has been developed by a number of experiments and could be shared
across the community. Smaller projects would probably
benefit from the increased exposure and larger development community from being hosted on GitHub.
For attribution and credit, GitHub also offers a more standard platform
that can be helpful to a developer's career. However, the key
requirement here is the ability to discuss and revise code changes in a
recorded fashion as they become part of the code repository.

\hypertarget{continuous-integration}{%
\subsubsection{Continuous integration}\label{continuous-integration}}

Proper testing of changes to code should always be done in advance of a
request being accepted. Continuous integration, where merge or pull
requests are built and tested in advance is now standard practice in the
open-source community and in industry. HEP software should not be
different. Not only should continuous integration run unit and
integration tests, it can also incorporate code quality checks and
policy checks that will help improve the consistency and quality of code
at low human cost.

For running continuous integration and build orchestration many options
exist (Jenkins, Gitlab CI, Travis, Bamboo) and the particular choice of
which one to use is a relatively pragmatic decision, with only a very
limited impact on the ability to cooperate around software.

\hypertarget{code-review}{%
\subsubsection{Code review}\label{code-review}}

Incorporation of code into the code base should usually also involve
sign off from at least one relevant expert who can check for
architectural violations, anti-patterns or even just simple coding
errors. Code review should be seen as a process of dialogue between the
developer and the reviewer (community expert) and it should be firm, but
friendly.

As well as a way to validate code, reviews can also be a way for
newcomers to learn about the architecture of a project. As part of a
mixed experience reviewing team, fresh eyes on a piece of software can
produce original comments, but the discussions occurring as a side
effect of the review effectively helps new developers understand the
issues at stake.

\hypertarget{agile-development}{%
\subsubsection{Agile development}\label{agile-development}}

HEP is a community where the developer of a piece of software is also
likely to be the `customer' as well. However, despite this, many useful
ideas can be taken from the various agile methodologies, in particular
with regard to solving only the problem at hand and keeping development
cycles as short as possible to find iterative and evolutionary
solutions. Agile development methodologies for physicists, and how to
select the most appropriate one, might be best taught by example rather
than as formalism.

One aspect of agile methodology that would serve HEP rather well is the
concept of a user story that describes the purpose of a system, rather
than requirements capture that was used in the past. Such a method is
more flexible in adapting to changing requirements, which can be quite a
common situation in HEP, especially in the analysis domain.

HEP would benefit from adopting agile retrospectives. A retrospective is
a formalised way for contributors to meet regularly to discuss ways to
improve their software development process. Since a retrospective occurs
regularly, there is a chance to catch issues affecting the development
process as they arise. Issues can be tracked and mitigated over time
rather than ballooning into daunting barriers. Additionally, the focus
is on learning from failures quickly, without placing blame. This
welcomes feedback, positive or negative, toward the goal of improving
the software development process.

Retrospectives can be structured as needed to suit HEP, but the standard
approach lends itself nicely to solving current problems since the last
retrospective:

\begin{itemize}
\item
    Gather data. What went well? What didn't go well? What needs to be
  improved? Each person who has input should have a chance to succinctly
  state their observations.
  \item
    Discuss top issues in prioritised order. Based on the feedback of the
  group, discuss top issues to reach an understanding about the root
  causes.
  \item
    Assign action points. Based on the pain points discussed, assign
  action items that are small enough in scope that they can be resolved
  before the next retrospective.
\end{itemize}

This setting of information sharing and collaboration can yield clear
productivity benefits. For example, it is possible that a developer
experiences issues that slow down their productivity without being aware
of it. Perhaps the developer became accustomed to the build time being
very high and thinks of it is as given constraint. In a retrospective, a
discussion about build times may arise along with known solutions for
reducing the build times. This will elicit awareness about the issue, a
solution, and a subsequent gain in productivity.

Agile development is very well supported by the modern development
infrastructure outlined above.~

\hypertarget{knowledge-sharing}{%
\subsection{Knowledge sharing}\label{knowledge-sharing}}

\hypertarget{documentation}{%
\subsubsection{Documentation}\label{documentation}}

As was mentioned above, proper training and documentation is key to
efficacious use of developer effort. For documentation that has to be
specific, favoured solutions would have a low barrier of entry for
contributors, but also allow and encourage the review of material.
Consequently it is very useful to host documentation sources in a
repository with a similar workflow to code and to use an engine that
translates the sources into modern web pages. An example would be using
the Jekyll engine, as the main HSF website currently does, or using
services such as Read The Docs.

Traditionally HEP has often used Doxygen for reference
documentation, especially given its emphasis on C++. New documentation tools are also
available that integrate Doxygen with better solutions in other
languages, e.g., the python Sphinx documentation tool can use
Breathe to integrate C++ and python documentation. Sites such as
Read The Docs also take much of the pain out of producing an attractive documentation site.

\hypertarget{training}{%
\subsubsection{Training}\label{training}}

Code development, as a critical and transferable skill, is well
recognised inside the HEP software community. Various schools exist that
train students in good coding practice, but only a small minority of
contributors to HEP code attend these. There is therefore scope for
broader and more accessible training that experiments can play a role in
encouraging collaborators to attend. Training events can start with
generic material, shared between experiments, then move to separate
experiment specific topics. This would be especially useful for
experiments sharing a home laboratory and provides an additional benefit
when different groups have managed to converge on common components.

Training and documentation are intimately linked and good training
materials can be studied offline by students in their own time.

As much use as possible should be made of generic resources that already
exist; there is a role here for the HSF to curate useful training
materials (where their ongoing usefulness and validity needs to be
managed) and encourage cooperation with other training initiatives, such
as Software Carpentry~\cite{SoftwareCarpentry}.

\hypertarget{seminars-and-discussion-forums}{%
\subsubsection{Seminars and discussion
forums}\label{seminars-and-discussion-forums}}

These are important to the community, in particular to discuss new
technologies or tools at an early stage of development. A good example
is the Software Technology
Forum~\cite{SoftwareTechForum}. These forums also allow questions to be posed
about problems that are encountered and identifying existing software that may provide
a solution (thus avoiding re-inventing the wheel yet again).

In terms of knowledge sharing among developers, solutions such
Stack Overflow are excellent for establishing
communication between software developers and publishing answers on
the web. In fact, this solution is used by many developers to find
solutions to the daily problems they face. It seems that in general
HEP software developers are consumers of these general help lists and we
encourage the community to take more active role in contributing as
well. Where a smaller community does not have the critical mass for
their own Stack Exchange service other products, like Discourse, can be
installed on-premises.

In addition, attempts have also been made to set up general software
discussion mailing lists in the past. However, it seems that after some
initial flurry of activity these are not really very successful. They
may be too far away from most developers day-to-day concerns and there
is probably some reluctance to pose questions to a wide list. Though
being a fast and attractive solution for discussions about a particular
issues they are not as viable for long term or broad knowledge sharing.
There may be some use in trying alternatives, such as a Slack channel,
at least to prototype with low investment.

\hypertarget{workshops-and-community-reviews}{%
\subsubsection{Workshops and community
reviews}\label{workshops-and-community-reviews}}

Hosting workshops on specific topics is a particularly useful way to
bring together experts from different experiments. These have an
essential role in introducing new ideas to the community and providing
an environment in which people can meet and interact. The HSF has also
organised community meetings on aspects of future HEP software
development that are of particular interest to the community in a
strategic sense. Two such meetings have been organised so far: one on
Geant V simulation~\cite{HSF-GeantV-Report}, and the other on the analysis
ecosystem~\cite{HSF-Analysis-Report}. These meetings can provide very valuable
feedback to developers on the priorities of the experiments and the most
critical features to develop and support. Organising further meetings on such topics provides
important strategic feedback on development.

\hypertarget{conferences}{%
\subsubsection{Conferences}\label{conferences}}

HEP has a strong tradition of software and computing conferences (ACAT
and CHEP are good examples). These continue to be a critical part of the
lifecycle of software and of the community that supports it, offering
recognition of work done as well as dissemination of knowledge about
available solutions that can, hopefully, be adopted by others. Although
they focus on completed pieces of work, they complement other activities
well and provide an important space for social interaction.

It would be beneficial for the field as a whole to increase its
participation in general conferences and software focused events. This
would help generate relationships and collaborations outside the direct
field with experts who may have interesting observations and
suggestions.

\hypertarget{journal-publication}{%
\subsubsection{Journal publication}\label{journal-publication}}

Journals disseminate information to the wider community in a
permanent way and so form an important part of the collective memory of the
community. Journals also have another significant role to play
in recognition of software work, which is discussed at greater length in
Section~\ref{recognition-of-software-development}.

\hypertarget{reusing-code-effectively}{%
\subsection{Reusing code effectively}\label{reusing-code-effectively}}

Effective reuse of code requires good design, documentation and support.
Good design starts from a good understanding of the problem to be
solved, with an appreciation of the important external interface to
provide to clients. APIs should be stable, with scope to change the internal
details of an implementation as needed. Training software developers in
good design principles will certainly help in this regard; however,
often the reuse potential of code is only discovered post-facto and then some
advice and expertise in refactoring will be useful for the longevity of
the package. Here Agile methodologies can help to focus on the key areas
of responsibility and avoid over designing interfaces for imagined
situations.

There is scope to improve the training in these areas within the HEP
community as discussed above.

In all cases good documentation for code will greatly help adoption (the
same solutions apply here as were discussed above) and this provides a
key aspect of user support. Utilising popular standard development
tools encourages collaboration and provides issue tracking capabilities.

\hypertarget{software-licenses}{%
\subsubsection{Software Licenses}\label{software-licenses}}

All code should have a clear copyright holder, which gives the legal
ownership of the software. Applying a license specifies how the software
can be used by others. Open-source licenses are now widely accepted by
the HEP experiments, which we believe is the correct approach. Open
source licenses fall into various categories

\begin{itemize}
\item
    Copyleft licenses mandate that any public releases with changes to the
  code must be licensed under the same terms and, in addition, that any
  software which uses the copyleft code must also itself be distributed
  under a copyleft license (which is why these licenses are sometimes
  referred to as `viral'). The General Public License, GPL, is the best
  known of these.
  \item
    Weak copyleft licenses (such as the Lesser General Public License,
  LGPL or the Mozilla Public License, MPL) remove the latter copyleft
  via use requirement, but still mandate that direct modifications to
  the package itself be redistributed under the same terms.
  \item
    Permissive licenses (Apache, MIT, BSD) generally allow for public
  modifications to be made without requiring them to be released under
  the same terms as the original code.
\end{itemize}

There is a modern report from the HSF on licensing 
topics~\cite{HSF-TN-2016-01} and we do not
repeat the discussion here. Instead we encourage code authors to ensure
that copyright and license for code is well established at the beginning
of the project, including the status of contributions to the code. The
implications of the chosen license should be properly understood,
especially in so far as it affects contributors to the code base and
users of the code who may incorporate it into a larger software stack,
together with their own code (the GPL has a fairly profound effect in
this regard).

\hypertarget{validation-of-complex-outputs}{%
\subsection{Validation of complex
outputs}\label{validation-of-complex-outputs}}

The outputs of HEP software packages can be extremely complex and thus
difficult to validate. This is especially true of the simulation of
modern high energy physics detectors, which are incredibly complex
machines and need to be accurately modeled to understand their
performance for rare decays and weak signals. Up to now most experiments
have developed their own infrastructure for comparing results.
Comparison of distributions is a relatively well understood topic in
statistics, so having standard tools that help compare outputs from
different versions of software would help save effort between
experiments and offers scope for improving the quality of our
validation.

\hypertarget{data-and-code-preservation}{%
\subsection{Data and code
preservation}\label{data-and-code-preservation}}

The longevity of code in HEP is becoming increasingly important. Large
experiments have multi-decade active lifetimes, with ``data
preservation'' now considered an important topic after an experiment has
finished taking data. Data taken in the early years of the experiment
may need to be reanalysed years afterwards with older software that has
not been built for a long time. Software virtual machines or containers
can help a great deal in preserving the working environment, including
the source code itself. Should changes need to be made to software to
correct for a long dormant bug or add a necessary new feature, having a
virtual machine or container with the source code and build environment,
in addition to any binary products, is important. Here the use of
standard tools, as encouraged above, can help, with the state of the
code and externals being encapsulated in a source repository tag. This
is becoming a prerequisite to publication in some journals, so being
able to trace the software used to produce some results, all the way
through to an analysis, is crucial. Experiments need to strongly
encourage their members to work in a preservable way.

In the design of code, relying on standard language features is more
sustainable than utilising external libraries that may become
deprecated. Of course, libraries that do tasks well are preferable to
inferior home grown implementations and achieving some uniformity in the
community will help for longer term support issues when they arise.

For further details on this topic we refer to the Data Preservation Community
White Paper group~\cite{HSF-CWP-2017-06}.

\hypertarget{code-deprecation}{%
\subsection{Code deprecation}\label{code-deprecation}}

The complexity of HEP software and the generic nature of many of the
sub-tasks that need to be performed mean that many external libraries
are essential to our workflows. It would be completely unfeasible and a
waste of effort to reproduce such code internally. However, there are
often many solutions to the same problem and picking the one that has
the best long term prospects or becomes the most widely adopted is not
possible a priori. Sometimes externals will simply be eclipsed by
alternatives and sometimes they will change their APIs in
backward-incompatible ways. The problem is often exacerbated by the fact
that the HEP-specific client may itself be only in maintenance mode and
the original author may not even work in the field anymore.~

In these cases evolving away from deprecated code is required. This can
be a labor intensive operation, but often many aspects of it are
automatable. There is a good example in the refactoring of C++, which is
a hard problem, but with the development of Clang and full knowledge of
the abstract syntax tree (AST) of the code, there are significant
possibilities for using this knowledge to tackle much of the drudge work
of refactoring. There have been successful examples of using Clang in
this way~\cite{ClangMR41342}
and we believe this is an area where investment could be very profitable,
both specifically for HEP and as a contribution to the wider open-source
community. Tools to further help validate the updated code can be shared
between experiments, e.g., support for any migration to new versions of
ROOT~\cite{Brun1996} or Geant4~\cite{Agostinelli2003}.

Changing software versions and evolving platforms is a natural part of
the lifecycle of production environments and can be helped by migration
to more self-contained deployment models, e.g., by using containers
instead of a site provided environment.

\hypertarget{hardware-and-software-evolution}{%
\subsection{Hardware and software
evolution}\label{hardware-and-software-evolution}}

The evolution of processors in recent years have given rise to highly
parallel architectures (from many core CPUs to GPUs); in addition,
thanks to the push coming from the mobile market, viable low-power
solutions are also now common (e.g., ARM architectures).

\hypertarget{effective-exploitation-of-parallel-hardware}{%
\subsubsection{Effective exploitation of parallel
hardware}\label{effective-exploitation-of-parallel-hardware}}

Implementation and optimisation of software on parallel architectures
requires different skills, tools and training (generally more advanced)
than the ones required for the serial world. In order to allow
physicists to effectively exploit parallel architecture, means are
required to insulate end user programmers from many of the low-level
difficult details of parallel programming.

Many specific details of parallel architectures can be masked by the
adoption of appropriate libraries. Generally HEP problems fit better
into a task based scheduling pattern, rather than to an optimiser of
kernels run in tight loops. The community has utilised Intel's Threaded
Building Blocks (TBB)~\cite{Pheatt:2008:ITB:1352079.1352134,TBB} successfully in
many areas. Should TBB continue to develop and be a good fit for our problems
this common solution also helps solve the problem of resource balancing between different parts of
the code. Should the solutions adopted become fragmented then overall
resource balancing becomes much harder and may become difficult to
solve. Evolution of the C++ standard will most likely bring more
concurrency primitives into the language, however, so some evolution
here is to be expected and should be planned for. This would mean
appropriate zero cost abstractions when possible.

\hypertarget{achieving-performance-portability}{%
\subsubsection{Achieving performance
portability~}\label{achieving-performance-portability}}

The fact that different parallel architectures (e.g., CPUs and GPUs) have
different sweet-spots for performance makes compile time performance
portability extremely difficult to obtain. This is true across
architectures, but also between different generations of the same
architecture. Past abstraction techniques might not work and development
needs to target specific hardware for best performance. However, it is
strongly advisable that the design of the code separates physics logic
from performance optimisation. This helps to avoid premature
optimisation and a harmful reduction in code maintainability. Options
for runtime optimisation would also be possible (e.g., fat libraries
with code for various vector architectures) and may help to avoid an
explosion in build target architectures. Such techniques are an area
where common community investigation would be worthwhile, especially
when common tools are used in multiple projects.

When aiming for cross-architecture performance portability, one
important concern is data layout, as different architectures have
different preferences in this respect. For example, GPUs perform best
when memory accesses from different threads are coalesced (as it
optimises memory bus utilisation), whereas CPUs perform best when writes
from different threads are relatively far apart in RAM (to avoid the
cache trashing effect known as ``false sharing''). These differences can
be abstracted away through the use of data structure generation mechanisms,
which are still in their early days but are begining to be offered by some
programming environments such as
Kokkos~\cite{CarterEdwards20143202},
and some domain specific languages, such as those used in the context of the
H.E.S.S.\ Observatory~\cite{HESS} and in the Cherenkov Telescope Array
reconstructions~\cite{CTA}.

\hypertarget{tracking-of-new-software-technologies-and-tools}{%
\subsubsection{Tracking of new software technologies and
tools}\label{tracking-of-new-software-technologies-and-tools}}

Compared to when HEP started as a field, the importance and ubiquity of
software development has increased hugely. If in the late 80s and 90s
many of the tools and practices were actually started inside our field
itself, or at least within academia, it is now true that a large amount
of development happens outside the traditional HEP field. Starting from
the contributions of ``bedroom coders'', to the scale of large
corporations like Google, Apple, or Facebook, a large number of high
quality tools and libraries are now available as open-source projects
for everyone to use. While this affords the field great opportunities,
the rate at which these tools and libraries change is something that
necessitates a continuous attention and an in-depth scrutiny of their
potential, as well as managing inevitable deprecations. As was
pointed out above, pooling community knowledge here will make optimal
choices in new technology more likely.

\hypertarget{dealing-with-hardware-diversity}{%
\subsubsection{Dealing with hardware
diversity}\label{dealing-with-hardware-diversity}}

The increasing number of different architectures, operating systems and
compilers with different versions give rise to a large number of
possible platforms that combine these parameters. Experiment software
should be tested on a large range of platforms to ensure a
competitive and effective use of these new technologies, which requires
good test cases and striving to minimise the operational and human costs
of multiple platforms. Use of open-source orchestration toolkits for
managing these stacks can help a lot. Jenkins has received much
attention, but alternatives, such as Travis, exist and there may be scope
for a community evaluation. Aspects such as application independence,
adaptation and modularity (a plugin-based structure) are important.
Basic platform types can share a lot of common infrastructure (e.g.,
x86\_64, AArch64 and PowerPC) that minimise the cost of adding new
flavors if common components are used.

There is an important link here when deciding what level of validation
is considered acceptable on each platform, especially if platforms do
not produce binary identical results. In order to evaluate and prepare
for the incorporation of such new platforms having early access is
extremely important. CERN's TechLab and OpenLab projects help provide
that, but there is a need to ensure that the whole community can benefit
and more sites could be involved in this activity.

\hypertarget{software-deployment}{%
\subsection{Software deployment}\label{software-deployment}}

Deployment of HEP software has been revolutionised by the widespread
adoption of CERN Virtual Machine File System (CVMFS)~\cite{7310920}
for flexible and ubiquitous access to a common
filesystem namespace with efficient distribution of content that can be
strongly cached at multiple levels. We expect that CVMFS will remain the
mainstay of software distribution to most computing sites in the
foreseeable future.

However, CVMFS cannot cover all possible use cases. Sites that have no
outbound network access (common in the HPC world) or are mobile and
perhaps temporarily disconnected (e.g., developer laptops) require
additional distribution mechanisms. Here, containerisation offers a
great deal of promise as a way to isolate the software environment from
the underlying hardware and to conveniently provide all the needed
runtime components. This is particularly useful when other constraints
drive the base operating system away from the standard one adopted and
validated by the experiment. At the moment Docker is favored as being
exceptionally convenient for development, with automated conversion to
other container implementations (e.g., Shifter or Singularity) being
possible. There should be no reason to deeply bind anything in the
install process into any particular container implementation.

Containers also play a role in data and code preservation as already
noted. While containers are architecture specific, there is currently no
serious risk that, e.g., x86\_64 platforms will be unavailable in the
foreseeable future.

\hypertarget{recognition-of-software-development}{%
\subsection{Recognition of software
development}\label{recognition-of-software-development}}

Proper recognition of the contribution that software makes to the
success of High Energy Physics is a key element of sustaining that
effort and establishing software as a `first class citizen' in the
field. As an academic discipline, HEP primarily recognises publications
in refereed academic journals and the citations of those publications as
indicators of merit. Therefore we strongly encourage authors of software
to submit publications of that software to such journals, of which there
are now many~\cite{SSI2017}.
It is also equally important that publications from the field
appropriately cite the software that they have used.

The correct level of granularity of software versions for publication
will need to be judged by authors themselves, as well as reviewers.
Significant improvements in algorithms, in either machine resource
consumption or in physics performance, would certainly merit a new
publication, as might a major release of software with significant new
features. There is also a strong case for having an overall publication
that encompasses a major piece of combined software used by an
experiment. One example is an article on an experiment's offline
software stack (that itself includes citations of the other software
used within the combined software). Such publications would then be
highly citable in physics papers, analogous to the common citations of
the detector design papers.

Currently, if HEP software has an article written about it, it will tend
to be in the conference proceedings of, e.g., CHEP or ACAT. Although
these are refereed, in general the barrier for entry is low, and as such
they are not considered as prestigious as refereed journals as the
quality is mixed. Without some reorganisation of the community, and a
more strict refereeing process, it is unlikely that conference
proceedings will become the journals of choice for recognition in our
field. However, conference journals (e.g., Journal of Physics:
Conference Series) usually do not allow resubmission of articles to
other journals without extended content. Software authors should be
aware of these issues when deciding where best to publish.

There are also now interesting projects to directly publish software
itself, for example to Zenodo~\cite{Zenodo} or
figshare\cite{figshare}, as has been recommended by the
FORCE11
Software Citation Working Group~\cite{10.7717/peerj-cs.86}. We are very
encouraging of such efforts, and believe it would be useful for HEP
software authors to participate, but at the moment the actual
recognition accrued from such a publication is unclear. Through its work
in reviewing papers, the HEP community could choose to suggest to
authors that they cite the software they use, whether directly or via
software papers. Over time, this small effort on the part of reviewers
could be sufficient to change community practice regarding software
credit.

For further discussion on this topic see also the Careers and
Training Community White Paper~\cite{HSF-CWP-2017-02}.

\hypertarget{roadmap-for-the-next-10-years}{%
\section{Roadmap for the next 10
years}\label{roadmap-for-the-next-10-years}}

In the areas covered by this white paper HEP must endeavor to be as
responsive as possible to developments outside of our field. In terms of
hardware and software tools there remains great uncertainty as to which
platforms will offer the best value for money in a decade. It therefore
behooves us to be as generic as possible in our technology choices, retaining
the necessary agility to adapt to this uncertain future.

Our vision is characterised by HEP being current with technologies and
para\-digms that are dominant in the wider software development world,
especially for open-source software, which we believe to be the right
model for our community. In order to achieve that aim we propose that
the community establishes a development forum that allows for technology
tracking and discussion of new opportunities. The HSF can play a key
role in marshalling this group and in ensuring its findings are widely
disseminated. In addition, having wider and more accessible training for
developers in the field, that will teach the core skills needed for
effective software development, would be of great benefit.

Given our agile focus, it is better to propose here projects and
objectives to be investigated in the short to medium term, alongside
establishing the means to continually review and refocus the community
on the most promising areas. The main idea is to investigate new tools
as demonstrator projects where clear metrics for success in reasonable
time should be established to avoid wasting community effort on
initially promising products that fail to live up to expectations.

\hypertarget{short-term-projects-1-2-years-and-ongoing-activities}{%
\subsection{Short term projects (1-2 years and ongoing
activities)}\label{short-term-projects-1-2-years-and-ongoing-activities}}

\begin{itemize}
\item
    Establish a common forum for the discussion of HEP software problems.
  This should be modeled along the lines of the Concurrency
  Forum~\cite{ConcurrencyForum}, which was very successful in establishing
  demonstrators and prototypes that were used as experiments started to develop
  multi-threading frameworks.
  \item
  Continue the HSF working group on packaging, with more prototype
  implementations based on the stronger candidates identified so far.
  \item
    Provide practical advice on how to best set up new software packages,
  developing on the current project template work and working to
  advertise this within the community.
  \item
    Work with HEP experiments and other training projects to provide
  accessible core skills training to the community. This training should
  be experiment neutral, but could be usefully combined with the current
  experiment specific training. Specifically this work can build on and
  collaborate with recent highly successful initiatives like the LHCb
  StarterKit~\cite{1742-6596-898-8-082054} and ALICE Juniors~\cite{ALICEJuniors}
  and with established generic training initiatives such as Software
  Carpentry~\cite{SoftwareCarpentry}.
  \item
    Strengthen links with software communities and conferences outside of
  the HEP domain, presenting papers on the HEP experience and problem
  domain. SciPy, SC (informally known as Supercomputing), RSE
  Conference, and Workshop on Sustainable Software for Science: Practice
  and Experiences (WSSSPE) would all be useful conferences to consider.
  \item
    Write a paper that looks at case studies of successful and
  unsuccessful HEP software developments and draws specific conclusions
  and advice for future projects.
\end{itemize}

\hypertarget{medium-term-projects-3-5-years}{%
\subsection{Medium term projects (3-5
years)}\label{medium-term-projects-3-5-years}}

\begin{itemize}
\item
    Prototype C++ refactoring tools, with specific use cases in migrating
  HEP code.
  \item
    Prototyping of portable solutions for exploiting modern vector
  hardware on heterogenous platforms.
  \item
    Develop tooling and instrumentation to measure software performance,
  especially in the domain of concurrency. This should primarily aim to
  further the developments of existing tools, e.g.,
  igprof~\cite{Eulisse:2005zz}, rather than developing new ones.
  \item
  Develop a common infrastructure to gather and analyse data about
  experiments' software, including profiling information and code
  metrics, to ease sharing of this information across experiments.
  \item 
  Develop tool(s) that allow developers to understand the usage of their
  software and to be credited for this usage, possibly via using these metrics
  associated with software citation.
  \item
    Undertake a feasibility study of a common toolkit for statistical
  analysis that would be of use in regression testing for experiment's
  simulation and reconstruction software.
\end{itemize}

It would be highly beneficial for the group to continue to function,
through the means of the forum proposed above, and to have at least one
meeting a year devoted to reviewing our activities and updating our
knowledge of best tools and practices.

\sloppy
\raggedright
\clearpage
\printbibliography[title={References},heading=bibintoc]

\end{document}

### 2018-01-16 CWP/papers/HSF-CWP-2017-01_roadmap/latex/authors.tex

Albrecht, Johannes$^{66}$;
Alves Jr, Antonio Augusto$^{79}$;
Amadio, Guilherme$^{5}$;
Anh-Ky, Nguyen$^{117}$;
Aphecetche, Laurent$^{63}$;
Apostolakis, John$^{5}$;
Asai, Makoto$^{61,p}$;
Atzori, Luca$^{5}$;
Babik, Marian$^{5}$;
Bagliesi, Giuseppe$^{31}$;
Bandieramonte, Marilena$^{5}$;
Barisits, Martin$^{5}$;
Bauerdick, Lothar A. T.$^{16,c}$;
Belforte, Stefano$^{34}$;
Benjamin, Douglas$^{80}$;
Bernius, Catrin$^{61}$;
Bhimji, Wahid$^{44}$;
Bianchi, Riccardo Maria$^{102}$;
Bird, Ian$^{5}$;
Biscarat, Catherine$^{50}$;
Blomer, Jakob$^{5}$;
Bloom, Kenneth$^{94}$;
Boccali, Tommaso$^{31}$;
Bockelman, Brian$^{94}$;
Bold, Tomasz$^{41}$;
Bonacorsi, Daniele$^{25}$;
Boveia, Antonio$^{98}$;
Bozzi, Concezio$^{27}$;
Bracko, Marko$^{91,39}$;
Britton, David$^{84}$;
Buckley, Andy$^{84}$;
Buncic, Predrag$^{5,a}$;
Calafiura, Paolo$^{44}$;
Campana, Simone$^{5,a}$;
Canal, Philippe$^{16,c}$;
Canali, Luca$^{5}$;
Carlino, Gianpaolo$^{30}$;
Castro, Nuno$^{45,93,d}$;
Cattaneo, Marco$^{5}$;
Cerminara, Gianluca$^{5}$;
Chang, Philip$^{73}$;
Chapman, John$^{74}$;
Chen, Gang$^{23}$;
Childers, Taylor$^{1}$;
Clarke, Peter$^{81}$;
Clemencic, Marco$^{5}$;
Cogneras, Eric$^{48}$;
Collier, Ian$^{59}$;
Corti, Gloria$^{5}$;
Cosmo, Gabriele$^{5}$;
Costanzo, Davide$^{108}$;
Couturier, Ben$^{5}$;
Cranmer, Kyle$^{55}$;
Cranshaw, Jack$^{1}$;
Cristella, Leonardo$^{26}$;
Crooks, David$^{84}$;
Crépé-Renaudin, Sabine$^{50}$;
Dallmeier-Tiessen, Sünje$^{5}$;
De, Kaushik$^{110}$;
De Cian, Michel$^{85}$;
Di Girolamo, Alessandro$^{5}$;
Dimitrov, Gancho$^{5}$;
Doglioni, Caterina$^{89,h}$;
Dotti, Andrea$^{61,p}$;
Duellmann, Dirk$^{5}$;
Duflot, Laurent$^{43}$;
Dykstra, Dave$^{16,c}$;
Dziedziniewicz-Wojcik, Katarzyna$^{5}$;
Dziurda, Agnieszka$^{5}$;
Egede, Ulrik$^{36}$;
Elmer, Peter$^{103,a}$;
Elmsheuser, Johannes$^{2}$;
Elvira, V. Daniel$^{16,c}$;
Eulisse, Giulio$^{5}$;
Ferber, Torben$^{71}$;
Filipcic, Andrej$^{39}$;
Fisk, Ian$^{62}$;
Fitzpatrick, Conor$^{14}$;
Flix, José$^{57,7,g}$;
Formica, Andrea$^{37}$;
Forti, Alessandra$^{90}$;
Frost, James$^{101}$;
Gaede, Frank$^{13}$;
Ganis, Gerardo$^{5}$;
Gardner, Robert$^{78}$;
Garonne, Vincent$^{99}$;
Gellrich, Andreas$^{13}$;
Genser, Krzysztof$^{16,c}$;
George, Simon$^{60}$;
Geurts, Frank$^{104}$;
Gheata, Andrei$^{5}$;
Gheata, Mihaela$^{5}$;
Giacomini, Francesco$^{9}$;
Giagu, Stefano$^{106,33}$;
Giffels, Manuel$^{40}$;
Gingrich, Douglas$^{67}$;
Girone, Maria$^{5}$;
Gligorov, Vladimir V.$^{49}$;
Glushkov, Ivan$^{110}$;
Gohn, Wesley$^{86}$;
Gonzalez Lopez, Jose Benito$^{5}$;
González Caballero, Isidro$^{100}$;
González Fernández, Juan R.$^{100}$;
Govi, Giacomo$^{16}$;
Grandi, Claudio$^{25}$;
Grasland, Hadrien$^{43}$;
Gray, Heather$^{44}$;
Grillo, Lucia$^{90}$;
Guan, Wen$^{115}$;
Gutsche, Oliver$^{16,c}$;
Gyurjyan, Vardan$^{38}$;
Hanushevsky, Andrew$^{61,p}$;
Hariri, Farah$^{5}$;
Hartmann, Thomas$^{13}$;
Harvey, John$^{5,a}$;
Hauth, Thomas$^{40}$;
Hegner, Benedikt$^{5,a}$;
Heinemann, Beate$^{13}$;
Heinrich, Lukas$^{55}$;
Hernández, José M.$^{7,g}$;
Hildreth, Michael$^{96,f}$;
Hodgkinson, Mark$^{108}$;
Hoeche, Stefan$^{61,p}$;
Hristov, Peter$^{5}$;
Huang, Xingtao$^{107}$;
Ivanchenko, Vladimir N.$^{5,111}$;
Ivanov, Todor$^{109}$;
Jashal, Brij$^{65}$;
Jayatilaka, Bodhitha$^{16,c}$;
Jones, Roger$^{87,a}$;
Jouvin, Michel$^{43,a}$;
Jun, Soon Yung$^{16,c}$;
Kagan, Michael$^{61,p}$;
Kalderon, Charles William$^{89}$;
Karavakis, Edward$^{5}$;
Katz, Daniel S.$^{77}$;
Kcira, Dorian$^{11}$;
Kersevan, Borut Paul$^{88}$;
Kirby, Michael$^{16,c}$;
Klimentov, Alexei$^{2}$;
Klute, Markus$^{51}$;
Komarov, Ilya$^{34,n}$;
Konstantinov, Dmitri$^{58}$;
Koppenburg, Patrick$^{54}$;
Kowalkowski, Jim$^{16,c}$;
Kreczko, Luke$^{70}$;
Kuhr, Thomas$^{47,a}$;
Kutschke, Robert$^{16,a,c}$;
Kuznetsov, Valentin$^{12}$;
Lampl, Walter$^{69}$;
Lancon, Eric$^{2}$;
Lange, David$^{103,a}$;
Lassnig, Mario$^{5}$;
Laycock, Paul$^{5}$;
Leggett, Charles$^{44}$;
Letts, James$^{73}$;
Lewendel, Birgit$^{13}$;
Li, Teng$^{81}$;
Lima, Guilherme$^{16}$;
Linacre, Jacob$^{59,m}$;
Linden, Tomas$^{18}$;
Lo Presti, Giuseppe$^{5}$;
Lopienski, Sebastian$^{5}$;
Love, Peter$^{87}$;
Magini, Nicolò$^{28}$;
Marshall, Zachary L$^{44}$;
Martelli, Edoardo$^{5}$;
Martin-Haugh, Stewart$^{59}$;
Mato, Pere$^{5}$;
Mazumdar, Kajari$^{65}$;
McCauley, Thomas$^{96}$;
McFayden, Josh$^{5}$;
McKee, Shawn$^{92,l}$;
McNab, Andrew$^{90}$;
Mehdiyev, Rashid$^{76}$;
Meinhard, Helge$^{5}$;
Menasce, Dario$^{29,a}$;
Mendez Lorenzo, Patricia$^{5}$;
Mete, Alaettin Serhan$^{72}$;
Michelotto, Michele$^{32}$;
Mitrevski, Jovan$^{47}$;
Moneta, Lorenzo$^{5}$;
Morgan, Ben$^{114}$;
Mount, Richard$^{61,p}$;
Moyse, Edward$^{68}$;
Murray, Sean$^{75,10}$;
Neubauer, Mark S$^{77,a,k}$;
Novaes, Sérgio$^{105}$;
Novak, Mihaly$^{5}$;
Oyanguren, Arantza$^{22}$;
Ozturk, Nurcan$^{110}$;
Pacheco Pages, Andres$^{57,20,j}$;
Paganini, Michela$^{116}$;
Pansanel, Jerome$^{35}$;
Pascuzzi, Vincent R.$^{112}$;
Patrick, Glenn$^{59}$;
Pearce, Alex$^{5}$;
Pearson, Ben$^{52}$;
Pedro, Kevin$^{16,c}$;
Perdue, Gabriel$^{16}$;
Perez-Calero Yzquierdo, Antonio$^{57,7,g}$;
Perrozzi, Luca$^{15}$;
Petersen, Troels$^{53}$;
Petric, Marko$^{5}$;
Piedra, Jónatan$^{21}$;
Piilonen, Leo$^{118,i}$;
Piparo, Danilo$^{5}$;
Pokorski, Witold$^{5}$;
Polci, Francesco$^{49}$;
Potamianos, Karolos$^{13}$;
Psihas, Fernanda$^{24}$;
Raven, Gerhard$^{54}$;
Reuter, Jürgen$^{13}$;
Ribon, Alberto$^{5}$;
Rinaldi, Lorenzo$^{25}$;
Ritter, Martin$^{47}$;
Robinson, James$^{13}$;
Rodrigues, Eduardo$^{79,a,e}$;
Roiser, Stefan$^{5,a}$;
Rousseau, David$^{43}$;
Roy, Gareth$^{84}$;
Sailer, Andre$^{5}$;
Sakuma, Tai$^{70}$;
Santana, Renato$^{3}$;
Sartirana, Andrea$^{46}$;
Schellman, Heidi$^{56}$;
Schovancová, Jaroslava$^{5}$;
Schramm, Steven$^{83}$;
Schulz, Markus$^{5}$;
Sciabà, Andrea$^{5}$;
Seidel, Sally$^{95}$;
Sekmen, Sezen$^{42}$;
Serfon, Cedric$^{99}$;
Severini, Horst$^{97}$;
Sexton-Kennedy, Elizabeth$^{16,a,c}$;
Seymour, Michael$^{90}$;
Shapoval, Illya$^{44}$;
Shiers, Jamie$^{5}$;
Shiu, Jing-Ge$^{64}$;
Short, Hannah$^{5}$;
Siroli, Gian Piero$^{25}$;
Skipsey, Sam$^{84}$;
Smith, Tim$^{5}$;
Snyder, Scott$^{2}$;
Sokoloff, Michael D$^{79,a}$;
Stadie, Hartmut$^{17}$;
Stark, Giordon$^{6}$;
Stewart, Gordon$^{84}$;
Stewart, Graeme$^{5,a}$;
Sánchez, Arturo$^{113,19}$;
Sánchez-Hernández, Alberto$^{8,o}$;
Taffard, Anyes$^{72}$;
Templon, Jeff$^{54}$;
Tenaglia, Giacomo$^{5}$;
Tsulaia, Vakhtang$^{44}$;
Tunnell, Christopher$^{6}$;
Vaandering, Eric$^{16,c}$;
Valassi, Andrea$^{5}$;
Vallecorsa, Sofia$^{82}$;
Valsan, Liviu$^{5}$;
Van Gemmeren, Peter$^{1}$;
Vernet, Renaud$^{4}$;
Viren, Brett$^{2}$;
Vlimant, Jean-Roch$^{11,a}$;
Voss, Christian$^{13}$;
Vuosalo, Carl$^{115}$;
Vázquez Sierra, Carlos$^{54}$;
Wartel, Romain$^{5}$;
Wenaus, Torre$^{2}$;
Wenzel, Sandro$^{5}$;
Winklmeier, Frank$^{56}$;
Wissing, Christoph$^{13}$;
Wuerthwein, Frank$^{73}$;
Wynne, Benjamin$^{81}$;
Xiaomei, Zhang$^{23}$;
Yang, Wei$^{61,p}$;
Yazgan, Efe$^{23}$
\bigskip
\par {\footnotesize $^{1}$ High Energy Physics Division, Argonne National Laboratory, Argonne IL, USA}
\par {\footnotesize $^{2}$ Physics Department, Brookhaven National Laboratory, Upton NY, USA}
\par {\footnotesize $^{3}$ Centro Brasileiro de Pesquisas Físicas, Rio de Janeiro, Brazil}
\par {\footnotesize $^{4}$ Centre de Calcul de l’IN2P3, Villeurbanne, Lyon, France}
\par {\footnotesize $^{5}$ CERN, Geneva, Switzerland}
\par {\footnotesize $^{6}$ Enrico Fermi Institute, University of Chicago, Chicago IL, USA}
\par {\footnotesize $^{7}$ Centro de Investigaciones Energéticas Medioambientales y Tecnológicas (CIEMAT), Madrid, Spain}
\par {\footnotesize $^{8}$ Cinvestav, Mexico City, Mexico}
\par {\footnotesize $^{9}$ Centro Nazionale Analisi Fotogrammi (CNAF), INFN, Bologna, Italy}
\par {\footnotesize $^{10}$ Center for High Performance Computing, Cape Town, South Africa}
\par {\footnotesize $^{11}$ California Institute of Technology, Pasadena, California, USA}
\par {\footnotesize $^{12}$ Cornell University, Ithaca, USA}
\par {\footnotesize $^{13}$ Deutsches Elektronen-Synchrotron, Hamburg, Germany}
\par {\footnotesize $^{14}$ Institute of Physics, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland}
\par {\footnotesize $^{15}$ ETH Zurich - Institute for Particle Physics and Astrophysics (IPA), Zurich, Switzerland}
\par {\footnotesize $^{16}$ Fermi National Accelerator Laboratory, Batavia, USA}
\par {\footnotesize $^{17}$ University of Hamburg, Hamburg, Germany}
\par {\footnotesize $^{18}$ Helsinki Institute of Physics, Helsinki, Finland}
\par {\footnotesize $^{19}$ International Center for Theoretical Physics, Trieste, Italy}
\par {\footnotesize $^{20}$ Institut de Física d’Altes Energies and Departament de Física de la Universitat Autònoma de Barcelona and ICREA, Barcelona, Spain}
\par {\footnotesize $^{21}$ Instituto de Física de Cantabria (IFCA), CSIC-Universidad de Cantabria, Santander, Spain}
\par {\footnotesize $^{22}$ Instituto de Física Corpuscular, Centro Mixto Universidad de Valencia - CSIC, Valencia, Spain}
\par {\footnotesize $^{23}$ Institute of High Energy Physics, Chinese Academy of Sciences, Beijing}
\par {\footnotesize $^{24}$ Department of Physics, Indiana University, Bloomington IN, USA}
\par {\footnotesize $^{25}$ INFN Sezione di Bologna, Università di Bologna, Bologna, Italy}
\par {\footnotesize $^{26}$ INFN Sezione di Bari, Università di Bari, Politecnico di Bari, Bari, Italy}
\par {\footnotesize $^{27}$ Università e INFN, Ferrara, Ferrara, Italy}
\par {\footnotesize $^{28}$ INFN Sezione di Genova, Genova, Italy}
\par {\footnotesize $^{29}$ INFN Sezione di Milano-Bicocca, Milano, Italy}
\par {\footnotesize $^{30}$ INFN Sezione di Napoli, Università di Napoli, Napoli, Italy}
\par {\footnotesize $^{31}$ INFN Sezione di Pisa, Università di Pisa, Scuola Normale Superiore di Pisa, Pisa, Italy}
\par {\footnotesize $^{32}$ INFN Sezione di Padova, Università di Padova b, Padova, Italy}
\par {\footnotesize $^{33}$ INFN Sezione di Roma I, Università La Sapienza, Roma, Italy}
\par {\footnotesize $^{34}$ INFN Sezione di Trieste, Università di Trieste, Trieste, Italy}
\par {\footnotesize $^{35}$ Université de Strasbourg, CNRS, IPHC UMR 7178, F-67000 Strasbourg, France}
\par {\footnotesize $^{36}$ Imperial College London, London, United Kingdom}
\par {\footnotesize $^{37}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat à l’Énergie Atomique), Gif-sur-Yvette, France}
\par {\footnotesize $^{38}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
\par {\footnotesize $^{39}$ Jožef Stefan Institute, Ljubljana, Slovenia}
\par {\footnotesize $^{40}$ Karlsruhe Institute of Technology, Karlsruhe, Germany}
\par {\footnotesize $^{41}$ AGH University of Science and Technology, Faculty of Physics and Applied Computer Science, Krakow, Poland}
\par {\footnotesize $^{42}$ Kyungpook National University, Daegu, Republic of Korea}
\par {\footnotesize $^{43}$ LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\par {\footnotesize $^{44}$ Lawrence Berkeley National Laboratory and University of California, Berkeley CA, USA}
\par {\footnotesize $^{45}$ Laboratório de Instrumentação e Física Experimental de Partículas (LIP), Lisboa, Portugal}
\par {\footnotesize $^{46}$ Laboratoire Leprince-Ringuet, École Polytechnique, CNRS/IN2P3, Université Paris-Saclay, Palaiseau, France}
\par {\footnotesize $^{47}$ Fakultät für Physik, Ludwig-Maximilians-Universität München, München, Germany}
\par {\footnotesize $^{48}$ Laboratoire de Physique Corpusculaire, Clermont Université and Université Blaise Pascal and CNRS/IN2P3, Clermont-Ferrand, France}
\par {\footnotesize $^{49}$ LPNHE, Université Pierre et Marie Curie, Université Paris Diderot, CNRS/IN2P3, Paris, France}
\par {\footnotesize $^{50}$ Laboratoire de Physique Subatomique et de Cosmologie, Université Joseph Fourier and CNRS/IN2P3 and Institut National Polytechnique de Grenoble, Grenoble, France}
\par {\footnotesize $^{51}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
\par {\footnotesize $^{52}$ Max-Planck-Institut für Physik (Werner-Heisenberg-Institut), München, Germany}
\par {\footnotesize $^{53}$ Niels Bohr Institute, University of Copenhagen, Kobenhavn, Denmark}
\par {\footnotesize $^{54}$ Nikhef National Institute for Subatomic Physics and University of Amsterdam, Amsterdam, Netherlands}
\par {\footnotesize $^{55}$ Department of Physics, New York University, New York, NY, USA}
\par {\footnotesize $^{56}$ Center for High Energy Physics, University of Oregon, Eugene OR, USA}
\par {\footnotesize $^{57}$ Port d’Informació Científica (PIC), Universitat Autònoma de Barcelona (UAB), Barcelona, Spain}
\par {\footnotesize $^{58}$ High Energy Physics (IHEP), Protvino, Russia}
\par {\footnotesize $^{59}$ STFC Rutherford Appleton Laboratory, Didcot, United Kingdom}
\par {\footnotesize $^{60}$ Department of Physics, Royal Holloway University of London, Surrey, United Kingdom}
\par {\footnotesize $^{61}$ SLAC National Accelerator Laboratory, Menlo Park CA, USA}
\par {\footnotesize $^{62}$ Simons Foundation, New York, USA}
\par {\footnotesize $^{63}$ SUBATECH, IMT Atlantique, Université de Nantes, CNRS-IN2P3, Nantes, France}
\par {\footnotesize $^{64}$ National Taiwan University, Taipei, Taiwan}
\par {\footnotesize $^{65}$ Tata Institute of Fundamental Research, Mumbai, India}
\par {\footnotesize $^{66}$ Technische Universit\"at Dortmund, Dortmund, Germany}
\par {\footnotesize $^{67}$ Department of Physics, University of Alberta, Edmonton, AB, Canada}
\par {\footnotesize $^{68}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
\par {\footnotesize $^{69}$ Department of Physics, University of Arizona, Tucson, AZ, USA}
\par {\footnotesize $^{70}$ H.H. Wills Physics Laboratory, University of Bristol, Bristol, United Kingdom}
\par {\footnotesize $^{71}$ Department of Physics, University of British Columbia, Vancouver, BC, Canada}
\par {\footnotesize $^{72}$ Department of Physics and Astronomy, University of California Irvine, Irvine, CA, USA}
\par {\footnotesize $^{73}$ University of California, San Diego, La Jolla, USA}
\par {\footnotesize $^{74}$ Cavendish Laboratory, University of Cambridge, Cambridge, United Kingdom}
\par {\footnotesize $^{75}$ Physics Department, University of Cape Town, Cape Town, South Africa}
\par {\footnotesize $^{76}$ Carleton University, Ottawa, ON, Canada}
\par {\footnotesize $^{77}$ University of Illinois Urbana-Champaign, Champaign, Illinois, USA}
\par {\footnotesize $^{78}$ Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
\par {\footnotesize $^{79}$ University of Cincinnati, Cincinnati, OH, USA}
\par {\footnotesize $^{80}$ Department of Physics, Duke University, Durham, NC, USA}
\par {\footnotesize $^{81}$ SUPA - School of Physics and Astronomy, University of Edinburgh, Edinburgh, United Kingdom}
\par {\footnotesize $^{82}$ Gangneung-Wonju National University, South Korea}
\par {\footnotesize $^{83}$ Section de Physique, Université de Genève, Geneva, Switzerland}
\par {\footnotesize $^{84}$ SUPA - School of Physics and Astronomy, University of Glasgow, Glasgow, United Kingdom}
\par {\footnotesize $^{85}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
\par {\footnotesize $^{86}$ Department of Physics and Astronomy, University of Kentucky, Lexington, USA}
\par {\footnotesize $^{87}$ Physics Department, Lancaster University, Lancaster, United Kingdom}
\par {\footnotesize $^{88}$ Department of Physics, Jožef Stefan Institute and University of Ljubljana, Ljubljana, Slovenia}
\par {\footnotesize $^{89}$ Fysiska institutionen, Lunds Universitet, Lund, Sweden}
\par {\footnotesize $^{90}$ School of Physics and Astronomy, University of Manchester, Manchester, United Kingdom}
\par {\footnotesize $^{91}$ University of Maribor, Ljubljana, Slovenia}
\par {\footnotesize $^{92}$ Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
\par {\footnotesize $^{93}$ Departamento de Física, Universidade do Minho, Braga, Portugal}
\par {\footnotesize $^{94}$ University of Nebraska-Lincoln, Lincoln, USA}
\par {\footnotesize $^{95}$ Department of Physics and Astronomy, University of New Mexico, Albuquerque NM, USA}
\par {\footnotesize $^{96}$ University of Notre Dame, Notre Dame, USA}
\par {\footnotesize $^{97}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman OK, USA}
\par {\footnotesize $^{98}$ The Ohio State University, Columbus, USA}
\par {\footnotesize $^{99}$ Department of Physics, University of Oslo, Oslo, Norway}
\par {\footnotesize $^{100}$ Universidad de Oviedo, Oviedo, Spain}
\par {\footnotesize $^{101}$ Department of Physics, Oxford University, Oxford, UK}
\par {\footnotesize $^{102}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
\par {\footnotesize $^{103}$ Princeton University, Princeton, USA}
\par {\footnotesize $^{104}$ Rice University, Houston, TX, USA}
\par {\footnotesize $^{105}$ Universidade Estadual Paulista, São Paulo, Brazil}
\par {\footnotesize $^{106}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
\par {\footnotesize $^{107}$ School of Physics, Shandong University, Shandong, China}
\par {\footnotesize $^{108}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
\par {\footnotesize $^{109}$ University of Sofia, Sofia, Bulgaria}
\par {\footnotesize $^{110}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
\par {\footnotesize $^{111}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
\par {\footnotesize $^{112}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
\par {\footnotesize $^{113}$ Dipartimento di Chimica, Fisica e Ambiente, Università di Udine, Udine, Italy}
\par {\footnotesize $^{114}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
\par {\footnotesize $^{115}$ University of Wisconsin - Madison, Madison, WI, USA}
\par {\footnotesize $^{116}$ Department of Physics, Yale University, New Haven, CT, USA}
\par {\footnotesize $^{117}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
\par {\footnotesize $^{118}$ Virginia Tech, Blacksburg, Virginia, USA}
\bigskip
\par {\footnotesize $^{a}$ Community White Paper Editorial Board Member}
\par {\footnotesize $^{b}$ Vladimir V. Gligorov acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 724777 “RECEPT”}
\par {\footnotesize $^{c}$ Supported by the US-DOE, DE-AC02-07CH11359}
\par {\footnotesize $^{d}$ Supported by FCT-Portugal, IF/00050/2013/CP1172/CT0002}
\par {\footnotesize $^{e}$ Supported by the US-NSF, ACI-1450319}
\par {\footnotesize $^{f}$ Supported by the US-NSF, PHY-1607578}
\par {\footnotesize $^{g}$ Supported by ES-MINECO, FPA2016-80994-c2-1-R \& MDM-2015-0509}
\par {\footnotesize $^{h}$ Caterina Doglioni acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 679305 “DARKJETS”}
\par {\footnotesize $^{i}$ Supported by the US-DOE, DE-SC0009973}
\par {\footnotesize $^{j}$ Supported by the ES-MINECO FPA2016-80994-C2-2-R}
\par {\footnotesize $^{k}$ Supported by the US-DOE, DE-SC0018098 and US-NSF ACI-1558233}
\par {\footnotesize $^{l}$ Supported by the US-DOE, DE-SC0007859 and US-NSF, 76749/1136652/2}
\par {\footnotesize $^{m}$ Supported by funding from the European Union’s Horizon 2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement No. 752730}
\par {\footnotesize $^{n}$ Supported by Swiss National Science Foundation, Early Postdoc Mobility Fellowship, project number P2ELP2\_168556}
\par {\footnotesize $^{o}$ Supported by CONACYT (Mexico)}
\par {\footnotesize $^{p}$ Supported by the US-DOE, DE-AC02-76SF0051}

### 2018-01-16 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp.tex

--- 2017-12-17 CWP/papers/roadmap/latex/cwp.tex
@@
 \item
-  Etudy the decays of $b$- and $c$-hadrons, and tau leptons, in the search
+  Study the decays of $b$- and $c$-hadrons, and tau leptons, in the search
   for manifestations of BSM physics, and investigate

### 2018-01-22 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
First import of Detector Simulation paper

% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.

% JHEP preprint template
\documentclass[12pt,a4paper]{article}
\usepackage{jheppub}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xspace}

\usepackage[utf8]{inputenc}

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{../../roadmap/latex/cwp.bib}
\addbibresource{../../roadmap/latex/cwp-chapters.bib}

%% Comment out before final submission
%\usepackage{lineno}  % for line numbering during review
%\linenumbers

\abstract{The High Energy Phyiscs community has developed and needs to maintain
many tens of millions of lines of code and to integrate effectively the
work of thousands of developers across large collaborations. 
Software needs to be built, validated, and deployed across hundreds of
sites. Software also has a lifetime of many years, frequently beyond that
of the original developer, it must be developed with sustainability in
mind. Adequate recognition of software development as a critical task in the HEP
community needs to be fostered and an appropriate publication and citation
strategy needs to be developed. As part of the HEP Softare Foundation's
Community White Paper process a working group on Software Development,
Deployment and Validation was formed to examine all of these
issues, identify best practice and to formulare recommendations for the next
decade. Its report is presented here.}

\begin{document}

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-13 \\
 & & December 21, 2017 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{HEP Software Foundation Community White Paper Working Group -- Software
Development, Deployment and Validation}

\author{HEP Software Foundation:}
\author[a]{Benjamin Couturier}
\author[a,1]{Giulio Eulisse}
\author[b]{Hadrien Grasland}
\author[a]{Benedikt Hegner}
\author[b]{Michel Jouvin}
\author[c]{Meghan Kane}
\author[d]{Daniel S. Katz}
\author[e]{Thomas Kuhr}
\author[f]{David Lange}
\author[a,1]{Patricia Mendez Lorenzo}
\author[e]{Martin Ritter}
\author[a,1]{Graeme Andrew Stewart}
\author[a]{Andrea Valassi}

\affiliation[a]{CERN, Geneva, Switzerland}
\affiliation[b]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\affiliation[c]{Soundcloud, Berlin, Germany}
\affiliation[d]{University of Illinois Urbana-Champaign, Champaign, Illinois,
USA}
\affiliation[e]{Fakultät für Physik, Ludwig-Maximilians-Universität München,
München, Germany}
\affiliation[f]{Princeton University, Princeton, USA}
\affiliation[1]{Paper Editor}

\maketitle

\newpage

\hypertarget{introduction}{%
\section{1 Introduction}\label{introduction}}

Detector simulation is an essential tool to design, build and commission
the highly sophisticated detectors utilised in modern particle physics
experiments and cosmology. It is also a fundamental tool for data
analysis and interpretation of the physics measurements. The typical
simulation software application in high-energy physics (HEP) experiments
consists of a set of packages executed in sequence, usually starting
with a generator of the physics processes under study. In particle
colliders, event generators provide the final state particles observed
as primary particles in the detector. A second module simulates the
passage of the generated primary particles through the detector material
and its magnetic field. In most contemporary experiments, this detector
simulation module is based on the Geant4 simulation toolkit, a software
package that provides the tools to describe the detector geometry and
materials, and incorporates a number of models to simulate
electromagnetic and hadronic interactions of particles with matter~\cite{ALLISON2016186}. 
When appropriate, various fast simulation options can be used
to save computing time with some penalty on physics accuracy. The next
module, often called digitisation, simulates the readout process,
including the signal processing by the detector electronics, as well as
the calibration of individual channels. At the end of the chain, the
same algorithms used to identify and reconstruct individual particles
and physics observables in real data are applied to simulated events.

In the case of accelerator-based neutrino experiments, the chain
typically starts with a Geant4 module to model the interaction of a
proton beam with a target to produce a neutrino beam which is driven to
a detector where the physics interaction of interest occurs. The
neutrino beam and the first neutrino-nucleus interaction in the detector
is typically modeled with the GENIE {[}7{]} package, while Geant4 is
utilised in a second step to simulate the nucleus de-excitation and
subsequent hadron production. Readout modeling, calibration, and
reconstruction is done as described for collider experiments.

During the last three decades, simulation has proven to be of critical
importance to the success of HEP experimental programmes. For example,
the detailed detector modeling and accurate physics of the CERN Large
Hadron Collider (LHC) CMS and ATLAS experiments Geant4-based simulation
software helped these collaborations deliver physics results of
outstanding quality faster than any previous hadron collider experiment.
The FLUKA simulation package has also been used extensively for studies
of shielding and radiation modeling, providing important input to the
design of the original LHC experiments and their upgrades. Simulation
software at the LHC experiments is more accurate and yet runs much
faster than its predecessor at the Tevatron. Simulation samples of
better quality and in larger quantities, evolving detector and computing
technology, and a wealth of experience from pre-LHC experiments on
calibration and data analysis techniques, improved significantly the
precision of the measurements in the current generation of experiments.
In parallel with this, simulation has been used to design a new
generation of neutrino and muon experiments and in searches for beyond
the standard model (BSM) physics in precision experiments.

For all its success so far, the challenges faced by the HEP field in the
simulation domain are daunting. During the first two runs, the LHC
experiments produced, reconstructed, stored, transferred, and analyzed
tens of billions of simulated events. This effort required more than
half of the total computing resources allocated to the experiments. As
part of the high-luminosity LHC physics program (HL-LHC) and through the
end of the 2030's, the upgraded experiments expect to collect 150 times
more data than in Run 1. The 50 PB of raw data produced in 2016 will
grow to approximately 600 PB in 2026 while the CPU needs will increase
by a factor of about 60. Demand for larger simulation samples to satisfy
analysis needs will grow accordingly. In addition, simulation tools have
to serve diverse communities, including accelerator-based particle
physics research utilizing proton-proton colliders, neutrino and muon
experiments, as well as the cosmic frontier. The complex detectors of
the future, with different module- or cell-level shapes, finer
segmentation, and novel materials and detection techniques, require
additional features in geometry tools and bring new demands on physics
coverage and accuracy within the constraints of the available computing
budget. More extensive use of Fast Simulation is a potential solution,
under the assumption that it is possible to improve time performance
without an unacceptable loss of physics accuracy.

The effort to improve detector simulation and modeling tools including
the accuracy of their physics, whilst saving memory allocation and
computing execution time, requires immediate attention. This report
addresses the future development of Full Simulation and Fast Simulation
applications of HEP experiments for detectors and beam lines with a view
to achieve improvements in software efficiency, scalability and
performance, and to make use of the advances in processor, storage and
network technologies. A plan must consider new approaches to computing
and software that could radically extend the physics reach of the
detectors, while ensuring the long-term sustainability of the software
through the lifetime of the experiments running through the 2020's and
2030's. One example is the GeantV {[}2{]} R\&D project, started in 2013.
Another example is the ongoing effort to port simulation code to run on
High Performance Computing (HPC) facilities, which provide massive CPU
power and modern computing architectures.

Components of the HEP simulation chain, such as physics generators and
event visualisation tools are discussed in other chapters, while the
modeling of the propagation and beam manipulation through accelerator
lattices (i.e. CERN and Fermilab accelerator complexes) are considered
to be outside the scope of this chapter. Space and medical applications
are referred to briefly where appropriate.

\hypertarget{section}{%
\section{}\label{section}}

\textbf{Draft executive summary document}:
\href{https://goo.gl/FHaAtm}{{https://goo.gl/FHaAtm}}

\hypertarget{challenges}{%
\section{2 Challenges}\label{challenges}}

The experimental programmes planned in the next decade are driving
developments in the simulation domain; they include the High Luminosity
LHC project (HL-LHC), neutrino and muon experiments, and studies towards
future colliders such as the Future Circular Collider (FCC) and the
Compact Linear Collider (CLIC). The requirement of improving precision
in the simulation implies production of larger Monte Carlo samples,
which scale with the number of real events recorded in future
experiments, and this places an additional burden on the computing
resources that will be needed to generate them. The diversification of
the physics programmes also requires new and improved physics models.

The main challenges to be addressed if the required physics and software
performance goals are to be achieved can be summarised as follows:

\begin{itemize}
\item
  reviewing the physics models assumptions, approximations and
  limitations in order to achieve higher precision, and to extend the
  validity of models up to energies of the order of 100 TeV foreseen
  with the Future Circular Collider (FCC) project;
\item
  redesigning, developing, commissioning detector simulation toolkits to
  be more efficient when executed on current vector CPUs and emerging
  new architectures, including GPGPUs where use of SIMD vectorisation is
  vital; this includes porting and optimising the experiments'
  simulation applications to allow exploitation of large HPC facilities;
\item
  exploring different Fast Simulation options, where the full detector
  simulation is replaced, in whole or in part, by computationally
  efficient techniques; an area of investigation is common frameworks
  for fast tuning and validation;
\item
  developing, improving, and optimising geometry tools that can be
  shared among experiments to make the modeling of complex detectors
  computationally more efficient, modular, and transparent;
\item
  developing techniques for background modeling, including contributions
  of multiple hard interactions overlapping the event of interest in
  collider experiments (pile-up);
\item
  revisiting digitisation algorithms to improve performance and
  exploring opportunities for code sharing among experiments;
\item
  recruiting, training, retaining, human resources in all areas of
  expertise pertaining to the simulation domain, including software and
  physics.
\end{itemize}

The complexity and diversity of future experiments results in an
enormous demand for CPU, and detailed detector simulation already
consumes the plurality of CPU resources of most current HEP experiments.
Some progress has already been made in improving software performance,
for example by adding multithreading capability to Geant4, and by
redesigning the geometry library into the VecGeom package that exploits
vectorisation. These R\&D programmes need to continue in order to
benefit from the various forms of parallelism offered by modern
architectures. For example, the goal of the GeantV R\&D project is to
build a realistic prototype application for a complex experiment with
the aim of demonstrating a big improvement in CPU time. The main
architectural feature of this prototype is the implementation of
particle-level, i.e. track-level, parallelisation for more efficient use
of SIMD vectorisation and data locality. Whenever possible, new
developments resulting from the R\&D programme will be tested in
realistic prototypes and then be integrated, validated, and deployed in
a timely fashion in Geant4 in order to maximise the benefits to
experiments at the earliest opportunity.

A flexible simulation strategy is needed that is capable of delivering a
broad spectrum of options offering different levels of physics accuracy,
from a pure Full Simulation approach to a combination of Geant4 showers,
parametrised response functions, and shower libraries. R\&D in the realm
of Fast Simulation includes the use of Machine Learning (ML) techniques
and embedded tools for tuning and validation in an effort to provide
faster and more accurate physics and to be able to safely expand the use
of Fast Simulation in the context of limited computing resources.

HPC facilities offer a valuable resource and work must continue to adapt
simulation applications to run efficiently on these systems, which may
produce a significant fraction of the simulation samples in future HEP
experiments.

Work also needs to continue to improve the accuracy of the physics
models, and the validity of the models needs to be verified for new
detectors that include, for example, liquid argon time projection
chambers (for DUNE) and silicon-absorber calorimeters (for CMS). The
models need to be improved and validated for a large variety of incident
particles and energy ranges, including muons, neutrinos, and the high
energies accessible in a Future Circular Collider (FCC).

In producing new models of physics interactions and improving the
fidelity of the models that exist, it is absolutely imperative that
high-quality data is available. Simulation model tuning often relies on
test beam data, and a program to improve the library of available data
could be invaluable to the community. Such data would ideally include
both thin-target test beams for improving interaction models and
calorimeter targets for improving shower models. These data could
potentially be used for directly tuning fast simulation models, as well.
At the least it is important that the experiments carefully consider
what measurements could be made that would be useful input to the
community of Monte Carlo program developers and tuners.

High pile-up simulation for the high-luminosity environment in future
data-taking at the LHC poses challenges in memory utilisation and data
transfer (I/O) for generating the large statistics samples needed to
compare with data from high trigger rates. It currently necessitates the
generation and handling of large minimum bias samples in order to
achieve an accurate description of high pile-up collision events. It
would be desirable to have tools that allow overlay of both simulated
and real data events. The latter poses a real challenge related to the
collection of unsuppressed zero-bias data samples that reflect the
luminosity profile of the physics run.

The digitisation component of the simulation chain offers opportunities
for parallelisation, given that the code between sub-detector components
is typically orthogonal. In the case of neutrino detectors based on
liquid argon technology, this step is computationally expensive, as
compared to other elements of the chain. Since digitisation is often
either memory- or I/O-intensive, there is a challenge to ensure that
parallelism is efficient. Another issue is that digitisation code,
including containers, data types and data structures, is typically
developed by the user. In order to exploit the SIMD opportunities
intrinsic to the digitisation process, digits must be constrained to a
subset of the possible data structures, with priority given to data
contiguity.

New areas of commonality across experiments need to be explored, for
example in the domains of Fast Simulation and geometry. A great deal of
common code is relied upon by many users for the benefit of the whole
HEP community for the long term. It will be important to identify these
pieces of code, remove the experiment dependency where appropriate, and
ensure that the pieces that cannot be removed are either going to be
maintained by a group that feels responsibility for them or will not
become a bottleneck for any detector simulation in the future.

The large reliance on simulations implies the need for careful
assessment of the systematic uncertainties associated to Geant4
predictions. New capabilities and tools are being developed to allow for
the calculation of systematic uncertainties on the physics observables
predicted by the Geant4 beam and detector simulations. In other words,
kinematic distributions, efficiencies, and calibration constants will be
predicted with systematic uncertainties resulting from the propagation
of the uncertainties associated with the individual Geant4 physics
models.

There are specific challenges associated with the Intensity Frontier
experimental programme. Properly simulating a neutrino experiment like
NOvA {[}3{]}, MicroBooNE{[}4{]}, MINERvA{[}5{]}, or DUNE {[}6{]}
involves a three-part software stack, the first and last of which are
relevant to this paper; the second, event generators such as GENIE
{[}7{]}, is discussed elsewhere. The first element of the stack concerns
the beamline and the neutrino flux prediction. Estimating the neutrino
flux is a notorious problem because as weakly interacting particles,
neutrinos offer no independent mechanisms for measuring the flux. A
central problem is in the difficulty of accurately simulating the
hadronic physics of meson production, especially in a thick target.
Additionally, the simulated neutrino flux depends very strongly on
minute details of the beamline, meaning that neutrino beam kinematic
distributions must be described in great detail. Geant4 is the most
commonly used software toolkit for this work, although FLUKA {[}8{]} and
MARS {[}9{]} are also used. The final element of the stack is the
detector simulation. Again there are numerous challenges in terms of
efficiently simulating large numbers of events inside complex detector
geometries and in handling subtle physics effects in the traversal of
radiation across matter. Neutrino experiments in particular rely heavily
on detector simulations to reconstruct neutrino energy, which requires
accurate modeling of energy deposition by a variety of particles across
a range of energies. Geant4 is the universal solution at this stage.

Muon experiments such as Muon g-2 and Mu2e also face large simulation
challenges. Because they are searching for extremely rare effects, they
must grapple with very low signal to background ratios, and the modeling
of low cross-section background processes. Often the physics of these
processes is not well understood, introducing large systematic
uncertainties. Additionally, the size of the computational problem is a
serious challenge, as large simulation runs are required to adequately
sample all relevant areas of experimental phase space, even when
techniques to minimise the required computations are used. Another
aspect related to the background events is the need to simulate the
effects of low energy neutrons, which requires large computational
resources. For example, Mu2e used a total of approximately 60 million
CPU hours, or 83 thousand core months, in a 2016 simulation campaign to
produce the Full Simulation sample needed to complete a technical design
report delivered in 2016. Geant4 is the primary simulation toolkit for
all stages of these experiments. For comparison, the CMS experiment
spent 860 thousand core months on simulation in the May 2015-May 2016
period.

Although the computing needs of the intensity frontier are currently
small compared to the LHC experiments, they are expected to grow
substantially in the next decade, and it is imperative that the
intensity frontier experiments make use of modern computing
architectures and algorithms. This transition will be complicated by the
fact that the intensity frontier is fractured into many small
experiments, some of which have as few as 50 collaborators, limiting the
manpower available for updating detector simulations to use new tools.
Efforts such as LArSoft{[}7{]} have recognised this challenge and aim to
pool resources of many experiments for the benefit of all.

\hypertarget{geometry}{%
\section{3 Geometry}\label{geometry}}

The geometry modeler is a key component in Monte Carlo detector
simulation, combining both solids modelling and efficient navigation
techniques. Stringent requirements apply to provide the right level of
flexibility, robustness and accuracy in modeling from the simplest to
the most complex setups and to offer adequate precision and efficiency
in the implemented algorithms. Geant4 {[}1{]} provides a wide variety of
tools and solutions for describing geometry setups from simple to highly
complex. Geometrical models of the LHC detectors, for instance, easily
reach millions of geometrical elements of different kinds combined
together in hierarchical structures. The geometry modeler provides
techniques by which memory consumption can be greatly reduced, allowing
regular or irregular patterns to be easily replicated and assembled.
This, combined with navigation and optimisation algorithms, allows the
efficient geometry computation of the simulated tracks with the elements
composing any geometry setup.

Recent extensions of the geometry modeler in Geant4 include specialised
navigation techniques and optimisation algorithms to aid also in medical
simulation studies. This has allowed complex geometrical models of the
human body to be developed. Extensions also include parallel navigation
and tracking in layered geometries which allow geometry setups to be
superimposed on one another with minimal impact on CPU time (Figure 1).
An important and rather useful construct for shapes delimited by any
kind of complex surface is offered by the tessellated-solid construct,
which allows complex geometrical shapes to be described by approximating
their surfaces as a set of planar facets (triangles), with tuneable
resolution. This technique is used nowadays for importing geometries
from CAD systems to generate surface-bounded solids. Recent developments
introduced in Geant4 along with the implementation of the Unified Solids
library {[}10{]} provide considerably improved CPU performance, making
it possible to use such constructs for very detailed and realistic
descriptions of surfaces, while optimally scaling with the number of
facets in use. Further improvements to the tessellated solid construct
are expected as part of the rewrite of most geometry primitives going on
in the VecGeom {[}12{]} package, initiated within the GeantV project
{[}2{]}.

The Unified Solids Library started as an AIDA project in 2011, with the
aim to unify and improve the algorithms provided by the Geant4 and ROOT
{[}11{]} approaches to solids modeling. The library which comes with a
large collection of primitives, has been extended and enhanced as part
of the AIDA-2020 project, and is now integrated in VecGeom, which
provides an independent geometry modeler library for use as alternative
replacement in either Geant4 or ROOT. In VecGeom, the interfaces to
geometrical primitives have been extended to support vector signatures,
which are essential for GeantV, and allow for parallel processing of
queries; algorithms are completely reviewed and in good part rewritten
to make use of SIMD instructions wherever possible, and support
different kernels and architectures, including accelerators such as
Intel Xeon Phi® and GPU devices. The code has been reengineered to make
use of C++ template techniques, allowing for more modular implementation
and for specializing efficiently on different topologies.

The benefits in CPU time are clearly visible with ray-tracing scalar
tests and also measurable in terms of few percent when full physics and
electromagnetic field integration is applied.

The VecGeom geometry modeler includes a highly efficient navigation
system. It has been developed and optimised to exploit vectorised
transport on CPU, GPGPUs and Intel Xeon Phi, making use of the full
potential of the vectorised features of the library. It is planned in
the near future to provide hooks within both Geant4 and ROOT for using
the navigation system of VecGeom, eventually boosting speed further also
for the scalar mode.

\includegraphics[width=6.26772in,height=4.76389in]{media/image1.png}

\textbf{Figure 1} - An interactive Geant4 example featuring parallel
geometries and scoring with VecGeom primitives and parameterisations

Geant4 offers the ability to import and export Detector geometrical
descriptions from text files according to the Geometry Description
Markup Language (GDML) {[}13{]} based on XML; the exchange format is
also supported by ROOT and the schema includes all the geometrical
primitives currently in use. GDML supports the modularisation of
geometry descriptions to multiple GDML files, allowing for rational
organisation of the modules for complex setups. Verification of the GDML
file against the latest version of the schema comes for free thanks to
Xerces-C++ {[}14{]}, with the possibility to turn it on or off in the
Geant4 GDML parser.

The DD4hep Geometry Toolkit {[}15{]}, being developed in the context of
the AIDA and AIDA-2020 projects aims to provide a generic detector
description toolkit built by reusing the existing components of widely
used packages and provide the missing functional elements and
interfaces, for offering a complete and coherent detector description
solution to experiments.

Computations performed by a geometry modeler in a simulation program
must be highly precise and efficient. Lots of effort has been spent in
the past decades to achieve the level of precision and reliability
offered today, still the evolution of the computing hardware, compilers
and programming languages require algorithms and related implementations
to evolve as well. It is therefore foreseeable that effort in this sense
will have to continue, in a way as it is currently being done for the
VecGeom package, to also favor modularity and reusability of the various
components.

A better integration of the existing tools used in simulation and the
adoption of CAD systems for designing and fast-prototyping the geometry
models is also desirable. A variety of solutions tailored for simulation
exist today, especially in tools developed for space science
applications, although none or few of them are fully open source. In the
past such a system was not desirable because of the difference in level
of detail required in CAD programs compared to standard detector
simulations. Individual pipes, bolts, and wires, often described in CAD,
are normally grouped together and modeled as single large volumes in
detector simulations. In some modern detectors, the precision of the
physics program necessitates a level of detail in the detector geometry
that is more comparable to that of a CAD design model.

Several of the systems for geometry description, while useful for
detector simulation, lack important features for use in data
reconstruction. The most important of these features is the ability to
include detector (mis-)alignment effects, some of which may cause
collisions between the 3-dimensional volumes in the detector. It is
critical to include these effects in reconstruction software, but often
in detector simulation such small alignment issues are omitted.

The programme of work for the near future in the field of geometry
includes the following:

\begin{itemize}
\item
  Promote the adoption of VecGeom by the experiments and users
  community, by adiabatically integrating and validating its use through
  the interfaces now existing and provided in the latest versions of the
  Geant4 toolkit.
\item
  Complete the porting and implementation in VecGeom of the missing
  shapes from the standard set of primitives of the GDML schema;
  deprecate the old scalar USolids implementation and tidy up code and
  configuration settings in VecGeom.
\item
  Provide VecGeom documentation and user guides.
\item
  Investigate adoption of VecGeom navigation algorithms in Geant4 and
  ROOT and provide first prototype implementations of a navigator
  adapter making full use of the scalar navigation capabilities in
  VecGeom.
\item
  Provide implementation of a GDML reader/writer in VecGeom to allow for
  easy detector description geometry exchange.
\end{itemize}

\hypertarget{section-1}{%
\section{\texorpdfstring{\\
}{ }}\label{section-1}}

\hypertarget{digitisation}{%
\section{4 Digitisation}\label{digitisation}}

Simulation toolkits like Geant4, in the end, only provide energy
depositions positioned in space and time in geometric elements. These
simulators often do not include some effects like charge drift in an
electric field, which would be prohibitively expensive, and often
unnecessary, to model in detail, and they do not include models of the
readout electronics of the experiments. They may not even have knowledge
of the detailed readout structure of the detector, for example
simulating charged particles passing through a block of silicon rather
than individual pixels or strips that correspond to readout channels.
Instead, these effects are normally taken into account in a separate
step called digitisation. The input to digitisation is the ``MC truth
information'' and energy deposits from the simulation, and the output
often includes a new, processed truth record, as well as a data format
that is conceptually similar to the data from the detector (e.g.
voltages, currents, times, and so on).

Digitisation is inherently local to a given sub-detector, and often even
to a given readout element, so that there are many opportunities for
parallelism if the code and the data objects are designed optimally.
Digitisation is also normally the stage at which pileup is introduced
and backgrounds, including detector noise, are included. Generally,
without pileup, the digitisation process is not particularly expensive
in terms of CPU time or memory. With pileup included, however, there can
be significant performance issues, particularly for HL-LHC
configurations of digitisation code. One exception to this performance
rule are large drift detectors like the ALICE TPC, where significant
``simulation'' occurs in the digitisation software. Again, parallelism
can provide a path to performance improvements, both in terms of
vectorisation and multiprocessing or multithreading.

Even without major performance challenges, digitisation presents one of
the most significant risks in terms of code aging and loss of expertise.
The detailed understanding of the readout structure of a detector is
normally most readily available when the detector is built, and as time
goes on the experts in these readout systems may move on, making code
maintenance and development, particularly for detector upgrades,
problematic. Similarly, because these codes often do not result in major
and obvious analysis-level issues, they are often left alone for many
years, without being updated to newer standards and best practices. Such
issues could be helped by greater code sharing, if it could be achieved.

Since digitisation code includes a number of effects that are specific
to the individual readout that is used, historically there has been very
limited sharing of code among digitisation frameworks. Recently, both
hardware and software projects have benefitted from an increased level
of sharing among experiments. LArSoft is an endeavor to provide common
digitisation code to all the experiments using large liquid argon
detectors. Similarly, the development of next generation silicon
detectors requires realistic simulation of the charge collection and
digitisation processes. Owing to the large variety of technologies,
common software frameworks need to be flexible and modular to cater the
different needs. The Allpix software and the more recent Allpix Squared
framework provide generic algorithms shared between many silicon pixel
detector projects, and Allpix Squared follows a modular approach and is
thus also suited to simulate silicon-based calorimeters {[}16{]}. The
frameworks also allow the implementation of effective radiation damage
models and therefore enable the simulation of the detector response
throughout the experiment lifetime.

As both CMS and ATLAS expect to use similar readout chips in their
future trackers, further code sharing might be possible. Such effects
will be more important in the coming years, and the experiments both at
the LHC and elsewhere are examining similar detector technologies for
future upgrades or implementations, so the possibility of common code
bases should not be ignored.

In summary, it is expected that within the next 3 years common
digitisation efforts are well-established among experiments, and
advanced high-performance generic digitisation examples, which
experiments could use as a basis to develop their own code, will become
available. For example, the development of next generation silicon
detectors requires realistic simulation of the charge collection and
digitisation processes. Owing to the large variety of technologies,
common software frameworks need to be flexible and modular to cater for
the different needs. Specific objectives are by:

\begin{itemize}
\item
  2020, to deliver advanced high-performance, SIMD-friendly generic
  digitisation examples that experiments can use as a basis to develop
  their own code.
\item
  2022, to fully test and validate optimised digitisation code that can
  be used by the HL-LHC and DUNE experiments.
\end{itemize}

\hypertarget{modeling-of-pileup}{%
\section{5 Modeling of Pileup}\label{modeling-of-pileup}}

For the LHC experiments, in particular ATLAS and CMS, the modeling of
pileup (proton-proton collisions simultaneous with the collision of
interest) represents a significant challenge over the next 10 years. The
amount of pileup is expected to grow from an average of
\textasciitilde{}30 collisions for every proton-proton bunch crossing in
2016, to \textasciitilde{}200 in the HL-LHC era in 2025. The individual
detector elements of the experiments are sensitive to pileup in
different ways, and over different time ranges. It is perhaps worth
noting that the discussion of pileup here can equally apply to the
high-multiplicity underlying event in central heavy-ion collisions at
the LHC.

There are a number of ways to model the effect of pileup that the
experiments have all examined. The simplest approach is to simulate a
large number of inclusive proton proton collision (minimum bias) events
with an event generator, such as Pythia8 {[}30{]}, and to overlay the
energy depositions from these events on top of the energy depositions
from the event of interest. Normally these minimum bias events are kept
in libraries to avoid frequent, CPU intensive re-simulation of all the
required pileup. With the pileup expected at the HL-LHC, these libraries
become very large if the reuse rate of events with high-pT jets is to be
kept small, demanding large amounts of disk storage for each job as they
copy a sufficient number of events locally. The copying introduced also
strains networks when large numbers of jobs are running.

Another option is to ``pre-mix'' together the minimum bias collisions
into individual events that have the full background expected for a
single collision of interest. This has the advantage that the disk usage
of a single event is reduced, but still runs a risk of background events
being reused for a large dataset.

The experiments have also explored the possibility of using data
directly to model the pileup and detector noise, a technique called
``overlay". This option has a number of advantages, including the
correct modeling of detector noise over the full run period. There are,
however, a few drawbacks that make this both technically and practically
difficult. One can only model detectors for which data has been taken,
so simulations of future layouts cannot be done this way. The detector
misalignments and simulation misalignments may not be the same, but the
events are normally run through a single reconstruction, so that hits
shift positions if different alignments are used for both. The data used
for overlay must also normally be taken without ``zero suppression'' in
order to ensure that low energy contributions from both the signal event
and background that are individually below the normal readout threshold
for a detector can coherently add to a deposition above that threshold.
In addition, some detector electronics have nonlinear responses to
deposited energy, implying that their output data cannot be added
trivially. Complications and approximations could be necessary in a data
overlay scheme to deal with nonlinear electronics.

With very fast simulation come additional possibilities. For
sufficiently fast simulation, it becomes feasible to perform the
simulation (and even the generation) of the additional pileup events at
the same time as the signal event. This ensures that the overlaid events
are unique, but has the disadvantage of the fast simulation's lower
fidelity. Under normal circumstances, however, analyses are not
sensitive to the details of the pileup simulation, and even a fast
simulation that is not sufficiently accurate for a signal event
simulation may be sufficient when simulating the pileup. Indeed, with
generative adversarial networks such as those discussed in the Fast
Simulation Section 10.3, the background from pileup could be generated
as a single background image, taking into account all backgrounds that
don't require special handling. Such techniques are obvious candidates
for exploration in the coming years.

Both experiments and theorists have also explored the possibility of
parameterisations of either the pileup or the detector response to the
pileup. Programs like DELPHES and PGS have simple parameterisations of
the response and resolution of analysis objects, and in general the
impact of pileup can be included, at least approximately, by modifying
those resolutions and lowering identification and isolation efficiencies
for leptons. Such techniques are rough, but are often good enough for
analysis prototyping or for far-future feasibility studies before the
experiments invest significant effort in the detailed simulation that
might be needed for the analysis. Such techniques are often employed
when performing large signal parameter space scans, as in the case of
searches for Supersymmetry, before detailed simulation of specific
signal models is undertaken.

In the parametric approach, any fast simulation technique used to model
the pileup must have a sufficiently correct model of the physics of
interest for the analysis at hand. For some analyses of long-lived
particles, this may not be the case, so multiple solutions should always
be kept in play.

As the LHC experiments further develop their data overlay techniques, it
may be useful to hold occasional discussions of challenges overcome in
these systems. Alignment issues, data-collection issues, and so on can
be addressed with the same philosophical approach, even if the technical
implementations differ somewhat. The development of fast simulations for
pileup should be followed as a part of the fast simulation discussions
explored in Section 10.3. As studies of physics analyses in
high-luminosity environments become more common, some of the current
parameterisation and fast simulation techniques will improve their
fidelity, but this will happen as a natural evolution with
experimentalists and phenomenologists, and likely does not require any
directed action.

In summary, an R\&D programme is needed to study different approaches to
managing the various backgrounds to hard-scattering events, including
in-time pileup, out-of-time pileup, cavern background and beam-gas
collisions. Progress is expected on the following within the next 3
years:

\begin{itemize}
\item
  Real zero-bias events can be collected, bypassing any zero
  suppression, and overlaid on the fully simulated hard scatters. This
  approach faces challenges related to the collection of
  non-zero-suppressed samples or the use of suppressed events,
  non-linear effects when adding electronic signals from different
  samples, and sub-detector misalignment consistency between the
  simulation and the real experiment. Collecting calibration and
  alignment data at the start of a new Run would necessarily incur
  delays such that this approach is mainly of use in the final analyses.
  The experiments are expected to invest in the development of the
  zero-bias overlay approach by 2020.
\item
  The baseline option is to ``pre-mix'' together the minimum bias
  collisions into individual events that have the full background
  expected for a single collision of interest. Experiments will invest
  effort on improving their pre-mixing techniques, which allow the
  mixing to be performed at the digitisation level reducing the disk and
  network usage for a single event.
\end{itemize}

\hypertarget{handling-of-monte-carlo-truth-information}{%
\section{6 Handling of Monte Carlo Truth
Information}\label{handling-of-monte-carlo-truth-information}}

All simulations keep some concept of Monte Carlo Truth, a record of what
was really going in the simulation of the detector, unaltered by the
detector readout. This is in some cases as simple as the event record
from the generator, which includes the particles that were input to the
simulation as well as some of the upstream particles from which they
derive. In most cases, however, more complex MC Truth is desired.
Interestingly, the standards for these records exist as text formats and
as C++ classes (HepMC and LHE), but there are not widely-used compressed
binary storage formats for the classes. Having such a standard might
facilitate the reusability of code and help the examples provided with
simulation toolkits be more practically useful.

One standard addition to the event generator record is a record of a
limited set of particle interactions in the detector. Often, for
example, the hadronic interaction of a pion within the tracker volume is
stored. There is not a standard format for such interactions, but it is
common to add these interactions directly to the original event
generator record, which is normally stored as a connected tree. This
tree is unique for a given event, but if multiple particles are
simultaneously in flight, as in the case of highly-parallel simulation,
then they may need to make simultaneous modifications to this tree. This
problem is not insurmountable, but requires some care.

The various other types of MC Truth are often detector-specific. These
can include records of true energy deposition, as well as records of
which particles in the generator record resulted in those energy
depositions, for example. Such records are more straightforward to store
in tracking detectors, where normally there is a one-to-one
correspondence between a ``hit'' used for track reconstruction and an
energy deposition by a particle from the event generator. Calorimeter
records are inherently more complicated, as showers from many particles
can overlap and merge.

The most significant challenge for MC Truth handling is keeping the
memory and disk consumption minimal, while still storing all the
information that is important to the analysis groups downstream. For
example, Geant4 has a concept of a ``Trajectory'', which can be useful
for keeping a particle history as a part of the truth record. If such
trajectories are kept for all particles produced in the calorimeter of
an LHC experiment, however, then the increase in memory usage would
become prohibitive. Similarly, while most interactions can be stored in
the truth record in the tracker, it would not be practical to do so in
the calorimeter, where each event results in millions of secondaries
generated. Ensuring that each experiment has the flexibility to save the
required MC Truth information, without inducing significant overheads in
memory or CPU usage, is critical for any future simulation toolkit.

For backgrounds that are included, as in the case of pileup, it is also
important for the experiments to be able to filter the truth
information, and indeed filter it differently than is done for a signal
event. What information is most of interest is both detector-specific
and analysis-specific, but frameworks must ensure the flexibility
required to satisfy even the most exotic use-case.

\hypertarget{physics-modeling}{%
\section{7 Physics Modeling}\label{physics-modeling}}

Geant4 physics libraries provide accurate simulation for a wide variety
of applications in HEP, space science, medical physics and other
scientific domains. HEP experiments tune their simulations by comparing
simulation results with both test beam data and collision data. By and
large, these comparisons show good agreement. Nevertheless,
discrepancies in comparisons between the results of simulation and
collider data do exist, and it can be extremely difficult to disentangle
whether the differences arise from misrepresentation of the detector
geometry and material or from inaccuracies in the physics modeling.

Big efforts continue to be invested by experimentalists to understand
the performance of their detectors, as well as by the physics developers
to improve and validate their physics simulation models. The comparison
against thin-target data has been very successfully used to develop and
to initially tune the physics models. However, it has been observed
recently that further tuning leading to improvements against thin target
data, does not necessarily lead to better simulation results when
compared to the thick-target data. Beyond the lack of enough thin target
data, this situation is, most likely, showing that the current models
are reaching their limits and in order to improve the overall detector
simulation, one needs to go beyond simple tuning of the existing models
and to develop new, more sophisticated ones. Another modeling challenge
that we will face at higher energies is related to the extreme closeness
of the central trackers to the interaction points, allowing short lived
particles such as hyperons to interact with the detector. Up to now,
these particles are decayed by the generators without being transported
by the simulation code as their decay happens in the vacuum of the beam
pipe. As beam energies increase, the line between particles dealt with
by generators and Geant4 will blur. These will survive long enough to
pass through the first few layers of the detectors leading to "missing
hits" in MC. As this situation becomes more common, the range of physics
models available for such particles will need to be expanded.

Our goal in detector simulation in the coming decade is to cope with
experimental needs for analysis of high-statistics data and for modeling
the next generation detectors (ILC/CLIC, FCC). Higher statistics in
future experiments will require smaller simulation errors, which in turn
implies higher precision models and larger Monte Carlo (MC) event
samples to be generated. This imposes two major challenges on physics
modeling. The first priority is to improve physics models to limit the
systematic error and match higher precision crucial for high luminosity
HL-LHC. The second priority is the extension of the validity of the
physics models up to FCC energies.

In the following, we summarise recent developments in electromagnetic
and hadronic physics modeling. We discuss physics accuracy and
performance improvements along with results of comparison of simulation
with experimental data.

\hypertarget{electromagnetic-physics}{%
\subsubsection{7.1 Electromagnetic
Physics}\label{electromagnetic-physics}}

Electromagnetic (EM) transport simulation is challenging as it occupies
a significant part of the computing resources used in full detector
simulation. More experimental data in the coming decade implies higher
statistics and this implies the need for large and accurate simulation
samples, which require both model accuracy and speed. Therefore the
physics models need to be improved for achieving higher accuracy and the
implementation of algorithms needs to be reviewed in order to optimise
performance. Significant efforts have been made in the recent past to
better describe the simulation of electromagnetic shower shapes, in
particular to model the H -\textgreater{} γγ signal accurately at the
LHC. This effort is being continued with emphasis on reviewing the
models assumptions, approximations and limitations, especially at very
high energy, and with a view to improving their respective software
implementations. For LHC Run-1 simulation Geant4 version 9.4 was used
{[}17{]}. For Run-2 significant improvements were introduced into the
bremsstrahlung, pair production and multiple scattering {[}18, 19{]}
further improving simulation of the EM shower shape; these improvements
were made available in Geant4 9.5 and 9.6. For validation purposes, an
extended set of tests versus data has been developed for high, moderate
and low-energies using test-beam data and published thin target data.
Comparisons with low-energy EM models from the Livermore and Penelope
packages were also carried out.

The process of multiple scattering of charged particles is a key
component of Monte Carlo transport codes. At high energy, it defines the
deviation of charged particles from ideal tracks, limiting the spatial
resolution of detectors. The scattering of low energy electrons defines
the energy flow across volume boundaries. This affects the sharing of
energy between absorbers and sensitive elements, directly affecting
shower shapes. A comprehensive description of recent improvements of the
Geant4 electromagnetic module can be found in {[}19{]}. Rather good
agreement was found when Geant4 predictions were compared with
experimental data collected at the LHC.

Figure 2 shows a comparison of CMS test beam data with predictions from
two Geant4 physics lists, QGSP\_FTFP\_BERT\_EML and FTFP\_BERT\_EMM. EML
utilises a simplified (faster) multiple scattering model for all
detectors, while EMM uses the detailed multiple scattering model for
sampling calorimeters and the simplified model for other detectors. The
same figure also shows the ratio of Monte Carlo to data as a function of
beam momentum. It is clear from the figure that the predictions from the
physics list FTFP\_BERT\_EMM are closer to the data, in particular at
higher beam energies, and therefore CMS has taken the decision to move
to this list for its 2017 MC productions. Recent results from the CMS
experiment were also presented showing discrepancies in the EM shower
width in the endcap region as shown in Figure 3. This is a fresh
observation that needs a much more detailed understanding, leaving us
with an open question as to whether there is a problem in the physics
model or an issue with the modelling of detector materials.

In the ATLAS experiment one area of disagreement concerns the
measurement of lateral shower shapes in the EM calorimeter. Figure 4
shows a measurement of R\textsubscript{η} for unconverted photons in the
ATLAS EM calorimeter, where R\textsubscript{η} is the ratio of shower
energy deposited in 3x7 cells to 7x7 cells in the forward hadron
calorimeter. There is a clear discrepancy between real data and the
results of MC simulation with Geant4 9.6 (red line). Shower shapes in
eta-direction are consistently wider in data than simulation, both for
electrons and photons. Detailed investigations suggest that this is not
a material issue and therefore work is continuing to better understand
the impact of the EM physics modeling on the results of the simulation.
A similar discrepancy is seen in the measurement of the lateral shower
width for electrons in the first layer of the EM calorimeter, as can be
seen from the lower plots in Figure 4 obtained with Geant4 9.2 and 9.4.
Although the simulation in the upper plot can be adjusted to better
match the data (``corrected'', in the figure), it is challenging to
understand correlations between mis-modeled effects, and a more accurate
simulation is desirable. A substantial amount of work and high level of
coordination with experiments is therefore needed to investigate the
pending discrepancies between data and simulation.

\includegraphics[width=6.70522in,height=4.94444in]{media/image2.png}\emph{Figure
2. The top plots correspond to the mean energy response for
p\textsuperscript{-} (left) and p\textsuperscript{+} (right) as a
function of the beam momentum. The bottom plots are ratios of Monte
Carlo predictions to the data for pions as a function of beam momentum.
The black points refer to the data, whilst the red and blue points are
Monte Carlo predictions from physics lists QGSP\_FTFP\_BERT\_EML of
Geant4 10.0.p02 and FTFP\_BERT\_EMM of Geant4 10.2.p02 respectively.}

The HL-LHC era will require higher precision simulations that can only
be achieved by continuous improvement of the electromagnetic physics
models used in the simulation. The further development of these models
is therefore still being actively pursued in order to improve their
precision and to extend their validity to higher energies. They include
the following.

\begin{itemize}
\item
  A new Goudsmit-Saunderson ``theory-based'' model for describing the
  \emph{multiple scattering} of electrons and positrons has been
  developed (Figure 5, blue line). It has been demonstrated to
  outperform, in terms of physics accuracy, the old theory-based model
  in Geant4 (grey line) and the Urban model that is the Geant4 default
  (Figure 5). The new implementation incorporates all the ingredients of
  the theoretical model resulting in extremely smooth angular
  distributions and has been developed with a view to fully exploiting
  track-level vectorisation in the future. At the same time it is
  important to note that this was achieved without incurring a CPU
  performance penalty.
\end{itemize}

\includegraphics[width=3.29292in,height=3.03646in]{media/image3.png}\includegraphics[width=3.29288in,height=3.07813in]{media/image4.png}

\emph{Figure 3. Crystal-based EM shower width from the CMS experiment in
the eta direction in both the barrel and the endcap. Simulation and data
are in good agreement in the barrel (figure on the left) but not in the
endcap region (figure on the right) where there is more material.}

\begin{itemize}
\item
  The models used to describe the \emph{bremsstrahlung} process have
  also been reviewed for Geant4 9.5 allowing to describe the shower
  shape for the CMS barrel (Figure 3). Recently an improved theoretical
  description of the Landau-Pomeranchuk-Migdal (LPM) effect was
  introduced which plays a significant role at high energies. We show in
  Figure 6 a comparison between the old (grey line) and improved
  bremsstrahlung implementations with respect to experimental data. The
  review of this model was beneficial and demonstrated excellent
  agreement with the data. Once again, these improvements were made
  without introducing any speed penalty. They are available with Geant4
  10.3.
\item
  The model describing the pair-production process has also been
  reviewed. Similar improvements were introduced in the LPM suppression
  functions. At very high energy the improved pair production model
  shows a significant difference in the differential cross-section. The
  old implementation (green plot in Figure 7) was over-suppressing gamma
  conversion i.e. the LPM effect was overestimated. There are currently
  no available experimental data for validation of our results. The use
  of the relativistic pair-production model is essential for accurate
  physics simulations for FCC.
\item
  At extreme high energies (important for FCC studies), nuclear size
  effects become important. Different elastic scattering models with
  nuclear form factors were developed taking into account finite nuclear
  size effects. Figure 8 shows that the newly developed single
  scattering model agrees with experimental data at relatively low
  energy. Also at high energies, since precision becomes increasingly
  important, formfactor description may affect various EM processes.
  Such effects may be important for high energy muons, for shower shape,
  and for simulation of background for a dark matter search experiment.
\end{itemize}

\includegraphics[width=3.45982in,height=2.73438in]{media/image5.png}

\includegraphics[width=6.66281in,height=2.57813in]{media/image6.png}\emph{Figure
4. The upper plot plot shows a measurement of} R\textsubscript{η}
\emph{for unconverted photons in the ATLAS EM calorimeter, where}
R\textsubscript{η} \emph{is the ratio of shower energy deposited in 3x7
cells to 7x7 cells in η-φ. The lower plot shows the lateral shower
width, W\textsubscript{η1}, for electrons in the first layer of the EM
calorimeter. Both plots show a clear discrepancy between real data and
the results of MC simulation.}

\includegraphics[width=5.84531in,height=4.08869in]{media/image7.png}

\emph{Figure 5. The new Goudsmit-Saunderson} \emph{multiple scattering
model implementation (GS - Geant4 10.2) shows no artefacts and excellent
agreement with experimental data. }

\includegraphics[width=6.03802in,height=4.2226in]{media/image8.png}

\emph{Figure 6. The improved bremsstrahlung model included in Geant4
10.3.beta and in GeantV shows better agreement with experimental data
with respect to the existing implementation.}

\includegraphics[width=5.88021in,height=3.99503in]{media/image9.png}

\emph{Figure 7. The improved pair production model included in both
Geant4 10.4.beta and in GeantV is compared with the existing Geant4
implementation with and without LPM. }

\begin{quote}
\includegraphics[width=4.31527in,height=3.89063in]{media/image10.png}
\end{quote}

\emph{Figure 8. New models were developed in Geant4 taking into account
finite nuclear size effect for elastic scattering of electrons at high
energies. }

\begin{quote}
Theoretical review of all electromagnetic models, including those of
hadrons and ions is therefore of high priority both for HL-LHC and even
more importantly for FCC studies. Efforts in this direction will
continue, but completion of this task by 2025 will demand a substantial
amount of work and continuous investments in effort. Avenues that should
be investigated include other channels that become important at extreme
energies, such as triplet production taking into account nuclear recoil
effects, size effects as well as γ conversion to muon and hadron pairs.
Energy fluctuation models should be developed based on theory, not just
parametrisation. For Geant4 10.3 the upper energy limit for all EM
physics models was extended to 100 TeV, which is important not only the
HL-LHC programme but becomes crucial for the work being carried out in
the context of the FCC detector design studies. At the same time,
further improvements for high energy EM models are possible. A
substantial amount of work is needed for preparing simulations of next
generation detectors.
\end{quote}

\hypertarget{hadronic-physics}{%
\subsubsection{7.2 Hadronic Physics}\label{hadronic-physics}}

Hadronic physics simulation is loosely defined to cover any reaction
producing hadrons in its final state. As such, it covers purely hadronic
interactions, lepton- and gamma-induced nuclear reactions, and
radioactive decay. In Geant4 the interaction is represented as a process
which consists of a cross-section to determine when the interaction will
occur, and a model which determines the final state of the interaction.
Models and cross-sections are provided which span an energy range from
sub-eV to TeV. More than one model is usually offered in any given
energy range in order to provide alternative approaches for different
applications. During the last several years, new models and
cross-sections have been added to the toolkit, while others have been
improved and some obsolete models have been removed.

it is not possible for a single model to describe all the physics
encountered in a simulation due to the large energy range that needs to
be covered and the simplified approximations that are used in hadronic
models to overcome the difficulty of solving the full theory (QCD). A
typical sequence of reactions may begin with a high energy
hadron--nucleon collision within a nucleus (parton string model),
followed by the propagation of the secondaries of the collision through
the nuclear medium (intra-nuclear cascade model), followed by the
de-excitation of the remnant nucleus (pre-compound model) and its
evaporation of particles (evaporation and breakup models) until it
reaches the ground state. Each of these stages is qualitatively
different from the other. Wherever possible a theory-based approach to
their implementation is followed as opposed to a phenomenological or
parameterised approach. The main motivation is to have more confidence
in the behaviour of the model outside the kinematic regions where it is
validated and tuned with experimental data, and also to preserve the
correlations between the different emitted particles.

Two models based on quark-parton concepts have been implemented in
Geant4, the quark--gluon string (QGS) model and the Fritiof (FTF) model.
There are also three intra-nuclear cascade models that are now offered:
Bertini, Binary and INCL++. As of release 10.0, the toolkit provides
nine reference physics lists whose names reflect the combination of
models used to describe the hadronic interactions necessary for various
applications over the whole energy range. Reference physics lists are
extensively and routinely validated. Currently the most-used reference
physics list for high energy and space applications is FTFP\_BERT. It
uses the Geant4 Bertini cascade for hadron--nucleus interactions from 0
to 5 GeV incident hadron energy, and the FTF parton string model for
hadron--nucleus interactions from 4 GeV upwards. The letter P indicates
that the Geant4 pre-compound model is used to de-excite the nucleus
after the high energy FTF interaction has been completed. The
FTFP-Bertini combination forms the backbone of many physics lists. It
successfully extends the use of the string model at lower energies, down
to 3 GeV, and at the same time the Bertini intra-nuclear cascade model
has been validated up to 15 GeV. The net result has been that the need
for using the LEP parameterised model at mid-range energies (around 10
GeV) has been eliminated. QGSP\_BERT is a popular alternative which
replaces the FTF model with the QGS model over the high energy range. It
is important to note that the existence of more than one model (for each
energy range) is very valuable in order to be able to determine the
systematics effects related to the approximations used. The situation is
similar to that of the Monte Carlo generators where the usage of
different generators provides an assessment of the systematic errors.

Detector response is an effective test of any model combination. It is
defined as the ratio of deposited energy visible to the detector, to the
incident beam energy. For the above combination of models (as in the
FTFP\_BERT physics list), the general agreement between the simulated
response and data for hadron-induced showers is at the level of a few
percent. Other useful data, such as shower shapes and energy resolution
are less precisely described and show agreement at a level of 10--20\%.

The use of highly granular calorimeters such as the ones being designed
by the CALICE collaboration for future linear colliders, allows a
detailed validation of the development of hadronic showers with
test-beam data. Preliminary results suggest that the lateral profiles of
Geant4 hadronic showers are too narrow. Comparisons with LHC test-beam
data have shown that a fundamental ingredient for improving the
description of the lateral development of showers is the use of
intermediate and low energy models that can describe the cascading of
hadrons in nuclear matter and the subsequent de-excitation of the
``wounded'' nucleus. The longitudinal development of hadron showers
mainly depends on the hadronic interactions at higher energies in the
forward direction: quasi-elastic scattering and diffraction. An
important effect recently introduced in Geant4 is the improvement of the
neutron capture cross-sections and final state generator. Based on the
high precision neutron library, it allows for an improved simulation of
the time structure and the lateral profile of hadronic showers in
neutron-rich materials {[}19{]}. Other recent developments include an
improved and re-tuned Fritiof model.

The LHC experiments are preparing detector upgrades for PHASE 2 of the
physics programme. For example, CMS will have a high granularity
calorimeter replacing the endcap electromagnetic and hadron
calorimeters. This will produce two main challenges. Firstly the EM
calorimeter will be of sampling type and so the longitudinal shower
profile becomes very important. Secondly, the sensitive detector
elements will be a mixture of silicon and scintillators and some of them
have hexagonal or half-hexagonal shape. This increases the complexity of
the detector geometry and will present a new challenge for the CPU
performance of the simulation code.

Work is ongoing to make a thorough review of the hadronics sector.
Additional work is currently being invested in the further improvement
of the QGS model, which is a more theory-based approach than the
phenomenological FTF model, and therefore offers better confidence at
high energies, up to a few TeV. This again is a large endeavour and
requires continuous effort over a long time. Further extension of the
coverage of hadronic interactions to higher energies, in the multi-TeV
region, will be necessary for FCC detector simulation studies. Moreover,
this would open a completely new domain of applications: the possibility
to simulate very high energy cosmic ray showers in the Earth's
atmosphere. The EPOS hadronic generator {[}20{]} offers currently the
most accurate simulations in this domain, as a fresh C++ rewrite of the
model, appears as a very interesting possibility for the future of
hadronic simulations in high energy and cosmic ray physics.

For the intra-nuclear cascade models, while Bertini and Binary are
likely to remain stable, further development is expected for the INCL++
model, besides the recent extension to higher energies (up to about 20
GeV). At the level of physics lists, the combined use of all the three
intra-nuclear cascade models - choosing between them according to
projectile hadron, projectile kinetic energy and target nucleus - will
be explored to optimise physics accuracy or simulation speed.

Low-energy hadronic physics - pre-equilibrium, nuclear de-excitation and
radiative decay models, as well as high-precision, data-driven
transportation of low-energy neutrons and charged particles (deuteron,
triton, He3 and alpha) - is and will remain a particular active area of
development, mostly driven by medical and nuclear physics applications.

The way hadronic cross-sections are implemented is critical for the CPU
performance of hadronic physics simulations and there are on-going
efforts to review them and make them more efficient via the development
of a common GeantV / Geant4 module. Total, inelastic and elastic
cross-sections for hadron--nucleus, nucleus--nucleus and
antinucleus--nucleus reactions are provided which cover energies up to
TeV in some cases. Proton-, neutron- and pion-nucleus cross-sections at
low to medium energies have been available in Geant4 since its
beginning. Recent work has focused on development of cross-sections for
other projectiles and at higher energies. The general behaviour of the
optical models is to predict constant cross-sections for very high
energies. However, experimental data show a moderate relativistic rise
of hadron--nucleus cross-sections. For this reason the Glauber model is
being used to describe hadron--nucleus cross-sections in the high energy
region (above 90 GeV). The current implementation of this model is under
review with a view to re-implementing it.

The above, ambitious program of development for the simulation of
hadronic physics relies heavily on an updated, extended, robust and
automated validation test-suite. Plenty of work has been done in the
recent past in this direction, but a lot more remains to be done. In
particular, a well-defined procedure to tune hadronic models needs to be
defined and deployed.

\hypertarget{physics-modelling-for-the-intensity-frontier}{%
\subsubsection{7.3 Physics Modelling for the Intensity
Frontier}\label{physics-modelling-for-the-intensity-frontier}}

Much of the text above concentrates on physics processes relevant to the
LHC experiments. Much of this overlaps the needs of muon and neutrino
experiments as well as dark matter and rare-decay experiments, which
have typically used Geant4 physics lists tuned based on LHC
requirements. However, as the precision needs of these experiments
increase, they increasingly need simulations responding to their
particular needs, which vary significantly with experiment but broadly
speaking involve precise simulations of hadrons and electromagnetic
particles from a few GeV to a few MeV. Precise neutron simulation is
particularly critical for accelerator-based neutrino experiments that
must reconstruct neutrino energy from final state particles. Also a
radioactive decay model for modelling the decay chain of isotopes is
essential for background studies.

Major components of the intensity frontier over the next decade will
include liquid Argon TPC detectors and liquid Xenon calorimeters.
Efforts to address common needs of the liquid Argon experiments
including the post Geant4 phase of simulation are undertaken in projects
such as LArSoft{[}21{]} and Qscan{[}22{]}, while extending Geant4 EM
physics models for liquid noble gases is undertaken by the NEST project
{[}23{]}. Detectors with large numbers of wires and photo-sensors may
naturally lend themselves to parallelisation techniques, provided that
the simulation frameworks are able to handle such cases in an efficient
way. Additionally, the Wire Cell Toolkit {[}24{]} provides simulation,
noise filtering, signal processing and reconstruction algorithms for
LArTPC detectors. It includes experimental support for parallel
processing following the data flow processing paradigm. More work and
input from parallel processing experts is needed to bring this support
to maturity.

Data-taking at current experiments, such as MicroBooNE, will greatly
inform the simulation needs of the Intensity Frontier (IF) community
moving forward. This might be most relevant for LArTPC technology as it
is the newest and least familiar. For example, if significant
non-uniformity of purity in data is observed in MicroBooNE, simulation
of non-uniform purity will be a high priority simulation need of future
LArTPC experiments.

Efforts are also underway to improve the accuracy of Geant4. While much
of this effort was motivated by requests from Intensity Frontier
experiments, especially the variation of the model parameters, it will
be of use to the entire HEP community. This work includes:

\begin{itemize}
\item
  Validation of models: a validation database known as DoSSiER {[}25{]}
  that contains data from experiments measuring particle cross-sections
  (e.g. NA61) is being developed. It was started as a Geant4 project,
  but is now also used by GeantV and should be suitable for other MC
  toolkits such as e.g. GENIE {[}7{]}.
\item
  Development of new physics lists: Geant4 physics lists utilised by IF
  experiments were originally designed to meet the needs of LHC
  experiments. For example, a ``Shielding'' physics list was originally
  configured for radiation shielding studies and cavern background
  simulations for LHC experiments, but it is now widely used by
  underground dark matter experiments such as CDMS and LZ. A variant of
  this Shielding physics list (ShieldingM) was created to address the
  needs of Mu2e. Geant4 physics lists tailored for LHC were not adequate
  to the generation of neutrino beams, they disagree with existing
  hadron-production data by up to 40\% in some areas of phase space. In
  light of this, a new physics list (``NuBeam'') has been created, aimed
  at meeting the specific needs of neutrino beam simulations.
\item
  Variation of model parameters: Equally as critical as having an
  accurate beam or detector simulation is having an ability to quantify
  that accuracy. Neutrino experiments go to great effort to estimate and
  propagate uncertainties on Geant4 models to physics measurements.This
  is currently done experiment-by-experiment, but an initiative within
  Geant4 was recently begun to provide methods of assessing model
  parameters. Additionally, procedures are being developed to modify
  model parameters and compare with data, with the aim of producing
  uncertainties and covariance matrices on these parameters that can
  then be propagated to physics measurements.
\end{itemize}

\hypertarget{non-hep-use-cases}{%
\subsubsection{7.4 Non-HEP use-cases}\label{non-hep-use-cases}}

The extensibility of Geant4 has been appreciated also outside of the HEP
domain and several applications and extensions have been created to
perform simulations in some additional domains. It is worthwhile to
mention:

\begin{itemize}
\item
  Crystal extensions to model the property of crystalline structures.
  Applications include the use of bent crystals for beams manipulations,
  phonons and ultra-cold neutrons beams {[}26{]}.
\item
  Geant4-DNA: a Geant4 extension for the calculation of physico-chemical
  damage to DNA at a nanoscale level {[}27{]}
\end{itemize}

\hypertarget{radiation-and-non-collision-background-modeling}{%
\subsubsection{7.5 Radiation and Non-Collision Background
Modeling}\label{radiation-and-non-collision-background-modeling}}

The LHC experiments have made use of the FLUKA simulation package, and
to a lesser extent GCALOR and MARS, for modeling of the radiation
backgrounds, primarily of low-energy neutrons and photons, present
during and after the running of the LHC. Understanding these backgrounds
is critical for estimating the required radiation hardness of
on-detector electronics, for developing shielding that might reduce
non-collision backgrounds in the detector, and for evaluating radiation
safety procedures when servicing or upgrading the detectors. The FLUKA
simulation package provides excellent modeling of these backgrounds, but
to date has not been used for any significant high-energy collision
modeling by any of the experiments. While its maintenance and continued
functionality is critical for the future of the LHC experiments and for
the development of the next generation of experiments, improvements in
both performance and physics are not as critical as they are in the case
of Geant4.

\hypertarget{section-2}{%
\section{}\label{section-2}}

\hypertarget{section-3}{%
\section{\texorpdfstring{\\
}{ }}\label{section-3}}

\hypertarget{software-performance}{%
\section{8 Software Performance}\label{software-performance}}

Simulation, including physics generation, interaction with matter
(detailed analogue or parametrised simulation), digitisation,
reconstruction and analysis takes a large fraction of the computing
resources utilised by HEP experiments. Depending on the experiments, and
the complexity of the detectors, each of these components of the
simulation chain contribute with different weights to the total. Setting
aside the reconstruction of simulated events, the modeling of the
interaction with matter is the most expensive module of a simulation
application, as is certainly the case for the LHC experiments. Neutrino
detectors are typically simpler, with a larger fraction of resources
spent in the simulation of the readout. Physics generation usually takes
the smallest fraction, except when generating signal samples that
require a scan of large regions of theory parameter space or for cutting
edge, high-accuracy calculations that are now becoming possible with new
event generation tools.

As an illustration of the above, the CMS experiment, from start-up in
2009 through May 2016, the Full Simulation chain including all elements
described above took 85\% of the total CPU time utilised by the
experiment, while the Geant4 module took about 40\%. This information
was obtained from the CMS Dashboard, which is a computing information
monitoring source available to CMS members. The rest of the CPU cycles
were primarily used to reconstruct and analyze real collider data. The
assumption for the 85\% figure is that the analysis of simulated data
consumes 75\% of the CPU time spent in analysis, including both
simulated and real data, and excludes the generation of signal samples
for BSM searches. The reason why the analysis of simulated data takes a
larger fraction of the total analysis CPU time than the analysis of real
collider data is that the design and optimisation of the measurements,
as well as the development and validation of data-driven methods, are
all based on MC samples.

An analysis of jobs running on both the grid and on HPC systems in
ATLAS, between July 2015 and July 2016, showed similar results with
simulation utilising 65\% of the total time available to the experiment.
The breakdown shows that Event Generation (Evgen) took 25\%, thanks to a
big burst of Mira throughput, whilst Full Simulation (Geant4) took 39\%
and Fast Simulation (Atlfast-II) just 2\%. In addition, Event
Reconstruction of data and simulated events took 19\%, whereas the
production of derived datasets for use in analysis took 5\% and
user-level Analysis took 10\% .

The Geant4 module of the ALICE Full Simulation application takes 55\% of
the total detector simulation CPU time, excluding event generation and
reconstruction, while digitisation takes 35\% of the time.

Although neutrino experiments are similar to LHC experiments in the
fraction of computing resources used for simulation, NOvA spent 70\% of
their CPU cycles on simulation in 2016, the total amount of computing
resources utilised by this experiment is two orders of magnitude less
than those spent by CMS.

The main drivers of the need to improve computing performance in the
next two decades are therefore the LHC experiments. It is expected that
their computing needs will increase by a factor of 10 to 100 in the
High-Luminosity LHC (HL-LHC) era depending on the solutions developed to
face simulation, pileup, and reconstruction challenges arising from the
high-luminosity environment. The challenge is daunting and the need to
explore both evolutionary and breakthrough solutions through R\&D
programs to speed up simulation and reconstruction is essential. The
computing needs of neutrino and muon experiments is expected to be
smaller than that of the LHC experiments, but will grow substantially
with time as more experiments come on-line and existing experiments
accumulate data. These experiments will therefore need to make use of
the solutions developed for the LHC.

The Geant4 Collaboration has invested significant effort into improving
the toolkit computing performance in the past decade, as code was
reviewed and optimised. In 2013, the release of event-level
multithreading capabilities in Geant4 brought significant memory
savings, as illustrated in Figure 10, which shows the CPU time (top) and
memory (bottom) consumption of a simplified CMS standalone simulation
application, based on Geant4 version 10.3.p01, runs for 50 GeV pions.
Geant4 achieves almost perfect strong scaling (top plot) with the number
of available cores (57) and still gains about +30\% in hyperthreading
regime {[}29,30{]}. The large memory reduction obtained with
multi-threading is clearly visible (bottom). While a multi-process
approach can reach the O(10GB) memory consumption with a relatively
small number of threads (Geant4 version 9.6), it is only with a well
designed multi-threading approach, which shares the read-only data among
threads, that it is possible to run with the maximum number of threads
remaining in a memory budget of approximately 1 GB. Recent versions of
Geant4 have put emphasis on further reducing memory consumption to the
current level of about 10MB needed for each additional thread.

Figure 11 shows the percentage change in CPU time performance taking
Geant4 version 10.0 as a reference, starting with Geant4 version 9.4.p02
(2010) and ending with Geant4 version 10.2 (2015) for the standalone CMS
application and a simple calorimeter configuration made of
Cu-Scintillator in a 4 Tesla magnetic field {[}29{]}.The study is
performed for 50 GeV e\textsuperscript{-}, p\textsuperscript{-}, and
protons, as well as for H → ZZ events. On average, the time performance
improvement through the life of the LHC experiments (2010-2015) is of
the order of 35\%. All tests were performed on the same hardware (AMC
Opteron 6128 HE @ 2 GHz) using the same compiler and operating system
(GCC4.9.2, Linux x86\_64). Remarkably, the percentage time performance
improvement during the period of time shown in the plots is in double
digits, even as the physics models were improved significantly for
accuracy, something that typically comes associated with a time
performance penalty. Although CPU profiling is a good starting point to
improve time performance, the biggest gains in Geant4 were achieved
while focusing on reducing the memory usage. Important speedup was
obtained as a direct result of both reducing the memory churn and
improving the physics algorithms. For example, the rewriting of the
Bertini physics model code yielded a reduced memory churn, making it
substantially faster.

Evolving compilers also offer opportunities for speed gains in
simulation applications. Figure 12 illustrates the impact of the use of
shared and static libraries in a simplified CMS application (HepExpMT
public benchmark) {[}31{]}. In this example, static linkage yields a
20\% gain, while Profile-Guided Optimisation (PGO) gives another 20\%.
Multithreaded applications benefit more from static linkage and PGO,
larger speedups as the number of threads increase. These options are
particularly interesting because they do not change the physics output
of the simulation. If this constraint can be relaxed additional speedups
can be obtained using additional Geant4 flags to speedup the simulation.

\includegraphics[width=6.5625in,height=3in]{media/image11.png}\includegraphics[width=6.48958in,height=2.9375in]{media/image12.png}\emph{Figure
10: CPU time (top) and memory (bottom) consumption of a CMS standalone
simulation application, based on Geant4 version 10.3.p01, run for 50 GeV
pions. The strong scaling (top) as a function of the number of threads
shows an almost perfect linearity (\textgreater{}90\%) with the number
of physical cores. The maximum number of cores (57) is denoted by the
dot-dash vertical line. The use of hyper-threading improves the
throughput by about 30\%. The use of multi-threading allows optimal use
of the limited memory budget of many-core systems. With multi-process
approaches (Geant4 9.6.p03 and 10.0.p04 sequential) it is possible to
use only a very limited set of the available parallelism. With
multithreading the memory requirements are reduced by a factor 10.}

\includegraphics[width=3.28646in,height=1.95745in]{media/image13.png}\includegraphics[width=3.26538in,height=1.95313in]{media/image14.png}\includegraphics[width=3.30325in,height=1.96354in]{media/image15.png}\includegraphics[width=3.2959in,height=1.95313in]{media/image16.png}

\emph{Figure 11: Percentage change in CPU time performance taking Geant4
version 10.0 as a reference, starting with Geant4 version 9.4.p02 (2010)
and ending with Geant4 version 10.2 (2015) for the standalone CMS
application and a simple calorimeter configuration.}

\includegraphics[width=6.70522in,height=6.73611in]{media/image17.png}

\emph{Figure 12: Impact of the use of shared and static libraries in a
simplified HEP application. Results obtained with a simplified version
of CMS with Geant4 version 10.0. Performances for different combinations
of compilation options are compared. There is an important performance
boost when using static builds (+20\%), the main reason being the lack
of PLT calls. Another +20\% can be obtained using Profiler Guided
Optimisations (PGO). These techniques do not change the physics output
of the simulation, if this constraint can be relaxed compilation and
runtime options of Geant4 can be used to obtain another +10\% speedup. }

In the following sections, we will discuss computing performance studies
by different HEP experiments, as well as their diverse approaches to
improve computing performance.

\hypertarget{cpu-time-and-memory-performance-measurements}{%
\subsubsection{8.1 CPU Time and Memory Performance
Measurements}\label{cpu-time-and-memory-performance-measurements}}

The largest improvements to computing performance experienced by ATLAS
and CMS have come from physics code reviews as part of the effort to
reduce memory churn. The speedup has been a side effect of this effort.
The lesson learned is that profiling is not a very useful exercise
unless it is related to physics observables. The table in Figure 13
shows the number of Geant4 steps in each ATLAS sub-detector and for each
particle type, providing useful quantitative information to optimise
Geant4 tracking and therefore improve time performance. It is apparent
that most of the steps occur in the forward and EM calorimeters, for
electrons, photons, and neutrons. Based on the experience described
above, the Geant4 Collaboration is developing a tool to monitor the
number of steps and tracks, and the time spent in a given detector
element. The tool will also provide this information for each Geant4
track in user defined energy ranges. The output will be displayed in the
console, or saved in histograms or ntuples.

\includegraphics[width=6.21354in,height=3.875in]{media/image18.png}

\emph{Figure 13: Number of Geant4 steps in each ATLAS sub-detector and
for each particle type. The sub-detector types are: Px (pixel tracker),
Sct (strip tracker), Trt (straw-tube tracker), Sev (tracker services),
Cry (calorimeter cryostat), Pre (calorimeter presampler), EMB (EM barrel
calorimeter), EMEC (EM endcap calorimeter), HCB (hadronic barrel
calorimeter), HEC (hadronic endcap calorimeter), FC1 (forward
calorimeter EM module), FC23 (forward calorimeter hadronic modules), FCO
(other parts of the forward calorimeter), LAr (other parts of the liquid
argon calorimeter system), Mu (muon system), and other (those not
otherwise accounted for).}

Figure 14 shows a chart of CPU time for the G4 module of the CMS Full
Simulation application {[}32{]}. About 60\% of the time is spent on
functions associated with the propagation of particles through the
detector geometry and magnetic field, while about 15\% is spent on EM
physics, and about 10\% followed by hadronic physics. Figure 15 shows
the CMS throughput and memory per event utilisation for QCD and ttbar
simulated events {[}32{]}. Maximum CPU efficiency is achieved after
simulating about 500 events, when contribution of initialisation of
Geant4 and CMSSW become negligible.

\includegraphics[width=4.37995in,height=3.08854in]{media/image19.png}

\emph{Figure 14: Chart of CPU time spent in different tasks of the CMS
Full Simulation application for Run-2 simulation}

Optimizing particle production cuts and replacing particle showers in
detectors with pre-generated showers or parameterisations are also part
of the strategy to improve computing performance. Both ATLAS and CMS
have these ``frozen showers'' or ``shower library'' options,
particularly to replace Full Simulation in the forward calorimeters.
GFLASH {[}33{]}, a Geant4 module for the fast simulation of
electromagnetic and hadronic showers using parameterisations for the
longitudinal and lateral profile, is not used at the moment due to the
peculiarities of the ATLAS accordion calorimeter and the limitations of
this approach to model longitudinal shower leakage in the CMS
calorimeters.

Geant4 offers the option to utilise the ``Russian Roulette'' technique
(see Sec. 10.1) to kill N-1 out of N particles within an ensemble of
neutrons and gamma below few MeV thresholds inside calorimeter geometry
regions. The surviving particle is assigned a weight of N. The use of
this technique by the CMS experiment yields significantly CPU time
savings, as discussed in Section 8.2.

\includegraphics[width=6.70522in,height=8.18056in]{media/image20.png}

\emph{Figure 15: Throughput and memory per event for ttbar simulated
events in CMS as a function of the event number. The version of Geant4
is 10.2p02, run in multi-threaded mode with 4 and 8 threads. Forced init
refers to the case when all data for photon evaporation is read at
initialisation time.}

8.2 ATLAS and CMS

Given that ATLAS and CMS are expected to be the largest consumers of
computing resources a decade from now, the two experiments have decided
to develop a computing performance assessment program that includes the
Run 1 and Run 2 detectors as a reference, the different HL-LHC detector
scenarios to predict future needs, and an evaluation on how much can be
gained by utilizing state of the art techniques, such as re-optimizing
simulation parameters and replacing Geant4 showers with libraries of
frozen showers. Towards that end, ATLAS and CMS agreed on a set of
simulation configurations and physics input samples:

\begin{itemize}
\item
  Machine: olhswep16.cern.ch (CERN's OpenLab), one thread runs
\end{itemize}

\begin{itemize}
\item
  Compiler: gcc 6.3
\item
  Geant4: version 10.2, FTFP\_BERT physics list
\item
  Pythia {[}34{]} events: 13 TeV Pythia minimum bias (300 events) and
  ttbar (300 events), pseudo rapidity cut
  \textbar{}η\textbar{}\textless{}5.5
\item
  Particle gun: 50 GeV e's, muons, pions with a flat η,φ distribution in
  η={[}-0.8,0.8{]}, η={[}2,2.7{]}, and φ={[}0,2π{]}
\item
  Geometry: 2015 or 2016
\end{itemize}

From this reference configuration, various options are incorporated to
study the impact on time and memory performance, including Russian
Roulette, shower libraries, timing cuts, and particle production cuts
per region. For the reference configuration, each experiment uses its
own timing and production cuts, as well as its own choice of stepper and
tracking parameter values. The experiments also study the evolution of
the time and memory performance of the Full Simulation application as
the detectors are upgraded from the Run 1 and Run 2 configurations to
several potential HL-LHC configurations.

The differing structures of the ATLAS and CMS detectors means that even
in stand-alone Geant4 the simulation times are very different. Different
detectors imply different CPU demands. This is clearly demonstrated by
the information from CMS in Table 1. This shows the CPU time and memory
measurements for different CMS HL-LHC candidate detector configurations
with respect to the 2015 reference geometry. HL-LHC CMS configurations
demand 28-36\% more CPU cycles than 2015 or 2016 configurations to
simulate minimum bias events, while they require 57-65\% more for ttbar
events. Given that the physics of the simulation at the core of the CMS
HL-LHC upgrade, the High Granularity Calorimeter (HGCAL), is still not
fully validated awaiting further test beam experiments, new physics
accuracy requirements might push the CPU needs higher. The reason is
that the HGCAL is a significantly more complex apparatus than the
current detector using different technology and materials, covering the
{[}1.5,3{]} pseudorapidity range, consisting of 40 layers of silicon and
copper/tungsten, brass, or steel, with a total of 6 million channels
with cell sizes of 0.5 cm\textsuperscript{2} and 1
cm\textsuperscript{2}. Owing to much more modest detector revisions, the
difference in simulation performance for the current Run 2 and upgraded
ATLAS detector is 5\% for ttbar events and 15\% for minimum bias events.
This is simply a reflection of the geometry differences in the inner
detector and muon system, in the case of ATLAS, and no additional
penalty is expected from significant revisions to physics models.

Tables 2, 3 show the CMS time performance of the Geant4 module of the
Full Simulation application, that is excluding physics generation,
digitisation, and reconstruction, for the default configuration and
several options to reduce CPU time consumption in production of Pythia
{[}34{]} generated minimum bias and ttbar events. Table 2 lists the
individual impact of each of the options, while Table 3 shows the
cumulative effect. The conclusion is that small gains may be achieved by
utilizing static libraries and by re-optimizing timing or particle
production cuts. In the CMS simulation Geant4 time cuts and tracking
cuts are optimised per detector region. The cut in range value for the
production of secondary particles is 1 mm in the electromagnetic and
hadronic calorimeters, 0.01 mm in the pixel detector, 0.1 mm in the
strip tracker, 0.002 mm in the muon sensitive volumes, and 1 cm in the
support structures. The propagation time cut is 500 ns. The tracking cut
is applied for charged particles below 2 MeV inside the vacuum chamber
to avoid looping electrons or positrons. The Russian roulette technique
applied inside calorimeter regions gives a 33\% speedup and is already
utilised in current detector configurations for Run-2, as is also the
case of the HF shower library, which yielded a 68\% gain in time
performance. Additional CPU time savings are achieved by optimizing
parameters associated with particle tracking in EM fields and production
of secondary particles.

Tables 4 and 5 shows the relative CPU performance of a number of
configurations of the ATLAS simulation application. The largest gain
comes from introducing shower libraries in the forward calorimeter,
which comes at a cost of 430MB of VMEM per job. No other option applied
here has an appreciable effect on memory requirements of the
application. The neutron time cut shows only a small CPU improvement,
but was shown to reduce the size of the output file by almost a factor
of two in previous tests. This file size issue can also be mitigated by
various approaches to hit storage, however. Removing range cuts
decreases the simulation time per event, because the thin volumes of the
sampling calorimeter demand a tighter range cut, including 30um in the
forward calorimeter and 300um in the EM calorimeters, to provide an
accurate physics description. This penalty is not present in the CMS
simulation, since the CMS EM calorimeter has much larger active volumes.
Still, the effect of range cuts on the two applications is similar.

Having exhausted the most obvious optimisations, CMS is exploring
various opportunities to improve computing performance in various ways
e.g. by testing the newest Geant4 versions, by using the most advanced
compilers, and by implementing an energy dependent ``smart tracking''
algorithm for propagation of particles in EM fields. However, the
biggest gains will most probably come from ongoing R\&D efforts, such as
the vectorised geometry library, VecGeom, which was recently released
for use with Geant4. Although VecGeom is optimised for GeantV, where it
promises a significant speedup by benefiting from data locality and SIMD
vectorisation, preliminary tests with Geant4 in CMS give a 5-10\% time
performance gain coming from improvements in the geometry algorithms. In
other words, although VecGeom can only be run in scalar mode with
Geant4, it still delivers significant time performance gains. The GeantV
project will offer more opportunities in the future, with the
possibility of integrating navigation and EM physics libraries to the
Geant4 CMS simulation application. The promise of a factor of 2-5 time
performance improvement using GeantV's full track-level parallelisation
approach and SIMD vectorisation is very enticing, and plans are being
made for a potential migration within the timeframe of the start of the
HL-LHC run in 2026, if the GeantV computing performance targets can be
met.

ATLAS is also exploring a number of technical and physics options to
improve simulation performance. Several of these options were delayed in
order to complete a migration to a new multi-threading friendly
infrastructure and to use a cmake build system. Russian Roulette for
neutrons and static library building are both currently being explored,
and it is expected that PGO will be investigated after those are
complete. Some further optimisation of range cuts is to be validated in
order to gain a few percent of CPU performance. ATLAS has also
investigated the new geometry modules available in the Geant4 10.4 beta
release and found in some cases significant CPU improvements resulting
from these (5\%-level). They are expected to be validated and moved into
production soon. VecGeom will be investigated as well.

When considering which optimisations to implement, or even the order in
which they should be attempted, the total amount of effort required to
bring each one into production should be taken into account, as well as
the overall impact on resource requirements.

An assessment of the effort required, in addition to the obvious cost of
how much person-power is required to implement an optimisation, should
also include how complicated it is to validate. For example static
compilation of Geant4 should come with no change in physics, so it is
easy, in principle, to validate. Conversely changing range cuts will
very often change the physics and so will require a lot more careful
examination by a larger group of experts before it can be signed off. A
modest CPU improvement may not be worth the risk of reproducing a large
Monte Carlo dataset if a problem is found with the physics of the
simulation.

When making an assessment of the overall impact on resources, the
potential gains in CPU usage and potential reductions in disk usage
should be considered alongside other potential issues. For example, if
samples simulated with the new optimisation cannot be combined with
older samples, then a new simulation campaign with new calibrations
might be required to benefit from the new optimisation.

\includegraphics[width=6.06094in,height=3.03532in]{media/image21.jpg}

\emph{Table 1: Time and memory performance of the Geant4 module of the
CMS Full Simulation application for minimum bias and ttbar events in
current and future detector configurations. The Geant4 version used in
the test is 10.2p02. With respect to the 2015 and 2016 configurations,
which are very similar, 2017 adds an upgraded pixel detector and
modifies the forward calorimeter layer configuration, 2023D4 and 2023D17
are different versions of the HL-LHC CMS detector with an upgraded
tracker and a High Granularity end cap calorimeter (HGCAL), the latter
sub-detector involving a significant increase in the number of Geant4
volumes. }

\includegraphics[width=7.59701in,height=3.65104in]{media/image22.jpg}\emph{Table
2: Relative time performance of the Geant4 module of the CMS Full
Simulation application for minimum bias and ttbar production.
Measurements of the individual effect of several options to reduce CPU
time consumption are listed. The Geant4 version used in the test is
10.2.p02. Values of range cuts and tracking cuts are specified in the
text.}\includegraphics[width=6.81771in,height=3.90017in]{media/image23.jpg}\emph{Table
3: Relative time performance of the Geant4 module of the CMS Full
Simulation application for minimum bias and ttbar production.
Measurements of the cumulative effect of several options to reduce CPU
time consumption are listed. The Geant4 version used in the test is
10.2p02. Values of range cuts and tracking cuts are specified in the
text.}

\begin{longtable}[]{@{}lll@{}}
\toprule
\endhead
Configuration & MinBias & ttbar\tabularnewline
Nominal production configuration: shower libraries in the forward
calorimeter, nominal range cuts, NystromRK4 stepper, FTFP\_BERT\_ATL
physics list, 250ns neutron time cut, simulation of primary particles
with pseudo-rapidity below 6.0 & 1.0 & 1.0\tabularnewline
No shower libraries & 1.5 & 1.3\tabularnewline
ClassicalRK4 stepper instead of NystromRK4 & 1.09 & 1.07\tabularnewline
No neutron time cut & 1.02 & 1.01\tabularnewline
FTFP\_BERT instead of FTFP\_BERT\_ATL physics list & No change & No
change\tabularnewline
No simulation of primaries with pseudo-rapidity above 5.5 & 0.94 &
0.95\tabularnewline
All range cuts set to 1mm & 0.92 & 0.90\tabularnewline
\bottomrule
\end{longtable}

\emph{Table 4: Performance of various configurations of the ATLAS
simulation for minimum bias and ttbar production events. The Geant4
version used for this test was G4 10.2p03. No significant performance
improvements were introduced in patch 03 with respect to patch 02.}

\begin{longtable}[]{@{}lll@{}}
\toprule
\endhead
Configuration & MinBias & ttbar\tabularnewline
Nominal production configuration: & 1.0 & 1.0\tabularnewline
No shower libraries & 1.5 & 1.3\tabularnewline
No shower libraries + use ClassicalRK4 stepper & 1.6 &
1.4\tabularnewline
No shower libraries + use ClassicalRK4 stepper + No neutron time cut &
1.6 & 1.4\tabularnewline
No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
use FTFP\_BERT physics list & 1.6 & 1.4\tabularnewline
No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
use FTFP\_BERT physics list + All range cuts set to 1mm & 1.5 &
1.3\tabularnewline
No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
use FTFP\_BERT physics list + All range cuts set to 1mm + No simulation
of primaries with pseudo-rapidity above 5.5 & 1.4 & 1.2\tabularnewline
\bottomrule
\end{longtable}

\emph{Table 5: Performance of various configurations of the ATLAS
simulation for minimum bias and ttbar production events. Measurements of
the cumulative effect of several options to reduce CPU time consumption
are listed. The Geant4 version used for this test was G4 10.2p03. No
significant performance improvements were introduced in patch 03 with
respect to patch 02.}

\hypertarget{lhcb-and-alice}{%
\subsubsection{8.3 LHCb and Alice}\label{lhcb-and-alice}}

Although the use of CPU grid resources by LHCb is two orders of
magnitude below ATLAS', the expectation for the 2020 LHC run and beyond
is to use up to 90\% of its allocation for Monte Carlo simulation. In
Run2 the experiment has adopted the concepts of split High Level
Triggers (HLT) and of the TURBO stream that integrate real-time
calibration and alignment into the data taking process, thereby
performing an online reconstruction that is equivalent to offline.
Nevertheless LHCb is currently exploiting all resources at its disposal
for simulation, including the use of its Online Farm when idle. The full
readout of the detector at 40 MHz in the LHCb upgrade will have a major
impact on trigger and offline computing systems and the experiment will
have to cope with higher rates, up to 2-5 GB/s from the detector. Hence
the concepts above will be further exploited moving the event
reconstruction and selection as close as possible to the online farm.
The experiment plans to match the amount of physics events collected
with the same order of simulated events. Exploring faster simulation
options is therefore a primary focus of the LHCb software efforts. The
experiment has yet to evolve its framework for parallelisation and is
currently migrating to a version of Geant4 that supports multithreading,
with the promise to achieve significant memory savings. Potential gains
from static library linkage and re-optimisation of simulation parameters
will also be investigated. The plan is to evolve the software into an
integrated simulation framework which allows to choose and combine
different fast simulation options. Table 4 shows the percentage of the
CPU time spent on each sub-detector systems for different particle types
in a given version of the software The RICH detector took almost 50\% of
the time followed by the Ecal with about 26\%. Optical photons took the
largest fraction of the total time, followed by photons, and electrons.
Re-optimisation of the RICH modeling has already been performed reducing
the time of the RICH simulation to 20-30\% in future production
versions.

\includegraphics[width=6.82813in,height=2.20063in]{media/image24.jpg}

\emph{Table 4: Percentage of the CPU time spent on each sub-detector
systems for different particle types for a given version of the
software.}

ALICE also consumes two orders of magnitude less CPU resources than
ATLAS. Figures 16 and 17 show charts of the percentage of time spent in
different elements of the detector simulation chain, including the
Geant4 and digitisation modules, and in the different library
components, respectively. The challenge for ALICE is to reduce the time
spent in Geant4 simulation (55\% of the total simulation time),
particularly in the geometry package and math functions, and re-optimise
the stepping algorithms. Digitisation takes a large fraction of the CPU
resources in ALICE (35\%) and is therefore also a target of ongoing R\&D
to improve the modeling of the TPC readout and the IO, and to explore
the benefits from SIMD vectorisation.

\includegraphics[width=5.37135in,height=2.64583in]{media/image25.jpg}

\emph{Figure 16: Approximate percentage of time spent in different
components of the ALICE detector simulation chain, including the Geant
and digitisation modules (for a representative Pb-Pb Hijing event)}

\includegraphics[width=4.52441in,height=2.60938in]{media/image26.jpg}

\emph{Figure 17: Approximate percentage of time spent in the different
library components of the ALICE detector simulation chain for the same
setting as in Figure 16.}

\hypertarget{neutrino-experiments}{%
\subsubsection{8.4 Neutrino Experiments}\label{neutrino-experiments}}

Accelerator-based neutrino experiments use ``detector'' simulation tools
in two different places. Beam simulations are used to model primary
proton beam interactions with the target and the creation, integration
and decay of hadrons that produce neutrinos. For this step, most
experiments use Geant4, although some use Fluka {[}8{]} or MARS {[}9{]}.
Neutrino-nucleus interactions inside the detectors are taken care of by
neutrino generator tools, such as GENIE {[}7{]}. The actual Geant4-based
detector simulation stage takes over from the generator to propagate the
final state particles through the detectors. Neutrino detectors divide
into three broad categories: segmented scintillator detectors, liquid
argon TPC's, and Cherenkov detectors.

Beam simulations generally do not require significant computing
resources. A total of 100 million events, produced at a rate of 0.01
sec/event and using much less than 1 GB per job, is generally sufficient
for a production pass. More resources are needed for assessing beam
uncertainties, simulating muons in muon monitors, and for beam design
optimisation.

For the MINERvA experiment, the whole simulation chain from neutrino
beam modeling to the production of the analysis files take about 20
sec/event, with 25-35\% of this time spent in the Geant4 module. The
required memory is typically less than 2 GB per job. The NOvA experiment
used 13 million CPU hours in 2016, 70\% of which were spent on
simulation, and less than 2 GB of memory per job.

For liquid argon TPC detectors, all Fermilab experiments use the LArSoft
software {[}21{]}, which contains the simulation and reconstruction
packages. LArSoft includes both the Geant4 module and a ``DetSim''
module that handles the conversion from energy to ionisation charge, the
scintillator light, and the propagation of charge and photons. The
MicroBooNE experiment spends approximately 3 min/event in the Geant4
module and 1 min/event the DetSim module, with a memory footprint per
job of 4.6 GB and 2.3 GB respectively. Table 5 shows a summary of CPU
time measurements for the ProtoDUNE and DUNE Far Detector (DUNE FD)
simulation applications. Although the DUNE FD is much larger (40 kTon)
than ProtoDUNE (770 tons), ProtoDUNE will operate on the surface and
requires simulation of cosmic backgrounds. It is also not necessary to
simulate the entire 40 kTon DUNE FD in every event. For these reasons,
Geant4 simulation of the DUNE FD is substantially faster than ProtoDUNE.
Memory consumption is 2 (3) GB per job for the Geant4 module and 1.3 (1)
GB per job for the DetSim module.

In summary, computing performance is not the dominant challenge for the
simulation of scintillator-based neutrino experiments. However, the
TPC's liquid argon and segmented detector technology under consideration
for the DUNE near detector (DUNE ND) may pose significant new demands on
computing resources. The issue of liquid argon simulations requiring
large amounts of memory will be mitigated by evolving the \emph{art}
software framework {[}35{]} to support multithreading.

\begin{longtable}[]{@{}llllll@{}}
\toprule
\endhead
\begin{minipage}[t]{0.14\columnwidth}\raggedright
\begin{quote}
\textbf{Experiment / Simulation Stage CPU} ( sec/event)
\end{quote}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
\textbf{Generation}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
\textbf{Geant4}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
\textbf{Geant4 with space charge}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
\textbf{DetSim}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
\textbf{Reco}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.14\columnwidth}\raggedright
\begin{quote}
\textbf{ProtoDUNE} (Cosmic events with Corsika {[}36{]})
\end{quote}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
10\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
180\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
480\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
25\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
1000\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.14\columnwidth}\raggedright
\begin{quote}
\textbf{DUNE FD} (v\textsubscript{e} events with GENIE)
\end{quote}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
0.04\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
4\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
44\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
160\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\emph{Table 5: Summary of CPU time measurements for the different
modules associated with the ProtoDUNE and DUNE Far Detector (DUNE FD)
simulation applications.}

\hypertarget{section-4}{%
\subsubsection{}\label{section-4}}

\hypertarget{direct-dark-matter-search-experiments}{%
\subsubsection{8.5 Direct dark-matter search
experiments}\label{direct-dark-matter-search-experiments}}

As is the case for neutrino and muon experiments, the direct dark-matter
search experiments do not face significant computing performance
challenges. Given the typical event topology (very few tracks of limited
energy), CPU and memory consumption are not considered a particular
concern, with the exception of some specific requirements. There is,
instead, a large interest in improved and extended physics modeling.

The issues associated with physics modeling can be summarised as:

\begin{itemize}
\item
  For the liquid noble gas experiments an extension to the Geant4
  scintillation process {[}37{]} is usually employed
\item
  For cryogenic crystal detectors the simulation of phonons and their
  interaction becomes important. Geant4 has been extended to support
  these processes {[}38{]}
\end{itemize}

The specific CPU-related elements are:

\begin{itemize}
\item
  High-precision neutron transport may be needed in some cases. The
  associated physics models are computationally more intensive than the
  default Geant4 models.
\item
  Optical photon transportation may become a CPU bottleneck for
  experiments with a detailed and complex optical model. Advanced
  techniques leveraging GPUs are under investigation {[}39{]}.
\end{itemize}

\hypertarget{rd-towards-improving-software-performance}{%
\section{9 R\&D Towards Improving Software Performance
}\label{rd-towards-improving-software-performance}}

To meet the challenge of improving the performance by a large factor, an
ambitious R\&D programme is underway to investigate each component of
the simulation software for the long term. Here we will discuss the
activities that use the latest state of the art techniques to improve
computing performance. In other words, we will include R\&D efforts
other than optimisation of compiler or library settings, shower library
replacements, and optimisation of Geant4 tracking parameters or
production cuts. The use of fine grained parallelisation techniques,
novel computing architectures, and HPC systems are within the scope of
this section.

The CPU time that is spent in the simulation of LHC detectors is
typically dominated by the physics components, and in particular by the
electromagnetic part. The development of electromagnetic showers in
calorimeters plays a major role for the CPU performance of simulations.
This is also true for jets because hadronic showers, induced by the
hadron components of jets, have electromagnetic components coming from
the decays of neutral pions. The geometry component as well as particle
propagation in fields fill also important slices in the overall
performance pie-chart of simulation, as can be observed in Figure 14.
These observations are showing in which areas the R\&D effort should be
mainly focused in the short and medium term from the perspective of
detailed simulation.

Improving the CPU performance is a major milestone of simulation R\&D
and the related program of work is mainly bound to the LHC schedule for
the high luminosity phase. Adapting the simulation workflow and
algorithms to make better profit from the FLOPS potential of modern
architectures is very important. In the following we describe in detail
some of the studies to be performed in the next 3-5 years.

\hypertarget{particle-transport-and-vectorisation}{%
\subsubsection{9.1 Particle Transport and
Vectorisation}\label{particle-transport-and-vectorisation}}

One of the most ambitious elements of the simulation R\&D programme is a
new approach to managing particle transport, which has been introduced
by the GeantV project. The aim is to deliver a multithreaded vectorised
transport engine that has the potential to deliver large performance
benefits. Its main feature is track-level parallelisation, bundling
particles with similar properties from different events to process them
in a single thread. This approach, combined with SIMD vectorisation
coding techniques and improved data locality, is expected to yield
significant speed-ups, which are to be measured in a realistic prototype
currently under development {[}40{]}.

For the GeantV transport engine to display its best computing
performance, it is necessary to vectorise and optimise the accompanying
modules, including geometry, navigation, and the physics models. These
are developed as independent libraries so that they can also be used
together with the current Geant4 transport engine. Of course, when used
with the current Geant4 they will not expose their full performance
potential, since transport in Geant4 is currently sequential, but this
allows for a preliminary validation and comparison with the existing
implementations. The benefit of this approach is that new developments
can be delivered as soon as they are available. The new vectorised
geometry package (VecGeom), developed as part of GeantV R\&D and
successfully integrated into Geant4, is an example that demonstrated the
benefit of this approach. By the end of 2018 it is intended to have a
proof-of-concept for the new particle transport engine that includes
vectorised EM physics, vectorised magnetic field propagation and that
uses the new vectorised geometry package. This will form a sound basis
for making performance comparisons for simulating EM showers in a
realistic detector. The \emph{beta} release of the GeantV transport
engine, expected by end 2019, will contain enough functionality to build
the first real applications. This will allow performance to be measured
and give sufficient time to prepare for HL-LHC running. It should
include the use of vectorisation in most of the components, including
physics modelling for electrons, gammas and positrons, whilst still
maintaining simulation reproducibility, and I/O in a concurrent
environment and multi-event user data management.

Making use of the SIMD pipelines available today even in commodity PC's
has been already investigated at large in the GeantV project. An
important lesson resulting from this R\&D is that vectorizing the
computing intensive algorithms of the program is required but not
sufficient for enabling significant overall gains due to SIMD pipelines.
The entire simulation workflow needs to be adapted to provide the
multiple data required by vectorisation. GeantV prototyped an approach
that splits the stepping procedure for tracks into stages, aiming to
accumulate many particles before actually performing the actions
involved by the stage. As illustrated In Figure 18, the generation of
the final states from discrete physics processes can be such a
simulation stage. Particles entering it are directed to different
physics model handlers that accumulate tracks into baskets, executing
the different model algorithms in multi-particle mode. This kind of
design changes the traditional stack-based execution into a multi-stream
execution pattern. It creates the premises for increasing the code and
data locality while allowing to feed multiple data to possibly
vectorised code.

Such reshaping of the simulation has several practical implications, out
of which many are affecting directly the user code. Mixing tracks from
multiple events is required for sustaining the basketised flow. To be
noted that several current or future experiment frameworks will support
track-level parallelism as well. This will affect the data management
and I/O on the user side, requiring concurrent bookkeeping and
management of multiple events in flight. Extending the user API with
vector signatures will allow vectorised user code to exploit the
particle-level parallelism supported by the framework..

The support for an alternative approach to particle transport will also
affect the strategy for the development and long term maintenance of
physics models. An important objective of the physics developers is
therefore to review the physics algorithms and their implementation in
order to better adapt them to a multi-particle flow. This will allow a
better use of the system caches even without vectorised algorithms. For
such revisited models, GeantV is offering a testbed allowing to turn on
the basket flow and benchmark the impact of vectorisation. It is of
utmost importance that newly developed or revisited physics models are
easily ported to both types of particle transport, and this would need
to be reflected in the program of work. This will also facilitate the
validation of the new developments and benchmarking their performance in
a ``basket flow'' environment.

\includegraphics[width=6.26772in,height=3.52778in]{media/image27.png}

Figure 18: \emph{The stepping procedure in particle transport is the
sequence of actions allowing to propagate a particle from a starting
point to the next point where the state of the particle changes
meaningfully (e.g. physics process occurrence or geometry crossing). In
GeantV, stepping is decomposed into stages, each stage selecting among a
set of models. Tracks can be buffered and sent in form of baskets to
vectorised models, which are allowed to coexist with scalar models. This
approach improves the locality and cache coherency compared to the
classical stack-based approach.}

The required improvement in CPU performance cannot be achieved only from
algorithmic improvements, structural changes and multiparticle flow.
This also involves the development of fast sampling algorithms as well
as Fast Simulation approaches that bypass the simulation of full
showers. The project aims to investigate further the ways in which Full
and Fast simulation capabilities can be integrated in a single detector
simulation framework.

The initial goal is to test the transport engine with the vectorised
geometry package and demonstrate a significant speedup factor. For
simplicity, the project started by following a ``tabulated physics''
approach using values derived from Geant4, thus enabling equal footing
comparisons between the two. Tests under this approach are performed to
ensure the consistency and validity of the time performance results.
Tabulated physics is now being replaced with a physics framework using
improved and vectorised physics models, which are being validated as
they become available. Full-scale physics validation will be performed
both on available ``thin target'' data and in the context of realistic
detector applications and in collaboration with the experiments.

The vectorised components show substantial speed-ups with respect to the
corresponding code in Geant4, sometimes exceeding by one order of
magnitude. This is the case for certain solids within the VecGeom
geometry package {[}41{]}, and is planned to be extended to field
propagators and certain EM physics models. While this result is very
promising, it still does not constitute an authoritative benchmark,
although it gives confidence that the stated project goal of reaching a
significant global speedup is well within reach.

Some of the speed-up will come from optimised, vectorised physics
models, which are being developed with the objective of improving
physics accuracy and to respond to the requirements of HL-LHC and the
future intensity and energy frontier experiments. Numerous branches in
the physics code, especially in hadronics physics, pose a major
challenge to vectorisation. Since EM processes dominate showers and have
relatively fewer branches, they are the first targets of optimisation
and vectorisation. Low energy neutron and elastic hadron scattering
processes are also more amenable to vectorisation.

The estimate for the remaining work to complete the first version of the
EM physics library is of order of 1.5 FTE-years, while the work to
implement a complete first version of hadronic physics is much larger.
Current plans are to ``wrap'' an existing Geant4 hadronic physics model
to allow meaningful physics and time comparisons. This effort is
estimated at 2 FTEs.

\hypertarget{hybrid-computing-and-simulation}{%
\subsubsection{9.2 Hybrid Computing and
simulation}\label{hybrid-computing-and-simulation}}

Hybrid computing refers to sharing work across a mixture of computers
with different architectures. This approach seeks to exploit the
benefits of multithreading and fine grained parallelism, and needs a
software framework for writing code that executes across heterogeneous
platforms, which at the moment may include CPUs, GPUs, MICs, etc. In
detector simulation, parallelism is achieved at the level of events, as
in Geant4, or at the level of tracks, as in GeantV. Monte Carlo
techniques for detector simulations are implemented as sequential
algorithms because each \emph{step} depends on the \emph{history of
previous steps}. Although the use of these emerging technologies in
detector simulation remains a challenge R\&D efforts are to be
aggressively pursued, given the potentially large gains to be obtained.

One of the most time consuming components common to the majority of
detector simulation applications is neutron tracking, which needs to be
precise. Neutron tracking has some characteristics that make it a good
candidate to explore acceleration with the use of co-processors:

\begin{itemize}
\item
  there are many particles of the same type in the simulation;
\item
  the number of physics branches (e.g. different physics processes)
  involved in the simulation of these particles is limited;
\item
  each particle goes through a large number of steps though the detector
  material before being absorbed or leaving the detector;
\item
  when eventually an interaction occurs, the number and species of
  generated secondary particles that will further contribute to the
  signal is relatively limited.
\end{itemize}

Previous experience {[}42{]} has shown that large speedup factors can be
achieved for tasks that are well-suited to GPGPU technology: simplified
geometry with only few materials, limited number of particle species and
physics processes. One example is the R\&D effort started by the Geant4
Collaboration to extend the existing code to the transport of neutrons.
One of the unique characteristics of this approach is that this system
can be integrated in existing applications, replacing only the specific
components with independent libraries. This approach is particularly
effective for HEP experiments because it allows for a non-disruptive
evolution of the simulation code that extends the thoroughly validated
current versions of the Geant4 neutron physics models (see Figure 19).

\includegraphics[width=6.70522in,height=1.38889in]{media/image28.png}

\emph{Figure 19: Proposed architecture of a specialised library to
leverage GPU for neutron transport.}

The first target of the described approach is the use of GPUs for
thermal neutron transportation. While the CPUs continue to process the
rest of the particles, the processing of all secondary thermal neutrons
would be offloaded to GPUs. Thermal neutrons propagate in the detector
and cavern for a long time before eventually interacting and generating
detector signals. Given that the latency is large, from 10 seconds to
minutes, thermal neutron depositions cannot always be associated with
the original event such that the neutron flux is the only meaningful
output information from thermal neutron transport. As granular I/O is
known to be one of the dominant obstacles against efficient use of GPUs,
a hybrid simulation with minimal I/O utilisation offers an ideal
opportunity for making efficient use of GPUs.

Along the same lines, the GeantV project achieved speedup factors of up
to 30 with respect to Geant4 when processing a basket of electrons and
photons in a GPU {[}43{]}. This is explained by the fact that EM
interactions in a detector evolve into a self-contained cascade of
electrons and photons that can exploit the benefits of vectorised code,
SIMD, and data locality. In other words, electrons and photons
traversing the same volume and getting applied the same instructions are
bundled in a single basket processed in a single thread on a GPU. The
caveat is that the baskets need to contain at least ten thousand tracks
in order for the factor of 30 to be achieved, due to the long transfer
times to and from the GPU.

\hypertarget{modularisation}{%
\subsubsection{9.3 Modularisation}\label{modularisation}}

Starting from the next release, a modularisation of Geant4 is being
pursued that will allow an easier integration in experimental
frameworks, with the possibility to include only the Geant4 modules that
are actually used. A further use case is the possibility to use one of
the Geant4 components in isolation, e.g., to use hadronic interaction
modeling without kernel components from a fast simulation framework. As
a first step a preliminary review of libraries granularity is being
pursued, which will be followed by a review of intra-library
dependencies with the final goal of reducing their dependencies. By
2019, it is intended to redesign some Geant4 kernel components to
improve the efficiency of the simulation on HPC systems, starting from
improved handling of Geant4 \emph{database}s on large core-count
systems. A review will be made of the multithreading design to be closer
to the task-based frameworks, such as Intel's TBB.

\hypertarget{simulation-on-high-performance-computing-systems}{%
\subsubsection{9.4 Simulation on High Performance Computing
Systems}\label{simulation-on-high-performance-computing-systems}}

HPC resources at supercomputer centers can provide a substantial
fraction of the needed computing power to address HL-LHC needs. The
Geant4 Collaboration has started an R\&D program of code optimisation
for a more efficient utilisation of HPC resources. One of the main
features of the next generation of supercomputers is the use of
massively parallel architectures that require the combination of novel
techniques to achieve scalability. The release of multi-threaded Geant4
in 2013 has shown that it is possible to obtain almost perfect linear
scaling, whilst keeping memory usage under control. Subsequent research
has focused on extending the parallelism to multi-node with HPC-friendly
techniques (MPI), using a mixed MT/MPI approach. A recent result has
shown that scaling for hundreds of thousands of concurrent threads can
be achieved for a semi-realistic HEP-Geant4 application. A test
performed on the MIRA supercomputer at ANL successfully ran a modified
version of Geant4 spawning more than 3 million threads {[}44{]}. The
test concluded that the main bottleneck to achieve scaling linearity is
concurrent access to I/O resources (see Figure 20). Extensions to Geant4
capabilities to improve the scalability of the code are underway. These
include:

\begin{itemize}
\item
  parallel merging of scorers/histograms and ntuples via MPI;
\item
  redesign of RNG Seeding in large parallel applications;
\item
  investigation of use of HPC-friendly data-storage for input data-base
  libraries.
\end{itemize}

A custom version of Geant4 developed for tests on MIRA showed
scalability for up to 3 million threads. This version is the basis for
further development to be included in future Geant4 public releases. A
spinoff result from the MIRA tests has been the validation of Geant4
physics to 1/10\textsuperscript{5}, a level of precision never achieved
before.

ATLAS is actively pursuing HPC resource utilisation to supplement grid
resources for simulation production. Multi-process jobs distributed with
YODA have been running successfully in production on supercomputers at
NERSC and machine utilisation is being studied {[}45{]}. ATLAS is
currently migrating its simulation application to multi-threading to
ensure good scaling performance on HPC architectures with minimal memory
footprint. Current measurements of multi-threaded event throughput and
memory consumption scaling on Intel Xeon Phi Knights Landing chips are
very promising, demonstrating ability to scale well up to the maximum
number of threads on a device with a factor five decrease in the
per-worker memory consumption relative to multi-process jobs {[}46{]}.
Overall throughput performance falls short of traditional Xeon
architectures, however, so studies are ongoing to identify and remove
the performance bottlenecks.

CMS has also been studying how best to exploit HPC resources by running
combined multi-step jobs that cover Generation, Simulation, Digitisation
and Reconstruction in a single workflow. The results when scaling to
high-pileup scenarios have shown that limitations are primarily due to
the reconstruction step, rather than the simulation step. Studies have
been made of various ways of filling a whole KNL many-core node with
Generation and Simulation jobs of varying sizes and the highest total
node event throughput was achieved with two jobs using 128 threads each.
The simulation application and its underlying simulation engine, Geant4,
is not limiting scaling behaviour.

\includegraphics[width=6.26772in,height=4.625in]{media/image29.png}\emph{Figure
20: Top: Initial testing of Geant4 on MIRA (blue diamonds) and VESTA
(green stars). Each node has 80 threads. The loss of scalability is due
to I/O and intra-node communication limitations (MIRA nodes do not have
disks and rely on MPI for every I/O operation). VESTA, a smaller
test-bed, shows better results because the number of I/O nodes is
larger. }

\hypertarget{section-5}{%
\section{}\label{section-5}}

\hypertarget{section-6}{%
\section{\texorpdfstring{\\
}{ }}\label{section-6}}

\hypertarget{fast-simulation}{%
\section{10. Fast Simulation}\label{fast-simulation}}

As pointed out earlier, simulation consumes the majority of the CPU time
used by many experiments. Optimisation studies on detailed simulation
codes are ongoing and can provide helpful performance improvements, but
in order to satisfy the needs of an entire experiment, other options
must be explored. The increase in size of the MC samples needed to match
those of the data collected by upgraded or future experiments, as well
as the increase in the complexity of the events, will put additional
strains on the CPU needs. While the availability of an accurate detailed
simulation will remain of paramount importance, often the full
high-fidelity description of the detector implemented by the experiments
is not necessary. There are many cases, including detector upgrade
studies, tests for evaluation of systematic uncertainties in physics
measurements, scans of large signal phase spaces in the context of new
theories, and prototyping of new analysis ideas, where the demand on
simulation physics accuracy may be relaxed. Even within a given physics
analysis, it may be good enough to evaluate some quantities with high
accuracy in a small MC sample and other quantities with lower accuracy
in larger samples. A combination of different fast simulation approaches
targeting parts or the whole simulation application are the optimal
solution for saving computing resources in these cases.

In general, simulation time scales with the energy of the particles
being simulated, and with the minimum energy down to which they are
simulated. This means that by far the most time is taken by showers in
calorimeters, where large numbers of low-energy electromagnetic
particles are produced and propagated by the simulation. Because the
simulation time scales with energy, rather than transverse energy, the
time consumed by LHC simulations is dominated by the forward regions,
which are the target of the highest energy particles in most collider
events. Fast simulations options for calorimeter showers have the
side-effect of reducing significantly the number of low energy neutrons
in the simulated event, which contribute significantly to the simulation
time due to their relatively long lifetimes. Particle propagation in
tracker detectors is typically the next most time-consuming step in a
simulation. This step consists mostly of charged particle propagation,
and the geometry detail is the most important aspect to be addressed by
the a fast simulation. Heavy Flavour physics experiments utilise Ring
Imaging Cherenkov (RICH) or Detection of Internally Reflected Cherenkov
light (DIRC) detector technology. The transport of Cherenkov photons
through optical processes needs to be simulated in detail in these
cases, and requires more CPU time than regular trackers. Fast simulation
of these processes is in general tied to the reconstruction technique
used in a given detector.

There are two main types of fast simulations. One is the replacement of
a particular physics process or set of processes with a module that
contains faster code. The other approach is the simulation of an entire
particle trajectory or shower, or even a more complex object like a
hadronic jet, with a faster software module. These faster modules often
consist of parameterised models of the physics processes or particle
trajectories. While the technique could be shared, the exact
implementation would be different for each experiment. Much of this
section focuses on the second technique, while the first is discussed
further in Section 10.3 on machine learning.

It is worth noting that the fast simulation techniques and software have
relevance outside the HEP domain. For example, radiation transport
techniques are extremely important for a wide range of applications:

\begin{itemize}
\item
  radiation protection in all industrial fields using ionizing
  radiations (sterilisation, imaging etc.);
\item
  radiation protection for space flights;
\item
  design and optimisation of imaging medical instruments;
\item
  treatment planning for radiation therapy;
\item
  radiation safety for nuclear power and fuel processing plants;
\end{itemize}

All these applications would greatly benefit from an increase in the
simulation speed with important benefits to their respective fields.

A way to achieve an additional speedup is to customise the simulation to
the needs of a specific physics analysis. This would allow to target a
given sub-detector or part of the event.

\hypertarget{fast-simulation-techniques-and-examples}{%
\subsubsection{10.1 Fast Simulation techniques and
examples}\label{fast-simulation-techniques-and-examples}}

There are many techniques to make simulation faster. Some of them are
applied to Geant4-based Full Simulation applications, such as for
example Russian Roulette and shower libraries, to speedup simulation in
a given sub-detector. In the literature, these applications are still
labeled Full Simulation and Geant4-based. Some experiments have
developed alternative fast simulation frameworks, based on fast track
propagation through a simplified geometry and parameterised showers,
which are less accurate and are typically used in detector upgrade
studies and to produce ``signal'' samples that require scans of a large
region of theory parameter space. Generally, the faster the simulation,
the more imprecise the physics is. Examples of fast simulation
techniques, used to speedup Geant4-based applications or in fast
simulation frameworks, are:

\begin{itemize}
\item
  \textbf{Russian Roulette:} Some fraction of the simulated particles
  are discarded and a weight is applied to the others to compensate for
  this effect. It is commonly used in the simulation of thermal
  neutrons, which have long lifetimes and consume a significant amount
  of CPU time while propagating through matter before being absorbed.
  Russian Roulette is used in CMS's Full Simulation application and
  yields 25\% (30\%) CPU savings when processing minimum bias (ttbar).
\end{itemize}

\begin{itemize}
\item
  \textbf{Shower libraries:} A particle of a given type and in a given
  energy range and geometry region is replaced with the simulation
  information of a similar particle stored in a library of showers
  pre-simulated with a detector simulation tool such as Geant4. Shower
  libraries are typically used to replace Full Simulation in
  sub-detectors with high particle occupancy, such as forward detectors
  in colliders. The are efficient to model high-energy showers where a
  large number of low-energy secondary particles are produced. If
  applied to the entire detector, they achieve speedup factors of 2-5.
  Both in ATLAS and CMS, shower libraries are used by default in the
  forward hadronic calorimeters, yielding a 25\%-40\% time performance
  gain in the case of CMS and a 30-50\% performance gain in the case of
  ATLAS. ATLAS tests indicated previously that extending the use of
  these libraries to the remainder of the calorimetry could bring gains
  of an additional 50\%. LHCb is currently exploring its use.
\item
  \textbf{Parameterisation of calorimeter showers}: The simplest
  implementation of a parameterised particle shower involves the
  parameterisation of the shower shapes of high- and low-energy EM
  particles. In sampling detectors with complex geometries, these
  parameterisations are difficult to tune. Additionally, with a
  relatively thick inner detector, and in the case of ATLAS a cryostat
  and solenoid in front of the active calorimeter region, there may not
  be many high-energy electromagnetic particles available to the fast
  simulation. This simplest approach, based on Ref. {[}34{]}, was
  abandoned by ATLAS about ten years ago. Current approaches used by
  both ATLAS and CMS involve more complex modeling of the showers than
  can be achieved by a simple parameterisation. Although not part of the
  current default simulation, CMS implemented shower parameterisations
  using GFLASH {[}33{]} to model the shower after the first interaction
  of each primary particle with matter, by tuning a set of parameters to
  fit the geometric distribution of the associated energy depositions
  obtained from a Geant4 simulation. The ATLAS approach uses histograms
  with energy deposit and shower shape information to directly treat
  particles before they enter the cryostat. All these approaches have
  the advantage that they can ignore, during energy deposition, the
  detailed geometry of the calorimeter, and deal only with the readout
  geometry. All have the option of building objects similar to the
  energy depositions from Geant4 or attempting to directly build
  summary-like objects that are used in digitisation. Typically,
  parametric simulation achieves speedup factors of 100-1000 in the part
  of the detector treated by the parameterisation.
\item
  \textbf{Low energy background parameterisation}: Muon detectors are
  located behind the calorimeter system preventing most high energy
  particles of a different nature to reach them. The main contribution
  to the occupancy in these detectors, apart from muons, is due to
  leakage from calorimetric showers and thermal neutrons. In order to
  obtain fully realistic occupancies in the muon system, photons,
  electrons and neutrons need to be transported to very low energy, and
  the surrounding infrastructure and accelerator elements must be
  modeled. In the LHCb experiment, the addition of this detail to the
  full Geant4 simulation would result in a CPU time increase of the
  order of a factor of 10. Instead, this low energy background is
  simulated via a parameterisation of the difference between a Full
  Simulation with and without the low energy processes. A similar method
  is being validated by the ATLAS collaboration.
\item
  \textbf{Fast tracker simulation}: Both ATLAS and CMS have custom fast
  track simulation algorithms based on simplified geometries. The time
  consumed by the Full Simulation of a tracking detector is primarily
  spent on the propagation of charged particles in a magnetic field,
  which is directly related to the number of times the field must be
  sampled. By simplifying the geometry, for example by considering
  infinitely thin layers, the number of samples is significantly
  reduced, resulting in a faster overall simulation. The other way to
  speed up the tracker simulation is by introducing fast, parameterised
  models for material interactions, particularly for the more common
  electromagnetic interactions. The fast track simulation code used by
  ATLAS has been made public as a part of the common tracking software
  project ACTS {[}47{]}, adopted by the FCC {[}48{]}. These fast track
  simulations offer speed improvements of 10-1000x over a detailed
  simulation, though for a Fast Simulation including both fast tracking
  and fast calorimeter simulation, tracking is typically the dominant
  resource consumer.
\item
  \textbf{Fast digitisation and reconstruction}: Especially for higher
  pileup at the LHC, once fast calorimeter simulation is used, and
  certainly when fast tracker simulation is included, digitisation and
  reconstruction become the dominant consumers of CPU time in the
  simulation chain. As a result, both CMS and ATLAS have implemented
  fast modules for specific pieces of the digitisation and
  reconstruction code that are most time-consuming, primarily in fast
  inner detector digitisation and fast tracking. Fast tracker hit
  reconstruction uses template hit smearing functions derived from full
  reconstruction, whereas fast tracking often uses generator (truth)
  particles as seeds to avoid the combinatorial issues of standard track
  reconstruction software. The speed improvements from these algorithms
  varies with pileup but are generally more than one order of magnitude.
  {[}49{]}
\item
  \textbf{Fully parameterised simulation:} Many programs are available
  in the public domain for the implementation of very fast parameterised
  simulation, the most popular being DELPHES and PGS {[}50,51{]}. These
  tools are extremely fast, running at hundreds of events per second,
  but they lack the accuracy of the alternatives discussed before. These
  programs are heavily used by phenomenologists for re-interpretation of
  LHC results, and by ATLAS, CMS, and LHCb for upgrade detector
  performance studies.
\item
  \textbf{Ultra fast self-tuning non-parametric simulations:} A
  re-emerging alternative trend in simulation is to develop ultra fast,
  self-tuning simulators based on lookup tables that directly map
  generator events into simulated events. The lookup tables in this case
  are formed directly from fully simulated events. A past example was
  Turbosim {[}52{]} developed at the Tevatron for the D0 and CDF
  experiments. A prototype for a redesign called Falcon {[}53{]} was
  recently released, with a more efficient algorithm for the
  multi-dimensional mapping of particle properties.
\end{itemize}

A natural evolution of having separate Full and Fast Simulation
frameworks in experiments is the concept of ``hybrid'' simulation (see
Sec. 9.2), where a single framework allows to treat some aspects of the
event with a fast or very fast simulation technique and others with a
detailed simulation approach. Such configurations might be useful for
handling pile-up and also for B-physics, where only a few particles are
of great interest to the analysis while it is helpful to have the others
simulated with some minimal accuracy. The difficulty with the hybrid
approach is that the optimal configuration is analysis dependent and
requires the derivation of customised calibration and efficiency
factors, activities which are person-power intensive.

LHCb has developed simulation code for reuse of the underlying event in
the case where only a few particles are of interest and the contribution
of the rest only degrades the `signal' measurement. The technique
consists of fully simulating the underlying event once, while the
`signal' is simulated N times independently and recombined with the one
underlying event. It allows a reduction in CPU time consumption
proportional to the number of times the underlying event is reused.

\hypertarget{physics-performance-of-fast-simulations}{%
\subsubsection{10.2 Physics Performance of Fast
Simulations}\label{physics-performance-of-fast-simulations}}

The fastest of the fast simulations are typically useful for situations
in which only approximate answers are needed, or where large
uncertainties are not a serious concern. Such situations are
surprisingly common. For example, studies of BSM limits on new particles
in LHC upgrade scenarios, where cross-sections fall rapidly with mass,
can afford a 100\% uncertainty on the analysis efficiency as it results
in a relatively small uncertainty on the excluded mass. The slower
simulation techniques generally provide higher fidelity, but they are
more resource intensive.

The performance of Fast Simulation in ATLAS, CMS, and LHCb is evaluated
with respect to the Geant4-based Full Simulation. For the fastest
version of ATLAS FastSim, and for CMS FastSim, the agreement is on
average quite good for kinematic distributions such as the momentum and
pseudorapidity of high level event objects, such as light jets, b jets,
missing transverse energy, muons, electrons, photons and taus. Larger
discrepancies are observed when comparing variables that depend on the
structure or shape properties, such as jet particle composition or
variables used for electron and photon identification. Fast simulations
have difficulty reproducing the detailed structure of hadronic shower
shapes, which create issues in modelling boosted objects and jet
substructure, which are very relevant signatures at higher energies.
Another common issue is in the treatment of exotic particles. Particles
with high charge (e.g. Q-balls), strange propagation (e.g. Quirks),
unusual nuclear interaction patterns (e.g. R-hadrons), and almost any
particles with late decays that result in displaced vertices or
disappearing tracks, for example, are difficult to treat with fast
simulations because of the assumptions built into the parameterisations.
For example, most fast simulations assume that particles are pointing,
while particles from late-decaying heavy particles may not be.

Fast simulations do have the advantage that they are generally more
tuneable than a detailed simulation toolkit like Geant4. The knobs for
tuning are generally better exposed to users (particularly for cases
like PGS and DELPHES), and they have a more clear connection with
physical observables. Thus, it can be easier to tune a fast simulation
to match the data than it is to tune a detailed simulation toolkit.

Figure 21 shows the speed of CMS Fast Simulation factored into different
subprocesses as a function of the number of pileup interactions. The
sub-processes are defined as follows:

\begin{itemize}
\item
  \emph{Trajectory sim.} refers to fast simulation of the particle
  trajectory through the detector, and formation of tracker and muon
  simulated hits.
\item
  \emph{Calorimeter sim.} refers to fast calorimeter simulation.
\item
  \emph{Tracker hits} refers to a FastSim specific simplified hit
  reconstruction in the tracker.
\item
  \emph{Pileup mixing} refers to the process of mixing hits from a
  physics process of interest with those from a minimum bias sample
  representing pileup.
\item
  \emph{Tracking} is the MC truth based, fast track finding algorithm.
\item
  \emph{Digitisation} is the simulation of readout, which is the same in
  fast and Full Simulation in CMS.
\item
  \emph{Calorimeter reco.} is the reconstruction of calorimeter hits,
  which is standard for all simulation and data.
\item
  Muon reco (not shown in the plot): is the reconstruction of muon
  detector hits, and takes on average \textasciitilde{}0.01s/event.
\item
  Finally, \emph{Object reco.} is the standard CMS reconstruction of all
  objects including vertices, jets, electrons, photons, muons and taus
  used in data and simulation.
\end{itemize}

\includegraphics[width=6.70522in,height=4.93056in]{media/image30.png}

\emph{Figure 21 Speed of CMS Fast Simulation factorised into different
subprocesses, as explained in Section 10.2, versus number of pileup
interactions. }

\hypertarget{machine-learning-in-fast-simulations}{%
\subsubsection{10.3 Machine Learning in Fast
Simulations}\label{machine-learning-in-fast-simulations}}

There has recently been a great deal of interest in the use of machine
learning in Fast Simulation, most of which has focused on the use of
multi-objective regression {[}54{]} and generative adversarial networks
(GANs) {[}55,56,57{]}. Since use of GANs allows for non-parametric
learning in cases such as calorimetric shower fluctuations, it is a
promising avenue for generating non-Gaussian and highly correlated
physical effects. This is an obvious area for future expansion and
development, as it is currently in its infancy. Just as with Fast
Simulation in general, one can imagine two types of machine learning
application to simulation. One is the replacement of a particular
physics process, or set of processes, with a machine learning module.
Such an approach has the clear advantage that it can be shared among the
experiments as a part of a general simulation toolkit like Geant4. The
approach has the disadvantage that, while it can be used to directly
replace a specific module of the toolkit by training against the
standard model, it otherwise might be rather difficult or cumbersome to
produce training data that can be used for the development of a clearly
distinct new model of physics. The other approach is the replacement of
a fast simulation technique with a machine learning module - that is, a
machine learning module that performs the entire simulation of a
particle, or even a more complex object like a hadronic jet. Machine
learning approaches also offer the opportunity for end-to-end simulation
where the output can represent composite objects resulting from multiple
simulation steps. While the technique, and perhaps even some details of
the network and training methods, could be shared, the exact
implementation would most likely have to differ among the experiments.
Discussions and sharing involved in this latter scenario is more
appropriate in an inter-experiment ``forum'' like the recently developed
Inter-experiment Machine Learning Working Group at CERN {[}58{]}. The
sharing of code and other technical details, in addition to ideas and
results, is essential in order to expand the benefits to the entire
community.

Most current work is focused on using multi-objective regression and
generative networks on simple events, such as single particle
calorimeter showers. Even though these are the most basic events to
simulate, more complex observables can be built as straightforward
combinations of single particle showers. In all cases, one must choose
an output, whether a general type like an energy deposit that might be
common to many experimental simulation applications, or a specific type
like the detector response that would presumably be applicable to only
one experiment. The technique and methodology, however, can be shared
among experiments even when their output formats differ.

The applications of machine learning are numerous, and the coming period
will involve a detailed exploration of many different approaches to
improving or speeding up the simulation using machine learning tools.
Within ATLAS, for example, neural networks are being explored as a
replacement for the histograms used in fast calorimeter simulation
mentioned earlier. This type of machine learning tool could also be
applicable to the storage of cross-sections within Geant4, for example.

While these techniques are quite promising, these are early days, and
detailed speculation about the adoption of machine learning across the
field is difficult. The fact that machine learning tools like
SciKitLearn {[}59{]} and TensorFlow {[}60{]} are being developed by
professionals and in an open-source community means that if the
simulation community takes full advantage of these publicly available
tools then the optimisations and developments of the broader community
will be immediately useful. Already, the parallelism of standard machine
learning tools used to implement GANs have made it possible to foresee
significant improvements from parallelism and vectorisation, even with
current simulation frameworks based on Geant4.

The first neural network-based fast simulations are now publicly
available, along with code, documentation, and training data
{[}56,57{]}. These first results are very promising and establish a
baseline for continuing to improve the performance in preparation for
integration into real experimental settings. Dialog between the
experiments has been, and will continue to be, important to push these
techniques forward.

There are two related approaches for replacing or augmenting fast
simulations with generative models. One approach is to construct
tailored architectures that are specifically designed for all or part of
a particular detector geometry and detector response. Another approach
is to develop generic tools that can be adapted for universal or common
aspects of simulation in many experiments. It is likely that a synergy
of both approaches will be required to extract the most performance from
state-of-the-art methods. The GeantV collaboration already has an active
program to develop generic tools for this second approach. Within the
GeantV project, it is planned to develop a general interface where a
user can specify a physics process and detector type and get back an
optimal trained machine-learning model that will simulate the detector
response. Such a tool, once developed and validated, will be usable with
any transport Monte Carlo, including Geant4 with its well-developed fast
simulation interface. The GeantV R\&D program also includes tests of the
generative neural networks in the context of Geant4. The tool will
incorporate various trained machine learning models, predictive
clustering trees and generative neural networks. Early studies have
shown that choosing heuristically a network layout and then training it
may not lead to the expected results. The fundamental parameters of the
network need to be adapted to the problem under study. This opens the
possibility of introducing some of the parameters defining the network
structure among the quantities to optimise in the training process.
Although this moves the whole optimisation process one step further by
introducing a meta-optimisation that can be accomplished in the same
step or in alternate steps to the more traditional network training, it
offers the possibility to go beyond the one-size-fits-all in Machine
Learning. This will remove part of the arbitrariness in choosing a given
network structure, and will provide the opportunity to control the
trade-offs between accuracy of the results and training and computation
time. In ML parlance, this is called an embedded algorithm, where the
information from the feature performance is fed back to the algorithm
building stage together with the meta-optimisation, usually called
\emph{hyper-parameter} \emph{tuning}, leading to problem-specific
optimised classifiers. If successful, this could provide an opportunity
for shared not only techniques but also work among the experiments, as
the meta-optimisation could account for some of the differences between
detectors and targets.

While the focus in this section and in the literature so far has been on
using machine learning to do fast simulation (e.g. learn Geant4), it may
be possible to train a generator directly on data to replace or augment
Full Simulation. This would require high fidelity testbeam data for a
variety of particles and energies. While this is beyond the scope of
this CWP, we suggest that this be pursued in parallel to improving the
speed and accuracy of fast simulators.

\hypertarget{sharing-of-fast-simulation-and-interfaces}{%
\subsubsection{10.4 Sharing of Fast Simulation and
Interfaces}\label{sharing-of-fast-simulation-and-interfaces}}

At the moment, simulation toolkits like Geant4 provide interfaces for
the experiments to ``hook in'' their fast simulation code. These have
proven sufficient for very simple fast simulations. Because many aspects
of Fast Simulation depend on experiment specific issues (not just in
terms of detector technologies, but in terms of actual C++ types used,
as in many cases intermediate classes are bypassed to save time), it can
be difficult to find commonalities. However, the techniques, approaches,
and methods, as well as some of the issues encountered during
implementation or understanding of the tuning of Fast Simulation, can be
common among the experiments. A forum in which such issues could be
discussed would be very useful, particularly as machine learning
techniques become more widely adopted.

In GeantV, Fast Simulation is being integrated as a user process,
offering the same flexibility as in Geant4 to combine fast with Full
Simulation. The Fast Simulation processes will get as input the full
track state information, allowing to limit the step length and therefore
select particles for performing the user-defined Fast Simulation based
on arbitrary selection criteria, such as particle type, energy,
detector, etc. As in the case of Geant4, it will be possible to have
physics lists per region, which in case of Fast Simulation gives extra
flexibility for implementing some use cases. The main difference however
will be the possibility to put Fast Simulation in the multi-particle
vectorised flow. As any physics model, Fast Simulation will provide not
only a scalar interface (allowing the user code working with Geant4 to
work also with GeantV), but also multi-particle interfaces. The Fast
Simulation user models that are vectorised will be able to benefit from
data and code locality, as well as instruction level parallelism as any
other component of GeantV, while the scalar models will be still usable.

\hypertarget{pseudorandom-number-generators}{%
\section{\texorpdfstring{\textbf{11. Pseudorandom number
generators}}{11. Pseudorandom number generators}}\label{pseudorandom-number-generators}}

The use of sequences of pseudo-random numbers is critical for all Monte
Carlo simulations. In simulating particle and radiation transport the
use of pseudorandom number generators is a vital component. The large
number of random numbers required in the simulation of LHC and other
High Energy and Nuclear Physics (HENP) detector events, as well as
simulation in other fields, and the need to obtain statistically correct
estimate while simulating with a large or very large degree of
parallelism present challenges in the selection and use of pseudorandom
number generators (PRNGs).

Reproducibility is a key requirement of HENP simulation i.e. when
repeating the simulation of an event with the same input particles, with
the same state of a PRNG (in the same geometry and with the same physics
models), the simulation must provide exactly the same results, i.e. the
same values for all observed quantities. Moreover it is required because
it allow problems to be located and debugged. It enables the
verification of new experiment software versions when based on the same
simulation release (e.g. Geant4 release 10.3 patch 1) and even of new
hardware, given the same binaries.

The results must be statistically equivalent for all PRNG, and obtaining
different results with any PRNG is a clear signal that at least one of
the PRNGs is not adequate for the simulation. Correct results require
that the sequences do not contain correlations.

Sequential runs of detector simulation in one process, e.g. using
Geant4, only require a single initialisation of a PRNG at the start of a
simulation run. The state of the PRNG at the start of an event can be
recorded, and stored with the event, to enable its reproduction.

In event-parallel multi-threaded simulation, such as multi-threaded
Geant4, reproducibility means that each event must be provided with a
deterministic seed (or state), typically created at the start of the
simulation. This method is a simple extension of event seeding used in
experiment grid productions.

Nevertheless only some PRNGs offer guarantees for the correlation of
such sequences. Thus tests to ensure that all simulation results are
unaffected by changing the type of PRNG engine is strongly advised. This
use is similar to the research on parallel PRNG for use in other fields
{[}61{]}. A recent survey {[}62{]} covers developments for parallel
PRNGs, including ones for GPUs.

When each track of an event can be simulated in different order, as in
the case of the GeantV simulation design, the state of the PRNG must be
attached to a track. This requires memory proportional to the size of
the state of the PRNG and to the number of tracks in flight. This
excludes the use of PRNGs with large state. A number of potentially
suitable PRNG classes have been identified, including counter-based
PRNGs, such as Random123 {[}63{]}, Combined Multiple Recursive Random
number Generators (CMRGs) {[}64{]} and the newest extension of the
MIXMAX PRNGs {[}65{]} with small matrix size.

These new families of PRNGs are largely untested in radiation transport
simulation, yet are at the forefront of PRNG 'technology' for use in
parallel applications. Thus a key aim should be to obtain
implementations compatible with current interfaces (CLHEP and ROOT) so
that they can be tested in large scale simulations, before their use
become necessary in the transition to large-scale or fine-grained
parallelism.

One of the methods for seeding parallel threads is to create
sub-sequences from a single PRNG with a very long period. The mechanism
for choosing a subsequence, whether it depends only on the sequencing of
tracks or on additional transport parameters, and its computational cost
are areas of active investigation. The resulting algorithms to generate
a new state for each daughter track, must be validated in trial
simulations.

In addition, the new C++11 \textless{}random\textgreater{} interface and
its implementations must be considered. These have not yet seen
widespread use in simulation. Trials within the Root project raised an
issue with the performance of distributions that output floating point
numbers (single and double precision) for several popular generators. If
these findings are confirmed, we must seek to create faster, lower
precision, distributions for floating point numbers for particular
generators. In current practice a single function call is used to obtain
many variates (output numbers) in performance critical code.
Investigations will be made as to whether it is possible to create
similar functionality by designing and implementing a new type of
distribution object, or obtain similar performance benefits in a
different way.

The future programme of work in the field of PRNGs includes the
following:

\begin{itemize}
\item
  Ensure that performant implementations of additional state-of-the-art
  PRNGs are made available, in particular PRNGs which have been shown to
  have none or minimal correlations between separate sequences or
  subsequences.
\item
  Promote the replacement of obsolete PRNGs currently in use in HEP
  experiments, including HepJamesRandom, RanECU, whose periods are small
  by the standards of today's computing power, and the number of output
  numbers required in typical event simulations.
\item
  Create vectorised implementations of PRNGs for use in sequential,
  event-parallel and track-parallel (fine grained) simulations.
\item
  Develop a single library containing sequential and vectorised
  implementations of the set of state-of-the-art PRNG, to replace the
  existing Root and CLHEP implementations within 3 years. This includes
  counter-based methods, the MIXMAX family of generators, improved
  implementations of RANLUX, implementations of CMRGs and other
  categories of sequential and parallel PRNGs. Promote a transition to
  the use of this library to replace existing implementations in ROOT,
  Geant4 and GeantV.
\item
  Collaborate with the authors of the state of the art PRNG testing
  suites, and in particular TestU01{[}66{]}, encouraging and seeking to
  contribute to the extension to increase the statistics of testing,
  extend testing to 64-bit variates, and expand the testing of
  correlations between sub-sequences.
\item
  Follow and contribute to the evolution of the field of PRNGs for
  parallel and highly parallel applications, collaborating with
  researchers in the development of PRNG, seeking to obtain generators
  which address better our challenging requirements.
\end{itemize}

\begin{itemize}
\item
  Investigate whether poor PRNGs, or PRNGs with few, known deficiencies
  can create erroneous results in the particle-transport simulation of
  simple setups.
\end{itemize}

\hypertarget{section-7}{%
\section{}\label{section-7}}

\hypertarget{section-8}{%
\section{\texorpdfstring{\\
}{ }}\label{section-8}}

\hypertarget{summary-and-outlook}{%
\section{12. Summary and outlook}\label{summary-and-outlook}}

In this note we have attempted to describe the main challenges and
opportunities for Full and Fast Simulation applications of HEP
experiments in view of planned future experimental programs in the next
decade; these include the HL-LHC, neutrino and muon experiments, and
studies towards future colliders such as FCC and CLIC. The combination
of additional physics models and higher precision in the simulation with
the need of larger Monte Carlo samples, which scale with the number of
real events recorded in future experiments, place an additional burden
on the computing resources that will be needed to generate them.

A multi-prong approach is being followed as a strategy for responding to
these challenges. Firstly Geant4 will continue to deliver new or refined
functionalities both in physics coverage and accuracy, whilst
introducing software performance improvements whenever possible. At the
same time, an R\&D programme is investigating new approaches that aim to
benefit from modern computing architectures. Its main feature is
track-level parallelisation, bundling particles with similar properties
from different events to process them in a single thread. This approach
combined with SIMD vectorisation coding techniques and use of data
locality is expected to yield large speed-ups, which are being measured
in a realistic prototype under development. In addition, the work on
Fast Simulation is accelerating with a view to producing a flexible
framework that permits Full and Fast simulation to be combined for
different particles in the same event. The overriding requirement is to
ensure the support of experiments through continuous maintenance,
support and improvement of the Geant4 simulation toolkit with minimal
API changes visible to these experiments at least until production
versions of potentially alternative engines, such as those resulting
from ongoing R\&D work, become available, integrated, and validated by
experiments. The agreed ongoing strategy to meet this goal is to ensure
that new developments resulting from the GeantV R\&D programme are
deployed in a timely way in Geant4.

Improving the speed of simulation codes by an order of magnitude
represents a huge challenge. However the gains that can be made by
speeding up critical elements of the common Geant4 toolkit can be
leveraged for all applications that use it and therefore will have a big
impact on computing resource requirements. It is important to note that
the current effort level available in the community is barely sufficient
to keep up with the maintenance and improvements required by current and
future experiments. It is therefore of critical importance that the
whole community of scientists working in the simulation domain continue
to work together in as efficient way as possible in order to deliver the
required improvements. Very specific expertise is required across all
simulation domains, such as the physics modeling, tracking through
complex geometries and magnetic fields, and building realistic
applications that accurately simulate highly complex detectors.
Continuous support is needed to recruit, train, and retain the
person-power with the unique set of skills needed to guarantee the
development, maintenance, and support of simulation codes over the
short, medium, and long timeframes foreseen in the HEP experimentals
programmes.

\hypertarget{references}{%
\section{References}\label{references}}

{[}1{]} J.Allison et al., NIM A 835 (2016), 186-225

{[}2{]} J.Apostolakis et al., J. Phys.: Conf. Ser., 396 (2012) 022014

{[}3{]} P. Adamson et al. {[}NOvA Collaboration{]}, ``First measurement
of electron neutrino appearance in NOvA'', Phys. Rev. Lett. 116 (2016)
no.15, 151806.
\href{https://arxiv.org/abs/1601.05022}{{arXiv:1601.05022}}.

{[}4{]} The MicroBooNE Collaboration, ``The MicroBooNE Technical Design
Report'',
\href{http://www-microboone.fnal.gov/publications}{{http://www-microboone.fnal.gov/publications}}.

{[}5{]} L. Aliaga, et al. {[}MINERvA Collaboration{]}, ``Design,
Calibration, and Performance of the MINERvA Detector''
Nucl.Instrum.Meth. A743 (2014) 130-159.
\href{https://arxiv.org/abs/1305.5199}{{arXiv:1305.5199}}

{[}6{]} R. Acciarri et al. {[}DUNE Collaboration{]}, ``Long-Baseline
Neutrino Facility (LBNF) and Deep Underground Neutrino Experiment (DUNE)
: Volume 2: The Physics Program for DUNE at LBNF'',
\href{https://arxiv.org/abs/1512.06148}{{arXiv:1512.06148}}.

{[}7{]} C.Andreopoulos et al., The GENIE Neutrino Monte Carlo Generator,
Nucl.Instrum.Meth.A614 (2010) 87-104.
\href{https://arxiv.org/abs/0905.2517}{{arXiv:09052517}}

{[}8{]} A. Ferrari, P. Sala, A. Fasso, J. Ranft, FLUKA: a multi-particle
transport code, CERN-2005-10 (2005), INFN/TC\_05/11, SLAC-R-773.

{[}9{]} The MARS Code System,
\href{http://mars.fnal.gov}{{http://mars.fnal.gov}}

{[}10{]} M.Gayer et al., New software library of geometrical primitives
for modeling of solids used in Monte Carlo detector simulations, J.
Phys.: Conf. Ser., 396 (2012)
\href{http://iopscience.iop.org/article/10.1088/1742-6596/396/5/052035/meta}{{052035}}

{[}11{]} R.Brun et al. NIM A 389 (1997) 81 -- 86

{[}12{]} J.Apostolakis et al., Towards a high performance geometry
library for particle-detector simulations, J. Phys.: Conf. Ser., 608
(2015)
\href{http://iopscience.iop.org/article/10.1088/1742-6596/608/1/012023}{{012023}}

{[}13{]} R.Chytracek et al.,
\href{http://ieeexplore.ieee.org/document/1710291/}{Geometry Description
Markup Language for Physics Simulation and Analysis Applications}, IEEE
Trans. Nucl. Sci., 53 (2006), p.
\href{http://ieeexplore.ieee.org/document/1710291/}{{2892}}

{[}14{]} Apache Xerces C Parser,

\href{http://xerces.apache.org/xerces-c/}{{http://xerces.apache.org/xerces-c/}}{[}15{]}
M.Frank et al, DD4hep: A Detector Description Toolkit for High Energy
Physics Experiments,
\href{http://iopscience.iop.org/journal/1742-6596}{{Journal of Physics:
Conference Series}},
\href{http://iopscience.iop.org/volume/1742-6596/664}{{Volume 664}},
\href{http://iopscience.iop.org/issue/1742-6596/664/7}{{Offline
software}}
(\href{http://iopscience.iop.org/article/10.1088/1742-6596/513/2/022010/meta}{{DD4HEP}})

M.Frank et al, DDG4 A Simulation Framework based on the DD4hep Detector
Description Toolkit,
\href{http://iopscience.iop.org/journal/1742-6596}{{Journal of Physics:
Conference Series}},
\href{http://iopscience.iop.org/volume/1742-6596/664}{{Volume 664}},
\href{http://iopscience.iop.org/issue/1742-6596/664/7}{{Offline
software}},
(\href{http://iopscience.iop.org/article/10.1088/1742-6596/664/7/072017/meta}{{DDG4}})

M.Petric et al, Detector Simulations with DD4hep,
\href{https://cds.cern.ch/record/2157890}{Computing in High Energy and
Nuclear Physics, CHEP 2016}, CERN preprint
\href{https://cds.cern.ch/record/2244362?ln=en}{{CLICdp-Conf-2017-001}}

{[}16{]} M.Benoit, J.Idarraga, The AllPix Simulation Framework, website:
\href{https://twiki.cern.ch/twiki/bin/view/Main/AllPix}{{https://twiki.cern.ch/twiki/bin/view/Main/AllPix}},

D.Hynds, S.Spannagel, K.Wolters, AllPix Squared Framework, website:
\href{http://project-allpix-squared.web.cern.ch/project-allpix-squared/}{{http://project-allpix-squared.web.cern.ch/project-allpix-squared/}}

{[}17{]} A. Schaeliske et al, J.Physics : Conf. Ser., 331 (2011) 032029

{[}18{]} J. Allison et al., J.Physics: Conf. Ser. 396 (2012) 022013

{[}19{]} V.N. Ivanchenko, J. Physics: Conf. Ser. 513 (2014) 0022015

{[}20{]} K.Werner, The hadronic interaction model EPOS,
\href{http://www.sciencedirect.com/science/journal/09205632}{Nuclear
Physics B - Proceedings Supplements},
\href{http://www.sciencedirect.com/science/journal/09205632/175/supp/C}{Volumes
175--176}, January 2008, Pages 81-87

{[}21{]} LArSoft: The Liquid Argon Software Collaboration
\href{http://larsoft.org}{{larsoft.org}}.

{[}22{]} D. Lussi, Study of the response of the novel LAr LEM-TPC
detector exposed to cosmic rays and a charged particle beam. PhD thesis,
ETH Zürich, 2013. Chapter 4.1

{[}23{]} M Szydagis, A Fyhrie, D Thorngren and M Tripathi, "Enhancement
of NEST capabilities for simulating low-energy recoils in liquid xenon"
doi:10.1088/1748-0221/8/10/C10003

{[}24{]} Wire-Cell, A 3D Reconstruction Package for LArTPC.
http://lar.bnl.gov/wire-cell.

{[}25{]}
\href{https://inspirehep.net/record/1500313}{{https://inspirehep.net/record/1500313}};
PoS(ICHEP2016)171

{[}26{]} Baglie, E. et al.,A model for the interaction of high-energy
particles in straight and bent crystals implemented in Geant4,
\href{https://arxiv.org/abs/1403.4984}{{arXiv:1403.4984}}

{[}27{]} Bernal, M.A. et al., Track structure modeling in liquid water:
A review of the Geant4-DNA very low energy extension of the Geant4 Monte
Carlo simulation toolkit, Phys. Med. 31 (2015) 861-874

{[}29{]}
\href{https://g4cpt.fnal.gov/g4p/prplots/cpu_by_version.html}{{Source}}:
Geant4 Testing and Quality Assurance Working Group, 2016.

{[}30{]} Dotti, A. et al "Extending Geant4 Parallelism with external
libraries (MPI, TBB) and its use on HPC resources," 2015 IEEE Nuclear
Science Symposium and Medical Imaging Conference (NSS/MIC), San Diego,
CA, 2015, pp. 1-2. doi: 10.1109/NSSMIC.2015.7581867

{[}31{]} Farrell, S. et al "Multi-threaded Geant4 on the Xeon-Phi with
complex high-energy physics geometry," 2015 IEEE Nuclear Science
Symposium and Medical Imaging Conference (NSS/MIC), San Diego, CA, 2015,
pp. 1-4. doi: 10.1109/NSSMIC.2015.7581868

{[}32{]} M. Hildreth, V. N. Ivanchenko and D. J. Lange, ``Upgrades for
the CMS simulation'', CHEP 2016.

{[}33{]} G. Grindhammer, et al., The Fast Simulation of Electromagnetic
and Hadronic Showers, Nucl. Instrum. Meth. A 290 (1990) 469.

{[}34{]} Torbjorn Sjostrand et al., An Introduction to PYTHIA 8.2,
arXiv:1410.3012 {[}hep-ph{]}doi:10.1016/j.cpc.2015.01.024; T. Sjostrand,
S. Mrenna, P. Skands, PYTHIA 6.4 Physics and Manual, JHEP05 (2006)
026doi:10.1088/1126-6708/2006/05/026.

{[}35{]} C. Green et. al., The art event processing framework,
\href{https://inspirehep.net/record/1211020}{J.Phys.Conf.Ser. 396 (2012)
022020}

{[}36{]} D. Heck, G. Schatz, T. Thouw, J. Knapp, J. N. Capdevielle,
CORSIKA: A Monte Carlo code to simulate extensive air showers, Tech.
Rep. 6019, FZKA (1998)

{[}37{]} Szydagis et al., NEST: A Comprehensive Model for Scintillation
Yield in Liquid Xenon,
\href{http://dx.doi.org/10.1088/1748-0221/6/10/P10002}{{2011 JINST 6
P10002}}; e-Print:
\href{http://arxiv.org/abs/1106.1613}{{arXiv:1106.1613
{[}physics.ins-det{]}}}

{[}38{]} Brandt, D. et al., Semiconductor phonon and charge transport
Monte Carlo simulation using Geant4,
\href{https://arxiv.org/abs/1403.4984}{{arXiv:1403.4984}}

{[}39{]} Blyth, S. et al., Opticks : GPU Optical Photon Simulation for
Particle Physics with NVIDIA OptiX, PSROC Annual Meeting, Tamkang
University, TKU, New Taipei City, Taiwan

{[}40{]} J. Apostolakis et al., ``Adaptive track scheduling to optimise
concurrency and vectorisation in GeantV'', Journal of Physics:
Conference Series 608 (2015) 012003

{[}41{]} J. Apostolakis et al, ``Towards a high performance geometry
library for particle-detector simulations'', Journal of Physics:
Conference Series 608 (2015) 012023

{[}42{]} Henderson N. et al.; A CUDA Monte Carlo simulator for radiation
therapy dosimetry based on Geant4; Joint International Conference on
Supercomputing in Nuclear Applications and Monte Carlo (2013)

{[}43{]} Amadio, G. et al.: GeantV: from CPU to accelerators; Journal of
Physics: Conference Series 762 (2016)
\href{http://iopscience.iop.org/article/10.1088/1742-6596/762/1/012019/meta}{{012019}}

{[}44{]} LeCompte, T. et al.; Geant4 at extreme scales, in preparation,
cfr \href{https://goo.gl/cgA92Y}{{https://goo.gl/cgA92Y}}

{[}45{]} Calafiura P et al. on behalf of the ATLAS Collaboration (2015)
"Fine grained event processing on HPCs with the ATLAS Yoda system", J.
Phys.: Conf. Ser., 664 092025,
\href{http://iopscience.iop.org/article/10.1088/1742-6596/664/9/092025}{{link}}

{[}46{]} S. Farrell, et al., "Multi-threaded ATLAS Simulation on Intel
Knights Landing Processors", CHEP 2016,
\href{http://cds.cern.ch/record/2242857}{{http://cds.cern.ch/record/2242857}}

{[}47{]} A Common Tracking Software: http://acts.web.cern.ch/

{[}48{]} V. Volkl, ``Applications of A-Common-Tracking-Software (ACTS)
in the Future Circular Collider Design Study.'' Submitted to ACAT 2017:
\href{https://indico.cern.ch/event/567550/contributions/2627161/contribution.pdf}{{https://indico.cern.ch/event/567550/contributions/2627161/contribution.pdf}}

{[}49{]} E. Ritsch for the ATLAS Collaboration, ``Concepts and plans
towards fast large-scale Monte Carlo production for the ATLAS
Experiment.'' J. Phys.: Conf. Ser. 523 (2014)
\href{http://iopscience.iop.org/article/10.1088/1742-6596/523/1/012035/meta}{{012035}}.

{[}50{]} The DELPHES 3 Collaboration, ``DELPHES 3: A modular framework
for fast simulation of a generic collider experiment.'' JHEP 02 (2014)
057.

{[}51{]} PGS-4,
\href{http://www.physics.ucdavis.edu:80/~conway/research/software/pgs/pgs4-general.htm}{{http://www.physics.ucdavis.edu:80/\textasciitilde{}conway/research/software/pgs/pgs4-general.htm}}

{[}52{]} B. Knuteson, ``Systematic Analysis of High Energy Collider
Data'', Nucl. Instrum. Meth. A534 (2004) 7--14, {[}hep-ex/0402029{]}

{[}53{]} H. B. Prosper et. al. ``Falcon: towards an ultra fast
non-parametric detector simulator'', contribution to Les Houches 2015:
Physics at TeV Colliders New Physics Working Group Report
{[}arXiv:1605.02684{]}.

{[}54{]} A. Hoecker et al., TMVA - Toolkit for Multivariate Data
Analysis, PoS ACAT 040 (2007), arXiv:physics/0703039

{[}55{]} I. J. Goodfellow et al., Generative Adversarial Networks,
arXiv:1406.2661 (2014).

{[}56{]} L. Oliveira, M. Paganini, B. Nachman, Learning Particle Physics
by Example: Location-Aware Generative Adversarial Networks for Physics
Synthesis, arXiv:1701.05927, 2017

{[}57{]} M. Paganini, L. Oliveira, B. Nachman, CaloGAN: Simulating 3D
High Energy Particle Showers in Multi-Layer Electromagnetic Calorimeters
with Generative Adversarial Networks, arXiv:1705.02355 2017.

{[}58{]} Inter-Experimental LHC Machine Learning Working Group :
\href{https://iml.cern.ch}{{https://iml.cern.ch}}

{[}59{]} F. Pedregosa et al., Scikit-learn: Machine learning in Python,
J.Machine Learning Res. 12 (2011) 2825-2830 {[}arXiv:1201.0490{]}

{[}60{]} M. Abudi et al., ``Tensorflow: A system for large-scale machine
learning'', in OSDI, vol. 16,

pp. 265--283, 2016

{[}61{]} H. Bauke and S. Mertens, ``Random numbers for large-scale
distributed Monte Carlo simulations,'' Phys. Rev. E, vol. 75, no. 6, p.
66701, Jun. 2007.

{[}62{]} P. L'Ecuyer, D. Munger, B. Oreshkin, and R. Simard, ``Random
numbers for parallel computers: Requirements and methods, with emphasis
on GPUs,'' Math. Comput. Simul., vol. 135, pp. 3--17, May 2017.

{[}63{]} J. K. Salmon, M. A. Moraes, R. O. Dror, and D. E. Shaw,
``Parallel random numbers: As easy as 1, 2, 3,'' in Proceedings of 2011
International Conference for High Performance Computing, Networking,
Storage and Analysis on - SC '11, 2011, p. 1.,
\href{http://dl.acm.org/citation.cfm?id=2063405\&CFID=810701954\&CFTOKEN=31644804}{{link}}

{[}64{]} P. L'Ecuyer, ``Good Parameters and Implementations for Combined
Multiple Recursive Random Number Generators,'' Operations Research, vol.
47, no. 1, pp. 159--164, Feb. 1999.

{[}65{]} K. Savvidy and G. Savvidy, ``Spectrum and entropy of C-systems
MIXMAX random number generator,'' Chaos, Solitons \& Fractals, vol. 91,
pp. 33--38, 2016.

{[}66{]} P. L'Ecuyer and R. Simard, ``TestU01,'' ACM Transactions on
Mathematical Software, vol. 33, no. 4, p. 22--es, Aug. 2007.

### 2018-01-22 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
Update to Abstract

@@
 
-\abstract{The High Energy Phyiscs community has developed and needs to maintain
-many tens of millions of lines of code and to integrate effectively the
-work of thousands of developers across large collaborations. 
-Software needs to be built, validated, and deployed across hundreds of
-sites. Software also has a lifetime of many years, frequently beyond that
-of the original developer, it must be developed with sustainability in
-mind. Adequate recognition of software development as a critical task in the HEP
-community needs to be fostered and an appropriate publication and citation
-strategy needs to be developed. As part of the HEP Softare Foundation's
-Community White Paper process a working group on Software Development,
-Deployment and Validation was formed to examine all of these
-issues, identify best practice and to formulare recommendations for the next
-decade. Its report is presented here.}
+\abstract{ A working group on detector simulation was formed as part of the HEP 
+Software Foundation's initiative to prepare a Community White Paper that describes 
+the main software challenges and opportunities to be faced in our field over the next 
+decade. The working group met over a period of several months in order to review the 
+current status of the Full and Fast simulation applications of HEP experiments 
+and the improvements that will need to be made in order to meet the goals of future HEP 
+experimental programmes.  The scope of the topics covered includes the main components 
+of a HEP simulation application, including MC truth handling, geometry modeling, particle 
+propagation in materials and fields, physics modeling of the interactions of particles 
+with matter, the treatment of pileup and other backgrounds, as well as signal processing 
+and digitisation. The work programme described in this document focuses on the need to 
+improve both software and physics performance, in view of the need to improve statistical precision
+and to address the diversification of physics programmes.}
 
@@
 
-\title{HEP Software Foundation Community White Paper Working Group -- Software
-Development, Deployment and Validation}
+\title{HEP Software Foundation Community White Paper Working Group -- Detector Simulation}
 
 \author{HEP Software Foundation:}
-\author[a]{Benjamin Couturier}
-\author[a,1]{Giulio Eulisse}
-\author[b]{Hadrien Grasland}
-\author[a]{Benedikt Hegner}
-\author[b]{Michel Jouvin}
-\author[c]{Meghan Kane}
-\author[d]{Daniel S. Katz}
-\author[e]{Thomas Kuhr}
-\author[f]{David Lange}
-\author[a,1]{Patricia Mendez Lorenzo}
-\author[e]{Martin Ritter}
-\author[a,1]{Graeme Andrew Stewart}
-\author[a]{Andrea Valassi}
+\author[f,1]{Victor D. Elvira}
+\author[a,1]{John Harvey}
 
@@
 \affiliation[b]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
-\affiliation[c]{Soundcloud, Berlin, Germany}
-\affiliation[d]{University of Illinois Urbana-Champaign, Champaign, Illinois,
+\affiliation[c]{University of Illinois Urbana-Champaign, Champaign, Illinois,
 USA}
-\affiliation[e]{Fakultät für Physik, Ludwig-Maximilians-Universität München,
-München, Germany}
+\affiliation[d]{FNAL, Fermilab, Batavia IL 60510- 5011, USA}
 \affiliation[f]{Princeton University, Princeton, USA}
@@
 
+\end{document}
+
 \hypertarget{references}{%

### 2018-01-23 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
incorporating Daniel's changes

@@
 
-\abstract{ A working group on detector simulation was formed as part of the HEP 
+\abstract{ A working group on detector simulation was formed as part of the high-energy physics (HEP) 
 Software Foundation's initiative to prepare a Community White Paper that describes 
-the main software challenges and opportunities to be faced in our field over the next 
+the main software challenges and opportunities to be faced in the HEP field over the next 
 decade. The working group met over a period of several months in order to review the 
@@
 experimental programmes.  The scope of the topics covered includes the main components 
-of a HEP simulation application, including MC truth handling, geometry modeling, particle 
+of a HEP simulation application, such as MC truth handling, geometry modeling, particle 
 propagation in materials and fields, physics modeling of the interactions of particles 
 with matter, the treatment of pileup and other backgrounds, as well as signal processing 
-and digitisation. The work programme described in this document focuses on the need to 
-improve both software and physics performance, in view of the need to improve statistical precision
-and to address the diversification of physics programmes.}
+and digitisation. The resulting work programme described in this document focuses on the need to 
+improve both the software and the physics of detector simulation. The goals are to increase the 
+accuracy of the physics models and expand their applicability to future physics programmes, 
+while achieving large factors in computing performance gains consistent with projections on 
+available computing resources.}
 
@@
 end of the 2030's, the upgraded experiments expect to collect 150 times
-more data than in Run 1. The 50 PB of raw data produced in 2016 will
-grow to approximately 600 PB in 2026 while the CPU needs will increase
-by a factor of about 60. Demand for larger simulation samples to satisfy
+more data than in Run 1. Demand for larger simulation samples to satisfy
 analysis needs will grow accordingly. In addition, simulation tools have
@@
 Muon experiments such as Muon g-2 and Mu2e also face large simulation
-challenges. Because they are searching for extremely rare effects, they
+challenges, being Geant4 the primary simulation toolkit for
+all stages of these experiments. Because they are searching for extremely rare effects, they
 must grapple with very low signal to background ratios, and the modeling
@@
 produce the Full Simulation sample needed to complete a technical design
-report delivered in 2016. Geant4 is the primary simulation toolkit for
-all stages of these experiments. For comparison, the CMS experiment
+report delivered in 2016.
+
+Although in comparison the current computing needs of neutrino and muon experiments
+are small compared to the LHC experiments, i.e. the CMS experiment
 spent 860 thousand core months on simulation in the May 2015-May 2016
-period.
-
-Although the computing needs of the intensity frontier are currently
-small compared to the LHC experiments, they are expected to grow
-substantially in the next decade, and it is imperative that the
+period, they are expected to grow substantially in the next decade.
+It is therefore imperative that the
 intensity frontier experiments make use of modern computing
@@
 Geant4 have put emphasis on further reducing memory consumption to the
-current level of about 10MB needed for each additional thread.
+current level of about 10 MB needed for each additional thread.
 
@@
 decreases the simulation time per event, because the thin volumes of the
-sampling calorimeter demand a tighter range cut, including 30um in the
-forward calorimeter and 300um in the EM calorimeters, to provide an
+sampling calorimeter demand a tighter range cut, including 30~$\mu$m in the
+forward calorimeter and 300~$\mu$m in the EM calorimeters, to provide an
 accurate physics description. This penalty is not present in the CMS
@@
 example Russian Roulette and shower libraries, to speedup simulation in
-a given sub-detector. In the literature, these applications are still
+a given sub-detector. These applications are still
 labeled Full Simulation and Geant4-based. Some experiments have
@@
 be possible to train a generator directly on data to replace or augment
-Full Simulation. This would require high fidelity testbeam data for a
+Full Simulation. This would require high fidelity test beam data for a
 variety of particles and energies. While this is beyond the scope of
@@
 studies towards future colliders such as FCC and CLIC. The combination
-of additional physics models and higher precision in the simulation with
+of additional physics models and higher accuracy in the simulation with
 the need of larger Monte Carlo samples, which scale with the number of
@@
 same time, an R\&D programme is investigating new approaches that aim to
-benefit from modern computing architectures. Its main feature is
+benefit from modern computing architectures. For example, 
+the main feature of the GeantV transport engine is
 track-level parallelisation, bundling particles with similar properties
@@
 different particles in the same event. The overriding requirement is to
-ensure the support of experiments through continuous maintenance,
-support and improvement of the Geant4 simulation toolkit with minimal
+ensure the support of experiments through continuous maintenance 
+and improvement of the Geant4 simulation toolkit with minimal
 API changes visible to these experiments at least until production
@@
 experiments. The agreed ongoing strategy to meet this goal is to ensure
-that new developments resulting from the GeantV R\&D programme are
+that new developments resulting from the GeantV R\&D programme,
+including the novel transport engine, are
 deployed in a timely way in Geant4.

### 2018-01-29 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
Many changes to put correct latex commands

@@
 \usepackage{xspace}
+\usepackage{verbatim}
 
@@
 and digitisation. The resulting work programme described in this document focuses on the need to 
-improve both the software and the physics of detector simulation. The goals are to increase the 
+improve both the software performance and the physics of detector simulation. The goals are to increase the 
 accuracy of the physics models and expand their applicability to future physics programmes, 
@@
  & & HSF-CWP-2017-13 \\
- & & December 21, 2017 \\ % use \date or hardwire e.g. December 15, 2017
+ & & January 24, 2018 \\ % use \date or hardwire e.g. December 15, 2017
  & & \\
 \end{tabular*}
-\vspace{2.0cm}
+\vspace{0.1cm}
 
@@
 
-\author{HEP Software Foundation:}
-\author[f,1]{Victor D. Elvira}
-\author[a,1]{John Harvey}
-
-\affiliation[a]{CERN, Geneva, Switzerland}
-\affiliation[b]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
-\affiliation[c]{University of Illinois Urbana-Champaign, Champaign, Illinois,
-USA}
-\affiliation[d]{FNAL, Fermilab, Batavia IL 60510- 5011, USA}
-\affiliation[f]{Princeton University, Princeton, USA}
+\author[d]{J Apostolakis}
+\author[q]{M Asai}
+\author[n]{S Banerjee}
+\author[m]{R Bianchi}
+\author[d]{F Carminati}
+\author[l]{R Cenci}
+\author[b]{J Chapman}
+\author[d]{G Corti}
+\author[d]{G Cosmo}
+\author[p]{S Easo}
+\author[q]{A Dotti}
+\author[f,1]{V Elvira}
+\author[f]{L Fields}
+\author[f]{K Genser}
+\author[d]{A Gheata}
+\author[d]{M Gheata}
+\author[d]{I Goulas}
+\author[d,1]{J Harvey}
+\author[d]{F Hariri}
+\author[f]{R Hatcher}
+\author[f]{K Herner}
+\author[d]{V Ivantchenko}
+\author[f]{T Junk}
+\author[f]{S-Y Jun}
+\author[f]{M Kirby}
+\author[f]{R Kutschke}
+\author[f]{P Lebrun}
+\author[f]{A Lyon}
+\author[j]{Z Marshall}
+\author[c]{J-F Marchand}
+\author[a]{M Mooney}
+\author[q]{R Mount}
+\author[k]{J Mousseau}
+\author[j]{B Nachman}
+\author[f]{A Norman}
+\author[f]{I Osborne}
+\author[f]{K Pedro}
+\author[d]{W Pokorski}
+\author[a]{X Qian}
+\author[f]{J Raaf}
+\author[l]{M Rama}
+\author[d]{A Ribon}
+\author[d]{S Roiser}
+\author[o]{D Ruterbories}
+\author[f]{S Sekmen}
+\author[h]{B Siddi}
+\author[f]{E Snider}
+\author[g]{S Vallecorsa}
+\author[e]{M Verderi}
+\author[f]{H Wenzel}
+\author[d]{S Wenzel}
+\author[d]{B Viren}
+\author[f]{T Yang}
+\author[f]{L Welty-Rieger}
+
+\vspace{0.5cm}
+
+\affiliation[a]{Brookhaven National Laboratory, Upton NY, USA}
+\affiliation[b]{University of Cambridge,Cambridge, UK}
+\affiliation[c]{Carleton University, Ottawa, ON, Canada}
+\affiliation[d]{CERN, Geneva, Switzerland}
+\affiliation[e]{IN2P3-LLR, Ecole Polytechnique, Palaiseau, France}
+\affiliation[f]{Fermilab National Accelerator Laboratory, Batavia IL , USA}
+\affiliation[g]{Gangneung-Wonju National University, Gangneung, Gangwon-do, South Korea}
+\affiliation[h]{Università di Ferrara and INFN Ferrara, Ferrara, Italy}
+\affiliation[j]{Lawrence Berkeley National Laboratory, Berkeley CA, USA}
+\affiliation[k]{University of Michigan, Ann Arbor MI, USA}
+\affiliation[l]{Università di Pisa and INFN Pisa, Pisa, Italy}
+\affiliation[m]{University of Pittsburgh, Pittsburgh PA, USA}
+\affiliation[n]{Princeton University, Princeton, NJ 08544, USA}
+\affiliation[o]{University of Rochester, Rochester NY, USA}
+\affiliation[p]{Rutherford Appleton Laboratory, Chilton, UK}
+\affiliation[q]{SLAC National Accelerator Laboratory, Menlo Park, CA 94025, USA}
+\affiliation[r]{University of Illinois Urbana-Champaign, Champaign IL, USA}
+
 \affiliation[1]{Paper Editor}
@@
 \hypertarget{introduction}{%
-\section{1 Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}}
 
@@
 
-\hypertarget{section}{%
-\section{}\label{section}}
-
-\textbf{Draft executive summary document}:
-\href{https://goo.gl/FHaAtm}{{https://goo.gl/FHaAtm}}
+%\textbf{Draft executive summary document}:
+%\href{https://goo.gl/FHaAtm}{{https://goo.gl/FHaAtm}}
 
 \hypertarget{challenges}{%
-\section{2 Challenges}\label{challenges}}
+\section{Challenges}\label{challenges}}
 
@@
 \hypertarget{geometry}{%
-\section{3 Geometry}\label{geometry}}
+\section{Geometry}\label{geometry}}
 
@@
 
-\includegraphics[width=6.26772in,height=4.76389in]{media/image1.png}
-
-\textbf{Figure 1} - An interactive Geant4 example featuring parallel
-geometries and scoring with VecGeom primitives and parameterisations
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image1.png}
+\caption{An interactive Geant4 example featuring parallel
+geometries and scoring with VecGeom primitives and parameterisations}
+\label{fig:geom}
+\end{figure}
 
@@
 
-\hypertarget{section-1}{%
-\section{\texorpdfstring{\\
-}{ }}\label{section-1}}
-
 \hypertarget{digitisation}{%
-\section{4 Digitisation}\label{digitisation}}
+\section{Digitisation}\label{digitisation}}
 
@@
 \hypertarget{modeling-of-pileup}{%
-\section{5 Modeling of Pileup}\label{modeling-of-pileup}}
+\section{Modeling of Pileup}\label{modeling-of-pileup}}
 
@@
 \hypertarget{handling-of-monte-carlo-truth-information}{%
-\section{6 Handling of Monte Carlo Truth
+\section{Handling of Monte Carlo Truth
 Information}\label{handling-of-monte-carlo-truth-information}}
@@
 \hypertarget{physics-modeling}{%
-\section{7 Physics Modeling}\label{physics-modeling}}
+\section{Physics Modeling}\label{physics-modeling}}
 
@@
 \hypertarget{electromagnetic-physics}{%
-\subsubsection{7.1 Electromagnetic
-Physics}\label{electromagnetic-physics}}
+\subsection{Electromagnetic Physics}\label{electromagnetic-physics}}
 
@@
 better describe the simulation of electromagnetic shower shapes, in
-particular to model the H -\textgreater{} γγ signal accurately at the
+particular to model the H $\rightarrow \gamma \gamma $ 
+signal accurately at the
 LHC. This effort is being continued with emphasis on reviewing the
@@
 same figure also shows the ratio of Monte Carlo to data as a function of
-beam momentum. It is clear from the figure that the predictions from the
+beam momentum.
+
+\begin{figure}[bthp]
+\centering
+\includegraphics[width=0.94\textwidth]{image2.png}
+\caption{The top plots correspond to the mean energy response for
+p\textsuperscript{-} (left) and p\textsuperscript{+} (right) as a
+function of the beam momentum. The bottom plots are ratios of Monte
+Carlo predictions to the data for pions as a function of beam momentum.
+The black points refer to the data, whilst the red and blue points are
+Monte Carlo predictions from physics lists QGSP\_FTFP\_BERT\_EML of
+Geant4 10.0.p02 and FTFP\_BERT\_EMM of Geant4 10.2.p02 respectively.}
+\label{fig:enresp}
+\end{figure}
+
+It is clear from the figure that the predictions from the
 physics list FTFP\_BERT\_EMM are closer to the data, in particular at
@@
 
+\begin{figure}[bthp]
+\centering
+\includegraphics[width=0.94\textwidth]{image3.png}
+\caption{Crystal-based EM shower width from the CMS experiment in the 
+eta direction in both the barrel and the endcap. Simulation and data
+are in good agreement in the barrel (figure on the left) but not in
+the endcap region (figure on the right) where there is more material.}
+\label{fig:emshwid}
+\end{figure}
+
+\begin{figure}[bthp]
+\centering
+\includegraphics[width=0.94\textwidth]{image4.png}
+\caption{The upper plot plot shows a measurement of R\textsubscript{$\eta$}
+for unconverted photons in the ATLAS EM calorimeter, where
+R\textsubscript{$\eta$} is the ratio of shower energy deposited in 3x7
+cells to 7x7 cells in $\eta$-$\varphi$. The lower plot shows the lateral shower
+width, W\textsubscript{$\eta1$}, for electrons in the first layer of the EM
+calorimeter. Both plots show a clear discrepancy between real data and
+the results of MC simulation.}
+\label{fig:R}
+\end{figure}
+
 In the ATLAS experiment one area of disagreement concerns the
 measurement of lateral shower shapes in the EM calorimeter. Figure 4
-shows a measurement of R\textsubscript{η} for unconverted photons in the
-ATLAS EM calorimeter, where R\textsubscript{η} is the ratio of shower
+shows a measurement of R\textsubscript{$\eta$}  for unconverted photons in the
+ATLAS EM calorimeter, where R\textsubscript{$\eta$} 
+is the ratio of shower
 energy deposited in 3x7 cells to 7x7 cells in the forward hadron
@@
 
-\includegraphics[width=6.70522in,height=4.94444in]{media/image2.png}\emph{Figure
-2. The top plots correspond to the mean energy response for
-p\textsuperscript{-} (left) and p\textsuperscript{+} (right) as a
-function of the beam momentum. The bottom plots are ratios of Monte
-Carlo predictions to the data for pions as a function of beam momentum.
-The black points refer to the data, whilst the red and blue points are
-Monte Carlo predictions from physics lists QGSP\_FTFP\_BERT\_EML of
-Geant4 10.0.p02 and FTFP\_BERT\_EMM of Geant4 10.2.p02 respectively.}
-
 The HL-LHC era will require higher precision simulations that can only
@@
 
-\includegraphics[width=3.29292in,height=3.03646in]{media/image3.png}\includegraphics[width=3.29288in,height=3.07813in]{media/image4.png}
-
-\emph{Figure 3. Crystal-based EM shower width from the CMS experiment in
-the eta direction in both the barrel and the endcap. Simulation and data
-are in good agreement in the barrel (figure on the left) but not in the
-endcap region (figure on the right) where there is more material.}
-
 \begin{itemize}
@@
   shows a significant difference in the differential cross-section. The
-  old implementation (green plot in Figure 7) was over-suppressing gamma
+  old implementation (green plot in Figure 7) was over-suppressing $ \gamma$
   conversion i.e. the LPM effect was overestimated. There are currently
@@
 
-\includegraphics[width=3.45982in,height=2.73438in]{media/image5.png}
-
-\includegraphics[width=6.66281in,height=2.57813in]{media/image6.png}\emph{Figure
-4. The upper plot plot shows a measurement of} R\textsubscript{η}
-\emph{for unconverted photons in the ATLAS EM calorimeter, where}
-R\textsubscript{η} \emph{is the ratio of shower energy deposited in 3x7
-cells to 7x7 cells in η-φ. The lower plot shows the lateral shower
-width, W\textsubscript{η1}, for electrons in the first layer of the EM
-calorimeter. Both plots show a clear discrepancy between real data and
-the results of MC simulation.}
-
-\includegraphics[width=5.84531in,height=4.08869in]{media/image7.png}
-
-\emph{Figure 5. The new Goudsmit-Saunderson} \emph{multiple scattering
+\begin{figure}[bthp]
+\centering
+\includegraphics[width=0.94\textwidth]{image5.png}
+\caption{The new Goudsmit-Saunderson multiple scattering
 model implementation (GS - Geant4 10.2) shows no artefacts and excellent
-agreement with experimental data. }
-
-\includegraphics[width=6.03802in,height=4.2226in]{media/image8.png}
-
-\emph{Figure 6. The improved bremsstrahlung model included in Geant4
+agreement with experimental data}
+\label{fig:gs}
+\end{figure}
+
+\begin{figure}[bthp]
+\centering
+\includegraphics[width=0.94\textwidth]{image6.png}
+\caption{The improved bremsstrahlung model included in Geant4
 10.3.beta and in GeantV shows better agreement with experimental data
-with respect to the existing implementation.}
-
-\includegraphics[width=5.88021in,height=3.99503in]{media/image9.png}
-
-\emph{Figure 7. The improved pair production model included in both
+with respect to the existing implementation}
+\label{fig:brem}
+\end{figure}
+
+\begin{figure}[bthp]
+\centering
+\includegraphics[width=0.94\textwidth]{image7.png}
+\caption{The improved pair production model included in both
 Geant4 10.4.beta and in GeantV is compared with the existing Geant4
-implementation with and without LPM. }
-
-\begin{quote}
-\includegraphics[width=4.31527in,height=3.89063in]{media/image10.png}
-\end{quote}
-
-\emph{Figure 8. New models were developed in Geant4 taking into account
+implementation with and without LPM}
+\label{fig:ppm}
+\end{figure}
+
+\begin{figure}[bthp]
+\centering
+\includegraphics[width=0.94\textwidth]{image8.png}
+\caption{New models were developed in Geant4 taking into account
 finite nuclear size effect for elastic scattering of electrons at high
-energies. }
-
-\begin{quote}
+energies}
+\label{fig:newm}
+\end{figure}
+
 Theoretical review of all electromagnetic models, including those of
@@
 energies, such as triplet production taking into account nuclear recoil
-effects, size effects as well as γ conversion to muon and hadron pairs.
+effects, size effects as well as $\gamma$ conversion to muon and hadron pairs.
 Energy fluctuation models should be developed based on theory, not just
@@
 generation detectors.
-\end{quote}
 
 \hypertarget{hadronic-physics}{%
-\subsubsection{7.2 Hadronic Physics}\label{hadronic-physics}}
+\subsection{Hadronic Physics}\label{hadronic-physics}}
 
@@
 after the high energy FTF interaction has been completed. The
-FTFP-Bertini combination forms the backbone of many physics lists. It
+FTFP--Bertini combination forms the backbone of many physics lists. It
 successfully extends the use of the string model at lower energies, down
@@
 \hypertarget{physics-modelling-for-the-intensity-frontier}{%
-\subsubsection{7.3 Physics Modelling for the Intensity
+\subsection{Physics Modelling for the Intensity
 Frontier}\label{physics-modelling-for-the-intensity-frontier}}
@@
 \hypertarget{non-hep-use-cases}{%
-\subsubsection{7.4 Non-HEP use-cases}\label{non-hep-use-cases}}
+\subsection{Non-HEP use-cases}\label{non-hep-use-cases}}
 
@@
 \hypertarget{radiation-and-non-collision-background-modeling}{%
-\subsubsection{7.5 Radiation and Non-Collision Background
+\subsection{Radiation and Non-Collision Background
 Modeling}\label{radiation-and-non-collision-background-modeling}}
@@
 
-\hypertarget{section-2}{%
-\section{}\label{section-2}}
-
-\hypertarget{section-3}{%
-\section{\texorpdfstring{\\
-}{ }}\label{section-3}}
-
 \hypertarget{software-performance}{%
-\section{8 Software Performance}\label{software-performance}}
+\section{Software Performance}\label{software-performance}}
 
@@
 multithreading capabilities in Geant4 brought significant memory
-savings, as illustrated in Figure 10, which shows the CPU time (top) and
+savings, as illustrated in Figure 9, which shows the CPU time (top) and
 memory (bottom) consumption of a simplified CMS standalone simulation
@@
 
-Figure 11 shows the percentage change in CPU time performance taking
+Figure 10 shows the percentage change in CPU time performance taking
 Geant4 version 10.0 as a reference, starting with Geant4 version 9.4.p02
@@
 performed for 50 GeV e\textsuperscript{-}, p\textsuperscript{-}, and
-protons, as well as for H → ZZ events. On average, the time performance
+protons, as well as for H $\rightarrow$ ZZ events. On average, the time performance
 improvement through the life of the LHC experiments (2010-2015) is of
@@
 Evolving compilers also offer opportunities for speed gains in
-simulation applications. Figure 12 illustrates the impact of the use of
+simulation applications. Figure 10 illustrates the impact of the use of
 shared and static libraries in a simplified CMS application (HepExpMT
@@
 
-\includegraphics[width=6.5625in,height=3in]{media/image11.png}\includegraphics[width=6.48958in,height=2.9375in]{media/image12.png}\emph{Figure
-10: CPU time (top) and memory (bottom) consumption of a CMS standalone
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image9.png}
+\caption{CPU time (top) and memory (bottom) consumption of a CMS standalone
 simulation application, based on Geant4 version 10.3.p01, run for 50 GeV
@@
 multithreading the memory requirements are reduced by a factor 10.}
-
-\includegraphics[width=3.28646in,height=1.95745in]{media/image13.png}\includegraphics[width=3.26538in,height=1.95313in]{media/image14.png}\includegraphics[width=3.30325in,height=1.96354in]{media/image15.png}\includegraphics[width=3.2959in,height=1.95313in]{media/image16.png}
-
-\emph{Figure 11: Percentage change in CPU time performance taking Geant4
+\label{fig:perfcms}
+\end{figure}
+
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image10.png}
+\caption{Percentage change in CPU time performance taking Geant4
 version 10.0 as a reference, starting with Geant4 version 9.4.p02 (2010)
@@
 application and a simple calorimeter configuration.}
-
-\includegraphics[width=6.70522in,height=6.73611in]{media/image17.png}
-
-\emph{Figure 12: Impact of the use of shared and static libraries in a
+\label{fig:perfchange}
+\end{figure}
+
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image11.png}
+\caption{Impact of the use of shared and static libraries in a
 simplified HEP application. Results obtained with a simplified version
@@
 runtime options of Geant4 can be used to obtain another +10\% speedup. }
+\label{fig:perflib}
+\end{figure}
 
@@
 \hypertarget{cpu-time-and-memory-performance-measurements}{%
-\subsubsection{8.1 CPU Time and Memory Performance
+\subsection{CPU Time and Memory Performance
 Measurements}\label{cpu-time-and-memory-performance-measurements}}
@@
 The lesson learned is that profiling is not a very useful exercise
-unless it is related to physics observables. The table in Figure 13
+unless it is related to physics observables. The table in Figure 12
 shows the number of Geant4 steps in each ATLAS sub-detector and for each
@@
 
-\includegraphics[width=6.21354in,height=3.875in]{media/image18.png}
-
-\emph{Figure 13: Number of Geant4 steps in each ATLAS sub-detector and
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image12.png}
+\caption{Number of Geant4 steps in each ATLAS sub-detector and
 for each particle type. The sub-detector types are: Px (pixel tracker),
@@
 otherwise accounted for).}
-
-Figure 14 shows a chart of CPU time for the G4 module of the CMS Full
+\label{fig:perfsteps}
+\end{figure}
+
+Figure 13 shows a chart of CPU time for the G4 module of the CMS Full
 Simulation application {[}32{]}. About 60\% of the time is spent on
@@
 detector geometry and magnetic field, while about 15\% is spent on EM
-physics, and about 10\% followed by hadronic physics. Figure 15 shows
+physics, and about 10\% followed by hadronic physics. Figure 14 shows
 the CMS throughput and memory per event utilisation for QCD and ttbar
@@
 
-\includegraphics[width=4.37995in,height=3.08854in]{media/image19.png}
-
-\emph{Figure 14: Chart of CPU time spent in different tasks of the CMS
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image13.png}
+\caption{Chart of CPU time spent in different tasks of the CMS
 Full Simulation application for Run-2 simulation}
+\label{fig:perfcpucms}
+\end{figure}
 
@@
 
-\includegraphics[width=6.70522in,height=8.18056in]{media/image20.png}
-
-\emph{Figure 15: Throughput and memory per event for ttbar simulated
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image14.png}
+\caption{Throughput and memory per event for ttbar simulated
 events in CMS as a function of the event number. The version of Geant4
@@
 initialisation time.}
+\label{fig:perfthroughput}
+\end{figure}
 
@@
   ttbar (300 events), pseudo rapidity cut
-  \textbar{}η\textbar{}\textless{}5.5
-\item
-  Particle gun: 50 GeV e's, muons, pions with a flat η,φ distribution in
-  η={[}-0.8,0.8{]}, η={[}2,2.7{]}, and φ={[}0,2π{]}
+  \textbar{}$\eta$\textbar{}\textless{}5.5
+\item
+  Particle gun: 50 GeV e's, muons, pions with a flat $\eta,\varphi$ distribution in
+  $\eta={[}-0.8,0.8{]}, \eta={[}2,2.7{]}, and \varphi={[}0,2\pi{]}$
 \item
@@
 several potential HL-LHC configurations.
+
+\begin{table}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image15.png}
+\caption{Time and memory performance of the Geant4 module of the
+CMS Full Simulation application for minimum bias and ttbar events in
+current and future detector configurations. The Geant4 version used in
+the test is 10.2p02. With respect to the 2015 and 2016 configurations,
+which are very similar, 2017 adds an upgraded pixel detector and
+modifies the forward calorimeter layer configuration, 2023D4 and 2023D17
+are different versions of the HL-LHC CMS detector with an upgraded
+tracker and a High Granularity end cap calorimeter (HGCAL), the latter
+sub-detector involving a significant increase in the number of Geant4
+volumes. }
+\label{label:table1}
+\end{table}
 
@@
 utilizing static libraries and by re-optimizing timing or particle
-production cuts. In the CMS simulation Geant4 time cuts and tracking
+production cuts.
+
+\begin{table}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image16.png}
+\caption{Relative time performance of the Geant4 module of the CMS Full
+Simulation application for minimum bias and ttbar production.
+Measurements of the individual effect of several options to reduce CPU
+time consumption are listed. The Geant4 version used in the test is
+10.2.p02. Values of range cuts and tracking cuts are specified in the
+text.}
+\label{table:table2}
+\end{table}
+
+\begin{table}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image17.png}
+\caption{Relative time performance of the Geant4 module of the CMS Full
+Simulation application for minimum bias and ttbar production.
+Measurements of the cumulative effect of several options to reduce CPU
+time consumption are listed. The Geant4 version used in the test is
+10.2p02. Values of range cuts and tracking cuts are specified in the text.}
+\label{table:table3}
+\end{table}
+
+In the CMS simulation Geant4 time cuts and tracking
 cuts are optimised per detector region. The cut in range value for the
@@
 Still, the effect of range cuts on the two applications is similar.
+
+\begin{table}[bhp]
+\centering
+\begin{tabular}{| p{9cm} | c | c |}
+\hline
+Configuration & MinBias & ttbar \\ \hline
+Nominal production configuration: shower libraries 
+in the forward calorimeter, nominal range cuts, 
+NystromRK4 stepper, FTFP\_BERT\_ATL
+physics list, 250ns neutron time cut, simulation 
+of primary particles with pseudo-rapidity below 6.0
+ & 1.0 & 1.0 \\ \hline
+No shower libraries & 1.5 & 1.3 \\ \hline
+ClassicalRK4 stepper instead of NystromRK4 & 1.09 & 1.07 \\ \hline
+No neutron time cut & 1.02 & 1.01 \\ \hline
+FTFP\_BERT instead of FTFP\_BERT\_ATL physics list & No change & No
+change \\ \hline
+No simulation of primaries with pseudo-rapidity above 5.5 & 0.94 & 0.95
+\\ \hline
+All range cuts set to 1mm & 0.92 & 0.90 \\ \hline
+\end{tabular}
+\caption{Performance of various configurations of the ATLAS
+simulation for minimum bias and ttbar production events. The Geant4
+version used for this test was G4 10.2p03. No significant performance
+improvements were introduced in patch 03 with respect to patch 02.}
+\label{table:table4}
+\end{table}
+
+\begin{table}[bthp]
+\centering
+\begin{tabular}{| p{9cm} | c | c |}
+\hline
+Configuration & MinBias & ttbar \\ \hline
+Nominal production configuration: & 1.0 & 1.0 \\ \hline
+No shower libraries & 1.5 & 1.3 \\ \hline
+No shower libraries + use ClassicalRK4 stepper & 1.6 & 1.4 \\ \hline
+No shower libraries + use ClassicalRK4 stepper + No neutron time cut &
+1.6 & 1.4 \\ \hline
+No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
+use FTFP\_BERT physics list & 1.6 & 1.4 \\ \hline
+No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
+use FTFP\_BERT physics list + All range cuts set to 1mm & 1.5 & 1.3
+\\ \hline
+No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
+use FTFP\_BERT physics list + All range cuts set to 1mm + No simulation
+of primaries with pseudo-rapidity above 5.5 & 1.4 & 1.2 \\ \hline
+\end{tabular}
+\caption{Performance of various configurations of the ATLAS
+simulation for minimum bias and ttbar production events. Measurements of
+the cumulative effect of several options to reduce CPU time consumption
+are listed. The Geant4 version used for this test was G4 10.2p03. No
+significant performance improvements were introduced in patch 03 with
+respect to patch 02.}
+\label{table:table5}
+\end{table}
 
@@
 
-\includegraphics[width=6.06094in,height=3.03532in]{media/image21.jpg}
-
-\emph{Table 1: Time and memory performance of the Geant4 module of the
-CMS Full Simulation application for minimum bias and ttbar events in
-current and future detector configurations. The Geant4 version used in
-the test is 10.2p02. With respect to the 2015 and 2016 configurations,
-which are very similar, 2017 adds an upgraded pixel detector and
-modifies the forward calorimeter layer configuration, 2023D4 and 2023D17
-are different versions of the HL-LHC CMS detector with an upgraded
-tracker and a High Granularity end cap calorimeter (HGCAL), the latter
-sub-detector involving a significant increase in the number of Geant4
-volumes. }
-
-\includegraphics[width=7.59701in,height=3.65104in]{media/image22.jpg}\emph{Table
-2: Relative time performance of the Geant4 module of the CMS Full
-Simulation application for minimum bias and ttbar production.
-Measurements of the individual effect of several options to reduce CPU
-time consumption are listed. The Geant4 version used in the test is
-10.2.p02. Values of range cuts and tracking cuts are specified in the
-text.}\includegraphics[width=6.81771in,height=3.90017in]{media/image23.jpg}\emph{Table
-3: Relative time performance of the Geant4 module of the CMS Full
-Simulation application for minimum bias and ttbar production.
-Measurements of the cumulative effect of several options to reduce CPU
-time consumption are listed. The Geant4 version used in the test is
-10.2p02. Values of range cuts and tracking cuts are specified in the
-text.}
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule
-\endhead
-Configuration & MinBias & ttbar\tabularnewline
-Nominal production configuration: shower libraries in the forward
-calorimeter, nominal range cuts, NystromRK4 stepper, FTFP\_BERT\_ATL
-physics list, 250ns neutron time cut, simulation of primary particles
-with pseudo-rapidity below 6.0 & 1.0 & 1.0\tabularnewline
-No shower libraries & 1.5 & 1.3\tabularnewline
-ClassicalRK4 stepper instead of NystromRK4 & 1.09 & 1.07\tabularnewline
-No neutron time cut & 1.02 & 1.01\tabularnewline
-FTFP\_BERT instead of FTFP\_BERT\_ATL physics list & No change & No
-change\tabularnewline
-No simulation of primaries with pseudo-rapidity above 5.5 & 0.94 &
-0.95\tabularnewline
-All range cuts set to 1mm & 0.92 & 0.90\tabularnewline
-\bottomrule
-\end{longtable}
-
-\emph{Table 4: Performance of various configurations of the ATLAS
-simulation for minimum bias and ttbar production events. The Geant4
-version used for this test was G4 10.2p03. No significant performance
-improvements were introduced in patch 03 with respect to patch 02.}
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule
-\endhead
-Configuration & MinBias & ttbar\tabularnewline
-Nominal production configuration: & 1.0 & 1.0\tabularnewline
-No shower libraries & 1.5 & 1.3\tabularnewline
-No shower libraries + use ClassicalRK4 stepper & 1.6 &
-1.4\tabularnewline
-No shower libraries + use ClassicalRK4 stepper + No neutron time cut &
-1.6 & 1.4\tabularnewline
-No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
-use FTFP\_BERT physics list & 1.6 & 1.4\tabularnewline
-No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
-use FTFP\_BERT physics list + All range cuts set to 1mm & 1.5 &
-1.3\tabularnewline
-No shower libraries + use ClassicalRK4 stepper + No neutron time cut +
-use FTFP\_BERT physics list + All range cuts set to 1mm + No simulation
-of primaries with pseudo-rapidity above 5.5 & 1.4 & 1.2\tabularnewline
-\bottomrule
-\end{longtable}
-
-\emph{Table 5: Performance of various configurations of the ATLAS
-simulation for minimum bias and ttbar production events. Measurements of
-the cumulative effect of several options to reduce CPU time consumption
-are listed. The Geant4 version used for this test was G4 10.2p03. No
-significant performance improvements were introduced in patch 03 with
-respect to patch 02.}
-
 \hypertarget{lhcb-and-alice}{%
-\subsubsection{8.3 LHCb and Alice}\label{lhcb-and-alice}}
+\subsection{LHCb and Alice}\label{lhcb-and-alice}}
 
@@
 integrated simulation framework which allows to choose and combine
-different fast simulation options. Table 4 shows the percentage of the
+different fast simulation options. Table 6 shows the percentage of the
 CPU time spent on each sub-detector systems for different particle types
@@
 
-\includegraphics[width=6.82813in,height=2.20063in]{media/image24.jpg}
-
-\emph{Table 4: Percentage of the CPU time spent on each sub-detector
-systems for different particle types for a given version of the
+\begin{table}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image18.png}
+\caption{Percentage of the CPU time spent on each sub-detector
+system in LHCb for different particle types for a given version of the
 software.}
+\label{table:table6}
+\end{table}
 
 ALICE also consumes two orders of magnitude less CPU resources than
-ATLAS. Figures 16 and 17 show charts of the percentage of time spent in
+ATLAS. Figures 15 and 16 show charts of the percentage of time spent in
 different elements of the detector simulation chain, including the
@@
 
-\includegraphics[width=5.37135in,height=2.64583in]{media/image25.jpg}
-
-\emph{Figure 16: Approximate percentage of time spent in different
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image19.png}
+\caption{Approximate percentage of time spent in different
 components of the ALICE detector simulation chain, including the Geant
 and digitisation modules (for a representative Pb-Pb Hijing event)}
-
-\includegraphics[width=4.52441in,height=2.60938in]{media/image26.jpg}
-
-\emph{Figure 17: Approximate percentage of time spent in the different
+\label{fig:alicecpu}
+\end{figure}
+
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image20.png}
+\caption{Approximate percentage of time spent in the different
 library components of the ALICE detector simulation chain for the same
-setting as in Figure 16.}
+setting as in Figure 16}
+\label{fig:alicesimchain}
+\end{figure}
 
 \hypertarget{neutrino-experiments}{%
-\subsubsection{8.4 Neutrino Experiments}\label{neutrino-experiments}}
+\subsection{Neutrino Experiments}\label{neutrino-experiments}}
 
@@
 
-\begin{longtable}[]{@{}llllll@{}}
-\toprule
-\endhead
-\begin{minipage}[t]{0.14\columnwidth}\raggedright
-\begin{quote}
-\textbf{Experiment / Simulation Stage CPU} ( sec/event)
-\end{quote}\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-\textbf{Generation}\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-\textbf{Geant4}\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-\textbf{Geant4 with space charge}\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-\textbf{DetSim}\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-\textbf{Reco}\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.14\columnwidth}\raggedright
-\begin{quote}
-\textbf{ProtoDUNE} (Cosmic events with Corsika {[}36{]})
-\end{quote}\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-10\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-180\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-480\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-25\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-1000\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.14\columnwidth}\raggedright
-\begin{quote}
-\textbf{DUNE FD} (v\textsubscript{e} events with GENIE)
-\end{quote}\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-0.04\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-44\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-160\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\emph{Table 5: Summary of CPU time measurements for the different
+\begin{table}[tbhp]
+\centering
+\begin{tabular}{| p{3.5cm} | c | c | p{2.5cm} | c | c |}
+\hline
+Experiment/ Simulation Stage CPU (sec/event) & Generation & Geant4 
+& Geant4 with space charge & DetSim & Reco \\ \hline
+ProtoDUNE (Cosmic events with Corsika {[}36{]}) & 10 & 180  & 
+480 & 25 & 1000 \\ \hline
+DUNE FD ($\nu$\textsubscript{e} events with GENIE) & 0.04 & 4 &   & 
+44 & 160 \\ \hline
+\end{tabular}
+\caption{Summary of CPU time measurements for the different
 modules associated with the ProtoDUNE and DUNE Far Detector (DUNE FD)
 simulation applications.}
-
-\hypertarget{section-4}{%
-\subsubsection{}\label{section-4}}
+\label{table:cpudune}
+\end{table}
 
 \hypertarget{direct-dark-matter-search-experiments}{%
-\subsubsection{8.5 Direct dark-matter search
+\subsection{Direct dark-matter search
 experiments}\label{direct-dark-matter-search-experiments}}
@@
 \hypertarget{rd-towards-improving-software-performance}{%
-\section{9 R\&D Towards Improving Software Performance
+\section{R\&D Towards Improving Software Performance
 }\label{rd-towards-improving-software-performance}}
@@
 propagation in fields fill also important slices in the overall
-performance pie-chart of simulation, as can be observed in Figure 14.
+performance pie-chart of simulation, as can be observed in Figure 13.
 These observations are showing in which areas the R\&D effort should be
@@
 \hypertarget{particle-transport-and-vectorisation}{%
-\subsubsection{9.1 Particle Transport and
+\subsection{Particle Transport and
 Vectorisation}\label{particle-transport-and-vectorisation}}
@@
 accumulate many particles before actually performing the actions
-involved by the stage. As illustrated In Figure 18, the generation of
+involved by the stage. As illustrated In Figure 17, the generation of
 the final states from discrete physics processes can be such a
@@
 
-\includegraphics[width=6.26772in,height=3.52778in]{media/image27.png}
-
-Figure 18: \emph{The stepping procedure in particle transport is the
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image22.png}
+\caption{The stepping procedure in particle transport is the
 sequence of actions allowing to propagate a particle from a starting
@@
 classical stack-based approach.}
+\label{fig:parttransp}
+\end{figure}
 
@@
 \hypertarget{hybrid-computing-and-simulation}{%
-\subsubsection{9.2 Hybrid Computing and
+\subsection{Hybrid Computing and
 simulation}\label{hybrid-computing-and-simulation}}
@@
 evolution of the simulation code that extends the thoroughly validated
-current versions of the Geant4 neutron physics models (see Figure 19).
-
-\includegraphics[width=6.70522in,height=1.38889in]{media/image28.png}
-
-\emph{Figure 19: Proposed architecture of a specialised library to
+current versions of the Geant4 neutron physics models (see Figure 18).
+
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image23.png}
+\caption{Proposed architecture of a specialised library to
 leverage GPU for neutron transport.}
+\label{fig:neutrontransport}
+\end{figure}
 
@@
 \hypertarget{modularisation}{%
-\subsubsection{9.3 Modularisation}\label{modularisation}}
+\subsection{Modularisation}\label{modularisation}}
 
@@
 \hypertarget{simulation-on-high-performance-computing-systems}{%
-\subsubsection{9.4 Simulation on High Performance Computing
+\subsection{Simulation on High Performance Computing
 Systems}\label{simulation-on-high-performance-computing-systems}}
@@
 test concluded that the main bottleneck to achieve scaling linearity is
-concurrent access to I/O resources (see Figure 20). Extensions to Geant4
+concurrent access to I/O resources (see Figure 19). Extensions to Geant4
 capabilities to improve the scalability of the code are underway. These
@@
 
-\includegraphics[width=6.26772in,height=4.625in]{media/image29.png}\emph{Figure
-20: Top: Initial testing of Geant4 on MIRA (blue diamonds) and VESTA
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image24.png}
+\caption{Top: Initial testing of Geant4 on MIRA (blue diamonds) and VESTA
 (green stars). Each node has 80 threads. The loss of scalability is due
@@
 larger. }
-
-\hypertarget{section-5}{%
-\section{}\label{section-5}}
-
-\hypertarget{section-6}{%
-\section{\texorpdfstring{\\
-}{ }}\label{section-6}}
+\label{fig:mira}
+\end{figure}
 
 \hypertarget{fast-simulation}{%
-\section{10. Fast Simulation}\label{fast-simulation}}
+\section{Fast Simulation}\label{fast-simulation}}
 
@@
 \hypertarget{fast-simulation-techniques-and-examples}{%
-\subsubsection{10.1 Fast Simulation techniques and
+\subsection{Fast Simulation techniques and
 examples}\label{fast-simulation-techniques-and-examples}}
@@
 frameworks in experiments is the concept of ``hybrid'' simulation (see
-Sec. 9.2), where a single framework allows to treat some aspects of the
+Section 9.2), where a single framework allows to treat some aspects of the
 event with a fast or very fast simulation technique and others with a
@@
 \hypertarget{physics-performance-of-fast-simulations}{%
-\subsubsection{10.2 Physics Performance of Fast
+\subsection{Physics Performance of Fast
 Simulations}\label{physics-performance-of-fast-simulations}}
@@
 
-Figure 21 shows the speed of CMS Fast Simulation factored into different
+Figure 20 shows the speed of CMS Fast Simulation factored into different
 subprocesses as a function of the number of pileup interactions. The
@@
 
-\includegraphics[width=6.70522in,height=4.93056in]{media/image30.png}
-
-\emph{Figure 21 Speed of CMS Fast Simulation factorised into different
+\begin{figure}[bthp]
+\vspace*{0.3cm}
+\centering
+\includegraphics[width=0.94\textwidth]{image25.png}
+\caption{Speed of CMS Fast Simulation factorised into different
 subprocesses, as explained in Section 10.2, versus number of pileup
 interactions. }
+\label{fig:speedfastsim}
+\end{figure}
 
 \hypertarget{machine-learning-in-fast-simulations}{%
-\subsubsection{10.3 Machine Learning in Fast
+\subsection{Machine Learning in Fast
 Simulations}\label{machine-learning-in-fast-simulations}}
@@
 \hypertarget{sharing-of-fast-simulation-and-interfaces}{%
-\subsubsection{10.4 Sharing of Fast Simulation and
+\subsection{Sharing of Fast Simulation and
 Interfaces}\label{sharing-of-fast-simulation-and-interfaces}}
@@
 \hypertarget{pseudorandom-number-generators}{%
-\section{\texorpdfstring{\textbf{11. Pseudorandom number
-generators}}{11. Pseudorandom number generators}}\label{pseudorandom-number-generators}}
+\section{\texorpdfstring{\textbf{Pseudorandom number
+generators}}{Pseudorandom number generators}}\label{pseudorandom-number-generators}}
 
@@
 
-\hypertarget{section-7}{%
-\section{}\label{section-7}}
-
-\hypertarget{section-8}{%
-\section{\texorpdfstring{\\
-}{ }}\label{section-8}}
-
 \hypertarget{summary-and-outlook}{%
-\section{12. Summary and outlook}\label{summary-and-outlook}}
+\section{Summary and outlook}\label{summary-and-outlook}}
 

### 2018-01-29 CWP/papers/HSF-CWP-2017-05_analysis/latex/cwp-wg-analysis-interpretation.tex
first version of analysis paper converted into HSF github repository

% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.

% JHEP preprint template
\documentclass[12pt,a4paper]{article}
\usepackage{jheppub}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xspace}

\usepackage[utf8]{inputenc}

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp.bib}
\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp-chapters.bib}

%% Comment out before final submission
\usepackage{lineno}  % for line numbering during review
\linenumbers

\abstract{To be written}

\begin{document}

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-05 \\
 & & February 15, 2018 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{HEP Software Foundation Community White Paper Working Group -- Data Analysis and Interpretation}

\author{HEP Software Foundation:}
\author[1]{Riccardo Maria Bianchi}
\author[2]{Brian Bockelman}
\author[3]{Nuno Castro}
\author[4]{Peter Elmer}
\author[16]{Robert Gardner}
\author[14]{Maria Girone}
\author[5]{Oliver Gutsche}
\author[14]{Benedikt Hegner}
\author[6]{Jos\'{e} M. Hern\'{a}ndez}
\author[5]{Bodhitha Jayatilaka}
\author[4]{David Lange}
\author[7]{Mark Neubauer}
\author[7]{Daniel S. Katz}
\author[8]{Lukasz Kreczko}
\author[9]{James Letts}
\author[10]{Shawn McKee}
\author[11]{Christoph Paus}
\author[5]{Kevin Pedro}
\author[12]{Eduardo Rodrigues}
\author[8]{Tai Sakuma}
\author[5]{Elizabeth Sexton-Kennedy}
\author[13]{Michael D. Sokoloff}
\author[9]{Frank W\"{u}rthwein}
\author[15]{Carl Vuosalo}

\affiliation[1]{University of Pittsburgh}
\affiliation[2]{University of Nebraska-Lincoln}
\affiliation[3]{LIP and University of Minho, Supported by FCT-Portugal, IF/00050/2013/CP1172/CT0002}
\affiliation[4]{Princeton University}
\affiliation[5]{Fermi National Accelerator Laboratory, Supported by the US-DOE, DE-AC02-07CH11359}
\affiliation[6]{CIEMAT, Supported by ES-MINECO, FPA2016-80994-c2-1-R and MDM-2015-0509}
\affiliation[7]{University of Illinois Urbana-Champaign}
\affiliation[8]{University of Bristol}
\affiliation[9]{UC San Diego}
\affiliation[10]{University of Michigan, Supported by the US-DOE, DE-SC0007859 and US-NSF, 76749/1136652/2}
\affiliation[11]{Massachusetts Institute of Technology}
\affiliation[12]{University of Cincinnati, Supported by the US-NSF, ACI-1450319}
\affiliation[13]{University of Cincinnati, Supported by the US-NSF, ACI-1558219}
\affiliation[14]{CERN}
\affiliation[15]{University of Wisconsin-Madison}
\affiliation[16]{University of Chicago}

\maketitle

\newpage

\section{Introduction}

High Energy Physics (HEP) answers scientific questions by analyzing the data obtained from detectors and sensors of suitably designed experiments, comparing measurements with predictions from models and theories. The number of different questions that can be answered by a single experiment range from a few for scientifically very focussed devices to a very diverse and large set of questions for multi-purpose devices. In all cases, data are analyzed by groups of researchers of varying sizes, from individual researchers to very large groups of scientists. ``Analysis'' as described here includes the reduction of processed detector data, from the trigger and event reconstruction algorithms, to create, analyze and interpret estimates of specific physics observables. The baseline analysis model utilizes successive stages of data reduction, finally analyzing a compact dataset with quick real time iteration. Experiments use a series of processing steps to reduce large input datasets down to sizes suitable for laptop-scale analysis. The line between managed production-like analysis processing and individual analysis, as well as the balance between harmonized versus individualized analysis data formats differs by experiment based on their needs and optimization level and the maturity of an experiment in its life cycle. The current baseline model stems from the goal to exploit the maximum possible scientific potential of the data while minimizing the ``time to insight'' for a large number of different analyses performed in parallel. It is a complicated product of diverse criteria ranging from computing resources and related innovation to management styles of the experiment collaborations. An evolution of the baseline approach is the ability to produce physics-ready data right from the output of the high-level trigger of the experiment, whereas the baseline approach also depends on further processing of the data with updated or new software algorithms or detector conditions. This could be a key enabler of a simplified analysis model that allows very efficient data reduction.

The goal for future analysis models is to reduce the ``time to insight'' while exploiting the maximum possible scientific potential of the data within the constraints of computing and human resources. Analysis models aim towards giving scientists access to the data in the most interactive way possible, and to enable quick turn-around in iteratively learning new insights from the data. Many analyses have common deadlines defined by conference schedules. The increased analysis activity before these deadlines requires the analysis system to be sufficiently elastic to guarantee a rich set of physics result. Also, heterogeneous computing hardware like GPUs, FPGAs, and new memory architectures are emerging and can be exploited to further reduce the ``time to insight''.

\section{HEP Analysis Software Ecosystem}

Over the past 20 years the HEP community has developed and gravitated around a single analysis ecosystem: ROOT~\cite{Brun1996}. This software ecosystem currently both dominates HEP analysis and impacts the full event processing chain, provides foundation libraries, I/O services, etc. It is an advantage for the HEP community compared to other science disciplines as ROOT provides an integrated and validated toolkit. This lowers the hurdle to start an analysis, enabling the community to talk a common analysis language, as well as making improvements and additions to the toolkit quickly available to the whole community, allowing a large number of analyses to benefit. The open source analysis tools landscape used primarily in industry is however evolving very fast and surpasses the HEP efforts in software development effort and community size.

The emergence and abundance of alternative and new analysis components and techniques coming from industry open source projects is a challenge for the HEP analysis software ecosystem. The community is very interested in using these new techniques and technologies and would like to use these together with established components of the ecosystem. It should  also be possible to interchange old components with new open source components. We propose in the first year to perform R\&D on enabling new open source tools to be dynamically plugged into the existing ecosystem and mechanisms to freely exchange parts of the ecosystem with new components. This could include investigating new ways of package management and distribution following open source approaches. For the 3-year time frame, we propose to research a comprehensive set of bridges and ferries between the HEP analysis ecosystem and the industry analysis tool landscape, where a bridge enables the ecosystem to use an open source analysis tool and a ferry allows to use data from the ecosystem in the tool and vice versa.

The maintenance and sustainability of the current analysis ecosystem is a challenge. The ecosystem supports a number of use cases and both integrates and maintains a wide variety of components. Components have to be prioritized to fit into the available effort envelope, which is provided by a few institutions and less distributed across the community. Legacy and less used parts of the ecosystem are hard to retire and their continued support strain the available effort. In the first year, we propose R\&D to evolve policies to minimize this effort by retiring less used components from the integration and validation efforts. We propose to enable individuals to continue to use retired components by taking over their maintenance and validation following the central efforts of the ecosystem, spending a little of their own effort. 
But not every component can just be retired if it is not used anymore by most of the ecosystem users. Therefore for the 3-year time frame, we propose to evolve our policies on how to adopt new alternatives in the analysis community and how to retire old components of the ecosystem.

\section{Analysis Languages}~\\
{\bf Python:} Given the large investment in existing physics software and the ROOT-based data models of the experiments, when considering reconstruction and other performance-critical code, the community will continue to look to C++. For analysis code, Python has developed into the language of choice in the data science community and also more and more within the HEP community. It is widely used in the data analytics community in other fields of science and industry and is liked for fast development cycles and ease-of-use. Python also has a rich ecosystem of well-maintained and advanced software packages. Python should become a first-class language in HEP, but performance-critical code should be offloaded to libraries, written in the most suitable language for the problem domain, thus allowing optimization ``under the hood'' by experts. Popular examples of such libraries are Numpy~\cite{5725236-numpy}, pandas~\cite{mckinney-proc-scipy-2010}, and Tensorflow~\cite{abadi2016tensorflow}.  Experiences from inside and outside the HEP community indicate that the simpler interfaces and code constructs of Python could reduce the complexity of analysis code and therefore contribute to the decrease in ``time to insight'' for HEP analyses. For example, data frames in pandas can represent arbitrary dimensions of histograms and offer a rich set of operations on the data, such as the split-apply-combine strategy~\cite{JSSv040i01}. The community needs to ensure that senior people are also included in the community language choice and not left behind, and to do so, education and documentation need to be a firm component of making Python a first-class language in HEP. We propose for the first year time frame to finalize full support of Python in our ecosystem including long term maintenance.\\ \\
{\bf Declarative Languages:} Further optimization could be gained by switching to a functional or declarative programming model. This would allow scientists to express the intended data transformation as a query on data. Instead of having to define and control the ``how'', the analyst would declare the ``what'' of his analysis, essentially removing the need to define the event loop in an analysis and leave it to underlying services and systems to optimally iterate over events. Analogously to how programming in C++ abstracts implementation features compared to programming in assembler, it appears that these high-level approaches will allow to abstract from the underlying implementations, allowing the computing systems more freedom in optimizing the utilization of diverse forms of computing resources. We propose to assess the advantages of the usage of a declarative programming model for data analysis over cognitive overhead for newcomers. More concretely, we propose to provide to the analysis community demonstrators to express their analysis following a declarative approach. Such demonstrators should not only help to assess the viability of the programming model but also attempt to fully leverage hardware resources at disposal, heterogeneous or not, both locally and in a distributed fashion. We propose on the 3-year time frame to conclude on the already ongoing R\&D projects (for example TDataFrame in ROOT) and to follow up with additional R\&D projects to develop a prototype functional or declarative programming language model.

\section{Analyzing Data}~\\
{\bf File format and compression algorithms:} The baseline approach to analyzing HEP data roughly follows a common pattern: The data are prepared for analysis by completing most of the CPU-intensive processing steps first and then making the data available to the community. In the following analysis steps the I/O performance for iterating over the data becomes one of the driving factors in minimizing the ``time to insight'' for analyses. There are many file format standards used by a wealth of data analytics tools in other science fields and industry, but the community feels that currently ROOT is best suited to the community's needs. Disk space is usually the key concern of the experiment computing models as disk is the most expensive hardware component. The community extensively uses compression to minimize storage costs. This reduces the performance of purely iterating over the data due to the cost of decompression. To improve on the current state and to enable analysis of much larger datasets than today, R\&D will be needed to continue in file formats, compression algorithms, and new ways of storing and accessing data for analysis. \\ \\
{\bf Analysis Facilities:} towards the High-Luminosity LHC (HL-LHC)~\cite{Apollinari:2284929} we envision dedicated data analysis facilities for experimenters, which provide an extendable environment that in turn provides fully functional analysis capabilities. Such end-to-end engineered facilities will give users easy and well-maintained access to the HEP analysis software ecosystem and to experiment-specific capabilities (software, documentation, databases, etc). They will support complex ``from beginning to end'' analysis workflows including support for provenance tracking, continuous integration, verification, reproducibility, etc. and will provide low-latency rapid-response computing capabilities with access to complete data sets without the user having to fall back to the high-latency ``batch system style'' analysis. ``Primitive'' versions of such analysis facilities are currently provided e.g. at CERN, Fermilab and other places. However for the HL-LHC, such dedicated analysis facilities would provide a much more end-to-end engineered latency-optimized and maintained environment, that then also would provide focus points for support and training of analysts.\\ \\
We propose to focus technology developments to support analysis by prototyping such analysis facilities to develop the capabilities and the required ecosystem to support the analysis use cases mentioned in section~\ref{sec:models}. This prototyping work  includes to continue to investigate optimizing the storage systems and used representation of data on disk together, and also to facilitate the utilization of new additional storage layers like SSD storage and NVRAM-like storage, which exhibit different characteristics than the currently dominant spinning disk installations. This should include a fresh look at the concept of ``virtual data'', optimizing the choice between stored versus re-computed data products. Regarding a work plan for designing and deploying data analysis facilities to support HL-LHC data analysis, much of the required developments can be efficiently driven by a small number of use cases, that, over the course of 5 years, should be refined and realigned with the help of demonstrators and analysis challenges, using an iterative approach towards prototyping such a facility (or facilities, for the different experiments). Detailed plans should be worked out regarding the prototyping of the capabilities of analysis facilities, iteratively scaling up both functionalities and capacities, up to the capabilities required for the HL-LHC. In years 1 to 3, the conceptualization of an end-to-end low-latency response high-capacity analysis facility should be started, with an initial prototype. In years 3 to 5, analysis of data from the 3rd running period of the LHC should be used to evaluate chosen architectures and verify design or provide input for corrective actions. At the end of this process starting in Year 5, a blueprint for remaining developments and system design should become available, in time for deployment.\\ \\
{\bf Non-event data handling:} another area that has not yet received the attention it deserves is the access to non-event data for analysis (cross section values, scale factors, tagging efficiencies, etc.). The community feels that the existing capabilities for event data, namely easy storage of event data of all sorts of different content, a similar way of saving and accessing non-event information during the analysis step is needed. There exist many ways of doing this now, but no commonly accepted and supported way has yet emerged. This could be expanded to think about event versus non-event data in general to support use cases from small data volumes (for example cross sections) to large data volumes (BDTs and NNs). We propose R\&D in the area of non-event information handling on the 3-year time scale, which would facilitate analysis at much higher scales than today.

\section{Analysis Models and Future Systems}\label{sec:models}

Methods for analyzing the data at the LHC experiments have been developed over the years and successfully applied to LHC data to produce physics results during Run-1 and Run-2. Analysis at the LHC experiments typically starts with users running code over centrally-managed data that is of O(100 kB/event) and contains all of information required to perform a typical analysis leading to publication. In this section, we describe some of the proposed analysis models for the future, building on the experiences of the past.

\subsection{Sequential Ntuple Reduction}

The most common approach to analyzing data is through a campaign of data reduction and refinement, ultimately producing data frames (flat ntuples) and histograms used to make plots and tables from which physics inference can be made. In the following, we are using Atlas and CMS computing model to illustrate data volumes involved in LHC analyses. The centrally-managed data are O(100 kB/event) and are typically too large (e.g. O(100 TBs) for 35 $fb^{-1}$ of Run-2 data) to be analyzed by users on their local computing infrastructures like their workgroup servers or laptops. An often stated aim of the data reduction steps is to arrive at a dataset that ``can fit on one's laptop'', presumably to facilitate low-latency, high-rate access to a manageable amount of data during the final stages of analysis.  At its core, creating and retaining intermediate datasets from a data reduction campaign, bringing and keeping them ``close'' (e.g. on laptop/desktop) to the analyzers, is designed to minimize latencies and risks related to resource contention.

\subsection{``Spark''-like Analysis}

A new model of data analysis, developed outside of HEP, maintains the concept of sequential ntuple reduction but mixes interactivity with batch processing. Spark is one such system, but TensorFlow~\cite{abadi2016tensorflow}, Dask~\cite{rocklin2015dask}, Blaze~\cite{wiebe2014blaze}, Parsl~\cite{babuji_yadu_2017_853492}, Pachyderm~\cite{thepachydermteam}, and Thrill~\cite{bingmann2016thrill} are others. Distributed processing is either launched as a part of user interaction at a command prompt or wrapped up for batch submission. The key differences from the above are:
\begin{itemize}
\item Parallelization is implicit through map/filter/reduce functionals,
\item Data are abstracted as remote, distributed datasets, rather than files,
\item Computation and storage are mixed for data locality: a specialized cluster must be prepared, but can yield higher throughput.
\end{itemize}

A Spark-like analysis facility would be a shared resource for exploratory data analysis (e.g. making quick plots on data subsets through the spark-shell) and batch submission with the same interface (e.g. substantial jobs through spark-submit). The primary advantage that software products like Spark introduce is in simplifying the user's access to data, lowering the cognitive overhead of setting up and running parallel jobs. Certain types of jobs may also be faster than batch processing, especially flat ntuple processing (which benefits from SQL-like optimization) and iterative procedures such as fits and machine learning (which benefit from cluster-wide cache).

Although Spark itself is the leading contender for this type of analysis, as it has a well developed ecosystem with many third-party tools developed by industry, it is the style of analysis workflow that we are distinguishing here rather than the specific technology present today. Spark itself is hard to interface with C++, but this might be alleviated by projects such as ROOT's TDataFrame, which presents a Spark-like interface in ROOT, and may allow for more streamlined interoperability.

\subsection{Query-based Analysis}

An alternative to the previously described models is to perform fast querying of  centrally-managed data and compute remotely on the queried data to produce the analysis products of interest, especially but not limited to histograms. This is similar to NoSQL databases tuned for exploratory data analysis, such as Spark-SQL~\cite{Armbrust:2015:SSR:2723372.2742797}, Impala~\cite{bittorf2015impala}, Kudu~\cite{t.lipcond.alvesd.burkertj.d.cryansa.dembom.percys.rusd.wangm.bertozzic.p.mccabea.wang.2015}, Hawq~\cite{Chang:2014:HMP:2588555.2595636}, Apache Drill~\cite{drill}, and Google Dremel/BigQuery~\cite{36632}, but with a focus on distribution plotting, rather than SQL table transformation.

In one vision for a query-based analysis approach, a series of analysis cycles, each of which provides minimal input (queries of data and code to execute), generates the essential output (histograms, ntuples, etc.) which can be retrieved by the user. The analysis workflow should be accomplished without focus on persistence of data traditionally associated with data reduction. However, transient data may be generated in order to efficiently accomplish this workflow and optionally could be retained to facilitate an analysis ``checkpoint'' for subsequent execution. In this approach, the focus is on obtaining the analysis end-products in a way that does not necessitate a data reduction campaign and associated provisioning of resources. A schematic depiction of a possible future query-based analysis system is shown in Fig~\ref{fig:query_system}.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{query_based_analysis_system.png}
\caption{\label{fig:query_system}Schematic depiction of a possible future query-based analysis system.}
\end{center}
\end{figure}

~\\Advantages of a query-based analysis include:\\ \\
{\bf Minimalistic Analysis:} A critical consideration of the Sequential Ntuple Reduction method might reasonably question why analyzers would bother to generate and store intermediate data to get to the same outcomes of interest (histograms, etc). A more economical approach is to provide only the minimal information -- code providing instructions for selecting the dataset, events of interest, and items to plot. \\ \\
{\bf Democratization of Analyses:} In the Sequential Ntuple Reduction method, as one gets further down the data reduction chain, the user (or small group of users) needs to figure out how to provision and manage the storage required to accommodate this intermediate data which in many cases is accessed with small ($<10^{-4}$) or zero duty cycle. For small groups, the resources required (both in personnel and hardware) to execute such a data reduction campaign might be prohibitive in the HL-LHC era, effectively ``pricing them out'' of contributing strongly to analyses -- possibly a lost opportunity for innovation and discovery. Removing the requirements on storing intermediate data in the analysis chain would help to ``democratize'' data analysis and streamline the overall analysis workflow.\\ \\
{\bf Ease of Provenance:} The query-based analysis provides an opportunity for autonomous storage of provenance information, as all processing in an analysis step from ``primary'' analysis-level data to the histograms is contained to a given facility. This information can be queried as well, for example.\\ \\
Key elements of the required infrastructure for a future query-based analysis system are expected to include:\\ \\
{\bf Sharing resources with traditional systems:} Unlike a traditional batch system, access to this query system is intermittent, so it would be hard to justify allocating exclusive resources to it. Even with a large number of users to smooth out the minute-by-minute load, a query system would have a strong day-night load difference effect, weekday-weekend effect, and pre-conference effect. Therefore, the query system must share resources with a traditional batch system (performing event reconstruction for instance). Then the query system could elastically scale in response to load, preempting the batch system.\\ \\
{\bf Columnar Partitioning of Analysis Data:} Organizing data to enable fast-access of hierarchical event information (``columnar'' data) is both a challenge and an opportunity. Presenting column partitions to analysis systems as the fundamental unit of data management as opposed to files containing collections of events would bring several advantages for HEP end-user analysis (not reconstruction). These column partitions would become first-class citizens in the same sense that files are today: either as single-column files or more likely as binary blobs in an object store. We note that columns are already a first-class citizen in the ROOT file system, however, appropriate data management and analysis software that leverages this capability is missing. Given a data store full of columns, datasets become loose associations among these columns, with metadata identifying a set of columns as mutually consistent and meaningful for analysis.\\ \\
{\bf Fast Columnar Data Caching:} Columnar cache is a key feature of the query system, retaining input data between queries, which are usually repeated with small modifications (intentionally as part of a systematics study or unplanned as part of normal data exploration). RAM cache would be a logical choice, given the speed of RAM memory, but the query system can't hold onto a large block of RAM if it is to share resources with a batch system. Furthermore, it can't even allocate large blocks of RAM temporarily, since this would trigger virtual memory swapping to a disk that is slower than the network it is getting the source data from. The query system must therefore stay within a tight RAM budget at all times. The query system's cache would therefore need to be implemented in SSD (or some future fast storage, such as X-Point). We can assume the query system would have exclusive access to an attached SSD disk, since caching is not required for the batch process.\\ \\
{\bf Provenance:} The query system should also attach enough provenance to each dataset that it could be recreated from the original source data, which is considered immutable. User datasets, while they can't be modified in-place, can be deleted, so a dataset's paper trail must extend all the way back to source data. This paper trail would take the form of the original dataset name followed by queries for each step of derivation: code and closure data.\\ \\
Additional details regarding the proposed query-based analysis approach, requirements and columnar management for HEP data can be found in the contributed CWP white paper ``Hierarchical Data Rolled Up in Columnar Arrays''~\cite{Pivarski2017}.

\section{Analysis Preservation}

Reproducibility is the cornerstone of scientific results and while HEP does not face a reproducibility crisis, it is currently difficult to repeat most HEP analyses after they have been completed. This difficulty mainly arises due to the number of scientists involved, the number of steps in a typical HEP analysis workflow, and the complex ecosystem of software that HEP analyses are based on. In addition to the over-arching desire for scientific reproducibility, the ability to preserve and reproduce analyses also has immediate practical benefits to the collaborations and the LHC program. In particular, collaborations often want to extend completed analyses with more data, but this typically involves a lot of turnover in the analysis team. Similarly, if something anomalous shows up in the data, we often want to go back to an earlier version of the analysis and scrutinize it. Traditionally, both of these tasks have been difficult.

Analysis preservation and reproducibility strategies are described in the Data and Software Preservation White Paper~\cite{HSF-CWP-2017-06}. Recent progress using workflow systems and containerization technology have rapidly transformed our ability to provide robust solutions. Nevertheless, analysis capture is best performed while the analysis is being developed in the first place. Thus, reproducibility needs to be considered in all new approaches under investigation and needs to be a fundamental component of the system as a whole. These considerations become even more critical as we explore analysis models with more heterogeneous hardware and analysis techniques.

\section{Analysis Interpretation}

The LHC collaborations are pursuing a vast number of searches for new physics. Interpretation of these analyses sits at the heart of the LHC physics priorities, and aligns with using the Higgs as a tool for discovery, identify the new physics of dark matter, and explore the unknown of new particles, interactions, and physical principles. The collaborations typically interpret these results in the context of specific models for new physics searches and sometimes reinterpret those same searches in the context of alternative theories. However, understanding the full implications of these searches requires the interpretation of the experimental results in the context of many more theoretical models than are currently explored by the experiments. This is a very active field, with close theory-experiment interaction and with several public tools in development.

A Forum~\cite{LHCInterpretationForum} on the interpretation of the LHC results for Beyond Standard Model (BSM) studies was thus initiated to discuss topics related to the BSM (re)interpretation of LHC data, including the development of the necessary public recasting tools\footnote{``Recasting'' is a term used for this style of reinterpretation~\cite{Cranmer:1299950}.} and related infrastructure, and to provide a platform for a continued interaction between the theorists and the experiments.

The infrastructure needed for analysis reinterpretation is a focal point of other cyber infrastructure components including the INSPIRE literature database~\cite{INSPIRE}, the HEPDatadata repository~\cite{HEPDataRepo} , the CERN Analysis Preservation framework~\cite{CAP}, and the REANA cloud-based workflow execution system~\cite{REANA}. Critically, this cyber infrastructure sits at the interface between the theoretical community and various experimental collaborations. As a result, this type of infrastructure is not funded through the experiments and tends to fall through the cracks. Thus, it is the perfect topic for a community-wide, cross-collaboration effort.

One specific piece of infrastructure that is currently missing is an analysis database able to represent the many-to-many mapping between publications, logical labels for the event selection defining signal and control regions, data products associated to the application of those event selections to specific datasets, the theoretical models associated to simulated datasets, the multiple implementations of those analyses from the experiments and theoretical community created for the purpose of analysis interpretation, and the results of those interpretations.

The protocol for (re)interpretation is clear and narrowly scoped, which makes it possible to offer it as a service. This type of activity lends itself to the Science Gateway\footnote{Science gateways allow science and engineering communities to access shared data, software, computing services, instruments, educational materials, and other resources specific to their disciplines~\cite{SciGateway}.} concept, which we refer to as the Interpretation Gateway. Such reinterpretation services have been foreseen for several years, and now most of the necessary infrastructure is in place to create it\footnote{RECAST: a reinterpretation service~\cite{Cranmer:1299950}, DPHEP Status Report~\cite{Akopov:2012bm}, ATLAS Data Access policy~\cite{ATL-CB-PUB-2015-001}.}. Such an interpretation service would greatly enhance the physics impact of the LHC and also enhance the legacy of the LHC well into the future.

\section{Analysis Roadmap}

\subsection{1-year Time Frame}

\begin{itemize}
\item Enable new open source tools to be plugged in dynamically in the existing ecosystem and mechanisms to dynamically exchange parts of the ecosystem with new components.
\item Develop requirements and design a next generation analysis facility concept, incorporating fast caching technologies to explore a query-based analysis approach and open-source cluster management tools.
\item Finalize full support of python in our ecosystem including long term maintenance.
\item Evolve policies to minimize this effort by retiring less used components from the integration and validation efforts.
\item Establish a schema for the analysis database.
\item Interpretation Gateway: conceptualization integrating the analysis facility, analysis preservation infrastructure, data repositories, and recasting tools. 
\end{itemize}

\subsection{3-year Time Frame}

\begin{itemize}
\item Research a comprehensive set of bridges and ferries, where a bridge enables the ecosystem to use an open source analysis tool and a ferry allows to use data from the ecosystem in the tool and vice versa.
\item Analysis facility: conceptualization of an end-to-end low-latency response high-capacity analysis facility with an initial prototype.
\item Develop a prototype functional or declarative programming language model.
\item Evolve our policies on how to replace components with new tools, maybe external, and solicit the community help in bridging and integrating them.
\item R\&D in the area of non-event information handling.
\item Develop a functional prototype for the Interpretation Gateway.
\end{itemize}

\subsection{5-year Time Frame}

\begin{itemize}
\item Analysis facility: evaluate chosen architectures and verify design or provide input for corrective actions (data from the 3rd running period of the LHC should be used).
\begin{itemize}
\item In year 5, a blueprint for remaining developments and system design should become available, in time for deployment.
\end{itemize}
\item Interpretation Gateway: evaluate design or provide input for corrective actions to enable an LHC legacy (re)interpretation gateway.
\end{itemize}

\sloppy
\raggedright
\clearpage
\printbibliography[title={References},heading=bibintoc]

\end{document}

### 2018-01-29 CWP/papers/HSF-CWP-2017-05_analysis/authors/a2tex.py

@@ -1,0 +1,205 @@
+#! /usr/bin/env python
+#
+# Process CWP signatory list into a LaTeX author list
+#
+# The data source is expected to be made of 3 text files:
+#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey [- Role ] (i)
+#      with i a footnote number. If there are multiple affiliations, separate AffiliationKeys with &.
+#    - footnotes:txt: the list of footnotes. There must be one footnote per line starting with 'i.'
+#    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
+#      AffiliationKey must not contain any space.
+#
+# These 3 files are typically built from the Google Docs https://docs.google.com/document/d/1tBXwlNnQsxxZA3gVS1_KSpa8wRXGyk250EIsJwJ2T34
+#
+# Note: this script requires Python >= 3.5
+#
+# Initial version written by Graeme Stewart (graeme.andrew.stewart@cern.ch) and Michel Jouvin (jouvin@lal.in2p3.fr)
+
+import sys
+PYTHON_MIN = (3, 5)
+major, minor, _, _, _ = sys.version_info
+if (major, minor) < PYTHON_MIN:
+    python_min_str = '.'.join(str(x) for x in PYTHON_MIN)
+    # Keep old formatter syntax to be sure it works with old version of Python
+    print ("This script requires Python %s or later" % python_min_str)
+    sys.exit(2)
+
+import os
+import re
+from operator import attrgetter
+import argparse
+
+# Define some constants related to application exit status
+EXIT_STATUS_SUCCESS = 0
+EXIT_STATUS_OPTION_ERROR = 3
+EXIT_STATUS_FAILURE = 5
+
+# Default input files
+AUTHOR_FILE_DEFAULT = "authors.txt"
+FOOTNOTE_FILE_DEFAULT = "footnotes.txt"
+AFFILIATION_ADDRESS_FILE_DEFAULT = "address.txt"
+LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
+ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
+
+
+class Footnote():
+    def __init__(self, number_str, text):
+        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
+        self.mark = footnote_symbols[int(number_str)-1]
+        self.text = text
+    def __repr__(self):
+        return repr((self.mark, self.text))
+
+class Author():
+    def __init__(self, author_line):
[…]

### 2018-01-29 CWP/papers/HSF-CWP-2017-05_analysis/authors/a2tex.py
remove unneeded author setup copy from roadmap

@@ -1,205 +1,0 @@
-#! /usr/bin/env python
-#
-# Process CWP signatory list into a LaTeX author list
-#
-# The data source is expected to be made of 3 text files:
-#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey [- Role ] (i)
-#      with i a footnote number. If there are multiple affiliations, separate AffiliationKeys with &.
-#    - footnotes:txt: the list of footnotes. There must be one footnote per line starting with 'i.'
-#    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
-#      AffiliationKey must not contain any space.
-#
-# These 3 files are typically built from the Google Docs https://docs.google.com/document/d/1tBXwlNnQsxxZA3gVS1_KSpa8wRXGyk250EIsJwJ2T34
-#
-# Note: this script requires Python >= 3.5
-#
-# Initial version written by Graeme Stewart (graeme.andrew.stewart@cern.ch) and Michel Jouvin (jouvin@lal.in2p3.fr)
-
-import sys
-PYTHON_MIN = (3, 5)
-major, minor, _, _, _ = sys.version_info
-if (major, minor) < PYTHON_MIN:
-    python_min_str = '.'.join(str(x) for x in PYTHON_MIN)
-    # Keep old formatter syntax to be sure it works with old version of Python
-    print ("This script requires Python %s or later" % python_min_str)
-    sys.exit(2)
-
-import os
-import re
-from operator import attrgetter
-import argparse
-
-# Define some constants related to application exit status
-EXIT_STATUS_SUCCESS = 0
-EXIT_STATUS_OPTION_ERROR = 3
-EXIT_STATUS_FAILURE = 5
-
-# Default input files
-AUTHOR_FILE_DEFAULT = "authors.txt"
-FOOTNOTE_FILE_DEFAULT = "footnotes.txt"
-AFFILIATION_ADDRESS_FILE_DEFAULT = "address.txt"
-LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
-ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
-
-
-class Footnote():
-    def __init__(self, number_str, text):
-        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
-        self.mark = footnote_symbols[int(number_str)-1]
-        self.text = text
-    def __repr__(self):
-        return repr((self.mark, self.text))
-
-class Author():
-    def __init__(self, author_line):
[…]

### 2018-01-31 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
starting to add references in bibtex format

@@
 neutrino beam and the first neutrino-nucleus interaction in the detector
-is typically modeled with the GENIE {[}7{]} package, while Geant4 is
+is typically modeled with the GENIE ~\cite{}{[}7{]} package, while Geant4 is
 utilised in a second step to simulate the nucleus de-excitation and

### 2018-01-31 CWP/papers/HSF-CWP-2017-08_framework/latex/df_fwk_summary.tex
as it is in google docs

% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.
%
% original source of this text was:
%  https://docs.google.com/document/d/1DYEHGgB3fanhpYRDJblE9NicX0OK7BNsjTCrq7gY-sU/edit#
%

% JHEP preprint template
\documentclass[12pt,a4paper]{article}
\usepackage{jheppub}

\usepackage{graphicx}
% \usepackage{subcaption}
\usepackage{xspace}

\usepackage[utf8]{inputenc}

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{../../../papers/HSF-CWP-2017-01_roadmap/latex/cwp.bib}
\addbibresource{../../../papers/HSF-CWP-2017-01_roadmap/latex/cwp-chapters.bib}

%% Comment out before final submission
%\usepackage{lineno}  % for line numbering during review
%\linenumbers

% Use \ifixme{Some short text} to typeset an identifiable brief note for
% your co-authors that will stand out in the text.
\newcommand{\ifixme}[1]{{\slshape\color{cyan}\textbf{FIXME: } #1}}
\newcommand{\etc}{\textit{etc}}

\abstract{Abstract goes here.}

\begin{document}

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-08 \\
 & & December 21, 2017 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{HEP Software Foundation Community White Paper -- Data Processing Framework WG Report}

\author{HEP Software Foundation:}
\author[d]{Paolo Calafiura}
\author[a,1]{Benadikt Hegner}
\author[c]{Chris Jones}
\author[c,1]{Jim Kowalkowski}
\author[c,1]{Elizabeth Sexton-Kennedy}
\author[b]{Michel Jouvin}
\author[d]{Several Others}

\affiliation[a]{CERN, Geneva, Switzerland}
\affiliation[b]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\affiliation[c]{Fermi National Accelerator Laboratory, Batavia, Illinois, USA}
\affiliation[d]{Lawence Berkeley National Laboratory, Berkeley, CA, USA}
\affiliation[e]{Other Places}
\affiliation[1]{Paper Editor}

\maketitle

\newpage

\section{Introduction}
The aim of this document is to promote a common vision and roadmap for
data processing frameworks that will allow enhanced collaboration
across experiments and will meet future challenges given the variety
of stakeholders which will be defined below. We describe the problem
domain and introduce the concepts and relationships needed to
formulate common data processing framework solutions for the future.

The time periods of interest for this document are DUNE and HL-LHC,
which will deliver on the order of 50 PB data events per year per
experiment . The results of the proposed R\&D ought to be used for
building the final software systems that will be utilized in
commissioning and operations of these experiments and the processing
of the data.

\section{Scope and Challenges}
\label{sec:scope-challenges}

Frameworks in HEP are used for the collaboration-wide data processing
tasks of triggering, reconstruction, and simulation, as well as other tasks that
subgroups of an experiment collaboration are responsible for, such as
detector alignment and calibration. 
Providing common framework services and libraries that will meet with
compute and data needs for HL-LHC experiments and the Intensity Frontier
experiments is a large challenge given the multi-decade legacy in this
domain. At the same time the computing landscape is changing requiring
enhanced framework functionality to help users adapt to these changes.
We see a number of upcoming challenges that need to be addressed in the
coming decade:

\begin{enumerate}
\item
    Changes needed in the programming model that are necessary to
    handle the massive parallelism that will be present throughout all
    layers in the available compute facilities. This is necessary
    because of the ever-increasing availability of specialized compute
    resources, and includes GPGPUs, Tensor Processing Units (TPUs),
    and tiered memory systems integrated with storage, ultra
    high-speed network interconnects.
\item
    Challenges related to advanced detector technology, like finer
    granularity high bandwidth continuous readout DAQ, and the need
    for a very tight feedback loop for conditions which in turn
    requires a focus on low-latency solutions.
\item
    Tighter integration with the computing model: informing
    higher-level systems (workflow, data management, workload
    management, job management) based on fundamental changes in
    compute facilities.
\item
    Support structures that permit frameworks to be collaboratively
    developed and maintained across a large number of
    experiments. This includes excellent integration with the wider
    ecosystem encompasses development, deployment, and runtime
    components.
\item
    Provide flexibility and the interfaces required to efficiently
    integrate the various parallelization efforts on simulation,
    reconstruction and other fields.
\end{enumerate}

The challenge present in all of these elements is to ensure
productivity given the increased complexity and scale of the upcoming
experiments and computing facilities. This means decreasing program
development and debugging time, and increasing efficiencies in diverse
facilities use.  Frameworks have accomplished these tasks in the past,
the challenge is to continue this into the next generation experiments
and facilities.

\subsection{Stakeholders}
\label{sec:stakeholders}

Understanding who influences the organization and behavior of the
framework is obviously important. This is typically accomplished by
collecting requirements and use cases. This paper does not pretend to
do this. We do, however, recognize the stakeholders that heavily
influence the requirements placed upon an event processing
framework. In this document we recognise that there are needs from
several kinds of stakeholders depicted in
Figure~\ref{fig:stakeholders} and listed below.  We attempt to address
the needs of all of these stakeholders.

\begin{enumerate}
\item
    Physics developers who write trigger, reconstruction, simulation,
    and analysis algorithms that plug into the framework. This
    category also includes providers of generalized infrastructures
    which are used by the physics developers to implement the above
    applications. Examples are the upgraded Geant simulation engine
    and online triggering infrastructures.These are important because
    interfacing to them may impose special requirements on the
    Framework.
\item
    Physics users who run framework programs for particular
    applications to produce collaboration wide datasets.
\item
    Funding agencies/laboratories, who mandate security requirements,
    or dictate the need for shared software infrastructure projects
\item
    Facilities who buy, operate, and build new systems that existing
    software infrastructures must be adapted to or rethought for
\end{enumerate}

\begin{figure}[hbt] %{0.85\textwidth}
\centering
\includegraphics[width=.75\textwidth]{stakeholders.pdf}
\caption{Our stakeholders. \label{fig:stakeholders}}
\end{figure}

\section{Current Practice}
\label{sec:current-practice}

Although most frameworks used in HEP share common concepts, there are,
for mainly historical reasons, a number of different implementations;
some of these are shared between experiments. A complete description
of framework use cases written for the Gaudi collaboration is
described in {[}2.fixme{]} and these are sufficiently general to apply
to all HEP experiments. The Gaudi framework {[}70{]} was originally
developed by LHCb, but is also used by ATLAS and various non-LHC
experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff},
which was forked to provide the art framework for the Fermilab
Intensity Frontier experiments~\cite{Green:2012gv}.  Belle II uses
basf2~\cite{1742-6596-331-3-032024}. The linear collider community
developed and uses Marlin~\cite{Gaede:2006pj}. The FAIR experiments
use FairROOT, closely related to ALICE's AliROOT. The FAIR experiments
and ALICE are now developing a new framework, which is called
O2~\cite{O2}. At the time of writing, most major frameworks support
basic parallelisation, both within and across events, based on a
task-based model~\cite{Jones:2015soc}~\cite{Clemencic:2015paa}.

The frameworks provide the necessary functionality like I/O,
scheduling, configuration, logging, etc to support the execution of
these processing components. The mentioned components provide
functionalities like pattern finding in a certain sub-detector or the
high-level identification of a given particle type. This layout allows
independent development and a high flexibility in the usage of physics
algorithms within an experiment collaboration.

The above defines frameworks in terms of its use. A more formal
definition is the framework holds the protocols, tools and concepts
for defining, developing, and deploying physics algorithms, along with
all the ancillary data and tools for providing services to the
algorithms.  This includes algorithm scheduling components, the event
data model, handling of input and output from physics applications
that utilize the framework, interfaces for non-event data, and
configuration of framework applications. Frameworks define processing
and programming models for a collaboration, as well as fulfill
requirements for interfacing to the computing model under which they
operate. The processing model is the mechanism used to execute and
apportion work. Mechanisms for this are threads, tasks, heavy-weight
processes along with interprocess communication. Programming model
elements, such as a well-defined logical design, scheduling, and
interactions with multiple languages that permit efficient and
maintainable algorithms, are also dictated by the processing
framework. The programming model also defines a well-thought out
physical code layout that minimizes coupling of logically independent
functionalities and libraries. This eases maintenance, extension and
refactoring, which is inevitable over the lifetime of HEP
experiments. The framework should provide hooks for doing monitoring
and logging services for performance and other purposes.

We have identified two kinds of required behavior that have directly
affected overall software system architecture, and therefore the
organization of the framework itself:

\begin{itemize}
\item
    Throughput maximizing: here it is most important to efficiently
    move data through all the available resources (memory, storage,
    and CPU), maximizing the number of events that are processed. The
    workload management systems used by experiments on the grid work
    towards this goal.
\item
    Latency minimizing (or reducing): online and interactive use
    cases where imposing constraints on how long it takes to calculate
    an answer for a particular datum is relevant and
    important. Dataflow and transaction processing systems work
    towards this goal.
\end{itemize}

Whether or not these differences necessitate fundamentally different
software systems remains to be seen, but both concepts need to be
accommodated and have relevant impact on the framework architecture
and design. Ways of accommodating both goals through the same system
architecture, or same software system with different configurations is
an area of research, and could enable us to meet some of the
challenges introduced in the previous section.

Current practice for throughput-maximizing system architectures have
constrained the scope of framework designs. Framework applications
have largely been viewed by the system as a batch job with complex
configuration, consuming resources according to rules dictated by the
computing model: one process using one core on one node, operating
independently with a fixed size memory space on a fixed set of files
(streamed or read directly). Only recently has CMS broken this
tradition starting in the beginning of Run 2, by utilizing all cores
on one virtual node in one process space using threading. ATLAS is
currently using a multi-process fork-and-copy-on-write solution to
remove the constraint of one core/process. Both experiments were
driven to solve this problem, by the ever growing needs for more
memory per process brought on by the increasing complexity of LHC
events. Current practice manages system-wide (or facility-wide)
scaling by dividing up datasets, generating a framework application
configuration, and scheduling jobs on nodes/cores to consume all
available resources. Given anticipated changes in hardware
(heterogeneity, connectivity, memory, storage) available at large
computing facilities, the interplay between workflow/workload
management systems and framework applications need to be carefully
examined. It may be advantageous to permit framework applications (or
systems) to span resources, permitting them to be first-class
participating in the business of scaling within a facility.

\section{Roadmap}
\label{sec:roadmap}

\subsection{One-year goals}
\label{sec:one-year-goals}

Our assumption is that a set of one-year goals will be completed by
the end of 2018. We want work completed here to produce results that
will be useful in refining and moving into satisfying the three-year
goals.  Below is a list of all the areas that we believe need work and
ought to be considered when establishing projects that will help meet
all or some subset of these goals. Since this is R\&D, overall goals
are papers, workshop reports, analysis results, and software
architectures. A major purpose being to set direction for further
R\&D, or defining a path towards a new production product development,
or integration of results into existing software components.

\paragraph{Concept refinement} Jointly identify key abstractions that
made frameworks good for HEP in detail beyond what can be described in
this paper. Identify and describe where individual frameworks have
similarly or uniquely implemented these concepts. A publishable paper
should come of this that will serve as an agreed-upon guide for where
we can hope to go.

\paragraph{Technology investigations} There are three key areas that
ought to be explored to help determine future direction with regards
to software technology. The areas are: (1) task-based programming
tools, (2) parallel number crunching libraries, and (3) framework
workflow management.

\paragraph{Functional programming} Conduct a study describing where we
currently are with functional programming. The study should address
the following questions. 1. What are the perceived benefits and
drawbacks?  2. Where can it be useful? 3. What does it mean in terms
of framework changes? 4. Where is language support headed for this
kind of programming, and recommendations for where to go next.

\paragraph{Support multiple concurrent scheduling tools, strategies,
  time scales, and granularities} Describe how an event-parallel
execution framework handles event-asynchronous I/O and processing
(e.g. alignment, offloading). Describe how underlying concurrency
libraries and tools can be shared. Describe how data blocks having
different granularities (e.g.  groups of particles, events) are
processed concurrently and synchronized. DNN (including simulation
GANs) offloading to GPU could be an early example to work on. A
parallelized Geant particle transport mechanism would be another one.

\paragraph{Domain-specific language} We believe it is useful to
describe the language that HEP uses to define and describe how to
simulate and reconstruct physics events. By language, we mean the
terms used to communicate and express how tasks are described and
carried out within a framework. This includes not only expressing data
dependencies, but also resources preferences and constraints, such a
GPU. The goal here is to provide enough information for a group to
take on development of domain-specific libraries components and tools
that will increase the efficiency of carrying out physics.
\ifixme{the wording here needs much improvement. Maybe a good example
  is how ML toolkits have evolved over the past few years - the
  abstractions that have been developed have greatly increased
  productivity and growth in the ML space}

\paragraph{Coupling with ongoing R\&D activities} Forward-looking work
is underway as part of projects funded through government agencies,
laboratories, and collaborations. We want to be sure that relevant
ideas and accomplishment are known, and that the groups doing this
work have a place to report to and receive feedback for everyone's
benefit.

\paragraph{Concluding workshop} Soon after the end of the first year,
our goal is to have a workshop. This workshop will be used for
reporting on results, reviewing alignment with written three-year
goals, and agreeing upon and developing next steps forward.

\subsection{Three-year goals}
\label{sec:three-year-goals}

Our assumption is that a set of three-year goals will be completed by
the end of 2020. Since LHC run three is targeted to start at this
time.  At least some of the focus ought to be on completing goals that
demonstrate advances over what will be in production during this
period.  An ongoing goal during this period ought to be incorporating
advancements in practices and tools into codes that will be available
during Run 3.

\paragraph{Common feature definition} Produce reports on how
frameworks ought to evolve to incorporate features of functional
programming, scheduling across heterogeneous resources, polyglot
programming, and addressing both necessary data model changes and I/O
handling.

\paragraph{Technology upgrades} Useful technologies identified during
the first phase ought to be further demonstrated and incorporated into
existing tools wherever possible.

\paragraph{Keeping up with evolving facilities} Memory, storage, and
network changes will be introduced during this timeframe. We want
studies that show how we might benefit from these changes. A goal is
to provide recommendations for how to react to ongoing facility
upgrades in the three areas listed above.

\paragraph{Common library integration} Plans for breaking out common
tools should be made during this period, along with demonstrating how
current frameworks might evolve to share more components.

\paragraph{Geant-framework cohesion} Make the next steps in bringing
underlying technology, toolkits, and designs together so Geant fits
well within the framework. A demonstration should be available in at
least one of the frameworks used in Run 3 of the LHC.

\paragraph{Blending of Workflow Management System functions}
Multi-node scheduling and heterogeneous resource management will start
to become more relevant during this period. Frameworks ought to share
responsibility with larger orchestration systems, and better exchange
information to increase flexibility in scheduling on diverse platforms
and architectures.

\paragraph{Facilities embedding} We have a consensus that the main
programming language we will be using in the near future is C++.  Much
of our community stresses use of modern language features. However,
some supercomputing facilities require the compilation of software
with dedicated compiler setups, thus making use of new language
features difficult. By the time of the milestone, we should have
members of the HEP community embedded in the decision-making process
for the provisioning of these machines. Since other programming
languages are also used, the same arguments can be made for them.

\paragraph{Continued coupling with ongoing R\&D activities} Projects
funded outside this work will continue through this period, and our
one-year goals ought to continue here.

\paragraph{Progress workshops} Workshops should happen at the end of
each year, with goals similar to those listed in the previous section.

\subsection{Five-year goals}
\label{sec:five-year-goals}

Our assumption is that a set of five-year goals will be completed by
the end of 2022. By this time we ought to have in place plans for
taking the R\&D results and incorporating them into the framework
software that will be further developed and utilized for HL-LHC. We
anticipate to work on these items:

\paragraph{Facilities readiness} A proven strategy for the integration
of HEP software frameworks with supercomputer centres and cloud
providers.  This work will be done in cooperation with the facilities
WG.

\paragraph{Common tools and practices demonstrated and in place where
  possible} Based on the common tools and components identified during
the \emph{common library integration} milestone, work on production
quality framework libraries for use by multiple experiments.

\paragraph{Interaction with Workflow and Data Management Systems} By
this time well-defined interfaces between those systems and frameworks
should be prepared, including at least one proven reference
implementation.

\paragraph{Incorporation of results from ongoing R\&D activities}
There will be independent progress on parallelization strategies and
implementations. At the time scale of 5 years we anticipate at least
one major paradigm shift to take place, which can not be incorporated
by continuous adjustment alone.

\paragraph{Defining what happens next} Based on the results of the
mentioned activities and the results of the other HSF working groups,
re-evaluate and define a concrete strategy for common framework
implementations for the coming years.

\subsection{Working towards these goals}
\label{sec:towards-goals}

To organize the community and to work towards these goals, one needs
to establish regular working group meetings, on a bi-monthly basis as
we did with the concurrency forum. Face to face workshops after at
least the 1st and the 3rd year can be co-hosted with events like CHEP
and/or the WLCG workshops. A future planning workshop for transforming
the results of the R\&D activities into a full development and
deployment project plan should happen at the 5-year timescale.

\sloppy
\raggedright
\clearpage
\printbibliography[title={References},heading=bibintoc]

\end{document}

### 2018-02-01 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
finished adding bibtex entries in cwp.bib file

--- 2018-01-29 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
@@
 \usepackage[style=numeric-comp,sorting=none]{biblatex}
-\addbibresource{../../roadmap/latex/cwp.bib}
-\addbibresource{../../roadmap/latex/cwp-chapters.bib}
+\addbibresource{../..//HSF-CWP-2017-01_roadmap/latex/cwp.bib}
+\addbibresource{../..//HSF-CWP-2017-01_roadmap/latex/cwp-chapters.bib}
 
@@
 neutrino beam and the first neutrino-nucleus interaction in the detector
-is typically modeled with the GENIE {[}7{]} package, while Geant4 is
+is typically modeled with the GENIE ~\cite{Andreopoulos:2009rq} package, while Geant4 is
 utilised in a second step to simulate the nucleus de-excitation and
@@
 through the lifetime of the experiments running through the 2020's and
-2030's. One example is the GeantV {[}2{]} R\&D project, started in 2013.
+2030's. One example is the GeantV ~\cite{1742-6596-396-2-022014} R\&D project, started in 2013.
 Another example is the ongoing effort to port simulation code to run on
@@
 experimental programme. Properly simulating a neutrino experiment like
-NOvA {[}3{]}, MicroBooNE{[}4{]}, MINERvA{[}5{]}, or DUNE {[}6{]}
+NOvA ~\cite{1742-6596-396-2-022014}, MicroBooNE~\cite{MicroBooNE}, 
+MINERvA~\cite{Aliaga:2013uqz}, or DUNE ~\cite{Acciarri:2015uup}
 involves a three-part software stack, the first and last of which are
-relevant to this paper; the second, event generators such as GENIE
-{[}7{]}, is discussed elsewhere. The first element of the stack concerns
+relevant to this paper; the second, event generators such as GENIE, 
+is discussed elsewhere. The first element of the stack concerns
 the beamline and the neutrino flux prediction. Estimating the neutrino
@@
 distributions must be described in great detail. Geant4 is the most
-commonly used software toolkit for this work, although FLUKA {[}8{]} and
-MARS {[}9{]} are also used. The final element of the stack is the
+commonly used software toolkit for this work, although FLUKA ~\cite{Ferrari:2005zk} 
+and MARS ~\cite{MARS} are also used. The final element of the stack is the
 detector simulation. Again there are numerous challenges in terms of
@@
 manpower available for updating detector simulations to use new tools.
-Efforts such as LArSoft{[}7{]} have recognised this challenge and aim to
-pool resources of many experiments for the benefit of all.
+Efforts such as LArSoft~\cite{LArSoft} have recognised this challenge and 
+aim to pool resources of many experiments for the benefit of all.
 
@@
 the most complex setups and to offer adequate precision and efficiency
-in the implemented algorithms. Geant4 {[}1{]} provides a wide variety of
+in the implemented algorithms. Geant4 provides a wide variety of
 tools and solutions for describing geometry setups from simple to highly
@@
 introduced in Geant4 along with the implementation of the Unified Solids
-library {[}10{]} provide considerably improved CPU performance, making
+library ~\cite{1742-6596-396-5-052035} provide considerably improved CPU performance, making
 it possible to use such constructs for very detailed and realistic
@@
 are expected as part of the rewrite of most geometry primitives going on
-in the VecGeom {[}12{]} package, initiated within the GeantV project
-{[}2{]}.
+in the VecGeom ~\cite{1742-6596-608-1-012023} package, initiated within 
+the GeantV project.
 
@@
 aim to unify and improve the algorithms provided by the Geant4 and ROOT
-{[}11{]} approaches to solids modeling. The library which comes with a
+~\cite{Brun1996} approaches to solids modeling. The library which comes with a
 large collection of primitives, has been extended and enhanced as part
@@
 descriptions from text files according to the Geometry Description
-Markup Language (GDML) {[}13{]} based on XML; the exchange format is
+Markup Language (GDML) ~\cite{1710291} based on XML; the exchange format is
 also supported by ROOT and the schema includes all the geometrical
@@
 file against the latest version of the schema comes for free thanks to
-Xerces-C++ {[}14{]}, with the possibility to turn it on or off in the
+Xerces-C++, with the possibility to turn it on or off in the
 Geant4 GDML parser.
 
-The DD4hep Geometry Toolkit {[}15{]}, being developed in the context of
+The DD4hep Geometry Toolkit ~\cite{1742-6596-513-2-022010}, being developed in the context of
 the AIDA and AIDA-2020 projects aims to provide a generic detector
@@
 detector projects, and Allpix Squared follows a modular approach and is
-thus also suited to simulate silicon-based calorimeters {[}16{]}. The
+thus also suited to simulate silicon-based calorimeters ~\cite{Allpix}. The
 frameworks also allow the implementation of effective radiation damage
@@
 large number of inclusive proton proton collision (minimum bias) events
-with an event generator, such as Pythia8 {[}30{]}, and to overlay the
+with an event generator, such as Pythia8 ~\cite{Sjostrand:2014zea}, and to overlay the
 energy depositions from these events on top of the energy depositions
@@
 particular to model the H $\rightarrow \gamma \gamma $ 
-signal accurately at the
-LHC. This effort is being continued with emphasis on reviewing the
+signal accurately at the LHC. This effort is being continued with 
+emphasis on reviewing the
 models assumptions, approximations and limitations, especially at very
@@
 implementations. For LHC Run-1 simulation Geant4 version 9.4 was used
-{[}17{]}. For Run-2 significant improvements were introduced into the
-bremsstrahlung, pair production and multiple scattering {[}18, 19{]}
+~\cite{1742-6596-331-3-032029}. For Run-2 significant improvements were introduced into the
+bremsstrahlung, pair production and multiple scattering 
+~\cite{1742-6596-396-2-022013} ~\cite{1742-6596-513-2-022015}
 further improving simulation of the EM shower shape; these improvements
@@
 shower shapes. A comprehensive description of recent improvements of the
-Geant4 electromagnetic module can be found in {[}19{]}. Rather good
+Geant4 electromagnetic module can be found in ~\cite{1742-6596-513-2-022015}. Rather good
 agreement was found when Geant4 predictions were compared with
@@
 the time structure and the lateral profile of hadronic showers in
-neutron-rich materials {[}19{]}. Other recent developments include an
+neutron-rich materials ~\cite{WERNER200881}. Other recent developments include an
 improved and re-tuned Fritiof model.
@@
 to simulate very high energy cosmic ray showers in the Earth's
-atmosphere. The EPOS hadronic generator {[}20{]} offers currently the
+atmosphere. The EPOS hadronic generator ~{20{]} offers currently the
 most accurate simulations in this domain, as a fresh C++ rewrite of the
@@
 including the post Geant4 phase of simulation are undertaken in projects
-such as LArSoft{[}21{]} and Qscan{[}22{]}, while extending Geant4 EM
+such as LArSoft~\cite{LArSoft} and Qscan~\cite{LArLEM-TPC}, while extending Geant4 EM
 physics models for liquid noble gases is undertaken by the NEST project
-{[}23{]}. Detectors with large numbers of wires and photo-sensors may
+~\cite{Szydagis:2013sih}. Detectors with large numbers of wires and photo-sensors may
 naturally lend themselves to parallelisation techniques, provided that
 the simulation frameworks are able to handle such cases in an efficient
-way. Additionally, the Wire Cell Toolkit {[}24{]} provides simulation,
+way. Additionally, the Wire Cell Toolkit ~\cite{Wire-Cell} provides simulation,
 noise filtering, signal processing and reconstruction algorithms for
@@
 \item
-  Validation of models: a validation database known as DoSSiER {[}25{]}
+  Validation of models: a validation database known as DoSSiER ~\cite{Wenzel:2017dqo}
   that contains data from experiments measuring particle cross-sections
@@
   but is now also used by GeantV and should be suitable for other MC
-  toolkits such as e.g. GENIE {[}7{]}.
+  toolkits such as e.g. GENIE.
 \item
@@
   Applications include the use of bent crystals for beams manipulations,
-  phonons and ultra-cold neutrons beams {[}26{]}.
+  phonons and ultra-cold neutrons beams ~\cite{Brandt:2014imy}.
 \item
   Geant4-DNA: a Geant4 extension for the calculation of physico-chemical
-  damage to DNA at a nanoscale level {[}27{]}
+  damage to DNA at a nanoscale level ~\cite{bernal:hal-01288764}
 \end{itemize}
@@
 of available cores (57) and still gains about +30\% in hyperthreading
-regime {[}29,30{]}. The large memory reduction obtained with
+regime ~\cite{G4CPUPerf} ~\cite{Dotti:2016ors}. The large memory reduction obtained with
 multi-threading is clearly visible (bottom). While a multi-process
@@
 application and a simple calorimeter configuration made of
-Cu-Scintillator in a 4 Tesla magnetic field {[}29{]}.The study is
+Cu-Scintillator in a 4 Tesla magnetic field ~\cite{G4CPUPerf}.The study is
 performed for 50 GeV e\textsuperscript{-}, p\textsuperscript{-}, and
@@
 shared and static libraries in a simplified CMS application (HepExpMT
-public benchmark) {[}31{]}. In this example, static linkage yields a
+public benchmark) ~\cite{Farrell:2016ovs}. In this example, static linkage yields a
 20\% gain, while Profile-Guided Optimisation (PGO) gives another 20\%.
@@
 Figure 13 shows a chart of CPU time for the G4 module of the CMS Full
-Simulation application {[}32{]}. About 60\% of the time is spent on
+Simulation application ~\cite{Hildreth:2017vpw}. About 60\% of the time is spent on
 functions associated with the propagation of particles through the
@@
 the CMS throughput and memory per event utilisation for QCD and ttbar
-simulated events {[}32{]}. Maximum CPU efficiency is achieved after
+simulated events. Maximum CPU efficiency is achieved after
 simulating about 500 events, when contribution of initialisation of
@@
 particularly to replace Full Simulation in the forward calorimeters.
-GFLASH {[}33{]}, a Geant4 module for the fast simulation of
+GFLASH ~\cite{GRINDHAMMER1990469}, a Geant4 module for the fast simulation of
 electromagnetic and hadronic showers using parameterisations for the
@@
 \item
-  Pythia {[}34{]} events: 13 TeV Pythia minimum bias (300 events) and
-  ttbar (300 events), pseudo rapidity cut
-  \textbar{}$\eta$\textbar{}\textless{}5.5
+  Pythia ~\cite{Sjostrand:2014zea} events: 13 TeV Pythia minimum bias (300 events) and
+  ttbar (300 events), pseudo rapidity cut \textbar{}$\eta$\textbar{}\textless{}5.5
 \item
   Particle gun: 50 GeV e's, muons, pions with a flat $\eta,\varphi$ distribution in
-  $\eta={[}-0.8,0.8{]}, \eta={[}2,2.7{]}, and \varphi={[}0,2\pi{]}$
+  $\eta={[}-0.8,0.8{]}, \eta={[}2,2.7{]}, and \varphi= {[}0,2\pi{]}$
 \item
@@
 several options to reduce CPU time consumption in production of Pythia
-{[}34{]} generated minimum bias and ttbar events. Table 2 lists the
+generated minimum bias and ttbar events. Table 2 lists the
 individual impact of each of the options, while Table 3 shows the
@@
 and decay of hadrons that produce neutrinos. For this step, most
-experiments use Geant4, although some use Fluka {[}8{]} or MARS {[}9{]}.
+experiments use Geant4, although some use Fluka or MARS.
 Neutrino-nucleus interactions inside the detectors are taken care of by
-neutrino generator tools, such as GENIE {[}7{]}. The actual Geant4-based
+neutrino generator tools, such as GENIE. The actual Geant4-based
 detector simulation stage takes over from the generator to propagate the
@@
 For liquid argon TPC detectors, all Fermilab experiments use the LArSoft
-software {[}21{]}, which contains the simulation and reconstruction
+software, which contains the simulation and reconstruction
 packages. LArSoft includes both the Geant4 module and a ``DetSim''
@@
 large amounts of memory will be mitigated by evolving the \emph{art}
-software framework {[}35{]} to support multithreading.
+software framework ~\cite{Green:2012gv} to support multithreading.
 
@@
 & Geant4 with space charge & DetSim & Reco \\ \hline
-ProtoDUNE (Cosmic events with Corsika {[}36{]}) & 10 & 180  & 
+ProtoDUNE (Cosmic events with Corsika ~\cite{Heck:1998vt}) & 10 & 180  & 
 480 & 25 & 1000 \\ \hline
@@
   For the liquid noble gas experiments an extension to the Geant4
-  scintillation process {[}37{]} is usually employed
+  scintillation process ~\cite{1748-0221-6-10-P10002} is usually employed
 \item
@@
   interaction becomes important. Geant4 has been extended to support
-  these processes {[}38{]}
+  these processes ~\cite{Brandt:2014imy}
 \end{itemize}
@@
   experiments with a detailed and complex optical model. Advanced
-  techniques leveraging GPUs are under investigation {[}39{]}.
+  techniques leveraging GPUs are under investigation ~\cite{1742-6596-898-4-042001}.
 \end{itemize}
@@
 significant speed-ups, which are to be measured in a realistic prototype
-currently under development {[}40{]}.
+currently under development ~\cite{1742-6596-608-1-012003}.
 
@@
 magnitude. This is the case for certain solids within the VecGeom
-geometry package {[}41{]}, and is planned to be extended to field
+geometry package ~\cite{1742-6596-608-1-012023}, and is planned to be extended to field
 propagators and certain EM physics models. While this result is very
@@
 
-Previous experience {[}42{]} has shown that large speedup factors can be
+Previous experience ~\cite{refId0} has shown that large speedup factors can be
 achieved for tasks that are well-suited to GPGPU technology: simplified
@@
 to 30 with respect to Geant4 when processing a basket of electrons and
-photons in a GPU {[}43{]}. This is explained by the fact that EM
+photons in a GPU ~\cite{1742-6596-762-1-012019}. This is explained by the fact that EM
 interactions in a detector evolve into a self-contained cascade of
@@
 performed on the MIRA supercomputer at ANL successfully ran a modified
-version of Geant4 spawning more than 3 million threads {[}44{]}. The
+version of Geant4 spawning more than 3 million threads. The
 test concluded that the main bottleneck to achieve scaling linearity is
@@
 YODA have been running successfully in production on supercomputers at
-NERSC and machine utilisation is being studied {[}45{]}. ATLAS is
+NERSC and machine utilisation is being studied ~\cite{1742-6596-664-9-092025}. ATLAS is
 currently migrating its simulation application to multi-threading to
@@
 number of threads on a device with a factor five decrease in the
-per-worker memory consumption relative to multi-process jobs {[}46{]}.
+per-worker memory consumption relative to multi-process jobs ~\cite{Farrell:2242857}.
 Overall throughput performance falls short of traditional Xeon
@@
   be many high-energy electromagnetic particles available to the fast
-  simulation. This simplest approach, based on Ref. {[}34{]}, was
+  simulation. This simplest approach, based on Ref. ~\cite{Sjostrand:2014zea}, was
   abandoned by ATLAS about ten years ago. Current approaches used by
@@
   current default simulation, CMS implemented shower parameterisations
-  using GFLASH {[}33{]} to model the shower after the first interaction
+  using GFLASH to model the shower after the first interaction
   of each primary particle with matter, by tuning a set of parameters to
@@
   ATLAS has been made public as a part of the common tracking software
-  project ACTS {[}47{]}, adopted by the FCC {[}48{]}. These fast track
+  project ACTS ~\cite{ACTS}, adopted by the FCC. These fast track
   simulations offer speed improvements of 10-1000x over a detailed
@@
   reconstruction software. The speed improvements from these algorithms
-  varies with pileup but are generally more than one order of magnitude.
-  {[}49{]}
+  varies with pileup but are generally more than one order of magnitude
+  ~\cite{1742-6596-523-1-012035}.
 \item
@@
   in the public domain for the implementation of very fast parameterised
-  simulation, the most popular being DELPHES and PGS {[}50,51{]}. These
+  simulation, the most popular being DELPHES ~\cite{deFavereau:2013fsa}
+  and PGS ~\cite{PGS4}. These
   tools are extremely fast, running at hundreds of events per second,
@@
   are formed directly from fully simulated events. A past example was
-  Turbosim {[}52{]} developed at the Tevatron for the D0 and CDF
-  experiments. A prototype for a redesign called Falcon {[}53{]} was
+  Turbosim ~\cite{Knuteson:2004nj} developed at the Tevatron for the D0 and CDF
+  experiments. A prototype for a redesign called Falcon ~\cite{Brooijmans:2016vro} was
   recently released, with a more efficient algorithm for the
@@
 learning in Fast Simulation, most of which has focused on the use of
-multi-objective regression {[}54{]} and generative adversarial networks
-(GANs) {[}55,56,57{]}. Since use of GANs allows for non-parametric
+multi-objective regression ~\cite{2007physics...3039H} and generative adversarial networks
+(GANs) ~\cite{2014arXiv1406.2661G},~\cite{deOliveira:2017pjk},
+~\cite{Paganini:2017hrr}. Since use of GANs allows for non-parametric
 learning in cases such as calorimetric shower fluctuations, it is a
@@
 appropriate in an inter-experiment ``forum'' like the recently developed
-Inter-experiment Machine Learning Working Group at CERN {[}58{]}. The
+Inter-experiment Machine Learning Working Group at CERN ~\cite{IML}. The
 sharing of code and other technical details, in addition to ideas and
@@
 field is difficult. The fact that machine learning tools like
-SciKitLearn {[}59{]} and TensorFlow {[}60{]} are being developed by
+SciKitLearn ~\cite{DBLP:journals/corr/abs-1201-0490} and 
+TensorFlow ~\cite{DBLP:journals/corr/AbadiABBCCCDDDG16} are being developed by
 professionals and in an open-source community means that if the
@@
 The first neural network-based fast simulations are now publicly
-available, along with code, documentation, and training data
-{[}56,57{]}. These first results are very promising and establish a
+available, along with code, documentation, and training data. 
+These first results are very promising and establish a
 baseline for continuing to improve the performance in preparation for
@@
 use is similar to the research on parallel PRNG for use in other fields
-{[}61{]}. A recent survey {[}62{]} covers developments for parallel
-PRNGs, including ones for GPUs.
+~\cite{PhysRevE.75.066701}. A recent survey ~\cite{lecuyer:hal-01399537} covers 
+developments for parallel PRNGs, including ones for GPUs.
 
@@
 suitable PRNG classes have been identified, including counter-based
-PRNGs, such as Random123 {[}63{]}, Combined Multiple Recursive Random
-number Generators (CMRGs) {[}64{]} and the newest extension of the
-MIXMAX PRNGs {[}65{]} with small matrix size.
+PRNGs, such as Random123 ~\cite{Salmon:2011:PRN:2063384.2063405}, 
+Combined Multiple Recursive Random number Generators (CMRGs) 
+~\cite{doi:10.1287/opre.47.1.159} and the newest extension of the
+MIXMAX PRNGs ~\cite{Savvidy:2015jva} with small matrix size.
 
@@
   Collaborate with the authors of the state of the art PRNG testing
-  suites, and in particular TestU01{[}66{]}, encouraging and seeking to
-  contribute to the extension to increase the statistics of testing,
-  extend testing to 64-bit variates, and expand the testing of
-  correlations between sub-sequences.
+  suites, and in particular TestU01~\cite{Ecuyer:2007:TCL:1268776.1268777}, 
+  encouraging and seeking to contribute to the extension to increase 
+  the statistics of testing, extend testing to 64-bit variates, 
+  and expand the testing of correlations between sub-sequences.
 \item
@@
 
+\sloppy
+\raggedright
+\clearpage
+\printbibliography[title={References},heading=bibintoc]
+
 \end{document}
@@
 
-{[}1{]} J.Allison et al., NIM A 835 (2016), 186-225
-
-{[}2{]} J.Apostolakis et al., J. Phys.: Conf. Ser., 396 (2012) 022014
-
-{[}3{]} P. Adamson et al. {[}NOvA Collaboration{]}, ``First measurement
+~\cite{1{]} J.Allison et al., NIM A 835 (2016), 186-225
+
+~\cite{2{]} J.Apostolakis et al., J. Phys.: Conf. Ser., 396 (2012) 022014
+
+~\cite{3{]} P. Adamson et al. ~\cite{NOvA Collaboration{]}, ``First measurement
 of electron neutrino appearance in NOvA'', Phys. Rev. Lett. 116 (2016)
@@
 
-{[}4{]} The MicroBooNE Collaboration, ``The MicroBooNE Technical Design
+~\cite{4{]} The MicroBooNE Collaboration, ``The MicroBooNE Technical Design
 Report'',
@@
 
-{[}5{]} L. Aliaga, et al. {[}MINERvA Collaboration{]}, ``Design,
+~\cite{5{]} L. Aliaga, et al. ~\cite{MINERvA Collaboration{]}, ``Design,
 Calibration, and Performance of the MINERvA Detector''
@@
 
-{[}6{]} R. Acciarri et al. {[}DUNE Collaboration{]}, ``Long-Baseline
+~\cite{6{]} R. Acciarri et al. ~\cite{DUNE Collaboration{]}, ``Long-Baseline
 Neutrino Facility (LBNF) and Deep Underground Neutrino Experiment (DUNE)
@@
 
-{[}7{]} C.Andreopoulos et al., The GENIE Neutrino Monte Carlo Generator,
+~\cite{7{]} C.Andreopoulos et al., The GENIE Neutrino Monte Carlo Generator,
 Nucl.Instrum.Meth.A614 (2010) 87-104.
@@
 
-{[}8{]} A. Ferrari, P. Sala, A. Fasso, J. Ranft, FLUKA: a multi-particle
+~\cite{8{]} A. Ferrari, P. Sala, A. Fasso, J. Ranft, FLUKA: a multi-particle
 transport code, CERN-2005-10 (2005), INFN/TC\_05/11, SLAC-R-773.
 
-{[}9{]} The MARS Code System,
+~\cite{9{]} The MARS Code System,
 \href{http://mars.fnal.gov}{{http://mars.fnal.gov}}
 
-{[}10{]} M.Gayer et al., New software library of geometrical primitives
+~\cite{10{]} M.Gayer et al., New software library of geometrical primitives
 for modeling of solids used in Monte Carlo detector simulations, J.
@@
 
-{[}11{]} R.Brun et al. NIM A 389 (1997) 81 -- 86
-
-{[}12{]} J.Apostolakis et al., Towards a high performance geometry
+~\cite{11{]} R.Brun et al. NIM A 389 (1997) 81 -- 86
+
+~\cite{12{]} J.Apostolakis et al., Towards a high performance geometry
 library for particle-detector simulations, J. Phys.: Conf. Ser., 608
@@
 
-{[}13{]} R.Chytracek et al.,
+~\cite{13{]} R.Chytracek et al.,
 \href{http://ieeexplore.ieee.org/document/1710291/}{Geometry Description
@@
 
-{[}14{]} Apache Xerces C Parser,
-
-\href{http://xerces.apache.org/xerces-c/}{{http://xerces.apache.org/xerces-c/}}{[}15{]}
+~\cite{14{]} Apache Xerces C Parser,
+
+\href{http://xerces.apache.org/xerces-c/}{{http://xerces.apache.org/xerces-c/}}~\cite{15{]}
 M.Frank et al, DD4hep: A Detector Description Toolkit for High Energy
@@
 
-{[}16{]} M.Benoit, J.Idarraga, The AllPix Simulation Framework, website:
+~\cite{16{]} M.Benoit, J.Idarraga, The AllPix Simulation Framework, website:
 \href{https://twiki.cern.ch/twiki/bin/view/Main/AllPix}{{https://twiki.cern.ch/twiki/bin/view/Main/AllPix}},
@@
 
-{[}17{]} A. Schaeliske et al, J.Physics : Conf. Ser., 331 (2011) 032029
-
-{[}18{]} J. Allison et al., J.Physics: Conf. Ser. 396 (2012) 022013
-
-{[}19{]} V.N. Ivanchenko, J. Physics: Conf. Ser. 513 (2014) 0022015
-
-{[}20{]} K.Werner, The hadronic interaction model EPOS,
+~\cite{17{]} A. Schaeliske et al, J.Physics : Conf. Ser., 331 (2011) 032029
+
+~\cite{18{]} J. Allison et al., J.Physics: Conf. Ser. 396 (2012) 022013
+
+~\cite{19{]} V.N. Ivanchenko, J. Physics: Conf. Ser. 513 (2014) 0022015
+
+~\cite{20{]} K.Werner, The hadronic interaction model EPOS,
 \href{http://www.sciencedirect.com/science/journal/09205632}{Nuclear
@@
 
-{[}21{]} LArSoft: The Liquid Argon Software Collaboration
+~\cite{21{]} LArSoft: The Liquid Argon Software Collaboration
 \href{http://larsoft.org}{{larsoft.org}}.
 
-{[}22{]} D. Lussi, Study of the response of the novel LAr LEM-TPC
+~\cite{22{]} D. Lussi, Study of the response of the novel LAr LEM-TPC
 detector exposed to cosmic rays and a charged particle beam. PhD thesis,
@@
 
-{[}23{]} M Szydagis, A Fyhrie, D Thorngren and M Tripathi, "Enhancement
+~\cite{23{]} M Szydagis, A Fyhrie, D Thorngren and M Tripathi, "Enhancement
 of NEST capabilities for simulating low-energy recoils in liquid xenon"
@@
 
-{[}24{]} Wire-Cell, A 3D Reconstruction Package for LArTPC.
+~\cite{24{]} Wire-Cell, A 3D Reconstruction Package for LArTPC.
 http://lar.bnl.gov/wire-cell.
 
-{[}25{]}
+~\cite{25{]}
 \href{https://inspirehep.net/record/1500313}{{https://inspirehep.net/record/1500313}};
@@
 
-{[}26{]} Baglie, E. et al.,A model for the interaction of high-energy
+~\cite{26{]} Baglie, E. et al.,A model for the interaction of high-energy
 particles in straight and bent crystals implemented in Geant4,
@@
 
-{[}27{]} Bernal, M.A. et al., Track structure modeling in liquid water:
+~\cite{27{]} Bernal, M.A. et al., Track structure modeling in liquid water:
 A review of the Geant4-DNA very low energy extension of the Geant4 Monte
@@
 
-{[}29{]}
+~\cite{29{]}
 \href{https://g4cpt.fnal.gov/g4p/prplots/cpu_by_version.html}{{Source}}:
@@
 
-{[}30{]} Dotti, A. et al "Extending Geant4 Parallelism with external
+~\cite{30{]} Dotti, A. et al "Extending Geant4 Parallelism with external
 libraries (MPI, TBB) and its use on HPC resources," 2015 IEEE Nuclear
@@
 
-{[}31{]} Farrell, S. et al "Multi-threaded Geant4 on the Xeon-Phi with
+~\cite{31{]} Farrell, S. et al "Multi-threaded Geant4 on the Xeon-Phi with
 complex high-energy physics geometry," 2015 IEEE Nuclear Science
@@
 
-{[}32{]} M. Hildreth, V. N. Ivanchenko and D. J. Lange, ``Upgrades for
+~\cite{32{]} M. Hildreth, V. N. Ivanchenko and D. J. Lange, ``Upgrades for
 the CMS simulation'', CHEP 2016.
 
-{[}33{]} G. Grindhammer, et al., The Fast Simulation of Electromagnetic
+~\cite{33{]} G. Grindhammer, et al., The Fast Simulation of Electromagnetic
 and Hadronic Showers, Nucl. Instrum. Meth. A 290 (1990) 469.
 
-{[}34{]} Torbjorn Sjostrand et al., An Introduction to PYTHIA 8.2,
-arXiv:1410.3012 {[}hep-ph{]}doi:10.1016/j.cpc.2015.01.024; T. Sjostrand,
+~\cite{34{]} Torbjorn Sjostrand et al., An Introduction to PYTHIA 8.2,
+arXiv:1410.3012 ~\cite{hep-ph{]}doi:10.1016/j.cpc.2015.01.024; T. Sjostrand,
 S. Mrenna, P. Skands, PYTHIA 6.4 Physics and Manual, JHEP05 (2006)
@@
 
-{[}35{]} C. Green et. al., The art event processing framework,
+~\cite{35{]} C. Green et. al., The art event processing framework,
 \href{https://inspirehep.net/record/1211020}{J.Phys.Conf.Ser. 396 (2012)
@@
 
-{[}36{]} D. Heck, G. Schatz, T. Thouw, J. Knapp, J. N. Capdevielle,
+~\cite{36{]} D. Heck, G. Schatz, T. Thouw, J. Knapp, J. N. Capdevielle,
 CORSIKA: A Monte Carlo code to simulate extensive air showers, Tech.
@@
 
-{[}37{]} Szydagis et al., NEST: A Comprehensive Model for Scintillation
+~\cite{37{]} Szydagis et al., NEST: A Comprehensive Model for Scintillation
 Yield in Liquid Xenon,
@@
 \href{http://arxiv.org/abs/1106.1613}{{arXiv:1106.1613
-{[}physics.ins-det{]}}}
-
-{[}38{]} Brandt, D. et al., Semiconductor phonon and charge transport
+~\cite{physics.ins-det{]}}}
+
+~\cite{38{]} Brandt, D. et al., Semiconductor phonon and charge transport
 Monte Carlo simulation using Geant4,
@@
 
-{[}39{]} Blyth, S. et al., Opticks : GPU Optical Photon Simulation for
+~\cite{39{]} Blyth, S. et al., Opticks : GPU Optical Photon Simulation for
 Particle Physics with NVIDIA OptiX, PSROC Annual Meeting, Tamkang
@@
 
-{[}40{]} J. Apostolakis et al., ``Adaptive track scheduling to optimise
+~\cite{40{]} J. Apostolakis et al., ``Adaptive track scheduling to optimise
 concurrency and vectorisation in GeantV'', Journal of Physics:
@@
 
-{[}41{]} J. Apostolakis et al, ``Towards a high performance geometry
+~\cite{41{]} J. Apostolakis et al, ``Towards a high performance geometry
 library for particle-detector simulations'', Journal of Physics:
@@
 
-{[}42{]} Henderson N. et al.; A CUDA Monte Carlo simulator for radiation
+~\cite{42{]} Henderson N. et al.; A CUDA Monte Carlo simulator for radiation
 therapy dosimetry based on Geant4; Joint International Conference on
@@
 
-{[}43{]} Amadio, G. et al.: GeantV: from CPU to accelerators; Journal of
+~\cite{43{]} Amadio, G. et al.: GeantV: from CPU to accelerators; Journal of
 Physics: Conference Series 762 (2016)
@@
 
-{[}44{]} LeCompte, T. et al.; Geant4 at extreme scales, in preparation,
+~\cite{44{]} LeCompte, T. et al.; Geant4 at extreme scales, in preparation,
 cfr \href{https://goo.gl/cgA92Y}{{https://goo.gl/cgA92Y}}
 
-{[}45{]} Calafiura P et al. on behalf of the ATLAS Collaboration (2015)
+~\cite{45{]} Calafiura P et al. on behalf of the ATLAS Collaboration (2015)
 "Fine grained event processing on HPCs with the ATLAS Yoda system", J.
@@
 
-{[}46{]} S. Farrell, et al., "Multi-threaded ATLAS Simulation on Intel
+~\cite{46{]} S. Farrell, et al., "Multi-threaded ATLAS Simulation on Intel
 Knights Landing Processors", CHEP 2016,
@@
 
-{[}47{]} A Common Tracking Software: http://acts.web.cern.ch/
-
-{[}48{]} V. Volkl, ``Applications of A-Common-Tracking-Software (ACTS)
+~\cite{47{]} A Common Tracking Software: http://acts.web.cern.ch/
+
+~\cite{48{]} V. Volkl, ``Applications of A-Common-Tracking-Software (ACTS)
 in the Future Circular Collider Design Study.'' Submitted to ACAT 2017:
@@
 
-{[}49{]} E. Ritsch for the ATLAS Collaboration, ``Concepts and plans
+~\cite{49{]} E. Ritsch for the ATLAS Collaboration, ``Concepts and plans
 towards fast large-scale Monte Carlo production for the ATLAS
@@
 
-{[}50{]} The DELPHES 3 Collaboration, ``DELPHES 3: A modular framework
+~\cite{50{]} The DELPHES 3 Collaboration, ``DELPHES 3: A modular framework
 for fast simulation of a generic collider experiment.'' JHEP 02 (2014)
@@
 
-{[}51{]} PGS-4,
+~\cite{51{]} PGS-4,
 \href{http://www.physics.ucdavis.edu:80/~conway/research/software/pgs/pgs4-general.htm}{{http://www.physics.ucdavis.edu:80/\textasciitilde{}conway/research/software/pgs/pgs4-general.htm}}
 
-{[}52{]} B. Knuteson, ``Systematic Analysis of High Energy Collider
-Data'', Nucl. Instrum. Meth. A534 (2004) 7--14, {[}hep-ex/0402029{]}
-
-{[}53{]} H. B. Prosper et. al. ``Falcon: towards an ultra fast
+~\cite{52{]} B. Knuteson, ``Systematic Analysis of High Energy Collider
+Data'', Nucl. Instrum. Meth. A534 (2004) 7--14, ~\cite{hep-ex/0402029{]}
+
+~\cite{53{]} H. B. Prosper et. al. ``Falcon: towards an ultra fast
 non-parametric detector simulator'', contribution to Les Houches 2015:
 Physics at TeV Colliders New Physics Working Group Report
-{[}arXiv:1605.02684{]}.
-
-{[}54{]} A. Hoecker et al., TMVA - Toolkit for Multivariate Data
+~\cite{arXiv:1605.02684{]}.
+
+~\cite{54{]} A. Hoecker et al., TMVA - Toolkit for Multivariate Data
 Analysis, PoS ACAT 040 (2007), arXiv:physics/0703039
 
-{[}55{]} I. J. Goodfellow et al., Generative Adversarial Networks,
+~\cite{55{]} I. J. Goodfellow et al., Generative Adversarial Networks,
 arXiv:1406.2661 (2014).
 
-{[}56{]} L. Oliveira, M. Paganini, B. Nachman, Learning Particle Physics
+~\cite{56{]} L. Oliveira, M. Paganini, B. Nachman, Learning Particle Physics
 by Example: Location-Aware Generative Adversarial Networks for Physics
@@
 
-{[}57{]} M. Paganini, L. Oliveira, B. Nachman, CaloGAN: Simulating 3D
+~\cite{57{]} M. Paganini, L. Oliveira, B. Nachman, CaloGAN: Simulating 3D
 High Energy Particle Showers in Multi-Layer Electromagnetic Calorimeters
@@
 
-{[}58{]} Inter-Experimental LHC Machine Learning Working Group :
+~\cite{58{]} Inter-Experimental LHC Machine Learning Working Group :
 \href{https://iml.cern.ch}{{https://iml.cern.ch}}
 
-{[}59{]} F. Pedregosa et al., Scikit-learn: Machine learning in Python,
-J.Machine Learning Res. 12 (2011) 2825-2830 {[}arXiv:1201.0490{]}
-
-{[}60{]} M. Abudi et al., ``Tensorflow: A system for large-scale machine
-learning'', in OSDI, vol. 16,
-
-pp. 265--283, 2016
-
-{[}61{]} H. Bauke and S. Mertens, ``Random numbers for large-scale
+~\cite{59{]} F. Pedregosa et al., Scikit-learn: Machine learning in Python,
+J.Machine Learning Res. 12 (2011) 2825-2830 {{}arXiv:1201.0490{]}
+
+~\cite{60{]} M. Abudi et al., ``Tensorflow: A system for large-scale machine
+learning'', in OSDI, vol. 16, pp. 265--283, 2016
+
+~\cite{61{]} H. Bauke and S. Mertens, ``Random numbers for large-scale
 distributed Monte Carlo simulations,'' Phys. Rev. E, vol. 75, no. 6, p.
@@
 
-{[}62{]} P. L'Ecuyer, D. Munger, B. Oreshkin, and R. Simard, ``Random
+~\cite{62{]} P. L'Ecuyer, D. Munger, B. Oreshkin, and R. Simard, ``Random
 numbers for parallel computers: Requirements and methods, with emphasis
@@
 
-{[}63{]} J. K. Salmon, M. A. Moraes, R. O. Dror, and D. E. Shaw,
+~\cite{63{]} J. K. Salmon, M. A. Moraes, R. O. Dror, and D. E. Shaw,
 ``Parallel random numbers: As easy as 1, 2, 3,'' in Proceedings of 2011
@@
 
-{[}64{]} P. L'Ecuyer, ``Good Parameters and Implementations for Combined
+~\cite{64{]} P. L'Ecuyer, ``Good Parameters and Implementations for Combined
 Multiple Recursive Random Number Generators,'' Operations Research, vol.
@@
 
-{[}65{]} K. Savvidy and G. Savvidy, ``Spectrum and entropy of C-systems
+~\cite{65{]} K. Savvidy and G. Savvidy, ``Spectrum and entropy of C-systems
 MIXMAX random number generator,'' Chaos, Solitons \& Fractals, vol. 91,
@@
 
-{[}66{]} P. L'Ecuyer and R. Simard, ``TestU01,'' ACM Transactions on
+~\cite{66{]} P. L'Ecuyer and R. Simard, ``TestU01,'' ACM Transactions on
 Mathematical Software, vol. 33, no. 4, p. 22--es, Aug. 2007.

### 2018-02-09 CWP/papers/HSF-CWP-2017-01_roadmap/latex/authors.tex
Update for arxiv v3 (#46)
* Updated author file

Add new author (Sunander) and correct place where footnote was for a few
of the Fermilab authors
Add a little more debug information when script has a problem
Regenerated tex file copied to CWP latex directory

* Update CMS figures

Correct error where MHS06s should be THS06s
Use PDFs instead of PNGs for improved quality

@@
 Amadio, Guilherme$^{5}$;
-Anh-Ky, Nguyen$^{117}$;
+Anh-Ky, Nguyen$^{119}$;
 Aphecetche, Laurent$^{63}$;
@@
 Babik, Marian$^{5}$;
-Bagliesi, Giuseppe$^{31}$;
+Bagliesi, Giuseppe$^{30}$;
 Bandieramonte, Marilena$^{5}$;
+Banerjee, Sunanda$^{15,c}$;
 Barisits, Martin$^{5}$;
-Bauerdick, Lothar A. T.$^{16,c}$;
-Belforte, Stefano$^{34}$;
+Bauerdick, Lothar A. T.$^{15,c}$;
+Belforte, Stefano$^{33}$;
 Benjamin, Douglas$^{80}$;
@@
 Bloom, Kenneth$^{94}$;
-Boccali, Tommaso$^{31}$;
+Boccali, Tommaso$^{30}$;
 Bockelman, Brian$^{94}$;
 Bold, Tomasz$^{41}$;
-Bonacorsi, Daniele$^{25}$;
+Bonacorsi, Daniele$^{24}$;
 Boveia, Antonio$^{98}$;
-Bozzi, Concezio$^{27}$;
+Bozzi, Concezio$^{26}$;
 Bracko, Marko$^{91,39}$;
@@
 Campana, Simone$^{5,a}$;
-Canal, Philippe$^{16,c}$;
+Canal, Philippe$^{15,c}$;
 Canali, Luca$^{5}$;
-Carlino, Gianpaolo$^{30}$;
+Carlino, Gianpaolo$^{29}$;
 Castro, Nuno$^{45,93,d}$;
@@
 Cerminara, Gianluca$^{5}$;
+Cervantes Villanueva, Javier$^{5}$;
 Chang, Philip$^{73}$;
 Chapman, John$^{74}$;
-Chen, Gang$^{23}$;
+Chen, Gang$^{22}$;
 Childers, Taylor$^{1}$;
@@
 Cogneras, Eric$^{48}$;
+Coles, Jeremy$^{74}$;
 Collier, Ian$^{59}$;
@@
 Cosmo, Gabriele$^{5}$;
-Costanzo, Davide$^{108}$;
+Costanzo, Davide$^{109}$;
 Couturier, Ben$^{5}$;
@@
 Cranshaw, Jack$^{1}$;
-Cristella, Leonardo$^{26}$;
+Cristella, Leonardo$^{25}$;
 Crooks, David$^{84}$;
@@
 Dallmeier-Tiessen, Sünje$^{5}$;
-De, Kaushik$^{110}$;
+De, Kaushik$^{111}$;
 De Cian, Michel$^{85}$;
+De Roeck, Albert$^{5}$;
+Delgado Peris, Antonio$^{6,g}$;
 Di Girolamo, Alessandro$^{5}$;
+Di Guida, Salvatore$^{28}$;
 Dimitrov, Gancho$^{5}$;
@@
 Duflot, Laurent$^{43}$;
-Dykstra, Dave$^{16,c}$;
+Dykstra, Dave$^{15,c}$;
 Dziedziniewicz-Wojcik, Katarzyna$^{5}$;
@@
 Elmsheuser, Johannes$^{2}$;
-Elvira, V. Daniel$^{16,c}$;
+Elvira, V. Daniel$^{15,c}$;
 Eulisse, Giulio$^{5}$;
+Farrell, Steven$^{44}$;
 Ferber, Torben$^{71}$;
@@
 Fisk, Ian$^{62}$;
-Fitzpatrick, Conor$^{14}$;
-Flix, José$^{57,7,g}$;
+Fitzpatrick, Conor$^{13}$;
+Flix, José$^{57,6,g}$;
 Formica, Andrea$^{37}$;
 Forti, Alessandra$^{90}$;
+Franzoni, Giovanni$^{5}$;
 Frost, James$^{101}$;
-Gaede, Frank$^{13}$;
+Fuess, Stu$^{15,c}$;
+Gaede, Frank$^{12}$;
 Ganis, Gerardo$^{5}$;
@@
 Garonne, Vincent$^{99}$;
-Gellrich, Andreas$^{13}$;
-Genser, Krzysztof$^{16,c}$;
+Gellrich, Andreas$^{12}$;
+Genser, Krzysztof$^{15,c}$;
 George, Simon$^{60}$;
@@
 Gheata, Mihaela$^{5}$;
-Giacomini, Francesco$^{9}$;
-Giagu, Stefano$^{106,33}$;
+Giacomini, Francesco$^{8}$;
+Giagu, Stefano$^{106,32}$;
 Giffels, Manuel$^{40}$;
@@
 Gligorov, Vladimir V.$^{49}$;
-Glushkov, Ivan$^{110}$;
+Glushkov, Ivan$^{111}$;
 Gohn, Wesley$^{86}$;
@@
 González Fernández, Juan R.$^{100}$;
-Govi, Giacomo$^{16}$;
-Grandi, Claudio$^{25}$;
+Govi, Giacomo$^{15}$;
+Grandi, Claudio$^{24}$;
 Grasland, Hadrien$^{43}$;
@@
 Grillo, Lucia$^{90}$;
-Guan, Wen$^{115}$;
-Gutsche, Oliver$^{16,c}$;
+Guan, Wen$^{116}$;
+Gutsche, Oliver$^{15,c}$;
 Gyurjyan, Vardan$^{38}$;
@@
 Hariri, Farah$^{5}$;
-Hartmann, Thomas$^{13}$;
+Hartmann, Thomas$^{12}$;
 Harvey, John$^{5,a}$;
@@
 Hegner, Benedikt$^{5,a}$;
-Heinemann, Beate$^{13}$;
+Heinemann, Beate$^{12}$;
 Heinrich, Lukas$^{55}$;
-Hernández, José M.$^{7,g}$;
+Hernández, José M.$^{6,g}$;
 Hildreth, Michael$^{96,f}$;
-Hodgkinson, Mark$^{108}$;
+Hodgkinson, Mark$^{109}$;
 Hoeche, Stefan$^{61,p}$;
+Holzman, Burt$^{15,c}$;
 Hristov, Peter$^{5}$;
-Huang, Xingtao$^{107}$;
-Ivanchenko, Vladimir N.$^{5,111}$;
-Ivanov, Todor$^{109}$;
+Huang, Xingtao$^{108}$;
+Ivanchenko, Vladimir N.$^{5,112}$;
+Ivanov, Todor$^{110}$;
 Jashal, Brij$^{65}$;
-Jayatilaka, Bodhitha$^{16,c}$;
+Jayatilaka, Bodhitha$^{15,c}$;
 Jones, Roger$^{87,a}$;
 Jouvin, Michel$^{43,a}$;
-Jun, Soon Yung$^{16,c}$;
+Jun, Soon Yung$^{15,c}$;
 Kagan, Michael$^{61,p}$;
@@
 Katz, Daniel S.$^{77}$;
-Kcira, Dorian$^{11}$;
+Kcira, Dorian$^{10}$;
 Kersevan, Borut Paul$^{88}$;
-Kirby, Michael$^{16,c}$;
+Kirby, Michael$^{15,c}$;
 Klimentov, Alexei$^{2}$;
 Klute, Markus$^{51}$;
-Komarov, Ilya$^{34,n}$;
+Komarov, Ilya$^{33,n}$;
 Konstantinov, Dmitri$^{58}$;
 Koppenburg, Patrick$^{54}$;
-Kowalkowski, Jim$^{16,c}$;
+Kowalkowski, Jim$^{15,c}$;
 Kreczko, Luke$^{70}$;
 Kuhr, Thomas$^{47,a}$;
-Kutschke, Robert$^{16,a,c}$;
-Kuznetsov, Valentin$^{12}$;
+Kutschke, Robert$^{15,a,c}$;
+Kuznetsov, Valentin$^{11}$;
 Lampl, Walter$^{69}$;
@@
 Letts, James$^{73}$;
-Lewendel, Birgit$^{13}$;
+Lewendel, Birgit$^{12}$;
 Li, Teng$^{81}$;
-Lima, Guilherme$^{16}$;
+Lima, Guilherme$^{15,c}$;
 Linacre, Jacob$^{59,m}$;
-Linden, Tomas$^{18}$;
+Linden, Tomas$^{17}$;
 Lo Presti, Giuseppe$^{5}$;
@@
 Love, Peter$^{87}$;
-Magini, Nicolò$^{28}$;
+Lyon, Adam$^{15,c}$;
+Magini, Nicolò$^{27}$;
 Marshall, Zachary L$^{44}$;
@@
 Meinhard, Helge$^{5}$;
-Menasce, Dario$^{29,a}$;
+Menasce, Dario$^{28,a}$;
 Mendez Lorenzo, Patricia$^{5}$;
 Mete, Alaettin Serhan$^{72}$;
-Michelotto, Michele$^{32}$;
+Michelotto, Michele$^{31}$;
 Mitrevski, Jovan$^{47}$;
 Moneta, Lorenzo$^{5}$;
-Morgan, Ben$^{114}$;
+Morgan, Ben$^{115}$;
 Mount, Richard$^{61,p}$;
 Moyse, Edward$^{68}$;
-Murray, Sean$^{75,10}$;
+Murray, Sean$^{75,9}$;
 Neubauer, Mark S$^{77,a,k}$;
+Norman, Andrew$^{15,c}$;
 Novaes, Sérgio$^{105}$;
 Novak, Mihaly$^{5}$;
-Oyanguren, Arantza$^{22}$;
-Ozturk, Nurcan$^{110}$;
-Pacheco Pages, Andres$^{57,20,j}$;
-Paganini, Michela$^{116}$;
+Oyanguren, Arantza$^{21}$;
+Ozturk, Nurcan$^{111}$;
+Pacheco Pages, Andres$^{57,19,j}$;
+Paganini, Michela$^{117}$;
 Pansanel, Jerome$^{35}$;
-Pascuzzi, Vincent R.$^{112}$;
+Pascuzzi, Vincent R.$^{113}$;
 Patrick, Glenn$^{59}$;
@@
 Pearson, Ben$^{52}$;
-Pedro, Kevin$^{16,c}$;
-Perdue, Gabriel$^{16}$;
-Perez-Calero Yzquierdo, Antonio$^{57,7,g}$;
-Perrozzi, Luca$^{15}$;
+Pedro, Kevin$^{15,c}$;
+Perdue, Gabriel$^{15,c}$;
+Perez-Calero Yzquierdo, Antonio$^{57,6,g}$;
+Perrozzi, Luca$^{14}$;
 Petersen, Troels$^{53}$;
 Petric, Marko$^{5}$;
-Piedra, Jónatan$^{21}$;
-Piilonen, Leo$^{118,i}$;
+Piedra, Jónatan$^{20}$;
+Piilonen, Leo$^{120,i}$;
 Piparo, Danilo$^{5}$;
@@
 Polci, Francesco$^{49}$;
-Potamianos, Karolos$^{13}$;
-Psihas, Fernanda$^{24}$;
+Potamianos, Karolos$^{12}$;
+Psihas, Fernanda$^{23}$;
+Puig Navarro, Albert$^{118}$;
 Raven, Gerhard$^{54}$;
-Reuter, Jürgen$^{13}$;
+Reuter, Jürgen$^{12}$;
 Ribon, Alberto$^{5}$;
-Rinaldi, Lorenzo$^{25}$;
+Rinaldi, Lorenzo$^{24}$;
 Ritter, Martin$^{47}$;
-Robinson, James$^{13}$;
+Robinson, James$^{12}$;
 Rodrigues, Eduardo$^{79,a,e}$;
@@
 Roy, Gareth$^{84}$;
+Rybkine, Grigori$^{43}$;
 Sailer, Andre$^{5}$;
@@
 Severini, Horst$^{97}$;
-Sexton-Kennedy, Elizabeth$^{16,a,c}$;
+Sexton-Kennedy, Elizabeth$^{15,a,c}$;
 Seymour, Michael$^{90}$;
+Sgalaberna, Davide$^{5}$;
 Shapoval, Illya$^{44}$;
@@
 Short, Hannah$^{5}$;
-Siroli, Gian Piero$^{25}$;
+Siroli, Gian Piero$^{24}$;
 Skipsey, Sam$^{84}$;
@@
 Sokoloff, Michael D$^{79,a}$;
-Stadie, Hartmut$^{17}$;
-Stark, Giordon$^{6}$;
+Spentzouris, Panagiotis$^{15,c}$;
+Stadie, Hartmut$^{16}$;
+Stark, Giordon$^{78}$;
 Stewart, Gordon$^{84}$;
 Stewart, Graeme$^{5,a}$;
-Sánchez, Arturo$^{113,19}$;
-Sánchez-Hernández, Alberto$^{8,o}$;
+Sánchez, Arturo$^{114,18}$;
+Sánchez-Hernández, Alberto$^{7,o}$;
 Taffard, Anyes$^{72}$;
+Tamponi, Umberto$^{34}$;
 Templon, Jeff$^{54}$;
@@
 Tsulaia, Vakhtang$^{44}$;
-Tunnell, Christopher$^{6}$;
-Vaandering, Eric$^{16,c}$;
+Tunnell, Christopher$^{78}$;
+Vaandering, Eric$^{15,c}$;
 Valassi, Andrea$^{5}$;
@@
 Viren, Brett$^{2}$;
-Vlimant, Jean-Roch$^{11,a}$;
-Voss, Christian$^{13}$;
-Vuosalo, Carl$^{115}$;
+Vlimant, Jean-Roch$^{10,a}$;
+Voss, Christian$^{12}$;
+Votava, Margaret$^{15,c}$;
+Vuosalo, Carl$^{116}$;
 Vázquez Sierra, Carlos$^{54}$;
 Wartel, Romain$^{5}$;
+Watts, Gordon T$^{107}$;
 Wenaus, Torre$^{2}$;
@@
 Winklmeier, Frank$^{56}$;
-Wissing, Christoph$^{13}$;
+Wissing, Christoph$^{12}$;
 Wuerthwein, Frank$^{73}$;
 Wynne, Benjamin$^{81}$;
-Xiaomei, Zhang$^{23}$;
+Xiaomei, Zhang$^{22}$;
 Yang, Wei$^{61,p}$;
-Yazgan, Efe$^{23}$
+Yazgan, Efe$^{22}$
 \bigskip
-\par {\footnotesize $^{1}$ High Energy Physics Division, Argonne National Laboratory, Argonne IL, USA}
-\par {\footnotesize $^{2}$ Physics Department, Brookhaven National Laboratory, Upton NY, USA}
+\par {\footnotesize $^{1}$ High Energy Physics Division, Argonne National Laboratory, Argonne, IL, USA}
+\par {\footnotesize $^{2}$ Physics Department, Brookhaven National Laboratory, Upton, NY, USA}
 \par {\footnotesize $^{3}$ Centro Brasileiro de Pesquisas Físicas, Rio de Janeiro, Brazil}
@@
 \par {\footnotesize $^{5}$ CERN, Geneva, Switzerland}
-\par {\footnotesize $^{6}$ Enrico Fermi Institute, University of Chicago, Chicago IL, USA}
-\par {\footnotesize $^{7}$ Centro de Investigaciones Energéticas Medioambientales y Tecnológicas (CIEMAT), Madrid, Spain}
-\par {\footnotesize $^{8}$ Cinvestav, Mexico City, Mexico}
-\par {\footnotesize $^{9}$ Centro Nazionale Analisi Fotogrammi (CNAF), INFN, Bologna, Italy}
-\par {\footnotesize $^{10}$ Center for High Performance Computing, Cape Town, South Africa}
-\par {\footnotesize $^{11}$ California Institute of Technology, Pasadena, California, USA}
-\par {\footnotesize $^{12}$ Cornell University, Ithaca, USA}
-\par {\footnotesize $^{13}$ Deutsches Elektronen-Synchrotron, Hamburg, Germany}
-\par {\footnotesize $^{14}$ Institute of Physics, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland}
-\par {\footnotesize $^{15}$ ETH Zurich - Institute for Particle Physics and Astrophysics (IPA), Zurich, Switzerland}
-\par {\footnotesize $^{16}$ Fermi National Accelerator Laboratory, Batavia, USA}
-\par {\footnotesize $^{17}$ University of Hamburg, Hamburg, Germany}
-\par {\footnotesize $^{18}$ Helsinki Institute of Physics, Helsinki, Finland}
-\par {\footnotesize $^{19}$ International Center for Theoretical Physics, Trieste, Italy}
-\par {\footnotesize $^{20}$ Institut de Física d’Altes Energies and Departament de Física de la Universitat Autònoma de Barcelona and ICREA, Barcelona, Spain}
-\par {\footnotesize $^{21}$ Instituto de Física de Cantabria (IFCA), CSIC-Universidad de Cantabria, Santander, Spain}
-\par {\footnotesize $^{22}$ Instituto de Física Corpuscular, Centro Mixto Universidad de Valencia - CSIC, Valencia, Spain}
-\par {\footnotesize $^{23}$ Institute of High Energy Physics, Chinese Academy of Sciences, Beijing}
-\par {\footnotesize $^{24}$ Department of Physics, Indiana University, Bloomington IN, USA}
-\par {\footnotesize $^{25}$ INFN Sezione di Bologna, Università di Bologna, Bologna, Italy}
-\par {\footnotesize $^{26}$ INFN Sezione di Bari, Università di Bari, Politecnico di Bari, Bari, Italy}
-\par {\footnotesize $^{27}$ Università e INFN, Ferrara, Ferrara, Italy}
-\par {\footnotesize $^{28}$ INFN Sezione di Genova, Genova, Italy}
-\par {\footnotesize $^{29}$ INFN Sezione di Milano-Bicocca, Milano, Italy}
-\par {\footnotesize $^{30}$ INFN Sezione di Napoli, Università di Napoli, Napoli, Italy}
-\par {\footnotesize $^{31}$ INFN Sezione di Pisa, Università di Pisa, Scuola Normale Superiore di Pisa, Pisa, Italy}
-\par {\footnotesize $^{32}$ INFN Sezione di Padova, Università di Padova b, Padova, Italy}
-\par {\footnotesize $^{33}$ INFN Sezione di Roma I, Università La Sapienza, Roma, Italy}
-\par {\footnotesize $^{34}$ INFN Sezione di Trieste, Università di Trieste, Trieste, Italy}
+\par {\footnotesize $^{6}$ Centro de Investigaciones Energéticas Medioambientales y Tecnológicas (CIEMAT), Madrid, Spain}
+\par {\footnotesize $^{7}$ Cinvestav, Mexico City, Mexico}
+\par {\footnotesize $^{8}$ Centro Nazionale Analisi Fotogrammi (CNAF), INFN, Bologna, Italy}
+\par {\footnotesize $^{9}$ Center for High Performance Computing, Cape Town, South Africa}
+\par {\footnotesize $^{10}$ California Institute of Technology, Pasadena, California, CA, USA}
+\par {\footnotesize $^{11}$ Cornell University, Ithaca, NY, USA}
+\par {\footnotesize $^{12}$ Deutsches Elektronen-Synchrotron, Hamburg, Germany}
+\par {\footnotesize $^{13}$ Institute of Physics, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland}
+\par {\footnotesize $^{14}$ ETH Zürich - Institute for Particle Physics and Astrophysics (IPA), Zürich, Switzerland}
+\par {\footnotesize $^{15}$ Fermi National Accelerator Laboratory, Batavia, IL, USA}
+\par {\footnotesize $^{16}$ University of Hamburg, Hamburg, Germany}
+\par {\footnotesize $^{17}$ Helsinki Institute of Physics, Helsinki, Finland}
+\par {\footnotesize $^{18}$ International Center for Theoretical Physics, Trieste, Italy}
+\par {\footnotesize $^{19}$ Institut de Física d’Altes Energies and Departament de Física de la Universitat Autònoma de Barcelona and ICREA, Barcelona, Spain}
+\par {\footnotesize $^{20}$ Instituto de Física de Cantabria (IFCA), CSIC-Universidad de Cantabria, Santander, Spain}
+\par {\footnotesize $^{21}$ Instituto de Física Corpuscular, Centro Mixto Universidad de Valencia - CSIC, Valencia, Spain}
+\par {\footnotesize $^{22}$ Institute of High Energy Physics, Chinese Academy of Sciences, Beijing}
+\par {\footnotesize $^{23}$ Department of Physics, Indiana University, Bloomington, IN, USA}
+\par {\footnotesize $^{24}$ INFN Sezione di Bologna, Università di Bologna, Bologna, Italy}
+\par {\footnotesize $^{25}$ INFN Sezione di Bari, Università di Bari, Politecnico di Bari, Bari, Italy}
+\par {\footnotesize $^{26}$ Università e INFN, Ferrara, Ferrara, Italy}
+\par {\footnotesize $^{27}$ INFN Sezione di Genova, Genova, Italy}
+\par {\footnotesize $^{28}$ INFN Sezione di Milano-Bicocca, Milano, Italy}
+\par {\footnotesize $^{29}$ INFN Sezione di Napoli, Università di Napoli, Napoli, Italy}
+\par {\footnotesize $^{30}$ INFN Sezione di Pisa, Università di Pisa, Scuola Normale Superiore di Pisa, Pisa, Italy}
+\par {\footnotesize $^{31}$ INFN Sezione di Padova, Università di Padova b, Padova, Italy}
+\par {\footnotesize $^{32}$ INFN Sezione di Roma I, Università La Sapienza, Roma, Italy}
+\par {\footnotesize $^{33}$ INFN Sezione di Trieste, Università di Trieste, Trieste, Italy}
+\par {\footnotesize $^{34}$ INFN Sezione di Torino, Torino, Italy}
 \par {\footnotesize $^{35}$ Université de Strasbourg, CNRS, IPHC UMR 7178, F-67000 Strasbourg, France}
@@
 \par {\footnotesize $^{37}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat à l’Énergie Atomique), Gif-sur-Yvette, France}
-\par {\footnotesize $^{38}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
+\par {\footnotesize $^{38}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, VA, USA}
 \par {\footnotesize $^{39}$ Jožef Stefan Institute, Ljubljana, Slovenia}
@@
 \par {\footnotesize $^{43}$ LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
-\par {\footnotesize $^{44}$ Lawrence Berkeley National Laboratory and University of California, Berkeley CA, USA}
+\par {\footnotesize $^{44}$ Lawrence Berkeley National Laboratory and University of California, Berkeley, CA, USA}
 \par {\footnotesize $^{45}$ Laboratório de Instrumentação e Física Experimental de Partículas (LIP), Lisboa, Portugal}
@@
 \par {\footnotesize $^{55}$ Department of Physics, New York University, New York, NY, USA}
-\par {\footnotesize $^{56}$ Center for High Energy Physics, University of Oregon, Eugene OR, USA}
+\par {\footnotesize $^{56}$ Center for High Energy Physics, University of Oregon, Eugene, OR, USA}
 \par {\footnotesize $^{57}$ Port d’Informació Científica (PIC), Universitat Autònoma de Barcelona (UAB), Barcelona, Spain}
@@
 \par {\footnotesize $^{60}$ Department of Physics, Royal Holloway University of London, Surrey, United Kingdom}
-\par {\footnotesize $^{61}$ SLAC National Accelerator Laboratory, Menlo Park CA, USA}
-\par {\footnotesize $^{62}$ Simons Foundation, New York, USA}
+\par {\footnotesize $^{61}$ SLAC National Accelerator Laboratory, Menlo Park, CA, USA}
+\par {\footnotesize $^{62}$ Simons Foundation, New York, NY, USA}
 \par {\footnotesize $^{63}$ SUBATECH, IMT Atlantique, Université de Nantes, CNRS-IN2P3, Nantes, France}
@@
 \par {\footnotesize $^{72}$ Department of Physics and Astronomy, University of California Irvine, Irvine, CA, USA}
-\par {\footnotesize $^{73}$ University of California, San Diego, La Jolla, USA}
+\par {\footnotesize $^{73}$ University of California, San Diego, La Jolla, CA, USA}
 \par {\footnotesize $^{74}$ Cavendish Laboratory, University of Cambridge, Cambridge, United Kingdom}
@@
 \par {\footnotesize $^{76}$ Carleton University, Ottawa, ON, Canada}
-\par {\footnotesize $^{77}$ University of Illinois Urbana-Champaign, Champaign, Illinois, USA}
+\par {\footnotesize $^{77}$ University of Illinois Urbana-Champaign, Champaign, Illinois, IL, USA}
 \par {\footnotesize $^{78}$ Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
@@
 \par {\footnotesize $^{85}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
-\par {\footnotesize $^{86}$ Department of Physics and Astronomy, University of Kentucky, Lexington, USA}
+\par {\footnotesize $^{86}$ Department of Physics and Astronomy, University of Kentucky, Lexington, KY, USA}
 \par {\footnotesize $^{87}$ Physics Department, Lancaster University, Lancaster, United Kingdom}
@@
 \par {\footnotesize $^{93}$ Departamento de Física, Universidade do Minho, Braga, Portugal}
-\par {\footnotesize $^{94}$ University of Nebraska-Lincoln, Lincoln, USA}
-\par {\footnotesize $^{95}$ Department of Physics and Astronomy, University of New Mexico, Albuquerque NM, USA}
-\par {\footnotesize $^{96}$ University of Notre Dame, Notre Dame, USA}
-\par {\footnotesize $^{97}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman OK, USA}
-\par {\footnotesize $^{98}$ The Ohio State University, Columbus, USA}
+\par {\footnotesize $^{94}$ University of Nebraska-Lincoln, Lincoln, NE, USA}
+\par {\footnotesize $^{95}$ Department of Physics and Astronomy, University of New Mexico, Albuquerque, NM, USA}
+\par {\footnotesize $^{96}$ University of Notre Dame, Notre Dame, IN, USA}
+\par {\footnotesize $^{97}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman, OK, USA}
+\par {\footnotesize $^{98}$ The Ohio State University, Columbus, OH, USA}
 \par {\footnotesize $^{99}$ Department of Physics, University of Oslo, Oslo, Norway}
@@
 \par {\footnotesize $^{102}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
-\par {\footnotesize $^{103}$ Princeton University, Princeton, USA}
+\par {\footnotesize $^{103}$ Princeton University, Princeton, NJ, USA}
 \par {\footnotesize $^{104}$ Rice University, Houston, TX, USA}
@@
 \par {\footnotesize $^{106}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
-\par {\footnotesize $^{107}$ School of Physics, Shandong University, Shandong, China}
-\par {\footnotesize $^{108}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
-\par {\footnotesize $^{109}$ University of Sofia, Sofia, Bulgaria}
-\par {\footnotesize $^{110}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
-\par {\footnotesize $^{111}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
-\par {\footnotesize $^{112}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
-\par {\footnotesize $^{113}$ Dipartimento di Chimica, Fisica e Ambiente, Università di Udine, Udine, Italy}
-\par {\footnotesize $^{114}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
-\par {\footnotesize $^{115}$ University of Wisconsin - Madison, Madison, WI, USA}
-\par {\footnotesize $^{116}$ Department of Physics, Yale University, New Haven, CT, USA}
-\par {\footnotesize $^{117}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
-\par {\footnotesize $^{118}$ Virginia Tech, Blacksburg, Virginia, USA}
+\par {\footnotesize $^{107}$ University of Washington, Seattle, WA, USA}
+\par {\footnotesize $^{108}$ School of Physics, Shandong University, Shandong, China}
+\par {\footnotesize $^{109}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
+\par {\footnotesize $^{110}$ University of Sofia, Sofia, Bulgaria}
+\par {\footnotesize $^{111}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
+\par {\footnotesize $^{112}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
+\par {\footnotesize $^{113}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
+\par {\footnotesize $^{114}$ Dipartimento di Chimica, Fisica e Ambiente, Università di Udine, Udine, Italy}
+\par {\footnotesize $^{115}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
+\par {\footnotesize $^{116}$ University of Wisconsin - Madison, Madison, WI, USA}
+\par {\footnotesize $^{117}$ Department of Physics, Yale University, New Haven, CT, USA}
+\par {\footnotesize $^{118}$ Physik-Institut, Universität Zürich, Zürich, Switzerland}
+\par {\footnotesize $^{119}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
+\par {\footnotesize $^{120}$ Virginia Tech, Blacksburg, Virginia, VA, USA}
 \bigskip

### 2018-02-09 CWP/papers/HSF-CWP-2017-01_roadmap/authors/a2tex.py

@@ -145,7 +145,9 @@ def main():
                 if affiliation in institute_address:
                     affiliation_map[affiliation] = Affiliation(institute_address[affiliation])
                 else:
-                    raise Exception('Affiliation {} not defined'.format(affiliation))
+                    raise Exception('Affiliation {} not defined for {}, {}'.format(affiliation, 
+                                                                                   author.surname,
+                                                                                   author.forename))
 
     # Alphabetise and assign footnote marks
     sorted_affiliations = sorted(affiliation_map)

### 2018-02-15 CWP/papers/HSF-CWP-2017-05_analysis/latex/cwp-wg-analysis-interpretation.tex
updates to affiliations

@@
 \author{HEP Software Foundation:}
-\author[1]{Riccardo Maria Bianchi}
-\author[2]{Brian Bockelman}
-\author[3]{Nuno Castro}
-\author[4]{Peter Elmer}
-\author[16]{Robert Gardner}
-\author[14]{Maria Girone}
-\author[5]{Oliver Gutsche}
-\author[14]{Benedikt Hegner}
-\author[6]{Jos\'{e} M. Hern\'{a}ndez}
-\author[5]{Bodhitha Jayatilaka}
-\author[4]{David Lange}
-\author[7]{Mark Neubauer}
-\author[7]{Daniel S. Katz}
-\author[8]{Lukasz Kreczko}
-\author[9]{James Letts}
-\author[10]{Shawn McKee}
-\author[11]{Christoph Paus}
-\author[5]{Kevin Pedro}
-\author[12]{Eduardo Rodrigues}
-\author[8]{Tai Sakuma}
-\author[5]{Elizabeth Sexton-Kennedy}
-\author[13]{Michael D. Sokoloff}
-\author[9]{Frank W\"{u}rthwein}
-\author[15]{Carl Vuosalo}
+\author[a]{Riccardo Maria Bianchi}
+\author[b]{Brian Bockelman}
+\author[c,d]{Nuno Castro}
+\author[e]{Peter Elmer}
+\author[f]{Robert Gardner}
+\author[g]{Maria Girone}
+\author[h,1]{Oliver Gutsche}
+\author[g]{Benedikt Hegner}
+\author[i]{Jos\'{e} M. Hern\'{a}ndez}
+\author[h]{Bodhitha Jayatilaka}
+\author[e]{David Lange}
+\author[j,1]{Mark Neubauer}
+\author[j]{Daniel S. Katz}
+\author[k]{Lukasz Kreczko}
+\author[l]{James Letts}
+\author[m]{Shawn McKee}
+\author[n]{Christoph Paus}
+\author[h]{Kevin Pedro}
+\author[o]{Eduardo Rodrigues}
+\author[k]{Tai Sakuma}
+\author[h]{Elizabeth Sexton-Kennedy}
+\author[o]{Michael D. Sokoloff}
+\author[l]{Frank W\"{u}rthwein}
+\author[p]{Carl Vuosalo}
 
-\affiliation[1]{University of Pittsburgh}
-\affiliation[2]{University of Nebraska-Lincoln}
-\affiliation[3]{LIP and University of Minho, Supported by FCT-Portugal, IF/00050/2013/CP1172/CT0002}
-\affiliation[4]{Princeton University}
-\affiliation[5]{Fermi National Accelerator Laboratory, Supported by the US-DOE, DE-AC02-07CH11359}
-\affiliation[6]{CIEMAT, Supported by ES-MINECO, FPA2016-80994-c2-1-R and MDM-2015-0509}
-\affiliation[7]{University of Illinois Urbana-Champaign}
-\affiliation[8]{University of Bristol}
-\affiliation[9]{UC San Diego}
-\affiliation[10]{University of Michigan, Supported by the US-DOE, DE-SC0007859 and US-NSF, 76749/1136652/2}
-\affiliation[11]{Massachusetts Institute of Technology}
-\affiliation[12]{University of Cincinnati, Supported by the US-NSF, ACI-1450319}
-\affiliation[13]{University of Cincinnati, Supported by the US-NSF, ACI-1558219}
-\affiliation[14]{CERN}
-\affiliation[15]{University of Wisconsin-Madison}
-\affiliation[16]{University of Chicago}
+\affiliation[a]{University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
+\affiliation[b]{University of Nebraska-Lincoln, Lincoln, Nebraska, USA}
+\affiliation[c]{Laboratório de Instrumentação e Física Experimental de Partículas, Lisbon, Portugal}
+\affiliation[d]{University of Minho, Braga, Portugal}
+\affiliation[e]{Princeton University, Princeton, New Jersey, USA}
+\affiliation[f]{University of Chicago, Chicago, Illinois, USA}
+\affiliation[g]{CERN, Geneva, Switzerland}
+\affiliation[h]{Fermi National Accelerator Laboratory, Batavia, Illinois USA}
+\affiliation[i]{Centro de Investigaciones Energéticas, Medioambientales y Tecnológicas, Madrid, Spain}
+\affiliation[j]{University of Illinois at Urbana-Champaign, Urbana, Illinois USA}
+\affiliation[k]{University of Bristol, Bristol, UK}
+\affiliation[l]{University of California at San Diego, La Jolla, California, USA}
+\affiliation[m]{University of Michigan, Ann Arbor, Michigan, USA}
+\affiliation[n]{Massachusetts Institute of Technology, Cambridge, Massachusetts, USA}
+\affiliation[o]{University of Cincinnati, Cincinnati, Ohio, USA}
+\affiliation[p]{University of Wisconsin-Madison, Madison, Wisconsin, USA}
+\affiliation[1]{Paper Editor}
+
+%\affiliation[c]{Laboratório de Instrumentação e Física Experimental de Partículas, Lisbon, Portugal and University of Minho, Braga, Portugal, Supported by FCT-Portugal, IF/00050/2013/CP1172/CT0002}
+%\affiliation[g]{Fermi National Accelerator Laboratory, Batavia, Illinois USA, Supported by the US-DOE, DE-AC02-07CH11359}
+%\affiliation[h]{Centro de Investigaciones Energéticas, Medioambientales y Tecnológicas (CIEMAT), Madrid, Spain, Supported by ES-MINECO, FPA2016-80994-c2-1-R and MDM-2015-0509}
+%\affiliation[i]{University of Illinois at Urbana-Champaign, Urbana, Illinois USA, Supported by the US-DOE, DE-SC0018098 and US-NSF ACI-1558233}
+%\affiliation[m]{University of Michigan, Supported by the US-DOE, DE-SC0007859 and US-NSF, 76749/1136652/2}
+%\affiliation[o]{University of Cincinnati, Supported by the US-NSF, ACI-1450319}
+%\affiliation[p]{University of Cincinnati, Supported by the US-NSF, ACI-1558219}
 

### 2018-02-20 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
updated author list
updated author and affiliation lists

@@
 
-\author[d]{J Apostolakis}
-\author[q]{M Asai}
-\author[n]{S Banerjee}
-\author[m]{R Bianchi}
-\author[d]{F Carminati}
-\author[l]{R Cenci}
+\author[c]{J Apostolakis}
+\author[p]{M Asai}
+\author[e]{S Banerjee}
+\author[l]{R Bianchi}
+\author[e]{P Canal}
+\author[k]{R Cenci}
 \author[b]{J Chapman}
-\author[d]{G Corti}
-\author[d]{G Cosmo}
-\author[p]{S Easo}
-\author[q]{A Dotti}
-\author[f,1]{V Elvira}
-\author[f]{L Fields}
-\author[f]{K Genser}
-\author[d]{A Gheata}
-\author[d]{M Gheata}
-\author[d]{I Goulas}
-\author[d,1]{J Harvey}
-\author[d]{F Hariri}
-\author[f]{R Hatcher}
-\author[f]{K Herner}
-\author[d]{V Ivantchenko}
-\author[f]{T Junk}
-\author[f]{S-Y Jun}
-\author[f]{M Kirby}
-\author[f]{R Kutschke}
-\author[f]{P Lebrun}
-\author[f]{A Lyon}
-\author[j]{Z Marshall}
-\author[c]{J-F Marchand}
+\author[c]{G Corti}
+\author[c]{G Cosmo}
+\author[o]{S Easo}
+\author[h]{L de Oliveira}
+\author[p]{A Dotti}
+\author[e,1]{V Elvira}
+\author[e]{L Fields}
+\author[e]{K Genser}
+\author[c]{A Gheata}
+\author[c]{M Gheata}
+\author[c,1]{J Harvey}
+\author[c]{F Hariri}
+\author[e]{R Hatcher}
+\author[e]{K Herner}
+\author[j]{M Hildreth}
+\author[c]{V Ivantchenko}
+\author[e]{T Junk}
+\author[e]{S-Y Jun}
+\author[e]{M Kirby}
+\author[m]{D Konstantinov}
+\author[e]{R Kutschke}
+\author[e]{P Lebrun}
+\author[e]{G Lima}
+\author[e]{A Lyon}
+\author[h]{Z Marshall}
+\author[c]{P Mato}
 \author[a]{M Mooney}
-\author[q]{R Mount}
-\author[k]{J Mousseau}
-\author[j]{B Nachman}
-\author[f]{A Norman}
-\author[f]{I Osborne}
-\author[f]{K Pedro}
-\author[d]{W Pokorski}
+\author[p]{R Mount}
+\author[i]{J Mousseau}
+\author[h]{B Nachman}
+\author[e]{A Norman}
+\author[c]{M Novak}
+\author[e]{I Osborne}
+\author[q]{M Paganini}
+\author[e]{K Pedro}
+\author[c]{W Pokorski}
 \author[a]{X Qian}
-\author[f]{J Raaf}
-\author[l]{M Rama}
-\author[d]{A Ribon}
-\author[d]{S Roiser}
-\author[o]{D Ruterbories}
-\author[f]{S Sekmen}
-\author[h]{B Siddi}
-\author[f]{E Snider}
-\author[g]{S Vallecorsa}
-\author[e]{M Verderi}
-\author[f]{H Wenzel}
-\author[d]{S Wenzel}
-\author[d]{B Viren}
-\author[f]{T Yang}
-\author[f]{L Welty-Rieger}
+\author[k]{M Rama}
+\author[c]{A Ribon}
+\author[c]{S Roiser}
+\author[n]{D Ruterbories}
+\author[e]{S Sekmen}
+\author[g]{B Siddi}
+\author[e]{E Snider}
+\author[f]{S Vallecorsa}
+\author[d]{M Verderi}
+\author[e]{H Wenzel}
+\author[c]{S Wenzel}
+\author[a]{B Viren}
+\author[e]{T Yang}
+\author[e]{J Yarba}
+\author[e]{L Welty-Rieger}
 
@@
 \affiliation[b]{University of Cambridge,Cambridge, UK}
-\affiliation[c]{Carleton University, Ottawa, ON, Canada}
-\affiliation[d]{CERN, Geneva, Switzerland}
-\affiliation[e]{IN2P3-LLR, Ecole Polytechnique, Palaiseau, France}
-\affiliation[f]{Fermilab National Accelerator Laboratory, Batavia IL , USA}
-\affiliation[g]{Gangneung-Wonju National University, Gangneung, Gangwon-do, South Korea}
-\affiliation[h]{Università di Ferrara and INFN Ferrara, Ferrara, Italy}
-\affiliation[j]{Lawrence Berkeley National Laboratory, Berkeley CA, USA}
-\affiliation[k]{University of Michigan, Ann Arbor MI, USA}
-\affiliation[l]{Università di Pisa and INFN Pisa, Pisa, Italy}
-\affiliation[m]{University of Pittsburgh, Pittsburgh PA, USA}
-\affiliation[n]{Princeton University, Princeton, NJ 08544, USA}
-\affiliation[o]{University of Rochester, Rochester NY, USA}
-\affiliation[p]{Rutherford Appleton Laboratory, Chilton, UK}
-\affiliation[q]{SLAC National Accelerator Laboratory, Menlo Park, CA 94025, USA}
-\affiliation[r]{University of Illinois Urbana-Champaign, Champaign IL, USA}
-
+\affiliation[c]{CERN, Geneva, Switzerland}
+\affiliation[d]{IN2P3-LLR, Ecole Polytechnique, Palaiseau, France}
+\affiliation[e]{Fermilab National Accelerator Laboratory, Batavia IL , USA}
+\affiliation[f]{Gangneung-Wonju National University, Gangneung, Gangwon-do, South Korea}
+\affiliation[g]{Università di Ferrara and INFN Ferrara, Ferrara, Italy}
+\affiliation[h]{Lawrence Berkeley National Laboratory, Berkeley CA, USA}
+\affiliation[i]{University of Michigan, Ann Arbor MI, USA}
+\affiliation[j]{University of Notre Dame, Notre Dame, IN 46556, USA}
+\affiliation[k]{Università di Pisa and INFN Pisa, Pisa, Italy}
+\affiliation[l]{University of Pittsburgh, Pittsburgh PA, USA}
+\affiliation[m]{Institute For High Energy Physics(IHEP), Protvino,Russia}
+\affiliation[n]{University of Rochester, Rochester NY, USA}
+\affiliation[o]{Rutherford Appleton Laboratory, Chilton, UK}
+\affiliation[p]{SLAC National Accelerator Laboratory, Menlo Park, CA 94025, USA}
+\affiliation[q]{Yale University, New Haven, CT 06520, USA}
 \affiliation[1]{Paper Editor}

### 2018-02-21 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
added text and reference on LiAr TPCs in section on Intensity Frontier

@@
 include liquid Argon TPC detectors and liquid Xenon calorimeters.
+For example, liquid Argon TPC's are equipped with optical readout systems 
+to detect optical photons from the scintillation process. The information 
+from the optical system is used to  provide a trigger and, potentially, 
+could be used to improve the energy resolution of showers by combining the 
+scintillation and charge collection signals. The modeling of optical photons 
+in liquid Argon detectors is a real challenge, since 50,000 scintillation 
+photons are produced per 1 MeV of energy deposited in liquid Argon. To track 
+all these photons in Geant4 would increase the computing time on a CPU by a 
+factor of  more than one thousand. Therefore, the full simulation is only used 
+to produce lookup tables with the probability that that an optical photon makes 
+it to the photo detector. These tables are an approximation and cannot replace 
+the full simulation without a significant loss in physics accuracy. Even the 
+lookup table solution requires a significant amount of CPU time to produce 
+enough statistics. In order to satisfy the physics accuracy and computing 
+performance requirements of future neutrino experiments, approaches based 
+on new computing platforms are under investigation. For example, 
+Opticks ~\cite{1742-6596-898-4-042001} 
+is a software package where the optical physics processes are implemented on the GPU. 
+This package is executed on NVIDIA machines to parallelize the
+tracing of optical photons and yields a drastic time performance improvement 
+with a speedup of the order of a thousand with respect to the original Geant4 
+package running on CPUs. Opticks has been used to simulate the optical photons 
+in the JUNO neutrino experiment.
+
 Efforts to address common needs of the liquid Argon experiments

### 2018-02-21 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
changed one sentence to read better

@@
 have typically used Geant4 physics lists tuned based on LHC
-requirements. However, as the precision needs of these experiments
-increase, they increasingly need simulations responding to their
-particular needs, which vary significantly with experiment but broadly
-speaking involve precise simulations of hadrons and electromagnetic
-particles from a few GeV to a few MeV. Precise neutron simulation is
+requirements. However, as the physics programs of these experiments evolve,
+they require simulation of increasing accuracy, which depends significantly
+on the experiment but in all cases involves accurate modeling of hadrons and
+electromagnetically interacting particles from a few MeV to a few GeV.
+Precise neutron simulation is
 particularly critical for accelerator-based neutrino experiments that

### 2018-02-28 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
final check of author list

@@
  & & HSF-CWP-2017-13 \\
- & & January 24, 2018 \\ % use \date or hardwire e.g. December 15, 2017
+ & & February 28, 2018 \\ % use \date or hardwire e.g. December 15, 2017
  & & \\
@@
 \author[a]{X Qian}
+\author[e]{J Raaf}
 \author[k]{M Rama}
@@
 \author[e]{L Welty-Rieger}
+\author[p]{D.H.Wright}
 

### 2018-02-28 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
2 more names on author list
corrected reference 4 to NOvA experiment

@@
 \author[e,1]{V Elvira}
+\author[h]{S Farrell}
 \author[e]{L Fields}
@@
 \author[e]{L Welty-Rieger}
-\author[p]{D.H.Wright}
+\author[p]{D H Wright}
+\author[a]{C Zhang}
 
@@
 experimental programme. Properly simulating a neutrino experiment like
-NOvA ~\cite{1742-6596-396-2-022014}, MicroBooNE~\cite{MicroBooNE}, 
+NOvA ~\cite{Adamson:2016tbq}, MicroBooNE~\cite{MicroBooNE}, 
 MINERvA~\cite{Aliaga:2013uqz}, or DUNE ~\cite{Acciarri:2015uup}

### 2018-03-01 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
final corrections to references and subsection numbering

@@
 Components of the HEP simulation chain, such as physics generators and
-event visualisation tools are discussed in other chapters, while the
+event visualisation tools are discussed elsewhere, while the
 modeling of the propagation and beam manipulation through accelerator
 lattices (i.e. CERN and Fermilab accelerator complexes) are considered
-to be outside the scope of this chapter. Space and medical applications
+to be outside the scope of this report. Space and medical applications
 are referred to briefly where appropriate.
@@
 bremsstrahlung, pair production and multiple scattering
-~\cite{1742-6596-396-2-022013} ~\cite{1742-6596-513-2-022015}
-further improving simulation of the EM shower shape; these improvements
+further improving simulation of the EM shower shape
+~\cite{1742-6596-396-2-022013}~\cite{1742-6596-513-2-022015}; these improvements
 were made available in Geant4 9.5 and 9.6. For validation purposes, an
@@
 shower shapes. A comprehensive description of recent improvements of the
-Geant4 electromagnetic module can be found in ~\cite{1742-6596-513-2-022015}. Rather good
+Geant4 electromagnetic module can be found in~\cite{1742-6596-513-2-022015}. Rather good
 agreement was found when Geant4 predictions were compared with
@@
 
+\begin{figure}[bthp]
+\centering
+\includegraphics[width=0.94\textwidth]{image5.png}
+\caption{The new Goudsmit-Saunderson multiple scattering
+model implementation (GS - Geant4 10.2) shows no artefacts and excellent
+agreement with experimental data}
+\label{fig:gs}
+\end{figure}
+
 \begin{itemize}
@@
 \end{itemize}
-
-\begin{figure}[bthp]
-\centering
-\includegraphics[width=0.94\textwidth]{image5.png}
-\caption{The new Goudsmit-Saunderson multiple scattering
-model implementation (GS - Geant4 10.2) shows no artefacts and excellent
-agreement with experimental data}
-\label{fig:gs}
-\end{figure}
 
@@
 
-8.2 ATLAS and CMS
+\hypertarget{ATLAS-and-CMS}{%
+\subsection{ATLAS and CMS}\label{ATLAS-and-CMS}}
 
@@
   Machine: olhswep16.cern.ch (CERN's OpenLab), one thread runs
-\end{itemize}
-
-\begin{itemize}
 \item

### 2018-03-01 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
moved position of 2 references 18 and 19

@@
 ~\cite{1742-6596-331-3-032029}. For Run-2 significant improvements were introduced into the
-bremsstrahlung, pair production and multiple scattering
+bremsstrahlung, pair production and multiple scattering,
 further improving simulation of the EM shower shape 

### 2018-03-02 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
fix the HSF document number, from 13 to 7

@@
 \begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
- & & HSF-CWP-2017-13 \\
+ & & HSF-CWP-2017-07 \\
  & & February 28, 2018 \\ % use \date or hardwire e.g. December 15, 2017

### 2018-03-11 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
Initial draft for HSF-CWP-2017-04 (DOMA)

% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.

% JHEP preprint template
\documentclass[12pt,a4paper]{article}
\usepackage{jheppub}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xspace}

\usepackage[utf8]{inputenc}

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp.bib}
\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp-chapters.bib}

%% Comment out before final submission
%\usepackage{lineno}  % for line numbering during review
%\linenumbers

\abstract{To be done.}

\begin{document}

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-04 \\
 & & March 11, 2018 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{HSF Community White Paper on Data Organization, Management and Access
(DOMA)}

Authors: D. Berzano, R.M. Bianchi, I. Bird, B. Bockelman, K. De, S.
Campana, D. Duellmann, P.Elmer, R. Gardner, V. Garonne, C. Grandi, O.
Gutsche, A. Hanushevsky, B. Holzman, B. Jayatilaka, I. Jimenez, M.
Jouvin, O. Keeble, A. Klimentov, V. Kuznetsov, E. Lançon, M. Lassnig, T.
Malik, C. Maltzahn, S. McKee, D. Menasce, M. Livny, B. Nitzberg, A.
Norman, J. Pivarski, M. Schulz, H. Severini, M. Sevilla, L.
Sexton-Kennedy, E. Vaandering, N. Watkins, T. Wenaus, F. Wuerthwein

\maketitle

\newpage

\section{Introduction}\label{introduction}

The coming decade will bring HEP to the exabyte scale. The expected data
volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
experiments such as SKA (Square Kilometre Array) will all enter this
regime. In devising experimental computing models for this era, many
factors have to be taken into account. ~In particular, the increasing
availability of very high-speed networks, which may reduce the need for
CPU and data co-location, need to be examined. Such networks may allow
for more extensive use of data access over the wide-area network (WAN),
which may provide failover capabilities, global and federated data
namespaces, and will have an impact on data caching. Shifts in the data
presentation/analysis models such as a potential move to event-based
data streaming from the more traditional dataset-based or file-based
data access, will be particularly important for optimizing the
utilization of opportunistic computing cycles on HPC facilities,
commercial cloud resources, and campus clusters, and can potentially
resolve currently limiting factors such as job eviction.

Planned analysis of HL-LHC data and data analysis from other large
international collaborations will need to adopt a distributed computing
model due to its scope. The data management systems that will enable
this type of analysis will need to provide cost-based optimization of
the data locations and delivery mechanisms.

\section{Challenges and
Opportunities}\label{challenges-and-opportunities}

The LHC experiments currently provision and manage about an Exabyte of
storage, approximately half of which is archival, and half is
traditional disk storage. The storage requirements per year are expected
to jump by a factor of 10 for the HL-LHC. This itself is faster than
projected Moore's Law gains and will present major challenges. Storage
will remain one of the visible cost drivers for HEP computing, however
the projected growth and cost of the computational resources needed to
analyze the data is also expected to grow even faster than the base
storage costs. The combination of storage and analysis computing costs
may restrict scientific output and potential physics reach of the
experiments. Thus, new techniques and algorithms are likely to be
required.

The three main challenges for data in the HL-LHC era can thus be
summarized:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  \textbf{Big Data:} The HL-LHC era will significantly increase both the
  date rate and the data volume. The computing systems will need to
  handle this without significant cost increases and within evolving
  storage technology limitations.
  \end{quote}
\item
  \begin{quote}
  \textbf{Dynamic Distributed Computing:} In addition, the significantly
  increased computational requirements for the HL-LHC era will also
  place new requirements on data. Specifically, the use of new types of
  compute resources (\emph{e.g.,} cloud, HPC) with different dynamic
  availability and characteristics are used will require more dynamic
  DOMA systems.
  \end{quote}
\item
  \begin{quote}
  \textbf{New Applications:} New applications, such as machine learning
  training or high-rate data-query systems for analysis, will likely be
  employed to meet the computational constraints and to extend the
  physics reach of the HL-LHC. These new applications will place new
  requirements on how and where data is accessed and produced. For
  example, specific applications (e.g. training for machine learning)
  may require use of specialized processor resources such as GPUs,
  placing further requirements on data.
  \end{quote}
\end{enumerate}

The rapid increase in recent years of data-intensive problems in both
the commercial world and in the rest of the research world also provides
a number of opportunities.

\section{Current Approaches}\label{current-approaches}

The original LHC computing models (circa 2005) were built up from the
simpler models used before distributed computing was a central part of
HEP computing. This allowed for a reasonably clean separation between
three different aspects of interacting with data: organization,
management and access.

\begin{itemize}
\item
  \begin{quote}
  \textbf{Data Organization:} This is essentially how data is structured
  as it is written. Most data is written in flat files, in ROOT format,
  typically with a column-wise organization of the data. The records
  corresponding to these columns are compressed. The internal details of
  this organization are typically visible only to individual software
  applications.
  \end{quote}
\item
  \begin{quote}
  \textbf{Data Management:} The key challenge here was the transition to
  the use of distributed computing in the form of the grid. The
  experiments developed dedicated data transfer and placement systems,
  along with catalogs, to move data between computing centers. To first
  order, the computing models were rather static: data was placed at
  sites and the relevant compute jobs were sent to the right locations.
  Applications might interact with catalogs or, at times, the workflow
  management systems does this on behalf of the applications.
  \end{quote}
\item
  \begin{quote}
  \textbf{Data Access:} Various protocols are used for direct reads
  (rfio, dcap, xrootd, etc.) with a given computer center and/or
  explicit local stage-in and caching for read by jobs. Application
  access may use different protocols than those used by the data
  transfers between sites.
  \end{quote}
\end{itemize}

Before the LHC turn-on and in the first years of the LHC, these three
areas were, to first order, optimized independently. Many of the
challenges were in the area of ''Data Management (DM)'' as the Worldwide
LHC Computing Grid was commissioned. As the LHC computing matured
through Run1 and Run2, the interest has turned to optimizations spanning
these three areas. For example, the recent use of ``Data Federations''
mixes up Data Management and Access. As we will see below, some of the
foreseen opportunities towards HL-LHC may require global optimizations.

Thus in this document we take a broader view than traditional ``DM'',
and consider the combination of ``Data Organization, Management and
Access (DOMA)'') together. We believe that this full picture of data
needs in HEP will provide important opportunities for efficiency and
scalability as we enter the many-Exabyte era.

\subsection{HEP Workflow in Context}\label{hep-workflow-in-context}

The way data are used at HEP experiments results in a dynamic range that
spans many orders of magnitude between the IOPS transactions for data
access and the bandwidth to storage that is required to store or
retrieve the information. The primary data analysis workflows that drive
these bandwidth requirements can be categorized into four main
activities. These are the contextual environment in which we present
future evolutions for HEP data access. In particular, characteristics
that describe these activities are enumerated for each category:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  \begin{quote}
  \textbf{Reconstruction:} Event reconstruction is CPU limited, due to
  the complexity of detector data and the computational algorithms
  needed to associate and disentangle the data. This is true of
  reconstruction methods used today at the LHC and throughout the HEP
  community. This will continue to be the driving characteristic of
  reconstruction at the HL-LHC and for liquid argon TPC experiments. The
  CPU-bound nature of the algorithms and the ratio of event size to
  network bandwidth make the option of streaming individual events
  records (\emph{e.g.} full collision events or beam spill triggers)
  across wide area networks to compute elements a feasible analysis
  strategy. The requirements of event streaming analysis models can be
  satisfied by a wide range of storage solutions.
  \end{quote}
\item
  \begin{quote}
  \textbf{Simulation:} HEP event simulation strategies focus on
  transforming a small set of key input parameters into event and
  detector response information that is representative of what would be
  seen in the physical detector systems. Simulation processes are
  typically characterized by highly asymmetric Input/Output workflows
  which have very small data ingest (typically just the key input
  parameters) and produce output events which are similar in size and
  structure to real detector data. The output of these simulation
  processes overlap heavily with other data storage and data replication
  processes. The requirements and challenges that this output presents
  are treated in those contexts. The primary drivers that differentiate
  this type of HEP workflow from the other categorizations are the
  increased complexities that the simulation stages often require. These
  simulations often rely on auxiliary event generation mechanisms and
  can, in some HEP domains, require large external data sets
  representing interaction cross sections or detector response functions
  to be available to each instance of the event simulation. This type of
  common input and overlay data may present a challenge in relation to
  its scaling to the runtime environments of future HPC facilities and
  to the design of caching layers at computing site.
  \end{quote}
\item
  \begin{quote}
  \textbf{Analysis:} Analysis level datasets and subsets are typically
  comprised of refined or reduced information that is most pertinent to
  the extraction of physics quantities. These analysis-focused event
  records can often be accepted or rejected through fast pre-filtered
  methods which do not require the reading and retrieval of the full
  event record. This fundamental difference in the way that the analysis
  data is consumed makes it more susceptible to storage bandwidth
  limitations and to IOPs transaction limits that underlying
  technologies can provide. In particular the internal structure of the
  event data can dramatically impact the efficiency of data retrieval
  and favor different access models and protocols (i.e. event
  streaming.) and the actual analysis calculations being performed can
  dramatically affect the CPU to IO ratios. These factors make this
  categorization of data the most challenging to project to future
  storage needs and models.
  \end{quote}
\item
  \begin{quote}
  \textbf{Replication:} Data replication, whether within a site or
  across multiple sites, is performed to enable efficient data access by
  improving data locality (the proximity of the data to compute
  resources) and exploiting the available data bandwidth between the
  storage and the compute locations. Data replication also serves to
  guard against data loss. Data replication places modest requirements
  on the rate of IOP transactions that need to be supported, but drives
  the bandwidth requirements that are needed in enacting site-to-site
  data transfers between WAN endpoints (e.g. 10-100Gbps.) This
  requirement on the available WAN bandwidth provides a corresponding
  performance requirement on on the underlying storage systems to match
  the network performance. Satisfying both requirements is necessary to
  provide an efficient point to point data flow.
  \end{quote}
\end{enumerate}

In addition to the characteristics described above, HEP workflows in
each category exhibit a high degree of parallelism, which allow the work
to map readily into both high throughput computing (HTC) and high
performance computing (HPC) environments. These domains span wide
dynamic ranges in their requirements for concurrent data access. If the
experiments of the HL-LHC, DUNE and other large international projects
are to utilize these facilities, they will need the ability to integrate
with multiple storage technologies. The storage technologies will need
to be tuned or matched to the corresponding computing resources that the
experiments are targeting their workflows to execute on.

These storage models will also need to be designed and optimized for
cost, in the context of the scientific workflows that they are enabling.
Simultaneously, these storage models must remain flexible enough to
support evolving physical storage options similar to the the shifts that
have occurred in recent years between mRAM, SSDs, shingled disks, tapes
and other technologies. ~

\section{Data Organization Models}\label{data-organization-models}

The HEP community has long established patterns that have been used for
data organization and have shaped the prevailing data access and storage
models that are in use today. These established patterns have focused
around sets of event records organized into discrete collections through
traditional file based storage. There are, however, potential gains that
can be made by exploring alternative data organization models and the
corresponding access and storage models that they can enable. These
alternative designs should be evaluated for the HL-LHC era and, if found
to be advantageous, implemented by the community. A shift in designs may
have the potential to reduce storage costs, shift network utilisation,
and result in more efficient use of compute resources.

There are two main avenues that these organizational changes can follow,
each with different impacts on the analysis methodologies. First, we
should reconsider the structure of the data, which can be row oriented,
column oriented, have varying levels of granularity, represent larger or
smaller aggregations of objects etc, boosting the efficiency of
particular applications.

Secondly, we should reconsider the granularity at which data management
systems operate. This is currently aggregations of events (files, or
larger sets such as blocks or datasets), and could shift towards being
the event (or even sub-event), potentially reducing storage volumes. In
that case, other types of ``container'' beyond files could be considered
(e.g. objects).

Design decisions will be subject to competing requirements from data
management and access optimisation, and will have consequences
throughout the system. We have to tackle a global optimization problem,
taking into account all use cases and their respective frequencies to
further optimise application performance for a given resource
investment.

\subsection{Optimizing Data Access}\label{optimizing-data-access}

\subsubsection{Data structures}\label{data-structures}

The structure of data has to be adapted to applications' requirements.
At one limit, the data could be organized simply as events, the other
extreme is to organize the data as columns which represent sub-event
level objects such as particles. Event-based organization favors
applications like event displays while column-based organization favors
large statistical analysis. Our current system of data tiers in files
draws intermediate (and overlapping) rectangles between these two
extremes. Finding the right balance will be an important factor in
efficient data analysis.

While it appears that organizing data in a more fine-grained fashion, as
events or columns, has the potential of opening additional avenues for
cost-savings and flexibility, the implications of such an organizational
shift are still not well understood and need to be investigated.

\subsubsection{Object and Cloud Storage
systems}\label{object-and-cloud-storage-systems}

Decisions on granularity and structure have the potential of broadening
the scope of technologies that can be used to store the data; for
example, exploiting object stores (such as AWS S3) as event stores. In
addition, this would support parallel reading and writing of events as
event objects from applications. This could allow applications to better
scale to large numbers of CPU core where now these applications
concurrency is limited by the requirement to sequentially write events
to files. Currently, the efficiency of of reading and writing events
from object stores is sensitive to the size of the event. The smaller
the event the worse is the efficiency.. Some of this may be mitigated by
creating event bundles (i.e. several events in one object). The
evolution of the ATLAS Event Service may provide some of the answers and
should be supported. As a parallel effort, the ROOT team is also looking
into increasing concurrency within the ROOT framework. This effort
should also be supported as it may have more immediate success in the
near term.

Commercial providers offer innovative storage services (``cloud
storage'') which may fit certain use cases well. The relevance of these
solutions to computing models must be understood, not only through
technical metrics such as performance, but with an understanding of the
underlying cost structure and the risks associated with such
procurements. Reliance on commercially procured cloud storage as a core
component of data organizational models would require a potential
fundamental shift in user policies and discussion with funding agencies.

\subsection{Optimizing Data Management
}\label{optimizing-data-management}

Data Management policy will directly affect data access. For example,
the decision to distribute multiple enriched data samples optimized for
different use cases would improve data access and thus compute
efficiency, at the expense of larger volumes transferred and stored.
Similarly, the unit of data management could be changed, and become the
event, which promises to reduce storage volumes and promote the use of
opportunistic resources through enabling pre-emptibility in jobs. This
would have knock-on effects on data catalogues, data distribution and
storage systems, and would affect data access. Where an experiment
positions itself amongst all these possibilities will be dictated by its
policy. Thus the potential gains in terms of efficient use of resources
should be realised through data management systems which allow this
policy to be expressed and implemented.

\subsubsection{Data catalog models}\label{data-catalog-models}

A key requirement in devising data catalog models is understanding the
required granularity of data to be cataloged and what data for an
experiment needs to be centrally cataloged or tracked, and for what
purpose (accounting, metadata, location). Analysis access to the data,
for instance, may benefit from being below the event level (e.g. physics
object). Any such catalogs may have to be external to the primary data
store (e.g., relational databases that work with metadata in the primary
data store). If cataloging requirements for production and
reconstruction differ considerably from analysis, an approach with
multiple, complementary catalogs can also be considered. The cataloging
schemes chosen will be tightly coupled with the data organization
models. For example, in the case where data organization is object-based
or content-addressable storage is used, the catalogs need only a single
handle to have a fully descriptive location of a given event or other
object being sought.

Our concern is that the size and complexity of a data catalog is
proportional to the granularity of the data organization. Given that the
organization is still in a state of flux (see previous section) it is
difficult to judge the impact of future dataset sizes on data catalog
models other than the simple fact that catalog sizes are likely to grow.
It does seem clear, however, that data catalog models need to evolve in
lockstep with the evolution of data organization models. A large and
diverse set of foundational technologies exist upon which such future
catalogs could be built. An informed technology choice which supports
the anticipated catalog use cases will be crucial in enabling future
scalability.

~

\subsubsection{Data delivery methods}\label{data-delivery-methods}

Data delivery models can become a limiting factor in performance,
especially as they heavily depend on shared resources such storage
systems and the network. Caching has to be considered to avoid data
delivery bottlenecks and unnecessary transfers. At present, HEP data
delivery models are designed solely with the traditional sequential
processing paradigm used in reconstruction in mind. Computing models
where the only form of persistent storage is on tape and all, or nearly
all, disk-based storage is considered as a caching layer should also be
considered, in which case work on optimising workflows for cache
efficiency and exploiting data popularity metrics would be valuable.

~

A useful analogy for data delivery methods might be content delivery
network services (CDNs) such as the one offered by Akamai. Without CDNs,
on-demand streaming of movies over the Internet or even the delivery of
the web sites of major news outlets like CNN or BBC would quickly
collapse. What is the equivalent of content delivery for accessing huge
datasets over distance, and for which a particular analysis might only
require a fraction of the dataset? CDNs use, to great benefit, highly
sophisticated, proprietary naming, caching, and placement techniques
that are customized to particular types of data and what is known about
their access patterns. Similarly intelligent techniques could be used to
place, compose, and deliver scientific data products that minimizes data
movement, latency, and other costs, taking advantage of data semantics
and information about the location of reusable intermediate results.
This is not unlike query planning in relational databases, except for
the scale and the different data type of scientific data: a query
optimizer might be able to recognize that data can be filtered remotely,
or combined from data products cached in the neighborhood instead of
having to fetch everything across continents. New approaches, like Named
Data Networking (NDN), could play a role: by allowing to name, at the
network level, the data you are looking for rather than its location, it
opens the paths for many optimizations behind the scenes based on
factors such as the client application location or the data type.

Currently, new data-caching strategies are being deployed (e.g., XRootD
caching proxy servers, server-less caching, ARC-style caching) and may
provide sufficient information on how such strategies can optimize
mammoth data delivery task for HL-LHC event reconstruction and analysis.
However, without a global view on how these strategies are used, they
will likely have a limited impact on improving data delivery efficiency.

\section{New Analysis Paradigms}\label{new-analysis-paradigms}

Data analysis is currently tied in with ROOT-based formats. In many
currently-used paradigms, physicists consider all events at an
equivalent level of detail and in the format offering the highest level
of detail that needs to be considered in an analysis. ~However, not
every event considered in analysis requires the same level of detail.
One consideration to improve event access throughput is to design event
tiers with different abstractions, and thus data sizes. All events can
be considered at a lighter-weight tier while events of interest only can
be accessed with a more information-rich tier.

For more scalable analysis, another opportunity to evaluate is how much
work can be offloaded to a storage system, for example caching
uncompressed or reordered data for fast access. The idea can be extended
to virtual data and to query interfaces which would perform some of the
transformation logic currently executed on CPU workers. Interactive
querying of large datasets is an active field in the Big Data industry;
examples include Spark-SQL, Impala, Kudu, Hawq, Apache Drill, and Google
Dremel/BigQuery. A key question is about the usability of these
techniques in HEP and we need to assess if our data transformations are
not too complex for the SQL-based query languages used by these
products. We also need to take into account that the adoption of these
techniques, if they prove to be beneficial, would represent a disruptive
change which directly impacts the end user and therefore promoting
acceptance through intermediate solutions would be desirable.

Many analyses may benefit from column-based data access instead of the
more traditional row-based access. Enabling data queries that consider
histogram indexing is another feature that could provide performance
increases in analysis.

When evaluating the potential benefits of moving to new techniques, for
analysis in particular but not only, we should not forget that some
techniques, like machine learning-based techniques, may have
contradictory requirements. On one end, machine learning should
dramatically reduce the pressure on the storage at the cost of using
more CPU but it requires a learning phase which, even if it is shorter
compared to the exploitation phase, can require a significant amount of
resources with a non-typical, iterative, data access pattern. Thus, the
challenge remains to find a good compromise to efficiently support a
large variety of access patterns.

In addition to assessing the techniques that could improve the
performance for the various use cases that we have, it is also important
to ensure that the resources required by these techniques are in line
with expected budget scenarios. Many of these techniques may imply
shifting some of the storage resource usage to CPU usage or vice-versa.
We need to ensure that the global cost will remain similar to what it is
today and that the resulting computing model evolution is compatible
with other constraints.

\section{Performance : Tools and
Metrics}\label{performance-tools-and-metrics}

As the size of available datasets grows, we must allow for the
possibility of increased user-driven selection and dataset generation.
In terms of reducing the strain on data access, techniques such as data
augmentation can prove useful. Cluster-computing frameworks (e.g.,
Spark) can also be leveraged to reduce the time of dataset reduction and
access. It will be necessary to study and classify IO patterns in
applications used in order to understand how data access methods can be
optimized across prevalent access patterns. Pattern data can then be
used in algorithms resulting from active research in computer science
that study payoffs between storage and cpu in various computing models.

A first correlated analysis of the IO patterns of different
computational tasks has already started at the CERN computing center and
within several experiments. In particular the combination of
infrastructure and experiment information (e.g. detailed knowledge of
hardware capabilities and network topology, per job CPU and storage
metrics) may allow us to define more meaningful performance metrics than
the traditionally used raw data rate and CPU utilization which both show
their limitations. These metrics should enable a joint quantitative
comparison of different storage and computing approaches across
individual system boundaries and hence lead to a more effective resource
investment.

~

\section{Common Challenges}\label{common-challenges}

The projected event complexity of data from future LHC runs and from
high resolution liquid argon detectors will require advanced
reconstruction algorithms and analysis tools to understand. The
precursors of these tools, in the form of new machine learning paradigms
and pattern recognition algorithms, already are proving to be drivers
for the CPU needs of the HEP community. As these techniques continue to
grow and blossom, they will place new requirements on the computational
resources that need to be leveraged by all of HEP. The storage systems
that are developed, and the data management techniques that are employed
will need to directly support this wide range of computational
facilities, and will need to be matched to the changes in the
computational work, so as not to impede the improvements that they are
bringing.

Storage will remain one of the visible cost drivers for HEP computing,
but the projected growth and cost of the computational resources needed
to analyze the data are expected to grow faster than the base storage
costs. The combination of storage and analysis computing costs may
restrict scientific output and potential physics reach of the
experiments. There must be R\&D efforts in data management on how to
minimize the impact of the data access and storage model on the overall
cost of doing scientific analysis. This R\&D should include an
optimization of both the capital costs of storage, as well as the
potential impacts the storage systems can have on the CPU requirements
for the experiments and their costs.

The ability to leverage new storage technologies as they become
available into existing data delivery models is a challenge that we must
be prepared for. New storage systems will present new interfaces and new
behaviour. A key decision in their successful exploitation will be
whether to encapsulate this within a more familiar service or whether to
present the new interface directly to applications which will thus have
to be adapted. As discussed in the preceding sections, much of this
change can be aided by active R\&D into our own IO patterns; an approach
which has not yet been adopted widely by the field.

HEP experiments should be prepared to leverage ``tactical storage''.

Storage that becomes most cost-effective as it becomes available (e.g.,
from a cloud provider) and have a data management and provisioning
system that can exploit such resources on short notice. Volatile data
sources would impact many aspects of the system; catalogs, job
brokering, monitoring/alerting, accounting, the applications themselves.

On the hardware side, R\&D is needed in alternative approaches to data
archiving to determine the possible cost/performance tradeoffs.
Currently, tape is extensively used to hold data that cannot be
economically made available online. While the data is still accessible,
it comes with a high latency penalty; limiting possible analysis. We
suggest investigating either separate direct access-based archives (e.g.
disk or optical) or new models that overlay online direct access volumes
with archive space. This is especially relevant when access latency is
proportional to storage density. Either approach would need to also
evaluate reliability risks and the effort needed to provide data
stability.

Cost reductions in maintenance and operation of the storage
infrastructure can be realised through convergence of the major
experiments and resource providers on shared solutions. This does not
necessarily mean promoting a monoculture, as different solutions will be
adapted to certain major classes of use-case, type of site or funding
environment. Indeed, there will always be a judgment to make on the
desirability of using a variety of specialised systems, or abstracting
the commonalities through a more limited but common interface (the SRM
story illustrates this point). Reduced costs and improved sustainability
will be further promoted by extending these concepts of convergence
beyond HEP and into the other large-scale scientific endeavours that
will share the infrastructure in the coming decade. Efforts must be made
as early as possible, during the formative design phases of such
projects, to create the necessary links.

Finally, any and all changes undertaken must not make the ease of access
to data any worse than it is under current computing models. We must
also be prepared to accept the fact that the best possible solution may
require significant changes in the way data is handled and analyzed.
What is clear is that what is being done today will not scale to the
needs of HL LHC.

\section{Research and Development Roadmap and
Goals}\label{research-and-development-roadmap-and-goals}

\protect\hypertarget{_y0jes1qd33m1}{}{}Event-based granularity - Can
this be implemented efficiently, scalably and in a cost-effective manner
for relevant use cases (reco, simulation, event-picking)? Does it
provide an advantage over current file-based granularity?

\begin{itemize}
\item
  \begin{quote}
  \protect\hypertarget{_xbbh0qorjpnn}{}{}Quantify impact on performances
  and resource utilization (storage, network) for the main type of
  access patterns (simulation, reconstruction, analysis)
  \end{quote}
\end{itemize}

\begin{itemize}
\item
  \begin{quote}
  \protect\hypertarget{_ox25y1bzr703}{}{}Assess impact on catalogs and
  data distribution
  \end{quote}
\end{itemize}

\begin{itemize}
\item
  \begin{quote}
  \protect\hypertarget{_hhvd696t4e1b}{}{}Assess whether
  event-granularity makes sense in object stores that tend to require
  large chunks of data for efficiency
  \end{quote}
\item
  \begin{quote}
  \protect\hypertarget{_hut1j5u624ly}{}{}Does it improve recoverability
  from preemption, in particular when using cloud spot resources and/or
  dynamic HPC resources
  \end{quote}
\end{itemize}

\emph{Should be completed by
2020: actions can be done in parallel}.

\protect\hypertarget{_txtjm296wo2a}{}{}Data organization and analysis
technologies - How can we benefit from technologies adopted by the
big-data world?

\begin{itemize}
\item
  \begin{quote}
  \protect\hypertarget{_2uzvesmlywml}{}{}row-based vs. column-based
  organization: impact of storage organization on the performance of
  each kind of access, potential storage format providing good
  performance for both
  \end{quote}
\item
  \begin{quote}
  \protect\hypertarget{_6xwv25emxnty}{}{}Can map-reduce, Spark-like
  analysis and their functional or declarative interfaces be adapted to
  HEP analysis needs?
  \end{quote}
\item
  \begin{quote}
  \protect\hypertarget{_vfqvjf8nrae}{}{}Evaluate just in time
  decompression schemes and mappings onto hardware architectures
  considering the flow of data from spinning disk to memory and
  application
  \end{quote}
\end{itemize}

\emph{Proof-of-concept by 2020, implementation if PoC successful in the following years}

\protect\hypertarget{_49uk6huxen3e}{}{}Data caching - What role will
caching play to use compute resources and what technologies will be
used?

\begin{itemize}
\item
  \begin{quote}
  \protect\hypertarget{_z9tvlvkzecc1}{}{}Quantify benefit of caching for
  main use cases (reconstruction, analysis, simulation)
  \end{quote}
\item
  \begin{quote}
  \protect\hypertarget{_vyj6hh4exb5m}{}{}Assess benefit of caching for
  Machine Learning-based applications, in particular for the learning
  phase
  \end{quote}
\item
  \begin{quote}
  \protect\hypertarget{_ob8u0lfv3fum}{}{}What are the benefits of a CDN
  or NDN approach beyond what HEP is doing today with their data
  delivery systems?
  \end{quote}
\end{itemize}

\emph{First two actions to be completed by 2020, CDN/NDN evaluation more long-term}

\protect\hypertarget{_rjxbgr6m4duh}{}{}Getting the most out of the
storage diversity - how to minimize HEP infrastructure cost by
exploiting varied quality of service from different storage
technologies?

\begin{itemize}
\item
  \begin{quote}
  \protect\hypertarget{_tmwkcbnhjw6g}{}{}Which role for
  opportunistic/tactical storage
  \end{quote}
\item
  \begin{quote}
  \protect\hypertarget{_tpsd2vti9z2o}{}{}Role of archival storage
  solutions
  \end{quote}
\end{itemize}

\emph{Proof-of-concept by 2020, implementation if PoC successful in the following years}

\protect\hypertarget{_kixm6ns2odgg}{}{}Efficiency / latency (global
optimization) - Can we globally optimize data access latency vs
efficiency of CPU use at a sustainable cost?

\begin{itemize}
\item
  \begin{quote}
  \protect\hypertarget{_ykqlttltg8mz}{}{}Impact of concentrating the
  data in fewer, larger locations (``data-lake'' approach)
  \end{quote}
\item
  \begin{quote}
  \protect\hypertarget{_ym4ou9y4kpdu}{}{}Impact of an increased use of
  opportunistic compute resources, further from the data
  \end{quote}
\end{itemize}

\emph{Proof-of-concept by 2020, implementation if PoC successful in the following years}

\section{Conclusions}\label{conclusions}

This document presents several
areas pertaining to data access and management in HEP that will need to
be addressed in the coming decade before the expected volume and
complexity of data becomes prohibitively expensive to store, access, and
analyze. Extending current data handling methods and methodologies will
prove intractable in the HL-LHC and DUNE eras. The development and
adoption of new data analysis paradigms gives the field, as a whole, a
window in which to adapt our data access and data management schemes to
ones which are more suited and optimally matched to a wide range of
advanced computing models and analysis applications. This type of shift
has the potential for enabling new analysis methods and allowing for an
increase in scientific output.

\sloppy
\raggedright
\clearpage
\printbibliography[title={References},heading=bibintoc]

\end{document}

### 2018-03-15 CWP/papers/HSF-CWP-2017-05_analysis/latex/cwp-wg-analysis-interpretation.tex
add abstract, text edits

@@
 
-\abstract{To be written}
+\abstract{At the heart of experimental high energy physics (HEP) is the development of facilities and instrumentation that provide sensitivity to new phenomena.
+Our understanding of Nature at is most fundamental level is advanced through the analysis and interpretation of data from sophisticated detectors in HEP experiments.
+The goal of data analysis systems is to reduce the “time-to-insight” while exploiting the maximum possible scientific potential of the data within the constraints of computing and human resources.
+To achieve this goal, future analysis systems should empower physicists to access the data with a high level of interactivity, reproducibility and throughput capability.
+As part of the HEP Software Foundation’s Community White Paper process, a working group on Data Analysis and Interpretation was formed to assess the challenges and opportunities in HEP data analysis and develop a roadmap for activities in this area over the next decade.
+In this report, the key findings and recommendations of the Data Analysis and Interpretation Working Group are presented.}
 
@@
 
-High Energy Physics (HEP) answers scientific questions by analyzing the data obtained from detectors and sensors of suitably designed experiments, comparing measurements with predictions from models and theories.
-The number of different questions that can be answered by a single experiment range from a few for scientifically very focussed devices to a very diverse and large set of questions for multi-purpose devices.
+High energy physics (HEP) answers scientific questions by analyzing the data obtained from detectors and sensors of suitably designed experiments, comparing measurements with predictions from models and theories.
+The number of different questions that can be answered by a single experiment range from a few for scientifically focused devices to a very diverse and large set of questions for multi-purpose devices.
 In all cases, data are analyzed by groups of researchers of varying sizes, from individual researchers to very large groups of scientists.
 ``Analysis'' as described here includes the reduction of processed detector data, from the trigger and event reconstruction algorithms, to create, analyze and interpret estimates of specific physics observables.
-The baseline analysis model utilizes successive stages of data reduction, finally analyzing a compact dataset with quick real time iteration.
+The baseline analysis model utilizes successive stages of data reduction, finally analyzing a compact dataset with quick real-time iteration.
 Experiments use a series of processing steps to reduce large input datasets down to sizes suitable for laptop-scale analysis.
-The line between managed production-like analysis processing and individual analysis, as well as the balance between harmonized versus individualized analysis data formats differs by experiment based on their needs and optimization level and the maturity of an experiment in its life cycle.
-The current baseline model stems from the goal to exploit the maximum possible scientific potential of the data while minimizing the ``time to insight'' for a large number of different analyses performed in parallel.
+The line between managed production-like analysis processing and individual analysis, as well as the balance between common versus individualized analysis data formats differs by experiment based on their needs, degree of optimization and the maturity of an experiment in its life cycle.
+The current baseline model stems from the goal to exploit the maximum possible scientific potential of the data while minimizing the ``time-to-insight'' for a large number of different analyses performed in parallel.
 It is a complicated product of diverse criteria ranging from computing resources and related innovation to management styles of the experiment collaborations.
@@
 
-The goal for future analysis models is to reduce the ``time to insight'' while exploiting the maximum possible scientific potential of the data within the constraints of computing and human resources.
+The goal for future analysis models is to reduce the ``time-to-insight'' while exploiting the maximum possible scientific potential of the data within the constraints of computing and human resources.
 Analysis models aim towards giving scientists access to the data in the most interactive way possible, and to enable quick turn-around in iteratively learning new insights from the data.
 Many analyses have common deadlines defined by conference schedules.
-The increased analysis activity before these deadlines requires the analysis system to be sufficiently elastic to guarantee a rich set of physics result.
-Also, heterogeneous computing hardware like GPUs, FPGAs, and new memory architectures are emerging and can be exploited to further reduce the ``time to insight''.
+The increased analysis activity before these deadlines requires the analysis system to be sufficiently elastic to guarantee a rich set of physics results.
+Also, heterogeneous computing hardware like GPUs, FPGAs, and new memory architectures are emerging and can be exploited to further reduce the ``time-to-insight''.
+
+As part of the HEP Software Foundation’s Community White Paper~\cite{HSF-CWP-2017-01} process, a working group on Data Analysis and Interpretation was formed to assess the challenges and opportunities in HEP data analysis and develop a roadmap for activities in this area over the next decade.
+In this report, the key findings and recommendations of the Data Analysis and Interpretation Working Group are presented.
 
@@
 Over the past 20 years the HEP community has developed and gravitated around a single analysis ecosystem: ROOT~\cite{Brun1996}.
-This software ecosystem currently both dominates HEP analysis and impacts the full event processing chain, provides foundation libraries, I/O services, etc.
+This software ecosystem currently both dominates HEP analysis and impacts the full event processing chain, provides foundation libraries, I/O services, {\it etc}.
 It is an advantage for the HEP community compared to other science disciplines as ROOT provides an integrated and validated toolkit.
 This lowers the hurdle to start an analysis, enabling the community to talk a common analysis language, as well as making improvements and additions to the toolkit quickly available to the whole community, allowing a large number of analyses to benefit.
-The open source analysis tools landscape used primarily in industry is however evolving very fast and surpasses the HEP efforts in software development effort and community size.
+The open source analysis tools landscape used primarily in industry is however evolving very fast and surpasses HEP in terms of software development effort and community size.
 
@@
 This could include investigating new ways of package management and distribution following open source approaches.
-For the 3-year time frame, we propose to research a comprehensive set of bridges and ferries between the HEP analysis ecosystem and the industry analysis tool landscape, where a bridge enables the ecosystem to use an open source analysis tool and a ferry allows to use data from the ecosystem in the tool and vice versa.
+For the 3-year time frame, we propose to research a comprehensive set of {\it bridges} and {\it ferries} between the HEP analysis ecosystem and the industry analysis tool landscape, where a bridge enables the ecosystem to use an open source analysis tool and a ferry allows to use data from the ecosystem in the tool and vice versa.
 
@@
 But not every component can just be retired if it is not used anymore by most of the ecosystem users.
-Therefore for the 3-year time frame, we propose to evolve our policies on how to adopt new alternatives in the analysis community and how to retire old components of the ecosystem.
+Therefore for the 3-year time frame, we propose to evolve our policies regarding adoption of alternatives in the analysis community and how to retire old components of the ecosystem.
 
@@
 Popular examples of such libraries are Numpy~\cite{5725236-numpy}, pandas~\cite{mckinney-proc-scipy-2010}, and Tensorflow~\cite{abadi2016tensorflow}.
-Experiences from inside and outside the HEP community indicate that the simpler interfaces and code constructs of Python could reduce the complexity of analysis code and therefore contribute to the decrease in ``time to insight'' for HEP analyses.
+Experiences from inside and outside the HEP community indicate that the simpler interfaces and code constructs of Python could reduce the complexity of analysis code and therefore contribute to the decrease in ``time-to-insight'' for HEP analyses.
 For example, data frames in pandas can represent arbitrary dimensions of histograms and offer a rich set of operations on the data, such as the split-apply-combine strategy~\cite{JSSv040i01}.
-The community needs to ensure that senior people are also included in the community language choice and not left behind, and to do so, education and documentation need to be a firm component of making Python a first-class language in HEP.
+The community needs to ensure that scientists at all levels are included in the community language choice and not left behind, and to do so, education and documentation need to be a strong component of making Python a first-class language in HEP.
 We propose for the first year time frame to finalize full support of Python in our ecosystem including long term maintenance.\\ \\
-{\bf Declarative Languages:} Further optimization could be gained by switching to a functional or declarative programming model.
+{\bf Declarative Languages:} Further optimization could be gained by employing a functional or declarative programming model.
 This would allow scientists to express the intended data transformation as a query on data.
-Instead of having to define and control the ``how'', the analyst would declare the ``what'' of his analysis, essentially removing the need to define the event loop in an analysis and leave it to underlying services and systems to optimally iterate over events.
+Instead of an analyzer defining and controlling the ``how'' of the analysis workflow, the analyst would declare the ``what'', essentially removing the need to define the event loop in an analysis and leave it to underlying services and systems to optimally iterate over events.
 Analogously to how programming in C++ abstracts implementation features compared to programming in assembler, it appears that these high-level approaches will allow to abstract from the underlying implementations, allowing the computing systems more freedom in optimizing the utilization of diverse forms of computing resources.
@@
 {\bf File format and compression algorithms:} The baseline approach to analyzing HEP data roughly follows a common pattern: The data are prepared for analysis by completing most of the CPU-intensive processing steps first and then making the data available to the community.
-In the following analysis steps the I/O performance for iterating over the data becomes one of the driving factors in minimizing the ``time to insight'' for analyses.
+In the following analysis steps the I/O performance for iterating over the data becomes one of the driving factors in minimizing the ``time-to-insight'' for analyses.
 There are many file format standards used by a wealth of data analytics tools in other science fields and industry, but the community feels that currently ROOT is best suited to the community's needs.
@@
 {\bf Analysis Facilities:} towards the High-Luminosity LHC (HL-LHC)~\cite{Apollinari:2284929} we envision dedicated data analysis facilities for experimenters, which provide an extendable environment that in turn provides fully functional analysis capabilities.
-Such end-to-end engineered facilities will give users easy and well-maintained access to the HEP analysis software ecosystem and to experiment-specific capabilities (software, documentation, databases, etc).
+Such end-to-end engineered facilities will give users easy and well-maintained access to the HEP analysis software ecosystem and to experiment-specific capabilities (software, documentation, databases, {\it etc}).
 They will support complex ``from beginning to end'' analysis workflows including support for provenance tracking, continuous integration, verification, reproducibility, etc.
@@
 There exist many ways of doing this now, but no commonly accepted and supported way has yet emerged.
-This could be expanded to think about event versus non-event data in general to support use cases from small data volumes (for example cross sections) to large data volumes (BDTs and NNs).
+This could be expanded to consider event versus non-event data in general to support use cases from small data volumes (for example cross sections) to large data volumes (BDTs and NNs).
 We propose R\&D in the area of non-event information handling on the 3-year time scale, which would facilitate analysis at much higher scales than today.
@@
 The most common approach to analyzing data is through a campaign of data reduction and refinement, ultimately producing data frames (flat ntuples) and histograms used to make plots and tables from which physics inference can be made.
-In the following, we are using Atlas and CMS computing model to illustrate data volumes involved in LHC analyses.
+In the following, we are using ATLAS and CMS computing model to illustrate data volumes involved in LHC analyses.
 The centrally-managed data are O(100 kB/event) and are typically too large (e.g.
-O(100 TBs) for 35 $fb^{-1}$ of Run-2 data) to be analyzed by users on their local computing infrastructures like their workgroup servers or laptops.
+O(100 TBs) for 35 fb$^{-1}$ of Run-2 data) to be analyzed by users on their local computing infrastructures like their workgroup servers or laptops.
 An often stated aim of the data reduction steps is to arrive at a dataset that ``can fit on one's laptop'', presumably to facilitate low-latency, high-rate access to a manageable amount of data during the final stages of analysis.
@@
 
-Although Spark itself is the leading contender for this type of analysis, as it has a well developed ecosystem with many third-party tools developed by industry, it is the style of analysis workflow that we are distinguishing here rather than the specific technology present today.
+Although Spark itself is the leading contender for this type of analysis, as it has a well-developed ecosystem with many third-party tools developed by industry, it is the style of analysis workflow that we are distinguishing here rather than the specific technology present today.
 Spark itself is hard to interface with C++, but this might be alleviated by projects such as ROOT's TDataFrame, which presents a Spark-like interface in ROOT, and may allow for more streamlined interoperability.
@@
 {\bf Columnar Partitioning of Analysis Data:} Organizing data to enable fast-access of hierarchical event information (``columnar'' data) is both a challenge and an opportunity.
-Presenting column partitions to analysis systems as the fundamental unit of data management as opposed to files containing collections of events would bring several advantages for HEP end-user analysis (not reconstruction).
+Presenting column partitions to analysis systems as the fundamental unit of data management as opposed to files containing collections of events would bring several advantages for HEP end-user analysis (in contrast to reconstruction jobs).
 These column partitions would become first-class citizens in the same sense that files are today: either as single-column files or more likely as binary blobs in an object store.
@@
 
-A Forum~\cite{LHCInterpretationForum} on the interpretation of the LHC results for Beyond Standard Model (BSM) studies was thus initiated to discuss topics related to the BSM (re)interpretation of LHC data, including the development of the necessary public recasting tools\footnote{``Recasting'' is a term used for this style of reinterpretation~\cite{Cranmer:1299950}.} and related infrastructure, and to provide a platform for a continued interaction between the theorists and the experiments.
+A Forum~\cite{LHCInterpretationForum} on the interpretation of the LHC results for Beyond Standard Model (BSM) studies was thus initiated to discuss topics related to the BSM reinterpretation of LHC data, including the development of the necessary public recasting tools\footnote{``Recasting'' is a term used for this style of reinterpretation~\cite{Cranmer:1299950}.} and related infrastructure, and to provide a platform for a continued interaction between the theorists and the experiments.
 
@@
 \begin{itemize}
-\item Research a comprehensive set of bridges and ferries, where a bridge enables the ecosystem to use an open source analysis tool and a ferry allows to use data from the ecosystem in the tool and vice versa.
+\item Research a comprehensive set of {\it bridges} and {\it ferries}, where a bridge enables the ecosystem to use an open source analysis tool and a ferry allows to use data from the ecosystem in the tool and vice versa.
 \item Analysis facility: conceptualization of an end-to-end low-latency response high-capacity analysis facility with an initial prototype.

### 2018-03-15 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex

@@
 \item
-  reviewing the physics models assumptions, approximations and
+  reviewing the physics models' assumptions, approximations and
   limitations in order to achieve higher precision, and to extend the
@@
 enormous demand for CPU, and detailed detector simulation already
-consumes the plurality of CPU resources of most current HEP experiments.
+consumes a large fraction of CPU resources of most current HEP experiments.
 Some progress has already been made in improving software performance,
@@
 Muon experiments such as Muon g-2 and Mu2e also face large simulation
-challenges, being Geant4 the primary simulation toolkit for
+challenges, as Geant4 is the primary simulation toolkit for
 all stages of these experiments. Because they are searching for extremely rare effects, they

### 2018-03-16 CWP/papers/HSF-CWP-2017-05_analysis/latex/cwp-wg-analysis-interpretation.tex

@@
 \abstract{At the heart of experimental high energy physics (HEP) is the development of facilities and instrumentation that provide sensitivity to new phenomena.
-Our understanding of Nature at is most fundamental level is advanced through the analysis and interpretation of data from sophisticated detectors in HEP experiments.
+Our understanding of Nature at its most fundamental level is advanced through the analysis and interpretation of data from sophisticated detectors in HEP experiments.
 The goal of data analysis systems is to reduce the “time-to-insight” while exploiting the maximum possible scientific potential of the data within the constraints of computing and human resources.

### 2018-03-19 CWP/papers/HSF-CWP-2017-05_analysis/latex/cwp-wg-analysis-interpretation.tex
implement suggestions in prep for archive submission

--- 2018-03-15 CWP/papers/HSF-CWP-2017-05_analysis/latex/cwp-wg-analysis-interpretation.tex
@@
 %% Comment out before final submission
-\usepackage{lineno}  % for line numbering during review
-\linenumbers
+%\usepackage{lineno}  % for line numbering during review
+%\linenumbers
 
 \abstract{At the heart of experimental high energy physics (HEP) is the development of facilities and instrumentation that provide sensitivity to new phenomena.
-Our understanding of Nature at is most fundamental level is advanced through the analysis and interpretation of data from sophisticated detectors in HEP experiments.
-The goal of data analysis systems is to reduce the “time-to-insight” while exploiting the maximum possible scientific potential of the data within the constraints of computing and human resources.
+Our understanding of nature at its most fundamental level is advanced through the analysis and interpretation of data from sophisticated detectors in HEP experiments.
+The goal of data analysis systems is to obtain the maximum possible scientific potential of the data within the constraints of computing and human resources in the least time.
 To achieve this goal, future analysis systems should empower physicists to access the data with a high level of interactivity, reproducibility and throughput capability.
@@
  & & HSF-CWP-2017-05 \\
- & & February 15, 2018 \\ % use \date or hardwire e.g. December 15, 2017
+ & & March 18, 2018 \\ % use \date or hardwire e.g. December 15, 2017
  & & \\
@@
 \author{HEP Software Foundation:}
-\author[a]{Riccardo Maria Bianchi}
-\author[b]{Brian Bockelman}
-\author[c,d]{Nuno Castro}
-\author[e]{Peter Elmer}
-\author[f]{Robert Gardner}
-\author[g]{Maria Girone}
-\author[h,1]{Oliver Gutsche}
-\author[g]{Benedikt Hegner}
-\author[i]{Jos\'{e} M. Hern\'{a}ndez}
-\author[h]{Bodhitha Jayatilaka}
-\author[e]{David Lange}
-\author[j,1]{Mark Neubauer}
-\author[j]{Daniel S. Katz}
-\author[k]{Lukasz Kreczko}
-\author[l]{James Letts}
-\author[m]{Shawn McKee}
-\author[n]{Christoph Paus}
-\author[h]{Kevin Pedro}
-\author[o]{Eduardo Rodrigues}
-\author[k]{Tai Sakuma}
-\author[h]{Elizabeth Sexton-Kennedy}
-\author[o]{Michael D. Sokoloff}
-\author[l]{Frank W\"{u}rthwein}
-\author[p]{Carl Vuosalo}
+\author[a]{Lothar Bauerdick}
+\author[b]{Riccardo Maria Bianchi}
+\author[c]{\\ Brian Bockelman}
+\author[d,e]{Nuno Castro}
+\author[f]{Kyle Cranmer}
+\author[g]{Peter Elmer}
+\author[h]{\\ Robert Gardner}
+\author[i]{Maria Girone}
+\author[a,1]{Oliver Gutsche}
+\author[i]{Benedikt Hegner}
+\author[j]{\\ Jos\'{e} M. Hern\'{a}ndez}
+\author[a]{Bodhitha Jayatilaka}
+\author[g]{David Lange}
+\author[k,1]{Mark Neubauer}
+\author[k]{Daniel S. Katz}
+\author[l]{Lukasz Kreczko}
+\author[m]{James Letts}
+\author[n]{Shawn McKee}
+\author[o]{\\ Christoph Paus}
+\author[a]{Kevin Pedro}
+\author[g]{Jim Pivarski}
+\author[p]{Eduardo Rodrigues}
+\author[l]{Tai Sakuma}
+\author[a]{Elizabeth Sexton-Kennedy}
+\author[p]{Michael D. Sokoloff}
+\author[m]{Frank W\"{u}rthwein}
+\author[q]{\\ Carl Vuosalo}
 
-\affiliation[a]{University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
-\affiliation[b]{University of Nebraska-Lincoln, Lincoln, Nebraska, USA}
-\affiliation[c]{Laboratório de Instrumentação e Física Experimental de Partículas, Lisbon, Portugal}
-\affiliation[d]{University of Minho, Braga, Portugal}
-\affiliation[e]{Princeton University, Princeton, New Jersey, USA}
-\affiliation[f]{University of Chicago, Chicago, Illinois, USA}
-\affiliation[g]{CERN, Geneva, Switzerland}
-\affiliation[h]{Fermi National Accelerator Laboratory, Batavia, Illinois USA}
-\affiliation[i]{Centro de Investigaciones Energéticas, Medioambientales y Tecnológicas, Madrid, Spain}
-\affiliation[j]{University of Illinois at Urbana-Champaign, Urbana, Illinois USA}
-\affiliation[k]{University of Bristol, Bristol, UK}
-\affiliation[l]{University of California at San Diego, La Jolla, California, USA}
-\affiliation[m]{University of Michigan, Ann Arbor, Michigan, USA}
-\affiliation[n]{Massachusetts Institute of Technology, Cambridge, Massachusetts, USA}
-\affiliation[o]{University of Cincinnati, Cincinnati, Ohio, USA}
-\affiliation[p]{University of Wisconsin-Madison, Madison, Wisconsin, USA}
+\affiliation[a]{Fermi National Accelerator Laboratory, Batavia, Illinois USA}
+\affiliation[b]{University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
+\affiliation[c]{University of Nebraska-Lincoln, Lincoln, Nebraska, USA}
+\affiliation[d]{Laboratório de Instrumentação e Física Experimental de Partículas, Lisbon, Portugal}
+\affiliation[e]{University of Minho, Braga, Portugal}
+\affiliation[f]{New York University, New York, New York, USA}
+\affiliation[g]{Princeton University, Princeton, New Jersey, USA}
+\affiliation[h]{University of Chicago, Chicago, Illinois, USA}
+\affiliation[i]{CERN, Geneva, Switzerland}
+\affiliation[j]{Centro de Investigaciones Energéticas, Medioambientales y Tecnológicas, Madrid, Spain}
+\affiliation[k]{University of Illinois at Urbana-Champaign, Urbana, Illinois USA}
+\affiliation[l]{University of Bristol, Bristol, UK}
+\affiliation[m]{University of California at San Diego, La Jolla, California, USA}
+\affiliation[n]{University of Michigan, Ann Arbor, Michigan, USA}
+\affiliation[o]{Massachusetts Institute of Technology, Cambridge, Massachusetts, USA}
+\affiliation[p]{University of Cincinnati, Cincinnati, Ohio, USA}
+\affiliation[q]{University of Wisconsin-Madison, Madison, Wisconsin, USA}
 \affiliation[1]{Paper Editor}
@@
 It is an advantage for the HEP community compared to other science disciplines as ROOT provides an integrated and validated toolkit.
-This lowers the hurdle to start an analysis, enabling the community to talk a common analysis language, as well as making improvements and additions to the toolkit quickly available to the whole community, allowing a large number of analyses to benefit.
-The open source analysis tools landscape used primarily in industry is however evolving very fast and surpasses HEP in terms of software development effort and community size.
+This lowers the bar for starting an analysis, enabling the community to talk a common analysis language, as well as making improvements and additions to the toolkit quickly available to the whole community, allowing a large number of analyses to benefit.
+However, the landscape of open source software (both academic and industrial) for data analysis is rapidly evolving and surpasses HEP in terms of software development effort and community size.
 
@@
 
-\section{Analysis Languages}~\\
-{\bf Python:} Given the large investment in existing physics software and the ROOT-based data models of the experiments, when considering reconstruction and other performance-critical code, the community will continue to look to C++.
-For analysis code, Python has developed into the language of choice in the data science community and also more and more within the HEP community.
+\section{Analysis Languages}
+
+Given the large investment in existing physics software and the ROOT-based data models of the experiments, when considering reconstruction and other performance-critical code, the community will continue to look to C++.
+Developments from the data science community leads us to additional considerations:
+
+\subsection{Python}
+For analysis code, Python has developed into the language of choice in the data science community and also increasingly within the HEP community.
 It is widely used in the data analytics community in other fields of science and industry and is liked for fast development cycles and ease-of-use.
@@
 The community needs to ensure that scientists at all levels are included in the community language choice and not left behind, and to do so, education and documentation need to be a strong component of making Python a first-class language in HEP.
-We propose for the first year time frame to finalize full support of Python in our ecosystem including long term maintenance.\\ \\
-{\bf Declarative Languages:} Further optimization could be gained by employing a functional or declarative programming model.
+We propose for the first year time frame to finalize full support of Python in our ecosystem including long term maintenance.
+
+\subsection{Declarative Languages} 
+Further optimization could be gained by employing a functional or declarative programming model.
 This would allow scientists to express the intended data transformation as a query on data.
@@
 
-\section{Analyzing Data}~\\
-{\bf File format and compression algorithms:} The baseline approach to analyzing HEP data roughly follows a common pattern: The data are prepared for analysis by completing most of the CPU-intensive processing steps first and then making the data available to the community.
-In the following analysis steps the I/O performance for iterating over the data becomes one of the driving factors in minimizing the ``time-to-insight'' for analyses.
+\section{Analyzing Data}
+
+\subsection{File format and compression algorithms} 
+The baseline approach to analyzing HEP data roughly follows a common pattern.
+The data are prepared for analysis by first completing most of the CPU-intensive processing steps and then making the data available to the community.
+In subsequent analysis steps, the I/O performance for iterating over the data becomes one of the driving factors in minimizing the ``time-to-insight'' for analyses.
 There are many file format standards used by a wealth of data analytics tools in other science fields and industry, but the community feels that currently ROOT is best suited to the community's needs.
@@
 To improve on the current state and to enable analysis of much larger datasets than today, R\&D will be needed to continue in file formats, compression algorithms, and new ways of storing and accessing data for analysis.
-\\ \\
-{\bf Analysis Facilities:} towards the High-Luminosity LHC (HL-LHC)~\cite{Apollinari:2284929} we envision dedicated data analysis facilities for experimenters, which provide an extendable environment that in turn provides fully functional analysis capabilities.
+
+\subsection{Analysis Facilities} 
+Towards the era of the High-Luminosity LHC (HL-LHC)~\cite{Apollinari:2284929}, we envision dedicated ``smart'' facilities optimized for data analysis, which provide an extendable environment that in turn provides fully functional analysis capabilities.
 Such end-to-end engineered facilities will give users easy and well-maintained access to the HEP analysis software ecosystem and to experiment-specific capabilities (software, documentation, databases, {\it etc}).
-They will support complex ``from beginning to end'' analysis workflows including support for provenance tracking, continuous integration, verification, reproducibility, etc.
-and will provide low-latency rapid-response computing capabilities with access to complete data sets without the user having to fall back to the high-latency ``batch system style'' analysis.
+They will support complex ``from beginning to end'' analysis workflows including support for provenance tracking, continuous integration, verification, and reproducibility.
+They will provide low-latency, high-throughput computing capabilities with access to complete data sets without the user having to fall back to the high-latency ``batch system style'' analysis.
 ``Primitive'' versions of such analysis facilities are currently provided e.g.
 at CERN, Fermilab and other places.
-However for the HL-LHC, such dedicated analysis facilities would provide a much more end-to-end engineered latency-optimized and maintained environment, that then also would provide focus points for support and training of analysts.\\ \\
-We propose to focus technology developments to support analysis by prototyping such analysis facilities to develop the capabilities and the required ecosystem to support the analysis use cases mentioned in section~\ref{sec:models}.
+However for the HL-LHC, such dedicated analysis facilities would provide a much more end-to-end engineered latency-optimized and maintained environment, that then also would provide focus points for support and training of analysts.
+
+We propose to focus technology developments to support analysis by prototyping such analysis facilities to develop the capabilities and the required ecosystem to support the analysis use cases mentioned in Section~\ref{sec:models}.
 This prototyping work  includes to continue to investigate optimizing the storage systems and used representation of data on disk together, and also to facilitate the utilization of new additional storage layers like SSD storage and NVRAM-like storage, which exhibit different characteristics than the currently dominant spinning disk installations.
@@
 Detailed plans should be worked out regarding the prototyping of the capabilities of analysis facilities, iteratively scaling up both functionalities and capacities, up to the capabilities required for the HL-LHC.
-In years 1 to 3, the conceptualization of an end-to-end low-latency response high-capacity analysis facility should be started, with an initial prototype.
-In years 3 to 5, analysis of data from the 3rd running period of the LHC should be used to evaluate chosen architectures and verify design or provide input for corrective actions.
-At the end of this process starting in Year 5, a blueprint for remaining developments and system design should become available, in time for deployment.\\ \\
-{\bf Non-event data handling:} another area that has not yet received the attention it deserves is the access to non-event data for analysis (cross section values, scale factors, tagging efficiencies, etc.).
+In Years 1 to 3, the conceptualization of an end-to-end low-latency response high-capacity analysis facility should be started, with an initial prototype.
+In Years 3 to 5, analysis of data from the 3rd running period of the LHC should be used to evaluate chosen architectures and verify design or provide input for corrective actions.
+At the end of this process starting in Year 5, a blueprint for remaining developments and system design should become available, in time for deployment.
+\subsection{Non-event data handling} 
+Another area that has not yet received the attention it deserves is the access to non-event data for analysis (cross section values, scale factors, tagging efficiencies, etc.).
 The community feels that the existing capabilities for event data, namely easy storage of event data of all sorts of different content, a similar way of saving and accessing non-event information during the analysis step is needed.
@@
 The centrally-managed data are O(100 kB/event) and are typically too large (e.g.
-O(100 TBs) for 35 fb$^{-1}$ of Run-2 data) to be analyzed by users on their local computing infrastructures like their workgroup servers or laptops.
+O(100 TB) for 35 fb$^{-1}$ of Run-2 data) to be analyzed by users on their local computing infrastructures like their workgroup servers or laptops.
 An often stated aim of the data reduction steps is to arrive at a dataset that ``can fit on one's laptop'', presumably to facilitate low-latency, high-rate access to a manageable amount of data during the final stages of analysis.
@@
 \begin{itemize}
-\item Parallelization is implicit through map/filter/reduce functionals,
+\item Parallelization is implicit through map/filter/reduce functionals or more general workflow patterns and constructs,
 \item Data are abstracted as remote, distributed datasets, rather than files,
@@
 
-~\\Advantages of a query-based analysis include:\\ \\
-{\bf Minimalistic Analysis:} A critical consideration of the Sequential Ntuple Reduction method might reasonably question why analyzers would bother to generate and store intermediate data to get to the same outcomes of interest (histograms, etc).
+\vskip 0.1in\noindent
+Advantages of a query-based analysis include:
+
+\vskip 0.1in\noindent
+\textbf{\textit{Minimalistic Analysis}}.
+A critical consideration of the Sequential Ntuple Reduction method might reasonably question why analyzers would bother to generate and store intermediate data to get to the same outcomes of interest (histograms, etc).
 A more economical approach is to provide only the minimal information -- code providing instructions for selecting the dataset, events of interest, and items to plot.
-\\ \\
-{\bf Democratization of Analyses:} In the Sequential Ntuple Reduction method, as one gets further down the data reduction chain, the user (or small group of users) needs to figure out how to provision and manage the storage required to accommodate this intermediate data which in many cases is accessed with small ($<10^{-4}$) or zero duty cycle.
+
+\vskip 0.1in\noindent
+\textbf{\textit{Democratization of Analyses}}.
+In the Sequential Ntuple Reduction method, as one gets further down the data reduction chain, the user (or small group of users) needs to figure out how to provision and manage the storage required to accommodate this intermediate data which in many cases is accessed with small ($<10^{-4}$) or zero duty cycle.
 For small groups, the resources required (both in personnel and hardware) to execute such a data reduction campaign might be prohibitive in the HL-LHC era, effectively ``pricing them out'' of contributing strongly to analyses -- possibly a lost opportunity for innovation and discovery.
-Removing the requirements on storing intermediate data in the analysis chain would help to ``democratize'' data analysis and streamline the overall analysis workflow.\\ \\
-{\bf Ease of Provenance:} The query-based analysis provides an opportunity for autonomous storage of provenance information, as all processing in an analysis step from ``primary'' analysis-level data to the histograms is contained to a given facility.
-This information can be queried as well, for example.\\ \\
-Key elements of the required infrastructure for a future query-based analysis system are expected to include:\\ \\
-{\bf Sharing resources with traditional systems:} Unlike a traditional batch system, access to this query system is intermittent, so it would be hard to justify allocating exclusive resources to it.
+Removing the requirements on storing intermediate data in the analysis chain would help to ``democratize'' data analysis and streamline the overall analysis workflow.
+
+\vskip 0.1in\noindent
+\textbf{\textit{Ease of Provenance}}.
+The query-based analysis provides an opportunity for autonomous storage of provenance information, as all processing in an analysis step from ``primary'' analysis-level data to the histograms is contained to a given facility.
+This information can be queried as well, for example.
+
+\vskip 0.1in\noindent
+Key elements of the required infrastructure for a future query-based analysis system are expected to include:
+
+\vskip 0.1in\noindent
+\textbf{\textit{Sharing resources with traditional systems}}.
+Unlike a traditional batch system, access to this query system is intermittent, so it would be hard to justify allocating exclusive resources to it.
 Even with a large number of users to smooth out the minute-by-minute load, a query system would have a strong day-night load difference effect, weekday-weekend effect, and pre-conference effect.
 Therefore, the query system must share resources with a traditional batch system (performing event reconstruction for instance).
-Then the query system could elastically scale in response to load, preempting the batch system.\\ \\
-{\bf Columnar Partitioning of Analysis Data:} Organizing data to enable fast-access of hierarchical event information (``columnar'' data) is both a challenge and an opportunity.
+Then the query system could elastically scale in response to load, preempting the batch system.
+
+\vskip 0.1in\noindent
+\textbf{\textit{Columnar Partitioning of Analysis Data}}.
+Organizing data to enable fast-access of hierarchical event information (``columnar'' data) is both a challenge and an opportunity.
 Presenting column partitions to analysis systems as the fundamental unit of data management as opposed to files containing collections of events would bring several advantages for HEP end-user analysis (in contrast to reconstruction jobs).
@@
 We note that columns are already a first-class citizen in the ROOT file system, however, appropriate data management and analysis software that leverages this capability is missing.
-Given a data store full of columns, datasets become loose associations among these columns, with metadata identifying a set of columns as mutually consistent and meaningful for analysis.\\ \\
-{\bf Fast Columnar Data Caching:} Columnar cache is a key feature of the query system, retaining input data between queries, which are usually repeated with small modifications (intentionally as part of a systematics study or unplanned as part of normal data exploration).
+Given a data store full of columns, datasets become loose associations among these columns, with metadata identifying a set of columns as mutually consistent and meaningful for analysis.
+
+\vskip 0.1in\noindent
+\textbf{\textit{Fast Columnar Data Caching}}.
+Columnar cache is a key feature of the query system, retaining input data between queries, which are usually repeated with small modifications (intentionally as part of a systematics study or unplanned as part of normal data exploration).
 RAM cache would be a logical choice, given the speed of RAM memory, but the query system can't hold onto a large block of RAM if it is to share resources with a batch system.
@@
 The query system's cache would therefore need to be implemented in SSD (or some future fast storage, such as X-Point).
-We can assume the query system would have exclusive access to an attached SSD disk, since caching is not required for the batch process.\\ \\
-{\bf Provenance:} The query system should also attach enough provenance to each dataset that it could be recreated from the original source data, which is considered immutable.
+We can assume the query system would have exclusive access to an attached SSD disk, since caching is not required for the batch process.
+
+\vskip 0.1in\noindent
+\textbf{\textit{Provenance}}.
+The query system should also attach enough provenance to each dataset that it could be recreated from the original source data, which is considered immutable.
 User datasets, while they can't be modified in-place, can be deleted, so a dataset's paper trail must extend all the way back to source data.
-This paper trail would take the form of the original dataset name followed by queries for each step of derivation: code and closure data.\\ \\
+This paper trail would take the form of the original dataset name followed by queries for each step of derivation: code and closure data.
+
+\vskip 0.1in\noindent
 Additional details regarding the proposed query-based analysis approach, requirements and columnar management for HEP data can be found in the contributed CWP white paper ``Hierarchical Data Rolled Up in Columnar Arrays''~\cite{Pivarski2017}.
@@
 
-A Forum~\cite{LHCInterpretationForum} on the interpretation of the LHC results for Beyond Standard Model (BSM) studies was thus initiated to discuss topics related to the BSM reinterpretation of LHC data, including the development of the necessary public recasting tools\footnote{``Recasting'' is a term used for this style of reinterpretation~\cite{Cranmer:1299950}.} and related infrastructure, and to provide a platform for a continued interaction between the theorists and the experiments.
+A Forum on the interpretation of the LHC results for Beyond Standard Model (BSM) studies~\cite{LHCInterpretationForum} was thus initiated to discuss topics related to the BSM reinterpretation of LHC data, including the development of the necessary public recasting tools\footnote{``Recasting'' is a term used for this style of reinterpretation~\cite{Cranmer:1299950}.} and related infrastructure, and to provide a platform for a continued interaction between the theorists and the experiments.
 
-The infrastructure needed for analysis reinterpretation is a focal point of other cyber infrastructure components including the INSPIRE literature database~\cite{INSPIRE}, the HEPDatadata repository~\cite{HEPDataRepo} , the CERN Analysis Preservation framework~\cite{CAP}, and the REANA cloud-based workflow execution system~\cite{REANA}.
-Critically, this cyber infrastructure sits at the interface between the theoretical community and various experimental collaborations.
+The infrastructure needed for analysis reinterpretation is a focal point of other cyberinfrastructure components including the INSPIRE literature database~\cite{INSPIRE}, the HEPDatadata repository~\cite{HEPDataRepo} , the CERN Analysis Preservation framework~\cite{CAP}, and the REANA cloud-based workflow execution system~\cite{REANA}.
+Critically, this cyberinfrastructure sits at the interface between the theoretical community and various experimental collaborations.
 As a result, this type of infrastructure is not funded through the experiments and tends to fall through the cracks.
@@
 \begin{itemize}
-\item In year 5, a blueprint for remaining developments and system design should become available, in time for deployment.
+\item In Year 5, a blueprint for remaining developments and system design should become available, in time for deployment.
 \end{itemize}

### 2018-03-21 CWP/papers/HSF-CWP-2017-04_doma/latex/authors.tex
DOMA CWP: add author list

\author[15]{Bianchi, Riccardo Maria}
\author[2]{Bird, Ian}
\author[12]{Bockelman, Brian}
\author[2]{Campana, Simone}
\author[17]{De, Kaushik}
\author[2]{Duellmann, Dirk}
\author[16]{Elmer, Peter}
\author[10]{Gardner, Robert}
\author[14]{Garonne, Vincent}
\author[5]{Grandi, Claudio}
\author[4]{Gutsche, Oliver}
\author[8]{Hanushevsky, Andrew}
\author[4]{Holzman, Burt}
\author[4]{Jayatilaka, Bodhitha}
\author[7]{Jouvin, Michel}
\author[2]{Keeble, Oliver}
\author[1]{Klimentov, Alexei}
\author[3]{Kuznetsov, Valentin}
\author[1]{Lancon, Eric}
\author[2]{Lassnig, Mario}
\author[11]{McKee, Shawn}
\author[6]{Menasce, Dario}
\author[4]{Norman, Andrew}
\author[16]{Pivarski, Jim}
\author[2]{Schulz, Markus}
\author[13]{Severini, Horst}
\author[4]{Sexton-Kennedy, Elizabeth}
\author[4]{Vaandering, Eric}
\author[1]{Wenaus, Torre}
\author[9]{Wuerthwein, Frank}
\affiliation[1]{Physics Department, Brookhaven National Laboratory, Upton, NY, USA}
\affiliation[2]{CERN, Geneva, Switzerland}
\affiliation[3]{Cornell University, Ithaca, NY, USA}
\affiliation[4]{Fermi National Accelerator Laboratory, Batavia, IL, USA}
\affiliation[5]{INFN Sezione di Bologna, Università di Bologna, Bologna, Italy}
\affiliation[6]{INFN Sezione di Milano-Bicocca, Milano, Italy}
\affiliation[7]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\affiliation[8]{SLAC National Accelerator Laboratory, Menlo Park, CA, USA}
\affiliation[9]{University of California, San Diego, La Jolla, CA, USA}
\affiliation[10]{Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
\affiliation[11]{Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
\affiliation[12]{University of Nebraska-Lincoln, Lincoln, NE, USA}
\affiliation[13]{Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman, OK, USA}
\affiliation[14]{Department of Physics, University of Oslo, Oslo, Norway}
\affiliation[15]{Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
\affiliation[16]{Princeton University, Princeton, NJ, USA}
\affiliation[17]{Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}

### 2018-03-21 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
DOMA CWP: add author list
DOMA CWP: fix roadmap formatting

@@
 
-Authors: D. Berzano, R.M. Bianchi, I. Bird, B. Bockelman, K. De, S.
-Campana, D. Duellmann, P.Elmer, R. Gardner, V. Garonne, C. Grandi, O.
-Gutsche, A. Hanushevsky, B. Holzman, B. Jayatilaka, I. Jimenez, M.
-Jouvin, O. Keeble, A. Klimentov, V. Kuznetsov, E. Lançon, M. Lassnig, T.
-Malik, C. Maltzahn, S. McKee, D. Menasce, M. Livny, B. Nitzberg, A.
-Norman, J. Pivarski, M. Schulz, H. Severini, M. Sevilla, L.
-Sexton-Kennedy, E. Vaandering, N. Watkins, T. Wenaus, F. Wuerthwein
+\input{authors.tex}
 
@@
 
-\protect\hypertarget{_y0jes1qd33m1}{}{}Event-based granularity - Can
-this be implemented efficiently, scalably and in a cost-effective manner
+\subsection{Event-based granularity}\label{event-based-granularity}
+Can this be implemented efficiently, scalably and in a cost-effective manner
 for relevant use cases (reco, simulation, event-picking)? Does it
@@
 
-\protect\hypertarget{_txtjm296wo2a}{}{}Data organization and analysis
-technologies - How can we benefit from technologies adopted by the
+\subsection{Data organization and analysis
+technologies}\label{data-organization-and-analysis}
+How can we benefit from technologies adopted by the
 big-data world?
@@
 
-\protect\hypertarget{_49uk6huxen3e}{}{}Data caching - What role will
+\subsection{Data caching}\label{data-caching}
+What role will
 caching play to use compute resources and what technologies will be
@@
 
-\protect\hypertarget{_rjxbgr6m4duh}{}{}Getting the most out of the
-storage diversity - how to minimize HEP infrastructure cost by
+\subsection{Getting the most out of the
+storage diversity}\label{storage-diversity}
+how to minimize HEP infrastructure cost by
 exploiting varied quality of service from different storage
@@
 
-\protect\hypertarget{_kixm6ns2odgg}{}{}Efficiency / latency (global
-optimization) - Can we globally optimize data access latency vs
+\subsection{Efficiency / latency (global
+optimization}\label{efficiency-latency}
+Can we globally optimize data access latency vs
 efficiency of CPU use at a sustainable cost?

### 2018-03-22 CWP/papers/HSF-CWP-2017-01_roadmap/authors/a2tex.py
DOMA CWP: initial version + improved author list generator (#57)
* Initial draft for HSF-CWP-2017-04 (DOMA)

* a2tex.py: add option to generate a JHEP-formatted author list

- Can be used in the title page of the jheppub.sty

* DOMA CWP: add author list

* DOMA CWP: fix roadmap formatting

@@ -105,10 +105,12 @@ def main():
         parser = argparse.ArgumentParser()
         parser.add_argument("--authors", dest="authors", default=AUTHOR_FILE_DEFAULT,
                             help="File containining author list (D: {})".format(AUTHOR_FILE_DEFAULT))
-        parser.add_argument("--footnotes", dest="footnotes", default=FOOTNOTE_FILE_DEFAULT,
-                            help="File containining footnote list (D: {})".format(FOOTNOTE_FILE_DEFAULT))
         parser.add_argument("--affiliations", dest="affiliations", default=AFFILIATION_ADDRESS_FILE_DEFAULT,
                             help="File containining affiliation list (D: {})".format(AFFILIATION_ADDRESS_FILE_DEFAULT))
+        parser.add_argument("--footnotes", dest="footnotes", default=FOOTNOTE_FILE_DEFAULT,
+                            help="File containining footnote list (D: {})".format(FOOTNOTE_FILE_DEFAULT))
+        parser.add_argument("--jhep", dest="jhep", action="store_true", default=False,
+                            help="Produce an author list that can be used in jheppub.sty")
         parser.add_argument("--output", dest="output_file", default=LATEX_AUTHOR_FILE_DEFAULT,
                             help="Output Latex file (D: {})".format(LATEX_AUTHOR_FILE_DEFAULT))
         parser.add_argument("--arxivoutput", dest="arxiv_output_file", default=ARXIV_AUTHOR_FILE_DEFAULT,
@@ -176,30 +178,43 @@ def main():
     with open(options.output_file, "w", encoding="utf-8") as output:
         for author in sorted(author_list, key=lambda author: author.surname):
             affiliation_list = ",".join([ affiliation_map[affiliation].mark for affiliation in author.affiliations ])
-            if author.footnotes:
-                footnote_str = ",".join( [ footnote_list[int(id)-1].mark for id in author.footnotes ])
-                affiliation_list = affiliation_list + "," + footnote_str
-            if author == author_list[-1]:
-                eol_str = ""
+
+            if options.jhep:
+                print ("\\author[{}]{{{}, {}}}".format(affiliation_list,
+                                                       author.surname,
+                                                       author.forename),
+                       file=output)
             else:
-                eol_str = ";"
-            print ("{}, {}$^{{{}}}${}".format(author.surname,
-                                               author.forename,
-                                               affiliation_list,
-                                               eol_str),
-                   file=output)
[…]

### 2018-03-22 CWP/papers/HSF-CWP-2017-05_analysis/latex/cwp-wg-analysis-interpretation.tex
implement suggestions to Analysis WG paper in preparation for archive submission (#55)
* fix bug in relative path to bib

* updates to affiliations

* add abstract, text edits

* correct typo in abstract

* implement suggestions in prep for archive submission

* add author and some minor fixes

@@
 Our understanding of nature at its most fundamental level is advanced through the analysis and interpretation of data from sophisticated detectors in HEP experiments.
-The goal of data analysis systems is to obtain the maximum possible scientific potential of the data within the constraints of computing and human resources in the least time.
+The goal of data analysis systems is to realize the maximum possible scientific potential of the data within the constraints of computing and human resources in the least time.
 To achieve this goal, future analysis systems should empower physicists to access the data with a high level of interactivity, reproducibility and throughput capability.
@@
 \end{tabular*}
-\vspace{2.0cm}
+\vspace{1.0cm}
 
@@
 \author[p]{Michael D. Sokoloff}
-\author[m]{Frank W\"{u}rthwein}
-\author[q]{\\ Carl Vuosalo}
+\author[q]{Carl Vuosalo}
+\author[m]{\\ Frank W\"{u}rthwein}
+\author[r]{Gordon Watts}
 
@@
 \affiliation[q]{University of Wisconsin-Madison, Madison, Wisconsin, USA}
+\affiliation[r]{University of Washington, Seattle, Washington, USA}
 \affiliation[1]{Paper Editor}
@@
 
-The goal for future analysis models is to reduce the ``time-to-insight'' while exploiting the maximum possible scientific potential of the data within the constraints of computing and human resources.
+The goal of data analysis systems is to realize the maximum possible scientific potential of the data within the constraints of computing and human resources in the least time.
 Analysis models aim towards giving scientists access to the data in the most interactive way possible, and to enable quick turn-around in iteratively learning new insights from the data.
@@
 
-Over the past 20 years the HEP community has developed and gravitated around a single analysis ecosystem: ROOT~\cite{Brun1996}.
-This software ecosystem currently both dominates HEP analysis and impacts the full event processing chain, provides foundation libraries, I/O services, {\it etc}.
-It is an advantage for the HEP community compared to other science disciplines as ROOT provides an integrated and validated toolkit.
-This lowers the bar for starting an analysis, enabling the community to talk a common analysis language, as well as making improvements and additions to the toolkit quickly available to the whole community, allowing a large number of analyses to benefit.
-However, the landscape of open source software (both academic and industrial) for data analysis is rapidly evolving and surpasses HEP in terms of software development effort and community size.
+Over the past 20 years, the HEP community has developed and gravitated around a single analysis ecosystem: ROOT~\cite{Brun1996}.
+This ecosystem currently dominates HEP analysis and impacts the full event processing chain, providing key elements such as foundation libraries and I/O services.
+It has been empowering for the HEP community as ROOT provides an integrated and validated toolkit which lowers the bar for analysis, enables the community to speak in a common analysis language and facilitates the generation of improvements and additions to the toolkit that are rapidly propogated to the whole community to benefit a large number of analyses.
+On the other hand, the landscape of open source software (both academic and industrial) for data analysis is rapidly evolving and surpasses HEP in terms of development effort and community size, compelling a community re-evaluation of this strategy.
 
 The emergence and abundance of alternative and new analysis components and techniques coming from industry open source projects is a challenge for the HEP analysis software ecosystem.
-The community is very interested in using these new techniques and technologies and would like to use these together with established components of the ecosystem.
+The HEP community is very interested in using these new techniques and technologies, especially together with established components of the ecosystem where possible.
 It should also be possible to interchange old components with new open source components.
@@
 This could include investigating new ways of package management and distribution following open source approaches.
-For the 3-year time frame, we propose to research a comprehensive set of {\it bridges} and {\it ferries} between the HEP analysis ecosystem and the industry analysis tool landscape, where a bridge enables the ecosystem to use an open source analysis tool and a ferry allows to use data from the ecosystem in the tool and vice versa.
+On a 3-year time frame, we propose to research a comprehensive set of {\it bridges} and {\it ferries} between the HEP analysis ecosystem and the industry analysis tool landscape, where a bridge enables the ecosystem to use an open source analysis tool and a ferry allows to use data from the ecosystem in the tool and vice versa.
 
@@
 
-Given the large investment in existing physics software and the ROOT-based data models of the experiments, when considering reconstruction and other performance-critical code, the community will continue to look to C++.
-Developments from the data science community leads us to additional considerations:
+Given the large prior investment in software and the predominance of ROOT-based data analysis models in HEP experiments, when considering reconstruction and other performance-critical code, the community will continue to look to C++.
+Recent developments within the data science community, however, lead us to additional considerations regarding software languages in support of data analysis.
 
@@
 Instead of an analyzer defining and controlling the ``how'' of the analysis workflow, the analyst would declare the ``what'', essentially removing the need to define the event loop in an analysis and leave it to underlying services and systems to optimally iterate over events.
-Analogously to how programming in C++ abstracts implementation features compared to programming in assembler, it appears that these high-level approaches will allow to abstract from the underlying implementations, allowing the computing systems more freedom in optimizing the utilization of diverse forms of computing resources.
+In analogy to how programming in C++ abstracts implementation features compared to programming in assembler, it appears that these high-level approaches will allow one to abstract from the underlying implementations, providing computing systems more freedom to optimize use of the underlying resources.
 We propose to assess the advantages of the usage of a declarative programming model for data analysis over cognitive overhead for newcomers.
@@
 Such demonstrators should not only help to assess the viability of the programming model but also attempt to fully leverage hardware resources at disposal, heterogeneous or not, both locally and in a distributed fashion.
-We propose on the 3-year time frame to conclude on the already ongoing R\&D projects (for example TDataFrame in ROOT) and to follow up with additional R\&D projects to develop a prototype functional or declarative programming language model.
+We propose on the 3-year time frame to conclude on the already ongoing R\&D projects (e.g.
+TDataFrame in ROOT) and to follow up with additional R\&D projects to develop a prototype functional or declarative programming language model.
 
@@
 The baseline approach to analyzing HEP data roughly follows a common pattern.
-The data are prepared for analysis by first completing most of the CPU-intensive processing steps and then making the data available to the community.
+The data are prepared for analysis by first completing most of the CPU-intensive processing steps and then making the resulting data available to the community.
 In subsequent analysis steps, the I/O performance for iterating over the data becomes one of the driving factors in minimizing the ``time-to-insight'' for analyses.
-There are many file format standards used by a wealth of data analytics tools in other science fields and industry, but the community feels that currently ROOT is best suited to the community's needs.
+There are many file format standards used by a wealth of data analytics tools in other science fields and industry which should be explored for use in HEP analyses, but it is generally felt that currently ROOT I/O is a good fit to the community's needs.
 Disk space is usually the key concern of the experiment computing models as disk is the most expensive hardware component.

### 2018-03-22 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
DOMA CWP: reintegrate lost modifications from Bo

@@
 
-\subsection{Event-based granularity}\label{event-based-granularity}
-Can this be implemented efficiently, scalably and in a cost-effective manner
-for relevant use cases (reco, simulation, event-picking)? Does it
-provide an advantage over current file-based granularity?
+\subsection{Sub-file granularity}\label{sub-file-granularity}
+Sub-file granularity ({\it e.g.} event-based granularity) should be studied to see
+whether it can be implemented efficiently and in a scalable, cost-effective manner for
+all use cases that rely on event selection and whether it offers an advantage over
+the current paradigm of file-based granularity.
 
@@
 
-\emph{Should be completed by
-2020: actions can be done in parallel}.
+\emph{The previous tasks should be completed by 2020 (and can be performed in parallel).}
 
@@
 technologies}\label{data-organization-and-analysis}
-How can we benefit from technologies adopted by the
-big-data world?
+Data organization and analysis technologies in use by other big data users should be
+studied to glean whether we can benefit from them.
 
@@
 
-\emph{Proof-of-concept by 2020, implementation if PoC successful in the following years}
+\emph{The proposed proof-of-concept involves the previous tasks that should be completed by 2020. If successful,
+implementation as a production system will be done in the following years.}
 
 \subsection{Data caching}\label{data-caching}
-What role will
-caching play to use compute resources and what technologies will be
-used?
+Investigate the role that data placement optimizations, such as caching,
+can play in optimizing the use of computing resources as well as the
+technologies that can be used for this.
 
@@
 
-\emph{First two actions to be completed by 2020, CDN/NDN evaluation more long-term}
+\emph{The first two actions should be completed by 2020 but CDN/NDN
+evaluation is more long-term}
 
@@
 storage diversity}\label{storage-diversity}
-how to minimize HEP infrastructure cost by
+How to minimize HEP infrastructure cost by
 exploiting varied quality of service from different storage

### 2018-03-22 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
DOMA CWP: LaTeX cleanup

@@
 \def\labelenumi{\arabic{enumi}.}
-\item
-  \begin{quote}
-  \textbf{Big Data:} The HL-LHC era will significantly increase both the
+\item \textbf{Big Data:} The HL-LHC era will significantly increase both the
   date rate and the data volume. The computing systems will need to
@@
   storage technology limitations.
-  \end{quote}
-\item
-  \begin{quote}
-  \textbf{Dynamic Distributed Computing:} In addition, the significantly
+\item \textbf{Dynamic Distributed Computing:} In addition, the significantly
   increased computational requirements for the HL-LHC era will also
@@
   DOMA systems.
-  \end{quote}
-\item
-  \begin{quote}
-  \textbf{New Applications:} New applications, such as machine learning
+\item \textbf{New Applications:} New applications, such as machine learning
   training or high-rate data-query systems for analysis, will likely be
@@
   placing further requirements on data.
-  \end{quote}
 \end{enumerate}
@@
 \begin{itemize}
-\item
-  \begin{quote}
-  \textbf{Data Organization:} This is essentially how data is structured
+\item \textbf{Data Organization:} This is essentially how data is structured
   as it is written. Most data is written in flat files, in ROOT format,
@@
   applications.
-  \end{quote}
-\item
-  \begin{quote}
-  \textbf{Data Management:} The key challenge here was the transition to
+\item \textbf{Data Management:} The key challenge here was the transition to
   the use of distributed computing in the form of the grid. The
@@
   management systems does this on behalf of the applications.
-  \end{quote}
-\item
-  \begin{quote}
-  \textbf{Data Access:} Various protocols are used for direct reads
+\item \textbf{Data Access:} Various protocols are used for direct reads
   (rfio, dcap, xrootd, etc.) with a given computer center and/or
@@
   transfers between sites.
-  \end{quote}
 \end{itemize}
@@
 \def\labelenumi{\arabic{enumi})}
-\item
-  \begin{quote}
-  \textbf{Reconstruction:} Event reconstruction is CPU limited, due to
+\item \textbf{Reconstruction:} Event reconstruction is CPU limited, due to
   the complexity of detector data and the computational algorithms
@@
   satisfied by a wide range of storage solutions.
-  \end{quote}
-\item
-  \begin{quote}
-  \textbf{Simulation:} HEP event simulation strategies focus on
+\item \textbf{Simulation:} HEP event simulation strategies focus on
   transforming a small set of key input parameters into event and
@@
   to the design of caching layers at computing site.
-  \end{quote}
-\item
-  \begin{quote}
-  \textbf{Analysis:} Analysis level datasets and subsets are typically
+\item \textbf{Analysis:} Analysis level datasets and subsets are typically
   comprised of refined or reduced information that is most pertinent to
@@
   storage needs and models.
-  \end{quote}
-\item
-  \begin{quote}
-  \textbf{Replication:} Data replication, whether within a site or
+\item \textbf{Replication:} Data replication, whether within a site or
   across multiple sites, is performed to enable efficient data access by
@@
   provide an efficient point to point data flow.
-  \end{quote}
 \end{enumerate}
@@
 \begin{itemize}
-\item
-  \begin{quote}
-  \protect\hypertarget{_xbbh0qorjpnn}{}{}Quantify impact on performances
+\item Quantify impact on performances
   and resource utilization (storage, network) for the main type of
   access patterns (simulation, reconstruction, analysis)
-  \end{quote}
-\end{itemize}
-
-\begin{itemize}
-\item
-  \begin{quote}
-  \protect\hypertarget{_ox25y1bzr703}{}{}Assess impact on catalogs and
+\item Assess impact on catalogs and
   data distribution
-  \end{quote}
-\end{itemize}
-
-\begin{itemize}
-\item
-  \begin{quote}
-  \protect\hypertarget{_hhvd696t4e1b}{}{}Assess whether
+\item Assess whether
   event-granularity makes sense in object stores that tend to require
   large chunks of data for efficiency
-  \end{quote}
-\item
-  \begin{quote}
-  \protect\hypertarget{_hut1j5u624ly}{}{}Does it improve recoverability
+\item Does it improve recoverability
   from preemption, in particular when using cloud spot resources and/or
   dynamic HPC resources
-  \end{quote}
 \end{itemize}
@@
 \begin{itemize}
-\item
-  \begin{quote}
-  \protect\hypertarget{_2uzvesmlywml}{}{}row-based vs. column-based
+\item Row-based vs. column-based
   organization: impact of storage organization on the performance of
@@
   performance for both
-  \end{quote}
-\item
-  \begin{quote}
-  \protect\hypertarget{_6xwv25emxnty}{}{}Can map-reduce, Spark-like
+\item Can map-reduce, Spark-like
   analysis and their functional or declarative interfaces be adapted to
   HEP analysis needs?
-  \end{quote}
-\item
-  \begin{quote}
-  \protect\hypertarget{_vfqvjf8nrae}{}{}Evaluate just in time
+\item Evaluate just in time
   decompression schemes and mappings onto hardware architectures
@@
   application
-  \end{quote}
 \end{itemize}
@@
 \begin{itemize}
-\item
-  \begin{quote}
-  \protect\hypertarget{_z9tvlvkzecc1}{}{}Quantify benefit of caching for
+\item Quantify benefit of caching for
   main use cases (reconstruction, analysis, simulation)
-  \end{quote}
-\item
-  \begin{quote}
-  \protect\hypertarget{_vyj6hh4exb5m}{}{}Assess benefit of caching for
+\item Assess benefit of caching for
   Machine Learning-based applications, in particular for the learning
   phase
-  \end{quote}
-\item
-  \begin{quote}
-  \protect\hypertarget{_ob8u0lfv3fum}{}{}What are the benefits of a CDN
+\item What are the benefits of a CDN
   or NDN approach beyond what HEP is doing today with their data
   delivery systems?
-  \end{quote}
 \end{itemize}
@@
 \begin{itemize}
-\item
-  \begin{quote}
-  \protect\hypertarget{_tmwkcbnhjw6g}{}{}Which role for
+\item Which role for
   opportunistic/tactical storage
-  \end{quote}
-\item
-  \begin{quote}
-  \protect\hypertarget{_tpsd2vti9z2o}{}{}Role of archival storage
+\item Role of archival storage
   solutions
-  \end{quote}
 \end{itemize}
@@
 \begin{itemize}
-\item
-  \begin{quote}
-  \protect\hypertarget{_ykqlttltg8mz}{}{}Impact of concentrating the
+\item Impact of concentrating the
   data in fewer, larger locations (``data-lake'' approach)
-  \end{quote}
-\item
-  \begin{quote}
-  \protect\hypertarget{_ym4ou9y4kpdu}{}{}Impact of an increased use of
+\item Impact of an increased use of
   opportunistic compute resources, further from the data
-  \end{quote}
 \end{itemize}

### 2018-03-23 CWP/papers/HSF-CWP-2017-04_doma/latex/authors.tex
CWP DOMA: more fixes to the roadmap section (#60)
* CWP DOMA: more fixes to the roadmap section

* CWP DOMA: missing authors added

- Authors from Google Doc not yet imported
CWP authors: add M. Livny
- Also fix affiliation in DOMA paper

@@
-\author[15]{Bianchi, Riccardo Maria}
+\author[2]{Berzano, Dario}
+\author[16]{Bianchi, Riccardo Maria}
 \author[2]{Bird, Ian}
-\author[12]{Bockelman, Brian}
+\author[13]{Bockelman, Brian}
 \author[2]{Campana, Simone}
-\author[17]{De, Kaushik}
+\author[18]{De, Kaushik}
 \author[2]{Duellmann, Dirk}
-\author[16]{Elmer, Peter}
-\author[10]{Gardner, Robert}
-\author[14]{Garonne, Vincent}
-\author[5]{Grandi, Claudio}
-\author[4]{Gutsche, Oliver}
-\author[8]{Hanushevsky, Andrew}
-\author[4]{Holzman, Burt}
-\author[4]{Jayatilaka, Bodhitha}
-\author[7]{Jouvin, Michel}
+\author[17]{Elmer, Peter}
+\author[11]{Gardner, Robert}
+\author[15]{Garonne, Vincent}
+\author[6]{Grandi, Claudio}
+\author[5]{Gutsche, Oliver}
+\author[9]{Hanushevsky, Andrew}
+\author[5]{Holzman, Burt}
+\author[5]{Jayatilaka, Bodhitha}
+\author[10]{Jimenez, Ivo}
+\author[8]{Jouvin, Michel}
 \author[2]{Keeble, Oliver}
 \author[1]{Klimentov, Alexei}
-\author[3]{Kuznetsov, Valentin}
+\author[4]{Kuznetsov, Valentin}
 \author[1]{Lancon, Eric}
 \author[2]{Lassnig, Mario}
-\author[11]{McKee, Shawn}
-\author[6]{Menasce, Dario}
-\author[4]{Norman, Andrew}
-\author[16]{Pivarski, Jim}
+\author[3]{Livny, Miron}
+\author[10]{Maltzahn, Carlos}
+\author[12]{McKee, Shawn}
+\author[7]{Menasce, Dario}
+\author[5]{Norman, Andrew}
+\author[17]{Pivarski, Jim}
 \author[2]{Schulz, Markus}
-\author[13]{Severini, Horst}
-\author[4]{Sexton-Kennedy, Elizabeth}
-\author[4]{Vaandering, Eric}
+\author[14]{Severini, Horst}
+\author[10]{Sevilla, Michael}
+\author[5]{Sexton-Kennedy, Elizabeth}
+\author[5]{Vaandering, Eric}
+\author[10]{Watkins, Noah}
 \author[1]{Wenaus, Torre}
-\author[9]{Wuerthwein, Frank}
+\author[10]{Wuerthwein, Frank}
 \affiliation[1]{Physics Department, Brookhaven National Laboratory, Upton, NY, USA}
 \affiliation[2]{CERN, Geneva, Switzerland}
-\affiliation[3]{Cornell University, Ithaca, NY, USA}
-\affiliation[4]{Fermi National Accelerator Laboratory, Batavia, IL, USA}
-\affiliation[5]{INFN Sezione di Bologna, Università di Bologna, Bologna, Italy}
-\affiliation[6]{INFN Sezione di Milano-Bicocca, Milano, Italy}
-\affiliation[7]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
-\affiliation[8]{SLAC National Accelerator Laboratory, Menlo Park, CA, USA}
-\affiliation[9]{University of California, San Diego, La Jolla, CA, USA}
-\affiliation[10]{Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
-\affiliation[11]{Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
-\affiliation[12]{University of Nebraska-Lincoln, Lincoln, NE, USA}
-\affiliation[13]{Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman, OK, USA}
-\affiliation[14]{Department of Physics, University of Oslo, Oslo, Norway}
-\affiliation[15]{Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
-\affiliation[16]{Princeton University, Princeton, NJ, USA}
-\affiliation[17]{Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
+\affiliation[3]{Center for High Throughput  Computing, Computer Sciences Department, University of Wisconsin-Madison, Madison, WI, USA}
+\affiliation[4]{Cornell University, Ithaca, NY, USA}
+\affiliation[5]{Fermi National Accelerator Laboratory, Batavia, IL, USA}
+\affiliation[6]{INFN Sezione di Bologna, Università di Bologna, Bologna, Italy}
+\affiliation[7]{INFN Sezione di Milano-Bicocca, Milano, Italy}
+\affiliation[8]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
+\affiliation[9]{SLAC National Accelerator Laboratory, Menlo Park, CA, USA}
+\affiliation[10]{University of California, San Diego, La Jolla, CA, USA}
+\affiliation[11]{Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
+\affiliation[12]{Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
+\affiliation[13]{University of Nebraska-Lincoln, Lincoln, NE, USA}
+\affiliation[14]{Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman, OK, USA}
+\affiliation[15]{Department of Physics, University of Oslo, Oslo, Norway}
+\affiliation[16]{Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
+\affiliation[17]{Princeton University, Princeton, NJ, USA}
+\affiliation[18]{Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}

### 2018-03-25 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex

@@
 \subsection{Sub-file granularity}\label{sub-file-granularity}
-Sub-file granularity ({\it e.g.} event-based granularity) should be studied to see
-whether it can be implemented efficiently and in a scalable, cost-effective manner for
-all use cases that rely on event selection and whether it offers an advantage over
-the current paradigm of file-based granularity.
+Sub-file granularity ({\it e.g.} event-based granularity) should be studied
+to see whether it can be implemented efficiently and in a scalable, cost-effective
+manner for all use cases that rely on event selection and whether it offers an advantage
+over the current paradigm of file-based granularity. Proposed actions are:
 
@@
   and resource utilization (storage, network) for the main type of
-  access patterns (simulation, reconstruction, analysis)
-\item Assess impact on catalogs and
-  data distribution
+  access patterns ({\it i.e.} simulation, reconstruction, analysis).
+\item Assess the impact on catalogs and
+  data distribution.
 \item Assess whether
   event-granularity makes sense in object stores that tend to require
-  large chunks of data for efficiency
-\item Does it improve recoverability
+  large chunks of data for efficiency.
+\item Test for improvement in recoverability
   from preemption, in particular when using cloud spot resources and/or
-  dynamic HPC resources
+  dynamic HPC resources.
 \end{itemize}
@@
 technologies}\label{data-organization-and-analysis}
-Data organization and analysis technologies in use by other big data users should be
-studied to glean whether we can benefit from them.
+Data organization and analysis technologies in use by other big data users should be studied
+to glean whether we can benefit from them. Proposed actions are:
 
 \begin{itemize}
-\item Row-based vs. column-based
-  organization: impact of storage organization on the performance of
-  each kind of access, potential storage format providing good
-  performance for both
-\item Can map-reduce, Spark-like
-  analysis and their functional or declarative interfaces be adapted to
-  HEP analysis needs?
-\item Evaluate just in time
+\item Evaluate row-based versus column-based
+  data organization on the performance of each type of access used.
+\item Investigate data storage and access solutions that support the use of map-reduce or
+Spark-like analysis tools and their adaptability to HEP analysis needs.
+\item Evaluate just-in-time
   decompression schemes and mappings onto hardware architectures
   considering the flow of data from spinning disk to memory and
-  application
+  application.
 \end{itemize}
 
-\emph{The proposed proof-of-concept involves the previous tasks that should be completed by 2020. If successful,
-implementation as a production system will be done in the following years.}
+\emph{The proposed proof-of-concept involves the previous tasks that should be completed by 2020.
+If successful, implementation as a production system will be done in the following years.}
 
 \subsection{Data caching}\label{data-caching}
-Investigate the role that data placement optimizations, such as caching,
-can play in optimizing the use of computing resources as well as the
-technologies that can be used for this.
+Investigate the role that data placement optimizations, such as caching, can play in optimizing
+the use of computing resources as well as the technologies that can be used for this. Proposed actions are:
 
 \begin{itemize}
-\item Quantify benefit of caching for
-  main use cases (reconstruction, analysis, simulation)
-\item Assess benefit of caching for
+\item Quantify the benefit of caching for reconstruction, analysis, and simulation.
+\item Assess the benefit of caching for
   Machine Learning-based applications, in particular for the learning
-  phase
-\item What are the benefits of a CDN
-  or NDN approach beyond what HEP is doing today with their data
-  delivery systems?
+  phase.
+\item Evaluate the benefits of using different approaches to data delivery from what is currently in use in HEP.
+Namely, Content Delivery Networks (CDN) and Named Data Networking (NDN) should be studied.
 \end{itemize}
 
-\emph{The first two actions should be completed by 2020 but CDN/NDN
-evaluation is more long-term}
+-\emph{The first two actions should be completed by 2020 but CDN/NDN
+-evaluation is more long-term}
 
@@
 storage diversity}\label{storage-diversity}
-How to minimize HEP infrastructure cost by
-exploiting varied quality of service from different storage
-technologies?
+
+The landscape of increasing diversity in storage service offerings and technologies should be studied and
+exploited in order to reduce HEP infrastructure costs. The proof-of-concept phase involves the following tasks:
 
 \begin{itemize}
-\item Which role for
-  opportunistic/tactical storage
-\item Role of archival storage
-  solutions
+\item Understand what role tactical or opportunistic (potentially short-term) storage can play in HEP.
+\item Re-evaluate the role of archival storage solutions for HEP.
 \end{itemize}
 
-\emph{Proof-of-concept by 2020, implementation if PoC successful in the following years}
+\emph{The  proof-of-concept should be completed by 2020 and, if successful, implementation in
+production will be done in the following years.}
 
-\subsection{Efficiency / latency (global
-optimization}\label{efficiency-latency}
-Can we globally optimize data access latency vs
-efficiency of CPU use at a sustainable cost?
+\subsection{Global optimization of efficiency and latency}\label{efficiency-latency}
+The inherent trade-offs between data access latency and efficiency of CPU use need to be studied and optimized
+on a global scale across the HEP infrastructure. The proof-of-concept phase involves the following tasks::
 
 \begin{itemize}
-\item Impact of concentrating the
-  data in fewer, larger locations (``data-lake'' approach)
-\item Impact of an increased use of
-  opportunistic compute resources, further from the data
+\item Understand the impact of concentrating the
+  data in fewer, larger locations (the ``data-lake'' approach).
+\item Understand the impact of an increased use of
+  opportunistic compute resources, specifically those further from where data are stored.
 \end{itemize}
 
-\emph{Proof-of-concept by 2020, implementation if PoC successful in the following years}
+\emph{The  proof-of-concept should be completed by 2020 and, if successful, implementation in
+production will be done in the following years.}
 

### 2018-03-28 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
CWP DOMA: language fixes and other cleanup by Bo
- Imported from Overleaf on April 5, 10 am CET

@@
 
-\abstract{To be done.}
+\abstract{Without significant changes to data organization, management, and access (DOMA), HEP experiments will find scientific output limited by how fast data can be accessed and digested by computational resources.
+In this white paper we discuss challenges in DOMA that HEP experiments such as the HL-LHC will face as well as potential ways to address them.
+A research and development timeline to assess these changes is also proposed.}
 
@@
  & & HSF-CWP-2017-04 \\
- & & March 11, 2018 \\ % use \date or hardwire e.g. December 15, 2017
+ & & April 3, 2018 \\ % use \date or hardwire e.g. December 15, 2017
  & & \\
@@
 
-The coming decade will bring HEP to the exabyte scale. The expected data
+The coming decade will bring HEP to the exabyte (EB) scale with the expected data
 volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
-experiments such as SKA (Square Kilometre Array) will all enter this
+experiments such as the Square Kilometre Array (SKA) entering this
 regime. In devising experimental computing models for this era, many
-factors have to be taken into account. ~In particular, the increasing
+factors have to be taken into account. In particular, the increasing
 availability of very high-speed networks, which may reduce the need for
+% * <briedel@uchicago.edu> 2018-04-04T17:59:27.886Z:
+%
+% > very high-speed networks
+% What is "very high-speed" here? 100G? 1000G?
+%
+% ^ <briedel@uchicago.edu> 2018-04-04T17:59:40.128Z.
 CPU and data co-location, need to be examined. Such networks may allow
 for more extensive use of data access over the wide-area network (WAN),
-which may provide failover capabilities, global and federated data
+which may provide fail-over capabilities, global and federated data
 namespaces, and will have an impact on data caching. Shifts in the data
-presentation/analysis models such as a potential move to event-based
-data streaming from the more traditional dataset-based or file-based
-data access, will be particularly important for optimizing the
-utilization of opportunistic computing cycles on HPC facilities,
-commercial cloud resources, and campus clusters, and can potentially
-resolve currently limiting factors such as job eviction.
+presentation and analysis models, such as a potential move to or
+sub-event-based or event-based data streaming from the more
+% * <briedel@uchicago.edu> 2018-04-04T18:00:17.686Z:
+%
+% > sub-event-based or event-based data
+% Added "sub-event-based" because we talk about it below.
+%
+% ^ <briedel@uchicago.edu> 2018-04-04T18:00:36.656Z.
+traditional dataset-based or file-based data access, will be
+particularly important for optimizing the utilization of
+opportunistic computing cycles on HPC facilities, commercial cloud resources,
+and campus clusters, and can potentially resolve currently
+limiting factors such as job eviction.
+% * <briedel@uchicago.edu> 2018-04-04T18:01:55.490Z:
+%
+% >  job eviction
+% Using "job evictions" as a reason to move to different DOMA model seems like an odd choice. Shouldn't it be something like "scheduling difficulties"
+%
+% ^.
 
 Planned analysis of HL-LHC data and data analysis from other large
+% * <briedel@uchicago.edu> 2018-04-04T18:09:21.988Z:
+%
+% > large
+% XENON1T has shown that the collaboration doesnt have to be "large" to require DOMA
+%
+% ^.
 international collaborations will need to adopt a distributed computing
@@
 
-The LHC experiments currently provision and manage about an Exabyte of
-storage, approximately half of which is archival, and half is
-traditional disk storage. The storage requirements per year are expected
-to jump by a factor of 10 for the HL-LHC. This itself is faster than
+The LHC experiments currently collectively provision and manage about an EB of
+storage of which approximately half is archival and half is
+traditional disk storage. The annual storage requirements are expected
+to jump by a factor of 10 for the HL-LHC. This increase is faster than
 projected Moore's Law gains and will present major challenges. Storage
-will remain one of the visible cost drivers for HEP computing, however
+% * <briedel@uchicago.edu> 2018-04-05T01:05:55.296Z:
+%
+% > Moore's Law gains
+% This is only a CPU.
+What about storage technology?
+Currently we are adding about 2 TB/year in disk capacity.
+It is unlikely will be at 100 TB/disk in 2023.
+We may have O(10) - O(100) TB SSDs  at a reasonable cost, but we can't be certain.
+What does the future of tape look like?
+%
+% ^.
+will remain one of the visible cost drivers for HEP computing with
 the projected growth and cost of the computational resources needed to
-analyze the data is also expected to grow even faster than the base
+analyze the data expected to grow even faster than the base
 storage costs. The combination of storage and analysis computing costs
@@
 \def\labelenumi{\arabic{enumi}.}
-\item \textbf{Big Data:} The HL-LHC era will significantly increase both the
-  date rate and the data volume. The computing systems will need to
+\item \textbf{Big Data:} The HL-LHC era will significantly increase both expected
+  date rates and data volumes. Computing systems will need to
   handle this without significant cost increases and within evolving
@@
   place new requirements on data. Specifically, the use of new types of
-  compute resources (\emph{e.g.,} cloud, HPC) with different dynamic
-  availability and characteristics are used will require more dynamic
+  compute resources (e.g., cloud, HPC) with different dynamic
+  availability and characteristics will require more dynamic
   DOMA systems.
@@
   requirements on how and where data is accessed and produced. For
-  example, specific applications (e.g. training for machine learning)
-  may require use of specialized processor resources such as GPUs,
+  example, specific applications (e.g., training for machine learning)
+  may require use of specialized processor resources, such as GPUs,
+% * <briedel@uchicago.edu> 2018-04-04T18:14:36.074Z:
+%
+% > ,  placing further requirements on data.
+% Not sure this is needed.
+%
+% ^.
   placing further requirements on data.
@@
 the commercial world and in the rest of the research world also provides
-a number of opportunities.
+a number of opportunities and solution to tackle these challenges.
 
@@
 
-The original LHC computing models (circa 2005) were built up from the
+The original LHC computing models (developed circa 2005) were built up from
 simpler models used before distributed computing was a central part of
@@
 \item \textbf{Data Organization:} This is essentially how data is structured
-  as it is written. Most data is written in flat files, in ROOT format,
+  as it is written. Most data is written in flat files, in the ROOT format, and
   typically with a column-wise organization of the data. The records
@@
   the use of distributed computing in the form of the grid. The
-  experiments developed dedicated data transfer and placement systems,
-  along with catalogs, to move data between computing centers. To first
+  experiments developed dedicated data catalog,  transfer and placement systems,
+  to move data between computing centers. To first
   order, the computing models were rather static: data was placed at
@@
 
-Before the LHC turn-on and in the first years of the LHC, these three
+Before the LHC turn-on and in the first years of data taking, these three
 areas were, to first order, optimized independently. Many of the
 challenges were in the area of ''Data Management (DM)'' as the Worldwide
-LHC Computing Grid was commissioned. As the LHC computing matured
-through Run1 and Run2, the interest has turned to optimizations spanning
+LHC Computing Grid was commissioned. As the LHC experiments' computing matured
+through Run1 and Run2, interest has turned to optimizations spanning
 these three areas. For example, the recent use of ``Data Federations''
-mixes up Data Management and Access. As we will see below, some of the
+must consider together both Data Management and Access. As we will see below, some of the
 foreseen opportunities towards HL-LHC may require global optimizations.
 
-Thus in this document we take a broader view than traditional ``DM'',
+Thus, in this document we take a broader view than traditional ``DM'',
 and consider the combination of ``Data Organization, Management and
-Access (DOMA)'') together. We believe that this full picture of data
+Access (DOMA)'' together. We believe that this full picture of data
 needs in HEP will provide important opportunities for efficiency and
-scalability as we enter the many-Exabyte era.
+scalability as we enter the many-EB era.
 
@@
   limitations and to IOPs transaction limits that underlying
-  technologies can provide. In particular the internal structure of the
+  technologies can have. In particular, the internal structure of the
   event data can dramatically impact the efficiency of data retrieval
   and favor different access models and protocols (i.e. event
-  streaming.) and the actual analysis calculations being performed can
-  dramatically affect the CPU to IO ratios. These factors make this
+  streaming) while the actual analysis calculations being performed can
+  dramatically affect the CPU-to-IO ratios. These factors make this
   categorization of data the most challenging to project to future
@@
   the bandwidth requirements that are needed in enacting site-to-site
-  data transfers between WAN endpoints (e.g. 10-100Gbps.) This
+  data transfers between WAN endpoints (e.g. 10-100~Gbps). This
   requirement on the available WAN bandwidth provides a corresponding
   performance requirement on on the underlying storage systems to match
-  the network performance. Satisfying both requirements is necessary to
-  provide an efficient point to point data flow.
+  network performance. Satisfying both requirements is necessary to
+  provide efficient point-to-point data flow.
 \end{enumerate}
@@
 to be tuned or matched to the corresponding computing resources that the
-experiments are targeting their workflows to execute on.
+experiments are targeting.
 
@@
 support evolving physical storage options similar to the the shifts that
-have occurred in recent years between mRAM, SSDs, shingled disks, tapes
-and other technologies. ~
+have occurred in recent years between mRAM, SSDs, shingled disks, tapes,
+and other technologies.
 
@@
 
-The HEP community has long established patterns that have been used for
-data organization and have shaped the prevailing data access and storage
-models that are in use today. These established patterns have focused
+The HEP community has long-established data organization patterns that have
+shaped the prevailing data access and storage
+models that are in use today. These established patterns are centered
 around sets of event records organized into discrete collections through
-traditional file based storage. There are, however, potential gains that
+traditional file-based storage. There are, however, potential gains that
 can be made by exploring alternative data organization models and the
@@
 to be advantageous, implemented by the community. A shift in designs may
-have the potential to reduce storage costs, shift network utilisation,
+have the potential to reduce storage costs, shift network utilization,
 and result in more efficient use of compute resources.
@@
 each with different impacts on the analysis methodologies. First, we
-should reconsider the structure of the data, which can be row oriented,
-column oriented, have varying levels of granularity, represent larger or
-smaller aggregations of objects etc, boosting the efficiency of
+should reconsider the structure of the data, which can be row-oriented,
+column-oriented, have varying levels of granularity, or represent larger or
+smaller aggregations of objects, boosting the efficiency of
 particular applications.
@@
 Secondly, we should reconsider the granularity at which data management
-systems operate. This is currently aggregations of events (files, or
-larger sets such as blocks or datasets), and could shift towards being
+systems operate. This is currently at aggregations of events (files, or
+larger sets, such as blocks or datasets), and could shift towards being
 the event (or even sub-event), potentially reducing storage volumes. In
-that case, other types of ``container'' beyond files could be considered
+that case, other types of ``containers'' beyond files could be considered
 (e.g. objects).
@@
 Design decisions will be subject to competing requirements from data
-management and access optimisation, and will have consequences
+management and access optimization, and will have consequences
 throughout the system. We have to tackle a global optimization problem,
 taking into account all use cases and their respective frequencies to
-further optimise application performance for a given resource
+further optimize application performance for a given resource
 investment.
@@
 
-The structure of data has to be adapted to applications' requirements.
-At one limit, the data could be organized simply as events, the other
+The structure of data has to be adapted to application requirements.
+At one limit, the data could be organized simply as events, while the other
 extreme is to organize the data as columns which represent sub-event
@@
 the scope of technologies that can be used to store the data; for
-example, exploiting object stores (such as AWS S3) as event stores. In
-addition, this would support parallel reading and writing of events as
+example, exploiting object stores (such as AWS S3 or Ceph) as event stores. Such an approach
+would support parallel reading and writing of events as
 event objects from applications. This could allow applications to better
-scale to large numbers of CPU core where now these applications
+scale to large numbers of CPU cores whereas in current models application
 concurrency is limited by the requirement to sequentially write events
 to files. Currently, the efficiency of of reading and writing events
-from object stores is sensitive to the size of the event. The smaller
-the event the worse is the efficiency.. Some of this may be mitigated by
-creating event bundles (i.e. several events in one object). The
+from object stores is sensitive to the size of the event---
+the smaller the event, the worse the efficiency is. Some of these potential inefficiencies
+may be mitigated by creating event bundles (i.e. several events in one object). The
 evolution of the ATLAS Event Service may provide some of the answers and
@@
 storage'') which may fit certain use cases well. The relevance of these
-solutions to computing models must be understood, not only through
+solutions to HEP computing models must be understood, not only through
 technical metrics such as performance, but with an understanding of the
@@
 
-\subsection{Optimizing Data Management
-}\label{optimizing-data-management}
+\subsection{Optimizing Data Management}\label{optimizing-data-management}
 
 Data Management policy will directly affect data access. For example,
-the decision to distribute multiple enriched data samples optimized for
+the decision to distribute multiple enriched data-samples optimized for
 different use cases would improve data access and thus compute
@@
 opportunistic resources through enabling pre-emptibility in jobs. This
-would have knock-on effects on data catalogues, data distribution and
+would have cascading effects on data catalogs, data distribution and
 storage systems, and would affect data access. Where an experiment
@@
 policy. Thus the potential gains in terms of efficient use of resources
-should be realised through data management systems which allow this
+should be realized through data management systems which allow this
 policy to be expressed and implemented.
@@
 purpose (accounting, metadata, location). Analysis access to the data,
-for instance, may benefit from being below the event level (e.g. physics
+for instance, may benefit from being below the event level (\emph{e.g.}, physics
 object). Any such catalogs may have to be external to the primary data
-store (e.g., relational databases that work with metadata in the primary
+store (\emph{e.g.}, relational databases that work with metadata in the primary
 data store). If cataloging requirements for production and
@@
 or content-addressable storage is used, the catalogs need only a single
+% * <briedel@uchicago.edu> 2018-04-05T02:04:26.292Z:
+%
+% > content-addressable
+% This comes a little out of nowhere. Where is this discussed before?
+%
+% ^.
 handle to have a fully descriptive location of a given event or other
 object being sought.
+% * <briedel@uchicago.edu> 2018-04-05T02:05:03.704Z:
+%
+% > or other object
+% This is the first time "other object" is mentioned. What other object? TBranches? It would be better to call this sub-event data like done before.
+%
+% ^.
 
@@
 all, disk-based storage is considered as a caching layer should also be
-considered, in which case work on optimising workflows for cache
+considered, in which case work on optimizing workflows for cache
 efficiency and exploiting data popularity metrics would be valuable.
@@
 network services (CDNs) such as the one offered by Akamai. Without CDNs,
+% * <briedel@uchicago.edu> 2018-04-05T02:08:36.870Z:
+%
+% > Akamai.
+% Google also offers one
+%
+% ^.
 on-demand streaming of movies over the Internet or even the delivery of
@@
 collapse. What is the equivalent of content delivery for accessing huge
-datasets over distance, and for which a particular analysis might only
+datasets over distance where a particular analysis might only
 require a fraction of the dataset? CDNs use, to great benefit, highly
-sophisticated, proprietary naming, caching, and placement techniques
+sophisticated proprietary naming, caching, and placement techniques
 that are customized to particular types of data and what is known about
@@
 having to fetch everything across continents. New approaches, like Named
-Data Networking (NDN), could play a role: by allowing to name, at the
-network level, the data you are looking for rather than its location, it
+Data Networking (NDN), could play a role. NDNs allow the naming, at the
+network level, of the data you are looking for rather than its location, which
 opens the paths for many optimizations behind the scenes based on
@@
 caching proxy servers, server-less caching, ARC-style caching) and may
+% * <briedel@uchicago.edu> 2018-04-05T02:09:39.626Z:
+%
+% > server-less caching
+% In-memory caching?
+%
+% ^.
 provide sufficient information on how such strategies can optimize
@@
 
-Data analysis is currently tied in with ROOT-based formats. In many
+Data analysis in most of HEP is currently tied to ROOT-based formats. In many
 currently-used paradigms, physicists consider all events at an
@@
 
-\section{Performance : Tools and
+\section{Performance: Tools and
 Metrics}\label{performance-tools-and-metrics}
@@
 possibility of increased user-driven selection and dataset generation.
-In terms of reducing the strain on data access, techniques such as data
+In terms of reducing the strain on data access mechanisms, techniques such as data
 augmentation can prove useful. Cluster-computing frameworks (e.g.,
 Spark) can also be leveraged to reduce the time of dataset reduction and
-access. It will be necessary to study and classify IO patterns in
+access. It will be necessary to study and classify I/O patterns in
 applications used in order to understand how data access methods can be
@@
 used in algorithms resulting from active research in computer science
-that study payoffs between storage and cpu in various computing models.
+that study trade-offs between storage and CPU in various computing models.
 
-A first correlated analysis of the IO patterns of different
-computational tasks has already started at the CERN computing center and
-within several experiments. In particular the combination of
+A first correlated analysis of the I/O patterns of different
+computational tasks has already started at the CERN Data Centre and
+within several experiments. In particular, the combination of
 infrastructure and experiment information (e.g. detailed knowledge of
@@
 metrics) may allow us to define more meaningful performance metrics than
-the traditionally used raw data rate and CPU utilization which both show
+the traditionally-used raw data rate and CPU utilization which both have
 their limitations. These metrics should enable a joint quantitative
 comparison of different storage and computing approaches across
-individual system boundaries and hence lead to a more effective resource
+individual system boundaries and thus lead to a more effective resource
 investment.
@@
 for the CPU needs of the HEP community. As these techniques continue to
-grow and blossom, they will place new requirements on the computational
+gain traction in our field and evolve, they will place new requirements on the computational
 resources that need to be leveraged by all of HEP. The storage systems
-that are developed, and the data management techniques that are employed
+that are developed and the data management techniques that are employed
 will need to directly support this wide range of computational
 facilities, and will need to be matched to the changes in the
-computational work, so as not to impede the improvements that they are
+computational work so as not to impede the improvements that they are
 bringing.
@@
 cost of doing scientific analysis. This R\&D should include an
-optimization of both the capital costs of storage, as well as the
+optimization of both the capital costs of storage as well as the
 potential impacts the storage systems can have on the CPU requirements
@@
 
-The ability to leverage new storage technologies as they become
-available into existing data delivery models is a challenge that we must
+The ability to leverage new storage technologies, as they become
+available, into existing data delivery models is a challenge that we must
 be prepared for. New storage systems will present new interfaces and new
-behaviour. A key decision in their successful exploitation will be
+behavior. A key decision in their successful exploitation will be
 whether to encapsulate this within a more familiar service or whether to
@@
 On the hardware side, R\&D is needed in alternative approaches to data
-archiving to determine the possible cost/performance tradeoffs.
-Currently, tape is extensively used to hold data that cannot be
-economically made available online. While the data is still accessible,
-it comes with a high latency penalty; limiting possible analysis. We
-suggest investigating either separate direct access-based archives (e.g.
-disk or optical) or new models that overlay online direct access volumes
+archiving to determine the possible cost/performance trade-offs.
+Currently, tape is extensively used to hold data that cannot economically be made available online. While the tape-stored data is still accessible,
+it comes with a high latency penalty and thus limiting possible analysis. We suggest investigating either separate direct-access-based archives (e.g.
+disk or optical) or new models that overlay online direct-access volumes
 with archive space. This is especially relevant when access latency is
@@
 Cost reductions in maintenance and operation of the storage
-infrastructure can be realised through convergence of the major
+infrastructure can be realized through convergence of the major
 experiments and resource providers on shared solutions. This does not
@@
 environment. Indeed, there will always be a judgment to make on the
-desirability of using a variety of specialised systems, or abstracting
+desirability of using a variety of specialized systems, or abstracting
 the commonalities through a more limited but common interface (the SRM
@@
 will be further promoted by extending these concepts of convergence
-beyond HEP and into the other large-scale scientific endeavours that
+beyond HEP and into the other large-scale scientific endeavors that
 will share the infrastructure in the coming decade. Efforts must be made
@@
 What is clear is that what is being done today will not scale to the
-needs of HL LHC.
+needs of HL-LHC experiments.
 
@@
 \subsection{Sub-file granularity}\label{sub-file-granularity}
-Sub-file granularity ({\it e.g.} event-based granularity) should be studied
+Sub-file granularity ({\it e.g.} sub-event-based or event-based granularity) should be studied
 to see whether it can be implemented efficiently and in a scalable, cost-effective
@@
 Spark-like analysis tools and their adaptability to HEP analysis needs.
-\item Evaluate just-in-time
+\item Evaluate new compression schemes, just-in-time
   decompression schemes and mappings onto hardware architectures
@@
   application.
+\item Evaluate possible alternative storage formats to the ROOT format, esp.\ storage formats optimized for storage efficiency and/or  data filterting through metdata
 \end{itemize}
@@
 \item Re-evaluate the role of archival storage solutions for HEP.
+\item Evaluate the role of "newer" storage architectures, such as object stores or key-value stores
 \end{itemize}
@@
 The inherent trade-offs between data access latency and efficiency of CPU use need to be studied and optimized
-on a global scale across the HEP infrastructure. The proof-of-concept phase involves the following tasks::
+on a global scale across the HEP infrastructure. The proof-of-concept phase involves the following tasks:
 
@@
 
+\subsection{Data Access and Data Access Patterns}\label{access-patterns}
+The data access pattern should inform the the use of storage media and data location. Proposed areas of inquiry are:
+
+\begin{itemize}
+\item Understand data access patterns, including investigating frequency of access of data
+\item Exploring ways to ease preliminary data filtering steps
+\item \end{itemize}
+\emph{The  proof-of-concept should be completed by 2020 and, if successful, implementation in
+production will be done in the following years.}
+
+A large portion of the LHC data is write-once, read-once. This would make it ideal to store this data on archieval media. Alternatively, a portion of the data is accessed frequently by analyzers.
+
 \section{Conclusions}\label{conclusions}
@@
 areas pertaining to data access and management in HEP that will need to
-be addressed in the coming decade before the expected volume and
+be re-examined in the coming decade before the expected volume and
 complexity of data becomes prohibitively expensive to store, access, and
 analyze. Extending current data handling methods and methodologies will
-prove intractable in the HL-LHC and DUNE eras. The development and
+prove intractable in the HL-LHC and DUNE era. The development and
 adoption of new data analysis paradigms gives the field, as a whole, a

### 2018-04-05 CWP/papers/HSF-CWP-2017-01_roadmap/authors/a2tex.py
a2tex: fix wrong \bigskip placement in affiliations
- Fixes #69

@@ -198,13 +198,16 @@ def main():
                                                    eol_str),
                        file=output)
 
+        # In JHEP style, space between authors and affiliations is managed by the style
+        if not options.jhep:
+            print ("\\bigskip", file=output)
+
         for affiliation in sorted_affiliations:
             if options.jhep:
                 print ("\\affiliation[{}]{{{}}}".format(str(affiliation_map[affiliation].mark),
                                                         latex_escape(affiliation_map[affiliation].address)),
                        file=output)
             else:
-                print ("\\bigskip", file=output)
                 print ("\\par {{\\footnotesize $^{{{}}}$ {}}}".format(str(affiliation_map[affiliation].mark),
                                                                       latex_escape(affiliation_map[affiliation].address)),
                        file=output)

### 2018-04-06 CWP/papers/HSF-CWP-2017-01_roadmap/authors/a2tex.py
Correct JHEP author stype formatting (#72)
Follow the style guide here:
https://jhep.sissa.it/jhep/help/JHEP/TeXclass/DOCS/JHEP-author-manual.pdf

@@ -176,13 +176,22 @@ def main():
         
     # Write out the latex author file
     with open(options.output_file, "w", encoding="utf-8") as output:
-        for author in sorted(author_list, key=lambda author: author.surname):
+        author_list = sorted(author_list, key=lambda author: author.surname)
+        for author in author_list:
             affiliation_list = ",".join([ affiliation_map[affiliation].mark for affiliation in author.affiliations ])
 
             if options.jhep:
-                print ("\\author[{}]{{{}, {}}}".format(affiliation_list,
-                                                       author.surname,
-                                                       author.forename),
+                formatted_name = "{} {}".format(author.forename,
+                                                author.surname)
+                author_position = author_list.index(author) - len(author_list)
+                # The fiddly bit to get "a, b, c and d" as in JHEP style
+                # N.B. For position -2 doing nothing *is correct*
+                if author_position < -2:
+                    formatted_name += ","
+                elif author_position == -1:
+                    formatted_name = "and {}".format(formatted_name)
+                print ("\\author[{}]{{{}}}".format(affiliation_list,
+                                                    formatted_name),
                        file=output)
             else:
                 if author.footnotes:

### 2018-04-10 CWP/papers/HSF-CWP-2017-13_soft-dev/latex/cwp-wg-software-development.tex
fix bug in relative path to bib (#47)
* fix bug in relative path to bib

* updates to affiliations

--- 2017-12-19 CWP/papers/HSF-CWP-2017-013_soft-dev/latex/cwp-wg-software-development.tex
@@
 \usepackage[style=numeric-comp,sorting=none]{biblatex}
-\addbibresource{../../roadmap/latex/cwp.bib}
-\addbibresource{../../roadmap/latex/cwp-chapters.bib}
+\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp.bib}
+\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp-chapters.bib}
 
@@
 
-\abstract{The High Energy Phyiscs community has developed and needs to maintain
+\abstract{The High Energy Physics community has developed and needs to maintain
 many tens of millions of lines of code and to integrate effectively the
@@
 community needs to be fostered and an appropriate publication and citation
-strategy needs to be developed. As part of the HEP Softare Foundation's
+strategy needs to be developed. As part of the HEP Software Foundation's
 Community White Paper process a working group on Software Development,
 Deployment and Validation was formed to examine all of these
-issues, identify best practice and to formulare recommendations for the next
+issues, identify best practice and to formulate recommendations for the next
 decade. Its report is presented here.}

### 2018-04-10 CWP/papers/HSF-CWP-2017-05_analysis/latex/cwp-wg-analysis-interpretation.tex
minor updates to sync repo with what was submitted to arXiv

@@
-% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.
+% Copyright (C) 2018, The HSF Community White Paper authors, licence CC-BY-4.0.
 
@@
 \usepackage[style=numeric-comp,sorting=none]{biblatex}
-\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp.bib}
-\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp-chapters.bib}
+\addbibresource{cwp.bib}
+\addbibresource{cwp-chapters.bib}
 
@@
  & & HSF-CWP-2017-05 \\
- & & March 18, 2018 \\ % use \date or hardwire e.g. December 15, 2017
+ & & March 25, 2018 \\ % use \date or hardwire e.g. December 15, 2017
  & & \\
@@
 \author[g]{David Lange}
-\author[k,1]{Mark Neubauer}
+\author[k,1]{Mark S. Neubauer}
 \author[k]{Daniel S. Katz}
@@
 \author[g]{Jim Pivarski}
-\author[p]{Eduardo Rodrigues}
+\author[p]{Martin Ritter}
+\author[q]{Eduardo Rodrigues}
 \author[l]{Tai Sakuma}
@@
 \author[p]{Michael D. Sokoloff}
-\author[q]{Carl Vuosalo}
-\author[m]{\\ Frank W\"{u}rthwein}
-\author[r]{Gordon Watts}
+\author[r]{\\ Carl Vuosalo}
+\author[m]{Frank W\"{u}rthwein}
+\author[s]{Gordon Watts}
 
@@
 \affiliation[o]{Massachusetts Institute of Technology, Cambridge, Massachusetts, USA}
-\affiliation[p]{University of Cincinnati, Cincinnati, Ohio, USA}
-\affiliation[q]{University of Wisconsin-Madison, Madison, Wisconsin, USA}
-\affiliation[r]{University of Washington, Seattle, Washington, USA}
+\affiliation[p]{Fakultät für Physik, Ludwig-Maximilians-Universität München, München, Germany}
+\affiliation[q]{University of Cincinnati, Cincinnati, Ohio, USA}
+\affiliation[r]{University of Wisconsin-Madison, Madison, Wisconsin, USA}
+\affiliation[s]{University of Washington, Seattle, Washington, USA}
 \affiliation[1]{Paper Editor}
-
-%\affiliation[c]{Laboratório de Instrumentação e Física Experimental de Partículas, Lisbon, Portugal and University of Minho, Braga, Portugal, Supported by FCT-Portugal, IF/00050/2013/CP1172/CT0002}
-%\affiliation[g]{Fermi National Accelerator Laboratory, Batavia, Illinois USA, Supported by the US-DOE, DE-AC02-07CH11359}
-%\affiliation[h]{Centro de Investigaciones Energéticas, Medioambientales y Tecnológicas (CIEMAT), Madrid, Spain, Supported by ES-MINECO, FPA2016-80994-c2-1-R and MDM-2015-0509}
-%\affiliation[i]{University of Illinois at Urbana-Champaign, Urbana, Illinois USA, Supported by the US-DOE, DE-SC0018098 and US-NSF ACI-1558233}
-%\affiliation[m]{University of Michigan, Supported by the US-DOE, DE-SC0007859 and US-NSF, 76749/1136652/2}
-%\affiliation[o]{University of Cincinnati, Supported by the US-NSF, ACI-1450319}
-%\affiliation[p]{University of Cincinnati, Supported by the US-NSF, ACI-1558219}
 

### 2018-04-14 CWP/papers/HSF-CWP-2017-01_roadmap/authors/a2tex.py
a2tex: add footnotes in author list (JHEP style)
Fixes #73

@@ -179,6 +179,9 @@ def main():
         author_list = sorted(author_list, key=lambda author: author.surname)
         for author in author_list:
             affiliation_list = ",".join([ affiliation_map[affiliation].mark for affiliation in author.affiliations ])
+            if author.footnotes:
+                footnote_str = ",".join([footnote_list[int(id) - 1].mark for id in author.footnotes])
+                affiliation_list = affiliation_list + "," + footnote_str
 
             if options.jhep:
                 formatted_name = "{} {}".format(author.forename,
@@ -194,9 +197,6 @@ def main():
                                                     formatted_name),
                        file=output)
             else:
-                if author.footnotes:
-                    footnote_str = ",".join( [ footnote_list[int(id)-1].mark for id in author.footnotes ])
-                    affiliation_list = affiliation_list + "," + footnote_str
                 if author == author_list[-1]:
                     eol_str = ""
                 else:
@@ -221,12 +221,18 @@ def main():
                                                                       latex_escape(affiliation_map[affiliation].address)),
                        file=output)
 
-        if not options.jhep:
-            print ("\\bigskip", file=output)
-            for footnote in sorted_footnotes:
-                print ("\\par {{\\footnotesize $^{{{}}}$ {}}}".format(footnote.mark,
-                                                                      latex_escape(footnote.text)),
-                       file=output)
+        if len(sorted_footnotes) > 0:
+            if options.jhep:
+                for footnote in sorted_footnotes:
+                    print ("\\note[{}]{{{}}}".format(footnote.mark,
+                                                         latex_escape(footnote.text)),
+                           file=output)
+            else:
+                print ("\\bigskip", file=output)
+                for footnote in sorted_footnotes:
+                    print ("\\par {{\\footnotesize $^{{{}}}$ {}}}".format(footnote.mark,
+                                                                          latex_escape(footnote.text)),
+                           file=output)
 
 
 if __name__ == '__main__':

### 2018-04-16 CWP/papers/HSF-CWP-2017-01_roadmap/authors/a2tex.py
a2tex: print only footnotes relevant to the actual author list

@@ -43,10 +43,14 @@ ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
 
 
 class Footnote():
-    def __init__(self, number_str, text):
-        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
-        self.mark = footnote_symbols[int(number_str)-1]
+    def __init__(self, text):
         self.text = text
+        self.mark = None
+
+    def set_mark(self, i):
+        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
+        self.mark = footnote_symbols[i-1]
+
     def __repr__(self):
         return repr((self.mark, self.text))
 
@@ -141,6 +145,7 @@ def main():
             institute_address[institute] = address
 
     affiliation_map = {}
+    referenced_foonotes = {}
     for author in author_list:
         for affiliation in author.affiliations:
             if affiliation not in affiliation_map:
@@ -150,23 +155,34 @@ def main():
                     raise Exception('Affiliation {} not defined for {}, {}'.format(affiliation, 
                                                                                    author.surname,
                                                                                    author.forename))
+        for footnote in author.footnotes:
+            referenced_foonotes[footnote] = ''
 
     # Alphabetise and assign footnote marks
     sorted_affiliations = sorted(affiliation_map)
     for footnote_mark, affiliation in enumerate(sorted_affiliations, start=1):
         affiliation_map[affiliation].set_mark(footnote_mark)
     
-    # Parse and assign footnotes
+    # Parse and assign footnotes: interprete the first field as a footnote key
+    # matched against author footnotes rather than the real mark
     with open(options.footnotes, encoding='utf-8') as footnote_fh:
-        footnote_list = []
+        footnote_map = {}
         for footnote_line in footnote_fh:
             footnote_line = footnote_line.strip()
             if footnote_line.startswith("#") or footnote_line == "":
                 continue
-            number, note = footnote_line.split(" ", 1)
-            number = number.rstrip(".")
-            footnote_list.append(Footnote(number, note))
-    sorted_footnotes = sorted(footnote_list, key=lambda footnote: footnote.mark)
+            note_key, note = footnote_line.split(" ", 1)
+            note_key = note_key.rstrip(".")
+            if note_key in referenced_foonotes:
+                if note_key not in footnote_map:
+                    footnote_map[note_key] = Footnote(note)
+                else:
+                    raise Exception('Duplicated footnote ({}'.format(note_key))
+
+    # Generate footnote marks (letters)
[…]

### 2018-04-16 CWP/papers/HSF-CWP-2017-01_roadmap/authors/a2tex.py
a2tex: code refactoring

@@ -42,14 +42,22 @@ LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
 ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
 
 
-class Footnote():
+class AffiliationFootnote():
     def __init__(self, text):
         self.text = text
         self.mark = None
 
-    def set_mark(self, i):
-        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
-        self.mark = footnote_symbols[i-1]
+    def set_mark(self, i, letter=False):
+        if letter:
+            if i > 26:
+                j = int(i / 26)
+                i = i % 26
+                j_str = chr(ord('a')+j-1)
+            else:
+                j_str = ''
+            self.mark = '{}{}'.format(j_str, chr(ord('a')+i-1))
+        else:
+            self.mark = str(i)
 
     def __repr__(self):
         return repr((self.mark, self.text))
@@ -74,14 +82,6 @@ class Author():
             raise Exeption("Got an exception({}) processing line: {}".format(e, author_line))
 
 
-class Affiliation():
-    def __init__(self,address):
-        self.address = address
-        self.mark = ''
-
-    def set_mark(self, i):
-        self.mark = str(i)
-
 def latex_escape(string):
     """
     Function to escape latex reserved characters
@@ -102,6 +102,35 @@ def latex_escape(string):
     for special_char, repl in SPECIAL_LATEX_CHARS.items():
        string = re.sub(re.escape(special_char), repl, string)
     return string
+
+
+def read_kv_file(file, separator=':', referenced_keys=None):
+    """
+    Read a text file defining key/values separated by separator, checking that
+    there is not duplicate key. If reference_keys is defined, only those keys are
+    loaded from the file.
+
+    :param file: file name
+    :param separator: separator string
+    :param referenced_keys: if defined, list of keys to consider (others are ignored)
+    :return: dictionnary of AffiliationNote object where the key is the file line key
+    """
+
+    with open(file, encoding='utf-8') as fh:
+        kv_map = {}
+        for kv_line in fh:
+            kv_line = kv_line.strip()
+            if kv_line.startswith("#") or kv_line == "":
+                continue
+            k, v = kv_line.split(separator, 1)
+            v = v.strip()
+            if k in kv_map:
+                raise Exception('Duplicated key ({}) found in {}'.format(k, file))
+            elif (referenced_keys is None) or (k in referenced_keys):
[…]

### 2018-04-16 CWP/papers/HSF-CWP-2017-01_roadmap/latex/authors.tex
CWP + DOMA paper: update author list with fixed a2tex

--- 2018-01-16 CWP/papers/HSF-CWP-2017-01_roadmap/latex/authors.tex
@@
-Albrecht, Johannes$^{66}$;
-Alves Jr, Antonio Augusto$^{79}$;
+Albrecht, Johannes$^{68}$;
+Alves Jr, Antonio Augusto$^{80}$;
 Amadio, Guilherme$^{5}$;
-Anh-Ky, Nguyen$^{117}$;
-Aphecetche, Laurent$^{63}$;
+Anh-Ky, Nguyen$^{121}$;
+Aphecetche, Laurent$^{65}$;
 Apostolakis, John$^{5}$;
-Asai, Makoto$^{61,p}$;
+Asai, Makoto$^{62,p}$;
 Atzori, Luca$^{5}$;
@@
 Bandieramonte, Marilena$^{5}$;
+Banerjee, Sunanda$^{16,c}$;
 Barisits, Martin$^{5}$;
@@
 Belforte, Stefano$^{34}$;
-Benjamin, Douglas$^{80}$;
-Bernius, Catrin$^{61}$;
-Bhimji, Wahid$^{44}$;
-Bianchi, Riccardo Maria$^{102}$;
+Benjamin, Douglas$^{81}$;
+Bernius, Catrin$^{62}$;
+Bhimji, Wahid$^{45}$;
+Bianchi, Riccardo Maria$^{104}$;
 Bird, Ian$^{5}$;
-Biscarat, Catherine$^{50}$;
+Biscarat, Catherine$^{51}$;
 Blomer, Jakob$^{5}$;
-Bloom, Kenneth$^{94}$;
+Bloom, Kenneth$^{96}$;
 Boccali, Tommaso$^{31}$;
-Bockelman, Brian$^{94}$;
-Bold, Tomasz$^{41}$;
+Bockelman, Brian$^{96}$;
+Bold, Tomasz$^{42}$;
 Bonacorsi, Daniele$^{25}$;
-Boveia, Antonio$^{98}$;
+Boveia, Antonio$^{100}$;
 Bozzi, Concezio$^{27}$;
-Bracko, Marko$^{91,39}$;
-Britton, David$^{84}$;
-Buckley, Andy$^{84}$;
+Bracko, Marko$^{92,40}$;
+Britton, David$^{85}$;
+Buckley, Andy$^{85}$;
 Buncic, Predrag$^{5,a}$;
-Calafiura, Paolo$^{44}$;
+Calafiura, Paolo$^{45}$;
 Campana, Simone$^{5,a}$;
@@
 Carlino, Gianpaolo$^{30}$;
-Castro, Nuno$^{45,93,d}$;
+Castro, Nuno$^{46,95,d}$;
 Cattaneo, Marco$^{5}$;
 Cerminara, Gianluca$^{5}$;
-Chang, Philip$^{73}$;
-Chapman, John$^{74}$;
+Cervantes Villanueva, Javier$^{5}$;
+Chang, Philip$^{74}$;
+Chapman, John$^{75}$;
 Chen, Gang$^{23}$;
 Childers, Taylor$^{1}$;
-Clarke, Peter$^{81}$;
+Clarke, Peter$^{82}$;
 Clemencic, Marco$^{5}$;
-Cogneras, Eric$^{48}$;
-Collier, Ian$^{59}$;
+Cogneras, Eric$^{49}$;
+Coles, Jeremy$^{75}$;
+Collier, Ian$^{60}$;
 Corti, Gloria$^{5}$;
 Cosmo, Gabriele$^{5}$;
-Costanzo, Davide$^{108}$;
+Costanzo, Davide$^{111}$;
 Couturier, Ben$^{5}$;
-Cranmer, Kyle$^{55}$;
+Cranmer, Kyle$^{56}$;
 Cranshaw, Jack$^{1}$;
 Cristella, Leonardo$^{26}$;
-Crooks, David$^{84}$;
-Crépé-Renaudin, Sabine$^{50}$;
+Crooks, David$^{85}$;
+Crépé-Renaudin, Sabine$^{51}$;
+Currie, Robert$^{82}$;
 Dallmeier-Tiessen, Sünje$^{5}$;
-De, Kaushik$^{110}$;
-De Cian, Michel$^{85}$;
+De, Kaushik$^{113}$;
+De Cian, Michel$^{86}$;
+De Roeck, Albert$^{5}$;
+Delgado Peris, Antonio$^{7,g}$;
+Derue, Frédéric$^{50}$;
 Di Girolamo, Alessandro$^{5}$;
+Di Guida, Salvatore$^{29}$;
 Dimitrov, Gancho$^{5}$;
-Doglioni, Caterina$^{89,h}$;
-Dotti, Andrea$^{61,p}$;
+Doglioni, Caterina$^{90,h}$;
+Dotti, Andrea$^{62,p}$;
 Duellmann, Dirk$^{5}$;
-Duflot, Laurent$^{43}$;
+Duflot, Laurent$^{44}$;
 Dykstra, Dave$^{16,c}$;
@@
 Dziurda, Agnieszka$^{5}$;
-Egede, Ulrik$^{36}$;
-Elmer, Peter$^{103,a}$;
+Egede, Ulrik$^{37}$;
+Elmer, Peter$^{105,a}$;
 Elmsheuser, Johannes$^{2}$;
@@
 Eulisse, Giulio$^{5}$;
-Ferber, Torben$^{71}$;
-Filipcic, Andrej$^{39}$;
-Fisk, Ian$^{62}$;
+Farrell, Steven$^{45}$;
+Ferber, Torben$^{72}$;
+Filipcic, Andrej$^{40}$;
+Fisk, Ian$^{63}$;
 Fitzpatrick, Conor$^{14}$;
-Flix, José$^{57,7,g}$;
-Formica, Andrea$^{37}$;
-Forti, Alessandra$^{90}$;
-Frost, James$^{101}$;
+Flix, José$^{58,7,g}$;
+Formica, Andrea$^{38}$;
+Forti, Alessandra$^{91}$;
+Franzoni, Giovanni$^{5}$;
+Frost, James$^{103}$;
+Fuess, Stu$^{16,c}$;
 Gaede, Frank$^{13}$;
 Ganis, Gerardo$^{5}$;
-Gardner, Robert$^{78}$;
-Garonne, Vincent$^{99}$;
+Gardner, Robert$^{79}$;
+Garonne, Vincent$^{101}$;
 Gellrich, Andreas$^{13}$;
 Genser, Krzysztof$^{16,c}$;
-George, Simon$^{60}$;
-Geurts, Frank$^{104}$;
+George, Simon$^{61}$;
+Geurts, Frank$^{106}$;
 Gheata, Andrei$^{5}$;
@@
 Giacomini, Francesco$^{9}$;
-Giagu, Stefano$^{106,33}$;
-Giffels, Manuel$^{40}$;
-Gingrich, Douglas$^{67}$;
+Giagu, Stefano$^{108,33}$;
+Giffels, Manuel$^{41}$;
+Gingrich, Douglas$^{69}$;
 Girone, Maria$^{5}$;
-Gligorov, Vladimir V.$^{49}$;
-Glushkov, Ivan$^{110}$;
-Gohn, Wesley$^{86}$;
+Gligorov, Vladimir V.$^{50,b}$;
+Glushkov, Ivan$^{113}$;
+Gohn, Wesley$^{87}$;
 Gonzalez Lopez, Jose Benito$^{5}$;
-González Caballero, Isidro$^{100}$;
-González Fernández, Juan R.$^{100}$;
+González Caballero, Isidro$^{102}$;
+González Fernández, Juan R.$^{102}$;
 Govi, Giacomo$^{16}$;
 Grandi, Claudio$^{25}$;
-Grasland, Hadrien$^{43}$;
-Gray, Heather$^{44}$;
-Grillo, Lucia$^{90}$;
-Guan, Wen$^{115}$;
+Grasland, Hadrien$^{44}$;
+Gray, Heather$^{45}$;
+Grillo, Lucia$^{91}$;
+Guan, Wen$^{118}$;
 Gutsche, Oliver$^{16,c}$;
-Gyurjyan, Vardan$^{38}$;
-Hanushevsky, Andrew$^{61,p}$;
+Gyurjyan, Vardan$^{39}$;
+Hanushevsky, Andrew$^{62,p}$;
 Hariri, Farah$^{5}$;
@@
 Harvey, John$^{5,a}$;
-Hauth, Thomas$^{40}$;
+Hauth, Thomas$^{41}$;
 Hegner, Benedikt$^{5,a}$;
 Heinemann, Beate$^{13}$;
-Heinrich, Lukas$^{55}$;
+Heinrich, Lukas$^{56}$;
+Heiss, Andreas$^{41}$;
 Hernández, José M.$^{7,g}$;
-Hildreth, Michael$^{96,f}$;
-Hodgkinson, Mark$^{108}$;
-Hoeche, Stefan$^{61,p}$;
+Hildreth, Michael$^{98,f}$;
+Hodgkinson, Mark$^{111}$;
+Hoeche, Stefan$^{62,p}$;
+Holzman, Burt$^{16,c}$;
 Hristov, Peter$^{5}$;
-Huang, Xingtao$^{107}$;
-Ivanchenko, Vladimir N.$^{5,111}$;
-Ivanov, Todor$^{109}$;
-Jashal, Brij$^{65}$;
+Huang, Xingtao$^{110}$;
+Ivanchenko, Vladimir N.$^{5,114}$;
+Ivanov, Todor$^{112}$;
+Iven, Jan$^{5}$;
+Jashal, Brij$^{67}$;
 Jayatilaka, Bodhitha$^{16,c}$;
-Jones, Roger$^{87,a}$;
-Jouvin, Michel$^{43,a}$;
+Jones, Roger$^{88,a}$;
+Jouvin, Michel$^{44,a}$;
 Jun, Soon Yung$^{16,c}$;
-Kagan, Michael$^{61,p}$;
-Kalderon, Charles William$^{89}$;
+Kagan, Michael$^{62,p}$;
+Kalderon, Charles William$^{90}$;
+Kane, Meghan$^{64}$;
 Karavakis, Edward$^{5}$;
-Katz, Daniel S.$^{77}$;
+Katz, Daniel S.$^{78}$;
 Kcira, Dorian$^{11}$;
-Kersevan, Borut Paul$^{88}$;
+Keeble, Oliver$^{5}$;
+Kersevan, Borut Paul$^{89}$;
 Kirby, Michael$^{16,c}$;
 Klimentov, Alexei$^{2}$;
-Klute, Markus$^{51}$;
+Klute, Markus$^{93}$;
 Komarov, Ilya$^{34,n}$;
-Konstantinov, Dmitri$^{58}$;
-Koppenburg, Patrick$^{54}$;
+Konstantinov, Dmitri$^{59}$;
+Koppenburg, Patrick$^{55}$;
 Kowalkowski, Jim$^{16,c}$;
-Kreczko, Luke$^{70}$;
-Kuhr, Thomas$^{47,a}$;
+Kreczko, Luke$^{71}$;
+Kuhr, Thomas$^{48,a}$;
 Kutschke, Robert$^{16,a,c}$;
 Kuznetsov, Valentin$^{12}$;
-Lampl, Walter$^{69}$;
+Lampl, Walter$^{70}$;
 Lancon, Eric$^{2}$;
-Lange, David$^{103,a}$;
+Lange, David$^{105,a}$;
 Lassnig, Mario$^{5}$;
 Laycock, Paul$^{5}$;
-Leggett, Charles$^{44}$;
-Letts, James$^{73}$;
+Leggett, Charles$^{45}$;
+Letts, James$^{74}$;
 Lewendel, Birgit$^{13}$;
-Li, Teng$^{81}$;
-Lima, Guilherme$^{16}$;
-Linacre, Jacob$^{59,m}$;
+Li, Teng$^{82}$;
+Lima, Guilherme$^{16,c}$;
+Linacre, Jacob$^{60,m}$;
 Linden, Tomas$^{18}$;
+Livny, Miron$^{6}$;
 Lo Presti, Giuseppe$^{5}$;
 Lopienski, Sebastian$^{5}$;
-Love, Peter$^{87}$;
+Love, Peter$^{88}$;
+Lyon, Adam$^{16,c}$;
 Magini, Nicolò$^{28}$;
-Marshall, Zachary L$^{44}$;
+Marshall, Zachary L$^{45}$;
 Martelli, Edoardo$^{5}$;
-Martin-Haugh, Stewart$^{59}$;
+Martin-Haugh, Stewart$^{60}$;
 Mato, Pere$^{5}$;
-Mazumdar, Kajari$^{65}$;
-McCauley, Thomas$^{96}$;
+Mazumdar, Kajari$^{67}$;
+McCauley, Thomas$^{98}$;
 McFayden, Josh$^{5}$;
-McKee, Shawn$^{92,l}$;
-McNab, Andrew$^{90}$;
-Mehdiyev, Rashid$^{76}$;
+McKee, Shawn$^{94,l}$;
+McNab, Andrew$^{91}$;
+Mehdiyev, Rashid$^{77}$;
 Meinhard, Helge$^{5}$;
@@
 Mendez Lorenzo, Patricia$^{5}$;
-Mete, Alaettin Serhan$^{72}$;
+Mete, Alaettin Serhan$^{73}$;
 Michelotto, Michele$^{32}$;
-Mitrevski, Jovan$^{47}$;
+Mitrevski, Jovan$^{48}$;
 Moneta, Lorenzo$^{5}$;
-Morgan, Ben$^{114}$;
-Mount, Richard$^{61,p}$;
-Moyse, Edward$^{68}$;
-Murray, Sean$^{75,10}$;
-Neubauer, Mark S$^{77,a,k}$;
-Novaes, Sérgio$^{105}$;
+Morgan, Ben$^{117}$;
+Mount, Richard$^{62,p}$;
+Moyse, Edward$^{93}$;
+Murray, Sean$^{76,10}$;
+Nairz, Armin$^{5}$;
+Neubauer, Mark S$^{78,a,k}$;
+Norman, Andrew$^{16,c}$;
+Novaes, Sérgio$^{107}$;
 Novak, Mihaly$^{5}$;
 Oyanguren, Arantza$^{22}$;
-Ozturk, Nurcan$^{110}$;
-Pacheco Pages, Andres$^{57,20,j}$;
-Paganini, Michela$^{116}$;
-Pansanel, Jerome$^{35}$;
-Pascuzzi, Vincent R.$^{112}$;
-Patrick, Glenn$^{59}$;
+Ozturk, Nurcan$^{113}$;
+Pacheco Pages, Andres$^{58,20,j}$;
+Paganini, Michela$^{119}$;
+Pansanel, Jerome$^{36}$;
+Pascuzzi, Vincent R.$^{115}$;
+Patrick, Glenn$^{60}$;
 Pearce, Alex$^{5}$;
-Pearson, Ben$^{52}$;
+Pearson, Ben$^{53}$;
 Pedro, Kevin$^{16,c}$;
-Perdue, Gabriel$^{16}$;
-Perez-Calero Yzquierdo, Antonio$^{57,7,g}$;
+Perdue, Gabriel$^{16,c}$;
+Perez-Calero Yzquierdo, Antonio$^{58,7,g}$;
 Perrozzi, Luca$^{15}$;
-Petersen, Troels$^{53}$;
+Petersen, Troels$^{54}$;
 Petric, Marko$^{5}$;
+Petzold, Andreas$^{41}$;
 Piedra, Jónatan$^{21}$;
-Piilonen, Leo$^{118,i}$;
+Piilonen, Leo$^{122,i}$;
 Piparo, Danilo$^{5}$;
+Pivarski, Jim$^{105}$;
 Pokorski, Witold$^{5}$;
-Polci, Francesco$^{49}$;
+Polci, Francesco$^{50}$;
 Potamianos, Karolos$^{13}$;
 Psihas, Fernanda$^{24}$;
-Raven, Gerhard$^{54}$;
+Puig Navarro, Albert$^{120}$;
+Quast, Günter$^{41}$;
+Raven, Gerhard$^{55,123}$;
 Reuter, Jürgen$^{13}$;
@@
 Rinaldi, Lorenzo$^{25}$;
-Ritter, Martin$^{47}$;
+Ritter, Martin$^{48}$;
 Robinson, James$^{13}$;
-Rodrigues, Eduardo$^{79,a,e}$;
+Rodrigues, Eduardo$^{80,a,e}$;
 Roiser, Stefan$^{5,a}$;
-Rousseau, David$^{43}$;
-Roy, Gareth$^{84}$;
+Rousseau, David$^{44}$;
+Roy, Gareth$^{85}$;
+Rybkine, Grigori$^{44}$;
 Sailer, Andre$^{5}$;
-Sakuma, Tai$^{70}$;
+Sakuma, Tai$^{71}$;
 Santana, Renato$^{3}$;
-Sartirana, Andrea$^{46}$;
-Schellman, Heidi$^{56}$;
+Sartirana, Andrea$^{47}$;
+Schellman, Heidi$^{57}$;
 Schovancová, Jaroslava$^{5}$;
-Schramm, Steven$^{83}$;
+Schramm, Steven$^{84}$;
 Schulz, Markus$^{5}$;
 Sciabà, Andrea$^{5}$;
-Seidel, Sally$^{95}$;
-Sekmen, Sezen$^{42}$;
-Serfon, Cedric$^{99}$;
-Severini, Horst$^{97}$;
+Seidel, Sally$^{97}$;
+Sekmen, Sezen$^{43}$;
+Serfon, Cedric$^{101}$;
+Severini, Horst$^{99}$;
 Sexton-Kennedy, Elizabeth$^{16,a,c}$;
-Seymour, Michael$^{90}$;
-Shapoval, Illya$^{44}$;
+Seymour, Michael$^{91}$;
+Sgalaberna, Davide$^{5}$;
+Shapoval, Illya$^{45}$;
 Shiers, Jamie$^{5}$;
-Shiu, Jing-Ge$^{64}$;
+Shiu, Jing-Ge$^{66}$;
 Short, Hannah$^{5}$;
 Siroli, Gian Piero$^{25}$;
-Skipsey, Sam$^{84}$;
+Skipsey, Sam$^{85}$;
 Smith, Tim$^{5}$;
 Snyder, Scott$^{2}$;
-Sokoloff, Michael D$^{79,a}$;
+Sokoloff, Michael D$^{80,a}$;
+Spentzouris, Panagiotis$^{16,c}$;
 Stadie, Hartmut$^{17}$;
-Stark, Giordon$^{6}$;
-Stewart, Gordon$^{84}$;
+Stark, Giordon$^{79}$;
+Stewart, Gordon$^{85}$;
 Stewart, Graeme$^{5,a}$;
-Sánchez, Arturo$^{113,19}$;
+Sánchez, Arturo$^{116,19}$;
 Sánchez-Hernández, Alberto$^{8,o}$;
-Taffard, Anyes$^{72}$;
-Templon, Jeff$^{54}$;
+Taffard, Anyes$^{73}$;
+Tamponi, Umberto$^{35}$;
+Templon, Jeff$^{55}$;
 Tenaglia, Giacomo$^{5}$;
-Tsulaia, Vakhtang$^{44}$;
-Tunnell, Christopher$^{6}$;
+Tsulaia, Vakhtang$^{45}$;
+Tunnell, Christopher$^{79}$;
 Vaandering, Eric$^{16,c}$;
 Valassi, Andrea$^{5}$;
-Vallecorsa, Sofia$^{82}$;
+Vallecorsa, Sofia$^{83}$;
 Valsan, Liviu$^{5}$;
@@
 Voss, Christian$^{13}$;
-Vuosalo, Carl$^{115}$;
-Vázquez Sierra, Carlos$^{54}$;
+Votava, Margaret$^{16,c}$;
+Vuosalo, Carl$^{118}$;
+Vázquez Sierra, Carlos$^{55}$;
 Wartel, Romain$^{5}$;
+Watts, Gordon T$^{109}$;
 Wenaus, Torre$^{2}$;
 Wenzel, Sandro$^{5}$;
-Winklmeier, Frank$^{56}$;
+Williams, Mike$^{52}$;
+Winklmeier, Frank$^{57}$;
 Wissing, Christoph$^{13}$;
-Wuerthwein, Frank$^{73}$;
-Wynne, Benjamin$^{81}$;
+Wuerthwein, Frank$^{74}$;
+Wynne, Benjamin$^{82}$;
 Xiaomei, Zhang$^{23}$;
-Yang, Wei$^{61,p}$;
+Yang, Wei$^{62,p}$;
 Yazgan, Efe$^{23}$
 \bigskip
-\par {\footnotesize $^{1}$ High Energy Physics Division, Argonne National Laboratory, Argonne IL, USA}
-\par {\footnotesize $^{2}$ Physics Department, Brookhaven National Laboratory, Upton NY, USA}
+\par {\footnotesize $^{1}$ High Energy Physics Division, Argonne National Laboratory, Argonne, IL, USA}
+\par {\footnotesize $^{2}$ Physics Department, Brookhaven National Laboratory, Upton, NY, USA}
 \par {\footnotesize $^{3}$ Centro Brasileiro de Pesquisas Físicas, Rio de Janeiro, Brazil}
@@
 \par {\footnotesize $^{5}$ CERN, Geneva, Switzerland}
-\par {\footnotesize $^{6}$ Enrico Fermi Institute, University of Chicago, Chicago IL, USA}
+\par {\footnotesize $^{6}$ Center for High Throughput  Computing, Computer Sciences Department, University of Wisconsin-Madison, Madison, WI, USA}
 \par {\footnotesize $^{7}$ Centro de Investigaciones Energéticas Medioambientales y Tecnológicas (CIEMAT), Madrid, Spain}
@@
 \par {\footnotesize $^{10}$ Center for High Performance Computing, Cape Town, South Africa}
-\par {\footnotesize $^{11}$ California Institute of Technology, Pasadena, California, USA}
-\par {\footnotesize $^{12}$ Cornell University, Ithaca, USA}
+\par {\footnotesize $^{11}$ California Institute of Technology, Pasadena, California, CA, USA}
+\par {\footnotesize $^{12}$ Cornell University, Ithaca, NY, USA}
 \par {\footnotesize $^{13}$ Deutsches Elektronen-Synchrotron, Hamburg, Germany}
 \par {\footnotesize $^{14}$ Institute of Physics, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland}
-\par {\footnotesize $^{15}$ ETH Zurich - Institute for Particle Physics and Astrophysics (IPA), Zurich, Switzerland}
-\par {\footnotesize $^{16}$ Fermi National Accelerator Laboratory, Batavia, USA}
+\par {\footnotesize $^{15}$ ETH Zürich - Institute for Particle Physics and Astrophysics (IPA), Zürich, Switzerland}
+\par {\footnotesize $^{16}$ Fermi National Accelerator Laboratory, Batavia, IL, USA}
 \par {\footnotesize $^{17}$ University of Hamburg, Hamburg, Germany}
@@
 \par {\footnotesize $^{23}$ Institute of High Energy Physics, Chinese Academy of Sciences, Beijing}
-\par {\footnotesize $^{24}$ Department of Physics, Indiana University, Bloomington IN, USA}
+\par {\footnotesize $^{24}$ Department of Physics, Indiana University, Bloomington, IN, USA}
 \par {\footnotesize $^{25}$ INFN Sezione di Bologna, Università di Bologna, Bologna, Italy}
@@
 \par {\footnotesize $^{34}$ INFN Sezione di Trieste, Università di Trieste, Trieste, Italy}
-\par {\footnotesize $^{35}$ Université de Strasbourg, CNRS, IPHC UMR 7178, F-67000 Strasbourg, France}
-\par {\footnotesize $^{36}$ Imperial College London, London, United Kingdom}
-\par {\footnotesize $^{37}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat à l’Énergie Atomique), Gif-sur-Yvette, France}
-\par {\footnotesize $^{38}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
-\par {\footnotesize $^{39}$ Jožef Stefan Institute, Ljubljana, Slovenia}
-\par {\footnotesize $^{40}$ Karlsruhe Institute of Technology, Karlsruhe, Germany}
-\par {\footnotesize $^{41}$ AGH University of Science and Technology, Faculty of Physics and Applied Computer Science, Krakow, Poland}
-\par {\footnotesize $^{42}$ Kyungpook National University, Daegu, Republic of Korea}
-\par {\footnotesize $^{43}$ LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
-\par {\footnotesize $^{44}$ Lawrence Berkeley National Laboratory and University of California, Berkeley CA, USA}
-\par {\footnotesize $^{45}$ Laboratório de Instrumentação e Física Experimental de Partículas (LIP), Lisboa, Portugal}
-\par {\footnotesize $^{46}$ Laboratoire Leprince-Ringuet, École Polytechnique, CNRS/IN2P3, Université Paris-Saclay, Palaiseau, France}
-\par {\footnotesize $^{47}$ Fakultät für Physik, Ludwig-Maximilians-Universität München, München, Germany}
-\par {\footnotesize $^{48}$ Laboratoire de Physique Corpusculaire, Clermont Université and Université Blaise Pascal and CNRS/IN2P3, Clermont-Ferrand, France}
-\par {\footnotesize $^{49}$ LPNHE, Université Pierre et Marie Curie, Université Paris Diderot, CNRS/IN2P3, Paris, France}
-\par {\footnotesize $^{50}$ Laboratoire de Physique Subatomique et de Cosmologie, Université Joseph Fourier and CNRS/IN2P3 and Institut National Polytechnique de Grenoble, Grenoble, France}
-\par {\footnotesize $^{51}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
-\par {\footnotesize $^{52}$ Max-Planck-Institut für Physik (Werner-Heisenberg-Institut), München, Germany}
-\par {\footnotesize $^{53}$ Niels Bohr Institute, University of Copenhagen, Kobenhavn, Denmark}
-\par {\footnotesize $^{54}$ Nikhef National Institute for Subatomic Physics and University of Amsterdam, Amsterdam, Netherlands}
-\par {\footnotesize $^{55}$ Department of Physics, New York University, New York, NY, USA}
-\par {\footnotesize $^{56}$ Center for High Energy Physics, University of Oregon, Eugene OR, USA}
-\par {\footnotesize $^{57}$ Port d’Informació Científica (PIC), Universitat Autònoma de Barcelona (UAB), Barcelona, Spain}
-\par {\footnotesize $^{58}$ High Energy Physics (IHEP), Protvino, Russia}
-\par {\footnotesize $^{59}$ STFC Rutherford Appleton Laboratory, Didcot, United Kingdom}
-\par {\footnotesize $^{60}$ Department of Physics, Royal Holloway University of London, Surrey, United Kingdom}
-\par {\footnotesize $^{61}$ SLAC National Accelerator Laboratory, Menlo Park CA, USA}
-\par {\footnotesize $^{62}$ Simons Foundation, New York, USA}
-\par {\footnotesize $^{63}$ SUBATECH, IMT Atlantique, Université de Nantes, CNRS-IN2P3, Nantes, France}
-\par {\footnotesize $^{64}$ National Taiwan University, Taipei, Taiwan}
-\par {\footnotesize $^{65}$ Tata Institute of Fundamental Research, Mumbai, India}
-\par {\footnotesize $^{66}$ Technische Universit\"at Dortmund, Dortmund, Germany}
-\par {\footnotesize $^{67}$ Department of Physics, University of Alberta, Edmonton, AB, Canada}
-\par {\footnotesize $^{68}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
-\par {\footnotesize $^{69}$ Department of Physics, University of Arizona, Tucson, AZ, USA}
-\par {\footnotesize $^{70}$ H.H. Wills Physics Laboratory, University of Bristol, Bristol, United Kingdom}
-\par {\footnotesize $^{71}$ Department of Physics, University of British Columbia, Vancouver, BC, Canada}
-\par {\footnotesize $^{72}$ Department of Physics and Astronomy, University of California Irvine, Irvine, CA, USA}
-\par {\footnotesize $^{73}$ University of California, San Diego, La Jolla, USA}
-\par {\footnotesize $^{74}$ Cavendish Laboratory, University of Cambridge, Cambridge, United Kingdom}
-\par {\footnotesize $^{75}$ Physics Department, University of Cape Town, Cape Town, South Africa}
-\par {\footnotesize $^{76}$ Carleton University, Ottawa, ON, Canada}
-\par {\footnotesize $^{77}$ University of Illinois Urbana-Champaign, Champaign, Illinois, USA}
-\par {\footnotesize $^{78}$ Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
-\par {\footnotesize $^{79}$ University of Cincinnati, Cincinnati, OH, USA}
-\par {\footnotesize $^{80}$ Department of Physics, Duke University, Durham, NC, USA}
-\par {\footnotesize $^{81}$ SUPA - School of Physics and Astronomy, University of Edinburgh, Edinburgh, United Kingdom}
-\par {\footnotesize $^{82}$ Gangneung-Wonju National University, South Korea}
-\par {\footnotesize $^{83}$ Section de Physique, Université de Genève, Geneva, Switzerland}
-\par {\footnotesize $^{84}$ SUPA - School of Physics and Astronomy, University of Glasgow, Glasgow, United Kingdom}
-\par {\footnotesize $^{85}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
-\par {\footnotesize $^{86}$ Department of Physics and Astronomy, University of Kentucky, Lexington, USA}
-\par {\footnotesize $^{87}$ Physics Department, Lancaster University, Lancaster, United Kingdom}
-\par {\footnotesize $^{88}$ Department of Physics, Jožef Stefan Institute and University of Ljubljana, Ljubljana, Slovenia}
-\par {\footnotesize $^{89}$ Fysiska institutionen, Lunds Universitet, Lund, Sweden}
-\par {\footnotesize $^{90}$ School of Physics and Astronomy, University of Manchester, Manchester, United Kingdom}
-\par {\footnotesize $^{91}$ University of Maribor, Ljubljana, Slovenia}
-\par {\footnotesize $^{92}$ Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
-\par {\footnotesize $^{93}$ Departamento de Física, Universidade do Minho, Braga, Portugal}
-\par {\footnotesize $^{94}$ University of Nebraska-Lincoln, Lincoln, USA}
-\par {\footnotesize $^{95}$ Department of Physics and Astronomy, University of New Mexico, Albuquerque NM, USA}
-\par {\footnotesize $^{96}$ University of Notre Dame, Notre Dame, USA}
-\par {\footnotesize $^{97}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman OK, USA}
-\par {\footnotesize $^{98}$ The Ohio State University, Columbus, USA}
-\par {\footnotesize $^{99}$ Department of Physics, University of Oslo, Oslo, Norway}
-\par {\footnotesize $^{100}$ Universidad de Oviedo, Oviedo, Spain}
-\par {\footnotesize $^{101}$ Department of Physics, Oxford University, Oxford, UK}
-\par {\footnotesize $^{102}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
-\par {\footnotesize $^{103}$ Princeton University, Princeton, USA}
-\par {\footnotesize $^{104}$ Rice University, Houston, TX, USA}
-\par {\footnotesize $^{105}$ Universidade Estadual Paulista, São Paulo, Brazil}
-\par {\footnotesize $^{106}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
-\par {\footnotesize $^{107}$ School of Physics, Shandong University, Shandong, China}
-\par {\footnotesize $^{108}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
-\par {\footnotesize $^{109}$ University of Sofia, Sofia, Bulgaria}
-\par {\footnotesize $^{110}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
-\par {\footnotesize $^{111}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
-\par {\footnotesize $^{112}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
-\par {\footnotesize $^{113}$ Dipartimento di Chimica, Fisica e Ambiente, Università di Udine, Udine, Italy}
-\par {\footnotesize $^{114}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
-\par {\footnotesize $^{115}$ University of Wisconsin - Madison, Madison, WI, USA}
-\par {\footnotesize $^{116}$ Department of Physics, Yale University, New Haven, CT, USA}
-\par {\footnotesize $^{117}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
-\par {\footnotesize $^{118}$ Virginia Tech, Blacksburg, Virginia, USA}
+\par {\footnotesize $^{35}$ INFN Sezione di Torino, Torino, Italy}
+\par {\footnotesize $^{36}$ Université de Strasbourg, CNRS, IPHC UMR 7178, F-67000 Strasbourg, France}
+\par {\footnotesize $^{37}$ Imperial College London, London, United Kingdom}
+\par {\footnotesize $^{38}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat à l’Énergie Atomique), Gif-sur-Yvette, France}
+\par {\footnotesize $^{39}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, VA, USA}
+\par {\footnotesize $^{40}$ Jožef Stefan Institute, Ljubljana, Slovenia}
+\par {\footnotesize $^{41}$ Karlsruhe Institute of Technology, Karlsruhe, Germany}
+\par {\footnotesize $^{42}$ AGH University of Science and Technology, Faculty of Physics and Applied Computer Science, Krakow, Poland}
+\par {\footnotesize $^{43}$ Kyungpook National University, Daegu, Republic of Korea}
+\par {\footnotesize $^{44}$ LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
+\par {\footnotesize $^{45}$ Lawrence Berkeley National Laboratory and University of California, Berkeley, CA, USA}
+\par {\footnotesize $^{46}$ Laboratório de Instrumentação e Física Experimental de Partículas (LIP), Lisboa, Portugal}
+\par {\footnotesize $^{47}$ Laboratoire Leprince-Ringuet, École Polytechnique, CNRS/IN2P3, Université Paris-Saclay, Palaiseau, France}
+\par {\footnotesize $^{48}$ Fakultät für Physik, Ludwig-Maximilians-Universität München, München, Germany}
+\par {\footnotesize $^{49}$ Laboratoire de Physique Corpusculaire, Clermont Université and Université Blaise Pascal and CNRS/IN2P3, Clermont-Ferrand, France}
+\par {\footnotesize $^{50}$ Laboratoire de Physique Nucléaire et de Hautes Energies (LPNHE), Sorbonne Université, Université Paris Diderot, CNRS/IN2P3, Paris, France}
+\par {\footnotesize $^{51}$ Laboratoire de Physique Subatomique et de Cosmologie, Université Joseph Fourier and CNRS/IN2P3 and Institut National Polytechnique de Grenoble, Grenoble, France}
+\par {\footnotesize $^{52}$ Massachusetts Institute of Technology, Cambridge, MA, USA}
+\par {\footnotesize $^{53}$ Max-Planck-Institut für Physik (Werner-Heisenberg-Institut), München, Germany}
+\par {\footnotesize $^{54}$ Niels Bohr Institute, University of Copenhagen, Kobenhavn, Denmark}
+\par {\footnotesize $^{55}$ Nikhef National Institute for Subatomic Physics, Amsterdam, Netherlands}
+\par {\footnotesize $^{56}$ Department of Physics, New York University, New York, NY, USA}
+\par {\footnotesize $^{57}$ Center for High Energy Physics, University of Oregon, Eugene, OR, USA}
+\par {\footnotesize $^{58}$ Port d’Informació Científica (PIC), Universitat Autònoma de Barcelona (UAB), Barcelona, Spain}
+\par {\footnotesize $^{59}$ High Energy Physics (IHEP), Protvino, Russia}
+\par {\footnotesize $^{60}$ STFC Rutherford Appleton Laboratory, Didcot, United Kingdom}
+\par {\footnotesize $^{61}$ Department of Physics, Royal Holloway University of London, Surrey, United Kingdom}
+\par {\footnotesize $^{62}$ SLAC National Accelerator Laboratory, Menlo Park, CA, USA}
+\par {\footnotesize $^{63}$ Simons Foundation, New York, NY, USA}
+\par {\footnotesize $^{64}$ SoundCloud, Berlin, Germany}
+\par {\footnotesize $^{65}$ SUBATECH, IMT Atlantique, Université de Nantes, CNRS-IN2P3, Nantes, France}
+\par {\footnotesize $^{66}$ National Taiwan University, Taipei, Taiwan}
+\par {\footnotesize $^{67}$ Tata Institute of Fundamental Research, Mumbai, India}
+\par {\footnotesize $^{68}$ Technische Universit\"at Dortmund, Dortmund, Germany}
+\par {\footnotesize $^{69}$ Department of Physics, University of Alberta, Edmonton, AB, Canada}
+\par {\footnotesize $^{70}$ Department of Physics, University of Arizona, Tucson, AZ, USA}
+\par {\footnotesize $^{71}$ H.H. Wills Physics Laboratory, University of Bristol, Bristol, United Kingdom}
+\par {\footnotesize $^{72}$ Department of Physics, University of British Columbia, Vancouver, BC, Canada}
+\par {\footnotesize $^{73}$ Department of Physics and Astronomy, University of California Irvine, Irvine, CA, USA}
+\par {\footnotesize $^{74}$ University of California, San Diego, La Jolla, CA, USA}
+\par {\footnotesize $^{75}$ Cavendish Laboratory, University of Cambridge, Cambridge, United Kingdom}
+\par {\footnotesize $^{76}$ Physics Department, University of Cape Town, Cape Town, South Africa}
+\par {\footnotesize $^{77}$ Carleton University, Ottawa, ON, Canada}
+\par {\footnotesize $^{78}$ University of Illinois Urbana-Champaign, Champaign, Illinois, IL, USA}
+\par {\footnotesize $^{79}$ Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
+\par {\footnotesize $^{80}$ University of Cincinnati, Cincinnati, OH, USA}
+\par {\footnotesize $^{81}$ Department of Physics, Duke University, Durham, NC, USA}
+\par {\footnotesize $^{82}$ SUPA - School of Physics and Astronomy, University of Edinburgh, Edinburgh, United Kingdom}
+\par {\footnotesize $^{83}$ Gangneung-Wonju National University, South Korea}
+\par {\footnotesize $^{84}$ Section de Physique, Université de Genève, Geneva, Switzerland}
+\par {\footnotesize $^{85}$ SUPA - School of Physics and Astronomy, University of Glasgow, Glasgow, United Kingdom}
+\par {\footnotesize $^{86}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
+\par {\footnotesize $^{87}$ Department of Physics and Astronomy, University of Kentucky, Lexington, KY, USA}
+\par {\footnotesize $^{88}$ Physics Department, Lancaster University, Lancaster, United Kingdom}
+\par {\footnotesize $^{89}$ Department of Physics, Jožef Stefan Institute and University of Ljubljana, Ljubljana, Slovenia}
+\par {\footnotesize $^{90}$ Fysiska institutionen, Lunds Universitet, Lund, Sweden}
+\par {\footnotesize $^{91}$ School of Physics and Astronomy, University of Manchester, Manchester, United Kingdom}
+\par {\footnotesize $^{92}$ University of Maribor, Ljubljana, Slovenia}
+\par {\footnotesize $^{93}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
+\par {\footnotesize $^{94}$ Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
+\par {\footnotesize $^{95}$ Departamento de Física, Universidade do Minho, Braga, Portugal}
+\par {\footnotesize $^{96}$ University of Nebraska-Lincoln, Lincoln, NE, USA}
+\par {\footnotesize $^{97}$ Department of Physics and Astronomy, University of New Mexico, Albuquerque, NM, USA}
+\par {\footnotesize $^{98}$ University of Notre Dame, Notre Dame, IN, USA}
+\par {\footnotesize $^{99}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman, OK, USA}
+\par {\footnotesize $^{100}$ The Ohio State University, Columbus, OH, USA}
+\par {\footnotesize $^{101}$ Department of Physics, University of Oslo, Oslo, Norway}
+\par {\footnotesize $^{102}$ Universidad de Oviedo, Oviedo, Spain}
+\par {\footnotesize $^{103}$ Department of Physics, Oxford University, Oxford, UK}
+\par {\footnotesize $^{104}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
+\par {\footnotesize $^{105}$ Princeton University, Princeton, NJ, USA}
+\par {\footnotesize $^{106}$ Rice University, Houston, TX, USA}
+\par {\footnotesize $^{107}$ Universidade Estadual Paulista, São Paulo, Brazil}
+\par {\footnotesize $^{108}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
+\par {\footnotesize $^{109}$ University of Washington, Seattle, WA, USA}
+\par {\footnotesize $^{110}$ School of Physics, Shandong University, Shandong, China}
+\par {\footnotesize $^{111}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
+\par {\footnotesize $^{112}$ University of Sofia, Sofia, Bulgaria}
+\par {\footnotesize $^{113}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
+\par {\footnotesize $^{114}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
+\par {\footnotesize $^{115}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
+\par {\footnotesize $^{116}$ Dipartimento di Chimica, Fisica e Ambiente, Università di Udine, Udine, Italy}
+\par {\footnotesize $^{117}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
+\par {\footnotesize $^{118}$ University of Wisconsin - Madison, Madison, WI, USA}
+\par {\footnotesize $^{119}$ Department of Physics, Yale University, New Haven, CT, USA}
+\par {\footnotesize $^{120}$ Physik-Institut, Universität Zürich, Zürich, Switzerland}
+\par {\footnotesize $^{121}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
+\par {\footnotesize $^{122}$ Virginia Tech, Blacksburg, Virginia, VA, USA}
+\par {\footnotesize $^{123}$ Vrije Universiteit Amsterdam, Amsterdam, The Netherlands}
 \bigskip

### 2018-04-16 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
CWP DOMA: import fixes from Overleaf
- Changes made up to 13/4 included

@@
 
-\abstract{Without significant changes to data organization, management, and access (DOMA), HEP experiments will find scientific output limited by how fast data can be accessed and digested by computational resources.
-In this white paper we discuss challenges in DOMA that HEP experiments such as the HL-LHC will face as well as potential ways to address them.
-A research and development timeline to assess these changes is also proposed.}
+\abstract{Without significant changes to data organization, management, and
+access (DOMA), HEP experiments will find scientific output limited by how
+fast data can be accessed and digested by computational resources. In this
+white paper we discuss challenges in DOMA that HEP experiments, such as
+the HL-LHC, will face as well as potential ways to address them. A research
+and development timeline to assess these changes is also proposed.}
 
@@
 
-\title{HSF Community White Paper on Data Organization, Management and Access
-(DOMA)}
+\title{HEP Software Foundation Community White Paper Working Group --
+Data Organization, Management and Access (DOMA)}
 
@@
 
-The coming decade will bring HEP to the exabyte (EB) scale with the expected data
+The coming decade will bring HEP to the exabyte scale with the expected data
 volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
-experiments such as the Square Kilometre Array (SKA) entering this
-regime. In devising experimental computing models for this era, many
+experiments such as the Square Kilometer Array (SKA) entering this
+regime. In devising computing models for this era, many
 factors have to be taken into account. In particular, the increasing
-availability of very high-speed networks, which may reduce the need for
+availability of very high-speed networks may reduce the need for
 % * <briedel@uchicago.edu> 2018-04-04T17:59:27.886Z:
@@
 % ^ <briedel@uchicago.edu> 2018-04-04T17:59:40.128Z.
-CPU and data co-location, need to be examined. Such networks may allow
-for more extensive use of data access over the wide-area network (WAN),
-which may provide fail-over capabilities, global and federated data
+CPU and data co-location. Such networks may allow
+for more extensive use of data access over the wide-area network (WAN) while providing fail-over capabilities, global and federated data
 namespaces, and will have an impact on data caching. Shifts in the data
-presentation and analysis models, such as a potential move to or
-sub-event-based or event-based data streaming from the more
-% * <briedel@uchicago.edu> 2018-04-04T18:00:17.686Z:
-%
-% > sub-event-based or event-based data
-% Added "sub-event-based" because we talk about it below.
-%
-% ^ <briedel@uchicago.edu> 2018-04-04T18:00:36.656Z.
-traditional dataset-based or file-based data access, will be
-particularly important for optimizing the utilization of
-opportunistic computing cycles on HPC facilities, commercial cloud resources,
-and campus clusters, and can potentially resolve currently
-limiting factors such as job eviction.
-% * <briedel@uchicago.edu> 2018-04-04T18:01:55.490Z:
-%
-% >  job eviction
-% Using "job evictions" as a reason to move to different DOMA model seems like an odd choice. Shouldn't it be something like "scheduling difficulties"
-%
-% ^.
+presentation and analysis models, such as a potential move to event-based or
+sub-event-based data streaming from the more traditional dataset-based or file-based
+data access, will be particularly important for optimizing the
+utilization of opportunistic computing cycles on HPC facilities,
+commercial cloud resources, and campus clusters, and can potentially
+resolve currently limiting factors such as job eviction.
 
-Planned analysis of HL-LHC data and data analysis from other large
+Planned analysis of HL-LHC data and data analysis from other
 % * <briedel@uchicago.edu> 2018-04-04T18:09:21.988Z:
@@
 %
-% ^.
+% ^ <briedel@uchicago.edu> 2018-04-09T20:36:43.846Z.
 international collaborations will need to adopt a distributed computing
@@
 
-The LHC experiments currently collectively provision and manage about an EB of
+The LHC experiments currently collectively provision and manage about an Exabyte of
 storage of which approximately half is archival and half is
@@
 to jump by a factor of 10 for the HL-LHC. This increase is faster than
-projected Moore's Law gains and will present major challenges. Storage
+projected density gains in current storage technologies and thus will present major challenges. Storage
 % * <briedel@uchicago.edu> 2018-04-05T01:05:55.296Z:
@@
 What does the future of tape look like?
-%
+% changed to "density gains in current storage technologies." Tape scaling has quite a future but hard drives, less so
 % ^.
-will remain one of the visible cost drivers for HEP computing with
-the projected growth and cost of the computational resources needed to
-analyze the data expected to grow even faster than the base
+will remain one of the visible cost drivers for HEP computing. The projected growth and cost of the computational resources needed to
+analyze the data is expected to grow even faster than the base
 storage costs. The combination of storage and analysis computing costs
@@
 
-The three main challenges for data in the HL-LHC era can thus be
-summarized:
+The three main challenges for data in the HL-LHC era can be
+summarized as:
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\item \textbf{Big Data:} The HL-LHC era will significantly increase both expected
-  date rates and data volumes. Computing systems will need to
+\item \textbf{Big Data:} The expected data volume will significantly increase in the HL-LHC era. Computing systems will need to
   handle this without significant cost increases and within evolving
@@
   place new requirements on data. Specifically, the use of new types of
-  compute resources (e.g., cloud, HPC) with different dynamic
+  compute resources (e.g., cloud, HPC) with different dynamic %don't want this to be just cloud and HPC, hence "new types"
   availability and characteristics will require more dynamic
@@
   example, specific applications (e.g., training for machine learning)
-  may require use of specialized processor resources, such as GPUs,
-% * <briedel@uchicago.edu> 2018-04-04T18:14:36.074Z:
-%
-% > ,  placing further requirements on data.
-% Not sure this is needed.
-%
-% ^.
-  placing further requirements on data.
+  may require use of specialized processor resources, such as GPUs.
 \end{enumerate}
@@
 the commercial world and in the rest of the research world also provides
-a number of opportunities and solution to tackle these challenges.
+a number of opportunities and solutions to tackle these challenges.
 
@@
 \item \textbf{Data Organization:} This is essentially how data is structured
-  as it is written. Most data is written in flat files, in the ROOT format, and
-  typically with a column-wise organization of the data. The records
-  corresponding to these columns are compressed. The internal details of
-  this organization are typically visible only to individual software
-  applications.
+  as it is written.
+Most data is written in the ROOT file format.
+ROOT serializes data objects, compresses, and writes them in a column-wise manner.
+The data written in this way can only be meaningfully read  by a software that includes data object libraries.
 \item \textbf{Data Management:} The key challenge here was the transition to
-  the use of distributed computing in the form of the grid. The
-  experiments developed dedicated data catalog,  transfer and placement systems,
-  to move data between computing centers. To first
+  the use of distributed computing in the form of the grid. To tackle this, the
+  experiments developed dedicated data catalog,  transfer and placement systems. To first
   order, the computing models were rather static: data was placed at
   sites and the relevant compute jobs were sent to the right locations.
-  Applications might interact with catalogs or, at times, the workflow
-  management systems does this on behalf of the applications.
 \item \textbf{Data Access:} Various protocols are used for direct reads
-  (rfio, dcap, xrootd, etc.) with a given computer center and/or
+  (RFIO, dCap, XRootD, etc.) within a given computer center and/or
   explicit local stage-in and caching for read by jobs. Application
@@
 areas were, to first order, optimized independently. Many of the
-challenges were in the area of ''Data Management (DM)'' as the Worldwide
+challenges were in the area of ``Data Management'' (DM) as the Worldwide
 LHC Computing Grid was commissioned. As the LHC experiments' computing matured
@@
 
-Thus, in this document we take a broader view than traditional ``DM'',
+Thus, in this document we take a broader view than traditional DM,
 and consider the combination of ``Data Organization, Management and
-Access (DOMA)'' together. We believe that this full picture of data
+Access '' together. We believe that this full picture of data
 needs in HEP will provide important opportunities for efficiency and
-scalability as we enter the many-EB era.
+scalability as we enter the many-Exabyte era.
 
@@
 
-The way data are used at HEP experiments results in a dynamic range that
-spans many orders of magnitude between the IOPS transactions for data
-access and the bandwidth to storage that is required to store or
-retrieve the information. The primary data analysis workflows that drive
+HEP data transfer and access patterns require network and storage architectures to be able to sustain a large dynamic range in IOPS. The primary data analysis workflows that drive
 these bandwidth requirements can be categorized into four main
-activities. These are the contextual environment in which we present
-future evolutions for HEP data access. In particular, characteristics
-that describe these activities are enumerated for each category:
+activities:
 
@@
   reconstruction methods used today at the LHC and throughout the HEP
-  community. This will continue to be the driving characteristic of
+  community and will continue to be the driving characteristic of
   reconstruction at the HL-LHC and for liquid argon TPC experiments. The
@@
   network bandwidth make the option of streaming individual events
-  records (\emph{e.g.} full collision events or beam spill triggers)
-  across wide area networks to compute elements a feasible analysis
-  strategy. The requirements of event streaming analysis models can be
+  records (e.g., full collision events or beam spill triggers)
+  across the WAN to compute elements a feasible
+  strategy. The requirements of event streaming can be
   satisfied by a wide range of storage solutions.
@@
   seen in the physical detector systems. Simulation processes are
-  typically characterized by highly asymmetric Input/Output workflows
-  which have very small data ingest (typically just the key input
-  parameters) and produce output events which are similar in size and
-  structure to real detector data. The output of these simulation
-  processes overlap heavily with other data storage and data replication
-  processes. The requirements and challenges that this output presents
-  are treated in those contexts. The primary drivers that differentiate
+  typically characterized by highly asymmetric IO. They typically have very small data ingest (typically just the key input
+  parameters), while output data are similar in size and
+  structure to real detector data.
+Due to their similar characteristics, output data from simulation processes share the requirements and associated challenges of detector data.
+The primary drivers that differentiate
   this type of HEP workflow from the other categorizations are the
@@
   data is consumed makes it more susceptible to storage bandwidth
-  limitations and to IOPs transaction limits that underlying
+  limitations and to IOPS transaction limits that underlying
   technologies can have. In particular, the internal structure of the
   event data can dramatically impact the efficiency of data retrieval
-  and favor different access models and protocols (i.e. event
-  streaming) while the actual analysis calculations being performed can
+  and favor different access models and protocols while the actual analysis calculations being performed can
   dramatically affect the CPU-to-IO ratios. These factors make this
@@
   guard against data loss. Data replication places modest requirements
-  on the rate of IOP transactions that need to be supported, but drives
+  on the IOPS that need to be supported, but drives
   the bandwidth requirements that are needed in enacting site-to-site
-  data transfers between WAN endpoints (e.g. 10-100~Gbps). This
+  data transfers between WAN endpoints (\emph{e.g.}, 10-100~Gbps). This
   requirement on the available WAN bandwidth provides a corresponding
@@
 each category exhibit a high degree of parallelism, which allow the work
-to map readily into both high throughput computing (HTC) and high
-performance computing (HPC) environments. These domains span wide
+to map readily into both high throughput computing (HTC) and HPC environments. These domains span wide
 dynamic ranges in their requirements for concurrent data access. If the
-experiments of the HL-LHC, DUNE and other large international projects
+experiments of the HL-LHC, DUNE and other international projects
 are to utilize these facilities, they will need the ability to integrate
@@
 support evolving physical storage options similar to the the shifts that
-have occurred in recent years between mRAM, SSDs, shingled disks, tapes,
+have occurred in recent years between NVMe, SSDs, shingled disks, tapes,
 and other technologies.
@@
 shaped the prevailing data access and storage
-models that are in use today. These established patterns are centered
-around sets of event records organized into discrete collections through
-traditional file-based storage. There are, however, potential gains that
-can be made by exploring alternative data organization models and the
-corresponding access and storage models that they can enable. These
-alternative designs should be evaluated for the HL-LHC era and, if found
-to be advantageous, implemented by the community. A shift in designs may
-have the potential to reduce storage costs, shift network utilization,
-and result in more efficient use of compute resources.
+models that are in use today.
+Today's HEP data organization is file-centric with sets of events stored in a given file.
+Alternative data organization models could yield gains depending on their respective data access and storage models.
+These
+alternative designs should be evaluated for the HL-LHC era, as they may produce cost reduction, better network utilization, and efficiency gains.
 
@@
 purpose (accounting, metadata, location). Analysis access to the data,
-for instance, may benefit from being below the event level (\emph{e.g.}, physics
+for instance, may benefit from being below the event level (e.g. physics
 object). Any such catalogs may have to be external to the primary data
-store (\emph{e.g.}, relational databases that work with metadata in the primary
+store (e.g., relational databases that work with metadata in the primary
 data store). If cataloging requirements for production and
@@
 their access patterns. Similarly intelligent techniques could be used to
-place, compose, and deliver scientific data products that minimizes data
+place, compose, and deliver scientific data products, minimizing data
 movement, latency, and other costs, taking advantage of data semantics
@@
 Currently, new data-caching strategies are being deployed (e.g., XRootD
-caching proxy servers, server-less caching, ARC-style caching) and may
+caching proxy servers, server-less caching, Adaptive Replacement Cache-style caching) and may
 % * <briedel@uchicago.edu> 2018-04-05T02:09:39.626Z:
@@
 acceptance through intermediate solutions would be desirable.
+One such solution - ``Service X'' - has been proposed in \href{https://bit.ly/2Ix9NIL}{Organizing Data Lakes}.
+The service would start with the data in its persistent format, perform one or more operations on it  (\emph{e.g.} joins, query, decompression, filtering to sub-event level), and finally deliver a data stream to a client.
 
@@
 used in algorithms resulting from active research in computer science
-that study trade-offs between storage and CPU in various computing models.
+that studies trade-offs between storage and CPU in various computing models.
 
@@
 HEP experiments should be prepared to leverage ``tactical storage''.
-
-Storage that becomes most cost-effective as it becomes available (e.g.,
+Such storage may be provisioned only when it becomes cost-effective
+(e.g.,
+% * <briedel@uchicago.edu> 2018-04-05T15:43:01.526Z:
+%
+% > most
+% "more" rather than "most"?
+%
+% ^.
 from a cloud provider) and have a data management and provisioning
 system that can exploit such resources on short notice. Volatile data
+% * <briedel@uchicago.edu> 2018-04-05T15:43:37.726Z:
+%
+% > Volatile data sources would impact many aspects of the system; catalogs, job brokering, monitoring/alerting, accounting, the applications themselves.
+% Aren't we dealing with somewhat volatile storage in terms of xrootd caches today already? The file _might_ be there, but it isnt guaranteed.
+%
+% ^.
 sources would impact many aspects of the system; catalogs, job
@@
 stability.
+% * <briedel@uchicago.edu> 2018-04-05T15:49:15.646Z:
+%
+% > stability.
+% This paragraph begs the question of whether one needs to move to "direct-access-based archives" if the data is only accessed once or twice. The plenty of tape is not that great in that case.
+%
+% ^.
 
@@
 experiments and resource providers on shared solutions. This does not
-necessarily mean promoting a monoculture, as different solutions will be
+necessarily mean promoting a mono-culture, as different solutions will be
 adapted to certain major classes of use-case, type of site or funding
@@
 \begin{itemize}
-\item Quantify impact on performances
+\item Quantify impact on performance
   and resource utilization (storage, network) for the main type of
@@
 
-\emph{The previous tasks should be completed by 2020 (and can be performed in parallel).}
+\emph{The above tasks should be completed by 2020 (and can be performed in parallel).}
 
@@
   application.
-\item Evaluate possible alternative storage formats to the ROOT format, esp.\ storage formats optimized for storage efficiency and/or  data filterting through metdata
+\item Evaluate possible alternative storage formats to the ROOT format, esp.\ storage formats optimized for storage efficiency and/or  data filtering through metadata
 \end{itemize}
 
-\emph{The proposed proof-of-concept involves the previous tasks that should be completed by 2020.
+\emph{The proposed proof-of-concept involves the above tasks that should be completed by 2020.
 If successful, implementation as a production system will be done in the following years.}
@@
 
--\emph{The first two actions should be completed by 2020 but CDN/NDN
--evaluation is more long-term}
+-\emph{The first two actions should be completed by 2020 but evaluation of CDN/NDN is more long-term}
 
@@
 \subsection{Data Access and Data Access Patterns}\label{access-patterns}
-The data access pattern should inform the the use of storage media and data location. Proposed areas of inquiry are:
+The data access needs and patterns should inform the the choice of storage media and data location. Proposed areas of inquiry are:
 
@@
 \item Exploring ways to ease preliminary data filtering steps
-\item \end{itemize}
+\item Exploring commercially-popular data transfer methods (\emph{e.g,} HTTP)
+\end{itemize}
 \emph{The  proof-of-concept should be completed by 2020 and, if successful, implementation in
 production will be done in the following years.}
-
-A large portion of the LHC data is write-once, read-once. This would make it ideal to store this data on archieval media. Alternatively, a portion of the data is accessed frequently by analyzers.
 

### 2018-04-17 CWP/papers/HSF-CWP-2017-01_roadmap/authors/a2tex.py
a2tex: add footnotes in author list (JHEP style) + improvements/fixes (#76)
* a2tex: add footnotes in author list (JHEP style)

Fixes #73

* a2tex: print only footnotes relevant to the actual author list

* CWP authors: fix missing footnote for V. Gligorov

* a2tex: code refactoring

* CWP + DOMA paper: update author list with fixed a2tex

* CWP DOMA: fix author list

- Add a specific footnote for DOMA editors
- Remove mention of CWP EB members

@@ -42,11 +42,23 @@ LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
 ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
 
 
-class Footnote():
-    def __init__(self, number_str, text):
-        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
-        self.mark = footnote_symbols[int(number_str)-1]
+class AffiliationFootnote():
+    def __init__(self, text):
         self.text = text
+        self.mark = None
+
+    def set_mark(self, i, letter=False):
+        if letter:
+            if i > 26:
+                j = int(i / 26)
+                i = i % 26
+                j_str = chr(ord('a')+j-1)
+            else:
+                j_str = ''
+            self.mark = '{}{}'.format(j_str, chr(ord('a')+i-1))
+        else:
+            self.mark = str(i)
+
     def __repr__(self):
         return repr((self.mark, self.text))
 
@@ -70,14 +82,6 @@ class Author():
             raise Exeption("Got an exception({}) processing line: {}".format(e, author_line))
 
 
-class Affiliation():
-    def __init__(self,address):
-        self.address = address
-        self.mark = ''
-
-    def set_mark(self, i):
-        self.mark = str(i)
-
 def latex_escape(string):
     """
     Function to escape latex reserved characters
@@ -98,6 +102,35 @@ def latex_escape(string):
     for special_char, repl in SPECIAL_LATEX_CHARS.items():
        string = re.sub(re.escape(special_char), repl, string)
     return string
+
+
+def read_kv_file(file, separator=':', referenced_keys=None):
+    """
+    Read a text file defining key/values separated by separator, checking that
+    there is not duplicate key. If reference_keys is defined, only those keys are
+    loaded from the file.
+
+    :param file: file name
+    :param separator: separator string
+    :param referenced_keys: if defined, list of keys to consider (others are ignored)
+    :return: dictionnary of AffiliationNote object where the key is the file line key
+    """
+
+    with open(file, encoding='utf-8') as fh:
+        kv_map = {}
+        for kv_line in fh:
+            kv_line = kv_line.strip()
+            if kv_line.startswith("#") or kv_line == "":
+                continue
+            k, v = kv_line.split(separator, 1)
+            v = v.strip()
+            if k in kv_map:
+                raise Exception('Duplicated key ({}) found in {}'.format(k, file))
[…]

### 2018-04-23 CWP/papers/HSF-CWP-2017-08_framework/latex/df_fwk_summary.tex
Update df_fwk_summary.tex
Sync the google doc to this input source for the final paper.  Everything is one except for the final 3-5 year goals.

@@
 which will deliver on the order of 50 PB data events per year per
-experiment . The results of the proposed R\&D ought to be used for
+experiment. The results of the proposed R\&D ought to be used for
 building the final software systems that will be utilized in
@@
     resources, and includes GPGPUs, Tensor Processing Units (TPUs),
-    and tiered memory systems integrated with storage, ultra
+    and tiered memory systems integrated with storage, and ultra
     high-speed network interconnects.
@@
     Challenges related to advanced detector technology, like finer
-    granularity high bandwidth continuous readout DAQ, and the need
-    for a very tight feedback loop for conditions which in turn
+    granularity, high bandwidth continuous readout DAQ, and the need
+    for a very tight feedback loop for conditions, which in turn
     requires a focus on low-latency solutions.
@@
 experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff},
-which was forked to provide the art framework for the Fermilab
-Intensity Frontier experiments~\cite{Green:2012gv}.  Belle II uses
-basf2~\cite{1742-6596-331-3-032024}. The linear collider community
-developed and uses Marlin~\cite{Gaede:2006pj}. The FAIR experiments
+{[}71{]} which was forked to provide the art framework for the Fermilab
+Intensity Frontier experiments~\cite{Green:2012gv},{[}72{]}.  Belle II uses
+basf2~\cite{1742-6596-331-3-032024},{[}73{]}. The linear collider community
+developed and uses Marlin~\cite{Gaede:2006pj},{[}74{]}. The FAIR experiments
 use FairROOT, closely related to ALICE's AliROOT. The FAIR experiments
 and ALICE are now developing a new framework, which is called
-O2~\cite{O2}. At the time of writing, most major frameworks support
+O2~\cite{O2},{[}75{]}. At the time of writing, most major frameworks support
 basic parallelisation, both within and across events, based on a
-task-based model~\cite{Jones:2015soc}~\cite{Clemencic:2015paa}.
+task-based model~\cite{Jones:2015soc}~\cite{Clemencic:2015paa}. O2 already 
+includes additional multi-node setups and communication.
 
@@
 currently using a multi-process fork-and-copy-on-write solution to
-remove the constraint of one core/process. Both experiments were
+remove the constraint of one core/process, and is now moving to the 
+multithreading approach too. Both experiments were
 driven to solve this problem, by the ever growing needs for more
@@
 systems) to span resources, permitting them to be first-class
-participating in the business of scaling within a facility.
+participatents in the business of scaling within a facility. O2 provides
+a successful proof-of-principle of this approach.
 
@@
 \label{sec:roadmap}
+Forward-looking work is underway as part of projects funded through government agencies, laboratories, and collaborations.
+We want to be sure that relevant ideas and accomplishment are known, and that the groups doing this work have a place to report to and receive feedback for everyone’s benefit.
+To organize the community, one needs to establish regular working group meetings, on a bi-monthly basis as we did with the concurrency forum.
+Face to face workshops after at least the 1st and the 3rd year can be co-hosted with events like CHEP and/or the WLCG workshops.
+A future planning workshop for transforming the results of the R&D activities into a full development and deployment project plan should happen at the 5-year timescale.
 
@@
 this paper. Identify and describe where individual frameworks have
-similarly or uniquely implemented these concepts. A publishable paper
-should come of this that will serve as an agreed-upon guide for where
-we can hope to go.
+similarly or uniquely implemented these concepts. It is important to 
+describe how these choices are connected to the concrete use-cases. A 
+publishable paper should come of this that will serve as an agreed-upon 
+guide for where we can hope to go.
 
-\paragraph{Technology investigations} There are three key areas that
+\paragraph{Technology investigations} There are four key areas that
 ought to be explored to help determine future direction with regards
-to software technology. The areas are: (1) task-based programming
-tools, (2) parallel number crunching libraries, and (3) framework
-workflow management.
+to software technology. The areas are: (1) task-based programming tools,
+(2) inter-process and inter-node communication tools, (3) parallel number 
+crunching libraries, and (4) framework workflow management.
 
@@
 carried out within a framework. This includes not only expressing data
-dependencies, but also resources preferences and constraints, such a
+dependencies, but also resource preferences and constraints, such as
 GPU. The goal here is to provide enough information for a group to
 take on development of domain-specific libraries components and tools
-that will increase the efficiency of carrying out physics.
-\ifixme{the wording here needs much improvement. Maybe a good example
-  is how ML toolkits have evolved over the past few years - the
-  abstractions that have been developed have greatly increased
-  productivity and growth in the ML space}
-
-\paragraph{Coupling with ongoing R\&D activities} Forward-looking work
-is underway as part of projects funded through government agencies,
-laboratories, and collaborations. We want to be sure that relevant
-ideas and accomplishment are known, and that the groups doing this
-work have a place to report to and receive feedback for everyone's
-benefit.
+that will increase the efficiency of carrying out physics. A good example
+is how ML toolkits have evolved over the past few years. The 
+abstractions that have been developed have greatly increased productivity 
+and growth in the ML space such as the abstractions in Tensor Flow that 
+allow a coding of the matrix algebra that then gets remapped internally to 
+match the shape of the data being operated on.  The user only has take care
+of getting the domain science functions correct.
 

### 2018-05-22 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-abstract.tex
Break apart LaTeX for arXiv style
Separate the main text body (cwp-main.tex) from the LaTeX for
the specific document class used for sumbission to arXiv in
JHEP style

--- 2018-01-16 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp.tex
@@
-% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.
-
-% JHEP preprint template
-\documentclass[12pt,a4paper]{article}
-\usepackage{jheppub}
-
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage[titletoc]{appendix}
-\usepackage{xspace}
-
-\usepackage[utf8]{inputenc}
-
-\usepackage[style=numeric-comp,sorting=none]{biblatex}
-\addbibresource{cwp.bib}
-\addbibresource{cwp-chapters.bib}
-
-%% Comment out before final submission
-%\usepackage{lineno}  % for line numbering during review
-%\linenumbers
-
 \abstract{Particle physics has an ambitious and broad experimental 
@@
 describes the R\&D activities required to prepare for this software upgrade.}
-
-\begin{document}
-
-\noindent
-\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
- & & HSF-CWP-2017-01 \\
- & & December 15, 2017 \\ % use \date or hardwire e.g. December 15, 2017
- & & \\
-\end{tabular*}
-\vspace{2.0cm}
-
-\title{A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s}
-
-\author{HEP Software Foundation\footnote{Authors are listed at the end of this report.}}
-
-\maketitle
-
-\newpage
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Particle physics has an ambitious experimental programme for the
-coming decades. The programme supports the strategic goals of the
-particle physics community that have been laid out by the European
-Strategy for Particle Physics~\cite{ESPP2013} and by
-the Particle Physics Project Prioritization Panel (P5)~\cite{DOE-P5} in the United
-States~\cite{2014pwa}. Broadly speaking, the
-scientific goals are:
-
-\begin{itemize}
-\item Exploit the discovery of the Higgs boson as a precision tool
-  for investigating Standard Model (SM) and Beyond the Standard Model
-  (BSM) physics.
-\item
-  Study the decays of $b$- and $c$-hadrons, and tau leptons, in the search
-  for manifestations of BSM physics, and investigate
-  matter-antimatter differences.
-\item
-  Search for signatures of dark matter.
-\item
-  Probe neutrino oscillations and masses.
-\item
-  Study the Quark Gluon Plasma state of matter in heavy-ion collisions.
-\item
-  Explore the unknown.
-\end{itemize}
-
-The High-Luminosity Large Hadron Collider
-(HL-LHC)~\cite{HL-LHC,1742-6596-515-1-012012,Apollinari:2284929} 
-will be a major
-upgrade of the current LHC~\cite{LHC} supporting the aim of 
-an in-depth investigation of the properties of the Higgs boson and its
-couplings to other particles (Figure~\ref{fig:hl-lhc}). The ATLAS~\cite{ATLAS} and CMS~\cite{CMS}
-collaborations will continue to make measurements in the Higgs sector,
-while searching for new physics Beyond the Standard Model (BSM). Should
-a BSM discovery be made, a full exploration of that physics will be
-pursued. Such BSM physics may help shed light on the nature of dark
-matter, which we know makes up the majority of gravitational matter in
-the universe, but which does not interact via the electromagnetic or
-strong nuclear forces~\cite{Mangano2016}.
-
-The LHCb experiment at the LHC~\cite{LHCb} and the Belle II experiment at
-KEK~\cite{BelleII} study various aspects of heavy
-flavour physics (b- and c-quark, and tau-lepton physics), where quantum
-influences of very high mass particles manifest themselves in lower
-energy phenomena. Their primary goal is to look for BSM physics, either
-by studying CP violation (that is, asymmetries in the behaviour of
-particles and their corresponding antiparticles) or modifications in
-rate or angular distributions in rare heavy-flavour decays. Current
-manifestations of such asymmetries do not explain why our universe is so
-matter dominated. These flavour physics programmes are related to BSM
-searches through effective field theory, and powerful constraints on new
-physics keep coming from such studies.
-
-\begin{figure}[bthp]
-    \centering
-    \includegraphics[width=0.94\textwidth]{HL-LHC-plan-2017-Plan-2.pdf}
-    \caption{The current schedule for the LHC and HL-LHC upgrade
-and run~\cite{HL-LHC}. Currently, the start of the HL-LHC run is foreseen for mid
-2026. The long shutdowns, LS2 and LS3, will be used to upgrade both the
-accelerator and the detector hardware.}
-    \label{fig:hl-lhc}
-\end{figure}
-
-The study of neutrinos, their mass and oscillations, can also shed light
-on matter-antimatter asymmetry. The DUNE experiment will provide a huge
-improvement in our ability to probe neutrino physics, detecting
-neutrinos from the Long Baseline Neutrino Facility at Fermilab, as well
-as linking to astro-particle physics programmes, in particular through
-the potential detection of supernovas and relic neutrinos. An overview
-of the experimental programme scheduled at the Fermilab facility is
-given in Figure~\ref{fig:fnal-if}.
-
-\begin{figure}
-\vspace*{0.3cm}
-    \centering
-    \includegraphics[width=0.94\textwidth]{10yr-PLAN-9Feb-2017-V7a.pdf}
-    \caption{Run schedule for the Fermilab facility until 2026.}
-    \label{fig:fnal-if}
-\end{figure}
-
-In the study of the early universe immediately after the Big Bang, it is
-critical to understand the phase transition between the highly
-compressed quark-gluon plasma and the nuclear matter in the universe
-today. The ALICE experiment at the LHC~\cite{ALICE} and
-the CBM~\cite{CBM} and PANDA~\cite{PANDA} experiments at the Facility for Antiproton and Ion
-Research (FAIR) are specifically designed to probe this aspect of
-nuclear and particle physics. In addition ATLAS, CMS and LHCb all
-contribute to the LHC heavy-ion programme.
-
-These experimental programmes require large investments in detector
-hardware, either to build new facilities and experiments (e.g., FAIR and DUNE) or to upgrade existing ones (HL-LHC, Belle II). Similarly, they
-require commensurate investment in the research and development
-necessary to deploy software to acquire, manage, process, and analyse
-the data recorded.
-
-For the HL-LHC, which is scheduled to begin taking data in 2026 (Figure~\ref{fig:hl-lhc}) and to run into the 2030s, some 30 times more data than
-the LHC has currently produced will be collected by ATLAS and CMS. As
-the total amount of LHC data already collected is close to an exabyte,
-it is clear that the problems to be solved require approaches beyond
-simply scaling current solutions, assuming Moore's Law and more or less
-constant operational budgets. The nature of computing hardware
-(processors, storage, networks) is evolving with radically new
-paradigms, the quantity of data to be processed is increasing
-dramatically, its complexity is increasing, and more sophisticated
-analyses will be required to maximise physics yield. Developing and
-deploying sustainable software for future and upgraded experiments,
-given these constraints, is both a technical and a social challenge, as
-detailed in this paper. An important message of this report is that a
-``software upgrade'' is needed to run in parallel with the hardware
-upgrades planned for the HL-LHC in order to take full advantage of these hardware upgrades and to complete the HL-LHC physics programme.
-
-In planning for the HL-LHC in particular, it is critical that all of the
-collaborating stakeholders agree on the software goals and priorities,
-and that the efforts complement each other. In this spirit, the HEP
-Software Foundation (HSF) began a planning exercise in late 2016 to
-prepare a Community White Paper (CWP)~\cite{HSF2017} at the behest of the Worldwide LHC
-Computing Grid (WLCG) project~\cite{WLCG2016}. The role of the HSF is to
-facilitate coordination and common efforts in HEP software and computing
-internationally and to provide a structure for the community to set
-goals and priorities for future work. The objective of the CWP is to
-provide a roadmap for software R\&D in preparation for the HL-LHC and
-for other HEP experiments on a similar timescale, which would identify
-and prioritise the software research and development investments
-required:
-
-\begin{itemize}
-\item
-  to achieve improvements in software efficiency, scalability and
-  performance, and to make use of advances in CPU, storage and network
-  technologies in order to cope with the challenges ahead;
-\item
-  to enable new approaches to computing and software that can radically
-  extend the physics reach of the detectors;
-\item
-  to ensure the long-term sustainability of the software through the
-  lifetime of the HL- LHC;
-\item
-  to ensure data and knowledge preservation beyond the lifetime of
-  individual experiments;
-\item
-  to attract the required new expertise by offering appropriate career
-  recognition to physicists specialising in software development, and by
-  an effective training effort to target all contributors in the
-  community.
-\end{itemize}
-
-The CWP process, organised by the HSF with the participation of the LHC
-experiments and the wider HEP software and computing community, began
-with a kick-off workshop at the San Diego Supercomputer Centre (SDSC),
-USA, in January 2017 and concluded after a final workshop in June 2017 at
-the Laboratoire d'Annecy de Physique des Particules (LAPP), France, with
-a large number of intermediate topical workshops and meetings
-(Appendix \ref{appendix-a---list-of-workshops}). The
-entire CWP process involved an estimated 250 participants.
-
-To reach more widely than the LHC experiments, specific contact was made
-with individuals with software and computing responsibilities in the
-Fermilab muon and neutrino experiments, Belle II, the Linear Collider
-community, as well as various national computing organisations. The CWP
-process was able to build on all the links established since the
-inception of the HSF in 2014.
-
-Working groups were established on various topics which were expected to
-be important parts of the HL-LHC roadmap: \emph{Careers, Staffing and
-Training; Conditions Database; Data Organisation, Management and Access;
-Data Analysis and Interpretation; Data and Software Preservation;
-Detector Simulation; Data-Flow Processing Frameworks; Facilities and
-Distributed Computing; Machine Learning; Physics Generators; Security;
-Software Development, Deployment and Validation/Verification; Software
-Trigger and Event Reconstruction;} and \emph{Visualisation.} The work of
-each working group is summarised in this document.
-
-This document is the result of the CWP process. Investing in the roadmap
-outlined here will be fruitful for the whole of the HEP programme and
-may also benefit other projects with similar technical challenges,
-particularly in astrophysics, e.g., the Square Kilometre Array (SKA)~\cite{SKA},
-the Cherenkov Telescope Array (CTA)~\cite{CTA} and the Large Synoptic Survey
-Telescope (LSST)~\cite{LSST}.
-
-\hypertarget{software-and-computing-challenges}{%
-\section{Software and Computing
-Challenges}\label{software-and-computing-challenges}}
-
-Run 2 for the LHC started in 2015 and delivered a proton-proton
-collision energy of
-\href{http://home.cern/about/engineering/restarting-lhc-why-13-tev}{13
-TeV}. By the end of LHC Run 2 in 2018, it is expected that about 150
-fb\textsuperscript{-1} of physics data will have been collected by both
-ATLAS and CMS. Together with ALICE and LHCb, the total size of LHC data
-storage pledged by sites for the year 2017 is around 1 exabyte, as shown
-in Table~\ref{tab:crsg2017} from the LHC's Computing Resource Scrutiny Group
-(CRSG)~\cite{Lucchesi:2284575}. The CPU allocation from the CRSG for 2017 to each
-experiment is also shown.
-
-\begin{table}
-    \centering
-\begin{tabular}{lp{0.15\textwidth}p{0.15\textwidth}p{0.15\textwidth}p{0.15\textwidth}}
-\hline
-\textbf{Experiment} & \textbf{2017 Disk Pledges (PB)} & \textbf{2017 Tape Pledges (PB)} & \textbf{Total Disk and Tape Pledges (PB)} & \textbf{2017 CPU Pledges (kHS06)}\\
-\hline
-ALICE & 67 & 68 & 138 & 807\\
-ATLAS & 172 & 251 & 423 & 2194\\
-CMS & 123 & 204 & 327 & 1729\\
-LHCb & 35 & 67 & 102 & 413\\
-\textbf{Total} & \textbf{400} & \textbf{591} & \textbf{990} &
-\textbf{5143}\\
-\hline
-\end{tabular}
-    \caption{Resources pledged by WLCG sites to the 4 LHC experiments for
-the year 2017 as described at the September 2017 session of the
-Computing Resources Scrutiny Group (CRSG).}
-    \label{tab:crsg2017}
-\end{table}
-
-Using an approximate conversion from HS06~\cite{HS06} to CPU cores of 10
-means that LHC computing in 2017 is supported by about 500k CPU cores.
-These resources are deployed ubiquitously, from close to the experiments
-themselves at CERN to a worldwide distributed computing infrastructure,
-the WLCG~\cite{WLCG}. Each experiment has developed its own workflow management and
-data management software to manage its share of WLCG resources.
-
-In order to process the data, the 4 largest LHC experiments have written
-more than 20 million lines of program code over the last 15 years. This
-has involved contributions from thousands of physicists and many
-computing professionals, encompassing a wide range of skills and
-abilities. The majority of this code was written for a single
-architecture (x86\_64) and with a serial processing model in mind. There
-is considerable anxiety in the experiments that much of this software is
-not sustainable, with the original authors no longer in the field and
-much of the code itself in a poorly maintained state, ill-documented, and
-lacking tests. This code, which is largely experiment-specific, manages
-the entire experiment data flow, including data acquisition, high-level
-triggering, calibration and alignment, simulation, reconstruction (of
-both real and simulated data), visualisation, and final data analysis.
-
-HEP experiments are typically served with a large set of integrated and
-configured common software components, which have been developed either
-in-house or externally. Well-known examples include ROOT~\cite{Brun1996}, which is a data analysis toolkit that
-also plays a critical role in the implementation of experiments' data
-storage systems, and Geant4~\cite{Agostinelli2003}, a
-simulation framework through which most detector simulation is achieved.
-Other packages provide tools for supporting the development process;
-they include compilers and scripting languages, as well as tools for
-integrating, building, testing, and generating documentation. Physics
-simulation is supported by a wide range of event generators provided by
-the theory community (PYTHIA~\cite{PYTHIA}, SHERPA~\cite{Gleisberg:2008ta}, ALPGEN~\cite{Mangano:565290}, MADGRAPH~\cite{MADGRAPH}, HERWIG~\cite{HERWIG},
-amongst many others). There is also code developed to support the
-computing infrastructure itself, such as the CVMFS distributed caching
-filesystem~\cite{1742-6596-331-4-042003}, the Frontier database caching 
-mechanism~\cite{Frontier}, the XRootD file access software~\cite{XRootD} and a
-number of storage systems (dCache, DPM, EOS). This list of packages is % REF?
-by no means exhaustive, but illustrates the range of software employed
-and its critical role in almost every aspect of the programme.
-
-Already in Run 3 LHCb will process more than 40 times the number of
-collisions that it does today, and ALICE will read out Pb-Pb collisions
-continuously at 50 kHz. The upgrade to the HL-LHC for Run 4 then
-produces a step change for ATLAS and CMS. The beam intensity will rise
-substantially, giving bunch crossings where the number of discrete
-proton-proton interactions (pileup) will rise to about 200, from about 60
-today. This has important consequences for the operation of the
-detectors and for the performance of the reconstruction software. The
-two experiments will upgrade their trigger systems to record 5-10 times
-as many events as they do today. It is anticipated that HL-LHC will
-deliver about 300 fb\textsuperscript{-1} of data each year.
-
-The steep rise in resources that are then required to manage this data
-can be estimated from an extrapolation of the Run 2 computing model and
-is shown in Figures~\ref{fig:cms-resources} and~\ref{fig:atlas-resources}.
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.47\textwidth}
-        \includegraphics[width=\textwidth]{cms-cpu.png}
-        \caption{}
-        \label{fig:cms-cpu}
-    \end{subfigure}
-    \quad
-    \begin{subfigure}[b]{0.47\textwidth}
-        \includegraphics[width=\textwidth]{cms-disk.png}
-        \caption{}
-        \label{fig:cms-disk}
-    \end{subfigure}
-    \caption{CMS estimated CPU (\ref{fig:cms-cpu}) and disk space (\ref{fig:cms-disk}) 
-    resources required into the HL-LHC era, using the current computing model 
-    with parameters projected out for the next 12 years.}
-    \label{fig:cms-resources}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.9\textwidth}
-    	    \centering
-        \includegraphics[width=0.6\textwidth]{atlas-cpuHLLHC.pdf}
-        \caption{Estimated CPU resources (in kHS06) needed for the years 2018 to 2028 for both data and simulation processing.
-The blue points are estimates based on the current software performance estimates and using the ATLAS computing model parameters from 2017.
-The solid line shows the amount of resources expected to be available if a flat funding scenario is assumed, which implies an increase of 20\% per year, based on the current technology trends.}
-        \label{fig:atlas-cpu}
-    \end{subfigure}
-    \quad
-    \begin{subfigure}[b]{0.9\textwidth}
-    	    \centering
-        \includegraphics[width=0.6\textwidth]{atlas-diskHLLHC.pdf}
-        \caption{Estimated total disk resources (in PB) needed for the years 2018 to 2028 for both data and simulation processing.
-The blue points are estimates based on the current event sizes estimates and using the ATLAS computing model parameters from 2017.
-The solid line shows the amount of resources expected to be available if a flat funding scenario is assumed, which implies an increase of 15\% per year, based on the current technology trends.}
-        \label{fig:atlas-disk}
-    \end{subfigure}
-    \caption{ATLAS resources required into the HL-LHC era, using the current computing model 
-    and software performance.\cite{ATLAS-HL-LHC-Plots}}
-    \label{fig:atlas-resources}
-\end{figure}
-
-In general, it can be said that the amount of data that experiments can
-collect and process in the future will be limited by affordable software
-and computing, and therefore the physics reach during HL-LHC will be
-limited by how efficiently these resources can be used.
-
-The ATLAS numbers, in Figure~\ref{fig:atlas-resources}, are particularly interesting as they
-estimate the resources that will be available to the experiment if a
-flat funding profile is maintained, taking into account the expected
-technology improvements given current trends~\cite{Panzer2017}. As can be
-seen, the shortfall between needs and bare technology gains is
-considerable: a factor 4 in CPU and a factor 7 in disk in 2027.
-
-While the density of transistors on silicon continues to increase
-following Moore's Law (albeit more slowly than in the past), power
-density constraints have limited the clock speed of processors for more
-than a decade. This has effectively stalled any progress in the
-processing capacity of a single CPU core. Instead, increases in
-potential processing capacity come from increases in the core count of
-CPUs and wide CPU registers. Alternative processing architectures have
-become more commonplace. These range from the many-core architecture
-based on standard x86\_64 cores to numerous alternatives such as GPUs.
-For GPUs, the processing model is very different, allowing a much
-greater fraction of the die to be dedicated to arithmetic calculations,
-but at a price in programming difficulty and memory handling for the
-developer that tends to be specific to each processor generation.
-Further developments may even see the use of FPGAs for more
-general-purpose tasks. Fully exploiting these evolutions requires a
-shift in programming model to one based on \emph{concurrency}.
-
-Even with the throttling of clock speed to limit power consumption,
-power remains a major issue. Low power architectures are in huge demand.
-At one level this might challenge the dominance of x86\_64 by simply
-replacing it with, for example, AArch64 devices that may achieve lower power
-costs for the scale of HEP computing needs than Intel has achieved with its
-Xeon architecture. More extreme is an architecture that would see
-specialised processing units dedicated to particular tasks, but with
-possibly large parts of the device switched off most of the time,
-so-called dark silicon.
-
-Limitations in affordable storage also pose a major challenge, as does
-the I/O rates of higher capacity hard disks. Network
-bandwidth will probably continue to increase at the required level, but
-the ability to use it efficiently will need a closer integration with
-applications. This will require software developments to
-support distributed computing (data and workload management, software
-distribution and data access) and an increasing awareness of the
-extremely hierarchical view of data, from long latency tape access and
-medium-latency network access through to the CPU memory hierarchy.
-
-Taking advantage of these new architectures and programming paradigms
-will be critical for HEP to increase the ability of our code to deliver
-physics results efficiently, and to meet the processing challenges of
-the future. Some of this work will be focused on re-optimised
-implementations of existing algorithms. This will be complicated by the
-fact that much of our code is written for the much simpler model of
-serial processing, and without the software engineering needed for
-sustainability. Proper support for taking advantage of concurrent
-programming techniques, such as vectorisation and thread-based
-programming, through frameworks and libraries, will be essential, as the
-majority of the code will still be written by physicists. Other
-approaches should examine new algorithms and techniques, including
-highly parallelised code that can run on GPUs or the use of machine
-learning techniques to replace computationally expensive pieces of
-simulation or pattern recognition. The ensemble of computing work that
-is needed by the experiments must remain sufficiently flexible to take
-advantage of different architectures that will provide computing to HEP
-in the future. The use of high performance computing sites and
-commercial cloud providers will very likely be a requirement for the
-community and will bring particular constraints and demand flexibility.
-
-These technical challenges are accompanied by significant human
-challenges. Software is written by many people in the collaborations,
-with varying levels of expertise, from a few experts with precious
-skills to novice coders. This implies organising training in effective
-coding techniques and providing excellent documentation, examples and
-support. Although it is inevitable that some developments will remain
-within the scope of a single experiment, tackling software problems
-coherently as a community will be critical to achieving success in the
-future. This will range from sharing knowledge of techniques and best
-practice to establishing common libraries and projects that will provide
-generic solutions to the community. Writing code that supports a wider
-subset of the community than just a single experiment will almost
-certainly be mandated upon HEP and presents a greater challenge, but the
-potential benefits are huge. Attracting, and retaining, people with the
-required skills who can provide leadership is another significant
-challenge, since it impacts on the need to give adequate recognition to
-physicists who specialise in software development. This is an important
-issue that is treated in more detail later in the report.
-
-Particle physics is no longer alone in facing these massive data
-challenges. Experiments in other fields, from astronomy to genomics,
-will produce huge amounts of data in the future, and will need to
-overcome the same challenges that we face, i.e., massive data handling and
-efficient scientific programming. Establishing links with these fields
-has already started. Additionally, interest from the computing science
-community in solving these data challenges exists, and mutually
-beneficial relationships would be possible where there are genuine
-research problems that are of academic interest to that community and
-provide practical solutions to ours. The efficient processing of massive
-data volumes is also a challenge faced by industry, in particular the
-internet economy, which developed novel and major new technologies under
-the banner of \emph{Big Data} that may be applicable to our use cases.
-
-Establishing a programme of investment in software for the HEP
-community, with a view to ensuring effective and sustainable software
-for the coming decades, will be essential to allow us to reap the
-physics benefits of the multi-exabyte data to come. It was in recognition of
-this fact that the HSF itself was set up and already works to promote
-these common projects and community developments~\cite{HSF2015}.
-
-\hypertarget{programme-of-work}{%
-\section{Programme of Work}\label{programme-of-work}}
-
-In the following we describe the programme of work being proposed for
-the range of topics covered by the CWP working groups. We summarise the
-main specific challenges each topic will face, describe current
-practices, and propose a number of R\&D tasks that should be undertaken
-in order to meet the challenges. R\&D tasks are grouped in two different
-timescales: short term (by 2020, in time for the HL-LHC Computing Technical Design Reports of
-ATLAS and CMS) and longer-term actions (by 2022, to be ready for testing
-or deployment during LHC Run 3).
-
-\hypertarget{physics-generators}{%
-\subsection{Physics Generators}\label{physics-generators}}
-
-\subsubsection*{Scope and Challenges}
-
-Monte-Carlo event generators are a vital part of modern particle
-physics, providing a key component of the understanding and
-interpretation of experiment data. Collider experiments have a need for
-theoretical QCD predictions at very high precision. Already in LHC Run
-2, experimental uncertainties for many analyses are at the same level as,
-or lower than, those from theory. Many analyses have irreducible
-QCD-induced backgrounds, where statistical extrapolation into the signal
-region can only come from theory calculations. With future experiment
-and machine upgrades, as well as reanalysis of current data, measured
-uncertainties will shrink even further, and this will increase the need
-to reduce the corresponding errors from theory.
-
-Increasing accuracy will compel the use of higher-order perturbation
-theory generators with challenging computational demands. Generating
-Monte Carlo events using leading order (LO) generators is only a small
-part of the overall computing requirements for HEP experiments.
-Next-to-leading order (NLO) event generation, used more during LHC Run
-2, is already using significant resources. Higher accuracy
-theoretical cross sections calculated at next-to-next-to-leading (NNLO),
-already important in some Run 2 analyses, are not widely used
-because of computational cost. By HL-LHC the use of NNLO event
-generation will be more widely required, so these obstacles to their
-adoption must be overcome.
-Increasing the order of the generators increases greatly the complexity
-of the phase space integration required to calculate the appropriate QCD
-matrix elements. The difficulty of this integration arises from the need
-to have sufficient coverage in a high-dimensional space (10-15
-dimensions, with numerous local maxima); the appearance of negative event
-weights; and the fact that many terms in the integration cancel, so that
-a very high degree of accuracy of each term is required. Memory demands
-for generators have generally been low and initialisation times have
-been fast, but an increase in order means that memory consumption
-becomes important and initialisation times can become a significant 
-fraction of the job's run time.
-
-For HEP experiments, in many cases, meaningful predictions can only be
-obtained by combining higher-order perturbative calculations with parton
-showers. This procedure is also needed as high-multiplicity final states
-become more interesting at higher luminosities and event rates. Matching
-(N)NLO fixed-order calculations to parton shower algorithms can have a
-very low efficiency, and increases further the computational load needed
-to generate the necessary number of particle-level events. In addition,
-many of the current models for the combination of parton-level event
-generators and parton shower codes are incompatible with requirements
-for concurrency on modern architectures. It is a major challenge to
-ensure that this software can run efficiently on next generation
-hardware and software systems.
-
-Developments in generator software are mainly done by the HEP theory
-community. Theorists typically derive career recognition and advancement from
-making contributions to theory itself, rather than by making improvements
-to the computational efficiency of generators \emph{per se}. So,
-improving the computational efficiency of event generators, and allowing
-them to run effectively on resources such as high performance computing
-facilities (HPCs), will mean engaging with experts in computational
-optimisation who can work with the theorists who develop generators.
-
-The challenge in the next decade is to advance the theory and practical
-implementation of event generators to support the needs of future
-experiments, reaching a new level of theory precision and recognising
-the demands for computation and computational efficiency that this will
-bring.
-
-\subsubsection*{Current Practice}
-
-Extensive use of LO generators and parton shower algorithms are still
-made by most HEP experiments. Each experiment has its own simulation
-needs, but for the LHC experiments tens of billions of generated events
-are now used each year for Monte Carlo simulations. During LHC Run 2 more and
-more NLO generators were used, because of their increased theoretical
-precision and stability. The raw computational complexity of NLO
-amplitudes, combined with many-body phase-space evaluations and the
-inefficiencies of the matching process, leads to a potentially much-increased CPU
-budget for physics event simulation for ATLAS and CMS.
-
-The use of NLO generators by the experiments today is also limited because of
-the way the generators are implemented, producing significant numbers of
-negative event weights. This means that the total number of events the
-experiments need to generate, simulate, and reconstruct can be many 
-times larger for NLO than for LO samples. At the same time, the
-experiments budget only a similar number of Monte Carlo simulation events as from the real data. Having large NLO samples is thus not consistent with
-existing computing budgets until a different scheme is developed that
-does not depend on negative event weights or produces them only at a
-significantly reduced rate.
-
-While most event generation is run on ``standard'' grid resources,
-effort is ongoing to run more demanding tasks on HPC resources, e.g.,
-W-boson + 5-jet events at the Argonne Mira HPC). However, scaling for
-efficient running on some of the existing HPC resources is not trivial
-and requires effort.
-
-Standard HEP libraries such as LHAPDF~\cite{LHAPDF}, HepMC\cite{HepMC}, 
-and Rivet~\cite{Rivet} are used by the generators for integration into
-the experiments' event generation workflows. These require extensions
-and sustained maintenance that should be considered a shared
-responsibility of the theoretical and experimental communities in the
-context of large-scale experiments. In practice, however, it has been
-difficult to achieve the level of support that is really needed as there
-has been a lack of recognition for this work. To help improve the
-capabilities and performance of generators as used by the experimental
-HEP programme, and to foster interaction between the communities, the 
-MCnet~\cite{MCnet} short-term studentship programme has been very useful.
-Interested experimental PhD students can join a generator group for
-several months to work on improving a physics aspect of the simulation
-that is relevant to their work, or to improve the integration of the
-generator into an experimental framework.
-
-\subsubsection*{Research and Development Programme}
-
-As the Monte Carlo projects are funded mainly to develop theoretical
-improvements, and not mainly as ``suppliers'' to the experimental HEP
-programme, any strong requests towards efficiency improvements from the
-experimental community would need to be backed up by plausible avenues
-of support that can fund contributions from software engineers with the
-correct technical skills in software optimisation to work within the
-generator author teams.
-
-In a similar way to the MCnet studentships, a matchmaking scheme could
-focus on the software engineering side, and transfer some of the
-expertise available in the experiments and facilities teams to the
-generator projects. Sustainable improvements are unlikely to be
-delivered by graduate students ``learning on the job'' and then leaving
-after a few months, so meeting the requirement of transferring technical
-expertise and effort will likely require placements for experienced
-optimisation specialists and a medium- to long-term connection to the
-generator project.
-
-HEP experiments, which are now managed by very large collaborations including many
-technical experts, can also play a key role in sustaining a healthy
-relationship between theory and experiment software. Effort to work on
-common tools that benefit both the experiment itself and the wider
-community would provide shared value that justifies direct investment
-from the stakeholders. This model would also be
-beneficial for core HEP tools like LHAPDF, HepMC and Rivet, where future
-improvements have no theoretical physics interest anymore, putting them
-in a similar situation to generator performance improvements. One
-structural issue blocking such a mode of operation is that some
-experiments do not currently recognise contributions to external
-projects as experiment service work --- a situation deserving of review
-in areas where external software tools are critical to experiment
-success.
-
-In the following we describe specific areas of R\&D for event generation 
-up to 2022 and beyond.
-
-\begin{itemize}
-\item
-  The development of new and improved theoretical algorithms provides
-  the largest potential for improving event generators. While it is not
-  guaranteed that simply increasing the effort dedicated to this task
-  will bring about the desired result, the long-term support of event
-  generator development, and the creation of career opportunities in
-  this research area, are critical given the commitment to experiments
-  on multi-decade scales.
-\item
-  Expand development in reweighting event samples, where new physics
-  signatures can be explored by updating the partonic weights according
-  to new matrix elements. It is necessary that the phase space for the
-  updated model be a subset of the original one, which is an important
-  limitation. The procedure is more complex at NLO and can require
-  additional information to be stored in the event files to properly
-  reweight in different cases. Overcoming the technical issues from
-  utilising negative event weights is crucial. Nevertheless, the method
-  can be powerful in many cases, and would hugely reduce the time needed
-  for the generation of BSM samples.
-\item
-  At a more technical level, concurrency is an avenue that has yet to be
-  explored in depth for event generation. As the calculation of matrix
-  elements requires VEGAS-style integration, this work would be helped
-  by the development of a new Monte-Carlo integrator. For multi-particle
-  interactions, factorising the full phase space integration into lower
-  dimensional integrals would be a powerful method of parallelising,
-  while the interference between different Feynman graphs can be handled
-  with known techniques.
-\item
-  For many widely used generators, basic problems of concurrency and
-  thread hostility need to be tackled, to make these packages suitable
-  for efficient large scale use on modern processors and within modern
-  HEP software frameworks. Providing appropriate common tools for
-  interfacing, benchmarking and optimising multithreaded code would
-  allow expertise to be shared effectively~\cite{HSF-CWP-2017-13}.
-\item
-  In most generators, parallelism was added post-facto, which leads to
-  scaling problems when the level of parallelism becomes very large,
-  e.g., on HPC machines. These HPC machines will be part of the
-  computing resource pool used by HEP, so solving scaling issues on
-  these resources for event generation is important, particularly as the
-  smaller generator code bases can make porting to non-x86\_64
-  architectures more tractable. The problem of long and inefficient
-  initialisation when a job utilises hundreds or thousands of cores on
-  an HPC needs to be tackled. While the memory consumption of event
-  generators is generally modest, the generation of tree-level
-  contributions to high multiplicity final states can use significant
-  memory, and gains would be expected from optimising here.
-\item
-  An underexplored avenue is the efficiency of event generation as
-  used by the experiments. An increasingly common usage is to generate
-  very large inclusive event samples, which are filtered on event
-  final-state criteria to decide which events are to be retained and
-  passed onto detector simulation and reconstruction. This naturally
-  introduces a large waste of very CPU-expensive event generation, which
-  could be reduced by developing filtering tools within the generators
-  themselves, designed for compatibility with the experiments'
-  requirements. A particularly wasteful example is where events are
-  separated into orthogonal subsamples by filtering, in which case the
-  same large inclusive sample is generated many times, with each stream
-  filtering the events into a different group: allowing a single
-  inclusive event generation to be filtered into several orthogonal
-  output streams would improve efficiency.
-\end{itemize}
-
-\hypertarget{detector-simulation}{%
-\subsection{Detector Simulation}\label{detector-simulation}}
-
-\subsubsection*{Scope and Challenges}
-
-For all its success so far, the challenges faced by the HEP field in the
-simulation domain are daunting. During the first two runs, the LHC
-experiments produced, reconstructed, stored, transferred, and analysed
-tens of billions of simulated events. This effort required more than
-half of the total computing resources allocated to the experiments. As
-part of the HL-LHC physics programme, the upgraded
-experiments expect to collect 150 times more data than in Run 1; demand
-for larger simulation samples to satisfy analysis needs will grow
-accordingly. In addition, simulation tools have to serve diverse
-communities, including accelerator-based particle physics research
-utilising proton-proton colliders, neutrino, dark matter, and muon
-experiments, as well as the cosmic frontier. The complex detectors of
-the future, with different module- or cell-level shapes, finer
-segmentation, and novel materials and detection techniques, require
-additional features in geometry tools and bring new demands on physics
-coverage and accuracy within the constraints of the available computing
-budget. The diversification of the physics programmes also requires new
-and improved physics models. More extensive use of Fast Simulation is a
-potential solution, under the assumption that it is possible to improve
-time performance without an unacceptable loss of physics accuracy.
-
-The gains that can be made by speeding up critical elements of the
-Geant4 simulation toolkit can be leveraged for all applications that use
-it, and it is therefore well worth the investment in effort needed to
-achieve it. The main challenges to be addressed if the required physics
-and software performance goals are to be achieved are:
-
-\begin{itemize}
-\item
-  Reviewing the physics models' assumptions, approximations, and
-  limitations in order to achieve higher precision, and to extend the
-  validity of models up to energies of the order of 100 TeV foreseen
-  with the Future Circular Collider (FCC) project~\cite{FCC}.
-\item
-  Redesigning, developing, and commissioning detector simulation
-  toolkits to be more efficient when executed on current vector CPUs and
-  emerging new architectures, including GPUs, where use of SIMD
-  vectorisation is vital; this includes porting and optimising the
-  experiments' simulation applications to allow exploitation of large
-  HPC facilities.
-\item
-  Exploring different Fast Simulation options, where the full detector
-  simulation is replaced, in whole or in part, by computationally
-  efficient techniques. An area of investigation is common frameworks
-  for fast tuning and validation.
-\item
-  Developing, improving and optimising geometry tools that can be shared
-  am\-ong experiments to make the modeling of complex detectors
-  computationally more efficient, modular, and transparent.
-\item
-  Developing techniques for background modeling, including contributions
-  of multiple hard interactions overlapping the event of interest in
-  collider experiments (pileup).
-\item
-  Revisiting digitisation algorithms to improve performance and
-  exploring opportunities for code sharing among experiments.
-\item
-  Recruiting, training, retaining human resources in all areas of
-  expertise pertaining to the simulation domain, including software and
-  physics.
-\end{itemize}
-
-It is obviously of critical importance that the whole community of
-scientists working in the simulation domain continue to work together in
-as efficient a way as possible in order to deliver the required
-improvements. Very specific expertise is required across all simulation
-domains, such as physics modeling, tracking through complex geometries
-and magnetic fields, and building realistic applications that accurately
-simulate highly complex detectors. Continuous support is needed to
-recruit, train, and retain people with a unique set of skills needed to
-guarantee the development, maintenance, and support of simulation codes
-over the long timeframes foreseen in the HEP experimental programme.
-
-\subsubsection*{Current Practices}
-
-The Geant4 detector simulation toolkit is at the core of simulation in
-almost every HEP experiment. Its continuous development, maintenance,
-and support for the experiments is of vital importance. New or refined
-functionality in physics coverage and accuracy continues to be delivered
-in the ongoing development programme and software
-performance improvements are introduced whenever possible.
-
-Physics models are a critical part of the detector simulation, and are
-continuously being reviewed, and in some cases reimplemented, in order
-to improve accuracy and software performance. Electromagnetic (EM)
-transport simulation is challenging as it occupies a large part of the
-computing resources used in full detector simulation. Significant
-efforts have been made in the recent past to better describe the
-simulation of electromagnetic shower shapes, in particular to model the
-$H \to \gamma \gamma$ signal and background accurately at the LHC. This
-effort is being continued with an emphasis on reviewing the models'
-assumptions, approximations, and limitations, especially at very high
-energy, with a view to improving their respective software
-implementations. In addition, a new ``theory-based'' model
-(Goudsmit-Saunderson), for describing the \emph{multiple scattering} of
-electrons and positrons, has been developed that has been demonstrated
-to outperform, in terms of physics accuracy and speed, the current
-models in Geant4. The models used to describe the \emph{bremsstrahlung}
-process have also been reviewed, and recently an improved theoretical
-description of the Landau-Pomeranchuk-Migdal effect was introduced
-that plays a significant role at high energies. Theoretical review of
-all electromagnetic models, including those of hadrons and ions, is
-therefore of high priority both for HL-LHC and for FCC studies.
-
-Hadronic physics simulation covers purely hadronic interactions. It is
-not possible for a single model to describe all the physics encountered
-in a simulation due to the large energy range that needs to be covered
-and the simplified approximations that are used to overcome the
-difficulty of solving the full theory (QCD). Currently the most-used
-reference physics list for high energy and space applications is
-FTFP\_BERT. It uses the Geant4 Bertini cascade for hadron--nucleus
-interactions from 0 to 12 GeV incident hadron energy and the FTF parton
-string model for hadron--nucleus interactions from 3 GeV upwards.
-QGSP\_BERT is a popular alternative which replaces the FTF model with
-the QGS model over the high energy range. The existence of more than one
-model (for each energy range) is very valuable in order to be able to
-determine the systematics effects related to the approximations used.
-The use of highly granular calorimeters, such as the ones being designed
-by the CALICE collaboration for future linear colliders, allows a
-detailed validation of the development of hadronic showers with
-test-beam data. Preliminary results suggest that the lateral profiles of
-Geant4 hadronic showers are too narrow. Comparisons with LHC test-beam
-data have shown that a fundamental ingredient for improving the
-description of the lateral development of showers is the use of
-intermediate and low energy models that can describe the cascading of
-hadrons in nuclear matter. Additional work is currently being invested
-in the further improvement of the QGS model, which is a more
-theory-based approach than the phenomenological FTF model, and therefore
-offers better confidence at high energies, up to a few TeV. This again
-is a large endeavour and requires continuous effort over a long time.
-
-The Geant4 collaboration is working closely with user communities to
-enrich the physics models' validation system with data acquired during
-physics runs and test beam campaigns. In producing new models of physics
-interactions and improving the fidelity of the models that exist, it is
-absolutely imperative that high-quality data are available. Simulation
-model tuning often relies on test beam data, and a program to improve
-the library of available data could be invaluable to the community. Such
-data would ideally include both thin-target test beams for improving
-interaction models and calorimeter targets for improving shower models.
-This data could potentially be used for directly tuning Fast Simulation
-models as well.
-
-There are specific challenges associated with the Intensity Frontier
-experimental programme, in particular simulation of the beamline and the
-neutrino flux. Neutrino experiments rely heavily on detector simulations
-to reconstruct neutrino energy, which requires accurate modelling of
-energy deposition by a variety of particles across a range of energies.
-Muon experiments such as Muon g-2 and Mu2e also face large simulation
-challenges; since they are searching for extremely rare effects, they
-must grapple with very low signal to background ratios and the modeling
-of low cross-section background processes. Additionally, the size of the
-computational problem is a serious challenge, as large simulation runs
-are required to adequately sample all relevant areas of experimental
-phase space, even when techniques to minimise the required computations
-are used. There is also a need to simulate the effects of low energy
-neutrons, which requires large computational resources. Geant4 is the
-primary simulation toolkit for all of these experiments.
-
-Simulation toolkits do not include effects like charge drift in an
-electric field or models of the readout electronics of the experiments.
-Instead, these effects are normally taken into account in a separate
-step called digitisation. Digitisation is inherently local to a given
-sub-detector and often even to a given readout element, so that there
-are many opportunities for parallelism in terms of vectorisation and
-multiprocessing or multithreading, if the code and the data objects are
-designed optimally. Recently, both hardware and software projects have
-benefitted from an increased level of sharing among experiments. The
-LArSoft Collaboration develops and supports a shared base of physics
-software across Liquid Argon (LAr) Time Projection Chamber (TPC)
-experiments, which includes providing common digitisation code.
-Similarly, an effort exists among the LHC experiments to share code for
-modeling radiation damage effects in silicon. As ATLAS and CMS expect
-to use similar readout chips in their future trackers, further code
-sharing might be possible.
-
-The Geant4 simulation toolkit will also evolve over the next decade to
-include contributions from various R\&D projects, as described in the
-following section. This is required to ensure the support of experiments
-through continuous maintenance and improvement of the Geant4 simulation
-toolkit. This is necessary until production versions of potentially
-alternative engines, such as those resulting from ongoing R\&D work,
-become available, integrated, and validated by experiments. The agreed
-ongoing strategy to make this adoption possible is to ensure that new
-developments resulting from the R\&D programme can be tested with
-realistic prototypes and then be integrated, validated, and deployed in
-a timely fashion in Geant4.
-
-\subsubsection*{Research and Development Programme}
-
-To meet the challenge of improving the performance by a large factor, an
-ambitious R\&D programme is underway to investigate each component of
-the simulation software for the long term. In the following we describe
-in detail some of the studies to be performed in the next 3-5 years.
-
-\begin{itemize}
-\item
-  Particle Transport and Vectorisation: the study of an efficient
-  transport of particles (tracks) in groups so as to maximise the
-  benefit of using SIMD operations.
-\item
-  Modularisation: improvement of Geant4 design to allow for a tighter
-  and easier integration of single sub-packages of the code into
-  experimental frameworks.
-\item
-  Physics Models: extensions and refinements of the physics algorithms
-  to provide new and more performant physics capabilities.
-\item
-  Other activities: integration of multi-threading capabilities in
-  experiment applications; experiment-agnostic software products to cope
-  with increased pile\-up, fast simulation, digitisation, and efficient
-  production of high-quality ran\-dom numbers.
-\end{itemize}
-
-\paragraph{Particle Transport and Vectorisation} One of the most
-ambitious elements of the simulation R\&D programme is a new approach to
-managing particle transport, which has been introduced by the GeantV
-project. The aim is to deliver a multithreaded vectorised transport
-engine that has the potential to deliver large performance benefits. Its
-main feature is track-level parallelisation, bundling particles with
-similar properties from different events to process them in a single
-thread. This approach, combined with SIMD vectorisation coding
-techniques and improved data locality, is expected to yield significant
-speed-ups, which are to be measured in a realistic prototype currently
-under development. For the GeantV transport engine to display its best
-computing performance, it is necessary to vectorise and optimise the
-accompanying modules, including geometry, navigation, and the physics
-models. These are developed as independent libraries so that they can
-also be used together with the current Geant4 transport engine. Of
-course, when used with the current Geant4 they will not expose their
-full performance potential, since transport in Geant4 is currently
-sequential, but this allows for a preliminary validation and comparison
-with the existing implementations. The benefit of this approach is that
-new developments can be delivered as soon as they are available. The new
-vectorised geometry package (VecGeom), developed as part of GeantV R\&D
-and successfully integrated into Geant4, is an example that demonstrated
-the benefit of this approach. By the end of 2018 it is intended to have
-a proof-of-concept for the new particle transport engine that includes
-vectorised EM physics, vectorised magnetic field propagation and that
-uses the new vectorised geometry package. This will form a sound basis
-for making performance comparisons for simulating EM showers in a
-realistic detector.
-
-\begin{itemize}
-\item
-  2019: the \emph{beta} release of the GeantV transport engine will
-  contain enough functionality to build the first real applications.
-  This will allow performance to be measured and give sufficient time to
-  prepare for HL-LHC running. It should include the use of vectorisation
-  in most of the components, including physics modelling for electrons,
-  gammas and positrons, whilst still maintaining simulation
-  reproducibility, and I/O in a concurrent environment and multi-event
-  user data management.
-\end{itemize}
-
-\paragraph{Modularisation} Starting from the next release, a
-modularisation of Geant4 is being pursued that will allow an easier
-integration in experimental frameworks, with the possibility to include
-only the Geant4 modules that are actually used. A further use case is
-the possibility to use one of the Geant4 components in isolation, e.g.,
-to use hadronic interaction modeling without kernel components from a
-fast simulation framework. As a first step a preliminary review of
-libraries' granularity is being pursued, which will be followed by a
-review of intra-library dependencies with the final goal of reducing
-their dependencies.
-
-\begin{itemize}
-\item
-  2019: Redesign of some Geant4 kernel components to improve the
-  efficiency of the simulation on HPC systems, starting from improved
-  handling of Geant4 \emph{database}s on large core-count systems. A
-  review will be made of the multithreading design to be closer to
-  task-based frameworks, such as Intel's Thread\-ed Building Blocks
-  (TBB)~\cite{TBB}.
-\end{itemize}
-
-\paragraph{Physics Models} It is intended to develop
-new and extended physics models to
-cover extended energy and physics processing of present and future
-colliders, Intensity Frontier experiments, and direct dark matter search
-experiments. The goal is to extend the missing models (e.g., neutrino
-interactions), improve models' physics accuracy and, at the same time,
-improve CPU and memory efficiency. The deliverables of these R\&D
-efforts include physics modules that produce equivalent quality physics,
-and will therefore require extensive validation in realistic
-applications.
-
-\begin{itemize}
-\item
-  2020: Improved implementation of hadronic cascade models for LHC and,
-  in particular, Liquid Argon detectors. Improved accuracy models of
-  EM interactions of photons and electrons. To address the needs of
-  cosmic frontier experiments, optical photon transport must be improved
-  and made faster.
-\item
-  2022: Implementation of EPOS string model for multi-GeV to multi-TeV
-  interactions, for FCC detector simulation and systematic studies of
-  HL-LHC detectors.
-\end{itemize}
-
-\paragraph{Experiment Applications} The experiment applications are
-essential for validating the software and physics performance of new
-versions of the simulation toolkit. ATLAS and CMS have already started
-to integrate Geant4 multithreading capability in their simulation
-applications; in the case of CMS the first Full Simulation production in
-multithreaded mode was delivered in the autumn of 2017. Specific
-milestones are as follows:
-
-\begin{itemize}
-\item
-  2020: LHC, Neutrino, Dark Matter, and Muon experiments to demonstrate
-  the ability to run their detector simulation in multithreaded mode,
-  using the improved navigation and electromagnetic physics packages.
-  This should bring experiments more accurate physics and improved
-  performance.
-\item
-  2020: Early integration of the beta release of the GeantV transport
-  engine in the experiments' simulation, including the implementation of
-  the new user interfaces, which will allow the first performance
-  measurements and physics validation to be made.
-\item
-  2022: The availability of a production version of the new track-level
-  parallelisation and fully vectorised geometry, navigation, and physics
-  libraries will offer the experiments the option to finalise
-  integration into their frameworks; intensive work will be needed in
-  physics validation and computing performance tests. If successful, the
-  new engine could be in production on the timescale of the start of the
-  HL-LHC run in 2026.
-\end{itemize}
-
-\paragraph{Pileup} Backgrounds to hard-scatter events have many
-components including in-time pileup, out-of-time pileup, cavern
-background and beam-gas collisions. All of these components can be
-simulated, but they present storage and I/O challenges related to the
-handling of the large simulated minimum bias samples used to model the
-extra interactions. An R\&D programme is needed to study different
-approaches to managing these backgrounds within the next 3 years:
-
-\begin{itemize}
-\item
-  Real zero-bias events can be collected, bypassing any zero
-  suppression, and overlaid on the fully simulated hard scatters. This
-  approach faces challenges related to the collection of
-  non-zero-suppressed samples or the use of suppressed events,
-  non-linear effects when adding electronic signals from different
-  samples, and sub-detector misalignment consistency between the
-  simulation and the real experiment. Collecting calibration and
-  alignment data at the start of a new Run would necessarily incur
-  delays such that this approach is mainly of use in the final analyses.
-  The experiments are expected to invest in the development of the
-  zero-bias overlay approach by 2020.
-\item
-  The baseline option is to ``pre-mix'' together the minimum bias
-  collisions into individual events that have the full background
-  expected for a single collision of interest. Experiments will invest
-  effort on improving their pre-mixing techniques, which allow the
-  mixing to be performed at the digitisation level, reducing the disk and
-  network usage for a single event.
-\end{itemize}
-
-\paragraph{Fast Simulation} The work on Fast Simulation is also
-accelerating with the objective of producing a flexible framework that
-permits Full and Fast simulation to be combined for different particles
-in the same event. Various approaches to Fast Simulation are being tried
-all with the same goal of saving computing time, under the assumption
-that it is possible to improve time performance without an unacceptable
-loss of physics accuracy. There has recently been a great deal of
-interest in the use of Machine Learning in Fast Simulation, most of
-which has focused on the use of multi-objective regression and
-generative adversarial networks (GANs). Since use of GANs allows for
-non-parametric learning in cases such as calorimetric shower
-fluctuations, it is a promising avenue for generating non-Gaussian and
-highly correlated physical effects. This is an obvious area for future
-expansion and development, as it is currently in its infancy.
-
-\begin{itemize}
-\item
-  2018: Assessment of the benefit of machine learning approach for Fast
-  Simulation.
-\item
-  2019: ML-based Fast Simulation for some physics observables.
-\item
-  2022: Demonstrate the potential of a common Fast Simulation
-  infrastructure applicable to the variety of detector configurations.
-\end{itemize}
-
-\paragraph{Digitisation} It is expected that, within the next 3
-years, common digitisation efforts are well-established among
-experiments, and advanced high-performance gener\-ic digitisation
-examples, which experiments could use as a basis to develop their own
-code, become available. For example, the development of next generation
-silicon detectors requires realistic simulation of the charge collection
-and digitisation processes. Owing to the large variety of technologies,
-common software frameworks need to be flexible and modular to cater for
-the different needs.
-
-\begin{itemize}
-\item
-  2020: Deliver advanced high-performance, SIMD-friendly generic
-  digitisation examples that experiments can use as a basis to develop
-  their own code.
-\item
-  2022: Fully tested and validated optimised digitisation code that can
-  be used by the HL-LHC and DUNE experiments.
-\end{itemize}
-
-\paragraph{Pseudorandom Number Generation} The selection of
-pseudorandom number generators (PRNGs) presents challenges when running
-on infrastructures with a large degree of parallelism, as
-reproducibility is a key requirement. HEP will collaborate with
-researchers in the development of PRNGs, seeking to obtain generators
-that address better our challenging requirements. Specific milestones
-are:
-
-\begin{itemize}
-\item
-  2020: Develop a single library containing sequential and vectorised
-  implementations of the set of state-of-the-art PRNGs, to replace the
-  existing ROOT and CLHEP implementations. Potential use of C++11 PRNG
-  interfaces and implementations, and their extension for our further
-  requirements (output of multiple values, vectorisation) will be
-  investigated.
-\item
-  2022: Promote a transition to the use of this library to replace
-  existing implementations in ROOT and Geant4.
-\end{itemize}
-
-\hypertarget{software-trigger-and-event-reconstruction}{%
-\subsection{Software Trigger and Event
-Reconstruction}\label{software-trigger-and-event-reconstruction}}
-
-\subsubsection*{Scope and Challenges}
-
-The reconstruction of raw detector data and simulated data, and its
-processing in real time, represent a major component of today's
-computing requirements in HEP. Advances in the capabilities of 
-facilities and future
-experiments bring the potential for a dramatic increase in physics
-reach, at the price of increased event complexities and rates. It is
-therefore essential that event reconstruction algorithms and software
-triggers continue to evolve so that they are able to efficiently exploit
-future computing architectures, and deal with the increase in data rates
-without loss of physics. Projections into future, e.g., at HL-LHC conditions,
-show that without significant changes in approach or algorithms
-the increase in resources needed would be incompatible with the 
-the expected budget.
-
-At the HL-LHC, the central challenge for object reconstruction is to
-maintain excellent efficiency and resolution in the face of high pileup
-values, especially at low transverse momentum ($p_T$).
-Detector upgrades, such as increases in channel density, high-precision
-timing, and improved detector geometric layouts, are essential to
-overcome these problems. In many cases these new technologies bring
-novel requirements to software trigger and/or event reconstruction
-algorithms, or require new algorithms to be developed. Ones of
-particular importance at the HL-LHC include high-granularity
-calorimetry, precision timing detectors, and hardware triggers based on
-tracking information, which may seed later software trigger and
-reconstruction algorithms.
-
-At the same time, trigger systems for next-generation experiments are
-evolving to be more capable, both in their ability to select a wider
-range of events of interest for the physics programme, and their ability
-to stream a larger rate of events for further processing. ATLAS and CMS
-both target systems where the output of the hardware trigger system is
-increased by an order of magnitude over the current capability, up to 1
-MHz~\cite{CERN-LHCC-2015-020,CMSCollaboration:2015zni}. In LHCb~\cite{CERN-LHCC-2014-016}
-and ALICE~\cite{Buncic:2011297}, the full collision
-rate (between 30 to 40 MHz for typical LHC proton-proton operations)
-will be streamed to real-time or quasi-real-time software trigger
-systems. The increase in event complexity also brings a ``problem'' of
-an overabundance of signals to the experiments, and specifically to the
-software trigger algorithms. The evolution towards a genuine real-time
-analysis of data has been driven by the need to analyse more signal than
-can be written out for traditional processing, and technological
-developments that enable this without reducing the analysis sensitivity
-or introducing biases.
-
-Evolutions in computing technologies are an opportunity to move
-beyond commodity x86\_64 technologies, which HEP has used very
-effectively over the past 20 years, but also represent a significant challenge 
-if we are to
-derive sufficient event processing throughput per cost to reasonably
-enable our physics programmes~\cite{Bird:1695401}.
-Among these challenges, important items identified include the increase
-of SIMD capabilities, the evolution towards multi- or many-core
-architectures, the slow increase in memory bandwidth relative to CPU
-capabilities, the rise of heterogeneous hardware, and the possible
-evolution in facilities available to HEP production systems.
-
-The move towards open source software development and continuous
-integration systems brings opportunities to assist developers of
-software trigger and event reconstruction algorithms. Continuous
-integration systems based on standard open-source tools have already
-allowed automated code quality and performance checks, both for
-algorithm developers and code integration teams. Scaling these up to
-allow for sufficiently high-statistics checks is still an
-outstanding challenge. Also, code quality demands increase as
-traditional offline analysis components migrate into trigger systems,
-where algorithms can only be run once, and any problem means losing data
-permanently.
-
-\subsubsection*{Current Practices}
-
-Substantial computing facilities are in use for both online and offline
-event processing across all experiments surveyed. In most experiments,
-online facilities are dedicated to the operation of the software
-trigger, but a recent trend has been to use them opportunistically
-for offline processing too, when the software trigger does not make them
-100\% busy. On the other hand, offline facilities are shared with
-event reconstruction, simulation, and analysis. CPU in use by
-experiments is typically measured at the scale of tens or hundreds of
-thousands of x86\_64 processing cores.
-
-The CPU needed for event reconstruction tends to be dominated by charged
-particle reconstruction (tracking), especially when the number of
-collisions per bunch crossing is high and an efficient
-reconstruction low p\textsubscript{T} particles is required.
-Calorimetric reconstruction, particle flow reconstruction, and particle
-identification algorithms also make up significant parts of the CPU
-budget in some experiments. Disk storage is typically 10s to 100s of PBs
-per experiment. It is dominantly used to make the output of the event
-reconstruction, both for real data and simulation, available for
-analysis.
-
-Current experiments have moved towards smaller, but still
-flexible, tiered data formats. These tiers are typically based on the
-ROOT~\cite{Brun1996} file format and constructed to
-facilitate both skimming of interesting events and the selection of
-interesting pieces of events by individual analysis groups or through
-centralised analysis processing systems. Initial implementations of
-real-time analysis systems are in use within several experiments. These
-approaches remove the detector data that typically makes up the raw data
-tier kept for offline reconstruction, and keep only final analysis
-objects~\cite{Aaij2016,ATL-DAQ-PUB-2017-003,Khachatryan:2149625}.
-
-Systems critical for reconstruction, calibration, and alignment
-generally implement a high level of automation in all experiments.
-They are an integral part of the data taking and data reconstruction processing chain, both in the online systems as well as the offline processing setup.
-
-\subsubsection*{Research and Development Programme}
-
-Seven key areas, itemised below, have been identified where research and
-development is necessary to enable the community to exploit the full
-power of the enormous datasets that we will be collecting. Three of
-these areas concern the increasingly parallel and heterogeneous
-computing architectures that we will have to write our code for. In
-addition to a general effort to vectorise our codebases, we must
-understand what kinds of algorithms are best suited to what kinds of
-hardware architectures, develop benchmarks that allow us to compare the
-physics-per-dollar-per-watt performance of different algorithms across a
-range of potential architectures, and find ways to optimally utilise
-heterogeneous processing centres. The consequent increase in the
-complexity and diversity of our codebase will necessitate both a
-determined push to educate physicists in modern coding
-practices, and a development of more sophisticated and automated quality
-assurance and control. The increasing granularity of
-our detectors, and the addition of timing information, which seems
-mandatory to cope with the extreme pileup conditions at the HL-LHC, will
-require new kinds of reconstruction algorithms that are sufficiently
-fast for use in real-time. Finally, the increased
-signal rates will mandate a push towards real-time analysis in many
-areas of HEP, in particular those with low-p\textbf{\textsubscript{T}}
-signatures.
-
-\begin{itemize}
-\item
-  HEP developed toolkits and algorithms typically make poor use of
-  vector units on commodity computing systems. Improving this will bring
-  speedups to applications running on both current computing systems and
-  most future architectures. The goal for work in this area is to evolve
-  current toolkit and algorithm implementations, and best programming
-  techniques, to better use SIMD capabilities of current and future
-  CPU architectures.
-\item
-  Computing platforms are generally evolving towards having more cores
-  in order to increase processing capability. This evolution has
-  resulted in multithreaded frameworks in use, or in development, across
-  HEP. Algorithm developers can improve throughput by being thread-safe
-  and enabling the use of fine-grained parallelism. The goal is to
-  evolve current event models, toolkits and algorithm implementations,
-  and best programming techniques, to improve the throughput of
-  multithreaded software trigger and event reconstruction applications.
-\item
-  Computing architectures using technologies beyond CPUs offer an
-  interesting alternative for increasing throughput of the most
-  time-consuming trigger or reconstruction algorithms. Examples
-  such as GPUs and FPGAs could be integrated into dedicated
-  trigger or specialised reconstruction processing facilities,
-  in particular
-  online computing farms. The goal is to demonstrate how the throughput
-  of toolkits or algorithms can be improved
-  in a production environment and to understand
-  how much these new architectures require rethinking the algorithms
-  used today. In addition, it is necessary to assess and minimise
-  possible additional costs coming from the maintenance of multiple
-  implementations of the same algorithm on different architectures.
-\item
-  HEP experiments have extensive continuous integration systems,
-  including varying code regression checks that have enhanced the
-  quality assurance (QA) and quality control (QC) procedures for
-  software development in recent years. These are typically maintained
-  by individual experiments and have not yet reached the point where
-  statistical regression, technical, and physics performance checks can
-  be performed for each proposed software change. The goal is to enable
-  the development, automation, and deployment of extended QA and QC
-  tools and facilities for software trigger and event reconstruction
-  algorithms.
-\item
-  Real-time analysis techniques are being adopted to enable a wider
-  range of physics signals to be saved by the trigger for final
-  analysis. As rates increase, these techniques can become more
-  important and widespread by enabling only the parts of an event
-  associated with the signal candidates to be saved, reducing the
-  disk space requirement. The goal is to evaluate and demonstrate the tools
-  needed to facilitate real-time analysis techniques. Research topics
-  include the study of compression and custom data formats, toolkits for real-time
-  detector calibration and validation that enable full offline analysis
-  chains to be ported into real-time, and frameworks that allow
-  non-expert offline analysts to design and deploy real-time analyses
-  without compromising data taking quality.
-\item
-  The central challenge for object reconstruction at the HL-LHC is to
-  maintain excellent efficiency and resolution in the face of high
-  pileup, especially at low object p\textsubscript{T}.
-  Trigger systems and reconstruction software need to exploit new techniques
-  and higher granularity detectors to maintain or even improve physics
-  measurements in the future. It is also becoming increasingly clear
-  that reconstruction in very high pileup environments, such as the
-  HL-LHC or FCC-hh, will not be possible without adding some timing
-  information to our detectors, in order to exploit the finite time
-  during which the beams cross and the interactions are produced. The
-  goal is to develop and demonstrate efficient techniques for physics
-  object reconstruction and identification in complex environments.
-\item
-  Future experimental facilities will bring a large increase in event
-  complexity. The performance scaling of current-generation algorithms with this
-  complexity must be improved to avoid a large increase in resource
-  needs. In addition, it may become necessary to deploy
-  new algorithms in order to solve these problems, including advanced
-  machine learning techniques. The goal is to evolve or rewrite existing
-  toolkits and algorithms focused on their physics and technical
-  performance at high event complexity, e.g., high pileup at HL-LHC.
-  Most important targets are those which limit expected throughput
-  performance at future facilities, e.g., charged-particle tracking. A
-  number of such efforts are already in progress.
-\end{itemize}
-
-\hypertarget{data-analysis-and-interpretation}{%
-\subsection{Data Analysis and
-Interpretation}\label{data-analysis-and-interpretation}}
-
-\subsubsection*{Scope and Challenges}
-
-Scientific questions are answered by analysing the data obtained from
-suitably designed experiments and comparing measurements with
-predictions from models and theories. Such comparisons are typically
-performed long after data taking, but can sometimes also be executed in
-quasi-real time on selected samples of reduced size.
-
-The final stages of analysis are undertaken by small groups or even
-individual researchers. The baseline analysis model utilises successive
-stages of data reduction, finally reaching a compact dataset for quick
-real-time iterations. This approach aims at exploiting the maximum
-possible scientific potential of the data, whilst minimising the ``time
-to insight'' for a large number of different analyses performed in
-parallel. It is a complicated combination of diverse criteria, ranging
-from the need to make efficient use of computing resources to the
-management styles of the experiment collaborations. Any analysis system
-has to be flexible enough to cope with deadlines imposed by conference
-schedules. Future analysis models must adapt to the massive increases in
-data taken by the experiments, while retaining this essential ``time to
-insight'' optimisation.
-
-Over the past 20 years the HEP community has developed and gravitated
-around a single analysis ecosystem based on ROOT~\cite{Brun1996}. 
-ROOT is a general-purpose object
-oriented framework that addresses the selection, integration,
-development, and support of a number of foundation and utility class
-libraries that can be used as a basis for developing HEP application
-codes. The added value to the HEP community is that it provides an
-integrated and validated toolkit, and its use encompasses the full event
-processing chain; it has a major impact on the way HEP analysis is
-performed. This lowers the hurdle to start an analysis, enabling the
-community to communicate using a common analysis language, as well as
-making common improvements as additions to the toolkit quickly become
-available. The ongoing ROOT programme of work addresses important new
-requirements, in both functionality and performance, and this is given
-a high priority by the HEP community.
-
-An important new development in the analysis domain has been the
-emergence of new analysis tools coming from industry and open source
-projects, and this presents new opportunities for improving the HEP
-analysis software ecosystem. The HEP community is very interested in
-using these software tools, together with established components, in an
-interchangeable way. The main challenge will be to enable new
-open-source tools to be plugged in dynamically to the existing ecosystem
-and to provide mechanisms that allow the existing and new components to
-interact and exchange data efficiently. To improve our ability to
-analyse much larger datasets, R\&D will be needed to investigate file
-formats, compression algorithms, and new ways of storing and accessing
-data for analysis and to adapt workflows to run on future computing
-infrastructures.
-
-Reproducibility is the cornerstone of scientific results. It is
-currently difficult to repeat most HEP analyses in exactly the manner
-they were originally performed. This difficulty mainly arises due to the
-number of scientists involved, the large number of steps in a typical
-HEP analysis workflow, and the complexity of the analyses themselves. A
-challenge specific to data analysis and interpretation is tracking the
-evolution of relationships between all the different components of an
-analysis.
-
-Robust methods for data reinterpretation are also critical.
-Collaborations typically interpret results in the context of specific
-models for new physics searches and sometimes reinterpret those same
-searches in the context of alternative theories. However, understanding
-the full implications of these searches requires the interpretation of
-the experimental results in the context of many more theoretical models
-than are currently explored at the time of publication. Analysis
-reproducibility and reinterpretation strategies need to be considered in
-all new approaches under investigation, so that they become a
-fundamental component of the system as a whole.
-
-Adapting to the rapidly evolving landscape of software tools, as well as
-to methodological approaches to data analysis, requires effort in
-continuous training, both for novices as well as for experienced
-researchers, as detailed in the Section 
-\ref{training-and-careers}. The
-maintenance and sustainability of the current analysis ecosystem also
-present a major challenge, as currently this effort is provided by just
-a few institutions. Legacy and less-used parts of the ecosystem need to
-be managed appropriately. New policies are needed to retire little used
-or obsolete components and free up effort for the development of new
-components. These new tools should be made attractive and useful to a
-significant part of the community to attract new contributors.
-
-\subsubsection*{Current Practices}
-
-Methods for analysing HEP data have been developed over many years and
-successfully applied to produce physics results, including more than
-2000 publications, during LHC Runs 1 and 2. Analysis at the LHC
-experiments typically starts with users running code over centrally
-managed data that is of O(100kB/event) and contains all of the
-information required to perform a typical analysis leading to
-publication. The most common approach is through a campaign of
-\emph{data reduction} and \emph{refinement}, ultimately producing
-simplified data structures of arrays of simple data types (``flat
-ntuples'') and histograms used to make plots and tables, from which
-physics results can be derived.
-
-The current centrally-managed data typically used by a Run 2 data
-analysis at the LHC (hundreds of TB) is far too large to be delivered
-locally to the user. An often-stated requirement of the data reduction
-steps is to arrive at a dataset that ``can fit on a laptop'', in order
-to facilitate low-latency, high-rate access to a manageable amount of
-data during the final stages of an analysis. Creating and retaining
-intermediate datasets produced by data reduction campaigns, bringing and
-keeping them ``close'' to the analysers, is designed to minimise latency
-and the risks related to resource contention. At the same time, disk
-space requirements are usually a key constraint of the experiment
-computing models. The
-LHC experiments have made a continuous effort to produce optimised
-analysis-oriented data formats with enough information to avoid the need
-to use intermediate formats. Another effective strategy has been to
-combine analyses from different users and execute them within the same
-batch jobs (so-called ``analysis trains''), thereby reducing the number
-of times data must be read from the storage systems. This has improved
-performance and usability, and simplified the task of the bookkeeping.
-
-There has been a huge investment in using C++ for performance-critical
-code, in particular in event reconstruction and simulation, and this
-will continue in the future. However, for analysis applications, Python
-has emerged as the language of choice in the data science community, and
-its use continues to grow within HEP. Python is highly
-appreciated for its ability to support fast development cycles, for its
-ease-of-use, and it offers an abundance of well-maintained and advanced
-open source software packages. Experience shows that the simpler
-interfaces and code constructs of Python could reduce the complexity of
-analysis code, and therefore contribute to decreasing the ``time to
-insight'' for HEP analyses, as well as increasing their sustainability.
-Increased HEP investment is needed to allow Python to become a first
-class supported language.
-
-One new model of data analysis, developed outside of HEP, maintains the
-concept of sequential reduction, but mixes interactivity with
-batch processing. These exploit new cluster management systems, most
-notably Apache Spark, which uses open-source tools contributed both by
-industry and the data-science community. Other products implementing the
-same analysis concepts and workflows are emerging, such as TensorFlow,
-Dask, Pachyderm, Blaze, Parsl, and Thrill. This approach can complement
-the present and widely adopted Grid processing of datasets. It may
-potentially simplify the access to data and the expression of
-parallelism, thereby improving the exploitation of cluster resources.
-
-An alternative approach, which was pioneered in astronomy but has become
-more widespread throughout the Big Data world, is to perform \emph{fast
-querying} of centrally managed data and compute remotely on the queried
-data to produce the analysis products of interest. The analysis workflow
-is accomplished without focus on persistence of data traditionally
-associated with data reduction, although transient data may be generated
-in order to efficiently accomplish this workflow and optionally can be
-retained to facilitate an analysis ``checkpoint'' for subsequent
-execution. In this approach, the focus is on obtaining the analysis
-end-products in a way that does not necessitate a data reduction
-campaign. It is of interest to understand the role
-that such an approach could have in the global analysis infrastructure,
-and if it can bring an optimisation of the global storage and computing
-resources required for the processing of raw data to analysis.
-
-Another active area regarding analysis in the world outside HEP is the
-switch to a functional or declarative programming model, as for example
-provided by Scala in the Spark environment. This allows scientists to
-express the intended data transformation as a query on data. Instead of
-having to define and control the ``how'', the analyst declares the
-``what'' of their analysis, essentially removing the need to define the
-event loop in an analysis, and leave it to underlying services and
-systems to optimally iterate over events. It appears that these
-high-level approaches will allow abstraction from the underlying
-implementations, allowing the computing systems more freedom in
-optimising the utilisation of diverse forms of computing resources. R\&D
-is already under way, e.g., TDataFrame~\cite{TDataFrame} in ROOT, and
-this needs to be continued with the ultimate goal of establishing a
-prototype functional or declarative programming paradigm.
-
-\subsubsection*{Research and Development Programme}
-
-Towards HL-LHC, we envisage dedicated data analysis facilities for
-experimenters, offering an extendable environment that can provide fully
-functional analysis capabilities, integrating all these technologies
-relevant for HEP. Initial prototypes of such analysis facilities are
-currently under development. On the time scale of HL-LHC, such dedicated
-analysis facilities would provide a complete system engineered for
-latency optimisation and stability.
-
-The following R\&D programme lists the tasks that need to be
-accomplished. By 2020:
-
-\begin{itemize}
-\item
-  Enable new open-source software tools to be plugged in dynamically to
-  the existing ecosystem, and provide mechanisms to dynamically exchange
-  parts of the ecosystem with new components.
-\item
-  Complete an advanced prototype of a low-latency response,
-  high-capacity analysis facility, incorporating fast caching
-  technologies to explore a query-based analysis approach and
-  open-source cluster-management tools. It should in particular include
-  an evaluation of additional storage layers, such as SSD storage and
-  NVRAM-like storage, and cloud and Big Data orchestration systems.
-\item
-  Expand support of Python in our ecosystem with a strategy for ensuring
-  long-term maintenance and sustainability. In particular in ROOT, the
-  current Python bindings should evolve to reach the ease of use of native Python modules.
-\item
-  Prototype a comprehensive set of mechanisms for interacting and
-  exchanging data between new open-source tools and the existing
-  analysis ecosystem.
-\item
-  Develop a prototype based on a functional or declarative programming
-  model for data analysis.
-\item
-  Conceptualise and prototype an analysis ``Interpretation Gateway'',
-  including data repositories, e.g., HEPData~\cite{HEPData,HEPDataRepo}, and
-  analysis preservation and reinterpretation tools.
-\end{itemize}
-
-By 2022:
-
-\begin{itemize}
-\item
-  Evaluate chosen architectures for analysis facilities, verify their
-  design and provide input for corrective actions to test them on a
-  larger scale during Run 3.
-\item
-  Develop a blueprint for remaining analysis facility developments,
-  system design and support model.
-\end{itemize}
-
-\hypertarget{machine-learning}{%
-\subsection{Machine Learning}\label{machine-learning}}
-
-Machine Learning (ML) is a rapidly evolving approach to characterising
-and describing data with the potential to radically change how data is
-reduced and analysed. Some applications will qualitatively improve the
-physics reach of datasets. Others will allow much more efficient use of
-processing and storage resources, effectively extending the physics
-reach of experiments. Many of the activities in this area will
-explicitly overlap with those in the other focus areas, whereas others
-will be more generic. As a first approximation, the HEP community will
-build domain-specific applications on top of existing toolkits and ML
-algorithms developed by computer scientists, data scientists, and
-scientific software developers from outside the HEP world. Work will
-also be done to understand where problems do not map well onto existing
-paradigms and how these problems can be recast into abstract
-formulations of more general interest.
-
-\subsubsection*{Scope and Challenges}
-
-The Machine Learning, Statistics, and Data Science communities have
-developed a variety of powerful ML approaches for classification (using
-pre-defined categories), clustering (where categories are discovered),
-regression (to produce continuous outputs), density estimation,
-dimensionality reduction, etc. Some of these have been used productively
-in HEP for more than 20 years, others have been introduced relatively
-recently. The portfolio of ML techniques and tools is in constant
-evolution, and a benefit is that many have well-documented open source
-software implementations. ML has already become ubiquitous in some
-HEP applications, most notably in classifiers used to discriminate
-between signals and backgrounds in final offline analyses. It is
-also increasingly used in both online and offline reconstruction and
-particle identification algorithms, as well as the classification of
-reconstruction-level objects, such as jets.
-
-The abundance of, and advancements in, ML algorithms and implementations
-present both opportunities and challenges for HEP. 
-The community
-needs to understand which are most appropriate for our use,
-tradeoffs for using one tool compared to another, and
-the tradeoffs of using ML algorithms compared to using more traditional software.
-These issues are not necessarily ``factorisable'', and a key goal will
-be to ensure that, as HEP research teams investigate the numerous
-approaches at hand, the expertise acquired and lessons learned, get
-adequately disseminated to the wider community. In general, each
-\emph{team}, typically a small group of scientists from a collaboration,
-will serve as a source of expertise, helping others develop and
-deploy experiment-specific ML-based algorithms in their software stacks.
-It should provide training to those developing new ML-based algorithms,
-as well as those planning to use established ML tools.
-
-With the advent of more powerful hardware and more performant ML
-algorithms, the ML toolset will be used to develop application software
-that could potentially, amongst other things:
-
-\begin{itemize}
-\item
-  Replace the most computationally expensive parts of pattern
-  recognition algorithms and parameter extraction algorithms for
-  characterising reconstructed objects. For example, investigating how
-  ML algorithms could improve the physics performance or execution speed
-  of charged track and vertex reconstruction, one of the most CPU
-  intensive elements of our current software.
-\item
-  Extend the use of ML algorithms for real-time event classification and
-  analysis, as discussed in more detail in Section \ref{software-trigger-and-event-reconstruction}.
-\item
-  Extend the physics reach of experiments by extending the role of ML at
-  the analysis stage: handling data/MC or control/signal region
-  differences, interpolating between mass points, training in a
-  systematics-aware way, etc.
-\item
-  Compress data significantly with negligible loss of fidelity in terms
-  of physics utility.
-\end{itemize}
-
-As already discussed, many particle physics detectors produce much more
-data than can be moved to permanent storage. The process of reducing the
-size of the datasets is managed by the trigger system. ML algorithms
-have already been used very successfully for triggering, to rapidly
-characterise which events should be selected for additional
-consideration and eventually saved to long-term storage. In the era
-of the HL-LHC, the challenges will increase both quantitatively and
-qualitatively as the number of proton-proton collisions per bunch
-crossing increases. The scope of ML applications in the trigger will
-need to expand in order to tackle the challenges to come.
-
-\subsubsection*{Current Practices}
-
-The use of ML in HEP analyses has become commonplace over the past two
-decades, and the most common use case has been in signal/background
-classification. The vast majority of HEP analyses published in recent
-years have used the HEP-specific software package TMVA~\cite{TMVA} included in ROOT. Recently, however, many HEP analysts have begun
-migrating to non-HEP ML packages such as scikit-learn~\cite{scikit-learn}
-and Keras~\cite{Keras}, although these efforts have yet to result
-in physics publications from major collaborations. Data scientists at
-Yandex created a Python package that provides a consistent API to most
-ML packages used in HEP~\cite{REP}. Packages like Spearmint~\cite{Spearmint} and scikit-optimize~\cite{scikit-optimize} perform Bayesian optimisation and can improve HEP Monte
-Carlo work.
-
-This shift in the set of ML techniques and packages utilised is
-especially strong in the neutrino physics community, where new
-experiments such as DUNE place ML at the very heart of their
-reconstruction algorithms and event selection. The shift is also
-occurring among LHC collaborations, where ML is becoming more and more
-commonplace in reconstruction and real-time applications. Examples where
-ML has already been deployed in a limited way include charged and
-neutral particle reconstruction and identification, jet reconstruction
-and identification, and determining a particle's production properties
-(flavour tagging), based on information from the rest of the event. In
-addition, ML algorithms have been developed that are
-insensitive to changing detector performance, for use in real-time
-applications, and algorithms that are minimally biased with respect to
-the physical observables of interest.
-
-At present, much of this development has happened in specific
-collaborations. While each experiment has, or is likely to have,
-different specific use cases, we expect that many of these will be
-sufficiently similar to each other that R\&D can be done in common. Even
-when this is not possible, experience with one type of problem will
-provide insights into how to approach other types of problem. This is
-why the Inter-experiment Machine Learning forum (IML~\cite{IML}) was
-created at CERN in 2016, as a compliment to experiment specific ML R\&D groups.
-It has already fostered closer collaboration between LHC and non-LHC
-collaborations in the ML field.
-
-\subsubsection*{Research and Development Roadmap and Goals}
-
-The R\&D roadmap presented here is based on the preliminary work done in
-recent years, coordinated by the HSF IML, which will remain the main
-forum to coordinate work in ML in HEP and ensure the proper links
-with the data science communities. The following programme of work is
-foreseen.
-
-By 2020:
-\begin{itemize}
-\item
-  \emph{Particle identification and particle properties:} in calorimeters or
-  time projection chambers (TPCs), where the data can be represented as
-  a 2D or 3D image (or even in 4D, including timing information), the
-  problems can be cast as a computer vision task. Deep Learning (DL),
-  one class of ML algorithm, in which neural networks are used to
-  reconstruct images from pixel intensities, is a good candidate to
-  identify particles and extract many parameters. Promising DL
-  architectures for these tasks include convolutional, recurrent, and
-  adversarial neural networks. A particularly important application is
-  to Liquid Argon TPCs (LArTPCs), which is the chosen detection
-  technology for DUNE, the new flagship experiment in the neutrino programme. 
-  A proof of
-  concept and comparison of DL architectures should be finalised by
-  2020. Particle identification can also be explored to tag the flavour
-  of jets in collider experiments (e.g., so-called b-tagging). The
-  investigation of these concepts, which connect to Natural Language
-  Processing, has started at the LHC and is to be pursued on the same
-  timescale.
-\item
-  \emph{ML middleware and data formats for offline usage:} HEP relies 
-  on the ROOT format for its data, wheras the ML community
-  has developed several other formats, often associated with specific ML
-  tools. A desirable data format for ML applications should have the
-  following attributes: high read-write speed for efficient training,
-  sparse readability without loading the entire dataset into RAM,
-  compressibility, and widespread adoption by the ML community. The thorough
-  evaluation of the different data formats and their impact on ML
-  performance in the HEP context must be continued, and it is necessary
-  to define a strategy for bridging or migrating HEP formats to the
-  chosen ML format(s), or vice-versa.
-\item
-  \emph{Computing resource optimisations:} managing large
-  volume data transfers is one
-  of the challenges facing current computing facilities. Networks 
-  play a crucial role in data exchange and so a
-  network-aware application layer may significantly improve experiment
-  operations. ML is a promising technology to identify anomalies in
-  network traffic, to predict and prevent network congestion, to detect
-  bugs via analysis of self-learning networks, and for WAN path
-  optimisation based on user access patterns.
-\item
-  \emph{ML as a Service (MLaaS):} current cloud providers rely on a MLaaS model
-  exploiting interactive
-  machine learning tools in order to make efficient use of resources, however,
-  this is not yet widely used in HEP. HEP
-  services for interactive analysis, such as CERN's Service for
-  Web-based Analysis, SWAN~\cite{SWAN}, may play an important role in adoption
-  of machine learning tools in HEP workflows. In order to use these tools
-  more efficiently, sufficient and appropriately tailored hardware and
-  instances other than SWAN will be identified.
-\end{itemize}
-
-By 2022:
-\begin{itemize}
-\item
-  \emph{Detector anomaly detection:} data taking is
-  continuously monitored by physicists taking shifts to monitor and
-  assess the quality of the incoming data, largely using reference
-  histograms produced by experts. A whole class of ML algorithms called anomaly detection
-  can be useful for automating this important task. Such unsupervised algorithms are able
-  to learn from data and produce an alert when deviations are observed.
-  By monitoring many variables at the same time, such algorithms are
-  sensitive to subtle signs forewarning of imminent failure, so that
-  pre-emptive maintenance can be scheduled. These techniques are already
-  used in industry.
-\item
-  \emph{Simulation:} recent progress in high fidelity fast generative models,
-  such as Generative Adversarial Networks (GANs) and Variational
-  Autoencoders (VAEs), which are able to sample high dimensional feature
-  distributions by learning from existing data samples, offer a
-  promising alternative for Fast Simulation. A simplified first attempt at
-  using such techniques in simulation saw orders of magnitude increase
-  in speed over existing Fast Simulation techniques, but has not yet
-  reached the required accuracy~\cite{CaloGAN}.
-\item
-  \emph{Triggering and real-time analysis:} one of the challenges is the
-  trade-off in algorithm complexity and performance under strict
-  inference time constraints. To deal with the increasing event
-  complexity at HL-LHC, the use of sophisticated ML
-  algorithms will be explored at all trigger levels, building on the pioneering work of
-  the LHC collaborations. A critical part of this work will be to
-  understand which ML techniques allow us to maximally exploit
-  future computing architectures.
-\item
-  \emph{Sustainable Matrix Element Method (MEM):} MEM is a powerful
-  technique that can be utilised for making measurements of physical model
-  parameters and direct searches for new phenomena. As it is
-  very computationally intensive its use in HEP is limited.
-  Although the use of neural networks for numerical integration is not new, 
-  it is a
-  technical challenge to design a network sufficiently rich
-  to encode the complexity of the ME calculation for a given process
-  over the phase space relevant to the signal process. Deep Neural
-  Networks (DNNs) are good candidates~\cite{Bendavid2017,BendavidIML2017}.
-\item
-  \emph{Tracking:} pattern recognition is always a computationally challenging
-  step. It becomes a huge challenge in the HL-LHC environment.
-  Adequate ML techniques may provide a solution that scales linearly
-  with LHC intensity. Several efforts in the HEP community have started
-  to investigate ML algorithms for track pattern recognition on
-  many-core processors.
-\end{itemize}
-
-\hypertarget{data-organisation-management-and-access}{%
-\subsection{Data Organisation, Management and
-Access}\label{data-organisation-management-and-access}}
-
-The scientific reach of data-intensive experiments is limited by how
-fast data can be accessed and digested by computational resources. Changes in
-computing technology and large increases in data volume require new
-computational models~\cite{Butler2013}, compatible with
-budget constraints. The integration of newly emerging data
-analysis paradigms into our computational model has the potential to
-enable new analysis methods and increase scientific output. The field,
-as a whole, has a window in which to adapt our data access and data
-management schemes to ones that are more suited and optimally matched to
-advanced computing models and a wide range of analysis applications.
-
-\subsubsection*{Scope and Challenges}
-
-The LHC experiments currently provision and manage about an exabyte of
-storage, approximately half of which is archival, and half is
-traditional disk storage. Other experiments that will soon start data taking have
-similar needs, e.g., Belle II has the same data volumes as ATLAS. The
-HL-LHC storage requirements per year are expected to jump by a factor
-close to 10, which is a growth rate faster than can be 
-accommodated by projected technology gains. Storage will remain one of the major
-cost drivers for HEP computing, at a level roughly equal to the cost
-of the computational resources. The combination of storage and analysis
-computing costs may restrict scientific output and the potential physics
-reach of the experiments, so new techniques and algorithms are likely to
-be required.
-
-In devising experiment computing models for this era many factors have
-to be taken into account. In particular, the increasing availability of
-very high-speed networks may reduce the need for CPU and data
-co-location. Such networks may allow for more
-extensive use of data access over the wide-area network (WAN), which may
-provide failover capabilities, global and federated data namespaces, and
-will have an impact on data caching. Shifts in data presentation and
-analysis models, such as the use of event-based data streaming along
-with more traditional dataset-based or file-based data access, will be
-particularly important for optimising the utilisation of opportunistic
-computing cycles on HPC facilities, commercial cloud resources, and
-campus clusters. This can potentially resolve currently limiting factors
-such as job eviction.
-
-The three main challenges for data management in the HL-LHC follow:
-
-\begin{itemize}
-\item
-  The experiments will significantly increase both
-  the data rate and the data volume. The computing systems will need to
-  handle this with as small a cost increase as possible and within
-  evolving storage technology limitations.
-\item
-  The significantly increased computational requirements for the HL-LHC
-  era will also place new requirements on data access. Specifically, the use of
-  new types of computing resources (cloud, HPC) that have different dynamic
-  availability and characteristics will require more dynamic data
-  management and access systems.
-\item
-  Applications employing new techniques, such as 
-  training for machine learning or high rate data query systems, will likely be employed to
-  meet the computational constraints and to extend physics reach.
-  These new applications will place new requirements
-  on how and where data is accessed and produced. Specific applications,
-  such as training for machine learning, may require use of specialised
-  processor resources, such as GPUs, placing further requirements on
-  data.
-\end{itemize}
-
-The projected event complexity of data from future HL-LHC
-runs with high pileup and from high resolution Liquid Argon detectors at
-DUNE will require advanced reconstruction algorithms and analysis tools
-to interpret the data. The precursors of these tools, in the form of
-new pattern recognition and tracking algorithms, 
-are already proving to be drivers for the
-compute needs of the HEP community. The storage systems that are
-developed, and the data management techniques that are employed,
-will need to be matched to these changes in computational work, so
-as not to hamper potential improvements.
-
-As with computing resources, the landscape of storage solutions
-is trending towards heterogeneity. The ability to
-leverage new storage technologies as they become available into existing
-data delivery models is a challenge that we must be prepared for. This
-also implies the need to leverage
-``tactical storage'', i.e., storage that becomes more cost-effective as
-it becomes available (e.g., from a cloud provider), and have a data
-management and provisioning system that can exploit such resources at
-short notice. Volatile data sources would impact many aspects of the
-system: catalogues, job brokering, monitoring and alerting, accounting,
-the applications themselves.
-
-On the hardware side, R\&D is needed in alternative approaches to data
-archiving to determine the possible cost/performance tradeoffs.
-Currently, tape is extensively used to hold data that cannot be
-economically made available online. While the data is still accessible,
-it comes with a high latency penalty, limiting effective data access. We
-suggest investigating either separate direct access-based archives
-(e.g., disk or optical) or new models that hierarchically overlay online
-direct access volumes with archive space. This is especially relevant
-when access latency is proportional to storage density. Either approach
-would need to also evaluate reliability risks and the effort needed to
-provide data stability. For this work, we should exchange experiences
-with communities that rely on large tape archives for their primary
-storage.
-
-Cost reductions in the maintenance and operation of storage
-infrastructure can be realised through convergence of the major
-experiments and resource providers on shared solutions. This does not
-necessarily mean promoting a monoculture, as different solutions will be
-adapted to certain major classes of use cases, type of site, or funding
-environment. There will always be a judgement to make on the
-desirability of using a variety of specialised systems, or of
-abstracting the commonalities through a more limited, but common,
-interface. Reduced costs and improved sustainability will be further
-promoted by extending these concepts of convergence beyond HEP and into
-the other large-scale scientific endeavours that will share the
-infrastructure in the coming decade (e.g., the SKA and CTA experiments).
-Efforts must be made as early as possible, during the formative design
-phases of such projects, to create the necessary links.
-
-Finally, all changes undertaken must not make the ease of access to data
-any worse than it is under current computing models. We must also be
-prepared to accept the fact that the best possible solution may require
-significant changes in the way data is handled and analysed. What is
-clear is that current practices will not scale to the needs of HL-LHC
-and other major HEP experiments of the coming era.
-
-\subsubsection*{Current Practices}
-
-The original LHC computing models were based on simpler models used
-before distributed computing was a central part of HEP computing. This
-allowed for a reasonably clean separation between four different aspects
-of interacting with data, namely data organisation, data management,
-data access, and data granularity. The meaning of these terms may be 
-summarised in what follows.
-
-\begin{itemize}
-\item
-  \emph{Data organisation} is essentially how data is structured as it
-  is written. Most data is written in files, in ROOT format, typically
-  with a column-wise organisation of the data. The records corresponding
-  to these columns are compressed. The internal details of this
-  organisation are visible only to individual software applications.
-\item
-  In the past, the key challenge for \emph{data management} was the transition to
-  use distributed computing in the form of the grid. The experiments
-  developed dedicated data transfer and placement systems, along with
-  catalogues, to move data between computing centres. Originally,
-  computing models were rather static: data was placed at sites, and the
-  relevant compute jobs were sent to the right locations. Since LHC
-  startup, this model has been made more flexible to limit non-optimal
-  pre-placement and to take into account data popularity. In addition,
-  applications might interact with catalogues or, at times, the workflow
-  management system does this on behalf of the applications.
-\item
-  \emph{Data access}: historically, various protocols have been used for
-  direct reads (rfio, dcap, xrootd, etc.) where jobs are reading data
-  explicitly staged-in or cached by the compute resource used or the
-  site it belongs to. A recent move has been the convergence towards
-  xrootd as the main protocol for direct access. With direct access,
-  applications may use alternative protocols to those used by data
-  transfers between sites. In addition, LHC experiments have been
-  increasingly using remote access to the data, without any stage-in
-  operations, using the possibilities offered by protocols such as xrootd
-  or http.
-\item
-  \emph{Data granularity}: the data is split into datasets, as defined
-  by physics selections and use cases, consisting of a set of individual
-  files. While individual files in datasets can be processed in
-  parallel, the files themselves are usually processed as a whole.
-\end{itemize}
-
-Before LHC turn-on, and in the first years of the LHC, these four
-areas were to first order optimised independently. As LHC computing
-matured, interest has turned to optimisations spanning multiple areas.
-For example, the recent use of ``Data Federations'' mixes up Data
-Management and Access. As we will see below, some of the foreseen
-opportunities towards HL-LHC may require global optimisations.
-
-Thus, in this section we take a broader view than traditional data
-management and consider the combination of ``Data Organisation,
-Management and Access'' (DOMA) together. We believe that this fuller
-picture will provide important opportunities for improving
-efficiency and scaleability, as we enter the many-exabyte era.
-
-\subsubsection*{Research and Development Programme}
-
-In the following, we describe
-tasks that will need to be carried out in order to demonstrate that the
-increased volume and complexity of data expected over the coming decade
-can be stored, accessed, and analysed at an affordable cost.
-
-\begin{itemize}
-\item
-  Sub-file granularity, e.g.,
-  event-based, will be studied to see whether it can be implemented
-  efficiently, and in a scalable, cost-effective manner, for all
-  applications making use of event selection, to see whether it offers
-  an advantage over current file-based granularity. The following tasks
-  should be completed by 2020:
-  \begin{itemize}
-  \item
-    Quantify the impact on
-    performance and resource utilisation  of the storage and network
-    for the main
-    access patterns, i.e., simulation, reconstruction, analysis.
-  \item
-    Assess the impact on
-    catalogues and data distribution.
-  \item
-    Assess whether
-    event-granularity makes sense in object stores that tend to require
-    large chunks of data for efficiency.
-  \item
-    Test for improvement in
-    recoverability from preemption, in particular when using cloud spot
-    resources and/or dynamic HPC resources.
-  \end{itemize}
-\item
-  We will seek to derive benefits
-  from data organisation and analysis technologies adopted by other big
-  data users. A proof-of-concept that involves the following tasks needs
-  to be established by 2020 to allow full implementations to be made in
-  the years that follow.
-  \begin{itemize}
-  \item
-    Study the impact of
-    column-wise, versus row-wise, organisation of data on the
-    performance of each kind of access.
-  \item
-    Investigate efficient data
-    storage and access solutions that support the use of map-reduce or
-    Spark-like analysis services.
-  \item
-    Evaluate just-in-time
-    decompression schemes and mappings onto hardware architectures
-    considering the flow of data, from spinning disk to memory and
-    application.
-  \end{itemize}
-\item
-  Investigate the role data
-  placement optimisations can play, such as caching, in order to use
-  computing resources effectively, and the technologies that can be used
-  for this. The following tasks should be completed by 2020:
-  \begin{itemize}
-  \item
-   Quantify the benefit of
-    placement optimisation for reconstruction,
-    analysis, and simulation.
-  \item
-    Assess the benefit of caching
-    for Machine Learning-based applications, in particular for the
-    learning phase, and follow-up the evolution of technology outside
-    HEP.
-  \end{itemize}
-  In the longer term the benefits
-  that can be derived from using different approaches to the way HEP is
-  currently managing its data delivery systems should be studied. Two
-  different content delivery methods will be looked at, namely Content
-  Delivery Networks (CDN) and Named Data Networking (NDN).
-\item
-  Study how to minimise HEP
-  infrastructure costs by exploiting varied quality of service from
-  different storage technologies. In particular, study the role that
-  opportunistic/tactical storage can play, as well as different archival
-  storage solutions. A proof-of-concept should be made by 2020, with a
-  full implementation to follow in the following years.
-\end{itemize}
-
-\begin{itemize}
-\item
-  Establish how to globally optimise data access latency, with respect
-  to the efficiency of using CPU, at a sustainable cost. This involves
-  studying the impact of concentrating data in fewer, larger locations
-  (the ``data-lake'' approach), and making increased use of opportunistic
-  compute resources located further from the data. Again, a
-  proof-of-concept should be made by 2020, with a full implementation in
-  the following years, if successful. This R\&D will be done in common
-  with the related actions planned as part of Facilities and Distributed
-  Computing.
-\end{itemize}
-
-\hypertarget{facilities-and-distributed-computing}{%
-\subsection{Facilities and Distributed
-Computing}\label{facilities-and-distributed-computing}}
-
-\subsubsection*{Scope and Challenges}
-
-As outlined in Section \ref{software-and-computing-challenges}, 
-huge resource requirements are anticipated for
-HL-LHC running. These need to be deployed and managed across the WLCG
-infrastructure, which has evolved from the original ideas on deployment
-before LHC data-taking started~\cite{MONARC}, to be a mature and
-effective infrastructure that is now exploited by LHC experiments.
-Currently, hardware costs are dominated by disk storage, closely
-followed by CPU, followed by tape and networking. Naive estimates of
-scaling to meet HL-LHC needs indicate that the current system would need
-almost an order of magnitude more resources than will be available from
-technology evolution alone. In addition, other initiatives such as Belle
-II and DUNE in particle physics, but also other science projects such as
-SKA, will require a comparable amount of resources on the same
-infrastructure. Even anticipating substantial software improvements, the
-major challenge in this area is to find the best configuration for
-facilities and computing sites that make HL-LHC computing feasible. This
-challenge is further complicated by substantial regional differences in
-funding models, meaning that any solution must be sensitive to these
-local considerations to be effective.
-
-There are a number of changes that can be anticipated on the timescale
-of the next decade that must be taken into account. There is an
-increasing need to use highly heterogeneous resources, including the
-use of HPC infrastructures (which can often have very particular setups
-and policies that make their exploitation challenging); volunteer
-computing (which is restricted in scope and unreliable, but can be a
-significant resource); and cloud computing,
-both commercial and research. All of these offer different resource
-provisioning interfaces and can be significantly more dynamic than
-directly funded HEP computing sites. In addition, diversity of computing
-architectures is expected to become the norm, with different CPU
-architectures, as well as more specialised GPUs and FPGAs.
-
-This increasingly dynamic environment for resources, particularly CPU,
-must be coupled with a highly reliable system for data storage and a
-suitable network infrastructure for delivering this data to where it
-will be processed. While CPU and disk capacity is expected to increase
-by respectively 15\% and 25\% per year for the same cost~\cite{CERN-COST-EST}, the trends of
-research network capacity increases show a much steeper growth, such as
-two orders of magnitude from now to HL-LHC times. Therefore, the
-evolution of the computing models would need to be more network centric.
-
-In the network domain, there are new technology developments, such as
-Software Defined Networks (SDNs), which enable user-defined high
-capacity network paths to be controlled via experiment software, and
-which could help manage these data flows. These new technologies require
-considerable R\&D to prove their utility and practicality. In addition,
-the networks used by HEP are likely to see large increases in traffic
-from other science domains.
-
-Underlying storage system technology will continue to evolve, for
-example towards object stores, and, as proposed in  
-Data Organisation, Management and Access  
-(Section \ref{data-organisation-management-and-access}),
-R\&D is also necessary to understand their usability and their role in
-the HEP infrastructures. There is also the continual challenge of
-assembling inhomogeneous systems and sites into an effective widely
-distributed worldwide data management infrastructure that is usable by
-experiments. This is particularly compounded by the scale increases for
-HL-LHC where multiple replicas of data (for redundancy and availability)
-will become extremely expensive.
-
-Evolutionary change towards HL-LHC is required, as the experiments will
-continue to use the current system. Mapping out a path for migration
-then requires a fuller understanding of the costs and benefits of the
-proposed changes. A model is needed in which the benefits of such
-changes can be evaluated, taking into account hardware and human costs,
-as well as the impact on software and workload performance that in turn
-leads to physics impact. Even if HL-LHC is the use case used to build
-this cost and performance model, because the ten years of experience
-running large-scale experiments helped to define the needs, it is
-believed that this work, and the resulting model, will be valuable for
-other upcoming data intensive scientific initiatives. This includes
-future HEP projects, such as Belle II, DUNE and possibly ILC
-experiments, but also non-HEP projects, such as SKA.
-
-\subsubsection*{Current Practices}
-
-While there are many particular exceptions, most resources incorporated
-into the current WLCG are done so in independently managed sites,
-usually with some regional organisation structure, and mostly offering
-both CPU and storage. The sites are usually funded directly to provide
-computing to WLCG, and are in some sense then ``owned'' by HEP, albeit
-often shared with others. Frequently substantial cost contributions are
-made indirectly, for example through funding of energy costs or
-additional staff effort, particularly at smaller centres. Tape is found
-only at CERN and at large national facilities, such as the WLCG Tier-1s~\cite{Bird:1695401}.
-
-Interfaces to these computing resources are defined by technical
-operations in WLCG. Frequently there are choices that sites can make
-among some limited set of approved options for interfaces. These can
-overlap in functionality. Some are very HEP specific and recognised as
-over-complex: work is in progress to get rid of them. The acceptable
-architectures and operating systems are also defined at the WLCG level
-(currently x86\_64, running Scientific Linux 6 and compatible), and
-sites can deploy these either directly onto ``bare metal'' or can use an
-abstraction layer, such as virtual machines or containers.
-
-There are different logical networks being used to connect sites: LHCOPN
-connects CERN with the Tier-1 centres and a mixture of LHCONE and
-generic academic networks connect other sites.
-
-Almost every experiment layers its own customised workload and data
-management system on top of the base WLCG provision, with several
-concepts, and a few lower-level components, in common. The pilot job
-model for workloads is ubiquitous, where a real workload is dispatched
-only once a job slot is secured. Data management layers aggregate files
-in the storage systems into datasets and manage experiment-specific
-metadata. In contrast to the MONARC model, sites are generally used more
-flexibly and homogeneously by experiments, both in workloads and in data
-stored.
-
-In total, WLCG currently provides experiments with resources distributed
-at about 170 sites, in 42 countries, which pledge every year the amount
-of CPU and disk resources they are committed to delivering. The pledge
-process is overseen by the Computing Resource Scrutiny Group (CRSG), mandated by
-the funding agencies to validate the experiment requests, and to
-identify mismatches with site pledges. These sites are connected by
-10-100 Gb links, and deliver approximately 500k CPU cores and 1 EB of
-storage, of which 400 PB is disk. More than 200M jobs are executed each day~\cite{Bird2017}.
-
-\subsubsection*{Research and Development programme}
-
-The following areas of study are ongoing, and will involve technology
-evaluations, prototyping, and scale tests. Several of the items below
-require some coordination with other topical areas discussed in this
-document, and some work is still needed to finalise the detailed action
-plan. These actions will need to be structured to meet the common
-milestones of informing the HL-LHC Computing Technical Design Reports (TDRs), and deploying
-advanced prototypes during LHC Run 3.
-
-\begin{itemize}
-\item
-  Understand better the relationship between the performance and costs
-  of the WLCG system, and how it delivers the necessary functionality to
-  support LHC physics. This will be an ongoing process, started by the
-  recently formed System Performance and Cost Modeling Working Group,
-  and aims to provide a quantitative assessment for any proposed
-  changes.
-\item
-  Define the functionality needed to implement a federated data centre
-  concept (``data lake'') that aims to reduce the operational cost of
-  storage for HL-LHC, and at the same time better manage network
-  capacity, whilst maintaining the overall CPU efficiency. This would
-  include the necessary qualities of service, and options for regionally
-  distributed implementations, including the ability to flexibly respond
-  to model changes in the balance between disk and tape. This work
-  should be done in conjunction with the existing Data Organisation,
-  Management and Access Working Group to evaluate the impact of the
-  different access patterns and data organisations envisaged.
-\item
-  Establish an agreement on the common data management functionality
-  that is required by experiments, targeting a consolidation and a lower
-  maintenance burden. The intimate relationship between the management
-  of elements in storage systems and metadata must be recognised. This
-  work requires coordination with the Data Processing Frameworks
-  Working Group. It
-  needs to address at least the following use cases:
-
-  \begin{itemize}
-  \item
-    processing sites that may have some small disk cache, but do not
-    manage primary data;
-  \item
-    fine grained processing strategies that may enable processing of
-    small chunks of data, with appropriate bookkeeping support;
-  \item
-    integration of heterogeneous processing resources, such as HPCs and
-    clou\-ds.
-  \end{itemize}
-\item
-  Explore scalable and uniform means of workload scheduling, which
-  incorporate dynamic heterogenous resources, and the capabilities of
-  finer grained processing that increases overall efficiency. The
-  optimal scheduling of special workloads that require particular
-  resources is clearly required.
-\item
-  Contribute to the prototyping and evaluation of a quasi-interactive
-  analysis facility that would offer a different model for physics
-  analysis, but would also need to be integrated into the data and
-  workload management of the experiments. This is work to be done in
-  collaboration with the Data Analysis and Interpretation Working Group.
-\end{itemize}
-
-\hypertarget{data-flow-processing-framework}{%
-\subsection{Data-Flow Processing
-Framework}\label{data-flow-processing-framework}}
-
-\subsubsection*{Scope and Challenges}
-
-Frameworks in HEP are used for the collaboration-wide
-data processing tasks of triggering, reconstruction, and simulation, as
-well as other tasks that subgroups of the collaboration are responsible
-for, such as detector alignment and calibration. Providing framework
-services and libraries that will satisfy the computing and data needs
-for future HEP experiments in the next decade, while maintaining our
-efficient exploitation of increasingly heterogeneous resources, is a
-huge challenge.
-
-To fully exploit the potential of modern processors, HEP data processing
-frameworks need to allow for the parallel execution of reconstruction or
-simulation algorithms on multiple events simultaneously. Frameworks face
-the challenge of handling the massive parallelism and heterogeneity that
-will be present in future computing facilities, including multi-core and
-many-core systems, GPUs, Tensor Processing Units (TPUs), and tiered
-memory systems, each integrated with storage and high-speed network
-interconnections. Efficient running on heterogeneous resources will
-require a tighter integration with the computing models' higher-level
-systems of workflow and data management. Experiment frameworks must also
-successfully integrate and marshall other HEP software that may have its
-own parallelisation model, such as physics generators and detector
-simulation.
-
-Common developments across experiments are desirable in this area, but
-are hampered by many decades of legacy work. Evolving our frameworks
-also has to be done recognising the needs of the different stakeholders
-in the system. This includes physicists, who are writing processing
-algorithms for triggering, reconstruction or analysis; production
-managers, who need to define processing workflows over massive datasets;
-and facility managers, who require their infrastructures to be used
-effectively. These frameworks are also constrained by security
-requirements, mandated by the groups and agencies in charge of it.
-
-\subsubsection*{Current Practices}
-
-Although most frameworks used in HEP share common concepts, there are,
-for mainly historical reasons, a number of different implementations;
-some of these are shared between experiments. The Gaudi framework~\cite{Barrand:2001ny} was
-originally developed by LHCb, but is also used by ATLAS and various
-non-LHC experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff}, which was forked
-to provide the art framework for the Fermilab Intensity Frontier experiments~\cite{Green:2012gv}. 
-Belle II uses basf2~\cite{1742-6596-331-3-032024}. The linear collider community developed and uses
-Marlin~\cite{Gaede:2006pj}. The FAIR experiments use FairROOT, closely related to ALICE's
-AliROOT. The FAIR experiments and ALICE are now developing a new
-framework, which is called O2~\cite{O2}. At the time of writing, most 
-major frameworks support basic parallelisation, both within and across
-events, based on a task-based model~\cite{Jones:2015soc}\cite{Clemencic:2015paa}.
-
-Each framework has a processing model, which provides the means to
-execute and apportion work. Mechanisms for this are threads, tasks,
-processes, and inter-process communication. The different strategies used
-reflect different trade-offs between constraints in the programming
-model, efficiency of execution, and ease of adapting to inhomogeneous
-resources. These concerns also reflect two different behaviours:
-firstly, maximising throughput, where it is most important to maximise
-the number of events that are processed by a given resource; secondly,
-minimising latency, where the primary constraint is on how long it takes
-to calculate an answer for a particular datum.
-
-Current practice for throughput maximising system architectures have
-con\-strain\-ed the scope of framework designs. Framework applications have
-largely been viewed by the system as a batch job with complex
-configuration, consuming resources according to rules dictated by the
-computing model: one process using one core on one node, operating
-independently with a fixed size memory space on a fixed set of files
-(streamed or read directly). Only recently has CMS broken this tradition
-starting at the beginning of Run 2, by utilising all available cores in one process space using threading. ATLAS is currently
-using a multi-process fork-and-copy-on-write solution to remove the
-constraint of one core/process. Both experiments were driven to solve
-this problem by the ever-growing need for more memory per process
-brought on by the increasing complexity of LHC events. Current practice
-manages systemwide (or facility-wide) scaling by dividing up datasets,
-generating a framework application configuration, and scheduling jobs on
-nodes/cores to consume all available resources. Given anticipated
-changes in hardware (heterogeneity, connectivity, memory, storage)
-available at computing facilities, the interplay between
-workflow and workload management systems and framework applications need to
-be carefully examined. It may be advantageous to permit framework
-applications (or systems) to span multi-node resources, allowing them
-to be first-class participants in the business of scaling within a
-facility. In our community some aspects of this approach, which maps
-features with microservices or function as a service, is being pioneered
-by the O2 framework.
-
-\subsubsection*{Research and Development programme}
-
-By the end of 2018: review the existing technologies that are the
-important building blocks for data processing frameworks and reach
-agreement on the main architectural concepts for the next generation of
-frameworks. Community meetings and workshops, along the lines of the
-original Concurrency Forum, are envisaged in order to foster collaboration in this work~\cite{ConcurrencyForum}. This includes the following:
-
-\begin{itemize}
-\item
-  Libraries used for concurrency, their likely evolution and the issues
-  in integrating the models used by detector simulation and physics
-  generators into the frameworks.
-\item
-  Functional programming, as well as domain specific languages, as a way
-  to describe the physics data processing that has to be undertaken
-  rather than how it has to be implemented. This approach is based on
-  the same concepts as the idea for functional approaches for
-  (statistical) analysis as described in Section \ref{data-analysis-and-interpretation}.
-\item
-  Analysis of the functional differences between the existing frameworks
-  and the different experiment use cases.
-\end{itemize}
-
-By 2020: prototype and demonstrator projects for the agreed
-architectural concepts and baseline to inform the HL-LHC Computing TDRs
-and to demonstrate advances over what is currently deployed. The
-following specific items will have to be taken into account:
-
-\begin{itemize}
-\item
-  These prototypes should be as common as possible between existing
-  frameworks, or at least several of them, as a proof-of-concept of
-  effort and component sharing between frameworks for their future
-  evolution. Possible migration paths to more common implementations
-  will be part of this activity.
-\item
-  In addition to covering the items mentioned for the review phase, they
-  should particularly demonstrate possible approaches for scheduling the
-  work across heterogeneous resources and using them efficiently, with a
-  particular focus on the efficient use of co-processors, such as GPUs.
-\item
-  They need to identify data model changes that are required for an
-  efficient use of new processor architectures (e.g., vectorisation),
-  and for scaling I/O performance in the context of concurrency.
-\item
-  Prototypes of a more advanced integration with workload management,
-  taking advantage in particular of the advanced features available at
-  facilities for a finer control of the interactions with storage and
-  network, and dealing efficiently with the specificities of HPC
-  resources.
-\end{itemize}
-
-By 2022: production-quality framework libraries usable by several
-experiment frameworks, covering the main areas successfully demonstrated
-in the previous phase. During these activities we expect at least one
-major paradigm shift to take place on this 5-year time scale. It will be
-important to continue discussing their impact within the community,
-which will be ensured through appropriate cross-experiment workshops
-dedicated to data processing frameworks.
-
-\hypertarget{conditions-data}{%
-\subsection{Conditions Data}\label{conditions-data}}
-
-\subsubsection*{Scope and Challenges}
-
-Conditions data is defined as the non-event data required by
-data-processing software to correctly simulate, digitise or reconstruct
-the raw detector event data. The non-event data discussed here consists
-mainly of detector calibration and alignment information, with some
-additional data describing the detector configuration, the machine
-parameters, as well as information from the detector control system.
-
-Conditions data is different from event data in many respects, but one
-of the important differences is that its volume scales with time rather
-than with the luminosity. As a consequence, its growth is limited, as
-compared to event data: conditions data volume is expected to be at the
-terabyte scale and the update rate is modest (typically O(1)Hz).
-However, conditions data is used by event processing applications
-running on a very large distributed computing infrastructure, resulting
-in tens of thousands of jobs that may try to access the conditions data
-at the same time, and leading to a very significant rate of reading
-(typically O(10) kHz).
-
-To successfully serve such rates, some form of caching is needed, either
-by using services such as web proxies (CMS and ATLAS use Frontier) or by
-delivering the conditions data as files distributed to the jobs. For the
-latter approach, CVMFS is an attractive solution due to its embedded
-caching, and its advanced snapshotting and branching features. ALICE
-have made some promising tests, and started to use this approach in Run
-2; Belle II already took the same approach~\cite{WoodACAT2017}, and NA62
-has also decided to adopt this solution. However, one particular
-challenge to be overcome with the filesystem approach is to design an
-efficient mapping of conditions data and metadata to files in order to
-use the CVMFS caching layers efficiently.
-
-Efficient caching is especially important in order to support the
-high-reading rates that will be necessary for ATLAS and CMS experiments
-starting with Run 4. For these experiments, a subset of the conditions
-data is linked to the luminosity, leading to an interval of granularity
-down to the order of a minute. Insufficient or inefficient caching may
-impact the efficiency of the reconstruction processing.
-
-Another important challenge is ensuring the long-term maintainability of
-the conditions data storage infrastructure. Shortcomings in the initial
-approach used in LHC Run 1 and Run 2, leading to complex
-implementations, helped to identify the key requirements for an
-efficient and sustainable condition data handling infrastructure. There
-is now a consensus among experiments on these requirements~\cite{Laycock2017}: 
-ATLAS and CMS are working on a common next-generation
-conditions database~\cite{1742-6596-898-4-042047}.
-The Belle II experiment, which is about to start its data taking, has already
-developed a solution based on the same concepts and architecture. One
-key point in this new design is to have a server mostly agnostic to the
-data content with most of the intelligence on the client side. This new
-approach should make it easier to rely on well-established open-source
-products (e.g., Boost) or software components developed for the
-processing of event data (e.g., CVMFS). With such an approach, it should
-be possible to leverage technologies such as REST interfaces to simplify
-insertion and read operations, and make them very efficient to reach the
-rate levels foreseen. Also, to provide a resilient service to jobs that
-depend on it, the client will be able to use multiple proxies or servers
-to access the data.
-
-One conditions data challenge may be linked to the use of an event
-service, as ATLAS is doing currently, to use efficiently HPC facilities
-for event simulation or processing. The event service allows better use
-of resources that may be volatile by allocating and bookkeeping the work
-done, not at the job granularity, but at the event granularity. This
-reduces the possibility for optimising access to the conditions data at
-the job level, and may lead to an increased pressure on the conditions
-data infrastructure. This approach is still at an early stage, and more
-experience is needed to better appreciate the exact impact on the
-conditions data.
-
-\subsubsection*{Current Practices}
-
-The data model for conditions data management is an area where the
-experiments have converged on something like a best common practice. The
-time information for the validity of the Payloads is specified with a
-parameter called an \emph{Interval Of Validity} (IOV), which can be
-represented by a Run number, the ID of a luminosity section or a
-universal timestamp. A fully qualified set of conditions data consists
-of a set of payloads and their associate IOVs covering the time span
-required by the workload. A label called a \emph{Tag} identifies the
-version of the set and the global tag is the top-level configuration of
-all conditions data. For a given detector subsystem and a given IOV, a
-global tag will resolve to one, and only one, conditions data payload.
-The global tag resolves to a particular system tag via the global tag
-map table. A system tag consists of many intervals of validity or
-entries in the IOV table. Finally, each entry in the IOV table maps to a
-payload via its unique hash key.
-
-A relational database is a good choice for implementing this design. One
-advantage of this approach is that a payload has a unique identifier,
-its hash key, and this identifier is the only way to access it. All
-other information, such as tags and IOV, is metadata used to select a
-particular payload. This allows a clear separation of the payload data
-from the metadata, and may allow use of a different backend technology
-to store the data and the metadata. This has potentially several
-advantages:
-
-\begin{itemize}
-\item
-  Payload objects can be cached independently of their metadata, using
-  the appropriate technology, without the constraints linked to metadata
-  queries.
-\item
-  Conditions data metadata are typically small compared to the
-  conditions data themselves, which makes it easy to export them as a
-  single file using technologies such as SQLite. This may help for
-  long-term data preservation.
-\item
-  IOVs, being independent of the payload, can also be cached on their
-  own.
-\end{itemize}
-
-A recent trend is the move to full reconstruction online, where the
-calibrations and alignment are computed and applied in the High Level
-Trigger (HLT). This is currently being tested by ALICE and LHCb, who
-will adopt it for use in Run 3. This will offer an opportunity to
-separate the distribution of conditions data to reconstruction jobs and
-analysis jobs, as they will not run on the same infrastructure. However,
-running reconstruction in the context of the HLT will put an increased
-pressure on the access efficiency to the conditions data, due to the HLT
-time budget constraints.
-
-\subsubsection*{Research and Development Programme}
-
-R\&D actions related to Conditions databases are already in progress,
-and all the activities described below should be completed by 2020. This
-will provide valuable input for the future HL-LHC TDRs, and allow these
-services to be deployed during Run 3 to overcome the limitations seen in
-today's solutions.
-
-\begin{itemize}
-\item
-  File-system view of conditions data for analysis jobs: study how to
-  leverage advanced snapshotting/branching features of CVMFS for
-  efficiently distributing conditions data as well as ways to optimise
-  data/metadata layout in order to benefit from CVMFS caching. Prototype
-  production of the file-system view from the conditions database.
-\item
-  Identify and evaluate industry technologies that could replace
-  HEP-specific components.
-\item
-  ATLAS: migrate current implementations based on COOL to the proposed
-  REST-based approach; study how to avoid moving too much complexity on
-  the client side, in particular for easier adoption by subsystems, e.g.,
-  possibility of common modules/libraries. ALICE is also planning to
-  explore this approach for the future, as an alternative or to
-  complement the current CVMFS-based implementation.
-\end{itemize}
-
-\hypertarget{visualisation}{%
-\subsection{Visualisation}\label{visualisation}}
-
-\subsubsection*{Scope and Challenges}
-
-In modern High Energy Physics (HEP) experiments, visualisation of data
-has a key role in many activities and tasks across the whole data
-processing chain: detector development, monitoring, event generation,
-reconstruction, detector simulation, data analysis, as well as outreach
-and education.
-
-\emph{Event displays} are the main tool to explore experimental data at
-the event level and to visualise the detector itself. There are two main
-types of application: firstly, those integrated in the experiments'
-frameworks, which are able to access and visualise all the experiments'
-data, but at a cost in terms of complexity; secondly,
-those designed as cross-platform applications, lightweight and fast,
-delivering only a simplified version or a subset of the event data. In
-the first case, access to data is tied intimately to an experiment's
-data model (for both event and geometry data) and this inhibits
-portability; in the second, processing the experiment data into a
-generic format usually loses some detail and is an extra processing
-step. In addition, there are various graphical backends that can be used
-to visualise the final product, either standalone or within a browser,
-and these can have a substantial impact on the types of devices
-supported.
-
-Beyond event displays, HEP also uses \emph{visualisation of statistical
-information}, typically histograms, which allow the analyst to quickly
-characterise the data. Unlike event displays, these visualisations are
-not strongly linked to the detector geometry, and often aggregate data
-from multiple events. Other types of visualisation are used to
-\emph{display non-spatial data}, such as graphs for describing the
-logical structure of the detector or for illustrating dependencies
-between the data products of different reconstruction algorithms.
-
-The main challenges in this domain are in the sustainability of the many
-experi\-ment specific visualisation tools when common projects could
-reduce duplication and increase quality and long-term maintenance. The
-ingestion of events and other data could be eased by common formats,
-which would need to be defined and satisfy all users. Changes to support
-a client-server architecture would help broaden the ability to support
-new devices, such as mobile phones. Making a good choice for the
-libraries used to render 3D shapes is also key, impacting on the range
-of output devices that can be supported and the level of interaction
-with the user. Reacting to a fast-changing technology landscape is very
-important -- HEP's effort is limited and generic solutions can often be
-used with modest effort. This applies strongly to non-event
-visualisation, where many open source and industry standard tools can be
-exploited.
-
-\subsubsection*{Current Practices}
-
-Three key features characterise almost all HEP event displays:
-
-\begin{itemize}
-\item
-  \textbf{Event-based workflow}: applications access experimental data
-  on an event-by-event basis, visualising the data collections belonging
-  to a particular event. Data can be related to the actual physics
-  events (e.g., physics objects such as jets or tracks) or to the
-  experimental conditions (e.g., detector descriptions, calibrations).
-\item
-  \textbf{Geometry visualisation}: The application can display the
-  geometry of the detector, as retrieved from the experiments' software
-  frameworks, or a simplified description, usually for the sake of speed
-  or portability.
-\item
-  \textbf{Interactivity}: applications offer different interfaces and
-  tools to users, in order to interact with the visualisation itself,
-  select event data, and set cuts on objects' properties.
-\end{itemize}
-
-Experiments have often developed multiple event displays that either
-take the full integration approach explained above or are standalone and
-rely on extracted and simplified data.
-
-The visualisation of data can be achieved through the low level OpenGL
-API, by the use of higher-level OpenGL-based libraries, or within a web
-browser using WebGL. Using OpenGL directly is robust and avoids other dependencies, but
-implies a significant effort. Instead of using the API directly, a
-library layer on top of OpenGL (e.g., Coin3D) can more closely match the
-underlying data, such as geometry, and offers a higher level API that
-simplifies development. However, this carries the risk that if the
-library itself becomes deprecated, as has happened with Coin3D, the
-experiment needs to migrate to a different solution or to take on the
-maintenance burden itself. Standalone applications often use WebGL
-technology to render 3D objects inside a web browser. This is a very
-convenient way of rendering 3D graphics, due to the cross-platform
-nature of web technologies, and offers many portability advantages (e.g.,
-easier support for mobile or virtual reality devices), but at some cost
-of not supporting the most complex visualisations requiring heavy
-interaction with the experiments' data.
-
-In recent years, video game engines, such as Unity or the Unreal Engine,
-have become particularly popular in the game and architectural
-visualisation industry. They provide very sophisticated graphics engines
-and offer a lot of tools for user interaction, such as menu systems or
-native handling of VR devices. They are well supported by industry and
-tend to have a long lifespan (Unreal Engine is now  20 years old and
-is still very popular). However, such engines are meant to be used as
-development frameworks and their usage in HEP code is not always
-evident. Code should be developed within them, while in HEP
-framework-based applications we often want to use graphics libraries
-that can be integrated in existing code. A number of HEP collaborations
-have started experimenting in building event display tools with such
-engines, among them Belle II and ATLAS, but their use is currently
-limited to the display of simplified data only.
-
-The new client-server architecture proposed as one of the visualisation
-R\&D activities will ease the usage of WebGL technologies and game
-engines in HEP.
-
-For statistical data, ROOT has been the tool of choice in HEP for many
-years and satisfies most use cases. However, increasing use of generic
-tools and data formats means Matplotlib (Python) or JavaScript based
-solutions (used, for example, in Jupyter notebooks) have made the
-landscape more diverse. For visualising trees or graphs interactively,
-there are many generic offerings and experiments have
-started to take advantage of them.
-
-\subsubsection*{Research and Development Roadmap}
-
-The main goal of R\&D projects in this area will be to develop
-techniques and tools that let visualisation applications and event
-displays be less dependent on specific experiments' software frameworks,
-leveraging common packages and common data formats. Exporters and
-interface packages will be designed as bridges between the experiments'
-frameworks, needed to access data at a high level of detail, and the
-common packages based on the community standards that this group will
-develop.
-
-As part of this development work, demonstrators will be designed to show
-the usability of our community solutions and tools. The goal will be to
-get a final design of those tools so that the experiments can depend on
-them in their future developments.
-
-The working group will also work towards a more convenient access to geometry and
-event data, through a client-server interface. In collaboration with the
-Data Access and Management Working Group, an API or a service to deliver streamed
-event data would be designed.
-
-The work above should be completed by 2020.
-
-Beyond that point, the focus will be on developing the actual
-community-driven tools, to be used by the experiments for their
-visualisation needs in production, potentially taking advantage of new
-data access services.
-
-The workshop that was held as part of the CWP process was felt to be
-extremely useful for exchanging knowledge between developers in
-different experiments, fostering collaboration and in bringing in ideas
-from outside the community. This will now be held as an annual event and
-will facilitate work on the common R\&D plan.
-
-\hypertarget{software-development-deployment-validation-and-verification}{%
-\subsection{Software Development, Deployment, Validation and
-Verification}\label{software-development-deployment-validation-and-verification}}
-
-\subsubsection*{Scope and Challenges}
-
-Modern HEP experiments are often large distributed collaborations
-with several hundred people actively writing software. It is
-therefore vital that the processes and tools used for development are
-streamlined to ease the process of contributing code and to facilitate
-collaboration between geographically separated peers. At the same time,
-we must properly manage the whole project, ensuring code quality,
-reproducibility, and maintainability with the least effort possible.
-Making sure this happens is largely a continuous process and shares a
-lot with non-HEP specific software industries.
-
-Work is ongoing to track and promote solutions in the following areas:
-
-\begin{itemize}
-\item
-  Distributed development of software components, including the tools
-  and processes required to do so (code organisation, documentation,
-  issue tracking, artefact building), and the best practices in terms of
-  code and people management.
-\item
-  Software quality, including aspects such as modularity and reusability
-  of the developed components, architectural and performance best
-  practices.
-\item
-  Software sustainability, including both development and maintenance
-  efforts, as well as best practices given long timescales of HEP
-  experiments.
-\item
-  Deployment of software and interaction with operations teams.
-\item
-  Validation of the software both at small scales (e.g., best practices
-  on how to write a unit test) and larger ones (large scale validation
-  of data produced by an experiment).
-\item
-  Software licensing and distribution, including their impact on
-  software interoperability.
-\item
-  Recognition of the significant contribution that software makes to HEP
-  as a field (also see Section \ref{training-and-careers} regarding
-  career recognition).
-\end{itemize}
-
-HEP-specific challenges derive from the fact that HEP is a large,
-inhomogeneous community with multiple sources of funding, mostly formed
-of people belonging to university groups and HEP-focused laboratories.
-Software development effort within an experiment usually encompasses a
-huge range of experience and skills, from a few more or less full-time
-experts to many physicist programmers with little formal software
-training. In addition, the community is split between different
-experiments that often diverge in timescales, size, and resources.
-Experiment software is usually divided in two separate use cases:
-production (being it data acquisition, data reconstruction or
-simulation) and user analysis, whose requirements and lifecycles are
-completely different. The former is very carefully managed in a
-centralised and slow-moving manner, following the schedule of the
-experiment itself. The latter is much more dynamic and strongly coupled
-with conferences or article publication timelines. Finding solutions
-that adapt well to both cases is not always obvious or even possible.
-
-\subsubsection*{Current Practices}
-
-Due to significant variations between experiments at various stages of
-their lifecycles, there is a huge variation in practice across the
-community. Thus, here we describe \emph{best practice}, with the
-understanding that this ideal may be far from the reality for some
-developers.
-
-It is important that developers can focus on the design and
-implementation of the code and do not have to spend a lot of time on
-technical issues. Clear procedures and policies must exist to perform
-administrative tasks in an easy and quick way. This starts with the
-setup of the development environment. Supporting different platforms not
-only allows developers to use their machines directly for 
-development, it also provides a check of code portability. Clear
-guidance and support for good design must be available in advance of
-actual coding.
-
-To maximise productivity, it is very beneficial to use development tools
-that are not HEP-specific. There are many open source projects that are
-of similar scale to large experiment software stacks and standard tools
-are usually well documented. For source control HEP has generally chosen
-to move to \emph{git}~\cite{git},
-which is very welcome, as it also brings an
-alignment with many open source projects and commercial organisations.
-Likewise, \emph{CMake}~\cite{CMake} 
-is widely used for the builds of software
-packages, both within HEP and outside. Packaging many build products
-together into a software stack is an area that still requires close
-attention with respect to active developments (the HSF has an active
-working group here).
-
-Proper testing of changes to code should always be done in advance of a
-change request to be accepted. Continuous integration, where `merge' or
-`pull' requests are built and tested in advance, is now standard practice
-in the open source community and in industry. Continuous integration can
-run unit and integration tests, and can also incorporate code quality
-checks and policy checks that help improve the consistency and quality
-of the code at low human cost. Further validation on different platforms
-and at large scales must be as automated as possible, including the
-deployment of build artefacts for production.
-
-Training (Section \ref{training-and-careers}) and documentation 
-are key to efficient use
-of developer effort. Documentation must cover best practices and
-conventions as well as technical issues. For documentation that has to
-be specific, the best solutions have a low barrier of entry for new
-contributors, but also allow and encourage review of material.
-Consequently, it is very useful to host documentation sources in a
-repository with a similar workflow to code, and to use an engine that
-translates the sources into modern web pages.
-
-Recognition of software work as a key part of science has resulted in a
-number of journals where developers can publish their work~\cite{SSI2017}. Journal publication also disseminates information to the
-wider community in a permanent way and is the most established mechanism
-for academic recognition. Publication in such journals provides proper
-peer review, beyond that provided in conference papers, so it is
-valuable for recognition as well as dissemination. However, this
-practice is not widespread enough in the community and needs further
-encouragement.
-
-\subsubsection*{Research and Development Programme}
-
-HEP must endeavour to be as responsive as possible to developments
-outside of our field. In terms of hardware and software tools, there
-remains great uncertainty as to what the platforms offering the best
-value for money will be on the timescale of a decade. It therefore
-behoves us to be as generic as possible in our technology choices,
-retaining the necessary agility to adapt to this uncertain future.
-
-Our vision is characterised by HEP being current with technologies and
-para\-digms that are dominant in the wider software development community,
-especially for open-source software, which we believe to be the right
-model for our community. In order to achieve that aim, we propose that
-the community establishes a development forum that allows for technology
-tracking and discussion of new opportunities. The HSF can play a key
-role in marshalling this group and in ensuring its findings are widely
-disseminated. In addition, having wider and more accessible training for
-developers in the field, that will teach the core skills needed for
-effective software development, would be of great benefit.
-
-Given our agile focus, it is better to propose here projects and
-objectives to be investigated in the short to medium term, alongside
-establishing the means to continually review and refocus the community
-on the most promising areas. The main idea is to investigate new tools
-as demonstrator projects where clear metrics for success in a reasonable
-time should be established to avoid wasting community effort on
-initially promising products that fail to live up to expectations.
-
-Ongoing activities and short-term projects, include the following:
-
-\begin{itemize}
-\item
-  Establish a common forum for the discussion of HEP software problems.
-  This should be modeled along the lines of the Concurrency Forum~\cite{ConcurrencyForum}, which was very successful in establishing
-  demonstrators and prototypes that were used as experiments started to
-  develop parallel data processing frameworks.
-\item
-  Continue the HSF working group on \emph{Packaging}, with more
-  prototype implementations based on the strongest candidates identified
-  so far.
-\item
-  Provide practical advice on how to best set up new software packages,
-  developing on the current project template work, and working to
-  advertise this within the community.
-\item
-  Work with HEP experiments and other training projects to provide
-  accessible core skills training to the community (see Section \ref{training-and-careers}). This
-  training should be experiment-neutral, but could be usefully combined
-  with the current experiment specific training. Specifically, this work
-  can build on, and collaborate with, recent highly successful
-  initiatives such as the LHCb \emph{Starterkit}~\cite{LHCbStarterkit}
-  and ALICE \emph{Juniors}~\cite{ALICEJuniors}, and with established
-  generic training initiatives such as \emph{Software Carpentry}~\cite{SoftwareCarpentry}.
-\item
-  Strengthen links with software communities and conferences outside of
-  the HEP domain, presenting papers on the HEP experience and problem
-  domain. The Scientific Computing with Python (SciPy), the
-  Supercomputing Conferences (SCxx), the Conference of
-  Research Software Engineers \emph{(RSE)}, and the Workshops on
-  Sustainable Software for Science: Practice and Experiences (WSSSPE)
-  would all be useful meetings to consider.
-\item
-  Write a paper that looks at case studies of successful and
-  unsuccessful HEP software developments and that draws specific
-  conclusions and advice for future projects.
-\item
-  Strengthen the publication record for important HEP software packages.
-  Both peer-reviewed journals~\cite{SSI2017} and citable software version
-  records (such as DOIs obtained via Zenodo~\cite{Zenodo}).
-\end{itemize}
-
-Medium term projects include the following:
-
-\begin{itemize}
-\item
-  Prototype C++ refactoring tools, with specific use cases in migrating
-  HEP code.
-\item
-  Prototyping of portable solutions for exploiting modern vector
-  hardware on heterogenous platforms.
-\item
-  Support the adoption of industry standards and solutions over
-  HEP-specific implementations whenever possible.
-\item
-  Develop tooling and instrumentation to measure software performance
-  where tools with sufficient capabilities are not available from
-  industry, especially in the domain of concurrency. This should
-  primarily aim to further developments of existing tools, such as
-  \emph{igprof}~\cite{Eulisse:2005zz},
-  rather than to develop new ones.
-\item
-  Develop a common infrastructure to gather and analyse data about
-  experiments' software, including profiling information and code
-  metrics, and to ease sharing across different user communities.
-\item
-  Undertake a feasibility study of a common toolkit for statistical
-  analysis that would be of use in regression testing for experiment's
-  simulation and reconstruction software.
-\end{itemize}
-
-\hypertarget{data-and-software-preservation}{%
-\subsection{Data and Software
-Preservation}\label{data-and-software-preservation}}
-
-\subsubsection*{Scope and Challenges}
-
-Given the very large investment in particle physics experiments, it is
-incumbent upon physicists to preserve the data and the knowledge that
-leads to scientific results in a manner such that this investment is not
-lost to future generations of scientists. For preserving ``data'', at
-whatever stage of production, many of the aspects of the low level
-bit-wise preservation have been covered by the Data Preservation for HEP
-group~\cite{DPHEP}. ``Knowledge'' preservation encompasses the more
-challenging aspects of retaining processing and analysis software,
-documentation, and other components necessary for reusing a given
-dataset. Preservation of this type can enable new analyses on older
-data, as well as a way to revisit the details of a result after
-publication. The latter can be especially important in resolving
-conflicts between published results, applying new theoretical
-assumptions, evaluating different theoretical models, or tuning new
-modeling techniques.
-
-Preservation enabling reuse can offer tangible benefits within a given
-experiment. The preservation of software and workflows such that they
-can be shared enhances collaborative work between analysts and analysis
-groups, providing a way of capturing the knowledge behind a given
-analysis during the review process. It enables easy transfer of knowledge
-to new students or analysis teams, and could establish a manner by which
-results can be generated automatically for submission to central
-repositories, such as HEPData~\cite{Maguire:2017ypu}.
-Preservation within an experiment can
-provide ways of reprocessing and reanalysing data that could have been
-collected more than a decade earlier. Benefits from preservation are
-derived internally whether or not analysis work is approved through the
-publication approval process for an experiment. Providing such immediate
-benefits makes the adoption of data preservation in experiment workflows
-particularly desirable.
-
-A final series of motivations comes from the potential re-use by others
-outside of the HEP experimental community. Significant outreach efforts
-to bring the excitement of analysis and discovery to younger students
-have been enabled by the preservation of experimental data and software
-in an accessible format. Many examples also exist of phenomenology
-papers reinterpreting the results of a particular analysis in a new
-context. This has been extended further with published results based on
-the reanalysis of processed data by scientists outside of the
-collaborations. Engagement of external communities, such as machine
-learning specialists, can be enhanced by providing the capability to
-process and understand low-level HEP data in portable and relatively
-platform-independent way, as happened with the Kaggle ML
-challenges~\cite{pmlr-v42-cowa14}. 
-This allows external users direct access to the same tools
-and data as the experimentalists working in the collaborations.
-Connections with industrial partners, such as those fostered by CERN
-OpenLab, can be facilitated in a similar manner.
-
-Preserving the knowledge of analysis, given the extremely wide scope of
-how analysts do their work and experiments manage their workflows, is
-far from easy. The level of reuse that is applicable needs to be
-identified, and so a variety of preservation systems will probably be
-appropriate given the different preservation needs between large central
-experiment workflows and the work of an individual analyst. The larger
-question is to what extent common low-level tools can be provided that
-address similar needs across a wide scale of preservation problems.
-These would range from capture tools, that preserve the details of an
-analysis and its requirements, to ensuring that software and services
-needed for a workflow would continue to function as required.
-
-The above-mentioned steps can be considered to be consistent with the
-FAIR data principles that are increasingly being mandated by funding
-agencies~\cite{FAIRdata}.
-
-\subsubsection*{Current Practices}
-
-Each of the LHC experiments has adopted a data access and/or data
-preservation policy, all of which can be found on the CERN Open Data
-Portal~\cite{ODP}. All of the LHC experiments support public access to
-some subset of the data in a highly reduced data format for the purposes
-of outreach and education. CMS has gone one step further, releasing
-substantial datasets in an Analysis Object Data (AOD) format that can be
-used for new analyses. The current data release includes simulated data,
-virtual machines that can instantiate the added analysis examples, and
-extensive documentation~\cite{CMS-OpenData}. ALICE has promised to
-release 10\% of their processed data after a five-year embargo and has
-released 2010 data at this time~\cite{ALICE-OpenData}.
-LHCb is willing to make access to reconstructed data available, but is
-unable to commit to a specific timescale due to resource limitations. A
-release of ntuple-level data for one high profile analysis, aimed
-primarily at educational activities, is currently in preparation. ATLAS
-has chosen a different direction for data release: data associated with
-journal publications is made available, and ATLAS also strives to make
-available additional material that allows reuse and reinterpretations of
-the data in the context of new theoretical models~\cite{ATL-CB-PUB-2015-001}.
-ATLAS is  exploring how
-to provide the capability for reinterpretation of searches in the future
-via a service such as 
-RECAST~\cite{Cranmer:1299950}, in which the original internal analysis
-code (including full detector simulation and reconstruction) is
-preserved, as opposed to the re-coding approach with object-efficiency
-calibrations used by external reinterpretation toolkits. All experiments
-frequently provide detailed supplemental data along with publications to
-allow for more detailed comparisons between results, or even
-reinterpretation.
-
-The LHC experiments have not yet set a formal policy addressing the new
-capabilities of the CERN Analysis Preservation Portal (CAP)~\cite{CAP}
-and whether or not some use of it will be required or merely encouraged.
-All of them support some mechanisms for internal preservation of the
-knowledge surrounding a physics publication~\cite{Shiers2017}.
-
-\subsubsection*{Research and Development Programme}
-
-There is a significant programme of work already happening in the data
-preservation area. The feasibility and cost of common base services have
-been studied for bit preservation, the preservation of executable
-software environments, and the structured capturing of analysis metadata~\cite{Berghaus2016}.
-
-The goals presented here should be orchestrated in conjunction with
-projects conducted by the R\&D programmes of other working groups, since
-the questions addressed are common. Goals to address on the timescale of
-2020 are:
-
-\begin{itemize}
-\item
-  Include embedded elements for the capture of preservation information
-  and metadata and tools for the archiving of this information in
-  developing a prototype analysis ecosystem(s). This should include an
-  early demonstration of the CAP analysis preservation portal with a
-  working UI.
-\item
-  Demonstrate the capability to provision and execute production
-  workflows for experiments that are composed of multiple independent
-  containers.
-\item
-  Collection of analysis use cases and elements that are necessary to
-  preserve in order to enable re-use and to ensure these analyses can be
-  captured in developing systems. This should track analysis evolution
-  towards possible \emph{Big Data} environments and determine any elements
-  that are difficult to capture, spawning further R\&D.
-\item
-  Evaluate, in the preservation area, the full potential and limitations
-  of sandbox and ``freezing'' technologies, possibly coupled with
-  version and history control software distribution systems.
-\item
-  Develop prototypes for the preservation and validation of large-scale
-  production executables and workflows.
-\item
-  Integrate preservation capabilities into newly developed computing
-  tools and workflows.
-\item
-  Extension and standardisation of the \emph{final data and analysis}
-  preservation sche\-me via HEPData, Rivet and/or other reinterpretation
-  tools. This could be used to preserve a sufficiently detailed
-  re-usable record of many LHC Run 2 research outputs.
-\end{itemize}
-
-This would then lead naturally to deployed solutions that support data
-preservation in the 2020-2022 time frame for the HEP experimental
-programmes, in particular an analysis ecosystem that enables reuse for
-any analysis that can be conducted in the ecosystem, and a system for
-the preservation and validation of large-scale production workflows.
-
-\hypertarget{security}{%
-\subsection{Security}\label{security}}
-
-\subsubsection*{Scope and Challenges}
-
-Security is a cross-cutting area that impacts our projects,
-collaborative work, users, and software infrastructure fundamentally. It
-crucially shapes our reputation, our collaboration, the trust between
-participants, and the users' perception of the quality and ease of use
-of our services.
-
-There are three key areas:
-
-\begin{itemize}
-\item
-  Trust and policies; this includes trust models, policies, compliance,
-  data protection issues.
-\item
-  Operational security; this includes threat intelligence, security
-  operations, incident response.
-\item
-  Authentication and Authorisation; this includes identity management,
-  identity federation, access control.
-\end{itemize}
-
-\paragraph{Trust and Policies}
-
-Data Protection defines the boundaries that enable HEP work to be
-conducted, in particular regarding data sharing aspects, for example
-between the EU and the US. It is essential to establish a trusted
-personal data exchange framework, minimising the amount of personal data
-to be processed and ensuring legal compliance.
-
-Beyond legal compliance and best practice, offering open access to
-scientific resources and achieving shared goals requires prioritising
-the protection of people and science, including the mitigation of the
-effects of surveillance programs on scientific collaborations.
-
-On the technical side, it is necessary to adapt the current, aging trust
-model and security architecture relying solely on X.509 (which is no longer
-the direction industry is taking), in order to include modern data
-exchange design, for example involving commercial providers or hybrid
-clouds. The future of our infrastructure involves increasingly diverse
-resource providers connected through cloud gateways. For example,
-HEPCloud~\cite{HEPCloud} at FNAL aims to connect Amazon,
-Google Clouds, and HPC centres with our traditional grid computing
-resources. The HNSciCloud European Project~\cite{HNSciCloud} aims to
-support the enhancement of commercial cloud providers in order to be
-leveraged by the scientific community. These are just two out of a
-number of endeavours. As part of this modernisation, a transition is
-needed from a model in which all participating organisations are bound
-by custom HEP security policies to a more flexible approach where some
-partners are not in a position to adopt such policies.
-
-\paragraph{Operational Security and Threat Intelligence}
-
-As attacks have become extremely sophisticated and costly to defend
-against, the only cost-effective strategy is to address security threats
-together, as a community. This involves constantly striving to liaise
-with external organisations, including security vendors and law
-enforcement entities, to enable the sharing of indicators of compromise
-and threat intelligence between all actors. For organisations from all
-sectors, including private companies, governments, and academia, threat
-intelligence has become the main means by which to detect and manage
-security breaches.
-
-In addition, a global forum for HEP and the larger Research and Education
-(R\&E) community needs to be built, where security experts feel
-confident enough to share threat intelligence and security expertise. A
-key to success is to ensure a closer collaboration between HEP security
-contacts and campus security. The current gap at many HEP organisations
-is both undermining the community's security posture and reducing the
-effectiveness of the HEP security strategy.
-
-There are several very active trust groups in the HEP community where
-HEP participants share threat intelligence and organise coordinated
-incident response \cite{EGI-CSIRT,REN-ISAC,XSEDE}.
-There is unfortunately still no global Research and Education forum for
-incident response, operational security, and threat intelligence
-sharing. With its mature security operations and dense, global network
-of HEP organisations, both of which are quite unique in the research
-sector, the HEP community is ideally positioned to contribute to such a
-forum and to benefit from the resulting threat intelligence, as it has
-exposure, sufficient expertise, and connections to lead such an
-initiative. It may play a key role in protecting multiple scientific
-domains at a very limited cost.
-
-There will be many technology evolutions as we start to take a serious
-look at the next generation internet. For example, IPv6 is one upcoming
-change that has yet to be fully understood from the security
-perspective. Another high impact area is the internet of things (IoT),
-connected devices on our networks that create new vectors of attack.
-
-It will become necessary to evaluate and maintain operational security
-in connected environments spanning public, private, and hybrid clouds.
-The trust relationship between our community and such providers has yet
-to be determined, including the allocation of responsibility for
-coordinating and performing vulnerability management and incident
-response. Incompatibilities between the e-Infrastructure approach to
-community-based incident response and the ``pay-for-what-you-break''
-model of certain commercial companies may come to light and must be
-resolved.
-
-\paragraph{Authentication and Authorisation Infrastructure}
-
-It is now largely acknowledged that end-user certificates are
-challenging to manage and create a certain entrance barrier to our
-infrastructure for early career researchers. Integrating our access
-control management system with new, user-friendly technologies and
-removing our dependency on X.509 certificates is a key area of interest
-for the HEP Community.
-
-An initial step is to identify other technologies that can satisfy
-traceability, isolation, privilege management and other requirements
-necessary for HEP workflows. The chosen solution should prioritise
-limiting the amount of change required to our services and follow
-accepted standards to ease integration with external entities, such as
-commercial clouds and HPC centres.
-
-Trust federations and inter-federations, such as the R\&E standard
-eduGAIN \cite{eduGAIN}, provide a needed functionality for
-Authentication. They can remove the burden of identity provisioning from
-our community and allow users to leverage their home organisation
-credentials to access distributed computing resources. Although certain
-web-based services have enabled authentication via such federations,
-uptake is not yet widespread. The challenge remains to have the
-necessary attributes published by each federation to provide robust
-authentication.
-
-The existing technologies leveraged by identity federations, e.g., the
-Security Assertion Markup Language (SAML), have not supported non-web
-applications historically. There is momentum within the wider community
-to develop next-generation identity federations that natively support a
-wider range of clients. In the meantime there are several viable interim
-solutions that are able to provision users with the token required to
-access a service (such as X.509) transparently, translated from their
-home organisation identity.
-
-Although federated identity provides a potential solution for our
-challenges in Authentication, Authorisation should continue to be
-tightly controlled by the HEP community. Enabling Virtual Organisation
-(VO) membership for federated credentials and integrating such a
-workflow with existing identity vetting processes is a major topic
-currently being worked on, in particular within the WLCG community.
-Commercial clouds and HPC centres have fundamentally different access
-control models and technologies from our grid environment. We shall need
-to enhance our access control model to ensure compatibility and
-translate our grid-based identity attributes into those consumable by
-such services.
-
-\subsubsection*{Current Activities}
-
-Multiple groups are working on policies and establishing a common trust
-framework, including the EGI Security Policy Group
-\cite{EGISecurityPolicyGroup} and the Security for Collaboration among
-Infrastructures working group~\cite{SCI_WG}.
-
-Operational security for the HEP community is being followed up in the
-WLCG Working Group on Security Operations Centres~\cite{WLCG-SOC-WG}. The
-HEP Community is actively involved in multiple operational security
-groups and trust groups, facilitating the exchange of threat
-intelligence and incident response communication. WISE~\cite{WISE}
-provides a forum for e-Infrastructures to share and develop security
-best practices and offers the opportunity to build relationships between
-security representatives at multiple e-infrastructures of interest to
-the HEP community.
-
-The evolution of Authentication and Authorisation is being evaluated in
-the recently created WLCG Working Group on Authorisation. 
-In parallel, HEP is contributing to a wider effort
-to document requirements for multiple Research Communities through the
-work of FIM4R~\cite{FIM4R}. CERN's participation in the European
-Authentication and Authorisation for Research and Collaboration (AARC)
-project~\cite{AARC} provides the opportunity to ensure
-that any directions chosen are consistent with those taken by the wider
-community of research collaborations. The flow of attributes between
-federated entities continues to be problematic, disrupting the
-authentication flow. Trust between service providers and identity
-providers is still evolving, and efforts within the R\&E Federations
-Group (REFEDS)~\cite{REFEDS} and the AARC project aim to address the
-visibility of both the level of assurance of identities and the security
-capability of federation participants (through Sirtfi~\cite{Sirtfi}).
-
-\subsubsection*{Research and Development Programme}
-
-Over the next decade, it is expected that considerable changes will be
-made to address security in the domains highlighted above. The
-individual groups, in particular those mentioned above, working in the
-areas of trust and policies, operational security, authentication and
-authorisation, and technology evolutions, are driving the R\&D
-activities. The list below summarises the most important actions:
-
-\paragraph{Trust and Policies}
-\begin{itemize}
-\item 
-By 2020:
-\begin{itemize}
-    \item
-      Define and adopt policies in line with new EU Data Protection
-      requirements.
-    \item
-      Develop frameworks to ensure trustworthy interoperability of
-      infrastructures and communities.
-\end{itemize}
-
-\item
-By 2022:
-\begin{itemize}
-    \item
-      Create and promote community driven incident response policies and
-      procedures.
-\end{itemize}
-\end{itemize}
-
-\paragraph{Operational Security and threat intelligence}
-\begin{itemize}
-\item
-By 2020:
-\begin{itemize}
-\item
-  Offer a reference implementation, or at least specific guidance, for a
-  Security Operation Centre deployment at HEP sites, enabling them to
-  take action based on threat intelligence shared within the HEP
-  community.
-\end{itemize}
-\item
-By 2022:
-\begin{itemize}
-\item
-  Participate in the founding of a global Research and Education Forum
-  for incident response, since responding as a global community is the only
-  effective solution against global security threats.
-\item
-  Build the capabilities to accommodate more participating organisations
-  and streamline communication workflows, within and outside HEP,
-  including maintaining a list of security contacts, secure communications
-  channels, and security incident response mechanisms.
-\item
-  Reinforce the integration of HEP security capabilities with their
-  respective home organisation, to ensure adequate integration of HEP
-  security teams and site security teams.
-\end{itemize}
-\item 
-By 2025:
-\begin{itemize}
-\item
-  Prepare adequately as a community, in order to enable HEP
-  organisations to operate defendable services against more
-  sophisticated threats, stemming both from global cyber-criminal gangs
-  targeting HEP resources (finance systems, intellectual property,
-  ransomware), as well as from state actors targeting the energy and
-  research sectors with advanced malware.
-\end{itemize}
-\end{itemize}
-
-\paragraph{Authentication and Authorisation}
-
-\begin{itemize}
-\item
-By 2020:
-\begin{itemize}
-\item
-  Ensure that ongoing efforts in trust frameworks are sufficient to
-  raise the level of confidence in federated identities to the
-  equivalent of X.509, at which stage they could be a viable alternative
-  to both grid certificates and CERN accounts.
-\item
-  Participate in setting directions for the future of identity
-  federations, through the FIM4R~\cite{FIM4R} community.
-\end{itemize}
-\item
-By 2022:
-\begin{itemize}
-\item
-  Overhaul the current Authentication and Authorisation infrastructure,
-  including Token Translation, integration with Community IdP-SP
-  Proxies, and Membership Management tools. Enhancements in this area
-  are needed to support a wider range of user identities for WLCG
-  services.
-\end{itemize}
-\end{itemize}
-
-\hypertarget{training-and-careers}{%
-\section{Training and Careers}\label{training-and-careers}}
-
-For HEP computing to be as successful as possible, the careers and
-skills of the individuals who participate must be considered. Ensuring
-that software developers can acquire the necessary skills and obtain
-successful careers is considered an essential goal of the HSF, which has
-the following specific objectives in its mission:
-
-\begin{itemize}
-\item
-  To provide training opportunities for developers; this should include
-  the support to the software schools for young scientists and computer
-  engineers, and of a permanent training infrastructure for accomplished
-  developers;
-\item
-  To provide career support for developers, for instance by listing job
-  opportunities and by helping to shape well-defined career paths that
-  provide advancement opportunities on a par with those in, for example,
-  detector construction;
-\item
-  To increase the visibility of the value of software developers in HEP,
-  recognising that it has scientific research value on an equal footing
-  with other activities, and acknowledging and promoting specific
-  ``champions'' in the field.
-\end{itemize}
-
-\subsection{Training Challenges}
-
-HEP is facing major challenges with its software and computing that
-require innovative solutions based on the proper adoption of new
-technologies. More and more technologies are emerging as scientific
-communities and industry face similar challenges and produce solutions
-relevant to us. Integrating such technologies in our software and
-computing infrastructure requires specialists, but it is also important that a
-large fraction of the community is able to use these new tools and
-paradigms. Specific solutions and optimisations must be implemented by
-the HEP community itself, since many advanced requirements are unique to
-our field.
-
-Unlike the situation that is traditional in some other fields, in
-which users express their requirements and computer specialists
-implement solutions, there is a close collaboration, even overlap,
-in HEP between users and developers that is essential for our success.
-Many details of experiment data
-cannot be known before data taking has started, and each change in
-detector technology or machine performance improvement can have
-important consequences for the software and computing infrastructure. In
-the case of detectors, engineers and physicists are required to have a
-good understanding of each other's field of expertise. In the same way,
-it is necessary that physicists understand some of the complexities of
-writing software, and that software experts are able to fathom the
-requirements of physics problems.
-
-Training must address an audience with very diverse computing skills,
-ranging from novice programmers to advanced developers and users.
-It must be used to spread best software engineering practices and
-software technologies to a very large number of people, including the
-physicists involved across the whole spectrum of data processing tasks,
-from triggering to analysis. It must be done by people who have a sound
-knowledge of the scientific and technical details, who prepare training
-material despite the many calls on their time. Training thus needs
-proper recognition to ensure that it happens and is carried out well.
-
-HEP is seen as an interesting, innovative, and challenging field. 
-This is a great advantage in attracting talented young people looking
-for experience in a challenging and diverse environment in which
-they can acquire skills that will be valuable, even in other fields.
-As discussed in Software Development (Section
-\ref{software-development-deployment-validation-and-verification}),
-using industry standard tools across different experiments, 
-and training people in how to use them properly,
-helps with people's later career prospects
-and makes our field even more attractive.
-At the same time, experiments have a scientific programme to accomplish
-and also to focus on the specific training required to
-accomplish their specific goals. The right balance must be found between
-these two requirements. It is necessary to find the right incentives to
-favour training activities that bring more benefits in the medium to
-long term, for the experiment, the community, and the careers of the
-trainees.
-
-\subsection{Possible Directions for Training}
-
-To increase training activities in the community, whilst taking into
-account the constraints of both the attendees and the trainers, we should
-explore new approaches to training. The current ``school''
-model is well established, as exemplified by three well-known successful
-schools, the CERN School of Computing~\cite{CSC}, 
-the Bertinoro School of Computing~\cite{ESC2017}
-and the GridKa School of Computing~\cite{GridKASchool}.
-They require a significant amount of
-dedicated time of all the participants, at the same time and location,
-and therefore are difficult to scale to meet the needs of a large number
-of students. In view of this, we should identify opportunities to work
-with HEP experiments and other training projects to provide accessible
-core skills training to the community by basing them at laboratories
-where students can easily travel. A number of highly successful
-experiment-specific examples exist, such as the LHCb StarterKit~\cite{LHCbStarterkit} 
-and ALICE Juniors~\cite{ALICEJuniors}, as well as
-established generic training initiatives, such as 
-Software Carpentry~\cite{SoftwareCarpentry}. As with hands-on tutorials organised during
-conferences and workshops, the resulting networking is an important and
-distinctive benefit of these events, where people build relationships
-with other colleagues and experts.
-
-In recent years, several R\&D projects, such as DIANA-HEP~\cite{DIANA-HEP} and
-AMVA4\-New\-Physics~\cite{AMVA4NewPhysics}, have had training as one 
-of their core activities. This
-has provided an incentive to organise training events and has resulted
-in the spread of expertise on advanced topics. We believe that training
-should become an integral part of future major R\&D projects.
-
-New pedagogical methods, such as active training and peer training, that
-are complementary to schools or topical tutorials, also deserve more
-attention. Online material can be shared by a student and a teacher to
-provide the exchange of real examples and practical exercises. For
-example, notebook technologies, such as Jupyter, support embedding of
-runnable code and comments into the same document. The initial material
-can be easily enriched by allowing other students and experts to add
-comments and more examples in a collaborative way. The HSF started to
-experiment with this approach with WikiToLearn~\cite{WikiToLearn}, a
-platform developed in Italy outside HEP that promotes this kind of
-training and collaborative enrichment of the training material. Projects
-such as ROOT~\cite{Brun1996} have also started to provide some
-training material based
-on notebooks.
-
-A lot of initiatives have been undertaken by the software community that
-HEP can benefit from, and materials have been made available in the form
-of online tutorials, active training, and Massive Open Online Courses
-(MOOCs). Some effort needs to be invested to evaluate existing courses
-and build a repository of selected ones that are appropriate to HEP
-needs. This is not a negligible task and would require some dedicated
-effort to reach the appropriate level of support. It should help to
-increase training efficiency by making it easier to identify appropriate
-courses or initiatives.
-
-A model that emerged in recent years as a very valuable means of sharing
-expertise is to use Question and Answer (Q\&A) systems, such as Stack
-Overflow. A few such systems are run by experiments for their own needs,
-but this is not necessarily optimal, as the value of these services is
-increased by a large number of contributors with diverse backgrounds.
-Running a cross-experiment Q\&A system has been discussed, but it has
-not yet been possible to converge on a viable approach, both technically
-and because of the effort required to run and support such a service.
-
-\subsection{Career Support and Recognition}
-
-Computer specialists in HEP are often physicists who have chosen to
-specialise in computing. This has always been the case and needs to
-continue. Nevertheless, for young people in particular, this leads to a
-career recognition problem, as software and computing activities are not
-well-recognised roles in various institutions supporting HEP research
-and recruiting people working in the field. The exact situation is
-highly dependent on policies and boundary conditions of the organisation
-or country, but recognition of physicists tends to be based generally on
-participation in data analysis or hardware developments. This is even a
-bigger problem if the person is spending time contributing to training
-efforts. This negatively impacts the future of these people and reduces
-the possibility of HEP engaging them in the training effort of the
-community when the community actually needs more people to participate
-in this activity. Recognition of training efforts, either by direct
-participation in training activities or by providing materials, is an
-important issue to address, complementary to the incentives mentioned
-above.
-
-There is no easy solution to this problem. Part of the difficulty is
-that organisations, and in particular the people inside them in charge
-of the candidate selections for new positions and promotions, need to
-adapt their expectations to these needs and to the importance of having
-computing experts with a strong physics background as permanent members
-of the community. Experts writing properly engineered and optimised
-software can significantly reduce resource consumption and increase
-physics reach, which provides huge financial value to modern HEP
-experiments. The actual path for improvements in career recognition, as
-the possible incentives for participating in the training efforts,
-depends on the local conditions.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Future challenges for High Energy Physics in the domain of software and
-computing are not simply an extrapolation of the challenges faced today.
-The needs of the HEP programme in the high luminosity era far exceed those
-that can be met by simply making incremental changes to today's code and
-scaling up computing facilities within the anticipated budget. At the
-same time, the limitation in single core CPU performance is making the
-landscape of computing hardware far more diverse and challenging to
-exploit, whilst offering huge performance boosts for suitable code.
-Exploiting parallelism and other new techniques, such as modern machine
-learning, offer great promise, but will require substantial work from
-the community to adapt to our problems. If there were any lingering
-notion that software or computing could be done cheaply by a few junior
-people for modern experimental programmes, it should now be thoroughly
-dispelled.
-
-We believe HEP Software and Computing requires a step change in its profile and
-effort to match the challenges ahead. We need investment in people who
-can understand the problems we face, the solutions employed today, and
-have the correct skills to provide innovative solutions for the future.
-There needs to be recognition from the whole community for the work done
-in this area, with a recognised career path for these experts. In
-addition, we will need to invest heavily in training for the whole
-software community as the contributions of the bulk of non-expert
-physicists are also vital for our success.
-
-We know that in any future scenario development effort will be
-constrained, so it is vital that successful R\&D projects provide
-sustainable software for the future. In many areas it is recognised 
-that different experiments could have adopted common solutions, 
-reducing overall development effort and increasing robustness and 
-functionality. That model of duplicated development is not sustainable. We must
-endeavour to achieve better coherence within HEP for future developments
-to build advanced, open-source projects that can be shared and supported
-in common. The HSF has already established itself as a forum that can
-facilitate this. Establishing links outside of HEP, to other academic
-disciplines, to industry, and to the computer science community, can
-strengthen both the research and production phases of new solutions. We
-should ensure that the best products are chosen, from inside and outside
-HEP, and that they receive support from all parties, aiming at technical
-excellence and economy of scale.
-
-We have presented programmes of work that the community has identified
-as being part of the roadmap for the future. While there is always some
-scope to reorient current effort in the field, we would highlight the
-following work programmes as being of the highest priority for
-investment to address the goals that were set in the introduction.
-
-\paragraph{Improvements in software efficiency, scalability and performance}
-
-\begin{quote}
-The bulk of CPU cycles consumed by experiments relate to the fundamental
-challenges of simulation and reconstruction. Thus, the work programmes
-in these areas, together with the frameworks that support them, are of
-critical importance. The sheer volumes of data involved make research
-into appropriate data formats and event content to reduce storage
-requirements vital. Optimisation of our distributed computing systems,
-including data and workload management, is paramount.
-\end{quote}
-
-\paragraph{Enable new approaches that can radically extend physics reach}
-
-\begin{quote}
-New techniques in simulation and reconstruction will be vital here.
-Phys\-ics analysis is an area where new ideas can be particularly
-fruitful. Exploring the full potential of machine learning is one common
-theme that underpins many new approaches and the community should
-endeavour to share knowledge widely across subdomains. New data analysis
-paradigms coming from the Big Data industry, based on innovative
-parallelised data processing on large computing farms, could transform
-data analysis.
-\end{quote}
-
-\paragraph{Ensure the long-term sustainability of the software}
-
-\begin{quote}
-Applying modern software development techniques to our codes has
-increased, and will continue to increase, developer productivity and
-code quality. There is ample scope for more common tools and common
-training to equip the community with the correct skills. Data
-Preservation makes sustainability an immediate goal of development and
-analysis and helps to reap the benefits of our experiments for decades
-to come. Support for common software used across the community needs to
-be recognised and accepted as a common task, borne by labs, institutes,
-experiments, and funding agencies.
-\end{quote}
-
-The R\&D actions proposed in this Roadmap have taken into account the
-charges that were laid down. When considering a specific project
-proposal addressing our computing challenges, that project's impact, measured
-against the charges, should be evaluated. Over the next decade, there
-will almost certainly be disruptive changes that cannot be planned for,
-and we must remain agile enough to adapt to these.
-
-The HEP community has many natural subdivisions, between different
-regional funding agencies, between universities and laboratories, and
-between different experiments. It was in an attempt to overcome these
-obstacles, and to encourage the community to work together in an
-efficient and effective way, that the HEP Software Foundation was
-established in 2014. This Community White Paper process has been
-possible only because of the success of that effort in bringing the
-community together. The need for more common developments in the future,
-as underlined here, reinforces the importance of the HSF as a common
-point of contact between all the parties involved, strengthening our
-community spirit and continuing to help share expertise and identify
-priorities. Even though this evolution will also require projects and
-experiments to define clear priorities about these common developments,
-we believe that the HSF, as a community effort, must be strongly
-supported as part of our roadmap to success.
-
-\newpage
-\begin{appendices}
-
-\hypertarget{appendix-a---list-of-workshops}{%
-\section{List of Workshops}\label{appendix-a---list-of-workshops}}
-
-\textbf{HEP Software Foundation Workshop}\\
-\emph{Date:} 23-26 Jan, 2017\\
-\emph{Location:} UCSD/SDSC (La Jolla, CA, USA)\\
-\emph{URL:}
-\href{http://indico.cern.ch/event/570249/}{{http://indico.cern.ch/event/570249/}}\\
-\emph{Description:} This HSF workshop at SDSC/UCSD was the first
-workshop supporting the CWP process. There were plenary sessions
-covering topics of general interest as well as parallel sessions for the
-many topical working groups in progress for the CWP.\\
-
-\noindent
-\textbf{Software Triggers and Event Reconstruction WG meeting}\\
-\emph{Date:} 9 Mar, 2017\\
-\emph{Location:} LAL-Orsay (Orsay, France)\\
-\emph{URL:}
-\href{https://indico.cern.ch/event/614111/}{{https://indico.cern.ch/event/614111/}}\\
-\emph{Description:} This was a meeting of the Software Triggers and Event
-Reconstruction CWP working group. It was held as a parallel session at
-the ``Connecting the Dots'' workshop, which focuses on forward-looking
-pattern recognition and machine learning algorithms for use in HEP.\\
-
-\noindent
-\textbf{IML Topical Machine Learning Workshop}\\
-\emph{Date:} 20-22 Mar, 2017\\
-\emph{Location:} CERN (Geneva, Switzerland)\\
-\emph{URL:}
-\href{https://indico.cern.ch/event/595059}{{https://indico.cern.ch/event/595059}}\\
-\emph{Description:} This was a meeting of the Machine Learning CWP
-working group. It was held as a parallel session at the
-``Inter-experimental Machine Learning (IML)'' workshop, an organisation
-formed in 2016 to facilitate communication regarding R\&D on ML
-applications in the LHC experiments.\\
-
-\noindent
-\textbf{Community White Paper Follow-up at FNAL}\\
-\emph{Date:} 23 Mar, 2017\\
-\emph{Location:} FNAL (Batavia, IL, USA)\\
-\emph{URL:}
-\href{https://indico.fnal.gov/conferenceDisplay.py?confId=14032}{{https://indico.fnal.gov/conferenceDisplay.py?confId=14032}}\\
-\emph{Description:} This one-day workshop was organised to engage with
-the experimental HEP community involved in computing and software for
-Intensity Frontier experiments at FNAL. Plans for the CWP were
-described, with discussion about commonalities between the HL-LHC
-challenges and the challenges of the FNAL neutrino and muon
-experiments\\
-
-\noindent
-\textbf{CWP Visualisation Workshop}\\
-\emph{Date:} 28-30 Mar, 2017\\
-\emph{Location:} CERN (Geneva, Switzerland)\\
-\emph{URL:}
-\href{https://indico.cern.ch/event/617054/}{{https://indico.cern.ch/event/617054/}}\\
-\emph{Description:} This workshop was organised by the Visualisation CWP
-working group. It explored the current landscape of HEP visualisation
-tools as well as visions for how these could evolve. There was
-participation both from HEP developers and industry.\\
-
-\noindent
-\textbf{DS@HEP 2017 (Data Science in High Energy Physics)}\\
-\emph{Date:} 8-12 May, 2017\\
-\emph{Location:} FNAL (Batava, IL, USA)\\
-\emph{URL:}
-\href{https://indico.fnal.gov/conferenceDisplay.py?confId=13497}{{https://indico.fnal.gov/conferenceDisplay.py?confId=13497}}\\
-\emph{Description:} This was a meeting of the Machine Learning CWP
-working group. It was held as a parallel session at the ``Data Science
-in High Energy Physics (DS@HEP)'' workshop, a workshop series begun in
-2015 to facilitate communication regarding R\&D on ML applications in
-HEP.\\
-
-\noindent
-\textbf{HEP Analysis Ecosystem Retreat}\\
-\emph{Date:} 22-24 May, 2017\\
-\emph{Location:} Amsterdam, the Netherlands\\
-\emph{URL:}
-\href{http://indico.cern.ch/event/613842/}{{http://indico.cern.ch/event/613842/}}\\
-\emph{Summary report:}
-\href{http://hepsoftwarefoundation.org/assets/AnalysisEcosystemReport20170804.pdf}{{http://cern.ch/go/mT8w}}\\
-\emph{Description:} This was a general workshop, organised about the
-HSF, about the ecosystem of analysis tools used in HEP and the ROOT
-software framework. The workshop focused both on the current status and
-the 5-10 year time scale covered by the CWP.\\
-
-\noindent
-\textbf{CWP Event Processing Frameworks Workshop}\\
-\emph{Date:} 5-6 Jun, 2017\\
-\emph{Location:} FNAL (Batavia, IL, USA)\\
-\emph{URL:}
-\href{https://indico.fnal.gov/conferenceDisplay.py?confId=14186}{{https://indico.fnal.gov/conferenceDisplay.py?confId=14186}}\\
-\emph{Description:} This was a workshop held by the Event Processing Frameworks
-CWP working group focused on writing an initial draft of the framework
-white paper. Representatives from most of the current practice
-frameworks participated.\\
-
-\noindent
-\textbf{HEP Software Foundation Workshop}\\
-\emph{Date:} 26-30 Jun, 2017\\
-\emph{Location:} LAPP (Annecy, France)\\
-\emph{URL:}
-\href{https://indico.cern.ch/event/613093/}{{https://indico.cern.ch/event/613093/}}\\
-\emph{Description:} This was the final general workshop for the CWP
-process. The CWP working groups came together to present their status
-and plans, and develop consensus on the organisation and context for the
-community roadmap. Plans were also made for the CWP writing phase that
-followed in the few months following this last workshop.
-
-\newpage
-\hypertarget{appendix-b-glossary}{%
-\section{Glossary}\label{appendix-b-glossary}}
-
-\begin{description}
-
-\item[AOD] Analysis Object Data is a summary of the reconstructed event and
-contains sufficient information for common physics analyses.
-
-\item[ALPGEN] An event generator designed for the generation of Standard Model
-processes in hadronic collisions, with emphasis on final states with
-large jet multiplicities. It is based on the exact LO evaluation of
-partonic matrix elements, as well as top quark and gauge boson decays
-with helicity correlations.
-
-\item[BSM] Physics beyond the Standard Model (BSM) refers to the theoretical
-developments needed to explain the deficiencies of the Standard Model
-(SM), such as the
-\href{https://en.wikipedia.org/wiki/Origin_of_mass}{origin of mass}, the
-\href{https://en.wikipedia.org/wiki/Strong_CP_problem}{strong CP
-problem},
-\href{https://en.wikipedia.org/wiki/Neutrino_oscillation}{neutrino
-oscillations},
-\href{https://en.wikipedia.org/wiki/Baryon_asymmetry}{matter--antimatter
-asymmetry}, and the nature of
-\href{https://en.wikipedia.org/wiki/Dark_matter}{dark matter} and
-\href{https://en.wikipedia.org/wiki/Dark_energy}{dark energy}.
-
-\item[Coin3D] A C++ object oriented retained mode 3D graphics API
-used to provide a higher layer of programming for OpenGL.
-
-\item[COOL] LHC Conditions Database Project, a subproject of the POOL persistency framework.
-
-Concurrency Forum: Software engineering is moving towards a paradigm shift in order to accommodate new CPU architectures with many cores, in which concurrency will play a more fundamental role in programming languages and libraries.
-The forum on concurrent programming models and frameworks aims to share knowledge among interested parties that work together to develop 'demonstrators' and agree on technology so that they can share code and compare results.
-
-\item[CRSG] Computing Resources Scrutiny Group, a WLCG committee in charge of scrutinizing and assessing LHC experiment yearly resource requests to prepare funding agency decisions.
-
-\item[CSIRT] Computer Security Incident Response Team.
-A CSIRT provides a reliable and trusted single point of contact for reporting computer security incidents and taking the appropriate measures in response tothem.
-
-\item[CVMFS] The CERN Virtual Machine File System is a network file system
-based on HTTP and optimised to deliver experiment software in a fast,
-scalable, and reliable way through sophisticated caching strategies.
-
-\item[CWP] The Community White Paper (this document) is the result of an
-organised effort to describe the community strategy and a roadmap for
-software and computing R\&D in HEP for the 2020s. This activity is
-organised under the umbrella of the HSF.
-
-\item[Deep Learning (DL)] one class of Machine Learning algorithms, based on a
-high number of neural network layers.
-
-\item[DNN] Deep Neural Network, class of neural networks with typically a
-large number of hidden layers through which data is processed.
-
-\item[DPHEP] The Data Preservation in HEP project is a collaboration for data
-preservation and long term analysis.
-
-\item[EGI] European Grid Initiative. A European organisation in charge of
-delivering advanced computing services to support scientists,
-multinational projects and research infrastructures, partially funded by
-the European Union. It is operating both a grid infrastructure (many
-WLCG sites in Europe are also EGI sites) and a federated cloud
-infrastructure. It is also responsible for security incident response
-for these infrastructures (CSIRT).
-
-\item[FAIR] The Facility for Antiproton and Ion Research (FAIR) is located at
-GSI Darmstadt. It is an international accelerator facility for research
-with antiprotons and ions.
-
-\item[FAIR] An abbreviation for a set of desirable data properties: Findable,
-Accessible, Interoperable, and Re-usable.
-
-\item[FCC] Future Circular Collider, a proposed new accelerator complex for
-CERN, presently under study.
-
-\item[FCC-hh] A 100 TeV proton-proton collider version of the FCC (the ``h''
-stands for ``hadron'').
-
-\item[GAN] Generative Adversarial Networks are a class of
-\href{https://en.wikipedia.org/wiki/Artificial_intelligence}{artificial
-intelligence} algorithms used in
-\href{https://en.wikipedia.org/wiki/Unsupervised_machine_learning}{unsupervised
-machine learning}, implemented by a system of two
-\href{https://en.wikipedia.org/wiki/Neural_network}{neural networks}
-contesting with each other in a
-\href{https://en.wikipedia.org/wiki/Zero-sum_game}{zero-sum game}
-framework.
-
-\item[Geant4] A toolkit for the simulation of the passage of
-particles through matter.
-
-\item[GeantV] An R\&D project that aims to fully exploit the
-parallelism, which is increasingly offered by the new generations of CPUs, in the field of detector simulation.
-
-\item[GPGPU] General-Purpose computing on Graphics Processing Units is the use of a Graphics Processing Unit (GPU), which typically handles computation only for computer graphics, to perform computation in applications traditionally handled by the Central Processing Unit (CPU).
-Programming for GPUs is typically more challenging, but can offer significant gains in arithmetic throughput.
-
-\item[HEPData] The Durham High Energy Physics Database is an open access
-repository for scattering data from experimental particle physics.
-
-\item[HERWIG] This is an event generator containing a wide range of Standard
-Model, Higgs and supersymmetric processes. It uses the parton-shower approach
-for initial- and final-state QCD radiation, including colour coherence effects and 
-azimuthal correlations both within and between jets.
-
-\item[HL-LHC] The High Luminosity Large Hadron Collider is a proposed upgrade to
-the Large Hadron Collider to be made in 2026. The upgrade aims at increasing the 
-luminosity of the machine by a factor of 10, up to 
-$10^{35}\mathrm{cm}^{-2}\mathrm{s}^{-1}$, 
-providing a better chance to see rare processes and improving
-statistically marginal measurements.
-
-\item[HLT] High Level Trigger. The computing resources, generally a large farm, close to the detector which process the events in real-time and select those who must be stored for further analysis.
-
-\item[HPC] High Performance Computing.
-
-\item[HS06] HEP-wide benchmark for measuring CPU performance based on the SPEC2006 benchmark
-(\href{https://www.spec.org}{{https://www.spec.org}}).
-
-\item[HSF] The HEP Software Foundation facilitates coordination and common
-efforts in high energy physics (HEP) software and computing
-internationally.
-
-\item[IML] The Inter-experimental LHC Machine Learning (IML) Working Group is
-focused on the development of modern state-of-the art machine learning
-methods, techniques and practices for high-energy physics problems.
-
-\item[IOV] Interval Of Validity, the period of time for which a specific piece
-of conditions data is valid.
-
-\item[JavaScript] A high-level, dynamic, weakly typed,
-prototype-based, multi-paradigm, and interpreted programming language.
-Alongside HTML and CSS, JavaScript is one of the three core technologies
-of World Wide Web content production.
-
-\item[Jupyter Notebook] This is a server-client application that allows
-editing and running notebook documents via a web browser. Notebooks are
-documents produced by the Jupyter Notebook App, which contain both
-computer code (e.g., python) and rich text elements (paragraph,
-equations, figures, links, etc...). Notebook documents are both
-human-readable documents containing the analysis description and the
-results (figures, tables, etc..) as well as executable documents which
-can be run to perform data analysis.
-
-\item[LHC] Large Hadron Collider, the main particle accelerator at CERN.
-
-\item[LHCONE] A set of network circuits, managed worldwide by the National
-Research and Education Networks, to provide dedicated transfer paths for
-LHC T1/T2/T3 sites on the standard academic and research physical
-network infrastructure.
-
-\item[LHCOPN] LHC Optical Private Network. It is the private physical and IP
-network that connects the Tier0 and the Tier1 sites of the WLCG.
-
-\item[MADEVENT] This is a multi-purpose tree-level event generator. It is
-powered by the matrix element event generator MADGRAPH, which generates
-the amplitudes for all relevant sub-processes and produces the mappings
-for the integration over the phase space.
-
-\item[Matplotlib] This is a Python 2D plotting library that provides
-publication quality figures in a variety of hardcopy formats and
-interactive environments across platforms.
-
-\item[ML] Machine learning is a field of computer science that gives computers
-the ability to learn without being explicitly programmed. It focuses on
-prediction making through the use of computers and emcompasses a lot of
-algorithm classes (boosted decision trees, neural networks\ldots{}).
-
-\item[MONARC] A model of large scale distributed computing based on many
-regional centers, with a focus on LHC experiments at CERN. As part of
-the MONARC project, a simulation framework was developed that provides a
-design and optimisation tool. The MONARC model has been the initial
-reference for building the WLCG infrastructure and to organise the data
-transfers around it.
-
-\item[OpenGL] Open Graphics Library is a cross-language, cross-platform
-application programming interface(API) for rendering 2D and 3D vector
-graphics. The API is typically used to interact with a graphics
-processing unit(GPU), to achieve hardware-accelerated rendering.
-
-\item[Openlab] CERN openlab is a public-private partnership that accelerates
-the development of cutting-edge solutions for the worldwide LHC
-community and wider scientific research.
-
-\item[P5] The Particle Physics Project Prioritization Panel is a scientific
-advisory panel tasked with recommending plans for U.S. investment in
-particle physics research over the next ten years.
-
-\item[PRNG] A PseudoRandom Number Generator is an algorithm for generating a
-sequence of numbers whose properties approximate the properties of
-sequences of random numbers.
-
-\item[PyROOT] A \href{http://www.python.org/}{Python} extension module that
-allows the user to interact with any ROOT class from the Python
-interpreter.
-
-\item[PYTHIA] A program for the generation of high-energy physics events, i.e.,
-for the description of collisions at high energies between elementary
-particles such as e+, e-, p and pbar in various combinations. It
-contains theory and models for a number of physics aspects, including
-hard and soft interactions, parton distributions, initial- and
-final-state parton showers, multiparton interactions, fragmentation and
-decay.
-
-\item[QCD] Quantum Chromodynamics, the theory describing the strong
-interaction between quarks and gluons.
-
-\item[REST] Representational State Transfer
-\href{https://en.wikipedia.org/wiki/Web_service}{web services} are a way
-of providing interoperability between computer systems on the Internet.
-One of its main features is stateless interactions between clients and
-servers (every interaction is totally independent of the others),
-allowing for very efficient caching.
-
-\item[ROOT] A modular scientific software framework widely used in HEP data
-processing applications.
-
-\item[SAML] Security Assertion Markup Language. It is an open, XML-based,
-standard for exchanging authentication and authorisation data between
-parties, in particular, between an identity provider and a service
-provider.
-
-\item[SDN] Software-defined networking is an umbrella term encompassing several kinds of network technology aimed at making the network as agile
-and flexible as the virtualised server and storage infrastructure of the modern data center.
-
-\item[SHERPA] Sherpa is a Monte Carlo event generator for the Simulation of
-High-Energy Reactions of PArticles in lepton-lepton, lepton-photon,
-photon-photon, lepton-hadron and hadron-hadron collisions.
-
-\item[SIMD] Single instruction, multiple data (\textbf{SIMD}), describes
-computers with multiple processing elements that perform the same
-operation on multiple data points simultaneously.
-
-\item[SM] The Standard Model is the name given in the 1970s to a theory of
-fundamental particles and how they interact. It is the currently
-dominant theory explaining the elementary particles and their dynamics.
-
-\item[SWAN] Service for Web based ANalysis is a platform for interactive data
-mining in the CERN cloud using the Jupyter notebook interface.
-
-\item[TBB] Intel Threading Building Blocks is a widely used C++ template
-library for task parallelism. It lets you easily write parallel C++
-programs that take full advantage of multicore performance.
-
-\item[TMVA] The Toolkit for Multivariate Data Analysis with ROOT is a
-standalone project that provides a ROOT-integrated machine learning
-environment for the processing and parallel evaluation of sophisticated
-multivariate classification techniques.
-
-\item[VecGeom] The vectorised geometry library for particle-detector simulation.
-
-\item[VO] Virtual Organisation. A group of users sharing a common interest
-(for example, each LHC experiment is a VO), centrally managed, and used
-in particular as the basis for authorisations in the WLCG
-infrastructure.
-
-\item[WebGL] The Web Graphics Library is a JavaScript API for rendering
-interactive 2D and 3D graphics within any compatible web browser without
-the use of plug-ins.
-
-\item[WLCG] The Worldwide LHC Computing Grid project is a global collaboration
-of more than 170 computing centres in 42 countries, linking up national
-and international grid infrastructures. The mission of the WLCG project
-is to provide global computing resources to store, distribute and
-analyse data generated by the Large Hadron Collider (LHC) at CERN.
-
-\item[X.509] A cryptographic standard which defines how to implement service
-security using electronic certificates, based on the use of a private
-and public key combination. It is widely used on web servers accessed
-using the https protocol and is the main authentication mechanism on the
-WLCG infrastructure.
-
-\item[x86\_64] 64-bit version of the x86 instruction set.
-
-\item[XRootD] Software framework that is a fully generic suite for fast, low latency and scalable data access.
-
-\end{description}
-
-\end{appendices}
-
-\sloppy
-\raggedright
-\clearpage
-\printbibliography[title={References},heading=bibintoc]
-
-\clearpage
-\section*{The HEP Software Foundation}
-\input{authors.tex}
-
-\end{document}

### 2018-05-22 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Break apart LaTeX for arXiv style
Separate the main text body (cwp-main.tex) from the LaTeX for
the specific document class used for sumbission to arXiv in
JHEP style

--- 2018-01-16 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp.tex
@@
+% This is the main body text for the HSF Community White Paper
+
 % Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.
-
-% JHEP preprint template
-\documentclass[12pt,a4paper]{article}
-\usepackage{jheppub}
-
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage[titletoc]{appendix}
-\usepackage{xspace}
-
-\usepackage[utf8]{inputenc}
-
-\usepackage[style=numeric-comp,sorting=none]{biblatex}
-\addbibresource{cwp.bib}
-\addbibresource{cwp-chapters.bib}
-
-%% Comment out before final submission
-%\usepackage{lineno}  % for line numbering during review
-%\linenumbers
-
-\abstract{Particle physics has an ambitious and broad experimental 
-programme for the coming decades. This programme requires large 
-investments in detector hardware, either to build new facilities 
-and experiments, or to upgrade existing ones.
-Similarly, it requires commensurate investment in the R\&D of software to acquire, manage, process, and analyse the shear amounts of data to be recorded.
-In planning for the HL-LHC in particular, it is critical that all of the
-collaborating stakeholders agree on the software goals and priorities,
-and that the efforts complement each other. In this spirit, this white paper
-describes the R\&D activities required to prepare for this software upgrade.}
-
-\begin{document}
-
-\noindent
-\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
- & & HSF-CWP-2017-01 \\
- & & December 15, 2017 \\ % use \date or hardwire e.g. December 15, 2017
- & & \\
-\end{tabular*}
-\vspace{2.0cm}
-
-\title{A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s}
-
-\author{HEP Software Foundation\footnote{Authors are listed at the end of this report.}}
-
-\maketitle
-
-\newpage
 
@@
     \begin{subfigure}[b]{0.47\textwidth}
-        \includegraphics[width=\textwidth]{cms-cpu.png}
+        \includegraphics[width=\textwidth]{cms-cpuHLLHC.pdf}
         \caption{}
@@
     \begin{subfigure}[b]{0.47\textwidth}
-        \includegraphics[width=\textwidth]{cms-disk.png}
+        \includegraphics[width=\textwidth]{cms-diskHLLHC.pdf}
         \caption{}
@@
 \end{appendices}
-
-\sloppy
-\raggedright
-\clearpage
-\printbibliography[title={References},heading=bibintoc]
-
-\clearpage
-\section*{The HEP Software Foundation}
-\input{authors.tex}
-
-\end{document}

### 2018-05-31 HSF-TN/draft-2015-NAM/draft-HSF-TN-2015-NAM.tex
update naming conventions w/ feedback from Graeme

@@
   \noindent
-  The note describes a proposal for a common platform naming scheme for HEP and tools to automatize the platform identification. 
+  The note describes a proposal for a common platform naming scheme for HEP and tools to automate the platform identification.
 \end{abstract}
@@
 
-In the view of the fact, that naming conventions are just agreements we propose to adopt existing naming conventions wherever possible and useful.
+In the view of the fact that naming conventions are just agreements we propose to adopt existing naming conventions wherever possible and useful.
 
@@
 
-Different OSes may call the same \emph{architecture} with different names.
+Different OSes may call the same \emph{architecture} by different names.
 We propose to rely on the abstraction of the Python package \emph{platform}, namely \texttt{platform.machine()}.
-This translates to e.g.
-x86\_64 for current Intel and AMD CPUs.
-In cases where optimized builds for a certain processor generation are required and software is not runable on the generic architecture, the generic architecture should be replaced by \emph{architecture+instructionset}.
-There exist no stable conventions for the instruction set, thus we propose to create ad-hoc conventions.
+This translates to, e.g., x86\_64 for current Intel and AMD CPUs.
+In cases where optimized builds for a certain processor generation are required and software is not runable on the generic architecture, the generic architecture should be replaced by \emph{architecture+instructionset1+instructionset2...}.
+For example \emph{x86\_64+avx2} denotes an x86\_64 processor supporting the AVX2 instruction set.
+There exist no explicit conventions for the instruction set across operating systems, thus we propose to create ad-hoc conventions.
+In the Linux case the \emph{flags} field of \emph{/proc/cpuinfo} could be used.
+The instruction sets mentioned should be ordered alphabetically to ensure consistency.
 
@@
 The \emph{compiler} is a combination of compiler name and compiler version.
-We propose to use the self-given names of the compilers like \emph{gcc, clang, msvc, icc} and adding major, minor version, and patch version numbers to it, e.g.
-\emph{clang350} or \emph{gcc511}.
+We propose to use the self-given names of the compilers like \emph{gcc, clang, msvc, icc} and adding the compiler version indicating feature relevant versions.
+That can be major, minor version, and patch version numbers for LLVM (e.g.
+\emph{clang350}) or just major and minor version for GCC (e.g.
+\emph{gcc73}).
 In case the system compiler is being used, the compiler should be denoted as \emph{native}.
@@
 The \emph{buildtype} denotes whether there is a debug build, an optimized build, or any other special setting.
-We propose to use \emph{opt} for optimized and \emph{dbg} for debug builds.
+We propose to use \emph{opt} for `optimized` and \emph{dbg} for debugging builds.
 The buildtype can be used to specify further custom build options, e.g.
-the used C++ standard, leading to opt11 or dbg14.
+the used C++ standard, leading to \emph{opt+std14} or \emph{dbg+std18}.
 In general, any special compile time setting can be tracked in this name component.
@@
 As to simplify the adoption of these conventions, we provide a minimal command line tool \emph{hsf\_get\_platform}, that is able to identify architecture, operating system, and compiler in a transparent way.
-It can be used to e.g.
-auto-derive everything in a given environment or to extract one part of the platform string
+It can be used to, e.g., auto-derive everything in a given environment or to extract one part of the platform string
 \\

### 2018-06-11 CWP/papers/HSF-CWP-2017-02_training/latex/preamble.tex
Training CWP Paper - Import from sharelatex
This is a verbatim import of the document sources from
sharedlatex. Note this version does not compile properly,
which will be fixed in the next commits.

\usepackage[utf8]{inputenc}
\usepackage[USenglish]{babel}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{verbatim}

\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-06 \\
 & & June 6, 2018 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{HEP Software Foundation Community White Paper Working Group --
Training, Staffing and Careers}

%\author{See page~\pageref{sec:author-list} for the full author list}
\date{March 2018}

\input{authors.tex}

\abstract{The rapidly increasing evolution of technology and the parallel increasing complexity of algorithmic analysis in HEP (and in science in general) require a much larger portfolio of specific skills by the developers with respect to the past. Young researchers, graduating from universities worldwide, currently do not receive adequate preparation in the very diverse fields of modern computing to respond to growing necessities posed by the most advanced experimental challenges. The vast majority of the HEP community strongly agrees on the need for an overarching umbrella of training programs to bring young researchers (but also more mature ones) up to date with new software technologies, such as, for example, concurrent programming and artificial intelligence. }

\usepackage{graphicx}
%\usepackage[a4paper,margin=2cm]{geometry}% A4 for a US project?
\usepackage{lineno}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,        % false: boxed links; true: colored links
  linkcolor=black,        % color of internal links
  citecolor=green,        % color of links to bibliography
  filecolor=magenta,      % color of file links
  urlcolor=blue           % color of external links
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{amssymb}               % http://ctan.org/pkg/amssymb
\usepackage{pifont}                % http://ctan.org/pkg/pifont
\providecommand{\xmark}{\ding{55}} % https://tex.stackexchange.com/questions/42619/x-mark-to-match-checkmark

%N.B.: If you are going to be defining your own commands please use \providecommand instead of \newcommand. c.f.: https://tex.stackexchange.com/questions/36175/what-do-newcommand-renewcommand-and-providecommand-do-and-how-do-they-differ - Matthew

% https://tex.stackexchange.com/questions/25701/bibtex-vs-biber-and-biblatex-vs-natbib
\usepackage[backend=biber,defernumbers=true,sorting=none,firstinits=true,natbib=true,style=numeric]{biblatex}
\addbibresource{bib/references.bib}

### 2018-06-11 CWP/papers/HSF-CWP-2017-02_training/latex/src/training.tex
Training CWP Paper - Import from sharelatex
This is a verbatim import of the document sources from
sharedlatex. Note this version does not compile properly,
which will be fixed in the next commits.

% ---------------------------------------------------
% CHAPTER SUMMARY MESSAGE: Motivate the need for training in the HEP community,
%  summarize existing training efforts to motivate that we already spend funding on
% this. Discuss initiatives and programs which can be developed in the coming years
% including a discussion of the known training needs of the community. 
%
%    Goals for this chapter 
%   - Motivation to dedicate resources to training
%   - The community does not recognize the importance
%   - people learn on the job
%   - are there objective criteria
%   - retaining people with skills? 
%
%   see e.g. https://github.com/particleist/inventingcarrots/blob/master/inventing_carrots_merged.md which tries to address the problem of "how do we incentivize affirm and reward learning about software in HEP"
%
%   - People should have incentives to 
%
% ---------------------------------------------------

%Contributors: Fernanda Psihas, Justin Vasel, Michela Paganini, Douglas Davis, Savannah
%Thais, Martin Vala, Jonas Eschle, Xavier Vilasís-Cardona, Jamal Rorie, Sean-Jiun Wang,
%Liz Sexton-Kennedy, Sergei Gleyzer, Laurent Basara, Kurt Rinnert, Bob Stienen, Matt Bellis, Meghan Kane (Industry - SoundCloud), Dario Menasce, Pat Skubic, Hadrien Grasland, Riccardo Maria Bianchi, Daniel S. Katz, Sudhir Malik, Albert Puig

%____________________________________________________________________
%____________________________________________________________________
\section{Introduction}

Throughout the last few decades, the field of high energy physics (HEP) has consistently either driven or utilized the latest innovations in computational tools and technologies for problem solving. 
The toolset of the particle physicist is vast and ever-growing, and the problem set increasingly complex.
A typical graduate student or postdoc working on HEP experiments will likely encounter a number of tools, software, and programming languages with which they are totally unfamiliar. 
This requires ``on-the-job'' training for the same people whose work drives the reach of our analyses, while they are developing the tools for these analyses.

A strong research community should consist of independent actors who ask research questions and direct their own analyses, instead of blindly following prescriptions and recipes. It is therefore imperative that the community be proactive about providing training resources and opportunities for the good of the community, and that we follow through on the promise that adequate career path possibilities exist for people in HEP who would otherwise leave the field due to limited advancement opportunities.

To produce the best possible science, it is important, for the broad HEP community, to invest in training physicists in essential software and software engineering skills, which exist at many different levels. The level of depth of software engineering competences in which a physicist should be trained will vary depending on the problems that they face, and the appropriate level may evolve over time according to the problem at hand. For example, for a researcher who only performs basic data analysis, it may be sufficient to be familiar with just Python and data analysis libraries. However, contributing to a large open source project, such as ROOT, requires greater training in software development and related practices. These higher-order skills may include an understanding of writing testable code, time and space complexity implications, and knowledge of common software architecture paradigms. The need for these skills cannot be fulfilled completely just by hiring specialists in computer science, since a thorough understanding of the physics problems is usually very important to the success of the effort, making it necessary to provide a career path for those young physicists willing to steer their interests in that particular direction.

In each HSF workshop, a discussion on the need for proper tutoring, training, and the relevant career possibilities in HEP computing emerged and was felt to be crucial and essential by almost all participants. There is a strong consensus that future software developments will benefit enormously from a successful implementation of such training programs, provided proper career prospects are put in place, as an incentive, to would-be trainers.

This document tries to address the most relevant questions related to the training in the HEP community, namely:
\begin{itemize}
    \item How to encourage and provide incentives to the members of the HEP community to train students and/or other collaborators?
    \item How to properly assign credit to software development (and training thereof) activities?
    \item What are the current practices in the field regarding hiring of researchers with a strong specialisation in computing and the policies to retain such an expertise in the long term? Is there room for improvement?
    \item How to address and bridge the gap in formal software training that is not always imparted by universities?
\end{itemize}

%____________________________________________________________________
%____________________________________________________________________
\section{Motivation}

%    DOES THIS SECTION ANSWER THE FOLLOWING: 
%    - How is training different in software? Specify that there is /no formal training for these tools?

The introduction and development of increasingly complex tools has led to increasing improvements in the performance of our analyses. However, the personnel who drive these advancements through novel implementations of the technology are the analyzers. In this context, the term {\it analyzers} includes all those researchers who participate in an HEP experiment whose contribution includes designing and implementing a high speed DAQ system, setting up a modern database to store the state of a detector, complex task of managing the data flow from the first trigger level to the final n-tuple dataset and creating sophisticated statistical and mathematical tools to extract publishable physical meaning from those same data.

The development of new tools is insufficient to drive advances in the field unless researchers who have the expertise apply them to physics questions. Furthermore, a larger and a better-connected community, with expertise in the tools and access to the new developments, can lead to an open-source model in HEP. The efforts to train the community in all relevant tools is, thus, a priority within the R\&D effort itself.

However, as tools become more complex, the learning rate of developers must keep up if we are to utilize them effectively.  At the same time, as in other branches of research, it is essential to sustain efforts in maintaining reproducibility of the usage of the tools. This again highlights the need for physicts to be appropriately trained in computing science, rather than professional computing scientists, whose goals and interests may not necessarily be aligned or focused on the physics results of a particular HEP experiment.

To avoid duplicate efforts within the community, it is important to find a way to approach training efforts holistically and cross-experimentally, whenever possible. This requires a certain level of organization and coordination in order to guarantee the delivery of efficient, non-redundant software to different experiments and to allow the different communities to agree on common standards and procedures.

Beyond the immediate benefit to the scientific community in having well-trained collaborators, most people who start a physics graduate program will have careers outside of academia. This trend is reflected in several calls for proposals from national scientific funding agencies like H2020 and those from the National Science Foundation (NSF) that promote training in computer infrastructure and information areas.

%\textcolor{red}{this call is more to develop cyberinfrastructure staff than scientists - for example, people at computing centers}

\begin{quote}
\it The overarching goal of these programs is to prepare, nurture and grow the national scientific workforce for creating, utilizing, and supporting advanced cyberinfrastructure (CI) that enables cutting-edge science and engineering and contributes to the Nation's overall economic competiveness and security.

{\bf - Training-based Workforce Development for Advanced Cyberinfrastructure (CyberTraining) Webinar}\footnote{ \url{https://www.nsf.gov/events/event_summ.jsp?cntn_id=190179&org=NSF}}
\end{quote}

\begin{quote}
\it New information, communication, and computational technologies have had profound impacts on the practice of science (in this solicitation, the term science includes the natural, mathematical, computing, and social sciences),  engineering, and education. This includes the means by which citizens of all ages use science  and engineering to enhance professional and private lives.  The systems, tools, and services emerging from these new technologies are linked to create a comprehensive cyberinfrastructure that is enabling individuals, groups, and organizations to advance research and education in ways that revolutionize who can participate, what they can do, and how they do it. Sustaining this revolution across all areas of science, engineering, and education requires the formation of a citizenry and workforce with the knowledge and skills needed to design and deploy as well as adopt and apply these cyber-based systems, tools and services over the long-term. The opportunity for such preparation should be available at all stages of formal and informal education (K-16 and lifelong), training and professional development, and must be extended to all individuals and communities.

{\bf - Cyberinfrastructure Training, Education, Advancement, and Mentoring for Our 21st Century Workforce (CI-TEAM)}\footnote{ \url{https://www.nsf.gov/funding/pgm_summ.jsp?pims_id=12782}}
\end{quote}

An open issue is the kind of incentive that our community can collectively leverage in order to increase the number of people (with strong expertise in computing) who decide to dedicate time and effort to train fellow members of their community. 
To be effective, the training must be done by people who work at the cutting edge of the technology, who in many cases are found among the youngest researchers. These are also the very same people who are most in need of officially recognized credits for the advancement of their careers. The current practice in HEP is that these young scientists only receive credit and recognition (in the vast majority of cases) by working on a data analysis project, and this discourages them from finding time to transfer their acquired know-how to others in an effective way. A possible solution to this conundrum could be the establishment of specific career paths for researchers with a computer science specialization. The practical implementation of such an approach, however, is extremely challenging as there are a wide range of institutions involved that belong to different countries with their own policies, priorities, and funding strategies. Broadly, two career paths could be envisaged. Both would recognise that domain-sepcific knowledge and continued engagement is vital to the success of HEP computing and software. While in an ideal world (for physicists, at least), a physicst could make a detailed plan of what they need (expressed in a tidy requirements document) and a computer scientist could use this document to provide the required solution, with very little additional exchange of information during the process (which is how the software development process sometimes works in industry). In the real HEP world, instead, most of the time the requirements are bound to change rather frequently for a variety of reasons (not only because of poor planning) forcing developers and physicsts to closely interact during the entire process of software development; this works smoothly only when both communities have the same goals and speak the same jargon, hence the need for physicists with a computer science specialisation.

The first possible career would be the path of a \emph{physicist researcher with computing science specialisation}, in some places corresponding to the role of a physicist-programmer. It would be conceptually different to that of a \emph{computer scientist}: the two have different goals, seek different paths to the solution of their problems, and usually do not even adopt the same jargon in expressing problems and solutions. Such a role would have an active role in physics analysis, and so meet many of the traditional metrics for an academic career path.

The second would be the path of a domain-specific \emph{researcher software engineer}. Such a person has the primary role of developing software and finding software solutions, but with a large domain-specific knowledge to both understand the use cases and clients, but also of the available solutions.  Such posts are almost impossible to develop in the current academic frameworks; but while RSEs are understood by many funding agencies and universities as a technical post, they are seen as available for short-term engagements with a researcher community, delivering a quick fix or a refactoring and then moving one. This does not deliver the domain specific expertise required.

Both career paths suffer in academic terms from few or poorly cited publications, few opportunities to win grants in their own right and little opportunity for impact with commerce and wider society. These three failings make an academic appointment challenging or disfavoured compared to other applicants. However, there are measures that an be taken to address all three weaknesses. 
%This makes the development of specific solutions even more difficult and cumbersome.

%Finally, the career of these two figures follows slightly different paths, with different requirements. 
%____________________________________________________________________
%\subsection{Existing Training efforts  \textcolor{pink}{is this needed?....}}
% why not, shows what is done already

%____________________________________________________________________
%____________________________________________________________________
\section{Training needs of the community}

% DOES THIS SECTION ANSWER THE FOLLOWING: 
% - Different levels of training are required

The HEP user community consists of a set of users with diverse software experience, interests, and time availablity to learn new techniques.  Any training program must take into account the variation in target audiences. For example, an undergraduate doing a summer research project has different needs and skills than their mentoring professor.
For purposes of training, it is useful to think of all our user community as \emph{students}. More specifically, we can broadly identify four different sets of students, classified by experience level, recognizing that even within these groups there will be people who will learn at different speeds and with differing depths of understanding:

\begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
   \item \emph{Beginners}: New collaborators with no knowledge of the tools or techniques they are expected to use. These might be undergraduates but they could also be senior researchers who have not yet or not recently engaged in software analysis. These are people in need of some kind of formal training in modern computing techniques (basically compiled and scripting languages together with operating system basics such as filesystems, version control, and command-line shells).
   \item \emph{Intermediate users}: HEP analyzers with some experience in either analysis concepts and/or tools, but  looking to supplement their experiences with more recent and modern approaches. 
   \item \emph{Advanced users}: HEP analysis experts who have mastered current technologies and implementations (at a certain point in time) and want to stay up-to-date with new advanced developments.
   \item \emph{Software developers}: HEP scientists in charge of software development in areas not limited only to analysis, such as DAQ systems, computing infrastructure, databases, pattern recongition, and complete frameworks.
\end{itemize}

Each of these groups has vastly different training needs. However, \emph{whenever possible}, any training program should take advantage of developments in pedagogy, such as active learning\footnote{\url{http://www.crlt.umich.edu/tstrategies/tsal}}, peer learning\footnote{\url{https://en.wikipedia.org/wiki/Peer_learning}}, and web-based training.\footnote{See \S\ref{sec:initiatives}.} In some cases, it may even be advantageous to hand out code samples that are purposely broken or flawed, and ask students to fix or improve them. Learning the material in a way that sticks with the students is difficult and challenging for both the students and the instructor and often takes more time than we would prefer. However, this is the best way to ensure an educated community of scientists who can fully contribute to the physics programs at large, which is really the ultimate goal of any training program.

%____________________________________________________________________
\subsection{Knowledge that needs to be transferred}

At all stages of software and computing training, we should take care to encourage \emph{Good Practices Across the Community} (GPAC), such as error checking, modularity of code design, version control, writing tests, etc. All the key concepts addressed in the training should not be specific to a particular experiment or field of application, but general enough to be useful to the whole HEP community and possibly beyond. In this section, we present a list of specific concepts that need to be taught to members of the community, in order to guarantee the base level of competence needed to write efficient code for the different tasks performed in HEP experiments.

An essential knowledge base that needs to be transferred includes basic programming concepts, data structures, basics of code design, error checking, code management tools, validation and debugging tools. More advanced topics include modularity of code design, advanced data structures, evaluation metrics, writing tests and working with different types of hardware accelerators. Cutting-edge approaches such as the use of hardware accelerators (whether GPUs or FPGAs), Deep Neural Networks, and Machine Learning should be considered important training topics. Additionally, special emphasis should be made on reporting results and documenting them.
The training subjects that the community considers important to pursue are listed as follows:

\begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
   \item Basic and Advanced Programming Concepts
   \begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
      \item Object oriented paradigm
      \item Compiled langauges (C++)
      \item Scripting languages (Python, Javascript,...)
   \end{itemize}
   \item Algorithmics 
   \begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
      \item Boost library
      \item STL algorithms for containers
      \item R and/or ROOT
   \end{itemize}
   \item Existing frameworks (development or application level)
   \begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
      \item Qt
      \item ROOT
      \item experiment specific framework (possibily if of potential interest outside the native experiment)
   \end{itemize}
   \item Code design (design patterns)
   \item Development tools
   \begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
      \item IDEs (Integrated Development Environment)
      \item Debuggers
      \item Profilers
   \end{itemize}
   \item Evaluation metrics
   \item “Trust” metrics such as data driven tests
   \item Specific software implementation training
   \item Good practices 
   \item Code style and clarity
   \item Scripting and data cleaning
   \item Reporting results reproducibly
   \item Writing Documentation
\end{itemize}

%____________________________________________________________________
%____________________________________________________________________
\section{Challenges}

Within a single experiment, different skill sets are needed. In addition to a skill set that contains basic programming language knowledge, testing and code management tools, and experiment-specific framework knowledge, there are more specialized skills that only a subset of the community needs to know, such as software optimization, or low-level hardware interfaces.
While there already are experiment-specific training efforts in place, there are many needs that are in common between experiments. Therefore, we should probably strive to extract the common knowledge and build a common training program. This would reduce duplication of efforts and enable growth of a shared training culture by accumulating and sharing expertise across different experiments.

To realise this goal, a possibility could be the institution of a {\em federation} of existing training initiatives, a very lightwheight umbrella just aimed at improving the efficiency and cost-effectivness  
of this important activity.

\section{Implementation Roadmap}
The implementation of training should target multiple training formats and knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and advanced visualizations, etc.
People who learn in non-conventional ways should not be excluded from the opportunity to learn. Since there are different stages of knowledge transfer, programs should be developed that encourage experts to share knowledge. Issues considered in this section include global training initiatives, strategies for leveraging existing forums, and training tool development.

An important point to consider is the difficulty, often experienced in the past, in developing large software programs across different experiment communities. It is always challenging to organize a Working Group for training across experiments or even across diverse communities (such as HEP and Astroparticle, for example). This is a point where an appropriate incentive program to reward those who train the community might help.

%____________________________________________________________________
\subsection{Initiatives for future training programs\label{sec:initiatives}}

Some methods that can be used for global training include Massive Online Open Courses, hands-on workshops, online knowledge bases, expert trainer volunteer networks, and web-based training approaches.

Massive Online Open Courses (MOOCs) can be used to develop an open-source set of tutorials and tools. Existing online courses such as Udacity and Coursera can be evaluated and recommended to the community. A novel approach such as WikiToLearn could also explored to assess potential benefits (as has already been attempted in the context of the GridKa School of Computing, see below).

Experiment-specific and global knowledge bases can be established with incentives for experts to contribute. They can be open-source so that a lot of knowledge can be added by the trainees themselves as information is learned (this is the particular context where an approach such as WikiToLearn could be of great help).

Question-and-answer websites such as Stack Overflow for HEP are also a very useful resource for common problems and questions. This approach has already been considered by HSF start-up members, and turned out to be difficult to pursue for a number of reasons, but in the future, boundary conditions might change, making this approach viable.

Hands-on workshops are an invaluable part of learning how to apply theoretical concepts in practice. Identifying which of the current workshops are productive and useful, and if they cover all the topics that need to be transferred and that are in demand by the students, is an important action item.

Creating an expert tutor volunteer network is another way to provide training and support to the community. 
This of course requires proper recognition, at least in terms of career prospects as an incentive for young researchers. It is clear that the best possible tutors are, in principle, those people who are actively engaged in modern software developments: these are in general young researchers in the first steps of their careers and, in order to be attracted to devoting substantial time to training and tutoring, they must be assured that such an effort would be properly recognised in an official way. Such kind of recognition is not currently implemented, at least not in a standardized or official way. A possible structure for such a network could be the establishment of a {\em federation} of existing schools.

It should be considered that some professors, who act as graduate student advisors, might need to be encouraged to make sure their students are properly trained. Sometimes, students are instead pushed to learn the bare minimum to get the work done, at the expense of a broader training/education curriculum that could actually yield improved results, but further down the line. One incentive would be to provide training programs that also count as course credit, perhaps as an elective. This model is already in limited use with some online solutions\footnote{\url{https://www.edx.org/credit}}, but this is not universal. Discussions should be started with collaborators at higher education institutions to see what the roadblocks or opportunities would be for these training sessions to serve double duty. 
It should also be considered that not all students will end their career in research or academia: their contribution to the research activity, as students, should therefore also provide them knowledge, know-how, and skills considered a valuable asset by industry, in order to increase their chances of a career outside research.

Difficulties that have emerged in the past with respect to implementing training courses are the lack of funding and the lack of available time by experts in the field. People with enough expertise or insight in the field usually don't have time to devote to prolonged periods of student training, and, even when they can find time, the cost of setting up a training course in an effective way is often beyond what's made available by funding agencies (funds for travel, hosting, setting up a room with a computing infrastructure to allow interactive hands-on session, etc.) A possible solution is a completely different approach to training (but complementary to the already existing and successful classical efforts such as the CERN School of Computing's Bertinoro and Kit ones): instead of directly teaching to students, trainees could make use of a web-based platform to provide training materials to students. This complementary approach has several advantages over traditional ones:
\begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
   \item Tutors can add material to the web site at a very slow pace (whenever they find time to do it, one slide a week or a chapter a day)
   \item Their material, made public on a web-site, can be further expanded by collaborators (also at their own pace) or, even better and more productively, by students who decide to contribute new additions, examples, exercises etc. Such a collaborative effort allows more people to be exposed to training at any given time, creates a sense of community, and creates bridges between people in contiguous areas or research. Students can use the same platform to exchange their own examples, make suggestions and point out interesting concepts. In such a model, the possible contribution from others to the training material needs to be moderated and validated with appropriate policies.
   \item If complemented by the availability of remote virtual machines (possibly through a browser), students could have access to examples and exercises that are already embedded in their own natural environment: all the necessary tools/libraries needed to implement the exercise will be already available in the virtual machine (possibly a Docker container). With just a web browser, students could run complex examples from home, taking advantage of a remote facility that provides some storage and computing power. Important here is the concept of ``environment'': a Docker container could be set up in such a way that students will work in an exact replica of the environment they will be exposed to in their experiment (or collaboration at large). Moreover, students could be provided with Docker containers that preserves their modified environment across sessions, allowing them to develop their skills over a prolonged period of time by accessing all the files that were made persistent day by day during the training.
   \item There would no longer be a need to find the resources to host a school and pay the tutor(s) (and eventually subsidize the students to participate in training in a remote location). Students could follow the training at their own pace from wherever they happen to be. A traditional school only lasts for 5 days (10 at most) and it is difficult to cover a  subject to any significant depth in such a short time. The web approach, instead, would allow for very long and in-depth coverage of any kind of subject, and in this sense it could be a {\it complementary} approach to a traditional school. Of particular interest could be courses such as, e.g., ``Machine Learning'', ``Statistical Analyis with ROOT'', or even just ``Good practices in C++'' or ``Python Programming for scientific computing''. 
   \item Finally, this approach could allow the creation of {\it browsable} repositories of all training materials, grouped by argument, by relevance, by experiment or whatever other critieria. Students from all over the world could be exposed to a large repository of examples, exercises and in general training material from their own home.
\end{itemize}

An example of a such a web-based platform already exists, and has been implemented as an Open Source project (backed by Wikimedia) by a group of more than 30 Italian students. The project is WikiToLearn \footnote{\url{https://it.wikitolearn.org/}}. It hosts training material in several languages, for several disciplines, ranging from Economics to Physics, Mathematics, and several others. Because it is based upon wikimedia\footnote{\url{https://www.wikimedia.org/}} software, it is very easy to add material to the site, and to make it appear under a specific topic (such as, e.g., Software/Techniques/Machine-Learning) and to manipulate it as if it were a single document. In the end, students can selectively choose individual chapters from the site and have the corresponding pdf sent them as a book, complete with index, content, and chapters.

The adoption of such an approach is made rather easy in WikiToLearn by the relative simplicity of the wikimedia-based toolset: users contribute their training material using just a web-browser, and in order to do this efficiently, the learning curve has been kept appropriately shallow.

An interesting exercise in this context has recently been made by colleagues of the GridKa School of computing: the material from this year (2017) has been made publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}. This constitutes an intersting example of what can be accomplished using this platform; it is just a first example of what's possible, but an inspiring one.

Finally it should be important to evaluate, if and to which extent, the complementary approaches to training (such as schools and dedicated web-site courses) could be of mutual benefit (how to make them cooperate in the development of a complete training program).

  % \item Question-and-answer web sites such as Stack Overflow 
   %(https://en.wikipedia.org/wiki/Stack_Overflow) 
   %/ Stack Exchange are very useful. 
   %(https://en.wikipedia.org/wiki/Stack_Exchange)
  % \item An expert/tutor volunteer network (office hours style) can be established.
   %\item Investing in periodic in-person, hands-on workshops can be continued.

%____________________________________________________________________
\subsection{Leveraging existing forums}

To achieve our goals for training the community, we can take advantage of existing training forums. Resources such as conferences, workshops, and schools (in person and online) can provide a lot of value for our training purposes with little effort to set up. We should leverage the existing training forums that most closely match the HEP community's needs.

Within the HEP community, there are already some working examples of dedicated training environments that alternate between general topics and experiment-specific topics. The LHC Physics Center (LPC) at Fermilab hosts Hands-on Advanced Tutorial Sessions (HATS)\footnote{\url{http://lpc.fnal.gov/programs/schools-workshops/}} throughout the year to introduce and train participants in topics as diverse as the latest $b$-tagging algorithms, Git/GitHub, and machine learning. These HATS provide face-to-face time with instructors and participants at Fermilab, and also allow remote collaborators to join in by video and complete the same online exercises. A similar approach is in use in the CMS Data Analysis Schools (CMSDAS)\footnote{\url{http://lpc.fnal.gov/programs/schools-workshops/cmsdas.shtml}}, a series of week-long workshops that now take place at multiple labs all over the world and are designed to ramp up new collaborators in CMS-specific analysis tools while providing some discussion of the physics as well.

Other examples are CERN School of Computing\footnote{\url{https://csc.web.cern.ch/}}, CERN OpenLab Software workshops\footnote{\url{http://openlab.cern/}} education in collaboration with industry partners, and a series of more advanced topical training courses provided by MPI Munich and DESY that focus on advanced programming, use of acceleration hardware and statistical tools including machine learning. This list includes the Bertinoro\footnote{\url{https://web.infn.it/esc17/index.php?lang=en}},  GridKa\footnote{\url{http://gridka-school.scc.kit.edu/2017/}}, and CODAS-HEP\footnote{\url{http://codas-hep.org}} Schools of Computing.

%Anecdotally, these are viewed as a great success by the CMS community in preparing new collaborators and might serve as a model for future efforts in ML training.

Over the past decade, massive open online courses (MOOCs) have been developed by universities and private organizations. They have been well received by industry and academia.% and are maintained by the outside community.
In addition, they provide a lot of flexibility in terms of cost and time constraints; they are typically free and open for enrollment at any time of the year. Since the material can be accessed at any time and revisited at any time, they can be completed at a pace that makes sense, for example, for a physicist who needs to learn machine leanrning in a piecemeal way.

There are a growing set of MOOCs teaching various subjects. Since there are many options, there is a wide variety with respect to the depth of the material and specific tools taught. Exploring these options allows us to choose which is the right offering for the knowledge needed to work on a specific HEP experiment. We can pick and choose modules to tailor an appropriate roadmap of skills to learn.
More difficult will be to assemble HEP specific training material not already avaible elsewhere in an efficient and organized way, since this requires adequate organization, volounteers and and a suitable infrastructure.

Several industry conferences already exist that bring together those in academia and industry who are at the cutting edge of these techniques. Conferences such as NIPS\footnote{\url{https://nips.cc}} and PyData\footnote{\url{https://pydata.org}} provide a focused place where attendees can learn a lot about machine learning in a short period of time. Machine learning concepts such as current methods, tools, and problems facing industry and academia can be learned at conferences. In addition, conferences are an excellent networking opportunity; attendees can meet and share ideas with fellow learners and experts. Bonds can be formed quickly at conferences that can be maintained after the duration of the conference. These connections to the outside community can be essential since we will be evolving training materials to ensure that they stay relevant over time.

For example, at the time of this writing, Coursera\footnote{\url{https://www.coursera.org/learn/machine-learning}} and Udacity\footnote{\url{https://www.udacity.com/course/intro-to-machine-learning--ud120}} both provide great machine learning massive open online courses at no cost. These two courses both provide a great foundation for learning machine learning fundamentals. However, Coursera's approach emphasizes more theory (with more math background necessary) and uses MATLAB/Octave while Udacity's approach emphasizes practical using machine learning techniques in a practical sense using Python tooling.

%____________________________________________________________________
\subsection{Resources and Incentives}

Training the community is something that a large cross-section of the community understands to be important, but finding time and effort to contribute to this effort is not actively on the radar of most potential contributors. Providing incentives for their participation and creating the appropriate platforms can go a long way to reach a productive training environment. It can be as simple as inviting someone to give a software tutorial on the subject that they are familiar with, give a lecture or seminar or contribute to a growing knowledge base.

Another important incentive is recognition. For younger members of the community, having the opportunity to create a training resource, such as a software tutorial or a knowledge base on a particular topic, is very empowering and motivational to continue the efforts of training others. Engaging younger members of the community is crucial to long-term success of HEP training endeavors.

It is also critical to incorporate training into grant proposals so that it can be connected with other areas such as research and development. Efforts like DIANA-HEP \footnote{\url{http://diana-hep.org/}} and aMVA4NewPhysics \footnote{\url{https://amva4newphysics.wordpress.com/}} that combine training and software development are good examples of such ideas in practice. More examples of such efforts are needed.

\section{Other resources}

Software Carpentry\footnote{\url{https://software-carpentry.org}} and Data Carpentry\footnote{\url{http://www.datacarpentry.org}} are two parts of The Carpentries organization that collaboratively build and teach some of the basic concepts in developing and maintaining software, and analyzing data, respectively.  The materials that they develop and use are open, and can be customized for science domains
(e.g., HEP), or participant groups (e.g., undergraduates). Their model is that they offer training of trainers, and then the trainers who have graduated can offer training under the SC/DC names, though of course, anyone can use the SC/DC materials without doing it under the SC/DC names.
Subjects covered in a typical Data Carpentry school are given below. This list is taken from the curriculum of the CODATA-RDA Research Data Science Summer School in progress in Trieste, Italy during July 2017. (http://indico.ictp.it/event/7974/):

\begin{itemize}
\item Introduction
\item UNIX Shell programming (Software Carpentry Module)
\item GitHub (Software Carpentry Module)
\item R (Software Carpentry Module)
\item BYOD (Bring Your Own Data) best practices
\item Data Management with SQL  (Software Carpentry Module)
\item RDM Storage Management
\item Visualisation
\item Machine Learning Overview - Recommendation
\item Recommender Systems
\item Artificial Neural Networks and other Machine Learning Systems 
\item Research Computational Infrastructure 
\end{itemize}

While Software Carpentry is leveraged to build a foundation of knowledge for later more advanced concepts, it is important to note much of this material is developed specifically for this course and not a part of a larger Software Carpentry program. This course focuses on a shallow but wide building of foundational skills approach, introducing many base concepts but not going deep into any one concept, with a through line of Open Science and Ethical Data Usage.

\section{Conclusions}

The HEP community by and large acknowledges and recognizes the great importance of training in the field of scientific computing. This activity should encompass several types of {\it students}, from undergraduates, to young researchers, up to senior physicists, all of them in need of an appropriately designed training path in order to be proficent in their scientific endevours.

We have identified a certain number of problems that need to be overcome in setting up an appropriate training program: 
\begin{itemize}
    \item Costs and relative funding
    \item Incentives
    \item Career paths
    \item Overall organization across experiments, countries and corresponding Funding Agencies
\end{itemize}

For each of these points we provide an overview of the current situation and possible proposals to build a roadmap.

%____________________________________________________________________
%____________________________________________________________________
%\begin{itemize}
%
%   \item How to motivate the community to participate in developing programs? Can some of it come from \& leadership?
%   \item How to deal with competitiveness in the field? Can’t force people to be interested in forming a community. How can we incentivize this? Building a %vibrant, welcoming community may help.
%\end{itemize}

\section*{Aknowledgements}
A.P. acknowledges support by SNF under contract 168169.

### 2018-06-11 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Restructure paper to standard layout
Move all content into a single LaTeX file for simplicity.
Remove all unused auxilliary files.
Rename principle file to cwp-wg-training-careers.tex.
Ignore PDF output file in git to avoid accidental checkin.

--- 2018-06-11 CWP/papers/HSF-CWP-2017-02_training/latex/src/training.tex
@@
+% Copyright (C) 2018, The HSF Community White Paper authors, licence CC-BY-4.0.
+
+% JHEP preprint template
+\documentclass[12pt,a4paper]{article}
+\usepackage{jheppub}
+
+\usepackage{graphicx}
+\usepackage{subcaption}
+\usepackage{xspace}
+
+\usepackage[utf8]{inputenc}
+
+\usepackage[style=numeric-comp,sorting=none]{biblatex}
+\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp.bib}
+\addbibresource{../../HSF-CWP-2017-01_roadmap/latex/cwp-chapters.bib}
+
+\abstract{The rapidly increasing evolution of technology and the parallel increasing complexity of algorithmic analysis in HEP (and in science in general) require a much larger portfolio of specific skills by the developers with respect to the past.
+Young researchers, graduating from universities worldwide, currently do not receive adequate preparation in the very diverse fields of modern computing to respond to growing necessities posed by the most advanced experimental challenges.
+The vast majority of the HEP community strongly agrees on the need for an overarching umbrella of training programs to bring young researchers (but also more mature ones) up to date with new software technologies, such as, for example, concurrent programming and artificial intelligence.
+}
+
+\begin{document}
+
+\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
+ & & HSF-CWP-2017-02 \\
+ & & June 6, 2018 \\ % use \date or hardwire e.g. December 15, 2017
+ & & \\
+\end{tabular*}
+\vspace{2.0cm}
+
+\title{HEP Software Foundation Community White Paper Working Group --
+Training, Staffing and Careers}
+
+%\linenumbers
+
+\author[1]{Dario Berzano,}
+\author[2]{Riccardo Maria Bianchi,}
+\author[3]{Peter Elmer,}
+\author[4]{Roger Jones,}
+\author[5]{Michel Jouvin,}
+\author[6]{Daniel S. Katz,}
+\author[7]{Sudhir Malik,}
+\author[8]{Dario Menasce,}
+\author[6]{Mark Neubauer,}
+\author[9,a]{Albert Puig}
+\author[1]{and Graeme Stewart}
+
+\affiliation[1]{CERN, Geneva, Switzerland}
+\affiliation[2]{Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
+\affiliation[3]{Princeton University, Princeton, NJ, USA}
+\affiliation[4]{Lancaster University, Lancaster, UK}
+\affiliation[5]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
+\affiliation[6]{University of Illinois Urbana-Champaign, Urbana, IL, USA}
+\affiliation[7]{University of Puerto Rico, Mayaguez, PR, USA}
+\affiliation[8]{INFN Sezione di Milano-Bicocca, Italy}
+\affiliation[9]{Physik-Institut, Universität Zürich, Zürich, Switzerland}
+\affiliation[a]{Supported by SNF under contract 168169.}
+
+\maketitle
+
+\tableofcontents
+\clearpage
+
+%-------------------------------------------------------------------------
+%\newpage
+%\section{Training the community  \textcolor{pink}{Will move into a NEW chapter} }
+%\label{sec:training}
+
 % ---------------------------------------------------
@@
 
-Throughout the last few decades, the field of high energy physics (HEP) has consistently either driven or utilized the latest innovations in computational tools and technologies for problem solving. 
-The toolset of the particle physicist is vast and ever-growing, and the problem set increasingly complex.
-A typical graduate student or postdoc working on HEP experiments will likely encounter a number of tools, software, and programming languages with which they are totally unfamiliar. 
-This requires ``on-the-job'' training for the same people whose work drives the reach of our analyses, while they are developing the tools for these analyses.
+Throughout the last few decades, the field of high energy physics (HEP) has
+consistently either driven or utilized the latest innovations in computational
+tools and technologies for problem solving.
+The toolset of the particle physicist is vast and ever-growing, and the problem
+set increasingly complex.
+A typical graduate student or postdoc working on HEP experiments will likely
+encounter a number of tools, software, and programming languages with which they
+are totally unfamiliar.
+This requires ``on-the-job'' training for the same people whose work drives the
+reach of our analyses, while they are developing the tools for these analyses.
 
-A strong research community should consist of independent actors who ask research questions and direct their own analyses, instead of blindly following prescriptions and recipes.
-It is therefore imperative that the community be proactive about providing training resources and opportunities for the good of the community, and that we follow through on the promise that adequate career path possibilities exist for people in HEP who would otherwise leave the field due to limited advancement opportunities.
+A strong research community should consist of independent actors who ask
+research questions and direct their own analyses, instead of blindly following
+prescriptions and recipes. It is therefore imperative that the community be
+proactive about providing training resources and opportunities for the good of
+the community, and that we follow through on the promise that adequate career
+path possibilities exist for people in HEP who would otherwise leave the field
+due to limited advancement opportunities.
 
-To produce the best possible science, it is important, for the broad HEP community, to invest in training physicists in essential software and software engineering skills, which exist at many different levels.
-The level of depth of software engineering competences in which a physicist should be trained will vary depending on the problems that they face, and the appropriate level may evolve over time according to the problem at hand.
-For example, for a researcher who only performs basic data analysis, it may be sufficient to be familiar with just Python and data analysis libraries.
-However, contributing to a large open source project, such as ROOT, requires greater training in software development and related practices.
-These higher-order skills may include an understanding of writing testable code, time and space complexity implications, and knowledge of common software architecture paradigms.
-The need for these skills cannot be fulfilled completely just by hiring specialists in computer science, since a thorough understanding of the physics problems is usually very important to the success of the effort, making it necessary to provide a career path for those young physicists willing to steer their interests in that particular direction.
+To produce the best possible science, it is important, for the broad HEP
+community, to invest in training physicists in essential software and software
+engineering skills, which exist at many different levels. The level of depth of
+software engineering competences in which a physicist should be trained will
+vary depending on the problems that they face, and the appropriate level may
+evolve over time according to the problem at hand. For example, for a researcher
+who only performs basic data analysis, it may be sufficient to be familiar with
+just Python and data analysis libraries. However, contributing to a large open
+source project, such as ROOT, requires greater training in software development
+and related practices. These higher-order skills may include an understanding of
+writing testable code, time and space complexity implications, and knowledge of
+common software architecture paradigms. The need for these skills cannot be
+fulfilled completely just by hiring specialists in computer science, since a
+thorough understanding of the physics problems is usually very important to the
+success of the effort, making it necessary to provide a career path for those
+young physicists willing to steer their interests in that particular direction.
 
-In each HSF workshop, a discussion on the need for proper tutoring, training, and the relevant career possibilities in HEP computing emerged and was felt to be crucial and essential by almost all participants.
-There is a strong consensus that future software developments will benefit enormously from a successful implementation of such training programs, provided proper career prospects are put in place, as an incentive, to would-be trainers.
+In each HSF workshop, a discussion on the need for proper tutoring, training,
+and the relevant career possibilities in HEP computing emerged and was felt to
+be crucial and essential by almost all participants. There is a strong consensus
+that future software developments will benefit enormously from a successful
+implementation of such training programs, provided proper career prospects are
+put in place, as an incentive, to would-be trainers.
 
-This document tries to address the most relevant questions related to the training in the HEP community, namely:
+This document tries to address the most relevant questions related to the
+training in the HEP community, namely:
 \begin{itemize}
-    \item How to encourage and provide incentives to the members of the HEP community to train students and/or other collaborators?
-    \item How to properly assign credit to software development (and training thereof) activities?
-    \item What are the current practices in the field regarding hiring of researchers with a strong specialisation in computing and the policies to retain such an expertise in the long term?
-Is there room for improvement?
-    \item How to address and bridge the gap in formal software training that is not always imparted by universities?
+    \item How to encourage and provide incentives to the members of the HEP 
+    community to train students and/or other collaborators?
+    \item How to properly assign credit to software development (and training 
+    thereof) activities?
+    \item What are the current practices in the field regarding hiring of 
+    researchers with a strong specialisation in computing and the policies 
+    to retain such an expertise in the long term? Is there room for improvement?
+    \item How to address and bridge the gap in formal software training that 
+    is not always imparted by universities?
 \end{itemize}
@@
 
-The introduction and development of increasingly complex tools has led to increasing improvements in the performance of our analyses.
-However, the personnel who drive these advancements through novel implementations of the technology are the analyzers.
-In this context, the term {\it analyzers} includes all those researchers who participate in an HEP experiment whose contribution includes designing and implementing a high speed DAQ system, setting up a modern database to store the state of a detector, complex task of managing the data flow from the first trigger level to the final n-tuple dataset and creating sophisticated statistical and mathematical tools to extract publishable physical meaning from those same data.
+The introduction and development of increasingly complex tools has led to
+increasing improvements in the performance of our analyses. However, the
+personnel who drive these advancements through novel implementations of the
+technology are the analyzers. In this context, the term {\it analyzers} includes
+all those researchers who participate in an HEP experiment whose contribution
+includes designing and implementing a high speed DAQ system, setting up a modern
+database to store the state of a detector, complex task of managing the data
+flow from the first trigger level to the final n-tuple dataset and creating
+sophisticated statistical and mathematical tools to extract publishable physical
+meaning from those same data.
 
-The development of new tools is insufficient to drive advances in the field unless researchers who have the expertise apply them to physics questions.
-Furthermore, a larger and a better-connected community, with expertise in the tools and access to the new developments, can lead to an open-source model in HEP.
-The efforts to train the community in all relevant tools is, thus, a priority within the R\&D effort itself.
+The development of new tools is insufficient to drive advances in the field
+unless researchers who have the expertise apply them to physics questions.
+Furthermore, a larger and a better-connected community, with expertise in the
+tools and access to the new developments, can lead to an open-source model in
+HEP. The efforts to train the community in all relevant tools is, thus, a
+priority within the R\&D effort itself.
 
-However, as tools become more complex, the learning rate of developers must keep up if we are to utilize them effectively.
-At the same time, as in other branches of research, it is essential to sustain efforts in maintaining reproducibility of the usage of the tools.
-This again highlights the need for physicts to be appropriately trained in computing science, rather than professional computing scientists, whose goals and interests may not necessarily be aligned or focused on the physics results of a particular HEP experiment.
+However, as tools become more complex, the learning rate of developers must keep
+up if we are to utilize them effectively.  At the same time, as in other
+branches of research, it is essential to sustain efforts in maintaining
+reproducibility of the usage of the tools. This again highlights the need for
+physicts to be appropriately trained in computing science, rather than
+professional computing scientists, whose goals and interests may not necessarily
+be aligned or focused on the physics results of a particular HEP experiment.
 
-To avoid duplicate efforts within the community, it is important to find a way to approach training efforts holistically and cross-experimentally, whenever possible.
-This requires a certain level of organization and coordination in order to guarantee the delivery of efficient, non-redundant software to different experiments and to allow the different communities to agree on common standards and procedures.
+To avoid duplicate efforts within the community, it is important to find a way
+to approach training efforts holistically and cross-experimentally, whenever
+possible. This requires a certain level of organization and coordination in
+order to guarantee the delivery of efficient, non-redundant software to
+different experiments and to allow the different communities to agree on common
+standards and procedures.
 
-Beyond the immediate benefit to the scientific community in having well-trained collaborators, most people who start a physics graduate program will have careers outside of academia.
-This trend is reflected in several calls for proposals from national scientific funding agencies like H2020 and those from the National Science Foundation (NSF) that promote training in computer infrastructure and information areas.
+Beyond the immediate benefit to the scientific community in having well-trained
+collaborators, most people who start a physics graduate program will have
+careers outside of academia. This trend is reflected in several calls for
+proposals from national scientific funding agencies like H2020 and those from
+the National Science Foundation (NSF) that promote training in computer
+infrastructure and information areas.
 
@@
 
-An open issue is the kind of incentive that our community can collectively leverage in order to increase the number of people (with strong expertise in computing) who decide to dedicate time and effort to train fellow members of their community.
-To be effective, the training must be done by people who work at the cutting edge of the technology, who in many cases are found among the youngest researchers.
-These are also the very same people who are most in need of officially recognized credits for the advancement of their careers.
-The current practice in HEP is that these young scientists only receive credit and recognition (in the vast majority of cases) by working on a data analysis project, and this discourages them from finding time to transfer their acquired know-how to others in an effective way.
-A possible solution to this conundrum could be the establishment of specific career paths for researchers with a computer science specialization.
-The practical implementation of such an approach, however, is extremely challenging as there are a wide range of institutions involved that belong to different countries with their own policies, priorities, and funding strategies.
-Broadly, two career paths could be envisaged.
-Both would recognise that domain-sepcific knowledge and continued engagement is vital to the success of HEP computing and software.
-While in an ideal world (for physicists, at least), a physicst could make a detailed plan of what they need (expressed in a tidy requirements document) and a computer scientist could use this document to provide the required solution, with very little additional exchange of information during the process (which is how the software development process sometimes works in industry).
-In the real HEP world, instead, most of the time the requirements are bound to change rather frequently for a variety of reasons (not only because of poor planning) forcing developers and physicsts to closely interact during the entire process of software development; this works smoothly only when both communities have the same goals and speak the same jargon, hence the need for physicists with a computer science specialisation.
+An open issue is the kind of incentive that our community can collectively
+leverage in order to increase the number of people (with strong expertise in
+computing) who decide to dedicate time and effort to train fellow members of
+their community.
+To be effective, the training must be done by people who work at the cutting
+edge of the technology, who in many cases are found among the youngest
+researchers. These are also the very same people who are most in need of
+officially recognized credits for the advancement of their careers. The current
+practice in HEP is that these young scientists only receive credit and
+recognition (in the vast majority of cases) by working on a data analysis
+project, and this discourages them from finding time to transfer their acquired
+know-how to others in an effective way. A possible solution to this conundrum
+could be the establishment of specific career paths for researchers with a
+computer science specialization. The practical implementation of such an
+approach, however, is extremely challenging as there are a wide range of
+institutions involved that belong to different countries with their own
+policies, priorities, and funding strategies. Broadly, two career paths could be
+envisaged. Both would recognise that domain-sepcific knowledge and continued
+engagement is vital to the success of HEP computing and software. While in an
+ideal world (for physicists, at least), a physicst could make a detailed plan of
+what they need (expressed in a tidy requirements document) and a computer
+scientist could use this document to provide the required solution, with very
+little additional exchange of information during the process (which is how the
+software development process sometimes works in industry). In the real HEP
+world, instead, most of the time the requirements are bound to change rather
+frequently for a variety of reasons (not only because of poor planning) forcing
+developers and physicsts to closely interact during the entire process of
+software development; this works smoothly only when both communities have the
+same goals and speak the same jargon, hence the need for physicists with a
+computer science specialisation.
 
-The first possible career would be the path of a \emph{physicist researcher with computing science specialisation}, in some places corresponding to the role of a physicist-programmer.
-It would be conceptually different to that of a \emph{computer scientist}: the two have different goals, seek different paths to the solution of their problems, and usually do not even adopt the same jargon in expressing problems and solutions.
-Such a role would have an active role in physics analysis, and so meet many of the traditional metrics for an academic career path.
+The first possible career would be the path of a \emph{physicist researcher with
+computing science specialisation}, in some places corresponding to the role of a
+physicist-programmer. It would be conceptually different to that of a
+\emph{computer scientist}: the two have different goals, seek different paths to
+the solution of their problems, and usually do not even adopt the same jargon in
+expressing problems and solutions. Such a role would have an active role in
+physics analysis, and so meet many of the traditional metrics for an academic
+career path.
 
-The second would be the path of a domain-specific \emph{researcher software engineer}.
-Such a person has the primary role of developing software and finding software solutions, but with a large domain-specific knowledge to both understand the use cases and clients, but also of the available solutions.
-Such posts are almost impossible to develop in the current academic frameworks; but while RSEs are understood by many funding agencies and universities as a technical post, they are seen as available for short-term engagements with a researcher community, delivering a quick fix or a refactoring and then moving one.
-This does not deliver the domain specific expertise required.
+The second would be the path of a domain-specific \emph{researcher software
+engineer}. Such a person has the primary role of developing software and finding
+software solutions, but with a large domain-specific knowledge to both
+understand the use cases and clients, but also of the available solutions.  Such
+posts are almost impossible to develop in the current academic frameworks; but
+while RSEs are understood by many funding agencies and universities as a
+technical post, they are seen as available for short-term engagements with a
+researcher community, delivering a quick fix or a refactoring and then moving
+one. This does not deliver the domain specific expertise required.
 
-Both career paths suffer in academic terms from few or poorly cited publications, few opportunities to win grants in their own right and little opportunity for impact with commerce and wider society.
-These three failings make an academic appointment challenging or disfavoured compared to other applicants.
-However, there are measures that an be taken to address all three weaknesses.
-%This makes the development of specific solutions even more difficult and cumbersome.
+Both career paths suffer in academic terms from few or poorly cited
+publications, few opportunities to win grants in their own right and little
+opportunity for impact with commerce and wider society. These three failings
+make an academic appointment challenging or disfavoured compared to other
+applicants. However, there are measures that an be taken to address all three
+weaknesses.
+
+% This makes the development of specific solutions even more difficult and
+% cumbersome.
 
@@
 
-The HEP user community consists of a set of users with diverse software experience, interests, and time availablity to learn new techniques.
-Any training program must take into account the variation in target audiences.
-For example, an undergraduate doing a summer research project has different needs and skills than their mentoring professor.
-For purposes of training, it is useful to think of all our user community as \emph{students}.
-More specifically, we can broadly identify four different sets of students, classified by experience level, recognizing that even within these groups there will be people who will learn at different speeds and with differing depths of understanding:
+The HEP user community consists of a set of users with diverse software
+experience, interests, and time availablity to learn new techniques.  Any
+training program must take into account the variation in target audiences. For
+example, an undergraduate doing a summer research project has different needs
+and skills than their mentoring professor.
+For purposes of training, it is useful to think of all our user community as
+\emph{students}. More specifically, we can broadly identify four different sets
+of students, classified by experience level, recognizing that even within these
+groups there will be people who will learn at different speeds and with
+differing depths of understanding:
 
-\begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
-   \item \emph{Beginners}: New collaborators with no knowledge of the tools or techniques they are expected to use.
-These might be undergraduates but they could also be senior researchers who have not yet or not recently engaged in software analysis.
-These are people in need of some kind of formal training in modern computing techniques (basically compiled and scripting languages together with operating system basics such as filesystems, version control, and command-line shells).
-   \item \emph{Intermediate users}: HEP analyzers with some experience in either analysis concepts and/or tools, but  looking to supplement their experiences with more recent and modern approaches. 
-   \item \emph{Advanced users}: HEP analysis experts who have mastered current technologies and implementations (at a certain point in time) and want to stay up-to-date with new advanced developments.
-   \item \emph{Software developers}: HEP scientists in charge of software development in areas not limited only to analysis, such as DAQ systems, computing infrastructure, databases, pattern recongition, and complete frameworks.
+\begin{itemize}
+   \item \emph{Beginners}: New collaborators with no knowledge of the 
+   tools or techniques they are expected to use. These might be undergraduates 
+   but they could also be senior researchers who have not yet or not recently 
+   engaged in software analysis. These are people in need of some kind of formal 
+   training in modern computing techniques (basically compiled and scripting 
+   languages together with operating system basics such as filesystems, version 
+   control, and command-line shells).
+   \item \emph{Intermediate users}: HEP analyzers with some experience in 
+   either analysis concepts and/or tools, but  looking to supplement their 
+   experiences with more recent and modern approaches. 
+   \item \emph{Advanced users}: HEP analysis experts who have mastered current 
+   technologies and implementations (at a certain point in time) and want to 
+   stay up-to-date with new advanced developments.
+   \item \emph{Software developers}: HEP scientists in charge of software 
+   development in areas not limited only to analysis, such as DAQ systems, 
+   computing infrastructure, databases, pattern recongition, and complete 
+   frameworks.
 \end{itemize}
 
-Each of these groups has vastly different training needs.
-However, \emph{whenever possible}, any training program should take advantage of developments in pedagogy, such as active learning\footnote{\url{http://www.crlt.umich.edu/tstrategies/tsal}}, peer learning\footnote{\url{https://en.wikipedia.org/wiki/Peer_learning}}, and web-based training.\footnote{See \S\ref{sec:initiatives}.} In some cases, it may even be advantageous to hand out code samples that are purposely broken or flawed, and ask students to fix or improve them.
-Learning the material in a way that sticks with the students is difficult and challenging for both the students and the instructor and often takes more time than we would prefer.
-However, this is the best way to ensure an educated community of scientists who can fully contribute to the physics programs at large, which is really the ultimate goal of any training program.
+Each of these groups has vastly different training needs. However,
+\emph{whenever possible}, any training program should take advantage of
+developments in pedagogy, such as active
+learning\footnote{\url{http://www.crlt.umich.edu/tstrategies/tsal}}, peer
+learning\footnote{\url{https://en.wikipedia.org/wiki/Peer_learning}}, and
+web-based training.\footnote{See \S\ref{sec:initiatives}.} In some cases, it may
+even be advantageous to hand out code samples that are purposely broken or
+flawed, and ask students to fix or improve them. Learning the material in a way
+that sticks with the students is difficult and challenging for both the students
+and the instructor and often takes more time than we would prefer. However, this
+is the best way to ensure an educated community of scientists who can fully
+contribute to the physics programs at large, which is really the ultimate goal
+of any training program.
 
@@
 
-At all stages of software and computing training, we should take care to encourage \emph{Good Practices Across the Community} (GPAC), such as error checking, modularity of code design, version control, writing tests, etc.
-All the key concepts addressed in the training should not be specific to a particular experiment or field of application, but general enough to be useful to the whole HEP community and possibly beyond.
-In this section, we present a list of specific concepts that need to be taught to members of the community, in order to guarantee the base level of competence needed to write efficient code for the different tasks performed in HEP experiments.
+At all stages of software and computing training, we should take care to
+encourage \emph{Good Practices Across the Community} (GPAC), such as error
+checking, modularity of code design, version control, writing tests, etc. All
+the key concepts addressed in the training should not be specific to a
+particular experiment or field of application, but general enough to be useful
+to the whole HEP community and possibly beyond. In this section, we present a
+list of specific concepts that need to be taught to members of the community, in
+order to guarantee the base level of competence needed to write efficient code
+for the different tasks performed in HEP experiments.
 
-An essential knowledge base that needs to be transferred includes basic programming concepts, data structures, basics of code design, error checking, code management tools, validation and debugging tools.
-More advanced topics include modularity of code design, advanced data structures, evaluation metrics, writing tests and working with different types of hardware accelerators.
-Cutting-edge approaches such as the use of hardware accelerators (whether GPUs or FPGAs), Deep Neural Networks, and Machine Learning should be considered important training topics.
-Additionally, special emphasis should be made on reporting results and documenting them.
-The training subjects that the community considers important to pursue are listed as follows:
+An essential knowledge base that needs to be transferred includes basic
+programming concepts, data structures, basics of code design, error checking,
+code management tools, validation and debugging tools. More advanced topics
+include modularity of code design, advanced data structures, evaluation metrics,
+writing tests and working with different types of hardware accelerators.
+Cutting-edge approaches such as the use of hardware accelerators (whether GPUs
+or FPGAs), Deep Neural Networks, and Machine Learning should be considered
+important training topics. Additionally, special emphasis should be made on
+reporting results and documenting them.
+The training subjects that the community considers important to pursue are
+listed as follows:
 
-\begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
+\begin{itemize}
    \item Basic and Advanced Programming Concepts
-   \begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
+   \begin{itemize}
       \item Object oriented paradigm
@@
    \item Algorithmics 
-   \begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
+   \begin{itemize}
       \item Boost library
@@
    \item Existing frameworks (development or application level)
-   \begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
+   \begin{itemize}
       \item Qt
@@
    \item Development tools
-   \begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
+   \begin{itemize}
       \item IDEs (Integrated Development Environment)
@@
 
-Within a single experiment, different skill sets are needed.
-In addition to a skill set that contains basic programming language knowledge, testing and code management tools, and experiment-specific framework knowledge, there are more specialized skills that only a subset of the community needs to know, such as software optimization, or low-level hardware interfaces.
-While there already are experiment-specific training efforts in place, there are many needs that are in common between experiments.
-Therefore, we should probably strive to extract the common knowledge and build a common training program.
-This would reduce duplication of efforts and enable growth of a shared training culture by accumulating and sharing expertise across different experiments.
+Within a single experiment, different skill sets are needed. In addition to a
+skill set that contains basic programming language knowledge, testing and code
+management tools, and experiment-specific framework knowledge, there are more
+specialized skills that only a subset of the community needs to know, such as
+software optimization, or low-level hardware interfaces.
+While there already are experiment-specific training efforts in place, there are
+many needs that are in common between experiments. Therefore, we should probably
+strive to extract the common knowledge and build a common training program. This
+would reduce duplication of efforts and enable growth of a shared training
+culture by accumulating and sharing expertise across different experiments.
 
-To realise this goal, a possibility could be the institution of a {\em federation} of existing training initiatives, a very lightwheight umbrella just aimed at improving the efficiency and cost-effectivness
-of this important activity.
+To realise this goal, a possibility could be the institution of a {\em
+federation} of existing training initiatives, a very lightwheight umbrella just
+aimed at improving the efficiency and cost-effectivness of this important
+activity.
 
@@
 The implementation of training should target multiple training formats and knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and advanced visualizations, etc.
-People who learn in non-conventional ways should not be excluded from the opportunity to learn.
-Since there are different stages of knowledge transfer, programs should be developed that encourage experts to share knowledge.
-Issues considered in this section include global training initiatives, strategies for leveraging existing forums, and training tool development.
+People who learn in non-conventional ways should not be excluded from the
+opportunity to learn. Since there are different stages of knowledge transfer,
+programs should be developed that encourage experts to share knowledge. Issues
+considered in this section include global training initiatives, strategies for
+leveraging existing forums, and training tool development.
 
-An important point to consider is the difficulty, often experienced in the past, in developing large software programs across different experiment communities.
-It is always challenging to organize a Working Group for training across experiments or even across diverse communities (such as HEP and Astroparticle, for example).
-This is a point where an appropriate incentive program to reward those who train the community might help.
+An important point to consider is the difficulty, often experienced in the past,
+in developing large software programs across different experiment communities.
+It is always challenging to organize a Working Group for training across
+experiments or even across diverse communities (such as HEP and Astroparticle,
+for example). This is a point where an appropriate incentive program to reward
+those who train the community might help.
 
@@
 
-Some methods that can be used for global training include Massive Online Open Courses, hands-on workshops, online knowledge bases, expert trainer volunteer networks, and web-based training approaches.
+Some methods that can be used for global training include Massive Online Open
+Courses, hands-on workshops, online knowledge bases, expert trainer volunteer
+networks, and web-based training approaches.
 
-Massive Online Open Courses (MOOCs) can be used to develop an open-source set of tutorials and tools.
-Existing online courses such as Udacity and Coursera can be evaluated and recommended to the community.
-A novel approach such as WikiToLearn could also explored to assess potential benefits (as has already been attempted in the context of the GridKa School of Computing, see below).
+Massive Online Open Courses (MOOCs) can be used to develop an open-source set of
+tutorials and tools. Existing online courses such as Udacity and Coursera can be
+evaluated and recommended to the community. A novel approach such as WikiToLearn
+could also explored to assess potential benefits (as has already been attempted
+in the context of the GridKa School of Computing, see below).
 
-Experiment-specific and global knowledge bases can be established with incentives for experts to contribute.
-They can be open-source so that a lot of knowledge can be added by the trainees themselves as information is learned (this is the particular context where an approach such as WikiToLearn could be of great help).
+Experiment-specific and global knowledge bases can be established with
+incentives for experts to contribute. They can be open-source so that a lot of
+knowledge can be added by the trainees themselves as information is learned
+(this is the particular context where an approach such as WikiToLearn could be
+of great help).
 
-Question-and-answer websites such as Stack Overflow for HEP are also a very useful resource for common problems and questions.
-This approach has already been considered by HSF start-up members, and turned out to be difficult to pursue for a number of reasons, but in the future, boundary conditions might change, making this approach viable.
+Question-and-answer websites such as Stack Overflow for HEP are also a very
+useful resource for common problems and questions. This approach has already
+been considered by HSF start-up members, and turned out to be difficult to
+pursue for a number of reasons, but in the future, boundary conditions might
+change, making this approach viable.
 
-Hands-on workshops are an invaluable part of learning how to apply theoretical concepts in practice.
-Identifying which of the current workshops are productive and useful, and if they cover all the topics that need to be transferred and that are in demand by the students, is an important action item.
+Hands-on workshops are an invaluable part of learning how to apply theoretical
+concepts in practice. Identifying which of the current workshops are productive
+and useful, and if they cover all the topics that need to be transferred and
+that are in demand by the students, is an important action item.
 
-Creating an expert tutor volunteer network is another way to provide training and support to the community. 
-This of course requires proper recognition, at least in terms of career prospects as an incentive for young researchers.
-It is clear that the best possible tutors are, in principle, those people who are actively engaged in modern software developments: these are in general young researchers in the first steps of their careers and, in order to be attracted to devoting substantial time to training and tutoring, they must be assured that such an effort would be properly recognised in an official way.
-Such kind of recognition is not currently implemented, at least not in a standardized or official way.
-A possible structure for such a network could be the establishment of a {\em federation} of existing schools.
+Creating an expert tutor volunteer network is another way to provide training
+and support to the community.
+This of course requires proper recognition, at least in terms of career
+prospects as an incentive for young researchers. It is clear that the best
+possible tutors are, in principle, those people who are actively engaged in
+modern software developments: these are in general young researchers in the
+first steps of their careers and, in order to be attracted to devoting
+substantial time to training and tutoring, they must be assured that such an
+effort would be properly recognised in an official way. Such kind of recognition
+is not currently implemented, at least not in a standardized or official way. A
+possible structure for such a network could be the establishment of a {\em
+federation} of existing schools.
 
-It should be considered that some professors, who act as graduate student advisors, might need to be encouraged to make sure their students are properly trained.
-Sometimes, students are instead pushed to learn the bare minimum to get the work done, at the expense of a broader training/education curriculum that could actually yield improved results, but further down the line.
-One incentive would be to provide training programs that also count as course credit, perhaps as an elective.
-This model is already in limited use with some online solutions\footnote{\url{https://www.edx.org/credit}}, but this is not universal.
-Discussions should be started with collaborators at higher education institutions to see what the roadblocks or opportunities would be for these training sessions to serve double duty.
-It should also be considered that not all students will end their career in research or academia: their contribution to the research activity, as students, should therefore also provide them knowledge, know-how, and skills considered a valuable asset by industry, in order to increase their chances of a career outside research.
+It should be considered that some professors, who act as graduate student
+advisors, might need to be encouraged to make sure their students are properly
+trained. Sometimes, students are instead pushed to learn the bare minimum to get
+the work done, at the expense of a broader training/education curriculum that
+could actually yield improved results, but further down the line. One incentive
+would be to provide training programs that also count as course credit, perhaps
+as an elective. This model is already in limited use with some online
+solutions\footnote{\url{https://www.edx.org/credit}}, but this is not universal.
+Discussions should be started with collaborators at higher education
+institutions to see what the roadblocks or opportunities would be for these
+training sessions to serve double duty.
+It should also be considered that not all students will end their career in
+research or academia: their contribution to the research activity, as students,
+should therefore also provide them knowledge, know-how, and skills considered a
+valuable asset by industry, in order to increase their chances of a career
+outside research.
 
-Difficulties that have emerged in the past with respect to implementing training courses are the lack of funding and the lack of available time by experts in the field.
-People with enough expertise or insight in the field usually don't have time to devote to prolonged periods of student training, and, even when they can find time, the cost of setting up a training course in an effective way is often beyond what's made available by funding agencies (funds for travel, hosting, setting up a room with a computing infrastructure to allow interactive hands-on session, etc.) A possible solution is a completely different approach to training (but complementary to the already existing and successful classical efforts such as the CERN School of Computing's Bertinoro and Kit ones): instead of directly teaching to students, trainees could make use of a web-based platform to provide training materials to students.
-This complementary approach has several advantages over traditional ones:
-\begin{itemize}[itemsep=0pt,parsep=0pt,topsep=5pt,partopsep=0pt]
-   \item Tutors can add material to the web site at a very slow pace (whenever they find time to do it, one slide a week or a chapter a day)
-   \item Their material, made public on a web-site, can be further expanded by collaborators (also at their own pace) or, even better and more productively, by students who decide to contribute new additions, examples, exercises etc.
-Such a collaborative effort allows more people to be exposed to training at any given time, creates a sense of community, and creates bridges between people in contiguous areas or research.
-Students can use the same platform to exchange their own examples, make suggestions and point out interesting concepts.
-In such a model, the possible contribution from others to the training material needs to be moderated and validated with appropriate policies.
-   \item If complemented by the availability of remote virtual machines (possibly through a browser), students could have access to examples and exercises that are already embedded in their own natural environment: all the necessary tools/libraries needed to implement the exercise will be already available in the virtual machine (possibly a Docker container).
-With just a web browser, students could run complex examples from home, taking advantage of a remote facility that provides some storage and computing power.
-Important here is the concept of ``environment'': a Docker container could be set up in such a way that students will work in an exact replica of the environment they will be exposed to in their experiment (or collaboration at large).
-Moreover, students could be provided with Docker containers that preserves their modified environment across sessions, allowing them to develop their skills over a prolonged period of time by accessing all the files that were made persistent day by day during the training.
-   \item There would no longer be a need to find the resources to host a school and pay the tutor(s) (and eventually subsidize the students to participate in training in a remote location).
-Students could follow the training at their own pace from wherever they happen to be.
-A traditional school only lasts for 5 days (10 at most) and it is difficult to cover a  subject to any significant depth in such a short time.
-The web approach, instead, would allow for very long and in-depth coverage of any kind of subject, and in this sense it could be a {\it complementary} approach to a traditional school.
-Of particular interest could be courses such as, e.g., ``Machine Learning'', ``Statistical Analyis with ROOT'', or even just ``Good practices in C++'' or ``Python Programming for scientific computing''.
-   \item Finally, this approach could allow the creation of {\it browsable} repositories of all training materials, grouped by argument, by relevance, by experiment or whatever other critieria.
-Students from all over the world could be exposed to a large repository of examples, exercises and in general training material from their own home.
+Difficulties that have emerged in the past with respect to implementing training
+courses are the lack of funding and the lack of available time by experts in the
+field. People with enough expertise or insight in the field usually don't have
+time to devote to prolonged periods of student training, and, even when they can
+find time, the cost of setting up a training course in an effective way is often
+beyond what's made available by funding agencies (funds for travel, hosting,
+setting up a room with a computing infrastructure to allow interactive hands-on
+session, etc.) A possible solution is a completely different approach to
+training (but complementary to the already existing and successful classical
+efforts such as the CERN School of Computing's Bertinoro and Kit ones): instead
+of directly teaching to students, trainees could make use of a web-based
+platform to provide training materials to students. This complementary approach
+has several advantages over traditional ones:
+\begin{itemize}
+   \item Tutors can add material to the web site at a very slow pace (whenever
+   they find time to do it, one slide a week or a chapter a day) \item Their
+   material, made public on a web-site, can be further expanded by collaborators
+   (also at their own pace) or, even better and more productively, by students
+   who decide to contribute new additions, examples, exercises etc. Such a
+   collaborative effort allows more people to be exposed to training at any
+   given time, creates a sense of community, and creates bridges between people
+   in contiguous areas or research. Students can use the same platform to
+   exchange their own examples, make suggestions and point out interesting
+   concepts. In such a model, the possible contribution from others to the
+   training material needs to be moderated and validated with appropriate
+   policies.
+   \item If complemented by the availability of remote virtual machines
+   (possibly through a browser), students could have access to examples and
+   exercises that are already embedded in their own natural environment: all the
+   necessary tools/libraries needed to implement the exercise will be already
+   available in the virtual machine (possibly a Docker container). With just a
+   web browser, students could run complex examples from home, taking advantage
+   of a remote facility that provides some storage and computing power.
+   Important here is the concept of ``environment'': a Docker container could be
+   set up in such a way that students will work in an exact replica of the
+   environment they will be exposed to in their experiment (or collaboration at
+   large). Moreover, students could be provided with Docker containers that
+   preserves their modified environment across sessions, allowing them to
+   develop their skills over a prolonged period of time by accessing all the
+   files that were made persistent day by day during the training.
+   \item There would no longer be a need to find the resources to host a school
+   and pay the tutor(s) (and eventually subsidize the students to participate in
+   training in a remote location). Students could follow the training at their
+   own pace from wherever they happen to be. A traditional school only lasts for
+   5 days (10 at most) and it is difficult to cover a  subject to any
+   significant depth in such a short time. The web approach, instead, would
+   allow for very long and in-depth coverage of any kind of subject, and in this
+   sense it could be a {\it complementary} approach to a traditional school. Of
+   particular interest could be courses such as, e.g., ``Machine Learning'',
+   ``Statistical Analyis with ROOT'', or even just ``Good practices in C++'' or
+   ``Python Programming for scientific computing''.
+   \item Finally, this approach could allow the creation of {\it browsable}
+   repositories of all training materials, grouped by argument, by relevance, by
+   experiment or whatever other critieria. Students from all over the world
+   could be exposed to a large repository of examples, exercises and in general
+   training material from their own home.
 \end{itemize}
 
-An example of a such a web-based platform already exists, and has been implemented as an Open Source project (backed by Wikimedia) by a group of more than 30 Italian students.
-The project is WikiToLearn \footnote{\url{https://it.wikitolearn.org/}}.
-It hosts training material in several languages, for several disciplines, ranging from Economics to Physics, Mathematics, and several others.
-Because it is based upon wikimedia\footnote{\url{https://www.wikimedia.org/}} software, it is very easy to add material to the site, and to make it appear under a specific topic (such as, e.g., Software/Techniques/Machine-Learning) and to manipulate it as if it were a single document.
-In the end, students can selectively choose individual chapters from the site and have the corresponding pdf sent them as a book, complete with index, content, and chapters.
+An example of a such a web-based platform already exists, and has been
+implemented as an Open Source project (backed by Wikimedia) by a group of more
+than 30 Italian students. The project is WikiToLearn
+\footnote{\url{https://it.wikitolearn.org/}}. It hosts training material in
+several languages, for several disciplines, ranging from Economics to Physics,
+Mathematics, and several others. Because it is based upon
+wikimedia\footnote{\url{https://www.wikimedia.org/}} software, it is very easy
+to add material to the site, and to make it appear under a specific topic (such
+as, e.g., Software/Techniques/Machine-Learning) and to manipulate it as if it
+were a single document. In the end, students can selectively choose individual
+chapters from the site and have the corresponding pdf sent them as a book,
+complete with index, content, and chapters.
 
-The adoption of such an approach is made rather easy in WikiToLearn by the relative simplicity of the wikimedia-based toolset: users contribute their training material using just a web-browser, and in order to do this efficiently, the learning curve has been kept appropriately shallow.
+The adoption of such an approach is made rather easy in WikiToLearn by the
+relative simplicity of the wikimedia-based toolset: users contribute their
+training material using just a web-browser, and in order to do this efficiently,
+the learning curve has been kept appropriately shallow.
 
-An interesting exercise in this context has recently been made by colleagues of the GridKa School of computing: the material from this year (2017) has been made publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}.
-This constitutes an intersting example of what can be accomplished using this platform; it is just a first example of what's possible, but an inspiring one.
+An interesting exercise in this context has recently been made by colleagues of
+the GridKa School of computing: the material from this year (2017) has been made
+publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}.
+This constitutes an intersting example of what can be accomplished using this
+platform; it is just a first example of what's possible, but an inspiring one.
 
-Finally it should be important to evaluate, if and to which extent, the complementary approaches to training (such as schools and dedicated web-site courses) could be of mutual benefit (how to make them cooperate in the development of a complete training program).
+Finally it should be important to evaluate, if and to which extent, the
+complementary approaches to training (such as schools and dedicated web-site
+courses) could be of mutual benefit (how to make them cooperate in the
+development of a complete training program).
 
@@
 
-To achieve our goals for training the community, we can take advantage of existing training forums.
-Resources such as conferences, workshops, and schools (in person and online) can provide a lot of value for our training purposes with little effort to set up.
-We should leverage the existing training forums that most closely match the HEP community's needs.
+To achieve our goals for training the community, we can take advantage of
+existing training forums. Resources such as conferences, workshops, and schools
+(in person and online) can provide a lot of value for our training purposes with
+little effort to set up. We should leverage the existing training forums that
+most closely match the HEP community's needs.
 
-Within the HEP community, there are already some working examples of dedicated training environments that alternate between general topics and experiment-specific topics.
-The LHC Physics Center (LPC) at Fermilab hosts Hands-on Advanced Tutorial Sessions (HATS)\footnote{\url{http://lpc.fnal.gov/programs/schools-workshops/}} throughout the year to introduce and train participants in topics as diverse as the latest $b$-tagging algorithms, Git/GitHub, and machine learning.
-These HATS provide face-to-face time with instructors and participants at Fermilab, and also allow remote collaborators to join in by video and complete the same online exercises.
-A similar approach is in use in the CMS Data Analysis Schools (CMSDAS)\footnote{\url{http://lpc.fnal.gov/programs/schools-workshops/cmsdas.shtml}}, a series of week-long workshops that now take place at multiple labs all over the world and are designed to ramp up new collaborators in CMS-specific analysis tools while providing some discussion of the physics as well.
+Within the HEP community, there are already some working examples of dedicated
+training environments that alternate between general topics and
+experiment-specific topics. The LHC Physics Center (LPC) at Fermilab hosts
+Hands-on Advanced Tutorial Sessions
+(HATS)\footnote{\url{http://lpc.fnal.gov/programs/schools-workshops/}}
+throughout the year to introduce and train participants in topics as diverse as
+the latest $b$-tagging algorithms, Git/GitHub, and machine learning. These HATS
+provide face-to-face time with instructors and participants at Fermilab, and
+also allow remote collaborators to join in by video and complete the same online
+exercises. A similar approach is in use in the CMS Data Analysis Schools
+(CMSDAS)\footnote{\url{http://lpc.fnal.gov/programs/schools-workshops/cmsdas.shtml}},
+a series of week-long workshops that now take place at multiple labs all over
+the world and are designed to ramp up new collaborators in CMS-specific analysis
+tools while providing some discussion of the physics as well.
 
-Other examples are CERN School of Computing\footnote{\url{https://csc.web.cern.ch/}}, CERN OpenLab Software workshops\footnote{\url{http://openlab.cern/}} education in collaboration with industry partners, and a series of more advanced topical training courses provided by MPI Munich and DESY that focus on advanced programming, use of acceleration hardware and statistical tools including machine learning.
-This list includes the Bertinoro\footnote{\url{https://web.infn.it/esc17/index.php?lang=en}},  GridKa\footnote{\url{http://gridka-school.scc.kit.edu/2017/}}, and CODAS-HEP\footnote{\url{http://codas-hep.org}} Schools of Computing.
+Other examples are CERN School of
+Computing\footnote{\url{https://csc.web.cern.ch/}}, CERN OpenLab Software
+workshops\footnote{\url{http://openlab.cern/}} education in collaboration with
+industry partners, and a series of more advanced topical training courses
+provided by MPI Munich and DESY that focus on advanced programming, use of
+acceleration hardware and statistical tools including machine learning. This
+list includes the
+Bertinoro\footnote{\url{https://web.infn.it/esc17/index.php?lang=en}}, 
+GridKa\footnote{\url{http://gridka-school.scc.kit.edu/2017/}}, and
+CODAS-HEP\footnote{\url{http://codas-hep.org}} Schools of Computing.
 
-%Anecdotally, these are viewed as a great success by the CMS community in preparing new collaborators and might serve as a model for future efforts in ML training.
+% Anecdotally, these are viewed as a great success by the CMS community in
+% preparing new collaborators and might serve as a model for future efforts in
+% ML training.
 
-Over the past decade, massive open online courses (MOOCs) have been developed by universities and private organizations.
-They have been well received by industry and academia.% and are maintained by the outside community.
-In addition, they provide a lot of flexibility in terms of cost and time constraints; they are typically free and open for enrollment at any time of the year.
-Since the material can be accessed at any time and revisited at any time, they can be completed at a pace that makes sense, for example, for a physicist who needs to learn machine leanrning in a piecemeal way.
+Over the past decade, massive open online courses (MOOCs) have been developed by
+universities and private organizations. They have been well received by industry
+and academia.% and are maintained by the outside community.
+In addition, they provide a lot of flexibility in terms of cost and time
+constraints; they are typically free and open for enrollment at any time of the
+year. Since the material can be accessed at any time and revisited at any time,
+they can be completed at a pace that makes sense, for example, for a physicist
+who needs to learn machine leanrning in a piecemeal way.
 
-There are a growing set of MOOCs teaching various subjects.
-Since there are many options, there is a wide variety with respect to the depth of the material and specific tools taught.
-Exploring these options allows us to choose which is the right offering for the knowledge needed to work on a specific HEP experiment.
-We can pick and choose modules to tailor an appropriate roadmap of skills to learn.
-More difficult will be to assemble HEP specific training material not already avaible elsewhere in an efficient and organized way, since this requires adequate organization, volounteers and and a suitable infrastructure.
+There are a growing set of MOOCs teaching various subjects. Since there are many
+options, there is a wide variety with respect to the depth of the material and
+specific tools taught. Exploring these options allows us to choose which is the
+right offering for the knowledge needed to work on a specific HEP experiment. We
+can pick and choose modules to tailor an appropriate roadmap of skills to learn.
+More difficult will be to assemble HEP specific training material not already
+avaible elsewhere in an efficient and organized way, since this requires
+adequate organization, volounteers and and a suitable infrastructure.
 
-Several industry conferences already exist that bring together those in academia and industry who are at the cutting edge of these techniques.
-Conferences such as NIPS\footnote{\url{https://nips.cc}} and PyData\footnote{\url{https://pydata.org}} provide a focused place where attendees can learn a lot about machine learning in a short period of time.
-Machine learning concepts such as current methods, tools, and problems facing industry and academia can be learned at conferences.
-In addition, conferences are an excellent networking opportunity; attendees can meet and share ideas with fellow learners and experts.
-Bonds can be formed quickly at conferences that can be maintained after the duration of the conference.
-These connections to the outside community can be essential since we will be evolving training materials to ensure that they stay relevant over time.
+Several industry conferences already exist that bring together those in academia
+and industry who are at the cutting edge of these techniques. Conferences such
+as NIPS\footnote{\url{https://nips.cc}} and
+PyData\footnote{\url{https://pydata.org}} provide a focused place where
+attendees can learn a lot about machine learning in a short period of time.
+Machine learning concepts such as current methods, tools, and problems facing
+industry and academia can be learned at conferences. In addition, conferences
+are an excellent networking opportunity; attendees can meet and share ideas with
+fellow learners and experts. Bonds can be formed quickly at conferences that can
+be maintained after the duration of the conference. These connections to the
+outside community can be essential since we will be evolving training materials
+to ensure that they stay relevant over time.
 
-For example, at the time of this writing, Coursera\footnote{\url{https://www.coursera.org/learn/machine-learning}} and Udacity\footnote{\url{https://www.udacity.com/course/intro-to-machine-learning--ud120}} both provide great machine learning massive open online courses at no cost.
-These two courses both provide a great foundation for learning machine learning fundamentals.
-However, Coursera's approach emphasizes more theory (with more math background necessary) and uses MATLAB/Octave while Udacity's approach emphasizes practical using machine learning techniques in a practical sense using Python tooling.
+For example, at the time of this writing,
+Coursera\footnote{\url{https://www.coursera.org/learn/machine-learning}} and
+Udacity\footnote{\url{https://www.udacity.com/course/intro-to-machine-learning--ud120}}
+both provide great machine learning massive open online courses at no cost.
+These two courses both provide a great foundation for learning machine learning
+fundamentals. However, Coursera's approach emphasizes more theory (with more
+math background necessary) and uses MATLAB/Octave while Udacity's approach
+emphasizes practical using machine learning techniques in a practical sense
+using Python tooling.
 
@@
 
-Training the community is something that a large cross-section of the community understands to be important, but finding time and effort to contribute to this effort is not actively on the radar of most potential contributors.
-Providing incentives for their participation and creating the appropriate platforms can go a long way to reach a productive training environment.
-It can be as simple as inviting someone to give a software tutorial on the subject that they are familiar with, give a lecture or seminar or contribute to a growing knowledge base.
+Training the community is something that a large cross-section of the community
+understands to be important, but finding time and effort to contribute to this
+effort is not actively on the radar of most potential contributors. Providing
+incentives for their participation and creating the appropriate platforms can go
+a long way to reach a productive training environment. It can be as simple as
+inviting someone to give a software tutorial on the subject that they are
+familiar with, give a lecture or seminar or contribute to a growing knowledge
+base.
 
-Another important incentive is recognition.
-For younger members of the community, having the opportunity to create a training resource, such as a software tutorial or a knowledge base on a particular topic, is very empowering and motivational to continue the efforts of training others.
-Engaging younger members of the community is crucial to long-term success of HEP training endeavors.
+Another important incentive is recognition. For younger members of the
+community, having the opportunity to create a training resource, such as a
+software tutorial or a knowledge base on a particular topic, is very empowering
+and motivational to continue the efforts of training others. Engaging younger
+members of the community is crucial to long-term success of HEP training
+endeavors.
 
-It is also critical to incorporate training into grant proposals so that it can be connected with other areas such as research and development.
-Efforts like DIANA-HEP \footnote{\url{http://diana-hep.org/}} and aMVA4NewPhysics \footnote{\url{https://amva4newphysics.wordpress.com/}} that combine training and software development are good examples of such ideas in practice.
-More examples of such efforts are needed.
+It is also critical to incorporate training into grant proposals so that it can
+be connected with other areas such as research and development. Efforts like
+DIANA-HEP \footnote{\url{http://diana-hep.org/}} and aMVA4NewPhysics
+\footnote{\url{https://amva4newphysics.wordpress.com/}} that combine training
+and software development are good examples of such ideas in practice. More
+examples of such efforts are needed.
 
@@
 
-Software Carpentry\footnote{\url{https://software-carpentry.org}} and Data Carpentry\footnote{\url{http://www.datacarpentry.org}} are two parts of The Carpentries organization that collaboratively build and teach some of the basic concepts in developing and maintaining software, and analyzing data, respectively.
-The materials that they develop and use are open, and can be customized for science domains
-(e.g., HEP), or participant groups (e.g., undergraduates).
-Their model is that they offer training of trainers, and then the trainers who have graduated can offer training under the SC/DC names, though of course, anyone can use the SC/DC materials without doing it under the SC/DC names.
-Subjects covered in a typical Data Carpentry school are given below.
-This list is taken from the curriculum of the CODATA-RDA Research Data Science Summer School in progress in Trieste, Italy during July 2017.
+Software Carpentry\footnote{\url{https://software-carpentry.org}} and Data
+Carpentry\footnote{\url{http://www.datacarpentry.org}} are two parts of The
+Carpentries organization that collaboratively build and teach some of the basic
+concepts in developing and maintaining software, and analyzing data,
+respectively.  The materials that they develop and use are open, and can be
+customized for science domains (e.g., HEP), or participant groups (e.g.,
+undergraduates). Their model is that they offer training of trainers, and then
+the trainers who have graduated can offer training under the SC/DC names, though
+of course, anyone can use the SC/DC materials without doing it under the SC/DC
+names.
+Subjects covered in a typical Data Carpentry school are given below. This list
+is taken from the curriculum of the CODATA-RDA Research Data Science Summer
+School in progress in Trieste, Italy during July 2017.
 (http://indico.ictp.it/event/7974/):
@@
 
-While Software Carpentry is leveraged to build a foundation of knowledge for later more advanced concepts, it is important to note much of this material is developed specifically for this course and not a part of a larger Software Carpentry program.
-This course focuses on a shallow but wide building of foundational skills approach, introducing many base concepts but not going deep into any one concept, with a through line of Open Science and Ethical Data Usage.
+While Software Carpentry is leveraged to build a foundation of knowledge for
+later more advanced concepts, it is important to note much of this material is
+developed specifically for this course and not a part of a larger Software
+Carpentry program. This course focuses on a shallow but wide building of
+foundational skills approach, introducing many base concepts but not going deep
+into any one concept, with a through line of Open Science and Ethical Data
+Usage.
 
@@
 
-The HEP community by and large acknowledges and recognizes the great importance of training in the field of scientific computing.
-This activity should encompass several types of {\it students}, from undergraduates, to young researchers, up to senior physicists, all of them in need of an appropriately designed training path in order to be proficent in their scientific endevours.
+The HEP community by and large acknowledges and recognizes the great importance
+of training in the field of scientific computing. This activity should encompass
+several types of {\it students}, from undergraduates, to young researchers, up
+to senior physicists, all of them in need of an appropriately designed training
+path in order to be proficent in their scientific endevours.
 
-We have identified a certain number of problems that need to be overcome in setting up an appropriate training program: 
+We have identified a certain number of problems that need to be overcome in
+setting up an appropriate training program:
 \begin{itemize}
@@
     \item Career paths
-    \item Overall organization across experiments, countries and corresponding Funding Agencies
+    \item Overall organization across experiments, countries and 
+    corresponding Funding Agencies
 \end{itemize}
@@
 
-\section*{Aknowledgements}
-A.P. acknowledges support by SNF under contract 168169.
+\sloppy
+\raggedright
+\clearpage
+\printbibliography[title={References},heading=bibintoc]
+
+\printbibliography[title={references}]
+
+\end{document}

### 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Changes from John H 2018-06-12

@@
 
-\abstract{The rapidly increasing evolution of technology and the parallel increasing complexity of algorithmic analysis in HEP (and in science in general) require a much larger portfolio of specific skills by the developers with respect to the past.
-Young researchers, graduating from universities worldwide, currently do not receive adequate preparation in the very diverse fields of modern computing to respond to growing necessities posed by the most advanced experimental challenges.
-The vast majority of the HEP community strongly agrees on the need for an overarching umbrella of training programs to bring young researchers (but also more mature ones) up to date with new software technologies, such as, for example, concurrent programming and artificial intelligence.
-}
+\abstract{The rapid evolution of technology and the parallel increasing 
+complexity of algorithmic analysis in HEP require developers 
+to acquire a much larger portfolio of programming skills. Young 
+researchers graduating from universities worldwide currently do not receive adequate 
+preparation in the very diverse fields of modern computing to respond to growing needs 
+of the most advanced experimental challenges. There is a growing cnsensus in the HEP community 
+on the need for training programmes to bring researchers up to date with new software technologies, 
+in particular in the domains of concurrent programming and artificial intelligence. We review some
+of the initiatives under way for introducing new training programmes and highlight some of the issues 
+that need to be taken into account for these to be successful.}
 
@@
 \author[3]{Peter Elmer,}
+\author[1]{John Harvey,}
 \author[4]{Roger Jones,}
@@
 \author[6]{Mark Neubauer,}
-\author[9,a]{Albert Puig}
-\author[1]{and Graeme Stewart}
+\author[9,a]{Albert Puig,}
+\author[1]{Graeme Stewart,}
+\author[10]{and Christopher Tunnell}
 
@@
 \affiliation[9]{Physik-Institut, Universität Zürich, Zürich, Switzerland}
+\affiliation[10]{University of Chicago, Chicago, IL, USA}
 \affiliation[a]{Supported by SNF under contract 168169.}
@@
 %  summarize existing training efforts to motivate that we already spend funding on
-% this. Discuss initiatives and programs which can be developed in the coming years
+% this. Discuss initiatives and programmes which can be developed in the coming years
 % including a discussion of the known training needs of the community. 
@@
 
-Throughout the last few decades, the field of high energy physics (HEP) has
-consistently either driven or utilized the latest innovations in computational
-tools and technologies for problem solving.
-The toolset of the particle physicist is vast and ever-growing, and the problem
-set increasingly complex.
-A typical graduate student or postdoc working on HEP experiments will likely
-encounter a number of tools, software, and programming languages with which they
-are totally unfamiliar.
-This requires ``on-the-job'' training for the same people whose work drives the
-reach of our analyses, while they are developing the tools for these analyses.
+The field of high energy physics (HEP) has consistently exploited 
+the latest innovations in computational tools and technologies for processing data.
+The software toolset of the particle physicist is ever-growing, and the problem
+set increasingly complex. Young researchers will likely encounter a number of 
+programming techniques with which they are totally unfamiliar, and will therefore 
+need "on-the-job" training in order for them to be productive.
 
-A strong research community should consist of independent actors who ask
+A strong research community comprises independent thinkers who ask
 research questions and direct their own analyses, instead of blindly following
-prescriptions and recipes. It is therefore imperative that the community be
-proactive about providing training resources and opportunities for the good of
-the community, and that we follow through on the promise that adequate career
+prescriptions and recipes. It is therefore imperative that training opportunities 
+and resources are made available, such that adequate career
 path possibilities exist for people in HEP who would otherwise leave the field
@@
 
-To produce the best possible science, it is important, for the broad HEP
-community, to invest in training physicists in essential software and software
-engineering skills, which exist at many different levels. The level of depth of
-software engineering competences in which a physicist should be trained will
-vary depending on the problems that they face, and the appropriate level may
-evolve over time according to the problem at hand. For example, for a researcher
-who only performs basic data analysis, it may be sufficient to be familiar with
-just Python and data analysis libraries. However, contributing to a large open
-source project, such as ROOT, requires greater training in software development
-and related practices. These higher-order skills may include an understanding of
-writing testable code, time and space complexity implications, and knowledge of
-common software architecture paradigms. The need for these skills cannot be
-fulfilled completely just by hiring specialists in computer science, since a
-thorough understanding of the physics problems is usually very important to the
-success of the effort, making it necessary to provide a career path for those
-young physicists willing to steer their interests in that particular direction.
+To produce the best possible science, it is important that physicists can easily acquire
+essential software engineering skills. The nature and level of the expertise a 
+physicist needs to acquire will vary according to the problems that need to be tackled. 
+For example, basic data analysis requires familiarity with the Python programming language 
+and with a broad range of data analysis libraries in current usage. However, those 
+wishing to contribute to large open source projects, such as ROOT, require greater 
+training in the use of software engineering practices that cover the full development 
+life-cycle. Clearly a thorough understanding of the physics problem domain is  
+essential, making it necessary to provide a career path for those
+young physicists willing to invest time and effort in becoming software specialists.
 
-In each HSF workshop, a discussion on the need for proper tutoring, training,
-and the relevant career possibilities in HEP computing emerged and was felt to
-be crucial and essential by almost all participants. There is a strong consensus
-that future software developments will benefit enormously from a successful
-implementation of such training programs, provided proper career prospects are
-put in place, as an incentive, to would-be trainers.
+In each HSF workshop, a consensus on the need for proper tutoring, training,
+and relevant career possibilities emerged as the basic ingredients for ensuring success.
+The main requirements and challenges to be faced were identified as follows:
 
-This document tries to address the most relevant questions related to the
-training in the HEP community, namely:
 \begin{itemize}
-    \item How to encourage and provide incentives to the members of the HEP 
-    community to train students and/or other collaborators?
-    \item How to properly assign credit to software development (and training 
-    thereof) activities?
-    \item What are the current practices in the field regarding hiring of 
-    researchers with a strong specialisation in computing and the policies 
-    to retain such an expertise in the long term? Is there room for improvement?
-    \item How to address and bridge the gap in formal software training that 
-    is not always imparted by universities?
+    \item to encourage and provide incentives to the members of the HEP 
+    community to train students and other collaborators,
+    \item to properly assign credit to software development as a scientific 
+    discipline, including training activities,
+    \item to establish policies for the hiring and long-term retention of researchers 
+    specialising computing,
+    \item to address the gap in formal software training that is not always 
+    given by universities as part of a physicist's education.
 \end{itemize}
@@
 %____________________________________________________________________
-\section{Motivation}
+\section{Motivation for some HSF coordinated actions}
 
@@
 
-The introduction and development of increasingly complex tools has led to
-increasing improvements in the performance of our analyses. However, the
-personnel who drive these advancements through novel implementations of the
-technology are the analyzers. In this context, the term {\it analyzers} includes
-all those researchers who participate in an HEP experiment whose contribution
-includes designing and implementing a high speed DAQ system, setting up a modern
-database to store the state of a detector, complex task of managing the data
-flow from the first trigger level to the final n-tuple dataset and creating
-sophisticated statistical and mathematical tools to extract publishable physical
-meaning from those same data.
+The introduction and development of sophisticated software tools has led to
+improvements in the performance of our data processing software. The
+personnel who drive these advances through novel implementations work on a broad 
+range of tasks, including the development of high speed DAQ systems, of modern databases for
+storing the state of a detector, the management of data flow from the first trigger 
+level to the final analysis dataset, as well as the development of algorithms and
+mathematical tools for extracting publishable physics results.
 
-The development of new tools is insufficient to drive advances in the field
-unless researchers who have the expertise apply them to physics questions.
-Furthermore, a larger and a better-connected community, with expertise in the
-tools and access to the new developments, can lead to an open-source model in
-HEP. The efforts to train the community in all relevant tools is, thus, a
-priority within the R\&D effort itself.
-
-However, as tools become more complex, the learning rate of developers must keep
-up if we are to utilize them effectively.  At the same time, as in other
-branches of research, it is essential to sustain efforts in maintaining
-reproducibility of the usage of the tools. This again highlights the need for
-physicts to be appropriately trained in computing science, rather than
-professional computing scientists, whose goals and interests may not necessarily
-be aligned or focused on the physics results of a particular HEP experiment.
-
-To avoid duplicate efforts within the community, it is important to find a way
-to approach training efforts holistically and cross-experimentally, whenever
-possible. This requires a certain level of organization and coordination in
-order to guarantee the delivery of efficient, non-redundant software to
-different experiments and to allow the different communities to agree on common
-standards and procedures.
+As tools become more complex, physicists must be continuously retrained in order
+to utilize them effectively. It is important to find a way to deploy training efforts
+holistically and cross-experimentally, whenever possible, in order to avoid unnecessary 
+duplication of the effort involved. This will require organization and coordination to
+ensure that different software teams can agree on common standards, thereby allowing different
+experiments to adopt the same olutions in an efficient way.
 
 Beyond the immediate benefit to the scientific community in having well-trained
-collaborators, most people who start a physics graduate program will have
+collaborators, most people who start a physics graduate programme will have
 careers outside of academia. This trend is reflected in several calls for
-proposals from national scientific funding agencies like H2020 and those from
-the National Science Foundation (NSF) that promote training in computer
-infrastructure and information areas.
+proposals from national scientific funding agencies that promote training in computer
+infrastructure areas, such as the EU's Horizon 2020 programme and those from
+the National Science Foundation (NSF) [Refs 1 and 2].
 
-%\textcolor{red}{this call is more to develop cyberinfrastructure staff than scientists - for example, people at computing centers}
-
-\begin{quote}
-\it The overarching goal of these programs is to prepare, nurture and grow the national scientific workforce for creating, utilizing, and supporting advanced cyberinfrastructure (CI) that enables cutting-edge science and engineering and contributes to the Nation's overall economic competiveness and security.
-
-{\bf - Training-based Workforce Development for Advanced Cyberinfrastructure (CyberTraining) Webinar}\footnote{ \url{https://www.nsf.gov/events/event_summ.jsp?cntn_id=190179&org=NSF}}
-\end{quote}
-
-\begin{quote}
-\it New information, communication, and computational technologies have had profound impacts on the practice of science (in this solicitation, the term science includes the natural, mathematical, computing, and social sciences),  engineering, and education.
-This includes the means by which citizens of all ages use science  and engineering to enhance professional and private lives.
-The systems, tools, and services emerging from these new technologies are linked to create a comprehensive cyberinfrastructure that is enabling individuals, groups, and organizations to advance research and education in ways that revolutionize who can participate, what they can do, and how they do it.
-Sustaining this revolution across all areas of science, engineering, and education requires the formation of a citizenry and workforce with the knowledge and skills needed to design and deploy as well as adopt and apply these cyber-based systems, tools and services over the long-term.
-The opportunity for such preparation should be available at all stages of formal and informal education (K-16 and lifelong), training and professional development, and must be extended to all individuals and communities.
-
-{\bf - Cyberinfrastructure Training, Education, Advancement, and Mentoring for Our 21st Century Workforce (CI-TEAM)}\footnote{ \url{https://www.nsf.gov/funding/pgm_summ.jsp?pims_id=12782}}
-\end{quote}
-
-An open issue is the kind of incentive that our community can collectively
-leverage in order to increase the number of people (with strong expertise in
-computing) who decide to dedicate time and effort to train fellow members of
-their community.
+An important challenge will be to discover incentives that will encourage 
+increasing numbers of people to dedicate time and effort to train their colleagues.
 To be effective, the training must be done by people who work at the cutting
@@
 researchers. These are also the very same people who are most in need of
-officially recognized credits for the advancement of their careers. The current
-practice in HEP is that these young scientists only receive credit and
-recognition (in the vast majority of cases) by working on a data analysis
-project, and this discourages them from finding time to transfer their acquired
-know-how to others in an effective way. A possible solution to this conundrum
-could be the establishment of specific career paths for researchers with a
-computer science specialization. The practical implementation of such an
-approach, however, is extremely challenging as there are a wide range of
+officially recognized credits for the advancement of their careers. Currently
+visibility and recognition is given only to those working on a data analysis project, 
+rather othan on the development and support of the underlying software. A possible 
+solution to this conundrum could be the establishment of specific career paths 
+for researchers specialising in scientific software development. However, the practical 
+implementation of such an approach is extremely challenging, as there are a wide range of
 institutions involved that belong to different countries with their own
-policies, priorities, and funding strategies. Broadly, two career paths could be
-envisaged. Both would recognise that domain-sepcific knowledge and continued
-engagement is vital to the success of HEP computing and software. While in an
-ideal world (for physicists, at least), a physicst could make a detailed plan of
-what they need (expressed in a tidy requirements document) and a computer
-scientist could use this document to provide the required solution, with very
-little additional exchange of information during the process (which is how the
-software development process sometimes works in industry). In the real HEP
-world, instead, most of the time the requirements are bound to change rather
-frequently for a variety of reasons (not only because of poor planning) forcing
-developers and physicsts to closely interact during the entire process of
+policies, priorities, and funding strategies.
+
+Broadly speaking, two scientific profiles of researchers could be
+envisaged; a physicst could make a detailed plan of what they need, expressed 
+in a tidy requirements document, and a computer scientist could use this document 
+to provide the required solution. This is how the software development process 
+typically works in the software industry. However, in HEP requirements are rapidly
+changing forcing developers and physicsts to closely interact during the entire process of
 software development; this works smoothly only when both communities have the
-same goals and speak the same jargon, hence the need for physicists with a
-computer science specialisation.
+same goals and speak the same language, hence the need for physicists with a good
+knowledge of computer science.
 
@@
 \emph{computer scientist}: the two have different goals, seek different paths to
-the solution of their problems, and usually do not even adopt the same jargon in
+the solution of their problems, and usually do not even adopt the same language in
 expressing problems and solutions. Such a role would have an active role in
@@
 researcher community, delivering a quick fix or a refactoring and then moving
-one. This does not deliver the domain specific expertise required.
+on. This does not deliver the domain specific expertise required.
 
@@
 experience, interests, and time availablity to learn new techniques.  Any
-training program must take into account the variation in target audiences. For
+training programme must take into account the variation in target audiences. For
 example, an undergraduate doing a summer research project has different needs
-and skills than their mentoring professor.
+and skills than their professor.
+
+\subsection{Classification of trainees}
+
 For purposes of training, it is useful to think of all our user community as
@@
    control, and command-line shells).
-   \item \emph{Intermediate users}: HEP analyzers with some experience in 
+   \item \emph{Intermediate level analysts}: people with some experience in 
    either analysis concepts and/or tools, but  looking to supplement their 
-   experiences with more recent and modern approaches. 
-   \item \emph{Advanced users}: HEP analysis experts who have mastered current 
-   technologies and implementations (at a certain point in time) and want to 
-   stay up-to-date with new advanced developments.
+   experience with more recent and modern approaches. 
+   \item \emph{Advanced level analysts}: experts who have mastered current 
+   technologies and implementations and who want to stay up-to-date with 
+   new advanced developments.
    \item \emph{Software developers}: HEP scientists in charge of software 
@@
 
-Each of these groups has vastly different training needs. However,
-\emph{whenever possible}, any training program should take advantage of
+Each of these groups has very different training needs. However,
+\emph{whenever possible}, any training programme should take advantage of
 developments in pedagogy, such as active
@@
 flawed, and ask students to fix or improve them. Learning the material in a way
-that sticks with the students is difficult and challenging for both the students
+that sticks is difficult and challenging for both the students
 and the instructor and often takes more time than we would prefer. However, this
-is the best way to ensure an educated community of scientists who can fully
-contribute to the physics programs at large, which is really the ultimate goal
-of any training program.
+is the best way to educate scientists who can fully
+contribute to the physics programmes at large, which is really the ultimate goal
+of any training programme.
 
@@
 to the whole HEP community and possibly beyond. In this section, we present a
-list of specific concepts that need to be taught to members of the community, in
-order to guarantee the base level of competence needed to write efficient code
+list of specific concepts that need to be taught, in
+order to guarantee the basic level of competence needed to write efficient code
 for the different tasks performed in HEP experiments.
@@
 include modularity of code design, advanced data structures, evaluation metrics,
-writing tests and working with different types of hardware accelerators.
+writing tests and working with different types of hardware accelerator.
 Cutting-edge approaches such as the use of hardware accelerators (whether GPUs
 or FPGAs), Deep Neural Networks, and Machine Learning should be considered
-important training topics. Additionally, special emphasis should be made on
+important training topics. Additionally, special emphasis should be given to
 reporting results and documenting them.
-The training subjects that the community considers important to pursue are
+The training subjects considered important to pursue are
 listed as follows:
@@
    \end{itemize}
-   \item Algorithmics 
+   \item Algorithms 
    \begin{itemize}
@@
    \end{itemize}
-   \item Existing frameworks (development or application level)
+   \item Frameworks (development or application level)
    \begin{itemize}
@@
       \item ROOT
-      \item experiment specific framework (possibily if of potential interest outside the native experiment)
+      \item experiment specific frameworks (possibly if of potential interest outside the originating experiment)
    \end{itemize}
@@
 %____________________________________________________________________
-\section{Challenges}
+\section{Implementation Roadmap}
+The implementation of training should target multiple training formats and knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and advanced visualizations, etc.
+People who learn in non-conventional ways should not be excluded from the
+opportunity to learn. Since there are different stages of knowledge transfer,
+programmes should be developed that encourage experts to share knowledge. Issues
+considered in this section include global training initiatives, strategies for
+leveraging existing forums, and training tool development.
 
-Within a single experiment, different skill sets are needed. In addition to a
-skill set that contains basic programming language knowledge, testing and code
-management tools, and experiment-specific framework knowledge, there are more
-specialized skills that only a subset of the community needs to know, such as
-software optimization, or low-level hardware interfaces.
+An important point to consider is the difficulty, often experienced in the past,
+in developing large software programs across different experiment communities.
 While there already are experiment-specific training efforts in place, there are
 many needs that are in common between experiments. Therefore, we should probably
-strive to extract the common knowledge and build a common training program. This
+strive to extract the common knowledge and build a common training programme. This
 would reduce duplication of efforts and enable growth of a shared training
@@
 aimed at improving the efficiency and cost-effectivness of this important
-activity.
-
-\section{Implementation Roadmap}
-The implementation of training should target multiple training formats and knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and advanced visualizations, etc.
-People who learn in non-conventional ways should not be excluded from the
-opportunity to learn. Since there are different stages of knowledge transfer,
-programs should be developed that encourage experts to share knowledge. Issues
-considered in this section include global training initiatives, strategies for
-leveraging existing forums, and training tool development.
-
-An important point to consider is the difficulty, often experienced in the past,
-in developing large software programs across different experiment communities.
-It is always challenging to organize a Working Group for training across
-experiments or even across diverse communities (such as HEP and Astroparticle,
-for example). This is a point where an appropriate incentive program to reward
-those who train the community might help.
+activity. An appropriate incentive programme to reward those who train the community 
+might help to facilitate the goals of such an initiative.
 
 %____________________________________________________________________
-\subsection{Initiatives for future training programs\label{sec:initiatives}}
+\subsection{Initiatives for future training programmes\label{sec:initiatives}}
 
 Some methods that can be used for global training include Massive Online Open
-Courses, hands-on workshops, online knowledge bases, expert trainer volunteer
+Courses (MOOCs), hands-on workshops, online knowledge bases, expert trainer volunteer
 networks, and web-based training approaches.
 
-Massive Online Open Courses (MOOCs) can be used to develop an open-source set of
-tutorials and tools. Existing online courses such as Udacity and Coursera can be
-evaluated and recommended to the community. A novel approach such as WikiToLearn
-could also explored to assess potential benefits (as has already been attempted
-in the context of the GridKa School of Computing, see below).
+MOOCs can be used to develop an open-source set of tutorials and tools. Existing 
+online courses such as Udacity and Coursera can be evaluated and recommended to 
+the community. A novel approach such as WikiToLearn
+could also explored to assess potential benefits, as has already been attempted
+in the context of the GridKa School of Computing (see below).
 
@@
 incentives for experts to contribute. They can be open-source so that a lot of
-knowledge can be added by the trainees themselves as information is learned
-(this is the particular context where an approach such as WikiToLearn could be
-of great help).
+knowledge can be added by the trainees themselves as information is learned;
+this is the particular context where an approach such as WikiToLearn could be
+of great help.
 
-Question-and-answer websites such as Stack Overflow for HEP are also a very
-useful resource for common problems and questions. This approach has already
+Question-and-answer websites, such as Stack Overflow for HEP, are also a very
+useful resource for addressing common problems and questions. This approach has already
 been considered by HSF start-up members, and turned out to be difficult to
@@
 the work done, at the expense of a broader training/education curriculum that
-could actually yield improved results, but further down the line. One incentive
-would be to provide training programs that also count as course credit, perhaps
+could actually yield improved results further down the line. One incentive
+would be to provide training programmes that also count as course credit, perhaps
 as an elective. This model is already in limited use with some online
@@
 courses are the lack of funding and the lack of available time by experts in the
-field. People with enough expertise or insight in the field usually don't have
+field. People with enough expertise or insight usually don't have
 time to devote to prolonged periods of student training, and, even when they can
@@
    \item Tutors can add material to the web site at a very slow pace (whenever
-   they find time to do it, one slide a week or a chapter a day) \item Their
-   material, made public on a web-site, can be further expanded by collaborators
+   they find time to do it, one slide a week or a chapter a day) 
+   \item Their material, made public on a web-site, can be further expanded by collaborators
    (also at their own pace) or, even better and more productively, by students
@@
    given time, creates a sense of community, and creates bridges between people
-   in contiguous areas or research. Students can use the same platform to
+   in contiguous areas of research. Students can use the same platform to
    exchange their own examples, make suggestions and point out interesting
@@
    set up in such a way that students will work in an exact replica of the
-   environment they will be exposed to in their experiment (or collaboration at
-   large). Moreover, students could be provided with Docker containers that
-   preserves their modified environment across sessions, allowing them to
+   environment they will be exposed to in their experiment. Moreover, students 
+   could be provided with Docker containers that
+   preserve their modified environment across sessions, allowing them to
    develop their skills over a prolonged period of time by accessing all the
@@
    sense it could be a {\it complementary} approach to a traditional school. Of
-   particular interest could be courses such as, e.g., ``Machine Learning'',
+   particular interest could be courses such as ``Machine Learning'',
    ``Statistical Analyis with ROOT'', or even just ``Good practices in C++'' or
@@
 to add material to the site, and to make it appear under a specific topic (such
-as, e.g., Software/Techniques/Machine-Learning) and to manipulate it as if it
+as Software/Techniques/Machine-Learning) and to manipulate it as if it
 were a single document. In the end, students can selectively choose individual
@@
 publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}.
-This constitutes an intersting example of what can be accomplished using this
+This constitutes an interesting example of what can be accomplished using this
 platform; it is just a first example of what's possible, but an inspiring one.
@@
 complementary approaches to training (such as schools and dedicated web-site
-courses) could be of mutual benefit (how to make them cooperate in the
-development of a complete training program).
+courses) could be of mutual benefit in the development of a complete training 
+programme.
 
@@
 
-Over the past decade, massive open online courses (MOOCs) have been developed by
-universities and private organizations. They have been well received by industry
-and academia.% and are maintained by the outside community.
-In addition, they provide a lot of flexibility in terms of cost and time
-constraints; they are typically free and open for enrollment at any time of the
+Over the past decade, MOOCs have been developed by universities and private organizations. 
+They have been well received by industry and academia.
+In addition, they provide a lot of flexibility in terms of cost and use of time; 
+they are typically free and open for enrollment at any time of the
 year. Since the material can be accessed at any time and revisited at any time,
@@
 specific tools taught. Exploring these options allows us to choose which is the
-right offering for the knowledge needed to work on a specific HEP experiment. We
+right offering for the knowledge needed to work on a specific experiment. We
 can pick and choose modules to tailor an appropriate roadmap of skills to learn.
-More difficult will be to assemble HEP specific training material not already
+More difficult will be to assemble specific training material not already
 avaible elsewhere in an efficient and organized way, since this requires
@@
 
-Training the community is something that a large cross-section of the community
+Training is something that a large cross-section of the community
 understands to be important, but finding time and effort to contribute to this
@@
 
-Software Carpentry\footnote{\url{https://software-carpentry.org}} and Data
-Carpentry\footnote{\url{http://www.datacarpentry.org}} are two parts of The
+Software Carpentry (SC) \footnote{\url{https://software-carpentry.org}} and Data
+Carpentry (DC) \footnote{\url{http://www.datacarpentry.org}} are two parts of The
 Carpentries organization that collaboratively build and teach some of the basic
@@
 developed specifically for this course and not a part of a larger Software
-Carpentry program. This course focuses on a shallow but wide building of
+Carpentry programme. This course focuses on a shallow but wide building of
 foundational skills approach, introducing many base concepts but not going deep
@@
 We have identified a certain number of problems that need to be overcome in
-setting up an appropriate training program:
+setting up an appropriate training programme:
 \begin{itemize}
@@
 %
-%   \item How to motivate the community to participate in developing programs? Can some of it come from \& leadership?
-%   \item How to deal with competitiveness in the field? Can’t force people to be interested in forming a community. How can we incentivize this? Building a %vibrant, welcoming community may help.
+%   \item How to motivate the community to participate in developing 
+programs? Can some of it come from \& leadership?
+%   \item How to deal with competitiveness in the field? Can’t force 
+people to be interested in forming a community. How can we incentivize this? 
+Building a %vibrant, welcoming community may help.
 %\end{itemize}

### 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Adding author/institute, and formatting (#85)
The following changes were made just to the author list:
* I added myself "Christopher Tunnell" and my institute "University of Chicago".
* Furthermore, there were no commas after a few other names toward the end that I added for consistency
* The "and" I assume happens with the last name in alphabetical order.  Since "Tunnell" is after "Stewart", I moved the 'and' to my name as the last alphabetically.
Mention IN2P3 gitlab set of tutorials.

@@
 \usepackage{xspace}
+\usepackage[titletoc]{appendix}
 
@@
 \abstract{The rapid evolution of technology and the parallel increasing 
-complexity of algorithmic analysis in HEP require developers 
+complexity of algorithmic analysis in HEP requires developers 
 to acquire a much larger portfolio of programming skills. Young 
@@
 preparation in the very diverse fields of modern computing to respond to growing needs 
-of the most advanced experimental challenges. There is a growing cnsensus in the HEP community 
-on the need for training programmes to bring researchers up to date with new software technologies, 
+of the most advanced experimental challenges. There is a growing consensus in
+the HEP community on the need for training programmes to bring researchers up to date with new software technologies, 
 in particular in the domains of concurrent programming and artificial intelligence. We review some
@@
  & & HSF-CWP-2017-02 \\
- & & June 6, 2018 \\ % use \date or hardwire e.g. December 15, 2017
+ & & June 20, 2018 \\ % use \date or hardwire e.g. December 15, 2017
  & & \\
@@
 
-\tableofcontents
 \clearpage
@@
 programming techniques with which they are totally unfamiliar, and will therefore 
-need "on-the-job" training in order for them to be productive.
+need ``on-the-job'' training in order for them to be productive.
 
-A strong research community comprises independent thinkers who ask
+Evidently a strong research community comprises independent thinkers who ask
 research questions and direct their own analyses, instead of blindly following
@@
 %____________________________________________________________________
-\section{Motivation for some HSF coordinated actions}
+\section{Career Paths for Software Experts}
 
@@
 ensure that different software teams can agree on common standards, thereby allowing different
-experiments to adopt the same olutions in an efficient way.
+experiments to adopt the same solutions in an efficient way.
 
@@
 infrastructure areas, such as the EU's Horizon 2020 programme and those from
-the National Science Foundation (NSF) [Refs 1 and 2].
+the National Science Foundation (NSF) \cite{NSF1,NSF2}.
 
-An important challenge will be to discover incentives that will encourage 
-increasing numbers of people to dedicate time and effort to train their colleagues.
+A key goal is to find incentives that will encourage 
+increasing numbers of people to dedicate their time and effort to train their
+colleagues.
+
 To be effective, the training must be done by people who work at the cutting
@@
 officially recognized credits for the advancement of their careers. Currently
-visibility and recognition is given only to those working on a data analysis project, 
-rather othan on the development and support of the underlying software. A possible 
-solution to this conundrum could be the establishment of specific career paths 
-for researchers specialising in scientific software development. However, the practical 
+visibility and recognition is given mainly to those working on data analysis
+projects, rather than to the development and support of the underlying
+software. One solution to this problem would be the establishment of specific
+career paths for researchers specialising in scientific software development. 
+%% GS Comment - is this actually ``motivation'' - it seems more like a solution
+% See suggested title change for this section...
+However, the practical 
 implementation of such an approach is extremely challenging, as there are a wide range of
@@
 typically works in the software industry. However, in HEP requirements are rapidly
-changing forcing developers and physicsts to closely interact during the entire process of
-software development; this works smoothly only when both communities have the
+changing, forcing developers and physicsts to interact closely during the entire
+process of software development; this works smoothly only when both communities have the
 same goals and speak the same language, hence the need for physicists with a good
@@
 make an academic appointment challenging or disfavoured compared to other
-applicants. However, there are measures that an be taken to address all three
+applicants. However, there are measures that can be taken to address all three
 weaknesses.
+%% GS Comment - This is an odd place to end a section. What are these measures?
+% Where do we list them? Even a forward reference would help.
 
@@
 \emph{students}. More specifically, we can broadly identify four different sets
-of students, classified by experience level, recognizing that even within these
-groups there will be people who will learn at different speeds and with
-differing depths of understanding:
+of students, classified by experience level:
 
@@
    \item \emph{Beginners}: New collaborators with no knowledge of the 
-   tools or techniques they are expected to use. These might be undergraduates 
-   but they could also be senior researchers who have not yet or not recently 
+   tools or techniques they are expected to use. These might be undergraduates, 
+   but they could also be senior researchers who have not yet, or not recently, 
    engaged in software analysis. These are people in need of some kind of formal 
    training in modern computing techniques (basically compiled and scripting 
-   languages together with operating system basics such as filesystems, version 
+   languages, together with operating system basics such as filesystems, version 
    control, and command-line shells).
@@
 developments in pedagogy, such as active
-learning\footnote{\url{http://www.crlt.umich.edu/tstrategies/tsal}}, peer
-learning\footnote{\url{https://en.wikipedia.org/wiki/Peer_learning}}, and
+learning~\cite{ActiveLearning}, peer learning~\cite{PeerLearning}, and
 web-based training.\footnote{See \S\ref{sec:initiatives}.} In some cases, it may
@@
 At all stages of software and computing training, we should take care to
-encourage \emph{Good Practices Across the Community} (GPAC), such as error
+encourage \emph{Good Practices Across the Community}, such as error
 checking, modularity of code design, version control, writing tests, etc. All
@@
 particular experiment or field of application, but general enough to be useful
-to the whole HEP community and possibly beyond. In this section, we present a
+to the whole HEP community and beyond. In this section, we present a
 list of specific concepts that need to be taught, in
@@
 
-An essential knowledge base that needs to be transferred includes basic
+Essential knowledge that needs to be transferred includes basic
 programming concepts, data structures, basics of code design, error checking,
@@
 writing tests and working with different types of hardware accelerator.
-Cutting-edge approaches such as the use of hardware accelerators (whether GPUs
-or FPGAs), Deep Neural Networks, and Machine Learning should be considered
-important training topics. Additionally, special emphasis should be given to
+Additionally, special emphasis should be given to
 reporting results and documenting them.
-The training subjects considered important to pursue are
-listed as follows:
 
-\begin{itemize}
-   \item Basic and Advanced Programming Concepts
-   \begin{itemize}
-      \item Object oriented paradigm
-      \item Compiled langauges (C++)
-      \item Scripting languages (Python, Javascript,...)
-   \end{itemize}
-   \item Algorithms 
-   \begin{itemize}
-      \item Boost library
-      \item STL algorithms for containers
-      \item R and/or ROOT
-   \end{itemize}
-   \item Frameworks (development or application level)
-   \begin{itemize}
-      \item Qt
-      \item ROOT
-      \item experiment specific frameworks (possibly if of potential interest outside the originating experiment)
-   \end{itemize}
-   \item Code design (design patterns)
-   \item Development tools
-   \begin{itemize}
-      \item IDEs (Integrated Development Environment)
-      \item Debuggers
-      \item Profilers
-   \end{itemize}
-   \item Evaluation metrics
-   \item “Trust” metrics such as data driven tests
-   \item Specific software implementation training
-   \item Good practices 
-   \item Code style and clarity
-   \item Scripting and data cleaning
-   \item Reporting results reproducibly
-   \item Writing Documentation
-\end{itemize}
+Some of the training subjects considered important to pursue are
+listed in Appendix \ref{appendix-a---training-topics-of-interest}.
 
-%____________________________________________________________________
-%____________________________________________________________________
+% ____________________________________________________________________
+% ____________________________________________________________________
 \section{Implementation Roadmap}
-The implementation of training should target multiple training formats and knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and advanced visualizations, etc.
+The implementation of training should target multiple training formats and
+knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and
+advanced visualizations, etc.
 People who learn in non-conventional ways should not be excluded from the
-opportunity to learn. Since there are different stages of knowledge transfer,
+opportunity to learn. 
+%% GS Comment - I'm not really sure what that previous sentence is supposed to
+% mean.
+Since there are different stages of knowledge transfer,
 programmes should be developed that encourage experts to share knowledge. Issues
@@
 
-MOOCs can be used to develop an open-source set of tutorials and tools. Existing 
-online courses such as Udacity and Coursera can be evaluated and recommended to 
-the community. A novel approach such as WikiToLearn
-could also explored to assess potential benefits, as has already been attempted
-in the context of the GridKa School of Computing (see below).
+MOOCs can be used to develop an open-source set of
+tutorials and tools. Existing online courses such as Udacity~\cite{Udacity} and 
+Coursera~\cite{Coursera} can be
+evaluated and recommended to the community. A novel approach such as WikiToLearn~\cite{WikiToLearn}
+could also be explored to assess potential benefits (as has already been attempted
+in the context of the GridKa School of Computing~\cite{GridKa}, see below).
 
@@
 
-Question-and-answer websites, such as Stack Overflow for HEP, are also a very
-useful resource for addressing common problems and questions. This approach has already
+Question-and-answer websites such as Stack Overflow~\cite{Stackoverflow} for HEP are also a very
+useful resource for common problems and questions. This approach has already
 been considered by HSF start-up members, and turned out to be difficult to
@@
 possible tutors are, in principle, those people who are actively engaged in
-modern software developments: these are in general young researchers in the
+modern software developments: these are often young researchers in the
 first steps of their careers and, in order to be attracted to devoting
@@
 is not currently implemented, at least not in a standardized or official way. A
-possible structure for such a network could be the establishment of a {\em
-federation} of existing schools.
+possible structure for such a network could be the establishment of a 
+{\em federation} of existing schools.
 
@@
 as an elective. This model is already in limited use with some online
-solutions\footnote{\url{https://www.edx.org/credit}}, but this is not universal.
+solutions~\cite{TrainingProgram}, but this is not universal.
 Discussions should be started with collaborators at higher education
@@
 training (but complementary to the already existing and successful classical
-efforts such as the CERN School of Computing's Bertinoro and Kit ones): instead
+efforts such as the CERN School of Computing's Bertinoro and KIT ones): instead
 of directly teaching to students, trainees could make use of a web-based
@@
    they find time to do it, one slide a week or a chapter a day) 
-   \item Their material, made public on a web-site, can be further expanded by collaborators
-   (also at their own pace) or, even better and more productively, by students
+   \item Their material, available public on a web-site, can be further expanded
+   by collaborators (also at their own pace) or, even better and more productively, by students
    who decide to contribute new additions, examples, exercises etc. Such a
@@
 implemented as an Open Source project (backed by Wikimedia) by a group of more
-than 30 Italian students. The project is WikiToLearn
-\footnote{\url{https://it.wikitolearn.org/}}. It hosts training material in
+than 30 Italian students. The project is WikiToLearn~\cite{WikiToLearn}. It hosts training material in
 several languages, for several disciplines, ranging from Economics to Physics,
 Mathematics, and several others. Because it is based upon
-wikimedia\footnote{\url{https://www.wikimedia.org/}} software, it is very easy
+wikimedia~\cite{WikiMedia} software, it is very easy
 to add material to the site, and to make it appear under a specific topic (such
@@
 training material using just a web-browser, and in order to do this efficiently,
-the learning curve has been kept appropriately shallow.
-
+the necessary learning curve has been kept appropriately shallow.
 An interesting exercise in this context has recently been made by colleagues of
+the GridKa School of computing: the material from this year (2017) has been made
+publicly available on WTL~\cite{WikiToLearnGridka}.
+This constitutes an intersting example of what can be accomplished using this
+the necessary learning curve has been kept appropriately shallow. An interesting exercise in
+this context has recently been made by colleagues of
 the GridKa School of computing: the material from this year (2017) has been made
@@
 
+Another example of web-based platform is a collection of online tutorials
+(mostly written in french) hosted on the Gitlab IN2P3
+server{\url{https://gitlab.in2p3.fr/MaitresNageurs}}. Taking profit of
+the Gitlab ability to host Docker images, those tutorials aims to avoid the
+"damned installation step" that often absorbs half of a training session.
+Similarly to the GridKa School for WikiToLearn, the annual IN2P3 Computing
+Days are an opportunity to refresh and extend the collection of
+tutorials every year. Future work will focus on english translation, and the
+development of a web site which will index the above tutorials, together with
+the best recommended external tutorials.
+
 Finally it should be important to evaluate, if and to which extent, the
-complementary approaches to training (such as schools and dedicated web-site
-courses) could be of mutual benefit in the development of a complete training 
-programme.
+complementary approaches to training, such as schools and dedicated web-site
+courses, could be of mutual benefit, in other words how to make them efficiently cooperate in the
+development of a complete training program.
 
@@
 Hands-on Advanced Tutorial Sessions
-(HATS)\footnote{\url{http://lpc.fnal.gov/programs/schools-workshops/}}
+(HATS)~\cite{LPCHandsOn}
 throughout the year to introduce and train participants in topics as diverse as
@@
 exercises. A similar approach is in use in the CMS Data Analysis Schools
-(CMSDAS)\footnote{\url{http://lpc.fnal.gov/programs/schools-workshops/cmsdas.shtml}},
+(CMSDAS)~\cite{LPCDataAnalysis},
 a series of week-long workshops that now take place at multiple labs all over
@@
 
-Other examples are CERN School of
-Computing\footnote{\url{https://csc.web.cern.ch/}}, CERN OpenLab Software
-workshops\footnote{\url{http://openlab.cern/}} education in collaboration with
+Other examples are CERN School of Computing~\cite{CERNSchoolOfComputing}, 
+CERN OpenLab Software workshops~\cite{OPENLab} education in collaboration with
 industry partners, and a series of more advanced topical training courses
@@
 list includes the
-Bertinoro\footnote{\url{https://web.infn.it/esc17/index.php?lang=en}}, 
-GridKa\footnote{\url{http://gridka-school.scc.kit.edu/2017/}}, and
-CODAS-HEP\footnote{\url{http://codas-hep.org}} Schools of Computing.
+Bertinoro~\cite{Bertinoro}, 
+GridKa~\cite{GridKa}, and
+CODAS-HEP~\cite{CODAS-HEP} Schools of Computing.
 
@@
 avaible elsewhere in an efficient and organized way, since this requires
-adequate organization, volounteers and and a suitable infrastructure.
+adequate organization, volounteers and a suitable infrastructure.
 
@@
 and industry who are at the cutting edge of these techniques. Conferences such
-as NIPS\footnote{\url{https://nips.cc}} and
-PyData\footnote{\url{https://pydata.org}} provide a focused place where
+as NIPS~\cite{NIPS} and
+PyData~\cite{PyData} provide a focused place where
 attendees can learn a lot about machine learning in a short period of time.
@@
 For example, at the time of this writing,
-Coursera\footnote{\url{https://www.coursera.org/learn/machine-learning}} and
-Udacity\footnote{\url{https://www.udacity.com/course/intro-to-machine-learning--ud120}}
+Coursera~\cite{Coursera} and
+Udacity~\cite{Udacity}
 both provide great machine learning massive open online courses at no cost.
-These two courses both provide a great foundation for learning machine learning
+These two courses both provide a great foundation for assimilating machine learning
 fundamentals. However, Coursera's approach emphasizes more theory (with more
 math background necessary) and uses MATLAB/Octave while Udacity's approach
-emphasizes practical using machine learning techniques in a practical sense
-using Python tooling.
+emphasizes more practical aspects using machine learning techniques 
+using Python tools.
 
@@
 understands to be important, but finding time and effort to contribute to this
-effort is not actively on the radar of most potential contributors. Providing
+project is not actively on the radar of most potential volunteers. Providing
 incentives for their participation and creating the appropriate platforms can go
@@
 be connected with other areas such as research and development. Efforts like
-DIANA-HEP \footnote{\url{http://diana-hep.org/}} and aMVA4NewPhysics
-\footnote{\url{https://amva4newphysics.wordpress.com/}} that combine training
+DIANA-HEP~\cite{DIANA-HEP} and AMVA4NewPhysics~\cite{AMVA4NewPhysics}
+that combine training
 and software development are good examples of such ideas in practice. More
@@
 
-Software Carpentry (SC) \footnote{\url{https://software-carpentry.org}} and Data
-Carpentry (DC) \footnote{\url{http://www.datacarpentry.org}} are two parts of The
+Software Carpentry~\cite{SoftwareCarpentry} and Data
+Carpentry~\cite{DataCarpentry} are two parts of The
 Carpentries organization that collaboratively build and teach some of the basic
@@
 names.
-Subjects covered in a typical Data Carpentry school are given below. This list
-is taken from the curriculum of the CODATA-RDA Research Data Science Summer
-School in progress in Trieste, Italy during July 2017.
-(http://indico.ictp.it/event/7974/):
-
-\begin{itemize}
-\item Introduction
-\item UNIX Shell programming (Software Carpentry Module)
-\item GitHub (Software Carpentry Module)
-\item R (Software Carpentry Module)
-\item BYOD (Bring Your Own Data) best practices
-\item Data Management with SQL  (Software Carpentry Module)
-\item RDM Storage Management
-\item Visualisation
-\item Machine Learning Overview - Recommendation
-\item Recommender Systems
-\item Artificial Neural Networks and other Machine Learning Systems 
-\item Research Computational Infrastructure 
-\end{itemize}
+Subjects covered in a typical Data Carpentry school are given in Appendix
+\ref{appendix-b}.
 
@@
 
-%____________________________________________________________________
-%____________________________________________________________________
-%\begin{itemize}
-%
-%   \item How to motivate the community to participate in developing 
-programs? Can some of it come from \& leadership?
-%   \item How to deal with competitiveness in the field? Can’t force 
-people to be interested in forming a community. How can we incentivize this? 
-Building a %vibrant, welcoming community may help.
-%\end{itemize}
+%% GS Comment - this section is incomplete. FIXME.
+
+\newpage
+\begin{appendices}
+
+\sloppy
+\raggedright
+
+\hypertarget{appendix-a---training-topics-of-interest}{%
+\section{Training Topics of Interest}\label{appendix-a---training-topics-of-interest}}
+
+\begin{itemize}
+   \item Basic and Advanced Programming Concepts
+   \begin{itemize}
+      \item Object oriented paradigm
+      \item Compiled langauges (C++)
+      \item Scripting languages (Python, Javascript,...)
+   \end{itemize}
+   \item Algorithms 
+   \begin{itemize}
+      \item Boost library
+      \item STL algorithms for containers
+      \item R and/or ROOT
+   \end{itemize}
+   \item Frameworks (development or application level)
+   \begin{itemize}
+      \item Qt
+      \item ROOT
+      \item experiment specific frameworks (possibly if of potential interest outside the originating experiment)
+   \end{itemize}
+   \item Code design (design patterns)
+   \item Development tools
+   \begin{itemize}
+      \item IDEs (Integrated Development Environment)
+      \item Debuggers
+      \item Profilers
+   \end{itemize}
+   \item Evaluation metrics
+   \item “Trust” metrics such as data driven tests
+   \item Specific software implementation training
+   \item Good practices 
+   \item Code style and clarity
+   \item Scripting and data cleaning
+   \item Reporting results reproducibly
+   \item Writing Documentation
+\end{itemize}
+
+\hypertarget{appendix-b}{%
+\section{Research Data Science Curriculum}\label{appendix-b}}
+
+This list is taken from the curriculum of the CODATA-RDA Research Data
+Science Summer School in progress in Trieste, Italy during July 2017.
+(http://indico.ictp.it/event/7974/):
+
+\begin{itemize}
+\item Introduction
+\item UNIX Shell programming (Software Carpentry Module)
+\item GitHub (Software Carpentry Module)
+\item R (Software Carpentry Module)
+\item BYOD (Bring Your Own Data) best practices
+\item Data Management with SQL  (Software Carpentry Module)
+\item RDM Storage Management
+\item Visualisation
+\item Machine Learning Overview - Recommendation
+\item Recommender Systems
+\item Artificial Neural Networks and other Machine Learning Systems 
+\item Research Computational Infrastructure 
+\end{itemize}
+
+\end{appendices}
 
@@
 
-\printbibliography[title={references}]
-
 \end{document}

### 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Tutorial of the month.

@@
 the necessary learning curve has been kept appropriately shallow.
+
 An interesting exercise in this context has recently been made by colleagues of
@@
 This constitutes an intersting example of what can be accomplished using this
-the necessary learning curve has been kept appropriately shallow. An interesting exercise in
-this context has recently been made by colleagues of
-the GridKa School of computing: the material from this year (2017) has been made
-publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}.
-This constitutes an interesting example of what can be accomplished using this
 platform; it is just a first example of what's possible, but an inspiring one.
-
-Another example of web-based platform is a collection of online tutorials
-(mostly written in french) hosted on the Gitlab IN2P3
-server{\url{https://gitlab.in2p3.fr/MaitresNageurs}}. Taking profit of
-the Gitlab ability to host Docker images, those tutorials aims to avoid the
-"damned installation step" that often absorbs half of a training session.
-Similarly to the GridKa School for WikiToLearn, the annual IN2P3 Computing
-Days are an opportunity to refresh and extend the collection of
-tutorials every year. Future work will focus on english translation, and the
-development of a web site which will index the above tutorials, together with
-the best recommended external tutorials.
 
@@
 endeavors.
+
+In the context of web-based training, if HSF helps to constitute a living
+collection of online tutorials, we could organize regular events such as
+a "Tutorial of the month", based on students votes. From a more general point
+of view, since the publications are the most universal recognized items
+in the research community, we should try hardly to create some sort of
+training journal, with reviewed training papers.
 

### 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Tutorial of the month.

--- 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 
-\abstract{The rapid evolution of technology and the parallel increasing 
-complexity of algorithmic analysis in HEP require developers 
-to acquire a much larger portfolio of programming skills. Young 
-researchers graduating from universities worldwide currently do not receive adequate 
-preparation in the very diverse fields of modern computing to respond to growing needs 
-of the most advanced experimental challenges. There is a growing cnsensus in the HEP community 
-on the need for training programmes to bring researchers up to date with new software technologies, 
-in particular in the domains of concurrent programming and artificial intelligence. We review some
-of the initiatives under way for introducing new training programmes and highlight some of the issues 
-that need to be taken into account for these to be successful.}
+\abstract{The rapidly increasing evolution of technology and the parallel increasing complexity of algorithmic analysis in HEP (and in science in general) require a much larger portfolio of specific skills by the developers with respect to the past.
+Young researchers, graduating from universities worldwide, currently do not receive adequate preparation in the very diverse fields of modern computing to respond to growing necessities posed by the most advanced experimental challenges.
+The vast majority of the HEP community strongly agrees on the need for an overarching umbrella of training programs to bring young researchers (but also more mature ones) up to date with new software technologies, such as, for example, concurrent programming and artificial intelligence.
+}
 
@@
 \author[3]{Peter Elmer,}
-\author[1]{John Harvey,}
 \author[4]{Roger Jones,}
@@
 %  summarize existing training efforts to motivate that we already spend funding on
-% this. Discuss initiatives and programmes which can be developed in the coming years
+% this. Discuss initiatives and programs which can be developed in the coming years
 % including a discussion of the known training needs of the community. 
@@
 
-The field of high energy physics (HEP) has consistently exploited 
-the latest innovations in computational tools and technologies for processing data.
-The software toolset of the particle physicist is ever-growing, and the problem
-set increasingly complex. Young researchers will likely encounter a number of 
-programming techniques with which they are totally unfamiliar, and will therefore 
-need "on-the-job" training in order for them to be productive.
+Throughout the last few decades, the field of high energy physics (HEP) has
+consistently either driven or utilized the latest innovations in computational
+tools and technologies for problem solving.
+The toolset of the particle physicist is vast and ever-growing, and the problem
+set increasingly complex.
+A typical graduate student or postdoc working on HEP experiments will likely
+encounter a number of tools, software, and programming languages with which they
+are totally unfamiliar.
+This requires ``on-the-job'' training for the same people whose work drives the
+reach of our analyses, while they are developing the tools for these analyses.
 
-A strong research community comprises independent thinkers who ask
+A strong research community should consist of independent actors who ask
 research questions and direct their own analyses, instead of blindly following
-prescriptions and recipes. It is therefore imperative that training opportunities 
-and resources are made available, such that adequate career
+prescriptions and recipes. It is therefore imperative that the community be
+proactive about providing training resources and opportunities for the good of
+the community, and that we follow through on the promise that adequate career
 path possibilities exist for people in HEP who would otherwise leave the field
@@
 
-To produce the best possible science, it is important that physicists can easily acquire
-essential software engineering skills. The nature and level of the expertise a 
-physicist needs to acquire will vary according to the problems that need to be tackled. 
-For example, basic data analysis requires familiarity with the Python programming language 
-and with a broad range of data analysis libraries in current usage. However, those 
-wishing to contribute to large open source projects, such as ROOT, require greater 
-training in the use of software engineering practices that cover the full development 
-life-cycle. Clearly a thorough understanding of the physics problem domain is  
-essential, making it necessary to provide a career path for those
-young physicists willing to invest time and effort in becoming software specialists.
+To produce the best possible science, it is important, for the broad HEP
+community, to invest in training physicists in essential software and software
+engineering skills, which exist at many different levels. The level of depth of
+software engineering competences in which a physicist should be trained will
+vary depending on the problems that they face, and the appropriate level may
+evolve over time according to the problem at hand. For example, for a researcher
+who only performs basic data analysis, it may be sufficient to be familiar with
+just Python and data analysis libraries. However, contributing to a large open
+source project, such as ROOT, requires greater training in software development
+and related practices. These higher-order skills may include an understanding of
+writing testable code, time and space complexity implications, and knowledge of
+common software architecture paradigms. The need for these skills cannot be
+fulfilled completely just by hiring specialists in computer science, since a
+thorough understanding of the physics problems is usually very important to the
+success of the effort, making it necessary to provide a career path for those
+young physicists willing to steer their interests in that particular direction.
 
-In each HSF workshop, a consensus on the need for proper tutoring, training,
-and relevant career possibilities emerged as the basic ingredients for ensuring success.
-The main requirements and challenges to be faced were identified as follows:
+In each HSF workshop, a discussion on the need for proper tutoring, training,
+and the relevant career possibilities in HEP computing emerged and was felt to
+be crucial and essential by almost all participants. There is a strong consensus
+that future software developments will benefit enormously from a successful
+implementation of such training programs, provided proper career prospects are
+put in place, as an incentive, to would-be trainers.
 
+This document tries to address the most relevant questions related to the
+training in the HEP community, namely:
 \begin{itemize}
-    \item to encourage and provide incentives to the members of the HEP 
-    community to train students and other collaborators,
-    \item to properly assign credit to software development as a scientific 
-    discipline, including training activities,
-    \item to establish policies for the hiring and long-term retention of researchers 
-    specialising computing,
-    \item to address the gap in formal software training that is not always 
-    given by universities as part of a physicist's education.
+    \item How to encourage and provide incentives to the members of the HEP 
+    community to train students and/or other collaborators?
+    \item How to properly assign credit to software development (and training 
+    thereof) activities?
+    \item What are the current practices in the field regarding hiring of 
+    researchers with a strong specialisation in computing and the policies 
+    to retain such an expertise in the long term? Is there room for improvement?
+    \item How to address and bridge the gap in formal software training that 
+    is not always imparted by universities?
 \end{itemize}
@@
 %____________________________________________________________________
-\section{Motivation for some HSF coordinated actions}
+\section{Motivation}
 
@@
 
-The introduction and development of sophisticated software tools has led to
-improvements in the performance of our data processing software. The
-personnel who drive these advances through novel implementations work on a broad 
-range of tasks, including the development of high speed DAQ systems, of modern databases for
-storing the state of a detector, the management of data flow from the first trigger 
-level to the final analysis dataset, as well as the development of algorithms and
-mathematical tools for extracting publishable physics results.
+The introduction and development of increasingly complex tools has led to
+increasing improvements in the performance of our analyses. However, the
+personnel who drive these advancements through novel implementations of the
+technology are the analyzers. In this context, the term {\it analyzers} includes
+all those researchers who participate in an HEP experiment whose contribution
+includes designing and implementing a high speed DAQ system, setting up a modern
+database to store the state of a detector, complex task of managing the data
+flow from the first trigger level to the final n-tuple dataset and creating
+sophisticated statistical and mathematical tools to extract publishable physical
+meaning from those same data.
 
-As tools become more complex, physicists must be continuously retrained in order
-to utilize them effectively. It is important to find a way to deploy training efforts
-holistically and cross-experimentally, whenever possible, in order to avoid unnecessary 
-duplication of the effort involved. This will require organization and coordination to
-ensure that different software teams can agree on common standards, thereby allowing different
-experiments to adopt the same olutions in an efficient way.
+The development of new tools is insufficient to drive advances in the field
+unless researchers who have the expertise apply them to physics questions.
+Furthermore, a larger and a better-connected community, with expertise in the
+tools and access to the new developments, can lead to an open-source model in
+HEP. The efforts to train the community in all relevant tools is, thus, a
+priority within the R\&D effort itself.
+
+However, as tools become more complex, the learning rate of developers must keep
+up if we are to utilize them effectively.  At the same time, as in other
+branches of research, it is essential to sustain efforts in maintaining
+reproducibility of the usage of the tools. This again highlights the need for
+physicts to be appropriately trained in computing science, rather than
+professional computing scientists, whose goals and interests may not necessarily
+be aligned or focused on the physics results of a particular HEP experiment.
+
+To avoid duplicate efforts within the community, it is important to find a way
+to approach training efforts holistically and cross-experimentally, whenever
+possible. This requires a certain level of organization and coordination in
+order to guarantee the delivery of efficient, non-redundant software to
+different experiments and to allow the different communities to agree on common
+standards and procedures.
 
 Beyond the immediate benefit to the scientific community in having well-trained
-collaborators, most people who start a physics graduate programme will have
+collaborators, most people who start a physics graduate program will have
 careers outside of academia. This trend is reflected in several calls for
-proposals from national scientific funding agencies that promote training in computer
-infrastructure areas, such as the EU's Horizon 2020 programme and those from
-the National Science Foundation (NSF) [Refs 1 and 2].
+proposals from national scientific funding agencies like H2020 and those from
+the National Science Foundation (NSF) that promote training in computer
+infrastructure and information areas.
 
-An important challenge will be to discover incentives that will encourage 
-increasing numbers of people to dedicate time and effort to train their colleagues.
+%\textcolor{red}{this call is more to develop cyberinfrastructure staff than scientists - for example, people at computing centers}
+
+\begin{quote}
+\it The overarching goal of these programs is to prepare, nurture and grow the national scientific workforce for creating, utilizing, and supporting advanced cyberinfrastructure (CI) that enables cutting-edge science and engineering and contributes to the Nation's overall economic competiveness and security.
+
+{\bf - Training-based Workforce Development for Advanced Cyberinfrastructure (CyberTraining) Webinar}\footnote{ \url{https://www.nsf.gov/events/event_summ.jsp?cntn_id=190179&org=NSF}}
+\end{quote}
+
+\begin{quote}
+\it New information, communication, and computational technologies have had profound impacts on the practice of science (in this solicitation, the term science includes the natural, mathematical, computing, and social sciences),  engineering, and education.
+This includes the means by which citizens of all ages use science  and engineering to enhance professional and private lives.
+The systems, tools, and services emerging from these new technologies are linked to create a comprehensive cyberinfrastructure that is enabling individuals, groups, and organizations to advance research and education in ways that revolutionize who can participate, what they can do, and how they do it.
+Sustaining this revolution across all areas of science, engineering, and education requires the formation of a citizenry and workforce with the knowledge and skills needed to design and deploy as well as adopt and apply these cyber-based systems, tools and services over the long-term.
+The opportunity for such preparation should be available at all stages of formal and informal education (K-16 and lifelong), training and professional development, and must be extended to all individuals and communities.
+
+{\bf - Cyberinfrastructure Training, Education, Advancement, and Mentoring for Our 21st Century Workforce (CI-TEAM)}\footnote{ \url{https://www.nsf.gov/funding/pgm_summ.jsp?pims_id=12782}}
+\end{quote}
+
+An open issue is the kind of incentive that our community can collectively
+leverage in order to increase the number of people (with strong expertise in
+computing) who decide to dedicate time and effort to train fellow members of
+their community.
 To be effective, the training must be done by people who work at the cutting
@@
 researchers. These are also the very same people who are most in need of
-officially recognized credits for the advancement of their careers. Currently
-visibility and recognition is given only to those working on a data analysis project, 
-rather othan on the development and support of the underlying software. A possible 
-solution to this conundrum could be the establishment of specific career paths 
-for researchers specialising in scientific software development. However, the practical 
-implementation of such an approach is extremely challenging, as there are a wide range of
+officially recognized credits for the advancement of their careers. The current
+practice in HEP is that these young scientists only receive credit and
+recognition (in the vast majority of cases) by working on a data analysis
+project, and this discourages them from finding time to transfer their acquired
+know-how to others in an effective way. A possible solution to this conundrum
+could be the establishment of specific career paths for researchers with a
+computer science specialization. The practical implementation of such an
+approach, however, is extremely challenging as there are a wide range of
 institutions involved that belong to different countries with their own
-policies, priorities, and funding strategies.
-
-Broadly speaking, two scientific profiles of researchers could be
-envisaged; a physicst could make a detailed plan of what they need, expressed 
-in a tidy requirements document, and a computer scientist could use this document 
-to provide the required solution. This is how the software development process 
-typically works in the software industry. However, in HEP requirements are rapidly
-changing forcing developers and physicsts to closely interact during the entire process of
+policies, priorities, and funding strategies. Broadly, two career paths could be
+envisaged. Both would recognise that domain-sepcific knowledge and continued
+engagement is vital to the success of HEP computing and software. While in an
+ideal world (for physicists, at least), a physicst could make a detailed plan of
+what they need (expressed in a tidy requirements document) and a computer
+scientist could use this document to provide the required solution, with very
+little additional exchange of information during the process (which is how the
+software development process sometimes works in industry). In the real HEP
+world, instead, most of the time the requirements are bound to change rather
+frequently for a variety of reasons (not only because of poor planning) forcing
+developers and physicsts to closely interact during the entire process of
 software development; this works smoothly only when both communities have the
-same goals and speak the same language, hence the need for physicists with a good
-knowledge of computer science.
+same goals and speak the same jargon, hence the need for physicists with a
+computer science specialisation.
 
@@
 \emph{computer scientist}: the two have different goals, seek different paths to
-the solution of their problems, and usually do not even adopt the same language in
+the solution of their problems, and usually do not even adopt the same jargon in
 expressing problems and solutions. Such a role would have an active role in
@@
 researcher community, delivering a quick fix or a refactoring and then moving
-on. This does not deliver the domain specific expertise required.
+one. This does not deliver the domain specific expertise required.
 
@@
 experience, interests, and time availablity to learn new techniques.  Any
-training programme must take into account the variation in target audiences. For
+training program must take into account the variation in target audiences. For
 example, an undergraduate doing a summer research project has different needs
-and skills than their professor.
-
-\subsection{Classification of trainees}
-
+and skills than their mentoring professor.
 For purposes of training, it is useful to think of all our user community as
@@
    control, and command-line shells).
-   \item \emph{Intermediate level analysts}: people with some experience in 
+   \item \emph{Intermediate users}: HEP analyzers with some experience in 
    either analysis concepts and/or tools, but  looking to supplement their 
-   experience with more recent and modern approaches. 
-   \item \emph{Advanced level analysts}: experts who have mastered current 
-   technologies and implementations and who want to stay up-to-date with 
-   new advanced developments.
+   experiences with more recent and modern approaches. 
+   \item \emph{Advanced users}: HEP analysis experts who have mastered current 
+   technologies and implementations (at a certain point in time) and want to 
+   stay up-to-date with new advanced developments.
    \item \emph{Software developers}: HEP scientists in charge of software 
@@
 
-Each of these groups has very different training needs. However,
-\emph{whenever possible}, any training programme should take advantage of
+Each of these groups has vastly different training needs. However,
+\emph{whenever possible}, any training program should take advantage of
 developments in pedagogy, such as active
@@
 flawed, and ask students to fix or improve them. Learning the material in a way
-that sticks is difficult and challenging for both the students
+that sticks with the students is difficult and challenging for both the students
 and the instructor and often takes more time than we would prefer. However, this
-is the best way to educate scientists who can fully
-contribute to the physics programmes at large, which is really the ultimate goal
-of any training programme.
+is the best way to ensure an educated community of scientists who can fully
+contribute to the physics programs at large, which is really the ultimate goal
+of any training program.
 
@@
 to the whole HEP community and possibly beyond. In this section, we present a
-list of specific concepts that need to be taught, in
-order to guarantee the basic level of competence needed to write efficient code
+list of specific concepts that need to be taught to members of the community, in
+order to guarantee the base level of competence needed to write efficient code
 for the different tasks performed in HEP experiments.
@@
 include modularity of code design, advanced data structures, evaluation metrics,
-writing tests and working with different types of hardware accelerator.
+writing tests and working with different types of hardware accelerators.
 Cutting-edge approaches such as the use of hardware accelerators (whether GPUs
 or FPGAs), Deep Neural Networks, and Machine Learning should be considered
-important training topics. Additionally, special emphasis should be given to
+important training topics. Additionally, special emphasis should be made on
 reporting results and documenting them.
-The training subjects considered important to pursue are
+The training subjects that the community considers important to pursue are
 listed as follows:
@@
    \end{itemize}
-   \item Algorithms 
+   \item Algorithmics 
    \begin{itemize}
@@
    \end{itemize}
-   \item Frameworks (development or application level)
+   \item Existing frameworks (development or application level)
    \begin{itemize}
@@
       \item ROOT
-      \item experiment specific frameworks (possibly if of potential interest outside the originating experiment)
+      \item experiment specific framework (possibily if of potential interest outside the native experiment)
    \end{itemize}
@@
 %____________________________________________________________________
-\section{Implementation Roadmap}
-The implementation of training should target multiple training formats and knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and advanced visualizations, etc.
-People who learn in non-conventional ways should not be excluded from the
-opportunity to learn. Since there are different stages of knowledge transfer,
-programmes should be developed that encourage experts to share knowledge. Issues
-considered in this section include global training initiatives, strategies for
-leveraging existing forums, and training tool development.
+\section{Challenges}
 
-An important point to consider is the difficulty, often experienced in the past,
-in developing large software programs across different experiment communities.
+Within a single experiment, different skill sets are needed. In addition to a
+skill set that contains basic programming language knowledge, testing and code
+management tools, and experiment-specific framework knowledge, there are more
+specialized skills that only a subset of the community needs to know, such as
+software optimization, or low-level hardware interfaces.
 While there already are experiment-specific training efforts in place, there are
 many needs that are in common between experiments. Therefore, we should probably
-strive to extract the common knowledge and build a common training programme. This
+strive to extract the common knowledge and build a common training program. This
 would reduce duplication of efforts and enable growth of a shared training
@@
 aimed at improving the efficiency and cost-effectivness of this important
-activity. An appropriate incentive programme to reward those who train the community 
-might help to facilitate the goals of such an initiative.
+activity.
+
+\section{Implementation Roadmap}
+The implementation of training should target multiple training formats and knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and advanced visualizations, etc.
+People who learn in non-conventional ways should not be excluded from the
+opportunity to learn. Since there are different stages of knowledge transfer,
+programs should be developed that encourage experts to share knowledge. Issues
+considered in this section include global training initiatives, strategies for
+leveraging existing forums, and training tool development.
+
+An important point to consider is the difficulty, often experienced in the past,
+in developing large software programs across different experiment communities.
+It is always challenging to organize a Working Group for training across
+experiments or even across diverse communities (such as HEP and Astroparticle,
+for example). This is a point where an appropriate incentive program to reward
+those who train the community might help.
 
 %____________________________________________________________________
-\subsection{Initiatives for future training programmes\label{sec:initiatives}}
+\subsection{Initiatives for future training programs\label{sec:initiatives}}
 
 Some methods that can be used for global training include Massive Online Open
-Courses (MOOCs), hands-on workshops, online knowledge bases, expert trainer volunteer
+Courses, hands-on workshops, online knowledge bases, expert trainer volunteer
 networks, and web-based training approaches.
 
-MOOCs can be used to develop an open-source set of tutorials and tools. Existing 
-online courses such as Udacity and Coursera can be evaluated and recommended to 
-the community. A novel approach such as WikiToLearn
-could also explored to assess potential benefits, as has already been attempted
-in the context of the GridKa School of Computing (see below).
+Massive Online Open Courses (MOOCs) can be used to develop an open-source set of
+tutorials and tools. Existing online courses such as Udacity and Coursera can be
+evaluated and recommended to the community. A novel approach such as WikiToLearn
+could also explored to assess potential benefits (as has already been attempted
+in the context of the GridKa School of Computing, see below).
 
@@
 incentives for experts to contribute. They can be open-source so that a lot of
-knowledge can be added by the trainees themselves as information is learned;
-this is the particular context where an approach such as WikiToLearn could be
-of great help.
+knowledge can be added by the trainees themselves as information is learned
+(this is the particular context where an approach such as WikiToLearn could be
+of great help).
 
-Question-and-answer websites, such as Stack Overflow for HEP, are also a very
-useful resource for addressing common problems and questions. This approach has already
+Question-and-answer websites such as Stack Overflow for HEP are also a very
+useful resource for common problems and questions. This approach has already
 been considered by HSF start-up members, and turned out to be difficult to
@@
 the work done, at the expense of a broader training/education curriculum that
-could actually yield improved results further down the line. One incentive
-would be to provide training programmes that also count as course credit, perhaps
+could actually yield improved results, but further down the line. One incentive
+would be to provide training programs that also count as course credit, perhaps
 as an elective. This model is already in limited use with some online
@@
 courses are the lack of funding and the lack of available time by experts in the
-field. People with enough expertise or insight usually don't have
+field. People with enough expertise or insight in the field usually don't have
 time to devote to prolonged periods of student training, and, even when they can
@@
    \item Tutors can add material to the web site at a very slow pace (whenever
-   they find time to do it, one slide a week or a chapter a day) 
-   \item Their material, made public on a web-site, can be further expanded by collaborators
+   they find time to do it, one slide a week or a chapter a day) \item Their
+   material, made public on a web-site, can be further expanded by collaborators
    (also at their own pace) or, even better and more productively, by students
@@
    given time, creates a sense of community, and creates bridges between people
-   in contiguous areas of research. Students can use the same platform to
+   in contiguous areas or research. Students can use the same platform to
    exchange their own examples, make suggestions and point out interesting
@@
    set up in such a way that students will work in an exact replica of the
-   environment they will be exposed to in their experiment. Moreover, students 
-   could be provided with Docker containers that
-   preserve their modified environment across sessions, allowing them to
+   environment they will be exposed to in their experiment (or collaboration at
+   large). Moreover, students could be provided with Docker containers that
+   preserves their modified environment across sessions, allowing them to
    develop their skills over a prolonged period of time by accessing all the
@@
    sense it could be a {\it complementary} approach to a traditional school. Of
-   particular interest could be courses such as ``Machine Learning'',
+   particular interest could be courses such as, e.g., ``Machine Learning'',
    ``Statistical Analyis with ROOT'', or even just ``Good practices in C++'' or
@@
 to add material to the site, and to make it appear under a specific topic (such
-as Software/Techniques/Machine-Learning) and to manipulate it as if it
+as, e.g., Software/Techniques/Machine-Learning) and to manipulate it as if it
 were a single document. In the end, students can selectively choose individual
@@
 publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}.
-This constitutes an interesting example of what can be accomplished using this
+This constitutes an intersting example of what can be accomplished using this
 platform; it is just a first example of what's possible, but an inspiring one.
@@
 complementary approaches to training (such as schools and dedicated web-site
-courses) could be of mutual benefit in the development of a complete training 
-programme.
+courses) could be of mutual benefit (how to make them cooperate in the
+development of a complete training program).
 
@@
 
-Over the past decade, MOOCs have been developed by universities and private organizations. 
-They have been well received by industry and academia.
-In addition, they provide a lot of flexibility in terms of cost and use of time; 
-they are typically free and open for enrollment at any time of the
+Over the past decade, massive open online courses (MOOCs) have been developed by
+universities and private organizations. They have been well received by industry
+and academia.% and are maintained by the outside community.
+In addition, they provide a lot of flexibility in terms of cost and time
+constraints; they are typically free and open for enrollment at any time of the
 year. Since the material can be accessed at any time and revisited at any time,
@@
 specific tools taught. Exploring these options allows us to choose which is the
-right offering for the knowledge needed to work on a specific experiment. We
+right offering for the knowledge needed to work on a specific HEP experiment. We
 can pick and choose modules to tailor an appropriate roadmap of skills to learn.
-More difficult will be to assemble specific training material not already
+More difficult will be to assemble HEP specific training material not already
 avaible elsewhere in an efficient and organized way, since this requires
@@
 
-Training is something that a large cross-section of the community
+Training the community is something that a large cross-section of the community
 understands to be important, but finding time and effort to contribute to this
@@
 
+In the context of web-based training, if HSF helps to constitute a living
+collection of online tutorials, we could organize regular events such as
+a "Tutorial of the month", based on students votes. From a more general point
+of view, since the publications are the most universal recognized items
+in the research community, we should try hardly to create some sort of
+training journal, with reviewed training papers.
+
 It is also critical to incorporate training into grant proposals so that it can
@@
 
-Software Carpentry (SC) \footnote{\url{https://software-carpentry.org}} and Data
-Carpentry (DC) \footnote{\url{http://www.datacarpentry.org}} are two parts of The
+Software Carpentry\footnote{\url{https://software-carpentry.org}} and Data
+Carpentry\footnote{\url{http://www.datacarpentry.org}} are two parts of The
 Carpentries organization that collaboratively build and teach some of the basic
@@
 developed specifically for this course and not a part of a larger Software
-Carpentry programme. This course focuses on a shallow but wide building of
+Carpentry program. This course focuses on a shallow but wide building of
 foundational skills approach, introducing many base concepts but not going deep
@@
 We have identified a certain number of problems that need to be overcome in
-setting up an appropriate training programme:
+setting up an appropriate training program:
 \begin{itemize}
@@
 %
-%   \item How to motivate the community to participate in developing 
-programs? Can some of it come from \& leadership?
-%   \item How to deal with competitiveness in the field? Can’t force 
-people to be interested in forming a community. How can we incentivize this? 
-Building a %vibrant, welcoming community may help.
+%   \item How to motivate the community to participate in developing programs? Can some of it come from \& leadership?
+%   \item How to deal with competitiveness in the field? Can’t force people to be interested in forming a community. How can we incentivize this? Building a %vibrant, welcoming community may help.
 %\end{itemize}

### 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Tutorial of the month.

--- 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 culture by accumulating and sharing expertise across different experiments.
-
 To realise this goal, a possibility could be the institution of a {\em
@@
 might help to facilitate the goals of such an initiative.
+
+When trying to exchange training materials within a group the first problem is
+to convince authors to contribute their work. This issue
+can be alleviated with cautious intellectual property clauses. Another
+problem is that trainers do not like to reuse given material as-is. They usually
+want to refactor it, building their own training story.
+This makes it difficult to have everyone agree on common material.
+It is even challenging to agree to host material in the same
+centralized place. A way to overcome this could be to settle for a centralized
+catalog, with each author being free to host their material in their location
+of choice.
+
+Another challenge is the wide heterogeneity in
+student competence. Special care must be given to setup
+a training structure that can cope with both beginners and advanced students.
+This may rest on a low granularity subdivision of independant topics.
+The IN2P3 authors try to restrict their tutorials to 20 minutes
+of material, as inspired by the Pomodoro technique, so that one can easily jump
+and compose one's own ``menu''. However, it turns out to be tricky to keep such
+small tutorials really independant and meaningful when they can be selected at
+will.
+
+All trainers have also faced the tremendous loss of time
+in software installation for any practical exercises as many students have
+not managed to prepare their machines in advance.
+Here containers may bring real
+progress, provided that everyone has at least Docker installed on their machine.
+JupyterHub, also, is a technology which enable training session with
+no need for installation (but with a need for network).
 
@@
 the necessary learning curve has been kept appropriately shallow.
-
 An interesting exercise in this context has recently been made by colleagues of
@@
 publicly available on WTL~\cite{WikiToLearnGridka}.
-This constitutes an intersting example of what can be accomplished using this
+This constitutes an interesting example of what can be accomplished using this
 platform; it is just a first example of what's possible, but an inspiring one.
+
+Another example of web-based platform is a collection of online tutorials
+(mostly written in French) hosted on the Gitlab IN2P3
+server~\cite{Code-Swim-Coaches}. Taking advantage of
+the Gitlab ability to host Docker images, those tutorials aim to avoid the
+``damned installation step'' that often absorbs half of a training session.
+Similarly to the GridKa School for WikiToLearn, the annual IN2P3 Computing
+Days are an opportunity to refresh and extend the collection of
+tutorials every year. Future work will focus on English translation, and the
+development of a web site which will index the above tutorials, together with
+the best recommended external tutorials.
 

### 2018-06-14 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Some more challenges.

--- 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 
+Yet, when trying to exchange training materials within a group, the problem
+is not only to convince authors to give their work. This first resistance
+can be facilitated with cautious intellectual property clauses. The other
+problem is that trainers do not like to reuse a given material as-is. They usually
+absolute want to refactor it, so to build their own training story.
+Thus, it is really difficult to have everyone agrees on common material.
+It is even challenging to have them all host their material in the same
+centralized place. A way to bypass this could be to settle for a centralized
+catalog, each author being free to host his material in his favorite place.
+
+Another challenge, as already stated, is the wide heterogeneity of the
+competences of the students. A special care must be given to setup
+training structure that can cope with both beginners and advanced students.
+This may rely on a low granularity subdivision of independant topics.
+The IN2P3 authors try to restrain their tutorials within a 20 minutes
+time frame, as inspired by the Pomodoro technique, so that one can easily jump
+and compose his own menu. But it turns to be tricky to keep such small
+tutorials really independant and meaningful when one choose them at will.
+
+Last but not least, any trainer has faced the tremendous lost of time
+with the softare installation step, at the beginning of any session, when
+many students have unhappilly not prepared their machine in advance,
+despites repeated injonctions... Here the containers may bring a real
+progress, provided everyone have at least Docker installed on his machine.
+JupyterHub, also, is a technology which enable training session with
+no need for installation (but need for network).
+
+%____________________________________________________________________
+%____________________________________________________________________
 \section{Implementation Roadmap}
@@
 
-In the context of web-based training, if HSF helps to constitute a living
-collection of online tutorials, we could organize regular events such as
-a "Tutorial of the month", based on students votes. From a more general point
-of view, since the publications are the most universal recognized items
-in the research community, we should try hardly to create some sort of
-training journal, with reviewed training papers.
-
 It is also critical to incorporate training into grant proposals so that it can

### 2018-06-14 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Two subsections for training needs.

--- 2018-06-13 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 example, an undergraduate doing a summer research project has different needs
-and skills than their mentoring professor.
-For purposes of training, it is useful to think of all our user community as
-\emph{students}. More specifically, we can broadly identify four different sets
-of students, classified by experience level, recognizing that even within these
-groups there will be people who will learn at different speeds and with
-differing depths of understanding:
+and skills than their mentoring professor. For purposes of training, we will
+name all our user community as \emph{students}.
+
+%____________________________________________________________________
+\subsection{Classification of trainees}
+
+We can broadly identify four different sets of students, classified by experience level,
+recognizing that even within these groups there will be people who will learn at different
+speeds and with differing depths of understanding:
 

### 2018-06-14 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
A more committed title

@@
 example, an undergraduate doing a summer research project has different needs
-and skills than their mentoring professor. For purposes of training, we will
-name all our user community as \emph{students}.
-
-%____________________________________________________________________
-\subsection{Classification of trainees}
-
-We can broadly identify four different sets of students, classified by experience level,
-recognizing that even within these groups there will be people who will learn at different
-speeds and with differing depths of understanding:
+and skills than their mentoring professor.
+For purposes of training, it is useful to think of all our user community as
+\emph{students}. More specifically, we can broadly identify four different sets
+of students, classified by experience level, recognizing that even within these
+groups there will be people who will learn at different speeds and with
+differing depths of understanding:
 

### 2018-06-20 CWP/papers/HSF-CWP-2017-01_roadmap/latex/authors.tex
Add Andronico as author and regenerated authors latex
* Add Giuseppe Andronico as authoe

Address taken from recent CMS publication, but not yet sure if the
University should be mentioned or not

* Add regenerated authors LaTeX file for arXiv

@@
-Albrecht, Johannes$^{68}$;
-Alves Jr, Antonio Augusto$^{80}$;
+Albrecht, Johannes$^{69}$;
+Alves Jr, Antonio Augusto$^{81}$;
 Amadio, Guilherme$^{5}$;
-Anh-Ky, Nguyen$^{121}$;
-Aphecetche, Laurent$^{65}$;
+Andronico, Giuseppe$^{27}$;
+Anh-Ky, Nguyen$^{122}$;
+Aphecetche, Laurent$^{66}$;
 Apostolakis, John$^{5}$;
-Asai, Makoto$^{62,p}$;
+Asai, Makoto$^{63,p}$;
 Atzori, Luca$^{5}$;
 Babik, Marian$^{5}$;
-Bagliesi, Giuseppe$^{31}$;
+Bagliesi, Giuseppe$^{32}$;
 Bandieramonte, Marilena$^{5}$;
@@
 Barisits, Martin$^{5}$;
-Bauerdick, Lothar A. T.$^{16,c}$;
-Belforte, Stefano$^{34}$;
-Benjamin, Douglas$^{81}$;
-Bernius, Catrin$^{62}$;
-Bhimji, Wahid$^{45}$;
-Bianchi, Riccardo Maria$^{104}$;
+Bauerdick, Lothar A.T.$^{16,c}$;
+Belforte, Stefano$^{35}$;
+Benjamin, Douglas$^{82}$;
+Bernius, Catrin$^{63}$;
+Bhimji, Wahid$^{46}$;
+Bianchi, Riccardo Maria$^{105}$;
 Bird, Ian$^{5}$;
-Biscarat, Catherine$^{51}$;
+Biscarat, Catherine$^{52}$;
 Blomer, Jakob$^{5}$;
-Bloom, Kenneth$^{96}$;
-Boccali, Tommaso$^{31}$;
-Bockelman, Brian$^{96}$;
-Bold, Tomasz$^{42}$;
+Bloom, Kenneth$^{97}$;
+Boccali, Tommaso$^{32}$;
+Bockelman, Brian$^{97}$;
+Bold, Tomasz$^{43}$;
 Bonacorsi, Daniele$^{25}$;
-Boveia, Antonio$^{100}$;
-Bozzi, Concezio$^{27}$;
-Bracko, Marko$^{92,40}$;
-Britton, David$^{85}$;
-Buckley, Andy$^{85}$;
+Boveia, Antonio$^{101}$;
+Bozzi, Concezio$^{28}$;
+Bracko, Marko$^{93,41}$;
+Britton, David$^{86}$;
+Buckley, Andy$^{86}$;
 Buncic, Predrag$^{5,a}$;
-Calafiura, Paolo$^{45}$;
+Calafiura, Paolo$^{46}$;
 Campana, Simone$^{5,a}$;
@@
 Canali, Luca$^{5}$;
-Carlino, Gianpaolo$^{30}$;
-Castro, Nuno$^{46,95,d}$;
+Carlino, Gianpaolo$^{31}$;
+Castro, Nuno$^{47,96,d}$;
 Cattaneo, Marco$^{5}$;
@@
 Cervantes Villanueva, Javier$^{5}$;
-Chang, Philip$^{74}$;
-Chapman, John$^{75}$;
+Chang, Philip$^{75}$;
+Chapman, John$^{76}$;
 Chen, Gang$^{23}$;
 Childers, Taylor$^{1}$;
-Clarke, Peter$^{82}$;
+Clarke, Peter$^{83}$;
 Clemencic, Marco$^{5}$;
-Cogneras, Eric$^{49}$;
-Coles, Jeremy$^{75}$;
-Collier, Ian$^{60}$;
+Cogneras, Eric$^{50}$;
+Coles, Jeremy$^{76}$;
+Collier, Ian$^{61}$;
+Colling, David$^{38}$;
 Corti, Gloria$^{5}$;
 Cosmo, Gabriele$^{5}$;
-Costanzo, Davide$^{111}$;
+Costanzo, Davide$^{112}$;
 Couturier, Ben$^{5}$;
-Cranmer, Kyle$^{56}$;
+Cranmer, Kyle$^{57}$;
 Cranshaw, Jack$^{1}$;
 Cristella, Leonardo$^{26}$;
-Crooks, David$^{85}$;
-Crépé-Renaudin, Sabine$^{51}$;
-Currie, Robert$^{82}$;
+Crooks, David$^{86}$;
+Crépé-Renaudin, Sabine$^{52}$;
+Currie, Robert$^{83}$;
 Dallmeier-Tiessen, Sünje$^{5}$;
-De, Kaushik$^{113}$;
-De Cian, Michel$^{86}$;
+De, Kaushik$^{114}$;
+De Cian, Michel$^{87}$;
 De Roeck, Albert$^{5}$;
 Delgado Peris, Antonio$^{7,g}$;
-Derue, Frédéric$^{50}$;
+Derue, Frédéric$^{51}$;
 Di Girolamo, Alessandro$^{5}$;
-Di Guida, Salvatore$^{29}$;
+Di Guida, Salvatore$^{30}$;
 Dimitrov, Gancho$^{5}$;
-Doglioni, Caterina$^{90,h}$;
-Dotti, Andrea$^{62,p}$;
+Doglioni, Caterina$^{91,h}$;
+Dotti, Andrea$^{63,p}$;
 Duellmann, Dirk$^{5}$;
-Duflot, Laurent$^{44}$;
+Duflot, Laurent$^{45}$;
 Dykstra, Dave$^{16,c}$;
@@
 Dziurda, Agnieszka$^{5}$;
-Egede, Ulrik$^{37}$;
-Elmer, Peter$^{105,a}$;
+Egede, Ulrik$^{38}$;
+Elmer, Peter$^{106,a}$;
 Elmsheuser, Johannes$^{2}$;
@@
 Eulisse, Giulio$^{5}$;
-Farrell, Steven$^{45}$;
-Ferber, Torben$^{72}$;
-Filipcic, Andrej$^{40}$;
-Fisk, Ian$^{63}$;
+Farrell, Steven$^{46}$;
+Ferber, Torben$^{73}$;
+Filipcic, Andrej$^{41}$;
+Fisk, Ian$^{64}$;
 Fitzpatrick, Conor$^{14}$;
-Flix, José$^{58,7,g}$;
-Formica, Andrea$^{38}$;
-Forti, Alessandra$^{91}$;
+Flix, José$^{59,7,g}$;
+Formica, Andrea$^{39}$;
+Forti, Alessandra$^{92}$;
 Franzoni, Giovanni$^{5}$;
-Frost, James$^{103}$;
+Frost, James$^{104}$;
 Fuess, Stu$^{16,c}$;
@@
 Ganis, Gerardo$^{5}$;
-Gardner, Robert$^{79}$;
-Garonne, Vincent$^{101}$;
+Gardner, Robert$^{80}$;
+Garonne, Vincent$^{102}$;
 Gellrich, Andreas$^{13}$;
 Genser, Krzysztof$^{16,c}$;
-George, Simon$^{61}$;
-Geurts, Frank$^{106}$;
+George, Simon$^{62}$;
+Geurts, Frank$^{107}$;
 Gheata, Andrei$^{5}$;
@@
 Giacomini, Francesco$^{9}$;
-Giagu, Stefano$^{108,33}$;
-Giffels, Manuel$^{41}$;
-Gingrich, Douglas$^{69}$;
+Giagu, Stefano$^{109,34}$;
+Giffels, Manuel$^{42}$;
+Gingrich, Douglas$^{70}$;
 Girone, Maria$^{5}$;
-Gligorov, Vladimir V.$^{50,b}$;
-Glushkov, Ivan$^{113}$;
-Gohn, Wesley$^{87}$;
+Gligorov, Vladimir V.$^{51,b}$;
+Glushkov, Ivan$^{114}$;
+Gohn, Wesley$^{88}$;
 Gonzalez Lopez, Jose Benito$^{5}$;
-González Caballero, Isidro$^{102}$;
-González Fernández, Juan R.$^{102}$;
+González Caballero, Isidro$^{103}$;
+González Fernández, Juan R.$^{103}$;
 Govi, Giacomo$^{16}$;
 Grandi, Claudio$^{25}$;
-Grasland, Hadrien$^{44}$;
-Gray, Heather$^{45}$;
-Grillo, Lucia$^{91}$;
-Guan, Wen$^{118}$;
+Grasland, Hadrien$^{45}$;
+Gray, Heather$^{46}$;
+Grillo, Lucia$^{92}$;
+Guan, Wen$^{119}$;
 Gutsche, Oliver$^{16,c}$;
-Gyurjyan, Vardan$^{39}$;
-Hanushevsky, Andrew$^{62,p}$;
+Gyurjyan, Vardan$^{40}$;
+Hanushevsky, Andrew$^{63,p}$;
 Hariri, Farah$^{5}$;
@@
 Harvey, John$^{5,a}$;
-Hauth, Thomas$^{41}$;
+Hauth, Thomas$^{42}$;
 Hegner, Benedikt$^{5,a}$;
 Heinemann, Beate$^{13}$;
-Heinrich, Lukas$^{56}$;
-Heiss, Andreas$^{41}$;
+Heinrich, Lukas$^{57}$;
+Heiss, Andreas$^{42}$;
 Hernández, José M.$^{7,g}$;
-Hildreth, Michael$^{98,f}$;
-Hodgkinson, Mark$^{111}$;
-Hoeche, Stefan$^{62,p}$;
+Hildreth, Michael$^{99,f}$;
+Hodgkinson, Mark$^{112}$;
+Hoeche, Stefan$^{63,p}$;
 Holzman, Burt$^{16,c}$;
 Hristov, Peter$^{5}$;
-Huang, Xingtao$^{110}$;
-Ivanchenko, Vladimir N.$^{5,114}$;
-Ivanov, Todor$^{112}$;
+Huang, Xingtao$^{111}$;
+Ivanchenko, Vladimir N.$^{5,115}$;
+Ivanov, Todor$^{113}$;
 Iven, Jan$^{5}$;
-Jashal, Brij$^{67}$;
+Jashal, Brij$^{68}$;
 Jayatilaka, Bodhitha$^{16,c}$;
-Jones, Roger$^{88,a}$;
-Jouvin, Michel$^{44,a}$;
+Jones, Roger$^{89,a}$;
+Jouvin, Michel$^{45,a}$;
 Jun, Soon Yung$^{16,c}$;
-Kagan, Michael$^{62,p}$;
-Kalderon, Charles William$^{90}$;
-Kane, Meghan$^{64}$;
+Kagan, Michael$^{63,p}$;
+Kalderon, Charles William$^{91}$;
+Kane, Meghan$^{65}$;
 Karavakis, Edward$^{5}$;
-Katz, Daniel S.$^{78}$;
+Katz, Daniel S.$^{79}$;
 Kcira, Dorian$^{11}$;
 Keeble, Oliver$^{5}$;
-Kersevan, Borut Paul$^{89}$;
+Kersevan, Borut Paul$^{90}$;
 Kirby, Michael$^{16,c}$;
 Klimentov, Alexei$^{2}$;
-Klute, Markus$^{93}$;
-Komarov, Ilya$^{34,n}$;
-Konstantinov, Dmitri$^{59}$;
-Koppenburg, Patrick$^{55}$;
+Klute, Markus$^{94}$;
+Komarov, Ilya$^{35,n}$;
+Konstantinov, Dmitri$^{60}$;
+Koppenburg, Patrick$^{56}$;
 Kowalkowski, Jim$^{16,c}$;
-Kreczko, Luke$^{71}$;
-Kuhr, Thomas$^{48,a}$;
+Kreczko, Luke$^{72}$;
+Kuhr, Thomas$^{49,a}$;
 Kutschke, Robert$^{16,a,c}$;
 Kuznetsov, Valentin$^{12}$;
-Lampl, Walter$^{70}$;
+Lampl, Walter$^{71}$;
 Lancon, Eric$^{2}$;
-Lange, David$^{105,a}$;
+Lange, David$^{106,a}$;
 Lassnig, Mario$^{5}$;
 Laycock, Paul$^{5}$;
-Leggett, Charles$^{45}$;
-Letts, James$^{74}$;
+Leggett, Charles$^{46}$;
+Letts, James$^{75}$;
 Lewendel, Birgit$^{13}$;
-Li, Teng$^{82}$;
+Li, Teng$^{83}$;
 Lima, Guilherme$^{16,c}$;
-Linacre, Jacob$^{60,m}$;
+Linacre, Jacob$^{61,m}$;
 Linden, Tomas$^{18}$;
@@
 Lopienski, Sebastian$^{5}$;
-Love, Peter$^{88}$;
+Love, Peter$^{89}$;
 Lyon, Adam$^{16,c}$;
-Magini, Nicolò$^{28}$;
-Marshall, Zachary L$^{45}$;
+Magini, Nicolò$^{29}$;
+Marshall, Zachary L.$^{46}$;
 Martelli, Edoardo$^{5}$;
-Martin-Haugh, Stewart$^{60}$;
+Martin-Haugh, Stewart$^{61}$;
 Mato, Pere$^{5}$;
-Mazumdar, Kajari$^{67}$;
-McCauley, Thomas$^{98}$;
+Mazumdar, Kajari$^{68}$;
+McCauley, Thomas$^{99}$;
 McFayden, Josh$^{5}$;
-McKee, Shawn$^{94,l}$;
-McNab, Andrew$^{91}$;
-Mehdiyev, Rashid$^{77}$;
+McKee, Shawn$^{95,l}$;
+McNab, Andrew$^{92}$;
+Mehdiyev, Rashid$^{78}$;
 Meinhard, Helge$^{5}$;
-Menasce, Dario$^{29,a}$;
+Menasce, Dario$^{30,a}$;
 Mendez Lorenzo, Patricia$^{5}$;
-Mete, Alaettin Serhan$^{73}$;
-Michelotto, Michele$^{32}$;
-Mitrevski, Jovan$^{48}$;
+Mete, Alaettin Serhan$^{74}$;
+Michelotto, Michele$^{33}$;
+Mitrevski, Jovan$^{49}$;
 Moneta, Lorenzo$^{5}$;
-Morgan, Ben$^{117}$;
-Mount, Richard$^{62,p}$;
-Moyse, Edward$^{93}$;
-Murray, Sean$^{76,10}$;
+Morgan, Ben$^{118}$;
+Mount, Richard$^{63,p}$;
+Moyse, Edward$^{94}$;
+Murray, Sean$^{77,10}$;
 Nairz, Armin$^{5}$;
-Neubauer, Mark S$^{78,a,k}$;
+Neubauer, Mark S.$^{79,a,k}$;
 Norman, Andrew$^{16,c}$;
-Novaes, Sérgio$^{107}$;
+Novaes, Sérgio$^{108}$;
 Novak, Mihaly$^{5}$;
 Oyanguren, Arantza$^{22}$;
-Ozturk, Nurcan$^{113}$;
-Pacheco Pages, Andres$^{58,20,j}$;
-Paganini, Michela$^{119}$;
-Pansanel, Jerome$^{36}$;
-Pascuzzi, Vincent R.$^{115}$;
-Patrick, Glenn$^{60}$;
+Ozturk, Nurcan$^{114}$;
+Pacheco Pages, Andres$^{59,20,j}$;
+Paganini, Michela$^{120}$;
+Pansanel, Jerome$^{37}$;
+Pascuzzi, Vincent R.$^{116}$;
+Patrick, Glenn$^{61}$;
 Pearce, Alex$^{5}$;
-Pearson, Ben$^{53}$;
+Pearson, Ben$^{54}$;
 Pedro, Kevin$^{16,c}$;
 Perdue, Gabriel$^{16,c}$;
-Perez-Calero Yzquierdo, Antonio$^{58,7,g}$;
+Perez-Calero Yzquierdo, Antonio$^{59,7,g}$;
 Perrozzi, Luca$^{15}$;
-Petersen, Troels$^{54}$;
+Petersen, Troels$^{55}$;
 Petric, Marko$^{5}$;
-Petzold, Andreas$^{41}$;
+Petzold, Andreas$^{42}$;
 Piedra, Jónatan$^{21}$;
-Piilonen, Leo$^{122,i}$;
+Piilonen, Leo$^{123,i}$;
 Piparo, Danilo$^{5}$;
-Pivarski, Jim$^{105}$;
+Pivarski, Jim$^{106}$;
 Pokorski, Witold$^{5}$;
-Polci, Francesco$^{50}$;
+Polci, Francesco$^{51}$;
 Potamianos, Karolos$^{13}$;
 Psihas, Fernanda$^{24}$;
-Puig Navarro, Albert$^{120}$;
-Quast, Günter$^{41}$;
-Raven, Gerhard$^{55,123}$;
+Puig Navarro, Albert$^{121}$;
+Quast, Günter$^{42}$;
+Raven, Gerhard$^{56,124}$;
 Reuter, Jürgen$^{13}$;
@@
 Rinaldi, Lorenzo$^{25}$;
-Ritter, Martin$^{48}$;
+Ritter, Martin$^{49}$;
 Robinson, James$^{13}$;
-Rodrigues, Eduardo$^{80,a,e}$;
+Rodrigues, Eduardo$^{81,a,e}$;
 Roiser, Stefan$^{5,a}$;
-Rousseau, David$^{44}$;
-Roy, Gareth$^{85}$;
-Rybkine, Grigori$^{44}$;
+Rousseau, David$^{45}$;
+Roy, Gareth$^{86}$;
+Rybkine, Grigori$^{45}$;
 Sailer, Andre$^{5}$;
-Sakuma, Tai$^{71}$;
+Sakuma, Tai$^{72}$;
 Santana, Renato$^{3}$;
-Sartirana, Andrea$^{47}$;
-Schellman, Heidi$^{57}$;
+Sartirana, Andrea$^{48}$;
+Schellman, Heidi$^{58}$;
 Schovancová, Jaroslava$^{5}$;
-Schramm, Steven$^{84}$;
+Schramm, Steven$^{85}$;
 Schulz, Markus$^{5}$;
 Sciabà, Andrea$^{5}$;
-Seidel, Sally$^{97}$;
-Sekmen, Sezen$^{43}$;
-Serfon, Cedric$^{101}$;
-Severini, Horst$^{99}$;
+Seidel, Sally$^{98}$;
+Sekmen, Sezen$^{44}$;
+Serfon, Cedric$^{102}$;
+Severini, Horst$^{100}$;
 Sexton-Kennedy, Elizabeth$^{16,a,c}$;
-Seymour, Michael$^{91}$;
+Seymour, Michael$^{92}$;
 Sgalaberna, Davide$^{5}$;
-Shapoval, Illya$^{45}$;
+Shapoval, Illya$^{46}$;
 Shiers, Jamie$^{5}$;
-Shiu, Jing-Ge$^{66}$;
+Shiu, Jing-Ge$^{67}$;
 Short, Hannah$^{5}$;
 Siroli, Gian Piero$^{25}$;
-Skipsey, Sam$^{85}$;
+Skipsey, Sam$^{86}$;
 Smith, Tim$^{5}$;
 Snyder, Scott$^{2}$;
-Sokoloff, Michael D$^{80,a}$;
+Sokoloff, Michael D.$^{81,a}$;
 Spentzouris, Panagiotis$^{16,c}$;
 Stadie, Hartmut$^{17}$;
-Stark, Giordon$^{79}$;
-Stewart, Gordon$^{85}$;
-Stewart, Graeme$^{5,a}$;
-Sánchez, Arturo$^{116,19}$;
+Stark, Giordon$^{80}$;
+Stewart, Gordon$^{86}$;
+Stewart, Graeme A.$^{5,a}$;
+Sánchez, Arturo$^{117,19}$;
 Sánchez-Hernández, Alberto$^{8,o}$;
-Taffard, Anyes$^{73}$;
-Tamponi, Umberto$^{35}$;
-Templon, Jeff$^{55}$;
+Taffard, Anyes$^{74}$;
+Tamponi, Umberto$^{36}$;
+Templon, Jeff$^{56}$;
 Tenaglia, Giacomo$^{5}$;
-Tsulaia, Vakhtang$^{45}$;
-Tunnell, Christopher$^{79}$;
+Tsulaia, Vakhtang$^{46}$;
+Tunnell, Christopher$^{80}$;
 Vaandering, Eric$^{16,c}$;
 Valassi, Andrea$^{5}$;
-Vallecorsa, Sofia$^{83}$;
+Vallecorsa, Sofia$^{84}$;
 Valsan, Liviu$^{5}$;
@@
 Votava, Margaret$^{16,c}$;
-Vuosalo, Carl$^{118}$;
-Vázquez Sierra, Carlos$^{55}$;
+Vuosalo, Carl$^{119}$;
+Vázquez Sierra, Carlos$^{56}$;
 Wartel, Romain$^{5}$;
-Watts, Gordon T$^{109}$;
+Watts, Gordon T.$^{110}$;
 Wenaus, Torre$^{2}$;
 Wenzel, Sandro$^{5}$;
-Williams, Mike$^{52}$;
-Winklmeier, Frank$^{57}$;
+Williams, Mike$^{53}$;
+Winklmeier, Frank$^{58}$;
 Wissing, Christoph$^{13}$;
-Wuerthwein, Frank$^{74}$;
-Wynne, Benjamin$^{82}$;
+Wuerthwein, Frank$^{75}$;
+Wynne, Benjamin$^{83}$;
 Xiaomei, Zhang$^{23}$;
-Yang, Wei$^{62,p}$;
+Yang, Wei$^{63,p}$;
 Yazgan, Efe$^{23}$
@@
 \par {\footnotesize $^{26}$ INFN Sezione di Bari, Università di Bari, Politecnico di Bari, Bari, Italy}
-\par {\footnotesize $^{27}$ Università e INFN, Ferrara, Ferrara, Italy}
-\par {\footnotesize $^{28}$ INFN Sezione di Genova, Genova, Italy}
-\par {\footnotesize $^{29}$ INFN Sezione di Milano-Bicocca, Milano, Italy}
-\par {\footnotesize $^{30}$ INFN Sezione di Napoli, Università di Napoli, Napoli, Italy}
-\par {\footnotesize $^{31}$ INFN Sezione di Pisa, Università di Pisa, Scuola Normale Superiore di Pisa, Pisa, Italy}
-\par {\footnotesize $^{32}$ INFN Sezione di Padova, Università di Padova b, Padova, Italy}
-\par {\footnotesize $^{33}$ INFN Sezione di Roma I, Università La Sapienza, Roma, Italy}
-\par {\footnotesize $^{34}$ INFN Sezione di Trieste, Università di Trieste, Trieste, Italy}
-\par {\footnotesize $^{35}$ INFN Sezione di Torino, Torino, Italy}
-\par {\footnotesize $^{36}$ Université de Strasbourg, CNRS, IPHC UMR 7178, F-67000 Strasbourg, France}
-\par {\footnotesize $^{37}$ Imperial College London, London, United Kingdom}
-\par {\footnotesize $^{38}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat à l’Énergie Atomique), Gif-sur-Yvette, France}
-\par {\footnotesize $^{39}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, VA, USA}
-\par {\footnotesize $^{40}$ Jožef Stefan Institute, Ljubljana, Slovenia}
-\par {\footnotesize $^{41}$ Karlsruhe Institute of Technology, Karlsruhe, Germany}
-\par {\footnotesize $^{42}$ AGH University of Science and Technology, Faculty of Physics and Applied Computer Science, Krakow, Poland}
-\par {\footnotesize $^{43}$ Kyungpook National University, Daegu, Republic of Korea}
-\par {\footnotesize $^{44}$ LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
-\par {\footnotesize $^{45}$ Lawrence Berkeley National Laboratory and University of California, Berkeley, CA, USA}
-\par {\footnotesize $^{46}$ Laboratório de Instrumentação e Física Experimental de Partículas (LIP), Lisboa, Portugal}
-\par {\footnotesize $^{47}$ Laboratoire Leprince-Ringuet, École Polytechnique, CNRS/IN2P3, Université Paris-Saclay, Palaiseau, France}
-\par {\footnotesize $^{48}$ Fakultät für Physik, Ludwig-Maximilians-Universität München, München, Germany}
-\par {\footnotesize $^{49}$ Laboratoire de Physique Corpusculaire, Clermont Université and Université Blaise Pascal and CNRS/IN2P3, Clermont-Ferrand, France}
-\par {\footnotesize $^{50}$ Laboratoire de Physique Nucléaire et de Hautes Energies (LPNHE), Sorbonne Université, Université Paris Diderot, CNRS/IN2P3, Paris, France}
-\par {\footnotesize $^{51}$ Laboratoire de Physique Subatomique et de Cosmologie, Université Joseph Fourier and CNRS/IN2P3 and Institut National Polytechnique de Grenoble, Grenoble, France}
-\par {\footnotesize $^{52}$ Massachusetts Institute of Technology, Cambridge, MA, USA}
-\par {\footnotesize $^{53}$ Max-Planck-Institut für Physik (Werner-Heisenberg-Institut), München, Germany}
-\par {\footnotesize $^{54}$ Niels Bohr Institute, University of Copenhagen, Kobenhavn, Denmark}
-\par {\footnotesize $^{55}$ Nikhef National Institute for Subatomic Physics, Amsterdam, Netherlands}
-\par {\footnotesize $^{56}$ Department of Physics, New York University, New York, NY, USA}
-\par {\footnotesize $^{57}$ Center for High Energy Physics, University of Oregon, Eugene, OR, USA}
-\par {\footnotesize $^{58}$ Port d’Informació Científica (PIC), Universitat Autònoma de Barcelona (UAB), Barcelona, Spain}
-\par {\footnotesize $^{59}$ High Energy Physics (IHEP), Protvino, Russia}
-\par {\footnotesize $^{60}$ STFC Rutherford Appleton Laboratory, Didcot, United Kingdom}
-\par {\footnotesize $^{61}$ Department of Physics, Royal Holloway University of London, Surrey, United Kingdom}
-\par {\footnotesize $^{62}$ SLAC National Accelerator Laboratory, Menlo Park, CA, USA}
-\par {\footnotesize $^{63}$ Simons Foundation, New York, NY, USA}
-\par {\footnotesize $^{64}$ SoundCloud, Berlin, Germany}
-\par {\footnotesize $^{65}$ SUBATECH, IMT Atlantique, Université de Nantes, CNRS-IN2P3, Nantes, France}
-\par {\footnotesize $^{66}$ National Taiwan University, Taipei, Taiwan}
-\par {\footnotesize $^{67}$ Tata Institute of Fundamental Research, Mumbai, India}
-\par {\footnotesize $^{68}$ Technische Universit\"at Dortmund, Dortmund, Germany}
-\par {\footnotesize $^{69}$ Department of Physics, University of Alberta, Edmonton, AB, Canada}
-\par {\footnotesize $^{70}$ Department of Physics, University of Arizona, Tucson, AZ, USA}
-\par {\footnotesize $^{71}$ H.H. Wills Physics Laboratory, University of Bristol, Bristol, United Kingdom}
-\par {\footnotesize $^{72}$ Department of Physics, University of British Columbia, Vancouver, BC, Canada}
-\par {\footnotesize $^{73}$ Department of Physics and Astronomy, University of California Irvine, Irvine, CA, USA}
-\par {\footnotesize $^{74}$ University of California, San Diego, La Jolla, CA, USA}
-\par {\footnotesize $^{75}$ Cavendish Laboratory, University of Cambridge, Cambridge, United Kingdom}
-\par {\footnotesize $^{76}$ Physics Department, University of Cape Town, Cape Town, South Africa}
-\par {\footnotesize $^{77}$ Carleton University, Ottawa, ON, Canada}
-\par {\footnotesize $^{78}$ University of Illinois Urbana-Champaign, Champaign, Illinois, IL, USA}
-\par {\footnotesize $^{79}$ Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
-\par {\footnotesize $^{80}$ University of Cincinnati, Cincinnati, OH, USA}
-\par {\footnotesize $^{81}$ Department of Physics, Duke University, Durham, NC, USA}
-\par {\footnotesize $^{82}$ SUPA - School of Physics and Astronomy, University of Edinburgh, Edinburgh, United Kingdom}
-\par {\footnotesize $^{83}$ Gangneung-Wonju National University, South Korea}
-\par {\footnotesize $^{84}$ Section de Physique, Université de Genève, Geneva, Switzerland}
-\par {\footnotesize $^{85}$ SUPA - School of Physics and Astronomy, University of Glasgow, Glasgow, United Kingdom}
-\par {\footnotesize $^{86}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
-\par {\footnotesize $^{87}$ Department of Physics and Astronomy, University of Kentucky, Lexington, KY, USA}
-\par {\footnotesize $^{88}$ Physics Department, Lancaster University, Lancaster, United Kingdom}
-\par {\footnotesize $^{89}$ Department of Physics, Jožef Stefan Institute and University of Ljubljana, Ljubljana, Slovenia}
-\par {\footnotesize $^{90}$ Fysiska institutionen, Lunds Universitet, Lund, Sweden}
-\par {\footnotesize $^{91}$ School of Physics and Astronomy, University of Manchester, Manchester, United Kingdom}
-\par {\footnotesize $^{92}$ University of Maribor, Ljubljana, Slovenia}
-\par {\footnotesize $^{93}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
-\par {\footnotesize $^{94}$ Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
-\par {\footnotesize $^{95}$ Departamento de Física, Universidade do Minho, Braga, Portugal}
-\par {\footnotesize $^{96}$ University of Nebraska-Lincoln, Lincoln, NE, USA}
-\par {\footnotesize $^{97}$ Department of Physics and Astronomy, University of New Mexico, Albuquerque, NM, USA}
-\par {\footnotesize $^{98}$ University of Notre Dame, Notre Dame, IN, USA}
-\par {\footnotesize $^{99}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman, OK, USA}
-\par {\footnotesize $^{100}$ The Ohio State University, Columbus, OH, USA}
-\par {\footnotesize $^{101}$ Department of Physics, University of Oslo, Oslo, Norway}
-\par {\footnotesize $^{102}$ Universidad de Oviedo, Oviedo, Spain}
-\par {\footnotesize $^{103}$ Department of Physics, Oxford University, Oxford, UK}
-\par {\footnotesize $^{104}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
-\par {\footnotesize $^{105}$ Princeton University, Princeton, NJ, USA}
-\par {\footnotesize $^{106}$ Rice University, Houston, TX, USA}
-\par {\footnotesize $^{107}$ Universidade Estadual Paulista, São Paulo, Brazil}
-\par {\footnotesize $^{108}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
-\par {\footnotesize $^{109}$ University of Washington, Seattle, WA, USA}
-\par {\footnotesize $^{110}$ School of Physics, Shandong University, Shandong, China}
-\par {\footnotesize $^{111}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
-\par {\footnotesize $^{112}$ University of Sofia, Sofia, Bulgaria}
-\par {\footnotesize $^{113}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
-\par {\footnotesize $^{114}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
-\par {\footnotesize $^{115}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
-\par {\footnotesize $^{116}$ Dipartimento di Chimica, Fisica e Ambiente, Università di Udine, Udine, Italy}
-\par {\footnotesize $^{117}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
-\par {\footnotesize $^{118}$ University of Wisconsin - Madison, Madison, WI, USA}
-\par {\footnotesize $^{119}$ Department of Physics, Yale University, New Haven, CT, USA}
-\par {\footnotesize $^{120}$ Physik-Institut, Universität Zürich, Zürich, Switzerland}
-\par {\footnotesize $^{121}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
-\par {\footnotesize $^{122}$ Virginia Tech, Blacksburg, Virginia, VA, USA}
-\par {\footnotesize $^{123}$ Vrije Universiteit Amsterdam, Amsterdam, The Netherlands}
+\par {\footnotesize $^{27}$ INFN Sezione di Catania, Università di Catania, Catania, Italy}
+\par {\footnotesize $^{28}$ Università e INFN, Ferrara, Ferrara, Italy}
+\par {\footnotesize $^{29}$ INFN Sezione di Genova, Genova, Italy}
+\par {\footnotesize $^{30}$ INFN Sezione di Milano-Bicocca, Milano, Italy}
+\par {\footnotesize $^{31}$ INFN Sezione di Napoli, Università di Napoli, Napoli, Italy}
+\par {\footnotesize $^{32}$ INFN Sezione di Pisa, Università di Pisa, Scuola Normale Superiore di Pisa, Pisa, Italy}
+\par {\footnotesize $^{33}$ INFN Sezione di Padova, Università di Padova b, Padova, Italy}
+\par {\footnotesize $^{34}$ INFN Sezione di Roma I, Università La Sapienza, Roma, Italy}
+\par {\footnotesize $^{35}$ INFN Sezione di Trieste, Università di Trieste, Trieste, Italy}
+\par {\footnotesize $^{36}$ INFN Sezione di Torino, Torino, Italy}
+\par {\footnotesize $^{37}$ Université de Strasbourg, CNRS, IPHC UMR 7178, F-67000 Strasbourg, France}
+\par {\footnotesize $^{38}$ Imperial College London, London, United Kingdom}
+\par {\footnotesize $^{39}$ DSM/IRFU (Institut de Recherches sur les Lois Fondamentales de l’Univers), CEA Saclay (Commissariat à l’Énergie Atomique), Gif-sur-Yvette, France}
+\par {\footnotesize $^{40}$ Thomas Jefferson National Accelerator Facility, Newport News, Virginia, VA, USA}
+\par {\footnotesize $^{41}$ Jožef Stefan Institute, Ljubljana, Slovenia}
+\par {\footnotesize $^{42}$ Karlsruhe Institute of Technology, Karlsruhe, Germany}
+\par {\footnotesize $^{43}$ AGH University of Science and Technology, Faculty of Physics and Applied Computer Science, Krakow, Poland}
+\par {\footnotesize $^{44}$ Kyungpook National University, Daegu, Republic of Korea}
+\par {\footnotesize $^{45}$ LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
+\par {\footnotesize $^{46}$ Lawrence Berkeley National Laboratory and University of California, Berkeley, CA, USA}
+\par {\footnotesize $^{47}$ Laboratório de Instrumentação e Física Experimental de Partículas (LIP), Lisboa, Portugal}
+\par {\footnotesize $^{48}$ Laboratoire Leprince-Ringuet, École Polytechnique, CNRS/IN2P3, Université Paris-Saclay, Palaiseau, France}
+\par {\footnotesize $^{49}$ Fakultät für Physik, Ludwig-Maximilians-Universität München, München, Germany}
+\par {\footnotesize $^{50}$ Laboratoire de Physique Corpusculaire, Clermont Université and Université Blaise Pascal and CNRS/IN2P3, Clermont-Ferrand, France}
+\par {\footnotesize $^{51}$ Laboratoire de Physique Nucléaire et de Hautes Energies (LPNHE), Sorbonne Université, Université Paris Diderot, CNRS/IN2P3, Paris, France}
+\par {\footnotesize $^{52}$ Laboratoire de Physique Subatomique et de Cosmologie, Université Joseph Fourier and CNRS/IN2P3 and Institut National Polytechnique de Grenoble, Grenoble, France}
+\par {\footnotesize $^{53}$ Massachusetts Institute of Technology, Cambridge, MA, USA}
+\par {\footnotesize $^{54}$ Max-Planck-Institut für Physik (Werner-Heisenberg-Institut), München, Germany}
+\par {\footnotesize $^{55}$ Niels Bohr Institute, University of Copenhagen, Kobenhavn, Denmark}
+\par {\footnotesize $^{56}$ Nikhef National Institute for Subatomic Physics, Amsterdam, Netherlands}
+\par {\footnotesize $^{57}$ Department of Physics, New York University, New York, NY, USA}
+\par {\footnotesize $^{58}$ Center for High Energy Physics, University of Oregon, Eugene, OR, USA}
+\par {\footnotesize $^{59}$ Port d’Informació Científica (PIC), Universitat Autònoma de Barcelona (UAB), Barcelona, Spain}
+\par {\footnotesize $^{60}$ High Energy Physics (IHEP), Protvino, Russia}
+\par {\footnotesize $^{61}$ STFC Rutherford Appleton Laboratory, Didcot, United Kingdom}
+\par {\footnotesize $^{62}$ Department of Physics, Royal Holloway University of London, Surrey, United Kingdom}
+\par {\footnotesize $^{63}$ SLAC National Accelerator Laboratory, Menlo Park, CA, USA}
+\par {\footnotesize $^{64}$ Simons Foundation, New York, NY, USA}
+\par {\footnotesize $^{65}$ SoundCloud, Berlin, Germany}
+\par {\footnotesize $^{66}$ SUBATECH, IMT Atlantique, Université de Nantes, CNRS-IN2P3, Nantes, France}
+\par {\footnotesize $^{67}$ National Taiwan University, Taipei, Taiwan}
+\par {\footnotesize $^{68}$ Tata Institute of Fundamental Research, Mumbai, India}
+\par {\footnotesize $^{69}$ Technische Universit\"at Dortmund, Dortmund, Germany}
+\par {\footnotesize $^{70}$ Department of Physics, University of Alberta, Edmonton, AB, Canada}
+\par {\footnotesize $^{71}$ Department of Physics, University of Arizona, Tucson, AZ, USA}
+\par {\footnotesize $^{72}$ H.H. Wills Physics Laboratory, University of Bristol, Bristol, United Kingdom}
+\par {\footnotesize $^{73}$ Department of Physics, University of British Columbia, Vancouver, BC, Canada}
+\par {\footnotesize $^{74}$ Department of Physics and Astronomy, University of California Irvine, Irvine, CA, USA}
+\par {\footnotesize $^{75}$ University of California, San Diego, La Jolla, CA, USA}
+\par {\footnotesize $^{76}$ Cavendish Laboratory, University of Cambridge, Cambridge, United Kingdom}
+\par {\footnotesize $^{77}$ Physics Department, University of Cape Town, Cape Town, South Africa}
+\par {\footnotesize $^{78}$ Carleton University, Ottawa, ON, Canada}
+\par {\footnotesize $^{79}$ University of Illinois Urbana-Champaign, Champaign, Illinois, IL, USA}
+\par {\footnotesize $^{80}$ Enrico Fermi Institute, University of Chicago, Chicago, IL, USA}
+\par {\footnotesize $^{81}$ University of Cincinnati, Cincinnati, OH, USA}
+\par {\footnotesize $^{82}$ Department of Physics, Duke University, Durham, NC, USA}
+\par {\footnotesize $^{83}$ SUPA - School of Physics and Astronomy, University of Edinburgh, Edinburgh, United Kingdom}
+\par {\footnotesize $^{84}$ Gangneung-Wonju National University, South Korea}
+\par {\footnotesize $^{85}$ Section de Physique, Université de Genève, Geneva, Switzerland}
+\par {\footnotesize $^{86}$ SUPA - School of Physics and Astronomy, University of Glasgow, Glasgow, United Kingdom}
+\par {\footnotesize $^{87}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
+\par {\footnotesize $^{88}$ Department of Physics and Astronomy, University of Kentucky, Lexington, KY, USA}
+\par {\footnotesize $^{89}$ Physics Department, Lancaster University, Lancaster, United Kingdom}
+\par {\footnotesize $^{90}$ Department of Physics, Jožef Stefan Institute and University of Ljubljana, Ljubljana, Slovenia}
+\par {\footnotesize $^{91}$ Fysiska institutionen, Lunds Universitet, Lund, Sweden}
+\par {\footnotesize $^{92}$ School of Physics and Astronomy, University of Manchester, Manchester, United Kingdom}
+\par {\footnotesize $^{93}$ University of Maribor, Ljubljana, Slovenia}
+\par {\footnotesize $^{94}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
+\par {\footnotesize $^{95}$ Department of Physics, The University of Michigan, Ann Arbor, MI, USA}
+\par {\footnotesize $^{96}$ Departamento de Física, Universidade do Minho, Braga, Portugal}
+\par {\footnotesize $^{97}$ University of Nebraska-Lincoln, Lincoln, NE, USA}
+\par {\footnotesize $^{98}$ Department of Physics and Astronomy, University of New Mexico, Albuquerque, NM, USA}
+\par {\footnotesize $^{99}$ University of Notre Dame, Notre Dame, IN, USA}
+\par {\footnotesize $^{100}$ Homer L. Dodge Department of Physics and Astronomy, University of Oklahoma, Norman, OK, USA}
+\par {\footnotesize $^{101}$ The Ohio State University, Columbus, OH, USA}
+\par {\footnotesize $^{102}$ Department of Physics, University of Oslo, Oslo, Norway}
+\par {\footnotesize $^{103}$ Universidad de Oviedo, Oviedo, Spain}
+\par {\footnotesize $^{104}$ Department of Physics, Oxford University, Oxford, UK}
+\par {\footnotesize $^{105}$ Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA, USA}
+\par {\footnotesize $^{106}$ Princeton University, Princeton, NJ, USA}
+\par {\footnotesize $^{107}$ Rice University, Houston, TX, USA}
+\par {\footnotesize $^{108}$ Universidade Estadual Paulista, São Paulo, Brazil}
+\par {\footnotesize $^{109}$ Dipartimento di Fisica, Università La Sapienza, Roma, Italy}
+\par {\footnotesize $^{110}$ University of Washington, Seattle, WA, USA}
+\par {\footnotesize $^{111}$ School of Physics, Shandong University, Shandong, China}
+\par {\footnotesize $^{112}$ Department of Physics and Astronomy, University of Sheffield, Sheffield, United Kingdom}
+\par {\footnotesize $^{113}$ University of Sofia, Sofia, Bulgaria}
+\par {\footnotesize $^{114}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
+\par {\footnotesize $^{115}$ National Research Tomsk Polytechnic University, Tomsk, Russia}
+\par {\footnotesize $^{116}$ Department of Physics, University of Toronto, Toronto, ON, Canada}
+\par {\footnotesize $^{117}$ Dipartimento di Chimica, Fisica e Ambiente, Università di Udine, Udine, Italy}
+\par {\footnotesize $^{118}$ Department of Physics, University of Warwick, Coventry, United Kingdom}
+\par {\footnotesize $^{119}$ University of Wisconsin - Madison, Madison, WI, USA}
+\par {\footnotesize $^{120}$ Department of Physics, Yale University, New Haven, CT, USA}
+\par {\footnotesize $^{121}$ Physik-Institut, Universität Zürich, Zürich, Switzerland}
+\par {\footnotesize $^{122}$ IOP and GUST, Vietnam Academy of Science and Technology (VAST), Hanoi, Vietnam}
+\par {\footnotesize $^{123}$ Virginia Tech, Blacksburg, Virginia, VA, USA}
+\par {\footnotesize $^{124}$ Vrije Universiteit Amsterdam, Amsterdam, The Netherlands}
 \bigskip

### 2018-06-21 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Improve English expression regarding challenegs

@@
 
+When trying to exchange training materials within a group the first problem is
+to convince authors to contribute their work. This issue
+can be alleviated with cautious intellectual property clauses. Another
+problem is that trainers do not like to reuse given material as-is. They usually
+want to refactor it, building their own training story.
+This makes it difficult to have everyone agree on common material.
+It is even challenging to agree to host material in the same
+centralized place. A way to overcome this could be to settle for a centralized
+catalog, with each author being free to host their material in their location
+of choice.
+
+Another challenge is the wide heterogeneity in
+student competence. Special care must be given to setup
+a training structure that can cope with both beginners and advanced students.
+This may rest on a low granularity subdivision of independant topics.
+The IN2P3 authors try to restrict their tutorials to 20 minutes
+of material, as inspired by the Pomodoro technique, so that one can easily jump
+and compose one's own ``menu''. However, it turns out to be tricky to keep such
+small tutorials really independant and meaningful when they can be selected at
+will.
+
+All trainers have also faced the tremendous loss of time
+in software installation for any practical exercises as many students have
+not managed to prepare their machines in advance.
+Here containers may bring real
+progress, provided that everyone has at least Docker installed on their machine.
+JupyterHub, also, is a technology which enable training session with
+no need for installation (but with a need for network).
+
+%____________________________________________________________________
+%____________________________________________________________________
 \section{Implementation Roadmap}

### 2018-06-21 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
A more committed title (#91)
Training john graeme edits (#92)
* Changes from John H 2018-06-12

* Additional fixes and improvements from Graeme

--- 2018-06-14 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 
-\abstract{The rapidly increasing evolution of technology and the parallel increasing complexity of algorithmic analysis in HEP (and in science in general) require a much larger portfolio of specific skills by the developers with respect to the past.
-Young researchers, graduating from universities worldwide, currently do not receive adequate preparation in the very diverse fields of modern computing to respond to growing necessities posed by the most advanced experimental challenges.
-The vast majority of the HEP community strongly agrees on the need for an overarching umbrella of training programs to bring young researchers (but also more mature ones) up to date with new software technologies, such as, for example, concurrent programming and artificial intelligence.
-}
+\abstract{The rapid evolution of technology and the parallel increasing 
+complexity of algorithmic analysis in HEP requires developers 
+to acquire a much larger portfolio of programming skills. Young 
+researchers graduating from universities worldwide currently do not receive adequate 
+preparation in the very diverse fields of modern computing to respond to growing needs 
+of the most advanced experimental challenges. There is a growing consensus in
+the HEP community on the need for training programmes to bring researchers up to date with new software technologies, 
+in particular in the domains of concurrent programming and artificial intelligence. We review some
+of the initiatives under way for introducing new training programmes and highlight some of the issues 
+that need to be taken into account for these to be successful.}
 
@@
 
-Throughout the last few decades, the field of high energy physics (HEP) has
-consistently either driven or utilized the latest innovations in computational
-tools and technologies for problem solving.
-The toolset of the particle physicist is vast and ever-growing, and the problem
-set increasingly complex.
-A typical graduate student or postdoc working on HEP experiments will likely
-encounter a number of tools, software, and programming languages with which they
-are totally unfamiliar.
-This requires ``on-the-job'' training for the same people whose work drives the
-reach of our analyses, while they are developing the tools for these analyses.
+The field of high energy physics (HEP) has consistently exploited 
+the latest innovations in computational tools and technologies for processing data.
+The software toolset of the particle physicist is ever-growing, and the problem
+set increasingly complex. Young researchers will likely encounter a number of 
+programming techniques with which they are totally unfamiliar, and will therefore 
+need ``on-the-job'' training in order for them to be productive.
 
-A strong research community should consist of independent actors who ask
+Evidently a strong research community comprises independent thinkers who ask
 research questions and direct their own analyses, instead of blindly following
-prescriptions and recipes. It is therefore imperative that the community be
-proactive about providing training resources and opportunities for the good of
-the community, and that we follow through on the promise that adequate career
+prescriptions and recipes. It is therefore imperative that training opportunities 
+and resources are made available, such that adequate career
 path possibilities exist for people in HEP who would otherwise leave the field
@@
 
-The introduction and development of increasingly complex tools has led to
-increasing improvements in the performance of our analyses. However, the
-personnel who drive these advancements through novel implementations of the
-technology are the analyzers. In this context, the term {\it analyzers} includes
-all those researchers who participate in an HEP experiment whose contribution
-includes designing and implementing a high speed DAQ system, setting up a modern
-database to store the state of a detector, complex task of managing the data
-flow from the first trigger level to the final n-tuple dataset and creating
-sophisticated statistical and mathematical tools to extract publishable physical
-meaning from those same data.
+The introduction and development of sophisticated software tools has led to
+improvements in the performance of our data processing software. The
+personnel who drive these advances through novel implementations work on a broad 
+range of tasks, including the development of high speed DAQ systems, of modern databases for
+storing the state of a detector, the management of data flow from the first trigger 
+level to the final analysis dataset, as well as the development of algorithms and
+mathematical tools for extracting publishable physics results.
 
-The development of new tools is insufficient to drive advances in the field
-unless researchers who have the expertise apply them to physics questions.
-Furthermore, a larger and a better-connected community, with expertise in the
-tools and access to the new developments, can lead to an open-source model in
-HEP. The efforts to train the community in all relevant tools is, thus, a
-priority within the R\&D effort itself.
-
-However, as tools become more complex, the learning rate of developers must keep
-up if we are to utilize them effectively.  At the same time, as in other
-branches of research, it is essential to sustain efforts in maintaining
-reproducibility of the usage of the tools. This again highlights the need for
-physicts to be appropriately trained in computing science, rather than
-professional computing scientists, whose goals and interests may not necessarily
-be aligned or focused on the physics results of a particular HEP experiment.
-
-To avoid duplicate efforts within the community, it is important to find a way
-to approach training efforts holistically and cross-experimentally, whenever
-possible. This requires a certain level of organization and coordination in
-order to guarantee the delivery of efficient, non-redundant software to
-different experiments and to allow the different communities to agree on common
-standards and procedures.
+As tools become more complex, physicists must be continuously retrained in order
+to utilize them effectively. It is important to find a way to deploy training efforts
+holistically and cross-experimentally, whenever possible, in order to avoid unnecessary 
+duplication of the effort involved. This will require organization and coordination to
+ensure that different software teams can agree on common standards, thereby allowing different
+experiments to adopt the same solutions in an efficient way.
 
 Beyond the immediate benefit to the scientific community in having well-trained
-collaborators, most people who start a physics graduate program will have
+collaborators, most people who start a physics graduate programme will have
 careers outside of academia. This trend is reflected in several calls for
-proposals from national scientific funding agencies like H2020 and those from
-the National Science Foundation (NSF) that promote training in computer
-infrastructure and information areas.
+proposals from national scientific funding agencies that promote training in computer
+infrastructure areas, such as the EU's Horizon 2020 programme and those from
+the National Science Foundation (NSF) [Refs 1 and 2].
 
-%\textcolor{red}{this call is more to develop cyberinfrastructure staff than scientists - for example, people at computing centers}
-
-\begin{quote}
-\it The overarching goal of these programs is to prepare, nurture and grow the national scientific workforce for creating, utilizing, and supporting advanced cyberinfrastructure (CI) that enables cutting-edge science and engineering and contributes to the Nation's overall economic competiveness and security.
-
-{\bf - Training-based Workforce Development for Advanced Cyberinfrastructure (CyberTraining) Webinar}\footnote{ \url{https://www.nsf.gov/events/event_summ.jsp?cntn_id=190179&org=NSF}}
-\end{quote}
-
-\begin{quote}
-\it New information, communication, and computational technologies have had profound impacts on the practice of science (in this solicitation, the term science includes the natural, mathematical, computing, and social sciences),  engineering, and education.
-This includes the means by which citizens of all ages use science  and engineering to enhance professional and private lives.
-The systems, tools, and services emerging from these new technologies are linked to create a comprehensive cyberinfrastructure that is enabling individuals, groups, and organizations to advance research and education in ways that revolutionize who can participate, what they can do, and how they do it.
-Sustaining this revolution across all areas of science, engineering, and education requires the formation of a citizenry and workforce with the knowledge and skills needed to design and deploy as well as adopt and apply these cyber-based systems, tools and services over the long-term.
-The opportunity for such preparation should be available at all stages of formal and informal education (K-16 and lifelong), training and professional development, and must be extended to all individuals and communities.
-
-{\bf - Cyberinfrastructure Training, Education, Advancement, and Mentoring for Our 21st Century Workforce (CI-TEAM)}\footnote{ \url{https://www.nsf.gov/funding/pgm_summ.jsp?pims_id=12782}}
-\end{quote}
-
-An open issue is the kind of incentive that our community can collectively
-leverage in order to increase the number of people (with strong expertise in
-computing) who decide to dedicate time and effort to train fellow members of
-their community.
+A key goal is to find incentives that will encourage 
+increasing numbers of people to dedicate their time and effort to train their
+colleagues.
 To be effective, the training must be done by people who work at the cutting
@@
 researchers. These are also the very same people who are most in need of
-officially recognized credits for the advancement of their careers. The current
-practice in HEP is that these young scientists only receive credit and
-recognition (in the vast majority of cases) by working on a data analysis
-project, and this discourages them from finding time to transfer their acquired
-know-how to others in an effective way. A possible solution to this conundrum
-could be the establishment of specific career paths for researchers with a
-computer science specialization. The practical implementation of such an
-approach, however, is extremely challenging as there are a wide range of
+officially recognized credits for the advancement of their careers. Currently
+visibility and recognition is given mainly to those working on data analysis
+projects, rather than to the development and support of the underlying
+software. One solution to this problem would be the establishment of specific
+career paths for researchers specialising in scientific software development. 
+%% GS Comment - is this actually ``motivation'' - it seems more like a solution
+% See suggested title change for this section...
+However, the practical 
+implementation of such an approach is extremely challenging, as there are a wide range of
 institutions involved that belong to different countries with their own
-policies, priorities, and funding strategies. Broadly, two career paths could be
-envisaged. Both would recognise that domain-sepcific knowledge and continued
-engagement is vital to the success of HEP computing and software. While in an
-ideal world (for physicists, at least), a physicst could make a detailed plan of
-what they need (expressed in a tidy requirements document) and a computer
-scientist could use this document to provide the required solution, with very
-little additional exchange of information during the process (which is how the
-software development process sometimes works in industry). In the real HEP
-world, instead, most of the time the requirements are bound to change rather
-frequently for a variety of reasons (not only because of poor planning) forcing
-developers and physicsts to closely interact during the entire process of
-software development; this works smoothly only when both communities have the
-same goals and speak the same jargon, hence the need for physicists with a
-computer science specialisation.
+policies, priorities, and funding strategies.
+
+Broadly speaking, two scientific profiles of researchers could be
+envisaged; a physicst could make a detailed plan of what they need, expressed 
+in a tidy requirements document, and a computer scientist could use this document 
+to provide the required solution. This is how the software development process 
+typically works in the software industry. However, in HEP requirements are rapidly
+changing, forcing developers and physicsts to interact closely during the entire
+process of software development; this works smoothly only when both communities have the
+same goals and speak the same language, hence the need for physicists with a good
+knowledge of computer science.
 
@@
 experience, interests, and time availablity to learn new techniques.  Any
-training program must take into account the variation in target audiences. For
+training programme must take into account the variation in target audiences. For
 example, an undergraduate doing a summer research project has different needs
-and skills than their mentoring professor.
+and skills than their professor.
+
+\subsection{Classification of trainees}
+
 For purposes of training, it is useful to think of all our user community as
 \emph{students}. More specifically, we can broadly identify four different sets
-of students, classified by experience level, recognizing that even within these
-groups there will be people who will learn at different speeds and with
-differing depths of understanding:
+of students, classified by experience level:
 
@@
    \item \emph{Beginners}: New collaborators with no knowledge of the 
-   tools or techniques they are expected to use. These might be undergraduates 
+   tools or techniques they are expected to use. These might be undergraduates, 
    but they could also be senior researchers who have not yet or not recently 
@@
    training in modern computing techniques (basically compiled and scripting 
-   languages together with operating system basics such as filesystems, version 
+   languages, together with operating system basics such as filesystems, version 
    control, and command-line shells).
-   \item \emph{Intermediate users}: HEP analyzers with some experience in 
+   \item \emph{Intermediate level analysts}: people with some experience in 
    either analysis concepts and/or tools, but  looking to supplement their 
-   experiences with more recent and modern approaches. 
-   \item \emph{Advanced users}: HEP analysis experts who have mastered current 
-   technologies and implementations (at a certain point in time) and want to 
-   stay up-to-date with new advanced developments.
+   experience with more recent and modern approaches. 
+   \item \emph{Advanced level analysts}: experts who have mastered current 
+   technologies and implementations and who want to stay up-to-date with 
+   new advanced developments.
    \item \emph{Software developers}: HEP scientists in charge of software 
@@
 particular experiment or field of application, but general enough to be useful
-to the whole HEP community and possibly beyond. In this section, we present a
-list of specific concepts that need to be taught to members of the community, in
-order to guarantee the base level of competence needed to write efficient code
+to the whole HEP community and beyond. In this section, we present a
+list of specific concepts that need to be taught, in
+order to guarantee the basic level of competence needed to write efficient code
 for the different tasks performed in HEP experiments.
 
-An essential knowledge base that needs to be transferred includes basic
+Essential knowledge that needs to be transferred includes basic
 programming concepts, data structures, basics of code design, error checking,
@@
 include modularity of code design, advanced data structures, evaluation metrics,
-writing tests and working with different types of hardware accelerators.
+writing tests and working with different types of hardware accelerator.
 Cutting-edge approaches such as the use of hardware accelerators (whether GPUs
 or FPGAs), Deep Neural Networks, and Machine Learning should be considered
-important training topics. Additionally, special emphasis should be made on
+important training topics. Additionally, special emphasis should be given to
 reporting results and documenting them.
-The training subjects that the community considers important to pursue are
-listed as follows:
+Some of the training subjects considered important to pursue are
+listed in Appendix \ref{appendix-a---training-topics-of-interest}.
 
-\begin{itemize}
-   \item Basic and Advanced Programming Concepts
-   \begin{itemize}
-      \item Object oriented paradigm
-      \item Compiled langauges (C++)
-      \item Scripting languages (Python, Javascript,...)
-   \end{itemize}
-   \item Algorithmics 
-   \begin{itemize}
-      \item Boost library
-      \item STL algorithms for containers
-      \item R and/or ROOT
-   \end{itemize}
-   \item Existing frameworks (development or application level)
-   \begin{itemize}
-      \item Qt
-      \item ROOT
-      \item experiment specific framework (possibily if of potential interest outside the native experiment)
-   \end{itemize}
-   \item Code design (design patterns)
-   \item Development tools
-   \begin{itemize}
-      \item IDEs (Integrated Development Environment)
-      \item Debuggers
-      \item Profilers
-   \end{itemize}
-   \item Evaluation metrics
-   \item “Trust” metrics such as data driven tests
-   \item Specific software implementation training
-   \item Good practices 
-   \item Code style and clarity
-   \item Scripting and data cleaning
-   \item Reporting results reproducibly
-   \item Writing Documentation
-\end{itemize}
+% ____________________________________________________________________
+% ____________________________________________________________________
+\section{Implementation Roadmap}
+The implementation of training should target multiple training formats and
+knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and
+advanced visualizations, etc.
+People who learn in non-conventional ways should not be excluded from the
+opportunity to learn. 
+%% GS Comment - I'm not really sure what that previous sentence is supposed to
+% mean.
+Since there are different stages of knowledge transfer,
+programmes should be developed that encourage experts to share knowledge. Issues
+considered in this section include global training initiatives, strategies for
+leveraging existing forums, and training tool development.
 
-%____________________________________________________________________
-%____________________________________________________________________
-\section{Challenges}
-
-Within a single experiment, different skill sets are needed. In addition to a
-skill set that contains basic programming language knowledge, testing and code
-management tools, and experiment-specific framework knowledge, there are more
-specialized skills that only a subset of the community needs to know, such as
-software optimization, or low-level hardware interfaces.
+An important point to consider is the difficulty, often experienced in the past,
+in developing large software programs across different experiment communities.
 While there already are experiment-specific training efforts in place, there are
 many needs that are in common between experiments. Therefore, we should probably
-strive to extract the common knowledge and build a common training program. This
+strive to extract the common knowledge and build a common training programme. This
 would reduce duplication of efforts and enable growth of a shared training
@@
    \item Tutors can add material to the web site at a very slow pace (whenever
-   they find time to do it, one slide a week or a chapter a day) \item Their
-   material, made public on a web-site, can be further expanded by collaborators
-   (also at their own pace) or, even better and more productively, by students
+   they find time to do it, one slide a week or a chapter a day) 
+   \item Their material, available public on a web-site, can be further expanded
+   by collaborators (also at their own pace) or, even better and more productively, by students
    who decide to contribute new additions, examples, exercises etc. Such a
@@
 names.
-Subjects covered in a typical Data Carpentry school are given below. This list
-is taken from the curriculum of the CODATA-RDA Research Data Science Summer
-School in progress in Trieste, Italy during July 2017.
+Subjects covered in a typical Data Carpentry school are given in Appendix
+\ref{appendix-b}.
+
+While Software Carpentry is leveraged to build a foundation of knowledge for
+later more advanced concepts, it is important to note much of this material is
+developed specifically for this course and not a part of a larger Software
+Carpentry programme. This course focuses on a shallow but wide building of
+foundational skills approach, introducing many base concepts but not going deep
+into any one concept, with a through line of Open Science and Ethical Data
+Usage.
+
+\section{Conclusions}
+
+The HEP community by and large acknowledges and recognizes the great importance
+of training in the field of scientific computing. This activity should encompass
+several types of {\it students}, from undergraduates, to young researchers, up
+to senior physicists, all of them in need of an appropriately designed training
+path in order to be proficent in their scientific endevours.
+
+We have identified a certain number of problems that need to be overcome in
+setting up an appropriate training programme:
+\begin{itemize}
+    \item Costs and relative funding
+    \item Incentives
+    \item Career paths
+    \item Overall organization across experiments, countries and 
+    corresponding Funding Agencies
+\end{itemize}
+
+For each of these points we provide an overview of the current situation and possible proposals to build a roadmap.
+
+%% GS Comment - this section is incomplete. FIXME.
+
+\newpage
+\begin{appendices}
+
+\sloppy
+\raggedright
+
+\hypertarget{appendix-a---training-topics-of-interest}{%
+\section{Training Topics of Interest}\label{appendix-a---training-topics-of-interest}}
+
+\begin{itemize}
+   \item Basic and Advanced Programming Concepts
+   \begin{itemize}
+      \item Object oriented paradigm
+      \item Compiled langauges (C++)
+      \item Scripting languages (Python, Javascript,...)
+   \end{itemize}
+   \item Algorithms 
+   \begin{itemize}
+      \item Boost library
+      \item STL algorithms for containers
+      \item R and/or ROOT
+   \end{itemize}
+   \item Frameworks (development or application level)
+   \begin{itemize}
+      \item Qt
+      \item ROOT
+      \item experiment specific frameworks (possibly if of potential interest outside the originating experiment)
+   \end{itemize}
+   \item Code design (design patterns)
+   \item Development tools
+   \begin{itemize}
+      \item IDEs (Integrated Development Environment)
+      \item Debuggers
+      \item Profilers
+   \end{itemize}
+   \item Evaluation metrics
+   \item “Trust” metrics such as data driven tests
+   \item Specific software implementation training
+   \item Good practices 
+   \item Code style and clarity
+   \item Scripting and data cleaning
+   \item Reporting results reproducibly
+   \item Writing Documentation
+\end{itemize}
+
+\hypertarget{appendix-b}{%
+\section{Research Data Science Curriculum}\label{appendix-b}}
+
+This list is taken from the curriculum of the CODATA-RDA Research Data
+Science Summer School in progress in Trieste, Italy during July 2017.
 (http://indico.ictp.it/event/7974/):
@@
 
-While Software Carpentry is leveraged to build a foundation of knowledge for
-later more advanced concepts, it is important to note much of this material is
-developed specifically for this course and not a part of a larger Software
-Carpentry program. This course focuses on a shallow but wide building of
-foundational skills approach, introducing many base concepts but not going deep
-into any one concept, with a through line of Open Science and Ethical Data
-Usage.
-
-\section{Conclusions}
-
-The HEP community by and large acknowledges and recognizes the great importance
-of training in the field of scientific computing. This activity should encompass
-several types of {\it students}, from undergraduates, to young researchers, up
-to senior physicists, all of them in need of an appropriately designed training
-path in order to be proficent in their scientific endevours.
-
-We have identified a certain number of problems that need to be overcome in
-setting up an appropriate training program:
-\begin{itemize}
-    \item Costs and relative funding
-    \item Incentives
-    \item Career paths
-    \item Overall organization across experiments, countries and 
-    corresponding Funding Agencies
-\end{itemize}
-
-For each of these points we provide an overview of the current situation and possible proposals to build a roadmap.
-
-%____________________________________________________________________
-%____________________________________________________________________
-%\begin{itemize}
-%
-%   \item How to motivate the community to participate in developing programs? Can some of it come from \& leadership?
-%   \item How to deal with competitiveness in the field? Can’t force people to be interested in forming a community. How can we incentivize this? Building a %vibrant, welcoming community may help.
-%\end{itemize}
+\end{appendices}
 

### 2018-06-21 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Changes from Dario
Add citations to be used in Training and Careers paper to CWP bib file
Fix many typos and minor errors in main paper

@@
 infrastructure areas, such as the EU's Horizon 2020 programme and those from
-the National Science Foundation (NSF) [Refs 1 and 2].
+the National Science Foundation (NSF) \cite{NSF1,NSF2}.
 
@@
 colleagues.
+
 To be effective, the training must be done by people who work at the cutting
@@
    tools or techniques they are expected to use. These might be undergraduates, 
-   but they could also be senior researchers who have not yet or not recently 
+   but they could also be senior researchers who have not yet, or not recently, 
    engaged in software analysis. These are people in need of some kind of formal 
@@
 writing tests and working with different types of hardware accelerator.
-Cutting-edge approaches such as the use of hardware accelerators (whether GPUs
-or FPGAs), Deep Neural Networks, and Machine Learning should be considered
-important training topics. Additionally, special emphasis should be given to
+Additionally, special emphasis should be given to
 reporting results and documenting them.
+
 Some of the training subjects considered important to pursue are
@@
 training material using just a web-browser, and in order to do this efficiently,
-the learning curve has been kept appropriately shallow.
+the necessary learning curve has been kept appropriately shallow.
 
@@
 the GridKa School of computing: the material from this year (2017) has been made
-publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}.
-This constitutes an interesting example of what can be accomplished using this
+publicly available on WTL~\cite{WikiToLearnGridka}.
+This constitutes an intersting example of what can be accomplished using this
 platform; it is just a first example of what's possible, but an inspiring one.
@@
 Finally it should be important to evaluate, if and to which extent, the
-complementary approaches to training (such as schools and dedicated web-site
-courses) could be of mutual benefit in the development of a complete training 
-programme.
+complementary approaches to training, such as schools and dedicated web-site
+courses, could be of mutual benefit, in other words how to make them efficiently cooperate in the
+development of a complete training program.
 
@@
 be connected with other areas such as research and development. Efforts like
-DIANA-HEP \footnote{\url{http://diana-hep.org/}} and aMVA4NewPhysics
-\footnote{\url{https://amva4newphysics.wordpress.com/}} that combine training
+DIANA-HEP~\cite{DIANA-HEP} and aMVA4NewPhysics~\cite{aMVA4NewPhysics}
+that combine training
 and software development are good examples of such ideas in practice. More

### 2018-06-21 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Proper citations for Training Paper (#94)
Add citations to be used in Training and Careers paper to CWP bib file
Fix many typos and minor errors in main paper

@@
 be connected with other areas such as research and development. Efforts like
-DIANA-HEP~\cite{DIANA-HEP} and aMVA4NewPhysics~\cite{aMVA4NewPhysics}
+DIANA-HEP~\cite{DIANA-HEP} and AMVA4NewPhysics~\cite{AMVA4NewPhysics}
 that combine training

### 2018-06-22 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex

--- 2018-04-16 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
@@
 volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
-experiments such as the Square Kilometer Array (SKA) entering this
+experiments such as the Square Kilometre Array (SKA) entering this
 regime. In devising computing models for this era, many
@@
 availability of very high-speed networks may reduce the need for
-% * <briedel@uchicago.edu> 2018-04-04T17:59:27.886Z:
-%
-% > very high-speed networks
-% What is "very high-speed" here? 100G? 1000G?
-%
-% ^ <briedel@uchicago.edu> 2018-04-04T17:59:40.128Z.
 CPU and data co-location. Such networks may allow
@@
 Planned analysis of HL-LHC data and data analysis from other
-% * <briedel@uchicago.edu> 2018-04-04T18:09:21.988Z:
-%
-% > large
-% XENON1T has shown that the collaboration doesnt have to be "large" to require DOMA
-%
-% ^ <briedel@uchicago.edu> 2018-04-09T20:36:43.846Z.
 international collaborations will need to adopt a distributed computing
-model due to its scope. The data management systems that will enable
+model due to their scope. The data management systems that will enable
 this type of analysis will need to provide cost-based optimization of
@@
 projected density gains in current storage technologies and thus will present major challenges. Storage
-% * <briedel@uchicago.edu> 2018-04-05T01:05:55.296Z:
-%
-% > Moore's Law gains
-% This is only a CPU.
-What about storage technology?
-Currently we are adding about 2 TB/year in disk capacity.
-It is unlikely will be at 100 TB/disk in 2023.
-We may have O(10) - O(100) TB SSDs  at a reasonable cost, but we can't be certain.
-What does the future of tape look like?
-% changed to "density gains in current storage technologies." Tape scaling has quite a future but hard drives, less so
-% ^.
 will remain one of the visible cost drivers for HEP computing. The projected growth and cost of the computational resources needed to
@@
   the bandwidth requirements that are needed in enacting site-to-site
-  data transfers between WAN endpoints (\emph{e.g.}, 10-100~Gbps). This
+  data transfers between WAN endpoints (e.g., 10-100~Gbps). This
   requirement on the available WAN bandwidth provides a corresponding
@@
 or content-addressable storage is used, the catalogs need only a single
-% * <briedel@uchicago.edu> 2018-04-05T02:04:26.292Z:
-%
-% > content-addressable
-% This comes a little out of nowhere. Where is this discussed before?
-%
-% ^.
 handle to have a fully descriptive location of a given event or other
 object being sought.
-% * <briedel@uchicago.edu> 2018-04-05T02:05:03.704Z:
-%
-% > or other object
-% This is the first time "other object" is mentioned. What other object? TBranches? It would be better to call this sub-event data like done before.
-%
-% ^.
 
@@
 network services (CDNs) such as the one offered by Akamai. Without CDNs,
-% * <briedel@uchicago.edu> 2018-04-05T02:08:36.870Z:
-%
-% > Akamai.
-% Google also offers one
-%
-% ^.
 on-demand streaming of movies over the Internet or even the delivery of
@@
 Currently, new data-caching strategies are being deployed (e.g., XRootD
-caching proxy servers, server-less caching, Adaptive Replacement Cache-style caching) and may
-% * <briedel@uchicago.edu> 2018-04-05T02:09:39.626Z:
-%
-% > server-less caching
-% In-memory caching?
-%
-% ^.
+caching proxy servers, server-less caching, ARC-style caching) and may
 provide sufficient information on how such strategies can optimize
@@
 One such solution - ``Service X'' - has been proposed in \href{https://bit.ly/2Ix9NIL}{Organizing Data Lakes}.
-The service would start with the data in its persistent format, perform one or more operations on it  (\emph{e.g.} joins, query, decompression, filtering to sub-event level), and finally deliver a data stream to a client.
+The service would start with the data in its persistent format, perform one or more operations on it  (e.g.
+joins, query, decompression, filtering to sub-event level), and finally deliver a data stream to a client.
 
@@
 (e.g.,
-% * <briedel@uchicago.edu> 2018-04-05T15:43:01.526Z:
-%
-% > most
-% "more" rather than "most"?
-%
-% ^.
 from a cloud provider) and have a data management and provisioning
 system that can exploit such resources on short notice. Volatile data
-% * <briedel@uchicago.edu> 2018-04-05T15:43:37.726Z:
-%
-% > Volatile data sources would impact many aspects of the system; catalogs, job brokering, monitoring/alerting, accounting, the applications themselves.
-% Aren't we dealing with somewhat volatile storage in terms of xrootd caches today already? The file _might_ be there, but it isnt guaranteed.
-%
-% ^.
 sources would impact many aspects of the system; catalogs, job
@@
 stability.
-% * <briedel@uchicago.edu> 2018-04-05T15:49:15.646Z:
-%
-% > stability.
-% This paragraph begs the question of whether one needs to move to "direct-access-based archives" if the data is only accessed once or twice. The plenty of tape is not that great in that case.
-%
-% ^.
 
@@
 experiments and resource providers on shared solutions. This does not
-necessarily mean promoting a mono-culture, as different solutions will be
+necessarily mean promoting a monoculture, as different solutions will be
 adapted to certain major classes of use-case, type of site or funding
@@
   application.
-\item Evaluate possible alternative storage formats to the ROOT format, esp.\ storage formats optimized for storage efficiency and/or  data filtering through metadata
+\item Evaluate possible alternative storage formats to the ROOT format, especially storage formats optimized for storage efficiency and/or  data filtering through metadata
 \end{itemize}
@@
 
--\emph{The first two actions should be completed by 2020 but evaluation of CDN/NDN is more long-term}
+\emph{The first two actions should be completed by 2020 but evaluation of CDN/NDN is more long-term}
 
@@
 \item Exploring ways to ease preliminary data filtering steps
-\item Exploring commercially-popular data transfer methods (\emph{e.g,} HTTP)
+\item Exploring commercially-popular data transfer methods (e.g. HTTP)
 \end{itemize}

### 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Apply Graeme suggestions

--- 2018-06-21 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 This constitutes an intersting example of what can be accomplished using this
+the necessary learning curve has been kept appropriately shallow. An interesting exercise in
+this context has recently been made by colleagues of
+the GridKa School of computing: the material from this year (2017) has been made
+publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}.
+This constitutes an interesting example of what can be accomplished using this
 platform; it is just a first example of what's possible, but an inspiring one.
+
+Another example of web-based platform is a collection of online tutorials
+(mostly written in French) hosted on the Gitlab IN2P3
+server~\cite{Code-Swim-Coaches}. Taking advantage of
+the Gitlab ability to host Docker images, those tutorials aim to avoid the
+"damned installation step" that often absorbs half of a training session.
+Similarly to the GridKa School for WikiToLearn, the annual IN2P3 Computing
+Days are an opportunity to refresh and extend the collection of
+tutorials every year. Future work will focus on English translation, and the
+development of a web site which will index the above tutorials, together with
+the best recommended external tutorials.
 

### 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Correct bad rebase effects

@@
 publicly available on WTL~\cite{WikiToLearnGridka}.
-This constitutes an intersting example of what can be accomplished using this
-the necessary learning curve has been kept appropriately shallow. An interesting exercise in
-this context has recently been made by colleagues of
-the GridKa School of computing: the material from this year (2017) has been made
-publicly available on WTL\footnote{\url{https://en.wikitolearn.org/GridKa2017}}.
 This constitutes an interesting example of what can be accomplished using this

### 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
rewrite

--- 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 endeavors.
+
+In the context of web-based training, if the HSF helps to constitute a living
+collection of online tutorials, we could organize regular events such as
+a ``Tutorial of the month'', or some sort of ``like'' system. The
+best authors must be recognized, so that they can showcase their most popular
+online tutorials, just the same as their publications.
 

### 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
rewrite

--- 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 publicly available on WTL~\cite{WikiToLearnGridka}.
-This constitutes an interesting example of what can be accomplished using this
+This constitutes an intersting example of what can be accomplished using this
 platform; it is just a first example of what's possible, but an inspiring one.
-
-Another example of web-based platform is a collection of online tutorials
-(mostly written in French) hosted on the Gitlab IN2P3
-server~\cite{Code-Swim-Coaches}. Taking advantage of
-the Gitlab ability to host Docker images, those tutorials aim to avoid the
-"damned installation step" that often absorbs half of a training session.
-Similarly to the GridKa School for WikiToLearn, the annual IN2P3 Computing
-Days are an opportunity to refresh and extend the collection of
-tutorials every year. Future work will focus on English translation, and the
-development of a web site which will index the above tutorials, together with
-the best recommended external tutorials.
 

### 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
rewrite

@@
 collection of online tutorials, we could organize regular events such as
-a ``Tutorial of the month'', or some sort of ``like'' system. The
-best authors must be recognized, so that they can showcase their most popular
-online tutorials, just the same as their publications.
+a ``Tutorial of the month'', or some sort of ``like'' system for the students
+to support their favorite tutorials. The best authors must be recognized, so
+that they can showcase their most popular online tutorials, just the same as
+their research publications.
 

### 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
fix latex quotes

--- 2018-06-22 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
@@
 endeavors.
-
-In the context of web-based training, if the HSF helps to constitute a living
-collection of online tutorials, we could organize regular events such as
-a ``Tutorial of the month'', or some sort of ``like'' system. The
-best authors must be recognized, so that they can showcase their most popular
-online tutorials, just the same as their publications.
 

### 2018-06-25 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Tutorial of the month 2.

@@
 
+In the context of web-based training, if the HSF helps to constitute a living
+collection of online tutorials, we could organize regular events such as
+a ``Tutorial of the month'', or some sort of ``like'' system for the students
+to support their favorite tutorials. The best authors must be recognized, so
+that they can showcase their most popular online tutorials, just the same as
+their research publications.
+
 It is also critical to incorporate training into grant proposals so that it can

### 2018-07-03 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Add some references and correct a few last points in the training paper

@@
 The IN2P3 authors try to restrict their tutorials to 20 minutes
-of material, as inspired by the Pomodoro technique, so that one can easily jump
-and compose one's own ``menu''. However, it turns out to be tricky to keep such
+of material, as inspired by the Pomodoro technique\cite{PomodoroTechnique}, so
+that one can easily jump and compose one's own ``menu''. However, it turns out to be tricky to keep such
 small tutorials really independant and meaningful when they can be selected at
@@
 Here containers may bring real
-progress, provided that everyone has at least Docker installed on their machine.
-JupyterHub, also, is a technology which enable training session with
-no need for installation (but with a need for network).
+progress, provided that everyone has at least Docker\cite{Docker} installed on
+their machine.
+JupyterHub\cite{JupyterHub}, also, is a technology which enable training session
+with no need for installation (but with a need for network).
 
@@
 to senior physicists, all of them in need of an appropriately designed training
-path in order to be proficent in their scientific endevours.
+path in order to be proficent in their scientific endeavours.
 
@@
 
-For each of these points we provide an overview of the current situation and possible proposals to build a roadmap.
+For each of these points we have provided an overview of the current situation
+and made proposals to improve the situation in HEP.
 

### 2018-07-03 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Update pomodoro time

@@
 This may rest on a low granularity subdivision of independant topics.
-The IN2P3 authors try to restrict their tutorials to 20 minutes
+The IN2P3 authors try to restrict their tutorials to 25 minutes
 of material, as inspired by the Pomodoro technique\cite{PomodoroTechnique}, so

### 2018-07-04 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Training last points (#97)
Commented by John and Graeme make the arguments more concise.

Reorganise the material in Sec 4 to present things in a more
coherent way, organised by topic. Reduce duplication of points
in places.

Remove "Roadmap" as a section 4 title as this is not
really what's prestented. Point to continuing work in the
auspices of the HSF Training WG in the conclusion.

Add some references and correct a few last points.

@@
     \item to establish policies for the hiring and long-term retention of researchers 
-    specialising computing,
+    specialising in computing,
     \item to address the gap in formal software training that is not always 
@@
 colleagues.
-
 To be effective, the training must be done by people who work at the cutting
-edge of the technology, who in many cases are found among the youngest
+edge of the technology and who in many cases are found among the youngest
 researchers. These are also the very same people who are most in need of
-officially recognized credits for the advancement of their careers. Currently
+officially recognized credits for the advancement of their careers. Currently,
 visibility and recognition is given mainly to those working on data analysis
@@
 career paths for researchers specialising in scientific software development. 
-%% GS Comment - is this actually ``motivation'' - it seems more like a solution
-% See suggested title change for this section...
 However, the practical 
@@
 Broadly speaking, two scientific profiles of researchers could be
-envisaged; a physicst could make a detailed plan of what they need, expressed 
+envisaged. A physicst could make a detailed plan of what they need, expressed 
 in a tidy requirements document, and a computer scientist could use this document 
@@
 changing, forcing developers and physicsts to interact closely during the entire
-process of software development; this works smoothly only when both communities have the
-same goals and speak the same language, hence the need for physicists with a good
+process of software development. This works smoothly only when both communities
+have the same goals and speak the same language, hence the need for physicists with a good
 knowledge of computer science.
 
-The first possible career would be the path of a \emph{physicist researcher with
-computing science specialisation}, in some places corresponding to the role of a
-physicist-programmer. It would be conceptually different to that of a
-\emph{computer scientist}: the two have different goals, seek different paths to
-the solution of their problems, and usually do not even adopt the same language in
-expressing problems and solutions. Such a role would have an active role in
-physics analysis, and so meet many of the traditional metrics for an academic
-career path.
+The first possible career path would be that of a \emph{physicist with computing
+science specialisation}, also known as a physicist-programmer in some
+communities. (It would be conceptually different to that of a \emph{computer
+scientist} as the two have different goals, seek different paths to the solution
+of their problems, and usually do not even share the same language.) Such people
+would have an active role in physics analysis, and so meet many of the criteria
+needed for an academic career path.
 
-The second would be the path of a domain-specific \emph{researcher software
+The second would be the path of a domain-specific \emph{software
 engineer}. Such a person has the primary role of developing software and finding
-software solutions, but with a large domain-specific knowledge to both
-understand the use cases and clients, but also of the available solutions.  Such
-posts are almost impossible to develop in the current academic frameworks; but
-while RSEs are understood by many funding agencies and universities as a
-technical post, they are seen as available for short-term engagements with a
-researcher community, delivering a quick fix or a refactoring and then moving
-on. This does not deliver the domain specific expertise required.
+software solutions, but with a large domain-specific knowledge to
+understand relevant use cases and the available solutions. 
+Such posts are almost impossible to establish in current academic
+institutions; rather being seen by many funding agencies and
+universities as short term technical positions. This does not allow domain
+specific expertise to be aquired or retained in our field.
 
@@
 opportunity for impact with commerce and wider society. These three failings
-make an academic appointment challenging or disfavoured compared to other
+make an academic appointment challenging or disfavoured as compared to other
 applicants. However, there are measures that can be taken to address all three
-weaknesses.
-%% GS Comment - This is an odd place to end a section. What are these measures?
-% Where do we list them? Even a forward reference would help.
-
-% This makes the development of specific solutions even more difficult and
-% cumbersome.
-
-%Finally, the career of these two figures follows slightly different paths, with different requirements. 
-%____________________________________________________________________
-%\subsection{Existing Training efforts  \textcolor{pink}{is this needed?....}}
-% why not, shows what is done already
+weaknesses, which we expand on in the remainder of this document.
 
@@
 
-% DOES THIS SECTION ANSWER THE FOLLOWING: 
-% - Different levels of training are required
-
-The HEP user community consists of a set of users with diverse software
+The HEP community consists of people with diverse software
 experience, interests, and time availablity to learn new techniques.  Any
@@
 
-For purposes of training, it is useful to think of all our user community as
-\emph{students}. More specifically, we can broadly identify four different sets
-of students, classified by experience level:
+For purposes of training we can broadly identify classify four
+different experience levels:
 
 \begin{itemize}
-   \item \emph{Beginners}: New collaborators with no knowledge of the 
-   tools or techniques they are expected to use. These might be undergraduates, 
-   but they could also be senior researchers who have not yet, or not recently, 
-   engaged in software analysis. These are people in need of some kind of formal 
-   training in modern computing techniques (basically compiled and scripting 
+   \item \emph{Beginner}: New collaborators with no knowledge of the 
+   tools or techniques they are expected to use. These are people in need of some kind of formal 
+   training in modern computing techniques, such as compiled and scripting 
    languages, together with operating system basics such as filesystems, version 
-   control, and command-line shells).
-   \item \emph{Intermediate level analysts}: people with some experience in 
-   either analysis concepts and/or tools, but  looking to supplement their 
+   control, and command-line shells.
+   \item \emph{Intermediate}: People with some experience in 
+   concepts and tools, but looking to supplement their 
    experience with more recent and modern approaches. 
-   \item \emph{Advanced level analysts}: experts who have mastered current 
+   \item \emph{Advanced}: Experts who have mastered current 
    technologies and implementations and who want to stay up-to-date with 
    new advanced developments.
-   \item \emph{Software developers}: HEP scientists in charge of software 
+   \item \emph{Software Specialists}: HEP scientists in charge of software 
    development in areas not limited only to analysis, such as DAQ systems, 
@@
 At all stages of software and computing training, we should take care to
-encourage \emph{Good Practices Across the Community}, such as error
+encourage \emph{good practices across the community}, such as ensuring error
 checking, modularity of code design, version control, writing tests, etc. All
-the key concepts addressed in the training should not be specific to a
-particular experiment or field of application, but general enough to be useful
-to the whole HEP community and beyond. In this section, we present a
-list of specific concepts that need to be taught, in
+the key concepts addressed in training should not be specific to a particular
+experiment or field of application, but general enough to be useful to the whole
+HEP community and beyond. A number of specific concepts need to be taught, in
 order to guarantee the basic level of competence needed to write efficient code
-for the different tasks performed in HEP experiments.
-
-Essential knowledge that needs to be transferred includes basic
+for the various tasks that need to be performed in HEP experiments.
+These include
 programming concepts, data structures, basics of code design, error checking,
@@
 % ____________________________________________________________________
-\section{Implementation Roadmap}
-The implementation of training should target multiple training formats and
-knowledge transfer, such as videos, wikis, lectures, jupyter notebooks and
-advanced visualizations, etc.
-People who learn in non-conventional ways should not be excluded from the
-opportunity to learn. 
-%% GS Comment - I'm not really sure what that previous sentence is supposed to
-% mean.
-Since there are different stages of knowledge transfer,
-programmes should be developed that encourage experts to share knowledge. Issues
-considered in this section include global training initiatives, strategies for
-leveraging existing forums, and training tool development.
+\section{Advancing Training in HEP}
+
+The implementation of training should employ different training formats, such as
+videos, wikis, lectures, jupyter notebooks and advanced visualizations, etc.,
+so that people can learn in a familiar and effective manner and in such a way
+that experts are encouraged to share their knowledge.
 
 An important point to consider is the difficulty, often experienced in the past,
-in developing large software programs across different experiment communities.
+in developing large software programs across different experiment
+collaborations.
 While there already are experiment-specific training efforts in place, there are
-many needs that are in common between experiments. Therefore, we should probably
-strive to extract the common knowledge and build a common training programme. This
+many needs that are in common. Establishing a common training programme could
+help facilitate the sharing of exprience amongst different experiments. This
 would reduce duplication of efforts and enable growth of a shared training
-culture by accumulating and sharing expertise across different experiments.
-To realise this goal, a possibility could be the institution of a {\em
-federation} of existing training initiatives, a very lightwheight umbrella just
+culture by accumulating and sharing expertise.
+To realise this goal, a possibility could be the creation of a 
+\emph{federation} of training initiatives,
 aimed at improving the efficiency and cost-effectivness of this important
@@
 to convince authors to contribute their work. This issue
-can be alleviated with cautious intellectual property clauses. Another
-problem is that trainers do not like to reuse given material as-is. They usually
-want to refactor it, building their own training story.
-This makes it difficult to have everyone agree on common material.
+can be alleviated by addressing the correct assignment of intellectual property.
+Another problem is that trainers do not like to reuse given material as-is. They usually
+want to refactor it, building their own training history.
+This makes it difficult to have everyone agree on common approaches.
 It is even challenging to agree to host material in the same
@@
 
-Another challenge is the wide heterogeneity in
-student competence. Special care must be given to setup
-a training structure that can cope with both beginners and advanced students.
-This may rest on a low granularity subdivision of independant topics.
-The IN2P3 authors try to restrict their tutorials to 25 minutes
-of material, as inspired by the Pomodoro technique\cite{PomodoroTechnique}, so
-that one can easily jump and compose one's own ``menu''. However, it turns out to be tricky to keep such
-small tutorials really independant and meaningful when they can be selected at
-will.
+Another challenge is the wide range of student competence. Special care must be
+given to setup a training structure that can manage both introductory and
+advanced material. This may be addressed by organising course material as a
+large number of independant topics. The IN2P3 authors try to restrict their
+tutorials to 25 minutes, as inspired by the Pomodoro
+technique\cite{PomodoroTechnique}, so that one can easily jump and compose one's
+own curriculum. However, it turns out to be tricky to keep such small tutorials
+really independant and meaningful when they can be selected at will.
 
@@
 their machine.
-JupyterHub\cite{JupyterHub}, also, is a technology which enable training session
-with no need for installation (but with a need for network).
+JupyterHub\cite{JupyterHub} is a technology which supports training sessions
+without the need for specilaist installations.
 
 %____________________________________________________________________
-\subsection{Initiatives for future training programmes\label{sec:initiatives}}
+\subsection{Initiatives for Future Training Programmes\label{sec:initiatives}}
 
-Some methods that can be used for global training include Massive Online Open
-Courses (MOOCs), hands-on workshops, online knowledge bases, expert trainer volunteer
+Some methods that can be used for location independent training include Massive
+Online Open Courses (MOOCs), hands-on workshops, online knowledge bases, expert trainer volunteer
 networks, and web-based training approaches.
@@
 Coursera~\cite{Coursera} can be
-evaluated and recommended to the community. A novel approach such as WikiToLearn~\cite{WikiToLearn}
-could also be explored to assess potential benefits (as has already been attempted
+evaluated and exploited by the community. A novel approach such as
+WikiToLearn~\cite{WikiToLearn} could also be explored to assess potential benefits (as has already been attempted
 in the context of the GridKa School of Computing~\cite{GridKa}, see below).
@@
 been considered by HSF start-up members, and turned out to be difficult to
-pursue for a number of reasons, but in the future, boundary conditions might
-change, making this approach viable.
+pursue due to the lack of a critical mass needed by Stack Overflow, but in the
+future, boundary conditions might change, making this approach viable.
 
@@
 possible structure for such a network could be the establishment of a 
-{\em federation} of existing schools.
+{\em federation} of existing schools, as discussed in \S\ref{sec:ehcurtrainprog}.
 
-It should be considered that some professors, who act as graduate student
-advisors, might need to be encouraged to make sure their students are properly
-trained. Sometimes, students are instead pushed to learn the bare minimum to get
-the work done, at the expense of a broader training/education curriculum that
-could actually yield improved results further down the line. One incentive
-would be to provide training programmes that also count as course credit, perhaps
-as an elective. This model is already in limited use with some online
-solutions~\cite{TrainingProgram}, but this is not universal.
-Discussions should be started with collaborators at higher education
-institutions to see what the roadblocks or opportunities would be for these
-training sessions to serve double duty.
-It should also be considered that not all students will end their career in
-research or academia: their contribution to the research activity, as students,
-should therefore also provide them knowledge, know-how, and skills considered a
-valuable asset by industry, in order to increase their chances of a career
-outside research.
+\subsection{Web Based Training}
 
@@
 courses are the lack of funding and the lack of available time by experts in the
-field. People with enough expertise or insight usually don't have
-time to devote to prolonged periods of student training, and, even when they can
-find time, the cost of setting up a training course in an effective way is often
-beyond what's made available by funding agencies (funds for travel, hosting,
-setting up a room with a computing infrastructure to allow interactive hands-on
-session, etc.) A possible solution is a completely different approach to
-training (but complementary to the already existing and successful classical
-efforts such as the CERN School of Computing's Bertinoro and KIT ones): instead
-of directly teaching to students, trainees could make use of a web-based
-platform to provide training materials to students. This complementary approach
-has several advantages over traditional ones:
+field. People with enough expertise or insight usually don't have time to devote
+to prolonged periods of student training, and, even when they can find time, the
+cost of setting up a training course in an effective way is often beyond what is
+made available by funding agencies (funds for travel, hosting, setting up a room
+with a computing infrastructure to allow interactive hands-on session, etc.) A
+possible solution is a completely different approach to training, using a
+web-based platform to provide training materials to students. This would be
+complementary to the already existing and successful efforts such as the CERN
+School of Computing's Bertinoro and KIT ones.
+
+The web based approach has several advantages over traditional ones:
 \begin{itemize}
    \item Tutors can add material to the web site at a very slow pace (whenever
-   they find time to do it, one slide a week or a chapter a day) 
+   they find time to do it, one slide a week or a chapter a day) .
    \item Their material, available public on a web-site, can be further expanded
@@
    \item If complemented by the availability of remote virtual machines
-   (possibly through a browser), students could have access to examples and
+   (possibly via a browser), students could have access to examples and
    exercises that are already embedded in their own natural environment: all the
-   necessary tools/libraries needed to implement the exercise will be already
-   available in the virtual machine (possibly a Docker container). With just a
+   necessary tools and libraries needed to implement the exercise will be
+   already available in the virtual machine (possibly a Docker container). With just a
    web browser, students could run complex examples from home, taking advantage
@@
 This constitutes an interesting example of what can be accomplished using this
-platform; it is just a first example of what's possible, but an inspiring one.
+platform; it is just a first example of what is possible, but an inspiring one.
 
@@
 
-  % \item Question-and-answer web sites such as Stack Overflow 
-   %(https://en.wikipedia.org/wiki/Stack_Overflow) 
-   %/ Stack Exchange are very useful. 
-   %(https://en.wikipedia.org/wiki/Stack_Exchange)
-  % \item An expert/tutor volunteer network (office hours style) can be established.
-   %\item Investing in periodic in-person, hands-on workshops can be continued.
-
 %____________________________________________________________________
-\subsection{Leveraging existing forums}
+\subsection{Enhancing Current Training Programmes}
+\label{sec:ehcurtrainprog}
 
@@
 
-% Anecdotally, these are viewed as a great success by the CMS community in
-% preparing new collaborators and might serve as a model for future efforts in
-% ML training.
-
 Over the past decade, MOOCs have been developed by universities and private organizations. 
@@
 avaible elsewhere in an efficient and organized way, since this requires
-adequate organization, volounteers and a suitable infrastructure.
+adequate organization, volunteers and a suitable infrastructure.
 
@@
 \subsection{Resources and Incentives}
+
+It should be considered that some graduate student
+advisors, might need to be encouraged to make sure their students are properly
+trained. Sometimes, students are instead pushed to learn the bare minimum to get
+the work done, at the expense of a broader training/education curriculum that
+could actually yield improved results further down the line. One incentive
+would be to provide training programmes that also count as course credit, perhaps
+as an elective. This model is already in limited use with some online
+solutions~\cite{TrainingProgram}, but this is not universal.
+Discussions should be started with collaborators at higher education
+institutions to see what the roadblocks or opportunities would be for these
+training sessions to serve double duty.
+It should also be considered that not all students will end their career in
+research or academia: their contribution to the research activity, as students,
+should therefore also provide them knowledge, know-how, and skills considered a
+valuable asset by industry, in order to increase their chances of a career
+outside research.
 
@@
 of training in the field of scientific computing. This activity should encompass
-several types of {\it students}, from undergraduates, to young researchers, up
-to senior physicists, all of them in need of an appropriately designed training
+several types of \emph{students}, from undergraduates, to young researchers,
+up to senior physicists, all of them in need of an appropriately designed training
 path in order to be proficent in their scientific endeavours.
@@
 For each of these points we have provided an overview of the current situation
-and made proposals to improve the situation in HEP.
-
-%% GS Comment - this section is incomplete. FIXME.
+and made proposals to improve the situation in HEP. The ideas presented need to
+be developed further and concrete actions in the community need to be
+implemented, which we will undertake in the HSF Training Working Group\cite{HSFTrainingWG}.
 

### 2018-07-04 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Small corrections from Dario

@@
 
-For purposes of training we can broadly identify classify four
+For purposes of training we can broadly classify four
 different experience levels:
@@
 JupyterHub\cite{JupyterHub} is a technology which supports training sessions
-without the need for specilaist installations.
+without the need for specialist installations.
 
@@
    they find time to do it, one slide a week or a chapter a day) .
-   \item Their material, available public on a web-site, can be further expanded
-   by collaborators (also at their own pace) or, even better and more productively, by students
+   \item Their material, publicly available on a web-site, can be further
+   expanded by collaborators (also at their own pace) or, even better and more productively, by students
    who decide to contribute new additions, examples, exercises etc. Such a

### 2018-07-18 CWP/papers/HSF-CWP-2017-02_training/latex/cwp-wg-training-careers.tex
Missed middle initial (#99)
Add new authors to the Training paper

@@
 \author[3]{Peter Elmer,}
+\author[4]{Sergei V. Glayzer}
 \author[1]{John Harvey,}
-\author[4]{Roger Jones,}
-\author[5]{Michel Jouvin,}
-\author[6]{Daniel S. Katz,}
-\author[7]{Sudhir Malik,}
-\author[8]{Dario Menasce,}
-\author[6]{Mark Neubauer,}
-\author[9,a]{Albert Puig,}
-\author[1]{Graeme Stewart,}
-\author[10]{and Christopher Tunnell}
+\author[5]{Roger Jones,}
+\author[6]{Michel Jouvin,}
+\author[7]{Daniel S. Katz,}
+\author[8]{Sudhir Malik,}
+\author[9]{Dario Menasce,}
+\author[7]{Mark Neubauer,}
+\author[10]{Fernanda Psihas}
+\author[11,a]{Albert Puig,}
+\author[1]{Graeme A. Stewart,}
+\author[12]{Christopher Tunnell,}
+\author[10]{Justin A. Vasel,}
+\author[4]{and Sean-Jiun Wang}
 
@@
 \affiliation[3]{Princeton University, Princeton, NJ, USA}
-\affiliation[4]{Lancaster University, Lancaster, UK}
-\affiliation[5]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
-\affiliation[6]{University of Illinois Urbana-Champaign, Urbana, IL, USA}
-\affiliation[7]{University of Puerto Rico, Mayaguez, PR, USA}
-\affiliation[8]{INFN Sezione di Milano-Bicocca, Italy}
-\affiliation[9]{Physik-Institut, Universität Zürich, Zürich, Switzerland}
-\affiliation[10]{University of Chicago, Chicago, IL, USA}
+\affiliation[4]{University of Florida, Gainesville, FL, USA}
+\affiliation[5]{Lancaster University, Lancaster, UK}
+\affiliation[6]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
+\affiliation[7]{University of Illinois Urbana-Champaign, Urbana, IL, USA}
+\affiliation[8]{University of Puerto Rico, Mayaguez, PR, USA}
+\affiliation[9]{INFN Sezione di Milano-Bicocca, Italy}
+\affiliation[10]{Department of Physics, Indiana University, Bloomington, IN,
+USA}
+\affiliation[11]{Physik-Institut, Universität Zürich, Zürich, Switzerland}
+\affiliation[12]{University of Chicago, Chicago, IL, USA}
 \affiliation[a]{Supported by SNF under contract 168169.}

### 2018-09-13 CWP/papers/HSF-CWP-2017-06_preservation/latex/PreservationWG_Arxiv.tex
Add files from WG
These were the files as passed by Mike Hildreth in
May 2018.

\documentclass[11pt]{article}
\usepackage{jheppub}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epsf,epsfig}
\usepackage{subcaption}
%\usepackage{authblk}
%\usepackage{etoolbox}

%\makeatletter
%\patchcmd{\maketitle}{\@fnsymbol}{\@alph}{}{}  % Footnote numbers from symbols to small letters
%\makeatother

%\textwidth = 6.5 in
%\textheight = 9.0 in
%\oddsidemargin = 0.0 in
%\evensidemargin = 1.5 in
%\topmargin = 0.0 in
%\headheight = -0.5 in
%\headheight = .25 in

\renewcommand{\floatpagefraction}{0.95}
\renewcommand{\textfraction}{0.05}

\abstract{In this chapter of the High Energy Physics Software Foundation Community Whitepaper, we discuss the current state of infrastructure, best practices, and ongoing developments in the area of data and software preservation in high energy physics. A re-framing of the motivation for preservation to enable re-use is presented.  A series of research and development goals in software and other cyberinfrastructure that will aid in the enabling of reuse of particle physics analyses and production software are presented and discussed.}

\title{High Energy Physics:\\ Data and Software Preservation to Enable Reuse}

\author[a,1]{M. D. Hildreth,}
\author[b]{A. Boehnlein,}
\author[c]{K. Cranmer,} 
\author[d]{S. Dallmeier-Tiessen,}
\author[e]{R. Gardner, }
\author[f]{T. Hacker,}
\author[c]{L. Heinrich,}
\author[g]{I. Jimenez, }
\author[h]{M. Kane,}
\author[i]{D. S. Katz,}
\author[j]{T. Malik,}
\author[g]{C. Maltzahn, }
\author[k]{M. Neubauer, }
\author[l]{S. Neubert,}
\author[m]{Jim Pivarski}
\author[n]{E. Sexton-Kennedy,}
\author[d]{ J. Shiers,}
\author[b]{T. Simko,}
\author[o]{S. Smith,}
\author[p]{D. South, }
\author[q]{A. Verbytskyi,}
\author[r]{G. Watts,}
\author[e]{J. Wozniak}

\affiliation[1]{Working Group Convener}
\affiliation[a]{University of Notre Dame, Notre Dame, IN, USA}
\affiliation[b]{Thomas Jefferson National Accelerator Facility, Newport News, VA, USA}
\affiliation[c]{New York University, New York, NY, USA}
\affiliation[d]{CERN, Geneva, Swizterland}
\affiliation[e]{University of Chicago, Chicago, IL, USA}
\affiliation[f]{Purdue University, West Lafayette, IN, USA}
\affiliation[g]{University of California Santa Cruz, Santa Cruz, CA, USA}
\affiliation[h]{Soundcloud, Berlin, Germany}
\affiliation[i]{NCSA, Urbana-Champaign, IL, USA}
\affiliation[j]{DePaul University, Chigago, IL, USA}
\affiliation[k]{University of Illinois Urbana-Champaign, Urbana-Champaign, IL, USA}
\affiliation[l]{University of Heidelberg, Heidelberg, Germany}
\affiliation[m]{Princeton University, Princeton NJ, USA}
\affiliation[n]{Fermi National Accelerator Laboratory, Batavia, IL, USA}
\affiliation[o]{McMaster University, Hamilton, Ontario, Canada}
\affiliation[p]{DESY, Hamburg, Germany}
\affiliation[q]{Max-Planck-Institut f{\"u}r Physik, Munich, Germany }
\affiliation[r]{University of Washington, Seattle, WA, USA}

%\renewcommand\Authands{ and }

\begin{document}

\maketitle

\newpage

\hypertarget{overview}{%
\section{Overview}\label{overview}}
As part of the process to produce a Community White Paper (CWP)\cite{CWP} within the efforts of the High Energy Physics Software Foundation (HSF)\cite{HSF}, a working group on Data and Software preservation was established to contribute to the broader discussion of how efforts in this area tied in with the overall goals of the HSF and the HEP community.  This document is essentially the chapter contributed to the CWP by this working group.

The Data and Software Preservation Working Group examined current preservation practices and technologies in HEP and beyond with the goal of describing a set of tools and procedures that will allow analysis preservation, discovery, and reuse. In addition, the group examined the issues around preserving the capabilities to read, simulate, and process the large experimental dataset from the raw data through physics analysis level data.  Clearly, the technical, sociological, and financial issues surrounding this topic are complex. This overview outlines the main points of the discussion, summarizes some interesting new developments, and suggests several areas of research and development that have the potential to advance the tools available for preservation as well as providing more of an impetus to do so.

Several recent documents produced by current projects in this area are readily available.  Because of this, a larger ``White Paper" on knowledge preservation seems redundant.  For more details, including use cases and a detailed description of the efforts within and among the various HEP experiments, the reader is urged to consult the 2015 DPHEP report\cite{DPHEP16}, documentation describing the CERN Open Data Portal\cite{CERNOpen} and the CERN Analysis Preservation Portal\cite{CAP}, the work of the DASPOS project\cite{DASPOS}, and a recent report on public access to data in the Mathematical and Physical Sciences in the United States\cite{MPSOpen}, which provides many other references.

\hypertarget{introduction}{%
\section{Introduction and Motivation}\label{introduction}}
It is easy to state a simple motivation for the preservation of the data and knowledge surrounding High Energy Physics (HEP) experiments. Given the large monetary investments that have been and will be made in particle physics experiments, it is incumbent upon physicists to preserve the data and the knowledge that can lead to scientific results in a manner such that this investment is not lost to future generations of analysts.  The word``knowledge" in this context includes processing and analysis software, documentation, and other components necessary for reusing a given dataset. This includes preserving processed data containing an appropriate level of detail to enable new analysis (``Level 3" data as defined in the 2009 DPHEP report\cite{DPHEP09} ). Preservation of this type can enable, for example, new analyses on older data, as well as a way to revisit the details of a result after publication.  Among other applications, the latter use can be especially important in resolving conflicts between published results, applying new theoretical assumptions, or evaluating different theoretical models.  Preservation enabling reuse offers specific benefits within a given experiment as well.  The preservation of software and workflows such that they can be shared enhances collaborative work between analysts and analysis groups, provides a way of capturing the knowledge behind a given analysis during the review process, enables easy transfer of knowledge to new students or analysis teams, and could establish a manner by which results can be generated automatically for submission to central repositories such as HEPData\cite{HEPDATA}, etc.  At a base level, preservation within an experiment could provide ways of re-processing and re-analyzing data that could have been collected more than a decade earlier. A final series of motivations comes from the potential re-use by others outside of the HEP experimental community.  Significant outreach efforts bringing the excitement of analysis and discovery to younger students has been enabled by the preservation of experimental data and software in an accessible format.  Many examples also exist of phenomenology papers reinterpreting the results of a particular analysis in a new context.  This has been extended further with published results based on the re-analysis of processed data by scientists outside of the collaborations\cite{MITQCD}.  Engagement of external communities, such as machine learning specialists, can be enhanced by providing the capability to process and understand low-level HEP data in portable and relatively platform-independent packages. This allows external users direct access to the same tools and data as the experimentalists working in the collaborations.  Connections with industrial partners, such as those fostered by CERN OpenLab\cite{Openlab} can be facilitated in a similar manner.  Note that the focus of the efforts described here is to both enable and facilitate knowledge preservation so that becomes easy to do.  Policy considerations on what workflows, analyses, and other knowledge should be preserved will necessarily be left to the scientists and the experiments to decide.

\hypertarget{WGOverview}{%
\section{Overview of Working Group Considerations}\label{WGOverview}}
In the process of evaluating what should be addressed within the scope of this working group, a consensus emerged on a set of elements that should form the basis of preservation systems that would enable HEP data to follow the FAIR principles as defined by the Force11 working group\cite{F11}, namely that they be findable, accessible, interoperable, and reusable.  Note that reusability implies, to a large extent, the other attributes, which is why we propose the goal of reuse.  Here a ``preservation system" is taken to mean a set of software tools coupled with a searchable archive that enables reuse. To the extent that they need to be developed, the elements that might comprise such a system are reflected in the research road map discussed below.  Throughout, the assumption is that the bit-wise preservation of data and other knowledge products, while neither a simple nor a zero-cost endeavor, is a solved problem in that items stored on digital media will continue to be retrievable with an acceptably small error rate long into the future.  Clearly, a variety of preservation systems will probably be appropriate given the different preservation needs between large central experiment workflows and the work of an individual analyst.  The larger question is to what extent common low-level tools can be provided that address similar needs across a wide scale of preservation problems.  The sections below contain some suggestions and examples of the kinds of elements that may be useful.

\subsection{Assessment Tools}\label{assessment}
When approaching the complex question of preservation, the question of ``what should be preserved to enable reuse?" must be answered\cite{MPS2}.  Of course, the answer depends on the level of reuse that is desired and the potential preservation framework in which the information will reside.  For any set of preservation systems, a set of guidelines should be developed that reflect the base assumptions on which preservation choices are made, that outline the features of each system, and that enable a user to choose a path toward preservation that is appropriate for the use case.  These guidelines or tools incorporating them can be complemented by different components of the preservation systems. One example could be a software tool executed at run time that is able to automatically provide the user with information on which external components the preserved elements depend. For example, to reuse a piece of analysis software, a potential new user may need access to data catalogues, conditions databases, software configurations, and maybe even experiment-specific software build systems. A tool that is able to observe the execution of the code before it is preserved could supply a list of these external elements, potentially while packaging the code up for preservation. A second example of a complementary tool is the design of elements that determine capture fidelity directly into the preservation framework.  These elements can audit captured information, data, executables, etc., to ensure that the stored elements are sufficient to satisfy the user's desired level of preservation.

\subsection{Capture Tools}\label{capture}
While a host of diverse tools exist for various aspects of digital preservation\cite{WflwTools}, very few provide a complete solution that enables scientists to package up their data, workflows, codes, executables, and documentation for preservation. And, none of these is suitable for the large-scale computational workflows typical of HEP.  The diversity of workflows, analysis practices, scales of computation, and computing techniques makes this a challenging area in which to suggest research.  It is likely that different tools may need to be developed to address the wide scale of preservation needs and scales within HEP and beyond.  It may be possible, however, to develop scalable infrastructure building-blocks that can be used to construct these tools.  No matter what forms these tools take, it is desirable that the capture of information, metadata, and the other necessary elements enabling preservation be done automatically, with as little effort from the user as possible.

The simplest need that should be served by these tools is the ability to run a given executable,``wrap it up", and then re-run it at any future time while obtaining the same results.  The manner in which this is done and the execution environments could be quite varied.  Techniques such as continuous integration could be brought into analysis workflows, for example.  Changes to code, workflows, and data products would be tracked and tested automatically, leaving a complete record of their evolution.  This would naturally serve as a "continuous preservation" environment, assuming there are sufficient personnel to intervene and resolve any problems raised by the integration framework.   At the opposite end of the spectrum, yet still served by continuous integration, is the infrastructure necessary for full-scale production of processed events.  Since reuse of data often implies a comparison of new theoretical models, the infrastructure required to regenerate simulated events must also be preserved in order to insure that these comparisons are possible.  Currently, only minimal central efforts are underway to preserve the capabilities for data reprocessing or the full simulation chain in large HEP experiments. Looking 10 or even 5 years in the future, this may render the current data unusable.  The working group considered this a serious problem that should be addressed before the expertise to do so becomes sparse.

While not directly related to capture tools, the availability and sustainability of infrastructure like CVMFS, which provides so many services to HEP experiments, is also a serious issue.  Can we be sure that CVMFS will provide the appropriate libraries needed to run older software versions several years from now?  Will CVMFS itself have a well-delineated future as computing platforms become more heterogeneous?  These are questions that should be addressed.

\begin{figure}[bthp]
    \centering
 \includegraphics[width=0.9\textwidth]{yadage_workflow_instance.pdf}
    \caption{A sample REANA workflow, constructed from elements specified in Yadage, showing the different components and the order of operation.}
    \label{fig:workflow}
\end{figure}

\subsection{Applications: Preservation-as-a-Service}\label{Applications_PaaS}
In the space of knowledge preservation, several preservation systems have been or are being built that provide infrastructure that users can employ in order to share and reuse elements of workflows, data, etc. One of the first of these systems in HEP is RECAST\cite{RECAST}, which provides the capability to reinterpret the constraints an analysis may place on new physics by comparing new models against a what was found in a completed search analysis on experimental data.  The novelty here is that the RECAST infrastructure maintains the capability to re-run the full simulation and analysis chain given the new physics model as input. If run as a service for theorists, for example, the individual components of the analysis workflow are not accessible.  The re-analysis is performed as a``black box" process, and the results are subsequently made available. However, the technology for preserving, composing, and orchestrating the processing of the different workflow steps within RECAST has evolved to a more general system called REANA  that is under construction within/behind the CERN Analysis Preservation Portal (CAP)\cite{CAP}.  In this framework, an analyst can preserve individual workflow steps in containers and string them together with a series of scripts that enable quite complex workflows.  An example workflow represented by these scripts can be seen in Figure \ref{fig:workflow}.  Within REANA, the preserved executables can be reinstantiated according to the instructions in the workflow scripts by a job scheduler and workflow system that can be attached to any generic computing resource that supports container execution. Other means of instantiating executables, such as the more general approach represented by the Umbrella project\cite{Umbrella}, have also been added to this infrastructure.  A diagram of the REANA system is shown in Figure \ref{fig:REANA}. If appropriate, the executables representing individual workflow steps can be shared, swapped out for different processes, and re-used in different workflows.  The infrastructure also provides a container registry and will eventually have discovery tools that will enable users to find executables that are suitable for their purposes.  Finally, the infrastructure under construction is built entirely from commodity elements, making it possible to install at any site with appropriate computing capability.

\begin{figure}[bthp]
    \centering
 \includegraphics[width=0.9\textwidth]{reana-architecture.pdf}
    \caption{The REANA framework being developed as part of the CERN Analysis Preservation Portal.}
    \label{fig:REANA}
\end{figure}

A second type of preservation-as-a-service infrastructure is represented by common analysis and workflow execution frameworks such as the ALICE``train model". In these projects, individual users can include their analysis workflows in a central process that automatically runs their code as it makes a pass through the experimental dataset. This requires more effort up front, as the analysis codes need to be validated and placed in official repositories before the analysis train is run.  It does, however, provide preservation and versioning of the analysis code and a means of automatic execution, something that is unusual for analysis codes.

A third example of a preservation infrastructure is represented by the work done at DESY\cite{DESY} in building a system that automatically performed periodic validation of a given experimental code base that has been targeted for preservation.  As computing technology, software libraries, compilers, etc., are updated, standard validation scripts are run to assess if changes to the computing environment have caused changes in the physics results or outputs, or if work needs to be done to allow the code to even run after the underlying changes.  All major experiments now use such systems to track changes between software releases during development.  Almost none use these techniques to monitor the viability of older code and software releases.

\hypertarget{Policies}{%
\section{Current Policies of the LHC Experiments}\label{Policies}}

Each of the LHC experiments has adopted a data access and/or data preservation policy, all of which can be found here.  All of the LHC experiments support public access to some subset of the data in a highly-reduced data format for the purposes of outreach and education.  CMS has gone one step further, releasing substantial datasets in an AOD format that can be used for new analysis.  The data release includes simulated data, virtual machines that can instantiate the added analysis examples, and extensive documentation\cite{CMSOpenData}.  ALICE has promised to release 10\% of their processed data after a five-year embargo and has released 2010 data at this time\cite{ALICEOpenData}.  LHCb has promised to release 50\% of the processed data and associated software after five years contingent on having sufficient manpower to produce this.  So far, that has not been the case. ATLAS has chosen a different direction for data release, as is evidenced by this excerpt from their public access policy\cite{ATLASOpenData}:

\begin{quote} ``Data associated with journal publications are also made available: tables and data from plots (e.g. cross section values, likelihood profiles, selection efficiencies, cross section limits, ...) are stored in appropriate repositories such as HEPDATA\cite{HEPDATA}. ATLAS also strives to make additional material related to the paper available that allows a reinterpretation of the data in the context of new theoretical models. For example, an extended encapsulation of the analysis is often provided for measurements in the framework of RIVET\cite{Rivet}. For searches information on signal acceptances is also made available to allow reinterpretation of these searches in the context of models developed by theorists after the publication. ATLAS is also exploring how to provide the capability for reinterpretation of searches in the future via a service such as RECAST. RECAST allows theorists to evaluate the sensitivity of a published analysis to a new model they have developed by submitting their model to ATLAS."
\end{quote}

None of the LHC experiments have made recent public statements addressing the new capabilities of the CERN Analysis Portal and whether or not some use of it will be required (or strongly encouraged).  All of them support some mechanisms for internal preservation of the knowledge surrounding a physics publication.

\hypertarget{Challenges}{%
\section{Key Challenges for Preservation and Reuse}\label{challenges}}
Creating the infrastructure and tools that will allow HEP data to adhere to the FAIR principles of being findable, accessible, interoperable, and, reusable presents several challenges, which are outlined in the sections below.
\subsection{Interoperability}\label{Interoperability}
Almost all of the systems and techniques discussed in the previous section rely on the capture of executables and their preservation for reuse. Today, the optimal technology to facilitate this is the linux container.  Even the systems such as the analysis train model which start from preserved code are generally compiled for a small number of hardware platforms.  However, with the proliferation of computing architectures and the potential customization required for optimal performance, it may be quite difficult to guarantee that any preserved executable can be run on an arbitrary compute node.  The large HEP experiments have already encountered many difficulties in this area when trying to use large HPC facilities in an opportunistic fashion.  As the computing hardware becomes more heterogeneous, new solutions may need to be found for sustainable computing.

An alternate approach would be to ensure that whatever tools are developed for executable and workflow capture function on most platforms where computing is done.  This would guarantee that preservation could be done at the time of execution. Of course, all such systems will face the challenges of tracking the evolution of the underlying hardware once they function at one point in time.

No matter what preservation tools are developed that might enable reuse of software, analysis techniques, and data, if they are not conceived from the beginning as an integral part of the standard frameworks, retrofitting will be nearly impossible.  Much of the information that is needed to enable reuse is transient and available only at the time of processing.  For example, an analysis framework with embedded preservation infrastructure that enables the automatic capture of appropriate details and metadata describing the workflows, the data processed, and the output products would allow the user to compile a record of that computation and preserve it for later reuse, archiving, or sharing with others.  Without automatic capture, assembling the necessary information is painful and time-consuming, which dramatically raises the effort for preservation.

\subsection{Findability}\label{Findability}
For reuse within an experiment or reuse more broadly, the elements that have been preserved need to be linked with some sort of metadata record in a searchable database. While simple search-able parameters can be attached to data and software objects, reuse places much more stringent requirements on documentation and demands a sophisticated metadata description for data objects, executables, workflow steps, and workflows themselves.  Substantial work has been done in this area in order to provide suitable metadata descriptions of computation, workflows, etc.\cite{Ontology} which will be incorporated into the CERN Analysis Preservation Portal, but the current situation represents a work in progress rather than a fully-developed solution.  Further investment in metadata schema implementation, prototyping, and testing will be required before a suitably-descriptive metadata vocabulary can exist.

Once a metadata description that is suitably complex to enable re-use is in place, a registry or database structure will also be required in order to both store analysis elements and to enable users to search for and find them.  While the effort to create such an infrastructure is probably small compared to that needed for the metadata description, it is still a necessary part of the system enabling preservation and reuse.

\subsection{Cost Justification}\label{Cost}
Clearly, adding the infrastructure necessary for preservation and reuse will require a substantial investment in development and, to a lesser extent, hardware for the implementation.  As discussed above, the motivations for doing so are strong.  The premise of this document is that the motivations do indeed justify the cost of development, especially since many of the basic elements of the entire preservation and reuse ecosystem already exist or are currently being envisioned, designed, and built. As the complexities of computational workflows increase and become increasingly ubiquitous across science, there is a exponentially growing need for the sorts of tools and infrastructure described here.

There is, however, a difference between mandated re-use and reproducibility and tools that enable these yet grow organically from the needs of the analysts. If an infrastructure that enables preservation and reuse can be embedded in the frameworks and analysis procedures in a way that makes the analyst's work more productive, then there is little cost to the analyst beyond the adoption of the analysis framework.  Increased productivity benefits everyone, resulting in more rapid production of results.  The barriers to preservation are lowered by making it automatic, and automatically useful to the individual physicist and the larger collaborations. Focusing on the internal benefits of such systems may be the simplest means to accelerate their acceptance.

\subsection{Adoption Challenges and Incentive Structure}\label{Adoption}
An additional challenge is the actual adoption of preservation tools and techniques by the community of physicists. An advantage to embedding them in a framework that is the implementation of a completely new analysis model is that they are there from the start.  As discussed elsewhere, many of the standard techniques of software, such as continuous integration, could have a profound impact on the way analysis is done as well as providing the means for automatic preservation.  Hence, efforts to design and implement new models of data analysis offer a unique opportunity to realize the benefits of preservation, sharing, and reuse as long as these elements are considered as integral parts of the infrastructure.

There has been much discussion about what sorts of modifications might be necessary to the scientific and academic incentive/reward structure in order to create a ``culture of preservation."  While such cultural and institutional changes might drive the adoption of preservation techniques and make re-use of scientific software and data widespread, a top-down approach is difficult to implement and to ensure that it meets the needs of the community.  An effort originating in the community itself, however, and one that makes the community more productive may well be the path of least resistance.

\hypertarget{Connections}{%
\section{Connections with Other Working Groups}\label{Connections}}
Aspects of preservation and reuse underlie many of the themes discussed across a wide variety of working groups whose conclusions appear in the CWP report. Some major overlapping themes are discussed here.

\subsection{Data Analysis and Interpretation WG}
One focus of the Data Analysis and Interpretation Working Group[{\bf cite report}] is designing an analysis framework that enables the analyst to remember and reproduce a piece of an analysis or an entire workflow at some time removed from its original creation. Here the emphasis is not so much on scientific integrity but just on being able to pause and reliably restart an analysis effort. Creating an infrastructure with this capability would incorporate essentially all aspects of a larger preservation system including: provenance tracking of results (which jobs ran when with which input and produced which output?), a means of capturing an executable and its computation environment, and some method of cataloguing the results.  A system of this type may be sufficiently modular to allow the arbitrary composability of different analysis executables, as seen in the REANA framework.  Clearly, the proposed elements of this framework would be very much in alignment with the broad outline of infrastructure discussed in this document. Many of the building blocks that need to be developed can be both widely applicable and shared among preservation systems.

\subsection{Other Groups}
Other elements of preservation and reuse infrastructure are clearly relevant for other working groups within the CWP effort.  For example, tools for executable and environment packaging, workflow preservation and reinstantiation have obvious overlaps with the interests of the Workflow and Resource Management 
WG\footnote{For the detailed overview and statements of each of the WGs, please see Reference \cite{CWP} and the individual WG chapters therein.} and the Facilities and Distributed Computing WG.  To the extent that continuous integration and other frameworks can contribute to software preservation, there is a direct connection to the Software Development WG.  Engagement with the Event/Data Processing Frameworks WG will be required in order to address how the production software for the experiments can be included in preservation frameworks for long-term use.  The Data Access, Organization, and Management WG clearly overlaps with long-term preservation as well.

\hypertarget{Roadmap}{%
\section{Meeting the Challenges: R\&D Roadmap}\label{Roadmap}}
While much of the infrastructure discussed in this document can be built today from existing technologies, several topics for research and development emerged from the discussions.  These are outlined below, followed by a roadmap for the next five years that will allow the exploration of many of these issues.  In proposing the R\&D roadmap, we note that, for many of the issues discussed, closer collaboration with industrial partners is absolutely necessary.  Not only have those in industry thought about many of these problems, it is likely that some of them have already developed paths to solutions.  The rapid advancement and proliferation of various container technologies is just one example of many. Whether these particular solutions are appropriate for HEP and other scientific applications is an open question and will need to be resolved on a case-by-case basis.

One issue that has broad implications for many of the projects in this report is the projected longevity of executables stored in containers.  This is, of course, directly related to the rapid evolution of computing architectures.  A series of important questions in this domain are currently unanswered and could form the basis of an important line of research.  For example, how much can one ``future-proof" executables and software against hardware evolution?  Is there an invariant set of operating system ``hooks" (similar to the POSIX standard\cite{POSIX}) that are necessary to guarantee that the executable within a container will still run some years into the future no matter what the changes in the hardware base?  Will it be possible to``nest" the execution of containers in VMs (in VMs) that preserve execution environments on new platforms? If so, what are the compromises in performance?

A second, related issue is one of software design and implementation, and the question of software sustainability in the modern era of heterogeneous computing architectures.  At what level is it possible to abstract the functional needs of a piece of software from the highly-optimized implementation that is required for computational speed and optimum memory use on a given hardware platform?  Is it possible (or desirable) to design low level libraries that can offer these implementations in a modular fashion, rather than at the level of single instructions?  One possible solution would be``toolkits" of numerical function libraries (e.g. a 21st century CERNLib) that are optimized for a particular class of hardware and that could be called from higher level code. Compilation systems could automatically recognize the platform (or kernel version) and choose the appropriate optimizations.

As mentioned above, an area of investigation that links preservation and reuse with questions raised by the Computing Models, Facilities, and Distributed Computing WG is one of container orchestration.  Clearly, it will be necessary in the future to distribute and execute multiple different containers for large multi-step HEP workflows.  While many container orchestration tools have been developed (e.g., Kubernetes, Helios), do they scale to the size of HEP production needs, what new infrastructure is required in order to run multi-step workflows within a typical HEP production system, and how can all of the elements of the system be preserved and catalogued for re-use?

The evolution of analysis techniques to include``big data" applications such as Apache Spark based on databases or dedicated big data "appliances" raises a host of new issues for preservation and reuse.  Analysis done in this environment follows a different paradigm from the skim/slim/ntuple processes that has dominated HEP analysis for decades.  However, there has been significant development of HEP-specific analysis tools on these platforms\cite{DIANA}.  In principle, analysis scripts or their sub-components can be reused, if they are preserved in a discoverable manner along with the interfaces and middleware that allow them to function. A separate question concerns the data itself.  In principle, one ingests or formats an entire dataset in order to prepare it for analysis.  Since there is no reduced data to associate with the final results, a snapshot of the full dataset may be the most appropriate to designate as the basis for the analysis.  This becomes extremely unwieldy, since either the entire (large) dataset in its "ingested" state needs to be kept, or the procedure by which it is ingested (and the hardware and database configuration on which it was analyzed) needs to be reproducible. Clearly, time-dependent datasets present additional challenges. The coupling of the data and the underlying storage hardware in this paradigm and its implications for preservation and reuse need to be better understood.

One concrete contribution that an R\&D effort in this arena can make is the development of tools that facilitate preservation and reuse.  The high-level goals addressed here require low-level tools that individuals can use to easily capture their executable and a description of their computational environment, for example, or their multi-step workflow. Currently, these common tools do not exist in the HEP domain. In order to make progress in this area, various use cases need to be developed so that tools can be tailored to a particular task. An example set of these have already been enumerated in the context of developing CAP/REANA\cite{CAPExamp}.  Possible uses include those mentioned in the introduction, such as having a complete set of code and documentation for internal review of an analysis, evaluating new theoretical models using the results of old analyses, preserving a benchmark analysis for posterity or to pass to a new student, etc.  Once several use cases are established, the base or common elements that need to be captured and described with metadata can be extracted and generic tools created.  Ideally, this would lead to metadata and provenance capture infrastructure in addition to the preservation of the computations themselves.  With diverse use cases, tools that are applicable to more than one discipline can be developed.  In the case of HEP, the CAP/REANA infrastructure under construction at CERN can be exploited as a test bed for different capture tools and discoverability, making it the ideal development platform.

A potential second instance of development to fill a needs gap is one that targets validation infrastructure.  As mentioned above, despite the work done at DESY directed at the continual evaluation of the viability of a preserved code base, none of the current HEP experiments have implemented such a system.  This may be more a question of motivation rather than technical challenge.  It is an important issue, however, if a given experiment wishes to preserve the capability to run the simulation software for an old detector configuration in order to reinterpret an existing result or to re-analyze ``old" data. An initial investment in what would amount to a small extension of the standard continuous integration and validation infrastructures that the experiments already possess could potentially result in a dramatic reduction in the effort required to resuscitate past software versions.

This leads to the larger question of the preservation of the production workflows and executables of the HEP experiments.  As mentioned above, none of the LHC experiments have projects to preserve the ability to re-use old software releases.  This presents a major obstacle to the re-use of the data for physics analysis, since, without further action, it will be impossible to re-run the detector simulation to understand the response of the detector to new physics signals.  While the infrastructure to capture the executables and the standard workflows exists, the lifetime of the candidate technologies is unknown.  That places this question in line with the rest of the R\&D topics presented in the above paragraphs.  There is an added complexity, here, in the external conditions databases and other external resources that are used to provide constants to the simulation that must also be sustained or separately preserved.

\subsection{Research and Development Roadmap}\label{RandD}
Here, we present a progression of research topics that address the issues discussed above.  Note that many of them should be orchestrated in conjunction with projects conducted by the R\&D programs of other working groups, since the questions addressed are common.  The timescales indicated extend from the start of the program.

\smallskip
\noindent{\bf Near term: (1 year)}
\begin{itemize}
\item 
Demonstrate a fully-functional version of the CERN Analysis Preservation Portal (CAP), the container-based solution for workflow and executable preservation.  This includes the creation of a UI that allows a user to reproduce a workflow with a single command, the implementation of a metadata schema that allows the stored workflows and executables to be discovered with a search engine, the implementation of this search engine, and the provisioning of a scalable back-end system for computation.  Demonstrate composing new workflows from preserved executable steps.
\item 
Demonstrate the capability to provision and execute production workflows for LHC experiments that are composed of multiple independent containers.  Investigate the modifications necessary to the underlying WLCG and workflow management infrastructure in order to construct and deploy these workflows. Begin work on large-scale executable/workflow preservation.
\item
Begin the investigation of limitations on the longevity and ubiquity of container-based solutions to executable preservation.
\item
Collection of analysis use cases and which elements are necessary to preserve to enable re-use.  Begin the creation of a taxonomy of information that should be captured during analysis steps.  Determine if there are analysis workflows that are so different that they represent examples that cannot be captured in a format amenable to CAP.
\item
Begin investigation of preservation issues in ``big data" analysis environments.
\end{itemize}

\noindent{\bf Medium Term: (3 years)}
\begin{itemize}
\item
Develop prototype analysis ecosystem(s), including embedded elements for the capture of preservation information and metadata, and tools for the archiving of this information.
\item
Evaluation of prototypes and suitability for re-use: determination of which elements are missing in order to enable re-use, which elements are difficult to capture.  Needs assessment in order to determine future R\&D necessary to evolve prototypes.
\item
Continue evaluation of the limits of container technologies in the preservation arena.
\item
Continue exploration of large-scale analytic workflows.
\item
Develop prototype for the preservation and validation of large-scale production executables/workflows.
\end{itemize}

\noindent{\bf Longer Term: (5 years)}
\begin{itemize}
\item
Deploy analysis ecosystem that enables re-use for any analysis that can be conducted in the ecosystem.
\item
Deploy system for the preservation and validation of large-scale production executables/workflows
\end{itemize}

\hypertarget{engage}{%
\section{Engaging Communities Beyond HEP}\label{engage}}
Clearly, the problems discussed in this document are common to many applications in other areas of science; many industrial applications may also present solutions or suggest ways forward to arrive at solutions. A particular example of this is represented by the series of tools and procedures developed by the DevOps community.  The modification and adoption of some of these and their application to physics analysis would naturally solve a series of problems related to versioning and workflow preservation.  Beyond this, many examples also exist of, for example, workflow systems in other scientific fields.  Many may not scale to LHC-sized applications, but the system architectures can be quite similar to those developed in HEP.  Unfortunately, few opportunities exist to discover and exploit commonalities in conceptualization and implementation.  This is one case where the strong HEP commitment to these technologies could serve as a basis for reaching out to other scientific disciplines to foster the creation of communities of development centered around the advancement of this work.

\newpage

\begin{thebibliography}{99}
\bibitem{CWP} 
  A.~A.~Alves, Jr {\it et al.},
  {\it ``A Roadmap for HEP Software and Computing R\&D for the 2020s,''}
  arXiv:1712.06982 [physics.comp-ph].
\bibitem{HSF} {\tt http://hepsoftwarefoundation.org/}
\bibitem{DPHEP16} S. Amera, {\it et al.}, {\it ``Status Report of the DPHEP Collaboration: A Global Effort for Sustainable Data Preservation in High Energy Physics,''} arXiv:1512.02019v2.
\bibitem{CERNOpen} Accessible at {\tt http://opendata.cern.ch/about} .
\bibitem{CAP} Accessible at {\tt https://analysispreservation.cern.ch/ } .
\bibitem{DASPOS} Information available at {\tt daspos.crc.nd.edu} .
\bibitem{MPSOpen} {\tt  doi:10.7274/R0542KN7}; information available at {\tt mpsopendata.crc.nd.edu} .
\bibitem{DPHEP09} Z. Akopov, {\it et al.}, {\it ``Status Report of the DPHEP Study Group: Towards a Global Effort for Sustainable Data Preservation in High Energy Physics''}, {\tt https://arxiv.org/ftp/arxiv/papers/0912/0912.0255.pdf} .
\bibitem{HEPDATA} {\tt http://hepdata.cedar.ac.uk/} .
\bibitem{MITQCD} J. Thaler et al., {\tt 10.1103/PhysRevD.96.074003} .
\bibitem{Openlab} {\tt http://openlab.cern} .
\bibitem{F11} M. D. Wilkinson, {\it et al.}, {\it ``The FAIR Guiding Principles for scientific data management and stewardship,''} Scientific Data {\bf 3}, 160018 (2016).  {\tt doi:10.1038/sdata.2016.18}.  Information available at {\tt https://www.force11.org/group/fairgroup/fairprinciples} .
\bibitem{MPS2} For a lengthy discussion, see, for example, https://mpsopendata.crc.nd.edu/index.php/final-report.
\bibitem{WflwTools} See, for example, the COPTR compilation at {\tt http://coptr.digipres.org/Main\_Page} or the POWRR Tool grid: {\tt http://digitalpowrr.niu.edu/tool-grid/} .
\bibitem{RECAST} Kyle Cranmer and Itay Yavin, {\it ``RECAST: Extending the Impact of Existing Analyses,''} JHEP 1104:038, 2011; {\tt https://arxiv.org/abs/1010.2506} .
\bibitem{REANA} {\tt http://reanahub.io/ }.
\bibitem{Umbrella} {\tt https://ccl.cse.nd.edu/software/umbrella/} .
\bibitem{DESY} Dmitri Ozerov and David M South, 2014 J. Phys.: Conf. Ser. {\bf 513} 042043.
\bibitem{CMSOpenData} {\tt http://opendata.cern.ch/research/CMS} .
\bibitem{ALICEOpenData} {\tt http://opendata.cern.ch/education/ALICE} .
\bibitem{ATLASOpenData} {\tt https://cds.cern.ch/record/2002139/files/ATL-CB-PUB-2015-001.pdf} .
\bibitem{Rivet}  {\tt http://rivet.hepforge.org/} .
\bibitem{Ontology} See, e.g., D. Carral {\it et al.}, "An Ontology Design Pattern for Particle Physics Analysis", Proceedings of the 6th Workshop on Ontology and Semantic Web Patterns (WOP 2015) co-located with the 14th International Semantic Web Conference (ISWC 2015), Bethlehem, Pensylvania, USA, October 11, 2015.
\bibitem{POSIX} {\tt http://standards.ieee.org/develop/wg/POSIX.html} .
\bibitem{DIANA} The DIANA-HEP project;  {\tt http://histogrammar.org/} .
\bibitem{CAPExamp} {\tt http://doi.org/10.5281/zenodo.33693} .

\end{thebibliography}

 \end{document}
 \end

### 2018-10-03 CWP/papers/HSF-CWP-2017-14_recotrigger/appendix.tex
Add files via upload

\section{Appendix}

This section compiles short descriptions of some on-going projects within the community of relevance to the research and development roadmap identified by the working group. We provide short descriptions (typically taken from the project web page when available), and links to code and/or recent references. In some cases the code is part of a larger experiment framework, but we nevertheless felt it was important to identify these on-going projects.

We do not attempt to cover two classes of software development projects. First, we do not include projects to develop new code or to improve the performance of existing code purely within a single experiment (or group of experiments) whose code is not easily shared with other parts of HEP either because it is not publically available, because it is strictly tied to use in the processing framework of a specific experiment, or for other reasons. Such projects are numerous in nature and are of critical importance to the success of future experiments, but  typically these works can not be easily made into a commonly available toolkit.

Second are toolkits developed outside of the HEP community. Software trigger and event reconstruction algorithms leverage numerous toolkits developed outside of HEP. Such toolkits are frequently the basis of new research and development projects and are one mechanism to ensure both good community support and efficient code that is likely to evolve with computing technology. These include packages designed for linear algebra, machine learning and other mathematical libraries. It is important that the HEP community encourage the use of these toolkits for future development, but unfortunately the breath of these toolkits make them too numerous to include here.

On-going community software projects identified by the working group:

\vskip 0.5 cm \noindent {\bf ACTS (A Common Tracking Software)}
This project is supposed to be an experiment-independent set of track reconstruction tools. The main philosophy is to provide high-level track reconstruction modules that can be used for any tracking detector. The description of the tracking detector's geometry is optimized for efficient navigation and quick extrapolation of tracks. 
\begin{itemize}
\item Project homepage: \href{https://gitlab.cern.ch/acts/a-common-tracking-sw}{GitLab homepage} 
\item References: ACTS-CDOT-Status-2017-03-07.pdf 
\end{itemize}

\vskip 0.5 cm \noindent {\bf AIDA Tracking Toolkit}
A generic, mostly framework independent, tracking toolkit. Development of this software package  is in the process of being merged with the ACTS project.
\begin{itemize}
\item Project homepage: \href{https://github.com/AIDASoft/aidaTT}{GitHub/AIDASoft/aidaTT} 
\item References: F. Gaede, et. al., “Software toolkit with tracking algorithms”, AIDA Delivery Report D2.8, (2015)  (http://cds.cern.ch/record/1982416). 
\end{itemize}

\vskip 0.5 cm \noindent {\bf Arbor}
ArborPFA is a C++ implementation of a Particle Flow Algorithm developed with the PandoraSDK framework. The idea under this clustering algorithm is based on the topological development of hadronic showers in high granularity sampling calorimeters follows an oriented-tree structure.
\begin{itemize}
\item Project homepage: \href{http://arborpfa.github.io/ArborPFA/}{GitHub homepage} 
\item References: M. Ruan and H. Videau, “Arbor, a new approach of the Particle Flow Algorithm”, in Proceedings, International Conference on Calorimetry for the High Energy Frontier (CHEF2013), p. 316. (2013) (https://arxiv.org/abs/1403.4784).
\end{itemize}

\vskip 0.5 cm \noindent {\bf Cross architecture Kalman Filter}
The aim of this project is to produce a fast and efficient Kalman Filter, while preserving correctness of results, across a variety of architectures.
\begin{itemize}
\item Project homepage: \href{https://gitlab.cern.ch/dcampora/cross_kalman}{GitLab/dcampora/cross\_kalman} 
\item References: D. H. C. Perez, Presentation at CHEP 2016: https://indico.cern.ch/event/505613/contributions/2227256/ 
\end{itemize}

\vskip 0.5 cm \noindent {\bf FastJet}
A software package for jet finding in pp and e+e− collisions. It includes fast native implementations of many sequential recombination clustering algorithms, plugins for access to a range of cone jet finders and tools for advanced jet manipulation.
\begin{itemize}
\item Project homepage: \href{http://fastjet.fr/}{Homepage} 
\item References: M. Cacciari, G.P. Salam and G. Soyez, Eur.Phys.J. C72 (2012) 1896 [arXiv:1111.6097].
\end{itemize}

\vskip 0.5 cm \noindent {\bf HEP.TrkX}
Project to evaluate and broaden the range of computational techniques and algorithms utilized in addressing HEP tracking challenges. Specifically the project will provide a framework to develop and evaluate new algorithms for track finding and classification, that will be demonstrated by applying advanced pattern recognition techniques to track candidate formation. On-going research includes deep neural networks applied to HL-LHC online and offline tracking.
\begin{itemize}
\item Project homepage: \href{https://heptrkx.github.io/}{GitHub} 
\item References: S. Farrell, et. al. “The HEP.TrkX project”, Presentation at the Connecting the Dots workshop, Orsay 2017 (Farrell\_HEPTrkX\_CTD2017.pdf). 
\end{itemize}

\vskip 0.5 cm \noindent {\bf Kalman-Filter tracking on parallel architectures}
This project aims to develop tracking algorithms based on the Kalman Filter for use in a collider experiment that are fully vectorized and parallelized. These will be usable with parallel processor architectures such as Intel's Xeon Phi and GPUs, but yet maintain and extend the physics performance required for the challenges for the High Luminosity LHC (HL-LHC) planned for the 2020s.
\begin{itemize}
\item Project homepage: \href{http://trackreco.github.io}{GitHub} 
\item References: Parallelized Kalman-Filter-Based Reconstruction of Particle Tracks on Many-Core Processors and GPUs - Submitted to proceedings of Connecting The Dots / Intelligent Trackers 2017 (Orsay) arXiv:1705.02876. Kalman filter tracking on parallel architectures - Proceedings of the 22nd International Conference on Computing in High Energy and Nuclear Physics (CHEP 2016) (San Francisco) arXiv:1702.06359.
\end{itemize}

\vskip 0.5 cm \noindent {\bf PandoraPFA}
Toolkit of particle flow algorithms and a framework for developing particle flow based reconstruction approaches.
\begin{itemize}
\item Project homepage: \href{https://github.com/PandoraPFA}{GitHub/PandoraPFA} 
\item References: M.A. Thomson, Particle Flow Calorimetry and the PandoraPFA Algorithm, Nucl. Instr. Meth. Phys. Res. A 611 (2009) 25; arXiv:0907.3577. 
\end{itemize}

\vskip 0.5 cm \noindent {\bf Pixel Tracking on GPUs} 
Fast and parallelizable algorithms for track seeding (in particular for Cellular Automation algorithm).
\begin{itemize}
\item Project homepage: Currently part of \href{https://github.com/cms-sw/cmssw}{GitHub/cms-sw/cmssw}. To be integrated into ACTS project.
\item References: \href{https://indico.cern.ch/event/567550/contributions/2627138/attachments/1512745/2359625/201708_Felice_ACAT17.pptx}{ACAT presentation} 
\end{itemize}

\vskip 0.5 cm \noindent {\bf PODIO}
C++ library to support the creation and handling of data models in particle physics. It is based on the idea of employing plain-old-data (POD) data structures wherever possible, while avoiding deep-object hierarchies and virtual inheritance. This is to both improve runtime performance and simplify the implementation of persistency services.
\begin{itemize}
\item Project homepage: \href{https://github.com/hegner/podio}{GitHub/hegner/podio} 
\item References: B. Hegner and F. Gaede, “PODIO: Design Document for the PODIO Event Data Model Toolkit”, AIDA-2020-NOTE-2016-004 (2016) (https://cds.cern.ch/record/2212785). 
\end{itemize}

### 2018-10-03 CWP/papers/HSF-CWP-2017-14_recotrigger/recoCWPFull.tex
Add files via upload

\documentclass[11pt,letterpaper,fleqn]{article}
\usepackage{graphicx}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{afterpage}
\usepackage{pdflscape}
\hypersetup{
  colorlinks=true,
  urlcolor=blue,
  citecolor=blue
}
\usepackage{subfigure}
\usepackage{cite}
\usepackage{xcolor,colortbl}
\usepackage{wrapfig}
\usepackage{enumitem} %%% uncommenting this line and associated
%% options only saves about 1/4 page
% enumitem options
%\setlist{nosep}
%\setitemize{wide}
%\setlist\bibitem{itemize}{leftmargin=*}
%\setlist\bibitem{enumerate}{leftmargin=*}

\selectcolormodel{rgb}
\definecolor{LHCb dark}{rgb}{0.0000,0.3412,0.6549}%From LHCb logo
\definecolor{UC red}{rgb}{0.8196,0.1176,0.2314} % from logo
\definecolor{brickred}{rgb}{0.8, 0.25, 0.33}
\definecolor{Gray}{gray}{0.85}

%KC see http://en.wikibooks.org/wiki/LaTeX/Page_Layout
\textwidth      6.5 in
\textheight     9.0 in
\hoffset = 0 in
\voffset = 0 in
\headheight = 0 in
\headsep = 0 in
\topmargin = 0 in
\oddsidemargin  = 0.0 in
\evensidemargin = 0.0 in

%KC For ~11 lines per 2 inches. smaller is not easily legible 
\linespread{0.95}

%adjust enumeration spacing
% see http://en.wikibooks.org/wiki/LaTeX/List_Structures#Line_spacing
\let\oldenumerate\itemize
\renewcommand{\itemize}{
  \oldenumerate
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{1pt}
  \setlength{\parsep}{1pt}
}

\def\s2i2{$S^2 I^2$}

\begin{document}
%\linenumbers
\thispagestyle{empty}

\input{titlepageFull.tex}
\newpage
%\tableofcontents
%\newpage
\setcounter{page}{1}

\input{allFull.tex}

%\newpage 
%\bibliographystyle{unsrt}
%\bibliography{recoCWP}

\begin{thebibliography}{99}

\bibitem{HSF2017}
HEP Software Foundation,
\textit{A Roadmap for HEP Software and Computing R\&D for the 2020s},
HSF-CWP-2017-01, arXiv:1712.06982v2 (\href{https://arxiv.org/abs/1712.06982}{Link}).
%\url{http://hepsoftwarefoundation.org/activities/cwp.html}.
%\href{http://hepsoftwarefoundation.org/activities/cwp.html}{HSF CWP page}.

%\bibitem{RECOCWP}
%Software trigger and event reconstruction CWP group, To be published (\href{https://docs.google.com/document/d/1QRO8RA488fwfSg5CSjmvm16-pZpGApSA0l666g_mS_0}{Link}).

\bibitem{Campana2016}
S. Campana,
\textit{Presentation to the 2016 Aix-les-Bains ECFA HL-LHC workshop},
Oct 3 2016 (\href{https://indico.cern.ch/event/524795/contributions/2236590/attachments/1347419/2032314/ECFA2016.pdf}{Link}).

\bibitem{Bordry2016} 
F. Bordry, 
\textit{presentation at the LHC Performance Workshop (Chamonix 2016)}, 
January 25th - 28th, 2016. (\href{http://chamonix-2016.web.cern.ch/}{Link} and \href{http://lhc-commissioning.web.cern.ch/lhc-commissioning/schedule/LHC-schedule-update.pdf}{Link})

\bibitem{LHCb2012} 
LHCb Collaboration, 
\textit{LHCb upgrade framework TDR},  
CERN-LHCC-2012-007 (2012) (\href{https://cds.cern.ch/record/1443882}{Link}).

\bibitem{ALICE2013} 
A. Kluge et al., 
\textit{Upgrade of the ALICE Readout and Trigger System}, 
CERN-LHCC-2013-019 (\href{https://cds.cern.ch/record/1603472}{Link}).

\bibitem{Zimmerman2009} 
F. Zimmerman, PoS EPS-HEP2009, 140 (2009). 
G. Apollinari et al., 
\textit{High-Luminosity Large Hadron Collider (HL-LHC): Preliminary Design Report}, 
CERN-2015-005.%, doi:10.5170/CERN-2015-005.

\bibitem{ATLAS2015} ATLAS Collaboration, \textit{ATLAS Phase-II Upgrade Scoping Document},
CERN-LHCC-2015-020, LHCC-G-166 (2015) (\href{https://cds.cern.ch/record/2055248}{Link}).

\bibitem{CMS2015}
CMS Collaboration,
\textit{Technical Proposal for the Phase-II Upgrade of the CMS Detector}, 
CERN-LHCC-2015-010, LHCC-P-008, CMS-TDR-15-02 (2015) (\href{https://cds.cern.ch/record/2020886}{Link}).

\bibitem{LHCb2017} 
LHCb Collaboration, 
\textit{Expression of Interest for a Phase-II LHCb Upgrade: Opportunities in flavour physics, and beyond, in the HL-LHC era},  
CERN-LHCC-2017-003 (2017) (\href{https://cds.cern.ch/record/2244311}{Link}).

\bibitem{Ohnishi2013} 
Y. Ohnishi, et. al., 
\textit{Accelerator design at SuperKEKB}, 
Prog. Theor. Exp. Phys. 2013 03A011 (2013) (\href{http://dx.doi.org/10.1093/ptep/pts083}{Link}).

\bibitem{BelleII2010} 
Belle II Collaboration, 
\textit{Belle II Technical Design Report}, 
KEK Report 2010-1 (2010) (\href{https://arxiv.org/abs/1011.0352}{Link}).

\bibitem{DUNE2016} 
DUNE Collaboration, 
\textit{Long-Baseline Neutrino Facility (LBNF) and Deep Underground Neutrino Experiment (DUNE) Conceptual Design Report Volume 1: The LBNF and DUNE Projects}, 
arXiv:1601.05471 (2016) (\href{https://arxiv.org/abs/1601.05471}{Link}).

\bibitem{Ball2014} 
A. Ball et al., 
\textit{Future Circular Collider Study Hadron Collider Parameters}, 
FCC-ACC-SPC-0001 1342402, CERN (2014) (\href{https://indico.cern.ch/event/298180/contributions/1658149/attachments/560575/772288/FCC-1401101315-DSC_HadronColliderParameters_V0.3.pdf}{Link}).

\bibitem{Ilten2016} 
P. Ilten, Y. Soreq, J. Thaler, M. Williams, and W. Xue, 
\textit{Proposed inclusive dark photon search at LHCb}, 
Phys. Rev. Lett. 116, 251803 (2016).

\bibitem{aTLAs2017}
%ATLAS Collaboration, \textit{Performance of the ATLAS Trigger System in 2015} ,
%Eur. Phys. J.C77 5 317 (2017),
%\href{https://inspirehep.net/record/1500696}{Link}. 
ATLAS Collaboration,
\textit{Trigger-object Level Analysis with the ATLAS detector at the Large Hadron Collider: summary and perspectives}, 
ATL-DAQ-PUB-2017-003 (2017) (\href{https://cds.cern.ch/record/2295739}{Link}).

\bibitem{CMS2016}
CMS Collaboration,
\textit{Search for narrow resonances in dijet final states at s√= 8 TeV with the novel CMS technique of data scouting},
Phys. Rev. Lett. 117 031802 (2016), (\href{https://cds.cern.ch/record/2149625}{Link}).

\bibitem{LHCb2014}
LHCb Collaboration, 
\textit{LHCb Trigger and Online Upgrade Technical Design Report},
CERN-LHCC-2014-016 (2014) (\href{https://cds.cern.ch/record/1701361}{Link}).

\bibitem{Marshall2015} The Pandora Software Development Kit for Pattern Recognition
J.S. Marshall, M.A. Thomson (Cambridge U.). Jun 16, 2015. 15 pp.
Published in Eur.Phys.J. C75 (2015) no.9, 439

\bibitem{Gray2016} 
L. Gray, 
\textit{Challenges of Particle Flow reconstruction in the CMS High-Granularity Calorimeter at the High-Luminosity LHC}, 
38th International Conference on High Energy Physics (ICHEP 2016), (\href{https://indico.cern.ch/event/432527/contributions/1071751/}{Link}).

\bibitem{Gray2017} 
L. Gray, 
\textit{4 Dimensional Trackers},
(2017) (\href{https://indico.cern.ch/event/577003/contributions/2476434}{Link}).

\bibitem{ATLAS2015} 
ATLAS Collaboration, 
\textit{ATLAS Phase-II Upgrade Scoping Document}, 
CERN-LHCC-2015-020, LHCC-G-166 (2015) (\href{https://cds.cern.ch/record/2055248}{Link})

\bibitem{ALICE2015}
ALICE Collaboration, 
\textit{Technical Design Report for the Upgrade of the Online-Offline Computing System}, 
(\href{https://cds.cern.ch/record/2011297}{Link}).

\bibitem{Abreu2014} 
R. Abreu, 
\textit{The upgrade of the ATLAS High Level Trigger and Data Acquisition systems and their integration}, 
Proceedings of 19th IEEE-NPSS Real-Time conference 2014, pp.7097414 (2014) (\href{https://cds.cern.ch/record/1702466/}{Link}).

\bibitem{Petrucciani2015} 
G. Petrucciani et. al., 
\textit{Mini-AOD: A New Analysis Data Format for CMS}, Journal of Physics: Conference Series 664 7 072052 (2015)

\bibitem{Eifert2015} 
T. Eifert, et. al., 
\textit{Implementation of the ATLAS Run 2 event data model}, 
Journal of Physics: Conference Series, Volume 664, Offline software (2015) (\href{http://iopscience.iop.org/article/10.1088/1742-6596/664/7/072045}{Link}).

\bibitem{Aaij2016}
R. Aaij, et. al., 
\textit{Tesla : an application for real-time data analysis in High Energy Physics}, 
Comput. Phys. Commun. 208 (2016) 35-42 (\href{https://cds.cern.ch/record/2147693}{Link}).

\bibitem{Bird2014}
I. Bird, et. al., 
\textit{Update of the Computing Models of the WLCG and the LHC Experiments},
CERN-LHCC-2014-014, (2014), (\href{http://cds.cern.ch/record/1695401/files/LCG-TDR-002.pdf}{Link}).

\bibitem{Allton2017} 
C. Allton, et. al.,  
\textit{Computing Resources Scrutiny Group}, 
CERN–RRB–2017–056 (2017) (\href{https://indico.cern.ch/event/617821/contributions/2493341/attachments/1446478/2233022/CERN-RRB-2017-056_V2.pdf}{Link}).

\bibitem{DUNE2015}, 
Dune Collaboration, 
\textit{LBNF/DUNE Conceptual Design Report, Annex 4B Expected Data Rates for the DUNE Detectors}, 
(2015) (\href{http://lbne2-docdb.fnal.gov/cgi-bin/ShowDocument?docid=10720}{Link}).

\bibitem{DUNE2017}, 
Dune Collaboration, 
\textit{The Single-Phase ProtoDUNE Technical Design Report},
(2017) (\href{https://arxiv.org/abs/1706.07081}{Link}).

\bibitem{Miyamoto2015} 
A.Miyamoto et al., 
\textit{Computing Requirements of the ILC Experiments}, 
LCC-Note (2015), (\href{http://www-jlc.kek.jp/~miyamoto/SoftwareCommonTask/docs/ILCComputing-EDMS1130485.A.1.1.pdf}{Link})

\bibitem{Richter2016} 
M. Richter, 
\textit{Online Data Compression in the ALICE O2 Facility}, 
Presented at the Computing in High Energy and Nuclear Physics conference (2016),
\href{https://indico.cern.ch/event/505613/contributions/2227264/}{Link}.

\bibitem{ATLASTwiki} 
\href{https://twiki.cern.ch/twiki/pub/AtlasPublic/TriggerOperationPublicResults/BandwidthPie2016.png}{BandwidthPie2016.png}, 
\href{https://twiki.cern.ch/twiki/pub/AtlasPublic/TriggerOperationPublicResults/Time_HLTRate_Stack_2016_07.png}{Time\_HLTRate\_Stack\_2016\_07.png}

\bibitem{CMS2014} 
CMS Collaboration, 
\textit{Description and performance of track and primary-vertex reconstruction with the CMS tracker},  
JINST 9 (2014) 10009 (\href{http://cds.cern.ch/record/1704291}{Link}).

\bibitem{CMS2017} 
CMS Collaboration, 
\textit{Particle-flow reconstruction and global event description with the CMS detector}, 
Submitted to JINST (2017) (\href{https://arxiv.org/abs/1706.04965}{Link}).

\bibitem{LHCbRTCalib2016} 
G. Dujani and B. Storaci, 
\textit{Real-time alignment and calibration of the LHCb Detector in Run II}, 
Journal of Physics: Conference Series, Volume 664, Online computing (\href{https://cds.cern.ch/record/2017839?ln=en}{Link}).

%\bibitem{Brun1996}
%R. Brun and F. Rademakers,
%\textit{ROOT - An Object Oriented Data Analysis Framework}, 
%Proceedings AIHENP'96 Workshop, Lausanne, Sep. 1996, Nucl. Inst. \& Meth. in Phys. Res. A 389 (1997) 81-86. See also \href{http://root.cern.ch}{ROOT homepage}.

\end{thebibliography}

\input{appendix.tex}

\end{document}

### 2018-10-03 CWP/papers/HSF-CWP-2017-14_recotrigger/titlepageFull.tex
Add files via upload

\begin{flushright}
HSF-CWP-2017-16
\end{flushright}

\Large
\begin{center}
{\bf Software trigger and event reconstruction}
\end{center}
\vskip 1cm

\normalsize

\hangindent=1cm
{\bf Abstract:} Realizing the physics programs of the planned and upgraded high-energy physics (HEP) experiments over the next 10 years will require the HEP community to address a number of challenges in the area of software and computing. For this reason, the HEP software community has engaged in a planning process over the past two years, with the objective of identifying and prioritizing the research and development required to enable the next generation of HEP detectors to fulfill their full physics potential. The aim is to produce a Community White Paper (CWP)~\cite{HSF2017} which will describe the community strategy and a roadmap for software and computing research and development in HEP for the 2020s. The topics of event reconstruction and software triggers were considered by a joint working group and are summarized together in this document.

\vskip 1cm

\hangindent=1cm
{\bf Editors}: Vladimir Vava Gligorov$^{12,a}$ and David Lange$^{20}$
%, LPNHE, Universite Pierre et Marie Curie, Universite Paris Diderot, CNRS/IN2P3 (vgligoro@lpnhe.in2p3.fr) and David Lange, Princeton University (David.Lange@cern.ch)

\vskip 0.2cm
\hangindent=1cm
{\bf Contributors}:
\input{authors.tex}

### 2018-10-03 CWP/papers/HSF-CWP-2017-14_recotrigger/a2tex.py

@@ -1,0 +1,231 @@
+#! /usr/bin/env python
+#
+# Process CWP signatory list into a LaTeX author list
+#
+# The data source is expected to be made of 3 text files:
+#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey [- Role ] (i)
+#      with i a footnote number. If there are multiple affiliations, separate AffiliationKeys with &.
+#    - footnotes:txt: the list of footnotes. There must be one footnote per line starting with 'i.'
+#    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
+#      AffiliationKey must not contain any space.
+#
+# These 3 files are typically built from the Google Docs https://docs.google.com/document/d/1tBXwlNnQsxxZA3gVS1_KSpa8wRXGyk250EIsJwJ2T34
+#
+# Note: this script requires Python >= 3.5
+#
+# Initial version written by Graeme Stewart (graeme.andrew.stewart@cern.ch) and Michel Jouvin (jouvin@lal.in2p3.fr)
+
+import sys
+PYTHON_MIN = (2, 6)
+major, minor, _, _, _ = sys.version_info
+if (major, minor) < PYTHON_MIN:
+    python_min_str = '.'.join(str(x) for x in PYTHON_MIN)
+    # Keep old formatter syntax to be sure it works with old version of Python
+    print ("This script requires Python %s or later" % python_min_str)
+    sys.exit(2)
+
+import os
+import re
+from operator import attrgetter
+import argparse
+
+# Define some constants related to application exit status
+EXIT_STATUS_SUCCESS = 0
+EXIT_STATUS_OPTION_ERROR = 3
+EXIT_STATUS_FAILURE = 5
+
+# Default input files
+AUTHOR_FILE_DEFAULT = "authors.txt"
+FOOTNOTE_FILE_DEFAULT = "footnotes.txt"
+AFFILIATION_ADDRESS_FILE_DEFAULT = "address.txt"
+LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
+ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
+
+
+class Footnote():
+    def __init__(self, number_str, text):
+        footnote_symbols = "abcdefghijklmnopqrstuvwxyz"
+        self.mark = footnote_symbols[int(number_str)-1]
+        self.text = text
+    def __repr__(self):
+        return repr((self.mark, self.text))
+
+class Author():
+    def __init__(self, author_line):
[…]

### 2018-10-04 CWP/papers/HSF-CWP-2017-14_recotrigger/latex/allFull.tex

\section{Introduction and Scope}

Realizing the physics programs of the planned and/or upgraded high-energy physics (HEP) experiments over the next 10 years will require the HEP community to address a number of 
challenges in the area of software and computing. For this reason, the HEP software community has engaged in a planning process over the past two years, with the objective of 
identifying and prioritizing the research and development required to enable the next generation of HEP detectors to fulfill their full physics potential. The aim is to produce 
a Community White Paper (CWP) \cite{HSF2017} which will describe the community strategy and a roadmap for software and computing research and development in HEP for the 2020s. 
This activity is organised under the umbrella of the HEP software foundation (HSF). The LHC experiments and HSF have been specifically charged by the WLCG project, 
but have reached out to other HEP experiments around the world throughout the community process in order to make it as representative as possible.

The CWP process was carried out by working groups centered on specific topics. The topics of event reconstruction and software triggers are covered together in 
this document and have resulted from discussions within a single working group. The reconstruction of raw detector data and simulated data and its processing in 
real time represent a major component of today's computing requirements in high-energy physics. A recent projection \cite{Campana2016} of the ATLAS 2016 computing model 
results in $>85\%$ of the HL-LHC CPU resources being spent on the reconstruction of data or simulated events. This working group evaluated the most important components 
of next generation algorithms, data structures, and code development and management paradigms needed to cope with highly complex environments expected in high-energy 
physics detector operations in the next decade. New approaches to data processing were also considered, including the use of novel, or at least, novel to HEP, algorithms, 
and the movement of data analysis tasks into real-time environments.

The remainder of this document is organized as follows. First we discuss how future changes including new and proposed facilities, detector designs, and evolutions in 
computing and software technologies change the requirements on software trigger and reconstruction applications. Second, we summarize current practices and identify the 
most limiting components in terms of both physics and computational performance. Finally we propose a research and development roadmap for the software trigger and event 
reconstruction areas including a survey of relevant on-going work for the topics identified in the roadmap.

Of course any discussion of the computing challenges and priorities for software triggers and reconstruction necessarily overlaps with software domains covered by other 
CWP documents. Indeed, the critical role of real-time reconstruction in allowing the LHC data to be collected in the first place means that the requirements set out here 
will drive much of the R\&D across other areas, whether that be in the development of more performant math libraries, simplified but accurate detector descriptions, or 
new reconstruction algorithms based on machine learning paradigms. Such areas of overlap are noted wherever relevant in the text, and the reader is encouraged to refer to 
the other CWP documents for more details.

\section{Nomenclature}

This document will discuss software algorithms essential to the interpretation of raw detector data into analysis level objects in several contexts. 
Specifically, these algorithms can be categorized as: 
\begin{enumerate}
\item {\bf Online}: Algorithms, or sequences of algorithms, executed on events read out from the detector in near-real-time as part of the software trigger, typically 
on a computing facility located close to the detector itself.
\item {\bf Offline}: As distinguished from online, any algorithm or sequence of algorithms executed on the subset of events preselected by the trigger system, or generated 
by a Monte Carlo simulation application, typically in a distributed computing system.
\item
{\bf Reconstruction}: The transformation of raw detector information into higher level objects used in physics analysis. 
Depending on the experiment in question, these 
higher level objects might be charged particle trajectories (``tracks''), neutral or charged particle calorimeter clusters, Cherenkov rings, jets, and so on. 
 A defining characteristic of “reconstruction” which separates it from “analysis” is that the quality criteria used in the reconstruction to, for example, minimize the number of fake tracks, should be general enough to be used in the full range of physics studies required by the experimental physics program.
Reconstruction algorithms are also typically run as part of the processing carried out by centralized computing facilities.
\item
{\bf Trigger}: the online classification of events, performed with the objective of reducing either the number of events which are kept for further ``offline'' analysis, 
the size of such events, or both. In this working group we were only concerned with software triggers, whose defining characteristic is that they process data without a 
fixed latency. Software triggers are part of the real-time processing path and must make processing decisions quickly enough to keep up with the incoming data, possibly 
with the benefit of substantial disk buffers.
\item
{\bf Real-time analysis}: The typical goal of a physics analysis is to combine the products of the reconstruction algorithms (tracks, clusters, jets...) into complex 
objects (hadrons, gauge bosons, new physics candidates...) which can then be used to infer some fundamental properties of nature (CP asymmetry, lepton universality, Higgs couplings...). 
We define as real-time analysis any physics analysis step that goes beyond object reconstruction and is performed online within the trigger system, in certain cases 
using simplified algorithms to fit within the trigger system constraints. Real-time analysis techniques are so far quite experiment-specific. Techniques may include 
the selection and classification of all objects crucial to the calibration of the detector performance, evaluation of backgrounds, as well as physics searches that are 
otherwise impossible given limitations of data samples passing the trigger and saved for offline work.
\end{enumerate}

The online and offline algorithms have traditionally been viewed as related, but at least partly separate due to their differing goals and requirements. Because the online 
algorithms have to run on all events read out from the detector,\footnote{Of course online algorithms are a sequence, and some of the algorithms in this sequence may select 
events for processing by later algorithms. So not every online algorithm must run on every event read out from the detector, but when grouped together the ensemble of ``online'' 
algorithms process all events read out from the detector in near-real-time.}
they typically must be executed on dedicated computer facilities (e.g. a server farm) located 
near to the detector in order to avoid prohibitive networking and data transfer costs.\footnote{Note that this is not an argument about latency. A hardware fixed latency trigger 
has a set maximum time to evaluate any single event. A software trigger has no such constraint, and instead has some average processing time to evaluate events, given by a 
combination of the processing power available, the size of the disk buffers which can store data waiting to be analyzed, and the network bandwidth available to move events to this disk buffer. } 
Such dedicated farms typically have a small (if any) amount of disk space 
to buffer events while waiting for them to be processed, and the online algorithms must therefore be tuned to strict timing and memory consumption requirements. In contrast 
the offline algorithms run only on a subset of events which have been saved to long term storage. They must still execute within a reasonable time so that their output is made 
available for analysis in a timely fashion, and to fit the computing resources available to the experiment, but these pressures are generally much less severe than for online 
processing. In addition, online algorithms often run in dedicated frameworks, with additional layers of safety or control compared to their offline counterparts.

Increasingly, however, the difficulties of maintaining these parallel software environments is driving online algorithms to become special cases of the offline ones configured 
for increased speed at the cost of precision, but otherwise relying on the same underlying codebase. This development is also driven by the desire to reduce systematic uncertainties 
introduced by having separate online and offline reconstruction, in particular for ``precision'' measurements.
This physics motivation to use offline algorithms online can lead to performance improvements in offline algorithms beyond what might have otherwise been achieved. In turn, such 
improvements free up offline resources, notably for producing the large samples of simulated events which will be needed in the HL-LHC period. We therefore assume that this trend 
will continue and intensify on the timescale considered in this document.

\section{New Challenges anticipated on the 5-10 year timescale}

This section summarizes the challenges identified by the working group for software trigger and event reconstruction techniques in the next decade. We have organized these 
challenges into those from new and upgrade accelerator facilities, from detector upgrades and new detector technologies, increases in anticipated event rates to be processed 
by algorithms (both online and offline), and from evolutions in software development practices.

\subsection{Challenges posed by Future Facilities }

Here we briefly describe some of the relevant accelerator facility upgrades and their impact on experimental data samples. These will be the basis of our discussion of how 
software trigger and event reconstruction algorithms must evolve.
\begin{enumerate}
\item 
LHC Run 3: Run 3 of the LHC is expected to last three years, starting in 2021, after the second long shutdown of the LHC. Having already exceeded the design luminosity of the 
LHC facility, there is no significant increase in instantaneous luminosity expected for the CMS and ATLAS experiments. The CMS and ATLAS experiments expect to accumulate up to 
300~fb$^{-1}$ of data each by the end of this run \cite{Bordry2016}. This is a nearly 10x increase over the samples of 13~TeV data collected through 2016. Both LHCb and ALICE will undergo 
major upgrades for Run 3 (described in the next section) : LHCb will have an instantaneous luminosity five times higher than in Run 2 \cite{LHCb2012}, while ALICE will 
upgrade \cite{ALICE2013} its readout and real-time data processing in order to enable the full 50~kHz Pb-Pb collision rate to be saved for offline analysis.
\item
High-luminosity LHC (HL-LHC): The HL-LHC project \cite{Zimmerman2009} is currently planned to begin operations in 2026. It is an upgrade to LHC, expected to result in an increase of up 
to a factor of 10 in instantaneous luminosity over the LHC design (so up to $10^{35}$~cm$^{-2}$s$^{-1}$). The beam energy of 14~TeV and 25~ns bunch spacing imply a considerable increase in the 
number of simultaneous collisions (pileup) seen by experiments. Operating scenarios under study are pileup of 140 or 200 for ATLAS \cite{ATLAS2015} and CMS \cite{CMS2015} at a 25~ns bunch 
spacing, both possibly with luminosity leveling techniques that would provide a relatively constant luminosity throughout a fill. In addition, LHCb \cite{LHCb2017} is planning a 
consolidation during LS3 for initial HL-LHC operations, with improvements to various detector components, followed by a potential later Upgrade II with a pileup of 50 to 60, to 
run from roughly 2030 onwards. The HL-LHC project is expected to run for at least 10 years. 
\item
Super KEK-B: The Super KEK-B facility \cite{Ohnishi2013}, together with the Belle-II experiment \cite{BelleII2010}, plans to achieve a 40 times increase in instantaneous luminosity over that 
achieved by the previous generation of $e^+e^-$ colliders operating on or near the Upsilon(4S) resonance (KEK-B and PEP-II). Super KEK-B should begin production data taking in 2018 and 
run until at least 2024 with a goal of 50~ab$^{-1}$ of data collected (e.g., nearly 60 billion $B$-$\bar B$ pair events recorded). 
\item
Long Baseline Neutrino Facility (LBNF): The LBNF project \cite{DUNE2016} is planning a high-intensity neutrino beamline from Fermilab to the SURF underground facility in South Dakota. 
Based on the NuMI beamline, LBNF plans a proton-beam power of 1.2~MW ($7.5\cdot 10^{13}$ protons per cycle) later upgraded to 2.4~MW ($1.5-2.0 \cdot 10^{14}$ protons per cycle). The facility expects 
to operate for 20 years starting 2025.
\item 
Linear Colliders (ILC and CLIC): two electron-positron collider projects are currently under study, the International Linear Collider (ILC), to be built in Japan and the Compact 
Linear Collider (CLIC) at CERN. The ILC will operate at a center of mass energy of 250-500~GeV. With a nominal luminosity of $1.47\cdot 10^{34}$~cm$^{-2}$s$^{-1}$ at 500~GeV and an expected raw data 
rate of 1~GB/s the two planned experiments will each accumulate up to 10~PB/year. Both colliders plan to run without any hardware trigger, requiring a fast and efficient prompt 
reconstruction and event building.
\item
Future Circular Collider (FCC): A 100 TeV facility is being studied as the next step in the energy frontier projects after HL-LHC \cite{Ball2014}. This could be realized as a 100 km 
circumference tunnel using 16 T magnets. Scenarios under discussion for such a facility include up to 1000 pileup events and the need to be hermetic up to a much larger eta 
(eg. $\eta=6$) than planned for the HL-LHC upgrades to the ATLAS or CMS experiments. 
\end{enumerate}
Several common themes are apparent from these facility plans. Accelerator operating conditions continue to evolve towards higher intensity and higher energy, as required to bring
new discovery potential. This means more complex and higher particle density environments from which signs of new physics must be extracted by trigger systems, event reconstruction 
algorithms and by analysis. This complexity brings new requirements and challenges to detector design as well as software algorithms, where detection efficiency needs to be maintained 
in more complex environments without increasing false-positive rates.

For HL-LHC, the increased pileup of many interactions in a single crossing leads to several critical problems. The higher particle multiplicities and detector occupancies will lead 
to a significant slowdown in all reconstruction algorithms, from the tracking itself to the reconstruction in other devices such as the electromagnetic calorimeter and RICH detectors. 
In addition to making the algorithms slower, pileup also leads to a loss of physics performance, for example :
\begin{itemize}
\item Reduced reconstruction efficiency in Electromagnetic calorimeters;
\item Increased association of tracks to wrong primary vertices;
\item Reduced efficiency of identifying isolated electrons, muons, taus, and photons;
\item Reduced selection efficiencies for electrons and photons;
\item Reconstruction efficiencies for hadronic tau decays and b-jets;
\item Worse energy resolution for electrons, photons, taus, jets, and missing transverse energy;
\item Worse reconstruction of jet properties (substructure for top/$W$-tagging, quark/gluon discrimination, etc);
\end{itemize}

The central challenge for object reconstruction at HL-LHC is thus to maintain excellent efficiency and resolution in the face of high pileup values, especially at low object transverse momenta. 
Detector upgrades such as increases in channel density, high precision timing and improved detector geometric layouts are essential to overcome these problems. For software, particularly for 
triggering and event reconstruction algorithms, there is an additional need not to dramatically increase the event processing time. A comprehensive program of studies is required to assess the 
efficiencies and resolutions for various approaches in events with up to 200 pileup interactions.

The increase in event complexity also brings a ``problem'' of overabundance of signal to the experiments, and specifically the software trigger algorithms. Traditional HEP triggers select a small 
subset of interesting events and store all information recorded in such events. This approach assumes first that only a very small fraction of collisions potentially contain interesting physics, 
second that the features of such events will be strikingly different than the features of ``uninteresting'' events, and third that discarding any information in an event would prevent later correcting 
any defects in the real-time processing, or preventing a full understanding of the process of interest. The evolution towards a genuine real-time processing of data has been driven by a breakdown 
in the first two assumptions and technological developments which means that the third assumption is no longer as worrying as it once was.

An illustrative example is the search for low-mass dark matter at the LHC, such as the  search for dark photons at LHCb \cite{Ilten2016}.  Since interactions between dark photons and Standard Model 
(SM) particles have very low cross sections, the probability of producing dark photons in proton-proton collisions is extremely small. Thus, discovering them at the LHC will require an immense 
number of proton-proton collisions and a highly efficient trigger. The key problem is that when the dark photon lifetime is small compared to the detector resolution, which is the case in much 
of the interesting parameter space, there is an overwhelming irreducible SM background from off-shell photons producing di-muon events. The current LHCb trigger configuration discards about 
90\% of potential dark photon decays, and a variety of other beyond the Standard Model and SM signals where the LHCb detector itself has good sensitivity. Once this stage is removed (and the 
luminosity is increased), the potential signal rate will increase by a factor of 50. However, the irreducible SM-background rate also increase significantly. Since offline storage resources 
are not expected to increase at nearly this rate, the data must be compressed much more than is now within the online system. In other words, techniques developed in the offline analysis for 
background must be integrated into the online software trigger.

Similar considerations apply to low-mass dijet searches at ATLAS/CMS, where the enormous background rate from quantum chromodynamics (QCD), which grows linearly with pileup, limits the study of 
physics at the electroweak scale in hadronic final states given current techniques. Trigger output bandwidth limitations mean that only a tiny fraction of lower-energy jets can be recorded, and 
hence the potential statistical precision of searches involving these low energy jets is vastly reduced. This is again relevant in the context of dark matter searches, here for light mediators 
between quarks and dark matter particles with low coupling strength. Further details are described in \cite{aTLAs2017, CMS2016}.

These considerations also apply to other areas of the LHC physics program. For example, by Run~3 \cite{LHCb2014}, most LHC bunch crossings will produce charm hadrons at least partially in the LHCb 
acceptance, and all bunch crossings will produce strange hadrons detectable in the LHCb acceptance. Even more dramatically, in the HL-LHC era, a potential Upgrade II of LHCb will have to cope 
with multiple reconstructible charm hadron signals per bunch crossing, and even semileptonic beauty-hadron decays will become more abundant than can be stored offline. The ability of the LHC to continue 
improving our knowledge in these areas will therefore entirely depend on the ability to select the specific signal decay modes in question at the trigger level. Taken together, the overabundance 
of interesting signals, which mandates more complex event reconstruction at the trigger level, and the increasing event complexity, which makes the event reconstruction ever more expensive, will 
influence the design and requirements of trigger and reconstruction algorithms over the next decade.

\subsection{Challenges and Opportunities from Evolution in Experimental apparatus}

A number of new detector and hardware trigger concepts are proposed on the 5-10 year timescale in order to help in overcoming the challenges identified above. In many cases, these new technologies
bring novel requirements to software trigger and event reconstruction algorithms. These include:
\begin{enumerate}
\item
High-granularity calorimetry: Experiments including CMS (for HL-LHC), ILC, CLIC and FCC are proposing very high granularity calorimeters in order to better separate showers from closeby particles in a 
high-density (i.e., high-pileup) environment and thereby improve the jet energy resolution. This granularity brings significant computational challenges as there is much more information to process in 
order to fully take advantage of these devices. Efficient algorithms \cite{Marshall2015,Gray2016} are needed to fully optimize the ambiguity reduction capability of the signals from millions of channels 
within finite computing times.

\item
Precision timing detectors \cite{Gray2017} for charged particles: Experiments including ATLAS, CMS and LHCb are pursuing timing detectors with precision approaching 30~ps. This information is another tool 
for dramatically reducing the effect of pileup on triggering and reconstruction algorithms in particular in the areas of tracking and vertex finding. Integrating timing information into Kalman filtering 
and vertex disambiguation algorithms can improve both physics performance and time-to-process events online and offline. Making current tracking plus new timing detectors effectively integrate into a 
performant 4D-detection system will necessitate the development of new reconstruction algorithms that make use of both spatial and timing information.

\item
Hardware triggers based on tracking information: Hardware trigger systems designed to identify tracks down to 2 GeV of transverse momentum are a valuable tool for ATLAS and CMS \cite{ATLAS2015,CMS2015}. This 
technology will enhance the ability to trigger on a range of physics signatures including isolated leptons, multi-jet signatures and displaced vertices, as well as mitigating the effects of pileup 
on these objects. Once an event is triggered, the results obtained from these triggers can also be a valuable tool in the software trigger, for example by seeding and therefore speeding up the 
subsequent reconstruction. Similar systems, with an ability to go below 500 MeV of transverse momentum are also under consideration for the LS3 Consolidation and Upgrade II of LHCb, and could be 
particularly valuable in the study of strange hadrons if the thresholds could be reduced down to 100 MeV of transverse momentum.

\item
Data streaming techniques: Experiments with no hardware trigger allow a software trigger to see all data before events are rejected from further processing. There is a clear advantage in physics 
capability if the data streaming capability is sufficient and if software triggers are efficient, effective and reliable enough for the task. There is always an advantage if additional algorithms 
can be run on an event before a decision must be taken about its importance. In the case of LHCb this means a facility and algorithms that are capable of processing with a sustained throughput 
of 30~MHz \cite{LHCb2014}. Similarly, the Alice experiment plans a 50~kHz interaction rate with no, or simple minimum bias, hardware trigger, and will stream 3~TB/s from the TPC in a common 
online-offline computing system \cite{ALICE2015}.
\end{enumerate}

\subsection{Challenges from Event rates and real-time processing}

Trigger systems for next-generation experiments are evolving to be more capable, both in their ability to select a wider range of events of interest for the physics program of their experiment, 
and their ability to stream a larger rate of events for further processing.  ATLAS and CMS both target systems where the output of the hardware trigger system is increased by 10x over the current 
capability, up to 1~MHz \cite{ATLAS2015,CMS2015}. In other cases, such as LHCb, the full collision rate (between 30 to 40~MHz for typical LHC operations) will be streamed to real-time or quasi-realtime 
software trigger systems. It is interesting to note that because the ATLAS/CMS events are O(10) times larger than those of LHCb (roughly O(1) vs O(0.1) MB/event), the resulting data rates are 
rather similar, namely 1-5~TB/s.

This enhanced capability naturally increases the demands on software trigger algorithms and offline reconstruction algorithms. In many cases, the current trigger bandwidth of experiments is limited 
by the offline processing and storage capabilities of an experiment rather than its ability to physically write out data to disk or tape for further processing.  This can be due either to the time 
to fully reconstruct events for analysis (corresponding to a fixed set of CPU resources)  or by the size of the analysis data itself (corresponding to a fixed set of disk resources). Current 
experiments are therefore constantly working to reduce the CPU needs to reconstruct events and storage needs for analyzing them, through a combination of code improvements (refactoring, vectorization, 
exploitation of optimized instruction sets) or through entirely rewriting algorithms.

This is an ongoing process that continues to yield throughput improvements, aided by improved code analysis tools, modern compilers and other tools. For many experiments, there is also a potential 
tradeoff between physics quality and CPU needs. A typical example is that track reconstruction requires less CPU if the transverse momentum threshold for tracks to be identified is raised, thereby reducing the 
overall combinatorics. This sort of CPU improvement is almost never desirable as it reduces the overall physics output. Instead software trigger and reconstruction applications are pressed to 
include more and more algorithms over time. Typical examples for ATLAS and CMS are new jet-finding, isolation, or pileup subtraction approaches, and, more generally, algorithms developed targeting 
specific physics use cases, for example in the trigger-level search for low-mass hadronic resonances in ATLAS described in \cite{Abreu2014}. Conversely, the need to reconstruct higher multiplicity, or 
increasingly soft, decays challenges LHCb’s applications.

Recent examples of significant storage reductions for the analysis data tier include the CMS MiniAOD \cite{Petrucciani2015} and the ATLAS xAOD \cite{Eifert2015}, both deployed for LHC Run 2 analysis. 
Improvements can be achieved by refinements in physics object selection (e.g., saving less uninteresting information), and by enhanced compression techniques (using lossless or lossy methods).

Real-time analysis is the only solution for those signals which are so abundant that they cannot all be saved to disk, or for which discriminating against backgrounds requires the best possible 
detector calibration, alignment, reconstruction, and analysis. In this case, some or all of the output of the software trigger system also serves as the final analysis format. Its development is 
justified by the need to conduct the broadest possible program of physics measurements with our existing detectors. This is critical for two reasons: first, because we do not want to miss any 
signatures of New Physics around the electroweak scale, whether direct or indirect, but also because, even if the New Physics scale lies beyond the reach of current detectors, we must probe the 
widest possible parameter space in order to motivate and guide the design of future colliders and experiments. This is particularly true given the cost and long timescale of such future facilities.

An early version of this approach consisted of keeping only a limited set of physics objects (e.g., jets) as computed in real-time processing, and proof-of-concept implementations exist since LHC 
Run~1 \cite{Aaij2016,Abreu2014,CMS2016}. In order to perform precision measurements in real-time, however, it is critical to be able to keep data long enough (hours or days, depending on the experiment) 
to perform quasi-real-time calibrations and a final offline analysis quality reconstruction in the trigger system itself. This approach was commissioned by LHCb in 2015. As a result, roughly one 
third of the LHCb experiment’s Run~2 trigger selections now use the real-time reconstruction to reduce the amount of data kept for further analysis. Advantages of these approaches include an order 
of magnitude reduction in data volume from not saving the raw detector data, potentially reduced systematic errors in analysis due to differences in algorithms or calibrations used in the online 
and offline processing, and a reduction or elimination of the need for offline processing.

%High data rates made possible by reducing event size and by processing online abundant data exposes the problem with Monte Carlo (MC) simulation support for those data. MC production is the primary 
%consumer of computing resources. Luminosity increase will proportionally increase the amount of necessary MC samples, especially for pileup contribution in forward regions of calorimeters. The full 
%range GEANT simulation of these dense events will require corresponding increase of computing resources. The alternative solution is based on the fact that actual readout granularity of calorimeters 
%and other detectors, which are heavy for simulation, is significantly lower than the spatial precision of detailed GEANT simulation. This gives an opportunity to substitute detailed GEANT simulation 
% much faster ML-based generative models, generating detector response with just right topology and granularity.

These advantages allow an entire category of physics to be probed by HL-LHC experiments that would not otherwise be considered. Among the challenges posed by these approaches are the need for 
very robust event reconstruction algorithms, the need to derive detector calibrations sufficient for final analysis within this short processing window, and the need to plan analyses sufficiently 
in advance of data taking so that the choices made in the real-time analysis will be robust against eventual systematics studies.

\subsection{Challenges from Evolutions in Computing technology}

This section summarizes recent, and expected, evolutions in computing technologies. These are both opportunities to move beyond commodity x86 technologies, which HEP has used very effectively over 
the past 20 years, and significant challenges to continue to derive sufficient event processing throughput per cost to enable our physics programs at reasonable computing cost.  A full description 
of this technology evolution and its effect on HEP is beyond the scope of this document \cite{Bird2014}. Here we identify the main technology changes identified as driving out research and development 
in the area of software trigger and event reconstruction:
\begin{itemize}
\item
{\bf Increase of SIMD capabilities}: The size of vector units on modern commodity processors are increasing rapidly. While not all algorithms can easily be adapted to benefit from this capability, 
large gains are possible where algorithms can be vectorized. Essentially all HEP codes need modifications, or large scale refactoring, to effectively utilize SIMD capabilities.
\item
{\bf Evolution towards multi- or many-core architectures}: The current trend is to move away from ever faster processing cores towards more power efficient and more numerous processing cores. This 
change has already broken the traditional ``one-core-one-event'' model in many HEP frameworks and algorithms. As core counts increase, a larger number of algorithm developers, instead of just those 
developing the most resource intensive algorithms, will need to incorporate parallelism techniques into their algorithm implementations.   
\item
{\bf Slow increase in memory bandwidth}: Software trigger and event reconstruction applications in HEP are very memory intensive, and I/O access to memory in commodity hardware has not kept up with 
CPU capabilities. To evolve towards modern architectures, including the effective use of hierarchical memory structures, HEP algorithms will need to be refactored or rewritten to considerably reduce 
the required memory per processing core. 
\item
{\bf Rise of heterogeneous hardware}: Evolution in HEP algorithms has long taken advantage of a single dominant commodity computing platform (x86 running Linux).  More and more studies have shown 
or are investigating the throughput benefits of using low-power systems, GPUs, FPGA systems for critical pieces of processing. Software trigger and event processing algorithms are those that may 
benefit the most in HEP from the effective use of these technologies if they are able to adapt to the requirements of using these systems effectively.
\item
{\bf Possible evolution in facilities}: HEP facilities are likely to evolve both due to the increased architectural variability of affordable hardware and due to evolution in data science techniques. 
One example of the latter is an analysis center whose design is driven by data science techniques and technologies. These technologies (e.g., Hadoop, Spark) are under investigation in HEP for analysis 
and may change the way HEP data centers are resourced. A particular example of how this will impact trigger or reconstruction algorithms is that of physics object identification algorithms. These 
are frequently rerun by analysts in order to include the most recent version developed by the collaboration in their analysis.
\end{itemize}
Evolution in computing technology is generally a slow but continual process. However, architectures available today can provide necessary development platforms for trigger and event reconstruction 
developers to adapt codes to be better suited to future architectures.

\subsection{Challenges from Evolutions in Software technology}

The status and evolution of software development in HEP is the subject of the Software Development CWP working group. In this section, we briefly discuss some of the issues and opportunities of 
particular relevance to software trigger and event reconstruction work.

The move towards open source software development and continuous integration systems brings a number of important opportunities to assist developers of software trigger and event reconstruction 
algorithms. Continuous integration systems have already brought the ability to automate code quality and performance checks, both for algorithm developers and code integration teams. Scaling 
these up to allow for sufficiently high statistics checks is among the still outstanding challenges. While it is straightforward to test changes where no regression is expected, fully developed 
infrastructure for assisting developers in confirming the physics and technical performance of their algorithms during development is a work in progress.

As the timescale for experimental data taking and analysis increases, the issues of legacy code support increase. In particular, it seems unaffordable to imagine rewriting all of the software 
developed by the LHC experiments during the long shutdown preceding HL-LHC operations. Thus, as the HL-LHC run progresses, much of the code base for software trigger and event reconstruction 
algorithms will be 15-30 years in age. This implies an increased need for sustainable software development and investment in software education for experimental teams.

Code quality demands increase as traditional offline analysis components migrate into trigger systems, or more generically into algorithms that can only be run once. As described above, 
this may be due to either the prohibitive cost of rerunning algorithms over large data sets, or due to not having retained sufficient data to rerun algorithms (e.g., not retaining the full 
raw data). Algorithms in the software trigger and event reconstruction areas are very frequently contributed to by a large community of physicists rather than expert programmers. In many 
cases, the most sensitive algorithms may be checked and optimized by programming experts. This has so far satisfied the need of reducing the total computing resources needed by experiments, 
as typically only a few algorithmic components dominate the overall computing need. However, this approach would be currently impossible to carry out across the entire reconstruction code 
stack. As the complexity (and number) of real-time algorithms increases, the need for training as well as code validation and regression checking increases considerably.

These challenges are further complicated by a growing diversity of developers in the software trigger and event reconstruction areas. There is a generally growing gap between the basic 
programming techniques which students learn in their undergraduate courses and the state-of-the-art programming techniques required to fully exploit the power of emerging parallel hardware 
architectures. Software development methods and programming techniques evolve particularly quickly and often developers are self taught on current techniques. The experiment software environment 
must facilitate contributions from a range of developers while adjusting to challenges of a more and more complex online and offline software toolkit.

\section{Current Approaches}

In this section, we summarize current practices and resource requirements for software trigger and event reconstruction implementation and infrastructure needs. These include the scale of resource 
requirements for these activities, the type of data structures and data contents kept for custodial storage and analysis use, and calibration technique requirements across different experiments in 
high-energy physics. For each topic, we will briefly introduce the issues and then summarize experiment specific implementation details where available. This is meant to provide a representative 
range of approaches and to illustrate where the most challenging aspects are. Approaches from all experiments are not always included, as this information is summarized 
from working-group contributions.

\subsection{Computing Resource Requirements}

The table below summarizes the online and offline computing requirements of current and future experiments. Not all quantities are relevant for all experiments, and many items evolve depending 
on where in the lifecycle an experiment is. For example, disk requirements depend both on the number of events collected in a given running period but also on how long the experiment has been 
collecting data still of relevance for current analysis activities. More information about each online and offline computing system can be found in the 
references \cite{ALICE2015, Allton2017, ATLAS2015, CMS2015, DUNE2015, DUNE2017, LHCb2017, Miyamoto2015, Richter2016}.

The columns are defined as follows:
\begin{enumerate}
\item Online CPUs: Approximate size (or fair-share in case of a shared resource) of the real-time processing facility available for use by the experiment.
\item Offline CPUs: Approximate size (or fair-share in case of a shared resource) of the offline processing facility available for use by the experiment. For LHC experiments we have given the 
2017 pledged CPU amount including both Tier0 and distributed Tier1 and Tier2 resources. These resources are used for Monte Carlo simulation, analysis and event reconstruction processing. 
Thus, these numbers only provide an indication of the CPU requirements for event reconstruction.
\item Input rate to software trigger: Approximate rate of events seen by software trigger. Typically this is the output rate of the hardware trigger.
\item Event rate for analysis: Approximate rate of events saved by the experiment for custodial storage. Typically this is the rate of events saved by the software trigger system.
\item RAW data size: Approximate size of typical physics events saved for offline processing. While the analysis tier size for an experiment is only loosely connected to the raw data size, 
this measure does provide a measure of the total data volume for the offline event processing. As discussed above, there are also frequently reduced data formats saved by trigger systems 
instead of the full raw data. These are typically much smaller than the sizes provided here.
\item Analysis data size: Approximate size of typical physics events in the format used most frequently for analysis (e.g., xAOD in ATLAS, miniAOD in CMS).
\item Offline disk: The total disk available for production and analysis activities. This is expected to scale primarily with the total size of the analysis data.
\end{enumerate}

\afterpage{
\clearpage
\begin{landscape}
\begin{table}
\begin{tabular}{|l|p{20mm}|p{20mm}|p{20mm}|p{20mm}|p{20mm}|p{20mm}|p{20mm}|} \hline
Experiment &
Online CPUs (kHS06) /GPUs  &
Offline CPUs (kHS06) &
Input rate to software trigger (kHz) &
Event rate for analysis (kHz or GB/s) &
RAW data size (MB/evt) &
Analysis data size (MB/evt) &
Offline disk (PB) \\ \hline
CMS (2017) &
500&
1729&
100&
1&
1.5&
~0.03&
123 \\  \hline
CMS (HL-LHC) &
&
&
750&
5-7.5&
4.2-4.6&
&
\\ \hline
ATLAS (2017) &
28k \newline CPU cores&
2194&
100&
1&
1&
&
172\\ \hline
ATLAS (HL-LHC) &
&
&
1000&
10&
5&
&
\\ \hline
LHCb (2017)&
25k \newline CPU cores&
413&
1000&
0.7 Gb/s&
0.07&
&
35\\ \hline
LHCb (Run 3-4) &
&
&
30000&
$>2$ GB/s&
0.13&
&
\\ \hline
ALICE (2017) &
9k \newline CPU cores, \newline 180 GPUs &
805&
0.3\newline (central \newline Pb-Pb)&
&
100\newline (central \newline Pb-Pb) &
&
66 \\ \hline
ALICE (Run 3) &
&
&
50&
20 GB/s&
60&
&
60/yr \\ \hline
Belle II (2021) &
6400 \newline CPU cores&
600&
30 kHz&
10 kHz&
0.1&
0.01&
60\\ \hline
ProtoDUNE&
&
5&
&
0.6 GB/s&
60&
&
\\ \hline
DUNE (0 supp)&
&
&
&
11 MHz&
$1.5e^{-4}$&
&
54/yr \\ \hline
ILC (ILD,SiD)&
250&
400&
1 GB/s&
1 GB/s&
1&
0.01&
150 \\ \hline
\end{tabular}
\caption{The LHCb resource requirements for Run 3 and 4 are currently under review, and will be agreed following publication of a computing model document later in 2018. 
The figures given here are minimal requirements below which physics performance is already known to degrade in an unacceptable manner.}
\end{table}
\end{landscape}
\clearpage
}

This table illustrates that there is substantial commonality in the scale of computing needed for current generation of experiments. It is also clear that there is a large increase 
in the scale of data expected for the next generation of experiments. This is true both in terms of event rate and event size.

\subsection{Analysis Data tiers and data structures}
Here we summarize data structures either consumed by, or produced by, the software trigger and event reconstruction applications in some current HEP experiments. These are meant to be 
representative of current practices and not inclusive of all approaches used in HEP. Historically, these are the raw data format and analysis data format, respectively. Recent work has 
led to some evolution in what sort of data structures are used in some experiments. We summarize the approach taken and issues observed by different experiments.

\vskip 0.5cm
\noindent
{\bf LHCb}: Major advances were made in the LHCb trigger system in Run 2. Increasing the number of (logical) cores used in the trigger farm to 50000, deploying faster reconstruction algorithms, 
and data-caching (described below) now allow the multilayer software trigger system to execute the offline reconstruction algorithms. LHCb now splits its 0.7~GB/s of data written to permanent 
storage into two distinct streams. The full stream persists all low-level data as in Run 1. The turbo stream persists a user-defined selection of high-level information, typically a small 
subset of reconstructed objects in each selected event. This makes it possible to keep a much larger number of events, with the cost being that the lower-level information is lost. 
Several LHCb publications have now demonstrated that this reduced-file-size approach works in practice, with typical event size reduction being one order of magnitude with respect to the raw event.

\vskip 0.5cm
\noindent
{\bf CMS}: The data formats used by CMS for event data are built from custom written C++ classes organized into a ROOT TTree-based structure by the CMS framework. The RAW data is largely 
composed of a packed byte format organized and segmented according to the detector readout electronics layout. Analysis data are made up of more complex objects, used also in the 
reconstruction and analysis algorithms themselves. Particle flow constituent objects and high level physics objects (e.g. muon, electrons, jets) are made of classes derived from a 
common base class in order to make combination of objects in a straightforward and efficient manner.

Formats in Run~2 have evolved to include more compact data structures (e.g., the miniAOD formats) and for a few trigger streams the concept of ``scouting''. Instead of keeping raw detector data, 
the scouting data structures keep only a small set of high-level objects as derived in the CMS software trigger. This makes the data format very compact (about 1\% of a full raw data event) 
and allows for a very high rate of data to be kept. CMS has also used data structures with lossy-compression algorithms (e.g., saving variables using fewer than 32 bits) to reduce the size of 
its ``MIniAOD'' data tier, now widely used in analysis. This type of approach is widely applicable where the detector configuration or reconstruction algorithm itself limits resolution 
to be far worse than that implied by the usual 32-bit or 64-bit data types. While the custom structures developed by each experiment so far are unlikely to be completely generalizable, 
we would still benefit from a field-standard, or data-science standard library containing those parts which are in common.  CMS is also investigating how less complex data structures 
can allow for even small data formats that match the needs of the final analysis stages.

\vskip 0.5cm
\noindent
{\bf ATLAS}:In Run 2, a reduced analysis object stream has been added, containing only high-level trigger jet objects, to be 
used for the search of low-mass hadronic resonances in trigger-level analyses. This stream requires roughly 1\% of the full HLT bandwidth, allowing for data taking rates comparable to the 
full offline physics stream \cite{ATLASTwiki} 
%\cite{computing experts should add here about other data formats}.

\subsection{Most resource consuming algorithm}
{\bf ALICE}: Reconstruction and compression of data from the ALICE TPC is the dominant component of the ALICE event reconstruction and the driving feature behind the design of the O2 
facility \cite{ALICE2015} for event reconstruction starting in LHC Run 3. The reconstruction of particle trajectories from the ALICE TPC is done by cellular automation and Kalman filter algorithms. 
In the Run 2 high-level trigger, these algorithms run on GPUs as part of a pipelined system.

\vskip 0.5cm
\noindent
{\bf LHCb}: The specific forward geometry of LHCb, with a dipole magnet, and its use of RICH detectors for particle identification, have a significant impact on both the time cost of 
reconstruction algorithms and their utility in triggering and real-time analysis. The largest components are pattern recognition and Kalman fitting of tracks, which run on all events, and 
calorimeter and RICH reconstruction, which run on a subset of events. The overall CPU cost for these algorithms is roughly the same, considering the different fractions of events they run on. 
The full tracking cost greatly depends on the transverse momentum range of tracks being searched for. The RICH reconstruction is expensive, but because many decays of interest include kaons, 
its results reduce the time spent in later processing for combining tracks into displaced vertices.

\vskip 0.5cm
\noindent
{\bf CMS}: The computing resource needs of trigger and reconstruction algorithms is a strong function of event complexity (or pileup) in CMS. In Run 2 conditions, the track pattern recognition 
is the largest single CPU contributor in CMS reconstruction. Kalman fitting \cite{CMS2014}, particle flow algorithms \cite{CMS2017} and HCAL local reconstruction algorithms are also a considerable 
fraction of the total offline reconstruction CPU budget. The importance of these algorithms is also true for online trigger algorithm configurations where a simplified particle flow technique
 is used in the final event selections. As they are carried out in each event online, the calorimetric raw data unpackers and algorithms for pileup rejection at the detector level (e.g. to form 
reconstructed detector clusters) are also quite important online. Both online and offline, CMS has implemented a low track transverse momentum threshold (200-300 MeV) in order to achieve the best possible particle 
flow performance. This has a large impact on the CPU needs of CMS, not only in tracking, but elsewhere in the reconstruction where the algorithm complexity scales with the number of tracks 
reconstructed. For HL-LHC, CMS has the additional computational challenge of clustering and particle flow with its high-granularity calorimeter.

\subsection{Calibration techniques and requirements}
{\bf LHCb}: the full detector alignment and calibration of all sub-detectors must continue to be performed continuously in real-time, and the real-time monitoring framework must also be maintained 
in order to enable problems to be quickly identified and fixed. The calibrations are performed \cite{LHCbRTCalib2016} at a certain frequency:
\begin{itemize}
\item
A few times per year: Muon system alignment, RICH mirror alignment, Fine calorimeter calibration 
\item
Start of each LHC fill: Vertex detector alignment, Full tracker alignment, Straw-tube tracker gas calibration, Coarse calorimeter calibration
\item
Multiple times per fill: RICH refractive index calibration
\end{itemize}
In order to perform the calibrations in real-time, LHCb exploits the trigger farm to run the calibration tasks in parallel with trigger tasks. This enables e.g. the full tracker alignment to be 
completed in around 8 minutes. The calibration jobs are automated and return an updated set of constants which are compared to the result of the previous calibration; if the change goes beyond a 
certain threshold the constants are updated and an expert is alerted to validate the update. The jobs themselves are built on the same codebase as the rest of the LHCb reconstruction, and the 
monitoring uses the same setup as the general detector monitoring.

\vskip 0.5cm
\noindent
{\bf CMS}: The prompt calibration loop is an automated process, runs as part of the CMS Tier-0 facility using a small subset of the total data volume saved by the CMS trigger. Example calibrations 
include the assessment of bad detector channels for each run as well as the global tracker alignment. Results are typically available within 24 hours and thus are used in the full prompt reconstruction 
processing of the CMS data. Other calibrations, including the calorimeter light output calibrations, are performed outside of the Tier-0 infrastructure but in time for the prompt reconstruction 
processing on a regular interval. Finally, calibrations requiring higher statistics are performed either as the LHC data taking period progresses (e.g. performed once adequate statistics are obtained) 
or after the end of the run to achieve the ultimate detector performance required by analysis (e.g, tracker module level alignment and full energy scale calibration of the electromagnetic calorimeter).

\vskip 0.5cm
\noindent
{\bf ATLAS}: For real-time analyses with trigger jets, reconstruction and calibration are  kept as close as possible to the full reconstruction. 
Dedicated calibrations mitigating the effect of pile-up and restoring the hadronic energy scale are applied to these partially recorded events, making the properties of jets reconstructed at 
the HLT comparable to those of jets reconstructed from full events. These calibrations account for differences between fully recorded jets as well as for missing information from detectors 
other than the calorimeters. In the case of jets during the first part of Run 2, only calorimeter information is available to the HLT for jet reconstruction and calibration. It is foreseen 
that tracking information can be added already during Run 2 using the Fast TracKer (FTK). The jet reconstruction procedure for trigger jets is summarized in \cite{Abreu2014}. 
%\cite{reco experts should add overall strategy here}.

\vskip 0.5cm
\noindent
{\bf ALICE}: Real-time calibration and data quality monitoring are critical parts of the O2 system for Run 3 as the data volume processed is too large to be redone in a later, truely offline, processing.

\section{Research and Development Roadmap}

This section describes the proposed research and development roadmap defined in the working group. We identify seven broad areas to be critical for software trigger and event reconstruction 
work over the next decade. These  are:
\begin{itemize}
\item Enhanced vectorization programming techniques
\item Algorithms and data structures to efficiently exploit many-core architectures
\item Algorithms and data structures for non-x86 computing architectures (e.g., GPUs, FPGAs)
\item Enhanced quality assurance (QA) and quality control (QC) for reconstruction techniques
\item Real-time analysis
\item Precision physics-object reconstruction, identification and measurement techniques
\item Fast software trigger and reconstruction algorithms for high-density environments
\end{itemize}
For each area, we identify the overall goals of the research, as well as short, medium and long term milestones. These can be viewed as goals that should be achieved in advance of next-generation 
experiments. The short and medium term milestones are intended to be achievable on a timescale to inform software, computing and trigger technical design reports where possible.

\subsection{Roadmap area 1: Enhanced vectorization programming techniques}
\noindent {\bf Motivation}: HEP developed toolkits and algorithms typically make poor use of vector units on commodity computing systems. Improving this will bring speedups to applications running 
on both current computing systems and most future architectures.

\noindent {\bf Overall goal}: To evolve current toolkit and algorithm implementations, and best programming techniques to better use SIMD capabilities of current and future computing architectures.

\vskip 0.5cm
\noindent {\bf Short-term goals}: Identify best practices and documented examples of how HEP code was improved to increase vectorization performance via series of developer meetings. Adopt and apply 
industry tools for code performance analysis to identify tools of particular importance for investigation.

\vskip 0.5cm
\noindent {\bf Medium-term goals}: Initiate work to make measurable improvement to vectorization performance of HEP code stacks via software toolkit improvements or rewrites. Continue developer 
discussions while toolkits are improved to facilitate co-development where common approaches to improving different toolkits are possible.

\vskip 0.5cm
\noindent {\bf Long-term goals}: Demonstrate and facilitate widespread experiment adoption of new or improved toolkits.

\subsection{Roadmap area 2: Algorithms and data structures to efficiently exploit many-core architectures}
\noindent {\bf Motivation}: Computing platforms are generally evolving towards having more cores in order to increase processing capability. This evolution has resulted in multi-threaded frameworks 
in use, or in development, across HEP. Algorithm developers can further improve throughput by being thread safe and enabling the use of fine-grained parallelism.

\vskip 0.5cm
\noindent {\bf Overall goal}: To evolve current event models, toolkits and algorithm implementations, and best programming techniques to improve the throughput of multithreaded software trigger and 
event reconstruction applications.

\vskip 0.5cm
\noindent {\bf Short-term goals}: Identify the key lessons of the work done to make current toolkits thread-safe or at least ``thread friendly''. Understand and document what conceptual limitations
a thread-safe framework imposes on reconstruction and selection logic, and what kind of workflow scheduling it requires. Identify what classes of reconstruction and selection algorithms are able 
to efficiently work in a multi-threaded framework and which are not.

\vskip 0.5cm
\noindent {\bf Medium-term goals}: Document best practice for thread safe computing in a way which allows new collaboration members to efficiently develop in such a framework. Develop a toolkit 
to logically express proposed new algorithms in terms of data sources, sinks, and consumers, which can automatically analyze an algorithm and establish how thread-safe its logic is.

\vskip 0.5cm
\noindent {\bf Long-term goals}: Introduce training on multi-threaded and more generally parallel algorithm development in physics syllabi, building on the experience gained and the toolkits developed earlier.

\subsection{Roadmap area 3: Algorithms and data structures for non-x86 computing architectures (e.g., GPUs, FPGAs)}
\noindent {\bf Motivation}: Computing architectures using technologies beyond CPUs offer an interesting alternative for increasing throughput of the most time consuming trigger or reconstruction 
algorithms. Such architectures (e.g., GPUs, FPGAs) could be easily integrated into dedicated trigger or specialized reconstruction processing facilities (e.g., online computing farms).

\vskip 0.5cm
\noindent {\bf Overall goal}: To demonstrate how the throughput of toolkits or algorithms can be improved through the use of new computing architectures in a production environment.

\vskip 0.5cm
\noindent {\bf Short-term goals}: Develop reliable and portable benchmarking for mixed architectures which properly accounts for I/O overheads between them. Using this, identify, and develop 
prototypes for, event reconstruction and software trigger algorithms where specialized hardware is likely to bring significant improvement in metrics such as total event throughput per cost or throughput per Watt.

\vskip 0.5cm
\noindent {\bf Medium-term goals}: Demonstrate programming models and software toolkits appropriate for a heterogeneous computing environment. Considerations include facilitating high-level 
code-reuse between architectures, construction of appropriate data structures, and adoption of externally developed toolkits, such as mathematical libraries, that provide significant performance 
improvements on certain architectures. Deploy prototypes for limited scale operational tests where resources are available in a controlled fashion (e.g., experiment trigger computing facilities).

\vskip 0.5cm
\noindent {\bf Long-term goals}: Demonstrate ability of event reconstruction applications to reliably use and benefit from heterogenous computing facilities on a distributed computing system. 
Define cost-benefit metrics to use to guide facility providers for providing the largest event throughput in cases where HEP controls the mix of hardware to be purchased.

\subsection{Roadmap area 4: Enhanced QA/QC for reconstruction techniques}
\noindent {\bf Motivation}: HEP experiments have extensive continuous integration systems, including varying types of code regression checks. These are typically maintained by individual experiments 
and have not yet reached the scale where statistical regression checks, as well as technical and physics performance can be enabled for each proposed software change.

\vskip 0.5cm
\noindent {\bf Overall goal}: Enable the development, automation, and deployment of extended QA and QC tools and facilities for software trigger and event reconstruction algorithms.

\vskip 0.5cm
\noindent {\bf Short-term goals}: Discuss integration and testing systems currently used by experiments to formulate requirements for scope and scale needed for a common integration system 
capable of providing developers feedback on trigger and event reconstruction outputs according to physics and technical metrics.

\vskip 0.5cm
\noindent {\bf Medium-term goals}: Develop and demonstrate a scalable system for use by multiple experiments based on industry standard continuous integration tools. Define requirements for 
regression and validation techniques (for Monte Carlo and data studies) given evolution towards heterogeneous hardware and real-time calibrations.

\vskip 0.5cm
\noindent {\bf Long-term goals}: Develop and demonstrate next-generation tools needed for regression testing, software integration and validation and data quality related tasks.

\subsection{Roadmap area 5: Real-time analysis}

\noindent {\bf Motivation}: Real-time analysis techniques are being adopted to enable a wider range of physics signals to be saved by the trigger for final analysis. As rates increase, these 
techniques can become more important and widespread by enabling only techniques such as saving only reconstructed event information or only
the parts of an event associated with the signal candidates, reducing the required disk space substantially.

\vskip 0.5cm
\noindent {\bf Overall goal} : Evaluate and demonstrate the tools needed to facilitate real-time analysis techniques. Research topics include compression and custom data formats; toolkits for real-time 
detector calibration and validation which will enable full offline analysis chains to be ported into real-time; and frameworks which will enable non-expert offline analysts to design and deploy 
real-time analyses without compromising data taking quality.

\vskip 0.5cm
\noindent {\bf Short-term goals}: Discuss the ongoing real-time analysis frameworks being built by different HEP experiments and establish areas of commonality. Understand the extent to which 
cross-experiment toolkits can help with these areas.

\vskip 0.5cm
\noindent {\bf Medium-term goals}: Develop common toolkits for enabling real-time analysis across experiments, drawing on experience and collaboration with real-time applications in industry 
wherever possible. Develop a framework which enables non-experts to design and deploy real-time analyses without threatening the integrity of data taking.

\vskip 0.5cm
\noindent {\bf Long-term goals}: Begin to include real-time analysis requirements in the design experiments (in particular high luminosity hadron colliders such as FCC). This means explicitly 
optimizing detector hardware not for physics which can be done with events we can afford to store to disk, but optimizing for physics which can be done in real-time.

\subsection{Roadmap area 6: Precision physics-object reconstruction, identification and measurement techniques}

\noindent {\bf Motivation}: The central challenge for object reconstruction at HL-LHC is thus to maintain excellent efficiency and resolution in the face of high pileup values, especially at 
low object transverse momentum. Both trigger and reconstruction approaches need to exploit new techniques and higher granularity detectors to maintain or even improve physics measurements in the future.
Reconstruction in very high pileup environments, such as the HL-LHC or FCC-hh, may also greatly benefit from adding timing information to our detectors, in order to exploit the finite beam crossing time during which interactions are produced.

\vskip 0.5cm
\noindent {\bf Overall goal}: Develop and demonstrate tools needed to efficient techniques for physics object reconstruction and identification in complex environments.

\vskip 0.5cm
\noindent {\bf Short-term goals}: Identify areas where either new toolkits based on either novel techniques or new detector designs, are likely to achieve significant physics quality 
improvements especially in very dense (e.g., high pileup) environments at facilities including HL-LHC and FCC-hh. Known candidates are charged tracking techniques including precision timing 
detector information, jet imaging techniques, and particle-flow algorithms that exploit high-precision calorimetry.

\vskip 0.5cm
\noindent {\bf Medium-term goals}: Development and demonstration of algorithms and integrated software packages that are efficient and performant in complex environments on planned detector 
configurations for HL-LHC. Understand interplay between new information and traditional observables to determine how physics measurables should be best derived. Development of algorithms to 
optimize splitting events containing different sets of objects to obtain the best balance between data storage overhead and data processing overhead during physics analyses.

\vskip 0.5cm
\noindent {\bf Long-term goals}: Deploy algorithms in experimental software stacks as they mature.

\subsection{Roadmap area 7: Fast software trigger and reconstruction algorithms for high-density environments}

\noindent {\bf Motivation}: Future experimental facilities will bring a large increase in event complexity. The scaling of current-generation algorithms with this complexity must be improved 
to avoid a large increase in resource needs. In addition, it may be desirable or indeed necessary to deploy new algorithms, including advanced machine learning techniques developed in other fields, 
in order to solve these problems.

\vskip 0.5cm
\noindent {\bf Overall goal}: Evolve or rewrite existing toolkits and algorithms focused on their physics and technical performance at high event complexity (e.g. high pileup at HL-LHC). Most 
important targets are those which limit expected throughput performance at future facilities (e.g., charged-particle tracking). A number of such efforts are already in progress across the community.

\vskip 0.5cm
\noindent {\bf Short-term goals}: Identify additional areas where substantial gains in event reconstruction and software trigger algorithms may be obtained by either a large-scale reimplementation 
of existing algorithms or by the use of a new algorithmic approach (including machine learning concepts). Possible areas of investigation include improved memory locality for algorithms and data structures.

\vskip 0.5cm
\noindent {\bf Medium-term goals}: Develop and demonstrate new toolkits. Evaluate their effectiveness against current approaches using both physics driven and event throughput per computing cost metrics. 
\vskip 0.5cm
\noindent {\bf Long-term goals}: Deploy algorithms in experimental software stacks as they mature. It is particularly important to test new approaches using data driven studies to demonstrate 
robustness against changing detector and accelerator operating conditions.

\section{Conclusions}

The next decade will see the volume and complexity of data being processed by HEP experiments increase by at least one order of magnitude. While much of this increase is driven by the planned 
upgrades to the four major LHC detectors, new experiments such as DUNE will also make significant demands on the HEP data processing infrastructure. It is therefore essential that event reconstruction 
algorithms and software triggers continue to evolve so that they are able to efficiently exploit future computing architectures and deal with this increase in data rates without loss of physics capability.

We have identified seven key areas where research and development is necessary to enable the community to exploit the full power of the enormous datasets which we will be collecting. Three of these 
areas concern the increasingly parallel and heterogeneous computing architectures which we will have to write our code for. In addition to a general effort to vectorize our codebases, we must understand 
what kinds of algorithms are best suited to what kinds of hardware architectures, develop benchmarks that allow us to compare the physics-per-dollar-per-watt performance of different algorithms across a 
range of potential architectures, and find ways to optimally utilise heterogeneous processing centres. The consequent increase in the complexity and diversity of our codebase will necessitate both a 
determined push to educate tomorrow’s physicists in modern coding practices, and a development of more sophisticated and automated quality assurance and control for our codebases. The increasing granularity 
of our detectors, and the addition of timing information to help cope with the extreme pileup conditions at the HL-LHC, will require us to both develop new kinds of reconstruction algorithms 
and to make them fast enough for use in real-time. Finally, the increased signal rates will mandate a push towards real-time analysis in many areas of HEP, in particular those with low transverse momentum 
signatures.

The success of this research and development program will be intimately linked to challenges confronted in other areas of HEP computing, most notably the development of software frameworks which are able 
to support heterogeneous parallel architectures, including the associated data structures and I/O, the development of lightweight detector models that maintain physics precision with minimal timing and 
memory consequences for the reconstruction, enabling the use of offline analysis toolkits and methods within real-time analysis, and an awareness of advances machine learning reconstruction algorithms 
being developed outside HEP and the ability to apply them to our problems. For this reason perhaps the most important task ahead of us is to maintain the community which has coalesced together in this 
CWP process, so that the work done in these sometimes disparate areas of HEP fuses coherently together into a solution to the problems facing us over the next decade.

%\section*{Acknowledgements}
%Vladimir V. Gligorov acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 724777 ``RECEPT''. Caterina Doglioni acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 679305  ``DARKJETS''.  Eduardo Rodrigues acknowledges support by the United States National Science Foundation under award ACI-1450319.

### 2018-10-04 CWP/papers/HSF-CWP-2017-14_recotrigger/latex/all.tex
Tex sources for reco & trigger chapter (#102)
* Add files via upload

* Move LaTeX files into standard subdirectory

* Add gitignore for generated PDFs

* Remove unneeded a2tex script

\section{Introduction}

Realizing the physics programs of the planned and/or upgraded high-energy physics (HEP) experiments over the next 10 years will require the HEP community to address a number of challenges in the area of software and computing. For this reason, the HEP software community has engaged in a planning process over the past two years, with the objective of identifying and prioritizing the research and development required to enable the next generation of HEP detectors to fulfill their full physics potential. The aim is to produce a Community White Paper (CWP)~\cite{HSF2017} which will describe the community strategy and a roadmap for software and computing research and development in HEP for the 2020s. This activity is organized under the umbrella of the HEP Software Foundation (HSF).

The CWP process was carried out by working groups centered on specific topics. The topics of event reconstruction and software triggers are summarized together in this document and in more detail elsewhere~\cite{RECOCWP}. The reconstruction of raw detector data and simulated data and its processing in real time represent a major component of today's computing requirements in HEP. A recent projection [Campana2016] of the ATLAS 2016 computing model results in $>$85\% of the HL-LHC CPU resources being spent on the reconstruction of data or simulated events. We have evaluated the most important components of next generation algorithms, data structures, and code development and management paradigms needed to cope with highly complex environments expected in HEP detector operations in the next decade. New approaches to data processing were also considered, including the use of novel, or at least, novel to HEP, algorithms, and the movement of data analysis into real-time environments.

This document will discuss software algorithms essential to the interpretation of raw detector data into analysis-level objects. Specifically, these algorithms can be broadly grouped: 
\begin{enumerate}[itemsep=0ex]
\item
{\bf Online}: Algorithms, or sequences of algorithms, executed on events read out from the detector in near-real-time as part of the software trigger, typically on a computing facility located close to the detector itself.
\item
{\bf Offline}: As distinguished from online, any algorithm or sequence of algorithms executed on the subset of events preselected by the trigger system, or generated by a Monte Carlo simulation application, typically in a distributed computing system.
\item
{\bf Reconstruction}: The transformation of raw detector information into higher level objects used in physics analysis. A defining characteristic of “reconstruction” which separates it from “analysis” is that the quality criteria used in the reconstruction to, for example, minimize the number of fake tracks, should be general enough to be used in the full range of physics studies required by the experimental physics program. This usually implies that reconstruction algorithms use the entirety of the detector information to attempt to create a full picture of each interaction in the detector. Reconstruction algorithms are also typically run as part of the processing carried out by centralized computing facilities.
\item
{\bf Trigger}: the part of the online system responsible for classification of events which reduces either the number of events which are kept for further “offline” analysis, the size of such events, or both. In this working group we were only concerned with software triggers, whose defining characteristic is that they process data without a fixed latency. Software triggers are part of the real-time processing path and must make decisions quickly enough to keep up with the incoming data, possibly using substantial disk buffers.
\item
{\bf Real-time analysis}: Data processing that goes beyond object reconstruction, and is performed online within the trigger system. The typical goal of real-time analysis is to combine the products of the reconstruction algorithms (tracks, clusters, jets...) into complex objects (hadrons, gauge bosons, new physics candidates...) which can then be used directly in analysis without an intermediate reconstruction step. 
\end{enumerate}

\section{Challenges}

Software trigger and event reconstruction techniques in HEP face a number of new challenges in the next decade. These are broadly categorized into 1) those from new and upgraded accelerator facilities, 2) from detector upgrades and new detector technologies, 3) increases in anticipated event rates to be processed by algorithms (both online and offline), and 4) from evolutions in software development practices.

Advancements in facilities and future experiments bring a dramatic increase in physics reach, as well as increased event complexity and rates. At the HL-LHC, the central challenge for object reconstruction is thus to maintain excellent efficiency and resolution in the face of high pileup values, especially at low transverse momenta. Detector upgrades such as increases in channel density, high precision timing and improved detector geometric layouts are essential to overcome these problems. For software, particularly for triggering and event reconstruction algorithms, there is a critical need not to dramatically increase the processing time per event.

A number of new detector concepts are proposed on the 5-10 year timescale in order to help in overcoming the challenges identified above. In many cases, these new technologies bring novel requirements to software trigger and event reconstruction algorithms or require new algorithms to be developed. Ones of particular importance at the HL-LHC include high-granularity calorimetry, precision timing detectors, and hardware triggers based on tracking information which may seed later software trigger and reconstruction algorithms. Longer term projects with sufficiently mature software infrastructure can include cost implications of the simulation and reconstruction algorithms into the detector design considerations. This is especially important when the computing cost is expected to be a substantial part of the total construction and operation cost for an experiment.

Trigger systems for next-generation experiments are evolving to be more capable, both in their ability to select a wider range of events of interest for the physics program of their experiment, and their ability to stream a larger rate of events for further processing.  ATLAS and CMS both target systems where the output of the hardware trigger system is increased by 10x over the current capability, up to 1 MHz~\cite{ATLAS2015,CMS2015}. In other cases, such as LHCb~\cite{LHCb2014} and ALICE~\cite{ALICE2015}, the full collision rate (between 30 to 40 MHz for typical LHC operations) will be streamed to real-time or quasi-realtime software trigger systems. The increase in event complexity also brings a “problem” of overabundance of signal to the experiments, and specifically the software trigger algorithms. The evolution towards a genuine real-time analysis of data has been driven by the need to analyze more signal than can be written out for traditional processing, and technological developments which make it possible to do this without reducing the analysis sensitivity or introducing biases.

Evolutions in computing technologies are both opportunities to move beyond commodity x86 technologies, which HEP has used very effectively over the past 20 years, and significant challenges to derive sufficient event processing throughput per cost to reasonably enable our physics programs~\cite{Bird2014}.  Specific items identified included 1) the increase of SIMD capabilities (processors capable of running a single instruction set simultaneously over multiple data), 2) the evolution towards multi- or many-core architectures, 3) the slow increase in memory bandwidth relative to CPU capabilities, 4) the rise of heterogeneous hardware, and 5) the possible evolution in facilities available to HEP production systems.

The move towards open source software development and continuous integration systems brings opportunities to assist developers of software trigger and event reconstruction algorithms. Continuous integration systems have already allowed automated code quality and performance checks, both for algorithm developers and code integration teams. Scaling these up to allow for sufficiently high statistics checks is among the still outstanding challenges. As the timescale for experimental data taking and analysis increases, the issues of legacy code support increase. Code quality demands increase as traditional offline analysis components migrate into trigger systems, or more generically into algorithms that can only be run once.

\section{Current approaches}

Substantial computing facilities are in use for both online and offline event processing across all experiments surveyed. Online facilities are dedicated to the operation of the software trigger, while offline facilities are shared for operational needs including event reconstruction, simulation (often the dominant component) and analysis. CPU in use by experiments is typically at the scale of tens or hundreds of thousands of x86 processing cores. Projections to future needs, such as for the HL-LHC, show the need for a substantial increase in scale of facilities without significant changes in approach or algorithms.

The CPU needed for event reconstruction tends to be dominated by charged particle reconstruction (tracking), especially when the need for efficiently reconstructing low transverse momentum particles is considered. Calorimetric reconstruction, particle flow reconstruction, particle identification algorithms also make up significant parts of the CPU budget in some experiments. The CPU required for event reconstruction and trigger area with challenges and substantial potential risk to the computing cost of experiments. In this respect, software for future experiments will continue to evolve, to improve both the physics and technical performance characteristics of algorithms, and the uncertainty and evolution due to detector performance and operating conditions will continue throughout the experimental program.

Disk storage is typically 10s to 100s of PB per experiment. It is dominantly used to make the output of the event reconstruction, both for real data and simulation, available for analysis. Current generation experiments have moved towards smaller, but still flexible, data tiers for analysis. These tiers are typically based on the ROOT~\cite{Brun1996} file format and constructed to facilitate both skimming of interesting events and the selection of interesting pieces of events by individual analysis groups or through centralized analysis processing systems. Initial implementations of real-time analysis systems are in use within several experiments. These approaches remove the detector data that typically makes up the raw data tier kept for offline reconstruction, and keep only final analysis objects~\cite{Aaij2016,ATLAS2017,CMS2016}.

Detector calibration and alignment requirements were surveyed. Generally a high level of automation is in place across experiments, both for very frequently updated measurements and more rarely updated measurements. Often automated procedures are integrated as part of the data taking and data reconstruction processing chain. Some longer term measurements, requiring significant data samples to be analyzed together remain as critical pieces of calibration and alignment work. These techniques are often most critical for a subset of precision measurements rather than for the entire physics program of an experiment.

\section{Research and development Roadmap and Goals}

We identify seven broad areas to be critical for software trigger and event reconstruction work over the next decade. These are:
\begin{enumerate}[itemsep=-1ex]
\item
Enhanced vectorization programming techniques
\item
Algorithms and data structures to efficiently exploit many-core architectures
\item
Algorithms and data structures for non-x86 architectures (e.g., GPUs, FGAs)
\item
Enhanced quality assurance (QA) and quality control (QC) for reconstruction techniques
\item
Real-time analysis
\item
Precision physics-object reconstruction, identification and measurement techniques
\item
Fast software trigger and reconstruction algorithms for high-density environments
\end{enumerate}
Not all roadmap areas are directly applicable to the event reconstruction and triggering approach taken by all experiments. However, we expect that each area of proposed research and development will be broadly applicable to future high-energy physics experimental programs.

\subsection*{Roadmap area 1: Enhanced vectorization programming techniques}

HEP developed toolkits and algorithms typically make poor use of vector units on commodity computing systems. Improving this will bring speedups to applications running on both current computing systems and most future architectures. The goal for work in this area is to evolve current toolkit and algorithm implementations, and best programming techniques to better use SIMD capabilities of current and future computing architectures.

\subsection*{Roadmap area 2: Algorithms and data structures to efficiently exploit many-core architectures}

Computing platforms are generally evolving towards having more cores in order to increase processing capability. This evolution has resulted in multi-threaded frameworks in use, or in development, across HEP. Algorithm developers can improve throughput by being thread safe and enabling the use of fine-grained parallelism. The goal is to evolve current event models, toolkits and algorithm implementations, and best programming techniques to improve the throughput of multithreaded software trigger and event reconstruction applications.

\subsection*{Roadmap area 3: Algorithms and data structures for non-x86 computing architectures (e.g., GPUs, FPGAs)}

Computing architectures using technologies beyond CPUs offer an interesting alternative for increasing throughput of the most time consuming trigger or reconstruction algorithms. Such architectures (e.g., GPUs, FPGAs) could be easily integrated into dedicated trigger or specialized reconstruction processing facilities (e.g., online computing farms). The goal is to demonstrate how the throughput of toolkits or algorithms can be improved through the use of new computing architectures in a production environment.

\subsection*{Roadmap area 4: Enhanced QA/QC for reconstruction techniques}

HEP experiments have extensive continuous integration systems, including varying code regression checks that have enhanced the quality assurance and quality control procedures for software development in recent years. These are typically maintained by individual experiments and have not yet reached the scale where statistical regression, technical, and physics performance checks can be performed for each proposed software change. The goal is to enable the development, automation, and deployment of extended QA and QC tools and facilities for software trigger and event reconstruction algorithms.

\subsection*{Roadmap area 5: Real-time analysis }

Real-time analysis techniques are being adopted to enable a wider range of physics signals to be saved by the trigger for final analysis. As rates increase, these techniques can become more important and widespread by enabling only the parts of an event associated with the signal candidates to be saved, reducing the required disk space. The goal is to evaluate and demonstrate the tools needed to facilitate real-time analysis techniques. Research topics include compression and custom data formats; toolkits for real-time detector calibration and validation which will enable full offline analysis chains to be ported into real-time; and frameworks which will enable non-expert offline analysts to design and deploy real-time analyses without compromising data taking quality.

\subsection*{Roadmap area 6: Precision physics-object reconstruction, identification and measurement techniques}

The central challenge for object reconstruction at HL-LHC is to maintain excellent efficiency and resolution in the face of high pileup values, especially at low transverse momenta. Both trigger and reconstruction approaches need to exploit new techniques and higher granularity detectors to maintain or even improve physics measurements in the future. Reconstruction in very high pileup environments, such as the HL-LHC or FCC-hh, may also greatly benefit from adding timing information to our detectors, in order to exploit the finite beam crossing time during which interactions are produced. The goal is to develop and demonstrate efficient techniques for physics object reconstruction and identification in complex environments.

\subsection*{Roadmap area 7: Fast software trigger and reconstruction algorithms for high-density environments }

Future experimental facilities will bring a large increase in event complexity. The scaling of current-generation algorithms with this complexity must be improved to avoid a large increase in resource needs. In addition, it may be desirable or indeed necessary to deploy new algorithms, including advanced machine learning techniques developed in other fields, in order to solve these problems. The goal is to evolve or rewrite existing toolkits and algorithms focused on their physics and technical performance at high event complexity (e.g. high pileup at HL-LHC). Most important targets are those which limit expected throughput performance at future facilities (e.g., charged-particle tracking). A number of such efforts are already in progress across the community.

\section{Conclusions}

The next decade will see the volume and complexity of data being processed by HEP experiments increase by at least one order of magnitude. While much of this increase is driven by planned upgrades to the four major LHC detectors, new experiments such as DUNE will also make significant demands on the HEP data processing infrastructure. It is essential that software triggers and event reconstruction algorithms continue to evolve so that they are able to efficiently exploit future computing architectures and deal with this increase in data rates without loss of physics capability.

We have identified seven key areas where R\&D is necessary to enable the community to exploit the full power of the enormous datasets which we will be collecting. Three of these areas concern the increasingly parallel and heterogeneous computing architectures which we will have to write our code for. In addition to a general effort to vectorize our codebases, we must understand what kinds of algorithms are best suited to what kinds of hardware architectures, develop benchmarks that allow us to compare the physics-per-dollar-per-watt performance of different algorithms across a range of potential architectures, and find ways to optimally utilise heterogeneous processing centres. The consequent increase in the complexity and diversity of our codebase will necessitate both a determined push to educate tomorrow’s physicists in modern coding practices, and a development of more sophisticated and automated quality assurance and control for our codebases. The increasing granularity of our detectors, and the addition of timing information to help cope with the extreme pileup conditions at the HL-LHC, will require us to both develop new kinds of reconstruction algorithms and to make them fast enough for use in real-time. Finally, the increased signal rates will mandate a push towards real-time analysis in many areas of HEP, in particular those with low transverse momentum signatures.

The success of this R\&D program will be intimately linked to challenges confronted in other areas of HEP computing, most notably the development of software frameworks which are able to support heterogeneous parallel architectures, including the associated data structures and I/O, the development of lightweight detector models which maintain physics precision with minimal timing and memory consequences for the reconstruction, enabling the use of offline analysis toolkits and methods within real-time analysis, and the ability to integrarte machine learning reconstruction algorithms being developed outside HEP into our workflows and apply them to our problems. For this reason perhaps the most important task ahead of us is to maintain the community which has coalesced together in this CWP process, so that the work done in these sometimes disparate areas of HEP fuses coherently together into a solution to the problems facing us over the next decade.

%\section*{Acknowledgements}
%Vladimir V. Gligorov acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 724777 ``RECEPT''. Caterina Doglioni acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 679305  ``DARKJETS''.  Eduardo Rodrigues acknowledges support by the United States National Science Foundation under award ACI-1450319.

### 2018-10-04 CWP/papers/HSF-CWP-2017-14_recotrigger/latex/authors.tex
Tex sources for reco & trigger chapter (#102)
* Add files via upload

* Move LaTeX files into standard subdirectory

* Add gitignore for generated PDFs

* Remove unneeded a2tex script

Albrecht, Johannes$^{1}$;
Bloom, Kenneth$^{2}$;
Boccali, Tommaso$^{3}$;
Boveia, Antonio$^{4,c}$;
De Cian, Michel$^{5}$;
Doglioni, Caterina$^{6,b}$;
Dziurda, Agnieszka$^{7}$;
Farbin, Amir$^{8}$;
Fitzpatrick, Conor$^{9}$;
Gaede, Frank$^{10}$;
George, Simon$^{11}$;
Gligorov, Vladimir$^{12,a}$;
Grasland, Hadrien$^{13}$;
Grillo, Lucia$^{14}$;
Hegner, Benedikt$^{7}$;
Kalderon, William$^{6}$;
Kama, Sami$^{15}$;
Koppenburg, Patrick$^{16}$;
Krutelyov, Slava$^{17}$;
Kutschke, Rob$^{18}$;
Lampl, Walter$^{19}$;
Lange, David$^{20}$;
Moyse, Ed$^{21}$;
Norman, Andrew$^{18}$;
Petric, Marko$^{7}$;
Polci, Francesco$^{12}$;
Potamianos, Karolos$^{10}$;
Ratnikov, Fedor$^{22}$;
Raven, Gerhard$^{16}$;
Ritter, Martin$^{23}$;
Rizzi, Andrea$^{3}$;
Rodrigues, Eduardo$^{24}$;
Rousseau, David$^{13}$;
Salzburger, Andy$^{7}$;
Sexton Kennedy, Liz$^{18}$;
Sokoloff, Michael D$^{24}$;
Stewart, Graeme$^{7}$;
Ustyuzhanin, Andrey$^{22}$;
Viren, Brett$^{25}$;
Williams, Mike$^{26}$;
Winklmeier, Frank$^{27}$;
Wuerthwein, Frank$^{17}$
\bigskip 
\par {\footnotesize $^{1}$ Technische Universit\"at Dortmund, Dortmund, Germany}
\par {\footnotesize $^{2}$ University of Nebraska-Lincoln, Lincoln, NE, USA}
\par {\footnotesize $^{3}$ INFN Sezione di Pisa, Università di Pisa, Scuola Normale Superiore di Pisa, Pisa, Italy}
\par {\footnotesize $^{4}$ The Ohio State University, Columbus, OH, USA}
\par {\footnotesize $^{5}$ Physikalisches Institut, Ruprecht-Karls-Universitat Heidelberg, Heidelberg, Germany}
\par {\footnotesize $^{6}$ Fysiska institutionen, Lunds Universitet, Lund, Sweden}
\par {\footnotesize $^{7}$ CERN, Geneva, Switzerland}
\par {\footnotesize $^{8}$ Department of Physics, The University of Texas at Arlington, Arlington, TX, USA}
\par {\footnotesize $^{9}$ Institute of Physics, École Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland}
\par {\footnotesize $^{10}$ Deutsches Elektronen-Synchrotron, Hamburg, Germany}
\par {\footnotesize $^{11}$ Department of Physics, Royal Holloway University of London, Surrey, United Kingdom}
\par {\footnotesize $^{12}$ LPNHE, Université Pierre et Marie Curie, Université Paris Diderot, CNRS/IN2P3, Paris, France}
\par {\footnotesize $^{13}$ LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\par {\footnotesize $^{14}$ School of Physics and Astronomy, University of Manchester, Manchester, United Kingdom}
\par {\footnotesize $^{15}$ Physics Department, Southern Methodist University, Dallas TX, United States of America}
\par {\footnotesize $^{16}$ Nikhef National Institute for Subatomic Physics and Vrije Universiteit Amsterdam, Amsterdam, The Netherlands}
\par {\footnotesize $^{17}$ University of California, San Diego, La Jolla, CA, USA}
\par {\footnotesize $^{18}$ Fermi National Accelerator Laboratory, Batavia, IL, USA}
\par {\footnotesize $^{19}$ Department of Physics, University of Arizona, Tucson, AZ, USA}
\par {\footnotesize $^{20}$ Princeton University, Princeton, NJ, USA}
\par {\footnotesize $^{21}$ Department of Physics, University of Massachusetts, Amherst, MA, USA}
\par {\footnotesize $^{22}$ National Research University Higher School of Economics, Moscow, Russia;  Yandex School of Data Analysis, Moscow, Russia}
\par {\footnotesize $^{23}$ Fakult\"at f\"ur Physik, Ludwig-Maximilians-Universit\"at M\"unchen, M\"unchen, Germany}
\par {\footnotesize $^{24}$ University of Cincinnati, Cincinnati, OH, USA}
\par {\footnotesize $^{25}$ Physics Department, Brookhaven National Laboratory, Upton, NY, USA}
\par {\footnotesize $^{26}$ Laboratory for Nuclear Science, Massachusetts Institute of Technology, Cambridge, MA, USA}
\par {\footnotesize $^{27}$ Center for High Energy Physics, University of Oregon, Eugene, OR, USA}
\bigskip
\par {\footnotesize $^{a}$ Vladimir V. Gligorov acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 724777 “RECEPT”}
\par {\footnotesize $^{b}$ Caterina Doglioni acknowledges funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme under grant agreement No 679305 “DARKJETS”}
\par {\footnotesize $^{c}$ Antonio Bovela acknowledges funding for US Department of Energy (Grant DE-SC0011726).}

### 2018-10-04 CWP/papers/HSF-CWP-2017-14_recotrigger/latex/recoCWP.tex
Tex sources for reco & trigger chapter (#102)
* Add files via upload

* Move LaTeX files into standard subdirectory

* Add gitignore for generated PDFs

* Remove unneeded a2tex script

--- 2018-10-03 CWP/papers/HSF-CWP-2017-14_recotrigger/recoCWPFull.tex
@@
 \usepackage{hyperref}
-\usepackage{afterpage}
-\usepackage{pdflscape}
 \hypersetup{
@@
 %\setitemize{wide}
-%\setlist\bibitem{itemize}{leftmargin=*}
-%\setlist\bibitem{enumerate}{leftmargin=*}
+%\setlist[itemize]{leftmargin=*}
+%\setlist[enumerate]{leftmargin=*}
 
@@
 
-\input{titlepageFull.tex}
+\input{titlepage.tex}
 \newpage
@@
 
-\input{allFull.tex}
+\input{all.tex}
 
@@
 
-\begin{thebibliography}{99}
+\begin{thebibliography}{9}
 
@@
 
-%\bibitem{RECOCWP}
-%Software trigger and event reconstruction CWP group, To be published (\href{https://docs.google.com/document/d/1QRO8RA488fwfSg5CSjmvm16-pZpGApSA0l666g_mS_0}{Link}).
+\bibitem{RECOCWP}
+Software trigger and event reconstruction CWP group, To be published as HSF-CWP-2017-16 (\href{https://github.com/HEP-SF/documents/tree/master/CWP/papers}{Link}).
 
@@
 Oct 3 2016 (\href{https://indico.cern.ch/event/524795/contributions/2236590/attachments/1347419/2032314/ECFA2016.pdf}{Link}).
-
-\bibitem{Bordry2016} 
-F. Bordry, 
-\textit{presentation at the LHC Performance Workshop (Chamonix 2016)}, 
-January 25th - 28th, 2016. (\href{http://chamonix-2016.web.cern.ch/}{Link} and \href{http://lhc-commissioning.web.cern.ch/lhc-commissioning/schedule/LHC-schedule-update.pdf}{Link})
-
-\bibitem{LHCb2012} 
-LHCb Collaboration, 
-\textit{LHCb upgrade framework TDR},  
-CERN-LHCC-2012-007 (2012) (\href{https://cds.cern.ch/record/1443882}{Link}).
-
-\bibitem{ALICE2013} 
-A. Kluge et al., 
-\textit{Upgrade of the ALICE Readout and Trigger System}, 
-CERN-LHCC-2013-019 (\href{https://cds.cern.ch/record/1603472}{Link}).
-
-\bibitem{Zimmerman2009} 
-F. Zimmerman, PoS EPS-HEP2009, 140 (2009). 
-G. Apollinari et al., 
-\textit{High-Luminosity Large Hadron Collider (HL-LHC): Preliminary Design Report}, 
-CERN-2015-005.%, doi:10.5170/CERN-2015-005.
 
@@
 
-\bibitem{LHCb2017} 
+\bibitem{LHCb2014}
 LHCb Collaboration, 
-\textit{Expression of Interest for a Phase-II LHCb Upgrade: Opportunities in flavour physics, and beyond, in the HL-LHC era},  
-CERN-LHCC-2017-003 (2017) (\href{https://cds.cern.ch/record/2244311}{Link}).
+\textit{LHCb Trigger and Online Upgrade Technical Design Report},
+CERN-LHCC-2014-016 (2014) (\href{https://cds.cern.ch/record/1701361}{Link}).
 
-\bibitem{Ohnishi2013} 
-Y. Ohnishi, et. al., 
-\textit{Accelerator design at SuperKEKB}, 
-Prog. Theor. Exp. Phys. 2013 03A011 (2013) (\href{http://dx.doi.org/10.1093/ptep/pts083}{Link}).
+\bibitem{ALICE2015}
+ALICE Collaboration, 
+\textit{Technical Design Report for the Upgrade of the Online-Offline Computing System}, 
+(\href{https://cds.cern.ch/record/2011297}{Link}).
 
-\bibitem{BelleII2010} 
-Belle II Collaboration, 
-\textit{Belle II Technical Design Report}, 
-KEK Report 2010-1 (2010) (\href{https://arxiv.org/abs/1011.0352}{Link}).
+\bibitem{Bird2014}
+I. Bird, et. al., 
+\textit{Update of the Computing Models of the WLCG and the LHC Experiments},
+CERN-LHCC-2014-014, (2014), (\href{http://cds.cern.ch/record/1695401/files/LCG-TDR-002.pdf}{Link}).
 
-\bibitem{DUNE2016} 
-DUNE Collaboration, 
-\textit{Long-Baseline Neutrino Facility (LBNF) and Deep Underground Neutrino Experiment (DUNE) Conceptual Design Report Volume 1: The LBNF and DUNE Projects}, 
-arXiv:1601.05471 (2016) (\href{https://arxiv.org/abs/1601.05471}{Link}).
+\bibitem{Brun1996}
+R. Brun and F. Rademakers,
+\textit{ROOT - An Object Oriented Data Analysis Framework}, 
+Proceedings AIHENP'96 Workshop, Lausanne, Sep. 1996, Nucl. Inst. \& Meth. in Phys. Res. A 389 (1997) 81-86. See also \href{http://root.cern.ch}{ROOT homepage}.
 
-\bibitem{Ball2014} 
-A. Ball et al., 
-\textit{Future Circular Collider Study Hadron Collider Parameters}, 
-FCC-ACC-SPC-0001 1342402, CERN (2014) (\href{https://indico.cern.ch/event/298180/contributions/1658149/attachments/560575/772288/FCC-1401101315-DSC_HadronColliderParameters_V0.3.pdf}{Link}).
+\bibitem{Aaij2016}
+R. Aaij, et. al., 
+\textit{Tesla : an application for real-time data analysis in High Energy Physics}, 
+Comput. Phys. Commun. 208 (2016) 35-42 (\href{https://cds.cern.ch/record/2147693}{Link}).
 
-\bibitem{Ilten2016} 
-P. Ilten, Y. Soreq, J. Thaler, M. Williams, and W. Xue, 
-\textit{Proposed inclusive dark photon search at LHCb}, 
-Phys. Rev. Lett. 116, 251803 (2016).
-
-\bibitem{aTLAs2017}
-%ATLAS Collaboration, \textit{Performance of the ATLAS Trigger System in 2015} ,
-%Eur. Phys. J.C77 5 317 (2017),
-%\href{https://inspirehep.net/record/1500696}{Link}. 
-ATLAS Collaboration,
+\bibitem{ATLAS2017}
+ATLAS Collaboration, \textit{Performance of the ATLAS Trigger System in 2015} ,
+Eur. Phys. J.C77 5 317 (2017),
+\href{https://inspirehep.net/record/1500696}. ATLAS Collaboration,
 \textit{Trigger-object Level Analysis with the ATLAS detector at the Large Hadron Collider: summary and perspectives}, 
@@
 
-\bibitem{LHCb2014}
-LHCb Collaboration, 
-\textit{LHCb Trigger and Online Upgrade Technical Design Report},
-CERN-LHCC-2014-016 (2014) (\href{https://cds.cern.ch/record/1701361}{Link}).
-
-\bibitem{Marshall2015} The Pandora Software Development Kit for Pattern Recognition
-J.S. Marshall, M.A. Thomson (Cambridge U.). Jun 16, 2015. 15 pp.
-Published in Eur.Phys.J. C75 (2015) no.9, 439
-
-\bibitem{Gray2016} 
-L. Gray, 
-\textit{Challenges of Particle Flow reconstruction in the CMS High-Granularity Calorimeter at the High-Luminosity LHC}, 
-38th International Conference on High Energy Physics (ICHEP 2016), (\href{https://indico.cern.ch/event/432527/contributions/1071751/}{Link}).
-
-\bibitem{Gray2017} 
-L. Gray, 
-\textit{4 Dimensional Trackers},
-(2017) (\href{https://indico.cern.ch/event/577003/contributions/2476434}{Link}).
-
-\bibitem{ATLAS2015} 
-ATLAS Collaboration, 
-\textit{ATLAS Phase-II Upgrade Scoping Document}, 
-CERN-LHCC-2015-020, LHCC-G-166 (2015) (\href{https://cds.cern.ch/record/2055248}{Link})
-
-\bibitem{ALICE2015}
-ALICE Collaboration, 
-\textit{Technical Design Report for the Upgrade of the Online-Offline Computing System}, 
-(\href{https://cds.cern.ch/record/2011297}{Link}).
-
-\bibitem{Abreu2014} 
-R. Abreu, 
-\textit{The upgrade of the ATLAS High Level Trigger and Data Acquisition systems and their integration}, 
-Proceedings of 19th IEEE-NPSS Real-Time conference 2014, pp.7097414 (2014) (\href{https://cds.cern.ch/record/1702466/}{Link}).
-
-\bibitem{Petrucciani2015} 
-G. Petrucciani et. al., 
-\textit{Mini-AOD: A New Analysis Data Format for CMS}, Journal of Physics: Conference Series 664 7 072052 (2015)
-
-\bibitem{Eifert2015} 
-T. Eifert, et. al., 
-\textit{Implementation of the ATLAS Run 2 event data model}, 
-Journal of Physics: Conference Series, Volume 664, Offline software (2015) (\href{http://iopscience.iop.org/article/10.1088/1742-6596/664/7/072045}{Link}).
-
-\bibitem{Aaij2016}
-R. Aaij, et. al., 
-\textit{Tesla : an application for real-time data analysis in High Energy Physics}, 
-Comput. Phys. Commun. 208 (2016) 35-42 (\href{https://cds.cern.ch/record/2147693}{Link}).
-
-\bibitem{Bird2014}
-I. Bird, et. al., 
-\textit{Update of the Computing Models of the WLCG and the LHC Experiments},
-CERN-LHCC-2014-014, (2014), (\href{http://cds.cern.ch/record/1695401/files/LCG-TDR-002.pdf}{Link}).
-
-\bibitem{Allton2017} 
-C. Allton, et. al.,  
-\textit{Computing Resources Scrutiny Group}, 
-CERN–RRB–2017–056 (2017) (\href{https://indico.cern.ch/event/617821/contributions/2493341/attachments/1446478/2233022/CERN-RRB-2017-056_V2.pdf}{Link}).
-
-\bibitem{DUNE2015}, 
-Dune Collaboration, 
-\textit{LBNF/DUNE Conceptual Design Report, Annex 4B Expected Data Rates for the DUNE Detectors}, 
-(2015) (\href{http://lbne2-docdb.fnal.gov/cgi-bin/ShowDocument?docid=10720}{Link}).
-
-\bibitem{DUNE2017}, 
-Dune Collaboration, 
-\textit{The Single-Phase ProtoDUNE Technical Design Report},
-(2017) (\href{https://arxiv.org/abs/1706.07081}{Link}).
-
-\bibitem{Miyamoto2015} 
-A.Miyamoto et al., 
-\textit{Computing Requirements of the ILC Experiments}, 
-LCC-Note (2015), (\href{http://www-jlc.kek.jp/~miyamoto/SoftwareCommonTask/docs/ILCComputing-EDMS1130485.A.1.1.pdf}{Link})
-
-\bibitem{Richter2016} 
-M. Richter, 
-\textit{Online Data Compression in the ALICE O2 Facility}, 
-Presented at the Computing in High Energy and Nuclear Physics conference (2016),
-\href{https://indico.cern.ch/event/505613/contributions/2227264/}{Link}.
-
-\bibitem{ATLASTwiki} 
-\href{https://twiki.cern.ch/twiki/pub/AtlasPublic/TriggerOperationPublicResults/BandwidthPie2016.png}{BandwidthPie2016.png}, 
-\href{https://twiki.cern.ch/twiki/pub/AtlasPublic/TriggerOperationPublicResults/Time_HLTRate_Stack_2016_07.png}{Time\_HLTRate\_Stack\_2016\_07.png}
-
-\bibitem{CMS2014} 
-CMS Collaboration, 
-\textit{Description and performance of track and primary-vertex reconstruction with the CMS tracker},  
-JINST 9 (2014) 10009 (\href{http://cds.cern.ch/record/1704291}{Link}).
-
-\bibitem{CMS2017} 
-CMS Collaboration, 
-\textit{Particle-flow reconstruction and global event description with the CMS detector}, 
-Submitted to JINST (2017) (\href{https://arxiv.org/abs/1706.04965}{Link}).
-
-\bibitem{LHCbRTCalib2016} 
-G. Dujani and B. Storaci, 
-\textit{Real-time alignment and calibration of the LHCb Detector in Run II}, 
-Journal of Physics: Conference Series, Volume 664, Online computing (\href{https://cds.cern.ch/record/2017839?ln=en}{Link}).
-
-%\bibitem{Brun1996}
-%R. Brun and F. Rademakers,
-%\textit{ROOT - An Object Oriented Data Analysis Framework}, 
-%Proceedings AIHENP'96 Workshop, Lausanne, Sep. 1996, Nucl. Inst. \& Meth. in Phys. Res. A 389 (1997) 81-86. See also \href{http://root.cern.ch}{ROOT homepage}.
-
 \end{thebibliography}
 
-\input{appendix.tex}
-
 \end{document}

### 2018-10-11 CWP/papers/HSF-CWP-2017-08_framework/latex/df_fwk_summary.tex
Add frameworks paper abstract
Correct some small mistakes in the text as well

@@
 
-\abstract{Abstract goes here.}
+\abstract{Data processing frameworks are an essential part of HEP
+experiments' software stacks. Frameworks provide a means by which code
+developers can undertake the essential tasks of physics data processing,
+accessing relevant inputs and storing their outputs, in a coherent way without
+needing to know the details of other domains. Frameworks provide essential
+core services for developers and help deliver a configurable working
+application to the experiments' production systems.
+Modern HEP processing frameworks are in the process of adapting to a new
+computing landscape dominated by parallel processing and heterogeneity,
+which pose many questions regarding enhanced functionality and scaling
+that must be faced without compromising the maintainability of the code.
+In this paper we identify a program of work that can help further clarify
+the key concepts of frameworks for HEP and then spawn R\&D activities that
+can focus the community's efforts in the most efficient manner to address
+the challenges of the upcoming experimental program.
+}
 
@@
 
-\title{HEP Software Foundation Community White Paper -- Data Processing Framework WG Report}
+\title{HEP Software Foundation Community White Paper Working Group -- Data Processing Frameworks}
 
@@
 \author[c]{Chris Jones}
+\author[b]{Michel Jouvin}
 \author[c,1]{Jim Kowalkowski}
 \author[c,1]{Elizabeth Sexton-Kennedy}
-\author[b]{Michel Jouvin}
-\author[d]{Several Others}
+\author[a]{Graeme A Stewart}
+\author[d]{Several Others - Charles, Marco, ?}
 
@@
 \affiliation[c]{Fermi National Accelerator Laboratory, Batavia, Illinois, USA}
-\affiliation[d]{Lawence Berkeley National Laboratory, Berkeley, CA, USA}
+\affiliation[d]{Lawrence Berkeley National Laboratory, Berkeley, CA, USA}
 \affiliation[e]{Other Places}
@@
 The time periods of interest for this document are DUNE and HL-LHC,
-which will deliver on the order of 50 PB data events per year per
+which will deliver on the order of 50 PB of data events per year per
 experiment. The results of the proposed R\&D ought to be used for
@@
     handle the massive parallelism that will be present throughout all
-    layers in the available compute facilities. This is necessary
+    layers in the available computing facilities. This is necessary
     because of the ever-increasing availability of specialized compute
-    resources, and includes GPGPUs, Tensor Processing Units (TPUs),
-    and tiered memory systems integrated with storage, and ultra
+    resources, including GPGPUs, Tensor Processing Units (TPUs),
+    tiered memory systems integrated with storage, and ultra
     high-speed network interconnects.
@@
 basic parallelisation, both within and across events, based on a
-task-based model~\cite{Jones:2015soc}~\cite{Clemencic:2015paa}. O2 already 
+task-based model~\cite{Jones:2015soc,Clemencic:2015paa}. O2 already
 includes additional multi-node setups and communication.
@@
 The frameworks provide the necessary functionality like I/O,
-scheduling, configuration, logging, etc to support the execution of
-these processing components. The mentioned components provide
+scheduling, configuration, logging, etc.\ to support the execution of
+these processing components. The aforementioned components provide
 functionalities like pattern finding in a certain sub-detector or the
@@
 multithreading approach too. Both experiments were
-driven to solve this problem, by the ever growing needs for more
+driven to solve this problem by the ever growing needs for more
 memory per process brought on by the increasing complexity of LHC
@@
 scaling by dividing up datasets, generating a framework application
-configuration, and scheduling jobs on nodes/cores to consume all
+configuration, and scheduling jobs on nodes/cores to utilize all
 available resources. Given anticipated changes in hardware
@@
 computing facilities, the interplay between workflow/workload
-management systems and framework applications need to be carefully
+management systems and framework applications needs to be carefully
 examined. It may be advantageous to permit framework applications (or
@@
 Face to face workshops after at least the 1st and the 3rd year can be co-hosted with events like CHEP and/or the WLCG workshops.
-A future planning workshop for transforming the results of the R&D activities into a full development and deployment project plan should happen at the 5-year timescale.
+A future planning workshop for transforming the results of the R\&D activities into a full development and deployment project plan should happen at the 5-year timescale.
 
@@
 dependencies, but also resource preferences and constraints, such as
-GPU. The goal here is to provide enough information for a group to
+GPUs. The goal here is to provide enough information for a group to
 take on development of domain-specific libraries components and tools

### 2018-10-12 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add CSBS first version
Use copy of class file and packages from Michel Jouvin.
Add class files for Springer style.
Convert floats into multi-column versions (harmess for single column
layout used with JHEP + arXiv).
CWP: appendices as a separate file, one column in cwp-csbs
- Also fix a non itemized entry in the glossary (Concurrency Forum)

@@
 
-\begin{figure}[bthp]
+\begin{figure*}[bthp]
     \centering
@@
     \label{fig:hl-lhc}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \vspace*{0.3cm}
@@
     \label{fig:fnal-if}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{table}
+\begin{table*}
     \centering
@@
     \label{tab:crsg2017}
-\end{table}
+\end{table*}
 
@@
 
-\begin{figure}
+\begin{figure*}
     \centering
@@
     \label{fig:cms-resources}
-\end{figure}
-
-\begin{figure}
+\end{figure*}
+
+\begin{figure*}
     \centering
@@
     \label{fig:atlas-resources}
-\end{figure}
+\end{figure*}
 
@@
 
-\newpage
-\begin{appendices}
-
-\hypertarget{appendix-a---list-of-workshops}{%
-\section{List of Workshops}\label{appendix-a---list-of-workshops}}
-
-\textbf{HEP Software Foundation Workshop}\\
-\emph{Date:} 23-26 Jan, 2017\\
-\emph{Location:} UCSD/SDSC (La Jolla, CA, USA)\\
-\emph{URL:}
-\href{http://indico.cern.ch/event/570249/}{{http://indico.cern.ch/event/570249/}}\\
-\emph{Description:} This HSF workshop at SDSC/UCSD was the first
-workshop supporting the CWP process. There were plenary sessions
-covering topics of general interest as well as parallel sessions for the
-many topical working groups in progress for the CWP.\\
-
-\noindent
-\textbf{Software Triggers and Event Reconstruction WG meeting}\\
-\emph{Date:} 9 Mar, 2017\\
-\emph{Location:} LAL-Orsay (Orsay, France)\\
-\emph{URL:}
-\href{https://indico.cern.ch/event/614111/}{{https://indico.cern.ch/event/614111/}}\\
-\emph{Description:} This was a meeting of the Software Triggers and Event
-Reconstruction CWP working group. It was held as a parallel session at
-the ``Connecting the Dots'' workshop, which focuses on forward-looking
-pattern recognition and machine learning algorithms for use in HEP.\\
-
-\noindent
-\textbf{IML Topical Machine Learning Workshop}\\
-\emph{Date:} 20-22 Mar, 2017\\
-\emph{Location:} CERN (Geneva, Switzerland)\\
-\emph{URL:}
-\href{https://indico.cern.ch/event/595059}{{https://indico.cern.ch/event/595059}}\\
-\emph{Description:} This was a meeting of the Machine Learning CWP
-working group. It was held as a parallel session at the
-``Inter-experimental Machine Learning (IML)'' workshop, an organisation
-formed in 2016 to facilitate communication regarding R\&D on ML
-applications in the LHC experiments.\\
-
-\noindent
-\textbf{Community White Paper Follow-up at FNAL}\\
-\emph{Date:} 23 Mar, 2017\\
-\emph{Location:} FNAL (Batavia, IL, USA)\\
-\emph{URL:}
-\href{https://indico.fnal.gov/conferenceDisplay.py?confId=14032}{{https://indico.fnal.gov/conferenceDisplay.py?confId=14032}}\\
-\emph{Description:} This one-day workshop was organised to engage with
-the experimental HEP community involved in computing and software for
-Intensity Frontier experiments at FNAL. Plans for the CWP were
-described, with discussion about commonalities between the HL-LHC
-challenges and the challenges of the FNAL neutrino and muon
-experiments\\
-
-\noindent
-\textbf{CWP Visualisation Workshop}\\
-\emph{Date:} 28-30 Mar, 2017\\
-\emph{Location:} CERN (Geneva, Switzerland)\\
-\emph{URL:}
-\href{https://indico.cern.ch/event/617054/}{{https://indico.cern.ch/event/617054/}}\\
-\emph{Description:} This workshop was organised by the Visualisation CWP
-working group. It explored the current landscape of HEP visualisation
-tools as well as visions for how these could evolve. There was
-participation both from HEP developers and industry.\\
-
-\noindent
-\textbf{DS@HEP 2017 (Data Science in High Energy Physics)}\\
-\emph{Date:} 8-12 May, 2017\\
-\emph{Location:} FNAL (Batava, IL, USA)\\
-\emph{URL:}
-\href{https://indico.fnal.gov/conferenceDisplay.py?confId=13497}{{https://indico.fnal.gov/conferenceDisplay.py?confId=13497}}\\
-\emph{Description:} This was a meeting of the Machine Learning CWP
-working group. It was held as a parallel session at the ``Data Science
-in High Energy Physics (DS@HEP)'' workshop, a workshop series begun in
-2015 to facilitate communication regarding R\&D on ML applications in
-HEP.\\
-
-\noindent
-\textbf{HEP Analysis Ecosystem Retreat}\\
-\emph{Date:} 22-24 May, 2017\\
-\emph{Location:} Amsterdam, the Netherlands\\
-\emph{URL:}
-\href{http://indico.cern.ch/event/613842/}{{http://indico.cern.ch/event/613842/}}\\
-\emph{Summary report:}
-\href{http://hepsoftwarefoundation.org/assets/AnalysisEcosystemReport20170804.pdf}{{http://cern.ch/go/mT8w}}\\
-\emph{Description:} This was a general workshop, organised about the
-HSF, about the ecosystem of analysis tools used in HEP and the ROOT
-software framework. The workshop focused both on the current status and
-the 5-10 year time scale covered by the CWP.\\
-
-\noindent
-\textbf{CWP Event Processing Frameworks Workshop}\\
-\emph{Date:} 5-6 Jun, 2017\\
-\emph{Location:} FNAL (Batavia, IL, USA)\\
-\emph{URL:}
-\href{https://indico.fnal.gov/conferenceDisplay.py?confId=14186}{{https://indico.fnal.gov/conferenceDisplay.py?confId=14186}}\\
-\emph{Description:} This was a workshop held by the Event Processing Frameworks
-CWP working group focused on writing an initial draft of the framework
-white paper. Representatives from most of the current practice
-frameworks participated.\\
-
-\noindent
-\textbf{HEP Software Foundation Workshop}\\
-\emph{Date:} 26-30 Jun, 2017\\
-\emph{Location:} LAPP (Annecy, France)\\
-\emph{URL:}
-\href{https://indico.cern.ch/event/613093/}{{https://indico.cern.ch/event/613093/}}\\
-\emph{Description:} This was the final general workshop for the CWP
-process. The CWP working groups came together to present their status
-and plans, and develop consensus on the organisation and context for the
-community roadmap. Plans were also made for the CWP writing phase that
-followed in the few months following this last workshop.
-
-\newpage
-\hypertarget{appendix-b-glossary}{%
-\section{Glossary}\label{appendix-b-glossary}}
-
-\begin{description}
-
-\item[AOD] Analysis Object Data is a summary of the reconstructed event and
-contains sufficient information for common physics analyses.
-
-\item[ALPGEN] An event generator designed for the generation of Standard Model
-processes in hadronic collisions, with emphasis on final states with
-large jet multiplicities. It is based on the exact LO evaluation of
-partonic matrix elements, as well as top quark and gauge boson decays
-with helicity correlations.
-
-\item[BSM] Physics beyond the Standard Model (BSM) refers to the theoretical
-developments needed to explain the deficiencies of the Standard Model
-(SM), such as the
-\href{https://en.wikipedia.org/wiki/Origin_of_mass}{origin of mass}, the
-\href{https://en.wikipedia.org/wiki/Strong_CP_problem}{strong CP
-problem},
-\href{https://en.wikipedia.org/wiki/Neutrino_oscillation}{neutrino
-oscillations},
-\href{https://en.wikipedia.org/wiki/Baryon_asymmetry}{matter--antimatter
-asymmetry}, and the nature of
-\href{https://en.wikipedia.org/wiki/Dark_matter}{dark matter} and
-\href{https://en.wikipedia.org/wiki/Dark_energy}{dark energy}.
-
-\item[Coin3D] A C++ object oriented retained mode 3D graphics API
-used to provide a higher layer of programming for OpenGL.
-
-\item[COOL] LHC Conditions Database Project, a subproject of the POOL persistency framework.
-
-Concurrency Forum: Software engineering is moving towards a paradigm shift in order to accommodate new CPU architectures with many cores, in which concurrency will play a more fundamental role in programming languages and libraries.
-The forum on concurrent programming models and frameworks aims to share knowledge among interested parties that work together to develop 'demonstrators' and agree on technology so that they can share code and compare results.
-
-\item[CRSG] Computing Resources Scrutiny Group, a WLCG committee in charge of scrutinizing and assessing LHC experiment yearly resource requests to prepare funding agency decisions.
-
-\item[CSIRT] Computer Security Incident Response Team.
-A CSIRT provides a reliable and trusted single point of contact for reporting computer security incidents and taking the appropriate measures in response tothem.
-
-\item[CVMFS] The CERN Virtual Machine File System is a network file system
-based on HTTP and optimised to deliver experiment software in a fast,
-scalable, and reliable way through sophisticated caching strategies.
-
-\item[CWP] The Community White Paper (this document) is the result of an
-organised effort to describe the community strategy and a roadmap for
-software and computing R\&D in HEP for the 2020s. This activity is
-organised under the umbrella of the HSF.
-
-\item[Deep Learning (DL)] one class of Machine Learning algorithms, based on a
-high number of neural network layers.
-
-\item[DNN] Deep Neural Network, class of neural networks with typically a
-large number of hidden layers through which data is processed.
-
-\item[DPHEP] The Data Preservation in HEP project is a collaboration for data
-preservation and long term analysis.
-
-\item[EGI] European Grid Initiative. A European organisation in charge of
-delivering advanced computing services to support scientists,
-multinational projects and research infrastructures, partially funded by
-the European Union. It is operating both a grid infrastructure (many
-WLCG sites in Europe are also EGI sites) and a federated cloud
-infrastructure. It is also responsible for security incident response
-for these infrastructures (CSIRT).
-
-\item[FAIR] The Facility for Antiproton and Ion Research (FAIR) is located at
-GSI Darmstadt. It is an international accelerator facility for research
-with antiprotons and ions.
-
-\item[FAIR] An abbreviation for a set of desirable data properties: Findable,
-Accessible, Interoperable, and Re-usable.
-
-\item[FCC] Future Circular Collider, a proposed new accelerator complex for
-CERN, presently under study.
-
-\item[FCC-hh] A 100 TeV proton-proton collider version of the FCC (the ``h''
-stands for ``hadron'').
-
-\item[GAN] Generative Adversarial Networks are a class of
-\href{https://en.wikipedia.org/wiki/Artificial_intelligence}{artificial
-intelligence} algorithms used in
-\href{https://en.wikipedia.org/wiki/Unsupervised_machine_learning}{unsupervised
-machine learning}, implemented by a system of two
-\href{https://en.wikipedia.org/wiki/Neural_network}{neural networks}
-contesting with each other in a
-\href{https://en.wikipedia.org/wiki/Zero-sum_game}{zero-sum game}
-framework.
-
-\item[Geant4] A toolkit for the simulation of the passage of
-particles through matter.
-
-\item[GeantV] An R\&D project that aims to fully exploit the
-parallelism, which is increasingly offered by the new generations of CPUs, in the field of detector simulation.
-
-\item[GPGPU] General-Purpose computing on Graphics Processing Units is the use of a Graphics Processing Unit (GPU), which typically handles computation only for computer graphics, to perform computation in applications traditionally handled by the Central Processing Unit (CPU).
-Programming for GPUs is typically more challenging, but can offer significant gains in arithmetic throughput.
-
-\item[HEPData] The Durham High Energy Physics Database is an open access
-repository for scattering data from experimental particle physics.
-
-\item[HERWIG] This is an event generator containing a wide range of Standard
-Model, Higgs and supersymmetric processes. It uses the parton-shower approach
-for initial- and final-state QCD radiation, including colour coherence effects and 
-azimuthal correlations both within and between jets.
-
-\item[HL-LHC] The High Luminosity Large Hadron Collider is a proposed upgrade to
-the Large Hadron Collider to be made in 2026. The upgrade aims at increasing the 
-luminosity of the machine by a factor of 10, up to 
-$10^{35}\mathrm{cm}^{-2}\mathrm{s}^{-1}$, 
-providing a better chance to see rare processes and improving
-statistically marginal measurements.
-
-\item[HLT] High Level Trigger. The computing resources, generally a large farm, close to the detector which process the events in real-time and select those who must be stored for further analysis.
-
-\item[HPC] High Performance Computing.
-
-\item[HS06] HEP-wide benchmark for measuring CPU performance based on the SPEC2006 benchmark
-(\href{https://www.spec.org}{{https://www.spec.org}}).
-
-\item[HSF] The HEP Software Foundation facilitates coordination and common
-efforts in high energy physics (HEP) software and computing
-internationally.
-
-\item[IML] The Inter-experimental LHC Machine Learning (IML) Working Group is
-focused on the development of modern state-of-the art machine learning
-methods, techniques and practices for high-energy physics problems.
-
-\item[IOV] Interval Of Validity, the period of time for which a specific piece
-of conditions data is valid.
-
-\item[JavaScript] A high-level, dynamic, weakly typed,
-prototype-based, multi-paradigm, and interpreted programming language.
-Alongside HTML and CSS, JavaScript is one of the three core technologies
-of World Wide Web content production.
-
-\item[Jupyter Notebook] This is a server-client application that allows
-editing and running notebook documents via a web browser. Notebooks are
-documents produced by the Jupyter Notebook App, which contain both
-computer code (e.g., python) and rich text elements (paragraph,
-equations, figures, links, etc...). Notebook documents are both
-human-readable documents containing the analysis description and the
-results (figures, tables, etc..) as well as executable documents which
-can be run to perform data analysis.
-
-\item[LHC] Large Hadron Collider, the main particle accelerator at CERN.
-
-\item[LHCONE] A set of network circuits, managed worldwide by the National
-Research and Education Networks, to provide dedicated transfer paths for
-LHC T1/T2/T3 sites on the standard academic and research physical
-network infrastructure.
-
-\item[LHCOPN] LHC Optical Private Network. It is the private physical and IP
-network that connects the Tier0 and the Tier1 sites of the WLCG.
-
-\item[MADEVENT] This is a multi-purpose tree-level event generator. It is
-powered by the matrix element event generator MADGRAPH, which generates
-the amplitudes for all relevant sub-processes and produces the mappings
-for the integration over the phase space.
-
-\item[Matplotlib] This is a Python 2D plotting library that provides
-publication quality figures in a variety of hardcopy formats and
-interactive environments across platforms.
-
-\item[ML] Machine learning is a field of computer science that gives computers
-the ability to learn without being explicitly programmed. It focuses on
-prediction making through the use of computers and emcompasses a lot of
-algorithm classes (boosted decision trees, neural networks\ldots{}).
-
-\item[MONARC] A model of large scale distributed computing based on many
-regional centers, with a focus on LHC experiments at CERN. As part of
-the MONARC project, a simulation framework was developed that provides a
-design and optimisation tool. The MONARC model has been the initial
-reference for building the WLCG infrastructure and to organise the data
-transfers around it.
-
-\item[OpenGL] Open Graphics Library is a cross-language, cross-platform
-application programming interface(API) for rendering 2D and 3D vector
-graphics. The API is typically used to interact with a graphics
-processing unit(GPU), to achieve hardware-accelerated rendering.
-
-\item[Openlab] CERN openlab is a public-private partnership that accelerates
-the development of cutting-edge solutions for the worldwide LHC
-community and wider scientific research.
-
-\item[P5] The Particle Physics Project Prioritization Panel is a scientific
-advisory panel tasked with recommending plans for U.S. investment in
-particle physics research over the next ten years.
-
-\item[PRNG] A PseudoRandom Number Generator is an algorithm for generating a
-sequence of numbers whose properties approximate the properties of
-sequences of random numbers.
-
-\item[PyROOT] A \href{http://www.python.org/}{Python} extension module that
-allows the user to interact with any ROOT class from the Python
-interpreter.
-
-\item[PYTHIA] A program for the generation of high-energy physics events, i.e.,
-for the description of collisions at high energies between elementary
-particles such as e+, e-, p and pbar in various combinations. It
-contains theory and models for a number of physics aspects, including
-hard and soft interactions, parton distributions, initial- and
-final-state parton showers, multiparton interactions, fragmentation and
-decay.
-
-\item[QCD] Quantum Chromodynamics, the theory describing the strong
-interaction between quarks and gluons.
-
-\item[REST] Representational State Transfer
-\href{https://en.wikipedia.org/wiki/Web_service}{web services} are a way
-of providing interoperability between computer systems on the Internet.
-One of its main features is stateless interactions between clients and
-servers (every interaction is totally independent of the others),
-allowing for very efficient caching.
-
-\item[ROOT] A modular scientific software framework widely used in HEP data
-processing applications.
-
-\item[SAML] Security Assertion Markup Language. It is an open, XML-based,
-standard for exchanging authentication and authorisation data between
-parties, in particular, between an identity provider and a service
-provider.
-
-\item[SDN] Software-defined networking is an umbrella term encompassing several kinds of network technology aimed at making the network as agile
-and flexible as the virtualised server and storage infrastructure of the modern data center.
-
-\item[SHERPA] Sherpa is a Monte Carlo event generator for the Simulation of
-High-Energy Reactions of PArticles in lepton-lepton, lepton-photon,
-photon-photon, lepton-hadron and hadron-hadron collisions.
-
-\item[SIMD] Single instruction, multiple data (\textbf{SIMD}), describes
-computers with multiple processing elements that perform the same
-operation on multiple data points simultaneously.
-
-\item[SM] The Standard Model is the name given in the 1970s to a theory of
-fundamental particles and how they interact. It is the currently
-dominant theory explaining the elementary particles and their dynamics.
-
-\item[SWAN] Service for Web based ANalysis is a platform for interactive data
-mining in the CERN cloud using the Jupyter notebook interface.
-
-\item[TBB] Intel Threading Building Blocks is a widely used C++ template
-library for task parallelism. It lets you easily write parallel C++
-programs that take full advantage of multicore performance.
-
-\item[TMVA] The Toolkit for Multivariate Data Analysis with ROOT is a
-standalone project that provides a ROOT-integrated machine learning
-environment for the processing and parallel evaluation of sophisticated
-multivariate classification techniques.
-
-\item[VecGeom] The vectorised geometry library for particle-detector simulation.
-
-\item[VO] Virtual Organisation. A group of users sharing a common interest
-(for example, each LHC experiment is a VO), centrally managed, and used
-in particular as the basis for authorisations in the WLCG
-infrastructure.
-
-\item[WebGL] The Web Graphics Library is a JavaScript API for rendering
-interactive 2D and 3D graphics within any compatible web browser without
-the use of plug-ins.
-
-\item[WLCG] The Worldwide LHC Computing Grid project is a global collaboration
-of more than 170 computing centres in 42 countries, linking up national
-and international grid infrastructures. The mission of the WLCG project
-is to provide global computing resources to store, distribute and
-analyse data generated by the Large Hadron Collider (LHC) at CERN.
-
-\item[X.509] A cryptographic standard which defines how to implement service
-security using electronic certificates, based on the use of a private
-and public key combination. It is widely used on web servers accessed
-using the https protocol and is the main authentication mechanism on the
-WLCG infrastructure.
-
-\item[x86\_64] 64-bit version of the x86 instruction set.
-
-\item[XRootD] Software framework that is a fully generic suite for fast, low latency and scalable data access.
-
-\end{description}
-
-\end{appendices}
+\onecolumn

### 2018-10-19 CWP/papers/HSF-CWP-2017-08_framework/latex/df_fwk_summary.tex
Improve references
Found what references I could and removed all of the
placeholders

@@
 \author[a]{Graeme A Stewart}
-\author[d]{Several Others - Charles, Marco, ?}
+%\author[e]{Several Others - Charles, Marco, ?}
 
@@
 \affiliation[d]{Lawrence Berkeley National Laboratory, Berkeley, CA, USA}
-\affiliation[e]{Other Places}
+%\affiliation[e]{Other Places}
 \affiliation[1]{Paper Editor}
@@
 of framework use cases written for the Gaudi collaboration is
-described in {[}2.fixme{]} and these are sufficiently general to apply
-to all HEP experiments. The Gaudi framework {[}70{]} was originally
+described in \cite{Maley:684466} and these are sufficiently general to apply
+to all HEP experiments. The Gaudi framework \cite{Barrand:2001ny,1742-6596-219-4-042006} was originally
 developed by LHCb, but is also used by ATLAS and various non-LHC
-experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff},
-{[}71{]} which was forked to provide the art framework for the Fermilab
-Intensity Frontier experiments~\cite{Green:2012gv},{[}72{]}.  Belle II uses
-basf2~\cite{1742-6596-331-3-032024},{[}73{]}. The linear collider community
-developed and uses Marlin~\cite{Gaede:2006pj},{[}74{]}. The FAIR experiments
+experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff}
+which was forked to provide the art framework for the Fermilab
+Intensity Frontier experiments~\cite{Green:2012gv}.  Belle II uses
+basf2~\cite{1742-6596-331-3-032024}. The linear collider community
+developed and uses Marlin~\cite{Gaede:2006pj}. The FAIR experiments
 use FairROOT, closely related to ALICE's AliROOT. The FAIR experiments
 and ALICE are now developing a new framework, which is called
-O2~\cite{O2},{[}75{]}. At the time of writing, most major frameworks support
+O2~\cite{O2,1742-6596-664-8-082046}. At the time of writing, most major frameworks support
 basic parallelisation, both within and across events, based on a

### 2018-10-19 CWP/papers/HSF-CWP-2017-01_roadmap/latex/appendices.tex
[WIP] Finalization of CWP version for CSBS journal (#104)
* Use one column for author list

* CWP: appendices as a separate file, one column in cwp-csbs

* Fix a non itemized entry in the glossary (Concurrency Forum)

--- 2018-01-16 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp.tex
@@
-% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.
-
-% JHEP preprint template
-\documentclass[12pt,a4paper]{article}
-\usepackage{jheppub}
-
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage[titletoc]{appendix}
-\usepackage{xspace}
-
-\usepackage[utf8]{inputenc}
-
-\usepackage[style=numeric-comp,sorting=none]{biblatex}
-\addbibresource{cwp.bib}
-\addbibresource{cwp-chapters.bib}
-
-%% Comment out before final submission
-%\usepackage{lineno}  % for line numbering during review
-%\linenumbers
-
-\abstract{Particle physics has an ambitious and broad experimental 
-programme for the coming decades. This programme requires large 
-investments in detector hardware, either to build new facilities 
-and experiments, or to upgrade existing ones.
-Similarly, it requires commensurate investment in the R\&D of software to acquire, manage, process, and analyse the shear amounts of data to be recorded.
-In planning for the HL-LHC in particular, it is critical that all of the
-collaborating stakeholders agree on the software goals and priorities,
-and that the efforts complement each other. In this spirit, this white paper
-describes the R\&D activities required to prepare for this software upgrade.}
-
-\begin{document}
-
-\noindent
-\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
- & & HSF-CWP-2017-01 \\
- & & December 15, 2017 \\ % use \date or hardwire e.g. December 15, 2017
- & & \\
-\end{tabular*}
-\vspace{2.0cm}
-
-\title{A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s}
-
-\author{HEP Software Foundation\footnote{Authors are listed at the end of this report.}}
-
-\maketitle
-
-\newpage
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Particle physics has an ambitious experimental programme for the
-coming decades. The programme supports the strategic goals of the
-particle physics community that have been laid out by the European
-Strategy for Particle Physics~\cite{ESPP2013} and by
-the Particle Physics Project Prioritization Panel (P5)~\cite{DOE-P5} in the United
-States~\cite{2014pwa}. Broadly speaking, the
-scientific goals are:
-
-\begin{itemize}
-\item Exploit the discovery of the Higgs boson as a precision tool
-  for investigating Standard Model (SM) and Beyond the Standard Model
-  (BSM) physics.
-\item
-  Study the decays of $b$- and $c$-hadrons, and tau leptons, in the search
-  for manifestations of BSM physics, and investigate
-  matter-antimatter differences.
-\item
-  Search for signatures of dark matter.
-\item
-  Probe neutrino oscillations and masses.
-\item
-  Study the Quark Gluon Plasma state of matter in heavy-ion collisions.
-\item
-  Explore the unknown.
-\end{itemize}
-
-The High-Luminosity Large Hadron Collider
-(HL-LHC)~\cite{HL-LHC,1742-6596-515-1-012012,Apollinari:2284929} 
-will be a major
-upgrade of the current LHC~\cite{LHC} supporting the aim of 
-an in-depth investigation of the properties of the Higgs boson and its
-couplings to other particles (Figure~\ref{fig:hl-lhc}). The ATLAS~\cite{ATLAS} and CMS~\cite{CMS}
-collaborations will continue to make measurements in the Higgs sector,
-while searching for new physics Beyond the Standard Model (BSM). Should
-a BSM discovery be made, a full exploration of that physics will be
-pursued. Such BSM physics may help shed light on the nature of dark
-matter, which we know makes up the majority of gravitational matter in
-the universe, but which does not interact via the electromagnetic or
-strong nuclear forces~\cite{Mangano2016}.
-
-The LHCb experiment at the LHC~\cite{LHCb} and the Belle II experiment at
-KEK~\cite{BelleII} study various aspects of heavy
-flavour physics (b- and c-quark, and tau-lepton physics), where quantum
-influences of very high mass particles manifest themselves in lower
-energy phenomena. Their primary goal is to look for BSM physics, either
-by studying CP violation (that is, asymmetries in the behaviour of
-particles and their corresponding antiparticles) or modifications in
-rate or angular distributions in rare heavy-flavour decays. Current
-manifestations of such asymmetries do not explain why our universe is so
-matter dominated. These flavour physics programmes are related to BSM
-searches through effective field theory, and powerful constraints on new
-physics keep coming from such studies.
-
-\begin{figure}[bthp]
-    \centering
-    \includegraphics[width=0.94\textwidth]{HL-LHC-plan-2017-Plan-2.pdf}
-    \caption{The current schedule for the LHC and HL-LHC upgrade
-and run~\cite{HL-LHC}. Currently, the start of the HL-LHC run is foreseen for mid
-2026. The long shutdowns, LS2 and LS3, will be used to upgrade both the
-accelerator and the detector hardware.}
-    \label{fig:hl-lhc}
-\end{figure}
-
-The study of neutrinos, their mass and oscillations, can also shed light
-on matter-antimatter asymmetry. The DUNE experiment will provide a huge
-improvement in our ability to probe neutrino physics, detecting
-neutrinos from the Long Baseline Neutrino Facility at Fermilab, as well
-as linking to astro-particle physics programmes, in particular through
-the potential detection of supernovas and relic neutrinos. An overview
-of the experimental programme scheduled at the Fermilab facility is
-given in Figure~\ref{fig:fnal-if}.
-
-\begin{figure}
-\vspace*{0.3cm}
-    \centering
-    \includegraphics[width=0.94\textwidth]{10yr-PLAN-9Feb-2017-V7a.pdf}
-    \caption{Run schedule for the Fermilab facility until 2026.}
-    \label{fig:fnal-if}
-\end{figure}
-
-In the study of the early universe immediately after the Big Bang, it is
-critical to understand the phase transition between the highly
-compressed quark-gluon plasma and the nuclear matter in the universe
-today. The ALICE experiment at the LHC~\cite{ALICE} and
-the CBM~\cite{CBM} and PANDA~\cite{PANDA} experiments at the Facility for Antiproton and Ion
-Research (FAIR) are specifically designed to probe this aspect of
-nuclear and particle physics. In addition ATLAS, CMS and LHCb all
-contribute to the LHC heavy-ion programme.
-
-These experimental programmes require large investments in detector
-hardware, either to build new facilities and experiments (e.g., FAIR and DUNE) or to upgrade existing ones (HL-LHC, Belle II). Similarly, they
-require commensurate investment in the research and development
-necessary to deploy software to acquire, manage, process, and analyse
-the data recorded.
-
-For the HL-LHC, which is scheduled to begin taking data in 2026 (Figure~\ref{fig:hl-lhc}) and to run into the 2030s, some 30 times more data than
-the LHC has currently produced will be collected by ATLAS and CMS. As
-the total amount of LHC data already collected is close to an exabyte,
-it is clear that the problems to be solved require approaches beyond
-simply scaling current solutions, assuming Moore's Law and more or less
-constant operational budgets. The nature of computing hardware
-(processors, storage, networks) is evolving with radically new
-paradigms, the quantity of data to be processed is increasing
-dramatically, its complexity is increasing, and more sophisticated
-analyses will be required to maximise physics yield. Developing and
-deploying sustainable software for future and upgraded experiments,
-given these constraints, is both a technical and a social challenge, as
-detailed in this paper. An important message of this report is that a
-``software upgrade'' is needed to run in parallel with the hardware
-upgrades planned for the HL-LHC in order to take full advantage of these hardware upgrades and to complete the HL-LHC physics programme.
-
-In planning for the HL-LHC in particular, it is critical that all of the
-collaborating stakeholders agree on the software goals and priorities,
-and that the efforts complement each other. In this spirit, the HEP
-Software Foundation (HSF) began a planning exercise in late 2016 to
-prepare a Community White Paper (CWP)~\cite{HSF2017} at the behest of the Worldwide LHC
-Computing Grid (WLCG) project~\cite{WLCG2016}. The role of the HSF is to
-facilitate coordination and common efforts in HEP software and computing
-internationally and to provide a structure for the community to set
-goals and priorities for future work. The objective of the CWP is to
-provide a roadmap for software R\&D in preparation for the HL-LHC and
-for other HEP experiments on a similar timescale, which would identify
-and prioritise the software research and development investments
-required:
-
-\begin{itemize}
-\item
-  to achieve improvements in software efficiency, scalability and
-  performance, and to make use of advances in CPU, storage and network
-  technologies in order to cope with the challenges ahead;
-\item
-  to enable new approaches to computing and software that can radically
-  extend the physics reach of the detectors;
-\item
-  to ensure the long-term sustainability of the software through the
-  lifetime of the HL- LHC;
-\item
-  to ensure data and knowledge preservation beyond the lifetime of
-  individual experiments;
-\item
-  to attract the required new expertise by offering appropriate career
-  recognition to physicists specialising in software development, and by
-  an effective training effort to target all contributors in the
-  community.
-\end{itemize}
-
-The CWP process, organised by the HSF with the participation of the LHC
-experiments and the wider HEP software and computing community, began
-with a kick-off workshop at the San Diego Supercomputer Centre (SDSC),
-USA, in January 2017 and concluded after a final workshop in June 2017 at
-the Laboratoire d'Annecy de Physique des Particules (LAPP), France, with
-a large number of intermediate topical workshops and meetings
-(Appendix \ref{appendix-a---list-of-workshops}). The
-entire CWP process involved an estimated 250 participants.
-
-To reach more widely than the LHC experiments, specific contact was made
-with individuals with software and computing responsibilities in the
-Fermilab muon and neutrino experiments, Belle II, the Linear Collider
-community, as well as various national computing organisations. The CWP
-process was able to build on all the links established since the
-inception of the HSF in 2014.
-
-Working groups were established on various topics which were expected to
-be important parts of the HL-LHC roadmap: \emph{Careers, Staffing and
-Training; Conditions Database; Data Organisation, Management and Access;
-Data Analysis and Interpretation; Data and Software Preservation;
-Detector Simulation; Data-Flow Processing Frameworks; Facilities and
-Distributed Computing; Machine Learning; Physics Generators; Security;
-Software Development, Deployment and Validation/Verification; Software
-Trigger and Event Reconstruction;} and \emph{Visualisation.} The work of
-each working group is summarised in this document.
-
-This document is the result of the CWP process. Investing in the roadmap
-outlined here will be fruitful for the whole of the HEP programme and
-may also benefit other projects with similar technical challenges,
-particularly in astrophysics, e.g., the Square Kilometre Array (SKA)~\cite{SKA},
-the Cherenkov Telescope Array (CTA)~\cite{CTA} and the Large Synoptic Survey
-Telescope (LSST)~\cite{LSST}.
-
-\hypertarget{software-and-computing-challenges}{%
-\section{Software and Computing
-Challenges}\label{software-and-computing-challenges}}
-
-Run 2 for the LHC started in 2015 and delivered a proton-proton
-collision energy of
-\href{http://home.cern/about/engineering/restarting-lhc-why-13-tev}{13
-TeV}. By the end of LHC Run 2 in 2018, it is expected that about 150
-fb\textsuperscript{-1} of physics data will have been collected by both
-ATLAS and CMS. Together with ALICE and LHCb, the total size of LHC data
-storage pledged by sites for the year 2017 is around 1 exabyte, as shown
-in Table~\ref{tab:crsg2017} from the LHC's Computing Resource Scrutiny Group
-(CRSG)~\cite{Lucchesi:2284575}. The CPU allocation from the CRSG for 2017 to each
-experiment is also shown.
-
-\begin{table}
-    \centering
-\begin{tabular}{lp{0.15\textwidth}p{0.15\textwidth}p{0.15\textwidth}p{0.15\textwidth}}
-\hline
-\textbf{Experiment} & \textbf{2017 Disk Pledges (PB)} & \textbf{2017 Tape Pledges (PB)} & \textbf{Total Disk and Tape Pledges (PB)} & \textbf{2017 CPU Pledges (kHS06)}\\
-\hline
-ALICE & 67 & 68 & 138 & 807\\
-ATLAS & 172 & 251 & 423 & 2194\\
-CMS & 123 & 204 & 327 & 1729\\
-LHCb & 35 & 67 & 102 & 413\\
-\textbf{Total} & \textbf{400} & \textbf{591} & \textbf{990} &
-\textbf{5143}\\
-\hline
-\end{tabular}
-    \caption{Resources pledged by WLCG sites to the 4 LHC experiments for
-the year 2017 as described at the September 2017 session of the
-Computing Resources Scrutiny Group (CRSG).}
-    \label{tab:crsg2017}
-\end{table}
-
-Using an approximate conversion from HS06~\cite{HS06} to CPU cores of 10
-means that LHC computing in 2017 is supported by about 500k CPU cores.
-These resources are deployed ubiquitously, from close to the experiments
-themselves at CERN to a worldwide distributed computing infrastructure,
-the WLCG~\cite{WLCG}. Each experiment has developed its own workflow management and
-data management software to manage its share of WLCG resources.
-
-In order to process the data, the 4 largest LHC experiments have written
-more than 20 million lines of program code over the last 15 years. This
-has involved contributions from thousands of physicists and many
-computing professionals, encompassing a wide range of skills and
-abilities. The majority of this code was written for a single
-architecture (x86\_64) and with a serial processing model in mind. There
-is considerable anxiety in the experiments that much of this software is
-not sustainable, with the original authors no longer in the field and
-much of the code itself in a poorly maintained state, ill-documented, and
-lacking tests. This code, which is largely experiment-specific, manages
-the entire experiment data flow, including data acquisition, high-level
-triggering, calibration and alignment, simulation, reconstruction (of
-both real and simulated data), visualisation, and final data analysis.
-
-HEP experiments are typically served with a large set of integrated and
-configured common software components, which have been developed either
-in-house or externally. Well-known examples include ROOT~\cite{Brun1996}, which is a data analysis toolkit that
-also plays a critical role in the implementation of experiments' data
-storage systems, and Geant4~\cite{Agostinelli2003}, a
-simulation framework through which most detector simulation is achieved.
-Other packages provide tools for supporting the development process;
-they include compilers and scripting languages, as well as tools for
-integrating, building, testing, and generating documentation. Physics
-simulation is supported by a wide range of event generators provided by
-the theory community (PYTHIA~\cite{PYTHIA}, SHERPA~\cite{Gleisberg:2008ta}, ALPGEN~\cite{Mangano:565290}, MADGRAPH~\cite{MADGRAPH}, HERWIG~\cite{HERWIG},
-amongst many others). There is also code developed to support the
-computing infrastructure itself, such as the CVMFS distributed caching
-filesystem~\cite{1742-6596-331-4-042003}, the Frontier database caching 
-mechanism~\cite{Frontier}, the XRootD file access software~\cite{XRootD} and a
-number of storage systems (dCache, DPM, EOS). This list of packages is % REF?
-by no means exhaustive, but illustrates the range of software employed
-and its critical role in almost every aspect of the programme.
-
-Already in Run 3 LHCb will process more than 40 times the number of
-collisions that it does today, and ALICE will read out Pb-Pb collisions
-continuously at 50 kHz. The upgrade to the HL-LHC for Run 4 then
-produces a step change for ATLAS and CMS. The beam intensity will rise
-substantially, giving bunch crossings where the number of discrete
-proton-proton interactions (pileup) will rise to about 200, from about 60
-today. This has important consequences for the operation of the
-detectors and for the performance of the reconstruction software. The
-two experiments will upgrade their trigger systems to record 5-10 times
-as many events as they do today. It is anticipated that HL-LHC will
-deliver about 300 fb\textsuperscript{-1} of data each year.
-
-The steep rise in resources that are then required to manage this data
-can be estimated from an extrapolation of the Run 2 computing model and
-is shown in Figures~\ref{fig:cms-resources} and~\ref{fig:atlas-resources}.
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.47\textwidth}
-        \includegraphics[width=\textwidth]{cms-cpu.png}
-        \caption{}
-        \label{fig:cms-cpu}
-    \end{subfigure}
-    \quad
-    \begin{subfigure}[b]{0.47\textwidth}
-        \includegraphics[width=\textwidth]{cms-disk.png}
-        \caption{}
-        \label{fig:cms-disk}
-    \end{subfigure}
-    \caption{CMS estimated CPU (\ref{fig:cms-cpu}) and disk space (\ref{fig:cms-disk}) 
-    resources required into the HL-LHC era, using the current computing model 
-    with parameters projected out for the next 12 years.}
-    \label{fig:cms-resources}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.9\textwidth}
-    	    \centering
-        \includegraphics[width=0.6\textwidth]{atlas-cpuHLLHC.pdf}
-        \caption{Estimated CPU resources (in kHS06) needed for the years 2018 to 2028 for both data and simulation processing.
-The blue points are estimates based on the current software performance estimates and using the ATLAS computing model parameters from 2017.
-The solid line shows the amount of resources expected to be available if a flat funding scenario is assumed, which implies an increase of 20\% per year, based on the current technology trends.}
-        \label{fig:atlas-cpu}
-    \end{subfigure}
-    \quad
-    \begin{subfigure}[b]{0.9\textwidth}
-    	    \centering
-        \includegraphics[width=0.6\textwidth]{atlas-diskHLLHC.pdf}
-        \caption{Estimated total disk resources (in PB) needed for the years 2018 to 2028 for both data and simulation processing.
-The blue points are estimates based on the current event sizes estimates and using the ATLAS computing model parameters from 2017.
-The solid line shows the amount of resources expected to be available if a flat funding scenario is assumed, which implies an increase of 15\% per year, based on the current technology trends.}
-        \label{fig:atlas-disk}
-    \end{subfigure}
-    \caption{ATLAS resources required into the HL-LHC era, using the current computing model 
-    and software performance.\cite{ATLAS-HL-LHC-Plots}}
-    \label{fig:atlas-resources}
-\end{figure}
-
-In general, it can be said that the amount of data that experiments can
-collect and process in the future will be limited by affordable software
-and computing, and therefore the physics reach during HL-LHC will be
-limited by how efficiently these resources can be used.
-
-The ATLAS numbers, in Figure~\ref{fig:atlas-resources}, are particularly interesting as they
-estimate the resources that will be available to the experiment if a
-flat funding profile is maintained, taking into account the expected
-technology improvements given current trends~\cite{Panzer2017}. As can be
-seen, the shortfall between needs and bare technology gains is
-considerable: a factor 4 in CPU and a factor 7 in disk in 2027.
-
-While the density of transistors on silicon continues to increase
-following Moore's Law (albeit more slowly than in the past), power
-density constraints have limited the clock speed of processors for more
-than a decade. This has effectively stalled any progress in the
-processing capacity of a single CPU core. Instead, increases in
-potential processing capacity come from increases in the core count of
-CPUs and wide CPU registers. Alternative processing architectures have
-become more commonplace. These range from the many-core architecture
-based on standard x86\_64 cores to numerous alternatives such as GPUs.
-For GPUs, the processing model is very different, allowing a much
-greater fraction of the die to be dedicated to arithmetic calculations,
-but at a price in programming difficulty and memory handling for the
-developer that tends to be specific to each processor generation.
-Further developments may even see the use of FPGAs for more
-general-purpose tasks. Fully exploiting these evolutions requires a
-shift in programming model to one based on \emph{concurrency}.
-
-Even with the throttling of clock speed to limit power consumption,
-power remains a major issue. Low power architectures are in huge demand.
-At one level this might challenge the dominance of x86\_64 by simply
-replacing it with, for example, AArch64 devices that may achieve lower power
-costs for the scale of HEP computing needs than Intel has achieved with its
-Xeon architecture. More extreme is an architecture that would see
-specialised processing units dedicated to particular tasks, but with
-possibly large parts of the device switched off most of the time,
-so-called dark silicon.
-
-Limitations in affordable storage also pose a major challenge, as does
-the I/O rates of higher capacity hard disks. Network
-bandwidth will probably continue to increase at the required level, but
-the ability to use it efficiently will need a closer integration with
-applications. This will require software developments to
-support distributed computing (data and workload management, software
-distribution and data access) and an increasing awareness of the
-extremely hierarchical view of data, from long latency tape access and
-medium-latency network access through to the CPU memory hierarchy.
-
-Taking advantage of these new architectures and programming paradigms
-will be critical for HEP to increase the ability of our code to deliver
-physics results efficiently, and to meet the processing challenges of
-the future. Some of this work will be focused on re-optimised
-implementations of existing algorithms. This will be complicated by the
-fact that much of our code is written for the much simpler model of
-serial processing, and without the software engineering needed for
-sustainability. Proper support for taking advantage of concurrent
-programming techniques, such as vectorisation and thread-based
-programming, through frameworks and libraries, will be essential, as the
-majority of the code will still be written by physicists. Other
-approaches should examine new algorithms and techniques, including
-highly parallelised code that can run on GPUs or the use of machine
-learning techniques to replace computationally expensive pieces of
-simulation or pattern recognition. The ensemble of computing work that
-is needed by the experiments must remain sufficiently flexible to take
-advantage of different architectures that will provide computing to HEP
-in the future. The use of high performance computing sites and
-commercial cloud providers will very likely be a requirement for the
-community and will bring particular constraints and demand flexibility.
-
-These technical challenges are accompanied by significant human
-challenges. Software is written by many people in the collaborations,
-with varying levels of expertise, from a few experts with precious
-skills to novice coders. This implies organising training in effective
-coding techniques and providing excellent documentation, examples and
-support. Although it is inevitable that some developments will remain
-within the scope of a single experiment, tackling software problems
-coherently as a community will be critical to achieving success in the
-future. This will range from sharing knowledge of techniques and best
-practice to establishing common libraries and projects that will provide
-generic solutions to the community. Writing code that supports a wider
-subset of the community than just a single experiment will almost
-certainly be mandated upon HEP and presents a greater challenge, but the
-potential benefits are huge. Attracting, and retaining, people with the
-required skills who can provide leadership is another significant
-challenge, since it impacts on the need to give adequate recognition to
-physicists who specialise in software development. This is an important
-issue that is treated in more detail later in the report.
-
-Particle physics is no longer alone in facing these massive data
-challenges. Experiments in other fields, from astronomy to genomics,
-will produce huge amounts of data in the future, and will need to
-overcome the same challenges that we face, i.e., massive data handling and
-efficient scientific programming. Establishing links with these fields
-has already started. Additionally, interest from the computing science
-community in solving these data challenges exists, and mutually
-beneficial relationships would be possible where there are genuine
-research problems that are of academic interest to that community and
-provide practical solutions to ours. The efficient processing of massive
-data volumes is also a challenge faced by industry, in particular the
-internet economy, which developed novel and major new technologies under
-the banner of \emph{Big Data} that may be applicable to our use cases.
-
-Establishing a programme of investment in software for the HEP
-community, with a view to ensuring effective and sustainable software
-for the coming decades, will be essential to allow us to reap the
-physics benefits of the multi-exabyte data to come. It was in recognition of
-this fact that the HSF itself was set up and already works to promote
-these common projects and community developments~\cite{HSF2015}.
-
-\hypertarget{programme-of-work}{%
-\section{Programme of Work}\label{programme-of-work}}
-
-In the following we describe the programme of work being proposed for
-the range of topics covered by the CWP working groups. We summarise the
-main specific challenges each topic will face, describe current
-practices, and propose a number of R\&D tasks that should be undertaken
-in order to meet the challenges. R\&D tasks are grouped in two different
-timescales: short term (by 2020, in time for the HL-LHC Computing Technical Design Reports of
-ATLAS and CMS) and longer-term actions (by 2022, to be ready for testing
-or deployment during LHC Run 3).
-
-\hypertarget{physics-generators}{%
-\subsection{Physics Generators}\label{physics-generators}}
-
-\subsubsection*{Scope and Challenges}
-
-Monte-Carlo event generators are a vital part of modern particle
-physics, providing a key component of the understanding and
-interpretation of experiment data. Collider experiments have a need for
-theoretical QCD predictions at very high precision. Already in LHC Run
-2, experimental uncertainties for many analyses are at the same level as,
-or lower than, those from theory. Many analyses have irreducible
-QCD-induced backgrounds, where statistical extrapolation into the signal
-region can only come from theory calculations. With future experiment
-and machine upgrades, as well as reanalysis of current data, measured
-uncertainties will shrink even further, and this will increase the need
-to reduce the corresponding errors from theory.
-
-Increasing accuracy will compel the use of higher-order perturbation
-theory generators with challenging computational demands. Generating
-Monte Carlo events using leading order (LO) generators is only a small
-part of the overall computing requirements for HEP experiments.
-Next-to-leading order (NLO) event generation, used more during LHC Run
-2, is already using significant resources. Higher accuracy
-theoretical cross sections calculated at next-to-next-to-leading (NNLO),
-already important in some Run 2 analyses, are not widely used
-because of computational cost. By HL-LHC the use of NNLO event
-generation will be more widely required, so these obstacles to their
-adoption must be overcome.
-Increasing the order of the generators increases greatly the complexity
-of the phase space integration required to calculate the appropriate QCD
-matrix elements. The difficulty of this integration arises from the need
-to have sufficient coverage in a high-dimensional space (10-15
-dimensions, with numerous local maxima); the appearance of negative event
-weights; and the fact that many terms in the integration cancel, so that
-a very high degree of accuracy of each term is required. Memory demands
-for generators have generally been low and initialisation times have
-been fast, but an increase in order means that memory consumption
-becomes important and initialisation times can become a significant 
-fraction of the job's run time.
-
-For HEP experiments, in many cases, meaningful predictions can only be
-obtained by combining higher-order perturbative calculations with parton
-showers. This procedure is also needed as high-multiplicity final states
-become more interesting at higher luminosities and event rates. Matching
-(N)NLO fixed-order calculations to parton shower algorithms can have a
-very low efficiency, and increases further the computational load needed
-to generate the necessary number of particle-level events. In addition,
-many of the current models for the combination of parton-level event
-generators and parton shower codes are incompatible with requirements
-for concurrency on modern architectures. It is a major challenge to
-ensure that this software can run efficiently on next generation
-hardware and software systems.
-
-Developments in generator software are mainly done by the HEP theory
-community. Theorists typically derive career recognition and advancement from
-making contributions to theory itself, rather than by making improvements
-to the computational efficiency of generators \emph{per se}. So,
-improving the computational efficiency of event generators, and allowing
-them to run effectively on resources such as high performance computing
-facilities (HPCs), will mean engaging with experts in computational
-optimisation who can work with the theorists who develop generators.
-
-The challenge in the next decade is to advance the theory and practical
-implementation of event generators to support the needs of future
-experiments, reaching a new level of theory precision and recognising
-the demands for computation and computational efficiency that this will
-bring.
-
-\subsubsection*{Current Practice}
-
-Extensive use of LO generators and parton shower algorithms are still
-made by most HEP experiments. Each experiment has its own simulation
-needs, but for the LHC experiments tens of billions of generated events
-are now used each year for Monte Carlo simulations. During LHC Run 2 more and
-more NLO generators were used, because of their increased theoretical
-precision and stability. The raw computational complexity of NLO
-amplitudes, combined with many-body phase-space evaluations and the
-inefficiencies of the matching process, leads to a potentially much-increased CPU
-budget for physics event simulation for ATLAS and CMS.
-
-The use of NLO generators by the experiments today is also limited because of
-the way the generators are implemented, producing significant numbers of
-negative event weights. This means that the total number of events the
-experiments need to generate, simulate, and reconstruct can be many 
-times larger for NLO than for LO samples. At the same time, the
-experiments budget only a similar number of Monte Carlo simulation events as from the real data. Having large NLO samples is thus not consistent with
-existing computing budgets until a different scheme is developed that
-does not depend on negative event weights or produces them only at a
-significantly reduced rate.
-
-While most event generation is run on ``standard'' grid resources,
-effort is ongoing to run more demanding tasks on HPC resources, e.g.,
-W-boson + 5-jet events at the Argonne Mira HPC). However, scaling for
-efficient running on some of the existing HPC resources is not trivial
-and requires effort.
-
-Standard HEP libraries such as LHAPDF~\cite{LHAPDF}, HepMC\cite{HepMC}, 
-and Rivet~\cite{Rivet} are used by the generators for integration into
-the experiments' event generation workflows. These require extensions
-and sustained maintenance that should be considered a shared
-responsibility of the theoretical and experimental communities in the
-context of large-scale experiments. In practice, however, it has been
-difficult to achieve the level of support that is really needed as there
-has been a lack of recognition for this work. To help improve the
-capabilities and performance of generators as used by the experimental
-HEP programme, and to foster interaction between the communities, the 
-MCnet~\cite{MCnet} short-term studentship programme has been very useful.
-Interested experimental PhD students can join a generator group for
-several months to work on improving a physics aspect of the simulation
-that is relevant to their work, or to improve the integration of the
-generator into an experimental framework.
-
-\subsubsection*{Research and Development Programme}
-
-As the Monte Carlo projects are funded mainly to develop theoretical
-improvements, and not mainly as ``suppliers'' to the experimental HEP
-programme, any strong requests towards efficiency improvements from the
-experimental community would need to be backed up by plausible avenues
-of support that can fund contributions from software engineers with the
-correct technical skills in software optimisation to work within the
-generator author teams.
-
-In a similar way to the MCnet studentships, a matchmaking scheme could
-focus on the software engineering side, and transfer some of the
-expertise available in the experiments and facilities teams to the
-generator projects. Sustainable improvements are unlikely to be
-delivered by graduate students ``learning on the job'' and then leaving
-after a few months, so meeting the requirement of transferring technical
-expertise and effort will likely require placements for experienced
-optimisation specialists and a medium- to long-term connection to the
-generator project.
-
-HEP experiments, which are now managed by very large collaborations including many
-technical experts, can also play a key role in sustaining a healthy
-relationship between theory and experiment software. Effort to work on
-common tools that benefit both the experiment itself and the wider
-community would provide shared value that justifies direct investment
-from the stakeholders. This model would also be
-beneficial for core HEP tools like LHAPDF, HepMC and Rivet, where future
-improvements have no theoretical physics interest anymore, putting them
-in a similar situation to generator performance improvements. One
-structural issue blocking such a mode of operation is that some
-experiments do not currently recognise contributions to external
-projects as experiment service work --- a situation deserving of review
-in areas where external software tools are critical to experiment
-success.
-
-In the following we describe specific areas of R\&D for event generation 
-up to 2022 and beyond.
-
-\begin{itemize}
-\item
-  The development of new and improved theoretical algorithms provides
-  the largest potential for improving event generators. While it is not
-  guaranteed that simply increasing the effort dedicated to this task
-  will bring about the desired result, the long-term support of event
-  generator development, and the creation of career opportunities in
-  this research area, are critical given the commitment to experiments
-  on multi-decade scales.
-\item
-  Expand development in reweighting event samples, where new physics
-  signatures can be explored by updating the partonic weights according
-  to new matrix elements. It is necessary that the phase space for the
-  updated model be a subset of the original one, which is an important
-  limitation. The procedure is more complex at NLO and can require
-  additional information to be stored in the event files to properly
-  reweight in different cases. Overcoming the technical issues from
-  utilising negative event weights is crucial. Nevertheless, the method
-  can be powerful in many cases, and would hugely reduce the time needed
-  for the generation of BSM samples.
-\item
-  At a more technical level, concurrency is an avenue that has yet to be
-  explored in depth for event generation. As the calculation of matrix
-  elements requires VEGAS-style integration, this work would be helped
-  by the development of a new Monte-Carlo integrator. For multi-particle
-  interactions, factorising the full phase space integration into lower
-  dimensional integrals would be a powerful method of parallelising,
-  while the interference between different Feynman graphs can be handled
-  with known techniques.
-\item
-  For many widely used generators, basic problems of concurrency and
-  thread hostility need to be tackled, to make these packages suitable
-  for efficient large scale use on modern processors and within modern
-  HEP software frameworks. Providing appropriate common tools for
-  interfacing, benchmarking and optimising multithreaded code would
-  allow expertise to be shared effectively~\cite{HSF-CWP-2017-13}.
-\item
-  In most generators, parallelism was added post-facto, which leads to
-  scaling problems when the level of parallelism becomes very large,
-  e.g., on HPC machines. These HPC machines will be part of the
-  computing resource pool used by HEP, so solving scaling issues on
-  these resources for event generation is important, particularly as the
-  smaller generator code bases can make porting to non-x86\_64
-  architectures more tractable. The problem of long and inefficient
-  initialisation when a job utilises hundreds or thousands of cores on
-  an HPC needs to be tackled. While the memory consumption of event
-  generators is generally modest, the generation of tree-level
-  contributions to high multiplicity final states can use significant
-  memory, and gains would be expected from optimising here.
-\item
-  An underexplored avenue is the efficiency of event generation as
-  used by the experiments. An increasingly common usage is to generate
-  very large inclusive event samples, which are filtered on event
-  final-state criteria to decide which events are to be retained and
-  passed onto detector simulation and reconstruction. This naturally
-  introduces a large waste of very CPU-expensive event generation, which
-  could be reduced by developing filtering tools within the generators
-  themselves, designed for compatibility with the experiments'
-  requirements. A particularly wasteful example is where events are
-  separated into orthogonal subsamples by filtering, in which case the
-  same large inclusive sample is generated many times, with each stream
-  filtering the events into a different group: allowing a single
-  inclusive event generation to be filtered into several orthogonal
-  output streams would improve efficiency.
-\end{itemize}
-
-\hypertarget{detector-simulation}{%
-\subsection{Detector Simulation}\label{detector-simulation}}
-
-\subsubsection*{Scope and Challenges}
-
-For all its success so far, the challenges faced by the HEP field in the
-simulation domain are daunting. During the first two runs, the LHC
-experiments produced, reconstructed, stored, transferred, and analysed
-tens of billions of simulated events. This effort required more than
-half of the total computing resources allocated to the experiments. As
-part of the HL-LHC physics programme, the upgraded
-experiments expect to collect 150 times more data than in Run 1; demand
-for larger simulation samples to satisfy analysis needs will grow
-accordingly. In addition, simulation tools have to serve diverse
-communities, including accelerator-based particle physics research
-utilising proton-proton colliders, neutrino, dark matter, and muon
-experiments, as well as the cosmic frontier. The complex detectors of
-the future, with different module- or cell-level shapes, finer
-segmentation, and novel materials and detection techniques, require
-additional features in geometry tools and bring new demands on physics
-coverage and accuracy within the constraints of the available computing
-budget. The diversification of the physics programmes also requires new
-and improved physics models. More extensive use of Fast Simulation is a
-potential solution, under the assumption that it is possible to improve
-time performance without an unacceptable loss of physics accuracy.
-
-The gains that can be made by speeding up critical elements of the
-Geant4 simulation toolkit can be leveraged for all applications that use
-it, and it is therefore well worth the investment in effort needed to
-achieve it. The main challenges to be addressed if the required physics
-and software performance goals are to be achieved are:
-
-\begin{itemize}
-\item
-  Reviewing the physics models' assumptions, approximations, and
-  limitations in order to achieve higher precision, and to extend the
-  validity of models up to energies of the order of 100 TeV foreseen
-  with the Future Circular Collider (FCC) project~\cite{FCC}.
-\item
-  Redesigning, developing, and commissioning detector simulation
-  toolkits to be more efficient when executed on current vector CPUs and
-  emerging new architectures, including GPUs, where use of SIMD
-  vectorisation is vital; this includes porting and optimising the
-  experiments' simulation applications to allow exploitation of large
-  HPC facilities.
-\item
-  Exploring different Fast Simulation options, where the full detector
-  simulation is replaced, in whole or in part, by computationally
-  efficient techniques. An area of investigation is common frameworks
-  for fast tuning and validation.
-\item
-  Developing, improving and optimising geometry tools that can be shared
-  am\-ong experiments to make the modeling of complex detectors
-  computationally more efficient, modular, and transparent.
-\item
-  Developing techniques for background modeling, including contributions
-  of multiple hard interactions overlapping the event of interest in
-  collider experiments (pileup).
-\item
-  Revisiting digitisation algorithms to improve performance and
-  exploring opportunities for code sharing among experiments.
-\item
-  Recruiting, training, retaining human resources in all areas of
-  expertise pertaining to the simulation domain, including software and
-  physics.
-\end{itemize}
-
-It is obviously of critical importance that the whole community of
-scientists working in the simulation domain continue to work together in
-as efficient a way as possible in order to deliver the required
-improvements. Very specific expertise is required across all simulation
-domains, such as physics modeling, tracking through complex geometries
-and magnetic fields, and building realistic applications that accurately
-simulate highly complex detectors. Continuous support is needed to
-recruit, train, and retain people with a unique set of skills needed to
-guarantee the development, maintenance, and support of simulation codes
-over the long timeframes foreseen in the HEP experimental programme.
-
-\subsubsection*{Current Practices}
-
-The Geant4 detector simulation toolkit is at the core of simulation in
-almost every HEP experiment. Its continuous development, maintenance,
-and support for the experiments is of vital importance. New or refined
-functionality in physics coverage and accuracy continues to be delivered
-in the ongoing development programme and software
-performance improvements are introduced whenever possible.
-
-Physics models are a critical part of the detector simulation, and are
-continuously being reviewed, and in some cases reimplemented, in order
-to improve accuracy and software performance. Electromagnetic (EM)
-transport simulation is challenging as it occupies a large part of the
-computing resources used in full detector simulation. Significant
-efforts have been made in the recent past to better describe the
-simulation of electromagnetic shower shapes, in particular to model the
-$H \to \gamma \gamma$ signal and background accurately at the LHC. This
-effort is being continued with an emphasis on reviewing the models'
-assumptions, approximations, and limitations, especially at very high
-energy, with a view to improving their respective software
-implementations. In addition, a new ``theory-based'' model
-(Goudsmit-Saunderson), for describing the \emph{multiple scattering} of
-electrons and positrons, has been developed that has been demonstrated
-to outperform, in terms of physics accuracy and speed, the current
-models in Geant4. The models used to describe the \emph{bremsstrahlung}
-process have also been reviewed, and recently an improved theoretical
-description of the Landau-Pomeranchuk-Migdal effect was introduced
-that plays a significant role at high energies. Theoretical review of
-all electromagnetic models, including those of hadrons and ions, is
-therefore of high priority both for HL-LHC and for FCC studies.
-
-Hadronic physics simulation covers purely hadronic interactions. It is
-not possible for a single model to describe all the physics encountered
-in a simulation due to the large energy range that needs to be covered
-and the simplified approximations that are used to overcome the
-difficulty of solving the full theory (QCD). Currently the most-used
-reference physics list for high energy and space applications is
-FTFP\_BERT. It uses the Geant4 Bertini cascade for hadron--nucleus
-interactions from 0 to 12 GeV incident hadron energy and the FTF parton
-string model for hadron--nucleus interactions from 3 GeV upwards.
-QGSP\_BERT is a popular alternative which replaces the FTF model with
-the QGS model over the high energy range. The existence of more than one
-model (for each energy range) is very valuable in order to be able to
-determine the systematics effects related to the approximations used.
-The use of highly granular calorimeters, such as the ones being designed
-by the CALICE collaboration for future linear colliders, allows a
-detailed validation of the development of hadronic showers with
-test-beam data. Preliminary results suggest that the lateral profiles of
-Geant4 hadronic showers are too narrow. Comparisons with LHC test-beam
-data have shown that a fundamental ingredient for improving the
-description of the lateral development of showers is the use of
-intermediate and low energy models that can describe the cascading of
-hadrons in nuclear matter. Additional work is currently being invested
-in the further improvement of the QGS model, which is a more
-theory-based approach than the phenomenological FTF model, and therefore
-offers better confidence at high energies, up to a few TeV. This again
-is a large endeavour and requires continuous effort over a long time.
-
-The Geant4 collaboration is working closely with user communities to
-enrich the physics models' validation system with data acquired during
-physics runs and test beam campaigns. In producing new models of physics
-interactions and improving the fidelity of the models that exist, it is
-absolutely imperative that high-quality data are available. Simulation
-model tuning often relies on test beam data, and a program to improve
-the library of available data could be invaluable to the community. Such
-data would ideally include both thin-target test beams for improving
-interaction models and calorimeter targets for improving shower models.
-This data could potentially be used for directly tuning Fast Simulation
-models as well.
-
-There are specific challenges associated with the Intensity Frontier
-experimental programme, in particular simulation of the beamline and the
-neutrino flux. Neutrino experiments rely heavily on detector simulations
-to reconstruct neutrino energy, which requires accurate modelling of
-energy deposition by a variety of particles across a range of energies.
-Muon experiments such as Muon g-2 and Mu2e also face large simulation
-challenges; since they are searching for extremely rare effects, they
-must grapple with very low signal to background ratios and the modeling
-of low cross-section background processes. Additionally, the size of the
-computational problem is a serious challenge, as large simulation runs
-are required to adequately sample all relevant areas of experimental
-phase space, even when techniques to minimise the required computations
-are used. There is also a need to simulate the effects of low energy
-neutrons, which requires large computational resources. Geant4 is the
-primary simulation toolkit for all of these experiments.
-
-Simulation toolkits do not include effects like charge drift in an
-electric field or models of the readout electronics of the experiments.
-Instead, these effects are normally taken into account in a separate
-step called digitisation. Digitisation is inherently local to a given
-sub-detector and often even to a given readout element, so that there
-are many opportunities for parallelism in terms of vectorisation and
-multiprocessing or multithreading, if the code and the data objects are
-designed optimally. Recently, both hardware and software projects have
-benefitted from an increased level of sharing among experiments. The
-LArSoft Collaboration develops and supports a shared base of physics
-software across Liquid Argon (LAr) Time Projection Chamber (TPC)
-experiments, which includes providing common digitisation code.
-Similarly, an effort exists among the LHC experiments to share code for
-modeling radiation damage effects in silicon. As ATLAS and CMS expect
-to use similar readout chips in their future trackers, further code
-sharing might be possible.
-
-The Geant4 simulation toolkit will also evolve over the next decade to
-include contributions from various R\&D projects, as described in the
-following section. This is required to ensure the support of experiments
-through continuous maintenance and improvement of the Geant4 simulation
-toolkit. This is necessary until production versions of potentially
-alternative engines, such as those resulting from ongoing R\&D work,
-become available, integrated, and validated by experiments. The agreed
-ongoing strategy to make this adoption possible is to ensure that new
-developments resulting from the R\&D programme can be tested with
-realistic prototypes and then be integrated, validated, and deployed in
-a timely fashion in Geant4.
-
-\subsubsection*{Research and Development Programme}
-
-To meet the challenge of improving the performance by a large factor, an
-ambitious R\&D programme is underway to investigate each component of
-the simulation software for the long term. In the following we describe
-in detail some of the studies to be performed in the next 3-5 years.
-
-\begin{itemize}
-\item
-  Particle Transport and Vectorisation: the study of an efficient
-  transport of particles (tracks) in groups so as to maximise the
-  benefit of using SIMD operations.
-\item
-  Modularisation: improvement of Geant4 design to allow for a tighter
-  and easier integration of single sub-packages of the code into
-  experimental frameworks.
-\item
-  Physics Models: extensions and refinements of the physics algorithms
-  to provide new and more performant physics capabilities.
-\item
-  Other activities: integration of multi-threading capabilities in
-  experiment applications; experiment-agnostic software products to cope
-  with increased pile\-up, fast simulation, digitisation, and efficient
-  production of high-quality ran\-dom numbers.
-\end{itemize}
-
-\paragraph{Particle Transport and Vectorisation} One of the most
-ambitious elements of the simulation R\&D programme is a new approach to
-managing particle transport, which has been introduced by the GeantV
-project. The aim is to deliver a multithreaded vectorised transport
-engine that has the potential to deliver large performance benefits. Its
-main feature is track-level parallelisation, bundling particles with
-similar properties from different events to process them in a single
-thread. This approach, combined with SIMD vectorisation coding
-techniques and improved data locality, is expected to yield significant
-speed-ups, which are to be measured in a realistic prototype currently
-under development. For the GeantV transport engine to display its best
-computing performance, it is necessary to vectorise and optimise the
-accompanying modules, including geometry, navigation, and the physics
-models. These are developed as independent libraries so that they can
-also be used together with the current Geant4 transport engine. Of
-course, when used with the current Geant4 they will not expose their
-full performance potential, since transport in Geant4 is currently
-sequential, but this allows for a preliminary validation and comparison
-with the existing implementations. The benefit of this approach is that
-new developments can be delivered as soon as they are available. The new
-vectorised geometry package (VecGeom), developed as part of GeantV R\&D
-and successfully integrated into Geant4, is an example that demonstrated
-the benefit of this approach. By the end of 2018 it is intended to have
-a proof-of-concept for the new particle transport engine that includes
-vectorised EM physics, vectorised magnetic field propagation and that
-uses the new vectorised geometry package. This will form a sound basis
-for making performance comparisons for simulating EM showers in a
-realistic detector.
-
-\begin{itemize}
-\item
-  2019: the \emph{beta} release of the GeantV transport engine will
-  contain enough functionality to build the first real applications.
-  This will allow performance to be measured and give sufficient time to
-  prepare for HL-LHC running. It should include the use of vectorisation
-  in most of the components, including physics modelling for electrons,
-  gammas and positrons, whilst still maintaining simulation
-  reproducibility, and I/O in a concurrent environment and multi-event
-  user data management.
-\end{itemize}
-
-\paragraph{Modularisation} Starting from the next release, a
-modularisation of Geant4 is being pursued that will allow an easier
-integration in experimental frameworks, with the possibility to include
-only the Geant4 modules that are actually used. A further use case is
-the possibility to use one of the Geant4 components in isolation, e.g.,
-to use hadronic interaction modeling without kernel components from a
-fast simulation framework. As a first step a preliminary review of
-libraries' granularity is being pursued, which will be followed by a
-review of intra-library dependencies with the final goal of reducing
-their dependencies.
-
-\begin{itemize}
-\item
-  2019: Redesign of some Geant4 kernel components to improve the
-  efficiency of the simulation on HPC systems, starting from improved
-  handling of Geant4 \emph{database}s on large core-count systems. A
-  review will be made of the multithreading design to be closer to
-  task-based frameworks, such as Intel's Thread\-ed Building Blocks
-  (TBB)~\cite{TBB}.
-\end{itemize}
-
-\paragraph{Physics Models} It is intended to develop
-new and extended physics models to
-cover extended energy and physics processing of present and future
-colliders, Intensity Frontier experiments, and direct dark matter search
-experiments. The goal is to extend the missing models (e.g., neutrino
-interactions), improve models' physics accuracy and, at the same time,
-improve CPU and memory efficiency. The deliverables of these R\&D
-efforts include physics modules that produce equivalent quality physics,
-and will therefore require extensive validation in realistic
-applications.
-
-\begin{itemize}
-\item
-  2020: Improved implementation of hadronic cascade models for LHC and,
-  in particular, Liquid Argon detectors. Improved accuracy models of
-  EM interactions of photons and electrons. To address the needs of
-  cosmic frontier experiments, optical photon transport must be improved
-  and made faster.
-\item
-  2022: Implementation of EPOS string model for multi-GeV to multi-TeV
-  interactions, for FCC detector simulation and systematic studies of
-  HL-LHC detectors.
-\end{itemize}
-
-\paragraph{Experiment Applications} The experiment applications are
-essential for validating the software and physics performance of new
-versions of the simulation toolkit. ATLAS and CMS have already started
-to integrate Geant4 multithreading capability in their simulation
-applications; in the case of CMS the first Full Simulation production in
-multithreaded mode was delivered in the autumn of 2017. Specific
-milestones are as follows:
-
-\begin{itemize}
-\item
-  2020: LHC, Neutrino, Dark Matter, and Muon experiments to demonstrate
-  the ability to run their detector simulation in multithreaded mode,
-  using the improved navigation and electromagnetic physics packages.
-  This should bring experiments more accurate physics and improved
-  performance.
-\item
-  2020: Early integration of the beta release of the GeantV transport
-  engine in the experiments' simulation, including the implementation of
-  the new user interfaces, which will allow the first performance
-  measurements and physics validation to be made.
-\item
-  2022: The availability of a production version of the new track-level
-  parallelisation and fully vectorised geometry, navigation, and physics
-  libraries will offer the experiments the option to finalise
-  integration into their frameworks; intensive work will be needed in
-  physics validation and computing performance tests. If successful, the
-  new engine could be in production on the timescale of the start of the
-  HL-LHC run in 2026.
-\end{itemize}
-
-\paragraph{Pileup} Backgrounds to hard-scatter events have many
-components including in-time pileup, out-of-time pileup, cavern
-background and beam-gas collisions. All of these components can be
-simulated, but they present storage and I/O challenges related to the
-handling of the large simulated minimum bias samples used to model the
-extra interactions. An R\&D programme is needed to study different
-approaches to managing these backgrounds within the next 3 years:
-
-\begin{itemize}
-\item
-  Real zero-bias events can be collected, bypassing any zero
-  suppression, and overlaid on the fully simulated hard scatters. This
-  approach faces challenges related to the collection of
-  non-zero-suppressed samples or the use of suppressed events,
-  non-linear effects when adding electronic signals from different
-  samples, and sub-detector misalignment consistency between the
-  simulation and the real experiment. Collecting calibration and
-  alignment data at the start of a new Run would necessarily incur
-  delays such that this approach is mainly of use in the final analyses.
-  The experiments are expected to invest in the development of the
-  zero-bias overlay approach by 2020.
-\item
-  The baseline option is to ``pre-mix'' together the minimum bias
-  collisions into individual events that have the full background
-  expected for a single collision of interest. Experiments will invest
-  effort on improving their pre-mixing techniques, which allow the
-  mixing to be performed at the digitisation level, reducing the disk and
-  network usage for a single event.
-\end{itemize}
-
-\paragraph{Fast Simulation} The work on Fast Simulation is also
-accelerating with the objective of producing a flexible framework that
-permits Full and Fast simulation to be combined for different particles
-in the same event. Various approaches to Fast Simulation are being tried
-all with the same goal of saving computing time, under the assumption
-that it is possible to improve time performance without an unacceptable
-loss of physics accuracy. There has recently been a great deal of
-interest in the use of Machine Learning in Fast Simulation, most of
-which has focused on the use of multi-objective regression and
-generative adversarial networks (GANs). Since use of GANs allows for
-non-parametric learning in cases such as calorimetric shower
-fluctuations, it is a promising avenue for generating non-Gaussian and
-highly correlated physical effects. This is an obvious area for future
-expansion and development, as it is currently in its infancy.
-
-\begin{itemize}
-\item
-  2018: Assessment of the benefit of machine learning approach for Fast
-  Simulation.
-\item
-  2019: ML-based Fast Simulation for some physics observables.
-\item
-  2022: Demonstrate the potential of a common Fast Simulation
-  infrastructure applicable to the variety of detector configurations.
-\end{itemize}
-
-\paragraph{Digitisation} It is expected that, within the next 3
-years, common digitisation efforts are well-established among
-experiments, and advanced high-performance gener\-ic digitisation
-examples, which experiments could use as a basis to develop their own
-code, become available. For example, the development of next generation
-silicon detectors requires realistic simulation of the charge collection
-and digitisation processes. Owing to the large variety of technologies,
-common software frameworks need to be flexible and modular to cater for
-the different needs.
-
-\begin{itemize}
-\item
-  2020: Deliver advanced high-performance, SIMD-friendly generic
-  digitisation examples that experiments can use as a basis to develop
-  their own code.
-\item
-  2022: Fully tested and validated optimised digitisation code that can
-  be used by the HL-LHC and DUNE experiments.
-\end{itemize}
-
-\paragraph{Pseudorandom Number Generation} The selection of
-pseudorandom number generators (PRNGs) presents challenges when running
-on infrastructures with a large degree of parallelism, as
-reproducibility is a key requirement. HEP will collaborate with
-researchers in the development of PRNGs, seeking to obtain generators
-that address better our challenging requirements. Specific milestones
-are:
-
-\begin{itemize}
-\item
-  2020: Develop a single library containing sequential and vectorised
-  implementations of the set of state-of-the-art PRNGs, to replace the
-  existing ROOT and CLHEP implementations. Potential use of C++11 PRNG
-  interfaces and implementations, and their extension for our further
-  requirements (output of multiple values, vectorisation) will be
-  investigated.
-\item
-  2022: Promote a transition to the use of this library to replace
-  existing implementations in ROOT and Geant4.
-\end{itemize}
-
-\hypertarget{software-trigger-and-event-reconstruction}{%
-\subsection{Software Trigger and Event
-Reconstruction}\label{software-trigger-and-event-reconstruction}}
-
-\subsubsection*{Scope and Challenges}
-
-The reconstruction of raw detector data and simulated data, and its
-processing in real time, represent a major component of today's
-computing requirements in HEP. Advances in the capabilities of 
-facilities and future
-experiments bring the potential for a dramatic increase in physics
-reach, at the price of increased event complexities and rates. It is
-therefore essential that event reconstruction algorithms and software
-triggers continue to evolve so that they are able to efficiently exploit
-future computing architectures, and deal with the increase in data rates
-without loss of physics. Projections into future, e.g., at HL-LHC conditions,
-show that without significant changes in approach or algorithms
-the increase in resources needed would be incompatible with the 
-the expected budget.
-
-At the HL-LHC, the central challenge for object reconstruction is to
-maintain excellent efficiency and resolution in the face of high pileup
-values, especially at low transverse momentum ($p_T$).
-Detector upgrades, such as increases in channel density, high-precision
-timing, and improved detector geometric layouts, are essential to
-overcome these problems. In many cases these new technologies bring
-novel requirements to software trigger and/or event reconstruction
-algorithms, or require new algorithms to be developed. Ones of
-particular importance at the HL-LHC include high-granularity
-calorimetry, precision timing detectors, and hardware triggers based on
-tracking information, which may seed later software trigger and
-reconstruction algorithms.
-
-At the same time, trigger systems for next-generation experiments are
-evolving to be more capable, both in their ability to select a wider
-range of events of interest for the physics programme, and their ability
-to stream a larger rate of events for further processing. ATLAS and CMS
-both target systems where the output of the hardware trigger system is
-increased by an order of magnitude over the current capability, up to 1
-MHz~\cite{CERN-LHCC-2015-020,CMSCollaboration:2015zni}. In LHCb~\cite{CERN-LHCC-2014-016}
-and ALICE~\cite{Buncic:2011297}, the full collision
-rate (between 30 to 40 MHz for typical LHC proton-proton operations)
-will be streamed to real-time or quasi-real-time software trigger
-systems. The increase in event complexity also brings a ``problem'' of
-an overabundance of signals to the experiments, and specifically to the
-software trigger algorithms. The evolution towards a genuine real-time
-analysis of data has been driven by the need to analyse more signal than
-can be written out for traditional processing, and technological
-developments that enable this without reducing the analysis sensitivity
-or introducing biases.
-
-Evolutions in computing technologies are an opportunity to move
-beyond commodity x86\_64 technologies, which HEP has used very
-effectively over the past 20 years, but also represent a significant challenge 
-if we are to
-derive sufficient event processing throughput per cost to reasonably
-enable our physics programmes~\cite{Bird:1695401}.
-Among these challenges, important items identified include the increase
-of SIMD capabilities, the evolution towards multi- or many-core
-architectures, the slow increase in memory bandwidth relative to CPU
-capabilities, the rise of heterogeneous hardware, and the possible
-evolution in facilities available to HEP production systems.
-
-The move towards open source software development and continuous
-integration systems brings opportunities to assist developers of
-software trigger and event reconstruction algorithms. Continuous
-integration systems based on standard open-source tools have already
-allowed automated code quality and performance checks, both for
-algorithm developers and code integration teams. Scaling these up to
-allow for sufficiently high-statistics checks is still an
-outstanding challenge. Also, code quality demands increase as
-traditional offline analysis components migrate into trigger systems,
-where algorithms can only be run once, and any problem means losing data
-permanently.
-
-\subsubsection*{Current Practices}
-
-Substantial computing facilities are in use for both online and offline
-event processing across all experiments surveyed. In most experiments,
-online facilities are dedicated to the operation of the software
-trigger, but a recent trend has been to use them opportunistically
-for offline processing too, when the software trigger does not make them
-100\% busy. On the other hand, offline facilities are shared with
-event reconstruction, simulation, and analysis. CPU in use by
-experiments is typically measured at the scale of tens or hundreds of
-thousands of x86\_64 processing cores.
-
-The CPU needed for event reconstruction tends to be dominated by charged
-particle reconstruction (tracking), especially when the number of
-collisions per bunch crossing is high and an efficient
-reconstruction low p\textsubscript{T} particles is required.
-Calorimetric reconstruction, particle flow reconstruction, and particle
-identification algorithms also make up significant parts of the CPU
-budget in some experiments. Disk storage is typically 10s to 100s of PBs
-per experiment. It is dominantly used to make the output of the event
-reconstruction, both for real data and simulation, available for
-analysis.
-
-Current experiments have moved towards smaller, but still
-flexible, tiered data formats. These tiers are typically based on the
-ROOT~\cite{Brun1996} file format and constructed to
-facilitate both skimming of interesting events and the selection of
-interesting pieces of events by individual analysis groups or through
-centralised analysis processing systems. Initial implementations of
-real-time analysis systems are in use within several experiments. These
-approaches remove the detector data that typically makes up the raw data
-tier kept for offline reconstruction, and keep only final analysis
-objects~\cite{Aaij2016,ATL-DAQ-PUB-2017-003,Khachatryan:2149625}.
-
-Systems critical for reconstruction, calibration, and alignment
-generally implement a high level of automation in all experiments.
-They are an integral part of the data taking and data reconstruction processing chain, both in the online systems as well as the offline processing setup.
-
-\subsubsection*{Research and Development Programme}
-
-Seven key areas, itemised below, have been identified where research and
-development is necessary to enable the community to exploit the full
-power of the enormous datasets that we will be collecting. Three of
-these areas concern the increasingly parallel and heterogeneous
-computing architectures that we will have to write our code for. In
-addition to a general effort to vectorise our codebases, we must
-understand what kinds of algorithms are best suited to what kinds of
-hardware architectures, develop benchmarks that allow us to compare the
-physics-per-dollar-per-watt performance of different algorithms across a
-range of potential architectures, and find ways to optimally utilise
-heterogeneous processing centres. The consequent increase in the
-complexity and diversity of our codebase will necessitate both a
-determined push to educate physicists in modern coding
-practices, and a development of more sophisticated and automated quality
-assurance and control. The increasing granularity of
-our detectors, and the addition of timing information, which seems
-mandatory to cope with the extreme pileup conditions at the HL-LHC, will
-require new kinds of reconstruction algorithms that are sufficiently
-fast for use in real-time. Finally, the increased
-signal rates will mandate a push towards real-time analysis in many
-areas of HEP, in particular those with low-p\textbf{\textsubscript{T}}
-signatures.
-
-\begin{itemize}
-\item
-  HEP developed toolkits and algorithms typically make poor use of
-  vector units on commodity computing systems. Improving this will bring
-  speedups to applications running on both current computing systems and
-  most future architectures. The goal for work in this area is to evolve
-  current toolkit and algorithm implementations, and best programming
-  techniques, to better use SIMD capabilities of current and future
-  CPU architectures.
-\item
-  Computing platforms are generally evolving towards having more cores
-  in order to increase processing capability. This evolution has
-  resulted in multithreaded frameworks in use, or in development, across
-  HEP. Algorithm developers can improve throughput by being thread-safe
-  and enabling the use of fine-grained parallelism. The goal is to
-  evolve current event models, toolkits and algorithm implementations,
-  and best programming techniques, to improve the throughput of
-  multithreaded software trigger and event reconstruction applications.
-\item
-  Computing architectures using technologies beyond CPUs offer an
-  interesting alternative for increasing throughput of the most
-  time-consuming trigger or reconstruction algorithms. Examples
-  such as GPUs and FPGAs could be integrated into dedicated
-  trigger or specialised reconstruction processing facilities,
-  in particular
-  online computing farms. The goal is to demonstrate how the throughput
-  of toolkits or algorithms can be improved
-  in a production environment and to understand
-  how much these new architectures require rethinking the algorithms
-  used today. In addition, it is necessary to assess and minimise
-  possible additional costs coming from the maintenance of multiple
-  implementations of the same algorithm on different architectures.
-\item
-  HEP experiments have extensive continuous integration systems,
-  including varying code regression checks that have enhanced the
-  quality assurance (QA) and quality control (QC) procedures for
-  software development in recent years. These are typically maintained
-  by individual experiments and have not yet reached the point where
-  statistical regression, technical, and physics performance checks can
-  be performed for each proposed software change. The goal is to enable
-  the development, automation, and deployment of extended QA and QC
-  tools and facilities for software trigger and event reconstruction
-  algorithms.
-\item
-  Real-time analysis techniques are being adopted to enable a wider
-  range of physics signals to be saved by the trigger for final
-  analysis. As rates increase, these techniques can become more
-  important and widespread by enabling only the parts of an event
-  associated with the signal candidates to be saved, reducing the
-  disk space requirement. The goal is to evaluate and demonstrate the tools
-  needed to facilitate real-time analysis techniques. Research topics
-  include the study of compression and custom data formats, toolkits for real-time
-  detector calibration and validation that enable full offline analysis
-  chains to be ported into real-time, and frameworks that allow
-  non-expert offline analysts to design and deploy real-time analyses
-  without compromising data taking quality.
-\item
-  The central challenge for object reconstruction at the HL-LHC is to
-  maintain excellent efficiency and resolution in the face of high
-  pileup, especially at low object p\textsubscript{T}.
-  Trigger systems and reconstruction software need to exploit new techniques
-  and higher granularity detectors to maintain or even improve physics
-  measurements in the future. It is also becoming increasingly clear
-  that reconstruction in very high pileup environments, such as the
-  HL-LHC or FCC-hh, will not be possible without adding some timing
-  information to our detectors, in order to exploit the finite time
-  during which the beams cross and the interactions are produced. The
-  goal is to develop and demonstrate efficient techniques for physics
-  object reconstruction and identification in complex environments.
-\item
-  Future experimental facilities will bring a large increase in event
-  complexity. The performance scaling of current-generation algorithms with this
-  complexity must be improved to avoid a large increase in resource
-  needs. In addition, it may become necessary to deploy
-  new algorithms in order to solve these problems, including advanced
-  machine learning techniques. The goal is to evolve or rewrite existing
-  toolkits and algorithms focused on their physics and technical
-  performance at high event complexity, e.g., high pileup at HL-LHC.
-  Most important targets are those which limit expected throughput
-  performance at future facilities, e.g., charged-particle tracking. A
-  number of such efforts are already in progress.
-\end{itemize}
-
-\hypertarget{data-analysis-and-interpretation}{%
-\subsection{Data Analysis and
-Interpretation}\label{data-analysis-and-interpretation}}
-
-\subsubsection*{Scope and Challenges}
-
-Scientific questions are answered by analysing the data obtained from
-suitably designed experiments and comparing measurements with
-predictions from models and theories. Such comparisons are typically
-performed long after data taking, but can sometimes also be executed in
-quasi-real time on selected samples of reduced size.
-
-The final stages of analysis are undertaken by small groups or even
-individual researchers. The baseline analysis model utilises successive
-stages of data reduction, finally reaching a compact dataset for quick
-real-time iterations. This approach aims at exploiting the maximum
-possible scientific potential of the data, whilst minimising the ``time
-to insight'' for a large number of different analyses performed in
-parallel. It is a complicated combination of diverse criteria, ranging
-from the need to make efficient use of computing resources to the
-management styles of the experiment collaborations. Any analysis system
-has to be flexible enough to cope with deadlines imposed by conference
-schedules. Future analysis models must adapt to the massive increases in
-data taken by the experiments, while retaining this essential ``time to
-insight'' optimisation.
-
-Over the past 20 years the HEP community has developed and gravitated
-around a single analysis ecosystem based on ROOT~\cite{Brun1996}. 
-ROOT is a general-purpose object
-oriented framework that addresses the selection, integration,
-development, and support of a number of foundation and utility class
-libraries that can be used as a basis for developing HEP application
-codes. The added value to the HEP community is that it provides an
-integrated and validated toolkit, and its use encompasses the full event
-processing chain; it has a major impact on the way HEP analysis is
-performed. This lowers the hurdle to start an analysis, enabling the
-community to communicate using a common analysis language, as well as
-making common improvements as additions to the toolkit quickly become
-available. The ongoing ROOT programme of work addresses important new
-requirements, in both functionality and performance, and this is given
-a high priority by the HEP community.
-
-An important new development in the analysis domain has been the
-emergence of new analysis tools coming from industry and open source
-projects, and this presents new opportunities for improving the HEP
-analysis software ecosystem. The HEP community is very interested in
-using these software tools, together with established components, in an
-interchangeable way. The main challenge will be to enable new
-open-source tools to be plugged in dynamically to the existing ecosystem
-and to provide mechanisms that allow the existing and new components to
-interact and exchange data efficiently. To improve our ability to
-analyse much larger datasets, R\&D will be needed to investigate file
-formats, compression algorithms, and new ways of storing and accessing
-data for analysis and to adapt workflows to run on future computing
-infrastructures.
-
-Reproducibility is the cornerstone of scientific results. It is
-currently difficult to repeat most HEP analyses in exactly the manner
-they were originally performed. This difficulty mainly arises due to the
-number of scientists involved, the large number of steps in a typical
-HEP analysis workflow, and the complexity of the analyses themselves. A
-challenge specific to data analysis and interpretation is tracking the
-evolution of relationships between all the different components of an
-analysis.
-
-Robust methods for data reinterpretation are also critical.
-Collaborations typically interpret results in the context of specific
-models for new physics searches and sometimes reinterpret those same
-searches in the context of alternative theories. However, understanding
-the full implications of these searches requires the interpretation of
-the experimental results in the context of many more theoretical models
-than are currently explored at the time of publication. Analysis
-reproducibility and reinterpretation strategies need to be considered in
-all new approaches under investigation, so that they become a
-fundamental component of the system as a whole.
-
-Adapting to the rapidly evolving landscape of software tools, as well as
-to methodological approaches to data analysis, requires effort in
-continuous training, both for novices as well as for experienced
-researchers, as detailed in the Section 
-\ref{training-and-careers}. The
-maintenance and sustainability of the current analysis ecosystem also
-present a major challenge, as currently this effort is provided by just
-a few institutions. Legacy and less-used parts of the ecosystem need to
-be managed appropriately. New policies are needed to retire little used
-or obsolete components and free up effort for the development of new
-components. These new tools should be made attractive and useful to a
-significant part of the community to attract new contributors.
-
-\subsubsection*{Current Practices}
-
-Methods for analysing HEP data have been developed over many years and
-successfully applied to produce physics results, including more than
-2000 publications, during LHC Runs 1 and 2. Analysis at the LHC
-experiments typically starts with users running code over centrally
-managed data that is of O(100kB/event) and contains all of the
-information required to perform a typical analysis leading to
-publication. The most common approach is through a campaign of
-\emph{data reduction} and \emph{refinement}, ultimately producing
-simplified data structures of arrays of simple data types (``flat
-ntuples'') and histograms used to make plots and tables, from which
-physics results can be derived.
-
-The current centrally-managed data typically used by a Run 2 data
-analysis at the LHC (hundreds of TB) is far too large to be delivered
-locally to the user. An often-stated requirement of the data reduction
-steps is to arrive at a dataset that ``can fit on a laptop'', in order
-to facilitate low-latency, high-rate access to a manageable amount of
-data during the final stages of an analysis. Creating and retaining
-intermediate datasets produced by data reduction campaigns, bringing and
-keeping them ``close'' to the analysers, is designed to minimise latency
-and the risks related to resource contention. At the same time, disk
-space requirements are usually a key constraint of the experiment
-computing models. The
-LHC experiments have made a continuous effort to produce optimised
-analysis-oriented data formats with enough information to avoid the need
-to use intermediate formats. Another effective strategy has been to
-combine analyses from different users and execute them within the same
-batch jobs (so-called ``analysis trains''), thereby reducing the number
-of times data must be read from the storage systems. This has improved
-performance and usability, and simplified the task of the bookkeeping.
-
-There has been a huge investment in using C++ for performance-critical
-code, in particular in event reconstruction and simulation, and this
-will continue in the future. However, for analysis applications, Python
-has emerged as the language of choice in the data science community, and
-its use continues to grow within HEP. Python is highly
-appreciated for its ability to support fast development cycles, for its
-ease-of-use, and it offers an abundance of well-maintained and advanced
-open source software packages. Experience shows that the simpler
-interfaces and code constructs of Python could reduce the complexity of
-analysis code, and therefore contribute to decreasing the ``time to
-insight'' for HEP analyses, as well as increasing their sustainability.
-Increased HEP investment is needed to allow Python to become a first
-class supported language.
-
-One new model of data analysis, developed outside of HEP, maintains the
-concept of sequential reduction, but mixes interactivity with
-batch processing. These exploit new cluster management systems, most
-notably Apache Spark, which uses open-source tools contributed both by
-industry and the data-science community. Other products implementing the
-same analysis concepts and workflows are emerging, such as TensorFlow,
-Dask, Pachyderm, Blaze, Parsl, and Thrill. This approach can complement
-the present and widely adopted Grid processing of datasets. It may
-potentially simplify the access to data and the expression of
-parallelism, thereby improving the exploitation of cluster resources.
-
-An alternative approach, which was pioneered in astronomy but has become
-more widespread throughout the Big Data world, is to perform \emph{fast
-querying} of centrally managed data and compute remotely on the queried
-data to produce the analysis products of interest. The analysis workflow
-is accomplished without focus on persistence of data traditionally
-associated with data reduction, although transient data may be generated
-in order to efficiently accomplish this workflow and optionally can be
-retained to facilitate an analysis ``checkpoint'' for subsequent
-execution. In this approach, the focus is on obtaining the analysis
-end-products in a way that does not necessitate a data reduction
-campaign. It is of interest to understand the role
-that such an approach could have in the global analysis infrastructure,
-and if it can bring an optimisation of the global storage and computing
-resources required for the processing of raw data to analysis.
-
-Another active area regarding analysis in the world outside HEP is the
-switch to a functional or declarative programming model, as for example
-provided by Scala in the Spark environment. This allows scientists to
-express the intended data transformation as a query on data. Instead of
-having to define and control the ``how'', the analyst declares the
-``what'' of their analysis, essentially removing the need to define the
-event loop in an analysis, and leave it to underlying services and
-systems to optimally iterate over events. It appears that these
-high-level approaches will allow abstraction from the underlying
-implementations, allowing the computing systems more freedom in
-optimising the utilisation of diverse forms of computing resources. R\&D
-is already under way, e.g., TDataFrame~\cite{TDataFrame} in ROOT, and
-this needs to be continued with the ultimate goal of establishing a
-prototype functional or declarative programming paradigm.
-
-\subsubsection*{Research and Development Programme}
-
-Towards HL-LHC, we envisage dedicated data analysis facilities for
-experimenters, offering an extendable environment that can provide fully
-functional analysis capabilities, integrating all these technologies
-relevant for HEP. Initial prototypes of such analysis facilities are
-currently under development. On the time scale of HL-LHC, such dedicated
-analysis facilities would provide a complete system engineered for
-latency optimisation and stability.
-
-The following R\&D programme lists the tasks that need to be
-accomplished. By 2020:
-
-\begin{itemize}
-\item
-  Enable new open-source software tools to be plugged in dynamically to
-  the existing ecosystem, and provide mechanisms to dynamically exchange
-  parts of the ecosystem with new components.
-\item
-  Complete an advanced prototype of a low-latency response,
-  high-capacity analysis facility, incorporating fast caching
-  technologies to explore a query-based analysis approach and
-  open-source cluster-management tools. It should in particular include
-  an evaluation of additional storage layers, such as SSD storage and
-  NVRAM-like storage, and cloud and Big Data orchestration systems.
-\item
-  Expand support of Python in our ecosystem with a strategy for ensuring
-  long-term maintenance and sustainability. In particular in ROOT, the
-  current Python bindings should evolve to reach the ease of use of native Python modules.
-\item
-  Prototype a comprehensive set of mechanisms for interacting and
-  exchanging data between new open-source tools and the existing
-  analysis ecosystem.
-\item
-  Develop a prototype based on a functional or declarative programming
-  model for data analysis.
-\item
-  Conceptualise and prototype an analysis ``Interpretation Gateway'',
-  including data repositories, e.g., HEPData~\cite{HEPData,HEPDataRepo}, and
-  analysis preservation and reinterpretation tools.
-\end{itemize}
-
-By 2022:
-
-\begin{itemize}
-\item
-  Evaluate chosen architectures for analysis facilities, verify their
-  design and provide input for corrective actions to test them on a
-  larger scale during Run 3.
-\item
-  Develop a blueprint for remaining analysis facility developments,
-  system design and support model.
-\end{itemize}
-
-\hypertarget{machine-learning}{%
-\subsection{Machine Learning}\label{machine-learning}}
-
-Machine Learning (ML) is a rapidly evolving approach to characterising
-and describing data with the potential to radically change how data is
-reduced and analysed. Some applications will qualitatively improve the
-physics reach of datasets. Others will allow much more efficient use of
-processing and storage resources, effectively extending the physics
-reach of experiments. Many of the activities in this area will
-explicitly overlap with those in the other focus areas, whereas others
-will be more generic. As a first approximation, the HEP community will
-build domain-specific applications on top of existing toolkits and ML
-algorithms developed by computer scientists, data scientists, and
-scientific software developers from outside the HEP world. Work will
-also be done to understand where problems do not map well onto existing
-paradigms and how these problems can be recast into abstract
-formulations of more general interest.
-
-\subsubsection*{Scope and Challenges}
-
-The Machine Learning, Statistics, and Data Science communities have
-developed a variety of powerful ML approaches for classification (using
-pre-defined categories), clustering (where categories are discovered),
-regression (to produce continuous outputs), density estimation,
-dimensionality reduction, etc. Some of these have been used productively
-in HEP for more than 20 years, others have been introduced relatively
-recently. The portfolio of ML techniques and tools is in constant
-evolution, and a benefit is that many have well-documented open source
-software implementations. ML has already become ubiquitous in some
-HEP applications, most notably in classifiers used to discriminate
-between signals and backgrounds in final offline analyses. It is
-also increasingly used in both online and offline reconstruction and
-particle identification algorithms, as well as the classification of
-reconstruction-level objects, such as jets.
-
-The abundance of, and advancements in, ML algorithms and implementations
-present both opportunities and challenges for HEP. 
-The community
-needs to understand which are most appropriate for our use,
-tradeoffs for using one tool compared to another, and
-the tradeoffs of using ML algorithms compared to using more traditional software.
-These issues are not necessarily ``factorisable'', and a key goal will
-be to ensure that, as HEP research teams investigate the numerous
-approaches at hand, the expertise acquired and lessons learned, get
-adequately disseminated to the wider community. In general, each
-\emph{team}, typically a small group of scientists from a collaboration,
-will serve as a source of expertise, helping others develop and
-deploy experiment-specific ML-based algorithms in their software stacks.
-It should provide training to those developing new ML-based algorithms,
-as well as those planning to use established ML tools.
-
-With the advent of more powerful hardware and more performant ML
-algorithms, the ML toolset will be used to develop application software
-that could potentially, amongst other things:
-
-\begin{itemize}
-\item
-  Replace the most computationally expensive parts of pattern
-  recognition algorithms and parameter extraction algorithms for
-  characterising reconstructed objects. For example, investigating how
-  ML algorithms could improve the physics performance or execution speed
-  of charged track and vertex reconstruction, one of the most CPU
-  intensive elements of our current software.
-\item
-  Extend the use of ML algorithms for real-time event classification and
-  analysis, as discussed in more detail in Section \ref{software-trigger-and-event-reconstruction}.
-\item
-  Extend the physics reach of experiments by extending the role of ML at
-  the analysis stage: handling data/MC or control/signal region
-  differences, interpolating between mass points, training in a
-  systematics-aware way, etc.
-\item
-  Compress data significantly with negligible loss of fidelity in terms
-  of physics utility.
-\end{itemize}
-
-As already discussed, many particle physics detectors produce much more
-data than can be moved to permanent storage. The process of reducing the
-size of the datasets is managed by the trigger system. ML algorithms
-have already been used very successfully for triggering, to rapidly
-characterise which events should be selected for additional
-consideration and eventually saved to long-term storage. In the era
-of the HL-LHC, the challenges will increase both quantitatively and
-qualitatively as the number of proton-proton collisions per bunch
-crossing increases. The scope of ML applications in the trigger will
-need to expand in order to tackle the challenges to come.
-
-\subsubsection*{Current Practices}
-
-The use of ML in HEP analyses has become commonplace over the past two
-decades, and the most common use case has been in signal/background
-classification. The vast majority of HEP analyses published in recent
-years have used the HEP-specific software package TMVA~\cite{TMVA} included in ROOT. Recently, however, many HEP analysts have begun
-migrating to non-HEP ML packages such as scikit-learn~\cite{scikit-learn}
-and Keras~\cite{Keras}, although these efforts have yet to result
-in physics publications from major collaborations. Data scientists at
-Yandex created a Python package that provides a consistent API to most
-ML packages used in HEP~\cite{REP}. Packages like Spearmint~\cite{Spearmint} and scikit-optimize~\cite{scikit-optimize} perform Bayesian optimisation and can improve HEP Monte
-Carlo work.
-
-This shift in the set of ML techniques and packages utilised is
-especially strong in the neutrino physics community, where new
-experiments such as DUNE place ML at the very heart of their
-reconstruction algorithms and event selection. The shift is also
-occurring among LHC collaborations, where ML is becoming more and more
-commonplace in reconstruction and real-time applications. Examples where
-ML has already been deployed in a limited way include charged and
-neutral particle reconstruction and identification, jet reconstruction
-and identification, and determining a particle's production properties
-(flavour tagging), based on information from the rest of the event. In
-addition, ML algorithms have been developed that are
-insensitive to changing detector performance, for use in real-time
-applications, and algorithms that are minimally biased with respect to
-the physical observables of interest.
-
-At present, much of this development has happened in specific
-collaborations. While each experiment has, or is likely to have,
-different specific use cases, we expect that many of these will be
-sufficiently similar to each other that R\&D can be done in common. Even
-when this is not possible, experience with one type of problem will
-provide insights into how to approach other types of problem. This is
-why the Inter-experiment Machine Learning forum (IML~\cite{IML}) was
-created at CERN in 2016, as a compliment to experiment specific ML R\&D groups.
-It has already fostered closer collaboration between LHC and non-LHC
-collaborations in the ML field.
-
-\subsubsection*{Research and Development Roadmap and Goals}
-
-The R\&D roadmap presented here is based on the preliminary work done in
-recent years, coordinated by the HSF IML, which will remain the main
-forum to coordinate work in ML in HEP and ensure the proper links
-with the data science communities. The following programme of work is
-foreseen.
-
-By 2020:
-\begin{itemize}
-\item
-  \emph{Particle identification and particle properties:} in calorimeters or
-  time projection chambers (TPCs), where the data can be represented as
-  a 2D or 3D image (or even in 4D, including timing information), the
-  problems can be cast as a computer vision task. Deep Learning (DL),
-  one class of ML algorithm, in which neural networks are used to
-  reconstruct images from pixel intensities, is a good candidate to
-  identify particles and extract many parameters. Promising DL
-  architectures for these tasks include convolutional, recurrent, and
-  adversarial neural networks. A particularly important application is
-  to Liquid Argon TPCs (LArTPCs), which is the chosen detection
-  technology for DUNE, the new flagship experiment in the neutrino programme. 
-  A proof of
-  concept and comparison of DL architectures should be finalised by
-  2020. Particle identification can also be explored to tag the flavour
-  of jets in collider experiments (e.g., so-called b-tagging). The
-  investigation of these concepts, which connect to Natural Language
-  Processing, has started at the LHC and is to be pursued on the same
-  timescale.
-\item
-  \emph{ML middleware and data formats for offline usage:} HEP relies 
-  on the ROOT format for its data, wheras the ML community
-  has developed several other formats, often associated with specific ML
-  tools. A desirable data format for ML applications should have the
-  following attributes: high read-write speed for efficient training,
-  sparse readability without loading the entire dataset into RAM,
-  compressibility, and widespread adoption by the ML community. The thorough
-  evaluation of the different data formats and their impact on ML
-  performance in the HEP context must be continued, and it is necessary
-  to define a strategy for bridging or migrating HEP formats to the
-  chosen ML format(s), or vice-versa.
-\item
-  \emph{Computing resource optimisations:} managing large
-  volume data transfers is one
-  of the challenges facing current computing facilities. Networks 
-  play a crucial role in data exchange and so a
-  network-aware application layer may significantly improve experiment
-  operations. ML is a promising technology to identify anomalies in
-  network traffic, to predict and prevent network congestion, to detect
-  bugs via analysis of self-learning networks, and for WAN path
-  optimisation based on user access patterns.
-\item
-  \emph{ML as a Service (MLaaS):} current cloud providers rely on a MLaaS model
-  exploiting interactive
-  machine learning tools in order to make efficient use of resources, however,
-  this is not yet widely used in HEP. HEP
-  services for interactive analysis, such as CERN's Service for
-  Web-based Analysis, SWAN~\cite{SWAN}, may play an important role in adoption
-  of machine learning tools in HEP workflows. In order to use these tools
-  more efficiently, sufficient and appropriately tailored hardware and
-  instances other than SWAN will be identified.
-\end{itemize}
-
-By 2022:
-\begin{itemize}
-\item
-  \emph{Detector anomaly detection:} data taking is
-  continuously monitored by physicists taking shifts to monitor and
-  assess the quality of the incoming data, largely using reference
-  histograms produced by experts. A whole class of ML algorithms called anomaly detection
-  can be useful for automating this important task. Such unsupervised algorithms are able
-  to learn from data and produce an alert when deviations are observed.
-  By monitoring many variables at the same time, such algorithms are
-  sensitive to subtle signs forewarning of imminent failure, so that
-  pre-emptive maintenance can be scheduled. These techniques are already
-  used in industry.
-\item
-  \emph{Simulation:} recent progress in high fidelity fast generative models,
-  such as Generative Adversarial Networks (GANs) and Variational
-  Autoencoders (VAEs), which are able to sample high dimensional feature
-  distributions by learning from existing data samples, offer a
-  promising alternative for Fast Simulation. A simplified first attempt at
-  using such techniques in simulation saw orders of magnitude increase
-  in speed over existing Fast Simulation techniques, but has not yet
-  reached the required accuracy~\cite{CaloGAN}.
-\item
-  \emph{Triggering and real-time analysis:} one of the challenges is the
-  trade-off in algorithm complexity and performance under strict
-  inference time constraints. To deal with the increasing event
-  complexity at HL-LHC, the use of sophisticated ML
-  algorithms will be explored at all trigger levels, building on the pioneering work of
-  the LHC collaborations. A critical part of this work will be to
-  understand which ML techniques allow us to maximally exploit
-  future computing architectures.
-\item
-  \emph{Sustainable Matrix Element Method (MEM):} MEM is a powerful
-  technique that can be utilised for making measurements of physical model
-  parameters and direct searches for new phenomena. As it is
-  very computationally intensive its use in HEP is limited.
-  Although the use of neural networks for numerical integration is not new, 
-  it is a
-  technical challenge to design a network sufficiently rich
-  to encode the complexity of the ME calculation for a given process
-  over the phase space relevant to the signal process. Deep Neural
-  Networks (DNNs) are good candidates~\cite{Bendavid2017,BendavidIML2017}.
-\item
-  \emph{Tracking:} pattern recognition is always a computationally challenging
-  step. It becomes a huge challenge in the HL-LHC environment.
-  Adequate ML techniques may provide a solution that scales linearly
-  with LHC intensity. Several efforts in the HEP community have started
-  to investigate ML algorithms for track pattern recognition on
-  many-core processors.
-\end{itemize}
-
-\hypertarget{data-organisation-management-and-access}{%
-\subsection{Data Organisation, Management and
-Access}\label{data-organisation-management-and-access}}
-
-The scientific reach of data-intensive experiments is limited by how
-fast data can be accessed and digested by computational resources. Changes in
-computing technology and large increases in data volume require new
-computational models~\cite{Butler2013}, compatible with
-budget constraints. The integration of newly emerging data
-analysis paradigms into our computational model has the potential to
-enable new analysis methods and increase scientific output. The field,
-as a whole, has a window in which to adapt our data access and data
-management schemes to ones that are more suited and optimally matched to
-advanced computing models and a wide range of analysis applications.
-
-\subsubsection*{Scope and Challenges}
-
-The LHC experiments currently provision and manage about an exabyte of
-storage, approximately half of which is archival, and half is
-traditional disk storage. Other experiments that will soon start data taking have
-similar needs, e.g., Belle II has the same data volumes as ATLAS. The
-HL-LHC storage requirements per year are expected to jump by a factor
-close to 10, which is a growth rate faster than can be 
-accommodated by projected technology gains. Storage will remain one of the major
-cost drivers for HEP computing, at a level roughly equal to the cost
-of the computational resources. The combination of storage and analysis
-computing costs may restrict scientific output and the potential physics
-reach of the experiments, so new techniques and algorithms are likely to
-be required.
-
-In devising experiment computing models for this era many factors have
-to be taken into account. In particular, the increasing availability of
-very high-speed networks may reduce the need for CPU and data
-co-location. Such networks may allow for more
-extensive use of data access over the wide-area network (WAN), which may
-provide failover capabilities, global and federated data namespaces, and
-will have an impact on data caching. Shifts in data presentation and
-analysis models, such as the use of event-based data streaming along
-with more traditional dataset-based or file-based data access, will be
-particularly important for optimising the utilisation of opportunistic
-computing cycles on HPC facilities, commercial cloud resources, and
-campus clusters. This can potentially resolve currently limiting factors
-such as job eviction.
-
-The three main challenges for data management in the HL-LHC follow:
-
-\begin{itemize}
-\item
-  The experiments will significantly increase both
-  the data rate and the data volume. The computing systems will need to
-  handle this with as small a cost increase as possible and within
-  evolving storage technology limitations.
-\item
-  The significantly increased computational requirements for the HL-LHC
-  era will also place new requirements on data access. Specifically, the use of
-  new types of computing resources (cloud, HPC) that have different dynamic
-  availability and characteristics will require more dynamic data
-  management and access systems.
-\item
-  Applications employing new techniques, such as 
-  training for machine learning or high rate data query systems, will likely be employed to
-  meet the computational constraints and to extend physics reach.
-  These new applications will place new requirements
-  on how and where data is accessed and produced. Specific applications,
-  such as training for machine learning, may require use of specialised
-  processor resources, such as GPUs, placing further requirements on
-  data.
-\end{itemize}
-
-The projected event complexity of data from future HL-LHC
-runs with high pileup and from high resolution Liquid Argon detectors at
-DUNE will require advanced reconstruction algorithms and analysis tools
-to interpret the data. The precursors of these tools, in the form of
-new pattern recognition and tracking algorithms, 
-are already proving to be drivers for the
-compute needs of the HEP community. The storage systems that are
-developed, and the data management techniques that are employed,
-will need to be matched to these changes in computational work, so
-as not to hamper potential improvements.
-
-As with computing resources, the landscape of storage solutions
-is trending towards heterogeneity. The ability to
-leverage new storage technologies as they become available into existing
-data delivery models is a challenge that we must be prepared for. This
-also implies the need to leverage
-``tactical storage'', i.e., storage that becomes more cost-effective as
-it becomes available (e.g., from a cloud provider), and have a data
-management and provisioning system that can exploit such resources at
-short notice. Volatile data sources would impact many aspects of the
-system: catalogues, job brokering, monitoring and alerting, accounting,
-the applications themselves.
-
-On the hardware side, R\&D is needed in alternative approaches to data
-archiving to determine the possible cost/performance tradeoffs.
-Currently, tape is extensively used to hold data that cannot be
-economically made available online. While the data is still accessible,
-it comes with a high latency penalty, limiting effective data access. We
-suggest investigating either separate direct access-based archives
-(e.g., disk or optical) or new models that hierarchically overlay online
-direct access volumes with archive space. This is especially relevant
-when access latency is proportional to storage density. Either approach
-would need to also evaluate reliability risks and the effort needed to
-provide data stability. For this work, we should exchange experiences
-with communities that rely on large tape archives for their primary
-storage.
-
-Cost reductions in the maintenance and operation of storage
-infrastructure can be realised through convergence of the major
-experiments and resource providers on shared solutions. This does not
-necessarily mean promoting a monoculture, as different solutions will be
-adapted to certain major classes of use cases, type of site, or funding
-environment. There will always be a judgement to make on the
-desirability of using a variety of specialised systems, or of
-abstracting the commonalities through a more limited, but common,
-interface. Reduced costs and improved sustainability will be further
-promoted by extending these concepts of convergence beyond HEP and into
-the other large-scale scientific endeavours that will share the
-infrastructure in the coming decade (e.g., the SKA and CTA experiments).
-Efforts must be made as early as possible, during the formative design
-phases of such projects, to create the necessary links.
-
-Finally, all changes undertaken must not make the ease of access to data
-any worse than it is under current computing models. We must also be
-prepared to accept the fact that the best possible solution may require
-significant changes in the way data is handled and analysed. What is
-clear is that current practices will not scale to the needs of HL-LHC
-and other major HEP experiments of the coming era.
-
-\subsubsection*{Current Practices}
-
-The original LHC computing models were based on simpler models used
-before distributed computing was a central part of HEP computing. This
-allowed for a reasonably clean separation between four different aspects
-of interacting with data, namely data organisation, data management,
-data access, and data granularity. The meaning of these terms may be 
-summarised in what follows.
-
-\begin{itemize}
-\item
-  \emph{Data organisation} is essentially how data is structured as it
-  is written. Most data is written in files, in ROOT format, typically
-  with a column-wise organisation of the data. The records corresponding
-  to these columns are compressed. The internal details of this
-  organisation are visible only to individual software applications.
-\item
-  In the past, the key challenge for \emph{data management} was the transition to
-  use distributed computing in the form of the grid. The experiments
-  developed dedicated data transfer and placement systems, along with
-  catalogues, to move data between computing centres. Originally,
-  computing models were rather static: data was placed at sites, and the
-  relevant compute jobs were sent to the right locations. Since LHC
-  startup, this model has been made more flexible to limit non-optimal
-  pre-placement and to take into account data popularity. In addition,
-  applications might interact with catalogues or, at times, the workflow
-  management system does this on behalf of the applications.
-\item
-  \emph{Data access}: historically, various protocols have been used for
-  direct reads (rfio, dcap, xrootd, etc.) where jobs are reading data
-  explicitly staged-in or cached by the compute resource used or the
-  site it belongs to. A recent move has been the convergence towards
-  xrootd as the main protocol for direct access. With direct access,
-  applications may use alternative protocols to those used by data
-  transfers between sites. In addition, LHC experiments have been
-  increasingly using remote access to the data, without any stage-in
-  operations, using the possibilities offered by protocols such as xrootd
-  or http.
-\item
-  \emph{Data granularity}: the data is split into datasets, as defined
-  by physics selections and use cases, consisting of a set of individual
-  files. While individual files in datasets can be processed in
-  parallel, the files themselves are usually processed as a whole.
-\end{itemize}
-
-Before LHC turn-on, and in the first years of the LHC, these four
-areas were to first order optimised independently. As LHC computing
-matured, interest has turned to optimisations spanning multiple areas.
-For example, the recent use of ``Data Federations'' mixes up Data
-Management and Access. As we will see below, some of the foreseen
-opportunities towards HL-LHC may require global optimisations.
-
-Thus, in this section we take a broader view than traditional data
-management and consider the combination of ``Data Organisation,
-Management and Access'' (DOMA) together. We believe that this fuller
-picture will provide important opportunities for improving
-efficiency and scaleability, as we enter the many-exabyte era.
-
-\subsubsection*{Research and Development Programme}
-
-In the following, we describe
-tasks that will need to be carried out in order to demonstrate that the
-increased volume and complexity of data expected over the coming decade
-can be stored, accessed, and analysed at an affordable cost.
-
-\begin{itemize}
-\item
-  Sub-file granularity, e.g.,
-  event-based, will be studied to see whether it can be implemented
-  efficiently, and in a scalable, cost-effective manner, for all
-  applications making use of event selection, to see whether it offers
-  an advantage over current file-based granularity. The following tasks
-  should be completed by 2020:
-  \begin{itemize}
-  \item
-    Quantify the impact on
-    performance and resource utilisation  of the storage and network
-    for the main
-    access patterns, i.e., simulation, reconstruction, analysis.
-  \item
-    Assess the impact on
-    catalogues and data distribution.
-  \item
-    Assess whether
-    event-granularity makes sense in object stores that tend to require
-    large chunks of data for efficiency.
-  \item
-    Test for improvement in
-    recoverability from preemption, in particular when using cloud spot
-    resources and/or dynamic HPC resources.
-  \end{itemize}
-\item
-  We will seek to derive benefits
-  from data organisation and analysis technologies adopted by other big
-  data users. A proof-of-concept that involves the following tasks needs
-  to be established by 2020 to allow full implementations to be made in
-  the years that follow.
-  \begin{itemize}
-  \item
-    Study the impact of
-    column-wise, versus row-wise, organisation of data on the
-    performance of each kind of access.
-  \item
-    Investigate efficient data
-    storage and access solutions that support the use of map-reduce or
-    Spark-like analysis services.
-  \item
-    Evaluate just-in-time
-    decompression schemes and mappings onto hardware architectures
-    considering the flow of data, from spinning disk to memory and
-    application.
-  \end{itemize}
-\item
-  Investigate the role data
-  placement optimisations can play, such as caching, in order to use
-  computing resources effectively, and the technologies that can be used
-  for this. The following tasks should be completed by 2020:
-  \begin{itemize}
-  \item
-   Quantify the benefit of
-    placement optimisation for reconstruction,
-    analysis, and simulation.
-  \item
-    Assess the benefit of caching
-    for Machine Learning-based applications, in particular for the
-    learning phase, and follow-up the evolution of technology outside
-    HEP.
-  \end{itemize}
-  In the longer term the benefits
-  that can be derived from using different approaches to the way HEP is
-  currently managing its data delivery systems should be studied. Two
-  different content delivery methods will be looked at, namely Content
-  Delivery Networks (CDN) and Named Data Networking (NDN).
-\item
-  Study how to minimise HEP
-  infrastructure costs by exploiting varied quality of service from
-  different storage technologies. In particular, study the role that
-  opportunistic/tactical storage can play, as well as different archival
-  storage solutions. A proof-of-concept should be made by 2020, with a
-  full implementation to follow in the following years.
-\end{itemize}
-
-\begin{itemize}
-\item
-  Establish how to globally optimise data access latency, with respect
-  to the efficiency of using CPU, at a sustainable cost. This involves
-  studying the impact of concentrating data in fewer, larger locations
-  (the ``data-lake'' approach), and making increased use of opportunistic
-  compute resources located further from the data. Again, a
-  proof-of-concept should be made by 2020, with a full implementation in
-  the following years, if successful. This R\&D will be done in common
-  with the related actions planned as part of Facilities and Distributed
-  Computing.
-\end{itemize}
-
-\hypertarget{facilities-and-distributed-computing}{%
-\subsection{Facilities and Distributed
-Computing}\label{facilities-and-distributed-computing}}
-
-\subsubsection*{Scope and Challenges}
-
-As outlined in Section \ref{software-and-computing-challenges}, 
-huge resource requirements are anticipated for
-HL-LHC running. These need to be deployed and managed across the WLCG
-infrastructure, which has evolved from the original ideas on deployment
-before LHC data-taking started~\cite{MONARC}, to be a mature and
-effective infrastructure that is now exploited by LHC experiments.
-Currently, hardware costs are dominated by disk storage, closely
-followed by CPU, followed by tape and networking. Naive estimates of
-scaling to meet HL-LHC needs indicate that the current system would need
-almost an order of magnitude more resources than will be available from
-technology evolution alone. In addition, other initiatives such as Belle
-II and DUNE in particle physics, but also other science projects such as
-SKA, will require a comparable amount of resources on the same
-infrastructure. Even anticipating substantial software improvements, the
-major challenge in this area is to find the best configuration for
-facilities and computing sites that make HL-LHC computing feasible. This
-challenge is further complicated by substantial regional differences in
-funding models, meaning that any solution must be sensitive to these
-local considerations to be effective.
-
-There are a number of changes that can be anticipated on the timescale
-of the next decade that must be taken into account. There is an
-increasing need to use highly heterogeneous resources, including the
-use of HPC infrastructures (which can often have very particular setups
-and policies that make their exploitation challenging); volunteer
-computing (which is restricted in scope and unreliable, but can be a
-significant resource); and cloud computing,
-both commercial and research. All of these offer different resource
-provisioning interfaces and can be significantly more dynamic than
-directly funded HEP computing sites. In addition, diversity of computing
-architectures is expected to become the norm, with different CPU
-architectures, as well as more specialised GPUs and FPGAs.
-
-This increasingly dynamic environment for resources, particularly CPU,
-must be coupled with a highly reliable system for data storage and a
-suitable network infrastructure for delivering this data to where it
-will be processed. While CPU and disk capacity is expected to increase
-by respectively 15\% and 25\% per year for the same cost~\cite{CERN-COST-EST}, the trends of
-research network capacity increases show a much steeper growth, such as
-two orders of magnitude from now to HL-LHC times. Therefore, the
-evolution of the computing models would need to be more network centric.
-
-In the network domain, there are new technology developments, such as
-Software Defined Networks (SDNs), which enable user-defined high
-capacity network paths to be controlled via experiment software, and
-which could help manage these data flows. These new technologies require
-considerable R\&D to prove their utility and practicality. In addition,
-the networks used by HEP are likely to see large increases in traffic
-from other science domains.
-
-Underlying storage system technology will continue to evolve, for
-example towards object stores, and, as proposed in  
-Data Organisation, Management and Access  
-(Section \ref{data-organisation-management-and-access}),
-R\&D is also necessary to understand their usability and their role in
-the HEP infrastructures. There is also the continual challenge of
-assembling inhomogeneous systems and sites into an effective widely
-distributed worldwide data management infrastructure that is usable by
-experiments. This is particularly compounded by the scale increases for
-HL-LHC where multiple replicas of data (for redundancy and availability)
-will become extremely expensive.
-
-Evolutionary change towards HL-LHC is required, as the experiments will
-continue to use the current system. Mapping out a path for migration
-then requires a fuller understanding of the costs and benefits of the
-proposed changes. A model is needed in which the benefits of such
-changes can be evaluated, taking into account hardware and human costs,
-as well as the impact on software and workload performance that in turn
-leads to physics impact. Even if HL-LHC is the use case used to build
-this cost and performance model, because the ten years of experience
-running large-scale experiments helped to define the needs, it is
-believed that this work, and the resulting model, will be valuable for
-other upcoming data intensive scientific initiatives. This includes
-future HEP projects, such as Belle II, DUNE and possibly ILC
-experiments, but also non-HEP projects, such as SKA.
-
-\subsubsection*{Current Practices}
-
-While there are many particular exceptions, most resources incorporated
-into the current WLCG are done so in independently managed sites,
-usually with some regional organisation structure, and mostly offering
-both CPU and storage. The sites are usually funded directly to provide
-computing to WLCG, and are in some sense then ``owned'' by HEP, albeit
-often shared with others. Frequently substantial cost contributions are
-made indirectly, for example through funding of energy costs or
-additional staff effort, particularly at smaller centres. Tape is found
-only at CERN and at large national facilities, such as the WLCG Tier-1s~\cite{Bird:1695401}.
-
-Interfaces to these computing resources are defined by technical
-operations in WLCG. Frequently there are choices that sites can make
-among some limited set of approved options for interfaces. These can
-overlap in functionality. Some are very HEP specific and recognised as
-over-complex: work is in progress to get rid of them. The acceptable
-architectures and operating systems are also defined at the WLCG level
-(currently x86\_64, running Scientific Linux 6 and compatible), and
-sites can deploy these either directly onto ``bare metal'' or can use an
-abstraction layer, such as virtual machines or containers.
-
-There are different logical networks being used to connect sites: LHCOPN
-connects CERN with the Tier-1 centres and a mixture of LHCONE and
-generic academic networks connect other sites.
-
-Almost every experiment layers its own customised workload and data
-management system on top of the base WLCG provision, with several
-concepts, and a few lower-level components, in common. The pilot job
-model for workloads is ubiquitous, where a real workload is dispatched
-only once a job slot is secured. Data management layers aggregate files
-in the storage systems into datasets and manage experiment-specific
-metadata. In contrast to the MONARC model, sites are generally used more
-flexibly and homogeneously by experiments, both in workloads and in data
-stored.
-
-In total, WLCG currently provides experiments with resources distributed
-at about 170 sites, in 42 countries, which pledge every year the amount
-of CPU and disk resources they are committed to delivering. The pledge
-process is overseen by the Computing Resource Scrutiny Group (CRSG), mandated by
-the funding agencies to validate the experiment requests, and to
-identify mismatches with site pledges. These sites are connected by
-10-100 Gb links, and deliver approximately 500k CPU cores and 1 EB of
-storage, of which 400 PB is disk. More than 200M jobs are executed each day~\cite{Bird2017}.
-
-\subsubsection*{Research and Development programme}
-
-The following areas of study are ongoing, and will involve technology
-evaluations, prototyping, and scale tests. Several of the items below
-require some coordination with other topical areas discussed in this
-document, and some work is still needed to finalise the detailed action
-plan. These actions will need to be structured to meet the common
-milestones of informing the HL-LHC Computing Technical Design Reports (TDRs), and deploying
-advanced prototypes during LHC Run 3.
-
-\begin{itemize}
-\item
-  Understand better the relationship between the performance and costs
-  of the WLCG system, and how it delivers the necessary functionality to
-  support LHC physics. This will be an ongoing process, started by the
-  recently formed System Performance and Cost Modeling Working Group,
-  and aims to provide a quantitative assessment for any proposed
-  changes.
-\item
-  Define the functionality needed to implement a federated data centre
-  concept (``data lake'') that aims to reduce the operational cost of
-  storage for HL-LHC, and at the same time better manage network
-  capacity, whilst maintaining the overall CPU efficiency. This would
-  include the necessary qualities of service, and options for regionally
-  distributed implementations, including the ability to flexibly respond
-  to model changes in the balance between disk and tape. This work
-  should be done in conjunction with the existing Data Organisation,
-  Management and Access Working Group to evaluate the impact of the
-  different access patterns and data organisations envisaged.
-\item
-  Establish an agreement on the common data management functionality
-  that is required by experiments, targeting a consolidation and a lower
-  maintenance burden. The intimate relationship between the management
-  of elements in storage systems and metadata must be recognised. This
-  work requires coordination with the Data Processing Frameworks
-  Working Group. It
-  needs to address at least the following use cases:
-
-  \begin{itemize}
-  \item
-    processing sites that may have some small disk cache, but do not
-    manage primary data;
-  \item
-    fine grained processing strategies that may enable processing of
-    small chunks of data, with appropriate bookkeeping support;
-  \item
-    integration of heterogeneous processing resources, such as HPCs and
-    clou\-ds.
-  \end{itemize}
-\item
-  Explore scalable and uniform means of workload scheduling, which
-  incorporate dynamic heterogenous resources, and the capabilities of
-  finer grained processing that increases overall efficiency. The
-  optimal scheduling of special workloads that require particular
-  resources is clearly required.
-\item
-  Contribute to the prototyping and evaluation of a quasi-interactive
-  analysis facility that would offer a different model for physics
-  analysis, but would also need to be integrated into the data and
-  workload management of the experiments. This is work to be done in
-  collaboration with the Data Analysis and Interpretation Working Group.
-\end{itemize}
-
-\hypertarget{data-flow-processing-framework}{%
-\subsection{Data-Flow Processing
-Framework}\label{data-flow-processing-framework}}
-
-\subsubsection*{Scope and Challenges}
-
-Frameworks in HEP are used for the collaboration-wide
-data processing tasks of triggering, reconstruction, and simulation, as
-well as other tasks that subgroups of the collaboration are responsible
-for, such as detector alignment and calibration. Providing framework
-services and libraries that will satisfy the computing and data needs
-for future HEP experiments in the next decade, while maintaining our
-efficient exploitation of increasingly heterogeneous resources, is a
-huge challenge.
-
-To fully exploit the potential of modern processors, HEP data processing
-frameworks need to allow for the parallel execution of reconstruction or
-simulation algorithms on multiple events simultaneously. Frameworks face
-the challenge of handling the massive parallelism and heterogeneity that
-will be present in future computing facilities, including multi-core and
-many-core systems, GPUs, Tensor Processing Units (TPUs), and tiered
-memory systems, each integrated with storage and high-speed network
-interconnections. Efficient running on heterogeneous resources will
-require a tighter integration with the computing models' higher-level
-systems of workflow and data management. Experiment frameworks must also
-successfully integrate and marshall other HEP software that may have its
-own parallelisation model, such as physics generators and detector
-simulation.
-
-Common developments across experiments are desirable in this area, but
-are hampered by many decades of legacy work. Evolving our frameworks
-also has to be done recognising the needs of the different stakeholders
-in the system. This includes physicists, who are writing processing
-algorithms for triggering, reconstruction or analysis; production
-managers, who need to define processing workflows over massive datasets;
-and facility managers, who require their infrastructures to be used
-effectively. These frameworks are also constrained by security
-requirements, mandated by the groups and agencies in charge of it.
-
-\subsubsection*{Current Practices}
-
-Although most frameworks used in HEP share common concepts, there are,
-for mainly historical reasons, a number of different implementations;
-some of these are shared between experiments. The Gaudi framework~\cite{Barrand:2001ny} was
-originally developed by LHCb, but is also used by ATLAS and various
-non-LHC experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff}, which was forked
-to provide the art framework for the Fermilab Intensity Frontier experiments~\cite{Green:2012gv}. 
-Belle II uses basf2~\cite{1742-6596-331-3-032024}. The linear collider community developed and uses
-Marlin~\cite{Gaede:2006pj}. The FAIR experiments use FairROOT, closely related to ALICE's
-AliROOT. The FAIR experiments and ALICE are now developing a new
-framework, which is called O2~\cite{O2}. At the time of writing, most 
-major frameworks support basic parallelisation, both within and across
-events, based on a task-based model~\cite{Jones:2015soc}\cite{Clemencic:2015paa}.
-
-Each framework has a processing model, which provides the means to
-execute and apportion work. Mechanisms for this are threads, tasks,
-processes, and inter-process communication. The different strategies used
-reflect different trade-offs between constraints in the programming
-model, efficiency of execution, and ease of adapting to inhomogeneous
-resources. These concerns also reflect two different behaviours:
-firstly, maximising throughput, where it is most important to maximise
-the number of events that are processed by a given resource; secondly,
-minimising latency, where the primary constraint is on how long it takes
-to calculate an answer for a particular datum.
-
-Current practice for throughput maximising system architectures have
-con\-strain\-ed the scope of framework designs. Framework applications have
-largely been viewed by the system as a batch job with complex
-configuration, consuming resources according to rules dictated by the
-computing model: one process using one core on one node, operating
-independently with a fixed size memory space on a fixed set of files
-(streamed or read directly). Only recently has CMS broken this tradition
-starting at the beginning of Run 2, by utilising all available cores in one process space using threading. ATLAS is currently
-using a multi-process fork-and-copy-on-write solution to remove the
-constraint of one core/process. Both experiments were driven to solve
-this problem by the ever-growing need for more memory per process
-brought on by the increasing complexity of LHC events. Current practice
-manages systemwide (or facility-wide) scaling by dividing up datasets,
-generating a framework application configuration, and scheduling jobs on
-nodes/cores to consume all available resources. Given anticipated
-changes in hardware (heterogeneity, connectivity, memory, storage)
-available at computing facilities, the interplay between
-workflow and workload management systems and framework applications need to
-be carefully examined. It may be advantageous to permit framework
-applications (or systems) to span multi-node resources, allowing them
-to be first-class participants in the business of scaling within a
-facility. In our community some aspects of this approach, which maps
-features with microservices or function as a service, is being pioneered
-by the O2 framework.
-
-\subsubsection*{Research and Development programme}
-
-By the end of 2018: review the existing technologies that are the
-important building blocks for data processing frameworks and reach
-agreement on the main architectural concepts for the next generation of
-frameworks. Community meetings and workshops, along the lines of the
-original Concurrency Forum, are envisaged in order to foster collaboration in this work~\cite{ConcurrencyForum}. This includes the following:
-
-\begin{itemize}
-\item
-  Libraries used for concurrency, their likely evolution and the issues
-  in integrating the models used by detector simulation and physics
-  generators into the frameworks.
-\item
-  Functional programming, as well as domain specific languages, as a way
-  to describe the physics data processing that has to be undertaken
-  rather than how it has to be implemented. This approach is based on
-  the same concepts as the idea for functional approaches for
-  (statistical) analysis as described in Section \ref{data-analysis-and-interpretation}.
-\item
-  Analysis of the functional differences between the existing frameworks
-  and the different experiment use cases.
-\end{itemize}
-
-By 2020: prototype and demonstrator projects for the agreed
-architectural concepts and baseline to inform the HL-LHC Computing TDRs
-and to demonstrate advances over what is currently deployed. The
-following specific items will have to be taken into account:
-
-\begin{itemize}
-\item
-  These prototypes should be as common as possible between existing
-  frameworks, or at least several of them, as a proof-of-concept of
-  effort and component sharing between frameworks for their future
-  evolution. Possible migration paths to more common implementations
-  will be part of this activity.
-\item
-  In addition to covering the items mentioned for the review phase, they
-  should particularly demonstrate possible approaches for scheduling the
-  work across heterogeneous resources and using them efficiently, with a
-  particular focus on the efficient use of co-processors, such as GPUs.
-\item
-  They need to identify data model changes that are required for an
-  efficient use of new processor architectures (e.g., vectorisation),
-  and for scaling I/O performance in the context of concurrency.
-\item
-  Prototypes of a more advanced integration with workload management,
-  taking advantage in particular of the advanced features available at
-  facilities for a finer control of the interactions with storage and
-  network, and dealing efficiently with the specificities of HPC
-  resources.
-\end{itemize}
-
-By 2022: production-quality framework libraries usable by several
-experiment frameworks, covering the main areas successfully demonstrated
-in the previous phase. During these activities we expect at least one
-major paradigm shift to take place on this 5-year time scale. It will be
-important to continue discussing their impact within the community,
-which will be ensured through appropriate cross-experiment workshops
-dedicated to data processing frameworks.
-
-\hypertarget{conditions-data}{%
-\subsection{Conditions Data}\label{conditions-data}}
-
-\subsubsection*{Scope and Challenges}
-
-Conditions data is defined as the non-event data required by
-data-processing software to correctly simulate, digitise or reconstruct
-the raw detector event data. The non-event data discussed here consists
-mainly of detector calibration and alignment information, with some
-additional data describing the detector configuration, the machine
-parameters, as well as information from the detector control system.
-
-Conditions data is different from event data in many respects, but one
-of the important differences is that its volume scales with time rather
-than with the luminosity. As a consequence, its growth is limited, as
-compared to event data: conditions data volume is expected to be at the
-terabyte scale and the update rate is modest (typically O(1)Hz).
-However, conditions data is used by event processing applications
-running on a very large distributed computing infrastructure, resulting
-in tens of thousands of jobs that may try to access the conditions data
-at the same time, and leading to a very significant rate of reading
-(typically O(10) kHz).
-
-To successfully serve such rates, some form of caching is needed, either
-by using services such as web proxies (CMS and ATLAS use Frontier) or by
-delivering the conditions data as files distributed to the jobs. For the
-latter approach, CVMFS is an attractive solution due to its embedded
-caching, and its advanced snapshotting and branching features. ALICE
-have made some promising tests, and started to use this approach in Run
-2; Belle II already took the same approach~\cite{WoodACAT2017}, and NA62
-has also decided to adopt this solution. However, one particular
-challenge to be overcome with the filesystem approach is to design an
-efficient mapping of conditions data and metadata to files in order to
-use the CVMFS caching layers efficiently.
-
-Efficient caching is especially important in order to support the
-high-reading rates that will be necessary for ATLAS and CMS experiments
-starting with Run 4. For these experiments, a subset of the conditions
-data is linked to the luminosity, leading to an interval of granularity
-down to the order of a minute. Insufficient or inefficient caching may
-impact the efficiency of the reconstruction processing.
-
-Another important challenge is ensuring the long-term maintainability of
-the conditions data storage infrastructure. Shortcomings in the initial
-approach used in LHC Run 1 and Run 2, leading to complex
-implementations, helped to identify the key requirements for an
-efficient and sustainable condition data handling infrastructure. There
-is now a consensus among experiments on these requirements~\cite{Laycock2017}: 
-ATLAS and CMS are working on a common next-generation
-conditions database~\cite{1742-6596-898-4-042047}.
-The Belle II experiment, which is about to start its data taking, has already
-developed a solution based on the same concepts and architecture. One
-key point in this new design is to have a server mostly agnostic to the
-data content with most of the intelligence on the client side. This new
-approach should make it easier to rely on well-established open-source
-products (e.g., Boost) or software components developed for the
-processing of event data (e.g., CVMFS). With such an approach, it should
-be possible to leverage technologies such as REST interfaces to simplify
-insertion and read operations, and make them very efficient to reach the
-rate levels foreseen. Also, to provide a resilient service to jobs that
-depend on it, the client will be able to use multiple proxies or servers
-to access the data.
-
-One conditions data challenge may be linked to the use of an event
-service, as ATLAS is doing currently, to use efficiently HPC facilities
-for event simulation or processing. The event service allows better use
-of resources that may be volatile by allocating and bookkeeping the work
-done, not at the job granularity, but at the event granularity. This
-reduces the possibility for optimising access to the conditions data at
-the job level, and may lead to an increased pressure on the conditions
-data infrastructure. This approach is still at an early stage, and more
-experience is needed to better appreciate the exact impact on the
-conditions data.
-
-\subsubsection*{Current Practices}
-
-The data model for conditions data management is an area where the
-experiments have converged on something like a best common practice. The
-time information for the validity of the Payloads is specified with a
-parameter called an \emph{Interval Of Validity} (IOV), which can be
-represented by a Run number, the ID of a luminosity section or a
-universal timestamp. A fully qualified set of conditions data consists
-of a set of payloads and their associate IOVs covering the time span
-required by the workload. A label called a \emph{Tag} identifies the
-version of the set and the global tag is the top-level configuration of
-all conditions data. For a given detector subsystem and a given IOV, a
-global tag will resolve to one, and only one, conditions data payload.
-The global tag resolves to a particular system tag via the global tag
-map table. A system tag consists of many intervals of validity or
-entries in the IOV table. Finally, each entry in the IOV table maps to a
-payload via its unique hash key.
-
-A relational database is a good choice for implementing this design. One
-advantage of this approach is that a payload has a unique identifier,
-its hash key, and this identifier is the only way to access it. All
-other information, such as tags and IOV, is metadata used to select a
-particular payload. This allows a clear separation of the payload data
-from the metadata, and may allow use of a different backend technology
-to store the data and the metadata. This has potentially several
-advantages:
-
-\begin{itemize}
-\item
-  Payload objects can be cached independently of their metadata, using
-  the appropriate technology, without the constraints linked to metadata
-  queries.
-\item
-  Conditions data metadata are typically small compared to the
-  conditions data themselves, which makes it easy to export them as a
-  single file using technologies such as SQLite. This may help for
-  long-term data preservation.
-\item
-  IOVs, being independent of the payload, can also be cached on their
-  own.
-\end{itemize}
-
-A recent trend is the move to full reconstruction online, where the
-calibrations and alignment are computed and applied in the High Level
-Trigger (HLT). This is currently being tested by ALICE and LHCb, who
-will adopt it for use in Run 3. This will offer an opportunity to
-separate the distribution of conditions data to reconstruction jobs and
-analysis jobs, as they will not run on the same infrastructure. However,
-running reconstruction in the context of the HLT will put an increased
-pressure on the access efficiency to the conditions data, due to the HLT
-time budget constraints.
-
-\subsubsection*{Research and Development Programme}
-
-R\&D actions related to Conditions databases are already in progress,
-and all the activities described below should be completed by 2020. This
-will provide valuable input for the future HL-LHC TDRs, and allow these
-services to be deployed during Run 3 to overcome the limitations seen in
-today's solutions.
-
-\begin{itemize}
-\item
-  File-system view of conditions data for analysis jobs: study how to
-  leverage advanced snapshotting/branching features of CVMFS for
-  efficiently distributing conditions data as well as ways to optimise
-  data/metadata layout in order to benefit from CVMFS caching. Prototype
-  production of the file-system view from the conditions database.
-\item
-  Identify and evaluate industry technologies that could replace
-  HEP-specific components.
-\item
-  ATLAS: migrate current implementations based on COOL to the proposed
-  REST-based approach; study how to avoid moving too much complexity on
-  the client side, in particular for easier adoption by subsystems, e.g.,
-  possibility of common modules/libraries. ALICE is also planning to
-  explore this approach for the future, as an alternative or to
-  complement the current CVMFS-based implementation.
-\end{itemize}
-
-\hypertarget{visualisation}{%
-\subsection{Visualisation}\label{visualisation}}
-
-\subsubsection*{Scope and Challenges}
-
-In modern High Energy Physics (HEP) experiments, visualisation of data
-has a key role in many activities and tasks across the whole data
-processing chain: detector development, monitoring, event generation,
-reconstruction, detector simulation, data analysis, as well as outreach
-and education.
-
-\emph{Event displays} are the main tool to explore experimental data at
-the event level and to visualise the detector itself. There are two main
-types of application: firstly, those integrated in the experiments'
-frameworks, which are able to access and visualise all the experiments'
-data, but at a cost in terms of complexity; secondly,
-those designed as cross-platform applications, lightweight and fast,
-delivering only a simplified version or a subset of the event data. In
-the first case, access to data is tied intimately to an experiment's
-data model (for both event and geometry data) and this inhibits
-portability; in the second, processing the experiment data into a
-generic format usually loses some detail and is an extra processing
-step. In addition, there are various graphical backends that can be used
-to visualise the final product, either standalone or within a browser,
-and these can have a substantial impact on the types of devices
-supported.
-
-Beyond event displays, HEP also uses \emph{visualisation of statistical
-information}, typically histograms, which allow the analyst to quickly
-characterise the data. Unlike event displays, these visualisations are
-not strongly linked to the detector geometry, and often aggregate data
-from multiple events. Other types of visualisation are used to
-\emph{display non-spatial data}, such as graphs for describing the
-logical structure of the detector or for illustrating dependencies
-between the data products of different reconstruction algorithms.
-
-The main challenges in this domain are in the sustainability of the many
-experi\-ment specific visualisation tools when common projects could
-reduce duplication and increase quality and long-term maintenance. The
-ingestion of events and other data could be eased by common formats,
-which would need to be defined and satisfy all users. Changes to support
-a client-server architecture would help broaden the ability to support
-new devices, such as mobile phones. Making a good choice for the
-libraries used to render 3D shapes is also key, impacting on the range
-of output devices that can be supported and the level of interaction
-with the user. Reacting to a fast-changing technology landscape is very
-important -- HEP's effort is limited and generic solutions can often be
-used with modest effort. This applies strongly to non-event
-visualisation, where many open source and industry standard tools can be
-exploited.
-
-\subsubsection*{Current Practices}
-
-Three key features characterise almost all HEP event displays:
-
-\begin{itemize}
-\item
-  \textbf{Event-based workflow}: applications access experimental data
-  on an event-by-event basis, visualising the data collections belonging
-  to a particular event. Data can be related to the actual physics
-  events (e.g., physics objects such as jets or tracks) or to the
-  experimental conditions (e.g., detector descriptions, calibrations).
-\item
-  \textbf{Geometry visualisation}: The application can display the
-  geometry of the detector, as retrieved from the experiments' software
-  frameworks, or a simplified description, usually for the sake of speed
-  or portability.
-\item
-  \textbf{Interactivity}: applications offer different interfaces and
-  tools to users, in order to interact with the visualisation itself,
-  select event data, and set cuts on objects' properties.
-\end{itemize}
-
-Experiments have often developed multiple event displays that either
-take the full integration approach explained above or are standalone and
-rely on extracted and simplified data.
-
-The visualisation of data can be achieved through the low level OpenGL
-API, by the use of higher-level OpenGL-based libraries, or within a web
-browser using WebGL. Using OpenGL directly is robust and avoids other dependencies, but
-implies a significant effort. Instead of using the API directly, a
-library layer on top of OpenGL (e.g., Coin3D) can more closely match the
-underlying data, such as geometry, and offers a higher level API that
-simplifies development. However, this carries the risk that if the
-library itself becomes deprecated, as has happened with Coin3D, the
-experiment needs to migrate to a different solution or to take on the
-maintenance burden itself. Standalone applications often use WebGL
-technology to render 3D objects inside a web browser. This is a very
-convenient way of rendering 3D graphics, due to the cross-platform
-nature of web technologies, and offers many portability advantages (e.g.,
-easier support for mobile or virtual reality devices), but at some cost
-of not supporting the most complex visualisations requiring heavy
-interaction with the experiments' data.
-
-In recent years, video game engines, such as Unity or the Unreal Engine,
-have become particularly popular in the game and architectural
-visualisation industry. They provide very sophisticated graphics engines
-and offer a lot of tools for user interaction, such as menu systems or
-native handling of VR devices. They are well supported by industry and
-tend to have a long lifespan (Unreal Engine is now  20 years old and
-is still very popular). However, such engines are meant to be used as
-development frameworks and their usage in HEP code is not always
-evident. Code should be developed within them, while in HEP
-framework-based applications we often want to use graphics libraries
-that can be integrated in existing code. A number of HEP collaborations
-have started experimenting in building event display tools with such
-engines, among them Belle II and ATLAS, but their use is currently
-limited to the display of simplified data only.
-
-The new client-server architecture proposed as one of the visualisation
-R\&D activities will ease the usage of WebGL technologies and game
-engines in HEP.
-
-For statistical data, ROOT has been the tool of choice in HEP for many
-years and satisfies most use cases. However, increasing use of generic
-tools and data formats means Matplotlib (Python) or JavaScript based
-solutions (used, for example, in Jupyter notebooks) have made the
-landscape more diverse. For visualising trees or graphs interactively,
-there are many generic offerings and experiments have
-started to take advantage of them.
-
-\subsubsection*{Research and Development Roadmap}
-
-The main goal of R\&D projects in this area will be to develop
-techniques and tools that let visualisation applications and event
-displays be less dependent on specific experiments' software frameworks,
-leveraging common packages and common data formats. Exporters and
-interface packages will be designed as bridges between the experiments'
-frameworks, needed to access data at a high level of detail, and the
-common packages based on the community standards that this group will
-develop.
-
-As part of this development work, demonstrators will be designed to show
-the usability of our community solutions and tools. The goal will be to
-get a final design of those tools so that the experiments can depend on
-them in their future developments.
-
-The working group will also work towards a more convenient access to geometry and
-event data, through a client-server interface. In collaboration with the
-Data Access and Management Working Group, an API or a service to deliver streamed
-event data would be designed.
-
-The work above should be completed by 2020.
-
-Beyond that point, the focus will be on developing the actual
-community-driven tools, to be used by the experiments for their
-visualisation needs in production, potentially taking advantage of new
-data access services.
-
-The workshop that was held as part of the CWP process was felt to be
-extremely useful for exchanging knowledge between developers in
-different experiments, fostering collaboration and in bringing in ideas
-from outside the community. This will now be held as an annual event and
-will facilitate work on the common R\&D plan.
-
-\hypertarget{software-development-deployment-validation-and-verification}{%
-\subsection{Software Development, Deployment, Validation and
-Verification}\label{software-development-deployment-validation-and-verification}}
-
-\subsubsection*{Scope and Challenges}
-
-Modern HEP experiments are often large distributed collaborations
-with several hundred people actively writing software. It is
-therefore vital that the processes and tools used for development are
-streamlined to ease the process of contributing code and to facilitate
-collaboration between geographically separated peers. At the same time,
-we must properly manage the whole project, ensuring code quality,
-reproducibility, and maintainability with the least effort possible.
-Making sure this happens is largely a continuous process and shares a
-lot with non-HEP specific software industries.
-
-Work is ongoing to track and promote solutions in the following areas:
-
-\begin{itemize}
-\item
-  Distributed development of software components, including the tools
-  and processes required to do so (code organisation, documentation,
-  issue tracking, artefact building), and the best practices in terms of
-  code and people management.
-\item
-  Software quality, including aspects such as modularity and reusability
-  of the developed components, architectural and performance best
-  practices.
-\item
-  Software sustainability, including both development and maintenance
-  efforts, as well as best practices given long timescales of HEP
-  experiments.
-\item
-  Deployment of software and interaction with operations teams.
-\item
-  Validation of the software both at small scales (e.g., best practices
-  on how to write a unit test) and larger ones (large scale validation
-  of data produced by an experiment).
-\item
-  Software licensing and distribution, including their impact on
-  software interoperability.
-\item
-  Recognition of the significant contribution that software makes to HEP
-  as a field (also see Section \ref{training-and-careers} regarding
-  career recognition).
-\end{itemize}
-
-HEP-specific challenges derive from the fact that HEP is a large,
-inhomogeneous community with multiple sources of funding, mostly formed
-of people belonging to university groups and HEP-focused laboratories.
-Software development effort within an experiment usually encompasses a
-huge range of experience and skills, from a few more or less full-time
-experts to many physicist programmers with little formal software
-training. In addition, the community is split between different
-experiments that often diverge in timescales, size, and resources.
-Experiment software is usually divided in two separate use cases:
-production (being it data acquisition, data reconstruction or
-simulation) and user analysis, whose requirements and lifecycles are
-completely different. The former is very carefully managed in a
-centralised and slow-moving manner, following the schedule of the
-experiment itself. The latter is much more dynamic and strongly coupled
-with conferences or article publication timelines. Finding solutions
-that adapt well to both cases is not always obvious or even possible.
-
-\subsubsection*{Current Practices}
-
-Due to significant variations between experiments at various stages of
-their lifecycles, there is a huge variation in practice across the
-community. Thus, here we describe \emph{best practice}, with the
-understanding that this ideal may be far from the reality for some
-developers.
-
-It is important that developers can focus on the design and
-implementation of the code and do not have to spend a lot of time on
-technical issues. Clear procedures and policies must exist to perform
-administrative tasks in an easy and quick way. This starts with the
-setup of the development environment. Supporting different platforms not
-only allows developers to use their machines directly for 
-development, it also provides a check of code portability. Clear
-guidance and support for good design must be available in advance of
-actual coding.
-
-To maximise productivity, it is very beneficial to use development tools
-that are not HEP-specific. There are many open source projects that are
-of similar scale to large experiment software stacks and standard tools
-are usually well documented. For source control HEP has generally chosen
-to move to \emph{git}~\cite{git},
-which is very welcome, as it also brings an
-alignment with many open source projects and commercial organisations.
-Likewise, \emph{CMake}~\cite{CMake} 
-is widely used for the builds of software
-packages, both within HEP and outside. Packaging many build products
-together into a software stack is an area that still requires close
-attention with respect to active developments (the HSF has an active
-working group here).
-
-Proper testing of changes to code should always be done in advance of a
-change request to be accepted. Continuous integration, where `merge' or
-`pull' requests are built and tested in advance, is now standard practice
-in the open source community and in industry. Continuous integration can
-run unit and integration tests, and can also incorporate code quality
-checks and policy checks that help improve the consistency and quality
-of the code at low human cost. Further validation on different platforms
-and at large scales must be as automated as possible, including the
-deployment of build artefacts for production.
-
-Training (Section \ref{training-and-careers}) and documentation 
-are key to efficient use
-of developer effort. Documentation must cover best practices and
-conventions as well as technical issues. For documentation that has to
-be specific, the best solutions have a low barrier of entry for new
-contributors, but also allow and encourage review of material.
-Consequently, it is very useful to host documentation sources in a
-repository with a similar workflow to code, and to use an engine that
-translates the sources into modern web pages.
-
-Recognition of software work as a key part of science has resulted in a
-number of journals where developers can publish their work~\cite{SSI2017}. Journal publication also disseminates information to the
-wider community in a permanent way and is the most established mechanism
-for academic recognition. Publication in such journals provides proper
-peer review, beyond that provided in conference papers, so it is
-valuable for recognition as well as dissemination. However, this
-practice is not widespread enough in the community and needs further
-encouragement.
-
-\subsubsection*{Research and Development Programme}
-
-HEP must endeavour to be as responsive as possible to developments
-outside of our field. In terms of hardware and software tools, there
-remains great uncertainty as to what the platforms offering the best
-value for money will be on the timescale of a decade. It therefore
-behoves us to be as generic as possible in our technology choices,
-retaining the necessary agility to adapt to this uncertain future.
-
-Our vision is characterised by HEP being current with technologies and
-para\-digms that are dominant in the wider software development community,
-especially for open-source software, which we believe to be the right
-model for our community. In order to achieve that aim, we propose that
-the community establishes a development forum that allows for technology
-tracking and discussion of new opportunities. The HSF can play a key
-role in marshalling this group and in ensuring its findings are widely
-disseminated. In addition, having wider and more accessible training for
-developers in the field, that will teach the core skills needed for
-effective software development, would be of great benefit.
-
-Given our agile focus, it is better to propose here projects and
-objectives to be investigated in the short to medium term, alongside
-establishing the means to continually review and refocus the community
-on the most promising areas. The main idea is to investigate new tools
-as demonstrator projects where clear metrics for success in a reasonable
-time should be established to avoid wasting community effort on
-initially promising products that fail to live up to expectations.
-
-Ongoing activities and short-term projects, include the following:
-
-\begin{itemize}
-\item
-  Establish a common forum for the discussion of HEP software problems.
-  This should be modeled along the lines of the Concurrency Forum~\cite{ConcurrencyForum}, which was very successful in establishing
-  demonstrators and prototypes that were used as experiments started to
-  develop parallel data processing frameworks.
-\item
-  Continue the HSF working group on \emph{Packaging}, with more
-  prototype implementations based on the strongest candidates identified
-  so far.
-\item
-  Provide practical advice on how to best set up new software packages,
-  developing on the current project template work, and working to
-  advertise this within the community.
-\item
-  Work with HEP experiments and other training projects to provide
-  accessible core skills training to the community (see Section \ref{training-and-careers}). This
-  training should be experiment-neutral, but could be usefully combined
-  with the current experiment specific training. Specifically, this work
-  can build on, and collaborate with, recent highly successful
-  initiatives such as the LHCb \emph{Starterkit}~\cite{LHCbStarterkit}
-  and ALICE \emph{Juniors}~\cite{ALICEJuniors}, and with established
-  generic training initiatives such as \emph{Software Carpentry}~\cite{SoftwareCarpentry}.
-\item
-  Strengthen links with software communities and conferences outside of
-  the HEP domain, presenting papers on the HEP experience and problem
-  domain. The Scientific Computing with Python (SciPy), the
-  Supercomputing Conferences (SCxx), the Conference of
-  Research Software Engineers \emph{(RSE)}, and the Workshops on
-  Sustainable Software for Science: Practice and Experiences (WSSSPE)
-  would all be useful meetings to consider.
-\item
-  Write a paper that looks at case studies of successful and
-  unsuccessful HEP software developments and that draws specific
-  conclusions and advice for future projects.
-\item
-  Strengthen the publication record for important HEP software packages.
-  Both peer-reviewed journals~\cite{SSI2017} and citable software version
-  records (such as DOIs obtained via Zenodo~\cite{Zenodo}).
-\end{itemize}
-
-Medium term projects include the following:
-
-\begin{itemize}
-\item
-  Prototype C++ refactoring tools, with specific use cases in migrating
-  HEP code.
-\item
-  Prototyping of portable solutions for exploiting modern vector
-  hardware on heterogenous platforms.
-\item
-  Support the adoption of industry standards and solutions over
-  HEP-specific implementations whenever possible.
-\item
-  Develop tooling and instrumentation to measure software performance
-  where tools with sufficient capabilities are not available from
-  industry, especially in the domain of concurrency. This should
-  primarily aim to further developments of existing tools, such as
-  \emph{igprof}~\cite{Eulisse:2005zz},
-  rather than to develop new ones.
-\item
-  Develop a common infrastructure to gather and analyse data about
-  experiments' software, including profiling information and code
-  metrics, and to ease sharing across different user communities.
-\item
-  Undertake a feasibility study of a common toolkit for statistical
-  analysis that would be of use in regression testing for experiment's
-  simulation and reconstruction software.
-\end{itemize}
-
-\hypertarget{data-and-software-preservation}{%
-\subsection{Data and Software
-Preservation}\label{data-and-software-preservation}}
-
-\subsubsection*{Scope and Challenges}
-
-Given the very large investment in particle physics experiments, it is
-incumbent upon physicists to preserve the data and the knowledge that
-leads to scientific results in a manner such that this investment is not
-lost to future generations of scientists. For preserving ``data'', at
-whatever stage of production, many of the aspects of the low level
-bit-wise preservation have been covered by the Data Preservation for HEP
-group~\cite{DPHEP}. ``Knowledge'' preservation encompasses the more
-challenging aspects of retaining processing and analysis software,
-documentation, and other components necessary for reusing a given
-dataset. Preservation of this type can enable new analyses on older
-data, as well as a way to revisit the details of a result after
-publication. The latter can be especially important in resolving
-conflicts between published results, applying new theoretical
-assumptions, evaluating different theoretical models, or tuning new
-modeling techniques.
-
-Preservation enabling reuse can offer tangible benefits within a given
-experiment. The preservation of software and workflows such that they
-can be shared enhances collaborative work between analysts and analysis
-groups, providing a way of capturing the knowledge behind a given
-analysis during the review process. It enables easy transfer of knowledge
-to new students or analysis teams, and could establish a manner by which
-results can be generated automatically for submission to central
-repositories, such as HEPData~\cite{Maguire:2017ypu}.
-Preservation within an experiment can
-provide ways of reprocessing and reanalysing data that could have been
-collected more than a decade earlier. Benefits from preservation are
-derived internally whether or not analysis work is approved through the
-publication approval process for an experiment. Providing such immediate
-benefits makes the adoption of data preservation in experiment workflows
-particularly desirable.
-
-A final series of motivations comes from the potential re-use by others
-outside of the HEP experimental community. Significant outreach efforts
-to bring the excitement of analysis and discovery to younger students
-have been enabled by the preservation of experimental data and software
-in an accessible format. Many examples also exist of phenomenology
-papers reinterpreting the results of a particular analysis in a new
-context. This has been extended further with published results based on
-the reanalysis of processed data by scientists outside of the
-collaborations. Engagement of external communities, such as machine
-learning specialists, can be enhanced by providing the capability to
-process and understand low-level HEP data in portable and relatively
-platform-independent way, as happened with the Kaggle ML
-challenges~\cite{pmlr-v42-cowa14}. 
-This allows external users direct access to the same tools
-and data as the experimentalists working in the collaborations.
-Connections with industrial partners, such as those fostered by CERN
-OpenLab, can be facilitated in a similar manner.
-
-Preserving the knowledge of analysis, given the extremely wide scope of
-how analysts do their work and experiments manage their workflows, is
-far from easy. The level of reuse that is applicable needs to be
-identified, and so a variety of preservation systems will probably be
-appropriate given the different preservation needs between large central
-experiment workflows and the work of an individual analyst. The larger
-question is to what extent common low-level tools can be provided that
-address similar needs across a wide scale of preservation problems.
-These would range from capture tools, that preserve the details of an
-analysis and its requirements, to ensuring that software and services
-needed for a workflow would continue to function as required.
-
-The above-mentioned steps can be considered to be consistent with the
-FAIR data principles that are increasingly being mandated by funding
-agencies~\cite{FAIRdata}.
-
-\subsubsection*{Current Practices}
-
-Each of the LHC experiments has adopted a data access and/or data
-preservation policy, all of which can be found on the CERN Open Data
-Portal~\cite{ODP}. All of the LHC experiments support public access to
-some subset of the data in a highly reduced data format for the purposes
-of outreach and education. CMS has gone one step further, releasing
-substantial datasets in an Analysis Object Data (AOD) format that can be
-used for new analyses. The current data release includes simulated data,
-virtual machines that can instantiate the added analysis examples, and
-extensive documentation~\cite{CMS-OpenData}. ALICE has promised to
-release 10\% of their processed data after a five-year embargo and has
-released 2010 data at this time~\cite{ALICE-OpenData}.
-LHCb is willing to make access to reconstructed data available, but is
-unable to commit to a specific timescale due to resource limitations. A
-release of ntuple-level data for one high profile analysis, aimed
-primarily at educational activities, is currently in preparation. ATLAS
-has chosen a different direction for data release: data associated with
-journal publications is made available, and ATLAS also strives to make
-available additional material that allows reuse and reinterpretations of
-the data in the context of new theoretical models~\cite{ATL-CB-PUB-2015-001}.
-ATLAS is  exploring how
-to provide the capability for reinterpretation of searches in the future
-via a service such as 
-RECAST~\cite{Cranmer:1299950}, in which the original internal analysis
-code (including full detector simulation and reconstruction) is
-preserved, as opposed to the re-coding approach with object-efficiency
-calibrations used by external reinterpretation toolkits. All experiments
-frequently provide detailed supplemental data along with publications to
-allow for more detailed comparisons between results, or even
-reinterpretation.
-
-The LHC experiments have not yet set a formal policy addressing the new
-capabilities of the CERN Analysis Preservation Portal (CAP)~\cite{CAP}
-and whether or not some use of it will be required or merely encouraged.
-All of them support some mechanisms for internal preservation of the
-knowledge surrounding a physics publication~\cite{Shiers2017}.
-
-\subsubsection*{Research and Development Programme}
-
-There is a significant programme of work already happening in the data
-preservation area. The feasibility and cost of common base services have
-been studied for bit preservation, the preservation of executable
-software environments, and the structured capturing of analysis metadata~\cite{Berghaus2016}.
-
-The goals presented here should be orchestrated in conjunction with
-projects conducted by the R\&D programmes of other working groups, since
-the questions addressed are common. Goals to address on the timescale of
-2020 are:
-
-\begin{itemize}
-\item
-  Include embedded elements for the capture of preservation information
-  and metadata and tools for the archiving of this information in
-  developing a prototype analysis ecosystem(s). This should include an
-  early demonstration of the CAP analysis preservation portal with a
-  working UI.
-\item
-  Demonstrate the capability to provision and execute production
-  workflows for experiments that are composed of multiple independent
-  containers.
-\item
-  Collection of analysis use cases and elements that are necessary to
-  preserve in order to enable re-use and to ensure these analyses can be
-  captured in developing systems. This should track analysis evolution
-  towards possible \emph{Big Data} environments and determine any elements
-  that are difficult to capture, spawning further R\&D.
-\item
-  Evaluate, in the preservation area, the full potential and limitations
-  of sandbox and ``freezing'' technologies, possibly coupled with
-  version and history control software distribution systems.
-\item
-  Develop prototypes for the preservation and validation of large-scale
-  production executables and workflows.
-\item
-  Integrate preservation capabilities into newly developed computing
-  tools and workflows.
-\item
-  Extension and standardisation of the \emph{final data and analysis}
-  preservation sche\-me via HEPData, Rivet and/or other reinterpretation
-  tools. This could be used to preserve a sufficiently detailed
-  re-usable record of many LHC Run 2 research outputs.
-\end{itemize}
-
-This would then lead naturally to deployed solutions that support data
-preservation in the 2020-2022 time frame for the HEP experimental
-programmes, in particular an analysis ecosystem that enables reuse for
-any analysis that can be conducted in the ecosystem, and a system for
-the preservation and validation of large-scale production workflows.
-
-\hypertarget{security}{%
-\subsection{Security}\label{security}}
-
-\subsubsection*{Scope and Challenges}
-
-Security is a cross-cutting area that impacts our projects,
-collaborative work, users, and software infrastructure fundamentally. It
-crucially shapes our reputation, our collaboration, the trust between
-participants, and the users' perception of the quality and ease of use
-of our services.
-
-There are three key areas:
-
-\begin{itemize}
-\item
-  Trust and policies; this includes trust models, policies, compliance,
-  data protection issues.
-\item
-  Operational security; this includes threat intelligence, security
-  operations, incident response.
-\item
-  Authentication and Authorisation; this includes identity management,
-  identity federation, access control.
-\end{itemize}
-
-\paragraph{Trust and Policies}
-
-Data Protection defines the boundaries that enable HEP work to be
-conducted, in particular regarding data sharing aspects, for example
-between the EU and the US. It is essential to establish a trusted
-personal data exchange framework, minimising the amount of personal data
-to be processed and ensuring legal compliance.
-
-Beyond legal compliance and best practice, offering open access to
-scientific resources and achieving shared goals requires prioritising
-the protection of people and science, including the mitigation of the
-effects of surveillance programs on scientific collaborations.
-
-On the technical side, it is necessary to adapt the current, aging trust
-model and security architecture relying solely on X.509 (which is no longer
-the direction industry is taking), in order to include modern data
-exchange design, for example involving commercial providers or hybrid
-clouds. The future of our infrastructure involves increasingly diverse
-resource providers connected through cloud gateways. For example,
-HEPCloud~\cite{HEPCloud} at FNAL aims to connect Amazon,
-Google Clouds, and HPC centres with our traditional grid computing
-resources. The HNSciCloud European Project~\cite{HNSciCloud} aims to
-support the enhancement of commercial cloud providers in order to be
-leveraged by the scientific community. These are just two out of a
-number of endeavours. As part of this modernisation, a transition is
-needed from a model in which all participating organisations are bound
-by custom HEP security policies to a more flexible approach where some
-partners are not in a position to adopt such policies.
-
-\paragraph{Operational Security and Threat Intelligence}
-
-As attacks have become extremely sophisticated and costly to defend
-against, the only cost-effective strategy is to address security threats
-together, as a community. This involves constantly striving to liaise
-with external organisations, including security vendors and law
-enforcement entities, to enable the sharing of indicators of compromise
-and threat intelligence between all actors. For organisations from all
-sectors, including private companies, governments, and academia, threat
-intelligence has become the main means by which to detect and manage
-security breaches.
-
-In addition, a global forum for HEP and the larger Research and Education
-(R\&E) community needs to be built, where security experts feel
-confident enough to share threat intelligence and security expertise. A
-key to success is to ensure a closer collaboration between HEP security
-contacts and campus security. The current gap at many HEP organisations
-is both undermining the community's security posture and reducing the
-effectiveness of the HEP security strategy.
-
-There are several very active trust groups in the HEP community where
-HEP participants share threat intelligence and organise coordinated
-incident response \cite{EGI-CSIRT,REN-ISAC,XSEDE}.
-There is unfortunately still no global Research and Education forum for
-incident response, operational security, and threat intelligence
-sharing. With its mature security operations and dense, global network
-of HEP organisations, both of which are quite unique in the research
-sector, the HEP community is ideally positioned to contribute to such a
-forum and to benefit from the resulting threat intelligence, as it has
-exposure, sufficient expertise, and connections to lead such an
-initiative. It may play a key role in protecting multiple scientific
-domains at a very limited cost.
-
-There will be many technology evolutions as we start to take a serious
-look at the next generation internet. For example, IPv6 is one upcoming
-change that has yet to be fully understood from the security
-perspective. Another high impact area is the internet of things (IoT),
-connected devices on our networks that create new vectors of attack.
-
-It will become necessary to evaluate and maintain operational security
-in connected environments spanning public, private, and hybrid clouds.
-The trust relationship between our community and such providers has yet
-to be determined, including the allocation of responsibility for
-coordinating and performing vulnerability management and incident
-response. Incompatibilities between the e-Infrastructure approach to
-community-based incident response and the ``pay-for-what-you-break''
-model of certain commercial companies may come to light and must be
-resolved.
-
-\paragraph{Authentication and Authorisation Infrastructure}
-
-It is now largely acknowledged that end-user certificates are
-challenging to manage and create a certain entrance barrier to our
-infrastructure for early career researchers. Integrating our access
-control management system with new, user-friendly technologies and
-removing our dependency on X.509 certificates is a key area of interest
-for the HEP Community.
-
-An initial step is to identify other technologies that can satisfy
-traceability, isolation, privilege management and other requirements
-necessary for HEP workflows. The chosen solution should prioritise
-limiting the amount of change required to our services and follow
-accepted standards to ease integration with external entities, such as
-commercial clouds and HPC centres.
-
-Trust federations and inter-federations, such as the R\&E standard
-eduGAIN \cite{eduGAIN}, provide a needed functionality for
-Authentication. They can remove the burden of identity provisioning from
-our community and allow users to leverage their home organisation
-credentials to access distributed computing resources. Although certain
-web-based services have enabled authentication via such federations,
-uptake is not yet widespread. The challenge remains to have the
-necessary attributes published by each federation to provide robust
-authentication.
-
-The existing technologies leveraged by identity federations, e.g., the
-Security Assertion Markup Language (SAML), have not supported non-web
-applications historically. There is momentum within the wider community
-to develop next-generation identity federations that natively support a
-wider range of clients. In the meantime there are several viable interim
-solutions that are able to provision users with the token required to
-access a service (such as X.509) transparently, translated from their
-home organisation identity.
-
-Although federated identity provides a potential solution for our
-challenges in Authentication, Authorisation should continue to be
-tightly controlled by the HEP community. Enabling Virtual Organisation
-(VO) membership for federated credentials and integrating such a
-workflow with existing identity vetting processes is a major topic
-currently being worked on, in particular within the WLCG community.
-Commercial clouds and HPC centres have fundamentally different access
-control models and technologies from our grid environment. We shall need
-to enhance our access control model to ensure compatibility and
-translate our grid-based identity attributes into those consumable by
-such services.
-
-\subsubsection*{Current Activities}
-
-Multiple groups are working on policies and establishing a common trust
-framework, including the EGI Security Policy Group
-\cite{EGISecurityPolicyGroup} and the Security for Collaboration among
-Infrastructures working group~\cite{SCI_WG}.
-
-Operational security for the HEP community is being followed up in the
-WLCG Working Group on Security Operations Centres~\cite{WLCG-SOC-WG}. The
-HEP Community is actively involved in multiple operational security
-groups and trust groups, facilitating the exchange of threat
-intelligence and incident response communication. WISE~\cite{WISE}
-provides a forum for e-Infrastructures to share and develop security
-best practices and offers the opportunity to build relationships between
-security representatives at multiple e-infrastructures of interest to
-the HEP community.
-
-The evolution of Authentication and Authorisation is being evaluated in
-the recently created WLCG Working Group on Authorisation. 
-In parallel, HEP is contributing to a wider effort
-to document requirements for multiple Research Communities through the
-work of FIM4R~\cite{FIM4R}. CERN's participation in the European
-Authentication and Authorisation for Research and Collaboration (AARC)
-project~\cite{AARC} provides the opportunity to ensure
-that any directions chosen are consistent with those taken by the wider
-community of research collaborations. The flow of attributes between
-federated entities continues to be problematic, disrupting the
-authentication flow. Trust between service providers and identity
-providers is still evolving, and efforts within the R\&E Federations
-Group (REFEDS)~\cite{REFEDS} and the AARC project aim to address the
-visibility of both the level of assurance of identities and the security
-capability of federation participants (through Sirtfi~\cite{Sirtfi}).
-
-\subsubsection*{Research and Development Programme}
-
-Over the next decade, it is expected that considerable changes will be
-made to address security in the domains highlighted above. The
-individual groups, in particular those mentioned above, working in the
-areas of trust and policies, operational security, authentication and
-authorisation, and technology evolutions, are driving the R\&D
-activities. The list below summarises the most important actions:
-
-\paragraph{Trust and Policies}
-\begin{itemize}
-\item 
-By 2020:
-\begin{itemize}
-    \item
-      Define and adopt policies in line with new EU Data Protection
-      requirements.
-    \item
-      Develop frameworks to ensure trustworthy interoperability of
-      infrastructures and communities.
-\end{itemize}
-
-\item
-By 2022:
-\begin{itemize}
-    \item
-      Create and promote community driven incident response policies and
-      procedures.
-\end{itemize}
-\end{itemize}
-
-\paragraph{Operational Security and threat intelligence}
-\begin{itemize}
-\item
-By 2020:
-\begin{itemize}
-\item
-  Offer a reference implementation, or at least specific guidance, for a
-  Security Operation Centre deployment at HEP sites, enabling them to
-  take action based on threat intelligence shared within the HEP
-  community.
-\end{itemize}
-\item
-By 2022:
-\begin{itemize}
-\item
-  Participate in the founding of a global Research and Education Forum
-  for incident response, since responding as a global community is the only
-  effective solution against global security threats.
-\item
-  Build the capabilities to accommodate more participating organisations
-  and streamline communication workflows, within and outside HEP,
-  including maintaining a list of security contacts, secure communications
-  channels, and security incident response mechanisms.
-\item
-  Reinforce the integration of HEP security capabilities with their
-  respective home organisation, to ensure adequate integration of HEP
-  security teams and site security teams.
-\end{itemize}
-\item 
-By 2025:
-\begin{itemize}
-\item
-  Prepare adequately as a community, in order to enable HEP
-  organisations to operate defendable services against more
-  sophisticated threats, stemming both from global cyber-criminal gangs
-  targeting HEP resources (finance systems, intellectual property,
-  ransomware), as well as from state actors targeting the energy and
-  research sectors with advanced malware.
-\end{itemize}
-\end{itemize}
-
-\paragraph{Authentication and Authorisation}
-
-\begin{itemize}
-\item
-By 2020:
-\begin{itemize}
-\item
-  Ensure that ongoing efforts in trust frameworks are sufficient to
-  raise the level of confidence in federated identities to the
-  equivalent of X.509, at which stage they could be a viable alternative
-  to both grid certificates and CERN accounts.
-\item
-  Participate in setting directions for the future of identity
-  federations, through the FIM4R~\cite{FIM4R} community.
-\end{itemize}
-\item
-By 2022:
-\begin{itemize}
-\item
-  Overhaul the current Authentication and Authorisation infrastructure,
-  including Token Translation, integration with Community IdP-SP
-  Proxies, and Membership Management tools. Enhancements in this area
-  are needed to support a wider range of user identities for WLCG
-  services.
-\end{itemize}
-\end{itemize}
-
-\hypertarget{training-and-careers}{%
-\section{Training and Careers}\label{training-and-careers}}
-
-For HEP computing to be as successful as possible, the careers and
-skills of the individuals who participate must be considered. Ensuring
-that software developers can acquire the necessary skills and obtain
-successful careers is considered an essential goal of the HSF, which has
-the following specific objectives in its mission:
-
-\begin{itemize}
-\item
-  To provide training opportunities for developers; this should include
-  the support to the software schools for young scientists and computer
-  engineers, and of a permanent training infrastructure for accomplished
-  developers;
-\item
-  To provide career support for developers, for instance by listing job
-  opportunities and by helping to shape well-defined career paths that
-  provide advancement opportunities on a par with those in, for example,
-  detector construction;
-\item
-  To increase the visibility of the value of software developers in HEP,
-  recognising that it has scientific research value on an equal footing
-  with other activities, and acknowledging and promoting specific
-  ``champions'' in the field.
-\end{itemize}
-
-\subsection{Training Challenges}
-
-HEP is facing major challenges with its software and computing that
-require innovative solutions based on the proper adoption of new
-technologies. More and more technologies are emerging as scientific
-communities and industry face similar challenges and produce solutions
-relevant to us. Integrating such technologies in our software and
-computing infrastructure requires specialists, but it is also important that a
-large fraction of the community is able to use these new tools and
-paradigms. Specific solutions and optimisations must be implemented by
-the HEP community itself, since many advanced requirements are unique to
-our field.
-
-Unlike the situation that is traditional in some other fields, in
-which users express their requirements and computer specialists
-implement solutions, there is a close collaboration, even overlap,
-in HEP between users and developers that is essential for our success.
-Many details of experiment data
-cannot be known before data taking has started, and each change in
-detector technology or machine performance improvement can have
-important consequences for the software and computing infrastructure. In
-the case of detectors, engineers and physicists are required to have a
-good understanding of each other's field of expertise. In the same way,
-it is necessary that physicists understand some of the complexities of
-writing software, and that software experts are able to fathom the
-requirements of physics problems.
-
-Training must address an audience with very diverse computing skills,
-ranging from novice programmers to advanced developers and users.
-It must be used to spread best software engineering practices and
-software technologies to a very large number of people, including the
-physicists involved across the whole spectrum of data processing tasks,
-from triggering to analysis. It must be done by people who have a sound
-knowledge of the scientific and technical details, who prepare training
-material despite the many calls on their time. Training thus needs
-proper recognition to ensure that it happens and is carried out well.
-
-HEP is seen as an interesting, innovative, and challenging field. 
-This is a great advantage in attracting talented young people looking
-for experience in a challenging and diverse environment in which
-they can acquire skills that will be valuable, even in other fields.
-As discussed in Software Development (Section
-\ref{software-development-deployment-validation-and-verification}),
-using industry standard tools across different experiments, 
-and training people in how to use them properly,
-helps with people's later career prospects
-and makes our field even more attractive.
-At the same time, experiments have a scientific programme to accomplish
-and also to focus on the specific training required to
-accomplish their specific goals. The right balance must be found between
-these two requirements. It is necessary to find the right incentives to
-favour training activities that bring more benefits in the medium to
-long term, for the experiment, the community, and the careers of the
-trainees.
-
-\subsection{Possible Directions for Training}
-
-To increase training activities in the community, whilst taking into
-account the constraints of both the attendees and the trainers, we should
-explore new approaches to training. The current ``school''
-model is well established, as exemplified by three well-known successful
-schools, the CERN School of Computing~\cite{CSC}, 
-the Bertinoro School of Computing~\cite{ESC2017}
-and the GridKa School of Computing~\cite{GridKASchool}.
-They require a significant amount of
-dedicated time of all the participants, at the same time and location,
-and therefore are difficult to scale to meet the needs of a large number
-of students. In view of this, we should identify opportunities to work
-with HEP experiments and other training projects to provide accessible
-core skills training to the community by basing them at laboratories
-where students can easily travel. A number of highly successful
-experiment-specific examples exist, such as the LHCb StarterKit~\cite{LHCbStarterkit} 
-and ALICE Juniors~\cite{ALICEJuniors}, as well as
-established generic training initiatives, such as 
-Software Carpentry~\cite{SoftwareCarpentry}. As with hands-on tutorials organised during
-conferences and workshops, the resulting networking is an important and
-distinctive benefit of these events, where people build relationships
-with other colleagues and experts.
-
-In recent years, several R\&D projects, such as DIANA-HEP~\cite{DIANA-HEP} and
-AMVA4\-New\-Physics~\cite{AMVA4NewPhysics}, have had training as one 
-of their core activities. This
-has provided an incentive to organise training events and has resulted
-in the spread of expertise on advanced topics. We believe that training
-should become an integral part of future major R\&D projects.
-
-New pedagogical methods, such as active training and peer training, that
-are complementary to schools or topical tutorials, also deserve more
-attention. Online material can be shared by a student and a teacher to
-provide the exchange of real examples and practical exercises. For
-example, notebook technologies, such as Jupyter, support embedding of
-runnable code and comments into the same document. The initial material
-can be easily enriched by allowing other students and experts to add
-comments and more examples in a collaborative way. The HSF started to
-experiment with this approach with WikiToLearn~\cite{WikiToLearn}, a
-platform developed in Italy outside HEP that promotes this kind of
-training and collaborative enrichment of the training material. Projects
-such as ROOT~\cite{Brun1996} have also started to provide some
-training material based
-on notebooks.
-
-A lot of initiatives have been undertaken by the software community that
-HEP can benefit from, and materials have been made available in the form
-of online tutorials, active training, and Massive Open Online Courses
-(MOOCs). Some effort needs to be invested to evaluate existing courses
-and build a repository of selected ones that are appropriate to HEP
-needs. This is not a negligible task and would require some dedicated
-effort to reach the appropriate level of support. It should help to
-increase training efficiency by making it easier to identify appropriate
-courses or initiatives.
-
-A model that emerged in recent years as a very valuable means of sharing
-expertise is to use Question and Answer (Q\&A) systems, such as Stack
-Overflow. A few such systems are run by experiments for their own needs,
-but this is not necessarily optimal, as the value of these services is
-increased by a large number of contributors with diverse backgrounds.
-Running a cross-experiment Q\&A system has been discussed, but it has
-not yet been possible to converge on a viable approach, both technically
-and because of the effort required to run and support such a service.
-
-\subsection{Career Support and Recognition}
-
-Computer specialists in HEP are often physicists who have chosen to
-specialise in computing. This has always been the case and needs to
-continue. Nevertheless, for young people in particular, this leads to a
-career recognition problem, as software and computing activities are not
-well-recognised roles in various institutions supporting HEP research
-and recruiting people working in the field. The exact situation is
-highly dependent on policies and boundary conditions of the organisation
-or country, but recognition of physicists tends to be based generally on
-participation in data analysis or hardware developments. This is even a
-bigger problem if the person is spending time contributing to training
-efforts. This negatively impacts the future of these people and reduces
-the possibility of HEP engaging them in the training effort of the
-community when the community actually needs more people to participate
-in this activity. Recognition of training efforts, either by direct
-participation in training activities or by providing materials, is an
-important issue to address, complementary to the incentives mentioned
-above.
-
-There is no easy solution to this problem. Part of the difficulty is
-that organisations, and in particular the people inside them in charge
-of the candidate selections for new positions and promotions, need to
-adapt their expectations to these needs and to the importance of having
-computing experts with a strong physics background as permanent members
-of the community. Experts writing properly engineered and optimised
-software can significantly reduce resource consumption and increase
-physics reach, which provides huge financial value to modern HEP
-experiments. The actual path for improvements in career recognition, as
-the possible incentives for participating in the training efforts,
-depends on the local conditions.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Future challenges for High Energy Physics in the domain of software and
-computing are not simply an extrapolation of the challenges faced today.
-The needs of the HEP programme in the high luminosity era far exceed those
-that can be met by simply making incremental changes to today's code and
-scaling up computing facilities within the anticipated budget. At the
-same time, the limitation in single core CPU performance is making the
-landscape of computing hardware far more diverse and challenging to
-exploit, whilst offering huge performance boosts for suitable code.
-Exploiting parallelism and other new techniques, such as modern machine
-learning, offer great promise, but will require substantial work from
-the community to adapt to our problems. If there were any lingering
-notion that software or computing could be done cheaply by a few junior
-people for modern experimental programmes, it should now be thoroughly
-dispelled.
-
-We believe HEP Software and Computing requires a step change in its profile and
-effort to match the challenges ahead. We need investment in people who
-can understand the problems we face, the solutions employed today, and
-have the correct skills to provide innovative solutions for the future.
-There needs to be recognition from the whole community for the work done
-in this area, with a recognised career path for these experts. In
-addition, we will need to invest heavily in training for the whole
-software community as the contributions of the bulk of non-expert
-physicists are also vital for our success.
-
-We know that in any future scenario development effort will be
-constrained, so it is vital that successful R\&D projects provide
-sustainable software for the future. In many areas it is recognised 
-that different experiments could have adopted common solutions, 
-reducing overall development effort and increasing robustness and 
-functionality. That model of duplicated development is not sustainable. We must
-endeavour to achieve better coherence within HEP for future developments
-to build advanced, open-source projects that can be shared and supported
-in common. The HSF has already established itself as a forum that can
-facilitate this. Establishing links outside of HEP, to other academic
-disciplines, to industry, and to the computer science community, can
-strengthen both the research and production phases of new solutions. We
-should ensure that the best products are chosen, from inside and outside
-HEP, and that they receive support from all parties, aiming at technical
-excellence and economy of scale.
-
-We have presented programmes of work that the community has identified
-as being part of the roadmap for the future. While there is always some
-scope to reorient current effort in the field, we would highlight the
-following work programmes as being of the highest priority for
-investment to address the goals that were set in the introduction.
-
-\paragraph{Improvements in software efficiency, scalability and performance}
-
-\begin{quote}
-The bulk of CPU cycles consumed by experiments relate to the fundamental
-challenges of simulation and reconstruction. Thus, the work programmes
-in these areas, together with the frameworks that support them, are of
-critical importance. The sheer volumes of data involved make research
-into appropriate data formats and event content to reduce storage
-requirements vital. Optimisation of our distributed computing systems,
-including data and workload management, is paramount.
-\end{quote}
-
-\paragraph{Enable new approaches that can radically extend physics reach}
-
-\begin{quote}
-New techniques in simulation and reconstruction will be vital here.
-Phys\-ics analysis is an area where new ideas can be particularly
-fruitful. Exploring the full potential of machine learning is one common
-theme that underpins many new approaches and the community should
-endeavour to share knowledge widely across subdomains. New data analysis
-paradigms coming from the Big Data industry, based on innovative
-parallelised data processing on large computing farms, could transform
-data analysis.
-\end{quote}
-
-\paragraph{Ensure the long-term sustainability of the software}
-
-\begin{quote}
-Applying modern software development techniques to our codes has
-increased, and will continue to increase, developer productivity and
-code quality. There is ample scope for more common tools and common
-training to equip the community with the correct skills. Data
-Preservation makes sustainability an immediate goal of development and
-analysis and helps to reap the benefits of our experiments for decades
-to come. Support for common software used across the community needs to
-be recognised and accepted as a common task, borne by labs, institutes,
-experiments, and funding agencies.
-\end{quote}
-
-The R\&D actions proposed in this Roadmap have taken into account the
-charges that were laid down. When considering a specific project
-proposal addressing our computing challenges, that project's impact, measured
-against the charges, should be evaluated. Over the next decade, there
-will almost certainly be disruptive changes that cannot be planned for,
-and we must remain agile enough to adapt to these.
-
-The HEP community has many natural subdivisions, between different
-regional funding agencies, between universities and laboratories, and
-between different experiments. It was in an attempt to overcome these
-obstacles, and to encourage the community to work together in an
-efficient and effective way, that the HEP Software Foundation was
-established in 2014. This Community White Paper process has been
-possible only because of the success of that effort in bringing the
-community together. The need for more common developments in the future,
-as underlined here, reinforces the importance of the HSF as a common
-point of contact between all the parties involved, strengthening our
-community spirit and continuing to help share expertise and identify
-priorities. Even though this evolution will also require projects and
-experiments to define clear priorities about these common developments,
-we believe that the HSF, as a community effort, must be strongly
-supported as part of our roadmap to success.
-
-\newpage
 \begin{appendices}
@@
 
-Concurrency Forum: Software engineering is moving towards a paradigm shift in order to accommodate new CPU architectures with many cores, in which concurrency will play a more fundamental role in programming languages and libraries.
-The forum on concurrent programming models and frameworks aims to share knowledge among interested parties that work together to develop 'demonstrators' and agree on technology so that they can share code and compare results.
+\item[Concurrency Forum] Software engineering is moving towards a paradigm shift in order to accommodate new CPU
+architectures with many cores, in which concurrency will play a more fundamental role in programming languages
+and libraries. The forum on concurrent programming models and frameworks aims to share knowledge among interested
+parties that work together to develop 'demonstrators' and agree on technology so that they can share code and compare results.
 

### 2018-10-19 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
[WIP] Finalization of CWP version for CSBS journal (#104)
* Use one column for author list

* CWP: appendices as a separate file, one column in cwp-csbs

* Fix a non itemized entry in the glossary (Concurrency Forum)

@@
   to attract the required new expertise by offering appropriate career
-  recognition to physicists specialising in software development, and by
-  an effective training effort to target all contributors in the
+  recognition to physicists specialising in software development and by
+  an effective training effort to target all software contributors in the
   community.

### 2018-10-23 HSF-TN/draft-2017-VACPLAT/draft-HSF-TN-2017-VACPLAT.tex
DOI on TN title page

--- 2016-10-04 HSF-TN/draft-2016-VACPLAT/draft-HSF-TN-2016-VACPLAT.tex
@@
 \\
- & & HSF-TN-2016-VACPLAT \\  % ID 
- & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
+ & & HSF-TN-2016-04 \\  % ID 
+ & & \today \\October 4, 2016 \\ % Date - Can also hardwire e.g.: 23 March 2010
  & & \\
@@
 
-This technical note describes components of the Vacuum Platform developed by
+This technical note describes components of the Vacuum Platform 2.0 developed by
 GridPP for managing VMs, including the \texttt{\$JOBOUTPUTS}, VacQuery, and 
-VacUserData interfaces.
+VacUserData interfaces. This updates the 1.0 specification explained in
+HSF-TN-2016-04\cite{HSFVACPLAT1}.
 
@@
 
-This technical note describes components of GridPP's Vacuum Platform for 
+This technical note describes components of GridPP's Vacuum Platform 2.0 for 
 managing virtual machines (VMs) to run jobs for WLCG and other HEP 
-experiments.
+experiments. It updates HSF-TN-2016-04\cite{HSFVACPLAT1} which described verison 1.0.
 
@@
 \item[message\_type] ``machine\_status''
-\item[vac\_version] Name and software version of Vac
+\item[daemon\_version] Name and software version of the factory daemon
 \item[vacquery\_version] Name and version of the VacQuery protocol
@@
 \item[space] Vac space name
+\item[site] The site name, either as used by GOCDB or a string derived from the space name.
 \item[factory] FQDN of the factory
@@
 \item[heartbeat\_time] Unix time when the VM was last observed to be running (this is not the same as any heartbeat generated within the VM)
+\item[num\_processors] Number of logical or virtual processors assigned to this machine
 \item[cpu\_seconds] CPU seconds used by the VM
@@
 \item[shutdown\_time] Unix time of the shutdown\_message
+\item[fqan] Fully Qualified Attribute Name associate with the VM, typically identifying a VO and perhaps a group or role within it.
 \end{description}
@@
 \item[message\_type] ``machinetype\_status''
-\item[vac\_version] Name and software version of Vac
+\item[daemon\_version] Name and software version of the factory daemon
 \item[vacquery\_version] Name and version of the VacQuery protocol
@@
 \item[running\_machines] Number of running VMs for this machinetype on this factory
-\item[running\_cpus] Number of CPUs allocated to running VMs for this machinetype on this factory
+\item[running\_processors] Number of logical or virtual processors allocated to running VMs for this machinetype on this factory
 \item[num\_before\_fizzle] Number of running VMs which have not yet reached fizzle\_seconds
@@
 \item[shutdown\_machine] Name of the VM slot associated with the shutdown\_message
+\item[fqan] Fully Qualified Attribute Name associate with the VM, typically identifying a VO and perhaps a group or role within it.
 \end{description}
@@
 As VacQuery messages are sent as JSON documents, they may be conveniently 
-recorded in data stores such as ElasticSearch.
+recorded in NoSQL data stores such as ElasticSearch.
 
@@
   an uppercase name for the VMLM software.
-\item[Processors] The number of virtual CPUs assigned to the VM.
+\item[Processors] The number of logical or virtual processors assigned to the VM.
 \end{description}
@@
 
-%\section{GLUE2}
-%\label{sec:glue2}
-%
-%VMLMs should publish status information as JSON documents using the GLUE 2.0 
-%schema, at an HTTPS URL advertised in the GOCDB service entry for the space,
-%in the Grid Information / URL field.
-
-%\clearpage
-
 \section{Summary}
@@
 
+\bibitem{HSFVACPLAT1} A.~McNab, HSF-TN-2016-04 ``The Vacuum Platform'' (HEP Software Foundation)
+
 \end{thebibliography}

### 2018-10-29 CWP/papers/HSF-CWP-2017-15_visualization/latex/cwp-wg-visualization.tex
Moving the final version of the "Visualization" CWP, from the working group's repository to the central CWP repository.

% Copyright (C) 2018, The HSF Community White Paper authors, licence CC-BY-4.0.

% !TeX spellcheck = en_US

% JHEP preprint template
\documentclass[12pt,a4paper]{article}
\usepackage{jheppub}

\usepackage{graphicx}
%Path relative to the main .tex file
\graphicspath{ {./img/} }

\usepackage{subcaption}
\usepackage{xspace}
\usepackage{todonotes}
\usepackage{epigraph}
\usepackage{subcaption}
\usepackage{mwe}

\usepackage[utf8]{inputenc}

%%% BIBLIOGRAPHY SETTINGS %%%
\usepackage[style=numeric-comp,sorting=none]{biblatex}
%\addbibresource{../../roadmap/latex/cwp.bib}
%\addbibresource{../../roadmap/latex/cwp-chapters.bib} % HSF's one, for final submission
\addbibresource{cwp-chapters.bib} % local one, remove it for final submission
\addbibresource{cwp-wg-visualization-bib.bib}
%%%%%%

%% Comment out before final submission
\usepackage{lineno}  % for line numbering during review
\linenumbers

\abstract{In modern High Energy Physics (HEP) experiments visualization of experimental data has a key role in many activities
and tasks across the whole data chain: from detector development to monitoring, from event generation to reconstruction of physics objects,
from detector simulation to data analysis, and all the way to outreach and education.
In this paper the definition, status, and evolution of data visualization for HEP experiments will be presented.
Suggestions for the upgrade of data visualization tools and techniques in current experiments will be outlined, along with
guidelines for future experiments. This paper expands on the summary content published in the
HSF \emph{Roadmap} Community White Paper~\cite{HSF-CWP-2017-01}.}

\begin{document}

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
& & HSF-CWP-2017-15 \\
& & October 26, 2018 \\ % use \date or hardwire e.g. December 15, 2017
& & \\
\end{tabular*}
\vspace{2.0cm}

\title{HEP Software Foundation Community White Paper Working Group -- Visualization}

\author{HEP Software Foundation:}

\author[a,b]{Matthew Bellis}
\author[c,1]{Riccardo Maria Bianchi\note{Paper Editors}}%this same note will be shared with all the other authors with the [1] (i.e. Thomas McCauley)
\author[d]{Sebastien Binet}
\author[e]{Ciril Bohak}
\author[f]{Benjamin Couturier}
\author[g]{Hadrien Grasland}
\author[h]{Oliver Gutsche}
\author[i]{Sergey Linev}
\author[j]{Alex Martyniuk}
\author[k,1]{Thomas McCauley}
\author[l]{Edward Moyse}
\author[m]{Alja Mrak Tadel}
\author[n]{Mark Neubauer}
\author[f]{Jeremi Niedziela}
\author[p]{Leo Piilonen}
\author[q]{Jim Pivarski}
\author[r]{Martin Ritter}
\author[s]{Tai Sakuma}
\author[m]{Matevz Tadel}
\author[f]{Barthélémy von Haller}
\author[t]{Ilija Vukotic}
\author[j]{Ben Waugh}

\affiliation[a]{Siena College, Loudonville NY, USA}
\affiliation[b]{Cornell University, Ithaca NY, USA}
\affiliation[c]{University of Pittsburgh, Pittsburgh PA, USA}
\affiliation[d]{CNRS/IN2P3, Clermont-Ferrand, France}%{LSST, Alice, LPC}
\affiliation[e]{University of Ljubljana, Ljubljana, Slovenia}
\affiliation[f]{CERN, Geneva, Switzerland}
\affiliation[g]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\affiliation[h]{FNAL, Batavia IL, USA}
\affiliation[i]{GSI Darmstadt, Germany}
\affiliation[j]{University College London, London, UK}
\affiliation[k]{University of Notre Dame, Notre Dame IN, USA}
\affiliation[l]{University of Massachusetts, Amherst MA, USA}
\affiliation[m]{University of California at San Diego, San Diego CA, USA}
\affiliation[n]{University of Illinois, IL, USA}
\affiliation[p]{Virginia Tech, VA, USA}
\affiliation[q]{Princeton University, Princeton PA, USA}
\affiliation[r]{LMU Munich, Munich, Germany}
\affiliation[s]{University of Bristol, Bristol, UK}
\affiliation[t]{University of Chicago, Chicago IL, USA}

\maketitle

\newpage

\hypertarget{scope}{%
\section{Scope}\label{scope}}

\epigraph{Visualization: Turning numbers into pixels}{\textit{Hadrien Grasland \\ HSF Workshop 2017, Annecy}}

This paper will describe three kinds of data visualization used in High-Energy Physics (HEP): interactive visualization of event data in
applications known commonly as event displays, statistical data visualization such as histograms,
and non-spatial data visualization such as networks and graphs.

Event displays are the main tool used to explore experimental data at the event level.
There are two main types of displays. The first type are those that are integrated into an experiment's software frameworks,
which are usually able to access and visualize all experimental data at the cost of greater application complexity and lesser portability.
The second type of displays are those designed as cross-platform applications, lightweight and fast, often
delivering a simplified version or a subset of the event data. All event displays show the detector geometry; the level of detail
displayed depends on the application's use-case and targeted audience as well as on the application's capability to
render geometries responsively.

Beyond event displays, HEP also uses statistical data visualizations such as histograms, which display the distributions of data variables
in aggregate over multiple events. These visualizations are not strongly linked to a detector geometry. Data
analysis tools and techniques used in HEP are
described in the HSF \textit{Data Analysis and Interpretation} Community White Paper~\cite{HSF-CWP-2017-05}.

The final types of visualization considered in this paper are those that visualize non-spatial data, such as the graphs used to
visually describe the structure of the detector description, that is
the representation of all geometrical volumes that compose the sub-detectors and the infrastructure of a HEP experiment.
More details about the detector geometry can be found in the HSF \textit{Detector Simulation} Community White Paper~\cite{HSF-CWP-2017-07}.

Other types of data visualization used in HEP experiments, such as visualization for slow control
or dashboards for data analytics, are considered out of scope and are not discussed.

The content of this paper is the summary of the direct experience of the authors in designing, building, and deploying
interactive data visualization applications for the experiments ALICE, ATLAS, Belle II, CMS, LHCb, and LSST, for scientific
software frameworks such as ROOT, and in other cross-experiment projects. It is the outcome of a number of discussions and workshops organized
under the auspices of the HEP Software Foundation,
in which the authors worked to build a common, shared view of the field, its current issues, and the potential ways it could and should evolve
in the context of HEP.

\hypertarget{current-landscape}{%
\section{Current landscape}\label{current-landscape}}

\hypertarget{event-displays}{%
\subsection{Event displays}\label{event-displays}}

Three key features characterize HEP event displays. The first is an \emph{event-based workflow}. Applications access
experimental data on an event-by-event basis,
visualizing the data collections belonging to a particular event. Data can be related to the
actual physics events ({\it e.g.}\ a collection of reconstructed physics objects, like jets and tracks) or to the experimental
conditions ({\it e.g.}\ different versions of the detector description and calibration data).

The second key feature is \emph{geometry visualization}.
The level of geometric detail displayed depends on the specific use-case, on the way the geometry information is stored and
fetched ({\it e.g.}\ from a database as part of a software framework or from an external file), and on
limitations of the application itself along with
considerations about speed, efficiency, and portability.

The third key feature is \emph{interactivity}. Applications offer different interfaces and tools for users to interact
with the visualization, select event data, and to set cuts on objects' properties.
In addition to the interactive usage, applications often store different settings to automate user actions.

In the following subsections several important aspects of data access, application development and distribution, and geometry description
and visualization, as they pertain to the current landscape of event displays, are discussed in more detail. Screenshots of several
event displays can be seen in Figure~\ref{fig:screenshots}.

%[ATLASED042016]
%[ATLASED072015]
%[OPEN-PHO-EXP-2015-013]
%[ALICE-EVENTDISPLAY-2017-007]
%{other images from other experiments ???}
\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{ATLAS_VP1_event_display_first_stable_beam_collisions_13TeV_many-vertices_run296942_evt34013839_2016-04-23T10-51-30_v4.png}
		\caption[ATLAS 3D event display]%
		{{\small ATLAS 3D event display}}
		\label{fig:ed-atlas}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{VBF2e2mu_v10}
		\caption[CMS event display]%
		{{\small CMS event display}}
		\label{fig:ed-cms}
	\end{subfigure}
	\vskip\baselineskip
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{run280235_ev341_2}
		\caption[ALICE event display]%
		{{\small ALICE event display}}
		\label{fig:ed-alice}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{2018Start2}
		\caption[LHCb event display]%
		{{\small LHCb event display}}
		\label{fig:ed-lhcb}
	\end{subfigure}
	\vskip\baselineskip
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{belleII_VR}
		\caption[Belle II VR event display]%
		{{\small Belle II VR event display}}
		\label{fig:ed-belleii}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{HI-ZEE-151058-747-4096951-large}
		\caption[CMS event display]%
		{{\small CMS event display}}
		\label{fig:ed-cms-fireworks}
	\end{subfigure}
	\caption[Examples of HEP event displays]
	{\small Different examples of HEP 3D event display applications: \subref{fig:ed-atlas}) ATLAS 3D event display \cite{ATLASED072015,ATLASVP12010};
	\subref{fig:ed-cms}) CMS 3D event display \cite{CMSED}; 
	\subref{fig:ed-alice}) ALICE 3D event display \cite{ALICEED};
	\subref{fig:ed-lhcb}) LHCb 3D event display \cite{LHCbED}; \subref{fig:ed-belleii}) Belle II VR event display \cite{BelleIIVRED}; 
	\subref{fig:ed-cms-fireworks}) A CMS 3D event display made with Fireworks~\cite{CMSFireworks}}
	\label{fig:screenshots}
\end{figure*}

\hypertarget{data-access}{%
\subsubsection{Data access}\label{data-access}}

Access to event data comes either natively or via intermediate formats. In the former case direct access of native event formats
is only possible for an application integrated with the experimental software framework.

There are several advantages to having access to the experiment's framework, such as full access to the experimental
data in its native format and to software tools, services, and databases. Through them, event display applications can
make use of the full detector simulation geometry, of conditions data, and of all the framework's application program interface (API).

One disadvantage of this approach is that full support for the display application is often limited to those platforms
on which the framework itself is supported, limiting cross-platform distribution and support.
One way to mitigate this is to distribute a light version of the framework along with the application; CMS Fireworks~\cite{CMSFireworks}
takes this approach. However, issues of platform support for the light framework and for the visualization application can still exist.
A further disadvantage to the full-framework (and even light-framework) approach is that one must also support various versions of
the data format along with the underlying framework API. In addition, users have to have knowledge of the framework in order to interactively
explore and visualize event-based data. Lastly often the user-interface to a
full-framework application is geared towards the expert.

The latter approach to data access is via an intermediate format. Usually the data needed for visualization is a subset of the
full information found in the native experimental format. Therefore one can extract what is needed from the framework through
the usage of dedicated exporting software tools and store in intermediate formats.

With the use of an intermediate data format (usually based on different flavors of the XML or JSON formats) the event display
application is potentially separate from the experimental software framework and therefore its deployment is not limited to
those platforms officially supported by the experiment. It is then possible to have both lightweight data and applications,
which can be easily and broadly distributed. %even for outreach and educational activities for the general public.

The primary drawback to this approach is that, with no direct access to the experimental data, some
information is necessarily not accessible. Moreover, every time there is the need to modify the content of the
intermediate data files, one needs to run the data extraction tools on the native data again.
In addition to that, only events which are identified as potentially interesting are extracted and their
data reduced to be stored in the intermediate data file; so, if the end user wants to analyze and visualize other
potentially interesting events, the extraction/reduction step must be performed again on the relevant data.

Regardless of the approach to data access one must consider the use-case: what is useful or necessary
in one use-case may not be for another. A graphical representation of the two approaches can be
seen in Figure~\ref{fig:eventdisplaysapplicationstypes}.

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{img/eventDisplays_applications_types}
	\caption[Different types of event display applications]{The two different types of event display applications. At the top,
	the framework-integrated event display application is able to access all experimental data, all services, APIs, and databases
	provided by the experiment's framework; as a drawback, the application must be run on specific platforms supported by the
	framework and it must use only graphics and GUI libraries compatible with them. At the bottom, the standalone approach, where
	experimental data are accessed, filtered, and extracted by using custom data exporters, which create intermediate data files
	containing only the interesting pieces of information; then, a standalone event display application reads those data in and it
	creates the required visualization; the advantage is having a cross-platform application which can use any graphics and GUI libraries,
	while the drawback is the lack of direct and full access to the experimental data and to the experiment's software tools,
	which prevents a detailed, full visualization.}
	\label{fig:eventdisplaysapplicationstypes}
\end{figure}

%%% MOVED TO SECTION "Suggested guidelines"
%But all these advantages come at a price, at least in the current landscape,
%as direct access to the experimental data is not possible. One has to export the required information from the experimental
%data into, for example, JSON, XML, or 3D-targeted file formats, and then render this information. However, in this process,
%one has distilled down from the experimental format and some information is lost or rather cannot be easily accessed.

%The access to experimental data is a problem in today’s experiments, and it is the most critical blocker when developing
%visualization tools for HEP. Developers have to choose from the beginning the data access pattern and the target of the
%application: either an integrated application which can visualize all data, but which can run only on specific platforms;
%or a cross-platform standalone tool, which can visualize only a subset of all the information. Moreover, the many
%differences in the access to data among the experiments, make very difficult to conceive common solutions and tools.

%Data, in fact, should be much more easily accessible to scientists, in a more transparent way and without the usage of
%complex software frameworks. That is an issue which is investigated and addressed in more details in the HSF
%"Data Organization, Management and Access" Community White Paper~\cite{HSF-CWP-2017-04}.

\hypertarget{application-development}{%
\subsubsection{Application development and distribution}\label{application-development}}

Currently, the two most common ways of distributing event display applications are as a desktop application and as a web
application running in the browser. Each approach has its advantages and disadvantages which are further described in this section.
The current landscape is summarized in Figure~\ref{fig:currentvisualizationlandscape} and is further described in this section.

Native mobile applications running on devices such as smartphones and virtual reality applications are less common in HEP.
However, they are a growing feature in the current landscape and many experiments are exploring the
possibilities of those emerging technologies. At the end of this section we describe briefly the mobile applications released so far.
Further developments will be described in
Section~\ref{modern-tech}.

{\bf Desktop applications} Many experiments have developed integrated event-display applications in C++, which is the main language
used for developing HEP software frameworks, on top of the OpenGL~\cite{OpenGL1992} APIs.
The choice of the OpenGL API, compared to other APIs like Direct3D~\cite{Direct3D}, resides in its cross-platform nature as
OpenGL is an open standard. The OpenGL consortium defines the API: the interface with which all the implementations have to
comply. The actual implementation is provided by vendors, usually targeting a specific hardware. Many hardware and
software companies such as Intel and NVIDIA are part of the OpenGL consortium, which assure
the support and the lifetime of the OpenGL API.

Some HEP visualization applications use OpenGL calls directly through custom graphics engines. This is the most
robust approach as the developers can take full control over the OpenGL interface
and the project can be independent of other software libraries. Two examples of HEP applications which follow this path are the
ATLAS Persint application~\cite{ATLASPersint2012} and the ROOT EVE toolkit~\cite{ROOTEVE2007}, which is used both by the CMS
Fireworks application~\cite{CMSFireworks} and the ALICE Event Visualisation Environment (AliEve)~\cite{alieve}.
One disadvantage of this approach is that developer time has to be assured to maintain the graphics engine itself,
on top of the effort needed for the development and maintanance of the event display application.

%\missingfigure[figwidth=12cm]{schematic B - OpenGL API access}

Other applications use higher-level interface libraries as graphics engines. This has the advantage of delegating
a large part of the lower-level development work to external software packages, leaving the developers to concentrate
on the application development itself. A popular graphics library used in HEP software has been
Open Inventor~\cite{OpenInventor1993}, used by the defunct CMS Iguana~\cite{CMSIguanaPaperNIM,CMSIguana} application, or
its clone implementation Coin (also known as Coin3D)~\cite{Coin3D}, which has been used by applications such
as ATLAS VP1~\cite{ATLASVP12010}, LHCb Panoramix~\cite{LHCbPanoramix} and the defunct desktop version of the
CMS iSpy application~\cite{CMSISpy}. Coin / Open Inventor was
chosen because of its integrability in C++ code, its performance, and its coding style. Moreover, the way Open Inventor handles
graphical volumes could be easily matched with the way geometry volumes are handled to describe the detectors in HEP experiments.
Open Inventor organizes geometry volumes as a series of nodes in a tree-like structure in the same way as some
HEP experiments do. ATLAS, for instance, developed their geometry library “GeoModel”~\cite{ATLASGeoModel2004} based on
the same tree-like structure of nodes used by Open Inventor.

The drawback of this approach is the dependency on external software projects, which could end up with a loss of functionality if
third-party library development and support are abandoned. Many scientific visualization applications, also in fields other than
HEP, faced this when the support of the Coin library was dropped by the company that led its development~\cite{CoinEndOfLifeLetter}.
The result is the aging of libraries which after a while show incompatibilities with modern hardware, compilers, and platforms. The time
spent by HEP developers to repair or to maintain those abandoned libraries results is time not spent on actual development
of the software applications themselves.

An additional approach to the development of event displays is to create and distribute an application using the Java programming language.
The Atlantis~\cite{ATLASAtlantis} program and its derivative MINERVA~\cite{ATLASMinerva}, which is used as an educational tool,
both developed for the ATLAS experiment, are based on the Java graphics libraries and can be run either online in a web browser
or stand-alone on a desktop machine.

{\bf Web-based applications} Several experiments at the LHC, notably CMS~\cite{CMSISpyWebGL}, LHCb~\cite{LHCbOnline2014}, and ATLAS
~\cite{ATLASTada2016, ATLASTracer2015} have created web-based event displays using WebGL (Web Graphics Library)~\cite{WebGL2011}.
WebGL is a JavaScript API that conforms to OpenGL ES (a subset of the OpenGL API for embedded systems) conceived for
rendering interactive 3D and 2D graphics within any compatible web browser without the need of external plug-ins.
With WebGL in the browser one can combine high-quality graphics with the functionality and accessibility of the browser.
This combination of graphics and user function was previously only available via bespoke desktop applications based on OpenGL and
graphical user interface toolkits such as Qt~\cite{QtFramework}. Browser-based event displays have several distinct advantages: they are
easy to distribute to the user, they can be prototyped quickly, and the client is often much lighter-weight,
as the need for building, packaging and distributing external libraries is greatly reduced.
There are also several mature and actively developed WebGL frameworks, such as the popular three.js~\cite{ThreeJS},
that provide straightforward and simplified APIs for ease of development.

\begin{figure}
	\centering
	\includegraphics[width=0.95\linewidth]{img/Current_Visualization_Landscape_v3}
	\caption[Current landscape of event display applications in HEP]{The image summarize the current landscape of event display applications in HEP.
	Many experiments developed full-framework desktop applications as well as light, web-based applications. As one can see from the plot,
	there are no examples, so far, of full-framework applications using web-based visualization graphics, due to the data access issues
	with today's experiments' frameworks. A new approach in that direction is what it is suggested in this paper, in Section \ref{client-server}.}
	\label{fig:currentvisualizationlandscape}
\end{figure}

{\bf Mobile applications} Mobile devices such as smart phones and tablets are more and more ubiquitous and these devices
are used more and more as substitutes for desktop and laptop machines.
However, mobile devices still do not have the computing power usually needed for HEP data analysis, where huge amount of experimental data
are retrieved and processed. In addition they usually run dedicated operating systems whose self-contained nature
makes their integration within the HEP workflow difficult, particularly for the statistical-based visualization used in
data analysis, described in Section~\ref{statistical-data-visualization}. Despite the current limitations for these use-cases mobile
devices can be found in the current landscape of event displays.

Several event display applications have been developed for, or at least can be run on, mobile devices.
An example of a native application is LHSee~\cite{LHSee} which live-streams ATLAS events, processed and extracted
through Atlantis~\cite{ATLASAtlantis},
to a user’s phone and provided contextual information on ATLAS and the
events being displayed. The CMS iSpy WebGL application~\cite{CMSISpyWebGL} runs on mobile devices in the browser.
The Camelia application~\cite{CERNCamelia}, % and its successor TEV~\cite{CERNTEV}
developed by the CERN Media Lab using the Unity game engine~\cite{Unity3D}, can be built as a mobile application
by using the tools provided by the game engine and run on mobile devices as well.
More details on game engines can be found in Section \ref{graphic-engines}.

{\bf Virtual and augmented reality applications} Virtual Reality (VR) simulates the user’s physical presence in a virtual environment,
and the application is typically run on a head-mounted display that provides visual and aural experience of the simulated environment.
Different degrees of realism and immersion are possible, depending on the targeted hardware.

VR technologies can be used to build immersive applications, to let the general public virtually visit HEP detectors and explore
experimental sites. Many HEP experiments started developing VR applications, mostly as an educational tool, for outreach events.
These include ATLASrift~\cite{ATLASRift} and Belle II VR~\cite{BelleIIVR}.
As many HEP experiments are not accessible during data taking or are classified as supervised areas due to security or safety issues,
VR applications let the HEP community open their sites and experiments to the general public. Also,
they let people look at simulated  collisions in a simplified yet realistic environment, which help people acquire the basic
concepts on which HEP experiments are built and run. Such applications are currently used in public events,
in museums and science centers, and in meetings with the governments and the funding agencies.

Augmented Reality (AR), instead, uses a camera to take a view of the real world around the user and screen where simulated
objects are rendered in 3D and shown on top of that image dynamically, following the user's interaction and motion.
This lets the user move within an environment where real and simulated objects live together. AR can be used in HEP as
an educational tool, for instance to dynamically show and describe a HEP detector to a group of people or a class.
Some HEP experiments have been started exploring AR technologies for HEP, particularly ALICE, ATLAS, and CMS.

Both VR and AR applications are usually developed in specialized graphics engines, which prevent the interaction
with the experiment's framework to access native data. More details on VR and AR applications for HEP are found in Section \ref{vr}.

%\missingfigure[figwidth=12cm]{Add ref/comment to https://twitter.com/acceleratARapp}

\hypertarget{geometry-description}{%
\subsubsection{Geometry description and visualization}\label{geometry-description}}

Geometry visualization provides important visual context for event displays and dedicated geometry displays are useful
applications by themselves. There are typically three levels of detail found in applications. The most detailed geometry
is typically called the \textit{simulation geometry} and can include the sensitive elements of the detector as well
as support structure. Detector experts are typical end-users of applications that display this level of geometrical detail. Less
detailed is the so-called \textit{reconstruction geometry}, which describes the sensitive elements of the detector
such as calorimeter crystals and wire and strip chambers. It is this level of detail that is typically found in
event displays. The least-detailed geometry descriptions are those that are simplified versions of the
detector geometries and are used to provide visual context only.

Currently, different geometry formats and libraries are used in HEP. 
Some experiments use their own custom format (\textit{e.g.}, \cite{ATLASGeoModel2004}), while others use the geometry tools provided by the ROOT framework~\cite{Root1997}.
More recently, some attempts have been made to build common formats and libraries for detector geometry, like DD4HEP~\cite{dd4hep},
adopted in conceptual design studies for future high-energy colliders, including the CLICdp~\cite{clicdp} and FCC~\cite{fcc} collaborations.
In all cases detector volumes are built from simpler geometrical entities: geometrical shapes like Tube, Cone, Box, and
more complex variations of these are combined in order to build the volumes of the experiment’s geometry.
Geometry libraries and formats are also described in the HSF \textit{Detector Simulation} Community White Paper~\cite{HSF-CWP-2017-07}.

The differences in geometry formats used by the different experiments, by detector simulation programs like Geant4~\cite{Geant4},
and by data analysis frameworks like ROOT, typically require developers of visualization applications to write converters
between the different formats. In addition it is often not easy to use a visualization tool developed for one experiment with another one as
current visualization tools are often tightly bound to the geometry library used by the experiment.

In framework-based applications the geometry information can come directly from the experiment's detector description and in
 many cases the hierarchical structure of the detector description is preserved and accessible. Standalone applications
typically use a geometry file with information exported from the software framework. An example hybrid solution is the one developed for the CMS experiment using the  
SketchUp application~\cite{sketchupSW}, described in Ref.~\cite{CMSSketchUp} (see Figure~\ref{fig:cms-sketchup}).
The CMS detector description \cite{cmsDetectorDescription} as written in XML is parsed using Ruby scripts and 3D models are built using the
SketchUp program via its Ruby API. SketchUp can then export to various standard 3D file formats. In this way
detailed simulation geometry can be available in a standalone application.

\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{cms-sketchup-shapes}
		\caption[Converting the CMS detector geometry data to 3D meshes]{{\small}}
		\label{fig:ed-sketchup-a}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{cms_160312_03}
		\caption[Rendering the CMS geometry in the SketchUp 3D software]%
		{{\small }}
		\label{fig:cms-sketchup-b}
	\end{subfigure}
	\caption[Using 3D editing software for HEP Geometry]
	{\small Use of a 3D software (SketchUp) to render the basic geometrical shapes imported from the CMS detector description and the final volumes \cite{CMSSketchUpImages}.}
	\label{fig:cms-sketchup}
\end{figure*}

%CMS created 3D models of the CMS detector in SketchUp~\cite{CMSSketchUp}. Ruby scripts were used to parse the
%description of the CMS detector geometry written in XML and to build 3D models in SketchUp via its Ruby API.
%Figures produced based on these 3D models have been widely used: in many conference presentations, many PhD theses,
%technical design reports, a few notable journal publications, books, magazines, posters, brochures, websites, and other media.
%Further, the 3D models were imported in iSpy WebGL mentioned above.

%In the case of framework-integrated applications, the geometry shown is usually the actual geometry used in the detector description,
%taken directly from the experiment’s software framework and used for all kinds of tasks related to geometry (see Figure [ATLASED042016]).
%While in the case of  java-based standalone or web-based applications the geometry shown is usually a simplified version of the actual one
%(see figure [ATLASED072015]) or, merely, a fake geometry.

\hypertarget{statistical-data-visualization}{%
\subsection{Statistical data visualization}\label{statistical-data-visualization}}

\epigraph{The simple graph has brought more information to the data analyst's mind than any other device.}{\textit{John Tukey \cite{Tukey1962}}}

Data visualization also means visualizing quantities and properties taken from a series of events, in order to extract statistical
meaning from them. Examples of statistical data visualizations are histograms and scatter plots.

In HEP, like most other scientific disciplines, visualization of the data and of the final results plays a key role in the
analysis pipeline. A new projection of the data may provide new insight, results must be summarized in a
clear and concise way, and multidimensional parameter spaces need to be visualized in an understandable fashion. The discussion
of how to properly display data is not new~\cite{Tufte1986}, but the tools are constantly evolving. Since its introduction 20 years ago,
ROOT~\cite{Root1997} has become the most widely used package in HEP to make plots, graphics, and even to build event displays.
It was developed at a time when there were few alternatives for the HEP community that did not have a significant
financial cost, and it has performed admirably.

However, the landscape is changing and there are several existing tools driven by non-HEP communities available. This section will look at
some of the current alternatives and comment on what options might be available in the future and what our needs are. The main focus is
on data exploration and presentation tools. The former describes tools with which one prepares and builds visualizations for the purpose
of exploring and attempting to understand one's dataset. The latter describes tools for the presentation of final plots in a convenient and accesible way.
For more details on data analysis itself, one should refer to the HSF \emph{Data Analysis and Interpretation} Community White Paper~\cite{HSF-CWP-2017-05}.

\hypertarget{stats-desktop}{%
\subsubsection{Desktop solutions}\label{stats-desktop}}

As it stands, ROOT is the most widely adopted plotting tool within the HEP and Nuclear Physics community.
It has even made some inroads to the astrophysics community and some small pockets within the financial community,
to where some physicists migrated. However, few other disciplines have adopted it. Still, it has many features beyond
the standard 1D/2D/3D histogram/graphing tools, such as 2D and 3D shapes, widgets for building a GUI, a JavaScript
implementation for web-based analysis~\cite{rootjs} and is available within a Jupyter notebook~\cite{JupyterNotebook}. But to
access the plotting features, an analyst must install the entire ROOT package which includes file I/O, scientific
libraries, fitting routines etc.,\ and often the installation process is non-trivial.

Many current HEP analysts make wide use of the Python programming language and the PyROOT libraries~\cite{PyROOT}.
Python is also very popular outside the HEP community and so it is worth looking at non-ROOT options available to Python users.
A recent (as of 2017) summary of the field was presented by Jake VanderPlas at PyCon 2017~\cite{VanderPlas2017}, a subset
of which will be presented here. It is emphasized that this is just a sampling and that the number of options available
is a function of time.

\begin{itemize}
\item The Python library \textit{matplotlib}~\cite{Hunter2007}, released in 2003, is the most mature plotting tool
for python and is the standard for most users. It can produce journal-quality graphics and there are some add-ons
that can improve the default plotting options~\cite{seaborn}. It does 1D, 2D, and 3D graphics with varying degrees
of success, but does not integrate with OpenGL libraries and so it can slow down when the number of data points gets very large.
It does produce most of the histograms found in HEP but some minimal, extra work must be done by the user
to make histograms with error bars. Plots are reactive in the sense that you can zoom in on different regions of the graph,
but you cannot do anything more significant with other mouseover commands (links, additional information, etc.).

\item The R programming language has several widely used graphics tools, both built-in or provided by external modules.
\textit{ggplot2}~\cite{Wickham2009} and \textit{lattice}~\cite{Sarkar2008} are particularly useful to visualize data in
multidimensional parameter spaces. \textit{ggplot2} is an implementation of the guidelines contained in the classic
text \textit{The Grammar of Graphics}~\cite{Wilkinson2005}, while \textit{lattice} is an implementation of the
so-called \textit{Trellis Display}~\cite{Trellis}. Both packages are very popular outside the HEP community and a wide range
of learning materials are available in books, online courses, and other media. Both packages are well developed and mature
and offer mechanisms for users to extend them by adding new features.
\textit{lattice} has a longer history: in 2005, the year \textit{ggplot2} first appeared, \textit{lattice} was already popular.
In fact, figures made with \textit{lattice} were shown in the
presentation in PHYSTAT05~\cite{phystat05} which introduced R to the
particle physics community.
\end{itemize}

\hypertarget{stats-web}{%
\subsubsection{Web-based solutions}\label{stats-web}}

Web-based data visualization is also being rapidly developed. Very sophisticated toolkits now provide tools to build web-based
fully-responsive visualization of data on all types of devices. In addition, they also offer other features, specially useful
for HEP, like full in-browser LaTeX rendering (with MathJAX) and real-time visualization of streamed data. Being JavaScript-based,
those libraries integrate with the overall ecosystem of web-based technologies, letting them use all the tools offered by other
web libraries. They are overall a good solution for data presentation, and can be combined with other tools such as Jupyter
in order to be used for data exploration. Some of the most used toolkits are described below:

\begin{itemize}
\item D3 (Data Driven Documents)~\cite{D32011} is perhaps the first web-based visualization toolkit which has been widely
adopted as the de-facto base solution for building interactive data visualization for the web. The strong point of D3 is the
link of the data to the DOM entities and the possibility to work with SVG objects natively. D3 is also the foundation layer
upon which many higher level toolkits are built.

\item Bokeh~\cite{Bokeh2014}. This is a plotting utility from Continuum~\cite{continuum}, the company behind the
\textit{Anaconda} Python distribution system and other Python modules. It is designed with the web in mind and builds in a high degree of
interactivity into the plots, making it useful to share results publicly and for building dashboards. However, it works by
writing HTML, which makes it difficult to work with unless you use specific IDEs like a Jupyter notebook. Exporting a figure
for a journal article ({\it e.g.}, in the PNG format) is non-trivial as well, as that is not currently the primary use-case for Bokeh.

\item Plotly~\cite{Plotly2015}. This is another web-oriented solution, similar to Bokeh and based on the D3 library, where plots
can be hosted in Plotly’s cloud service or viewed in a Jupyter notebook. The plots are similarly very interactive and there
are ways to export figure images, but that is not the goal. Dashboards can be built with relative ease and plotly offers
libraries in R and JavaScript, in addition to Python. They offer both a free and enterprise business model.
\end{itemize}

The so-called notebooks are a rapidly evolving way of using web-based technology for both online and offline data analysis and
visualization, with access to local resources as well. After having started from a Mathematica-like notebook user interface
paradigm mixing server-side code snippet execution, structured text, and (mostly) static visualizations, the Jupyter community
is now exploring more interactive user interface paradigms, including in the area of visualization. The JupyterLab project is
exploring a more MATLAB-like IDE user experience inside the web browser, with features such as multiple source editing tabs
and interactive python consoles. Its ipywidgets sub-project tries to make Jupyter more interactive by moving more visualization
work to client-side Javascript and introducing classic GUI widgets such as sliders and checkboxes for interacting with the live visualization.
The Belle II experiment has started to use Jupyter notebooks to train new users in data analysis; the learning curve is much gentler than in
traditional terminal-based tutorials and the time to useful visualization of results is much faster.

\hypertarget{stats-issues}{%
\subsubsection{Issues}\label{stats-issues}}

{\bf Separating data visualization from the data analysis:}
The suite of statistical plotting tools in ROOT, Matplotlib, etc.\ are adequate for analysis, and their development is
very responsive to analysts’ needs. However, it is often hard to separate plot-making abilities from the data analysis framework.
As a consequence, if a physicist’s data can only be found on a particular server, the plot-generating code must also be located
there and the outcome is sometimes hard to bring to the physicist’s laptop screen. In the worst cases, graphics files (PNGs) must be copied
from the server to the laptop for viewing. This causes a high interaction latency, discouraging exploration. That’s why the
development of new tools should go towards a sharper separation between the computation on data and the interactive data visualization
routines, pushing the latter to the client side as much as possible.

{\bf Separating the plotting functions and content from the plotting style:}
Another symptom of the tight coupling between data analysis infrastructure and plotting is that trivial changes to the plot— axis labels,
colors, and such— are so deeply buried in the analysis script that persistifying changes to them often requires a full recalculation
of the statistics. Changes to the final plot through the usage of on-display user interfaces, in fact, are overwritten and lost if a
plot is updated for other reasons (new version of data upstream, for example). Here, one could use inspiration from the increasing
separation of logic and presentation that is occurring in GUI toolkits (see {\it e.g.}\ use of CSS stylesheets in the GTK/GNOME environment).
%\end{itemize}

A looser coupling between style and content, as well as a looser coupling between locality of computation and locality of rendering,
would benefit the physics community.

\hypertarget{non-spatial-visualization}{%
\subsection{Non-spatial visualization}\label{non-spatial-visualization}}

In HEP, there are data which are organized in a tree-like structure, and for which a graph or a network visualization is the best choice.
The Detector Description is an example of a source of such data: it describes all the pieces which compose a HEP experiment detector.
The different pieces of the Detector Description are interconnected through different relationships: geometrical volumes can be organized
in a parent-child relationship, or a property node can be shared among many volumes. The visualization of those data in a network helps
developers in the understanding and the debugging of the Detector Description, by visualizing the relationships among all the nodes and
their properties. An example, from the ATLAS experiment \cite{ATLASGeoModel2017}, of a graph visualizing the inner
structure of a HEP detector description can be seen in Figure~\ref{fig:atlasgeotree}.

Networks and graphs are very effective ways of visualizing tree-like data, because they are able to show all the 	nodes, their relationships and their properties in a proper way. Some degree of interactivity can let the scientists applying different filters and layout, helping them to get rid of the clutter, to better understand and analyze the data.

\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.6\textwidth}
		\centering
		\includegraphics[width=\textwidth]{atlas-geometry-tree-example}
		\caption[Tree-structured geometry data, an example from ATLAS]%
		{{\small {}}}
		\label{fig:atlasgeo1}
	\end{subfigure}
	\vskip\baselineskip
	\begin{subfigure}[b]{0.6\textwidth}
		\centering
		\includegraphics[width=\textwidth]{graph_top_LogVols}
		\caption[A graph visualizing the ATLAS geometry data]{}
		\label{fig:atlasgeo2}
	\end{subfigure}
	\caption[Using 3D editing software for HEP Geometry]
	{\small \subref{fig:atlasgeo1}) A schematic drawing depicting the tree-structure of the data describing the geometry of the ATLAS detector. Such data structures are better visualized using graphs and networks. \subref{fig:atlasgeo2}) A graph visualizing the first layer of the nodes of the ATLAS Detector Description. Different colors indicate different types of nodes; also, the labels along the lines state the different types of relationship between two data nodes \cite{ATLASGeoModel2017}.}
	\label{fig:atlasgeotree}
\end{figure*}

Another example of HEP data that can benefit from a graph-based visualization is that one describing the execution chain of the jobs
used to filter and reconstruct the experimental data. Very recently HEP experiments began to develop new parallel frameworks
to concurrently handle analysis or reconstruction jobs, to efficiently exploit the parallelism offered by the modern hardware
(more details can be found in the HSF \textit{Event/Data Processing Frameworks} Community White Paper~\cite{HSF-CWP-2017-08}).
The jobs are handled by a scheduler, which organizes them according to their needed input and output data. The outcome of the scheduler
is a directed acyclic graph (DAG). The visualization by means of a graph helps the developer understanding and debugging the
reconstruction and framework code.

Other HEP experiments use graph-based tools to analyze and visualize other types of data, like geographical distribution and load of computer
networks used to transfer data between GRID sites \cite{neo4jnetworkstudy} (see Picture \ref{fig:networkstudy})
 or to store and query conditions data \cite{Clemencic:2012cw}.

\begin{figure*}
	\centering
	\includegraphics[width=0.95\linewidth]{proximity.png}
	\caption[Graphs used to visualize network topologies]
	{\small The image shows a graph used to visualize particular snapshot of the network topology between four WLCG sites, which was used to perform network path analysis. [Image provided by the authors of \cite{neo4jnetworkstudy}].}
	\label{fig:networkstudy}
\end{figure*}

All those data are not space- nor time-dependent, and they are better visualized through a graph or a network. Graph-based visualization,
as well as graph-databases, are fairly new in the HEP landscape but they can be very powerful tools to effectively visualize
non-spatial data which are by their nature organized with a network layout. We suggest the community further explores those tools,
to better understand the possibility offered by graph-based solutions for HEP needs.

% More examples from the other experiments, if any and if possible... ==> I didn't find any other example]

\hypertarget{suggested-guidelines}{%
\section{Suggested guidelines and future development}\label{suggested-guidelines}}

As a community what we want to suggest here is the design and the usage of common base visualization guidelines,
to be able to share knowledge and best practices among the different groups, and to foster collaboration among the HEP experiments.

\hypertarget{common-format}{%
\subsection{A common community-defined format}\label{common-format}}

Visualization has a key role in the lifecycle of a HEP experiment, addressing input data from many sources and in many different formats.
The input data are often quite tightly bound to a specific experiment's software framework and because of that many visualization tools
are integrated into the frameworks to some extent. However, the visualization is often the last step on the experiment data chain and the
output of a visualization application is often not used by any other tool in the software framework. So while the direct interaction with experimental data
formats is highly experiment specific, there is a real possibility of having the final stages of the visualization pipeline
shared between several experiments.

Let us take the example of the detector geometry. There are very many different geometry libraries and formats in use among the HEP experiments.
However, geometry libraries are all different ways to describe and handle basic geometrical entities and combinations of them.
From a visualization point of view, the output of all geometry libraries are mere descriptions of 3D shapes, which could be abstracted
from the underlying actual implementation. A shape like a box, or a cone, or a tube, or some boolean combination of them, could be
interpreted and handled the same way by a visualization tool in all experiments.

The same reasoning made for the geometry can be done for the event data. Of course different experiments detect different objects
and measure different quantities, but there are many common entities, especially among experiments within the same research field.
For example, all experiments working on hadron colliders use the notion of particle track, which is usually constructed translating
the track measurements into space points or points and angles, and use the notion of particle jet, usually
visualized as a cone whose length is related to its energy and whose radius is linked to the specific algorithm used for the jet
reconstruction. Both of these objects are currently often handled and visualized differently in different experiments, but could in principle
be the target of a common definition within the community. If so, experiments could share best practices or snippets of code,
if not complete basic tools, to build and handle their visualization. In this way, the know-how and the tools linked to visualization needs
could be shared as well, and developed in a collaborative way.

%Visualization applications should not be aware,
%and should not care, of how the given 3D volume has been created in the experiment’s code; they should only be able to
%correctly interpret the final information about it and convert that to pixel values, to be displayed on the screen.

%Of course, the usage of a common geometry library would solve many portability issues. But we know that this is not a viable solution: experiments
%have very different needs, in terms of coding languages, available know-how, integration with other software, portability, and so forth;
%and so they have to choose the geometry library which can fit the most of their needs.
%Moreover, the experiments can decide to extend the geometry library they use, in order to add custom shapes or specific functionalities.
%So, a common geometry library for all HEP needs cannot work. But we could imagine common guidelines and, perhaps, tools to visualize final geometry volumes.
%If we can imagine of having basic functions to visualize a community-defined Box shape, for example, we could imagine sharing the knowledge and
%the workload among different experiments and communities. This does not mean that all experiments have to use a common definition
%of a Box shape; but all experiments, whatever definition of a Box shape they use internally in their code for other reasons, should be able to
%provide exporters from their internal definition to the community-agreed definition of a Box shape.

We as a community still need to propose and design such common definitions and guidelines. This will
be addressed in the second phase of this Community action, following the completion of the Community White Paper.

For the moment, we observe that several event displays in experiments have exporters to translate geometry information to standard
formats used in the communities outside HEP, mainly in computer graphics and engineering. For example the Belle II experiment has
written exporters from Geant4 data to different formats, including VRML~\cite{vrml} and FBX \cite{BelleIIGeoExporters}, which are two of the most
common formats used to store and share 3D graphics data. The Unity game-development engine~\cite{Unity3D}, in turn, can export the
FBX geometry to the glTF format~\cite{glTF}, an emerging royalty-free specification for 3D objects and scenes, for fast web distribution
via the SketchFab community repository~\cite{SketchFabBelleII,SketchFab}. Displays based on three.js have access to multiple importers and exporters
of several geometry formats, including the ones mentioned above. %For the near future, we will foster similar experiments within the community.

In order to start sharing knowledge and to start working on demonstrators to show and share best practices, we propose to start
defining a common format to exchange data among the experiments. We should start by finding and listing common shared objects from
geometry and event data. After that, we should start converging on a shared definition of those objects, to build a common design
toward a data model to handle and serve them. The idea, in fact, is to enable usage of this common format to visualize data from
the different experiments with the same shared best practices, if not the same foundation software tools.

We think that community-developed common formats and tools should also be extendable, to let the experiments add their own custom
content and objects. As an example, calorimeter cells can be of very different shapes, and an experiment might need to add its own
custom shapes to the common format to visualize them properly. Thus, in addition to the part handling the common objects, there should
be a part of the format targeted at storing extended custom content, specific to a given experiment. For such custom content,
experiments will have to develop custom visualization tools as well; but they could build them upon the foundation
of the community-driven part.

Some experiments in that direction have been performed within the community in the already. For example, the
ALICE experiment made use of the mini ``Visualization Summary Data'' (VSD) set of classes, contained in the ROOT Event
Visualization Environment (EVE), to make ALICE data visualization decoupled from the AliROOT experiment’s framework \cite{TadelALICE}.

\hypertarget{serving-data}{%
\subsection{Serving the geometry and event data through services}\label{serving-data}}

Once a common format for shared objects is defined, we believe that the design and the development of online services to query and serve the geometry
data would be a very useful addition to the landscape. %, compared to what the HEP experiments currently feature.
The main driving force is the realization that detector description should be much more accessible than it is today.
For many experiments, accessing the detector description means starting and running at least parts of the experiment’s framework. The need of accessing a specific geometry version, in fact, is critical for reconstruction and simulation. However, the geometry
data needed for event visualization can often be simpler: even when showing the actual geometry of the experiment, accessing the latest
alignment constants is not crucial for visualization purposes, because small differences in the geometry are rarely visible in an event display.
So, we think that serving a “frozen” version of the experiment’s geometry would be enough, and that a simpler way to retrieve it
should be designed, to ease data access for visualization: for example, an online service could remotely serve the geometry data to the visualization applications in a standardized format.

It would be desirable for the new mechanism to have a search/filter functionality too, to let client applications query for a specific
subset of information and for a way to select the level of details, to set the desired accuracy and complexity of the retrieved geometry.

The same could be envisaged for event data, even though that is a more complicated task, involving many different layers and services:
very often event data are stored on the Grid and very often they need to be processed in order to be usable for event displays.
However, experimental data should be made more accessible for visualization. Thus, in collaboration with
the HSF \textit{Data Access and Management} working group, an API or a service to get streamed event data will be designed.
In addition, simulation data description for visualization could be handled in the same way, by implementing converters from
generators and simulation applications.

After a first phase of development and stand-alone testing, the streamed data could be used by the current visualization tools as well,
as the first step of their modernization and towards the usage of common community-developed techniques and best practices.

\hypertarget{client-server}{%
\subsection{Client-server architecture for geometry and event data visualization}\label{client-server}}

After common data formats and mechanisms to serve them are designed, together with a set of exporters required to translate the experiments’
data to the common format, we are proposing to build a client-server architecture, upon which next generation visualization applications can be built.

The idea behind that is that if we can send commands from the client to the server, and get the answer back in the data stream, then
we will be able to interact with the experiment’s framework as well, in addition to using common visualization applications to visualize
the common objects. In this way, we can develop a modular architecture where HEP experiments can share the design, the
development and the maintenance of common visualization tools, while maintaining a certain degree of freedom to add custom content
and objects and to interact with their own framework to retrieve specific content.

\hypertarget{modern-tech}{%
\subsection{Exploring modern technologies}\label{modern-tech}}

\hypertarget{graphic-engines}{%
\subsubsection{Graphics and game engines}\label{graphic-engines}}

So far direct OpenGL, its derivatives (like WebGL), or old graphics libraries have been used in HEP visualization applications. Nowadays,
another type of graphics library is rapidly evolving, those embedded in so-called game engines. Game engines are software frameworks targeted
at the gaming industry, and they feature very efficient, optimized, and modern 3D graphics.
Integration with existing code is not easy, however, because they are usually meant to be used as development environments,
and not as embedded libraries like those commonly used in HEP. So they would probably require some major changes in the usual
software architecture used in HEP. But they offer very optimized graphics and modern features, like tools for Virtual Reality,
which could be exploited in our applications.

Some HEP experiments have recently started to successfully use them to build visualization applications and event displays,
like Belle II~\cite{BelleIIVR} and the Total Event Visualiser (TEV) of the CERN Media Lab~\cite{CERNTEV},
which used the Unity game engine~\cite{Unity3D}, and ATLAS which used the Unreal Engine~\cite{EpicUnreal} for
its virtual reality application ATLASrift~\cite{ATLASRift}. These two game engines are the most popular ones on the market, and they are
free for educational and non-commercial projects.

Unreal Engine is fully open source and it supports two modes of development (C++ and the so-called Blueprints~\cite{UnrealEngineBlueprints}) that can be used
interchangeably even in the same project. It produces extremely performant executables for basically all platforms
(Windows, OSX, Linux, iOS, Android, Web, all VR platforms). All parts of development cycle are fast even for a novice,
thanks to powerful tools implemented as plugins. They have large developer communities and are very fast in supporting the latest technologies;
for example it already supports Vulkan~\cite{vulkanAPI}, the cross-platform 3D and computing API.

The Unity development platform~\cite{Unity3D} is very intuitive for novices as
well as experts and provides rapid turnaround during the development cycle: the project can be executed immediately without having to
compile and link an executable. All of the platforms supported by Unreal are supported as targets by Unity as well. Presently, user code is written in either
C\# or an adaptation of Javascript.

%These languages are somewhat foreign to the majority of HEP-trained programmers who are more comfortable with
%C++; however, this accounts in part for the broader use of Unity than Unreal.

%and VR-animation performance is reported to be significantly better on low-end devices.
%(Notably, the emerging WebVR standard~\cite{WebVR} (recently replaced by WebXR \cite{WebXR}) is not yet supported for ``VR in the browser''. It is possible in Unity to create a
%WebGL~\cite{WebGL2011} executable of a VR app; however, no browser will run this app successfully because of its need to load the
%VR-hardware’s object library, which violates security restrictions in the browser.)

Another game engine that has gained attention in recent years is the open-source engine Godot~\cite{Godot}.
While it is still not quite at the same level as Unity or Unreal Engine, it allows the deployment to similar platforms as Unity and Unreal
Engine but is very lightweight.
It offers support for 2D and 3D graphics and for multiple programming languages {\it e.g.}\ GDScript (a Python-like scripting language),
 C\# 7.0 (by using Mono), and C++.
It also offers visual scripting using blocks and connections and support for additional languages with
community-provided support for Python, Nim~\cite{nimLang}, D and other languages.

As a community, we would like to explore further the features those modern game engines can offer. Also, we would like to take a
look at possible usage patterns in the context and within the workflow of HEP visualization.

Finally, another new entry in the 3D graphics engines landscape is Qt3D~\cite{Qt3d}. The key feature of Qt3D is that it is natively integrated with the Qt framework, which eliminates a layer which was needed until now, \textit{i.e.}, a glue
package to connect the Qt GUI with the window showing the 3D content (like, for example, the SoQt~\cite{soqt} package used by ATLAS). By eliminating that, we could simplify the architecture
of our visualization tools and lower the maintenance workload. Qt3D is still in development, but it shows an initial set of
features which are worth a further consideration of the new toolkit. We plan to take a look at its development in the near future,
to see if it can satisfy the HEP requirements. Also, being open source, we could consider contributing to the Qt3D software
project as a community, by providing the pieces we need for our applications.

\hypertarget{web-based}{%
\subsubsection{Web-based applications}\label{web-based}}

Web-based graphics have traditionally been considered not powerful enough to handle the thousands of volumes that can be
shown in HEP event displays, for example, when visualizing hits in a very busy event. But the technology has rapidly
evolved and web-based graphics can now visualize very complex scenes.
For example, the glTF~\cite{glTF} 3D model of the Belle II detector~\cite{BelleII}, with tens of thousands of elements,
can be loaded, viewed and manipulated in a web browser, even on a smartphone, very effectively~\cite{SketchFabBelleII}.

The advantages of visualization in the browser have been mentioned
previously and several application have already been developed.
There is therefore already a strong interest in the community in supporting the usage and the development of web-based tools.
In particular, JSROOT~\cite{rootjs} could be used as an
underlying layer for event data visualization as well as three.js~\cite{ThreeJS}, which have been
used successfully by different experiments (\textit{e.g.}, \cite{ATLASTada2016,ATLASTracer2015,CMSISpyWebGL}) to visualize geometry and event data.

\hypertarget{vr}{%
\subsubsection{Virtual and augmented reality}\label{vr}}

As briefly described in Section \ref{application-development}, Virtual Reality (VR) describes the simulation of the user’s
physical presence in a virtual environment. This simulation
is typically delivered via a Head Mounted Display (HMD) that provides visual and aural experience of the simulated world.
Rotational and positional tracking of the user’s head and hand motion (when available) allow for interaction and navigation
in the virtual environment. This lets the user live an immersive experience of the virtual world offering many degrees of freedom. Purely rotational motion is usually refered to as
3 degree-of-freedom motion; when combined with positional motion support, an application is said to support 6 degrees-of-freedom.

There are several ways to deliver VR to the user with varying levels of functionality, accessibility, and cost.
They range from applications running on a mobile phone viewed through simple headsets to the most realistic and
immersive VR experiences provided by the combination of advanced HMDs and desktop computers.

The simplest and most inexpensive way to deliver VR is via the web browser on a mobile phone and viewed through a
Google Cardboard~\cite{GoogleCardboard} headset (which can itself be literally made from cardboard).
Rotational tracking is achieved through device orientation
controls, either in a native application or using the HTML5 device orientation API in the browser. No hand controller is
used in the Cardboard but for native applications a click event is available via a magnet attached to the Cardboard viewer.

The Google Daydream~\cite{GoogleDaydream} is the next iteration of viewers developed by Google for their Google VR technology.
Content is still delivered by a mobile phone (thus, the performance is limited by the computing power of the phone) but a
Bluetooth-connected hand controller with rotational tracking is available. The Samsung Gear VR platform~\cite{SamsungGearVR}
is another headset powered by an inserted smartphone targeted for the Oculus platform.

Currently, the most immersive and interactive VR environments are provided by the Oculus Rift~\cite{OculusRift} and the
HTC Vive~\cite{HTCVive}, which combine the computing resources of a desktop machine with sensors, controllers, and high-quality HMDs.
Those sophisticated devices are relatively expensive (even if prices are lowering in the last months) and require a fairly powerful computer to run.
The presence of a proper computer solves the performance issue of phone-based viewers, since the 3D graphics computations are handled
by the computer, but that also increases the required initial budget for people willing to test such technologies, and that limits the number of
people getting access to such platforms, for example in public events organized by HEP institutes.

Meanwhile, a new type of device has been recently developed: a standalone viewer, equipped with an on-board CPU, able to run
medium computation-intensive applications. Those devices lower the budget needed to develop and deploy VR applications significantly.
Examples of those new devices are Oculus Go\cite{OculusGo} and the stand-alone Lenovo headset for the
Daydream environment~\cite{LenovoMirageSolo}.

Game engines, described in Section~\ref{graphic-engines},  provide powerful integrated development environments for
creation of VR applications targeting multiple devices. Thanks to engines' abstraction of third party VR libraries,
most HEP experiments should be able to develop VR applications which natively support both standard displays and all VR hardware.
Currently, both ATLASrift~\cite{ATLASRift} and Belle II VR~\cite{BelleIIVR} support Oculus, HTC Vive,
and standard 2D displays. CMS, using the Unity game engine, is currently working on CMS.VR (to be released), targeting Oculus and HTC.

Augmented Reality (AR) applications use the device's camera to looks at the world around the user, to use that as an underlying layer,
over which, based on the user's interaction and motion, they dynamically render simulated virtual entities. The user can navigate in the real world, while
looking at and interacting with the virtual objects. AR technology can be used in HEP as an educational tool, for instance to
dynamically show and describe a HEP detector to a group of people or a class.

ATLAS and ALICE researchers have started to explore the possibilities offered by augmented reality for outreach and education,
by using Unity~\cite{Unity3D} and the Vuforia framework~\cite{VuforiaAR} (commercial, but free for development). The ATLAS-in-Your-Pocket~\cite{AtlasPocket} application
uses printed marks to place a rendered geometry of the ATLAS detector on top of a view of the real world in front of the user.
The More-Than-ALICE~\cite{MoreThanALICE} application allows users to superimpose a description of the detector or event visualizations to the camera image of the actual
ALICE detector (for example, on a screen or during public visits to the experimental site) or its paper model.

Development of web-based VR and AR applications for mobile browser can be done using a WebGL library such as three.js~\cite{ThreeJS} and using the
HTML device orientation control API. The viewport is split into two views, one for each eye, each with a dedicated camera view separated
by an appropriate distance to create a stereoscopic effect ({\it e.g.}\ iSpy WebGL~\cite{CMSISpyWebGL} has a stereo mode for Google Cardboard). 
The developing WebVR specification~\cite{WebVR} provides interfaces to VR hardware via the browser. A powerful framework for
development of VR applications for various devices using the browser is A-Frame~\cite{AFrame}. A-Frame has support for several device controllers as well,
such as for the Daydream and Oculus. CMS is exploring the possibilites of A-Frame in the
browser for both VR and AR ({\it e.g.}\ CMS A-Frame prototype~\cite{CMSAFrame}).

VR and AR applications could be used, in principle, to build event displays and data visualization tools for research work as well.
However, the current data access model prevents a straightforward use of those technologies together with the experiments' software frameworks.
A future client-server approach to data access could fill the current gap and let developers build new VR and AR tools targeting HEP physicists as end users.

\hypertarget{mobile-tech}{%
\subsubsection{Mobile technologies}\label{mobile-tech}}

Portability and simplicity of usage are the strong points of mobile devices. More than as ``mobile'' devices, smartphones,
tablets and ultrabooks can be considered, as devices ``close to people''. As such, the usage of such devices
should be exploited more in the final steps of the visualization chain, where heavy batch data processing is not needed. For instance,
their usage should be leveraged for the production and visualization of event displays.
Ideally, a user should be able to easily retrieve interesting events from the experiment and interactively visualize them on all
kinds of devices.

That is why we strongly promote the usage of the server-client architecture described and supported in this paper in Section~\ref{client-server}
and the new data access patterns presented and supported in the the HSF \textit{Data Access and Management} Community White Paper~\cite{HSF-CWP-2017-04}. %%% TODO: Check the link to the DOMA paper when out!
This would open up new possibilities for interactive visualization on mobile devices: it would let visualization clients running on mobile devices
connect to server tools running in the experiment's framework to easily and interactively retrieve the desired data.

It is worth noting that in other areas of science, for instance in astronomy, researchers have worked to facilitate data
access and to migrate to more standard data formats. This allowed for the possibility of having data visualization tools
on mobile devices, in addition to desktop and laptop machines. Moreover, this not only helped the researchers in accessing and visualizing
their data, but it also helped in making science accessible by the public, having eased the development of programs used in
Outreach and Education activities and events. It is true that HEP data are usually much more complex than astronomy data, and so it
will be harder to achieve, but we think that an effort in simplifying the access to experimental data would be worthwhile.

Therefore, the leverage of the usage of mobile devices in HEP adds a strong point to the development and the support of common
client-server tools and data exchange formats among HEP experiments in the near future.

\hypertarget{multi-user}{%
\subsubsection{Multi-user applications}\label{multi-user}}

Nowadays multi-user technology is used in many applications: for example in GoogleDocs, where many users can simultaneously
interact with the same document. What we would like to provide is multi-user support for visualization to let several users
explore and interact with events at the same time. Beside being a useful feature for expert users (for example, when asking for advice on the visualization of a piece of detector to another person at a distant institute), it could be important for
outreach and education activities, where students and other people could interact together with an event display, or for virtual guided tours.
Game engines offer multi-user support natively, thus we could starting explore their usage.
An example of such collaborative features in a 3D environment is integrated in the Med3D visualization framework~\cite{Bohak2017}.

\hypertarget{sharing-knowledge}{%
\section{Sharing knowledge and fostering collaboration}\label{sharing-knowledge}}

During the kick-starter meetings and the different workshops organized to start and develop the present Community White Paper,
the whole HSF Visualization Working Group has agreed on the importance of sharing knowledge among the whole HEP community,
as well as best practices and know-how. Too often, in fact, solutions and tools developed for one HEP experiment are
not sufficiently advertised to the rest of the HEP Visualization community, with the result that the community base knowledge
is fragmented and not efficiently exploited.

The focus of this Working Group in fact is not limited to the preparation of this white paper. Instead, a longer term collaboration
among the experiments is foreseen, in order to collaborate on common visualization projects.
To foster collaboration and sharing, we agree on following courses-of-action which are described below.

\hypertarget{workshop}{%
\subsection{Yearly workshop}\label{workshop}}

On March 2017 the first HSF Visualization Workshop was organized at CERN to let all the experts from the different experiments and projects
show their work and share their solutions. In addition, external experts from industry were invited to present the latest
advancements in the field and best practices \cite{HSFViz2017}.

It was the first topical workshop focused on HEP Visualization in many years. Many HEP
experiments and communities showed their latest developments. Given the high number of presentations, a second mini-workshop~\cite{HSFViz2017Mini} has been organized as a follow-up, to let the remaining communities to present their work.

The Working Group agreed on the importance of meeting to share findings, knowledge and solutions; and it was decided to
try to organize a topical workshop on HEP visualization once per year.

An important point was raised while organizing the first Workshop: other scientific fields have visualization and
graphics needs similar to HEP, for example Geophysics. In future workshops we will try to
have presentations from other communities as well, in order to try to foster friendly and fruitful collaborations
which could benefit the whole scientific community.

\hypertarget{repo}{%
\subsection{Code repository}\label{repo}}

The HSF Visualization Working Group also agreed on the importance of fostering collaborative work.
As a start a new dedicated project has been created within the HSF GitHub repository~\cite{HSFVizRepo}.
It is intended to be the space where members of the Working Group can share their work-in-progress studies and their solutions, and where
community-driven projects will be stored.

\hypertarget{roadmap}{%
\section{Roadmap}\label{roadmap}}

\hypertarget{one-year}{%
\subsection{One year}\label{one-year}}

In the first year the Visualization Working Group will work on defining R{\&}D projects, based on the key
points and ideas discussed in this community white paper.

The main goal will be developing techniques and tools which let visualization applications and event displays be
less dependent on specific experiments’ software frameworks, leveraging the usage of common packages and common data formats.

In a first phase, the community will identify common objects and will agree on common definitions, as described in
section~\ref{common-format}. Then, a common data exchange format, either based on custom data formats or, if possible, on
open standards, will be designed by the community. After that, exporters and interface packages would be designed
as bridges from the experiments’ frameworks, which are needed to access data at a high level of detail, to the common packages.

\hypertarget{three-year}{%
\subsection{Three years: ATLAS and CMS Computing TDRs}\label{three-year}}

In the second and third year the Visualization WG will work on designing and building demonstrators to show the feasibility of
the community-driven best practices and tools. The goal will be to get a final design of those tools, to be included in the development
plans of the different experiments. Moreover, the WG will work towards a more convenient access to geometry and event data.
In collaboration with the HSF \textit{Data Access and Management} working group, an API or a service to get streamed event data would be designed.

\hypertarget{five-year}{%
\subsection{Five years: Towards HL-LHC}\label{five-year}}

In the fourth and fifth year, the focus will be on developing the actual community-driven tools, to be used by the experiments
for their visualization needs in production.

The goal will be the usage of the community-developed tools within the experiments’ visualization applications; and
perhaps the usage of a simplified data access, but that depends on the actual feasibility, which will be established
after an initial study.

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

Modern and modular visualization tools, which will feature simplified data access and retrieval as well, would leverage
the accessibility, letting end users exploit all the possibilities offered by modern visualization solutions, without the need
of running them on specific platforms, running them within the experiments’ software frameworks, or being bound to specific solutions.
And a better experience will reflect to a better usage, which will positively affect the usage of such visualization tools for
detector development and simulation of new experiments, as well as the data analysis and the upgrade studies of the current ones.

In the end, common community-driven tools will let users of all experiments make use of the latest and best tools, while sharing the development,
the maintenance and the workload among all the experiments.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

The authors would like to thank Guy Barrand \textit{(Geant4, LSST, LAL)} for the fruitful discussions and input about what is done in astronomy in terms of data access and visualization on mobile platforms.

\sloppy
\raggedright
\clearpage
\printbibliography[title={References},heading=bibintoc]

\end{document}

### 2018-11-05 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add dark silicon reference

@@
 In order to process the data, the 4 largest LHC experiments have written
-more than 20 million lines of program code over the last 15 years. This
+more than 20 million lines of program code over the last 15 years
+\cite{1742-6596-898-7-072013}. This
 has involved contributions from thousands of physicists and many
@@
 costs for the scale of HEP computing needs than Intel has achieved with its
-Xeon architecture. More extreme is an architecture that would see
+Xeon architecture~\cite{1742-6596-608-1-012001}. More extreme is an architecture that would see
 specialised processing units dedicated to particular tasks, but with
 possibly large parts of the device switched off most of the time,
-so-called dark silicon.
+so-called dark silicon~\cite{martin2014,IntelCSA}.
 
@@
 performed long after data taking, but can sometimes also be executed in
-quasi-real time on selected samples of reduced size.
+near-real time on selected samples of reduced size.
 
@@
 An important new development in the analysis domain has been the
-emergence of new analysis tools coming from industry and open source
-projects, and this presents new opportunities for improving the HEP
+emergence of new analysis tools coming from industry and open-source
+projects (e.g. Jupyter notebooks~\cite{http://jupyter.org/}, the scikit-learn package~\cite{scikit-learn}), and this presents new opportunities for improving the HEP
 analysis software ecosystem. The HEP community is very interested in
@@
 evolution of relationships between all the different components of an
-analysis.
+analysis, i.e. the provenance of each step.
+
+Reproducibility of scientific results goes in hand with the need to preserve both the data and the software.
+Section~\ref{data-and-software-preservation}} develops this latter topic where the FAIR principles of data management are embraced.
 
@@
 batch processing. These exploit new cluster management systems, most
-notably Apache Spark, which uses open-source tools contributed both by
+notably Apache Spark~\cite{Spark}, which uses open-source tools contributed both by
 industry and the data-science community. Other products implementing the
-same analysis concepts and workflows are emerging, such as TensorFlow,
-Dask, Pachyderm, Blaze, Parsl, and Thrill. This approach can complement
-the present and widely adopted Grid processing of datasets. It may
+same analysis concepts and workflows are emerging, such as
+TensorFlow~\cite{abadi2016tensorflow}, Dask~\cite{rocklin2015dask}, Pachyderm~\cite{thepachydermteam}, Blaze~\cite{wiebe2014blaze}, Parsl~\cite{babuji_yadu_2017_853492}, and Thrill~\cite{bingmann2016thrill}.
+This approach can complement the present and widely adopted Grid processing of datasets. It may
 potentially simplify the access to data and the expression of
@@
 switch to a functional or declarative programming model, as for example
-provided by Scala in the Spark environment. This allows scientists to
+provided by Scala~\cite{Scala} in the Spark environment. This allows scientists to
 express the intended data transformation as a query on data. Instead of
@@
   \item
+    Prototype a comprehensive set of mechanisms for interacting and
+    exchanging data between new open-source tools and the existing
+    analysis ecosystem.
+\item
   Complete an advanced prototype of a low-latency response,
@@
   current Python bindings should evolve to reach the ease of use of native Python modules.
-\item
-  Prototype a comprehensive set of mechanisms for interacting and
-  exchanging data between new open-source tools and the existing
-  analysis ecosystem.
 \item
@@
 capacity network paths to be controlled via experiment software, and
-which could help manage these data flows. These new technologies require
-considerable R\&D to prove their utility and practicality. In addition,
+which could help manage these data flows~\cite{Blikra:2221659}.
+Some projects already started
+to explore the potential of these technologies~\cite{OSiRIS} but a
+considerable R\&D is required to prove their utility and practicality.
+In addition,
 the networks used by HEP are likely to see large increases in traffic
@@
   different access patterns and data organisations envisaged.
+\item
+  Building upon the experience of projects currently exploring SDN
+  potential, define the role for this technology in managing data transfers and
+  access and the integration strategy into experiment frameworks.
 \item
@@
 
-The above-mentioned steps can be considered to be consistent with the
+The above-mentioned steps are consistent with the
 FAIR data principles that are increasingly being mandated by funding
@@
 
+The evolution in the security domain requires the HEP community to work in
+collaboration with the various national security organisations and
+policy groups, building on many relationships that are already
+established.
+
 \paragraph{Trust and Policies}
@@
 applications historically. There is momentum within the wider community
-to develop next-generation identity federations that natively support a
+to develop next-generation identity federations~\cite{OICFederation}
+that natively support a
 wider range of clients. In the meantime there are several viable interim
@@
 
-Although federated identity provides a potential solution for our
+Although non-X509 federated identity provides a potential solution for our
 challenges in Authentication, Authorisation should continue to be
@@
 authorisation, and technology evolutions, are driving the R\&D
-activities. The list below summarises the most important actions:
+activities. These groups are generally much broader than just the HEP
+community. The list below summarises the most important actions:
 
@@
   Ensure that ongoing efforts in trust frameworks are sufficient to
-  raise the level of confidence in federated identities to the
+  raise the level of confidence in non-X509 federated identities to the
   equivalent of X.509, at which stage they could be a viable alternative
@@
 constrained, so it is vital that successful R\&D projects provide
-sustainable software for the future. In many areas it is recognised 
+sustainable software for the future. It is important to emphasise that the goal
+is to support the HEP physics programme in a cost effective manner, so the
+deployment consequences of a particular technology choice or direction must be
+understood with partners in distributed computing. In many areas it is recognised
 that different experiments could have adopted common solutions,

### 2018-11-05 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add some Big Data tools references

@@
 internet economy, which developed novel and major new technologies under
-the banner of \emph{Big Data} that may be applicable to our use cases.
+the banner of \emph{Big Data} that may be applicable to our use cases
+\cite{47224,Shanahan:2015:LSD:2783258.2789993,Armbrust:2015:SSR:2723372.2742797}.
 

### 2018-11-05 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Remove out of place item about people from simulation section
Correct double "the"

@@
   exploring opportunities for code sharing among experiments.
-\item
-  Recruiting, training, retaining human resources in all areas of
-  expertise pertaining to the simulation domain, including software and
-  physics.
 \end{itemize}
@@
 show that without significant changes in approach or algorithms
-the increase in resources needed would be incompatible with the
+the increase in resources needed would be incompatible with
 the expected budget.

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Use more standard language for "near-real" time (referee comment)

@@
 rate (between 30 to 40 MHz for typical LHC proton-proton operations)
-will be streamed to real-time or quasi-real-time software trigger
+will be streamed to real-time or near-real-time software trigger
 systems. The increase in event complexity also brings a ``problem'' of

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add comment about GPUs to ML section

@@
 
-With the advent of more powerful hardware and more performant ML
+With the advent of more powerful hardware, particularly GPUs and ML dedicated
+processors, as well as more performant ML
 algorithms, the ML toolset will be used to develop application software

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Remove HSF label from IML group

@@
 The R\&D roadmap presented here is based on the preliminary work done in
-recent years, coordinated by the HSF IML, which will remain the main
+recent years, coordinated by the IML, which will remain the main
 forum to coordinate work in ML in HEP and ensure the proper links

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add network traffic growth citation

@@
 research network capacity increases show a much steeper growth, such as
-two orders of magnitude from now to HL-LHC times. Therefore, the
+two orders of magnitude from now to HL-LHC times~\cite{Roberts:17}. Therefore, the
 evolution of the computing models would need to be more network centric.

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add WG references

@@
   support LHC physics. This will be an ongoing process, started by the
-  recently formed System Performance and Cost Modeling Working Group,
+  recently formed System Performance and Cost Modeling Working Group~\cite{SPCMWG},
   and aims to provide a quantitative assessment for any proposed
@@
   should be done in conjunction with the existing Data Organisation,
-  Management and Access Working Group to evaluate the impact of the
+  Management and Access Working Group~\cite{WLCG-DOMA} to evaluate the impact of the
   different access patterns and data organisations envisaged.
@@
   workload management of the experiments. This is work to be done in
-  collaboration with the Data Analysis and Interpretation Working Group.
+  collaboration with groups working on new data analysis models.
 \end{itemize}

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add reference to client/server geometry developments
Add visualisation workshop reference to Appendix A

@@
 The working group will also work towards a more convenient access to geometry and
-event data, through a client-server interface. In collaboration with the
+event data, through a client-server interface~\cite{1742-6596-898-7-072015}. In collaboration with the
 Data Access and Management Working Group, an API or a service to deliver streamed
@@
 
-The workshop that was held as part of the CWP process was felt to be
+The workshop that was held as part of the CWP process
+(\emph{HSF Visualization Workshop}, see Appendix \ref{appendix-a---list-of-workshops})
+was felt to be
 extremely useful for exchanging knowledge between developers in

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Consistent dates for short/long term software development goals
Add mention and references to GitHub and GitLab

@@
 alignment with many open source projects and commercial organisations.
+A major benefit that has come with this technical choice is the use
+of social coding sites, such as \emph{GitHub} ~\cite{GitHub}
+and \emph{GitLab}~\cite{GitLab}, where
+code sharing and code review are far superior compared to previous
+solutions.
 Likewise, \emph{CMake}~\cite{CMake}
@@
 
-Ongoing activities and short-term projects, include the following:
+Ongoing activities and short-term projects, to complete by 2020,
+include the following:
 
@@
 
-Medium term projects include the following:
+Longer term projects, to conclude by 2022, include the following:
 

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Avoid confusion of "promoting champions", use clearer phrase

@@
   recognising that it has scientific research value on an equal footing
-  with other activities, and acknowledging and promoting specific
-  ``champions'' in the field.
+  with other activities, and acknowledging and advocating for researchers
+  who choose this as their speciality.
 \end{itemize}

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Avoid unwarrented assertion about other fields' development practice

@@
 
-Unlike the situation that is traditional in some other fields, in
-which users express their requirements and computer specialists
-implement solutions, there is a close collaboration, even overlap,
-in HEP between users and developers that is essential for our success.
+There is a very close collaboration, even overlap,
+in HEP between users of software and developers. This has given experiments
+an agility that was often essential for success in the past.
 Many details of experiment data

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add reference to Intel CSA about dark-silicon

--- 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
@@
 In order to process the data, the 4 largest LHC experiments have written
-more than 20 million lines of program code over the last 15 years
-\cite{1742-6596-898-7-072013}. This
+more than 20 million lines of program code over the last 15 years. This
 has involved contributions from thousands of physicists and many
@@
 costs for the scale of HEP computing needs than Intel has achieved with its
-Xeon architecture~\cite{1742-6596-608-1-012001}. More extreme is an architecture that would see
+Xeon architecture. More extreme is an architecture that would see
 specialised processing units dedicated to particular tasks, but with
 possibly large parts of the device switched off most of the time,
-so-called dark silicon~\cite{martin2014,IntelCSA}.
+so-called dark silicon~\cite{IntelCSA}.
 
@@
 internet economy, which developed novel and major new technologies under
-the banner of \emph{Big Data} that may be applicable to our use cases
-\cite{47224,Shanahan:2015:LSD:2783258.2789993,Armbrust:2015:SSR:2723372.2742797}.
+the banner of \emph{Big Data} that may be applicable to our use cases.
 
@@
   exploring opportunities for code sharing among experiments.
+\item
+  Recruiting, training, retaining human resources in all areas of
+  expertise pertaining to the simulation domain, including software and
+  physics.
 \end{itemize}
@@
 show that without significant changes in approach or algorithms
-the increase in resources needed would be incompatible with
+the increase in resources needed would be incompatible with the 
 the expected budget.
@@
 rate (between 30 to 40 MHz for typical LHC proton-proton operations)
-will be streamed to real-time or near-real-time software trigger
+will be streamed to real-time or quasi-real-time software trigger
 systems. The increase in event complexity also brings a ``problem'' of
@@
 performed long after data taking, but can sometimes also be executed in
-near-real time on selected samples of reduced size.
+quasi-real time on selected samples of reduced size.
 
@@
 An important new development in the analysis domain has been the
-emergence of new analysis tools coming from industry and open-source
-projects (e.g. Jupyter notebooks~\cite{http://jupyter.org/}, the scikit-learn package~\cite{scikit-learn}), and this presents new opportunities for improving the HEP
+emergence of new analysis tools coming from industry and open source
+projects, and this presents new opportunities for improving the HEP
 analysis software ecosystem. The HEP community is very interested in
@@
 evolution of relationships between all the different components of an
-analysis, i.e. the provenance of each step.
-
-Reproducibility of scientific results goes in hand with the need to preserve both the data and the software.
-Section~\ref{data-and-software-preservation}} develops this latter topic where the FAIR principles of data management are embraced.
+analysis.
 
@@
 batch processing. These exploit new cluster management systems, most
-notably Apache Spark~\cite{Spark}, which uses open-source tools contributed both by
+notably Apache Spark, which uses open-source tools contributed both by
 industry and the data-science community. Other products implementing the
-same analysis concepts and workflows are emerging, such as
-TensorFlow~\cite{abadi2016tensorflow}, Dask~\cite{rocklin2015dask}, Pachyderm~\cite{thepachydermteam}, Blaze~\cite{wiebe2014blaze}, Parsl~\cite{babuji_yadu_2017_853492}, and Thrill~\cite{bingmann2016thrill}.
-This approach can complement the present and widely adopted Grid processing of datasets. It may
+same analysis concepts and workflows are emerging, such as TensorFlow,
+Dask, Pachyderm, Blaze, Parsl, and Thrill. This approach can complement
+the present and widely adopted Grid processing of datasets. It may
 potentially simplify the access to data and the expression of
@@
 switch to a functional or declarative programming model, as for example
-provided by Scala~\cite{Scala} in the Spark environment. This allows scientists to
+provided by Scala in the Spark environment. This allows scientists to
 express the intended data transformation as a query on data. Instead of
@@
 \item
-    Prototype a comprehensive set of mechanisms for interacting and
-    exchanging data between new open-source tools and the existing
-    analysis ecosystem.
-\item
   Complete an advanced prototype of a low-latency response,
@@
   current Python bindings should evolve to reach the ease of use of native Python modules.
+\item
+  Prototype a comprehensive set of mechanisms for interacting and
+  exchanging data between new open-source tools and the existing
+  analysis ecosystem.
 \item
@@
 
-With the advent of more powerful hardware, particularly GPUs and ML dedicated
-processors, as well as more performant ML
+With the advent of more powerful hardware and more performant ML
 algorithms, the ML toolset will be used to develop application software
@@
 The R\&D roadmap presented here is based on the preliminary work done in
-recent years, coordinated by the IML, which will remain the main
+recent years, coordinated by the HSF IML, which will remain the main
 forum to coordinate work in ML in HEP and ensure the proper links
@@
 research network capacity increases show a much steeper growth, such as
-two orders of magnitude from now to HL-LHC times~\cite{Roberts:17}. Therefore, the
+two orders of magnitude from now to HL-LHC times. Therefore, the
 evolution of the computing models would need to be more network centric.
@@
 capacity network paths to be controlled via experiment software, and
-which could help manage these data flows~\cite{Blikra:2221659}.
-Some projects already started
-to explore the potential of these technologies~\cite{OSiRIS} but a
-considerable R\&D is required to prove their utility and practicality.
-In addition,
+which could help manage these data flows. These new technologies require
+considerable R\&D to prove their utility and practicality. In addition,
 the networks used by HEP are likely to see large increases in traffic
@@
   support LHC physics. This will be an ongoing process, started by the
-  recently formed System Performance and Cost Modeling Working Group~\cite{SPCMWG},
+  recently formed System Performance and Cost Modeling Working Group,
   and aims to provide a quantitative assessment for any proposed
@@
   should be done in conjunction with the existing Data Organisation,
-  Management and Access Working Group~\cite{WLCG-DOMA} to evaluate the impact of the
+  Management and Access Working Group to evaluate the impact of the
   different access patterns and data organisations envisaged.
-\item
-  Building upon the experience of projects currently exploring SDN
-  potential, define the role for this technology in managing data transfers and
-  access and the integration strategy into experiment frameworks.
 \item
@@
   workload management of the experiments. This is work to be done in
-  collaboration with groups working on new data analysis models.
+  collaboration with the Data Analysis and Interpretation Working Group.
 \end{itemize}
@@
 The working group will also work towards a more convenient access to geometry and
-event data, through a client-server interface~\cite{1742-6596-898-7-072015}. In collaboration with the
+event data, through a client-server interface. In collaboration with the
 Data Access and Management Working Group, an API or a service to deliver streamed
@@
 
-The workshop that was held as part of the CWP process
-(\emph{HSF Visualization Workshop}, see Appendix \ref{appendix-a---list-of-workshops})
-was felt to be
+The workshop that was held as part of the CWP process was felt to be
 extremely useful for exchanging knowledge between developers in
@@
 alignment with many open source projects and commercial organisations.
-A major benefit that has come with this technical choice is the use
-of social coding sites, such as \emph{GitHub} ~\cite{GitHub}
-and \emph{GitLab}~\cite{GitLab}, where
-code sharing and code review are far superior compared to previous
-solutions.
 Likewise, \emph{CMake}~\cite{CMake} 
@@
 
-Ongoing activities and short-term projects, to complete by 2020,
-include the following:
+Ongoing activities and short-term projects, include the following:
 
@@
 
-Longer term projects, to conclude by 2022, include the following:
+Medium term projects include the following:
 
@@
 
-The above-mentioned steps are consistent with the
+The above-mentioned steps can be considered to be consistent with the
 FAIR data principles that are increasingly being mandated by funding
@@
 
-The evolution in the security domain requires the HEP community to work in
-collaboration with the various national security organisations and
-policy groups, building on many relationships that are already
-established.
-
 \paragraph{Trust and Policies}
@@
 applications historically. There is momentum within the wider community
-to develop next-generation identity federations~\cite{OICFederation}
-that natively support a
+to develop next-generation identity federations that natively support a
 wider range of clients. In the meantime there are several viable interim
@@
 
-Although non-X509 federated identity provides a potential solution for our
+Although federated identity provides a potential solution for our
 challenges in Authentication, Authorisation should continue to be
@@
 authorisation, and technology evolutions, are driving the R\&D
-activities. These groups are generally much broader than just the HEP
-community. The list below summarises the most important actions:
+activities. The list below summarises the most important actions:
 
@@
   Ensure that ongoing efforts in trust frameworks are sufficient to
-  raise the level of confidence in non-X509 federated identities to the
+  raise the level of confidence in federated identities to the
   equivalent of X.509, at which stage they could be a viable alternative
@@
 constrained, so it is vital that successful R\&D projects provide
-sustainable software for the future. It is important to emphasise that the goal
-is to support the HEP physics programme in a cost effective manner, so the
-deployment consequences of a particular technology choice or direction must be
-understood with partners in distributed computing. In many areas it is recognised
+sustainable software for the future. In many areas it is recognised 
 that different experiments could have adopted common solutions, 

### 2018-11-07 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
add references

--- 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
@@
 density constraints have limited the clock speed of processors for more
-than a decade. This has effectively stalled any progress in the
+than a decade~\cite{NAP12980}. This has effectively stalled any progress in the
 processing capacity of a single CPU core. Instead, increases in
@@
 based on standard x86\_64 cores to numerous alternatives such as GPUs.
-For GPUs, the processing model is very different, allowing a much
+For GPUs, the processing model is very different~\cite{Cook:2012:CPD:2430671}, allowing a much
 greater fraction of the die to be dedicated to arithmetic calculations,
@@
 possibly large parts of the device switched off most of the time,
-so-called dark silicon~\cite{IntelCSA}.
+so-called dark silicon.
 
@@
 Particle physics is no longer alone in facing these massive data
-challenges. Experiments in other fields, from astronomy to genomics,
+challenges. Experiments in other fields, from astronomy~\cite{10.1371/journal.pbio.1002195} 
+to genomics~\cite{ijms18020412},
 will produce huge amounts of data in the future, and will need to
@@
 which has focused on the use of multi-objective regression and
-generative adversarial networks (GANs). Since use of GANs allows for
+generative adversarial networks (GANs)~\cite{}. Since use of GANs allows for
 non-parametric learning in cases such as calorimetric shower
@@
 flexible, tiered data formats. These tiers are typically based on the
-ROOT~\cite{Brun1996} file format and constructed to
+ROOT file format and constructed to
 facilitate both skimming of interesting events and the selection of
@@
 Over the past 20 years the HEP community has developed and gravitated
-around a single analysis ecosystem based on ROOT~\cite{Brun1996}. 
+around a single analysis ecosystem based on ROOT. 
 ROOT is a general-purpose object
@@
 batch processing. These exploit new cluster management systems, most
-notably Apache Spark, which uses open-source tools contributed both by
+notably Apache Spark~\cite{Armbrust:2015:SSR:2723372.2742797}, which uses open-source tools contributed both by
 industry and the data-science community. Other products implementing the
-same analysis concepts and workflows are emerging, such as TensorFlow,
-Dask, Pachyderm, Blaze, Parsl, and Thrill. This approach can complement
+same analysis concepts and workflows are emerging, such as TensorFlow~\cite{tensorflow2015-whitepaper},
+Dask~\cite{dask_ref}, Pachyderm~\cite{thepachydermteam}, 
+Blaze~\cite{blaze}, Parsl~\cite{babuji_yadu_2017_853492}, 
+and Thrill~\cite{bingmann2016thrill}. This approach can complement
 the present and widely adopted Grid processing of datasets. It may
@@
   investigation of these concepts, which connect to Natural Language
-  Processing, has started at the LHC and is to be pursued on the same
+  Processing~\cite{Collobert:2011:NLP:1953048.2078186}, has started at the LHC and is to be pursued on the same
   timescale.
@@
   \emph{Simulation:} recent progress in high fidelity fast generative models,
-  such as Generative Adversarial Networks (GANs) and Variational
-  Autoencoders (VAEs), which are able to sample high dimensional feature
+  such as Generative Adversarial Networks (GANs)~\cite{NIPS2014_5423} and Variational
+  Autoencoders (VAEs)~\cite{2013arXiv1312.6114K}, 
+  which are able to sample high dimensional feature
   distributions by learning from existing data samples, offer a
@@
 infrastructure, which has evolved from the original ideas on deployment
-before LHC data-taking started~\cite{MONARC}, to be a mature and
+before LHC data-taking started~\cite{Aderholz:510694}, to be a mature and
 effective infrastructure that is now exploited by LHC experiments.
@@
 
-In recent years, video game engines, such as Unity or the Unreal Engine,
+In recent years, video game engines, such as Unity~\cite{unity} or 
+the Unreal Engine~\cite{Sanders:2016:IUE:3099885},
 have become particularly popular in the game and architectural
@@
 training and collaborative enrichment of the training material. Projects
-such as ROOT~\cite{Brun1996} have also started to provide some
+such as ROOT have also started to provide some
 training material based
@@
 expertise is to use Question and Answer (Q\&A) systems, such as Stack
-Overflow. A few such systems are run by experiments for their own needs,
+Overflow~\cite{Stackoverflow}. A few such systems are run by experiments for their own needs,
 but this is not necessarily optimal, as the value of these services is

### 2018-11-08 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Finished dealing with CSBS journal referees feedback on CWP analysis chapter

--- 2018-11-07 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
@@
 possibly large parts of the device switched off most of the time,
-so-called dark silicon.
+so-called dark silicon~\cite{IntelCSA}.
 

### 2018-11-08 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Address review comments
- Add extra citation for HEP SLOC and rephrase
- Refer to appendix for visualisation workshop*s*
- Correct LaTeX formatting error

--- 2018-11-07 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
@@
 In order to process the data, the 4 largest LHC experiments have written
-more than 20 million lines of program code over the last 15 years. This
+tens of millions of lines of program code over the last 15 years
+\cite{1742-6596-898-7-072013,Elmer2014}. This
 has involved contributions from thousands of physicists and many
@@
 costs for the scale of HEP computing needs than Intel has achieved with its
-Xeon architecture. More extreme is an architecture that would see
+Xeon architecture~\cite{1742-6596-608-1-012001}. More extreme is an architecture that would see
 specialised processing units dedicated to particular tasks, but with
 possibly large parts of the device switched off most of the time,
-so-called dark silicon~\cite{IntelCSA}.
+so-called dark silicon~\cite{martin2014,IntelCSA}.
 
@@
 
-The workshop that was held as part of the CWP process was felt to be
+The workshops that were held as part of the CWP process
+(\emph{HSF Visualization Workshop}, see Appendix \ref{appendix-a---list-of-workshops})
+were felt to be
 extremely useful for exchanging knowledge between developers in
 different experiments, fostering collaboration and in bringing in ideas
-from outside the community. This will now be held as an annual event and
+from outside the community. These will now be held as an annual events and
 will facilitate work on the common R\&D plan.
@@
 alignment with many open source projects and commercial organisations.
+A major benefit that has come with this technical choice is the use
+of social coding sites, such as \emph{GitHub}~\cite{GitHub}
+and \emph{GitLab}~\cite{GitLab}, where
+code sharing and code review are far superior compared to previous
+solutions.
 Likewise, \emph{CMake}~\cite{CMake}
@@
 constrained, so it is vital that successful R\&D projects provide
-sustainable software for the future. In many areas it is recognised
+sustainable software for the future. It is important to emphasise that the goal
+is to support the HEP physics programme in a cost effective manner, so the
+deployment consequences of a particular technology choice or direction must be
+understood with partners in distributed computing. In many areas it is recognised
 that different experiments could have adopted common solutions,

### 2018-11-08 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Update cwp-main.tex
add missed reference

--- 2018-11-08 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
@@
 which has focused on the use of multi-objective regression and
-generative adversarial networks (GANs)~\cite{}. Since use of GANs allows for
+generative adversarial networks (GANs)~\cite{NIPS2014_5423}. Since use of GANs allows for
 non-parametric learning in cases such as calorimetric shower

### 2018-11-08 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex

@@
 around a single analysis ecosystem based on ROOT.
+
 ROOT is a general-purpose object
@@
 batch processing. These exploit new cluster management systems, most
-notably Apache Spark~\cite{Armbrust:2015:SSR:2723372.2742797}, which uses open-source tools contributed both by
+
+notably Apache Spark~\cite{Spark,Armbrust:2015:SSR:2723372.2742797}, which uses open-source tools contributed both by
 industry and the data-science community. Other products implementing the
 same analysis concepts and workflows are emerging, such as TensorFlow~\cite{tensorflow2015-whitepaper},
-Dask~\cite{dask_ref}, Pachyderm~\cite{thepachydermteam}, 
-Blaze~\cite{blaze}, Parsl~\cite{babuji_yadu_2017_853492}, 
+Dask~\cite{rocklin2015dask,dask_ref}, Pachyderm~\cite{thepachydermteam}, 
+Blaze~\cite{wiebe2014blaze}, Parsl~\cite{babuji_yadu_2017_853492}, 
 and Thrill~\cite{bingmann2016thrill}. This approach can complement
 the present and widely adopted Grid processing of datasets. It may
+
 potentially simplify the access to data and the expression of

### 2018-11-09 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
CWP Referee Comment improvements from Graeme (#110)
* Address first CSBS review comment about deployment consequences

It's quite hard to squeeze in changes like this, but what I did was minimal and should
be enough to satisfy the review point

* Address comment on Security and working with groups wider than HEP

* Add citation for ATLAS SLOC

* Add ATLAS ARM events/J reference

* Add dark silicon reference

* Add some Big Data tools references

* Remove out of place item about people from simulation section

* Correct double "the"

* Use more standard language for "near-real" time (referee co

--- 2018-11-08 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
@@
 \item
-  Reviewing the physics models' assumptions, approximations, and
-  limitations in order to achieve higher precision, and to extend the
-  validity of models up to energies of the order of 100 TeV foreseen
-  with the Future Circular Collider (FCC) project~\cite{FCC}.
+  Review the implementations of physics models, including the
+  assumptions, approximations, and
+  limitations. In the best cases this can achieve higher precision and
+  improve runtime performance through code modernisation\cite{Novak2018}.
+  The extension of the validity of models up to energies of 100 TeV is foreseen
+  for the Future Circular Collider (FCC) project~\cite{FCC} and provides
+  a good opportunity for this modernisation.
 \item

### 2018-11-09 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Fixup a few reference issues with #113

@@
 emergence of new analysis tools coming from industry and open-source
-projects (e.g. Jupyter notebooks~\cite{http://jupyter.org/}, the scikit-learn package~\cite{scikit-learn}), and this presents new opportunities for improving the HEP
+projects (e.g. Jupyter notebooks~\cite{Jupyter}, the scikit-learn package~\cite{scikit-learn}), and this presents new opportunities for improving the HEP
 analysis software ecosystem. The HEP community is very interested in
@@
 Reproducibility of scientific results goes in hand with the need to preserve both the data and the software.
-Section~\ref{data-and-software-preservation}} develops this latter topic where the FAIR principles of data management are embraced.
+Section~\ref{data-and-software-preservation} develops this latter topic where the FAIR principles of data management are embraced.
 

### 2018-11-09 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add citations to ALICE SLOC counts
Add a couple more AArch64 references

@@
 tens of millions of lines of program code over the last 15 years
-\cite{1742-6596-898-7-072013,Elmer2014}. This
+\cite{1742-6596-898-7-072013,Elmer2014,OHAliRoot,OHAliPhysics}. This
 has involved contributions from thousands of physicists and many
@@
 costs for the scale of HEP computing needs than Intel has achieved with its
-Xeon architecture~\cite{1742-6596-608-1-012001}. More extreme is an architecture that would see
+Xeon architecture~\cite{1742-6596-608-1-012001,1742-6596-513-5-052014,2014arXiv1404.6929A}.
+More extreme is an architecture that would see
 specialised processing units dedicated to particular tasks, but with

### 2018-11-09 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add reference for FNAL acceleator schedule
add references (#112)
* add references

* Update cwp-main.tex

add missed reference

@@
     \includegraphics[width=0.94\textwidth]{10yr-PLAN-9Feb-2017-V7a.pdf}
-    \caption{Run schedule for the Fermilab facility until 2026.}
+    \caption{Run schedule for the Fermilab facility until 2026~\cite{fnal-schedule}.}
     \label{fig:fnal-if}
@@
 which has focused on the use of multi-objective regression and
-generative adversarial networks (GANs). Since use of GANs allows for
+generative adversarial networks (GANs)~\cite{NIPS2014_5423}. Since use of GANs allows for
 non-parametric learning in cases such as calorimetric shower
@@
 Over the past 20 years the HEP community has developed and gravitated
-around a single analysis ecosystem based on ROOT~\cite{Brun1996}.
+around a single analysis ecosystem based on ROOT.
+
 ROOT is a general-purpose object
@@
 batch processing. These exploit new cluster management systems, most
-notably Apache Spark~\cite{Spark}, which uses open-source tools contributed both by
+
+notably Apache Spark~\cite{Spark,Armbrust:2015:SSR:2723372.2742797}, which uses open-source tools contributed both by
 industry and the data-science community. Other products implementing the
-same analysis concepts and workflows are emerging, such as
-TensorFlow~\cite{abadi2016tensorflow}, Dask~\cite{rocklin2015dask}, Pachyderm~\cite{thepachydermteam}, Blaze~\cite{wiebe2014blaze}, Parsl~\cite{babuji_yadu_2017_853492}, and Thrill~\cite{bingmann2016thrill}.
-This approach can complement the present and widely adopted Grid processing of datasets. It may
+same analysis concepts and workflows are emerging, such as TensorFlow~\cite{tensorflow2015-whitepaper},
+Dask~\cite{rocklin2015dask,dask_ref}, Pachyderm~\cite{thepachydermteam}, 
+Blaze~\cite{wiebe2014blaze}, Parsl~\cite{babuji_yadu_2017_853492}, 
+and Thrill~\cite{bingmann2016thrill}. This approach can complement
+the present and widely adopted Grid processing of datasets. It may
+
 potentially simplify the access to data and the expression of

### 2018-11-10 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
add comment about reco cwp doc
CWP CSBS: fix comment related to AARC project participants

@@
   Most important targets are those which limit expected throughput
-  performance at future facilities, e.g., charged-particle tracking. A
-  number of such efforts are already in progress.
+  performance at future facilities, most significantly 
+  charged-particle tracking. A
+  number of efforts in this area are already in progress~\cite{recocwp}.
 \end{itemize}
@@
 to document requirements for multiple Research Communities through the
-work of FIM4R~\cite{FIM4R}. CERN's participation in the European
+work of FIM4R~\cite{FIM4R}. Participation of CERN and a few other major
+WLCG sites in the European
 Authentication and Authorisation for Research and Collaboration (AARC)

### 2018-11-10 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
CWP CSBS: address comments about algorithms in 3.3

--- 2018-11-09 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
@@
 understand what kinds of algorithms are best suited to what kinds of
-hardware architectures, develop benchmarks that allow us to compare the
+hardware architectures. It is an area where collaboration with the
+computer science community is required. We also need to develop
+benchmarks that allow us to compare the
 physics-per-dollar-per-watt performance of different algorithms across a
@@
 complexity and diversity of our codebase will necessitate both a
-determined push to educate physicists in modern coding
-practices, and a development of more sophisticated and automated quality
+determined push to educate physicists in modern algorithmic approaches and coding practices,
+and a development of more sophisticated and automated quality
 assurance and control. The increasing granularity of

### 2018-11-12 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Small fixes to references and LaTeX formatting
Added OS X friendly like to .gitignore (.DS_Store is the HPFS metadata directory)

@@
   limitations. In the best cases this can achieve higher precision and
-  improve runtime performance through code modernisation\cite{Novak2018}.
+  improve runtime performance through code modernisation~\cite{Novak2018}.
   The extension of the validity of models up to energies of 100 TeV is foreseen
@@
 emergence of new analysis tools coming from industry and open-source
-projects (e.g. Jupyter notebooks~\cite{Jupyter}, the scikit-learn package~\cite{scikit-learn}), and this presents new opportunities for improving the HEP
+projects (e.g. Jupyter notebooks~\cite{Jupyter}, the scikit-learn package~\cite{Pedregosa:2011:SML:1953048.2078195}), and this presents new opportunities for improving the HEP
 analysis software ecosystem. The HEP community is very interested in
@@
 batch processing. These exploit new cluster management systems, most
-
 notably Apache Spark~\cite{Spark,Armbrust:2015:SSR:2723372.2742797}, which uses open-source tools contributed both by
@@
 years have used the HEP-specific software package TMVA~\cite{TMVA} included in ROOT. Recently, however, many HEP analysts have begun
-migrating to non-HEP ML packages such as scikit-learn~\cite{scikit-learn}
+migrating to non-HEP ML packages such as scikit-learn~\cite{Pedregosa:2011:SML:1953048.2078195}
 and Keras~\cite{Keras}, although these efforts have yet to result

### 2018-11-13 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add reference to CMS plots from ACAT

@@
     resources required into the HL-LHC era, using the current computing model
-    with parameters projected out for the next 12 years.}
+    with parameters projected out for the next 12 years~\cite{1742-6596-1085-2-022006}.}
     \label{fig:cms-resources}

### 2018-11-30 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
Update LaTeX adding references
This version taken from overleaf, with some references added by
Bo and Michel
Add one reference to global cwp.bib file
Add a gitignore for the PDF output

@@
 The coming decade will bring HEP to the exabyte scale with the expected data
-volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
-experiments such as the Square Kilometre Array (SKA) entering this
+volumes of the HL-LHC~\cite{HL-LHC,1742-6596-515-1-012012,Apollinari:2284929} experiments, the DUNE~\cite{Acciarri:2015uup} experiment, and even non-HEP
+experiments such as the Square Kilometre Array (SKA)~\cite{SKA} entering this
 regime. In devising computing models for this era, many
@@
   as it is written.
-Most data is written in the ROOT file format.
+Most data is written in the ROOT~\cite{Brun1996} file format.
 ROOT serializes data objects, compresses, and writes them in a column-wise manner.
@@
 challenges were in the area of ``Data Management'' (DM) as the Worldwide
-LHC Computing Grid was commissioned. As the LHC experiments' computing matured
+LHC Computing Grid~\cite{WLCG} was commissioned. As the LHC experiments' computing matured
 through Run1 and Run2, interest has turned to optimizations spanning
@@
 the scope of technologies that can be used to store the data; for
-example, exploiting object stores (such as AWS S3 or Ceph) as event stores. Such an approach
+example, exploiting object stores (such as AWS S3 or Ceph~\cite{Weil:2006:CSH:1298455.1298485}) as event stores. Such an approach
 would support parallel reading and writing of events as

### 2018-12-17 HSF-Notes/2018-01/espp-hsf-submission.tex
HSF Sumbission for ESPP in LaTeX format

% This is the styling for submission to arXiv using JHEP Preprint Style

% Copyright (C) 2018, the authors, licence CC-BY-4.0.

% JHEP preprint template
\documentclass[11pt,a4paper]{article}
\usepackage{jheppub}
\notoc

\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage[titletoc]{appendix}
\usepackage{xspace}

\usepackage[utf8]{inputenc}

%% Comment out before final submission
\usepackage{lineno}  % for line numbering during review
\linenumbers

\abstract{In 2017 the experimental High-Energy Physics community wrote a
\emph{Roadmap for HEP Software and Computing R\&D for the
2020s}\footnote{\href{https://arxiv.org/abs/1712.06982}{\emph{arXiv:1712.06982}}}.
This effort was organised by the HEP Software
Foundation\footnote{\href{https://hepsoftwarefoundation.org/}{\emph{https://hepsoftwarefoundation.org/}}}
(HSF) and was supported by more than 300 physicists from more than 100
institutes worldwide. It delivered a strategy outlining the most
important areas in which investment is needed to ensure the success of
our experimental programme. This contribution to the ESPP is an
executive summary of the most critical and relevant points raised in
that white paper.
}

\begin{document}

% HSF Document Number (only for arXiv)
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 %& & HSF-NOTE-2018-01 \\
 & & December 17, 2018 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{The Importance of Software and Computing to Particle Physics}

\author{A contribution from the High-Energy Physics Software Foundation
to the European Strategy for Particle Physics}

\maketitle

{
\setlength{\parindent}{0cm}
Contact: Graeme A Stewart,
\href{mailto:graeme.andrew.stewart@cern.ch}{\emph{graeme.andrew.stewart@cern.ch}}
}

\newpage

\section*{The Importance of Software to the HEP Experimental
Programme}\label{the-importance-of-software-to-the-hep-experimental-programme}

The current success of the experimental High-Energy Physics programme
relies hugely on software and computing. We have developed many
innovative and unique pieces of software as a necessity, as no
off-the-shelf solutions were available. The HEP software landscape
consists of tens of millions of lines of code, from widely used community
supported packages, like ROOT and Geant4, to experiment and detector
specific codes. Thousands of physicists spend time writing this code and
the lifetimes of HEP experiments require maintaining and evolving these
codes over decades. These codes are run at a massive scale. For the LHC
experiments, in particular, about 750k CPU cores are constantly employed
on the Worldwide LHC Grid running event generation, simulation,
reconstruction and analysis jobs that are critical to our physics
output. The costs of this part of our programme is very high - to
purchase just this CPU resource in a commercial cloud would cost hundreds of
millions of Euros every year. It is incumbent on us to use these
resources as efficiently as possible for physics.

\section*{Future Challenges in an Evolving Technology
Landscape}\label{future-challenges-in-an-evolving-technology-landscape}

A programme of new and upgraded accelerators and experiments will result
in far higher data rates, for both trigger and offline. We need to adapt
our data processing code and data handling infrastructure to cope with
these rates, but we do so in a landscape where the core technologies
and the scientific computing infrastructure will not be the
same as before and violate many of the assumptions made in the past.

HEP benefited for many years from the rapid evolution of microprocessor
technology and a homogeneous data centre model (x86 and Linux). While
shrinking feature sizes still lead to increases in transistor density
(Moore's Law), improvements in clock speeds have stopped and only
internal processor parallelism is improving. This fundamental shift
means that software now has to exploit more and more parallelism to
reduce processing time. Controlling memory usage and bandwidth becomes
critical to maintaining throughput. This paradigm shift from serial to
parallel execution is far complete for HEP and requires more software
developers with both specific technical expertise and physics domain
knowledge to work together.

Modern complex CPU architectures are now no longer seen as optimal for many
workloads and the performance gains on different platforms, such as GPUs or
FPGAs, have been much greater than on CPUs. Adapting our large code bases to
this new hardware landscape often requires a fundamental redesign of algorithms
from traditional serial implementations. The current set of C++ abstractions
used for two decades in both the algorithmic and framework codes needs to be
substantially rethought.

Currently HEP handles data across many geographically dispersed, but
rather homogeneous, computing centres. Our grid infrastructure is now
evolving to incorporate other resources, such as HPC centres with GPUs.
Other resources may be ephemeral and
our new data handling strategies must be able to accommodate these
resources easily. Production systems must also adapt to new workflows,
e.g. related to large scale machine learning training. Much of this new
infrastructure will be shared with other science communities as well, so
assumptions of dedicated network and storage resources only for HEP must
be relaxed.

\subsection*{Software for Future
Detectors}\label{software-for-future-detectors}

Efficient working software is required years in advance in order to
properly optimise detector design for efficient reconstruction. This is
true across all the frontiers of particle physics (energy, intensity and
cosmic). One example is future hadron colliders, which present a huge
challenge. Such accelerators would produce collisions at a pile-up of up
to 1000, requiring new detector technologies and significant advances in
reconstruction software to extract signals in a reasonable time.
Efficient reconstruction and simulation is needed and must be extremely
accurate to support a precision physics search programme. We must
anticipate analyses which deal with orders of magnitude more events,
requiring more efficient approaches to data storage and processing,
while not burdening analysts with complex interfaces. We note that such
challenges are faced now in many sectors of `big data' society and HEP
can both learn from and contribute to these domains.

\section*{Necessary Investments in
People}\label{necessary-investments-in-people}

To face the computing challenges of the future HEP programme substantial
effort is required to improve our software and computing. Ongoing
investment in R\&D, with increased funding, will be critical to extract
as much physics as possible from the data delivered by the upgraded LHC
and other detectors.

Areas for investment must focus on developing and evolving our software
so that it can adapt to the changing and uncertain hardware landscape,
that it can be sustained for decades, and that it can incorporate
advances from other fields, such as machine learning and data science.
This will require cooperation with software engineers and computer
scientists who are not traditionally from our field. The necessary
skills we need require proper, well-funded training for students and
postdocs; this training will make the contribution from HEP even more
valuable when people move outside our field. We have long lead times and
lifetimes for experiments, meaning we need long lifetimes for software
and long-term support for those who develop it. It is thus crucial that
we offer well defined career paths, including new roles such as Research
Software Engineers, for those who chose to work in these software areas,
so as to retain the necessary talent for our discipline. This is
especially true for those who chose to focus on the software engineering
of scientific code, where long-term career prospects must be viable.
This requires endorsing new initiatives to attract funding for a
sustainable and cooperative scientific software ecosystem at European
universities and labs.

\end{document}

### 2018-12-18 HSF-Notes/2018-01/espp-hsf-submission.tex
Final small updates

@@
 %% Comment out before final submission
-\usepackage{lineno}  % for line numbering during review
-\linenumbers
+%\usepackage{lineno}  % for line numbering during review
+%\linenumbers
 
@@
 \author{A contribution from the High-Energy Physics Software Foundation
-to the European Strategy for Particle Physics}
+to the European Particle Physics Strategy Update 2018-2020}
 
@@
 specific codes. Thousands of physicists spend time writing this code and
-the lifetimes of HEP experiments require maintaining and evolving these
-codes over decades. These codes are run at a massive scale. For the LHC
+the lifetimes of HEP experiments require maintaining and evolving this
+software over decades. These codes are run at a massive scale. For the LHC
 experiments, in particular, about 750k CPU cores are constantly employed
@@
 reconstruction and analysis jobs that are critical to our physics
-output. The costs of this part of our programme is very high - to
+output. The cost of this part of our programme is very high - to
 purchase just this CPU resource in a commercial cloud would cost hundreds of

### 2018-12-19 CWP/papers/HSF-CWP-2017-08_framework/latex/df_fwk_summary.tex
Fix name and affiliation (#122)
frameworks: author update and ALFA reference (#124)

@@
 \author[d]{Paolo Calafiura}
-\author[a,1]{Benadikt Hegner}
+\author[a]{Marco Clemencic}
+\author[b]{Hadrien Grasland}
+\author[c]{Chris Green}
+\author[a,e,1]{Benedikt Hegner}
 \author[c]{Chris Jones}
 \author[b]{Michel Jouvin}
+\author[c]{Kyle Knoepfel}
+\author[g]{Thomas Kuhr}
 \author[c,1]{Jim Kowalkowski}
+\author[d]{Charles Leggett}
+\author[c]{Adam Lyon}
+\author[e]{David Malon}
+\author[c]{Marc Paterno}
+\author[d]{Simon Patton}
 \author[c,1]{Elizabeth Sexton-Kennedy}
 \author[a]{Graeme A Stewart}
+\author[d]{Vakho Tsulaia}
+
+% below are the best of the names from the original list
+% that either indicated they wanted to be included or silently
+% let their names be added.
+
 %\author[e]{Several Others - Charles, Marco, ?}
+% Marc F Paterno <paterno@fnal.gov>;  
+% Simon Patton <sjpatton@lbl.gov>;  
+% Christopher H Green <greenc@fnal.gov>; 
+% Kyle Knoepfel <knoepfel@fnal.gov>; 
+% Adam L Lyon <lyon@fnal.gov>; 
+% David Malon <malon@anl.gov>; 
+% Vakho Tsulaia <vtsulaia@lbl.gov>; 
+% Thomas Kuhr <Thomas.Kuhr@lmu.de>; 
+% Hadrien Grasland <grasland@lal.in2p3.fr>; 
+% Charles Leggett <cgleggett@lbl.gov>; 
+% Marco Clemencic <marco.clemencic@cern.ch>
 
@@
 \affiliation[d]{Lawrence Berkeley National Laboratory, Berkeley, CA, USA}
-%\affiliation[e]{Other Places}
+\affiliation[e]{Brookhaven National Laboratory, Upton, NY, USA}
+\affiliation[f]{Argonne National Laboratory, Lemont, IL, USA}
+\affiliation[g]{Ludwig-Maximilians-Universität München, Munich, Germany} 
 \affiliation[1]{Paper Editor}
@@
 and ALICE are now developing a new framework, which is called
-O2~\cite{O2,1742-6596-664-8-082046}. At the time of writing, most major frameworks support
+ALFA~\cite{ALFA:ALICE-FAIR:2015}. At the time of writing, most major frameworks support
 basic parallelisation, both within and across events, based on a
-task-based model~\cite{Jones:2015soc,Clemencic:2015paa}. O2 already
+task-based model~\cite{Jones:2015soc,Clemencic:2015paa}. ALFA already
 includes additional multi-node setups and communication.

### 2018-12-19 CWP/papers/HSF-CWP-2017-08_framework/latex/df_fwk_summary.tex
Final tweaks for arXiv publication (#125)
- Readded the subcaption package as the document would not process without it
- Updated date to date of submisison

@@
 \usepackage{graphicx}
-% \usepackage{subcaption}
+\usepackage{subcaption}
 \usepackage{xspace}
@@
  & & HSF-CWP-2017-08 \\
- & & December 21, 2017 \\ % use \date or hardwire e.g. December 15, 2017
+ & & December 19, 2018 \\ % use \date or hardwire e.g. December 15, 2017
  & & \\

### 2019-02-04 CWP/papers/HSF-CWP-2017-11_generators/latex/main.tex
Add generator CWP source files as-is from Overleaf (https://www.overleaf.com/read/wyyybnvxyfyn)

 \documentclass{article}
\usepackage{graphicx,amsmath}
\usepackage[utf8]{inputenc}
\title{Generator \& Theory Working Group Chapter for CWP}
\author{
R. Boughezal (Argonne National Laboratory) \\ 
J.T. Childers (Argonne National Laboratory) \\
S. H{\"o}che (SLAC National Accelerator Laboratory) \\
O. Mattelaer (Universit\'e Catholique de Louvain) \\
J. McFayden (CERN)\\
S. Mrenna (Fermi National Accelerator Laboratory)\\
F. Petriello (Argonne National Laboratory \& Norwestern University)\\
S. Prestel (Fermi National Accelerator Laboratory)\\
C. Reuschle (Florida State University)}

\date{\today}

\def \Herwig {Herwig}
\def \Pythia {Pythia}
\def \mglo {MadGraph5\_aMC@NLO}
\def \mgnlo {MadGraph5\_aMC@NLO}
\graphicspath{{figures/}}

\begin{document}
\maketitle

\section{Introduction}

LHC experimental results continue to show excellent agreement with Standard Model predictions driving High Energy Physics into a new era of precision collider physics measurements. Theoretical predictions at the LHC and other colliders require perturbative calculations in Quantum Chromodynamics (QCD), the theory of the strong force that binds quarks and gluons into the colliding protons.  As a rough guideline to the required order in the QCD perturbative expansion, leading order (LO) only gets the order of magnitude of the result correct.  Next-to-leading order (NLO) gets results accurate to tens of percent.  Percent-level precision is possible at next-to-next-to-leading order (NNLO).

Event generators use Monte Carlo methods to calculate the interaction cross-sections of LHC proton-proton collisions, and allow to interpret each unweighted Monte-Carlo sampling point as collision event in the measurement. This feature greatly alleviates a direct comparison between theory and experiment, and it makes event generators a necessity for interpreting measurements and identifying new physics beyond the Standard Model.

The precision of LHC Run 2 analyses already rivals or exceeds the precision of theoretical predictions from NLO perturbative QCD \cite{AtlasZxs2017,AtlasWZxs2017}. As the LHC continues to break records in luminosity delivered and the High-Luminosity LHC (HL-LHC) comes online in 2025 experimental precision will continue to improve and the need for larger and larger simulated event samples will increase. The particle multiplicity per event continues to rise as well. These conditions drive an increase in the computing requirements of theoretical calculations and event generators. This chapter addresses plans for mitigating these increases and proposes solutions to ensure theory calculations and event generators are prepared to provide results in LHC Run 4 and beyond.

\subsection{Importance of Generators for LHC Experiments}

Event generators are the first step in the simulation chain of LHC experiments such as ATLAS and CMS. They calculate parton interaction cross-sections for LHC proton-proton interactions, then generate collision events based on these cross-sections. 
Figure~\ref{fig:atlas_grid_jobs_2016} shows the distribution of work performed by the ATLAS experiment on the grid in 2016 with 20\% going to event generation. ATLAS uses approximately 1.5 billion core-hours per year on the Grid and elsewhere. During the Run-1 (Run-2) MC production campaign, ATLAS generated 6.1 (12.5) billion events using 1.6 (91.4) CPU-time per event.

There are two main contributing factors in the difference between Run-1 and Run-2 CPU consumption. During Run-1 the usage of Les Houches event (LHE) files~\cite{Alwall:2006yp,Andersen:2014efa} (produced with off-grid resources) was commonplace whereas this has largely been replaced with centralised LHE file production by ATLAS in Run-2. Also, in Run-2 higher precision calculations, for example using NLO merging, are more prevalent and such calculations naturally take up signifacntly more resources.

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{atlas_grid_usage_2016}
\caption{Distribution of job types run by the ATLAS experiment in 2016.}
\label{fig:atlas_grid_jobs_2016}
\end{figure}

%\subsection{Challenges with current generators}

%\subsection{Characteristics of generators (dividing ME and PS, orders, etc) relevant for this discussion (not a detailed overview)}

\section{Current Landscape}

LHC experiments depend on many different physics event generators and theory calculations for both their similarities and their differences. The fact that theoretical approaches as well as implementations can vary offers an opportunity to validate and cross-check the predictions in a similar way that ATLAS and CMS provide systematic checks of experimental measurements.

The following section reviews some of the more heavily used generators in LHC experiments and some leading edge theory calculations.

\subsection{Leading-Order Event Generators}
\label{ref:lo_generators}
Ever since the first detailed, phenomenological studies of
fragmentation by Feynman and Field \cite{Field:1977fa}, computer
simulations of ``events'' have played an increasingly important role
in understanding and validating the Standard Model.   
The roots of event generators begin
in the attempt to describe the non-perturbative hadronization process. 
With the acceptance of QCD, the perturbative evolution of partons was
added, yielding the parton shower description.   
The combination of leading order differential cross sections for 
the simplest Standard Model processes with the parton
shower and hadronization
model gave a good description of jet and weak boson physics.
To understand physics at hadron colliders, models of beam remnant
evolution (so-called underlying event models) were developed 
analogously to the first models of hadronization.
The challenging search for
the top quark at the Tevatron collider, and the inherent weakness of
the parton shower approach in predicting high-$p_T$ jets, led to
interfaces with tree-level matrix-element based predictions.
By the end of the Tevatron era, precise measurements of the top quark
cross section had necessitated the development of consistent NLO +
parton shower predictions.

At the start of the LHC era, two of the main event generators had
migrated to C++ code (Herwig++ and Pythia8) while a third had started
from the beginning in C++ (Sherpa).

In recent years, our ability to make predictions at NLO and NNLO has
increased by leaps and bounds.   Nonetheless, event generators are
still the main workhorses of LHC experiments, for direct physics
measurements, but also for the important tasks of
estimating trigger rates, occupancy, detector performance, etc.
Event generator predictions are used almost exclusively to describe
jet physics.    Many exotic physics scenarios beyond the Standard
Model can also be studied with event generators.
However, as a shift has been made to making matrix element based
predictions, event generators take on the role of handling 
the matching/merging of these predictions with the parton shower,
underlying event, hadronization, etc. In any event, lowest-order 
event generators remain a crucial part of high-energy-physics
computing.   Besides modeling hadronization in general, they are
needed for a detailed description of pile-up, minimum bias events,
the underlying event, and jet structure and sub-structure.

To give an example, the modeling of hadronization, perturbative QCD
and soft QCD effects dominate the uncertainty on the top quark mass derived
from data \cite{Khachatryan:2015hba} and are a significant part of
the uncertainty on the $W$ boson mass \cite{Aaboud:2017svj}.
Data-driven pile-up modelling needs to be validated and cross-checked 
with event generators (see e.g.~\cite{TheATLAScollaboration:2013pia}), 
theoretical uncertainties on the (sub-)structure of jets need to be 
assessed by parton showering~\cite{Bellm:2016rhh,Mrenna:2016sih,Bothmann:2016nao} and 
the effects of high-jet-multiplicity background and signal events 
must be modeled using multi-jet-merged calculations \cite{Catani:2001cc,Lonnblad:2001iq,Krauss:2002up,Mrenna:2003if,Mangano:2006rw,
  Alwall:2007fs,Hoeche:2009rj,Hamilton:2010wh,Hoeche:2012yf,Platzer:2012bs,
  Lonnblad:2012ix,Lonnblad:2012ng,Frederix:2012ps}.

As another example, the simulation of pile-up events with event generators requires detailed models
of how the total proton-proton scattering cross section is distributed over
phase-space, and how different scattering processes (elastic scattering, 
diffractive scattering producing rapidity gaps, single- or multi-parton 
scattering) compete for the available phase space. These processes can
yield any number of detectable particles, with rare events containing 
a very large number of particles. The event generator model of pile-up-like
scatterings needs to avoid and bias like the bias that is e.g.\ introduced 
by cuts that are necessary to avoid singularities in perturbative
scattering calculations. This then means that a simultaneous integration of
processes with cross sections differing by many orders of magnitude is
necessary. In turn, this requires the generation of very large samples of
phase-space points. Currently, high-throughput computing is used to address this
problem.

Another computational frontier of leading-order event generators is the 
generation of perturbative variations that allow to assess the uncertainties
of the calculations. This includes both uncertainties in the modelling of
the hardest scattering that produces the partonic seeds of jets as well as
uncertainties in the evolution of partons into jets of hadrons. Both of these
facets have recently seen a move away from schemes to produce individual
uncorrelated variations based on high-throughput computation towards producing
statistically correlated systematic variations on-the-fly in a single 
event-generator run~\cite{Bellm:2016rhh,Mrenna:2016sih,Bothmann:2016nao}. The 
latter means that automated uncertainty estimates might
become amenable to high-performance computing in the near future.

Finally, the omnipresence of matching- and merging approaches to 
enable an accurate modeling of multi-jet final states (for both
signal- and background processes) has shifted much of the computational
challenges of modern event-generator calculations to the production of
fixed-order multi-parton scattering events. The phase space in which 
parton-showering is allowed to produce additional partons is severely 
restricted by matching/merging. Thus, for matched/merged calculations, the 
efficiency of the calculation is almost directly inherited from the
efficiency of the fixed-order (LO, NLO, \dots) calculation,
%~\cite{Mcfayden:2016}
such that the bottlenecks 
described in Section 2.2 impact the event generator calculation as well.
An example for the capabilities of modern tree-level event generators
is given in Tab~\ref{tab:lo_gen_perf}.
\begin{table}
  \begin{center}
  \begin{tabular}{lcccc}
    \hline
    Process & $W^-$+0j & $W^-$+1j & $W^-$+2j & $W^-$+3j \\\hline
%    Processes (mapped) & 1 (1) & 3 (3) & 18 (42) & 28 (68) \\
    RAM Usage & $<$1MB & $<$1MB & 1MB & 2 MB \\
    Initialization time & $<$1s & $<$1s & $<$1s & 2s \\
    Startup time & $<$1s & $<$1s & $<$1s & $<$1s \\
    Integration time & 8s & 2m 4s & 22m 8s & 2h 3m \\
    MC uncertainty [\%] & 0.18 & 0.14 & 0.25 & 0.44 \\
    \hline\\[1mm]
    \hline
    Process & $W^-$+4j & $W^-$+5j & $W^-$+6j & $W^-$+7j \\\hline
%    Processes (mapped) & 88 (324) & 116 (436) & 280 (1332) & 344 (1648) \\
    RAM Usage & 23 MB & 81 MB & 435 MB & 1.51 GB \\
    Initialization time & 33s & 3m 17s & 51m 52s & 6h 50m \\
    Startup time & $<$1s & $<$1s & 2s & 12s \\
    Integration time & 1d 5h & 6d 23h & 32d 19h & 64d 15h \\
    MC uncertainty [\%] & 0.66 & 0.78 & 1.29 & 2.70 \\
    \hline
  \end{tabular}
  \caption{Example performance measures for the computation of $pp\to W^-+n$-jets production at $\mathcal{O}(\alpha^2\alpha_s^n)$. 
    The center-of-mass energy is 13~TeV and jets are
    defined as $p_{T,j}>30$~GeV, $|y_j|<4.5$. All partonic channels are included
    in the calculation, and the CKM matrix is chosen to be diagonal.
    Numbers are generated on a dual 18-core Intel$^\text{\textregistered}$
    Xeon$^\text{\textregistered}$ 2.30GHz CPU. Integration times are cumulative for all MPI ranks in the parallel calculation.
    \label{tab:lo_gen_perf}}
  \end{center}
\end{table}
Note that a staged approach of pre-tabulating and storing fixed-order results 
before showering, as e.g.\ used in the 
\textsc{MadGraph5\_aMC@NLO} + \textsc{Pythia} or 
\textsc{Alpgen} + \textsc{Pythia} event-generation chain, can leads to 
computational overhead since certain pieces of the fixed-order calculation 
(i.e.~splitting histories) have to be reconstructed at shower run-time to 
guarantee a consistent combination. This overhead can become significant
at high parton multiplicity, but may be amendable to speed gains from
high-performance computing. The approach of generating fixed-order
results and showering simultaneously also suffers from computational 
challenges, as it leads to increased executable size and the loss of 
information about the partonic scattering event. It is therefore obvious
clear that developments that make fixed-order calculations amenable to 
high-performance gains alone will not improve the speed and efficiency of 
matched/merged event generator calculations, Instead, developments in 
fixed-order parts as well as showers (and how they are used to enable 
matching/merging) have to go hand-in-hand with the integration of codes to
allow for maximal computational gain.

\subsection{Next-to-Leading Order Event Generators}

Calculations and event simulation at NLO in perturbative QCD are used both 
for direct comparison with experimental data that are corrected to the 
parton level, and as an input to particle-level event simulation.
At the tree level, calculations have long been performed completely automatically 
using programs like Alpgen \cite{Mangano:2002ea}, Amegic \cite{Krauss:2001iv}, 
Comix \cite{Gleisberg:2008fv}, CompHEP \cite{Boos:2004kh}, HELAC \cite{Kanaki:2000ey}, 
MadGraph \cite{Alwall:2011uj} and Whizard \cite{Kilian:2007gr}.
At the next-to-leading order, such a level of automation required two main ingredients: 
The implementation of known generic methods to perform the subtraction of infrared 
singularities and the automated computation of one-loop amplitudes.

The calculation of virtual corrections for processes with a large number
of hard, light jets in the final state has been one of the focal points 
in the past decades. The field received a boost from generalized unitarity 
\cite{Bern:1994cg,Bern:1994zx,Bern:1997sc}, which is used in this context
to determine one-loop amplitudes by decomposing them into known scalar 
one-loop integrals and rational coefficients determined from on-shell 
tree amplitudes, plus a rational piece \cite{Ossola:2006us,Forde:2007mi,
  Ellis:2007br,Ossola:2008xq,Ellis:2008ir,Peraro:2014cba,Hirschi:2016mdz}. A large number of programs have been 
developed that automate these methods~\cite{Berger:2008sj,Ellis:2008qc,Ellis:2009zw,Bevilacqua:2011xh,Hirschi:2011pa,Alwall:2014hca,Badger:2010nx,Cascioli:2011va,Cullen:2011kv,Fleischer:2011bi} based on library computing the scalar integral \cite{Ellis:2007qk,vanHameren:2010cp,Hahn:1998yk,Patel:2016fam,vanOldenborgh:1990yc} 
and that supplement well-established codes based on analytic calculations
\cite{MCFM,Campbell:2010ff,Campbell:2015qma} and on improved automated 
tensor reduction \cite{Denner:2005nn,Binoth:2005ff,Cullen:2011xs,Cullen:2014yla,
  Cascioli:2011va,Actis:2012qn}. A completely numerical approach exists \cite{Becker:2010ng,Becker:2011vg}. 
All these developments have brought about the construction of ever faster 
numerical programs for phenomenological applications. Monte-Carlo methods 
can be used to reduce the number of loop-evaluations when computing NLO computation \cite{Alwall:2014hca} allowing to have NLO computations not dominated by loop-computations. An example for the arising timing distribution
is shown in Table \ref{table:MG5NLO}. Despite all these advantages, NLO
computations typically require significantly more computing power
than their leading-order counterparts. In addition, the numerical methods
to evaluate NLO corrections can suffer from numerical instabilities, such that re-evaluation in higher precision arithmetics becomes necessary. In this case the computation can be slowed down significantly.

\begin{table}
  \begin{center}
    \begin{tabular}{lccc}
\hline
process & Virtual & Tree-Level & Subtraction \\
\hline\hline
$p p \to t \bar t$ & 18\% & 17\% & 37\% \\
$p p \to t  \bar t j$ & 35\% & 23\% & 38\% \\
$p p \to t  \bar t j j $ & 31\% & 32\% & 34\% \\
\hline
$p p \to W^+ W^- $ & 10\% & 20\% & 48\% \\
$p p \to W^+ W^- j$ & 15\% & 36\% & 36\% \\
$p p \to W^+ W^- j j$ & 35\% & 38\% & 23\% \\
\hline
$p p \to e^+ e- $ & 4\% & 14\% &  26\% \\
$p p \to e^+ e- j $ & 3\% & 51\% & 34\% \\
$p p \to e^+ e- j j$ & 14\% & 40\% & 31\% \\

\hline
    \end{tabular}
    \caption{\label{table:MG5NLO} Relative CPU time dedicated to the evaluation of the virtual matrix-element, of the tree-level matrix-element (Born and Real) and of the various subtraction terms. Numbers are computed with {\sc MadGraph5\_aMC@NLO}. The difference between 100\% and the sums of virtual, tree-level and subtraction is due to other parts of the code (like generating/writing events, computation of systematics, etc.)}
   \end{center}
\end{table}

A major bottleneck is nowadays often found in the computation of 
real corrections and the corresponding infrared subtraction terms
\cite{Catani:1996vz,Catani:2002hc}.
The subtraction terms consist of color-correlated tree-level 
matrix elements multiplied by spin-dependent splitting operators, 
hence the aforementioned programs for leading order calculations 
are ideally suited for this part of the computation. Correspondingly,
infrared subtraction techniques have been implemented in various
general-purpose matrix element generators \cite{Gleisberg:2007md,
  Czakon:2009ss,Frederix:2008hu,Frederix:2010cj,Frederix:2009yq}.
Their reach is expanded sinificantly by parallel computing \cite{Hoche:2013zja,Bern:2013gka,Hoche:2016elu}, and it is expected
that this trend will continue in the future.

\subsection{NNLO Fixed-Order Calculations}

Achieving NNLO precision is an enormously difficult theoretical challenge.  Individual contributions to the expansion are separately divergent, necessitating regularization at intermediate stages.  It is also an enormous computational challenge.

These theoretical and computational challenges have received significant attention in the community  New methods have proven capable of providing the NNLO predictions needed to match the exacting experimental precision.  Those which have proven capable of handling the full complexity of $2 \to 2$ scattering with colored final states at the LHC are: antennae subtraction~\cite{GehrmannDeRidder:2005cm}; $N$-jettiness subtraction~\cite{Boughezal:2015dva,Gaunt:2015pea}; sector-improved residue subtraction~\cite{Czakon:2010td,Boughezal:2011jf}.  Together these techniques have made possible NNLO predictions for $W$+jet~\cite{Boughezal:2015dva}, $Z$+jet~\cite{Ridder:2015dxa,Boughezal:2015ded}, Higgs+jet~\cite{Boughezal:2015dra,Boughezal:2015aha,Chen:2014gva}, and $t\bar{t}$ production~\cite{Czakon:2013goa}, as well as partial results for di-jet production~\cite{Ridder:2013mf}.   We demonstrate the utility of these higher-order calculations using a result for the $Z$+jet process obtained using $N$-jettiness subtraction~\cite{Boughezal:2016yfp}.   The figure shows a plot for the $H_T$ distribution in the $Z$+jet process.  This observable provides an important control sample in the search for dark matter at colliders.  The data points as measured by the experimental collaborations are shown as points with error bars.  Theory predictions at NLO and NNLO are shown as red and blue bands respectively, where the bands denote an estimate of the residual theoretical error.  The NLO result falls below the measured data by nearly a factor of two, and fails to properly describe the distribution shape.  The NNLO prediction obtains both the correct normalization and shape.  More examples and a detailed discussion of the theoretical precision are given in Ref.~\cite{Boughezal:2016yfp}, but the message of this plot is clear: one simply cannot describe this important LHC data without a NNLO prediction.

\begin{figure}
\centering
\includegraphics[width=3.21in]{ATLAS-Z-7TeV_HT}
\includegraphics[width=3.21in]{CMS-Z-7TeV_HT}
\caption{Comparison of the $H_T$ distribution in the $Z$+jet process measured by ATLAS (left panel) and CMS (right panel) to theoretical predictions at NLO and NNLO.  The data points denote the experimental measurements with the reported errors, while the red and blue bands respectively denote the estimated NLO and NNLO theoretical errors.}
\label{fig:CMSZ}
\end{figure}

We discuss the computational aspects and challenges at NNLO using $N$-jettiness subtraction as an example.  The $N$-jettiness subtraction idea utilizes a novel computational approach that relies critically on parallel high-performance computing (HPC) systems.  The $N$-jettiness subtraction scheme achieves NNLO precision for LHC scattering processes by splitting the necessary integrations over the phase space of the $N$ final-state particles according to a resolution criterion on the $N$-jettiness event shape variable:
%
%\vspace{-.15in}
\begin{equation} \label{eq:PSsplit} 
\sigma_{{\rm NNLO}}=\int d\Omega \, \left[ \frac{{\rm d}\sigma_{{\rm NNLO}}}{{\rm d} {\cal T}_N} 
	\theta({\cal T}_N^{cut}-{\cal T}_N)
	+ \frac{{\rm d}\sigma_{{\rm NNLO}}}{{\rm d} {\cal T}_N} \theta({\cal T}_N-{\cal T}_N^{cut}) \right].
\end{equation} 
%
The $N$-jettiness variable ${\cal T}_N$ was defined in Ref.~\cite{Stewart:2010tn}.  For our purposes we note that it divides the phase space into an `above-cut' region (the second term in Eq.~\ref{eq:PSsplit}, where ${\cal T}_N>{\cal T}_N^{cut}$) which can be obtained from a simpler NLO calculation for $N+1$ jets.  The `below-cut' region (the first term in Eq.~\ref{eq:PSsplit}, where ${\cal T}_N<{\cal T}_N^{cut}$) can be obtained from an all-orders result in the strong coupling constant~\cite{Stewart:2009yx}.  A detailed discussion of how this technique is practically implemented was presented in the initial work.  The end result is the reduction of the number of required integrations to three: a computationally-simple below-cut integration, and two computationally-intensive above-cut integrations.  These integrations typically live in 10-15 dimensions and contain numerous peaks throughout phase space, and are approached with Monte-Carlo techniques such as the VEGAS algorithm~\cite{Lepage:1977sw}.

The primary computational challenge is that each integral in Eq.~(\ref{eq:PSsplit}) contains logarithms of the resolution 
parameter ${\cal T}_N^{cut}$ that cancel upon addition of all three terms.  Each separate term has up to ${\rm log}^4 ({\cal T}_N^{cut})$ dependence.  Since ${\cal T}_N^{cut}$ must be taken small for theoretical reasons, we must dig out the residual result from the numerical noise induced by the cancellation between large logarithms in the numerical integration.  The figure-of-merit to consider is the degree of cancellation between the above-cut and below-cut integrations.  In applications considered so far this ranges from 1 part in 10 to 1 part in 1000.   The desired numerical precision on the final result is sub-percent, implying that some of the separate integrals must be computed to $10^{-6}$ relative accuracy.  To reach this target for a Monte Carlo integration with a large number of dimensions requires an enormous number of integrand evaluations.  We split these between many threads, which must be coupled in order to use this information to appropriately refine the integration grid.  This explains the need for a massively parallel HPC resource.

The  $N$-jettiness subtraction method is a general theoretical framework that can be implemented in publicly available  codes for NLO calculations.  Several independent codes incorporating $N$-jettiness subtraction have been successfully run on the Mira supercomputer at the Argonne Leadership Computing Facility (ALCF)~\cite{Boughezal:2016wmq,Abelof:2016pby}.  These are written in a combination of Fortran77 and Fortran90.   All codes performs a Monte-Carlo integration of multi-dimensional integrands.  The many required integrand evaluations are the most computationally expensive part of the code, are parallelized using a hybrid MPI-OpenMP approach.  Reasonably good scaling up to the million-thread level is obtained on the Mira supercomputer at the ALCF.  As more complex scattering processes are studied, this scaling will become increasingly important.

\subsection{Parton Showers at Higher Order}
Parton shower algorithms describing QCD and QED multiple radiation have been
a central ingredient of simulation programs for particle physics experiments
at the energy frontier~\cite{Webber:1986mc,Buckley:2011ms}. Invented about
three decades ago, they have undergone tremendous developments since then, 
with some of the most important steps being the introduction of QCD coherent evolution~\cite{Webber:1983if,Marchesini:1983bm,Marchesini:1987cf} and the 
incorporation of process-independent soft-gluon enhacement~\cite{Catani:1990rr}.

About a decade of work has been spent on matching~\cite{Frixione:2002ik,
  Hoeche:2011fd,Hamilton:2012rf,Hamilton:2013fea,Hoeche:2014aia,Hoche:2014dla}
and merging~\cite{Catani:2001cc,Lonnblad:2001iq,Krauss:2002up,Mangano:2006rw,
  Alwall:2007fs,Hoeche:2009rj,Hamilton:2010wh,Hoeche:2012yf,Platzer:2012bs,
  Lonnblad:2012ix,Lonnblad:2012ng,Frederix:2012ps} algorithms to include 
fixed-order higher-order corrections. More recently, the need for improved 
overall simulations triggered a resurgence of interest in improving 
parton-shower algorithms themselves. Several new parton showers have been 
constructed~\cite{Gieseke:2003rz,Sjostrand:2004ef,Schumann:2007mg,Dinsdale:2007mf,
  Winter:2007ye,Giele:2007di,Platzer:2009jq,Ritzmann:2012ca,Hoche:2015sya}, 
and the possibility of including next-to-leading order corrections into 
the evolution has been explored~\cite{Hartgring:2013jma,Li:2016yez,
  Hoche:2017iem,Hoche:2017hno,Nagy:2017ggp}. In addition, NLO splitting functions 
have been recomputed~\cite{Jadach:2011kc,Gituliar:2014eba}, with the aim 
to improve parton-shower simulations, and the dependence of NLO matching terms on the 
parton-shower evolution variable has been investigated~\cite{Jadach:2016zgk}.

While all these developments may lead to theoretically improved simulations rather soon,
they also require a significant increase in computation time for the generation
of each event. As a simple example, the implementation of next-to-leading order
accurate splitting functions in the parton shower effectively turns the previously
probabilistic simulation of the parton branching process into a true NLO
calculation. The occurrence of negative weights for the splitting kernels
is characteristic and leads to additional weight fluctuations in the overall 
simulation that have to be compensated by a higher number of generated events,
while each event also comes at an additional computational cost. This situation 
is very much alike to fixed-order perturbation theory.

%\subsection{Current Performance Measures}

%Table~\ref{tab:gen_perf} shows the average time (in seconds) it takes to generate one ttbar event on the Grid. The averages include number of sites to make them representative.

%\begin{table}
%\centering
%\caption{My caption}
%\label{tab:gen_perf}
%\begin{tabular}{l|c|c}
%generator & process & seconds per event \\
%\hline
%Sherpa & ttbar all-hadronic & $49\pm13$ \\
%Pythia8 & ttbar & $0.012\pm0.005$ \\
%Powheg+Pythia & ttbar & $82\times10^{-6}\pm24\times10^{-6}$ \\
%MadGraph+Pythia & ttbar & $0.0012\pm0.0005$ \\
%aMC@NLO+Pythia & ttbar & $0.014\pm0.004$ 
%\end{tabular}
%\end{table}

%\begin{itemize}
%\item events/core-hour ?
%\item  LO vs NLO ? NNLO?
%\item Integration vs Generation ?
%\item Production time spent by the experiments ?
%\end{itemize}

\section{Future Strategies}

The following techniques will be investigated with the goal of increasing the precision of generators needed for analyses of the HL-LHC while aiming to reduce our computing requirements.

%\subsection{Evolutionary Techniques}

\subsection{Scalable Adaptive Monte-Carlo Integrators}

As generators move to higher orders in perturbative QCD and the HL-LHC probes higher multiplicity processes, the time needed to integrate the interaction cross-section becomes prohibitively large on a single CPU core. Parallel processing becomes a requirement and being able to scale on a supercomputer enables ever more aggressive analyses.

Existing generators were not developed in the mulit-core age that has become the industry standard in the past five years and therefore parallel processing was added after frameworks had largely been implemented. This leads to meager performance improvements when running in parallel, especially when running in parallel across many compute nodes of a supercomputer. Work has been done to improve scalability of existing generators such as Sherpa and Alpgen~\cite{AlpgenScaling2017}.

In the case of advanced frameworks such as MadGraph and Sherpa, the architecture of the framework was not designed to take advantage of highly parallel computers and maintain high CPU and memory efficiency. The development of new scalable adaptive Monte-Carlo integrators that can be used by existing frameworks would be a big step toward addressing HL-LHC computing requirements. In the US, the DOE is planning to deploy supercomputers over the next five years that will increase the scientific computing capacity by a factor of 200. Developing algorithms that target these machines will offload computation from Grid resources and will in general lead to the development of better algorithms.

% The full phase-space integral for multi-particle interactions factorizes into lower-dimensional integrals corresponding to $2\rightarrow2$-scattering and $1\rightarrow2$-decay processes. This can be used as a guiding principle to create integrators for high-multiplicity processes. The factorization is typically performed for each tree graph in each partonic channel in order to map out all known peaks of the integrand. Interferences between the Feynman graphs can be dealt with using the multi-channel method~\cite{Kleiss:1994qy} and the single-diagram enhanced integration technique~\cite{Maltoni:2002qb}. The latter may be well suited for parallelization if the integrand can be divided into a large number of graphs, which is always the case in high-multiplicity processes.

% Creating a new common integration tool which scales from laptop to HPC is a goal of the Generator \& Theory community. Such a tool will require evaluating the existing methods and studying the latest Monte-Carlo and quasi-Monte-Carlo techniques. The goal should be to introduce more parallelism and provide intelligence to determine at which scale different methods may lead to faster results for a target accuracy. Such a tool should use standard libraries that are compatible on clusters and HPCs, such as MPI or OpenMP (the latest version of which can target both CPUs and GPUs). It must be compatible with existing generator frameworks for use as a plugin.

\subsection{Reweighting of Event Samples}

Re-weighting method consists to attach to a given partonic events a new weight corresponding to a new theory. Those new weights allow to predict accurately (up to statistical precision) all the differential distributions at parton-level and to perform a single detector simulation for all the models under consideration.

Different types of re-weighting exist. First there is the reweighting mentioned in Sec.~\ref{ref:lo_generators}, which relates to the estimation of the theoretical uncertainties of a simulation. 
Second the same techniques can be used to accomodate BSM re-weighting for both LO \cite{Gainer:2014bta,Baglio:2014uba,Kalinowski:2016qcd,Mattelaer:2016gcx} and  NLO event samples, both approximatively \cite{Alioli:2010xd,REPOLO} and exactly \cite{Mattelaer:2016gcx}.

At LO accuracy the new weight ($W_{new}$) can be easily obtained from the original one ($W_{orig}$) by simply multiplying it by the ratio of the matrix-elements estimated on that event for both models (noted respectively $|M_{orig}|^2$ and $|M_{new}|^2$):
\begin{equation}\label{me_re-weighting}
W_{new} = \frac{|M_{new}|^2}{|M_{orig}|^2}W_{orig}.
\end{equation}
While at NLO, one needs to track the dependence in the various event and counter-event and reweight by the associated matrix-element.

A few remarks are in order regarding the range of validity of this method. First, even if the method returns the correct weight, it requires that the event sampling related to $W_{orig}$ covers appropriately the phase-space for the new theory. In particular, $W_{orig}$ must be non-zero in all regions where $W_{new}$ is non-vanishing. Though obvious, this requirement is in fact the most important and critical one. 
In other words, the phase-space where the new theoretical hypothesis contributes should be a subset of the original one. For example, re-weighing can not be used for scanning over different mass values of the final state particles\footnote{For intermediate particle a small variation of the mass --order of the width-- is reasonable.}, yet it is typically well-suited for probing different types of spin and/or coupling structures.

Second, the parton-level configuration feeder to parton-shower programs not only depends of the four-momenta but also of additional information, which is commonly encoded in the Les Houches Event File (LHEF) \cite{Alwall:2006yp,Andersen:2014efa}. Consequently, re-weighting  by an hypothesis that does not preserve such  additional information is not accurate. In general, such informations are related to:

\begin{itemize}
\item {\bf Helicity:} The helicity state of the external states of a parton-level event is optional in the LHEF convention, yet some programs (e.g. \cite{Meade:2007js}) use this information to decay the heavy state with an approximated spin-correlation matrix. \cite{Mattelaer:2016gcx,Artoisenet:2012st}
proposes a dedicated re-weihting method in this case to fix such issue.

\item {\bf Color-flow:} A second piece of information presented in the LHEF is the color assignment in the large $N_c$ limit. This information is used as the starting point for the dipole emission of the parton shower and therefore determines the result of the QCD evolution  and hadronisation.  Such information is untouched by the re-weigthing limiting the validity of the method.

\item {\bf Internal resonances:} In presence of on-shell propagators, the associated internal particle is written in the LHEF. This is used by the parton-shower program to guarantee that the associated invariant mass is preserved during the re-shuffling procedure intrinsic to the showering process. Consequently, modifying the mass/width of internal propagator should be done with caution since it can impact the parton-shower behaviour. This information can not be corrected via a re-weighting formula, as it links in a non-trivial way short-distance with long-distance physics.
\end{itemize}

Even with such limitation the application of such re-weighting will allow to decrease by huge factor the time spent on the generation of BSM samples.

\bibliographystyle{abbrv}
\bibliography{main}

\end{document}

### 2020-04-16 LHCC/2020/2020-01/ctcs.tex
Proof of principle conversion to LaTeX

% This is the styling for submission to arXiv using JHEP Preprint Style

% Copyright (C) 2020, the authors, licence CC-BY-4.0.

% JHEP preprint template
\documentclass[11pt,a4paper]{article}
\usepackage{jheppub}
\notoc

\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage[titletoc]{appendix}
\usepackage{xspace}

\usepackage[utf8]{inputenc}

%% Comment out before final submission
\usepackage{lineno}  % for line numbering during review
\linenumbers

\begin{document}

% HSF Document Number (only for arXiv)
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-DOC-2020-01 \\
 & & Zenodo DOI Here \\
 & & May 1, 2020 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{Common Tools and Community Software}

\author{High-Energy Physics Software Foundation}

\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The challenges for future software for high-energy physics are driven by
an ambitious physics programme, notably the LHC accelerator upgrade to
high-luminosity, HL-LHC, and the corresponding detector upgrades of
ATLAS and CMS. The General Purpose Detectors describe their specific
challenges elsewhere; in this document we address the challenges for
software that is used in multiple experiments (usually even more widely
than ATLAS and CMS) and maintained by teams of developers who are either
not linked to a particular experiment or who contribute to common
software within the context of their experiment activity. We also give
space to general considerations for future software and projects that
tackle upcoming challenges, no matter who writes it, which is an area
where community convergence on best practice is extremely useful.

ATLAS and CMS will undergo major detector upgrades and will also
increase their trigger rates for HL-LHC by about a factor of 10; event
complexity will rise, with peak pile-up of 200, far higher than in
Run-2. This places an enormous burden on storage and processing
resources. Current CMOS microprocessor technology is clock speed limited
(due to the failure of Dennard scaling) and, while processors
per-integrated circuit still keeps rising, Moore's Law is expected to
stall during the 2020s. More importantly, the effective runtime related
improvements in computing from CPU servers at sites is likely to be only
in the 5-10\% per year range, making the shortfall in processing
resources more severe. As manufacturers struggle to deliver ever-more
effective computing through CPUs the drive to different architectures
intensifies, with GPUs becoming commonplace and increasing interest in
even more specialised architectures, such as TPUs, IPUs and developments
that make FPGA devices more user friendly. These pose huge challenges
for our community as the programming models and APIs vary widely here
and possible lock-in to a particular vendor's devices is a significant
problem for code sustainability and preservation. Huge work has been
done already to adapt to this changing landscape, e.g. multi-threading
software and GPU codes have been used in production by some experiments
for years. In other areas migration is still ongoing. In others it is
not even yet started. Generic heterogeneous programming models have
existed for some time, and new ones are arising, but there is, as yet,
no clear winner, in-part because the performance obtained can be highly
application-dependent. HEP itself will not drive the success of one
model or another, so even if the community converged on a prefered
model, its long term success would not be assured. The C++ standard
itself is likely to lag for many years behind what is required by us in
the area of heterogeneous or even distributed computing. Further,
experiment frameworks (and with knock-on effects to systems like
workload management) will need to adapt to sites that provide
heterogeneous resources. How to do this, and make effective use of
different heterogeneous resources across different sites, remains far
from settled.

For storage systems (see the DOMA and WLCG documents) the pressure on
software is to store as much physics information in as few bytes as
possible, but also to be able to read at very high rates to deliver data
from modern storage technologies into the processing hardware. This
requires critical developments in the storage formats and libraries used
in HEP, e.g. developments like RNTuple for ROOT I/O is likely to be of
great importance for the community. The likelihood of finding an
off-the-shelf compact and efficient storage format for HEP data is
remote, so investment in smart software to support our PB sized science
data is simply cost effective. Particularly for analysis, which is
usually I/O bound, we have an end-to-end problem from storage
technology, through the software layers, to processing resources that
may well span multiple nodes. Other workflows, which are less dependent
on I/O rates will, nevertheless, have to be adapted to using remote data
where the I/O layer must optimise data transfers and hide latency, e.g.
taking advantage of xrootd's single column streaming ability.

In this increasingly complex environment in which to write software,
there are important problems where sharing information at the community
level is far more efficient. Providing a high level of support in the
build environment for developers, sharing knowledge about how to
measure, and then improve, performance (especially on multiple different
architectures) and sharing best practice for code development can have a
large integrated benefit. This requires improved training and the
development of a curriculum for all developer levels. In the field, such
initiatives are often warmly received, but real support is needed for
those who can put this into practice, also ensuring that their work in
training contributes to their career development and a long term future
in the field. HEP software stacks are already deep and wide and building
these consistently and coherently is also an area where knowledge can be
shared. Support is needed for multiple architectural targets and
ensuring the correct runtime on heterogeneous environments.

This brings up the important question of validation and the need to
improve the security of physics results, which is even more complicated
on heterogeneous platforms, when exact binary compatibility often cannot
be assured. Currently almost every experiment and project has its own
infrastructure for this.

Once software is built, it needs to be shipped worldwide so that the
production workflows can run. CernVM-FS was a huge leap forward for
software distribution and has even been widely adopted outside HEP.
However, new challenges arise, with container based payloads, scaling
issues and disconnected supercomputer sites. So maintenance and
development needs to be undertaken to support and extend this key
supporting infrastructure for software.

Finally, over the multi-decade lifetimes of HEP experiments, we need to
preserve both the core and analysis software so that results can be
confirmed and updated as the field moves on. There are many exciting
developments based around CernVM-FS, containers and things like analysis
description languages, but these are not yet at the stage of being
settled nor integrated into our day-to-day workflows.

In the rest of this document the main issues associated with the key
parts of the software workflow in high-energy physics are presented,
focusing on those that dominate current resource consumption: physics
event generation, detector simulation, reconstruction and analysis.

\hypertarget{physics-event-generators}{%
\section{Physics Event Generators}\label{physics-event-generators}}

\hypertarget{introduction-1}{%
\subsection{Introduction}\label{introduction-1}}

Physics event generators are essential in HEP. All of the LHC scientific
results, both precision measurements or searches for new physics, depend
significantly on the comparison of experimental measurements to
theoretical predictions computed using generator software.

Using Monte Carlo (MC) techniques, generators allow both the generation
of unweighted events for experimental studies of differential
distributions and the prediction of total cross sections. The
large-scale event generation campaigns of the LHC experiments have
significant computational costs, mainly in terms of CPU resources. The
limited size of simulated samples is a source of major uncertainty in
many analyses and is therefore a limiting factor on the potential
physics output of the LHC programme. This situation will get
significantly worse in the high luminosity running phase of the LHC
(HL-LHC). The fraction of the CPU resources available to the LHC
experiments which is used for event generation today is currently
estimated to be approximately 5-15\%. As is the case for the other big
consumers of CPU (detector simulation and reconstruction, that have
until now been the main focus), speedups in generator software are also
needed to address the overall resource problem expected at the HL-LHC,
also because more accurate predictions, requiring more complex
calculations (e.g. beyond NLO or with higher jet multiplicities) will be
needed from generators. Many other issues, both technical and
non-technical (e.g. funding, training, careers for those working in this
area), exist {[}Alb19,Gen18{]}.

A HSF Working Group (WG) on generators {[}Gen19{]} was set up at the
beginning of 2019. The main focus of the WG so far has been on gaining a
better understanding of the current situation, and identifying and
prioritising the areas where computing costs can be reduced. In
particular, the WG has been active in analysing the ATLAS and CMS
compute budgets in detail, in profiling MG5\_aMC {[}Alw14{]}, Sherpa
{[}Bot19{]} and Powheg {[}Fri17{]}, in discussing the possible sharing
of common parton-level samples by ATLAS and CMS, and in reviewing and
supporting the efforts for porting generators to modern architectures
(e.g., MG5\_aMC to GPUs). This last activity is particularly important,
as it has become increasingly clear that being able to run
compute-intensive WLCG software workloads on GPUs would allow the
exploitation of modern GPU-based supercomputers at High Performance
Computing (HPC) centers, and generators look like a natural candidate
for this.

This document gives an overview of the many technical and non-technical
challenges in the generator area and of the work that can be done to
address them. This is finally condensed into a list of a few
high-priority items, for the next 18 months before the LHCC review that
is currently scheduled for fall 2021.

\hypertarget{collaboration-challenges}{%
\subsection{Collaboration challenges}\label{collaboration-challenges}}

1. {The landscape of generator software is extremely varied}. Different
generators are used for different processes. Generating a sample also
involves choices of precision (e.g. Leading Order, LO, or
Next-to-Leading-Order, NLO), hadronization and Parton Shower (PS)
models, underlying event tunes, prescriptions for matching/merging and
simulating particle decays, and other input parameters, chiefly among
them the parton density functions, for which different interfaces exist.
Various combinations of software libraries are thus possible, often
written by different authors and some dating back many years. For a
given process, the LHC experiments often use different software packages
and settings from each other, and a single experiment can generate
events using more than one choice. Many different packages and
configurations may therefore need to be worked on to get cumulative CPU
cost reductions. The large number of external packages also complicates
their long-term maintenance and integration in the experiments software
and workflows, sometimes leading to job failures and computing
inefficiencies. Other packages are also absolutely critical for the
whole generator community and must be maintained, even if their CPU cost
is relatively low (LHAPDF, Rivet, HepMC, etc.).

2. {A very diverse combination of skills/profiles are needed}: theorists
(who create fundamental physics models, and design, develop and optimize
most generator code), experimentalists working in research (who request
types/sizes of samples) and in computing (who implement, monitor and
account execution of workflows on computing resources), software
engineers and system performance experts. This is a richness and
opportunity, as some technical problems are best addressed by people
with specific skills, but it also poses some challenges:

- {Training challenges}\textbf{.} Theorists and experimentalists often
lack formal training in software development and optimization. Software
engineers and experimentalists are often not experts in the theoretical
physics models implemented in MC codes.

- {Communication challenges}. It is difficult to find a shared
terminology and set of concepts to understand one another: notions and
practices that are taken for granted in one domain may be obscure for
others. As an example, there are many articles about the physics in
generators, but software engineers need papers describing the software
modules and overall data/control flow.

- {Career challenges}. Those working in the development, optimization or
execution of generator software provide essential contributions to the
success of the HL-LHC physics programme and it is critical that they get
the right recognition and motivation. However, theorists get recognition
on published papers, and may not be motivated to work on software
optimizations that are not "theoretical" enough to advance their
careers. Generator support tasks in the experiments may also not be
enough to secure jobs/funding to experimentalists pursuing a career in
research.

- {Mismatch in optimization focus}. The goals and metrics of software
optimization work may also differ. Theorists are mainly interested in
calculating cross sections and focus on minimising the phase space
integration time for a given statistical precision. The LHC experiments
run large scale productions of unweighted event generation, and mainly
need to maximize the throughput of events generated per unit time on a
given node.

- {Programming languages}: Attracting collaborators with a computer
science background to work on generators, especially students, may also
be complicated by the fact that critical components of some generator
packages are written in Fortran. In some cases, alternative
implementations of these components in other languages exist, but their
speed is significantly lower than the optimized Fortran version. Some of
the generators also do not use industry standard version control
systems, making it harder to contribute code.

\hypertarget{technical-challenges-software-and-computing}{%
\subsection{Technical challenges (software and
computing)}\label{technical-challenges-software-and-computing}}

The event generation workflow presents several challenges and
opportunities for improvement.

1. There are many sources of inefficiency in unweighted event
generation:

- {Phase space sampling inefficiency}. Efficient sampling is the most
critical ingredient for efficient unweighted event generation. Many
generic algorithms exist (e.g. VEGAS {[}Lep80{]}, BASES / SPRING
{[}Kaw86{]}, MINT {[}Nas07{]}, FOAM {[}Jad03{]}), as well as others
developed specifically for a given generator (e.g. MadEvent {[}Mal03{]},
itself based on a modified version of VEGAS, in MG5\_aMC, or COMIX
{[}Gle08{]} in Sherpa). In general, the larger the dimensionality of the
phase space, the lower the unweighting efficiency that can be achieved:
in W+jets, for instance, the Sherpa efficiency is 30\% for W+0 jets and
0.08\% for W+3jets {[}Gao20{]}. This is an area where research is very
active, and should be actively encouraged, as significant cost
reductions in WLCG compute budgets could be achieved. Improvements in
this area can only start from physics-motivated approaches based on the
knowledge of phase space peaks, but they can be complemented by
brute-force machine learning (ML) algorithmic methods {[}Ben17, Bot20,
Gao20, Kli18{]}.

- {Merging inefficiency}. Merging prescriptions (e.g. MLM {[}Man02{]},
CKKW-L {[}Lon05{]} at LO and FxFx {[}Fre12{]} at NLO) imply the
rejection of some events, to avoid double counting between events
produced with n\textsubscript{jets}+1 MEs and with n\textsubscript{jets}
MEs plus parton showers. The resulting inefficiencies can be relatively
low depending on the process, but they are unavoidable in the
algorithmic strategy used by the underlying physics modeling
{[}Alw08{]}. However, a method like shower-kt MLM can reduce the merging
inefficiency of MLM {[}Alw09{]}.

- {Filtering inefficiency}. An additional large source of inefficiency
is due to the way the experiments simulate some processes, where they
generate large inclusive event samples, which are then ﬁltered on
ﬁnal-state criteria to decide which events are passed on to detector
simulation and reconstruction (e.g. CMS simulations of specific
$\Lambda_{B}$ decays have a 0.01\% efficiency and ATLAS B-hadron
filtering has \textasciitilde10\% efficiency for V+jets). This
inefficiency could be reduced by developing ﬁltering tools within the
generators themselves, designed for compatibility with the requirements
of the experiments. Filtering is an area where LHCb has a lot of
experience and already obtained significant speedups through various
techniques. The speed of color reconnection algorithms is a limiting
factor for simulating rare hadron decays in LHCb.

- {Sample sharing and reweighting}. In addition to removing
inefficiencies, other ways could be explored to make maximal use of the
CPU spent for generation by reusing samples for more than one purpose.
Sharing parton-level samples between ATLAS and CMS is being discussed
for some physics analyses. Event reweighting is already commonly used
(e.g. for new physics searches and some systematic uncertainty
calculations) and could be explored further, though this may require
more theoretical work for samples involving merging or NLO matching
{[}Mat16{]}.

- {Negative weights}. Matching prescriptions, like MC@NLO, are required
in (N)NLO calculations to avoid double counting between (N)NLO matrix
elements and parton showers, leading to the appearance of events with
negative weights. This causes a large inefficiency, as larger event
samples must be generated and passed through the experiment simulation,
reconstruction and analysis codes, increasing the compute and storage
requirements. For a fraction r of negative weight events, the number of
events to generate increases by a factor 1/(1-2r)\^{}2: for instance,
with r=25\% (which may be regarded as a worst-case scenario in ttbar
production {[}Fre20{]}), one needs to generate 4 times as many events.
Negative weights can instead be almost completely avoided, by design, in
another popular matching prescription, POWHEG, which however is only
available for a limited number of processes and describes the relevant
physics to a different degree of precision than MC@NLO (see {[}Fre20{]}
for a more in-depth discussion). Progress in this area can only be
achieved by theorists: for instance, a new approach for MC@NLO-type
matching with reduced negative weights has recently been proposed
{[}Fre20{]}. It should be noted that negative weights can also happen at
LO because of not-positive-definite parton density function sets and
interference terms, e.g. in effective field theory calculations.

2. While progress has been made to better understand which areas of
generator software have the highest computational cost, more detailed
accounting of the experiment workloads and profiling of the main
generator software packages would help to further refine R\&D
priorities:

- {Accounting of CPU budgets for generators in ATLAS/CMS}. Thanks to a
lot of effort from the generator teams in both experiments, a lot of
insight into the settings used to support each experiment's physics
programme was gained within the WG, and it is now clear that the
fraction of CPU that ATLAS spends for event generation is somewhat
higher than that in CMS. More detailed analysis of the different
strategies is ongoing. A practical issue is that these figures had to be
harvested from logs and production system databases, which was
particularly difficult for CMS, requiring significant person hours to
extract the required information. It is important to establish better
mechanisms to collect this information, to allow for an easy comparison
between different experiments.

- {Profiling of the generator software setups used for production}.
Another area where the WG has been active is the definition and
profiling of standard generator setups, reproducing those used in
production. Detailed profiling could also be useful to assess the CPU
cost of external parton distribution function libraries {[}Kon20{]}, or
the memory requirements of the software (which may motivate a move to
multithreading or multiprocessing approaches).

3. More generally, as is the case in other areas of HEP, some R\&D on
generator software would certainly be needed to modernise it and make it
more efficient, or even port it to more modern computing architectures
{[}Bau13, Alb19{]}:

- {Data parallelism, GPUs and vectorization}. The data flow of an MC
generator, where the same (matrix element) function is computed
repeatedly at many phase space points, lends itself naturally to the
data parallel approaches found in CPU vectorized code and in GPU compute
kernels. Porting and optimizing generators on GPUs is essential to be
able to use modern GPU-based HPCs (such as SUMMIT, where 95\% of the
compute capacity comes from GPUs). The work done in this direction in
the past on MG5\_aMC, that never reached production quality, is now
being revamped by the WG, in collaboration with the MG5\_aMC team, and
represents one of the main R\&D priorities of the WG. This work is
presently focusing on Nvidia CUDA, but abstraction libraries will also
be investigated. GPUs may also be relevant to ML-based sampling
algorithms and to the pseudo-random number generation libraries used in
all MC generators.

- {Task parallelism, multithreading and multiprocessing}. Generators are
generally executed as single threaded software units. In most cases,
this is not a problem, as the memory footprint of unweighted event
generation is small and usually fits within the 2 GB per core available
on WLCG nodes. However, there are cases where memory is higher than 2 GB
(e.g. DY+jets using Sherpa); this leads to inefficiencies as some
processor cores remain unused, which could be avoided using
multithreading approaches. The fact that some generators are not even
thread safe, may also be a problem, for instance to embed them in
multi-threaded event processing frameworks, such as that of CMS.
Multi-processing approaches may also be useful to speed up the
integration and optimization step for complex high-dimensional final
states. In particular, a lot of work has been done to implement
MPI-based multi-processing workflows for generators in recent years. For
instance, the scaling of LO-based generation of merged many-jet samples
has been successfully tested and benchmarked on HPC architectures using
both Alpgen {[}Chi17{]} and Sherpa {[}Hoe19{]}; in the latter case, new
event formats based on HDF5, replacing LHEF, have also been instrumental
in the success of these tests. MPI integration has also been completed
for MG5\_aMC {[}Mat18{]}. It should also be noted that, even if HPCs
offer extremely high-speed inter-node connectivity, it is perfectly ok
for WLCG workflows to use these systems as clusters of unrelated nodes.

- {Generic code optimizations}. A speedup of generators may also be
achievable by more generic optimizations, not involving concurrency. It
should be studied, for instance, if data caching {[}Kon20{]} or
different compilers and build strategies may lead to any improvements.

\hypertarget{physics-challenges-e.g.-nnlo}{%
\subsection{Physics challenges (e.g.
NNLO)}\label{physics-challenges-e.g.-nnlo}}

In addition to software issues, important physics questions should also
be addressed about more accurate theoretical predictions (above all NNLO
QCD calculations, but also electroweak corrections) and their potential
impact on the computational cost of event generators at HL-LHC. Some
specific NNLO calculations are already available and used today by the
LHC experiments in their data analysis. With a view to HL-LHC, however,
some open questions remain to be answered, in particular:

1. {NNLO: status of theoretical physics research}. The first question is
for which processes NNLO precision will be available at the time of the
HL-LHC. For example, when would NNLO be expected for 2-\textgreater3 or
even higher multiplicity final states? And for lower multiplicity final
states, where differential NNLO predictions exist but the generation of
unweighted NNLO+PS events is not yet possible. It is important to
clarify the theoretical and more practical challenges in these
calculations, and the corresponding computational strategies and impact
on CPU time needs.

2. {NNLO: experimental requirements for data analysis at HL-LHC}. The
second question is for which final states unweighted event generation
with NNLO precision would actually be required, and how many events
would be needed. One should also ask if reweighting LO event samples to
NNLO would not be an acceptable cheaper alternative to address the
experimental needs.

3. {Size of unweighted event samples required for experimental analysis
at HL-LHC}. Another question is in which regions of phase space the
number of unweighted events must be strictly proportional to the
luminosity. For example, in the low pT regions of W boson production it
is probably impossible to keep up with the data, due to the huge cross
section. Alternative techniques should be investigated, to avoid the
generation of huge event samples.

\hypertarget{summary---generator-community-current-priorities}{%
\subsection{Summary - generator community current
priorities}\label{summary---generator-community-current-priorities}}

Out of the issues outlined above, the WG identifies the following as its
current main priorities:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Gain a more detailed understanding of current CPU costs by accounting
  and profiling.
  \end{quote}
\item
  \begin{quote}
  Survey generator codes to understand the best way to move to GPUs and
  vectorized code, and prototype the port of the software to GPUs using
  data-parallel paradigms.
  \end{quote}
\item
  \begin{quote}
  Support efforts to optimize phase space sampling and integration
  algorithms, including the use of Machine Learning techniques such as
  neural networks.
  \end{quote}
\item
  \begin{quote}
  Promote research on how to reduce the cost associated with negative
  weight events, using new theoretical or experimental approaches.
  \end{quote}
\item
  \begin{quote}
  Promote collaboration, training, funding and career opportunities in
  the generator area.
  \end{quote}
\end{enumerate}

\hypertarget{section-1}{%
\subsection{}\label{section-1}}

\hypertarget{references}{%
\subsection{References}\label{references}}

{[}Alb19{]} J. Albrecht et al. ``A Roadmap for HEP Software and
Computing R\&D for the 2020s'' Comput Softw Big Sci (2019) 3, 7,
\href{https://doi.org/10.1007/s41781-018-0018-8}{{doi:10.1007/s41781-018-0018-8}},
\href{https://arxiv.org/abs/1712.06982}{{arXiv:1712.06982}}

{[}Alw08{]} J. Alwall et al. ``Comparative study of various algorithms
for the merging of parton showers and matrix elements in hadronic
collisions'', Eur. Phys. J. C 53 (2008) 53,
\href{https://doi.org/10.1140/epjc/s10052-007-0490-5}{{doi:10.1140/epjc/s10052-007-0490-5}}

{[}Alw09{]} J. Alwall, S. de Visscher, F. Maltoni, ``QCD radiation in
the production of heavy colored particles at the LHC'', JHEP 0902 (2009)
017,
\href{https://doi.org/10.1088/1126-6708/2009/02/017}{{doi:10.1088/1126-6708/2009/02/017}}

{[}Alw14{]} J. Alwall et al. ``The automated computation of tree-level
and next-to-leading order differential cross sections, and their
matching to parton shower simulations'', JHEP 07 (2014) 079,
\href{https://doi.org/10.1007/JHEP07(2014)079}{{doi:10.1007/JHEP07(2014)079}}

{[}Bau13{]} C. Bauer et al., "Computing for perturbative QCD - a
Snowmass White Paper",
\href{https://arxiv.org/abs/1309.3598}{{arxiv:1309.3598}}

{[}Ben17{]} J. Bendavid, ``Efficient Monte Carlo Integration Using
Boosted Decision Trees and Generative Deep Neural Networks'',
\href{https://arxiv.org/abs/1707.00028}{{arXiv:1707.00028}}

{[}Bot19{]} E. Bothmann et al., ``Event generation with Sherpa 2.2'',
\href{https://scipost.org/SciPostPhys.7.3.034}{{SciPost Phys. 7 (2019)
034}}

{[}Bot20{]} E. Bothmann et al. ``Exploring phase space with Neural
Importance Sampling'',
\href{https://arxiv.org/abs/2001.05478}{{arXiv:2001.05478}}

{[}Chi17{]} J. T. Childers et al., "Adapting the serial Alpgen
parton-interaction generator to simulate LHC collisions on millions of
parallel threads", Comp. Phys. Comm. 210 (2017) 5154.
\href{https://doi.org/10.1016/j.cpc.2016.09.0135286}{{https://doi.org/10.1016/j.cpc.2016.09.0135286}}.

{[}Fre12{]} R. Frederix, S. Frixione, "Merging meets matching in
MC@NLO", JHEP 2012 (2012) 61.
\href{https://doi.org/10.1007/JHEP12(2012)061}{{doi:10.1007/JHEP12(2012)061}}

{[}Fre20{]} R. Frederix, S. Frixione, S. Prestel. P. Torrielli, ``On the
reduction of negative weights in MC@NLO-type matching procedures'',
\href{https://arxiv.org/abs/2002.12716}{{arXiv:2002.12716}}

{[}Fri17{]} S. Frixione, P. Nason, C. Oleari, "Matching NLO QCD
computations with parton shower simulations: the POWHEG method", JHEP11
(2007) 070,
\href{https://doi.org/10.1088/1126-6708/2007/11/070}{{doi:10.1088/1126-6708/2007/11/070}}

{[}Gao20{]} C. Gao. et al. ``Event Generation with Normalizing Flows'',
\href{https://arxiv.org/abs/2001.10028}{{arXiv:2001.10028}}

{[}Gen18{]} HSF Physics Event Generator Computing Workshop, CERN,
November 2018,
\href{https://indico.cern.ch/event/751693}{{https://indico.cern.ch/event/751693}}

{[}Gen19{]} HSF Physics Generators Working Group,
\href{https://hepsoftwarefoundation.org/workinggroups/generators.html}{{https://hepsoftwarefoundation.org/workinggroups/generators.html}}

{[}Gle08{]} T. Gleisberg, S. Hoeche, "Comix, a new matrix element
generator", JHEP 12 (2008) 039.
\href{https://doi.org/10.1088/1126-6708/2008/12/039}{{doi:10.1088/1126-6708/2008/12/039}}

{[}Hoe19{]} S. Hoeche, S. Prestel, H. Schulz, "Simulation of vector
boson plus many final jets at the high luminosity LHC", Phys. Rev. D
100, 0140124 (2019),
\href{http://dx.doi.org/10.1103/PhysRevD.100.014024}{{doi:10.1103/PhysRevD.100.014024}}

{[}Jad03{]} S. Jadach, Foam: a general-purpose cellular Monte Carlo
event generator, Comp. Phys. Comm. 152 (2003) 55.
\href{https://doi.org/10.1016/S0010-4655(02)00755-5}{{doi:10.1016/S0010-4655(02)00755-5}}

{[}Kaw86{]} S.Kawabata, A new Monte Carlo event generator for high
energy physics, Comp. Phys. Comm. 41 (1986) 127.
\href{https://doi.org/10.1016/0010-4655(86)90025-1}{{doi:10.1016/0010-4655(86)90025-1}}

{[}Kli18{]} M.D. Klimek, M. Perelstein, ``Neural Network-Based Approach
to Phase Space Integration'',
\href{https://arxiv.org/abs/1810.11509}{{arXiv:1810.11509}}

{[}Kon20{]} D. Konstantinov, "Optimization of Pythia8", EP-SFT group
meeting, CERN, January 2020,
\href{https://indico.cern.ch/event/890670}{{https://indico.cern.ch/event/890670}}

{[}Lep80{]} G. P. Lepage, "VEGAS: an adaptive multi-dimensional
integration program", Cornell report CLNS-447 (1980).
\href{https://cds.cern.ch/record/123074}{{https://cds.cern.ch/record/123074}}

{[}Lon05{]} L. Lonnblad, "Correcting the colour-dipole cascade model
with fixed order matrix elements", JHEP 05 (2002) 046.
\href{https://doi.org/10.1088/1126-6708/2002/05/046}{{doi:10.1088/1126-6708/2002/05/046}}

{[}Mal03{]} F. Maltoni, T. Stelzer, "MadEvent: automatic event
generation with MadGraph", JHEP 02 (2003) 027.
\href{https://doi.org/10.1088/1126-6708/2003/02/027}{{doi:10.1088/1126-6708/2003/02/027}}

{[}Man02{]} M. L. Mangano, M. Moretti, R. Pittau, "Multijet matrix
elements and shower evolution in hadronic collisions: Wbb + n-jets as a
case study", Nucl. Phys. B 632 (2002) 343.
\href{https://doi.org/10.1016/S0550-3213(02)00249-3}{{doi:10.1016/S0550-3213(02)00249-3}}

{[}Mat16{]} O. Mattelaer, "On the maximal use of Monte Carlo samples:
re-weighting events at NLO accuracy", \emph{Eur. Phys. J. C}
76\textbf{,} 674 (2016),
\href{https://doi.org/10.1140/epjc/s10052-016-4533-7}{{doi:10.1140/epjc/s10052-016-4533-7}}

{[}Mat18{]} O. Mattelaer, "MG5aMC status and plans", in {[}Gen19{]}

{[}Nas07{]} P. Nason, MINT: a Computer Program for Adaptive Monte Carlo
Integration and Generation of Unweighted Distributions Bicocca-FT-07-13
(2007). \href{https://arxiv.org/abs/0709.2085}{{arxiv:0709.2085}}

{[}SKe18{]} L. Sexton-Kennedy, G. Stewart, "CWP challenges and workshop
aims", in {[}Gen19{]}

\hypertarget{detector-simulation}{%
\section{Detector Simulation}\label{detector-simulation}}

\hypertarget{introduction-2}{%
\subsection{Introduction}\label{introduction-2}}

In HEP, data analysis, theory model evaluation and detector design
choices rely on detailed detector simulation. Since the start of the
first run the LHC experiments have produced, reconstructed, stored,
transferred and analysed hundreds of billions of simulated events. This
effort required a major fraction of the WLCG computing resources {[}Ref
to Exp Comp TDRs{]}.

The increase in delivered luminosity in future LHC runs, in particular
with the HL-LHC, will create unique research opportunities by collecting
orders of magnitude more data. So at the same time, the demand for
detector simulation will grow accordingly, in order to keep the
statistical uncertainties as low as possible. However, the expected
computing resources will not suffice if current detector simulation is
to be used; the availability of simulated samples of sufficient size
will soon become a major limiting factor as far as new physics
discoveries, for example through precision measurements, are concerned.
Development of faster simulation, therefore, is of crucial importance
and different techniques need to be explored under the assumption that
they will provide a sufficient gain in time performance with a
negligible or acceptable loss in physics accuracy.

The Geant4 simulation toolkit has been the de facto standard for HEP
experiment simulation for physics studies over the last two decades.
Designed in the late 90' to cope with the simulation challenges of the
LHC era, Geant4 introduced flexible physics modeling, exploiting layers
of virtuality. This fostered progress in physics by comparison of
competing models on same energy ranges, and allowed complementary models
to be combined to cover large energy ranges. The simulation applications
of the LHC experiments make use of this toolkit and, as such, most of
the LHC physics program relies on it. It has delivered physics precision
adequate to allow the experiments to perform precision analysis of the
collected data and to produce new results pushing the boundaries of our
understanding in HEP. At the same time, the code of the toolkit is
becoming old, with some parts written almost thirty years ago, making it
more and more difficult to run efficiently on modern computing
architectures. Any improvement in the critical elements of Geant4 can be
leveraged by the applications that use it.

Improvements of physics models need to continue to take place, however,
their exact influence on the precision of physics analysis is not always
easy to quantify. On the other hand, the overall CPU performance of the
current code will not improve drastically just by relying on modern
hardware or better compilers. Taking all those elements into account,
three main development paths have emerged to be undertaken
simultaneously: these have started to be pursued, however, effort in any
one of them is far from sufficient to exploit their individual
potential. On one hand, it is needed to continue to modernize and
refactor the simulation code by using more optimal, modern programming
techniques. Those improvements, requiring considerable effort but no
major paradigm changes, can reasonably bring several tens of percent of
speed-up, by improving the throughput on modern CPUs, avoiding constant
churn in data and instruction caches. This will certainly help to
produce more simulated events, but it will not be sufficient to meet the
high luminosity needs of the LHC experiments. On the other hand,
different fast simulation techniques, where tracking of individual
particles in given detectors is replaced by some sort of
parameterization, have already been extensively used for LHC simulation
campaigns, where some precision is traded to achieve higher event
throughput (thus lowering the cost of samples overall). Techniques to
re-use part of the simulated events multiple times have also emerged.
New technologies (like Machine Learning), should certainly be
investigated as fast simulation approaches. All of these will become
more and more significant and will need to be seamlessly integrated with
the detailed simulation provided by Geant4. Finally, the use of compute
accelerators (like GPUs or FPGAa) is the third emerging path that needs
to be undertaken and that requires intense R\&D effort. We will discuss
more in detail those three development axes in the following sections.

\hypertarget{geant4-rd}{%
\subsection{Geant4 R\&D}\label{geant4-rd}}

The Geant4 particle transport toolkit contains advanced models for
electromagnetic and hadronic physics processes, as well as the
geometrical description of the detectors. The toolkit design principles
was to allow flexibility in future extensions of its own code, as well
as choices for particular applications. It is clear that this
flexibility comes at the price of a performance penalty. We observe that
writing more compact and more specialized code can lead to several
percent speed-up. The technical improvements to explore here consist of
avoiding too many virtual layers and improving the instruction and data
locality, leading to better cache use. Going even further as far as
specialization is concerned, studies suggest that implementing compact,
self-contained transport libraries with a minimal set of physics models
(for instance only electromagnetic interaction for selected particles)
with predefined scoring (to simulate, for instance, the response of an
EM calorimeter) can also bring speed-ups to the complete
application{[}Ref to GeantV RD{]}. As far as architectural modifications
are concerned, the design of the Geant4 track-status machinery prevents
the introduction of fine-grained track parallelism. It is still not
clear whether this kind of parallelism can bring any substantial
speed-up, due to the overheads; however, making the Geant4 kernel and
physics processes stateless would certainly provide the necessary
starting points for further R\&D in that direction. Most of these
investigations have started and it is hoped to see some first results on
a year time-scale. More effort, however, will be required to integrate
any eventual changes in the production code in a timely manner. The
GeantV Vector prototype has allowed the assessment of the achievable
speedup using a novel, vectorized approach to particle transport. The
demonstrator of a full electromagnetic shower in realistic (CMS)
geometry has been implemented and the comparison to a similar Geant4
application has been performed. The conclusion of that work showed that
the main factors in the speedup seem to include better cache use and
tighter code, while the vectorization impact was much smaller than hoped
for. This unimpressive impact is due to a set of factors, including the
fraction of the current algorithms that could be vectorized, unresolved
challenges for the gather/scatter and tail handling (to keep the number
of events in flight within bound) needed for large number of shapes, and
of course Amdahl's Law applied to vectorized where some bottleneck (for
example geometry navigation) could not be tackled without a major
additional long term effort. {[}Ref-GeantV-Final-Report{]}. At the same
time libraries developed in the context of the GeantV R\&D, e.g.
VecGeom, have been successfully integrated in Geant4 and ROOT
benefitting the whole HEP software community {[}CMS01{]}. The
cost/benefit of more vectorization of the code has not been deemed worth
the necessary investment also due to the speculation that a complete
rewrite of the physics base could become necessary to fully exploit it.
As a result investing in the optimization and modernization of the
Geant4 code has gained even more relevance.

\hypertarget{experiments-optimization}{%
\subsection{Experiments optimization}\label{experiments-optimization}}

All LHC experiments have dedicated simulation frameworks making use of
the Geant4 toolkit for modelling their detectors and the physics
processes occuring in the experimental setups. Every experiment
configures and uses what is provided by Geant4 for their specific needs.
Effort has been spent by each experiment since the publication of the
Community Whitepaper to benchmark their simulation code and to maximally
exploit the options provided by the Geant4 toolkit to optimise the
performance of their simulation code. All of the various handles
provided are being continuously explored and adopted when deemed useful:
range and physics cuts have been reviewed and customised by all
experiments, shower libraries adopted as baseline in the forward region
by ATLAS and CMS, stepping granularity optimized, magnetic field caching
investigated, neutron russian roulette used for dedicated simulations.
All of them can have major impacts on simulation time and sharing
knowledge and experience between experiments can be of high benefit even
if each experiment has to find its optimal solution.

ATLAS, CMS and LHCb have integrated the Geant4 event-based
multi-threading features within their own multi-threaded frameworks,
harmonizing the different architectural choices made {[}CMS02, LHCb01,
ATLAS?{]}, and have or are planning to use them in their production
environment to gain access to a larger set of distributed resources.
ATLAS has been running their multi-process simulation framework on HPCs
systems for production in the last years {[}ATLAS02{]} while CMS has
pioneered the use of their multi-threaded simulation on available
resources including HPCs {[}CMS03{]}. Both approaches have allowed to
exploit a higher number of computing resources by gaining access to
systems with limited memory.

\emph{Should we put here a bit on experiments fast simulation in the
sense of all doing things about it and then give details in next
session?}

\hypertarget{fast-simulations}{%
\subsection{Fast Simulations}\label{fast-simulations}}

The so-called `fast simulation' techniques consist of replacing the
tracking of individual particles through (a part of) the detector,
including all the physics processes they would undergo, by a
parameterization where the detector response is produced directly as a
function of the incoming particle type and energy. An example of such a
parameterization was implemented several years ago in the context of the
H1 experiment. This parameterization, available within the Geant4
toolkit under the name of the GFLASH library, became the starting idea
for several `custom' implementations, specific for other calorimeters.
Those were interfaced to experiments' simulation applications through
the available `hook' in the Geant4 toolkit. The LHC experiments
implemented, for example, dedicated parametrized response libraries for
some of their calorimeters. Recently, an effort has been invested in
Geant4 to generalize the parameterization formulae, and to implement
automatic procedures of tuning the parameters for specific detector
geometries. This kind of `classical' fast simulation, which describes
the particle shower shape using some complex mathematical functions with
several parameters, will remain an important tool; however, it is also
clear that their overall precision for the different energy values,
especially for highly granular and complex calorimeters will always be
relatively low. The recent developments of Deep Learning-based
techniques have opened an exciting possibility of replacing those
`classical' parameterisations by trained neural networks that would
reproduce the detector response. This approach consists of training
generative models such as Generative Adversarial Networks (GAN),
Variational Auto-Encoders (VAE) or Autoregressive Generative Networks on
the `images' of particle showers {[}ML001{]}. The energy deposition
values in the calorimeter cells are considered as `pixels' of the images
that the network is supposed to reproduce. The first studies and
prototypes have shown very promising results, however, the
generalization of the developed models (for different calorimeters, with
different particles incident at different angles with different
energies), still requires further effort. Detailed physics validation of
those tools and understanding of the capability to reproduce the
fluctuations is a prerequisite towards `production-quality' fast
simulation libraries. This work is already ongoing, but will require,
over the next few years, more effort to be invested combining the
physics and Machine Learningexpertise. Given that the training of the
deep learning algorithms usually requires using sufficiently large MC
samples produced using standard techniques, the development of the novel
tools does not necessarily remove the need to speed up Geant4.
Experiments are in the process of investigating the use of fast
simulations for other types of detectors than calorimeters, e.g.
Cerenkov based systems, as well as a fully parametric response at the
level of reconstructed objects (tracks and clusters) for fast
systematics verifications. In LHCb the re-use of part of the underlying
events for specific signals of interest have been established when
appropriate with particular care paid to ensure to keep under control
bias on statistical uncertainties {[}LHCb02{]}, its applicability in
ATLAS and CMS could be explored for specific cases. Similarly, the
re-use of simulated or real events to mimic the background to
hard-scatter events is also under investigation, where additional
challenges exist concerning storage and I/O due to the handling of the
large minimum bias samples needed to model the additional interactions.

It is reasonable to expect that the full variety of simulation options,
from `smeared reconstructed quantities' to parameterisation for specific
detectors to the detailed Geant4, will need to be exploited by the
experiments for different tasks. Seamless ways of providing them in a
transparent integrated way in the experiments' simulation frameworks
should continue to be pursued with Geant4 supporting hooks for such
integrations.

\hypertarget{technical-challenges-software-and-computing-1}{%
\subsection{Technical Challenges (Software and
Computing)}\label{technical-challenges-software-and-computing-1}}

The hardware landscape has been always setting direction as far as
software evolution was concerned. The recent adaptation of the
simulation code to multithreading (MT) turned out to be relatively
straightforward, but still took several years to adopt and validate.
Geant4 can now run in the MT mode efficiently, distributing the events
between the threads and sharing resources like the geometry description
or physics data and thus reduce the memory footprint of a simulation
while maintaining its throughput performance. The efficient utilization
of Single Instruction Multiple Data (SIMD) architectures, on the other
hand turned out to be more challenging. Not only was a limited part of
the code `vectorizable' but also, as the already mentioned GeantV R\&D
has shown, overheads related to gathering the data for vector processing
presented a significant challenges (tail handling, work balancing across
threads, etc.) to efficiently (development time wise and run time wise)
profit from it {[}Ref to HSF final review of GeantV{]}.

The most recent `revolution' in the computing hardware systems consists
of using Graphics Processing Units (GPUs) for general computing. Several
attempts have been made to port specific simulation code to GPUs. A few
of them turned out to be successful leading to factors of several
hundred or thousand of speedup, {[}GPU-``MPEXS-DNA{]},
{[}Hybrid-Gate{]}, {[}G4CU{]}, {[}Opticks{]} however, they were always
limited to a very specific simulation problem, far from what is
addressed by a general HEP simulation. The application of GPUs has been
very successful in domains like medical physics simulation, neutron
transport or optical photon transport. In those applications, the type
of particles considered is very limited (often just one type, with no
secondary particle production), the set of physics processes is reduced
and the geometry is often extremely simple compared to the LHC
detectors. A natural `extrapolation' of those approaches to a general
HEP simulation is very difficult to imagine, because the stochasticity
of the simulated events would immediately lead to divergences as far as
different GPU threads are concerned, not to mention, simply the
feasibility of porting the whole simulation to the GPU code for which
specific expertise would be needed. On the other hand, running only very
small pieces of the simulation application on GPUs does not seem to be
efficient either, as gathering data and transferring it from the host to
the device and back may strongly limit any possible performance gain,
similarly to what was seen with the GeantV vectorisation R\&D. The most
plausible approaches seem therefore to lead in the direction of
specialized libraries (like those described above in the context of
speeding up Geant4 simulation on the CPUs) that would perform the
complete simulation of specific sub-detectors (for instance EM
calorimeter, for specific incoming particles or Cherenkov-based
detectors for the optical processes) on the device. Such libraries could
be the right compromise between the complexity of the algorithm that
needs to be ported to the device and the overall time that is now spent
(and could be gained) on the corresponding of the full simulation on the
CPU. The investigation in those directions are starting now, but will
certainly require a considerable effort to be further invested. The
first steps that are currently being undertaken consist of implementing,
on GPUs, prototypes based on VecGeom geometry and navigation
capabilities. If those prototypes turn out to be successful, the next
challenge will consist of adding tracking and eventually a limited
subset of physics models. While Geant4 is a toolkit capable of
addressing different modeling and simulation problems and contains many
features and capabilities allowing for user access to detailed
information for a large variety of use cases and scientific communities,
this GPU development might turn into specialized transport modules,
stripped of some features which would be expensive to implement or
support efficiently on GPUs. Another interesting avenue consists of
exploring the application of some vendor libraries (like Nvidia Optix).
Those libraries, originally developed for ray tracing, have several
similarities with general particle transport and if applied
successfully, could lead to major speed-ups. All those efforts certainly
require a major investment and new developers, expert in those
technologies, to join the R\&D work.

\hypertarget{other-activities}{%
\subsection{Other Activities}\label{other-activities}}

Common digitisation efforts would be desirable among experiments, with
advanced high-performance generic examples, which experiments could use
as a basis to develop their own code. Nevertheless the large variety of
detectors' technologies used by the experiments reduces its
applicability. Digitisation not yet being a limiting factor in terms of
CPU requirements also does not urge developments in this area.

Simulated samples are often processed through the reconstruction as real
data. As such in some cases they require the emulation of hardware
triggers. Hardware triggers are based on very specific custom devices
and a general approach does not seem feasible even if some
parametrization could be generalised.

\hypertarget{outlook}{%
\subsection{Outlook}\label{outlook}}

To conclude, it is important to realise that without investing a serious
effort in the simulation R\&D, a considerable risk of this being a major
limiting factor as far as new physics discoveries are concerned. We need
the concerted effort of different experts in physics, Machine Learning
and GPUs. There are too many unknowns to focus on only one direction, so
a lot of prototyping is needed. We need to further develop Geant4 while
working on fast simulation algorithms and doing extensive R\&D on the
compute accelerators. While the R\&D activities are taking place, one
should not forget that one still needs to maintain the existing code and
make sure that sufficient funding and staffing is provided for
maintenance and development of the physics algorithms, as well as for
adapting the code to the ongoing hardware, operating system and compiler
evolution.

It is also important to stress once again the need of not only ongoing
R\&D investigations and independent prototyping to explore novel ideas,
but also a continuous integration activity of new solutions into the
Geant4 code base and experiments applications to profit from each
improvement as quickly as possible. An effort needs to be invested into
making it possible to evolve it to meet the HL-LHC needs without
compromising the production quality of the Geant4 code used by the
experiments. We need to ensure that new developments resulting from the
R\&D programmes can be tested with realistic prototypes and, if
successful, then integrated, validated, and deployed in a timely fashion
in Geant4 and adopted by the experiments. The strategy adopted in the
successful integration in Geant4 of VecGeom libraries developed in the
context of the GeantV R\&D can provide a working example of how to
proceed to provide some incremental improvements to the existing
libraries.

\hypertarget{references-1}{%
\subsection{References}\label{references-1}}

{[}CMS01{]} V. Ivanchenko and S. Banerjee on behalf of the CMS
Collaboration, EPJ Web of Conferences 214, 02012 (2019), The PDF is
here:
\url{https://www.epj-conferences.org/articles/epjconf/pdf/2019/19/epjconf_chep2018_02012.pdf}

{[}GPU-``MPEXS-DNA{]} S. Okada et al. ``MPEXS-DNA: a new GPU-based Monte
Carlo simulator for track structures and radiation chemistry at
subcellular scale,'' Med Phys. vol 46 no. 3, pp. 1483-1500 (2019)

{[}Hybrid-Gate{]} J. Bert et al., "Hybrid GATE: A GPU/CPU implementation
for imaging and therapy applications," 2012 IEEE Nuclear Science
Symposium and Medical Imaging Conference Record (NSS/MIC), Anaheim, CA,
pp. 2247-2250 (2012)

{[}G4CU{]}
\url{https://kds.kek.jp/indico/event/15926/session/30/contribution/110/material/slides/0.pdf}

{[}Opticks{]} S. Blyth, ``Opticks : GPU Optical Photon Simulation for
Particle Physics using NVIDIA® OptiXTM,'' EPJ Web Conf., 214, pp. 02027
(2019) 02027

{[}LHCb01{]} CHEP for Gaussino and Gauss of Gaussino MT

{[}CMS02{]} Use of MT in CMSSW

{[}ML001{]} Adversarial Networks (GAN), Variational Auto-Encoders (VAE)
or Autoregressive Generative Networks on the `images' of particle
showers

{[}LHCb02{]} ReDecay paper

\hypertarget{reconstruction-and-software-triggers}{%
\section{Reconstruction and Software
Triggers}\label{reconstruction-and-software-triggers}}

Software trigger and event reconstruction techniques in HEP face a
number of new challenges in the next decade. Advancements in facilities
and future experiments bring a dramatic increase in physics reach, as
well as increased event complexity and rates.

At the HL-LHC, the central challenge for high-level triggers (eg,
software triggers) and object reconstruction is to maintain excellent
efficiency and resolution in the face of high pileup values, especially
at low transverse momenta. Detector upgrades such as increases in
channel density, high precision timing and improved detector layouts are
essential to overcome these problems.

The subsequent increase of event complexity at the HL-LHC also requires
the development of software algorithms that can process events with a
similar cost per event to Run-2 and Run-3. At the same time, algorithmic
approaches need to effectively take advantage of evolutions in computing
technologies, including increased SIMD capabilities, increased core
counts in CPUs, and heterogeneous hardware.

This section focuses on the challenges identified in the Community White
Paper {[}1{]} and the development of solutions that. It also includes
mentions of open source software efforts that have developed within or
outside the LHC collaborations that could be adapted by the LHC
community to improve trigger and reconstruction algorithms.

\textbf{The evolution of triggers and real-time analysis}

Trigger systems are evolving to be more capable, both in their ability
to select a wider range of events of interest for the physics program of
their experiment, and their ability to stream a larger rate of events
for further processing.

The event rates that will be processed by experiments at the HL-LHC will
increase by up to a factor 10 with respect to Run-3, owing to upgraded
trigger systems.

ATLAS and CMS plan to maintain a two-tiered trigger system, where a
hardware trigger makes use of coarse event information to reduce the
event rate to 10x over the current capability, up to 1 MHz {[}cite
TDR{]}. The high level trigger system, implemented in software, selects
up to 100 kHz of events to be saved in full for subsequent
analysis\footnote{LHCb {[}cite TDR{]} and ALICE {[}cite TDR{]} will both
  stream the full collision rate to real-time or quasi-realtime software
  trigger systems.}.

Experiments have also been working towards minimizing the differences
between trigger (online) and offline software for reconstruction and
calibration, so that more refined physics objects can be obtained
directly within the HLT farm for a more efficient event selection. This
is also in-line with enhancing the experiment's capabilities for
real-time data analysis of events accepted by the hardware trigger
system, driven by use cases where the physics potential improves when
analyzing more events than what can be written out in full with
traditional data processing.

Implementations of real-time analysis systems, where raw data is
processed in its final form as close as possible to the detector (e.g.
in the high-level trigger farm), are in use within several experiments.
These approaches remove the detector data that typically makes up the
raw data tier kept for offline reconstruction, and keep only a limited
set of analysis objects reconstructed within the high level trigger.

Real-time analysis techniques are being adopted to enable a wider range
of physics signals to be saved by the trigger for final analysis. As
rates increase, these techniques can become more important and
widespread by enabling only the parts of an event associated with the
signal candidates to be saved, reducing the required disk space. The
experiments are focusing on the technological developments that make it
possible to do this with acceptable reduction of the analysis
sensitivity and with minimal biases.

Active research topics include compression and custom data formats;
toolkits for real-time detector calibration and validation which will
enable full offline analysis chains to be ported into real-time; and
frameworks which will enable non-expert offline analysts to design and
deploy real-time analyses without compromising data taking quality.
Use-cases for real-time analysis techniques have expanded during the
last years of Run 2 (cite LHCb-TESLA?) and their use appears likely to
grow already during Run 3 data taking for all experiments owing to
improvements in the HLT software and farms (cite TDAQ TDR, TLA pub note
ATLAS). Further ideas include retaining very reduced data from as much
of the data taking rate as possible from selected regions and
subdetectors, even up to the 40 MHz bunch crossing rate (cite CHEP
proceedings CMS).

\textbf{Challenges and improvements foreseen in event reconstruction}

Processing and reducing raw data into analysis-level formats, event
reconstruction, is a major component of offline computing resource
needs, and in light of the previous section it is relevant for online
resources as well. This is an essential step towards precision
reconstruction, identification and measurement of physics-objects at
HL-LHC.

Algorithmic areas of particular importance at HL-LHC experiments are
charged-particle trajectory reconstruction (``tracking''), including
hardware triggers based on tracking information which may seed later
software trigger and reconstruction algorithms; jet reconstruction,
including the use of high-granularity calorimetry; and the use of
precision timing detectors.

\emph{\textbf{Tracking in high pile-up environments}}

The CPU needed for event reconstruction in Run-2 and Run-3 has been
dominated by charged particle reconstruction (tracking). This is still a
focus for HL-LHC triggering and event reconstruction, especially when
the need for efficiently reconstructing low transverse momentum
particles is considered.

The huge increase in the number of charged particles and hence the
combinatorial load for tracking algorithms at future colliders will put
great strain on CPU resources. To minimize the memory footprint,
tracking software needs to be thread-safe to support multi-threaded
execution per core. At the same time, the software has to be efficient
and accurate to meet the physics requirements.

Since the CWP, experiments have made progress towards improving software
tracking efficiency in HL-LHC simulation, see e.g. Ref. {[}2{]}. A
number of collaboration and community initiatives have been focusing on
tracking, targeting both offline and online, and we list them below.

The \emph{\textbf{ACTS}} project {[}5{]} is an attempt to encapsulate
the current ATLAS track reconstruction software into an
experiment-independent and framework-independent tracking software
designed to fully exploit modern computing architectures. It builds on
the tracking experience already obtained at the LHC, and targets the
HL-LHC as well as future hadron colliders, providing support for other
HEP experiments, such as FASER, Belle-II, EIC, CEPC, etc. ACTS provides
a set of track reconstruction tools designed for parallel architectures,
with a particular emphasis on thread-safety and concurrent event
reconstruction.

The aim of the \textbf{mkFit} project is to speed up Kalman filter (KF)
tracking algorithms using highly parallel architectures and to deliver
track building (and possibly fitting) software for the HL-LHC. Recent
activities focused on delivering a production quality software setup to
perform track building in the context of the LHC Run-3 (or Run-2). The
implementation relies on single-precision floating point math and is
available for multicore CPUs and achieves significant gains in compute
performance from the use of vector instructions set extending to
AVX-512, based on Matriplex library (a part of the mkFIt project).

\textbf{Exa.TrkX} is a cross-experiment collaboration of data scientists
and computational physicists from ATLAS, CMS and DUNE. It develops
production-quality deep neural network models, in particular Graph
Neural Networks, for charged particle tracking on diverse detectors
employing next-generation computing architectures such as HPCs. It is
also exploring distributed training and optimization of Graph Neural
Networks (GNN) on HPCs, and the deployment of GNNs with microsecond
latencies on Level-1 trigger systems.

\emph{\textbf{\hfill\break
Addition of timing information in reconstruction}}

Physics performance in very high pileup environments, such as the HL-LHC
or FCC-hh, may also benefit from adding timing information to the
reconstruction. This permits to mitigate effects of pile-up by
exploiting the time-separation of collision products. {[}cite HGTD/CMS
timing TPs,
\href{https://cds.cern.ch/record/2296612}{{https://cds.cern.ch/record/2296612}}
{]}\\
The experimental communities have been working on timing detector
reconstruction and object identification techniques in complex
environments. Since the CWP, initial tracking and vertexing algorithms
that include timing information have been developed and incorporated
into experimental software stacks {[}cite ATLAS/CMS CDRs?, CTD
contribution
\href{https://indico.cern.ch/event/658267/contributions/2870293/attachments/1620243/2577412/TimingCTD_LindseyGray_20032018.pdf}{{First
Steps Towards Four-Dimensional Tracking: Timing Layers at the
HL-LHC}}{]}. \emph{Missing: common software, wrote to Ivan Kisel (CBM)
about how portable this is
\href{https://inspirehep.net/literature/1414007}{{4-Dimensional Event
Building in the First-Level Event Selection of the CBM Experiment -
INSPIRE}}. Also public reference from LHCb on how preselection of PV can
improve the reconstruction time for one event.}

\textbf{Enhanced data quality and software monitoring for trigger and
reconstruction}.

At HL-LHC, the development, automation, and deployment of extended and
efficient monitoring tools for software trigger and event reconstruction
algorithms will be crucial for the success of the experimental physics
programme.

HEP experiments have extensive continuous integration systems, including
code regression checks that have enhanced the monitoring procedures for
software development in recent years. They also have automated
procedures to check trigger rates as well as the performance of the low-
and high-level physics objects in the data.

While we are not aware of updates in this area since the CWP, we would
like to stress the importance of investment in this area as the
monitoring of the quality of the recorded/reconstructed data is
paramount for further physics analysis.

\textbf{Hardware and software improvements towards reconstruction
software}

\emph{\hfill\break
\textbf{General reconstruction software improvements: vectorization}}

Improving the ability of HEP developed toolkits to use vector units on
commodity hardware will bring speedups to applications running on both
current and future hardware. The goal for work in this area is to evolve
current toolkit and algorithm implementations, and best programming
techniques to better use SIMD capabilities of current and future
computing architectures. Since the CWP, algorithm development projects
(e.g. {[}cite
\url{https://lhcbproject.web.cern.ch/lhcbproject/Publications/f/p/LHCb-FIGURE-2019-002.html}
and mkfit{]}) have demonstrated success in increasing the use of vector
units. In addition, industry developed software, including machine
learning toolkits (e.g., tensorflow) typically make excellent use of
vector units.

\emph{\textbf{Use of machine learning for software trigger and
reconstruction algorithms}}

It may be desirable or even necessary to deploy new algorithms,
including advanced machine learning techniques, in order to face the
increase in event complexity without increasing the per-event
reconstruction resource needs.

Work is already undergoing in the collaborations to evolve or rewrite
existing toolkits and algorithms focused on their physics and technical
performance at high event complexity (e.g. high pileup at HL-LHC), and
efforts in this area have expanded. Cross-collaboration developments are
focusing on the availability of ML algorithms for experimental software,
especially in persistifying models and running inferences in production
(e.g.
\emph{\href{https://github.com/lwtnn/lwtnn}{{https://github.com/lwtnn/lwtnn}}
and \href{https://onnx.ai}{{https://onnx.ai}}}) and in enhancing the
training capacities of collaborations by submitting jobs to the grid and
to facilities offering large CPU/GPU resources. One can also foresee
that HEP will benefit from techniques and cross-talk from industry.

\emph{\textbf{Algorithms and data structures to efficiently exploit}}
\emph{\textbf{many-core and non-x86 architectures (e.g. GPUs, FPGAs)}:}

Computing platforms are generally evolving towards having more cores in
order to increase processing capability. The goal is to evolve current
event models, toolkits and algorithm implementations, and best
programming techniques to improve the throughput of multithreaded
software trigger and event reconstruction applications. Additionally,
computing architectures using technologies beyond CPUs offer an
interesting alternative for increasing the throughput of the most time
consuming trigger or reconstruction algorithms.

Since the CWP, HEP now has a number of algorithmic projects that are
designed for, or have successfully adapted to, hardware accelerators,
most notably to NVidia GPUs using CUDA. Two projects providing open
source software implementations are Patatrack (from CMS) {[}3{]} and
Allen (from LHCb) {[}4{]}.

\textbf{Patatrack} is a CMS initiative aimed to use heterogeneous
computing for the charged particle reconstruction. Within the CMS
codebase (CMSSW), this project has demonstrated that physics
reconstruction code can be written to leverage heterogeneous
architectures, like GPUs, to achieve a significant speed-up while
reducing the overall cost and power consumption of a system. An example
is the CMS Pixel local reconstruction and the track and vertex
reconstruction: these algorithms can run on an NVIDIA T4 GPU with the
same performance as 52-56 Xeon cores, at a fraction of the cost and
power consumption of an equivalent CPU-only system; further improvements
may come pairing one or more high end GPUs with a low power ARM system.

\textbf{Allen} is a data processing framework for GPUs, as well as a
specific implementation of a first-level trigger for LHCb Run 3
data-taking. Allen is optimized for sustaining the rate required by
real-time processing in LHCb, but can be used as a more general GPU data
processing as it includes a scheduler and a memory manager that can be
integrated into Gaudi. This permits the extension and implementation of
Allen within other experimental software using the same underlying
framework.

The experience built with these projects shows that GPUs can be used for
increasing the throughput in cases of large-size (ALICE) {[}cite Alice
O2{]} and small-size events per second (LHCb), and to speed up specific
parts of the data processing (CMS, Patatrack).

It is clear that the technology is improving rapidly in this area. The
adoption of portability toolkits is needed to avoid vendor lock-in
and/or the need to evolve algorithms by hand to adapt to each new type
of computing architecture, as well as to increase the sustainability of
the codebase avoiding multiple implementations of the same algorithm.

Heterogeneous computing is still a rapidly developing and changing
field. The burden of writing architecture-specific code repeatedly will
certainly exceed the available effort from the experiments. Efforts are
ongoing in this direction, since the CWP, trying to take advantage of
the latest development in synergy with industry and attempting to
increase the use of centers with heterogeneous resources by experiments.

The \textbf{HLS4ML} project {[}7{]} targets the implementation of
machine learning algorithms on FPGA for low-latency applications useful
for e.g. Level-1 triggers. The aim of the HLS4ML project is the
translation of trained ML models into FPGA firmware. HLS4ML also extends
to the production of coprocessor kernels for FPGAs for longer latency
applications, and can be used as a tool to design AI-powered ASICs. The
\textbf{SONIC} application {[}8{]} is being developed in parallel, with
the goal of facilitating and accelerating the inference of deep neural
networks for triggering, reconstruction, and analysis, by providing
software to use heterogeneous computing resources as a service targeting
next-generation facilities at the energy and intensity frontier (HL-LHC,
LBNF).

\textbf{References (not properly formatted, this will go on Overleaf):}

\emph{{[}1{]}
\href{https://arxiv.org/abs/1802.08640}{{https://arxiv.org/abs/1802.08640}}}

\emph{{[}2{]}
\href{https://cds.cern.ch/record/2693670}{{https://cds.cern.ch/record/2693670}}}

\emph{{[}3{]}} - Andrea Bocci \emph{at al.}, ``\emph{Heterogeneous
online reconstruction at CMS}'', presented at CHEP 2019\\
- Andrea Bocci \emph{at al.}, ``\emph{Heterogeneous reconstruction:
combining ARM processors with GPUs}'', presented at CHEP 2019

\emph{{[}4{]}}
\href{https://arxiv.org/abs/1912.09161}{{https://arxiv.org/abs/1912.09161}}

\emph{{[}5{]}}
\href{http://dx.doi.org/10.1088/1742-6596/898/4/042011}{{http://dx.doi.org/10.1088/1742-6596/898/4/042011}},
\href{https://arxiv.org/abs/1910.03128}{{https://arxiv.org/abs/1910.03128}},
The ACTS project: track reconstruction software for HL-LHC and beyond,
Proceedings of CHEP2019, in progress

{[}6{]} Xiangyang Ju et al, "Graph Neural Networks for Particle
Reconstruction in High Energy Physics Detectors"
\href{https://arxiv.org/abs/2003.11603}{{(arxiv link}})
(\href{https://github.com/exatrkx/exatrkx-neurips19}{{Software and
documentation}})

{[}7{]} J. Duarte \emph{et al}, ``Fast inference of deep neural networks
in FPGAs for particle physics'', \emph{JINST} \textbf{13} P07027, 2018,
doi:
\href{https://iopscience.iop.org/article/10.1088/1748-0221/13/07/P07027/meta}{{10.1088/1748-0221/13/07/P07027}}\\
S. Summers \emph{et al}, ``Fast inference of Boosted Decision Trees in
FPGAs for particle physics'', 2020, {[}Online{]} available:
\href{https://arxiv.org/abs/2002.02534}{{arXiv:2002.02534}}\\
G. Gi Guglielmo \emph{et al}, ``Compressing deep neural networks on
FPGAs to binary and ternary precision with HLS4ML'', 2020, {[}Online{]}
available: \href{https://arxiv.org/abs/2003.06308}{{arXiv:2003.06308}}

{[}8{]} SONIC:
\href{https://arxiv.org/abs/1904.08986}{{https://arxiv.org/abs/1904.08986}}

\hypertarget{data-analysis}{%
\section{Data Analysis}\label{data-analysis}}

\hypertarget{key-analysis-computing-challenges-at-the-hl-lhc}{%
\subsubsection{Key analysis computing challenges at the
HL-LHC}\label{key-analysis-computing-challenges-at-the-hl-lhc}}

Examination of collision data is in essence the primary objective of the
experimental collaborations, coming at the end of the data
preparation/simulation and reconstruction chain. The stage of analysis,
starting in most cases from a standard data format generically referred
to as ``Analysis Object Data'' (AOD) containing reconstructed physics
object data and producing physics results as the end product, poses
unique computing challenges. The scope of analysis processing is broad,
encompassing the production of derived data formats as well as end stage
analysis actions such as producing histograms from those intermediate
datasets and visualising the final results. In the following, attention
is focused on the computing challenges related to analysis for the ATLAS
and CMS experiments.

Today, ROOT format AOD files and derived datasets take up the lion's
share of disk resources, filling approximately 80\% of the total data
volume for both ATLAS and CMS {[}1,2{]}. To serve precision analyses
that require large event statistics, the HL-LHC will increase the
recorded event rate by an order of magnitude, compared with a projected
15\% annual growth of storage resources. The large pile-up in HL-LHC
collisions will compound the storage challenge by inflating the data
size per event. Reducing the storage footprint of data analysis is
therefore of paramount importance.

The effective CPU needs for end-stage analysis payloads are typically
orders of magnitude lower than those in simulation and reconstruction.
When scaled up to O(100) analyses per year, each processing the input
data dozens of times a year, the total CPU consumption at the HL-LHC is
projected to be around 10\% of total experimental computing {[}1,2{]}.
Central production of analysis data formats may account for another
10-30\%. It is worth noting at this point that analysis data access
patterns tend to be more chaotic than preceding stages, which can
increase the effective resource needs by large factors.

More significant than the global storage and computational costs is the
job turnaround time, closely tied to the ``time to insight'', which is a
strong limitation on the speed of analysis publication. One significant
constraint is the need for full coverage of the input data, which is
inextricably linked to weaknesses in book-keeping tools that make it
difficult, if not impossible, to make meaningful studies on subsets of
the data. This full coverage requirement is clearly limited by computing
infrastructure load and downtime. Another challenge is the
multiplication of resource demands from the assessment of systematic
uncertainties, which involves processing many alternate configurations
and input data, inflating both the CPU load and storage footprint.
Improvement of community standards for metadata handling and uncertainty
handling will be important for HL-LHC.

A last concern is that, while simulation and reconstruction code is
mostly optimised and developed under greater scrutiny, analysis code is
often produced with emphasis on quick results and less emphasis on code
quality or performance. Over time, this may result in inefficient,
under-documented code that exacerbates the disk and CPU shortfall and
poses a major difficulty when software is repurposed for a new analysis.
Add to this a growing array of alternative toolkits from the data
science community, in particular for machine learning, and it is clear
that the entropy of the analysis code ecosystem has become a major
challenge in itself.

The following sections expand on the major challenges for analysis
software, including those identified above. An overview is given of
ongoing work on common tools that alleviate these problems, followed by
an outlook on R\&D prospects that would more significantly transform the
HL-LHC analysis model for major resource savings.

\hypertarget{analysis-processing-and-workflows-current-problems-and-solutions}{%
\subsubsection{Analysis processing and workflows: current problems and
solutions}\label{analysis-processing-and-workflows-current-problems-and-solutions}}

To make data analysis tractable, AOD data are typically reduced both by
saving only relevant information for each event, which may require
transformations (object calibration, computation of derived variables)
beyond simply discarding information, and by skimming out only events
that are interesting. In post-reconstruction data formats, lossy
compression is also in use and being optimised. Coordination of the data
reduction phase is a key point for the organization of data analysis
processing at HL-LHC.

Two approaches have been followed during LHC Run2 by ATLAS and CMS:
ATLAS analysis trains {[}4{]} and the CMS mini-AOD, a highly reduced and
standardised event content. Analysis trains place software payloads
tailored for specific analyses (carriages) into a periodically scheduled
centralised execution, amortising data staging costs and sharing some
common processing steps {[}3{]}. Standardised event contents are instead
common reductions that satisfy the needs of the bulk of analyses,
avoiding duplication of commonly selected events in multiple datasets.
Neither of these approaches provides exhaustive coverage of analysis use
cases on the experiments and alternative data formats are needed for a
small but significant (10-20\%) fraction of analyses requiring custom
reconstruction that may be CPU-intensive. These exceptions to the rule
are expected to remain at the HL-LHC.

It is worth considering the viability of the arguably more flexible
option to have analysis trains run over the reconstruction output at the
HL-LHC. Extrapolating using a simple model, ATLAS finds that they would
produce 35 PB/year of data AOD and 215 PB/year of MC {[}1{]}. In
comparison, a reduced format (10 kB/event) made from the AOD would
result in 0.5 PB/year for data and 2 PB/year for MC. Experience on CMS
shows that further reduction (down to \textasciitilde2 kB/event for the
nano-AOD) could be possible while still supporting a majority of physics
analyses {[}5{]}. File size reductions also permit multiple copies to be
held on disk for more efficient parallel processing. It is clear that as
many analyses as possible need to be fed into this standardized analysis
event-content pipeline if analysis computing resources are to be kept
under control. Nevertheless, analysis trains or any similar means of
synchronising data access could be applied to this highly reduced
format.

Careful attention needs to be paid to the assessment of systematic
uncertainties, particularly with regard to event content
standardisation, to avoid creating many derived outputs that differ only
minimally. Analysis models that do not require all event information to
be present in a single file, instead leveraging ROOT ``friend trees'' or
columnar storage, may be a way to reduce this duplication both in the
case of uncertainty calculations and reconstruction augmentations, but
will require development of robust strategies for keeping the
augmentations in sync with the core events. It is noted that the ROOT
data format is quantifiably very efficient for the processing model in
HEP {[}16{]}, and any competitors would have to achieve a similar level
of performance.

Access to metadata, defined as cross-event information, is clearly a
weak point in the handling of the multiple analysis datasets originating
from the LHC. Metadata may include information such as book-keeping of
processed events, detector conditions and data quality, \emph{a
posteriori} measured scale factors, and software versioning. This
information is often scattered in multiple locations such as:
``in-file'' data, remote database resources, shared-area txt/root files,
twiki pages or even in e-mail. The lack of proper metadata integration
and access interfaces leaves analyses exposed to any problems in data
processing. This in turn results in data completeness demands that
simply will not work at the HL-LHC, where the huge datasets all but
guarantee that some fraction of the data will be unavailable at any
given point in time.

At HL-LHC, as energy and luminosity conditions are quasi stable across
the years, datasets from multiple years with different conditions will
need to be analyzed in a coherent way. The analysis software will then
need to be able to fetch and use the proper metadata automatically while
today's analysis software often requires dedicated configuration and
tuning for each data-taking period. Belle II have taken some steps to
address this by using the ``global tag'' concept commonly used in
reconstruction and applying it to analysis, allowing users to better
organise and store their analysis metadata. Nevertheless, until all
metadata is organised under the same global tag umbrella, which is
extremely challenging, the problem remains. Improvements here are
necessary both to permit efficient studies on partial datasets, and to
reduce the risk of user error in metadata access.

The growing complexity of analysis codes makes them extremely fragile
(i.e. bug-prone), hardly reusable, and unsuited for analysis
preservation needs. The current best efforts at analysis preservation
are based on a snapshot of the full analysis setup that can ensure the
possibility to re-run the code on the original dataset in a few years.
This may allow a future analyst to reproduce previous results, but will
likely not provide any understanding, for the future analyser, about
what the analysis was effectively doing. Another use case is to re-use
the analysis code on new data, be that the same dataset with improved
calibrations or additional data that will improve the precision of the
measurement. This brings additional difficulties, not least related to
metadata.

Nevertheless, these preservation efforts encourage better organisation
of the analysis as a workflow, and promote the use of version control
tools together with continuous integration, which offers a natural route
to improving analysis code quality. In addition to these code quality
measures, a longer-term solution for code complexity may be the adoption
of a Domain-Specific Language (DSL) {[}10{]}, discussed in more detail
later, a model that would have physicists write logical descriptions of
their analyses, leaving low-level code generation to a backend. Thus
analysis design could be quickly understood and shared, sidestepping the
usual dearth of documentation, while simultaneously isolating analysis
design from implementation and hardware, providing a natural means of
analysis preservation.

\hypertarget{plans-and-recent-progress}{%
\subsubsection{Plans and recent
progress}\label{plans-and-recent-progress}}

Most of the problems identified above were already identified in the
community white paper {[}17{]} from 2017 and since then some progress
was made prototyping new technologies to enable data analysis at the
HL-LHC.

A first branch of new technologies is that of efficient data analysis
processing in the last steps of the analysis, i.e. when event data needs
to be selected, derived and aggregated into histograms. Several
platforms developed by industry or data science have emerged in the past
years to quickly aggregate large datasets with parallel execution
(either on clusters of computers or simply on multi-core servers). Many
of those tools have been tested to understand the feasibility of usage
in the HEP context {[}references here to spark tests, google cloud,
pyhep stuff, etc\ldots{]}.

In particular, the Python ecosystem, implementing many such solutions
(e.g. numpy, pandas dataframes), is emerging as a possible alternative
to HEP's traditional C++ environment. This is thanks largely to a
combination of its versatility as a language, allowing rapid
proto-typing, and the ability to out-source compute-intensive work to
more performant languages like C and C++. In fact this is similar to the
experimental software frameworks where Python is used as steering code
and C/C++ is used as an efficient backend implementation for the Python
modules. The typical Belle II analysis starts with Python using a custom
DSL to specify particle selections and algorithms implemented in C++ and
output ntuple content. The ntuple is typically analysed by the PyHEP
toolkit {[}6{]}, and also benefits from leveraging packages from the
wider data science community. Deeper integration with python tools is
also particularly important for enabling state-of-the-art machine
learning frameworks such as PyTorch and TensorFlow, whose relevance in
analysis is likely to grow in the long term. It is important to provide
enough usable functionality in the optimised layer to avoid the use of
slow loops in Python, which would cripple performance. Tools like
PyROOT, which allows the use of any C++ class from Python, can play an
instrumental role in enlarging the whole analysis ecosystem with an
effortless integration of Python and C++ components, in particular
consolidating HEP specific tools written in C++ with industry python big
data functionality.

In addition to adopting externally available tools, HEP developers are
prototyping similar concepts in more domain-specific implementations. A
core requirement of the HEP analysis toolkit is the representation and
manipulation of physics data where each event contains scalars (i.e.
single event variables), objects (i.e. variables related to the same
physics object) and vectors of objects (i.e. a variable number of
physics objects). In addition the Lorentz vector algebra that is bread
and butter for HEP analysis needs to be naturally accommodated.

The ROOT package, i.e. the foundation and workhorse for LHC analysis,
has been the focus of substantial development. On the data storage
front, the ROOT team demonstrated better data compression and
accelerated reading for a wide variety of data formats. Information
density drives compression and this can vary massively between
experiments and analyses. This new format called RNTuple {[}9{]}, an
evolution of the TTree file format, shows robust and significant
performance improvements (1-5x faster), which could potentially save
significant storage space (10-20\%) for the HL-LHC. Another area of
progress, and potential consolidation in ROOT I/O, is lossy compression,
where the number of significant bits for a quantity may be far fewer
than a standard storage type's mantissa, enabling further savings in
storage space. The event processing framework was extended with
DataFrame-like functionality (RDataFrame) implementing a declarative
approach to analysis processing in either Python or C++ (always executed
in C++), which natively exploits multi-core architectures {[}7{]} and
could be extended to accelerators. Flexibility in interfacing RDataFrame
to non-ROOT data formats has allowed the ALICE collaboration to
prototype its Run 3 analysis model on this new technology. Growing use
of Machine Learning was also anticipated, and the TMVA toolkit has been
improved with substantial optimisations, more complex deep learning
models and enhanced interoperability with common data science tools
{[}14,15{]} which is of particular importance for inference.

A complementary approach provided in Python is the toolchain of uproot,
awkward arrays and coffea {[}8{]} for efficient columnar operations and
transformations of the data. These function as a lightweight end-stage
analysis environment that provides only the elements necessary for
pre-processing ROOT inputs and converting them into formats used by the
standard ML frameworks, typically numpy. Lightweight distribution using
package managers such as pip or conda allows for rapid setup and
extension with other Python tools. Good performance appears to have been
achieved within the scope of these projects, which has been kept focused
-\/- in the event of uptake by a large portion of the community,
long-term support for these modules would need to be established.

A further feature linked to the growing use of Python is the use of
``notebook'' technology such as Jupyter. This permits quasi-interactive
exploration where the annotated history including formatted outputs and
graphics can be saved and shared with collaborators for reproduction and
adaptation. Although not a substitute for command line scripts in
well-tested and complex workflows, notebooks make an effective vessel
for software education and have been leveraged as a powerful frontend
for access to facilities including CERN's Swan.

\hypertarget{prospects-and-rd-needs}{%
\subsubsection{Prospects and R\&D needs}\label{prospects-and-rd-needs}}

As the complexity of analysis data handling and processing grows,
developing efficient and robust code becomes harder and harder. In
addition, analysis code is typically reimplemented by tens of
individuals, in different experiments, analysis teams or institutes
mostly performing the same kind of operations (data reduction, plotting,
variation of systematic uncertainty, fitting etc\ldots). An appealing
idea, circulating for a long time in the HEP world, is to develop a
domain specific analysis language to describe any analysis in a concise
way {[}10{]}. An analysis backend would then translate this into
optimised code to effectively run the analysis exploiting the full power
of different kinds of platforms and architectures. At a more basic
level, declarative rather than imperative programming is visibly
becoming more prevalent in experimental software configuration, with
consequent improvement in the transparency of steering code, setting a
good example for analysis.

For HL-LHC, the growing analysis complexity, rapidly evolving compute
architectures and the need for analysis code to be highly optimised
could make the usage of analysis languages a must.

Prototypes of analysis languages have been developed in the past years
{[}refs, 10?{]} for the event processing and histogram production parts
of the analysis workflow and probably with some additional effort the
HEP community can converge on broadly usable tools. Similar efforts can
also be investigated in the context of data interpretation (e.g.
fitting) standardizing the description of what to fit and how (an
interesting example here is the so called ``combine tools'' developed by
CMS and based on RooFit for the Higgs discovery) {[}ref{]}. There is an
obvious need for good interfacing with analysis backends, whether these
be local CPU, batch or grid resources, for which there is some
experience within the experimental collaborations as well as in data
science frameworks.

Eight initial benchmark challenges have been defined to set the scope
that should be addressed by analysis languages {[}11{]}. Several
prototypes have already been developed, some of which solve non-trivial
challenges with concise descriptions. What remains to be demonstrated is
the capability of such systems to scale to full analysis complexity,
including tasks such as handling of systematic uncertainties,
data-driven background estimation, which are yet to be incorporated into
the benchmarks. Their performance in terms of CPU and memory consumption
also has to be fairly assessed. It is also a point of concern that
widespread adoption of analysis languages could pigeonhole HEP analysers
into the role of expert users of domain-specific languages, and
resulting in stratification between backend developers and users.

A similar, and somewhat related, topic for R\&D is that of hardware
resources and infrastructures needed for analysis. At some stage in the
analysis processing interactivity or fast turn-around is needed.
Quantitatively speaking, while typical grid task completion has a time
scale of one day to one week, in the context of analysis exploration and
developments answers are needed on timescales of a few seconds to a few
hours in order to keep people productive. As analysis tasks are often
IO-intensive and not necessarily-CPU intensive, the right tradeoff
between CPU and high bandwidth storage should be carefully studied when
designing the computing infrastructure of HL-LHC. As a proof of
principle, an unoptimized Higgs discovery analysis has been rerun on
Google Cloud in a few minutes as the infrastructure was able to quickly
provide tens of thousands of CPUs with granted bandwidth to storage of 2
Gb/coresec {[}12{]}. Scaling this to sustained, diverse analysis loads
will be far from trivial. R\&D is needed to understand current usage
patterns and project them to HL-LHC in the last steps of the analysis so
that the right balance of resources can be procured. An ongoing topic of
discussion is how to best provision computing resources for end stage
analysis, where there are many open questions {[}13{]}.

There has already been substantial experimentation with different grid
site architectures, in particular with regards to I/O, e.g. purely
remote data access as opposed to local storage. Various job allocation
and management models, many incorporating industry solutions, have been
trialled, but the volatility of analysis loads remains a challenge to be
overcome. More detailed analysis of resource usage and its evolution,
especially as pertains to the need for heterogeneous compute, is needed
to correctly design hardware allocation.

\hypertarget{priority-topics}{%
\subsubsection{Priority topics}\label{priority-topics}}

In summary, the following goals in analysis computing stand out as foci
for development towards the HL-LHC era:

\begin{itemize}
\item
  \begin{quote}
  Identify a CPU-efficient data reduction and storage model that serves
  the majority of analyses with a footprint of O(1kb/event), retaining
  the flexibility to accommodate the needs of specialised analyses;
  \end{quote}
\item
  \begin{quote}
  Streamline analysis metadata access and calibration schemes, and
  provide effective book-keeping for fractional datasets;
  \end{quote}
\item
  \begin{quote}
  Develop cross-experimental tools establishing declarative syntax
  and/or languages for analysis description, interfaced with distributed
  computing backends; and
  \end{quote}
\item
  \begin{quote}
  Define and improve schemes for interoperability of end-stage
  experimental software with data science/machine learning frameworks.
  \end{quote}
\end{itemize}

\hypertarget{references-2}{%
\subsubsection{References}\label{references-2}}

{[}1{]} *ATLAS Collaboration, \emph{Evolution of the ATLAS analysis
model for Run-3 and prospects for HL-LHC}, ATL-SOFT-PROC-2020-002,
\href{https://cds.cern.ch/record/2708664/}{{https://cds.cern.ch/record/2708664/}}

{[}2{]} CMS Collaboration, HL-LHC resource projections,
\href{https://twiki.cern.ch/twiki/bin/view/CMSPublic/CMSOfflineComputingResults}{{https://twiki.cern.ch/twiki/bin/view/CMSPublic/CMSOfflineComputingResults}}

{[}3{]} *ALICE Collaboration, \emph{The ALICE analysis train system},
J.Phys.Conf.Ser. 608 (2015) no.1, 012019,
\href{http://arxiv.org/abs/arXiv:1502.06381}{{http://arxiv.org/abs/arXiv:1502.06381}}

{[}4{]} *ATLAS Collaboration, \emph{A new petabyte-scale data derivation
framework for ATLAS. Journal of Physics: Conference Series}, 664.
072007.

{[}5{]} *CMS Collaboration, \emph{A further reduction in CMS event data
for analysis: the NANOAOD format}, EPJ Web of Conferences. 214. 06021.

{[}6{]}
\href{https://github.com/hsf-training/PyHEP-resources}{{https://github.com/hsf-training/PyHEP-resources}}

{[}7{]} D Piparo et al, \emph{RDataFrame: Easy Parallel ROOT Analysis at
100 Threads}, EPJ Web of Conferences 214, 06029 (2019),
\href{https://www.epj-conferences.org/articles/epjconf/abs/2019/19/epjconf_chep2018_06029/epjconf_chep2018_06029.html}{{https://www.epj-conferences.org/articles/epjconf/abs/2019/19/epjconf\_chep2018\_06029/epjconf\_chep2018\_06029.html}}

{[}8{]}
\href{https://github.com/scikit-hep/uproot}{{https://github.com/scikit-hep/uproot}},
\href{https://github.com/scikit-hep/awkward-array}{{https://github.com/scikit-hep/awkward-array}},
\href{https://github.com/CoffeaTeam}{{https://github.com/CoffeaTeam}},
\href{https://github.com/diana-hep/pyhf}{{https://github.com/diana-hep/pyhf}}

{[}9{]} J Blomer et al, \emph{Evolution of the ROOT Tree I/O},
\href{https://arxiv.org/abs/2003.07669}{{https://arxiv.org/abs/2003.07669}}

{[}10{]} ADL Workshop at Fermilab,
\href{https://indico.cern.ch/event/769263/timetable/?view=standard}{{https://indico.cern.ch/event/769263/}},
Gordon's CHEP talk (good overview of links, is there something more
official?)
\href{https://indico.cern.ch/event/773049/contributions/3476174/attachments/1938123/3220511/Declarative.pdf}{{https://indico.cern.ch/event/773049/contributions/3476174/attachments/1938123/3220511/Declarative.pdf}}

{[}11{]}
\href{https://github.com/iris-hep/adl-benchmarks-index}{{https://github.com/iris-hep/adl-benchmarks-index}}

{[}12{]} L Heinrich \& R Rocha, \emph{Reperforming a Nobel Prize
discovery on Kubernetes}, talk at CHEP 2019,
\href{https://indico.cern.ch/event/773049/contributions/3581373}{{https://indico.cern.ch/event/773049/contributions/3581373}}
(originally at KubeCon)

{[}13{]} *ALICE Collaboration, \emph{The ALICE analysis framework for
LHC Run 3}, EPJ Web Conf. 214 (2019) 05045
\href{http://cds.cern.ch/record/2699856}{{http://cds.cern.ch/record/2699856}}

{[}14{]} \href{http://paperpile.com/b/2mqsuS/IDej}{L Moneta et al,
\emph{Machine Learning with ROOT/TMVA}, EPJ Web of Conferences;
Proceedings of CHEP 2019 (in print). 2020.}

{[}15{]} \href{http://paperpile.com/b/2mqsuS/6uIv}{K Albertsson et al,
\emph{Fast Inference for Machine Learning in ROOT/TMVA}, EPJ Web of
Conferences; Proceedings of CHEP 2019 (in print). 2020.}

{[}16{]} J Blomer, \emph{A quantitative review of data formats for HEP
analyses}, Journal of Physics: Conference Series. 2018. p. 032020.

{[}17{]} HEP Software Foundation Collaboration, \emph{HEP Software
Foundation Community White Paper Working Group - Data Analysis and
Interpretation},
\href{https://arxiv.org/abs/1804.03983}{{arXiv:1804.03983}}

Proceedings marked `*' will be credited to the shorter author lists

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

Final wrap-up of the most important messages and make some final points
that expand the scope. For now collect points that we may want to
mention here:

\begin{itemize}
\item
  \begin{quote}
  May want to mention quantum computing and other disruptive technology
  for later on in the HL-LHC programme
  \end{quote}

  \begin{itemize}
  \item
    \begin{quote}
    ``In the more distant future devices such as quantum or neuromorphic
    devices may provide a substantial processing power.'' (From
    Federico)
    \end{quote}
  \end{itemize}
\item
  \begin{quote}
  Considerations to be made about attracting scientists to sw
  development or sw teaching
  \end{quote}
\end{itemize}

In total selfish self-aggrandizing (but not only), I propose to mention
more extensively (and not necessarily in the conclusion).

\begin{itemize}
\item
  \begin{quote}
  R\&D on Quantum Computing. At the moment any statement on QC is
  unsubstantiated, both those that foresee a bright tomorrow and those
  that predict it will never work. However my personal opinion is that
  some limited R\&D in this field is necessary to inform the Physics
  community of its potential and its limitations. Should it really
  become a viable production technology, it will be essential to have
  few people in the community that will guide the colleagues on how to
  make the best use of it. Had we had more of these people for GPUs, we
  would be in a better position today.
  \end{quote}
\item
  \begin{quote}
  Relations with IT industry are important. The CERN openlab model was
  cited even by Fabiola in her conclusions in Grenada. We should mention
  that this kind of collaborations is fruitful and that we have a well
  established and efficient model for that (which of course does not
  mean that it is not perfectible!). May be a short section describing
  how it works and how it could be developed would fit into this
  document.
  \end{quote}
\item
  \begin{quote}
  I would explicitly mention Neuromorphic computing and DNA storage as
  ``exotic'' possiblities, and other too.
  \end{quote}
\end{itemize}

Remember\ldots{} the ``start'' of HL-LHC is 2027, but the end is 2040!
Will QC and other more exotic possibilities be production ready for
2027? The jury is out. Will they be in 2035? This is an entirely
different question and the answer may well be yes.

\end{document}

### 2020-04-23 LHCC/2020/2020-01/ctcs.tex
Update to current version of the document (with no references!)

@@
 % JHEP preprint template
-\documentclass[11pt,a4paper]{article}
+\documentclass[10pt,a4paper]{article}
 \usepackage{jheppub}
@@
 \usepackage{xspace}
+
+% Stop lists from using so much space
+\usepackage{enumitem}
+\setlist{noitemsep}
 
@@
 
-The challenges for future software for high-energy physics are driven by
-an ambitious physics programme, notably the LHC accelerator upgrade to
-high-luminosity, HL-LHC, and the corresponding detector upgrades of
-ATLAS and CMS. The General Purpose Detectors describe their specific
-challenges elsewhere; in this document we address the challenges for
-software that is used in multiple experiments (usually even more widely
-than ATLAS and CMS) and maintained by teams of developers who are either
-not linked to a particular experiment or who contribute to common
-software within the context of their experiment activity. We also give
-space to general considerations for future software and projects that
-tackle upcoming challenges, no matter who writes it, which is an area
-where community convergence on best practice is extremely useful.
+Common and community software packages, such as ROOT, Geant4 and event
+generators have been a key part of the LHC's success so far and
+continued development and optimisation will be critical in the future
+\textbackslash cite\{HSF-EPPSU-Document, EPPSU-Briefingbook\}. The
+challenges are driven by an ambitious physics programme, notably the LHC
+accelerator upgrade to high-luminosity, HL-LHC, and the corresponding
+detector upgrades of ATLAS and CMS. The General Purpose Detectors
+describe their specific challenges elsewhere; in this document we
+address the issues for software that is used in multiple experiments
+(usually even more widely than ATLAS and CMS) and maintained by teams of
+developers who are either not linked to a particular experiment or who
+contribute to common software within the context of their experiment
+activity. We also give space to general considerations for future
+software and projects that tackle upcoming challenges, no matter who
+writes it, which is an area where community convergence on best practice
+is extremely useful.
 
@@
 physics output of the LHC programme. This situation will get
-significantly worse in the high luminosity running phase of the LHC
+significantly worse in the high-luminosity running phase of the LHC
 (HL-LHC). The fraction of the CPU resources available to the LHC
@@
 
-This document gives an overview of the many technical and non-technical
+This section gives an overview of the many technical and non-technical
 challenges in the generator area and of the work that can be done to
@@
 high-priority items, for the next 18 months before the LHCC review that
-is currently scheduled for fall 2021.
+is currently scheduled for fall 2021. A more detailed version of this
+contribution, including a more complete list of references, will soon be
+uploaded on arxiv and submitted for publication {[}Gen20{]}.
 
@@
 on published papers, and may not be motivated to work on software
-optimizations that are not "theoretical" enough to advance their
+optimisations that are not "theoretical" enough to advance their
 careers. Generator support tasks in the experiments may also not be
@@
 
-- {Mismatch in optimization focus}. The goals and metrics of software
-optimization work may also differ. Theorists are mainly interested in
-calculating cross sections and focus on minimising the phase space
-integration time for a given statistical precision. The LHC experiments
-run large scale productions of unweighted event generation, and mainly
-need to maximize the throughput of events generated per unit time on a
-given node.
+- {Mismatch in usage patterns and optimisation focus}. The way
+generators are built and used by their authors is often different from
+the way in which they are deployed and integrated by the experiments in
+their software frameworks and computing infrastructure.The goals and
+metrics of software optimisation work may also differ. Theorists are
+mainly interested in calculating cross sections and focus on minimising
+the phase space integration time for a given statistical precision. The
+LHC experiments run large scale productions of unweighted event
+generation, and mainly need to maximize the throughput of events
+generated per unit time on a given node.
 
@@
 - {Merging inefficiency}. Merging prescriptions (e.g. MLM {[}Man02{]},
-CKKW-L {[}Lon05{]} at LO and FxFx {[}Fre12{]} at NLO) imply the
-rejection of some events, to avoid double counting between events
-produced with n\textsubscript{jets}+1 MEs and with n\textsubscript{jets}
-MEs plus parton showers. The resulting inefficiencies can be relatively
-low depending on the process, but they are unavoidable in the
-algorithmic strategy used by the underlying physics modeling
-{[}Alw08{]}. However, a method like shower-kt MLM can reduce the merging
-inefficiency of MLM {[}Alw09{]}.
+CKKW-L {[}Lon05{]} at LO and FxFx {[}Fre12{]}, MEPS@NLO {[}Hoe14{]} at
+NLO) imply the rejection of some events, to avoid double counting
+between events produced with n\textsubscript{jets}+1 MEs and with
+n\textsubscript{jets} MEs plus parton showers. The resulting
+inefficiencies can be relatively low depending on the process, but they
+are unavoidable in the algorithmic strategy used by the underlying
+physics modeling {[}Alw08{]}. However, a method like shower-kt MLM can
+reduce the merging inefficiency of MLM {[}Alw09{]}.
 
@@
 repeatedly at many phase space points, lends itself naturally to the
-data parallel approaches found in CPU vectorized code and in GPU compute
-kernels. Porting and optimizing generators on GPUs is essential to be
+data parallel approaches found in CPU vectorised code and in GPU compute
+kernels. Porting and optimising generators on GPUs is essential to be
 able to use modern GPU-based HPCs (such as SUMMIT, where 95\% of the
@@
 Multi-processing approaches may also be useful to speed up the
-integration and optimization step for complex high-dimensional final
+integration and optimisation step for complex high-dimensional final
 states. In particular, a lot of work has been done to implement
@@
 
-Out of the issues outlined above, the WG identifies the following as its
-current main priorities:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  \begin{quote}
-  Gain a more detailed understanding of current CPU costs by accounting
-  and profiling.
-  \end{quote}
-\item
-  \begin{quote}
-  Survey generator codes to understand the best way to move to GPUs and
-  vectorized code, and prototype the port of the software to GPUs using
-  data-parallel paradigms.
-  \end{quote}
-\item
-  \begin{quote}
-  Support efforts to optimize phase space sampling and integration
-  algorithms, including the use of Machine Learning techniques such as
-  neural networks.
-  \end{quote}
-\item
-  \begin{quote}
-  Promote research on how to reduce the cost associated with negative
-  weight events, using new theoretical or experimental approaches.
-  \end{quote}
-\item
-  \begin{quote}
-  Promote collaboration, training, funding and career opportunities in
-  the generator area.
-  \end{quote}
-\end{enumerate}
-
-\hypertarget{section-1}{%
-\subsection{}\label{section-1}}
-
-\hypertarget{references}{%
-\subsection{References}\label{references}}
-
-{[}Alb19{]} J. Albrecht et al. ``A Roadmap for HEP Software and
-Computing R\&D for the 2020s'' Comput Softw Big Sci (2019) 3, 7,
-\href{https://doi.org/10.1007/s41781-018-0018-8}{{doi:10.1007/s41781-018-0018-8}},
-\href{https://arxiv.org/abs/1712.06982}{{arXiv:1712.06982}}
-
-{[}Alw08{]} J. Alwall et al. ``Comparative study of various algorithms
-for the merging of parton showers and matrix elements in hadronic
-collisions'', Eur. Phys. J. C 53 (2008) 53,
-\href{https://doi.org/10.1140/epjc/s10052-007-0490-5}{{doi:10.1140/epjc/s10052-007-0490-5}}
-
-{[}Alw09{]} J. Alwall, S. de Visscher, F. Maltoni, ``QCD radiation in
-the production of heavy colored particles at the LHC'', JHEP 0902 (2009)
-017,
-\href{https://doi.org/10.1088/1126-6708/2009/02/017}{{doi:10.1088/1126-6708/2009/02/017}}
-
-{[}Alw14{]} J. Alwall et al. ``The automated computation of tree-level
-and next-to-leading order differential cross sections, and their
-matching to parton shower simulations'', JHEP 07 (2014) 079,
-\href{https://doi.org/10.1007/JHEP07(2014)079}{{doi:10.1007/JHEP07(2014)079}}
-
-{[}Bau13{]} C. Bauer et al., "Computing for perturbative QCD - a
-Snowmass White Paper",
-\href{https://arxiv.org/abs/1309.3598}{{arxiv:1309.3598}}
-
-{[}Ben17{]} J. Bendavid, ``Efficient Monte Carlo Integration Using
-Boosted Decision Trees and Generative Deep Neural Networks'',
-\href{https://arxiv.org/abs/1707.00028}{{arXiv:1707.00028}}
-
-{[}Bot19{]} E. Bothmann et al., ``Event generation with Sherpa 2.2'',
-\href{https://scipost.org/SciPostPhys.7.3.034}{{SciPost Phys. 7 (2019)
-034}}
-
-{[}Bot20{]} E. Bothmann et al. ``Exploring phase space with Neural
-Importance Sampling'',
-\href{https://arxiv.org/abs/2001.05478}{{arXiv:2001.05478}}
-
-{[}Chi17{]} J. T. Childers et al., "Adapting the serial Alpgen
-parton-interaction generator to simulate LHC collisions on millions of
-parallel threads", Comp. Phys. Comm. 210 (2017) 5154.
-\href{https://doi.org/10.1016/j.cpc.2016.09.0135286}{{https://doi.org/10.1016/j.cpc.2016.09.0135286}}.
-
-{[}Fre12{]} R. Frederix, S. Frixione, "Merging meets matching in
-MC@NLO", JHEP 2012 (2012) 61.
-\href{https://doi.org/10.1007/JHEP12(2012)061}{{doi:10.1007/JHEP12(2012)061}}
-
-{[}Fre20{]} R. Frederix, S. Frixione, S. Prestel. P. Torrielli, ``On the
-reduction of negative weights in MC@NLO-type matching procedures'',
-\href{https://arxiv.org/abs/2002.12716}{{arXiv:2002.12716}}
-
-{[}Fri17{]} S. Frixione, P. Nason, C. Oleari, "Matching NLO QCD
-computations with parton shower simulations: the POWHEG method", JHEP11
-(2007) 070,
-\href{https://doi.org/10.1088/1126-6708/2007/11/070}{{doi:10.1088/1126-6708/2007/11/070}}
-
-{[}Gao20{]} C. Gao. et al. ``Event Generation with Normalizing Flows'',
-\href{https://arxiv.org/abs/2001.10028}{{arXiv:2001.10028}}
-
-{[}Gen18{]} HSF Physics Event Generator Computing Workshop, CERN,
-November 2018,
-\href{https://indico.cern.ch/event/751693}{{https://indico.cern.ch/event/751693}}
-
-{[}Gen19{]} HSF Physics Generators Working Group,
-\href{https://hepsoftwarefoundation.org/workinggroups/generators.html}{{https://hepsoftwarefoundation.org/workinggroups/generators.html}}
-
-{[}Gle08{]} T. Gleisberg, S. Hoeche, "Comix, a new matrix element
-generator", JHEP 12 (2008) 039.
-\href{https://doi.org/10.1088/1126-6708/2008/12/039}{{doi:10.1088/1126-6708/2008/12/039}}
-
-{[}Hoe19{]} S. Hoeche, S. Prestel, H. Schulz, "Simulation of vector
-boson plus many final jets at the high luminosity LHC", Phys. Rev. D
-100, 0140124 (2019),
-\href{http://dx.doi.org/10.1103/PhysRevD.100.014024}{{doi:10.1103/PhysRevD.100.014024}}
-
-{[}Jad03{]} S. Jadach, Foam: a general-purpose cellular Monte Carlo
-event generator, Comp. Phys. Comm. 152 (2003) 55.
-\href{https://doi.org/10.1016/S0010-4655(02)00755-5}{{doi:10.1016/S0010-4655(02)00755-5}}
-
-{[}Kaw86{]} S.Kawabata, A new Monte Carlo event generator for high
-energy physics, Comp. Phys. Comm. 41 (1986) 127.
-\href{https://doi.org/10.1016/0010-4655(86)90025-1}{{doi:10.1016/0010-4655(86)90025-1}}
-
-{[}Kli18{]} M.D. Klimek, M. Perelstein, ``Neural Network-Based Approach
-to Phase Space Integration'',
-\href{https://arxiv.org/abs/1810.11509}{{arXiv:1810.11509}}
-
-{[}Kon20{]} D. Konstantinov, "Optimization of Pythia8", EP-SFT group
-meeting, CERN, January 2020,
-\href{https://indico.cern.ch/event/890670}{{https://indico.cern.ch/event/890670}}
-
-{[}Lep80{]} G. P. Lepage, "VEGAS: an adaptive multi-dimensional
-integration program", Cornell report CLNS-447 (1980).
-\href{https://cds.cern.ch/record/123074}{{https://cds.cern.ch/record/123074}}
-
-{[}Lon05{]} L. Lonnblad, "Correcting the colour-dipole cascade model
-with fixed order matrix elements", JHEP 05 (2002) 046.
-\href{https://doi.org/10.1088/1126-6708/2002/05/046}{{doi:10.1088/1126-6708/2002/05/046}}
-
-{[}Mal03{]} F. Maltoni, T. Stelzer, "MadEvent: automatic event
-generation with MadGraph", JHEP 02 (2003) 027.
-\href{https://doi.org/10.1088/1126-6708/2003/02/027}{{doi:10.1088/1126-6708/2003/02/027}}
-
-{[}Man02{]} M. L. Mangano, M. Moretti, R. Pittau, "Multijet matrix
-elements and shower evolution in hadronic collisions: Wbb + n-jets as a
-case study", Nucl. Phys. B 632 (2002) 343.
-\href{https://doi.org/10.1016/S0550-3213(02)00249-3}{{doi:10.1016/S0550-3213(02)00249-3}}
-
-{[}Mat16{]} O. Mattelaer, "On the maximal use of Monte Carlo samples:
-re-weighting events at NLO accuracy", \emph{Eur. Phys. J. C}
-76\textbf{,} 674 (2016),
-\href{https://doi.org/10.1140/epjc/s10052-016-4533-7}{{doi:10.1140/epjc/s10052-016-4533-7}}
-
-{[}Mat18{]} O. Mattelaer, "MG5aMC status and plans", in {[}Gen19{]}
-
-{[}Nas07{]} P. Nason, MINT: a Computer Program for Adaptive Monte Carlo
-Integration and Generation of Unweighted Distributions Bicocca-FT-07-13
-(2007). \href{https://arxiv.org/abs/0709.2085}{{arxiv:0709.2085}}
-
-{[}SKe18{]} L. Sexton-Kennedy, G. Stewart, "CWP challenges and workshop
-aims", in {[}Gen19{]}
+{[}Text moved to final summary{]}
 
@@
 with the HL-LHC, will create unique research opportunities by collecting
-orders of magnitude more data. So at the same time, the demand for
+an order of magnitude more data. So at the same time, the demand for
 detector simulation will grow accordingly, in order to keep the
@@
 experiment simulation for physics studies over the last two decades.
-Designed in the late 90' to cope with the simulation challenges of the
+Designed in the late 1990s to cope with the simulation challenges of the
 LHC era, Geant4 introduced flexible physics modeling, exploiting layers
@@
 of the LHC experiments make use of this toolkit and, as such, most of
-the LHC physics program relies on it. It has delivered physics precision
-adequate to allow the experiments to perform precision analysis of the
-collected data and to produce new results pushing the boundaries of our
-understanding in HEP. At the same time, the code of the toolkit is
+the LHC physics program relies on it. It has delivered a physics
+precision that allows the experiments to perform precision analysis of
+the collected data and to produce new results pushing the boundaries of
+our understanding in HEP. At the same time, the code of the toolkit is
 becoming old, with some parts written almost thirty years ago, making it
@@
 
-Improvements of physics models need to continue to take place, however,
-their exact influence on the precision of physics analysis is not always
-easy to quantify. On the other hand, the overall CPU performance of the
-current code will not improve drastically just by relying on modern
-hardware or better compilers. Taking all those elements into account,
-three main development paths have emerged to be undertaken
-simultaneously: these have started to be pursued, however, effort in any
-one of them is far from sufficient to exploit their individual
-potential. On one hand, it is needed to continue to modernize and
-refactor the simulation code by using more optimal, modern programming
-techniques. Those improvements, requiring considerable effort but no
-major paradigm changes, can reasonably bring several tens of percent of
-speed-up, by improving the throughput on modern CPUs, avoiding constant
-churn in data and instruction caches. This will certainly help to
-produce more simulated events, but it will not be sufficient to meet the
-high luminosity needs of the LHC experiments. On the other hand,
-different fast simulation techniques, where tracking of individual
-particles in given detectors is replaced by some sort of
-parameterization, have already been extensively used for LHC simulation
+Improvements of physics models need to continue to take place to prevent
+systematic uncertainties from the simulation from becoming a dominant
+factor, however, their exact influence on the precision of physics
+analysis is not always easy to quantify. On the other hand, the overall
+CPU performance of the current code will not improve drastically just by
+relying on modern hardware or better compilers. Taking all those
+elements into account, three main development paths have emerged to be
+undertaken simultaneously: these have started to be pursued, however,
+effort in any one of them is far from sufficient to exploit their
+individual potential. On one hand, it is needed to continue to modernise
+and refactor the simulation code by using more optimal, modern
+programming techniques. Those improvements, requiring considerable
+effort but no major paradigm changes, can reasonably bring several tens
+of percent of speed-up, by improving the throughput on modern CPUs,
+avoiding constant churn in data and instruction caches. This will
+certainly help to produce more simulated events, but it will not be
+sufficient to meet the high luminosity needs of the LHC experiments. On
+the other hand, different fast simulation techniques, where tracking of
+individual particles in given detectors is replaced by some sort of
+parameterisation, have already been extensively used for LHC simulation
 campaigns, where some precision is traded to achieve higher event
@@
 geometrical description of the detectors. The toolkit design principles
-was to allow flexibility in future extensions of its own code, as well
+were to allow flexibility in future extensions of its own code, as well
 as choices for particular applications. It is clear that this
 flexibility comes at the price of a performance penalty. We observe that
-writing more compact and more specialized code can lead to several
-percent speed-up. The technical improvements to explore here consist of
-avoiding too many virtual layers and improving the instruction and data
-locality, leading to better cache use. Going even further as far as
-specialization is concerned, studies suggest that implementing compact,
-self-contained transport libraries with a minimal set of physics models
-(for instance only electromagnetic interaction for selected particles)
-with predefined scoring (to simulate, for instance, the response of an
-EM calorimeter) can also bring speed-ups to the complete
-application{[}Ref to GeantV RD{]}. As far as architectural modifications
-are concerned, the design of the Geant4 track-status machinery prevents
-the introduction of fine-grained track parallelism. It is still not
-clear whether this kind of parallelism can bring any substantial
-speed-up, due to the overheads; however, making the Geant4 kernel and
-physics processes stateless would certainly provide the necessary
-starting points for further R\&D in that direction. Most of these
-investigations have started and it is hoped to see some first results on
-a year time-scale. More effort, however, will be required to integrate
-any eventual changes in the production code in a timely manner. The
-GeantV Vector prototype has allowed the assessment of the achievable
-speedup using a novel, vectorized approach to particle transport. The
-demonstrator of a full electromagnetic shower in realistic (CMS)
-geometry has been implemented and the comparison to a similar Geant4
-application has been performed. The conclusion of that work showed that
-the main factors in the speedup seem to include better cache use and
-tighter code, while the vectorization impact was much smaller than hoped
-for. This unimpressive impact is due to a set of factors, including the
-fraction of the current algorithms that could be vectorized, unresolved
-challenges for the gather/scatter and tail handling (to keep the number
-of events in flight within bound) needed for large number of shapes, and
-of course Amdahl's Law applied to vectorized where some bottleneck (for
-example geometry navigation) could not be tackled without a major
-additional long term effort. {[}Ref-GeantV-Final-Report{]}. At the same
-time libraries developed in the context of the GeantV R\&D, e.g.
-VecGeom, have been successfully integrated in Geant4 and ROOT
-benefitting the whole HEP software community {[}CMS01{]}. The
-cost/benefit of more vectorization of the code has not been deemed worth
-the necessary investment also due to the speculation that a complete
-rewrite of the physics base could become necessary to fully exploit it.
-As a result investing in the optimization and modernization of the
-Geant4 code has gained even more relevance.
-
-\hypertarget{experiments-optimization}{%
-\subsection{Experiments optimization}\label{experiments-optimization}}
+writing more compact and more specialised code can lead to several
+percent speed-up {[}GeantV RD{]}. The technical improvements to explore
+here consist of avoiding too many virtual layers and improving the
+instruction and data locality, leading to better cache use. Going even
+further as far as specialisation is concerned, studies suggest that
+implementing compact, self-contained transport libraries with a minimal
+set of physics models (for instance only electromagnetic interaction for
+selected particles) with predefined scoring (to simulate, for instance,
+the response of an EM calorimeter) can also bring speed-ups to the
+complete application {[}GeantV RD{]}. As far as architectural
+modifications are concerned, the design of the Geant4 track-status
+machinery prevents the introduction of fine-grained track parallelism.
+It is still not clear whether this kind of parallelism can bring any
+substantial speed-up, due to the overheads; however, making the Geant4
+kernel and physics processes stateless would certainly provide the
+necessary starting points for further R\&D in that direction. Most of
+these investigations have started and it is hoped to see some first
+results on a year time-scale. More effort, however, will be required to
+integrate any eventual changes in the production code in a timely
+manner. The GeantV Vector prototype has allowed the assessment of the
+achievable speedup using a novel, vectorized approach to particle
+transport. The demonstrator of a full electromagnetic shower in
+realistic (CMS) geometry has been implemented and the comparison to a
+similar Geant4 application has been performed. The conclusion of that
+work showed that the main factors in the speedup seem to include better
+cache use and tighter code, while the vectorisation impact was much
+smaller than hoped for. This unimpressive impact is due to a set of
+factors, including the fraction of the current algorithms that could be
+vectorised, unresolved challenges for the gather/scatter and tail
+handling (to keep the number of events in flight within bound) needed
+for large number of shapes, and of course Amdahl's Law applied to
+vectorisation where some bottleneck (for example geometry navigation)
+could not be tackled without a major additional long term effort.
+{[}GeantV{]}. At the same time libraries developed in the context of the
+GeantV R\&D, e.g. VecGeom, have been successfully integrated in Geant4
+and ROOT benefitting the whole HEP software community {[}CMS01{]}. The
+cost-over-benefit of more vectorisation of the code has not been deemed
+worth the necessary investment also due to the speculation that a
+complete rewrite of the physics base could become necessary to fully
+exploit it. As a result investing in the optimisation and modernisation
+of the Geant4 code has gained even more relevance.
+
+\hypertarget{experiments-applications-and-optimisation-of-the-use-of-geant4}{%
+\subsection{Experiments applications and optimisation of the use of
+Geant4}\label{experiments-applications-and-optimisation-of-the-use-of-geant4}}
 
@@
 exploit the options provided by the Geant4 toolkit to optimise the
-performance of their simulation code. All of the various handles
-provided are being continuously explored and adopted when deemed useful:
-range and physics cuts have been reviewed and customised by all
-experiments, shower libraries adopted as baseline in the forward region
-by ATLAS and CMS, stepping granularity optimized, magnetic field caching
+performance of their applications. All of the various handles provided
+are being continuously explored and adopted when deemed useful: range
+and physics cuts have been reviewed and customised by all experiments,
+shower libraries adopted as baseline in the forward region by ATLAS and
+CMS, stepping granularity optimized, magnetic field caching
 investigated, neutron russian roulette used for dedicated simulations.
@@
 multi-threading features within their own multi-threaded frameworks,
-harmonizing the different architectural choices made {[}CMS02, LHCb01,
-ATLAS?{]}, and have or are planning to use them in their production
-environment to gain access to a larger set of distributed resources.
-ATLAS has been running their multi-process simulation framework on HPCs
-systems for production in the last years {[}ATLAS02{]} while CMS has
-pioneered the use of their multi-threaded simulation on available
-resources including HPCs {[}CMS03{]}. Both approaches have allowed to
-exploit a higher number of computing resources by gaining access to
-systems with limited memory.
-
-\emph{Should we put here a bit on experiments fast simulation in the
-sense of all doing things about it and then give details in next
-session?}
+harmonizing their different architectural choices with the Geant4
+implementation {[}CMS02, LHCb01, ATLAS?{]}, and have or are planning to
+use them in their production environment to gain access to a larger set
+of distributed resources. ATLAS has been running their multi-process
+simulation framework on HPCs systems for production in the last years
+{[}ATLAS02{]} while CMS has pioneered the use of their multi-threaded
+simulation on available resources including HPCs {[}CMS03{]}. Both
+approaches have allowed to exploit a higher number of computing
+resources by gaining access to systems with limited memory.
 
@@
 over the next few years, more effort to be invested combining the
-physics and Machine Learningexpertise. Given that the training of the
+physics and Machine Learning expertise. Given that the training of the
 deep learning algorithms usually requires using sufficiently large MC
@@
 tools does not necessarily remove the need to speed up Geant4.
-Experiments are in the process of investigating the use of fast
-simulations for other types of detectors than calorimeters, e.g.
-Cerenkov based systems, as well as a fully parametric response at the
-level of reconstructed objects (tracks and clusters) for fast
-systematics verifications. In LHCb the re-use of part of the underlying
-events for specific signals of interest have been established when
-appropriate with particular care paid to ensure to keep under control
-bias on statistical uncertainties {[}LHCb02{]}, its applicability in
-ATLAS and CMS could be explored for specific cases. Similarly, the
-re-use of simulated or real events to mimic the background to
-hard-scatter events is also under investigation, where additional
-challenges exist concerning storage and I/O due to the handling of the
-large minimum bias samples needed to model the additional interactions.
+
+It is worth pointing out that different fast simulation techniques have
+already been extensively used for LHC simulation campaigns. Effort has
+been spent in the experiments frameworks to combine as much as possible
+fast and detailed implementation. In particular ATLAS and CMS are using
+in their baseline production shower libraries for the modelling of
+calorimeters in the forward region in combination with the detailed
+simulation of the rest of the sub-detectors. In LHCb the re-use of part
+of the underlying events for specific signals of interest has been
+established in production of samples when appropriate, with particular
+care paid to ensure keeping under control bias on statistical
+uncertainties {[}LHCb02{]}, its applicability in ATLAS and CMS could be
+explored for specific cases. Similarly, the re-use of simulated or real
+events to mimic the background to hard-scatter events is also under
+investigation, where additional challenges exist concerning storage and
+I/O due to the handling of the large minimum bias samples needed to
+model the additional interactions.
+
+All LHC experiments have explored the use of multi-objective regression
+and generative adversarial networks (GANs) {[}GAN-ATLAS, GAN-CMS,
+GAN-LHCb{]}, they are now in the process of investigating the use of
+fast simulations for other types of detectors than calorimeters, e.g.
+Cerenkov based systems {[}LAMARR{]} and Time Projection Chambers
+{[}ALICE01{]}. While implementations of fast simulations of given
+sub-detectors is specific to their technology and experimental
+environment and as such the responsibility of each LHC experiment, an
+area of potential common investigation is frameworks for fast tuning and
+validation.
+
+The use of fully parametric response of experimental setups at the level
+of reconstructed objects (tracks and clusters) for rapid evaluation of
+physic's reach is also exploited by the experiments with generic
+frameworks for fast simulation of typical collider detectors being used
+{[}Delphes, PGS4{]}. The advantage of such frameworks is to allow
+simultaneous studies for different experiments as well as their
+potential use by the theory community. Experiments specific fully
+parameterized simulations providing reconstructed objects compatible
+with the experiment's analysis frameworks for systematic verifications
+have also been emerging {[}Lamarr{]}.
 
@@
 threads, etc.) to efficiently (development time wise and run time wise)
-profit from it {[}Ref to HSF final review of GeantV{]}.
+profit from it {[}GeantV{]}.
 
@@
 addressed by a general HEP simulation. The application of GPUs has been
-very successful in domains like medical physics simulation, neutron
-transport or optical photon transport. In those applications, the type
-of particles considered is very limited (often just one type, with no
-secondary particle production), the set of physics processes is reduced
-and the geometry is often extremely simple compared to the LHC
-detectors. A natural `extrapolation' of those approaches to a general
-HEP simulation is very difficult to imagine, because the stochasticity
-of the simulated events would immediately lead to divergences as far as
-different GPU threads are concerned, not to mention, simply the
-feasibility of porting the whole simulation to the GPU code for which
-specific expertise would be needed. On the other hand, running only very
-small pieces of the simulation application on GPUs does not seem to be
-efficient either, as gathering data and transferring it from the host to
-the device and back may strongly limit any possible performance gain,
-similarly to what was seen with the GeantV vectorisation R\&D. The most
-plausible approaches seem therefore to lead in the direction of
-specialized libraries (like those described above in the context of
-speeding up Geant4 simulation on the CPUs) that would perform the
-complete simulation of specific sub-detectors (for instance EM
-calorimeter, for specific incoming particles or Cherenkov-based
-detectors for the optical processes) on the device. Such libraries could
-be the right compromise between the complexity of the algorithm that
-needs to be ported to the device and the overall time that is now spent
-(and could be gained) on the corresponding of the full simulation on the
-CPU. The investigation in those directions are starting now, but will
-certainly require a considerable effort to be further invested. The
-first steps that are currently being undertaken consist of implementing,
-on GPUs, prototypes based on VecGeom geometry and navigation
-capabilities. If those prototypes turn out to be successful, the next
-challenge will consist of adding tracking and eventually a limited
-subset of physics models. While Geant4 is a toolkit capable of
-addressing different modeling and simulation problems and contains many
-features and capabilities allowing for user access to detailed
-information for a large variety of use cases and scientific communities,
-this GPU development might turn into specialized transport modules,
-stripped of some features which would be expensive to implement or
-support efficiently on GPUs. Another interesting avenue consists of
+very successful in restricted domains like medical physics simulation,
+neutron transport {[}SHIFT01{]} or optical photon transport
+{[}CHEP-Juno{]}. In those applications, the type of particles considered
+is very limited (often just one type, with no secondary particle
+production), the set of physics processes is reduced and the geometry is
+often extremely simple compared to the LHC detectors. A natural
+`extrapolation' of those approaches to a general HEP simulation is very
+difficult to imagine, because the stochasticity of the simulated events
+would immediately lead to divergences as far as different GPU threads
+are concerned, not to mention, simply the feasibility of porting the
+whole simulation to the GPU code for which specific expertise would be
+needed. On the other hand, running only very small pieces of the
+simulation application on GPUs does not seem to be efficient either, as
+gathering data and transferring it from the host to the device and back
+may strongly limit any possible performance gain, similarly to what was
+seen with the GeantV vectorisation R\&D. The most plausible approaches
+seem therefore to lead in the direction of specialized libraries (like
+those described above in the context of speeding up Geant4 simulation on
+the CPUs) that would perform the complete simulation of specific
+sub-detectors (for instance EM calorimeter, for specific incoming
+particles or Cherenkov-based detectors for the optical processes) on the
+device. Such libraries could be the right compromise between the
+complexity of the algorithm that needs to be ported to the device and
+the overall time that is now spent (and could be gained) on the
+corresponding of the full simulation on the CPU. The investigation in
+those directions are starting now, but will certainly require a
+considerable effort to be further invested before being able to judge
+their real impact. The first steps that are currently being undertaken
+consist of implementing, on GPUs, prototypes based on VecGeom geometry
+and navigation capabilities. If those prototypes turn out to be
+successful, the next challenge will consist of adding tracking and
+eventually a limited subset of physics models. While Geant4 is a toolkit
+capable of addressing different modeling and simulation problems and
+contains many features and capabilities allowing for user access to
+detailed information for a large variety of use cases and scientific
+communities, this GPU development might turn into specialized transport
+modules, stripped of some features which would be expensive to implement
+or support efficiently on GPUs. Another interesting avenue consists of
 exploring the application of some vendor libraries (like Nvidia Optix).
@@
 
-To conclude, it is important to realise that without investing a serious
-effort in the simulation R\&D, a considerable risk of this being a major
-limiting factor as far as new physics discoveries are concerned. We need
-the concerted effort of different experts in physics, Machine Learning
-and GPUs. There are too many unknowns to focus on only one direction, so
-a lot of prototyping is needed. We need to further develop Geant4 while
-working on fast simulation algorithms and doing extensive R\&D on the
-compute accelerators. While the R\&D activities are taking place, one
-should not forget that one still needs to maintain the existing code and
-make sure that sufficient funding and staffing is provided for
-maintenance and development of the physics algorithms, as well as for
-adapting the code to the ongoing hardware, operating system and compiler
-evolution.
+To conclude, it is important to realise that there is a considerable
+risk for the simulation to become a major limiting factor as far as new
+physics discoveries are concerned if no serious effort is invested in
+R\&D. We need the concerted effort of different experts in physics,
+Machine Learning and GPUs. There are too many unknowns to focus on only
+one direction, so a lot of different prototyping is needed. We need to
+further develop Geant4 while working on fast simulation algorithms and
+doing extensive R\&D on leveraging compute accelerators. While the R\&D
+activities are taking place, one can not forget that it is still
+necessary to support the existing code and make sure that sufficient
+funding and staffing is provided for maintenance and development of the
+physics algorithms, as well as for adapting the code to the ongoing
+hardware, operating system and compiler evolution.
 
@@
 R\&D investigations and independent prototyping to explore novel ideas,
-but also a continuous integration activity of new solutions into the
-Geant4 code base and experiments applications to profit from each
-improvement as quickly as possible. An effort needs to be invested into
-making it possible to evolve it to meet the HL-LHC needs without
-compromising the production quality of the Geant4 code used by the
-experiments. We need to ensure that new developments resulting from the
-R\&D programmes can be tested with realistic prototypes and, if
-successful, then integrated, validated, and deployed in a timely fashion
-in Geant4 and adopted by the experiments. The strategy adopted in the
-successful integration in Geant4 of VecGeom libraries developed in the
-context of the GeantV R\&D can provide a working example of how to
-proceed to provide some incremental improvements to the existing
-libraries.
-
-\hypertarget{references-1}{%
-\subsection{References}\label{references-1}}
-
-{[}CMS01{]} V. Ivanchenko and S. Banerjee on behalf of the CMS
-Collaboration, EPJ Web of Conferences 214, 02012 (2019), The PDF is
-here:
-\url{https://www.epj-conferences.org/articles/epjconf/pdf/2019/19/epjconf_chep2018_02012.pdf}
-
-{[}GPU-``MPEXS-DNA{]} S. Okada et al. ``MPEXS-DNA: a new GPU-based Monte
-Carlo simulator for track structures and radiation chemistry at
-subcellular scale,'' Med Phys. vol 46 no. 3, pp. 1483-1500 (2019)
-
-{[}Hybrid-Gate{]} J. Bert et al., "Hybrid GATE: A GPU/CPU implementation
-for imaging and therapy applications," 2012 IEEE Nuclear Science
-Symposium and Medical Imaging Conference Record (NSS/MIC), Anaheim, CA,
-pp. 2247-2250 (2012)
-
-{[}G4CU{]}
-\url{https://kds.kek.jp/indico/event/15926/session/30/contribution/110/material/slides/0.pdf}
-
-{[}Opticks{]} S. Blyth, ``Opticks : GPU Optical Photon Simulation for
-Particle Physics using NVIDIA® OptiXTM,'' EPJ Web Conf., 214, pp. 02027
-(2019) 02027
-
-{[}LHCb01{]} CHEP for Gaussino and Gauss of Gaussino MT
-
-{[}CMS02{]} Use of MT in CMSSW
-
-{[}ML001{]} Adversarial Networks (GAN), Variational Auto-Encoders (VAE)
-or Autoregressive Generative Networks on the `images' of particle
-showers
-
-{[}LHCb02{]} ReDecay paper
+but also a continuous integration activity of new runtime performance
+improvements and solutions into the Geant4 code base and experiments
+applications to profit from each improvement as quickly as possible. An
+effort needs to be invested into making it possible to evolve it to meet
+the HL-LHC needs without compromising the production quality of the
+Geant4 code used by the experiments. We need to ensure that new
+developments resulting from the R\&D programs can be tested with
+realistic prototypes and, if successful, then integrated, validated, and
+deployed in a timely fashion in Geant4 and adopted by the experiments.
+The strategy adopted in the successful integration in Geant4 of VecGeom
+libraries developed in the context of the GeantV R\&D can provide a
+working example of how to proceed to provide some incremental
+improvements to existing applications.
+
+No single solution appears at the moment to provide by itself the
+processing gains required by HL-LHC, nevertheless if a novel approach
+emerges that could do so it should be pursued and carefully evaluated in
+terms of gains against the disruption of a complete reimplementation of
+how events are simulated.
 
@@
 Software trigger and event reconstruction techniques in HEP face a
-number of new challenges in the next decade. Advancements in facilities
-and future experiments bring a dramatic increase in physics reach, as
-well as increased event complexity and rates.
-
-At the HL-LHC, the central challenge for high-level triggers (eg,
+number of new challenges in the next decade. Advances in facilities and
+future experiments bring a dramatic increase in physics reach, as well
+as increased event complexity and rates.
+
+At the HL-LHC, the central challenge for high-level triggers (e.g.,
 software triggers) and object reconstruction is to maintain excellent
@@
 channel density, high precision timing and improved detector layouts are
-essential to overcome these problems.
-
-The subsequent increase of event complexity at the HL-LHC also requires
-the development of software algorithms that can process events with a
-similar cost per event to Run-2 and Run-3. At the same time, algorithmic
-approaches need to effectively take advantage of evolutions in computing
-technologies, including increased SIMD capabilities, increased core
-counts in CPUs, and heterogeneous hardware.
+essential to overcome these problems. The subsequent increase of event
+complexity at the HL-LHC also requires the development of software
+algorithms that can process events with a similar cost per event to
+Run-2 and Run-3. At the same time, algorithmic approaches need to
+effectively take advantage of evolutions in computing technologies,
+including increased SIMD capabilities, increased core counts in CPUs,
+and heterogeneous hardware.
 
@@
 mentions of open source software efforts that have developed within or
-outside the LHC collaborations that could be adapted by the LHC
-community to improve trigger and reconstruction algorithms.
-
-\textbf{The evolution of triggers and real-time analysis}
+outside LHC collaborations that could be adapted by the LHC community to
+improve trigger and reconstruction algorithms. Nearly all software
+solutions presented in this chapter are tailored to a specific
+experiment, and would require additional work to be adapted to different
+environments. Nevertheless, it is still useful to maintain open
+communication channels to discuss design choices and compare performance
+assessments of different solutions to guide future software and
+reconstruction design choices.
+
+\hypertarget{the-evolution-of-triggers-and-real-time-analysis}{%
+\subsection{\texorpdfstring{The evolution of triggers and real-time
+analysis
+}{The evolution of triggers and real-time analysis }}\label{the-evolution-of-triggers-and-real-time-analysis}}
 
@@
 their experiment, and their ability to stream a larger rate of events
-for further processing.
-
-The event rates that will be processed by experiments at the HL-LHC will
-increase by up to a factor 10 with respect to Run-3, owing to upgraded
-trigger systems.
-
-ATLAS and CMS plan to maintain a two-tiered trigger system, where a
-hardware trigger makes use of coarse event information to reduce the
-event rate to 10x over the current capability, up to 1 MHz {[}cite
-TDR{]}. The high level trigger system, implemented in software, selects
-up to 100 kHz of events to be saved in full for subsequent
-analysis\footnote{LHCb {[}cite TDR{]} and ALICE {[}cite TDR{]} will both
-  stream the full collision rate to real-time or quasi-realtime software
-  trigger systems.}.
+for further processing. The event rates that will be processed by
+experiments at the HL-LHC will increase by up to a factor 10 with
+respect to Run-3, owing to upgraded trigger systems. ATLAS and CMS plan
+to maintain a two-tiered trigger system, where a hardware trigger makes
+use of coarse event information to reduce the event rate to 10x over the
+current capability, up to 1 MHz {[}cite TDR{]}. The high level trigger
+system, implemented in software, selects up to 100 kHz of events to be
+saved in full for subsequent analysis\footnote{LHCb {[}cite TDR{]} and
+  ALICE {[}cite TDR{]} will both stream the full collision rate to
+  real-time or quasi-realtime software trigger systems.}.
 
@@
 system, driven by use cases where the physics potential improves when
-analyzing more events than what can be written out in full with
+analysing more events than what can be written out in full with
 traditional data processing.
@@
 of the data taking rate as possible from selected regions and
-subdetectors, even up to the 40 MHz bunch crossing rate (cite CHEP
+sub-detectors, even up to the 40 MHz bunch crossing rate (cite CHEP
 proceedings CMS).
 
-\textbf{Challenges and improvements foreseen in event reconstruction}
+\hypertarget{challenges-and-improvements-foreseen-in-event-reconstruction}{%
+\subsection{Challenges and improvements foreseen in event
+reconstruction}\label{challenges-and-improvements-foreseen-in-event-reconstruction}}
 
@@
 
-\emph{\textbf{Tracking in high pile-up environments}}
+\hypertarget{tracking-in-high-pile-up-environments}{%
+\subsubsection{Tracking in high pile-up
+environments}\label{tracking-in-high-pile-up-environments}}
 
@@
 combinatorial load for tracking algorithms at future colliders will put
-great strain on CPU resources. To minimize the memory footprint,
+great strain on compute resources. To minimize the memory footprint,
 tracking software needs to be thread-safe to support multi-threaded
@@
 the tracking experience already obtained at the LHC, and targets the
-HL-LHC as well as future hadron colliders, providing support for other
-HEP experiments, such as FASER, Belle-II, EIC, CEPC, etc. ACTS provides
-a set of track reconstruction tools designed for parallel architectures,
-with a particular emphasis on thread-safety and concurrent event
-reconstruction.
+HL-LHC as well as future hadron colliders. ACTS provides a set of track
+reconstruction tools designed for parallel architectures, with a
+particular emphasis on thread-safety and concurrent event
+reconstruction. ACTS has active collaborators from other HEP
+experiments, such as FASER, and provides support to Belle-II, EIC, CEPC,
+etc.
 
@@
 
-\emph{\textbf{\hfill\break
-Addition of timing information in reconstruction}}
+\hypertarget{addition-of-timing-information-in-reconstruction}{%
+\subsubsection{Addition of timing information in
+reconstruction}\label{addition-of-timing-information-in-reconstruction}}
 
@@
 or FCC-hh, may also benefit from adding timing information to the
-reconstruction. This permits to mitigate effects of pile-up by
+reconstruction. This allows the mitigation of the effects of pile-up by
 exploiting the time-separation of collision products. {[}cite HGTD/CMS
@@
 {]}\\
-The experimental communities have been working on timing detector
+The experiment communities have been working on timing detector
 reconstruction and object identification techniques in complex
@@
 that include timing information have been developed and incorporated
-into experimental software stacks {[}cite ATLAS/CMS CDRs?, CTD
+into experimental software stacks {[}cite ATLAS/CMS CDRs, CTD
 contribution
@@
 Steps Towards Four-Dimensional Tracking: Timing Layers at the
-HL-LHC}}{]}. \emph{Missing: common software, wrote to Ivan Kisel (CBM)
-about how portable this is
-\href{https://inspirehep.net/literature/1414007}{{4-Dimensional Event
-Building in the First-Level Event Selection of the CBM Experiment -
-INSPIRE}}. Also public reference from LHCb on how preselection of PV can
-improve the reconstruction time for one event.}
-
-\textbf{Enhanced data quality and software monitoring for trigger and
-reconstruction}.
+HL-LHC}}{]}\footnote{Tracking algorithms where events overlap in time
+  within time slices are employed by the CBM experiment {[}ref
+  \href{https://indico.gsi.de/event/6198/contributions/28511/attachments/20671/26101/Akishina_thesis_corrected.pdf}{{https://indico.gsi.de/event/6198/contributions/28511/attachments/20671/26101/Akishina\_thesis\_corrected.pdf}},
+  \href{https://inspirehep.net/literature/1414007}{\emph{{4-Dimensional
+  Event Building in the First-Level Event Selection of the CBM
+  Experiment - INSPIRE}}}{]}. The use of a cellular automaton makes the
+  algorithm less dependent on the specific detector geometry, and there
+  may be the possibility of cross-talk with LHC experiments that are
+  investigating 4D reconstruction.}.
+
+\hypertarget{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}{%
+\subsection{Enhanced data quality and software monitoring for trigger
+and
+reconstruction}\label{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}}
 
@@
 procedures to check trigger rates as well as the performance of the low-
-and high-level physics objects in the data.
-
-While we are not aware of updates in this area since the CWP, we would
-like to stress the importance of investment in this area as the
-monitoring of the quality of the recorded/reconstructed data is
-paramount for further physics analysis.
-
-\textbf{Hardware and software improvements towards reconstruction
-software}
-
-\emph{\hfill\break
-\textbf{General reconstruction software improvements: vectorization}}
+and high-level physics objects in the data {[}ref ATLAS DQ and CMS
+DQ{]}.
+
+Since the CWP, experiments have started making limited use of machine
+learning algorithms for anomaly detection {[}ref LHCb RoboShifter:
+\href{https://cds.cern.ch/record/2298467/files/pdf.pdf}{{https://cds.cern.ch/record/2298467/files/pdf.pdf}},
+ref CMS machine learning muons paper - finding out whether it is in
+production?{]}.
+
+\hypertarget{general-reconstruction-software-improvements-vectorization}{%
+\subsubsection{General reconstruction software improvements:
+vectorization}\label{general-reconstruction-software-improvements-vectorization}}
 
@@
 
-\emph{\textbf{Use of machine learning for software trigger and
-reconstruction algorithms}}
+\hypertarget{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}{%
+\subsubsection{Use of machine learning for software trigger and
+reconstruction
+algorithms}\label{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}}
 
@@
 
-\emph{\textbf{Algorithms and data structures to efficiently exploit}}
-\emph{\textbf{many-core and non-x86 architectures (e.g. GPUs, FPGAs)}:}
+\hypertarget{algorithms-and-data-structures-to-efficiently-exploit-many-core-and-non-x86-architectures-e.g.-gpus-fpgas}{%
+\subsubsection{\texorpdfstring{Algorithms and data structures to
+efficiently exploit many-core and non-x86 architectures (e.g. GPUs,
+FPGAs):
+}{Algorithms and data structures to efficiently exploit many-core and non-x86 architectures (e.g.
+GPUs, FPGAs): }}\label{algorithms-and-data-structures-to-efficiently-exploit-many-core-and-non-x86-architectures-e.g.-gpus-fpgas}}
 
@@
 designed for, or have successfully adapted to, hardware accelerators,
-most notably to NVidia GPUs using CUDA. Two projects providing open
-source software implementations are Patatrack (from CMS) {[}3{]} and
+most notably to NVidia GPUs using CUDA. Two projects targeting Run-3 and
+with prospects for use in Run-4 are Patatrack (from CMS) {[}3{]} and
 Allen (from LHCb) {[}4{]}.
@@
 
-The experience built with these projects shows that GPUs can be used for
-increasing the throughput in cases of large-size (ALICE) {[}cite Alice
-O2{]} and small-size events per second (LHCb), and to speed up specific
-parts of the data processing (CMS, Patatrack).
+ATLAS is also investigating GPU solutions for reconstruction software
+and its acceleration at HL-LHC, in particular in terms of highly
+parallel algorithms {[}3a{]} and to enable the use of machine learning
+algorithms in object reconstruction and identification {[}\emph{waiting
+for an email from ATLAS with references}{]}.
+
+The experience acquired with these projects shows that GPUs can be used
+for increasing the throughput in cases of large-size (ALICE) {[}cite
+Alice O2{]} and small-size events per second (LHCb), and to speed up
+specific parts of the data processing (CMS, Patatrack).
 
@@
 
-\textbf{References (not properly formatted, this will go on Overleaf):}
-
-\emph{{[}1{]}
-\href{https://arxiv.org/abs/1802.08640}{{https://arxiv.org/abs/1802.08640}}}
-
-\emph{{[}2{]}
-\href{https://cds.cern.ch/record/2693670}{{https://cds.cern.ch/record/2693670}}}
-
-\emph{{[}3{]}} - Andrea Bocci \emph{at al.}, ``\emph{Heterogeneous
-online reconstruction at CMS}'', presented at CHEP 2019\\
-- Andrea Bocci \emph{at al.}, ``\emph{Heterogeneous reconstruction:
-combining ARM processors with GPUs}'', presented at CHEP 2019
-
-\emph{{[}4{]}}
-\href{https://arxiv.org/abs/1912.09161}{{https://arxiv.org/abs/1912.09161}}
-
-\emph{{[}5{]}}
-\href{http://dx.doi.org/10.1088/1742-6596/898/4/042011}{{http://dx.doi.org/10.1088/1742-6596/898/4/042011}},
-\href{https://arxiv.org/abs/1910.03128}{{https://arxiv.org/abs/1910.03128}},
-The ACTS project: track reconstruction software for HL-LHC and beyond,
-Proceedings of CHEP2019, in progress
-
-{[}6{]} Xiangyang Ju et al, "Graph Neural Networks for Particle
-Reconstruction in High Energy Physics Detectors"
-\href{https://arxiv.org/abs/2003.11603}{{(arxiv link}})
-(\href{https://github.com/exatrkx/exatrkx-neurips19}{{Software and
-documentation}})
-
-{[}7{]} J. Duarte \emph{et al}, ``Fast inference of deep neural networks
-in FPGAs for particle physics'', \emph{JINST} \textbf{13} P07027, 2018,
-doi:
-\href{https://iopscience.iop.org/article/10.1088/1748-0221/13/07/P07027/meta}{{10.1088/1748-0221/13/07/P07027}}\\
-S. Summers \emph{et al}, ``Fast inference of Boosted Decision Trees in
-FPGAs for particle physics'', 2020, {[}Online{]} available:
-\href{https://arxiv.org/abs/2002.02534}{{arXiv:2002.02534}}\\
-G. Gi Guglielmo \emph{et al}, ``Compressing deep neural networks on
-FPGAs to binary and ternary precision with HLS4ML'', 2020, {[}Online{]}
-available: \href{https://arxiv.org/abs/2003.06308}{{arXiv:2003.06308}}
-
-{[}8{]} SONIC:
-\href{https://arxiv.org/abs/1904.08986}{{https://arxiv.org/abs/1904.08986}}
-
 \hypertarget{data-analysis}{%
@@
 \hypertarget{key-analysis-computing-challenges-at-the-hl-lhc}{%
-\subsubsection{Key analysis computing challenges at the
+\subsection{Key analysis computing challenges at the
 HL-LHC}\label{key-analysis-computing-challenges-at-the-hl-lhc}}
@@
 to as ``Analysis Object Data'' (AOD) containing reconstructed physics
-object data and producing physics results as the end product, poses
-unique computing challenges. The scope of analysis processing is broad,
-encompassing the production of derived data formats as well as end stage
-analysis actions such as producing histograms from those intermediate
-datasets and visualising the final results. In the following, attention
-is focused on the computing challenges related to analysis for the ATLAS
-and CMS experiments.
+object data, and producing physics results as the end product, poses
+unique computing challenges. The scope of analysis data processing is
+broad, encompassing the production of derived data formats as well as
+end stage analysis actions such as producing histograms from those
+intermediate datasets and visualising the final results. In the
+following, attention is focused on the computing challenges related to
+analysis for the ATLAS and CMS experiments.
 
@@
 recorded event rate by an order of magnitude, compared with a projected
-15\% annual growth of storage resources. The large pile-up in HL-LHC
+5-10\% annual growth of storage resources. The large pile-up in HL-LHC
 collisions will compound the storage challenge by inflating the data
@@
 infrastructure load and downtime. Another challenge is the
-multiplication of resource demands from the assessment of systematic
-uncertainties, which involves processing many alternate configurations
-and input data, inflating both the CPU load and storage footprint.
-Improvement of community standards for metadata handling and uncertainty
-handling will be important for HL-LHC.
+multiplication of computing resource demands from the assessment of
+systematic uncertainties, which involves processing many alternate
+configurations and input data, inflating both the CPU load and storage
+footprint. Improvement of community standards for metadata handling and
+uncertainty handling will be very important for HL-LHC.
 
@@
 \hypertarget{analysis-processing-and-workflows-current-problems-and-solutions}{%
-\subsubsection{Analysis processing and workflows: current problems and
+\subsection{Analysis processing and workflows: current problems and
 solutions}\label{analysis-processing-and-workflows-current-problems-and-solutions}}
@@
 
-Two approaches have been followed during LHC Run2 by ATLAS and CMS:
+Two approaches have been followed during LHC Run 2 by ATLAS and CMS:
 ATLAS analysis trains {[}4{]} and the CMS mini-AOD, a highly reduced and
@@
 comparison, a reduced format (10 kB/event) made from the AOD would
-result in 0.5 PB/year for data and 2 PB/year for MC. Experience on CMS
-shows that further reduction (down to \textasciitilde2 kB/event for the
-nano-AOD) could be possible while still supporting a majority of physics
-analyses {[}5{]}. File size reductions also permit multiple copies to be
-held on disk for more efficient parallel processing. It is clear that as
-many analyses as possible need to be fed into this standardized analysis
-event-content pipeline if analysis computing resources are to be kept
-under control. Nevertheless, analysis trains or any similar means of
-synchronising data access could be applied to this highly reduced
-format.
+result in 0.5 PB/year for data and 2 PB/year for MC. Production of this
+reduced format using a Data Carousel model {[}1{]} allows the much
+larger AOD to reside on tape, providing significant savings in disk
+storage. Experience on CMS shows that further reduction (down to
+\textasciitilde2 kB/event for the nano-AOD) could be possible while
+still supporting a majority of physics analyses {[}5{]}. File size
+reductions also permit multiple copies to be held on disk for more
+efficient parallel processing. It is clear that as many analyses as
+possible need to be fed into this standardized analysis event-content
+pipeline if analysis computing resources are to be kept under control.
+Nevertheless, analysis trains or any similar means of synchronising data
+access could be applied to this highly reduced format. It is worth
+noting the LHCb analysis model already in Run 3 has to forego the luxury
+of reprocessing Raw data. This constraint from sheer event rate is not
+without merit as it greatly simplifies the data processing and analysis
+models and removes one tier of offline data products.
 
@@
 twiki pages or even in e-mail. The lack of proper metadata integration
-and access interfaces leaves analyses exposed to any problems in data
+and a coherent interface leaves analyses exposed to any problems in data
 processing. This in turn results in data completeness demands that
@@
 
-At HL-LHC, as energy and luminosity conditions are quasi stable across
-the years, datasets from multiple years with different conditions will
-need to be analyzed in a coherent way. The analysis software will then
-need to be able to fetch and use the proper metadata automatically while
-today's analysis software often requires dedicated configuration and
-tuning for each data-taking period. Belle II have taken some steps to
-address this by using the ``global tag'' concept commonly used in
+At the HL-LHC, as energy and luminosity conditions are quasi stable
+across the years, datasets from multiple years with different conditions
+will need to be analyzed in a coherent way. The analysis software will
+then need to be able to fetch and use the proper metadata automatically
+while today's analysis software often requires dedicated configuration
+and tuning for each data-taking period. Belle II have taken some steps
+to address this by using the ``global tag'' concept commonly used in
 reconstruction and applying it to analysis, allowing users to better
 organise and store their analysis metadata. Nevertheless, until all
-metadata is organised under the same global tag umbrella, which is
-extremely challenging, the problem remains. Improvements here are
-necessary both to permit efficient studies on partial datasets, and to
-reduce the risk of user error in metadata access.
+metadata is organised under the same global tag umbrella and accessed
+through a coherent interface, which is extremely challenging, the
+problem remains. Improvements here are necessary both to permit
+efficient studies on partial datasets, and to reduce the risk of user
+error in metadata access.
 
@@
 possibility to re-run the code on the original dataset in a few years.
-This may allow a future analyst to reproduce previous results, but will
+This may allow a future analyser to reproduce previous results, but will
 likely not provide any understanding, for the future analyser, about
@@
 calibrations or additional data that will improve the precision of the
-measurement. This brings additional difficulties, not least related to
-metadata.
+measurement. This brings significant additional challenges, not least
+related to metadata.
 
@@
 later, a model that would have physicists write logical descriptions of
-their analyses, leaving low-level code generation to a backend. Thus
-analysis design could be quickly understood and shared, sidestepping the
-usual dearth of documentation, while simultaneously isolating analysis
-design from implementation and hardware, providing a natural means of
-analysis preservation.
+their analyses, leaving low-level code generation and hardware
+optimisation to a backend. Thus analysis design could be quickly
+understood and shared, sidestepping the usual dearth of documentation,
+while simultaneously isolating analysis design from implementation and
+hardware, providing a natural means of analysis preservation. An
+interesting development in this direction is the Reana platform {[}21{]}
+that supports multiple backends and uses a DSL to provide a framework to
+run an analysis. Already a valuable resource as a tool for analysing
+open data, it is interesting to ask what would happen if every analyst
+first had to structure their analysis and capture their workflows before
+they started submitting jobs.
 
 \hypertarget{plans-and-recent-progress}{%
-\subsubsection{Plans and recent
-progress}\label{plans-and-recent-progress}}
+\subsection{Plans and recent progress}\label{plans-and-recent-progress}}
 
@@
 of those tools have been tested to understand the feasibility of usage
-in the HEP context {[}references here to spark tests, google cloud,
-pyhep stuff, etc\ldots{]}.
+in the HEP context.
 
@@
 
-In addition to adopting externally available tools, HEP developers are
-prototyping similar concepts in more domain-specific implementations. A
-core requirement of the HEP analysis toolkit is the representation and
-manipulation of physics data where each event contains scalars (i.e.
-single event variables), objects (i.e. variables related to the same
-physics object) and vectors of objects (i.e. a variable number of
-physics objects). In addition the Lorentz vector algebra that is bread
-and butter for HEP analysis needs to be naturally accommodated.
-
 The ROOT package, i.e. the foundation and workhorse for LHC analysis,
@@
 in C++), which natively exploits multi-core architectures {[}7{]} and
-could be extended to accelerators. Flexibility in interfacing RDataFrame
-to non-ROOT data formats has allowed the ALICE collaboration to
-prototype its Run 3 analysis model on this new technology. Growing use
-of Machine Learning was also anticipated, and the TMVA toolkit has been
-improved with substantial optimisations, more complex deep learning
+could be extended to accelerators. Used from Python, it allows to
+pre-process data in C++, and export to numpy. Flexibility in interfacing
+RDataFrame to non-ROOT data formats has allowed the ALICE collaboration
+to prototype its Run 3 analysis model on this new technology. Growing
+use of Machine Learning was also anticipated, and the TMVA toolkit has
+been improved with substantial optimisations, more complex deep learning
 models and enhanced interoperability with common data science tools
@@
 extension with other Python tools. Good performance appears to have been
-achieved within the scope of these projects, which has been kept focused
--\/- in the event of uptake by a large portion of the community,
-long-term support for these modules would need to be established.
+achieved within the scope of these projects, which has been kept
+focused, but for the purposes of fair comparison with other existing or
+emerging options, defining clear benchmarks for I/O and processing speed
+would be beneficial.
 
@@
 for software education and have been leveraged as a powerful frontend
-for access to facilities including CERN's Swan.
+for access to facilities including CERN's SWAN {[}20{]}.
 
 \hypertarget{prospects-and-rd-needs}{%
-\subsubsection{Prospects and R\&D needs}\label{prospects-and-rd-needs}}
+\subsection{Prospects and R\&D needs}\label{prospects-and-rd-needs}}
 
 As the complexity of analysis data handling and processing grows,
-developing efficient and robust code becomes harder and harder. In
-addition, analysis code is typically reimplemented by tens of
-individuals, in different experiments, analysis teams or institutes
-mostly performing the same kind of operations (data reduction, plotting,
-variation of systematic uncertainty, fitting etc\ldots). An appealing
-idea, circulating for a long time in the HEP world, is to develop a
-domain specific analysis language to describe any analysis in a concise
-way {[}10{]}. An analysis backend would then translate this into
-optimised code to effectively run the analysis exploiting the full power
-of different kinds of platforms and architectures. At a more basic
-level, declarative rather than imperative programming is visibly
-becoming more prevalent in experimental software configuration, with
-consequent improvement in the transparency of steering code, setting a
-good example for analysis.
-
-For HL-LHC, the growing analysis complexity, rapidly evolving compute
-architectures and the need for analysis code to be highly optimised
-could make the usage of analysis languages a must.
-
-Prototypes of analysis languages have been developed in the past years
-{[}refs, 10?{]} for the event processing and histogram production parts
-of the analysis workflow and probably with some additional effort the
-HEP community can converge on broadly usable tools. Similar efforts can
-also be investigated in the context of data interpretation (e.g.
-fitting) standardizing the description of what to fit and how (an
-interesting example here is the so called ``combine tools'' developed by
-CMS and based on RooFit for the Higgs discovery) {[}ref{]}. There is an
-obvious need for good interfacing with analysis backends, whether these
-be local CPU, batch or grid resources, for which there is some
-experience within the experimental collaborations as well as in data
-science frameworks.
-
-Eight initial benchmark challenges have been defined to set the scope
-that should be addressed by analysis languages {[}11{]}. Several
-prototypes have already been developed, some of which solve non-trivial
-challenges with concise descriptions. What remains to be demonstrated is
-the capability of such systems to scale to full analysis complexity,
-including tasks such as handling of systematic uncertainties,
-data-driven background estimation, which are yet to be incorporated into
-the benchmarks. Their performance in terms of CPU and memory consumption
-also has to be fairly assessed. It is also a point of concern that
-widespread adoption of analysis languages could pigeonhole HEP analysers
-into the role of expert users of domain-specific languages, and
-resulting in stratification between backend developers and users.
-
-A similar, and somewhat related, topic for R\&D is that of hardware
-resources and infrastructures needed for analysis. At some stage in the
-analysis processing interactivity or fast turn-around is needed.
-Quantitatively speaking, while typical grid task completion has a time
-scale of one day to one week, in the context of analysis exploration and
-developments answers are needed on timescales of a few seconds to a few
-hours in order to keep people productive. As analysis tasks are often
-IO-intensive and not necessarily-CPU intensive, the right tradeoff
-between CPU and high bandwidth storage should be carefully studied when
-designing the computing infrastructure of HL-LHC. As a proof of
-principle, an unoptimized Higgs discovery analysis has been rerun on
-Google Cloud in a few minutes as the infrastructure was able to quickly
-provide tens of thousands of CPUs with granted bandwidth to storage of 2
-Gb/coresec {[}12{]}. Scaling this to sustained, diverse analysis loads
-will be far from trivial. R\&D is needed to understand current usage
-patterns and project them to HL-LHC in the last steps of the analysis so
-that the right balance of resources can be procured. An ongoing topic of
-discussion is how to best provision computing resources for end stage
-analysis, where there are many open questions {[}13{]}.
-
-There has already been substantial experimentation with different grid
-site architectures, in particular with regards to I/O, e.g. purely
-remote data access as opposed to local storage. Various job allocation
-and management models, many incorporating industry solutions, have been
-trialled, but the volatility of analysis loads remains a challenge to be
-overcome. More detailed analysis of resource usage and its evolution,
-especially as pertains to the need for heterogeneous compute, is needed
-to correctly design hardware allocation.
-
-\hypertarget{priority-topics}{%
-\subsubsection{Priority topics}\label{priority-topics}}
-
-In summary, the following goals in analysis computing stand out as foci
-for development towards the HL-LHC era:
-
-\begin{itemize}
+developing efficient and robust code using a steadily increasing number
+of tools running on ever more heterogeneous hardware becomes more and
+more difficult. In addition, analysis code is typically reimplemented by
+tens of individuals, in different experiments, analysis teams or
+institutes mostly performing the same kind of operations, i.e. data
+reduction, plotting, variation of systematic uncertainty, fitting etc.
+It is clear that tools and frameworks for performing these repeated
+operations need to provide very efficient interfaces to avoid analysis
+code bloat. Declarative rather than imperative programming is becoming
+visibly more prevalent both in and outside of the field as it naturally
+provides efficient syntax.
+
+DSLs are a generalisation of this declarative concept and have already
+demonstrated their ability to simplify analysis code {[}{]}. Not only do
+they allow users to express operations concisely, improving
+comprehension and reusability, DSLs provide a natural layer of
+insulation against hardware evolution as low-level, optimised code
+generation for multiple different platforms and architectures is
+delegated to the tool and framework experts. Prototypes of DSLs have
+been developed in recent years {[}10{]} for the event processing and
+histogram production parts of the analysis workflow, and even built into
+experimental frameworks to perform high-level analysis {[}19{]}. Similar
+efforts can also be investigated in the context of data interpretation
+and statistical analysis and an interesting example here is the so
+called ``combine tools'' developed by CMS and based on RooFit for the
+Higgs discovery) {[}ref{]}. Descriptions of a full analysis in terms of
+workflows could then tie together the analysis at the top level {[}18,
+21{]}. With some effort, the HEP community may be able to converge on
+effective tools that provide a common solution, ideally across
+experiments.
+
+There is an obvious need for good integration with analysis backends,
+whether these be local CPU, batch or grid resources. At some stage in
+the analysis process, interactivity or fast turn-around is needed. While
+typical grid task completion has a time scale of one day to one week,
+fast turn-around exploration requires answers within a few seconds up to
+perhaps a few hours in order to keep people productive. As analysis
+tasks are often IO-intensive and not necessarily-CPU intensive, the
+right tradeoff between CPU and high bandwidth storage should be
+carefully studied. More detailed studies of resource usage and its
+evolution, especially as pertains to the need for heterogeneous compute,
+is needed to ensure the right resource balance is available in the
+HL-LHC computing infrastructure. This will require more interaction
+between the analysis and infrastructure communities.
+
+As a proof of principle, an unoptimized Higgs discovery analysis has
+been rerun on Google Cloud in a few minutes as the infrastructure was
+able to quickly provide tens of thousands of CPUs with guaranteed
+bandwidth to storage of 2 Gb/coresec {[}12{]}. Scaling this to
+sustained, diverse analysis payloads will be far from trivial, so the
+question becomes how we can compare different approaches to provisioning
+analysis. Eight initial benchmark challenges have been defined to set
+the scope that should be addressed by analysis systems {[}11{]}. What
+remains to be demonstrated is the capability of such systems to scale to
+full analysis complexity, including tasks such as the handling of
+systematic uncertainties and data-driven background estimation, and
+these are yet to be incorporated into the benchmarks.
+
+In contrast to the activity on analysis languages and infrastructure,
+metadata is a topic that is almost entirely neglected by the analysis
+community, even while computing experts have long understood its
+importance. That is likely due to the absence of a single coherent
+stake-holder for and provider of analysis metadata. Event data comes in
+a ROOT file, but already within a single experiment there are numerous
+sources of metadata that operate on vastly different timescales and this
+poses a significant challenge to standardisation. As analyses and
+detectors grow in complexity, further divergence may amplify the
+challenge, making it critical to begin addressing this problem promptly.
+Nevertheless, the drive to capture a complete analysis must drag
+metadata with it, and the gleeful adoption of the declarative paradigm
+will make database-like interfaces seem less scary than they once were
+and may finally allow a more complete integration with analysis code. I
+have a dream.
+
+Finally, attention must be given to ensuring the longevity of any
+projects that the HEP community decides to adopt en masse. Whether by
+specific institutes, funding schemes or other collective programmes,
+responsibility for key software will need to be ensured in the long
+term. The development of a secure support base will therefore need to be
+considered for emerging projects that show significant promise.
+
+\hypertarget{summary}{%
+\section{Summary}\label{summary}}
+
+The main challenges and R\&D lines of interest have been discussed
+above. There are clearly identified priority topics for the different
+areas in the next few years. We note that longer term R\&D, into areas
+such as quantum or neuromorphic computing, should happen at a low level,
+but are almost certainly too far from production to deliver improvements
+for Run 4 of the LHC.
+
+\hypertarget{physics-event-generators-1}{%
+\subsection{Physics Event Generators}\label{physics-event-generators-1}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  \begin{quote}
+  Gain a more detailed understanding of current CPU costs by accounting
+  and profiling.
+  \end{quote}
+\item
+  \begin{quote}
+  Survey generator codes to understand the best way to move to GPUs and
+  vectorized code, and prototype the port of the software to GPUs using
+  data-parallel paradigms.
+  \end{quote}
+\item
+  \begin{quote}
+  Support efforts to optimize phase space sampling and integration
+  algorithms, including the use of Machine Learning techniques such as
+  neural networks.
+  \end{quote}
+\item
+  \begin{quote}
+  Promote research on how to reduce the cost associated with negative
+  weight events, using new theoretical or experimental approaches.
+  \end{quote}
+\item
+  \begin{quote}
+  Promote collaboration, training, funding and career opportunities in
+  the generator area.
+  \end{quote}
+\end{enumerate}
+
+\hypertarget{detector-simulation-1}{%
+\subsection{Detector Simulation}\label{detector-simulation-1}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  \begin{quote}
+  Undertake a detailed performance analysis of current production
+  detector simulation code, to better understand where performance
+  limitations in data and cache locality arise from.
+  \end{quote}
+\item
+  \begin{quote}
+  Improve current codes through refactoring and specialised HEP models
+  to avoid bottlenecks identified,
+  \end{quote}
+\item
+  \begin{quote}
+  Develop generic fast simulation models and techniques, including
+  machine learning based options in addition to optimised parametric
+  approaches.
+  \end{quote}
+\item
+  \begin{quote}
+  Develop GPU codes for tackling very specific, time consuming, parts of
+  the simulation for HEP. Specifically calorimetry, including geometry,
+  field and electro-magnetic physics processes.
+  \end{quote}
+\end{enumerate}
+
+\hypertarget{reconstruction-and-software-triggers-1}{%
+\subsection{Reconstruction and Software
+Triggers}\label{reconstruction-and-software-triggers-1}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  \begin{quote}
+  Support uniformity between online and offline software wherever
+  possible, so the experiments can benefit from consistent event
+  selection between trigger and final analysis.
+  \end{quote}
+\item
+  \begin{quote}
+  Continue encouraging cross-talk between LHC experiments (as well as
+  with other disciplines) on use cases for real-time analysis, as it
+  enables ATLAS and CMS to increase the physics output for selected
+  physics cases without adding significant overhead to the current
+  resources.
+  \end{quote}
+\item
+  \begin{quote}
+  Develop fast and performant tracking solutions (including timing
+  information, if available) that can withstand the increase of pile-up
+  at HL-LHC.
+  \end{quote}
+\item
+  \begin{quote}
+  Vectorisation (is this bundled with vectorisation?)
+  \end{quote}
+\item
+  \begin{quote}
+  Further promote activities in the area of data quality and software
+  monitoring, as the quality of the recorded/reconstructed data is
+  paramount for all physics analysis.
+  \end{quote}
+\item
+  \begin{quote}
+  Support efforts towards the implementation of machine learning models
+  in experimental framework (C++) to enable more widespread use in
+  trigger and reconstruction software.
+  \end{quote}
+\item
+  \begin{quote}
+  Develop solutions that can exploit accelerators and HPCs for trigger
+  and reconstruction, with a focus to mitigating their dependencies on
+  architecture-specific codes.
+  \end{quote}
+\item
+  \begin{quote}
+  Encourage the use of open-source elements of the LHC software stack
+  for trigger and reconstruction by smaller experiments, especially in
+  case of common experimental design\footnote{An example of software
+    sharing happens in the case of the FASER experiment, whose offline
+    software
+    (\href{https://gitlab.cern.ch/faser/calypso}{{https://gitlab.cern.ch/faser/calypso}})
+    is based on a derivative of the ATLAS Athena open-source software
+    {[}ref{]}, which in turn uses the Gaudi framework {[}ref{]}. FASER
+    also shares tracking detector components with ATLAS, facilitating
+    the adoption of software relying on similar detector descriptions.}.
+  This benefits experiments with a smaller developer base than LHC
+  experiments, and increases the return on investment for LHC experiment
+  software. For this reason, we also encourage the addition of common
+  open software projects to the ESCAPE software catalogue {[}ref
+  ESCAPE{]}.
+  \end{quote}
+\end{enumerate}
+
+\hypertarget{data-analysis-1}{%
+\subsection{Data Analysis}\label{data-analysis-1}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \item
@@
   the majority of analyses with a footprint of O(1kb/event), retaining
-  the flexibility to accommodate the needs of specialised analyses;
+  the flexibility to accommodate the needs of specialised analyses.
   \end{quote}
@@
   Streamline analysis metadata access and calibration schemes, and
-  provide effective book-keeping for fractional datasets;
+  provide effective book-keeping for fractional datasets.
   \end{quote}
@@
   and/or languages for analysis description, interfaced with distributed
-  computing backends; and
+  computing backends.
   \end{quote}
@@
   \end{quote}
-\end{itemize}
-
-\hypertarget{references-2}{%
-\subsubsection{References}\label{references-2}}
-
-{[}1{]} *ATLAS Collaboration, \emph{Evolution of the ATLAS analysis
-model for Run-3 and prospects for HL-LHC}, ATL-SOFT-PROC-2020-002,
-\href{https://cds.cern.ch/record/2708664/}{{https://cds.cern.ch/record/2708664/}}
-
-{[}2{]} CMS Collaboration, HL-LHC resource projections,
-\href{https://twiki.cern.ch/twiki/bin/view/CMSPublic/CMSOfflineComputingResults}{{https://twiki.cern.ch/twiki/bin/view/CMSPublic/CMSOfflineComputingResults}}
-
-{[}3{]} *ALICE Collaboration, \emph{The ALICE analysis train system},
-J.Phys.Conf.Ser. 608 (2015) no.1, 012019,
-\href{http://arxiv.org/abs/arXiv:1502.06381}{{http://arxiv.org/abs/arXiv:1502.06381}}
-
-{[}4{]} *ATLAS Collaboration, \emph{A new petabyte-scale data derivation
-framework for ATLAS. Journal of Physics: Conference Series}, 664.
-072007.
-
-{[}5{]} *CMS Collaboration, \emph{A further reduction in CMS event data
-for analysis: the NANOAOD format}, EPJ Web of Conferences. 214. 06021.
-
-{[}6{]}
-\href{https://github.com/hsf-training/PyHEP-resources}{{https://github.com/hsf-training/PyHEP-resources}}
-
-{[}7{]} D Piparo et al, \emph{RDataFrame: Easy Parallel ROOT Analysis at
-100 Threads}, EPJ Web of Conferences 214, 06029 (2019),
-\href{https://www.epj-conferences.org/articles/epjconf/abs/2019/19/epjconf_chep2018_06029/epjconf_chep2018_06029.html}{{https://www.epj-conferences.org/articles/epjconf/abs/2019/19/epjconf\_chep2018\_06029/epjconf\_chep2018\_06029.html}}
-
-{[}8{]}
-\href{https://github.com/scikit-hep/uproot}{{https://github.com/scikit-hep/uproot}},
-\href{https://github.com/scikit-hep/awkward-array}{{https://github.com/scikit-hep/awkward-array}},
-\href{https://github.com/CoffeaTeam}{{https://github.com/CoffeaTeam}},
-\href{https://github.com/diana-hep/pyhf}{{https://github.com/diana-hep/pyhf}}
-
-{[}9{]} J Blomer et al, \emph{Evolution of the ROOT Tree I/O},
-\href{https://arxiv.org/abs/2003.07669}{{https://arxiv.org/abs/2003.07669}}
-
-{[}10{]} ADL Workshop at Fermilab,
-\href{https://indico.cern.ch/event/769263/timetable/?view=standard}{{https://indico.cern.ch/event/769263/}},
-Gordon's CHEP talk (good overview of links, is there something more
-official?)
-\href{https://indico.cern.ch/event/773049/contributions/3476174/attachments/1938123/3220511/Declarative.pdf}{{https://indico.cern.ch/event/773049/contributions/3476174/attachments/1938123/3220511/Declarative.pdf}}
-
-{[}11{]}
-\href{https://github.com/iris-hep/adl-benchmarks-index}{{https://github.com/iris-hep/adl-benchmarks-index}}
-
-{[}12{]} L Heinrich \& R Rocha, \emph{Reperforming a Nobel Prize
-discovery on Kubernetes}, talk at CHEP 2019,
-\href{https://indico.cern.ch/event/773049/contributions/3581373}{{https://indico.cern.ch/event/773049/contributions/3581373}}
-(originally at KubeCon)
-
-{[}13{]} *ALICE Collaboration, \emph{The ALICE analysis framework for
-LHC Run 3}, EPJ Web Conf. 214 (2019) 05045
-\href{http://cds.cern.ch/record/2699856}{{http://cds.cern.ch/record/2699856}}
-
-{[}14{]} \href{http://paperpile.com/b/2mqsuS/IDej}{L Moneta et al,
-\emph{Machine Learning with ROOT/TMVA}, EPJ Web of Conferences;
-Proceedings of CHEP 2019 (in print). 2020.}
-
-{[}15{]} \href{http://paperpile.com/b/2mqsuS/6uIv}{K Albertsson et al,
-\emph{Fast Inference for Machine Learning in ROOT/TMVA}, EPJ Web of
-Conferences; Proceedings of CHEP 2019 (in print). 2020.}
-
-{[}16{]} J Blomer, \emph{A quantitative review of data formats for HEP
-analyses}, Journal of Physics: Conference Series. 2018. p. 032020.
-
-{[}17{]} HEP Software Foundation Collaboration, \emph{HEP Software
-Foundation Community White Paper Working Group - Data Analysis and
-Interpretation},
-\href{https://arxiv.org/abs/1804.03983}{{arXiv:1804.03983}}
-
-Proceedings marked `*' will be credited to the shorter author lists
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Final wrap-up of the most important messages and make some final points
-that expand the scope. For now collect points that we may want to
-mention here:
-
-\begin{itemize}
-\item
-  \begin{quote}
-  May want to mention quantum computing and other disruptive technology
-  for later on in the HL-LHC programme
-  \end{quote}
-
-  \begin{itemize}
-  \item
-    \begin{quote}
-    ``In the more distant future devices such as quantum or neuromorphic
-    devices may provide a substantial processing power.'' (From
-    Federico)
-    \end{quote}
-  \end{itemize}
-\item
-  \begin{quote}
-  Considerations to be made about attracting scientists to sw
-  development or sw teaching
-  \end{quote}
-\end{itemize}
-
-In total selfish self-aggrandizing (but not only), I propose to mention
-more extensively (and not necessarily in the conclusion).
-
-\begin{itemize}
-\item
-  \begin{quote}
-  R\&D on Quantum Computing. At the moment any statement on QC is
-  unsubstantiated, both those that foresee a bright tomorrow and those
-  that predict it will never work. However my personal opinion is that
-  some limited R\&D in this field is necessary to inform the Physics
-  community of its potential and its limitations. Should it really
-  become a viable production technology, it will be essential to have
-  few people in the community that will guide the colleagues on how to
-  make the best use of it. Had we had more of these people for GPUs, we
-  would be in a better position today.
-  \end{quote}
-\item
-  \begin{quote}
-  Relations with IT industry are important. The CERN openlab model was
-  cited even by Fabiola in her conclusions in Grenada. We should mention
-  that this kind of collaborations is fruitful and that we have a well
-  established and efficient model for that (which of course does not
-  mean that it is not perfectible!). May be a short section describing
-  how it works and how it could be developed would fit into this
-  document.
-  \end{quote}
-\item
-  \begin{quote}
-  I would explicitly mention Neuromorphic computing and DNA storage as
-  ``exotic'' possiblities, and other too.
-  \end{quote}
-\end{itemize}
-
-Remember\ldots{} the ``start'' of HL-LHC is 2027, but the end is 2040!
-Will QC and other more exotic possibilities be production ready for
-2027? The jury is out. Will they be in 2035? This is an entirely
-different question and the answer may well be yes.
+\end{enumerate}
+
+Undertaking this programme of priority topics and R\&D requires, first
+and foremost, investment from funding agencies to support developments.
+A new generation of developers is also needed to refresh and regenerate
+the long lived software projects on which the field so heavily relies.
+The transition from R\&D to production ready software is usually long
+and so it is urgent to undertake this investment now, with the positive
+outcomes ready to become integrated in time for HL-LHC. Regenerating the
+cadre of software developers in HEP also requires support for realistic
+long term career prospects for those who specialise in this vital area.
+Collaboration with industry can certainly be fruitful and is to be
+encouraged to allow early access to technology and communication of HEP
+problems and priorities.
+
+Almost all the domain areas identify the use of compute accelerators,
+particularly GPUs, as being a priority item, which matches our
+expectation of how computing hardware will evolve. As a common problem
+area, this is an obvious area in which expertise should be developed.
+
+The stakeholders in HEP must be able to give their priorities and
+feedback on R\&D areas. Each project has mechanisms for doing this,
+while the HSF can help to provide overall input on prioritisation.
 

### 2020-04-24 LHCC/2020/2020-01/ctcs.tex
Final version of text from Google Doc

@@
-% This is the styling for submission to arXiv using JHEP Preprint Style
-
 % Copyright (C) 2020, the authors, licence CC-BY-4.0.
@@
 \usepackage{jheppub}
-\notoc
+%\notoc
+
+\usepackage[utf8]{inputenc}
 
@@
 \usepackage{subcaption}
-%\usepackage[titletoc]{appendix}
+\usepackage[titletoc]{appendix}
 \usepackage{xspace}
@@
 
-\usepackage[utf8]{inputenc}
-
-%% Comment out before final submission
+% References
+\usepackage[style=numeric-comp,sorting=none]{biblatex}
+\addbibresource{ctcs.bib}
+
+%% Comment out before final submission %%
 \usepackage{lineno}  % for line numbering during review
@@
 continued development and optimisation will be critical in the future
-\textbackslash cite\{HSF-EPPSU-Document, EPPSU-Briefingbook\}. The
+\cite{stewart_graeme_a_2018_2413005, Ellis:2691414}. The
 challenges are driven by an ambitious physics programme, notably the LHC
@@
 improvements in computing from CPU servers at sites is likely to be only
-in the 5-10\% per year range, making the shortfall in processing
-resources more severe. As manufacturers struggle to deliver ever-more
-effective computing through CPUs the drive to different architectures
-intensifies, with GPUs becoming commonplace and increasing interest in
-even more specialised architectures, such as TPUs, IPUs and developments
-that make FPGA devices more user friendly. These pose huge challenges
-for our community as the programming models and APIs vary widely here
-and possible lock-in to a particular vendor's devices is a significant
+about 10\% per year, making the shortfall in processing resources more
+severe. As manufacturers struggle to deliver ever-more effective
+computing through CPUs the drive to different architectures intensifies,
+with GPUs becoming commonplace and increasing interest in even more
+specialised architectures, such as TPUs, IPUs and developments that make
+FPGA devices more user friendly. These pose huge challenges for our
+community as the programming models and APIs vary widely here and
+possible lock-in to a particular vendor's devices is a significant
 problem for code sustainability and preservation. Huge work has been
@@
 execution of generator software provide essential contributions to the
-success of the HL-LHC physics programme and it is critical that they get
-the right recognition and motivation. However, theorists get recognition
-on published papers, and may not be motivated to work on software
-optimisations that are not "theoretical" enough to advance their
-careers. Generator support tasks in the experiments may also not be
-enough to secure jobs/funding to experimentalists pursuing a career in
-research.
+success of the (HL-)LHC physics programme and it is critical that they
+get the right recognition and motivation. However, theorists get
+recognition on published papers, and may not be motivated to work on
+software optimisations that are not "theoretical" enough to advance
+their careers. Generator support tasks in the experiments may also not
+be enough to secure jobs/funding to experimentalists pursuing a career
+in research.
 
@@
 be complicated by the fact that critical components of some generator
-packages are written in Fortran. In some cases, alternative
-implementations of these components in other languages exist, but their
-speed is significantly lower than the optimized Fortran version. Some of
+packages are written in Fortran, which is rarely used in industry and
+less popular among developers than other programming languages. Some of
 the generators also do not use industry standard version control
@@
 NLO) imply the rejection of some events, to avoid double counting
-between events produced with n\textsubscript{jets}+1 MEs and with
-n\textsubscript{jets} MEs plus parton showers. The resulting
-inefficiencies can be relatively low depending on the process, but they
+between events produced with n\textsubscript{jets}+1 matrix elements and
+with n\textsubscript{jets} MEs plus parton showers. The resulting
+inefficiencies can be relatively high depending on the process, but they
 are unavoidable in the algorithmic strategy used by the underlying
@@
 events to generate increases by a factor 1/(1-2r)\^{}2: for instance,
-with r=25\% (which may be regarded as a worst-case scenario in ttbar
-production {[}Fre20{]}), one needs to generate 4 times as many events.
-Negative weights can instead be almost completely avoided, by design, in
-another popular matching prescription, POWHEG, which however is only
-available for a limited number of processes and describes the relevant
-physics to a different degree of precision than MC@NLO (see {[}Fre20{]}
-for a more in-depth discussion). Progress in this area can only be
-achieved by theorists: for instance, a new approach for MC@NLO-type
-matching with reduced negative weights has recently been proposed
-{[}Fre20{]}. It should be noted that negative weights can also happen at
-LO because of not-positive-definite parton density function sets and
-interference terms, e.g. in effective field theory calculations.
+with r=25\% (which may be regarded as a worst-case scenario in top quark
+pair production {[}Fre20{]}), one needs to generate 4 times as many
+events. Negative weights can instead be almost completely avoided, by
+design, in another popular matching prescription, POWHEG, which however
+is only available for a limited number of processes and describes the
+relevant physics to a different degree of precision than MC@NLO (see
+{[}Fre20{]} for a more in-depth discussion). Progress in this area can
+only be achieved by theorists: for instance, a new approach for
+MC@NLO-type matching with reduced negative weights has recently been
+proposed {[}Fre20{]}. It should be noted that negative weights can also
+happen at LO because of not-positive-definite parton density function
+sets and interference terms, e.g. in effective field theory
+calculations.
 
@@
 3. {Size of unweighted event samples required for experimental analysis
-at HL-LHC}. Another question is in which regions of phase space the
-number of unweighted events must be strictly proportional to the
-luminosity. For example, in the low pT regions of W boson production it
-is probably impossible to keep up with the data, due to the huge cross
-section. Alternative techniques should be investigated, to avoid the
-generation of huge event samples.
-
-\hypertarget{summary---generator-community-current-priorities}{%
-\subsection{Summary - generator community current
-priorities}\label{summary---generator-community-current-priorities}}
-
-{[}Text moved to final summary{]}
+at HL-LHC}. Another question, unrelated to NNLO, is in which regions of
+phase space the number of unweighted events must be strictly
+proportional to the luminosity. For example, in the low pT regions of W
+boson production it is probably impossible to keep up with the data, due
+to the huge cross section. Alternative techniques should be
+investigated, to avoid the generation of huge event samples.
+
+\subsection{Generator References}\label{references}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TO BE CONVERTED TO BibTeX, but left here for "working" %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+{[}Alb19{]} J. Albrecht et al. ``A Roadmap for HEP Software and
+Computing R\&D for the 2020s'' Comput Softw Big Sci (2019) 3, 7,
+\href{https://doi.org/10.1007/s41781-018-0018-8}{{doi:10.1007/s41781-018-0018-8}},
+\href{https://arxiv.org/abs/1712.06982}{{arXiv:1712.06982}}
+
+{[}Alw08{]} J. Alwall et al. ``Comparative study of various algorithms
+for the merging of parton showers and matrix elements in hadronic
+collisions'', Eur. Phys. J. C 53 (2008) 53,
+\href{https://doi.org/10.1140/epjc/s10052-007-0490-5}{{doi:10.1140/epjc/s10052-007-0490-5}}
+
+{[}Alw09{]} J. Alwall, S. de Visscher, F. Maltoni, ``QCD radiation in
+the production of heavy colored particles at the LHC'', JHEP 0902 (2009)
+017,
+\href{https://doi.org/10.1088/1126-6708/2009/02/017}{{doi:10.1088/1126-6708/2009/02/017}}
+
+{[}Alw14{]} J. Alwall et al. ``The automated computation of tree-level
+and next-to-leading order differential cross sections, and their
+matching to parton shower simulations'', JHEP 07 (2014) 079,
+\href{https://doi.org/10.1007/JHEP07(2014)079}{{doi:10.1007/JHEP07(2014)079}}
+
+{[}Bau13{]} C. Bauer et al., "Computing for perturbative QCD - a
+Snowmass White Paper",
+\href{https://arxiv.org/abs/1309.3598}{{arxiv:1309.3598}}
+
+{[}Ben17{]} J. Bendavid, ``Efficient Monte Carlo Integration Using
+Boosted Decision Trees and Generative Deep Neural Networks'',
+\href{https://arxiv.org/abs/1707.00028}{{arXiv:1707.00028}}
+
+{[}Bot19{]} E. Bothmann et al., ``Event generation with Sherpa 2.2'',
+\href{https://scipost.org/SciPostPhys.7.3.034}{{SciPost Phys. 7 (2019)
+034}}
+
+{[}Bot20{]} E. Bothmann et al. ``Exploring phase space with Neural
+Importance Sampling'',
+\href{https://arxiv.org/abs/2001.05478}{{arXiv:2001.05478}}
+
+{[}Chi17{]} J. T. Childers et al., "Adapting the serial Alpgen
+parton-interaction generator to simulate LHC collisions on millions of
+parallel threads", Comp. Phys. Comm. 210 (2017) 5154.
+\href{https://doi.org/10.1016/j.cpc.2016.09.0135286}{{https://doi.org/10.1016/j.cpc.2016.09.0135286}}.
+
+{[}Fre12{]} R. Frederix, S. Frixione, "Merging meets matching in
+MC@NLO", JHEP 2012 (2012) 61.
+\href{https://doi.org/10.1007/JHEP12(2012)061}{{doi:10.1007/JHEP12(2012)061}}
+
+{[}Fre20{]} R. Frederix, S. Frixione, S. Prestel. P. Torrielli, ``On the
+reduction of negative weights in MC@NLO-type matching procedures'',
+\href{https://arxiv.org/abs/2002.12716}{{arXiv:2002.12716}}
+
+{[}Fri17{]} S. Frixione, P. Nason, C. Oleari, "Matching NLO QCD
+computations with parton shower simulations: the POWHEG method", JHEP11
+(2007) 070,
+\href{https://doi.org/10.1088/1126-6708/2007/11/070}{{doi:10.1088/1126-6708/2007/11/070}}
+
+{[}Gao20{]} C. Gao. et al. ``Event Generation with Normalizing Flows'',
+\href{https://arxiv.org/abs/2001.10028}{{arXiv:2001.10028}}
+
+{[}Gen18{]} HSF Physics Event Generator Computing Workshop, CERN,
+November 2018,
+\href{https://indico.cern.ch/event/751693}{{https://indico.cern.ch/event/751693}}
+
+{[}Gen19{]} HSF Physics Generators Working Group,
+\href{https://hepsoftwarefoundation.org/workinggroups/generators.html}{{https://hepsoftwarefoundation.org/workinggroups/generators.html}}
+
+{[}Gen20{]} HSF Physics Generators Working Group, "Challenges in Monte
+Carlo event generator software for High-Luminosity LHC", in preparation
+(2020).
+
+{[}Gle08{]} T. Gleisberg, S. Hoeche, "Comix, a new matrix element
+generator", JHEP 12 (2008) 039.
+\href{https://doi.org/10.1088/1126-6708/2008/12/039}{{doi:10.1088/1126-6708/2008/12/039}}
+
+{[}Hoe14{]} S. Hoeche, F. Krauss, M. Schoenherr, "Uncertainties in
+MEPS@NLO calculations of h+jets", Phys. Rev. D 90 (2014) 014012.
+\href{https://doi.org/10.1103/PhysRevD.90.014012}{{doi:10.1103/PhysRevD.90.014012}}
+
+{[}Hoe19{]} S. Hoeche, S. Prestel, H. Schulz, "Simulation of vector
+boson plus many final jets at the high luminosity LHC", Phys. Rev. D
+100, 0140124 (2019).
+\href{http://dx.doi.org/10.1103/PhysRevD.100.014024}{{doi:10.1103/PhysRevD.100.014024}}
+
+{[}Jad03{]} S. Jadach, Foam: a general-purpose cellular Monte Carlo
+event generator, Comp. Phys. Comm. 152 (2003) 55.
+\href{https://doi.org/10.1016/S0010-4655(02)00755-5}{{doi:10.1016/S0010-4655(02)00755-5}}
+
+{[}Kaw86{]} S.Kawabata, A new Monte Carlo event generator for high
+energy physics, Comp. Phys. Comm. 41 (1986) 127.
+\href{https://doi.org/10.1016/0010-4655(86)90025-1}{{doi:10.1016/0010-4655(86)90025-1}}
+
+{[}Kli18{]} M.D. Klimek, M. Perelstein, ``Neural Network-Based Approach
+to Phase Space Integration'',
+\href{https://arxiv.org/abs/1810.11509}{{arXiv:1810.11509}}
+
+{[}Kon20{]} D. Konstantinov, "Optimization of Pythia8", EP-SFT group
+meeting, CERN, January 2020,
+\href{https://indico.cern.ch/event/890670}{{https://indico.cern.ch/event/890670}}
+
+{[}Lep80{]} G. P. Lepage, "VEGAS: an adaptive multi-dimensional
+integration program", Cornell report CLNS-447 (1980).
+\href{https://cds.cern.ch/record/123074}{{https://cds.cern.ch/record/123074}}
+
+{[}Lon05{]} L. Lonnblad, "Correcting the colour-dipole cascade model
+with fixed order matrix elements", JHEP 05 (2002) 046.
+\href{https://doi.org/10.1088/1126-6708/2002/05/046}{{doi:10.1088/1126-6708/2002/05/046}}
+
+{[}Mal03{]} F. Maltoni, T. Stelzer, "MadEvent: automatic event
+generation with MadGraph", JHEP 02 (2003) 027.
+\href{https://doi.org/10.1088/1126-6708/2003/02/027}{{doi:10.1088/1126-6708/2003/02/027}}
+
+{[}Man02{]} M. L. Mangano, M. Moretti, R. Pittau, "Multijet matrix
+elements and shower evolution in hadronic collisions: Wbb + n-jets as a
+case study", Nucl. Phys. B 632 (2002) 343.
+\href{https://doi.org/10.1016/S0550-3213(02)00249-3}{{doi:10.1016/S0550-3213(02)00249-3}}
+
+{[}Mat16{]} O. Mattelaer, "On the maximal use of Monte Carlo samples:
+re-weighting events at NLO accuracy", \emph{Eur. Phys. J. C}
+76\textbf{,} 674 (2016),
+\href{https://doi.org/10.1140/epjc/s10052-016-4533-7}{{doi:10.1140/epjc/s10052-016-4533-7}}
+
+{[}Mat18{]} O. Mattelaer, "MG5aMC status and plans", in {[}Gen19{]}
+
+{[}Nas07{]} P. Nason, MINT: a Computer Program for Adaptive Monte Carlo
+Integration and Generation of Unweighted Distributions Bicocca-FT-07-13
+(2007). \href{https://arxiv.org/abs/0709.2085}{{arxiv:0709.2085}}
+
+{[}SKe18{]} L. Sexton-Kennedy, G. Stewart, "CWP challenges and workshop
+aims", in {[}Gen19{]}
 
@@
 flexibility comes at the price of a performance penalty. We observe that
-writing more compact and more specialised code can lead to several
-percent speed-up {[}GeantV RD{]}. The technical improvements to explore
-here consist of avoiding too many virtual layers and improving the
-instruction and data locality, leading to better cache use. Going even
-further as far as specialisation is concerned, studies suggest that
-implementing compact, self-contained transport libraries with a minimal
-set of physics models (for instance only electromagnetic interaction for
-selected particles) with predefined scoring (to simulate, for instance,
-the response of an EM calorimeter) can also bring speed-ups to the
-complete application {[}GeantV RD{]}. As far as architectural
-modifications are concerned, the design of the Geant4 track-status
-machinery prevents the introduction of fine-grained track parallelism.
-It is still not clear whether this kind of parallelism can bring any
-substantial speed-up, due to the overheads; however, making the Geant4
-kernel and physics processes stateless would certainly provide the
-necessary starting points for further R\&D in that direction. Most of
-these investigations have started and it is hoped to see some first
-results on a year time-scale. More effort, however, will be required to
-integrate any eventual changes in the production code in a timely
-manner. The GeantV Vector prototype has allowed the assessment of the
-achievable speedup using a novel, vectorized approach to particle
+writing more compact and more specialised code can lead in certain
+places to up to several tens percent speed-up {[}GeantV RD{]}. The
+technical improvements to explore here consist of avoiding too many
+virtual layers and improving the instruction and data locality, leading
+to better cache use. Going even further as far as specialisation is
+concerned, studies suggest that implementing compact, self-contained
+transport libraries with a minimal set of physics models (for instance
+only electromagnetic interaction for selected particles) with predefined
+scoring (to simulate, for instance, the response of an EM calorimeter)
+can also bring speed-ups to the complete application {[}GeantV RD{]}. As
+far as architectural modifications are concerned, the design of the
+Geant4 track-status machinery prevents the introduction of fine-grained
+track parallelism. It is still not clear whether this kind of
+parallelism can bring any substantial speed-up, due to the overheads;
+however, making the Geant4 kernel and physics processes stateless would
+certainly provide the necessary starting points for further R\&D in that
+direction. Most of these investigations have started and it is hoped to
+see some first results on a year time-scale. More effort, however, will
+be required to integrate any eventual changes in the production code in
+a timely manner. The GeantV Vector prototype has allowed the assessment
+of the achievable speedup using a novel, vectorized approach to particle
 transport. The demonstrator of a full electromagnetic shower in
@@
 function of the incoming particle type and energy. An example of such a
-parameterization was implemented several years ago in the context of the
-H1 experiment. This parameterization, available within the Geant4
-toolkit under the name of the GFLASH library, became the starting idea
-for several `custom' implementations, specific for other calorimeters.
-Those were interfaced to experiments' simulation applications through
-the available `hook' in the Geant4 toolkit. The LHC experiments
-implemented, for example, dedicated parametrized response libraries for
-some of their calorimeters. Recently, an effort has been invested in
-Geant4 to generalize the parameterization formulae, and to implement
-automatic procedures of tuning the parameters for specific detector
-geometries. This kind of `classical' fast simulation, which describes
-the particle shower shape using some complex mathematical functions with
-several parameters, will remain an important tool; however, it is also
-clear that their overall precision for the different energy values,
-especially for highly granular and complex calorimeters will always be
-relatively low. The recent developments of Deep Learning-based
-techniques have opened an exciting possibility of replacing those
-`classical' parameterisations by trained neural networks that would
-reproduce the detector response. This approach consists of training
-generative models such as Generative Adversarial Networks (GAN),
-Variational Auto-Encoders (VAE) or Autoregressive Generative Networks on
-the `images' of particle showers {[}ML001{]}. The energy deposition
-values in the calorimeter cells are considered as `pixels' of the images
-that the network is supposed to reproduce. The first studies and
-prototypes have shown very promising results, however, the
+parameterization was implemented many years ago in the context of the H1
+experiment {[}H1Ref{]}. This parameterization, available within the
+Geant4 toolkit under the name of the GFLASH library, became the starting
+idea for several `custom' implementations, specific for other
+calorimeters. Those were interfaced to experiments' simulation
+applications through the available `hook' in the Geant4 toolkit. The LHC
+experiments implemented, for example, dedicated parametrized response
+libraries for some of their calorimeters. Recently, an effort has been
+invested in Geant4 to generalize the parameterization formulae, and to
+implement automatic procedures of tuning the parameters for specific
+detector geometries. This kind of `classical' fast simulation, which
+describes the particle shower shape using some complex mathematical
+functions with several parameters, will remain an important tool;
+however, it is also clear that their overall precision for the different
+energy values, especially for highly granular and complex calorimeters
+will always be relatively low. The recent developments of Deep
+Learning-based techniques have opened an exciting possibility of
+replacing those `classical' parameterisations by trained neural networks
+that would reproduce the detector response. This approach consists of
+training generative models such as Generative Adversarial Networks
+(GAN), Variational Auto-Encoders (VAE) or Autoregressive Generative
+Networks on the `images' of particle showers {[}ML001{]}. The energy
+deposition values in the calorimeter cells are considered as `pixels' of
+the images that the network is supposed to reproduce. The first studies
+and prototypes have shown very promising results, however, the
 generalization of the developed models (for different calorimeters, with
@@
 explored for specific cases. Similarly, the re-use of simulated or real
-events to mimic the background to hard-scatter events is also under
-investigation, where additional challenges exist concerning storage and
-I/O due to the handling of the large minimum bias samples needed to
-model the additional interactions.
+events to mimic the background to hard-scatter events is also exploited,
+where additional challenges exist concerning storage and I/O due to the
+handling of the large minimum bias samples needed to model the
+additional interactions.
 
@@
 
+\subsection{Simulation References}\label{references-1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TO BE CONVERTED TO BibTeX, but left here for "working" %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+{[}CMS01{]} V. Ivanchenko and S. Banerjee on behalf of the CMS
+Collaboration, EPJ Web of Conferences 214, 02012 (2019), The PDF is
+here:
+\url{https://www.epj-conferences.org/articles/epjconf/pdf/2019/19/epjconf_chep2018_02012.pdf}
+
+{[}CMS02{]} M Hildreth et al 2015 J. Phys.: Conf. Ser. 664 072022
+
+{[}CMS03{]} M Hildreth et al 2017 J. Phys.: Conf. Ser. 898 042040
+
+{[}GPU-``MPEXS-DNA{]} S. Okada et al. ``MPEXS-DNA: a new GPU-based Monte
+Carlo simulator for track structures and radiation chemistry at
+subcellular scale,'' Med Phys. vol 46 no. 3, pp. 1483-1500 (2019)
+
+{[}Hybrid-Gate{]} J. Bert et al., "Hybrid GATE: A GPU/CPU implementation
+for imaging and therapy applications," 2012 IEEE Nuclear Science
+Symposium and Medical Imaging Conference Record (NSS/MIC), Anaheim, CA,
+pp. 2247-2250 (2012)
+
+{[}G4CU{]}
+\url{https://kds.kek.jp/indico/event/15926/session/30/contribution/110/material/slides/0.pdf}
+
+{[}Opticks{]} S. Blyth, ``Opticks : GPU Optical Photon Simulation for
+Particle Physics using NVIDIA® OptiXTM,'' EPJ Web Conf., 214, pp. 02027
+(2019) 02027
+
+{[}LHCb01{]} D. Muller, ``Adopting new technologies in the LHCb Gauss
+simulation framework'', EPJ Web of Conferences \textbf{214}, 02004
+(2019); B.Siddi and D. Muller, ``Gaussino a Gaudi-based core simulation
+framework'', CERN-LHCb-PROC-2019-010;
+
+{[}ML001{]} Adversarial Networks (GAN) (3DGAN reference), Variational
+Auto-Encoders (VAE) or Autoregressive Generative Networks on the
+`images' of particle showers
+
+{[}GAN-ATLAS, GAN-CMS, GAN-LHCb{]} Reference to GAN calorimeters for
+ATLAS, CMS and LHCb
+
+{[}LHCb02{]} D. Muller et al. ``ReDecay: A novel approach to speed up
+the simulation at LHCb'' , Eur. Phys. J. C 78 (2018)
+
+{[}GeantV RD{]}
+\href{https://indico.cern.ch/event/818702/}{{https://indico.cern.ch/event/818702/}}
+
+{[}CHEP-Juno{]} S. Blyth, ``Meeting the challenge of JUNO simulation
+with Opticks GPU Optical Photon Accelerator'', presented at CHEP 2019 in
+Adelaide, Australia
+
+{[}SHIFT01{]} Hamilton, S.P., Evans, T.M., 2019. Continuous-energy Monte
+Carlo neutron transport on GPUs in the Shift code. \emph{Annals of
+Nuclear Energy} \textbf{128}, 236--247.
+\url{https://doi.org/10.1016/j.anucene.2019.01.012}
+
+{[}DELPHES{]} de Favereau, J., Delaere, C., Demin, P. \emph{et al.}
+DELPHES 3: a modular framework for fast simulation of a generic collider
+experiment. \emph{J. High Energ. Phys.} 2014, 57 (2014).
+\href{https://doi.org/10.1007/JHEP02(2014)057}{{https://doi.org/10.1007/JHEP02(2014)057}}
+
+{[}PGS4{]} J. Conway, R. Culbertson, R. Demina, B. Kilminster, M. Kruse,
+S. Mrenna, J. Nielsen, M. Roco, A. Pierce, J. Thaler, T. Wizansky,
+``Pretty Good Simulation'',
+\href{http://conway.physics.ucdavis.edu/research/software/pgs/pgs4-general.htm}{{http://conway.physics.ucdavis.edu/research/software/pgs/pgs4-general.htm}}
+
+{[}LAMARR{]} A.Davis, ``Fast simulations in LHCb'', CHEP 2019,
+\href{https://indico.cern.ch/event/773049/contributions/3474742/attachments/1937708/3211775/davis_fast_simulations_at_lhcb_v3.pdf}{{https://indico.cern.ch/event/773049/contributions/3474742/attachments/1937708/3211775/davis\_fast\_simulations\_at\_lhcb\_v3.pdf}}
+
+{[}ALICE01{]} B. Volker, HSF Simulation Working Group meeting,
+\href{https://indico.cern.ch/event/782507/contributions/3293465/}{{https://indico.cern.ch/event/782507/contributions/3293465/}}
+
+{[}H1Ref{]} G. Grindhammer, M. Rudowicz and S. Peters, NIM A290 (1990)
+469-488
+
 \hypertarget{reconstruction-and-software-triggers}{%
@@
 outside LHC collaborations that could be adapted by the LHC community to
-improve trigger and reconstruction algorithms. Nearly all software
-solutions presented in this chapter are tailored to a specific
-experiment, and would require additional work to be adapted to different
-environments. Nevertheless, it is still useful to maintain open
-communication channels to discuss design choices and compare performance
-assessments of different solutions to guide future software and
-reconstruction design choices.
+improve trigger and reconstruction algorithms.
 
@@
 experiments at the HL-LHC will increase by up to a factor 10 with
-respect to Run-3, owing to upgraded trigger systems. ATLAS and CMS plan
-to maintain a two-tiered trigger system, where a hardware trigger makes
-use of coarse event information to reduce the event rate to 10x over the
-current capability, up to 1 MHz {[}cite TDR{]}. The high level trigger
-system, implemented in software, selects up to 100 kHz of events to be
-saved in full for subsequent analysis\footnote{LHCb {[}cite TDR{]} and
-  ALICE {[}cite TDR{]} will both stream the full collision rate to
-  real-time or quasi-realtime software trigger systems.}.
+respect to Run-3, owing to upgraded trigger systems and expanded physics
+programs. ATLAS and CMS plan to maintain a two-tiered trigger system,
+where a hardware trigger makes use of coarse event information to reduce
+the event rate to 10x over the current capability, up to 1 MHz {[}cite
+TDR{]}. The high level trigger system, implemented in software, selects
+up to 100 kHz of events to be saved in full for subsequent
+analysis\footnote{LHCb {[}cite TDR{]} and ALICE {[}cite TDR{]} will both
+  stream the full collision rate to real-time or quasi-realtime software
+  trigger systems.}.
 
@@
 Implementations of real-time analysis systems, where raw data is
-processed in its final form as close as possible to the detector (e.g.
+processed into its final form as close as possible to the detector (e.g.
 in the high-level trigger farm), are in use within several experiments.
@@
 possible to do this with acceptable reduction of the analysis
-sensitivity and with minimal biases.
+sensitivity and with minimal biases. The HSF trigger and reconstruction
+group has been encouraging cross-talk between LHC experiments and beyond
+on real-time analysis, as these kinds of non-standard trigger and
+reconstruction workflows increase the physics output for selected
+physics cases without adding significant overhead to the current
+resources.
 
@@
 and mkfit{]}) have demonstrated success in increasing the use of vector
-units. In addition, industry developed software, including machine
-learning toolkits (e.g., tensorflow) typically make excellent use of
-vector units.
+units. In addition, HEP has increasingly benefited from industry
+developed software, including machine learning toolkits (e.g.,
+tensorflow), that typically make excellent use of vector units.
+Challenges remain in this area. For example, full exploitation of this
+must also account for many generations of hardware that experiments must
+exploit on the grid. In addition, to realize full performance gains, a
+large fraction of the application must be improved to use SIMD processor
+capabilities (partly due to turbo-boost capabilities of processors)
 
@@
 to facilities offering large CPU/GPU resources. One can also foresee
-that HEP will benefit from techniques and cross-talk from industry.
+that HEP will increase the benefit from techniques and cross-talk from
+industry.
 
@@
 
+\hypertarget{general-remarks-on-trigger-and-reconstruction-software-sharing}{%
+\subsection{General remarks on trigger and reconstruction software
+sharing}\label{general-remarks-on-trigger-and-reconstruction-software-sharing}}
+
+Nearly all software solutions presented in this chapter are tailored to
+a specific experiment, and would require additional work to be adapted
+to different environments. Nevertheless, it is still useful to maintain
+open communication channels to discuss design choices and compare
+performance assessments of different solutions to guide future software
+and reconstruction design choices.
+
+The use of open-source elements of the LHC software stack for trigger
+and reconstruction by smaller experiments, especially in case of common
+experimental design\footnote{An example of software sharing happens in
+  the case of the FASER experiment, whose offline software
+  (\href{https://gitlab.cern.ch/faser/calypso}{{https://gitlab.cern.ch/faser/calypso}})
+  is based on a derivative of the ATLAS Athena open-source software
+  {[}ref{]}, which in turn uses the Gaudi framework {[}ref{]}. FASER
+  also shares tracking detector components with ATLAS, facilitating the
+  adoption of software relying on similar detector descriptions.}, is
+still worthwhile This benefits experiments with a smaller developer base
+than LHC experiments, and increases the return on investment for LHC
+experiment software. For this reason, we also encourage the addition of
+common open software projects to the ESCAPE software catalogue {[}ref
+ESCAPE{]}.
+
+\subsection{Reconstruction and Software Trigger References}\label{references-rst}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TO BE CONVERTED TO BibTeX, but left here for "working" %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\emph{{[}1{]}
+\href{https://arxiv.org/abs/1802.08640}{{https://arxiv.org/abs/1802.08640}}}
+
+\emph{{[}2{]}
+\href{https://cds.cern.ch/record/2693670}{{https://cds.cern.ch/record/2693670}}}
+
+\emph{{[}3{]}} - Andrea Bocci \emph{at al.}, ``\emph{Heterogeneous
+online reconstruction at CMS}'', presented at CHEP 2019\\
+- Andrea Bocci \emph{at al.}, ``\emph{Heterogeneous reconstruction:
+combining ARM processors with GPUs}'', presented at CHEP 2019
+
+{[}3a{]}
+\href{https://indico.cern.ch/event/773049/contributions/3473268/attachments/1937520/3211376/GPU_Usage_in_ATLAS_Reconstruction_and_Analysis_2019.11.04..pdf}{{https://indico.cern.ch/event/773049/contributions/3473268/attachments/1937520/3211376/GPU\_Usage\_in\_ATLAS\_Reconstruction\_and\_Analysis\_2019.11.04..pdf}}
+
+\emph{{[}4{]}}
+\href{https://arxiv.org/abs/1912.09161}{{https://arxiv.org/abs/1912.09161}}
+
+\emph{{[}5{]}}
+\href{http://dx.doi.org/10.1088/1742-6596/898/4/042011}{{http://dx.doi.org/10.1088/1742-6596/898/4/042011}},
+\href{https://arxiv.org/abs/1910.03128}{{https://arxiv.org/abs/1910.03128}},
+The ACTS project: track reconstruction software for HL-LHC and beyond,
+Proceedings of CHEP2019, in progress
+
+{[}6{]} Xiangyang Ju et al, "Graph Neural Networks for Particle
+Reconstruction in High Energy Physics Detectors"
+\href{https://arxiv.org/abs/2003.11603}{{(arxiv link}})
+(\href{https://github.com/exatrkx/exatrkx-neurips19}{{Software and
+documentation}})
+
+{[}7{]} J. Duarte \emph{et al}, ``Fast inference of deep neural networks
+in FPGAs for particle physics'', \emph{JINST} \textbf{13} P07027, 2018,
+doi:
+\href{https://iopscience.iop.org/article/10.1088/1748-0221/13/07/P07027/meta}{{10.1088/1748-0221/13/07/P07027}}\\
+S. Summers \emph{et al}, ``Fast inference of Boosted Decision Trees in
+FPGAs for particle physics'', 2020, {[}Online{]} available:
+\href{https://arxiv.org/abs/2002.02534}{{arXiv:2002.02534}}\\
+G. Gi Guglielmo \emph{et al}, ``Compressing deep neural networks on
+FPGAs to binary and ternary precision with HLS4ML'', 2020, {[}Online{]}
+available: \href{https://arxiv.org/abs/2003.06308}{{arXiv:2003.06308}}
+
+{[}8{]} SONIC:
+\href{https://arxiv.org/abs/1904.08986}{{https://arxiv.org/abs/1904.08986}}
+
 \hypertarget{data-analysis}{%
@@
 recorded event rate by an order of magnitude, compared with a projected
-5-10\% annual growth of storage resources. The large pile-up in HL-LHC
+\textasciitilde 10\% annual growth of storage resources. The large pile-up in HL-LHC
 collisions will compound the storage challenge by inflating the data
@@
 to be selected, derived and aggregated into histograms. Several
-platforms developed by industry or data science have emerged in the past
+platforms developed by industry or data science have emerged in recent
 years to quickly aggregate large datasets with parallel execution
@@
 of those tools have been tested to understand the feasibility of usage
-in the HEP context.
+in the HEP context providing fresh ideas and new paradigms that have
+stimulated development within the HEP toolkit.
 
@@
 proto-typing, and the ability to out-source compute-intensive work to
-more performant languages like C and C++. In fact this is similar to the
+more performant languages like C and C++. This is similar to the
 experimental software frameworks where Python is used as steering code
-and C/C++ is used as an efficient backend implementation for the Python
-modules. The typical Belle II analysis starts with Python using a custom
-DSL to specify particle selections and algorithms implemented in C++ and
-output ntuple content. The ntuple is typically analysed by the PyHEP
-toolkit {[}6{]}, and also benefits from leveraging packages from the
-wider data science community. Deeper integration with python tools is
-also particularly important for enabling state-of-the-art machine
-learning frameworks such as PyTorch and TensorFlow, whose relevance in
-analysis is likely to grow in the long term. It is important to provide
-enough usable functionality in the optimised layer to avoid the use of
-slow loops in Python, which would cripple performance. Tools like
-PyROOT, which allows the use of any C++ class from Python, can play an
-instrumental role in enlarging the whole analysis ecosystem with an
-effortless integration of Python and C++ components, in particular
-consolidating HEP specific tools written in C++ with industry python big
-data functionality.
+and C/C++ is used as an efficient backend implementation. The typical
+Belle II analysis starts with Python using a custom DSL to specify
+particle selections and algorithms implemented in C++ and define output
+ntuples. The ntuples are typically analysed by the PyHEP toolkit
+{[}6{]}, and packages from the wider data science community. Deeper
+integration with external python tools is also particularly important
+for enabling state-of-the-art machine learning frameworks such as
+PyTorch and TensorFlow, whose relevance in analysis is likely to grow in
+the long term. Meanwhile PyROOT, which allows the use of any C++ class
+from Python, has the potential to provide a coherent analysis ecosystem
+with an effortless integration of HEP specific tools written in C++ with
+industry python big data toolkits.
 
@@
 than a standard storage type's mantissa, enabling further savings in
-storage space. The event processing framework was extended with
-DataFrame-like functionality (RDataFrame) implementing a declarative
-approach to analysis processing in either Python or C++ (always executed
-in C++), which natively exploits multi-core architectures {[}7{]} and
-could be extended to accelerators. Used from Python, it allows to
-pre-process data in C++, and export to numpy. Flexibility in interfacing
-RDataFrame to non-ROOT data formats has allowed the ALICE collaboration
-to prototype its Run 3 analysis model on this new technology. Growing
-use of Machine Learning was also anticipated, and the TMVA toolkit has
-been improved with substantial optimisations, more complex deep learning
-models and enhanced interoperability with common data science tools
-{[}14,15{]} which is of particular importance for inference.
+storage space.
+
+Inspired by data science tools, the event processing framework was
+extended with DataFrame-like functionality (RDataFrame) implementing a
+declarative approach to analysis processing in either Python or C++
+(always executed in C++), which natively exploits multi-core
+architectures {[}7{]} and could be extended to accelerators. Used from
+Python, it allows to pre-process data in C++, and export to numpy.
+Flexibility in interfacing RDataFrame to non-ROOT data formats has
+allowed the ALICE collaboration to prototype its Run 3 analysis model on
+this new technology. Growing use of Machine Learning was also
+anticipated, and the TMVA toolkit has been improved with substantial
+optimisations, more complex deep learning models and enhanced
+interoperability with common data science tools {[}14,15{]} which is of
+particular importance for inference.
 
@@
 institutes mostly performing the same kind of operations, i.e. data
-reduction, plotting, variation of systematic uncertainty, fitting etc.
+reduction, plotting, variation of systematic uncertainty, fitting, etc.
 It is clear that tools and frameworks for performing these repeated
@@
 
+One final comment in reviewing recent developments is worth noting.
+Exposure of the HEP community to industry-standard data science tools
+has undoubtedly been a force for good and a net gain for the community,
+providing fresh ideas and new ways of working - RDataFrame is an
+excellent example. It is clear that the larger and better-supported data
+science community will continue to innovate and produce tools that HEP
+will benefit from, both by using directly and by borrowing their ideas
+for our own tools. Taking machine learning as an example, it is
+extremely likely that industry will blaze a trail that HEP would do well
+to follow and we should continue to build bridges to those tools.
+Equally though, HEP will continue to have its own unique challenges and
+it's less likely that ultra-performant machine learning inference will
+be top of the data scientists wishlist. Given the challenges presented
+by extreme-throughput analysis at the HL-LHC, the right balance must be
+found between software development within the community and relying on
+external toolkits.
+
 Finally, attention must be given to ensuring the longevity of any
@@
 considered for emerging projects that show significant promise.
+
+\subsection{Analysis References}\label{references-2}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TO BE CONVERTED TO BibTeX, but left here for "working" %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+{[}1{]} *ATLAS Collaboration, \emph{Evolution of the ATLAS analysis
+model for Run-3 and prospects for HL-LHC}, ATL-SOFT-PROC-2020-002,
+\href{https://cds.cern.ch/record/2708664/}{{https://cds.cern.ch/record/2708664/}}
+
+{[}2{]} CMS Collaboration, HL-LHC resource projections,
+\href{https://twiki.cern.ch/twiki/bin/view/CMSPublic/CMSOfflineComputingResults}{{https://twiki.cern.ch/twiki/bin/view/CMSPublic/CMSOfflineComputingResults}}
+
+{[}3{]} *ALICE Collaboration, \emph{The ALICE analysis train system},
+J.Phys.Conf.Ser. 608 (2015) no.1, 012019,
+\href{http://arxiv.org/abs/arXiv:1502.06381}{{http://arxiv.org/abs/arXiv:1502.06381}}
+
+{[}4{]} *ATLAS Collaboration, \emph{A new petabyte-scale data derivation
+framework for ATLAS. Journal of Physics: Conference Series}, 664.
+072007.
+
+{[}5{]} *CMS Collaboration, \emph{A further reduction in CMS event data
+for analysis: the NANOAOD format}, EPJ Web of Conferences. 214. 06021.
+
+{[}6{]}
+\href{https://github.com/hsf-training/PyHEP-resources}{{https://github.com/hsf-training/PyHEP-resources}}
+
+{[}7{]} D Piparo et al, \emph{RDataFrame: Easy Parallel ROOT Analysis at
+100 Threads}, EPJ Web of Conferences 214, 06029 (2019),
+\href{https://www.epj-conferences.org/articles/epjconf/abs/2019/19/epjconf_chep2018_06029/epjconf_chep2018_06029.html}{{https://www.epj-conferences.org/articles/epjconf/abs/2019/19/epjconf\_chep2018\_06029/epjconf\_chep2018\_06029.html}}
+
+{[}8{]}
+\href{https://github.com/scikit-hep/uproot}{{https://github.com/scikit-hep/uproot}},
+\href{https://github.com/scikit-hep/awkward-array}{{https://github.com/scikit-hep/awkward-array}},
+\href{https://github.com/CoffeaTeam}{{https://github.com/CoffeaTeam}},
+\href{https://github.com/diana-hep/pyhf}{{https://github.com/diana-hep/pyhf}}
+
+{[}9{]} J Blomer et al, \emph{Evolution of the ROOT Tree I/O},
+\href{https://arxiv.org/abs/2003.07669}{{https://arxiv.org/abs/2003.07669}}
+
+{[}10{]} ADL Workshop at Fermilab,
+\href{https://indico.cern.ch/event/769263/timetable/?view=standard}{{https://indico.cern.ch/event/769263/}},
+Gordon's CHEP talk (good overview of links, is there something more
+official?)
+\href{https://indico.cern.ch/event/773049/contributions/3476174/attachments/1938123/3220511/Declarative.pdf}{{https://indico.cern.ch/event/773049/contributions/3476174/attachments/1938123/3220511/Declarative.pdf}}
+
+{[}11{]}
+\href{https://github.com/iris-hep/adl-benchmarks-index}{{https://github.com/iris-hep/adl-benchmarks-index}}
+
+{[}12{]} L Heinrich \& R Rocha, \emph{Reperforming a Nobel Prize
+discovery on Kubernetes}, talk at CHEP 2019,
+\href{https://indico.cern.ch/event/773049/contributions/3581373}{{https://indico.cern.ch/event/773049/contributions/3581373}}
+(originally at KubeCon)
+
+{[}13{]} *ALICE Collaboration, \emph{The ALICE analysis framework for
+LHC Run 3}, EPJ Web Conf. 214 (2019) 05045
+\href{http://cds.cern.ch/record/2699856}{{http://cds.cern.ch/record/2699856}}
+
+{[}14{]} \href{http://paperpile.com/b/2mqsuS/IDej}{L Moneta et al,
+\emph{Machine Learning with ROOT/TMVA}, EPJ Web of Conferences;
+Proceedings of CHEP 2019 (in print). 2020.}
+
+{[}15{]} \href{http://paperpile.com/b/2mqsuS/6uIv}{K Albertsson et al,
+\emph{Fast Inference for Machine Learning in ROOT/TMVA}, EPJ Web of
+Conferences; Proceedings of CHEP 2019 (in print). 2020.}
+
+{[}16{]} J Blomer, \emph{A quantitative review of data formats for HEP
+analyses}, Journal of Physics: Conference Series. 2018. p. 032020.
+
+{[}17{]} HEP Software Foundation Collaboration, \emph{HEP Software
+Foundation Community White Paper Working Group - Data Analysis and
+Interpretation},
+\href{https://arxiv.org/abs/1804.03983}{{arXiv:1804.03983}}
+
+{[}18{]} M Rieger et al, \emph{Design and Execution of make-like,
+distributed Analyses based on Spotify's Pipelining Package Luigi},
+\href{https://arxiv.org/abs/1706.00955}{{arXiv:1706.00955}}
+
+{[}19{]} T Kuhr et al, \emph{The Belle II Core Software},
+\href{https://docs.belle2.org/record/1044}{{BELLE2-PUB-TE-2018-002}}
+
+{[}20{]} Swan homepage:
+\href{https://swan.web.cern.ch/}{{https://swan.web.cern.ch/}}
+
+{[}21{]} Reana homepage:
+\href{http://reanahub.io/}{{http://reanahub.io/}}
+
+Proceedings marked `*' will be credited to the shorter author lists
 
@@
   \begin{quote}
-  Develop generic fast simulation models and techniques, including
+  Develop further fast simulation models and techniques, including
   machine learning based options in addition to optimised parametric
-  approaches.
+  approaches as well as common frameworks for their tuning and
+  validation.
   \end{quote}
@@
   \end{quote}
+\item
+  \begin{quote}
+  Develop integration prototypes to exercise and benchmark the
+  simultaneous use of GPU specific libraries with CPU-based software
+  (e.g. Geant4) to cover all particles and processes required in
+  experiments simulation frameworks.
+  \end{quote}
 \end{enumerate}
@@
   \begin{quote}
-  Support uniformity between online and offline software wherever
-  possible, so the experiments can benefit from consistent event
-  selection between trigger and final analysis.
+  Develop fast and performant reconstruction software, making optimal
+  use of hardware resources. A particularly crucial development for
+  HL-LHC is that of pattern recognition solutions (including timing
+  information, if available) that can withstand the increase of pile-up
+  at HL-LHC. Wherever possible, solutions should be uniform for online
+  and offline software, so the experiments can benefit from consistent
+  event selection between trigger and final analysis and from more
+  effective real-time analysis results.
   \end{quote}
@@
   \begin{quote}
-  Continue encouraging cross-talk between LHC experiments (as well as
-  with other disciplines) on use cases for real-time analysis, as it
-  enables ATLAS and CMS to increase the physics output for selected
-  physics cases without adding significant overhead to the current
-  resources.
-  \end{quote}
-\item
-  \begin{quote}
-  Develop fast and performant tracking solutions (including timing
-  information, if available) that can withstand the increase of pile-up
-  at HL-LHC.
-  \end{quote}
-\item
-  \begin{quote}
-  Vectorisation (is this bundled with vectorisation?)
+  Develop solutions that can exploit accelerators (FPGA, GPU) for
+  trigger and reconstruction, including general-purpose interoperability
+  libraries with a focus to generalizing and mitigating their
+  dependencies on architecture-specific codes.
   \end{quote}
@@
   trigger and reconstruction software.
-  \end{quote}
-\item
-  \begin{quote}
-  Develop solutions that can exploit accelerators and HPCs for trigger
-  and reconstruction, with a focus to mitigating their dependencies on
-  architecture-specific codes.
-  \end{quote}
-\item
-  \begin{quote}
-  Encourage the use of open-source elements of the LHC software stack
-  for trigger and reconstruction by smaller experiments, especially in
-  case of common experimental design\footnote{An example of software
-    sharing happens in the case of the FASER experiment, whose offline
-    software
-    (\href{https://gitlab.cern.ch/faser/calypso}{{https://gitlab.cern.ch/faser/calypso}})
-    is based on a derivative of the ATLAS Athena open-source software
-    {[}ref{]}, which in turn uses the Gaudi framework {[}ref{]}. FASER
-    also shares tracking detector components with ATLAS, facilitating
-    the adoption of software relying on similar detector descriptions.}.
-  This benefits experiments with a smaller developer base than LHC
-  experiments, and increases the return on investment for LHC experiment
-  software. For this reason, we also encourage the addition of common
-  open software projects to the ESCAPE software catalogue {[}ref
-  ESCAPE{]}.
   \end{quote}

### 2020-04-28 LHCC/2020/2020-01/ctcs.tex
Snapshot, trigger and reco properly referenced
Was moved to a separate input file to avoid editing mistakes

@@
 % JHEP preprint template
-\documentclass[10pt,a4paper]{article}
+\documentclass[11pt,a4paper]{article}
 \usepackage{jheppub}
@@
 non-technical (e.g. funding, training, careers for those working in this
-area), exist {[}Alb19,Gen18{]}.
+area), exist~\cite{Alves:2017she,Gen18}.
 
@@
 \hypertarget{collaboration-challenges}{%
-\subsection{Collaboration challenges}\label{collaboration-challenges}}
+\subsection{Collaborative challenges}\label{collaboration-challenges}}
 
@@
 {[}Gen20{]} HSF Physics Generators Working Group, "Challenges in Monte
-Carlo event generator software for High-Luminosity LHC", in preparation
-(2020).
+Carlo event generator software for High-Luminosity LHC", CERN-LPCC-2020-002 (2020).
 
@@
 
-\hypertarget{reconstruction-and-software-triggers}{%
-\section{Reconstruction and Software
-Triggers}\label{reconstruction-and-software-triggers}}
-
-Software trigger and event reconstruction techniques in HEP face a
-number of new challenges in the next decade. Advances in facilities and
-future experiments bring a dramatic increase in physics reach, as well
-as increased event complexity and rates.
-
-At the HL-LHC, the central challenge for high-level triggers (e.g.,
-software triggers) and object reconstruction is to maintain excellent
-efficiency and resolution in the face of high pileup values, especially
-at low transverse momenta. Detector upgrades such as increases in
-channel density, high precision timing and improved detector layouts are
-essential to overcome these problems. The subsequent increase of event
-complexity at the HL-LHC also requires the development of software
-algorithms that can process events with a similar cost per event to
-Run-2 and Run-3. At the same time, algorithmic approaches need to
-effectively take advantage of evolutions in computing technologies,
-including increased SIMD capabilities, increased core counts in CPUs,
-and heterogeneous hardware.
-
-This section focuses on the challenges identified in the Community White
-Paper {[}1{]} and the development of solutions that. It also includes
-mentions of open source software efforts that have developed within or
-outside LHC collaborations that could be adapted by the LHC community to
-improve trigger and reconstruction algorithms.
-
-\hypertarget{the-evolution-of-triggers-and-real-time-analysis}{%
-\subsection{\texorpdfstring{The evolution of triggers and real-time
-analysis
-}{The evolution of triggers and real-time analysis }}\label{the-evolution-of-triggers-and-real-time-analysis}}
-
-Trigger systems are evolving to be more capable, both in their ability
-to select a wider range of events of interest for the physics program of
-their experiment, and their ability to stream a larger rate of events
-for further processing. The event rates that will be processed by
-experiments at the HL-LHC will increase by up to a factor 10 with
-respect to Run-3, owing to upgraded trigger systems and expanded physics
-programs. ATLAS and CMS plan to maintain a two-tiered trigger system,
-where a hardware trigger makes use of coarse event information to reduce
-the event rate to 10x over the current capability, up to 1 MHz {[}cite
-TDR{]}. The high level trigger system, implemented in software, selects
-up to 100 kHz of events to be saved in full for subsequent
-analysis\footnote{LHCb {[}cite TDR{]} and ALICE {[}cite TDR{]} will both
-  stream the full collision rate to real-time or quasi-realtime software
-  trigger systems.}.
-
-Experiments have also been working towards minimizing the differences
-between trigger (online) and offline software for reconstruction and
-calibration, so that more refined physics objects can be obtained
-directly within the HLT farm for a more efficient event selection. This
-is also in-line with enhancing the experiment's capabilities for
-real-time data analysis of events accepted by the hardware trigger
-system, driven by use cases where the physics potential improves when
-analysing more events than what can be written out in full with
-traditional data processing.
-
-Implementations of real-time analysis systems, where raw data is
-processed into its final form as close as possible to the detector (e.g.
-in the high-level trigger farm), are in use within several experiments.
-These approaches remove the detector data that typically makes up the
-raw data tier kept for offline reconstruction, and keep only a limited
-set of analysis objects reconstructed within the high level trigger.
-
-Real-time analysis techniques are being adopted to enable a wider range
-of physics signals to be saved by the trigger for final analysis. As
-rates increase, these techniques can become more important and
-widespread by enabling only the parts of an event associated with the
-signal candidates to be saved, reducing the required disk space. The
-experiments are focusing on the technological developments that make it
-possible to do this with acceptable reduction of the analysis
-sensitivity and with minimal biases. The HSF trigger and reconstruction
-group has been encouraging cross-talk between LHC experiments and beyond
-on real-time analysis, as these kinds of non-standard trigger and
-reconstruction workflows increase the physics output for selected
-physics cases without adding significant overhead to the current
-resources.
-
-Active research topics include compression and custom data formats;
-toolkits for real-time detector calibration and validation which will
-enable full offline analysis chains to be ported into real-time; and
-frameworks which will enable non-expert offline analysts to design and
-deploy real-time analyses without compromising data taking quality.
-Use-cases for real-time analysis techniques have expanded during the
-last years of Run 2 (cite LHCb-TESLA?) and their use appears likely to
-grow already during Run 3 data taking for all experiments owing to
-improvements in the HLT software and farms (cite TDAQ TDR, TLA pub note
-ATLAS). Further ideas include retaining very reduced data from as much
-of the data taking rate as possible from selected regions and
-sub-detectors, even up to the 40 MHz bunch crossing rate (cite CHEP
-proceedings CMS).
-
-\hypertarget{challenges-and-improvements-foreseen-in-event-reconstruction}{%
-\subsection{Challenges and improvements foreseen in event
-reconstruction}\label{challenges-and-improvements-foreseen-in-event-reconstruction}}
-
-Processing and reducing raw data into analysis-level formats, event
-reconstruction, is a major component of offline computing resource
-needs, and in light of the previous section it is relevant for online
-resources as well. This is an essential step towards precision
-reconstruction, identification and measurement of physics-objects at
-HL-LHC.
-
-Algorithmic areas of particular importance at HL-LHC experiments are
-charged-particle trajectory reconstruction (``tracking''), including
-hardware triggers based on tracking information which may seed later
-software trigger and reconstruction algorithms; jet reconstruction,
-including the use of high-granularity calorimetry; and the use of
-precision timing detectors.
-
-\hypertarget{tracking-in-high-pile-up-environments}{%
-\subsubsection{Tracking in high pile-up
-environments}\label{tracking-in-high-pile-up-environments}}
-
-The CPU needed for event reconstruction in Run-2 and Run-3 has been
-dominated by charged particle reconstruction (tracking). This is still a
-focus for HL-LHC triggering and event reconstruction, especially when
-the need for efficiently reconstructing low transverse momentum
-particles is considered.
-
-The huge increase in the number of charged particles and hence the
-combinatorial load for tracking algorithms at future colliders will put
-great strain on compute resources. To minimize the memory footprint,
-tracking software needs to be thread-safe to support multi-threaded
-execution per core. At the same time, the software has to be efficient
-and accurate to meet the physics requirements.
-
-Since the CWP, experiments have made progress towards improving software
-tracking efficiency in HL-LHC simulation, see e.g. Ref. {[}2{]}. A
-number of collaboration and community initiatives have been focusing on
-tracking, targeting both offline and online, and we list them below.
-
-The \emph{\textbf{ACTS}} project {[}5{]} is an attempt to encapsulate
-the current ATLAS track reconstruction software into an
-experiment-independent and framework-independent tracking software
-designed to fully exploit modern computing architectures. It builds on
-the tracking experience already obtained at the LHC, and targets the
-HL-LHC as well as future hadron colliders. ACTS provides a set of track
-reconstruction tools designed for parallel architectures, with a
-particular emphasis on thread-safety and concurrent event
-reconstruction. ACTS has active collaborators from other HEP
-experiments, such as FASER, and provides support to Belle-II, EIC, CEPC,
-etc.
-
-The aim of the \textbf{mkFit} project is to speed up Kalman filter (KF)
-tracking algorithms using highly parallel architectures and to deliver
-track building (and possibly fitting) software for the HL-LHC. Recent
-activities focused on delivering a production quality software setup to
-perform track building in the context of the LHC Run-3 (or Run-2). The
-implementation relies on single-precision floating point math and is
-available for multicore CPUs and achieves significant gains in compute
-performance from the use of vector instructions set extending to
-AVX-512, based on Matriplex library (a part of the mkFIt project).
-
-\textbf{Exa.TrkX} is a cross-experiment collaboration of data scientists
-and computational physicists from ATLAS, CMS and DUNE. It develops
-production-quality deep neural network models, in particular Graph
-Neural Networks, for charged particle tracking on diverse detectors
-employing next-generation computing architectures such as HPCs. It is
-also exploring distributed training and optimization of Graph Neural
-Networks (GNN) on HPCs, and the deployment of GNNs with microsecond
-latencies on Level-1 trigger systems.
-
-\hypertarget{addition-of-timing-information-in-reconstruction}{%
-\subsubsection{Addition of timing information in
-reconstruction}\label{addition-of-timing-information-in-reconstruction}}
-
-Physics performance in very high pileup environments, such as the HL-LHC
-or FCC-hh, may also benefit from adding timing information to the
-reconstruction. This allows the mitigation of the effects of pile-up by
-exploiting the time-separation of collision products. {[}cite HGTD/CMS
-timing TPs,
-\href{https://cds.cern.ch/record/2296612}{{https://cds.cern.ch/record/2296612}}
-{]}\\
-The experiment communities have been working on timing detector
-reconstruction and object identification techniques in complex
-environments. Since the CWP, initial tracking and vertexing algorithms
-that include timing information have been developed and incorporated
-into experimental software stacks {[}cite ATLAS/CMS CDRs, CTD
-contribution
-\href{https://indico.cern.ch/event/658267/contributions/2870293/attachments/1620243/2577412/TimingCTD_LindseyGray_20032018.pdf}{{First
-Steps Towards Four-Dimensional Tracking: Timing Layers at the
-HL-LHC}}{]}\footnote{Tracking algorithms where events overlap in time
-  within time slices are employed by the CBM experiment {[}ref
-  \href{https://indico.gsi.de/event/6198/contributions/28511/attachments/20671/26101/Akishina_thesis_corrected.pdf}{{https://indico.gsi.de/event/6198/contributions/28511/attachments/20671/26101/Akishina\_thesis\_corrected.pdf}},
-  \href{https://inspirehep.net/literature/1414007}{\emph{{4-Dimensional
-  Event Building in the First-Level Event Selection of the CBM
-  Experiment - INSPIRE}}}{]}. The use of a cellular automaton makes the
-  algorithm less dependent on the specific detector geometry, and there
-  may be the possibility of cross-talk with LHC experiments that are
-  investigating 4D reconstruction.}.
-
-\hypertarget{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}{%
-\subsection{Enhanced data quality and software monitoring for trigger
-and
-reconstruction}\label{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}}
-
-At HL-LHC, the development, automation, and deployment of extended and
-efficient monitoring tools for software trigger and event reconstruction
-algorithms will be crucial for the success of the experimental physics
-programme.
-
-HEP experiments have extensive continuous integration systems, including
-code regression checks that have enhanced the monitoring procedures for
-software development in recent years. They also have automated
-procedures to check trigger rates as well as the performance of the low-
-and high-level physics objects in the data {[}ref ATLAS DQ and CMS
-DQ{]}.
-
-Since the CWP, experiments have started making limited use of machine
-learning algorithms for anomaly detection {[}ref LHCb RoboShifter:
-\href{https://cds.cern.ch/record/2298467/files/pdf.pdf}{{https://cds.cern.ch/record/2298467/files/pdf.pdf}},
-ref CMS machine learning muons paper - finding out whether it is in
-production?{]}.
-
-\hypertarget{general-reconstruction-software-improvements-vectorization}{%
-\subsubsection{General reconstruction software improvements:
-vectorization}\label{general-reconstruction-software-improvements-vectorization}}
-
-Improving the ability of HEP developed toolkits to use vector units on
-commodity hardware will bring speedups to applications running on both
-current and future hardware. The goal for work in this area is to evolve
-current toolkit and algorithm implementations, and best programming
-techniques to better use SIMD capabilities of current and future
-computing architectures. Since the CWP, algorithm development projects
-(e.g. {[}cite
-\url{https://lhcbproject.web.cern.ch/lhcbproject/Publications/f/p/LHCb-FIGURE-2019-002.html}
-and mkfit{]}) have demonstrated success in increasing the use of vector
-units. In addition, HEP has increasingly benefited from industry
-developed software, including machine learning toolkits (e.g.,
-tensorflow), that typically make excellent use of vector units.
-Challenges remain in this area. For example, full exploitation of this
-must also account for many generations of hardware that experiments must
-exploit on the grid. In addition, to realize full performance gains, a
-large fraction of the application must be improved to use SIMD processor
-capabilities (partly due to turbo-boost capabilities of processors)
-
-\hypertarget{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}{%
-\subsubsection{Use of machine learning for software trigger and
-reconstruction
-algorithms}\label{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}}
-
-It may be desirable or even necessary to deploy new algorithms,
-including advanced machine learning techniques, in order to face the
-increase in event complexity without increasing the per-event
-reconstruction resource needs.
-
-Work is already undergoing in the collaborations to evolve or rewrite
-existing toolkits and algorithms focused on their physics and technical
-performance at high event complexity (e.g. high pileup at HL-LHC), and
-efforts in this area have expanded. Cross-collaboration developments are
-focusing on the availability of ML algorithms for experimental software,
-especially in persistifying models and running inferences in production
-(e.g.
-\emph{\href{https://github.com/lwtnn/lwtnn}{{https://github.com/lwtnn/lwtnn}}
-and \href{https://onnx.ai}{{https://onnx.ai}}}) and in enhancing the
-training capacities of collaborations by submitting jobs to the grid and
-to facilities offering large CPU/GPU resources. One can also foresee
-that HEP will increase the benefit from techniques and cross-talk from
-industry.
-
-\hypertarget{algorithms-and-data-structures-to-efficiently-exploit-many-core-and-non-x86-architectures-e.g.-gpus-fpgas}{%
-\subsubsection{\texorpdfstring{Algorithms and data structures to
-efficiently exploit many-core and non-x86 architectures (e.g. GPUs,
-FPGAs):
-}{Algorithms and data structures to efficiently exploit many-core and non-x86 architectures (e.g.
-GPUs, FPGAs): }}\label{algorithms-and-data-structures-to-efficiently-exploit-many-core-and-non-x86-architectures-e.g.-gpus-fpgas}}
-
-Computing platforms are generally evolving towards having more cores in
-order to increase processing capability. The goal is to evolve current
-event models, toolkits and algorithm implementations, and best
-programming techniques to improve the throughput of multithreaded
-software trigger and event reconstruction applications. Additionally,
-computing architectures using technologies beyond CPUs offer an
-interesting alternative for increasing the throughput of the most time
-consuming trigger or reconstruction algorithms.
-
-Since the CWP, HEP now has a number of algorithmic projects that are
-designed for, or have successfully adapted to, hardware accelerators,
-most notably to NVidia GPUs using CUDA. Two projects targeting Run-3 and
-with prospects for use in Run-4 are Patatrack (from CMS) {[}3{]} and
-Allen (from LHCb) {[}4{]}.
-
-\textbf{Patatrack} is a CMS initiative aimed to use heterogeneous
-computing for the charged particle reconstruction. Within the CMS
-codebase (CMSSW), this project has demonstrated that physics
-reconstruction code can be written to leverage heterogeneous
-architectures, like GPUs, to achieve a significant speed-up while
-reducing the overall cost and power consumption of a system. An example
-is the CMS Pixel local reconstruction and the track and vertex
-reconstruction: these algorithms can run on an NVIDIA T4 GPU with the
-same performance as 52-56 Xeon cores, at a fraction of the cost and
-power consumption of an equivalent CPU-only system; further improvements
-may come pairing one or more high end GPUs with a low power ARM system.
-
-\textbf{Allen} is a data processing framework for GPUs, as well as a
-specific implementation of a first-level trigger for LHCb Run 3
-data-taking. Allen is optimized for sustaining the rate required by
-real-time processing in LHCb, but can be used as a more general GPU data
-processing as it includes a scheduler and a memory manager that can be
-integrated into Gaudi. This permits the extension and implementation of
-Allen within other experimental software using the same underlying
-framework.
-
-ATLAS is also investigating GPU solutions for reconstruction software
-and its acceleration at HL-LHC, in particular in terms of highly
-parallel algorithms {[}3a{]} and to enable the use of machine learning
-algorithms in object reconstruction and identification {[}\emph{waiting
-for an email from ATLAS with references}{]}.
-
-The experience acquired with these projects shows that GPUs can be used
-for increasing the throughput in cases of large-size (ALICE) {[}cite
-Alice O2{]} and small-size events per second (LHCb), and to speed up
-specific parts of the data processing (CMS, Patatrack).
-
-It is clear that the technology is improving rapidly in this area. The
-adoption of portability toolkits is needed to avoid vendor lock-in
-and/or the need to evolve algorithms by hand to adapt to each new type
-of computing architecture, as well as to increase the sustainability of
-the codebase avoiding multiple implementations of the same algorithm.
-
-Heterogeneous computing is still a rapidly developing and changing
-field. The burden of writing architecture-specific code repeatedly will
-certainly exceed the available effort from the experiments. Efforts are
-ongoing in this direction, since the CWP, trying to take advantage of
-the latest development in synergy with industry and attempting to
-increase the use of centers with heterogeneous resources by experiments.
-
-The \textbf{HLS4ML} project {[}7{]} targets the implementation of
-machine learning algorithms on FPGA for low-latency applications useful
-for e.g. Level-1 triggers. The aim of the HLS4ML project is the
-translation of trained ML models into FPGA firmware. HLS4ML also extends
-to the production of coprocessor kernels for FPGAs for longer latency
-applications, and can be used as a tool to design AI-powered ASICs. The
-\textbf{SONIC} application {[}8{]} is being developed in parallel, with
-the goal of facilitating and accelerating the inference of deep neural
-networks for triggering, reconstruction, and analysis, by providing
-software to use heterogeneous computing resources as a service targeting
-next-generation facilities at the energy and intensity frontier (HL-LHC,
-LBNF).
-
-\hypertarget{general-remarks-on-trigger-and-reconstruction-software-sharing}{%
-\subsection{General remarks on trigger and reconstruction software
-sharing}\label{general-remarks-on-trigger-and-reconstruction-software-sharing}}
-
-Nearly all software solutions presented in this chapter are tailored to
-a specific experiment, and would require additional work to be adapted
-to different environments. Nevertheless, it is still useful to maintain
-open communication channels to discuss design choices and compare
-performance assessments of different solutions to guide future software
-and reconstruction design choices.
-
-The use of open-source elements of the LHC software stack for trigger
-and reconstruction by smaller experiments, especially in case of common
-experimental design\footnote{An example of software sharing happens in
-  the case of the FASER experiment, whose offline software
-  (\href{https://gitlab.cern.ch/faser/calypso}{{https://gitlab.cern.ch/faser/calypso}})
-  is based on a derivative of the ATLAS Athena open-source software
-  {[}ref{]}, which in turn uses the Gaudi framework {[}ref{]}. FASER
-  also shares tracking detector components with ATLAS, facilitating the
-  adoption of software relying on similar detector descriptions.}, is
-still worthwhile This benefits experiments with a smaller developer base
-than LHC experiments, and increases the return on investment for LHC
-experiment software. For this reason, we also encourage the addition of
-common open software projects to the ESCAPE software catalogue {[}ref
-ESCAPE{]}.
-
-\subsection{Reconstruction and Software Trigger References}\label{references-rst}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TO BE CONVERTED TO BibTeX, but left here for "working" %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\emph{{[}1{]}
-\href{https://arxiv.org/abs/1802.08640}{{https://arxiv.org/abs/1802.08640}}}
-
-\emph{{[}2{]}
-\href{https://cds.cern.ch/record/2693670}{{https://cds.cern.ch/record/2693670}}}
-
-\emph{{[}3{]}} - Andrea Bocci \emph{at al.}, ``\emph{Heterogeneous
-online reconstruction at CMS}'', presented at CHEP 2019\\
-- Andrea Bocci \emph{at al.}, ``\emph{Heterogeneous reconstruction:
-combining ARM processors with GPUs}'', presented at CHEP 2019
-
-{[}3a{]}
-\href{https://indico.cern.ch/event/773049/contributions/3473268/attachments/1937520/3211376/GPU_Usage_in_ATLAS_Reconstruction_and_Analysis_2019.11.04..pdf}{{https://indico.cern.ch/event/773049/contributions/3473268/attachments/1937520/3211376/GPU\_Usage\_in\_ATLAS\_Reconstruction\_and\_Analysis\_2019.11.04..pdf}}
-
-\emph{{[}4{]}}
-\href{https://arxiv.org/abs/1912.09161}{{https://arxiv.org/abs/1912.09161}}
-
-\emph{{[}5{]}}
-\href{http://dx.doi.org/10.1088/1742-6596/898/4/042011}{{http://dx.doi.org/10.1088/1742-6596/898/4/042011}},
-\href{https://arxiv.org/abs/1910.03128}{{https://arxiv.org/abs/1910.03128}},
-The ACTS project: track reconstruction software for HL-LHC and beyond,
-Proceedings of CHEP2019, in progress
-
-{[}6{]} Xiangyang Ju et al, "Graph Neural Networks for Particle
-Reconstruction in High Energy Physics Detectors"
-\href{https://arxiv.org/abs/2003.11603}{{(arxiv link}})
-(\href{https://github.com/exatrkx/exatrkx-neurips19}{{Software and
-documentation}})
-
-{[}7{]} J. Duarte \emph{et al}, ``Fast inference of deep neural networks
-in FPGAs for particle physics'', \emph{JINST} \textbf{13} P07027, 2018,
-doi:
-\href{https://iopscience.iop.org/article/10.1088/1748-0221/13/07/P07027/meta}{{10.1088/1748-0221/13/07/P07027}}\\
-S. Summers \emph{et al}, ``Fast inference of Boosted Decision Trees in
-FPGAs for particle physics'', 2020, {[}Online{]} available:
-\href{https://arxiv.org/abs/2002.02534}{{arXiv:2002.02534}}\\
-G. Gi Guglielmo \emph{et al}, ``Compressing deep neural networks on
-FPGAs to binary and ternary precision with HLS4ML'', 2020, {[}Online{]}
-available: \href{https://arxiv.org/abs/2003.06308}{{arXiv:2003.06308}}
-
-{[}8{]} SONIC:
-\href{https://arxiv.org/abs/1904.08986}{{https://arxiv.org/abs/1904.08986}}
+\input{input_recoTrigger}
 

### 2020-04-28 LHCC/2020/2020-01/input_recoTrigger.tex
Snapshot, trigger and reco properly referenced
Was moved to a separate input file to avoid editing mistakes

\hypertarget{reconstruction-and-software-triggers}{%
\section{Reconstruction and Software
Triggers}\label{reconstruction-and-software-triggers}}

Software trigger and event reconstruction techniques in HEP face a
number of new challenges in the next decade. Advances in facilities and
future experiments bring a dramatic increase in physics reach, as well
as increased event complexity and rates.

At the HL-LHC, the central challenge for high-level triggers (e.g.,
software triggers) and object reconstruction is to maintain excellent
efficiency and resolution in the face of high pileup values, especially
at low transverse momenta. Detector upgrades such as increases in
channel density, high precision timing and improved detector layouts are
essential to overcome these problems. The subsequent increase of event
complexity at the HL-LHC also requires the development of software
algorithms that can process events with a similar cost per event to
Run-2 and Run-3. At the same time, algorithmic approaches need to
effectively take advantage of evolutions in computing technologies,
including increased SIMD capabilities, increased core counts in CPUs,
and heterogeneous hardware.

This section focuses on the challenges identified in the Community White
Paper~\cite{Alves:2017she,albrecht2018hepexec,albrecht2018hep} and the development of solutions that. It also includes
mentions of open source software efforts that have developed within or
outside LHC collaborations that could be adapted by the LHC community to
improve trigger and reconstruction algorithms.

\hypertarget{the-evolution-of-triggers-and-real-time-analysis}{%
\subsection{\texorpdfstring{The evolution of triggers and real-time
analysis
}{The evolution of triggers and real-time analysis }}\label{the-evolution-of-triggers-and-real-time-analysis}}

Trigger systems are evolving to be more capable, both in their ability
to select a wider range of events of interest for the physics program of
their experiment, and their ability to stream a larger rate of events
for further processing. The event rates that will be processed by
experiments at the HL-LHC will increase by up to a factor 10 with
respect to Run-3, owing to upgraded trigger systems and expanded physics
programs. ATLAS and CMS plan to maintain a two-tiered trigger system,
where a hardware trigger makes use of coarse event information to reduce
the event rate to 10x over the current capability, up to 1 MHz \cite{ATLAS-TDR-29,collaboration:2714892}. The high level trigger system, implemented in software, selects
up to 100 kHz of events to be saved in full for subsequent
analysis\footnote{LHCb \cite{Aaij:2019uij} and ALICE \cite{Buncic:2011297} will both
  stream the full collision rate to real-time or quasi-realtime software
  trigger systems.}.

Experiments have also been working towards minimizing the differences
between trigger (online) and offline software for reconstruction and
calibration, so that more refined physics objects can be obtained
directly within the HLT farm for a more efficient event selection. This
is also in-line with enhancing the experiment's capabilities for
real-time data analysis of events accepted by the hardware trigger
system, driven by use cases where the physics potential improves when
analysing more events than what can be written out in full with
traditional data processing.

Implementations of real-time analysis systems, where raw data is
processed into its final form as close as possible to the detector (e.g.
in the high-level trigger farm), are in use within several experiments.
These approaches remove the detector data that typically makes up the
raw data tier kept for offline reconstruction, and keep only a limited
set of analysis objects reconstructed within the high level trigger.

Real-time analysis techniques are being adopted to enable a wider range
of physics signals to be saved by the trigger for final analysis. As
rates increase, these techniques can become more important and
widespread by enabling only the parts of an event associated with the
signal candidates to be saved, reducing the required disk space. The
experiments are focusing on the technological developments that make it
possible to do this with acceptable reduction of the analysis
sensitivity and with minimal biases. The HSF trigger and reconstruction
group has been encouraging cross-talk between LHC experiments and beyond
on real-time analysis, as these kinds of non-standard trigger and
reconstruction workflows increase the physics output for selected
physics cases without adding significant overhead to the current
resources.

Active research topics include compression and custom data formats;
toolkits for real-time detector calibration and validation which will
enable full offline analysis chains to be ported into real-time; and
frameworks which will enable non-expert offline analysts to design and
deploy real-time analyses without compromising data taking quality.
Use-cases for real-time analysis techniques have expanded during the
last years of Run 2 and their use appears likely to
grow already during Run 3 data taking for all experiments owing to
improvements in the HLT software and farms \textbf{Cite CMS and ATLAS CDRs}. Further ideas include retaining very reduced data from as much
of the data taking rate as possible from selected regions and
sub-detectors, even up to the 40 MHz bunch crossing rate~\cite{hannes_sakulin_2019_3598769}.

\hypertarget{challenges-and-improvements-foreseen-in-event-reconstruction}{%
\subsection{Challenges and improvements foreseen in event
reconstruction}\label{challenges-and-improvements-foreseen-in-event-reconstruction}}

Processing and reducing raw data into analysis-level formats, event
reconstruction, is a major component of offline computing resource
needs, and in light of the previous section it is relevant for online
resources as well. This is an essential step towards precision
reconstruction, identification and measurement of physics-objects at
HL-LHC.

Algorithmic areas of particular importance at HL-LHC experiments are
charged-particle trajectory reconstruction (\textit{tracking}), including
hardware triggers based on tracking information which may seed later
software trigger and reconstruction algorithms; jet reconstruction,
including the use of high-granularity calorimetry; and the use of
precision timing detectors.

\hypertarget{tracking-in-high-pile-up-environments}{%
\subsubsection{Tracking in high pile-up
environments}\label{tracking-in-high-pile-up-environments}}

The CPU needed for event reconstruction in Run-2 and Run-3 has been
dominated by charged particle reconstruction (tracking). This is still a
focus for HL-LHC triggering and event reconstruction, especially when
the need for efficiently reconstructing low transverse momentum
particles is considered.

The huge increase in the number of charged particles and hence the
combinatorial load for tracking algorithms at future colliders will put
great strain on compute resources. To minimize the memory footprint,
tracking software needs to be thread-safe to support multi-threaded
execution per core. At the same time, the software has to be efficient
and accurate to meet the physics requirements.

Since the CWP, experiments have made progress towards improving software
tracking efficiency in HL-LHC simulation, see e.g. Ref. \cite{ATL-PHYS-PUB-2019-041}. A
number of collaboration and community initiatives have been focusing on
tracking, targeting both offline and online, and we list them below.

The \emph{\textbf{ACTS}} project \cite{Ai:2019kze,Gumpert_2017} is an attempt to encapsulate
the current ATLAS track reconstruction software into an
experiment-independent and framework-independent tracking software
designed to fully exploit modern computing architectures. It builds on
the tracking experience already obtained at the LHC, and targets the
HL-LHC as well as future hadron colliders. ACTS provides a set of track
reconstruction tools designed for parallel architectures, with a
particular emphasis on thread-safety and concurrent event
reconstruction. ACTS has active collaborators from other HEP
experiments, such as FASER, and provides support to Belle-II, EIC, CEPC,
etc.

The aim of the \textbf{mkFit} project is to speed up Kalman filter (KF)
tracking algorithms using highly parallel architectures and to deliver
track building (and possibly fitting) software for the HL-LHC. Recent
activities focused on delivering a production quality software setup to
perform track building in the context of the LHC Run-3 (or Run-2). The
implementation relies on single-precision floating point math and is
available for multicore CPUs and achieves significant gains in compute
performance from the use of vector instructions set extending to
AVX-512, based on Matriplex library (a part of the mkFIt project).

\textbf{Exa.TrkX}~\cite{Ju:2020xty} is a cross-experiment collaboration of data scientists
and computational physicists from ATLAS, CMS and DUNE. It develops
production-quality deep neural network models, in particular Graph
Neural Networks, for charged particle tracking on diverse detectors
employing next-generation computing architectures such as HPCs. It is
also exploring distributed training and optimization of Graph Neural
Networks (GNN) on HPCs, and the deployment of GNNs with microsecond
latencies on Level-1 trigger systems.

\hypertarget{addition-of-timing-information-in-reconstruction}{%
\subsubsection{Addition of timing information in
reconstruction}\label{addition-of-timing-information-in-reconstruction}}

Physics performance in very high pileup environments, such as the HL-LHC
or FCC-hh, may also benefit from adding timing information to the
reconstruction. This allows the mitigation of the effects of pile-up by
exploiting the time-separation of collision products~\cite{Collaboration:2623663,CMS:2667167}\\
The experiment communities have been working on timing detector
reconstruction and object identification techniques in complex
environments. Since the CWP, initial tracking and vertexing algorithms
that include timing information have been developed and incorporated
into experimental software stacks \cite{LindseyGreyMIPCTD}\footnote{Tracking algorithms where events overlap in time
within time slices are employed by the CBM experiment~\cite{Akishina:2015ghv,AkishinaThesis}. The use of a cellular automaton makes the
  algorithm less dependent on the specific detector geometry, and there
  may be the possibility of cross-talk with LHC experiments that are
  investigating 4D reconstruction.}.

\hypertarget{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}{%
\subsection{Enhanced data quality and software monitoring for trigger
and
reconstruction}\label{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}}

At HL-LHC, the development, automation, and deployment of extended and
efficient monitoring tools for software trigger and event reconstruction
algorithms will be crucial for the success of the experimental physics
programme.

HEP experiments have extensive continuous integration systems, including
code regression checks that have enhanced the monitoring procedures for
software development in recent years. They also have automated
procedures to check trigger rates as well as the performance of the low-
and high-level physics objects in the data {[}ref ATLAS DQ and CMS
DQ{]}.

Since the CWP, experiments have started making limited use of machine
learning algorithms for anomaly detection \cite{Adinolfi:2298467,CMSMonitoring}.

\hypertarget{general-reconstruction-software-improvements-vectorization}{%
\subsubsection{General reconstruction software improvements:
vectorization}\label{general-reconstruction-software-improvements-vectorization}}

Improving the ability of HEP developed toolkits to use vector units on
commodity hardware will bring speedups to applications running on both
current and future hardware. The goal for work in this area is to evolve
current toolkit and algorithm implementations, and best programming
techniques to better use SIMD capabilities of current and future
computing architectures. Since the CWP, algorithm development projects
(e.g. \cite{LHCB-FIGURE-2019-002}
as well as mkFit) have demonstrated success in increasing the use of vector
units. In addition, HEP has increasingly benefited from industry
developed software, including machine learning toolkits (e.g.,
tensorflow), that typically make excellent use of vector units.
Challenges remain in this area. For example, full exploitation of this
must also account for many generations of hardware that experiments must
exploit on the grid. In addition, to realize full performance gains, a
large fraction of the application must be improved to use SIMD processor
capabilities (partly due to turbo-boost capabilities of processors)

\hypertarget{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}{%
\subsubsection{Use of machine learning for software trigger and
reconstruction
algorithms}\label{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}}

It may be desirable or even necessary to deploy new algorithms,
including advanced machine learning techniques, in order to face the
increase in event complexity without increasing the per-event
reconstruction resource needs.

Work is already undergoing in the collaborations to evolve or rewrite
existing toolkits and algorithms focused on their physics and technical
performance at high event complexity (e.g. high pileup at HL-LHC), and
efforts in this area have expanded. Cross-collaboration developments are
focusing on the availability of ML algorithms for experimental software,
especially in persistifying models and running inferences in production
(e.g.
\emph{\href{https://github.com/lwtnn/lwtnn}{{https://github.com/lwtnn/lwtnn}}
and \href{https://onnx.ai}{{https://onnx.ai}}}) and in enhancing the
training capacities of collaborations by submitting jobs to the grid and
to facilities offering large CPU/GPU resources. One can also foresee
that HEP will increase the benefit from techniques and cross-talk from
industry.

\hypertarget{algorithms-and-data-structures-to-efficiently-exploit-many-core-and-non-x86-architectures-e.g.-gpus-fpgas}{%
\subsubsection{\texorpdfstring{Algorithms and data structures to
efficiently exploit many-core and non-x86 architectures (e.g. GPUs,
FPGAs):
}{Algorithms and data structures to efficiently exploit many-core and non-x86 architectures (e.g. GPUs, FPGAs): }}\label{algorithms-and-data-structures-to-efficiently-exploit-many-core-and-non-x86-architectures-e.g.-gpus-fpgas}}

Computing platforms are generally evolving towards having more cores in
order to increase processing capability. The goal is to evolve current
event models, toolkits and algorithm implementations, and best
programming techniques to improve the throughput of multithreaded
software trigger and event reconstruction applications. Additionally,
computing architectures using technologies beyond CPUs offer an
interesting alternative for increasing the throughput of the most time
consuming trigger or reconstruction algorithms.

Since the CWP, HEP now has a number of algorithmic projects that are
designed for, or have successfully adapted to, hardware accelerators,
most notably to NVidia GPUs using CUDA. Two projects targeting Run-3 and
with prospects for use in Run-4 are Patatrack (from CMS) \cite{andrea_bocci_2019_3598824} and
Allen (from LHCb) \cite{Aaij:2019zbu}.

\textbf{Patatrack} is a CMS initiative aimed to use heterogeneous
computing for the charged particle reconstruction. Within the CMS
codebase (CMSSW), this project has demonstrated that physics
reconstruction code can be written to leverage heterogeneous
architectures, like GPUs, to achieve a significant speed-up while
reducing the overall cost and power consumption of a system. An example
is the CMS Pixel local reconstruction and the track and vertex
reconstruction: these algorithms can run on an NVIDIA T4 GPU with the
same performance as 52-56 Xeon cores, at a fraction of the cost and
power consumption of an equivalent CPU-only system; further improvements
may come pairing one or more high end GPUs with a low power ARM system.

\textbf{Allen} is a data processing framework for GPUs, as well as a
specific implementation of a first-level trigger for LHCb Run 3
data-taking. Allen is optimized for sustaining the rate required by
real-time processing in LHCb, but can be used as a more general GPU data
processing as it includes a scheduler and a memory manager that can be
integrated into Gaudi. This permits the extension and implementation of
Allen within other experimental software using the same underlying
framework.

ATLAS is also investigating GPU solutions for reconstruction software
and its acceleration at HL-LHC, in particular in terms of highly
parallel algorithms \cite{attila_krasznahorkay_2019_3599103} and to enable the use of machine learning
algorithms in object reconstruction and identification {[}\emph{waiting
for an email from ATLAS with references}{]}.

The experience acquired with these projects shows that GPUs can be used
for increasing the throughput in cases of large-size (ALICE) \cite{david_rohr_2019_3599418} and small-size events per second (LHCb), and to speed up
specific parts of the data processing (CMS, Patatrack).

It is clear that the technology is improving rapidly in this area. The
adoption of portability toolkits is needed to avoid vendor lock-in
and/or the need to evolve algorithms by hand to adapt to each new type
of computing architecture, as well as to increase the sustainability of
the codebase avoiding multiple implementations of the same algorithm.

Heterogeneous computing is still a rapidly developing and changing
field. The burden of writing architecture-specific code repeatedly will
certainly exceed the available effort from the experiments. Efforts are
ongoing in this direction, since the CWP, trying to take advantage of
the latest development in synergy with industry and attempting to
increase the use of centers with heterogeneous resources by experiments.

The \textbf{HLS4ML} project \cite{Duarte:2018ite,Summers:2020xiy,DiGuglielmo:2020eqx} targets the implementation of
machine learning algorithms on FPGA for low-latency applications useful
for e.g. Level-1 triggers. The aim of the HLS4ML project is the
translation of trained ML models into FPGA firmware. HLS4ML also extends
to the production of coprocessor kernels for FPGAs for longer latency
applications, and can be used as a tool to design AI-powered ASICs. The
\textbf{SONIC} application \cite{Duarte:2019fta} is being developed in parallel, with
the goal of facilitating and accelerating the inference of deep neural
networks for triggering, reconstruction, and analysis, by providing
software to use heterogeneous computing resources as a service targeting
next-generation facilities at the energy and intensity frontier (HL-LHC,
LBNF).

\hypertarget{general-remarks-on-trigger-and-reconstruction-software-sharing}{%
\subsection{General remarks on trigger and reconstruction software
sharing}\label{general-remarks-on-trigger-and-reconstruction-software-sharing}}

Nearly all software solutions presented in this chapter are tailored to
a specific experiment, and would require additional work to be adapted
to different environments. Nevertheless, it is still useful to maintain
open communication channels to discuss design choices and compare
performance assessments of different solutions to guide future software
and reconstruction design choices.

The use of open-source elements of the LHC software stack for trigger
and reconstruction by smaller experiments, especially in case of common
experimental design\footnote{An example of software sharing happens in
  the case of the FASER experiment, whose offline software
  (\href{https://gitlab.cern.ch/faser/calypso}{{https://gitlab.cern.ch/faser/calypso}})
  is based on a derivative of the ATLAS Athena open-source software, which in turn uses the Gaudi framework. FASER
  also shares tracking detector components with ATLAS, facilitating the
  adoption of software relying on similar detector descriptions.}, is
still worthwhile This benefits experiments with a smaller developer base
than LHC experiments, and increases the return on investment for LHC
experiment software. For this reason, we also encourage the addition of
common open software projects to the \href{https://projectescape.eu/services/open-source-scientific-software-and-service-repository}{ESCAPE software catalogue}.

### 2020-04-30 LHCC/2020/2020-01/ctcs.tex
References converted to BibTeX
Some additional minor fixes

@@
 \linenumbers
+
+\usepackage{comment}
 
@@
 continued development and optimisation will be critical in the future
-\cite{stewart_graeme_a_2018_2413005, Ellis:2691414}. The
+\cite{stewart_graeme_a_2018_2413005, Ellis:2691414, Albrecht2019}. The
 challenges are driven by an ambitious physics programme, notably the LHC
@@
 specialised architectures, such as TPUs, IPUs and developments that make
-FPGA devices more user friendly. These pose huge challenges for our
+FPGA devices more user friendly \cite{acm10.1145/3282307}.
+These pose huge challenges for our
 community as the programming models and APIs vary widely here and
@@
 software and GPU codes have been used in production by some experiments
-for years. In other areas migration is still ongoing. In others it is
+for years~\cite{calafiura2018hep, albrecht2018hepexec}.
+In other areas migration is still ongoing. In others it is
 not even yet started. Generic heterogeneous programming models have
@@
 application-dependent. HEP itself will not drive the success of one
-model or another, so even if the community converged on a prefered
+model or another, so even if the community converged on a preferred
 model, its long term success would not be assured. The C++ standard
@@
 in HEP, e.g. developments like RNTuple for ROOT I/O is likely to be of
-great importance for the community. The likelihood of finding an
+great importance for the community \cite{ROOT-2020-HL-LHC}. The likelihood of finding an
 off-the-shelf compact and efficient storage format for HEP data is
@@
 where the I/O layer must optimise data transfers and hide latency, e.g.
-taking advantage of xrootd's single column streaming ability.
+taking advantage of xrootd's single column streaming ability \cite{xrootd}.
 
@@
 architectures) and sharing best practice for code development can have a
-large integrated benefit. This requires improved training and the
+large integrated benefit \cite{couturier2017hep}. This requires improved training and the
 development of a curriculum for all developer levels. In the field, such
@@
 training contributes to their career development and a long term future
-in the field. HEP software stacks are already deep and wide and building
+in the field \cite{foundation2018hep}. HEP software stacks are already deep and wide and building
 these consistently and coherently is also an area where knowledge can be
@@
 production workflows can run. CernVM-FS was a huge leap forward for
-software distribution and has even been widely adopted outside HEP.
+software distribution and has even been widely adopted outside HEP 
+\cite{Blomer_2011, 7310920}.
 However, new challenges arise, with container based payloads, scaling
@@
 confirmed and updated as the field moves on. There are many exciting
-developments based around CernVM-FS, containers and things like analysis
-description languages, but these are not yet at the stage of being
+developments based around CernVM-FS
+\cite{10.1007/978-3-319-67630-2_52, serverless-cvmfs}, containers and things like analysis
+description languages \cite{Ref10}, but these are not yet at the stage of being
 settled nor integrated into our day-to-day workflows.
@@
 
-A HSF Working Group (WG) on generators {[}Gen19{]} was set up at the
+A HSF Working Group (WG) on generators~\cite{Gen19} was set up at the
 beginning of 2019. The main focus of the WG so far has been on gaining a
@@
 particular, the WG has been active in analysing the ATLAS and CMS
-compute budgets in detail, in profiling MG5\_aMC {[}Alw14{]}, Sherpa
-{[}Bot19{]} and Powheg {[}Fri17{]}, in discussing the possible sharing
+compute budgets in detail, in profiling MG5\_aMC~\cite{Alw14}, 
+Sherpa~\cite{Bot19} and Powheg~\cite{Fri07}, in discussing the possible sharing
 of common parton-level samples by ATLAS and CMS, and in reviewing and
@@
 contribution, including a more complete list of references, will soon be
-uploaded on arxiv and submitted for publication {[}Gen20{]}.
+uploaded on arxiv and submitted for publication~\cite{Gen20}.
 
@@
 critical ingredient for efficient unweighted event generation. Many
-generic algorithms exist (e.g. VEGAS {[}Lep80{]}, BASES / SPRING
-{[}Kaw86{]}, MINT {[}Nas07{]}, FOAM {[}Jad03{]}), as well as others
-developed specifically for a given generator (e.g. MadEvent {[}Mal03{]},
-itself based on a modified version of VEGAS, in MG5\_aMC, or COMIX
-{[}Gle08{]} in Sherpa). In general, the larger the dimensionality of the
+generic algorithms exist (e.g. VEGAS~\cite{Lep80}, BASES / SPRING~\cite{Kaw86}, 
+MINT~\cite{Nas07}, FOAM~\cite{Jad03}), as well as others
+developed specifically for a given generator (e.g. MadEvent~\cite{Mal03},
+itself based on a modified version of VEGAS, in MG5\_aMC, or COMIX~\cite{Gle08} 
+in Sherpa). In general, the larger the dimensionality of the
 phase space, the lower the unweighting efficiency that can be achieved:
 in W+jets, for instance, the Sherpa efficiency is 30\% for W+0 jets and
-0.08\% for W+3jets {[}Gao20{]}. This is an area where research is very
+0.08\% for W+3jets~\cite{Gao20}. This is an area where research is very
 active, and should be actively encouraged, as significant cost
@@
 knowledge of phase space peaks, but they can be complemented by
-brute-force machine learning (ML) algorithmic methods {[}Ben17, Bot20,
-Gao20, Kli18{]}.
-
-- {Merging inefficiency}. Merging prescriptions (e.g. MLM {[}Man02{]},
-CKKW-L {[}Lon05{]} at LO and FxFx {[}Fre12{]}, MEPS@NLO {[}Hoe14{]} at
+brute-force machine learning (ML) algorithmic methods~\cite{Ben17,Bot20,Gao20,Kli18}.
+
+- {Merging inefficiency}. Merging prescriptions (e.g. MLM~\cite{Man02},
+CKKW-L~\cite{Lon02} at LO and FxFx~\cite{Fre12}, MEPS@NLO~\cite{Hoe14} at
 NLO) imply the rejection of some events, to avoid double counting
@@
 are unavoidable in the algorithmic strategy used by the underlying
-physics modeling {[}Alw08{]}. However, a method like shower-kt MLM can
-reduce the merging inefficiency of MLM {[}Alw09{]}.
+physics modeling~\cite{Alw08}. However, a method like shower-kt MLM can
+reduce the merging inefficiency of MLM~\cite{Alw09}.
 
@@
 calculations) and could be explored further, though this may require
-more theoretical work for samples involving merging or NLO matching
-{[}Mat16{]}.
+more theoretical work for samples involving merging or NLO matching~\cite{Mat16}.
 
@@
 with r=25\% (which may be regarded as a worst-case scenario in top quark
-pair production {[}Fre20{]}), one needs to generate 4 times as many
+pair production~\cite{Fre20}), one needs to generate 4 times as many
 events. Negative weights can instead be almost completely avoided, by
@@
 is only available for a limited number of processes and describes the
-relevant physics to a different degree of precision than MC@NLO (see
-{[}Fre20{]} for a more in-depth discussion). Progress in this area can
+relevant physics to a different degree of precision than MC@NLO 
+(see~\cite{Fre20} for a more in-depth discussion). Progress in this area can
 only be achieved by theorists: for instance, a new approach for
 MC@NLO-type matching with reduced negative weights has recently been
-proposed {[}Fre20{]}. It should be noted that negative weights can also
+proposed~\cite{Fre20}. It should be noted that negative weights can also
 happen at LO because of not-positive-definite parton density function
@@
 production. Detailed profiling could also be useful to assess the CPU
-cost of external parton distribution function libraries {[}Kon20{]}, or
+cost of external parton distribution function libraries~\cite{Kon20}, or
 the memory requirements of the software (which may motivate a move to
@@
 generator software would certainly be needed to modernise it and make it
-more efficient, or even port it to more modern computing architectures
-{[}Bau13, Alb19{]}:
+more efficient, or even port it to more modern computing 
+architectures~\cite{Bau13,Alves:2017she}:
 
@@
 has been successfully tested and benchmarked on HPC architectures using
-both Alpgen {[}Chi17{]} and Sherpa {[}Hoe19{]}; in the latter case, new
+both Alpgen~\cite{Chi17} and Sherpa~\cite{Hoe19}; in the latter case, new
 event formats based on HDF5, replacing LHEF, have also been instrumental
 in the success of these tests. MPI integration has also been completed
-for MG5\_aMC {[}Mat18{]}. It should also be noted that, even if HPCs
+for MG5\_aMC~\cite{Mat18}. It should also be noted that, even if HPCs
 offer extremely high-speed inter-node connectivity, it is perfectly ok
@@
 achievable by more generic optimizations, not involving concurrency. It
-should be studied, for instance, if data caching {[}Kon20{]} or
+should be studied, for instance, if data caching~\cite{Kon20} or
 different compilers and build strategies may lead to any improvements.
@@
 
-\subsection{Generator References}\label{references}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TO BE CONVERTED TO BibTeX, but left here for "working" %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-{[}Alb19{]} J. Albrecht et al. ``A Roadmap for HEP Software and
-Computing R\&D for the 2020s'' Comput Softw Big Sci (2019) 3, 7,
-\href{https://doi.org/10.1007/s41781-018-0018-8}{{doi:10.1007/s41781-018-0018-8}},
-\href{https://arxiv.org/abs/1712.06982}{{arXiv:1712.06982}}
-
-{[}Alw08{]} J. Alwall et al. ``Comparative study of various algorithms
-for the merging of parton showers and matrix elements in hadronic
-collisions'', Eur. Phys. J. C 53 (2008) 53,
-\href{https://doi.org/10.1140/epjc/s10052-007-0490-5}{{doi:10.1140/epjc/s10052-007-0490-5}}
-
-{[}Alw09{]} J. Alwall, S. de Visscher, F. Maltoni, ``QCD radiation in
-the production of heavy colored particles at the LHC'', JHEP 0902 (2009)
-017,
-\href{https://doi.org/10.1088/1126-6708/2009/02/017}{{doi:10.1088/1126-6708/2009/02/017}}
-
-{[}Alw14{]} J. Alwall et al. ``The automated computation of tree-level
-and next-to-leading order differential cross sections, and their
-matching to parton shower simulations'', JHEP 07 (2014) 079,
-\href{https://doi.org/10.1007/JHEP07(2014)079}{{doi:10.1007/JHEP07(2014)079}}
-
-{[}Bau13{]} C. Bauer et al., "Computing for perturbative QCD - a
-Snowmass White Paper",
-\href{https://arxiv.org/abs/1309.3598}{{arxiv:1309.3598}}
-
-{[}Ben17{]} J. Bendavid, ``Efficient Monte Carlo Integration Using
-Boosted Decision Trees and Generative Deep Neural Networks'',
-\href{https://arxiv.org/abs/1707.00028}{{arXiv:1707.00028}}
-
-{[}Bot19{]} E. Bothmann et al., ``Event generation with Sherpa 2.2'',
-\href{https://scipost.org/SciPostPhys.7.3.034}{{SciPost Phys. 7 (2019)
-034}}
-
-{[}Bot20{]} E. Bothmann et al. ``Exploring phase space with Neural
-Importance Sampling'',
-\href{https://arxiv.org/abs/2001.05478}{{arXiv:2001.05478}}
-
-{[}Chi17{]} J. T. Childers et al., "Adapting the serial Alpgen
-parton-interaction generator to simulate LHC collisions on millions of
-parallel threads", Comp. Phys. Comm. 210 (2017) 5154.
-\href{https://doi.org/10.1016/j.cpc.2016.09.0135286}{{https://doi.org/10.1016/j.cpc.2016.09.0135286}}.
-
-{[}Fre12{]} R. Frederix, S. Frixione, "Merging meets matching in
-MC@NLO", JHEP 2012 (2012) 61.
-\href{https://doi.org/10.1007/JHEP12(2012)061}{{doi:10.1007/JHEP12(2012)061}}
-
-{[}Fre20{]} R. Frederix, S. Frixione, S. Prestel. P. Torrielli, ``On the
-reduction of negative weights in MC@NLO-type matching procedures'',
-\href{https://arxiv.org/abs/2002.12716}{{arXiv:2002.12716}}
-
-{[}Fri17{]} S. Frixione, P. Nason, C. Oleari, "Matching NLO QCD
-computations with parton shower simulations: the POWHEG method", JHEP11
-(2007) 070,
-\href{https://doi.org/10.1088/1126-6708/2007/11/070}{{doi:10.1088/1126-6708/2007/11/070}}
-
-{[}Gao20{]} C. Gao. et al. ``Event Generation with Normalizing Flows'',
-\href{https://arxiv.org/abs/2001.10028}{{arXiv:2001.10028}}
-
-{[}Gen18{]} HSF Physics Event Generator Computing Workshop, CERN,
-November 2018,
-\href{https://indico.cern.ch/event/751693}{{https://indico.cern.ch/event/751693}}
-
-{[}Gen19{]} HSF Physics Generators Working Group,
-\href{https://hepsoftwarefoundation.org/workinggroups/generators.html}{{https://hepsoftwarefoundation.org/workinggroups/generators.html}}
-
-{[}Gen20{]} HSF Physics Generators Working Group, "Challenges in Monte
-Carlo event generator software for High-Luminosity LHC", CERN-LPCC-2020-002 (2020).
-
-{[}Gle08{]} T. Gleisberg, S. Hoeche, "Comix, a new matrix element
-generator", JHEP 12 (2008) 039.
-\href{https://doi.org/10.1088/1126-6708/2008/12/039}{{doi:10.1088/1126-6708/2008/12/039}}
-
-{[}Hoe14{]} S. Hoeche, F. Krauss, M. Schoenherr, "Uncertainties in
-MEPS@NLO calculations of h+jets", Phys. Rev. D 90 (2014) 014012.
-\href{https://doi.org/10.1103/PhysRevD.90.014012}{{doi:10.1103/PhysRevD.90.014012}}
-
-{[}Hoe19{]} S. Hoeche, S. Prestel, H. Schulz, "Simulation of vector
-boson plus many final jets at the high luminosity LHC", Phys. Rev. D
-100, 0140124 (2019).
-\href{http://dx.doi.org/10.1103/PhysRevD.100.014024}{{doi:10.1103/PhysRevD.100.014024}}
-
-{[}Jad03{]} S. Jadach, Foam: a general-purpose cellular Monte Carlo
-event generator, Comp. Phys. Comm. 152 (2003) 55.
-\href{https://doi.org/10.1016/S0010-4655(02)00755-5}{{doi:10.1016/S0010-4655(02)00755-5}}
-
-{[}Kaw86{]} S.Kawabata, A new Monte Carlo event generator for high
-energy physics, Comp. Phys. Comm. 41 (1986) 127.
-\href{https://doi.org/10.1016/0010-4655(86)90025-1}{{doi:10.1016/0010-4655(86)90025-1}}
-
-{[}Kli18{]} M.D. Klimek, M. Perelstein, ``Neural Network-Based Approach
-to Phase Space Integration'',
-\href{https://arxiv.org/abs/1810.11509}{{arXiv:1810.11509}}
-
-{[}Kon20{]} D. Konstantinov, "Optimization of Pythia8", EP-SFT group
-meeting, CERN, January 2020,
-\href{https://indico.cern.ch/event/890670}{{https://indico.cern.ch/event/890670}}
-
-{[}Lep80{]} G. P. Lepage, "VEGAS: an adaptive multi-dimensional
-integration program", Cornell report CLNS-447 (1980).
-\href{https://cds.cern.ch/record/123074}{{https://cds.cern.ch/record/123074}}
-
-{[}Lon05{]} L. Lonnblad, "Correcting the colour-dipole cascade model
-with fixed order matrix elements", JHEP 05 (2002) 046.
-\href{https://doi.org/10.1088/1126-6708/2002/05/046}{{doi:10.1088/1126-6708/2002/05/046}}
-
-{[}Mal03{]} F. Maltoni, T. Stelzer, "MadEvent: automatic event
-generation with MadGraph", JHEP 02 (2003) 027.
-\href{https://doi.org/10.1088/1126-6708/2003/02/027}{{doi:10.1088/1126-6708/2003/02/027}}
-
-{[}Man02{]} M. L. Mangano, M. Moretti, R. Pittau, "Multijet matrix
-elements and shower evolution in hadronic collisions: Wbb + n-jets as a
-case study", Nucl. Phys. B 632 (2002) 343.
-\href{https://doi.org/10.1016/S0550-3213(02)00249-3}{{doi:10.1016/S0550-3213(02)00249-3}}
-
-{[}Mat16{]} O. Mattelaer, "On the maximal use of Monte Carlo samples:
-re-weighting events at NLO accuracy", \emph{Eur. Phys. J. C}
-76\textbf{,} 674 (2016),
-\href{https://doi.org/10.1140/epjc/s10052-016-4533-7}{{doi:10.1140/epjc/s10052-016-4533-7}}
-
-{[}Mat18{]} O. Mattelaer, "MG5aMC status and plans", in {[}Gen19{]}
-
-{[}Nas07{]} P. Nason, MINT: a Computer Program for Adaptive Monte Carlo
-Integration and Generation of Unweighted Distributions Bicocca-FT-07-13
-(2007). \href{https://arxiv.org/abs/0709.2085}{{arxiv:0709.2085}}
-
-{[}SKe18{]} L. Sexton-Kennedy, G. Stewart, "CWP challenges and workshop
-aims", in {[}Gen19{]}
 
@@
 transferred and analysed hundreds of billions of simulated events. This
-effort required a major fraction of the WLCG computing resources {[}Ref
-to Exp Comp TDRs{]}.
+effort required a major fraction of the WLCG computing 
+resources~\cite{ALICE-TDR-12, ATLAS-TDR-17, CMS-TDR-7, LHCb-TDR-11, WLCG-CM-Update, ALICE-TDR-019, 
+LHCb-TDR-018}
 
@@
 with the HL-LHC, will create unique research opportunities by collecting
-an order of magnitude more data. So at the same time, the demand for
+an order of magnitude more data. At the same time, the demand for
 detector simulation will grow accordingly, in order to keep the
@@
 The Geant4 simulation toolkit has been the de facto standard for HEP
-experiment simulation for physics studies over the last two decades.
+experiment simulation for physics studies over the last two decades~\cite{Geant4-2003, Geant4-2006, Geant4-2016}.
 Designed in the late 1990s to cope with the simulation challenges of the
@@
 leveraged by the applications that use it. 
-
 Improvements of physics models need to continue to take place to prevent
@@
 CPU performance of the current code will not improve drastically just by
-relying on modern hardware or better compilers. Taking all those
-elements into account, three main development paths have emerged to be
+relying on modern hardware or better compilers.
+
+Taking all those elements into account, three main development paths have emerged to be
 undertaken simultaneously: these have started to be pursued, however,
 effort in any one of them is far from sufficient to exploit their
-individual potential. On one hand, it is needed to continue to modernise
+individual potential.
+
+On one hand, it is needed to continue to modernise
 and refactor the simulation code by using more optimal, modern
@@
 certainly help to produce more simulated events, but it will not be
-sufficient to meet the high luminosity needs of the LHC experiments. On
-the other hand, different fast simulation techniques, where tracking of
+sufficient to meet the high luminosity needs of the LHC experiments.
+
+On the other hand, different fast simulation techniques, where tracking of
 individual particles in given detectors is replaced by some sort of
@@
 more and more significant and will need to be seamlessly integrated with
-the detailed simulation provided by Geant4. Finally, the use of compute
+the detailed simulation provided by Geant4.
+
+Finally, the use of compute
 accelerators (like GPUs or FPGAa) is the third emerging path that needs
@@
 more in detail those three development axes in the following sections.
+
+In this section we will outline the developments identified in and since the
+Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} focusing
+in particular in these three area.
 
@@
 as choices for particular applications. It is clear that this
-flexibility comes at the price of a performance penalty. We observe that
+flexibility comes at the price of a performance penalty. It was observed that
 writing more compact and more specialised code can lead in certain
-places to up to several tens percent speed-up {[}GeantV RD{]}. The
+places to up to several tens percent speed-up~\cite{GeantV}. The
 technical improvements to explore here consist of avoiding too many
@@
 scoring (to simulate, for instance, the response of an EM calorimeter)
-can also bring speed-ups to the complete application {[}GeantV RD{]}. As
+can also bring speed-ups to the complete application~\cite{GeantV}. As
 far as architectural modifications are concerned, the design of the
@@
 vectorisation where some bottleneck (for example geometry navigation)
-could not be tackled without a major additional long term effort.
-{[}GeantV{]}. At the same time libraries developed in the context of the
+could not be tackled without a major additional long term effort~\cite{GeantV}.
+At the same time libraries developed in the context of the
 GeantV R\&D, e.g. VecGeom, have been successfully integrated in Geant4
-and ROOT benefitting the whole HEP software community {[}CMS01{]}. The
+and ROOT benefitting the whole HEP software community~\cite{CMS01}. The
 cost-over-benefit of more vectorisation of the code has not been deemed
@@
 Effort has been spent by each experiment since the publication of the
-Community Whitepaper to benchmark their simulation code and to maximally
+Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} to benchmark their 
+simulation code and to maximally
 exploit the options provided by the Geant4 toolkit to optimise the
-performance of their applications. All of the various handles provided
+performance of their applications~\cite{ATLAS-G4OPT}. All of the various handles provided
 are being continuously explored and adopted when deemed useful: range
@@
 harmonizing their different architectural choices with the Geant4
-implementation {[}CMS02, LHCb01, ATLAS?{]}, and have or are planning to
+implementation~\cite{CMS02, LHCb01, ATLAS01}, and have or are planning to
 use them in their production environment to gain access to a larger set
 of distributed resources. ATLAS has been running their multi-process
-simulation framework on HPCs systems for production in the last years
-{[}ATLAS02{]} while CMS has pioneered the use of their multi-threaded
-simulation on available resources including HPCs {[}CMS03{]}. Both
+simulation framework on HPCs systems for production in the
+last years~\cite{ATLAS02, Benjamin:2696330} 
+while CMS has pioneered the use of their multi-threaded
+simulation on available resources including HPCs~\cite{CMS03}. Both
 approaches have allowed to exploit a higher number of computing
@@
 parameterization was implemented many years ago in the context of the H1
-experiment {[}H1Ref{]}. This parameterization, available within the
+experiment~\cite{H1Gflash}. This parameterization, available within the
 Geant4 toolkit under the name of the GFLASH library, became the starting
@@
 (GAN), Variational Auto-Encoders (VAE) or Autoregressive Generative
-Networks on the `images' of particle showers {[}ML001{]}. The energy
+Networks on the `images' of particle 
+showers~\cite{ML001-GAN, aishik_ghosh_2019_3599705, GAN-ATLAS}. The energy
 deposition values in the calorimeter cells are considered as `pixels' of
@@
 care paid to ensure keeping under control bias on statistical
-uncertainties {[}LHCb02{]}, its applicability in ATLAS and CMS could be
+uncertainties~\cite{LHCb02}, its applicability in ATLAS and CMS could be
 explored for specific cases. Similarly, the re-use of simulated or real
@@
 All LHC experiments have explored the use of multi-objective regression
-and generative adversarial networks (GANs) {[}GAN-ATLAS, GAN-CMS,
-GAN-LHCb{]}, they are now in the process of investigating the use of
+and generative adversarial networks (GANs)~\cite{GAN-ATLAS, GAN-LHCb}, they are now in the process of investigating the use of
 fast simulations for other types of detectors than calorimeters, e.g.
-Cerenkov based systems {[}LAMARR{]} and Time Projection Chambers
-{[}ALICE01{]}. While implementations of fast simulations of given
+Cerenkov based systems~\cite{Lamarr} and Time Projection Chambers~\cite{ALICE-TPC}. While implementations of fast simulations of given
 sub-detectors is specific to their technology and experimental
@@
 physic's reach is also exploited by the experiments with generic
-frameworks for fast simulation of typical collider detectors being used
-{[}Delphes, PGS4{]}. The advantage of such frameworks is to allow
+frameworks for fast simulation of typical collider detectors being used~\cite{bib-Delphes, PGS4}. 
+The advantage of such frameworks is to allow
 simultaneous studies for different experiments as well as their
@@
 with the experiment's analysis frameworks for systematic verifications
-have also been emerging {[}Lamarr{]}.
+have also been emerging~\cite{Lamarr} .
 
@@
 threads, etc.) to efficiently (development time wise and run time wise)
-profit from it {[}GeantV{]}.
+profit from it~\cite{GeantV}.
 
@@
 of them turned out to be successful leading to factors of several
-hundred or thousand of speedup, {[}GPU-``MPEXS-DNA{]},
-{[}Hybrid-Gate{]}, {[}G4CU{]}, {[}Opticks{]} however, they were always
+hundred or thousand of speedup~\cite{GPU-MPEXS-DNA, Hybrid-Gate, Opticks}, however, they were always
 limited to a very specific simulation problem, far from what is
@@
 very successful in restricted domains like medical physics simulation,
-neutron transport {[}SHIFT01{]} or optical photon transport
-{[}CHEP-Juno{]}. In those applications, the type of particles considered
+neutron transport~\cite{SHIFT01} or optical photon transport~\cite{CHEP-Juno}. 
+In those applications, the type of particles considered
 is very limited (often just one type, with no secondary particle
@@
 
-\subsection{Simulation References}\label{references-1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TO BE CONVERTED TO BibTeX, but left here for "working" %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-{[}CMS01{]} V. Ivanchenko and S. Banerjee on behalf of the CMS
-Collaboration, EPJ Web of Conferences 214, 02012 (2019), The PDF is
-here:
-\url{https://www.epj-conferences.org/articles/epjconf/pdf/2019/19/epjconf_chep2018_02012.pdf}
-
-{[}CMS02{]} M Hildreth et al 2015 J. Phys.: Conf. Ser. 664 072022
-
-{[}CMS03{]} M Hildreth et al 2017 J. Phys.: Conf. Ser. 898 042040
-
-{[}GPU-``MPEXS-DNA{]} S. Okada et al. ``MPEXS-DNA: a new GPU-based Monte
-Carlo simulator for track structures and radiation chemistry at
-subcellular scale,'' Med Phys. vol 46 no. 3, pp. 1483-1500 (2019)
-
-{[}Hybrid-Gate{]} J. Bert et al., "Hybrid GATE: A GPU/CPU implementation
-for imaging and therapy applications," 2012 IEEE Nuclear Science
-Symposium and Medical Imaging Conference Record (NSS/MIC), Anaheim, CA,
-pp. 2247-2250 (2012)
-
-{[}G4CU{]}
-\url{https://kds.kek.jp/indico/event/15926/session/30/contribution/110/material/slides/0.pdf}
-
-{[}Opticks{]} S. Blyth, ``Opticks : GPU Optical Photon Simulation for
-Particle Physics using NVIDIA® OptiXTM,'' EPJ Web Conf., 214, pp. 02027
-(2019) 02027
-
-{[}LHCb01{]} D. Muller, ``Adopting new technologies in the LHCb Gauss
-simulation framework'', EPJ Web of Conferences \textbf{214}, 02004
-(2019); B.Siddi and D. Muller, ``Gaussino a Gaudi-based core simulation
-framework'', CERN-LHCb-PROC-2019-010;
-
-{[}ML001{]} Adversarial Networks (GAN) (3DGAN reference), Variational
-Auto-Encoders (VAE) or Autoregressive Generative Networks on the
-`images' of particle showers
-
-{[}GAN-ATLAS, GAN-CMS, GAN-LHCb{]} Reference to GAN calorimeters for
-ATLAS, CMS and LHCb
-
-{[}LHCb02{]} D. Muller et al. ``ReDecay: A novel approach to speed up
-the simulation at LHCb'' , Eur. Phys. J. C 78 (2018)
-
-{[}GeantV RD{]}
-\href{https://indico.cern.ch/event/818702/}{{https://indico.cern.ch/event/818702/}}
-
-{[}CHEP-Juno{]} S. Blyth, ``Meeting the challenge of JUNO simulation
-with Opticks GPU Optical Photon Accelerator'', presented at CHEP 2019 in
-Adelaide, Australia
-
-{[}SHIFT01{]} Hamilton, S.P., Evans, T.M., 2019. Continuous-energy Monte
-Carlo neutron transport on GPUs in the Shift code. \emph{Annals of
-Nuclear Energy} \textbf{128}, 236--247.
-\url{https://doi.org/10.1016/j.anucene.2019.01.012}
-
-{[}DELPHES{]} de Favereau, J., Delaere, C., Demin, P. \emph{et al.}
-DELPHES 3: a modular framework for fast simulation of a generic collider
-experiment. \emph{J. High Energ. Phys.} 2014, 57 (2014).
-\href{https://doi.org/10.1007/JHEP02(2014)057}{{https://doi.org/10.1007/JHEP02(2014)057}}
-
-{[}PGS4{]} J. Conway, R. Culbertson, R. Demina, B. Kilminster, M. Kruse,
-S. Mrenna, J. Nielsen, M. Roco, A. Pierce, J. Thaler, T. Wizansky,
-``Pretty Good Simulation'',
-\href{http://conway.physics.ucdavis.edu/research/software/pgs/pgs4-general.htm}{{http://conway.physics.ucdavis.edu/research/software/pgs/pgs4-general.htm}}
-
-{[}LAMARR{]} A.Davis, ``Fast simulations in LHCb'', CHEP 2019,
-\href{https://indico.cern.ch/event/773049/contributions/3474742/attachments/1937708/3211775/davis_fast_simulations_at_lhcb_v3.pdf}{{https://indico.cern.ch/event/773049/contributions/3474742/attachments/1937708/3211775/davis\_fast\_simulations\_at\_lhcb\_v3.pdf}}
-
-{[}ALICE01{]} B. Volker, HSF Simulation Working Group meeting,
-\href{https://indico.cern.ch/event/782507/contributions/3293465/}{{https://indico.cern.ch/event/782507/contributions/3293465/}}
-
-{[}H1Ref{]} G. Grindhammer, M. Rudowicz and S. Peters, NIM A290 (1990)
-469-488
-
 \input{input_recoTrigger}
@@
 share of disk resources, filling approximately 80\% of the total data
-volume for both ATLAS and CMS {[}1,2{]}. To serve precision analyses
+volume for both ATLAS and CMS \cite{Ref1,Ref2}. To serve precision analyses
 that require large event statistics, the HL-LHC will increase the
@@
 data dozens of times a year, the total CPU consumption at the HL-LHC is
-projected to be around 10\% of total experimental computing {[}1,2{]}.
+projected to be around 10\% of total experimental computing \cite{Ref1,Ref2}.
 Central production of analysis data formats may account for another
@@
 Two approaches have been followed during LHC Run 2 by ATLAS and CMS:
-ATLAS analysis trains {[}4{]} and the CMS mini-AOD, a highly reduced and
+ATLAS analysis trains \cite{Ref4} and the CMS mini-AOD, a highly reduced and
 standardised event content. Analysis trains place software payloads
@@
 centralised execution, amortising data staging costs and sharing some
-common processing steps {[}3{]}. Standardised event contents are instead
+common processing steps \cite{Ref3}. Standardised event contents are instead
 common reductions that satisfy the needs of the bulk of analyses,
@@
 HL-LHC. Extrapolating using a simple model, ATLAS finds that they would
-produce 35 PB/year of data AOD and 215 PB/year of MC {[}1{]}. In
+produce 35 PB/year of data AOD and 215 PB/year of MC \cite{Ref1}. In
 comparison, a reduced format (10 kB/event) made from the AOD would
 result in 0.5 PB/year for data and 2 PB/year for MC. Production of this
-reduced format using a Data Carousel model {[}1{]} allows the much
+reduced format using a Data Carousel model \cite{Ref1} allows the much
 larger AOD to reside on tape, providing significant savings in disk
@@
 \textasciitilde2 kB/event for the nano-AOD) could be possible while
-still supporting a majority of physics analyses {[}5{]}. File size
+still supporting a majority of physics analyses \cite{Ref5}. File size
 reductions also permit multiple copies to be held on disk for more
@@
 data format is quantifiably very efficient for the processing model in
-HEP {[}16{]}, and any competitors would have to achieve a similar level
+HEP \cite{Ref16}, and any competitors would have to achieve a similar level
 of performance.
@@
 measures, a longer-term solution for code complexity may be the adoption
-of a Domain-Specific Language (DSL) {[}10{]}, discussed in more detail
+of a Domain-Specific Language (DSL) \cite{Ref10}, discussed in more detail
 later, a model that would have physicists write logical descriptions of
@@
 hardware, providing a natural means of analysis preservation. An
-interesting development in this direction is the Reana platform {[}21{]}
+interesting development in this direction is the Reana platform \cite{Ref21}
 that supports multiple backends and uses a DSL to provide a framework to
@@
 Most of the problems identified above were already identified in the
-community white paper {[}17{]} from 2017 and since then some progress
+community white paper \cite{Ref17} from 2017 and since then some progress
 was made prototyping new technologies to enable data analysis at the
@@
 ntuples. The ntuples are typically analysed by the PyHEP toolkit
-{[}6{]}, and packages from the wider data science community. Deeper
+\cite{Ref6}, and packages from the wider data science community. Deeper
 integration with external python tools is also particularly important
@@
 density drives compression and this can vary massively between
-experiments and analyses. This new format called RNTuple {[}9{]}, an
+experiments and analyses. This new format called RNTuple \cite{Ref9}, an
 evolution of the TTree file format, shows robust and significant
@@
 (always executed in C++), which natively exploits multi-core
-architectures {[}7{]} and could be extended to accelerators. Used from
+architectures \cite{Ref7} and could be extended to accelerators. Used from
 Python, it allows to pre-process data in C++, and export to numpy.
@@
 optimisations, more complex deep learning models and enhanced
-interoperability with common data science tools {[}14,15{]} which is of
+interoperability with common data science tools \cite{Ref14,Ref15}  which is of
 particular importance for inference.
@@
 A complementary approach provided in Python is the toolchain of uproot,
-awkward arrays and coffea {[}8{]} for efficient columnar operations and
+awkward arrays and coffea \cite{Ref8} for efficient columnar operations and
 transformations of the data. These function as a lightweight end-stage
@@
 for software education and have been leveraged as a powerful frontend
-for access to facilities including CERN's SWAN {[}20{]}.
+for access to facilities including CERN's SWAN \cite{Ref20}.
 
@@
 delegated to the tool and framework experts. Prototypes of DSLs have
-been developed in recent years {[}10{]} for the event processing and
+been developed in recent years \cite{Ref10} for the event processing and
 histogram production parts of the analysis workflow, and even built into
-experimental frameworks to perform high-level analysis {[}19{]}. Similar
+experimental frameworks to perform high-level analysis \cite{Ref19}. Similar
 efforts can also be investigated in the context of data interpretation
@@
 Higgs discovery) {[}ref{]}. Descriptions of a full analysis in terms of
-workflows could then tie together the analysis at the top level {[}18,
-21{]}. With some effort, the HEP community may be able to converge on
+workflows could then tie together the analysis at the top level \cite{Ref18,Ref21}. With some effort, the HEP community may be able to converge on
 effective tools that provide a common solution, ideally across
@@
 able to quickly provide tens of thousands of CPUs with guaranteed
-bandwidth to storage of 2 Gb/coresec {[}12{]}. Scaling this to
+bandwidth to storage of 2 Gb/coresec \cite{Ref12}. Scaling this to
 sustained, diverse analysis payloads will be far from trivial, so the
@@
 analysis. Eight initial benchmark challenges have been defined to set
-the scope that should be addressed by analysis systems {[}11{]}. What
+the scope that should be addressed by analysis systems \cite{Ref11}. What
 remains to be demonstrated is the capability of such systems to scale to
@@
 
-\subsection{Analysis References}\label{references-2}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TO BE CONVERTED TO BibTeX, but left here for "working" %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-{[}1{]} *ATLAS Collaboration, \emph{Evolution of the ATLAS analysis
-model for Run-3 and prospects for HL-LHC}, ATL-SOFT-PROC-2020-002,
-\href{https://cds.cern.ch/record/2708664/}{{https://cds.cern.ch/record/2708664/}}
-
-{[}2{]} CMS Collaboration, HL-LHC resource projections,
-\href{https://twiki.cern.ch/twiki/bin/view/CMSPublic/CMSOfflineComputingResults}{{https://twiki.cern.ch/twiki/bin/view/CMSPublic/CMSOfflineComputingResults}}
-
-{[}3{]} *ALICE Collaboration, \emph{The ALICE analysis train system},
-J.Phys.Conf.Ser. 608 (2015) no.1, 012019,
-\href{http://arxiv.org/abs/arXiv:1502.06381}{{http://arxiv.org/abs/arXiv:1502.06381}}
-
-{[}4{]} *ATLAS Collaboration, \emph{A new petabyte-scale data derivation
-framework for ATLAS. Journal of Physics: Conference Series}, 664.
-072007.
-
-{[}5{]} *CMS Collaboration, \emph{A further reduction in CMS event data
-for analysis: the NANOAOD format}, EPJ Web of Conferences. 214. 06021.
-
-{[}6{]}
-\href{https://github.com/hsf-training/PyHEP-resources}{{https://github.com/hsf-training/PyHEP-resources}}
-
-{[}7{]} D Piparo et al, \emph{RDataFrame: Easy Parallel ROOT Analysis at
-100 Threads}, EPJ Web of Conferences 214, 06029 (2019),
-\href{https://www.epj-conferences.org/articles/epjconf/abs/2019/19/epjconf_chep2018_06029/epjconf_chep2018_06029.html}{{https://www.epj-conferences.org/articles/epjconf/abs/2019/19/epjconf\_chep2018\_06029/epjconf\_chep2018\_06029.html}}
-
-{[}8{]}
-\href{https://github.com/scikit-hep/uproot}{{https://github.com/scikit-hep/uproot}},
-\href{https://github.com/scikit-hep/awkward-array}{{https://github.com/scikit-hep/awkward-array}},
-\href{https://github.com/CoffeaTeam}{{https://github.com/CoffeaTeam}},
-\href{https://github.com/diana-hep/pyhf}{{https://github.com/diana-hep/pyhf}}
-
-{[}9{]} J Blomer et al, \emph{Evolution of the ROOT Tree I/O},
-\href{https://arxiv.org/abs/2003.07669}{{https://arxiv.org/abs/2003.07669}}
-
-{[}10{]} ADL Workshop at Fermilab,
-\href{https://indico.cern.ch/event/769263/timetable/?view=standard}{{https://indico.cern.ch/event/769263/}},
-Gordon's CHEP talk (good overview of links, is there something more
-official?)
-\href{https://indico.cern.ch/event/773049/contributions/3476174/attachments/1938123/3220511/Declarative.pdf}{{https://indico.cern.ch/event/773049/contributions/3476174/attachments/1938123/3220511/Declarative.pdf}}
-
-{[}11{]}
-\href{https://github.com/iris-hep/adl-benchmarks-index}{{https://github.com/iris-hep/adl-benchmarks-index}}
-
-{[}12{]} L Heinrich \& R Rocha, \emph{Reperforming a Nobel Prize
-discovery on Kubernetes}, talk at CHEP 2019,
-\href{https://indico.cern.ch/event/773049/contributions/3581373}{{https://indico.cern.ch/event/773049/contributions/3581373}}
-(originally at KubeCon)
-
-{[}13{]} *ALICE Collaboration, \emph{The ALICE analysis framework for
-LHC Run 3}, EPJ Web Conf. 214 (2019) 05045
-\href{http://cds.cern.ch/record/2699856}{{http://cds.cern.ch/record/2699856}}
-
-{[}14{]} \href{http://paperpile.com/b/2mqsuS/IDej}{L Moneta et al,
-\emph{Machine Learning with ROOT/TMVA}, EPJ Web of Conferences;
-Proceedings of CHEP 2019 (in print). 2020.}
-
-{[}15{]} \href{http://paperpile.com/b/2mqsuS/6uIv}{K Albertsson et al,
-\emph{Fast Inference for Machine Learning in ROOT/TMVA}, EPJ Web of
-Conferences; Proceedings of CHEP 2019 (in print). 2020.}
-
-{[}16{]} J Blomer, \emph{A quantitative review of data formats for HEP
-analyses}, Journal of Physics: Conference Series. 2018. p. 032020.
-
-{[}17{]} HEP Software Foundation Collaboration, \emph{HEP Software
-Foundation Community White Paper Working Group - Data Analysis and
-Interpretation},
-\href{https://arxiv.org/abs/1804.03983}{{arXiv:1804.03983}}
-
-{[}18{]} M Rieger et al, \emph{Design and Execution of make-like,
-distributed Analyses based on Spotify's Pipelining Package Luigi},
-\href{https://arxiv.org/abs/1706.00955}{{arXiv:1706.00955}}
-
-{[}19{]} T Kuhr et al, \emph{The Belle II Core Software},
-\href{https://docs.belle2.org/record/1044}{{BELLE2-PUB-TE-2018-002}}
-
-{[}20{]} Swan homepage:
-\href{https://swan.web.cern.ch/}{{https://swan.web.cern.ch/}}
-
-{[}21{]} Reana homepage:
-\href{http://reanahub.io/}{{http://reanahub.io/}}
-
-Proceedings marked `*' will be credited to the shorter author lists
-
 \hypertarget{summary}{%

### 2020-04-30 LHCC/2020/2020-01/ctcs.tex
Add references and British English spelling changes

@@
 
-\title{Common Tools and Community Software}
+\title{HL-LHC Computing Review:\\Common Tools and Community Software}
 
@@
 where the I/O layer must optimise data transfers and hide latency, e.g.
-taking advantage of xrootd's single column streaming ability \cite{xrootd}.
+taking advantage of XRootD's single column streaming ability \cite{xrootd}.
 
@@
 exploitation of modern GPU-based supercomputers at High Performance
-Computing (HPC) centers, and generators look like a natural candidate
+Computing (HPC) centres, and generators look like a natural candidate
 for this.
@@
 is currently scheduled for fall 2021. A more detailed version of this
-contribution, including a more complete list of references, will soon be
-uploaded on arxiv and submitted for publication~\cite{Gen20}.
+contribution, including a more complete list of references, is
+uploaded on arXiv and will be submitted for publication~\cite{Gen20}.
 
@@
 involves choices of precision (e.g. Leading Order, LO, or
-Next-to-Leading-Order, NLO), hadronization and Parton Shower (PS)
+Next-to-Leading-Order, NLO), hadronisation and Parton Shower (PS)
 models, underlying event tunes, prescriptions for matching/merging and
@@
 2. {A very diverse combination of skills/profiles are needed}: theorists
-(who create fundamental physics models, and design, develop and optimize
+(who create fundamental physics models, and design, develop and optimise
 most generator code), experimentalists working in research (who request
@@
 - {Training challenges}\textbf{.} Theorists and experimentalists often
-lack formal training in software development and optimization. Software
+lack formal training in software development and optimisation. Software
 engineers and experimentalists are often not experts in the theoretical
@@
 
-- {Career challenges}. Those working in the development, optimization or
+- {Career challenges}. Those working in the development, optimisation or
 execution of generator software provide essential contributions to the
@@
 LHC experiments run large scale productions of unweighted event
-generation, and mainly need to maximize the throughput of events
+generation, and mainly need to maximise the throughput of events
 generated per unit time on a given node.
@@
 experience and already obtained significant speedups through various
-techniques. The speed of color reconnection algorithms is a limiting
+techniques. The speed of colour reconnection algorithms is a limiting
 factor for simulating rare hadron decays in LHCb.
@@
 requirements. For a fraction r of negative weight events, the number of
-events to generate increases by a factor 1/(1-2r)\^{}2: for instance,
+events to generate increases by a factor $1/(1-2r)^2$: for instance,
 with r=25\% (which may be regarded as a worst-case scenario in top quark
@@
 
-- {Data parallelism, GPUs and vectorization}. The data flow of an MC
+- {Data parallelism, GPUs and vectorisation}. The data flow of an MC
 generator, where the same (matrix element) function is computed
@@
 
-- {Generic code optimizations}. A speedup of generators may also be
-achievable by more generic optimizations, not involving concurrency. It
+- {Generic code optimisations}. A speedup of generators may also be
+achievable by more generic optimisations, not involving concurrency. It
 should be studied, for instance, if data caching~\cite{Kon20} or
@@
 effort required a major fraction of the WLCG computing 
-resources~\cite{ALICE-TDR-12, ATLAS-TDR-17, CMS-TDR-7, LHCb-TDR-11, WLCG-CM-Update, ALICE-TDR-019, 
-LHCb-TDR-018}
+resources~\cite{ALICE-TDR-12, ATLAS-TDR-17, CMS-TDR-7, LHCb-TDR-11, WLCG-CM-Update, 
+ALICE-TDR-019, LHCb-TDR-018, ATLAS-LHCC-2019-02, CMS-LHCC-2019-09}
 
@@
 Designed in the late 1990s to cope with the simulation challenges of the
-LHC era, Geant4 introduced flexible physics modeling, exploiting layers
+LHC era, Geant4 introduced flexible physics modelling, exploiting layers
 of virtuality. This fostered progress in physics by comparison of
@@
 Finally, the use of compute
-accelerators (like GPUs or FPGAa) is the third emerging path that needs
+accelerators (like GPUs or FPGAs) is the third emerging path that needs
 to be undertaken and that requires intense R\&D effort. We will discuss
@@
 a timely manner. The GeantV Vector prototype has allowed the assessment
-of the achievable speedup using a novel, vectorized approach to particle
+of the achievable speedup using a novel, vectorised approach to particle
 transport. The demonstrator of a full electromagnetic shower in
@@
 GeantV R\&D, e.g. VecGeom, have been successfully integrated in Geant4
-and ROOT benefitting the whole HEP software community~\cite{CMS01}. The
+and ROOT benefiting the whole HEP software community~\cite{CMS01}. The
 cost-over-benefit of more vectorisation of the code has not been deemed
@@
 the Geant4 toolkit for modelling their detectors and the physics
-processes occuring in the experimental setups. Every experiment
+processes occurring in the experimental setups. Every experiment
 configures and uses what is provided by Geant4 for their specific needs.
@@
 shower libraries adopted as baseline in the forward region by ATLAS and
-CMS, stepping granularity optimized, magnetic field caching
-investigated, neutron russian roulette used for dedicated simulations.
+CMS, stepping granularity optimised, magnetic field caching
+investigated, neutron Russian roulette used for dedicated simulations.
 All of them can have major impacts on simulation time and sharing
@@
 multi-threading features within their own multi-threaded frameworks,
-harmonizing their different architectural choices with the Geant4
+harmonising their different architectural choices with the Geant4
 implementation~\cite{CMS02, LHCb01, ATLAS01}, and have or are planning to
@@
 including all the physics processes they would undergo, by a
-parameterization where the detector response is produced directly as a
+parameterisation where the detector response is produced directly as a
 function of the incoming particle type and energy. An example of such a
-parameterization was implemented many years ago in the context of the H1
-experiment~\cite{H1Gflash}. This parameterization, available within the
+parameterisation was implemented many years ago in the context of the H1
+experiment~\cite{H1Gflash}. This parameterisation, available within the
 Geant4 toolkit under the name of the GFLASH library, became the starting
@@
 applications through the available `hook' in the Geant4 toolkit. The LHC
-experiments implemented, for example, dedicated parametrized response
+experiments implemented, for example, dedicated parametrised response
 libraries for some of their calorimeters. Recently, an effort has been
-invested in Geant4 to generalize the parameterization formulae, and to
+invested in Geant4 to generalise the parameterisation formulae, and to
 implement automatic procedures of tuning the parameters for specific
@@
 and prototypes have shown very promising results, however, the
-generalization of the developed models (for different calorimeters, with
+generalisation of the developed models (for different calorimeters, with
 different particles incident at different angles with different
@@
 potential use by the theory community. Experiments specific fully
-parameterized simulations providing reconstructed objects compatible
+parameterised simulations providing reconstructed objects compatible
 with the experiment's analysis frameworks for systematic verifications
@@
 or physics data and thus reduce the memory footprint of a simulation
-while maintaining its throughput performance. The efficient utilization
+while maintaining its throughput performance. The efficient utilisation
 of Single Instruction Multiple Data (SIMD) architectures, on the other
 hand turned out to be more challenging. Not only was a limited part of
-the code `vectorizable' but also, as the already mentioned GeantV R\&D
+the code `vectorisable' but also, as the already mentioned GeantV R\&D
 has shown, overheads related to gathering the data for vector processing
@@
 seen with the GeantV vectorisation R\&D. The most plausible approaches
-seem therefore to lead in the direction of specialized libraries (like
+seem therefore to lead in the direction of specialised libraries (like
 those described above in the context of speeding up Geant4 simulation on
@@
 eventually a limited subset of physics models. While Geant4 is a toolkit
-capable of addressing different modeling and simulation problems and
+capable of addressing different modelling and simulation problems and
 contains many features and capabilities allowing for user access to
 detailed information for a large variety of use cases and scientific
-communities, this GPU development might turn into specialized transport
+communities, this GPU development might turn into specialised transport
 modules, stripped of some features which would be expensive to implement
@@
 and a general approach does not seem feasible even if some
-parametrization could be generalised.
+parametrisation could be generalised.
 
@@
 emerges that could do so it should be pursued and carefully evaluated in
-terms of gains against the disruption of a complete reimplementation of
+terms of gains against the disruption of a complete re-implementation of
 how events are simulated.
@@
 under-documented code that exacerbates the disk and CPU shortfall and
-poses a major difficulty when software is repurposed for a new analysis.
+poses a major difficulty when software is re-purposed for a new analysis.
 Add to this a growing array of alternative toolkits from the data
@@
 compression is also in use and being optimised. Coordination of the data
-reduction phase is a key point for the organization of data analysis
+reduction phase is a key point for the organisation of data analysis
 processing at HL-LHC.
@@
 HL-LHC. Extrapolating using a simple model, ATLAS finds that they would
-produce 35 PB/year of data AOD and 215 PB/year of MC \cite{Ref1}. In
-comparison, a reduced format (10 kB/event) made from the AOD would
+produce 35 PB/year of data AOD and 213 PB/year of MC \cite{Ref1}. In
+comparison, a reduced DAOD\_PHYSLITE format (10 kB/event) made from the AOD would
 result in 0.5 PB/year for data and 2 PB/year for MC. Production of this
@@
 efficient parallel processing. It is clear that as many analyses as
-possible need to be fed into this standardized analysis event-content
+possible need to be fed into this standardised analysis event-content
 pipeline if analysis computing resources are to be kept under control.
@@
 information is often scattered in multiple locations such as:
-``in-file'' data, remote database resources, shared-area txt/root files,
-twiki pages or even in e-mail. The lack of proper metadata integration
+``in-file'' data, remote database resources, shared-area text or ROOT files,
+TWiki pages or even in e-mail. The lack of proper metadata integration
 and a coherent interface leaves analyses exposed to any problems in data
@@
 across the years, datasets from multiple years with different conditions
-will need to be analyzed in a coherent way. The analysis software will
+will need to be analysed in a coherent way. The analysis software will
 then need to be able to fetch and use the proper metadata automatically
@@
 hardware, providing a natural means of analysis preservation. An
-interesting development in this direction is the Reana platform \cite{Ref21}
+interesting development in this direction is the REANA platform \cite{Ref21}
 that supports multiple backends and uses a DSL to provide a framework to
@@
 combination of its versatility as a language, allowing rapid
-proto-typing, and the ability to out-source compute-intensive work to
+prototyping, and the ability to out-source compute-intensive work to
 more performant languages like C and C++. This is similar to the
@@
 density drives compression and this can vary massively between
-experiments and analyses. This new format called RNTuple \cite{Ref9}, an
+experiments and analyses. This new format called RNTuple \cite{Ref9, ROOT-2020-HL-LHC}, an
 evolution of the TTree file format, shows robust and significant
@@
 than a standard storage type's mantissa, enabling further savings in
-storage space.
+storage space~\cite{ROOT-2020-HL-LHC}.
 
@@
 of tools running on ever more heterogeneous hardware becomes more and
-more difficult. In addition, analysis code is typically reimplemented by
+more difficult. In addition, analysis code is typically re-implemented by
 tens of individuals, in different experiments, analysis teams or
@@
 DSLs are a generalisation of this declarative concept and have already
-demonstrated their ability to simplify analysis code {[}{]}. Not only do
+demonstrated their ability to simplify analysis code. Not only do %% Missing reference?
 they allow users to express operations concisely, improving
@@
 called ``combine tools'' developed by CMS and based on RooFit for the
-Higgs discovery) {[}ref{]}. Descriptions of a full analysis in terms of
+Higgs discovery). Descriptions of a full analysis in terms of %% Missing reference?
 workflows could then tie together the analysis at the top level \cite{Ref18,Ref21}. With some effort, the HEP community may be able to converge on
@@
 tasks are often IO-intensive and not necessarily-CPU intensive, the
-right tradeoff between CPU and high bandwidth storage should be
+right trade off between CPU and high bandwidth storage should be
 carefully studied. More detailed studies of resource usage and its
@@
 
-As a proof of principle, an unoptimized Higgs discovery analysis has
+As a proof of principle, an unoptimised Higgs discovery analysis has
 been rerun on Google Cloud in a few minutes as the infrastructure was
@@
   Survey generator codes to understand the best way to move to GPUs and
-  vectorized code, and prototype the port of the software to GPUs using
+  vectorised code, and prototype the port of the software to GPUs using
   data-parallel paradigms.
@@
   \begin{quote}
-  Support efforts to optimize phase space sampling and integration
+  Support efforts to optimise phase space sampling and integration
   algorithms, including the use of Machine Learning techniques such as
@@
   trigger and reconstruction, including general-purpose interoperability
-  libraries with a focus to generalizing and mitigating their
+  libraries with a focus to generalising and mitigating their
   dependencies on architecture-specific codes.

### 2020-04-30 LHCC/2020/2020-01/input_recoTrigger.tex
Add references and British English spelling changes

@@
 
-Experiments have also been working towards minimizing the differences
+Experiments have also been working towards minimising the differences
 between trigger (online) and offline software for reconstruction and
@@
 combinatorial load for tracking algorithms at future colliders will put
-great strain on compute resources. To minimize the memory footprint,
+great strain on compute resources. To minimise the memory footprint,
 tracking software needs to be thread-safe to support multi-threaded
@@
 perform track building in the context of the LHC Run-3 (or Run-2). The
-implementation relies on single-precision floating point math and is
+implementation relies on single-precision floating point mathematics and is
 available for multicore CPUs and achieves significant gains in compute
@@
 employing next-generation computing architectures such as HPCs. It is
-also exploring distributed training and optimization of Graph Neural
+also exploring distributed training and optimisation of Graph Neural
 Networks (GNN) on HPCs, and the deployment of GNNs with microsecond
@@
 
-\hypertarget{general-reconstruction-software-improvements-vectorization}{%
+\hypertarget{general-reconstruction-software-improvements-vectorisation}{%
 \subsubsection{General reconstruction software improvements:
-vectorization}\label{general-reconstruction-software-improvements-vectorization}}
+vectorisation}\label{general-reconstruction-software-improvements-vectorization}}
 
@@
 developed software, including machine learning toolkits (e.g.,
-tensorflow), that typically make excellent use of vector units.
+TensorFlow), that typically make excellent use of vector units.
 Challenges remain in this area. For example, full exploitation of this
 must also account for many generations of hardware that experiments must
-exploit on the grid. In addition, to realize full performance gains, a
+exploit on the grid. In addition, to realise full performance gains, a
 large fraction of the application must be improved to use SIMD processor
@@
 especially in persistifying models and running inferences in production
-(e.g.
-\emph{\href{https://github.com/lwtnn/lwtnn}{{https://github.com/lwtnn/lwtnn}}
-and \href{https://onnx.ai}{{https://onnx.ai}}}) and in enhancing the
+and in enhancing the
 training capacities of collaborations by submitting jobs to the grid and
-to facilities offering large CPU/GPU resources. One can also foresee
+to facilities offering large CPU/GPU resources~\cite{LWTNN, ONNX}. One can also foresee
 that HEP will increase the benefit from techniques and cross-talk from
@@
 specific implementation of a first-level trigger for LHCb Run 3
-data-taking. Allen is optimized for sustaining the rate required by
+data-taking. Allen is optimised for sustaining the rate required by
 real-time processing in LHCb, but can be used as a more general GPU data
@@
 the latest development in synergy with industry and attempting to
-increase the use of centers with heterogeneous resources by experiments.
+increase the use of centres with heterogeneous resources by experiments.
 

### 2020-04-30 LHCC/2020/2020-01/ctcs.tex
Rationalise logical document structure
In particular use more logical LaTeX constructs in the generators document,
avoid "home made" numberings

Shorten many of the overlong subsection titles and consistently capitalise

@@
 % Copyright (C) 2020, the authors, licence CC-BY-4.0.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% Document now locked in Overleaf, transition to GitHub in progress %%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \usepackage[titletoc]{appendix}
+\setcounter{tocdepth}{2} % Only to subsection level in ToC
 \usepackage{xspace}
@@
 \hypertarget{collaboration-challenges}{%
-\subsection{Collaborative challenges}\label{collaboration-challenges}}
-
-1. {The landscape of generator software is extremely varied}. Different
+\subsection{Collaborative Challenges}\label{collaboration-challenges}}
+
+\subsubsection{Generator Software Landscape}
+
+The landscape of generator software is extremely varied. Different
 generators are used for different processes. Generating a sample also
@@
 
-2. {A very diverse combination of skills/profiles are needed}: theorists
+\subsubsection{Skills and Profiles}
+
+A very diverse combination of skills are needed in generator software,
+thus people with many distinct profiles need to contribute. Theorists
 (who create fundamental physics models, and design, develop and optimise
@@
 
-- {Training challenges}\textbf{.} Theorists and experimentalists often
+\paragraph{Training challenges.} Theorists and experimentalists often
 lack formal training in software development and optimisation. Software
@@
 
-- {Communication challenges}. It is difficult to find a shared
+\paragraph{Communication challenges.} It is difficult to find a shared
 terminology and set of concepts to understand one another: notions and
@@
 
-- {Career challenges}. Those working in the development, optimisation or
+\paragraph{Career challenges.} Those working in the development, optimisation or
 execution of generator software provide essential contributions to the
@@
 
-- {Mismatch in usage patterns and optimisation focus}. The way
+\paragraph{Mismatch in usage patterns and optimisation focus.} The way
 generators are built and used by their authors is often different from
@@
 
-- {Programming languages}: Attracting collaborators with a computer
+\paragraph{Programming languages.} Attracting collaborators with a computer
 science background to work on generators, especially students, may also
@@
 \hypertarget{technical-challenges-software-and-computing}{%
-\subsection{Technical challenges (software and
-computing)}\label{technical-challenges-software-and-computing}}
+\subsection{Technical Challenges}\label{technical-challenges-software-and-computing}}
 
@@
 
-1. There are many sources of inefficiency in unweighted event
-generation:
-
-- {Phase space sampling inefficiency}. Efficient sampling is the most
+\subsubsection{Inefficiency in Unweighted Event Generation}
+
+\paragraph{Phase space sampling inefficiency.} Efficient sampling is the most
 critical ingredient for efficient unweighted event generation. Many
@@
 
-- {Merging inefficiency}. Merging prescriptions (e.g. MLM~\cite{Man02},
+\paragraph{Merging inefficiency.} Merging prescriptions (e.g. MLM~\cite{Man02},
 CKKW-L~\cite{Lon02} at LO and FxFx~\cite{Fre12}, MEPS@NLO~\cite{Hoe14} at
@@
 
-- {Filtering inefficiency}. An additional large source of inefficiency
+\paragraph{Filtering inefficiency.} An additional large source of inefficiency
 is due to the way the experiments simulate some processes, where they
@@
 
-- {Sample sharing and reweighting}. In addition to removing
+\paragraph{Sample sharing and reweighting.} In addition to removing
 inefficiencies, other ways could be explored to make maximal use of the
@@
 
-- {Negative weights}. Matching prescriptions, like MC@NLO, are required
+\paragraph{Negative weights.} Matching prescriptions, like MC@NLO, are required
 in (N)NLO calculations to avoid double counting between (N)NLO matrix
@@
 
-2. While progress has been made to better understand which areas of
+\subsubsection{Accounting and Profiling}
+
+While progress has been made to better understand which areas of
 generator software have the highest computational cost, more detailed
@@
 
-- {Accounting of CPU budgets for generators in ATLAS/CMS}. Thanks to a
+\paragraph{Accounting of CPU budgets for generators in ATLAS/CMS.} Thanks to a
 lot of effort from the generator teams in both experiments, a lot of
@@
 
-- {Profiling of the generator software setups used for production}.
+\paragraph{Profiling of the generator software setups used for production.}
 Another area where the WG has been active is the definition and
@@
 
-3. More generally, as is the case in other areas of HEP, some R\&D on
+\subsubsection{Software Modernisation}
+
+More generally, as is the case in other areas of HEP, some R\&D on
 generator software would certainly be needed to modernise it and make it
@@
 
-- {Data parallelism, GPUs and vectorisation}. The data flow of an MC
+\paragraph{Data parallelism, GPUs and vectorisation.} The data flow of an MC
 generator, where the same (matrix element) function is computed
@@
 
-- {Task parallelism, multithreading and multiprocessing}. Generators are
+\paragraph{Task parallelism, multithreading and multiprocessing.} Generators are
 generally executed as single threaded software units. In most cases,
@@
 
-- {Generic code optimisations}. A speedup of generators may also be
+\paragraph{Generic code optimisations.} A speedup of generators may also be
 achievable by more generic optimisations, not involving concurrency. It
@@
 
-\hypertarget{physics-challenges-e.g.-nnlo}{%
-\subsection{Physics challenges (e.g.
-NNLO)}\label{physics-challenges-e.g.-nnlo}}
+\hypertarget{physics-challenges}{%
+\subsection{Physics Challenges}\label{physics-challenges}}
 
@@
 
-1. {NNLO: status of theoretical physics research}. The first question is
+\paragraph{NNLO: status of theoretical physics research.} The first question is
 for which processes NNLO precision will be available at the time of the
@@
 
-2. {NNLO: experimental requirements for data analysis at HL-LHC}. The
+\paragraph{NNLO: experimental requirements for data analysis at HL-LHC.} The
 second question is for which final states unweighted event generation
@@
 
-3. {Size of unweighted event samples required for experimental analysis
-at HL-LHC}. Another question, unrelated to NNLO, is in which regions of
+\paragraph{Size of unweighted event samples required for experimental analysis
+at HL-LHC.} Another question, unrelated to NNLO, is in which regions of
 phase space the number of unweighted events must be strictly
@@
 \hypertarget{experiments-applications-and-optimisation-of-the-use-of-geant4}{%
-\subsection{Experiments applications and optimisation of the use of
+\subsection{Experiment Applications and Optimised Use of
 Geant4}\label{experiments-applications-and-optimisation-of-the-use-of-geant4}}
@@
 \hypertarget{technical-challenges-software-and-computing-1}{%
-\subsection{Technical Challenges (Software and
-Computing)}\label{technical-challenges-software-and-computing-1}}
+\subsection{Technical Challenges}\label{technical-challenges-software-and-computing-1}}
 
@@
 \hypertarget{key-analysis-computing-challenges-at-the-hl-lhc}{%
-\subsection{Key analysis computing challenges at the
+\subsection{Key Analysis Computing Challenges at the
 HL-LHC}\label{key-analysis-computing-challenges-at-the-hl-lhc}}
@@
 \hypertarget{analysis-processing-and-workflows-current-problems-and-solutions}{%
-\subsection{Analysis processing and workflows: current problems and
-solutions}\label{analysis-processing-and-workflows-current-problems-and-solutions}}
+\subsection{Analysis Processing and Workflows: Current Problems and
+Solutions}\label{analysis-processing-and-workflows-current-problems-and-solutions}}
 
@@
 \hypertarget{plans-and-recent-progress}{%
-\subsection{Plans and recent progress}\label{plans-and-recent-progress}}
+\subsection{Plans and Recent Progress}\label{plans-and-recent-progress}}
 

### 2020-04-30 LHCC/2020/2020-01/input_recoTrigger.tex
Rationalise logical document structure
In particular use more logical LaTeX constructs in the generators document,
avoid "home made" numberings

Shorten many of the overlong subsection titles and consistently capitalise

@@
 \hypertarget{the-evolution-of-triggers-and-real-time-analysis}{%
-\subsection{\texorpdfstring{The evolution of triggers and real-time
-analysis
-}{The evolution of triggers and real-time analysis }}\label{the-evolution-of-triggers-and-real-time-analysis}}
+\subsection{Evolution of Triggers and Real-Time Analysis}\label{the-evolution-of-triggers-and-real-time-analysis}}
 
@@
 \hypertarget{challenges-and-improvements-foreseen-in-event-reconstruction}{%
-\subsection{Challenges and improvements foreseen in event
-reconstruction}\label{challenges-and-improvements-foreseen-in-event-reconstruction}}
+\subsection{Challenges and Improvements Foreseen in
+Reconstruction}\label{challenges-and-improvements-foreseen-in-event-reconstruction}}
 
@@
 \hypertarget{tracking-in-high-pile-up-environments}{%
-\subsubsection{Tracking in high pile-up
-environments}\label{tracking-in-high-pile-up-environments}}
+\subsubsection{Tracking in High Pile-up
+Environments}\label{tracking-in-high-pile-up-environments}}
 
@@
 \hypertarget{addition-of-timing-information-in-reconstruction}{%
-\subsubsection{Addition of timing information in
-reconstruction}\label{addition-of-timing-information-in-reconstruction}}
+\subsubsection{Adding Timing Information to
+Reconstruction}\label{addition-of-timing-information-in-reconstruction}}
 
@@
 \hypertarget{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}{%
-\subsection{Enhanced data quality and software monitoring for trigger
-and
-reconstruction}\label{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}}
+\subsection{Enhanced Data Quality and Software Monitoring}\label{enhanced-data-quality-and-software-monitoring-for-trigger-and-reconstruction}}
 
@@
 \hypertarget{general-reconstruction-software-improvements-vectorisation}{%
-\subsubsection{General reconstruction software improvements:
-vectorisation}\label{general-reconstruction-software-improvements-vectorization}}
+\subsubsection{General Reconstruction Software Improvements:
+Vectorisation}\label{general-reconstruction-software-improvements-vectorization}}
 
@@
 \hypertarget{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}{%
-\subsubsection{Use of machine learning for software trigger and
-reconstruction
-algorithms}\label{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}}
+\subsubsection{Use of Machine Learning}\label{use-of-machine-learning-for-software-trigger-and-reconstruction-algorithms}}
 
@@
 
-\hypertarget{algorithms-and-data-structures-to-efficiently-exploit-many-core-and-non-x86-architectures-e.g.-gpus-fpgas}{%
-\subsubsection{\texorpdfstring{Algorithms and data structures to
-efficiently exploit many-core and non-x86 architectures (e.g. GPUs,
-FPGAs):
-}{Algorithms and data structures to efficiently exploit many-core and non-x86 architectures (e.g.
-GPUs, FPGAs): }}\label{algorithms-and-data-structures-to-efficiently-exploit-many-core-and-non-x86-architectures-e.g.-gpus-fpgas}}
+\hypertarget{algorithms-and-data-structures}{%
+\subsubsection{Algorithms and Data Structures}\label{algorithms-and-data-structures}}
 
 Computing platforms are generally evolving towards having more cores in
-order to increase processing capability. The goal is to evolve current
-event models, toolkits and algorithm implementations, and best
-programming techniques to improve the throughput of multithreaded
-software trigger and event reconstruction applications. Additionally,
-computing architectures using technologies beyond CPUs offer an
-interesting alternative for increasing the throughput of the most time
-consuming trigger or reconstruction algorithms.
+order to increase processing capability. In addition different
+architectural models, e.g. GPUs, FPGAs are available.
+The goal is to improve the throughput of multithreaded
+software trigger and event reconstruction applications, so current
+event models, toolkits and algorithm implementations have to evolve to
+efficiently exploit these opportunities. The first algorithms that should be
+targetted are those which are most time consuming.
 
@@
 
-\hypertarget{general-remarks-on-trigger-and-reconstruction-software-sharing}{%
-\subsection{General remarks on trigger and reconstruction software
-sharing}\label{general-remarks-on-trigger-and-reconstruction-software-sharing}}
+\hypertarget{trigger-and-reconstruction-software-sharing}{%
+\subsection{Trigger and Reconstruction Software
+Sharing}\label{trigger-and-reconstruction-software-sharing}}
 

### 2020-04-30 LHCC/2020/2020-01/ctcs.tex
Break sources into an input file per section

--- 2020-04-30 LHCC/2020/2020-01/ctcs.tex
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+% Content is stored in separate TeX files for managability
+\input{input_introduction}
 
-Common and community software packages, such as ROOT, Geant4 and event
-generators have been a key part of the LHC's success so far and
-continued development and optimisation will be critical in the future
-\cite{stewart_graeme_a_2018_2413005, Ellis:2691414, Albrecht2019}. The
-challenges are driven by an ambitious physics programme, notably the LHC
-accelerator upgrade to high-luminosity, HL-LHC, and the corresponding
-detector upgrades of ATLAS and CMS. The General Purpose Detectors
-describe their specific challenges elsewhere; in this document we
-address the issues for software that is used in multiple experiments
-(usually even more widely than ATLAS and CMS) and maintained by teams of
-developers who are either not linked to a particular experiment or who
-contribute to common software within the context of their experiment
-activity. We also give space to general considerations for future
-software and projects that tackle upcoming challenges, no matter who
-writes it, which is an area where community convergence on best practice
-is extremely useful.
+\input{input_generators}
 
-ATLAS and CMS will undergo major detector upgrades and will also
-increase their trigger rates for HL-LHC by about a factor of 10; event
-complexity will rise, with peak pile-up of 200, far higher than in
-Run-2. This places an enormous burden on storage and processing
-resources. Current CMOS microprocessor technology is clock speed limited
-(due to the failure of Dennard scaling) and, while processors
-per-integrated circuit still keeps rising, Moore's Law is expected to
-stall during the 2020s. More importantly, the effective runtime related
-improvements in computing from CPU servers at sites is likely to be only
-about 10\% per year, making the shortfall in processing resources more
-severe. As manufacturers struggle to deliver ever-more effective
-computing through CPUs the drive to different architectures intensifies,
-with GPUs becoming commonplace and increasing interest in even more
-specialised architectures, such as TPUs, IPUs and developments that make
-FPGA devices more user friendly \cite{acm10.1145/3282307}.
-These pose huge challenges for our
-community as the programming models and APIs vary widely here and
-possible lock-in to a particular vendor's devices is a significant
-problem for code sustainability and preservation. Huge work has been
-done already to adapt to this changing landscape, e.g. multi-threading
-software and GPU codes have been used in production by some experiments
-for years~\cite{calafiura2018hep, albrecht2018hepexec}.
-In other areas migration is still ongoing. In others it is
-not even yet started. Generic heterogeneous programming models have
-existed for some time, and new ones are arising, but there is, as yet,
-no clear winner, in-part because the performance obtained can be highly
-application-dependent. HEP itself will not drive the success of one
-model or another, so even if the community converged on a preferred
-model, its long term success would not be assured. The C++ standard
-itself is likely to lag for many years behind what is required by us in
-the area of heterogeneous or even distributed computing. Further,
-experiment frameworks (and with knock-on effects to systems like
-workload management) will need to adapt to sites that provide
-heterogeneous resources. How to do this, and make effective use of
-different heterogeneous resources across different sites, remains far
-from settled.
-
-For storage systems (see the DOMA and WLCG documents) the pressure on
-software is to store as much physics information in as few bytes as
-possible, but also to be able to read at very high rates to deliver data
-from modern storage technologies into the processing hardware. This
-requires critical developments in the storage formats and libraries used
-in HEP, e.g. developments like RNTuple for ROOT I/O is likely to be of
-great importance for the community \cite{ROOT-2020-HL-LHC}. The likelihood of finding an
-off-the-shelf compact and efficient storage format for HEP data is
-remote, so investment in smart software to support our PB sized science
-data is simply cost effective. Particularly for analysis, which is
-usually I/O bound, we have an end-to-end problem from storage
-technology, through the software layers, to processing resources that
-may well span multiple nodes. Other workflows, which are less dependent
-on I/O rates will, nevertheless, have to be adapted to using remote data
-where the I/O layer must optimise data transfers and hide latency, e.g.
-taking advantage of xrootd's single column streaming ability \cite{xrootd}.
-
-In this increasingly complex environment in which to write software,
-there are important problems where sharing information at the community
-level is far more efficient. Providing a high level of support in the
-build environment for developers, sharing knowledge about how to
-measure, and then improve, performance (especially on multiple different
-architectures) and sharing best practice for code development can have a
-large integrated benefit \cite{couturier2017hep}. This requires improved training and the
-development of a curriculum for all developer levels. In the field, such
-initiatives are often warmly received, but real support is needed for
-those who can put this into practice, also ensuring that their work in
-training contributes to their career development and a long term future
-in the field \cite{foundation2018hep}. HEP software stacks are already deep and wide and building
-these consistently and coherently is also an area where knowledge can be
-shared. Support is needed for multiple architectural targets and
-ensuring the correct runtime on heterogeneous environments.
-
-This brings up the important question of validation and the need to
-improve the security of physics results, which is even more complicated
-on heterogeneous platforms, when exact binary compatibility often cannot
-be assured. Currently almost every experiment and project has its own
-infrastructure for this.
-
-Once software is built, it needs to be shipped worldwide so that the
-production workflows can run. CernVM-FS was a huge leap forward for
-software distribution and has even been widely adopted outside HEP 
-\cite{Blomer_2011, 7310920}.
-However, new challenges arise, with container based payloads, scaling
-issues and disconnected supercomputer sites. So maintenance and
-development needs to be undertaken to support and extend this key
-supporting infrastructure for software.
-
-Finally, over the multi-decade lifetimes of HEP experiments, we need to
-preserve both the core and analysis software so that results can be
-confirmed and updated as the field moves on. There are many exciting
-developments based around CernVM-FS
-\cite{10.1007/978-3-319-67630-2_52, serverless-cvmfs}, containers and things like analysis
-description languages \cite{Ref10}, but these are not yet at the stage of being
-settled nor integrated into our day-to-day workflows.
-
-In the rest of this document the main issues associated with the key
-parts of the software workflow in high-energy physics are presented,
-focusing on those that dominate current resource consumption: physics
-event generation, detector simulation, reconstruction and analysis.
-
-\hypertarget{physics-event-generators}{%
-\section{Physics Event Generators}\label{physics-event-generators}}
-
-\hypertarget{introduction-1}{%
-\subsection{Introduction}\label{introduction-1}}
-
-Physics event generators are essential in HEP. All of the LHC scientific
-results, both precision measurements or searches for new physics, depend
-significantly on the comparison of experimental measurements to
-theoretical predictions computed using generator software.
-
-Using Monte Carlo (MC) techniques, generators allow both the generation
-of unweighted events for experimental studies of differential
-distributions and the prediction of total cross sections. The
-large-scale event generation campaigns of the LHC experiments have
-significant computational costs, mainly in terms of CPU resources. The
-limited size of simulated samples is a source of major uncertainty in
-many analyses and is therefore a limiting factor on the potential
-physics output of the LHC programme. This situation will get
-significantly worse in the high-luminosity running phase of the LHC
-(HL-LHC). The fraction of the CPU resources available to the LHC
-experiments which is used for event generation today is currently
-estimated to be approximately 5-15\%. As is the case for the other big
-consumers of CPU (detector simulation and reconstruction, that have
-until now been the main focus), speedups in generator software are also
-needed to address the overall resource problem expected at the HL-LHC,
-also because more accurate predictions, requiring more complex
-calculations (e.g. beyond NLO or with higher jet multiplicities) will be
-needed from generators. Many other issues, both technical and
-non-technical (e.g. funding, training, careers for those working in this
-area), exist~\cite{Alves:2017she,Gen18}.
-
-A HSF Working Group (WG) on generators~\cite{Gen19} was set up at the
-beginning of 2019. The main focus of the WG so far has been on gaining a
-better understanding of the current situation, and identifying and
-prioritising the areas where computing costs can be reduced. In
-particular, the WG has been active in analysing the ATLAS and CMS
-compute budgets in detail, in profiling MG5\_aMC~\cite{Alw14}, 
-Sherpa~\cite{Bot19} and Powheg~\cite{Fri07}, in discussing the possible sharing
-of common parton-level samples by ATLAS and CMS, and in reviewing and
-supporting the efforts for porting generators to modern architectures
-(e.g., MG5\_aMC to GPUs). This last activity is particularly important,
-as it has become increasingly clear that being able to run
-compute-intensive WLCG software workloads on GPUs would allow the
-exploitation of modern GPU-based supercomputers at High Performance
-Computing (HPC) centers, and generators look like a natural candidate
-for this.
-
-This section gives an overview of the many technical and non-technical
-challenges in the generator area and of the work that can be done to
-address them. This is finally condensed into a list of a few
-high-priority items, for the next 18 months before the LHCC review that
-is currently scheduled for fall 2021. A more detailed version of this
-contribution, including a more complete list of references, will soon be
-uploaded on arxiv and submitted for publication~\cite{Gen20}.
-
-\hypertarget{collaboration-challenges}{%
-\subsection{Collaborative challenges}\label{collaboration-challenges}}
-
-1. {The landscape of generator software is extremely varied}. Different
-generators are used for different processes. Generating a sample also
-involves choices of precision (e.g. Leading Order, LO, or
-Next-to-Leading-Order, NLO), hadronization and Parton Shower (PS)
-models, underlying event tunes, prescriptions for matching/merging and
-simulating particle decays, and other input parameters, chiefly among
-them the parton density functions, for which different interfaces exist.
-Various combinations of software libraries are thus possible, often
-written by different authors and some dating back many years. For a
-given process, the LHC experiments often use different software packages
-and settings from each other, and a single experiment can generate
-events using more than one choice. Many different packages and
-configurations may therefore need to be worked on to get cumulative CPU
-cost reductions. The large number of external packages also complicates
-their long-term maintenance and integration in the experiments software
-and workflows, sometimes leading to job failures and computing
-inefficiencies. Other packages are also absolutely critical for the
-whole generator community and must be maintained, even if their CPU cost
-is relatively low (LHAPDF, Rivet, HepMC, etc.).
-
-2. {A very diverse combination of skills/profiles are needed}: theorists
-(who create fundamental physics models, and design, develop and optimize
-most generator code), experimentalists working in research (who request
-types/sizes of samples) and in computing (who implement, monitor and
-account execution of workflows on computing resources), software
-engineers and system performance experts. This is a richness and
-opportunity, as some technical problems are best addressed by people
-with specific skills, but it also poses some challenges:
-
-- {Training challenges}\textbf{.} Theorists and experimentalists often
-lack formal training in software development and optimization. Software
-engineers and experimentalists are often not experts in the theoretical
-physics models implemented in MC codes.
-
-- {Communication challenges}. It is difficult to find a shared
-terminology and set of concepts to understand one another: notions and
-practices that are taken for granted in one domain may be obscure for
-others. As an example, there are many articles about the physics in
-generators, but software engineers need papers describing the software
-modules and overall data/control flow.
-
-- {Career challenges}. Those working in the development, optimization or
-execution of generator software provide essential contributions to the
-success of the (HL-)LHC physics programme and it is critical that they
-get the right recognition and motivation. However, theorists get
-recognition on published papers, and may not be motivated to work on
-software optimisations that are not "theoretical" enough to advance
-their careers. Generator support tasks in the experiments may also not
-be enough to secure jobs/funding to experimentalists pursuing a career
-in research.
-
-- {Mismatch in usage patterns and optimisation focus}. The way
-generators are built and used by their authors is often different from
-the way in which they are deployed and integrated by the experiments in
-their software frameworks and computing infrastructure.The goals and
-metrics of software optimisation work may also differ. Theorists are
-mainly interested in calculating cross sections and focus on minimising
-the phase space integration time for a given statistical precision. The
-LHC experiments run large scale productions of unweighted event
-generation, and mainly need to maximize the throughput of events
-generated per unit time on a given node.
-
-- {Programming languages}: Attracting collaborators with a computer
-science background to work on generators, especially students, may also
-be complicated by the fact that critical components of some generator
-packages are written in Fortran, which is rarely used in industry and
-less popular among developers than other programming languages. Some of
-the generators also do not use industry standard version control
-systems, making it harder to contribute code.
-
-\hypertarget{technical-challenges-software-and-computing}{%
-\subsection{Technical challenges (software and
-computing)}\label{technical-challenges-software-and-computing}}
-
-The event generation workflow presents several challenges and
-opportunities for improvement.
-
-1. There are many sources of inefficiency in unweighted event
-generation:
-
-- {Phase space sampling inefficiency}. Efficient sampling is the most
-critical ingredient for efficient unweighted event generation. Many
-generic algorithms exist (e.g. VEGAS~\cite{Lep80}, BASES / SPRING~\cite{Kaw86}, 
-MINT~\cite{Nas07}, FOAM~\cite{Jad03}), as well as others
-developed specifically for a given generator (e.g. MadEvent~\cite{Mal03},
-itself based on a modified version of VEGAS, in MG5\_aMC, or COMIX~\cite{Gle08} 
-in Sherpa). In general, the larger the dimensionality of the
-phase space, the lower the unweighting efficiency that can be achieved:
-in W+jets, for instance, the Sherpa efficiency is 30\% for W+0 jets and
-0.08\% for W+3jets~\cite{Gao20}. This is an area where research is very
-active, and should be actively encouraged, as significant cost
-reductions in WLCG compute budgets could be achieved. Improvements in
-this area can only start from physics-motivated approaches based on the
-knowledge of phase space peaks, but they can be complemented by
-brute-force machine learning (ML) algorithmic methods~\cite{Ben17,Bot20,Gao20,Kli18}.
-
-- {Merging inefficiency}. Merging prescriptions (e.g. MLM~\cite{Man02},
-CKKW-L~\cite{Lon02} at LO and FxFx~\cite{Fre12}, MEPS@NLO~\cite{Hoe14} at
-NLO) imply the rejection of some events, to avoid double counting
-between events produced with n\textsubscript{jets}+1 matrix elements and
-with n\textsubscript{jets} MEs plus parton showers. The resulting
-inefficiencies can be relatively high depending on the process, but they
-are unavoidable in the algorithmic strategy used by the underlying
-physics modeling~\cite{Alw08}. However, a method like shower-kt MLM can
-reduce the merging inefficiency of MLM~\cite{Alw09}.
-
-- {Filtering inefficiency}. An additional large source of inefficiency
-is due to the way the experiments simulate some processes, where they
-generate large inclusive event samples, which are then ﬁltered on
-ﬁnal-state criteria to decide which events are passed on to detector
-simulation and reconstruction (e.g. CMS simulations of specific
-$\Lambda_{B}$ decays have a 0.01\% efficiency and ATLAS B-hadron
-filtering has \textasciitilde10\% efficiency for V+jets). This
-inefficiency could be reduced by developing ﬁltering tools within the
-generators themselves, designed for compatibility with the requirements
-of the experiments. Filtering is an area where LHCb has a lot of
-experience and already obtained significant speedups through various
-techniques. The speed of color reconnection algorithms is a limiting
-factor for simulating rare hadron decays in LHCb.
-
-- {Sample sharing and reweighting}. In addition to removing
-inefficiencies, other ways could be explored to make maximal use of the
-CPU spent for generation by reusing samples for more than one purpose.
-Sharing parton-level samples between ATLAS and CMS is being discussed
-for some physics analyses. Event reweighting is already commonly used
-(e.g. for new physics searches and some systematic uncertainty
-calculations) and could be explored further, though this may require
-more theoretical work for samples involving merging or NLO matching~\cite{Mat16}.
-
-- {Negative weights}. Matching prescriptions, like MC@NLO, are required
-in (N)NLO calculations to avoid double counting between (N)NLO matrix
-elements and parton showers, leading to the appearance of events with
-negative weights. This causes a large inefficiency, as larger event
-samples must be generated and passed through the experiment simulation,
-reconstruction and analysis codes, increasing the compute and storage
-requirements. For a fraction r of negative weight events, the number of
-events to generate increases by a factor 1/(1-2r)\^{}2: for instance,
-with r=25\% (which may be regarded as a worst-case scenario in top quark
-pair production~\cite{Fre20}), one needs to generate 4 times as many
-events. Negative weights can instead be almost completely avoided, by
-design, in another popular matching prescription, POWHEG, which however
-is only available for a limited number of processes and describes the
-relevant physics to a different degree of precision than MC@NLO 
-(see~\cite{Fre20} for a more in-depth discussion). Progress in this area can
-only be achieved by theorists: for instance, a new approach for
-MC@NLO-type matching with reduced negative weights has recently been
-proposed~\cite{Fre20}. It should be noted that negative weights can also
-happen at LO because of not-positive-definite parton density function
-sets and interference terms, e.g. in effective field theory
-calculations.
-
-2. While progress has been made to better understand which areas of
-generator software have the highest computational cost, more detailed
-accounting of the experiment workloads and profiling of the main
-generator software packages would help to further refine R\&D
-priorities:
-
-- {Accounting of CPU budgets for generators in ATLAS/CMS}. Thanks to a
-lot of effort from the generator teams in both experiments, a lot of
-insight into the settings used to support each experiment's physics
-programme was gained within the WG, and it is now clear that the
-fraction of CPU that ATLAS spends for event generation is somewhat
-higher than that in CMS. More detailed analysis of the different
-strategies is ongoing. A practical issue is that these figures had to be
-harvested from logs and production system databases, which was
-particularly difficult for CMS, requiring significant person hours to
-extract the required information. It is important to establish better
-mechanisms to collect this information, to allow for an easy comparison
-between different experiments.
-
-- {Profiling of the generator software setups used for production}.
-Another area where the WG has been active is the definition and
-profiling of standard generator setups, reproducing those used in
-production. Detailed profiling could also be useful to assess the CPU
-cost of external parton distribution function libraries~\cite{Kon20}, or
-the memory requirements of the software (which may motivate a move to
-multithreading or multiprocessing approaches).
-
-3. More generally, as is the case in other areas of HEP, some R\&D on
-generator software would certainly be needed to modernise it and make it
-more efficient, or even port it to more modern computing 
-architectures~\cite{Bau13,Alves:2017she}:
-
-- {Data parallelism, GPUs and vectorization}. The data flow of an MC
-generator, where the same (matrix element) function is computed
-repeatedly at many phase space points, lends itself naturally to the
-data parallel approaches found in CPU vectorised code and in GPU compute
-kernels. Porting and optimising generators on GPUs is essential to be
-able to use modern GPU-based HPCs (such as SUMMIT, where 95\% of the
-compute capacity comes from GPUs). The work done in this direction in
-the past on MG5\_aMC, that never reached production quality, is now
-being revamped by the WG, in collaboration with the MG5\_aMC team, and
-represents one of the main R\&D priorities of the WG. This work is
-presently focusing on Nvidia CUDA, but abstraction libraries will also
-be investigated. GPUs may also be relevant to ML-based sampling
-algorithms and to the pseudo-random number generation libraries used in
-all MC generators.
-
-- {Task parallelism, multithreading and multiprocessing}. Generators are
-generally executed as single threaded software units. In most cases,
-this is not a problem, as the memory footprint of unweighted event
-generation is small and usually fits within the 2 GB per core available
-on WLCG nodes. However, there are cases where memory is higher than 2 GB
-(e.g. DY+jets using Sherpa); this leads to inefficiencies as some
-processor cores remain unused, which could be avoided using
-multithreading approaches. The fact that some generators are not even
-thread safe, may also be a problem, for instance to embed them in
-multi-threaded event processing frameworks, such as that of CMS.
-Multi-processing approaches may also be useful to speed up the
-integration and optimisation step for complex high-dimensional final
-states. In particular, a lot of work has been done to implement
-MPI-based multi-processing workflows for generators in recent years. For
-instance, the scaling of LO-based generation of merged many-jet samples
-has been successfully tested and benchmarked on HPC architectures using
-both Alpgen~\cite{Chi17} and Sherpa~\cite{Hoe19}; in the latter case, new
-event formats based on HDF5, replacing LHEF, have also been instrumental
-in the success of these tests. MPI integration has also been completed
-for MG5\_aMC~\cite{Mat18}. It should also be noted that, even if HPCs
-offer extremely high-speed inter-node connectivity, it is perfectly ok
-for WLCG workflows to use these systems as clusters of unrelated nodes.
-
-- {Generic code optimizations}. A speedup of generators may also be
-achievable by more generic optimizations, not involving concurrency. It
-should be studied, for instance, if data caching~\cite{Kon20} or
-different compilers and build strategies may lead to any improvements.
-
-\hypertarget{physics-challenges-e.g.-nnlo}{%
-\subsection{Physics challenges (e.g.
-NNLO)}\label{physics-challenges-e.g.-nnlo}}
-
-In addition to software issues, important physics questions should also
-be addressed about more accurate theoretical predictions (above all NNLO
-QCD calculations, but also electroweak corrections) and their potential
-impact on the computational cost of event generators at HL-LHC. Some
-specific NNLO calculations are already available and used today by the
-LHC experiments in their data analysis. With a view to HL-LHC, however,
-some open questions remain to be answered, in particular:
-
-1. {NNLO: status of theoretical physics research}. The first question is
-for which processes NNLO precision will be available at the time of the
-HL-LHC. For example, when would NNLO be expected for 2-\textgreater3 or
-even higher multiplicity final states? And for lower multiplicity final
-states, where differential NNLO predictions exist but the generation of
-unweighted NNLO+PS events is not yet possible. It is important to
-clarify the theoretical and more practical challenges in these
-calculations, and the corresponding computational strategies and impact
-on CPU time needs.
-
-2. {NNLO: experimental requirements for data analysis at HL-LHC}. The
-second question is for which final states unweighted event generation
-with NNLO precision would actually be required, and how many events
-would be needed. One should also ask if reweighting LO event samples to
-NNLO would not be an acceptable cheaper alternative to address the
-experimental needs.
-
-3. {Size of unweighted event samples required for experimental analysis
-at HL-LHC}. Another question, unrelated to NNLO, is in which regions of
-phase space the number of unweighted events must be strictly
-proportional to the luminosity. For example, in the low pT regions of W
-boson production it is probably impossible to keep up with the data, due
-to the huge cross section. Alternative techniques should be
-investigated, to avoid the generation of huge event samples.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\hypertarget{detector-simulation}{%
-\section{Detector Simulation}\label{detector-simulation}}
-
-\hypertarget{introduction-2}{%
-\subsection{Introduction}\label{introduction-2}}
-
-In HEP, data analysis, theory model evaluation and detector design
-choices rely on detailed detector simulation. Since the start of the
-first run the LHC experiments have produced, reconstructed, stored,
-transferred and analysed hundreds of billions of simulated events. This
-effort required a major fraction of the WLCG computing 
-resources~\cite{ALICE-TDR-12, ATLAS-TDR-17, CMS-TDR-7, LHCb-TDR-11, WLCG-CM-Update, ALICE-TDR-019, 
-LHCb-TDR-018}
-
-The increase in delivered luminosity in future LHC runs, in particular
-with the HL-LHC, will create unique research opportunities by collecting
-an order of magnitude more data. At the same time, the demand for
-detector simulation will grow accordingly, in order to keep the
-statistical uncertainties as low as possible. However, the expected
-computing resources will not suffice if current detector simulation is
-to be used; the availability of simulated samples of sufficient size
-will soon become a major limiting factor as far as new physics
-discoveries, for example through precision measurements, are concerned.
-Development of faster simulation, therefore, is of crucial importance
-and different techniques need to be explored under the assumption that
-they will provide a sufficient gain in time performance with a
-negligible or acceptable loss in physics accuracy.
-
-The Geant4 simulation toolkit has been the de facto standard for HEP
-experiment simulation for physics studies over the last two decades~\cite{Geant4-2003, Geant4-2006, Geant4-2016}.
-Designed in the late 1990s to cope with the simulation challenges of the
-LHC era, Geant4 introduced flexible physics modeling, exploiting layers
-of virtuality. This fostered progress in physics by comparison of
-competing models on same energy ranges, and allowed complementary models
-to be combined to cover large energy ranges. The simulation applications
-of the LHC experiments make use of this toolkit and, as such, most of
-the LHC physics program relies on it. It has delivered a physics
-precision that allows the experiments to perform precision analysis of
-the collected data and to produce new results pushing the boundaries of
-our understanding in HEP. At the same time, the code of the toolkit is
-becoming old, with some parts written almost thirty years ago, making it
-more and more difficult to run efficiently on modern computing
-architectures. Any improvement in the critical elements of Geant4 can be
-leveraged by the applications that use it. 
-Improvements of physics models need to continue to take place to prevent
-systematic uncertainties from the simulation from becoming a dominant
-factor, however, their exact influence on the precision of physics
-analysis is not always easy to quantify. On the other hand, the overall
-CPU performance of the current code will not improve drastically just by
-relying on modern hardware or better compilers.
-
-Taking all those elements into account, three main development paths have emerged to be
-undertaken simultaneously: these have started to be pursued, however,
-effort in any one of them is far from sufficient to exploit their
-individual potential.
-
-On one hand, it is needed to continue to modernise
-and refactor the simulation code by using more optimal, modern
-programming techniques. Those improvements, requiring considerable
-effort but no major paradigm changes, can reasonably bring several tens
-of percent of speed-up, by improving the throughput on modern CPUs,
-avoiding constant churn in data and instruction caches. This will
-certainly help to produce more simulated events, but it will not be
-sufficient to meet the high luminosity needs of the LHC experiments.
-
-On the other hand, different fast simulation techniques, where tracking of
-individual particles in given detectors is replaced by some sort of
-parameterisation, have already been extensively used for LHC simulation
-campaigns, where some precision is traded to achieve higher event
-throughput (thus lowering the cost of samples overall). Techniques to
-re-use part of the simulated events multiple times have also emerged.
-New technologies (like Machine Learning), should certainly be
-investigated as fast simulation approaches. All of these will become
-more and more significant and will need to be seamlessly integrated with
-the detailed simulation provided by Geant4.
-
-Finally, the use of compute
-accelerators (like GPUs or FPGAa) is the third emerging path that needs
-to be undertaken and that requires intense R\&D effort. We will discuss
-more in detail those three development axes in the following sections.
-
-In this section we will outline the developments identified in and since the
-Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} focusing
-in particular in these three area.
-
-\hypertarget{geant4-rd}{%
-\subsection{Geant4 R\&D}\label{geant4-rd}}
-
-The Geant4 particle transport toolkit contains advanced models for
-electromagnetic and hadronic physics processes, as well as the
-geometrical description of the detectors. The toolkit design principles
-were to allow flexibility in future extensions of its own code, as well
-as choices for particular applications. It is clear that this
-flexibility comes at the price of a performance penalty. It was observed that
-writing more compact and more specialised code can lead in certain
-places to up to several tens percent speed-up~\cite{GeantV}. The
-technical improvements to explore here consist of avoiding too many
-virtual layers and improving the instruction and data locality, leading
-to better cache use. Going even further as far as specialisation is
-concerned, studies suggest that implementing compact, self-contained
-transport libraries with a minimal set of physics models (for instance
-only electromagnetic interaction for selected particles) with predefined
-scoring (to simulate, for instance, the response of an EM calorimeter)
-can also bring speed-ups to the complete application~\cite{GeantV}. As
-far as architectural modifications are concerned, the design of the
-Geant4 track-status machinery prevents the introduction of fine-grained
-track parallelism. It is still not clear whether this kind of
-parallelism can bring any substantial speed-up, due to the overheads;
-however, making the Geant4 kernel and physics processes stateless would
-certainly provide the necessary starting points for further R\&D in that
-direction. Most of these investigations have started and it is hoped to
-see some first results on a year time-scale. More effort, however, will
-be required to integrate any eventual changes in the production code in
-a timely manner. The GeantV Vector prototype has allowed the assessment
-of the achievable speedup using a novel, vectorized approach to particle
-transport. The demonstrator of a full electromagnetic shower in
-realistic (CMS) geometry has been implemented and the comparison to a
-similar Geant4 application has been performed. The conclusion of that
-work showed that the main factors in the speedup seem to include better
-cache use and tighter code, while the vectorisation impact was much
-smaller than hoped for. This unimpressive impact is due to a set of
-factors, including the fraction of the current algorithms that could be
-vectorised, unresolved challenges for the gather/scatter and tail
-handling (to keep the number of events in flight within bound) needed
-for large number of shapes, and of course Amdahl's Law applied to
-vectorisation where some bottleneck (for example geometry navigation)
-could not be tackled without a major additional long term effort~\cite{GeantV}.
-At the same time libraries developed in the context of the
-GeantV R\&D, e.g. VecGeom, have been successfully integrated in Geant4
-and ROOT benefitting the whole HEP software community~\cite{CMS01}. The
-cost-over-benefit of more vectorisation of the code has not been deemed
-worth the necessary investment also due to the speculation that a
-complete rewrite of the physics base could become necessary to fully
-exploit it. As a result investing in the optimisation and modernisation
-of the Geant4 code has gained even more relevance.
-
-\hypertarget{experiments-applications-and-optimisation-of-the-use-of-geant4}{%
-\subsection{Experiments applications and optimisation of the use of
-Geant4}\label{experiments-applications-and-optimisation-of-the-use-of-geant4}}
-
-All LHC experiments have dedicated simulation frameworks making use of
-the Geant4 toolkit for modelling their detectors and the physics
-processes occuring in the experimental setups. Every experiment
-configures and uses what is provided by Geant4 for their specific needs.
-Effort has been spent by each experiment since the publication of the
-Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} to benchmark their 
-simulation code and to maximally
-exploit the options provided by the Geant4 toolkit to optimise the
-performance of their applications~\cite{ATLAS-G4OPT}. All of the various handles provided
-are being continuously explored and adopted when deemed useful: range
-and physics cuts have been reviewed and customised by all experiments,
-shower libraries adopted as baseline in the forward region by ATLAS and
-CMS, stepping granularity optimized, magnetic field caching
-investigated, neutron russian roulette used for dedicated simulations.
-All of them can have major impacts on simulation time and sharing
-knowledge and experience between experiments can be of high benefit even
-if each experiment has to find its optimal solution.
-
-ATLAS, CMS and LHCb have integrated the Geant4 event-based
-multi-threading features within their own multi-threaded frameworks,
-harmonizing their different architectural choices with the Geant4
-implementation~\cite{CMS02, LHCb01, ATLAS01}, and have or are planning to
-use them in their production environment to gain access to a larger set
-of distributed resources. ATLAS has been running their multi-process
-simulation framework on HPCs systems for production in the
-last years~\cite{ATLAS02, Benjamin:2696330} 
-while CMS has pioneered the use of their multi-threaded
-simulation on available resources including HPCs~\cite{CMS03}. Both
-approaches have allowed to exploit a higher number of computing
-resources by gaining access to systems with limited memory.
-
-\hypertarget{fast-simulations}{%
-\subsection{Fast Simulations}\label{fast-simulations}}
-
-The so-called `fast simulation' techniques consist of replacing the
-tracking of individual particles through (a part of) the detector,
-including all the physics processes they would undergo, by a
-parameterization where the detector response is produced directly as a
-function of the incoming particle type and energy. An example of such a
-parameterization was implemented many years ago in the context of the H1
-experiment~\cite{H1Gflash}. This parameterization, available within the
-Geant4 toolkit under the name of the GFLASH library, became the starting
-idea for several `custom' implementations, specific for other
-calorimeters. Those were interfaced to experiments' simulation
-applications through the available `hook' in the Geant4 toolkit. The LHC
-experiments implemented, for example, dedicated parametrized response
-libraries for some of their calorimeters. Recently, an effort has been
-invested in Geant4 to generalize the parameterization formulae, and to
-implement automatic procedures of tuning the parameters for specific
-detector geometries. This kind of `classical' fast simulation, which
-describes the particle shower shape using some complex mathematical
-functions with several parameters, will remain an important tool;
-however, it is also clear that their overall precision for the different
-energy values, especially for highly granular and complex calorimeters
-will always be relatively low. The recent developments of Deep
-Learning-based techniques have opened an exciting possibility of
-replacing those `classical' parameterisations by trained neural networks
-that would reproduce the detector response. This approach consists of
-training generative models such as Generative Adversarial Networks
-(GAN), Variational Auto-Encoders (VAE) or Autoregressive Generative
-Networks on the `images' of particle 
-showers~\cite{ML001-GAN, aishik_ghosh_2019_3599705, GAN-ATLAS}. The energy
-deposition values in the calorimeter cells are considered as `pixels' of
-the images that the network is supposed to reproduce. The first studies
-and prototypes have shown very promising results, however, the
-generalization of the developed models (for different calorimeters, with
-different particles incident at different angles with different
-energies), still requires further effort. Detailed physics validation of
-those tools and understanding of the capability to reproduce the
-fluctuations is a prerequisite towards `production-quality' fast
-simulation libraries. This work is already ongoing, but will require,
-over the next few years, more effort to be invested combining the
-physics and Machine Learning expertise. Given that the training of the
-deep learning algorithms usually requires using sufficiently large MC
-samples produced using standard techniques, the development of the novel
-tools does not necessarily remove the need to speed up Geant4.
-
-It is worth pointing out that different fast simulation techniques have
-already been extensively used for LHC simulation campaigns. Effort has
-been spent in the experiments frameworks to combine as much as possible
-fast and detailed implementation. In particular ATLAS and CMS are using
-in their baseline production shower libraries for the modelling of
-calorimeters in the forward region in combination with the detailed
-simulation of the rest of the sub-detectors. In LHCb the re-use of part
-of the underlying events for specific signals of interest has been
-established in production of samples when appropriate, with particular
-care paid to ensure keeping under control bias on statistical
-uncertainties~\cite{LHCb02}, its applicability in ATLAS and CMS could be
-explored for specific cases. Similarly, the re-use of simulated or real
-events to mimic the background to hard-scatter events is also exploited,
-where additional challenges exist concerning storage and I/O due to the
-handling of the large minimum bias samples needed to model the
-additional interactions.
-
-All LHC experiments have explored the use of multi-objective regression
-and generative adversarial networks (GANs)~\cite{GAN-ATLAS, GAN-LHCb}, they are now in the process of investigating the use of
-fast simulations for other types of detectors than calorimeters, e.g.
-Cerenkov based systems~\cite{Lamarr} and Time Projection Chambers~\cite{ALICE-TPC}. While implementations of fast simulations of given
-sub-detectors is specific to their technology and experimental
-environment and as such the responsibility of each LHC experiment, an
-area of potential common investigation is frameworks for fast tuning and
-validation.
-
-The use of fully parametric response of experimental setups at the level
-of reconstructed objects (tracks and clusters) for rapid evaluation of
-physic's reach is also exploited by the experiments with generic
-frameworks for fast simulation of typical collider detectors being used~\cite{bib-Delphes, PGS4}. 
-The advantage of such frameworks is to allow
-simultaneous studies for different experiments as well as their
-potential use by the theory community. Experiments specific fully
-parameterized simulations providing reconstructed objects compatible
-with the experiment's analysis frameworks for systematic verifications
-have also been emerging~\cite{Lamarr} .
-
-It is reasonable to expect that the full variety of simulation options,
-from `smeared reconstructed quantities' to parameterisation for specific
-detectors to the detailed Geant4, will need to be exploited by the
-experiments for different tasks. Seamless ways of providing them in a
-transparent integrated way in the experiments' simulation frameworks
-should continue to be pursued with Geant4 supporting hooks for such
-integrations.
-
-\hypertarget{technical-challenges-software-and-computing-1}{%
-\subsection{Technical Challenges (Software and
-Computing)}\label{technical-challenges-software-and-computing-1}}
-
-The hardware landscape has been always setting direction as far as
-software evolution was concerned. The recent adaptation of the
-simulation code to multithreading (MT) turned out to be relatively
-straightforward, but still took several years to adopt and validate.
-Geant4 can now run in the MT mode efficiently, distributing the events
-between the threads and sharing resources like the geometry description
-or physics data and thus reduce the memory footprint of a simulation
-while maintaining its throughput performance. The efficient utilization
-of Single Instruction Multiple Data (SIMD) architectures, on the other
-hand turned out to be more challenging. Not only was a limited part of
-the code `vectorizable' but also, as the already mentioned GeantV R\&D
-has shown, overheads related to gathering the data for vector processing
-presented a significant challenges (tail handling, work balancing across
-threads, etc.) to efficiently (development time wise and run time wise)
-profit from it~\cite{GeantV}.
-
-The most recent `revolution' in the computing hardware systems consists
-of using Graphics Processing Units (GPUs) for general computing. Several
-attempts have been made to port specific simulation code to GPUs. A few
-of them turned out to be successful leading to factors of several
-hundred or thousand of speedup~\cite{GPU-MPEXS-DNA, Hybrid-Gate, Opticks}, however, they were always
-limited to a very specific simulation problem, far from what is
-addressed by a general HEP simulation. The application of GPUs has been
-very successful in restricted domains like medical physics simulation,
-neutron transport~\cite{SHIFT01} or optical photon transport~\cite{CHEP-Juno}. 
-In those applications, the type of particles considered
-is very limited (often just one type, with no secondary particle
-production), the set of physics processes is reduced and the geometry is
-often extremely simple compared to the LHC detectors. A natural
-`extrapolation' of those approaches to a general HEP simulation is very
-difficult to imagine, because the stochasticity of the simulated events
-would immediately lead to divergences as far as different GPU threads
-are concerned, not to mention, simply the feasibility of porting the
-whole simulation to the GPU code for which specific expertise would be
-needed. On the other hand, running only very small pieces of the
-simulation application on GPUs does not seem to be efficient either, as
-gathering data and transferring it from the host to the device and back
-may strongly limit any possible performance gain, similarly to what was
-seen with the GeantV vectorisation R\&D. The most plausible approaches
-seem therefore to lead in the direction of specialized libraries (like
-those described above in the context of speeding up Geant4 simulation on
-the CPUs) that would perform the complete simulation of specific
-sub-detectors (for instance EM calorimeter, for specific incoming
-particles or Cherenkov-based detectors for the optical processes) on the
-device. Such libraries could be the right compromise between the
-complexity of the algorithm that needs to be ported to the device and
-the overall time that is now spent (and could be gained) on the
-corresponding of the full simulation on the CPU. The investigation in
-those directions are starting now, but will certainly require a
-considerable effort to be further invested before being able to judge
-their real impact. The first steps that are currently being undertaken
-consist of implementing, on GPUs, prototypes based on VecGeom geometry
-and navigation capabilities. If those prototypes turn out to be
-successful, the next challenge will consist of adding tracking and
-eventually a limited subset of physics models. While Geant4 is a toolkit
-capable of addressing different modeling and simulation problems and
-contains many features and capabilities allowing for user access to
-detailed information for a large variety of use cases and scientific
-communities, this GPU development might turn into specialized transport
-modules, stripped of some features which would be expensive to implement
-or support efficiently on GPUs. Another interesting avenue consists of
-exploring the application of some vendor libraries (like Nvidia Optix).
-Those libraries, originally developed for ray tracing, have several
-similarities with general particle transport and if applied
-successfully, could lead to major speed-ups. All those efforts certainly
-require a major investment and new developers, expert in those
-technologies, to join the R\&D work.
-
-\hypertarget{other-activities}{%
-\subsection{Other Activities}\label{other-activities}}
-
-Common digitisation efforts would be desirable among experiments, with
-advanced high-performance generic examples, which experiments could use
-as a basis to develop their own code. Nevertheless the large variety of
-detectors' technologies used by the experiments reduces its
-applicability. Digitisation not yet being a limiting factor in terms of
-CPU requirements also does not urge developments in this area.
-
-Simulated samples are often processed through the reconstruction as real
-data. As such in some cases they require the emulation of hardware
-triggers. Hardware triggers are based on very specific custom devices
-and a general approach does not seem feasible even if some
-parametrization could be generalised.
-
-\hypertarget{outlook}{%
-\subsection{Outlook}\label{outlook}}
-
-To conclude, it is important to realise that there is a considerable
-risk for the simulation to become a major limiting factor as far as new
-physics discoveries are concerned if no serious effort is invested in
-R\&D. We need the concerted effort of different experts in physics,
-Machine Learning and GPUs. There are too many unknowns to focus on only
-one direction, so a lot of different prototyping is needed. We need to
-further develop Geant4 while working on fast simulation algorithms and
-doing extensive R\&D on leveraging compute accelerators. While the R\&D
-activities are taking place, one can not forget that it is still
-necessary to support the existing code and make sure that sufficient
-funding and staffing is provided for maintenance and development of the
-physics algorithms, as well as for adapting the code to the ongoing
-hardware, operating system and compiler evolution.
-
-It is also important to stress once again the need of not only ongoing
-R\&D investigations and independent prototyping to explore novel ideas,
-but also a continuous integration activity of new runtime performance
-improvements and solutions into the Geant4 code base and experiments
-applications to profit from each improvement as quickly as possible. An
-effort needs to be invested into making it possible to evolve it to meet
-the HL-LHC needs without compromising the production quality of the
-Geant4 code used by the experiments. We need to ensure that new
-developments resulting from the R\&D programs can be tested with
-realistic prototypes and, if successful, then integrated, validated, and
-deployed in a timely fashion in Geant4 and adopted by the experiments.
-The strategy adopted in the successful integration in Geant4 of VecGeom
-libraries developed in the context of the GeantV R\&D can provide a
-working example of how to proceed to provide some incremental
-improvements to existing applications.
-
-No single solution appears at the moment to provide by itself the
-processing gains required by HL-LHC, nevertheless if a novel approach
-emerges that could do so it should be pursued and carefully evaluated in
-terms of gains against the disruption of a complete reimplementation of
-how events are simulated.
+\input{input_simulation}
 
@@
 
-\hypertarget{data-analysis}{%
-\section{Data Analysis}\label{data-analysis}}
+\input{input_analysis}
 
-\hypertarget{key-analysis-computing-challenges-at-the-hl-lhc}{%
-\subsection{Key analysis computing challenges at the
-HL-LHC}\label{key-analysis-computing-challenges-at-the-hl-lhc}}
-
-Examination of collision data is in essence the primary objective of the
-experimental collaborations, coming at the end of the data
-preparation/simulation and reconstruction chain. The stage of analysis,
-starting in most cases from a standard data format generically referred
-to as ``Analysis Object Data'' (AOD) containing reconstructed physics
-object data, and producing physics results as the end product, poses
-unique computing challenges. The scope of analysis data processing is
-broad, encompassing the production of derived data formats as well as
-end stage analysis actions such as producing histograms from those
-intermediate datasets and visualising the final results. In the
-following, attention is focused on the computing challenges related to
-analysis for the ATLAS and CMS experiments.
-
-Today, ROOT format AOD files and derived datasets take up the lion's
-share of disk resources, filling approximately 80\% of the total data
-volume for both ATLAS and CMS \cite{Ref1,Ref2}. To serve precision analyses
-that require large event statistics, the HL-LHC will increase the
-recorded event rate by an order of magnitude, compared with a projected
-\textasciitilde 10\% annual growth of storage resources. The large pile-up in HL-LHC
-collisions will compound the storage challenge by inflating the data
-size per event. Reducing the storage footprint of data analysis is
-therefore of paramount importance.
-
-The effective CPU needs for end-stage analysis payloads are typically
-orders of magnitude lower than those in simulation and reconstruction.
-When scaled up to O(100) analyses per year, each processing the input
-data dozens of times a year, the total CPU consumption at the HL-LHC is
-projected to be around 10\% of total experimental computing \cite{Ref1,Ref2}.
-Central production of analysis data formats may account for another
-10-30\%. It is worth noting at this point that analysis data access
-patterns tend to be more chaotic than preceding stages, which can
-increase the effective resource needs by large factors.
-
-More significant than the global storage and computational costs is the
-job turnaround time, closely tied to the ``time to insight'', which is a
-strong limitation on the speed of analysis publication. One significant
-constraint is the need for full coverage of the input data, which is
-inextricably linked to weaknesses in book-keeping tools that make it
-difficult, if not impossible, to make meaningful studies on subsets of
-the data. This full coverage requirement is clearly limited by computing
-infrastructure load and downtime. Another challenge is the
-multiplication of computing resource demands from the assessment of
-systematic uncertainties, which involves processing many alternate
-configurations and input data, inflating both the CPU load and storage
-footprint. Improvement of community standards for metadata handling and
-uncertainty handling will be very important for HL-LHC.
-
-A last concern is that, while simulation and reconstruction code is
-mostly optimised and developed under greater scrutiny, analysis code is
-often produced with emphasis on quick results and less emphasis on code
-quality or performance. Over time, this may result in inefficient,
-under-documented code that exacerbates the disk and CPU shortfall and
-poses a major difficulty when software is repurposed for a new analysis.
-Add to this a growing array of alternative toolkits from the data
-science community, in particular for machine learning, and it is clear
-that the entropy of the analysis code ecosystem has become a major
-challenge in itself.
-
-The following sections expand on the major challenges for analysis
-software, including those identified above. An overview is given of
-ongoing work on common tools that alleviate these problems, followed by
-an outlook on R\&D prospects that would more significantly transform the
-HL-LHC analysis model for major resource savings.
-
-\hypertarget{analysis-processing-and-workflows-current-problems-and-solutions}{%
-\subsection{Analysis processing and workflows: current problems and
-solutions}\label{analysis-processing-and-workflows-current-problems-and-solutions}}
-
-To make data analysis tractable, AOD data are typically reduced both by
-saving only relevant information for each event, which may require
-transformations (object calibration, computation of derived variables)
-beyond simply discarding information, and by skimming out only events
-that are interesting. In post-reconstruction data formats, lossy
-compression is also in use and being optimised. Coordination of the data
-reduction phase is a key point for the organization of data analysis
-processing at HL-LHC.
-
-Two approaches have been followed during LHC Run 2 by ATLAS and CMS:
-ATLAS analysis trains \cite{Ref4} and the CMS mini-AOD, a highly reduced and
-standardised event content. Analysis trains place software payloads
-tailored for specific analyses (carriages) into a periodically scheduled
-centralised execution, amortising data staging costs and sharing some
-common processing steps \cite{Ref3}. Standardised event contents are instead
-common reductions that satisfy the needs of the bulk of analyses,
-avoiding duplication of commonly selected events in multiple datasets.
-Neither of these approaches provides exhaustive coverage of analysis use
-cases on the experiments and alternative data formats are needed for a
-small but significant (10-20\%) fraction of analyses requiring custom
-reconstruction that may be CPU-intensive. These exceptions to the rule
-are expected to remain at the HL-LHC.
-
-It is worth considering the viability of the arguably more flexible
-option to have analysis trains run over the reconstruction output at the
-HL-LHC. Extrapolating using a simple model, ATLAS finds that they would
-produce 35 PB/year of data AOD and 215 PB/year of MC \cite{Ref1}. In
-comparison, a reduced format (10 kB/event) made from the AOD would
-result in 0.5 PB/year for data and 2 PB/year for MC. Production of this
-reduced format using a Data Carousel model \cite{Ref1} allows the much
-larger AOD to reside on tape, providing significant savings in disk
-storage. Experience on CMS shows that further reduction (down to
-\textasciitilde2 kB/event for the nano-AOD) could be possible while
-still supporting a majority of physics analyses \cite{Ref5}. File size
-reductions also permit multiple copies to be held on disk for more
-efficient parallel processing. It is clear that as many analyses as
-possible need to be fed into this standardized analysis event-content
-pipeline if analysis computing resources are to be kept under control.
-Nevertheless, analysis trains or any similar means of synchronising data
-access could be applied to this highly reduced format. It is worth
-noting the LHCb analysis model already in Run 3 has to forego the luxury
-of reprocessing Raw data. This constraint from sheer event rate is not
-without merit as it greatly simplifies the data processing and analysis
-models and removes one tier of offline data products.
-
-Careful attention needs to be paid to the assessment of systematic
-uncertainties, particularly with regard to event content
-standardisation, to avoid creating many derived outputs that differ only
-minimally. Analysis models that do not require all event information to
-be present in a single file, instead leveraging ROOT ``friend trees'' or
-columnar storage, may be a way to reduce this duplication both in the
-case of uncertainty calculations and reconstruction augmentations, but
-will require development of robust strategies for keeping the
-augmentations in sync with the core events. It is noted that the ROOT
-data format is quantifiably very efficient for the processing model in
-HEP \cite{Ref16}, and any competitors would have to achieve a similar level
-of performance.
-
-Access to metadata, defined as cross-event information, is clearly a
-weak point in the handling of the multiple analysis datasets originating
-from the LHC. Metadata may include information such as book-keeping of
-processed events, detector conditions and data quality, \emph{a
-posteriori} measured scale factors, and software versioning. This
-information is often scattered in multiple locations such as:
-``in-file'' data, remote database resources, shared-area txt/root files,
-twiki pages or even in e-mail. The lack of proper metadata integration
-and a coherent interface leaves analyses exposed to any problems in data
-processing. This in turn results in data completeness demands that
-simply will not work at the HL-LHC, where the huge datasets all but
-guarantee that some fraction of the data will be unavailable at any
-given point in time.
-
-At the HL-LHC, as energy and luminosity conditions are quasi stable
-across the years, datasets from multiple years with different conditions
-will need to be analyzed in a coherent way. The analysis software will
-then need to be able to fetch and use the proper metadata automatically
-while today's analysis software often requires dedicated configuration
-and tuning for each data-taking period. Belle II have taken some steps
-to address this by using the ``global tag'' concept commonly used in
-reconstruction and applying it to analysis, allowing users to better
-organise and store their analysis metadata. Nevertheless, until all
-metadata is organised under the same global tag umbrella and accessed
-through a coherent interface, which is extremely challenging, the
-problem remains. Improvements here are necessary both to permit
-efficient studies on partial datasets, and to reduce the risk of user
-error in metadata access.
-
-The growing complexity of analysis codes makes them extremely fragile
-(i.e. bug-prone), hardly reusable, and unsuited for analysis
-preservation needs. The current best efforts at analysis preservation
-are based on a snapshot of the full analysis setup that can ensure the
-possibility to re-run the code on the original dataset in a few years.
-This may allow a future analyser to reproduce previous results, but will
-likely not provide any understanding, for the future analyser, about
-what the analysis was effectively doing. Another use case is to re-use
-the analysis code on new data, be that the same dataset with improved
-calibrations or additional data that will improve the precision of the
-measurement. This brings significant additional challenges, not least
-related to metadata.
-
-Nevertheless, these preservation efforts encourage better organisation
-of the analysis as a workflow, and promote the use of version control
-tools together with continuous integration, which offers a natural route
-to improving analysis code quality. In addition to these code quality
-measures, a longer-term solution for code complexity may be the adoption
-of a Domain-Specific Language (DSL) \cite{Ref10}, discussed in more detail
-later, a model that would have physicists write logical descriptions of
-their analyses, leaving low-level code generation and hardware
-optimisation to a backend. Thus analysis design could be quickly
-understood and shared, sidestepping the usual dearth of documentation,
-while simultaneously isolating analysis design from implementation and
-hardware, providing a natural means of analysis preservation. An
-interesting development in this direction is the Reana platform \cite{Ref21}
-that supports multiple backends and uses a DSL to provide a framework to
-run an analysis. Already a valuable resource as a tool for analysing
-open data, it is interesting to ask what would happen if every analyst
-first had to structure their analysis and capture their workflows before
-they started submitting jobs.
-
-\hypertarget{plans-and-recent-progress}{%
-\subsection{Plans and recent progress}\label{plans-and-recent-progress}}
-
-Most of the problems identified above were already identified in the
-community white paper \cite{Ref17} from 2017 and since then some progress
-was made prototyping new technologies to enable data analysis at the
-HL-LHC.
-
-A first branch of new technologies is that of efficient data analysis
-processing in the last steps of the analysis, i.e. when event data needs
-to be selected, derived and aggregated into histograms. Several
-platforms developed by industry or data science have emerged in recent
-years to quickly aggregate large datasets with parallel execution
-(either on clusters of computers or simply on multi-core servers). Many
-of those tools have been tested to understand the feasibility of usage
-in the HEP context providing fresh ideas and new paradigms that have
-stimulated development within the HEP toolkit.
-
-In particular, the Python ecosystem, implementing many such solutions
-(e.g. numpy, pandas dataframes), is emerging as a possible alternative
-to HEP's traditional C++ environment. This is thanks largely to a
-combination of its versatility as a language, allowing rapid
-proto-typing, and the ability to out-source compute-intensive work to
-more performant languages like C and C++. This is similar to the
-experimental software frameworks where Python is used as steering code
-and C/C++ is used as an efficient backend implementation. The typical
-Belle II analysis starts with Python using a custom DSL to specify
-particle selections and algorithms implemented in C++ and define output
-ntuples. The ntuples are typically analysed by the PyHEP toolkit
-\cite{Ref6}, and packages from the wider data science community. Deeper
-integration with external python tools is also particularly important
-for enabling state-of-the-art machine learning frameworks such as
-PyTorch and TensorFlow, whose relevance in analysis is likely to grow in
-the long term. Meanwhile PyROOT, which allows the use of any C++ class
-from Python, has the potential to provide a coherent analysis ecosystem
-with an effortless integration of HEP specific tools written in C++ with
-industry python big data toolkits.
-
-The ROOT package, i.e. the foundation and workhorse for LHC analysis,
-has been the focus of substantial development. On the data storage
-front, the ROOT team demonstrated better data compression and
-accelerated reading for a wide variety of data formats. Information
-density drives compression and this can vary massively between
-experiments and analyses. This new format called RNTuple \cite{Ref9}, an
-evolution of the TTree file format, shows robust and significant
-performance improvements (1-5x faster), which could potentially save
-significant storage space (10-20\%) for the HL-LHC. Another area of
-progress, and potential consolidation in ROOT I/O, is lossy compression,
-where the number of significant bits for a quantity may be far fewer
-than a standard storage type's mantissa, enabling further savings in
-storage space.
-
-Inspired by data science tools, the event processing framework was
-extended with DataFrame-like functionality (RDataFrame) implementing a
-declarative approach to analysis processing in either Python or C++
-(always executed in C++), which natively exploits multi-core
-architectures \cite{Ref7} and could be extended to accelerators. Used from
-Python, it allows to pre-process data in C++, and export to numpy.
-Flexibility in interfacing RDataFrame to non-ROOT data formats has
-allowed the ALICE collaboration to prototype its Run 3 analysis model on
-this new technology. Growing use of Machine Learning was also
-anticipated, and the TMVA toolkit has been improved with substantial
-optimisations, more complex deep learning models and enhanced
-interoperability with common data science tools \cite{Ref14,Ref15}  which is of
-particular importance for inference.
-
-A complementary approach provided in Python is the toolchain of uproot,
-awkward arrays and coffea \cite{Ref8} for efficient columnar operations and
-transformations of the data. These function as a lightweight end-stage
-analysis environment that provides only the elements necessary for
-pre-processing ROOT inputs and converting them into formats used by the
-standard ML frameworks, typically numpy. Lightweight distribution using
-package managers such as pip or conda allows for rapid setup and
-extension with other Python tools. Good performance appears to have been
-achieved within the scope of these projects, which has been kept
-focused, but for the purposes of fair comparison with other existing or
-emerging options, defining clear benchmarks for I/O and processing speed
-would be beneficial.
-
-A further feature linked to the growing use of Python is the use of
-``notebook'' technology such as Jupyter. This permits quasi-interactive
-exploration where the annotated history including formatted outputs and
-graphics can be saved and shared with collaborators for reproduction and
-adaptation. Although not a substitute for command line scripts in
-well-tested and complex workflows, notebooks make an effective vessel
-for software education and have been leveraged as a powerful frontend
-for access to facilities including CERN's SWAN \cite{Ref20}.
-
-\hypertarget{prospects-and-rd-needs}{%
-\subsection{Prospects and R\&D needs}\label{prospects-and-rd-needs}}
-
-As the complexity of analysis data handling and processing grows,
-developing efficient and robust code using a steadily increasing number
-of tools running on ever more heterogeneous hardware becomes more and
-more difficult. In addition, analysis code is typically reimplemented by
-tens of individuals, in different experiments, analysis teams or
-institutes mostly performing the same kind of operations, i.e. data
-reduction, plotting, variation of systematic uncertainty, fitting, etc.
-It is clear that tools and frameworks for performing these repeated
-operations need to provide very efficient interfaces to avoid analysis
-code bloat. Declarative rather than imperative programming is becoming
-visibly more prevalent both in and outside of the field as it naturally
-provides efficient syntax.
-
-DSLs are a generalisation of this declarative concept and have already
-demonstrated their ability to simplify analysis code {[}{]}. Not only do
-they allow users to express operations concisely, improving
-comprehension and reusability, DSLs provide a natural layer of
-insulation against hardware evolution as low-level, optimised code
-generation for multiple different platforms and architectures is
-delegated to the tool and framework experts. Prototypes of DSLs have
-been developed in recent years \cite{Ref10} for the event processing and
-histogram production parts of the analysis workflow, and even built into
-experimental frameworks to perform high-level analysis \cite{Ref19}. Similar
-efforts can also be investigated in the context of data interpretation
-and statistical analysis and an interesting example here is the so
-called ``combine tools'' developed by CMS and based on RooFit for the
-Higgs discovery) {[}ref{]}. Descriptions of a full analysis in terms of
-workflows could then tie together the analysis at the top level \cite{Ref18,Ref21}. With some effort, the HEP community may be able to converge on
-effective tools that provide a common solution, ideally across
-experiments.
-
-There is an obvious need for good integration with analysis backends,
-whether these be local CPU, batch or grid resources. At some stage in
-the analysis process, interactivity or fast turn-around is needed. While
-typical grid task completion has a time scale of one day to one week,
-fast turn-around exploration requires answers within a few seconds up to
-perhaps a few hours in order to keep people productive. As analysis
-tasks are often IO-intensive and not necessarily-CPU intensive, the
-right tradeoff between CPU and high bandwidth storage should be
-carefully studied. More detailed studies of resource usage and its
-evolution, especially as pertains to the need for heterogeneous compute,
-is needed to ensure the right resource balance is available in the
-HL-LHC computing infrastructure. This will require more interaction
-between the analysis and infrastructure communities.
-
-As a proof of principle, an unoptimized Higgs discovery analysis has
-been rerun on Google Cloud in a few minutes as the infrastructure was
-able to quickly provide tens of thousands of CPUs with guaranteed
-bandwidth to storage of 2 Gb/coresec \cite{Ref12}. Scaling this to
-sustained, diverse analysis payloads will be far from trivial, so the
-question becomes how we can compare different approaches to provisioning
-analysis. Eight initial benchmark challenges have been defined to set
-the scope that should be addressed by analysis systems \cite{Ref11}. What
-remains to be demonstrated is the capability of such systems to scale to
-full analysis complexity, including tasks such as the handling of
-systematic uncertainties and data-driven background estimation, and
-these are yet to be incorporated into the benchmarks.
-
-In contrast to the activity on analysis languages and infrastructure,
-metadata is a topic that is almost entirely neglected by the analysis
-community, even while computing experts have long understood its
-importance. That is likely due to the absence of a single coherent
-stake-holder for and provider of analysis metadata. Event data comes in
-a ROOT file, but already within a single experiment there are numerous
-sources of metadata that operate on vastly different timescales and this
-poses a significant challenge to standardisation. As analyses and
-detectors grow in complexity, further divergence may amplify the
-challenge, making it critical to begin addressing this problem promptly.
-Nevertheless, the drive to capture a complete analysis must drag
-metadata with it, and the gleeful adoption of the declarative paradigm
-will make database-like interfaces seem less scary than they once were
-and may finally allow a more complete integration with analysis code. I
-have a dream.
-
-One final comment in reviewing recent developments is worth noting.
-Exposure of the HEP community to industry-standard data science tools
-has undoubtedly been a force for good and a net gain for the community,
-providing fresh ideas and new ways of working - RDataFrame is an
-excellent example. It is clear that the larger and better-supported data
-science community will continue to innovate and produce tools that HEP
-will benefit from, both by using directly and by borrowing their ideas
-for our own tools. Taking machine learning as an example, it is
-extremely likely that industry will blaze a trail that HEP would do well
-to follow and we should continue to build bridges to those tools.
-Equally though, HEP will continue to have its own unique challenges and
-it's less likely that ultra-performant machine learning inference will
-be top of the data scientists wishlist. Given the challenges presented
-by extreme-throughput analysis at the HL-LHC, the right balance must be
-found between software development within the community and relying on
-external toolkits.
-
-Finally, attention must be given to ensuring the longevity of any
-projects that the HEP community decides to adopt en masse. Whether by
-specific institutes, funding schemes or other collective programmes,
-responsibility for key software will need to be ensured in the long
-term. The development of a secure support base will therefore need to be
-considered for emerging projects that show significant promise.
-
-\hypertarget{summary}{%
-\section{Summary}\label{summary}}
-
-The main challenges and R\&D lines of interest have been discussed
-above. There are clearly identified priority topics for the different
-areas in the next few years. We note that longer term R\&D, into areas
-such as quantum or neuromorphic computing, should happen at a low level,
-but are almost certainly too far from production to deliver improvements
-for Run 4 of the LHC.
-
-\hypertarget{physics-event-generators-1}{%
-\subsection{Physics Event Generators}\label{physics-event-generators-1}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  \begin{quote}
-  Gain a more detailed understanding of current CPU costs by accounting
-  and profiling.
-  \end{quote}
-\item
-  \begin{quote}
-  Survey generator codes to understand the best way to move to GPUs and
-  vectorized code, and prototype the port of the software to GPUs using
-  data-parallel paradigms.
-  \end{quote}
-\item
-  \begin{quote}
-  Support efforts to optimize phase space sampling and integration
-  algorithms, including the use of Machine Learning techniques such as
-  neural networks.
-  \end{quote}
-\item
-  \begin{quote}
-  Promote research on how to reduce the cost associated with negative
-  weight events, using new theoretical or experimental approaches.
-  \end{quote}
-\item
-  \begin{quote}
-  Promote collaboration, training, funding and career opportunities in
-  the generator area.
-  \end{quote}
-\end{enumerate}
-
-\hypertarget{detector-simulation-1}{%
-\subsection{Detector Simulation}\label{detector-simulation-1}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  \begin{quote}
-  Undertake a detailed performance analysis of current production
-  detector simulation code, to better understand where performance
-  limitations in data and cache locality arise from.
-  \end{quote}
-\item
-  \begin{quote}
-  Improve current codes through refactoring and specialised HEP models
-  to avoid bottlenecks identified,
-  \end{quote}
-\item
-  \begin{quote}
-  Develop further fast simulation models and techniques, including
-  machine learning based options in addition to optimised parametric
-  approaches as well as common frameworks for their tuning and
-  validation.
-  \end{quote}
-\item
-  \begin{quote}
-  Develop GPU codes for tackling very specific, time consuming, parts of
-  the simulation for HEP. Specifically calorimetry, including geometry,
-  field and electro-magnetic physics processes.
-  \end{quote}
-\item
-  \begin{quote}
-  Develop integration prototypes to exercise and benchmark the
-  simultaneous use of GPU specific libraries with CPU-based software
-  (e.g. Geant4) to cover all particles and processes required in
-  experiments simulation frameworks.
-  \end{quote}
-\end{enumerate}
-
-\hypertarget{reconstruction-and-software-triggers-1}{%
-\subsection{Reconstruction and Software
-Triggers}\label{reconstruction-and-software-triggers-1}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  \begin{quote}
-  Develop fast and performant reconstruction software, making optimal
-  use of hardware resources. A particularly crucial development for
-  HL-LHC is that of pattern recognition solutions (including timing
-  information, if available) that can withstand the increase of pile-up
-  at HL-LHC. Wherever possible, solutions should be uniform for online
-  and offline software, so the experiments can benefit from consistent
-  event selection between trigger and final analysis and from more
-  effective real-time analysis results.
-  \end{quote}
-\item
-  \begin{quote}
-  Develop solutions that can exploit accelerators (FPGA, GPU) for
-  trigger and reconstruction, including general-purpose interoperability
-  libraries with a focus to generalizing and mitigating their
-  dependencies on architecture-specific codes.
-  \end{quote}
-\item
-  \begin{quote}
-  Further promote activities in the area of data quality and software
-  monitoring, as the quality of the recorded/reconstructed data is
-  paramount for all physics analysis.
-  \end{quote}
-\item
-  \begin{quote}
-  Support efforts towards the implementation of machine learning models
-  in experimental framework (C++) to enable more widespread use in
-  trigger and reconstruction software.
-  \end{quote}
-\end{enumerate}
-
-\hypertarget{data-analysis-1}{%
-\subsection{Data Analysis}\label{data-analysis-1}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  \begin{quote}
-  Identify a CPU-efficient data reduction and storage model that serves
-  the majority of analyses with a footprint of O(1kb/event), retaining
-  the flexibility to accommodate the needs of specialised analyses.
-  \end{quote}
-\item
-  \begin{quote}
-  Streamline analysis metadata access and calibration schemes, and
-  provide effective book-keeping for fractional datasets.
-  \end{quote}
-\item
-  \begin{quote}
-  Develop cross-experimental tools establishing declarative syntax
-  and/or languages for analysis description, interfaced with distributed
-  computing backends.
-  \end{quote}
-\item
-  \begin{quote}
-  Define and improve schemes for interoperability of end-stage
-  experimental software with data science/machine learning frameworks.
-  \end{quote}
-\end{enumerate}
-
-Undertaking this programme of priority topics and R\&D requires, first
-and foremost, investment from funding agencies to support developments.
-A new generation of developers is also needed to refresh and regenerate
-the long lived software projects on which the field so heavily relies.
-The transition from R\&D to production ready software is usually long
-and so it is urgent to undertake this investment now, with the positive
-outcomes ready to become integrated in time for HL-LHC. Regenerating the
-cadre of software developers in HEP also requires support for realistic
-long term career prospects for those who specialise in this vital area.
-Collaboration with industry can certainly be fruitful and is to be
-encouraged to allow early access to technology and communication of HEP
-problems and priorities.
-
-Almost all the domain areas identify the use of compute accelerators,
-particularly GPUs, as being a priority item, which matches our
-expectation of how computing hardware will evolve. As a common problem
-area, this is an obvious area in which expertise should be developed.
-
-The stakeholders in HEP must be able to give their priorities and
-feedback on R\&D areas. Each project has mechanisms for doing this,
-while the HSF can help to provide overall input on prioritisation.
+\input{input_summary}
 

### 2020-04-30 LHCC/2020/2020-01/input_analysis.tex
Break sources into an input file per section

\hypertarget{data-analysis}{%
\section{Data Analysis}\label{data-analysis}}

\hypertarget{key-analysis-computing-challenges-at-the-hl-lhc}{%
\subsection{Key Analysis Computing Challenges at the
HL-LHC}\label{key-analysis-computing-challenges-at-the-hl-lhc}}

Examination of collision data is in essence the primary objective of the
experimental collaborations, coming at the end of the data
preparation/simulation and reconstruction chain. The stage of analysis,
starting in most cases from a standard data format generically referred
to as ``Analysis Object Data'' (AOD) containing reconstructed physics
object data, and producing physics results as the end product, poses
unique computing challenges. The scope of analysis data processing is
broad, encompassing the production of derived data formats as well as
end stage analysis actions such as producing histograms from those
intermediate datasets and visualising the final results. In the
following, attention is focused on the computing challenges related to
analysis for the ATLAS and CMS experiments.

Today, ROOT format AOD files and derived datasets take up the lion's
share of disk resources, filling approximately 80\% of the total data
volume for both ATLAS and CMS \cite{Ref1,Ref2}. To serve precision analyses
that require large event statistics, the HL-LHC will increase the
recorded event rate by an order of magnitude, compared with a projected
\textasciitilde 10\% annual growth of storage resources. The large pile-up in HL-LHC
collisions will compound the storage challenge by inflating the data
size per event. Reducing the storage footprint of data analysis is
therefore of paramount importance.

The effective CPU needs for end-stage analysis payloads are typically
orders of magnitude lower than those in simulation and reconstruction.
When scaled up to O(100) analyses per year, each processing the input
data dozens of times a year, the total CPU consumption at the HL-LHC is
projected to be around 10\% of total experimental computing \cite{Ref1,Ref2}.
Central production of analysis data formats may account for another
10-30\%. It is worth noting at this point that analysis data access
patterns tend to be more chaotic than preceding stages, which can
increase the effective resource needs by large factors.

More significant than the global storage and computational costs is the
job turnaround time, closely tied to the ``time to insight'', which is a
strong limitation on the speed of analysis publication. One significant
constraint is the need for full coverage of the input data, which is
inextricably linked to weaknesses in book-keeping tools that make it
difficult, if not impossible, to make meaningful studies on subsets of
the data. This full coverage requirement is clearly limited by computing
infrastructure load and downtime. Another challenge is the
multiplication of computing resource demands from the assessment of
systematic uncertainties, which involves processing many alternate
configurations and input data, inflating both the CPU load and storage
footprint. Improvement of community standards for metadata handling and
uncertainty handling will be very important for HL-LHC.

A last concern is that, while simulation and reconstruction code is
mostly optimised and developed under greater scrutiny, analysis code is
often produced with emphasis on quick results and less emphasis on code
quality or performance. Over time, this may result in inefficient,
under-documented code that exacerbates the disk and CPU shortfall and
poses a major difficulty when software is re-purposed for a new analysis.
Add to this a growing array of alternative toolkits from the data
science community, in particular for machine learning, and it is clear
that the entropy of the analysis code ecosystem has become a major
challenge in itself.

The following sections expand on the major challenges for analysis
software, including those identified above. An overview is given of
ongoing work on common tools that alleviate these problems, followed by
an outlook on R\&D prospects that would more significantly transform the
HL-LHC analysis model for major resource savings.

\hypertarget{analysis-processing-and-workflows-current-problems-and-solutions}{%
\subsection{Analysis Processing and Workflows: Current Problems and
Solutions}\label{analysis-processing-and-workflows-current-problems-and-solutions}}

To make data analysis tractable, AOD data are typically reduced both by
saving only relevant information for each event, which may require
transformations (object calibration, computation of derived variables)
beyond simply discarding information, and by skimming out only events
that are interesting. In post-reconstruction data formats, lossy
compression is also in use and being optimised. Coordination of the data
reduction phase is a key point for the organisation of data analysis
processing at HL-LHC.

Two approaches have been followed during LHC Run 2 by ATLAS and CMS:
ATLAS analysis trains \cite{Ref4} and the CMS mini-AOD, a highly reduced and
standardised event content. Analysis trains place software payloads
tailored for specific analyses (carriages) into a periodically scheduled
centralised execution, amortising data staging costs and sharing some
common processing steps \cite{Ref3}. Standardised event contents are instead
common reductions that satisfy the needs of the bulk of analyses,
avoiding duplication of commonly selected events in multiple datasets.
Neither of these approaches provides exhaustive coverage of analysis use
cases on the experiments and alternative data formats are needed for a
small but significant (10-20\%) fraction of analyses requiring custom
reconstruction that may be CPU-intensive. These exceptions to the rule
are expected to remain at the HL-LHC.

It is worth considering the viability of the arguably more flexible
option to have analysis trains run over the reconstruction output at the
HL-LHC. Extrapolating using a simple model, ATLAS finds that they would
produce 35 PB/year of data AOD and 213 PB/year of MC \cite{Ref1}. In
comparison, a reduced DAOD\_PHYSLITE format (10 kB/event) made from the AOD would
result in 0.5 PB/year for data and 2 PB/year for MC. Production of this
reduced format using a Data Carousel model \cite{Ref1} allows the much
larger AOD to reside on tape, providing significant savings in disk
storage. Experience on CMS shows that further reduction (down to
\textasciitilde2 kB/event for the nano-AOD) could be possible while
still supporting a majority of physics analyses \cite{Ref5}. File size
reductions also permit multiple copies to be held on disk for more
efficient parallel processing. It is clear that as many analyses as
possible need to be fed into this standardised analysis event-content
pipeline if analysis computing resources are to be kept under control.
Nevertheless, analysis trains or any similar means of synchronising data
access could be applied to this highly reduced format. It is worth
noting the LHCb analysis model already in Run 3 has to forego the luxury
of reprocessing Raw data. This constraint from sheer event rate is not
without merit as it greatly simplifies the data processing and analysis
models and removes one tier of offline data products.

Careful attention needs to be paid to the assessment of systematic
uncertainties, particularly with regard to event content
standardisation, to avoid creating many derived outputs that differ only
minimally. Analysis models that do not require all event information to
be present in a single file, instead leveraging ROOT ``friend trees'' or
columnar storage, may be a way to reduce this duplication both in the
case of uncertainty calculations and reconstruction augmentations, but
will require development of robust strategies for keeping the
augmentations in sync with the core events. It is noted that the ROOT
data format is quantifiably very efficient for the processing model in
HEP \cite{Ref16}, and any competitors would have to achieve a similar level
of performance.

Access to metadata, defined as cross-event information, is clearly a
weak point in the handling of the multiple analysis datasets originating
from the LHC. Metadata may include information such as book-keeping of
processed events, detector conditions and data quality, \emph{a
posteriori} measured scale factors, and software versioning. This
information is often scattered in multiple locations such as:
``in-file'' data, remote database resources, shared-area text or ROOT files,
TWiki pages or even in e-mail. The lack of proper metadata integration
and a coherent interface leaves analyses exposed to any problems in data
processing. This in turn results in data completeness demands that
simply will not work at the HL-LHC, where the huge datasets all but
guarantee that some fraction of the data will be unavailable at any
given point in time.

At the HL-LHC, as energy and luminosity conditions are quasi stable
across the years, datasets from multiple years with different conditions
will need to be analysed in a coherent way. The analysis software will
then need to be able to fetch and use the proper metadata automatically
while today's analysis software often requires dedicated configuration
and tuning for each data-taking period. Belle II have taken some steps
to address this by using the ``global tag'' concept commonly used in
reconstruction and applying it to analysis, allowing users to better
organise and store their analysis metadata. Nevertheless, until all
metadata is organised under the same global tag umbrella and accessed
through a coherent interface, which is extremely challenging, the
problem remains. Improvements here are necessary both to permit
efficient studies on partial datasets, and to reduce the risk of user
error in metadata access.

The growing complexity of analysis codes makes them extremely fragile
(i.e. bug-prone), hardly reusable, and unsuited for analysis
preservation needs. The current best efforts at analysis preservation
are based on a snapshot of the full analysis setup that can ensure the
possibility to re-run the code on the original dataset in a few years.
This may allow a future analyser to reproduce previous results, but will
likely not provide any understanding, for the future analyser, about
what the analysis was effectively doing. Another use case is to re-use
the analysis code on new data, be that the same dataset with improved
calibrations or additional data that will improve the precision of the
measurement. This brings significant additional challenges, not least
related to metadata.

Nevertheless, these preservation efforts encourage better organisation
of the analysis as a workflow, and promote the use of version control
tools together with continuous integration, which offers a natural route
to improving analysis code quality. In addition to these code quality
measures, a longer-term solution for code complexity may be the adoption
of a Domain-Specific Language (DSL) \cite{Ref10}, discussed in more detail
later, a model that would have physicists write logical descriptions of
their analyses, leaving low-level code generation and hardware
optimisation to a backend. Thus analysis design could be quickly
understood and shared, sidestepping the usual dearth of documentation,
while simultaneously isolating analysis design from implementation and
hardware, providing a natural means of analysis preservation. An
interesting development in this direction is the REANA platform \cite{Ref21}
that supports multiple backends and uses a DSL to provide a framework to
run an analysis. Already a valuable resource as a tool for analysing
open data, it is interesting to ask what would happen if every analyst
first had to structure their analysis and capture their workflows before
they started submitting jobs.

\hypertarget{plans-and-recent-progress}{%
\subsection{Plans and Recent Progress}\label{plans-and-recent-progress}}

Most of the problems identified above were already identified in the
community white paper \cite{Ref17} from 2017 and since then some progress
was made prototyping new technologies to enable data analysis at the
HL-LHC.

A first branch of new technologies is that of efficient data analysis
processing in the last steps of the analysis, i.e. when event data needs
to be selected, derived and aggregated into histograms. Several
platforms developed by industry or data science have emerged in recent
years to quickly aggregate large datasets with parallel execution
(either on clusters of computers or simply on multi-core servers). Many
of those tools have been tested to understand the feasibility of usage
in the HEP context providing fresh ideas and new paradigms that have
stimulated development within the HEP toolkit.

In particular, the Python ecosystem, implementing many such solutions
(e.g. numpy, pandas dataframes), is emerging as a possible alternative
to HEP's traditional C++ environment. This is thanks largely to a
combination of its versatility as a language, allowing rapid
prototyping, and the ability to out-source compute-intensive work to
more performant languages like C and C++. This is similar to the
experimental software frameworks where Python is used as steering code
and C/C++ is used as an efficient backend implementation. The typical
Belle II analysis starts with Python using a custom DSL to specify
particle selections and algorithms implemented in C++ and define output
ntuples. The ntuples are typically analysed by the PyHEP toolkit
\cite{Ref6}, and packages from the wider data science community. Deeper
integration with external python tools is also particularly important
for enabling state-of-the-art machine learning frameworks such as
PyTorch and TensorFlow, whose relevance in analysis is likely to grow in
the long term. Meanwhile PyROOT, which allows the use of any C++ class
from Python, has the potential to provide a coherent analysis ecosystem
with an effortless integration of HEP specific tools written in C++ with
industry python big data toolkits.

The ROOT package, i.e. the foundation and workhorse for LHC analysis,
has been the focus of substantial development. On the data storage
front, the ROOT team demonstrated better data compression and
accelerated reading for a wide variety of data formats. Information
density drives compression and this can vary massively between
experiments and analyses. This new format called RNTuple \cite{Ref9, ROOT-2020-HL-LHC}, an
evolution of the TTree file format, shows robust and significant
performance improvements (1-5x faster), which could potentially save
significant storage space (10-20\%) for the HL-LHC. Another area of
progress, and potential consolidation in ROOT I/O, is lossy compression,
where the number of significant bits for a quantity may be far fewer
than a standard storage type's mantissa, enabling further savings in
storage space~\cite{ROOT-2020-HL-LHC}.

Inspired by data science tools, the event processing framework was
extended with DataFrame-like functionality (RDataFrame) implementing a
declarative approach to analysis processing in either Python or C++
(always executed in C++), which natively exploits multi-core
architectures \cite{Ref7} and could be extended to accelerators. Used from
Python, it allows to pre-process data in C++, and export to numpy.
Flexibility in interfacing RDataFrame to non-ROOT data formats has
allowed the ALICE collaboration to prototype its Run 3 analysis model on
this new technology. Growing use of Machine Learning was also
anticipated, and the TMVA toolkit has been improved with substantial
optimisations, more complex deep learning models and enhanced
interoperability with common data science tools \cite{Ref14,Ref15}  which is of
particular importance for inference.

A complementary approach provided in Python is the toolchain of uproot,
awkward arrays and coffea \cite{Ref8} for efficient columnar operations and
transformations of the data. These function as a lightweight end-stage
analysis environment that provides only the elements necessary for
pre-processing ROOT inputs and converting them into formats used by the
standard ML frameworks, typically numpy. Lightweight distribution using
package managers such as pip or conda allows for rapid setup and
extension with other Python tools. Good performance appears to have been
achieved within the scope of these projects, which has been kept
focused, but for the purposes of fair comparison with other existing or
emerging options, defining clear benchmarks for I/O and processing speed
would be beneficial.

A further feature linked to the growing use of Python is the use of
``notebook'' technology such as Jupyter. This permits quasi-interactive
exploration where the annotated history including formatted outputs and
graphics can be saved and shared with collaborators for reproduction and
adaptation. Although not a substitute for command line scripts in
well-tested and complex workflows, notebooks make an effective vessel
for software education and have been leveraged as a powerful frontend
for access to facilities including CERN's SWAN \cite{Ref20}.

\hypertarget{prospects-and-rd-needs}{%
\subsection{Prospects and R\&D needs}\label{prospects-and-rd-needs}}

As the complexity of analysis data handling and processing grows,
developing efficient and robust code using a steadily increasing number
of tools running on ever more heterogeneous hardware becomes more and
more difficult. In addition, analysis code is typically re-implemented by
tens of individuals, in different experiments, analysis teams or
institutes mostly performing the same kind of operations, i.e. data
reduction, plotting, variation of systematic uncertainty, fitting, etc.
It is clear that tools and frameworks for performing these repeated
operations need to provide very efficient interfaces to avoid analysis
code bloat. Declarative rather than imperative programming is becoming
visibly more prevalent both in and outside of the field as it naturally
provides efficient syntax.

DSLs are a generalisation of this declarative concept and have already
demonstrated their ability to simplify analysis code. Not only do %% Missing reference?
they allow users to express operations concisely, improving
comprehension and reusability, DSLs provide a natural layer of
insulation against hardware evolution as low-level, optimised code
generation for multiple different platforms and architectures is
delegated to the tool and framework experts. Prototypes of DSLs have
been developed in recent years \cite{Ref10} for the event processing and
histogram production parts of the analysis workflow, and even built into
experimental frameworks to perform high-level analysis \cite{Ref19}. Similar
efforts can also be investigated in the context of data interpretation
and statistical analysis and an interesting example here is the so
called ``combine tools'' developed by CMS and based on RooFit for the
Higgs discovery). Descriptions of a full analysis in terms of %% Missing reference?
workflows could then tie together the analysis at the top level \cite{Ref18,Ref21}. With some effort, the HEP community may be able to converge on
effective tools that provide a common solution, ideally across
experiments.

There is an obvious need for good integration with analysis backends,
whether these be local CPU, batch or grid resources. At some stage in
the analysis process, interactivity or fast turn-around is needed. While
typical grid task completion has a time scale of one day to one week,
fast turn-around exploration requires answers within a few seconds up to
perhaps a few hours in order to keep people productive. As analysis
tasks are often IO-intensive and not necessarily-CPU intensive, the
right trade off between CPU and high bandwidth storage should be
carefully studied. More detailed studies of resource usage and its
evolution, especially as pertains to the need for heterogeneous compute,
is needed to ensure the right resource balance is available in the
HL-LHC computing infrastructure. This will require more interaction
between the analysis and infrastructure communities.

As a proof of principle, an unoptimised Higgs discovery analysis has
been rerun on Google Cloud in a few minutes as the infrastructure was
able to quickly provide tens of thousands of CPUs with guaranteed
bandwidth to storage of 2 Gb/coresec \cite{Ref12}. Scaling this to
sustained, diverse analysis payloads will be far from trivial, so the
question becomes how we can compare different approaches to provisioning
analysis. Eight initial benchmark challenges have been defined to set
the scope that should be addressed by analysis systems \cite{Ref11}. What
remains to be demonstrated is the capability of such systems to scale to
full analysis complexity, including tasks such as the handling of
systematic uncertainties and data-driven background estimation, and
these are yet to be incorporated into the benchmarks.

In contrast to the activity on analysis languages and infrastructure,
metadata is a topic that is almost entirely neglected by the analysis
community, even while computing experts have long understood its
importance. That is likely due to the absence of a single coherent
stake-holder for and provider of analysis metadata. Event data comes in
a ROOT file, but already within a single experiment there are numerous
sources of metadata that operate on vastly different timescales and this
poses a significant challenge to standardisation. As analyses and
detectors grow in complexity, further divergence may amplify the
challenge, making it critical to begin addressing this problem promptly.
Nevertheless, the drive to capture a complete analysis must drag
metadata with it, and the gleeful adoption of the declarative paradigm
will make database-like interfaces seem less scary than they once were
and may finally allow a more complete integration with analysis code. I
have a dream.

One final comment in reviewing recent developments is worth noting.
Exposure of the HEP community to industry-standard data science tools
has undoubtedly been a force for good and a net gain for the community,
providing fresh ideas and new ways of working - RDataFrame is an
excellent example. It is clear that the larger and better-supported data
science community will continue to innovate and produce tools that HEP
will benefit from, both by using directly and by borrowing their ideas
for our own tools. Taking machine learning as an example, it is
extremely likely that industry will blaze a trail that HEP would do well
to follow and we should continue to build bridges to those tools.
Equally though, HEP will continue to have its own unique challenges and
it's less likely that ultra-performant machine learning inference will
be top of the data scientists wishlist. Given the challenges presented
by extreme-throughput analysis at the HL-LHC, the right balance must be
found between software development within the community and relying on
external toolkits.

Finally, attention must be given to ensuring the longevity of any
projects that the HEP community decides to adopt en masse. Whether by
specific institutes, funding schemes or other collective programmes,
responsibility for key software will need to be ensured in the long
term. The development of a secure support base will therefore need to be
considered for emerging projects that show significant promise.

### 2020-04-30 LHCC/2020/2020-01/input_generators.tex
Break sources into an input file per section

\hypertarget{physics-event-generators}{%
\section{Physics Event Generators}\label{physics-event-generators}}

\hypertarget{introduction-1}{%
\subsection{Introduction}\label{introduction-1}}

Physics event generators are essential in HEP. All of the LHC scientific
results, both precision measurements or searches for new physics, depend
significantly on the comparison of experimental measurements to
theoretical predictions computed using generator software.

Using Monte Carlo (MC) techniques, generators allow both the generation
of unweighted events for experimental studies of differential
distributions and the prediction of total cross sections. The
large-scale event generation campaigns of the LHC experiments have
significant computational costs, mainly in terms of CPU resources. The
limited size of simulated samples is a source of major uncertainty in
many analyses and is therefore a limiting factor on the potential
physics output of the LHC programme. This situation will get
significantly worse in the high-luminosity running phase of the LHC
(HL-LHC). The fraction of the CPU resources available to the LHC
experiments which is used for event generation today is currently
estimated to be approximately 5-15\%. As is the case for the other big
consumers of CPU (detector simulation and reconstruction, that have
until now been the main focus), speedups in generator software are also
needed to address the overall resource problem expected at the HL-LHC,
also because more accurate predictions, requiring more complex
calculations (e.g. beyond NLO or with higher jet multiplicities) will be
needed from generators. Many other issues, both technical and
non-technical (e.g. funding, training, careers for those working in this
area), exist~\cite{Alves:2017she,Gen18}.

A HSF Working Group (WG) on generators~\cite{Gen19} was set up at the
beginning of 2019. The main focus of the WG so far has been on gaining a
better understanding of the current situation, and identifying and
prioritising the areas where computing costs can be reduced. In
particular, the WG has been active in analysing the ATLAS and CMS
compute budgets in detail, in profiling MG5\_aMC~\cite{Alw14}, 
Sherpa~\cite{Bot19} and Powheg~\cite{Fri07}, in discussing the possible sharing
of common parton-level samples by ATLAS and CMS, and in reviewing and
supporting the efforts for porting generators to modern architectures
(e.g., MG5\_aMC to GPUs). This last activity is particularly important,
as it has become increasingly clear that being able to run
compute-intensive WLCG software workloads on GPUs would allow the
exploitation of modern GPU-based supercomputers at High Performance
Computing (HPC) centres, and generators look like a natural candidate
for this.

This section gives an overview of the many technical and non-technical
challenges in the generator area and of the work that can be done to
address them. This is finally condensed into a list of a few
high-priority items, for the next 18 months before the LHCC review that
is currently scheduled for fall 2021. A more detailed version of this
contribution, including a more complete list of references, is
uploaded on arXiv and will be submitted for publication~\cite{Gen20}.

\hypertarget{collaboration-challenges}{%
\subsection{Collaborative Challenges}\label{collaboration-challenges}}

\subsubsection{Generator Software Landscape}

The landscape of generator software is extremely varied. Different
generators are used for different processes. Generating a sample also
involves choices of precision (e.g. Leading Order, LO, or
Next-to-Leading-Order, NLO), hadronisation and Parton Shower (PS)
models, underlying event tunes, prescriptions for matching/merging and
simulating particle decays, and other input parameters, chiefly among
them the parton density functions, for which different interfaces exist.
Various combinations of software libraries are thus possible, often
written by different authors and some dating back many years. For a
given process, the LHC experiments often use different software packages
and settings from each other, and a single experiment can generate
events using more than one choice. Many different packages and
configurations may therefore need to be worked on to get cumulative CPU
cost reductions. The large number of external packages also complicates
their long-term maintenance and integration in the experiments software
and workflows, sometimes leading to job failures and computing
inefficiencies. Other packages are also absolutely critical for the
whole generator community and must be maintained, even if their CPU cost
is relatively low (LHAPDF, Rivet, HepMC, etc.).

\subsubsection{Skills and Profiles}

A very diverse combination of skills are needed in generator software,
thus people with many distinct profiles need to contribute. Theorists
(who create fundamental physics models, and design, develop and optimise
most generator code), experimentalists working in research (who request
types/sizes of samples) and in computing (who implement, monitor and
account execution of workflows on computing resources), software
engineers and system performance experts. This is a richness and
opportunity, as some technical problems are best addressed by people
with specific skills, but it also poses some challenges:

\paragraph{Training challenges.} Theorists and experimentalists often
lack formal training in software development and optimisation. Software
engineers and experimentalists are often not experts in the theoretical
physics models implemented in MC codes.

\paragraph{Communication challenges.} It is difficult to find a shared
terminology and set of concepts to understand one another: notions and
practices that are taken for granted in one domain may be obscure for
others. As an example, there are many articles about the physics in
generators, but software engineers need papers describing the software
modules and overall data/control flow.

\paragraph{Career challenges.} Those working in the development, optimisation or
execution of generator software provide essential contributions to the
success of the (HL-)LHC physics programme and it is critical that they
get the right recognition and motivation. However, theorists get
recognition on published papers, and may not be motivated to work on
software optimisations that are not "theoretical" enough to advance
their careers. Generator support tasks in the experiments may also not
be enough to secure jobs/funding to experimentalists pursuing a career
in research.

\paragraph{Mismatch in usage patterns and optimisation focus.} The way
generators are built and used by their authors is often different from
the way in which they are deployed and integrated by the experiments in
their software frameworks and computing infrastructure.The goals and
metrics of software optimisation work may also differ. Theorists are
mainly interested in calculating cross sections and focus on minimising
the phase space integration time for a given statistical precision. The
LHC experiments run large scale productions of unweighted event
generation, and mainly need to maximise the throughput of events
generated per unit time on a given node.

\paragraph{Programming languages.} Attracting collaborators with a computer
science background to work on generators, especially students, may also
be complicated by the fact that critical components of some generator
packages are written in Fortran, which is rarely used in industry and
less popular among developers than other programming languages. Some of
the generators also do not use industry standard version control
systems, making it harder to contribute code.

\hypertarget{technical-challenges-software-and-computing}{%
\subsection{Technical Challenges}\label{technical-challenges-software-and-computing}}

The event generation workflow presents several challenges and
opportunities for improvement.

\subsubsection{Inefficiency in Unweighted Event Generation}

\paragraph{Phase space sampling inefficiency.} Efficient sampling is the most
critical ingredient for efficient unweighted event generation. Many
generic algorithms exist (e.g. VEGAS~\cite{Lep80}, BASES / SPRING~\cite{Kaw86}, 
MINT~\cite{Nas07}, FOAM~\cite{Jad03}), as well as others
developed specifically for a given generator (e.g. MadEvent~\cite{Mal03},
itself based on a modified version of VEGAS, in MG5\_aMC, or COMIX~\cite{Gle08} 
in Sherpa). In general, the larger the dimensionality of the
phase space, the lower the unweighting efficiency that can be achieved:
in W+jets, for instance, the Sherpa efficiency is 30\% for W+0 jets and
0.08\% for W+3jets~\cite{Gao20}. This is an area where research is very
active, and should be actively encouraged, as significant cost
reductions in WLCG compute budgets could be achieved. Improvements in
this area can only start from physics-motivated approaches based on the
knowledge of phase space peaks, but they can be complemented by
brute-force machine learning (ML) algorithmic methods~\cite{Ben17,Bot20,Gao20,Kli18}.

\paragraph{Merging inefficiency.} Merging prescriptions (e.g. MLM~\cite{Man02},
CKKW-L~\cite{Lon02} at LO and FxFx~\cite{Fre12}, MEPS@NLO~\cite{Hoe14} at
NLO) imply the rejection of some events, to avoid double counting
between events produced with n\textsubscript{jets}+1 matrix elements and
with n\textsubscript{jets} MEs plus parton showers. The resulting
inefficiencies can be relatively high depending on the process, but they
are unavoidable in the algorithmic strategy used by the underlying
physics modeling~\cite{Alw08}. However, a method like shower-kt MLM can
reduce the merging inefficiency of MLM~\cite{Alw09}.

\paragraph{Filtering inefficiency.} An additional large source of inefficiency
is due to the way the experiments simulate some processes, where they
generate large inclusive event samples, which are then ﬁltered on
ﬁnal-state criteria to decide which events are passed on to detector
simulation and reconstruction (e.g. CMS simulations of specific
$\Lambda_{B}$ decays have a 0.01\% efficiency and ATLAS B-hadron
filtering has \textasciitilde10\% efficiency for V+jets). This
inefficiency could be reduced by developing ﬁltering tools within the
generators themselves, designed for compatibility with the requirements
of the experiments. Filtering is an area where LHCb has a lot of
experience and already obtained significant speedups through various
techniques. The speed of colour reconnection algorithms is a limiting
factor for simulating rare hadron decays in LHCb.

\paragraph{Sample sharing and reweighting.} In addition to removing
inefficiencies, other ways could be explored to make maximal use of the
CPU spent for generation by reusing samples for more than one purpose.
Sharing parton-level samples between ATLAS and CMS is being discussed
for some physics analyses. Event reweighting is already commonly used
(e.g. for new physics searches and some systematic uncertainty
calculations) and could be explored further, though this may require
more theoretical work for samples involving merging or NLO matching~\cite{Mat16}.

\paragraph{Negative weights.} Matching prescriptions, like MC@NLO, are required
in (N)NLO calculations to avoid double counting between (N)NLO matrix
elements and parton showers, leading to the appearance of events with
negative weights. This causes a large inefficiency, as larger event
samples must be generated and passed through the experiment simulation,
reconstruction and analysis codes, increasing the compute and storage
requirements. For a fraction r of negative weight events, the number of
events to generate increases by a factor $1/(1-2r)^2$: for instance,
with r=25\% (which may be regarded as a worst-case scenario in top quark
pair production~\cite{Fre20}), one needs to generate 4 times as many
events. Negative weights can instead be almost completely avoided, by
design, in another popular matching prescription, POWHEG, which however
is only available for a limited number of processes and describes the
relevant physics to a different degree of precision than MC@NLO 
(see~\cite{Fre20} for a more in-depth discussion). Progress in this area can
only be achieved by theorists: for instance, a new approach for
MC@NLO-type matching with reduced negative weights has recently been
proposed~\cite{Fre20}. It should be noted that negative weights can also
happen at LO because of not-positive-definite parton density function
sets and interference terms, e.g. in effective field theory
calculations.

\subsubsection{Accounting and Profiling}

While progress has been made to better understand which areas of
generator software have the highest computational cost, more detailed
accounting of the experiment workloads and profiling of the main
generator software packages would help to further refine R\&D
priorities:

\paragraph{Accounting of CPU budgets for generators in ATLAS/CMS.} Thanks to a
lot of effort from the generator teams in both experiments, a lot of
insight into the settings used to support each experiment's physics
programme was gained within the WG, and it is now clear that the
fraction of CPU that ATLAS spends for event generation is somewhat
higher than that in CMS. More detailed analysis of the different
strategies is ongoing. A practical issue is that these figures had to be
harvested from logs and production system databases, which was
particularly difficult for CMS, requiring significant person hours to
extract the required information. It is important to establish better
mechanisms to collect this information, to allow for an easy comparison
between different experiments.

\paragraph{Profiling of the generator software setups used for production.}
Another area where the WG has been active is the definition and
profiling of standard generator setups, reproducing those used in
production. Detailed profiling could also be useful to assess the CPU
cost of external parton distribution function libraries~\cite{Kon20}, or
the memory requirements of the software (which may motivate a move to
multithreading or multiprocessing approaches).

\subsubsection{Software Modernisation}

More generally, as is the case in other areas of HEP, some R\&D on
generator software would certainly be needed to modernise it and make it
more efficient, or even port it to more modern computing 
architectures~\cite{Bau13,Alves:2017she}:

\paragraph{Data parallelism, GPUs and vectorisation.} The data flow of an MC
generator, where the same (matrix element) function is computed
repeatedly at many phase space points, lends itself naturally to the
data parallel approaches found in CPU vectorised code and in GPU compute
kernels. Porting and optimising generators on GPUs is essential to be
able to use modern GPU-based HPCs (such as SUMMIT, where 95\% of the
compute capacity comes from GPUs). The work done in this direction in
the past on MG5\_aMC, that never reached production quality, is now
being revamped by the WG, in collaboration with the MG5\_aMC team, and
represents one of the main R\&D priorities of the WG. This work is
presently focusing on Nvidia CUDA, but abstraction libraries will also
be investigated. GPUs may also be relevant to ML-based sampling
algorithms and to the pseudo-random number generation libraries used in
all MC generators.

\paragraph{Task parallelism, multithreading and multiprocessing.} Generators are
generally executed as single threaded software units. In most cases,
this is not a problem, as the memory footprint of unweighted event
generation is small and usually fits within the 2 GB per core available
on WLCG nodes. However, there are cases where memory is higher than 2 GB
(e.g. DY+jets using Sherpa); this leads to inefficiencies as some
processor cores remain unused, which could be avoided using
multithreading approaches. The fact that some generators are not even
thread safe, may also be a problem, for instance to embed them in
multi-threaded event processing frameworks, such as that of CMS.
Multi-processing approaches may also be useful to speed up the
integration and optimisation step for complex high-dimensional final
states. In particular, a lot of work has been done to implement
MPI-based multi-processing workflows for generators in recent years. For
instance, the scaling of LO-based generation of merged many-jet samples
has been successfully tested and benchmarked on HPC architectures using
both Alpgen~\cite{Chi17} and Sherpa~\cite{Hoe19}; in the latter case, new
event formats based on HDF5, replacing LHEF, have also been instrumental
in the success of these tests. MPI integration has also been completed
for MG5\_aMC~\cite{Mat18}. It should also be noted that, even if HPCs
offer extremely high-speed inter-node connectivity, it is perfectly ok
for WLCG workflows to use these systems as clusters of unrelated nodes.

\paragraph{Generic code optimisations.} A speedup of generators may also be
achievable by more generic optimisations, not involving concurrency. It
should be studied, for instance, if data caching~\cite{Kon20} or
different compilers and build strategies may lead to any improvements.

\hypertarget{physics-challenges}{%
\subsection{Physics Challenges}\label{physics-challenges}}

In addition to software issues, important physics questions should also
be addressed about more accurate theoretical predictions (above all NNLO
QCD calculations, but also electroweak corrections) and their potential
impact on the computational cost of event generators at HL-LHC. Some
specific NNLO calculations are already available and used today by the
LHC experiments in their data analysis. With a view to HL-LHC, however,
some open questions remain to be answered, in particular:

\paragraph{NNLO: status of theoretical physics research.} The first question is
for which processes NNLO precision will be available at the time of the
HL-LHC. For example, when would NNLO be expected for 2-\textgreater3 or
even higher multiplicity final states? And for lower multiplicity final
states, where differential NNLO predictions exist but the generation of
unweighted NNLO+PS events is not yet possible. It is important to
clarify the theoretical and more practical challenges in these
calculations, and the corresponding computational strategies and impact
on CPU time needs.

\paragraph{NNLO: experimental requirements for data analysis at HL-LHC.} The
second question is for which final states unweighted event generation
with NNLO precision would actually be required, and how many events
would be needed. One should also ask if reweighting LO event samples to
NNLO would not be an acceptable cheaper alternative to address the
experimental needs.

\paragraph{Size of unweighted event samples required for experimental analysis
at HL-LHC.} Another question, unrelated to NNLO, is in which regions of
phase space the number of unweighted events must be strictly
proportional to the luminosity. For example, in the low pT regions of W
boson production it is probably impossible to keep up with the data, due
to the huge cross section. Alternative techniques should be
investigated, to avoid the generation of huge event samples.

### 2020-04-30 LHCC/2020/2020-01/input_introduction.tex
Break sources into an input file per section

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Common and community software packages, such as ROOT, Geant4 and event
generators have been a key part of the LHC's success so far and
continued development and optimisation will be critical in the future
\cite{stewart_graeme_a_2018_2413005, Ellis:2691414, Albrecht2019}. The
challenges are driven by an ambitious physics programme, notably the LHC
accelerator upgrade to high-luminosity, HL-LHC, and the corresponding
detector upgrades of ATLAS and CMS. The General Purpose Detectors
describe their specific challenges elsewhere; in this document we
address the issues for software that is used in multiple experiments
(usually even more widely than ATLAS and CMS) and maintained by teams of
developers who are either not linked to a particular experiment or who
contribute to common software within the context of their experiment
activity. We also give space to general considerations for future
software and projects that tackle upcoming challenges, no matter who
writes it, which is an area where community convergence on best practice
is extremely useful.

ATLAS and CMS will undergo major detector upgrades and will also
increase their trigger rates for HL-LHC by about a factor of 10; event
complexity will rise, with peak pile-up of 200, far higher than in
Run-2. This places an enormous burden on storage and processing
resources. Current CMOS microprocessor technology is clock speed limited
(due to the failure of Dennard scaling) and, while processors
per-integrated circuit still keeps rising, Moore's Law is expected to
stall during the 2020s. More importantly, the effective runtime related
improvements in computing from CPU servers at sites is likely to be only
about 10\% per year, making the shortfall in processing resources more
severe. As manufacturers struggle to deliver ever-more effective
computing through CPUs the drive to different architectures intensifies,
with GPUs becoming commonplace and increasing interest in even more
specialised architectures, such as TPUs, IPUs and developments that make
FPGA devices more user friendly \cite{acm10.1145/3282307}.
These pose huge challenges for our
community as the programming models and APIs vary widely here and
possible lock-in to a particular vendor's devices is a significant
problem for code sustainability and preservation. Huge work has been
done already to adapt to this changing landscape, e.g. multi-threading
software and GPU codes have been used in production by some experiments
for years~\cite{calafiura2018hep, albrecht2018hepexec}.
In other areas migration is still ongoing. In others it is
not even yet started. Generic heterogeneous programming models have
existed for some time, and new ones are arising, but there is, as yet,
no clear winner, in-part because the performance obtained can be highly
application-dependent. HEP itself will not drive the success of one
model or another, so even if the community converged on a preferred
model, its long term success would not be assured. The C++ standard
itself is likely to lag for many years behind what is required by us in
the area of heterogeneous or even distributed computing. Further,
experiment frameworks (and with knock-on effects to systems like
workload management) will need to adapt to sites that provide
heterogeneous resources. How to do this, and make effective use of
different heterogeneous resources across different sites, remains far
from settled.

For storage systems (see the DOMA and WLCG documents) the pressure on
software is to store as much physics information in as few bytes as
possible, but also to be able to read at very high rates to deliver data
from modern storage technologies into the processing hardware. This
requires critical developments in the storage formats and libraries used
in HEP, e.g. developments like RNTuple for ROOT I/O is likely to be of
great importance for the community \cite{ROOT-2020-HL-LHC}. The likelihood of finding an
off-the-shelf compact and efficient storage format for HEP data is
remote, so investment in smart software to support our PB sized science
data is simply cost effective. Particularly for analysis, which is
usually I/O bound, we have an end-to-end problem from storage
technology, through the software layers, to processing resources that
may well span multiple nodes. Other workflows, which are less dependent
on I/O rates will, nevertheless, have to be adapted to using remote data
where the I/O layer must optimise data transfers and hide latency, e.g.
taking advantage of XRootD's single column streaming ability \cite{xrootd}.

In this increasingly complex environment in which to write software,
there are important problems where sharing information at the community
level is far more efficient. Providing a high level of support in the
build environment for developers, sharing knowledge about how to
measure, and then improve, performance (especially on multiple different
architectures) and sharing best practice for code development can have a
large integrated benefit \cite{couturier2017hep}. This requires improved training and the
development of a curriculum for all developer levels. In the field, such
initiatives are often warmly received, but real support is needed for
those who can put this into practice, also ensuring that their work in
training contributes to their career development and a long term future
in the field \cite{foundation2018hep}. HEP software stacks are already deep and wide and building
these consistently and coherently is also an area where knowledge can be
shared. Support is needed for multiple architectural targets and
ensuring the correct runtime on heterogeneous environments.

This brings up the important question of validation and the need to
improve the security of physics results, which is even more complicated
on heterogeneous platforms, when exact binary compatibility often cannot
be assured. Currently almost every experiment and project has its own
infrastructure for this.

Once software is built, it needs to be shipped worldwide so that the
production workflows can run. CernVM-FS was a huge leap forward for
software distribution and has even been widely adopted outside HEP 
\cite{Blomer_2011, 7310920}.
However, new challenges arise, with container based payloads, scaling
issues and disconnected supercomputer sites. So maintenance and
development needs to be undertaken to support and extend this key
supporting infrastructure for software.

Finally, over the multi-decade lifetimes of HEP experiments, we need to
preserve both the core and analysis software so that results can be
confirmed and updated as the field moves on. There are many exciting
developments based around CernVM-FS
\cite{10.1007/978-3-319-67630-2_52, serverless-cvmfs}, containers and things like analysis
description languages \cite{Ref10}, but these are not yet at the stage of being
settled nor integrated into our day-to-day workflows.

In the rest of this document the main issues associated with the key
parts of the software workflow in high-energy physics are presented,
focusing on those that dominate current resource consumption: physics
event generation, detector simulation, reconstruction and analysis.

### 2020-04-30 LHCC/2020/2020-01/input_simulation.tex
Break sources into an input file per section

\hypertarget{detector-simulation}{%
\section{Detector Simulation}\label{detector-simulation}}

\hypertarget{introduction-2}{%
\subsection{Introduction}\label{introduction-2}}

In HEP, data analysis, theory model evaluation and detector design
choices rely on detailed detector simulation. Since the start of the
first run the LHC experiments have produced, reconstructed, stored,
transferred and analysed hundreds of billions of simulated events. This
effort required a major fraction of the WLCG computing 
resources~\cite{ALICE-TDR-12, ATLAS-TDR-17, CMS-TDR-7, LHCb-TDR-11, WLCG-CM-Update, 
ALICE-TDR-019, LHCb-TDR-018, ATLAS-LHCC-2019-02, CMS-LHCC-2019-09}

The increase in delivered luminosity in future LHC runs, in particular
with the HL-LHC, will create unique research opportunities by collecting
an order of magnitude more data. At the same time, the demand for
detector simulation will grow accordingly, in order to keep the
statistical uncertainties as low as possible. However, the expected
computing resources will not suffice if current detector simulation is
to be used; the availability of simulated samples of sufficient size
will soon become a major limiting factor as far as new physics
discoveries, for example through precision measurements, are concerned.
Development of faster simulation, therefore, is of crucial importance
and different techniques need to be explored under the assumption that
they will provide a sufficient gain in time performance with a
negligible or acceptable loss in physics accuracy.

The Geant4 simulation toolkit has been the de facto standard for HEP
experiment simulation for physics studies over the last two decades~\cite{Geant4-2003, Geant4-2006, Geant4-2016}.
Designed in the late 1990s to cope with the simulation challenges of the
LHC era, Geant4 introduced flexible physics modelling, exploiting layers
of virtuality. This fostered progress in physics by comparison of
competing models on same energy ranges, and allowed complementary models
to be combined to cover large energy ranges. The simulation applications
of the LHC experiments make use of this toolkit and, as such, most of
the LHC physics program relies on it. It has delivered a physics
precision that allows the experiments to perform precision analysis of
the collected data and to produce new results pushing the boundaries of
our understanding in HEP. At the same time, the code of the toolkit is
becoming old, with some parts written almost thirty years ago, making it
more and more difficult to run efficiently on modern computing
architectures. Any improvement in the critical elements of Geant4 can be
leveraged by the applications that use it. 
Improvements of physics models need to continue to take place to prevent
systematic uncertainties from the simulation from becoming a dominant
factor, however, their exact influence on the precision of physics
analysis is not always easy to quantify. On the other hand, the overall
CPU performance of the current code will not improve drastically just by
relying on modern hardware or better compilers.

Taking all those elements into account, three main development paths have emerged to be
undertaken simultaneously: these have started to be pursued, however,
effort in any one of them is far from sufficient to exploit their
individual potential.

On one hand, it is needed to continue to modernise
and refactor the simulation code by using more optimal, modern
programming techniques. Those improvements, requiring considerable
effort but no major paradigm changes, can reasonably bring several tens
of percent of speed-up, by improving the throughput on modern CPUs,
avoiding constant churn in data and instruction caches. This will
certainly help to produce more simulated events, but it will not be
sufficient to meet the high luminosity needs of the LHC experiments.

On the other hand, different fast simulation techniques, where tracking of
individual particles in given detectors is replaced by some sort of
parameterisation, have already been extensively used for LHC simulation
campaigns, where some precision is traded to achieve higher event
throughput (thus lowering the cost of samples overall). Techniques to
re-use part of the simulated events multiple times have also emerged.
New technologies (like Machine Learning), should certainly be
investigated as fast simulation approaches. All of these will become
more and more significant and will need to be seamlessly integrated with
the detailed simulation provided by Geant4.

Finally, the use of compute
accelerators (like GPUs or FPGAs) is the third emerging path that needs
to be undertaken and that requires intense R\&D effort. We will discuss
more in detail those three development axes in the following sections.

In this section we will outline the developments identified in and since the
Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} focusing
in particular in these three area.

\hypertarget{geant4-rd}{%
\subsection{Geant4 R\&D}\label{geant4-rd}}

The Geant4 particle transport toolkit contains advanced models for
electromagnetic and hadronic physics processes, as well as the
geometrical description of the detectors. The toolkit design principles
were to allow flexibility in future extensions of its own code, as well
as choices for particular applications. It is clear that this
flexibility comes at the price of a performance penalty. It was observed that
writing more compact and more specialised code can lead in certain
places to up to several tens percent speed-up~\cite{GeantV}. The
technical improvements to explore here consist of avoiding too many
virtual layers and improving the instruction and data locality, leading
to better cache use. Going even further as far as specialisation is
concerned, studies suggest that implementing compact, self-contained
transport libraries with a minimal set of physics models (for instance
only electromagnetic interaction for selected particles) with predefined
scoring (to simulate, for instance, the response of an EM calorimeter)
can also bring speed-ups to the complete application~\cite{GeantV}. As
far as architectural modifications are concerned, the design of the
Geant4 track-status machinery prevents the introduction of fine-grained
track parallelism. It is still not clear whether this kind of
parallelism can bring any substantial speed-up, due to the overheads;
however, making the Geant4 kernel and physics processes stateless would
certainly provide the necessary starting points for further R\&D in that
direction. Most of these investigations have started and it is hoped to
see some first results on a year time-scale. More effort, however, will
be required to integrate any eventual changes in the production code in
a timely manner. The GeantV Vector prototype has allowed the assessment
of the achievable speedup using a novel, vectorised approach to particle
transport. The demonstrator of a full electromagnetic shower in
realistic (CMS) geometry has been implemented and the comparison to a
similar Geant4 application has been performed. The conclusion of that
work showed that the main factors in the speedup seem to include better
cache use and tighter code, while the vectorisation impact was much
smaller than hoped for. This unimpressive impact is due to a set of
factors, including the fraction of the current algorithms that could be
vectorised, unresolved challenges for the gather/scatter and tail
handling (to keep the number of events in flight within bound) needed
for large number of shapes, and of course Amdahl's Law applied to
vectorisation where some bottleneck (for example geometry navigation)
could not be tackled without a major additional long term effort~\cite{GeantV}.
At the same time libraries developed in the context of the
GeantV R\&D, e.g. VecGeom, have been successfully integrated in Geant4
and ROOT benefiting the whole HEP software community~\cite{CMS01}. The
cost-over-benefit of more vectorisation of the code has not been deemed
worth the necessary investment also due to the speculation that a
complete rewrite of the physics base could become necessary to fully
exploit it. As a result investing in the optimisation and modernisation
of the Geant4 code has gained even more relevance.

\hypertarget{experiments-applications-and-optimisation-of-the-use-of-geant4}{%
\subsection{Experiment Applications and Optimised Use of
Geant4}\label{experiments-applications-and-optimisation-of-the-use-of-geant4}}

All LHC experiments have dedicated simulation frameworks making use of
the Geant4 toolkit for modelling their detectors and the physics
processes occurring in the experimental setups. Every experiment
configures and uses what is provided by Geant4 for their specific needs.
Effort has been spent by each experiment since the publication of the
Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} to benchmark their 
simulation code and to maximally
exploit the options provided by the Geant4 toolkit to optimise the
performance of their applications~\cite{ATLAS-G4OPT}. All of the various handles provided
are being continuously explored and adopted when deemed useful: range
and physics cuts have been reviewed and customised by all experiments,
shower libraries adopted as baseline in the forward region by ATLAS and
CMS, stepping granularity optimised, magnetic field caching
investigated, neutron Russian roulette used for dedicated simulations.
All of them can have major impacts on simulation time and sharing
knowledge and experience between experiments can be of high benefit even
if each experiment has to find its optimal solution.

ATLAS, CMS and LHCb have integrated the Geant4 event-based
multi-threading features within their own multi-threaded frameworks,
harmonising their different architectural choices with the Geant4
implementation~\cite{CMS02, LHCb01, ATLAS01}, and have or are planning to
use them in their production environment to gain access to a larger set
of distributed resources. ATLAS has been running their multi-process
simulation framework on HPCs systems for production in the
last years~\cite{ATLAS02, Benjamin:2696330} 
while CMS has pioneered the use of their multi-threaded
simulation on available resources including HPCs~\cite{CMS03}. Both
approaches have allowed to exploit a higher number of computing
resources by gaining access to systems with limited memory.

\hypertarget{fast-simulations}{%
\subsection{Fast Simulations}\label{fast-simulations}}

The so-called `fast simulation' techniques consist of replacing the
tracking of individual particles through (a part of) the detector,
including all the physics processes they would undergo, by a
parameterisation where the detector response is produced directly as a
function of the incoming particle type and energy. An example of such a
parameterisation was implemented many years ago in the context of the H1
experiment~\cite{H1Gflash}. This parameterisation, available within the
Geant4 toolkit under the name of the GFLASH library, became the starting
idea for several `custom' implementations, specific for other
calorimeters. Those were interfaced to experiments' simulation
applications through the available `hook' in the Geant4 toolkit. The LHC
experiments implemented, for example, dedicated parametrised response
libraries for some of their calorimeters. Recently, an effort has been
invested in Geant4 to generalise the parameterisation formulae, and to
implement automatic procedures of tuning the parameters for specific
detector geometries. This kind of `classical' fast simulation, which
describes the particle shower shape using some complex mathematical
functions with several parameters, will remain an important tool;
however, it is also clear that their overall precision for the different
energy values, especially for highly granular and complex calorimeters
will always be relatively low. The recent developments of Deep
Learning-based techniques have opened an exciting possibility of
replacing those `classical' parameterisations by trained neural networks
that would reproduce the detector response. This approach consists of
training generative models such as Generative Adversarial Networks
(GAN), Variational Auto-Encoders (VAE) or Autoregressive Generative
Networks on the `images' of particle 
showers~\cite{ML001-GAN, aishik_ghosh_2019_3599705, GAN-ATLAS}. The energy
deposition values in the calorimeter cells are considered as `pixels' of
the images that the network is supposed to reproduce. The first studies
and prototypes have shown very promising results, however, the
generalisation of the developed models (for different calorimeters, with
different particles incident at different angles with different
energies), still requires further effort. Detailed physics validation of
those tools and understanding of the capability to reproduce the
fluctuations is a prerequisite towards `production-quality' fast
simulation libraries. This work is already ongoing, but will require,
over the next few years, more effort to be invested combining the
physics and Machine Learning expertise. Given that the training of the
deep learning algorithms usually requires using sufficiently large MC
samples produced using standard techniques, the development of the novel
tools does not necessarily remove the need to speed up Geant4.

It is worth pointing out that different fast simulation techniques have
already been extensively used for LHC simulation campaigns. Effort has
been spent in the experiments frameworks to combine as much as possible
fast and detailed implementation. In particular ATLAS and CMS are using
in their baseline production shower libraries for the modelling of
calorimeters in the forward region in combination with the detailed
simulation of the rest of the sub-detectors. In LHCb the re-use of part
of the underlying events for specific signals of interest has been
established in production of samples when appropriate, with particular
care paid to ensure keeping under control bias on statistical
uncertainties~\cite{LHCb02}, its applicability in ATLAS and CMS could be
explored for specific cases. Similarly, the re-use of simulated or real
events to mimic the background to hard-scatter events is also exploited,
where additional challenges exist concerning storage and I/O due to the
handling of the large minimum bias samples needed to model the
additional interactions.

All LHC experiments have explored the use of multi-objective regression
and generative adversarial networks (GANs)~\cite{GAN-ATLAS, GAN-LHCb}, they are now in the process of investigating the use of
fast simulations for other types of detectors than calorimeters, e.g.
Cerenkov based systems~\cite{Lamarr} and Time Projection Chambers~\cite{ALICE-TPC}. While implementations of fast simulations of given
sub-detectors is specific to their technology and experimental
environment and as such the responsibility of each LHC experiment, an
area of potential common investigation is frameworks for fast tuning and
validation.

The use of fully parametric response of experimental setups at the level
of reconstructed objects (tracks and clusters) for rapid evaluation of
physic's reach is also exploited by the experiments with generic
frameworks for fast simulation of typical collider detectors being used~\cite{bib-Delphes, PGS4}. 
The advantage of such frameworks is to allow
simultaneous studies for different experiments as well as their
potential use by the theory community. Experiments specific fully
parameterised simulations providing reconstructed objects compatible
with the experiment's analysis frameworks for systematic verifications
have also been emerging~\cite{Lamarr} .

It is reasonable to expect that the full variety of simulation options,
from `smeared reconstructed quantities' to parameterisation for specific
detectors to the detailed Geant4, will need to be exploited by the
experiments for different tasks. Seamless ways of providing them in a
transparent integrated way in the experiments' simulation frameworks
should continue to be pursued with Geant4 supporting hooks for such
integrations.

\hypertarget{technical-challenges-software-and-computing-1}{%
\subsection{Technical Challenges}\label{technical-challenges-software-and-computing-1}}

The hardware landscape has been always setting direction as far as
software evolution was concerned. The recent adaptation of the
simulation code to multithreading (MT) turned out to be relatively
straightforward, but still took several years to adopt and validate.
Geant4 can now run in the MT mode efficiently, distributing the events
between the threads and sharing resources like the geometry description
or physics data and thus reduce the memory footprint of a simulation
while maintaining its throughput performance. The efficient utilisation
of Single Instruction Multiple Data (SIMD) architectures, on the other
hand turned out to be more challenging. Not only was a limited part of
the code `vectorisable' but also, as the already mentioned GeantV R\&D
has shown, overheads related to gathering the data for vector processing
presented a significant challenges (tail handling, work balancing across
threads, etc.) to efficiently (development time wise and run time wise)
profit from it~\cite{GeantV}.

The most recent `revolution' in the computing hardware systems consists
of using Graphics Processing Units (GPUs) for general computing. Several
attempts have been made to port specific simulation code to GPUs. A few
of them turned out to be successful leading to factors of several
hundred or thousand of speedup~\cite{GPU-MPEXS-DNA, Hybrid-Gate, Opticks}, however, they were always
limited to a very specific simulation problem, far from what is
addressed by a general HEP simulation. The application of GPUs has been
very successful in restricted domains like medical physics simulation,
neutron transport~\cite{SHIFT01} or optical photon transport~\cite{CHEP-Juno}. 
In those applications, the type of particles considered
is very limited (often just one type, with no secondary particle
production), the set of physics processes is reduced and the geometry is
often extremely simple compared to the LHC detectors. A natural
`extrapolation' of those approaches to a general HEP simulation is very
difficult to imagine, because the stochasticity of the simulated events
would immediately lead to divergences as far as different GPU threads
are concerned, not to mention, simply the feasibility of porting the
whole simulation to the GPU code for which specific expertise would be
needed. On the other hand, running only very small pieces of the
simulation application on GPUs does not seem to be efficient either, as
gathering data and transferring it from the host to the device and back
may strongly limit any possible performance gain, similarly to what was
seen with the GeantV vectorisation R\&D. The most plausible approaches
seem therefore to lead in the direction of specialised libraries (like
those described above in the context of speeding up Geant4 simulation on
the CPUs) that would perform the complete simulation of specific
sub-detectors (for instance EM calorimeter, for specific incoming
particles or Cherenkov-based detectors for the optical processes) on the
device. Such libraries could be the right compromise between the
complexity of the algorithm that needs to be ported to the device and
the overall time that is now spent (and could be gained) on the
corresponding of the full simulation on the CPU. The investigation in
those directions are starting now, but will certainly require a
considerable effort to be further invested before being able to judge
their real impact. The first steps that are currently being undertaken
consist of implementing, on GPUs, prototypes based on VecGeom geometry
and navigation capabilities. If those prototypes turn out to be
successful, the next challenge will consist of adding tracking and
eventually a limited subset of physics models. While Geant4 is a toolkit
capable of addressing different modelling and simulation problems and
contains many features and capabilities allowing for user access to
detailed information for a large variety of use cases and scientific
communities, this GPU development might turn into specialised transport
modules, stripped of some features which would be expensive to implement
or support efficiently on GPUs. Another interesting avenue consists of
exploring the application of some vendor libraries (like Nvidia Optix).
Those libraries, originally developed for ray tracing, have several
similarities with general particle transport and if applied
successfully, could lead to major speed-ups. All those efforts certainly
require a major investment and new developers, expert in those
technologies, to join the R\&D work.

\hypertarget{other-activities}{%
\subsection{Other Activities}\label{other-activities}}

Common digitisation efforts would be desirable among experiments, with
advanced high-performance generic examples, which experiments could use
as a basis to develop their own code. Nevertheless the large variety of
detectors' technologies used by the experiments reduces its
applicability. Digitisation not yet being a limiting factor in terms of
CPU requirements also does not urge developments in this area.

Simulated samples are often processed through the reconstruction as real
data. As such in some cases they require the emulation of hardware
triggers. Hardware triggers are based on very specific custom devices
and a general approach does not seem feasible even if some
parametrisation could be generalised.

\hypertarget{outlook}{%
\subsection{Outlook}\label{outlook}}

To conclude, it is important to realise that there is a considerable
risk for the simulation to become a major limiting factor as far as new
physics discoveries are concerned if no serious effort is invested in
R\&D. We need the concerted effort of different experts in physics,
Machine Learning and GPUs. There are too many unknowns to focus on only
one direction, so a lot of different prototyping is needed. We need to
further develop Geant4 while working on fast simulation algorithms and
doing extensive R\&D on leveraging compute accelerators. While the R\&D
activities are taking place, one can not forget that it is still
necessary to support the existing code and make sure that sufficient
funding and staffing is provided for maintenance and development of the
physics algorithms, as well as for adapting the code to the ongoing
hardware, operating system and compiler evolution.

It is also important to stress once again the need of not only ongoing
R\&D investigations and independent prototyping to explore novel ideas,
but also a continuous integration activity of new runtime performance
improvements and solutions into the Geant4 code base and experiments
applications to profit from each improvement as quickly as possible. An
effort needs to be invested into making it possible to evolve it to meet
the HL-LHC needs without compromising the production quality of the
Geant4 code used by the experiments. We need to ensure that new
developments resulting from the R\&D programs can be tested with
realistic prototypes and, if successful, then integrated, validated, and
deployed in a timely fashion in Geant4 and adopted by the experiments.
The strategy adopted in the successful integration in Geant4 of VecGeom
libraries developed in the context of the GeantV R\&D can provide a
working example of how to proceed to provide some incremental
improvements to existing applications.

No single solution appears at the moment to provide by itself the
processing gains required by HL-LHC, nevertheless if a novel approach
emerges that could do so it should be pursued and carefully evaluated in
terms of gains against the disruption of a complete re-implementation of
how events are simulated.

### 2020-04-30 LHCC/2020/2020-01/input_summary.tex
Break sources into an input file per section

\hypertarget{summary}{%
\section{Summary}\label{summary}}

The main challenges and R\&D lines of interest have been discussed
above. There are clearly identified priority topics for the different
areas in the next few years. We note that longer term R\&D, into areas
such as quantum or neuromorphic computing, should happen at a low level,
but are almost certainly too far from production to deliver improvements
for Run 4 of the LHC.

\hypertarget{physics-event-generators-1}{%
\subsection{Physics Event Generators}\label{physics-event-generators-1}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Gain a more detailed understanding of current CPU costs by accounting
  and profiling.
  \end{quote}
\item
  \begin{quote}
  Survey generator codes to understand the best way to move to GPUs and
  vectorised code, and prototype the port of the software to GPUs using
  data-parallel paradigms.
  \end{quote}
\item
  \begin{quote}
  Support efforts to optimise phase space sampling and integration
  algorithms, including the use of Machine Learning techniques such as
  neural networks.
  \end{quote}
\item
  \begin{quote}
  Promote research on how to reduce the cost associated with negative
  weight events, using new theoretical or experimental approaches.
  \end{quote}
\item
  \begin{quote}
  Promote collaboration, training, funding and career opportunities in
  the generator area.
  \end{quote}
\end{enumerate}

\hypertarget{detector-simulation-1}{%
\subsection{Detector Simulation}\label{detector-simulation-1}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Undertake a detailed performance analysis of current production
  detector simulation code, to better understand where performance
  limitations in data and cache locality arise from.
  \end{quote}
\item
  \begin{quote}
  Improve current codes through refactoring and specialised HEP models
  to avoid bottlenecks identified,
  \end{quote}
\item
  \begin{quote}
  Develop further fast simulation models and techniques, including
  machine learning based options in addition to optimised parametric
  approaches as well as common frameworks for their tuning and
  validation.
  \end{quote}
\item
  \begin{quote}
  Develop GPU codes for tackling very specific, time consuming, parts of
  the simulation for HEP. Specifically calorimetry, including geometry,
  field and electro-magnetic physics processes.
  \end{quote}
\item
  \begin{quote}
  Develop integration prototypes to exercise and benchmark the
  simultaneous use of GPU specific libraries with CPU-based software
  (e.g. Geant4) to cover all particles and processes required in
  experiments simulation frameworks.
  \end{quote}
\end{enumerate}

\hypertarget{reconstruction-and-software-triggers-1}{%
\subsection{Reconstruction and Software
Triggers}\label{reconstruction-and-software-triggers-1}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Develop fast and performant reconstruction software, making optimal
  use of hardware resources. A particularly crucial development for
  HL-LHC is that of pattern recognition solutions (including timing
  information, if available) that can withstand the increase of pile-up
  at HL-LHC. Wherever possible, solutions should be uniform for online
  and offline software, so the experiments can benefit from consistent
  event selection between trigger and final analysis and from more
  effective real-time analysis results.
  \end{quote}
\item
  \begin{quote}
  Develop solutions that can exploit accelerators (FPGA, GPU) for
  trigger and reconstruction, including general-purpose interoperability
  libraries with a focus to generalising and mitigating their
  dependencies on architecture-specific codes.
  \end{quote}
\item
  \begin{quote}
  Further promote activities in the area of data quality and software
  monitoring, as the quality of the recorded/reconstructed data is
  paramount for all physics analysis.
  \end{quote}
\item
  \begin{quote}
  Support efforts towards the implementation of machine learning models
  in experimental framework (C++) to enable more widespread use in
  trigger and reconstruction software.
  \end{quote}
\end{enumerate}

\hypertarget{data-analysis-1}{%
\subsection{Data Analysis}\label{data-analysis-1}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Identify a CPU-efficient data reduction and storage model that serves
  the majority of analyses with a footprint of O(1kb/event), retaining
  the flexibility to accommodate the needs of specialised analyses.
  \end{quote}
\item
  \begin{quote}
  Streamline analysis metadata access and calibration schemes, and
  provide effective book-keeping for fractional datasets.
  \end{quote}
\item
  \begin{quote}
  Develop cross-experimental tools establishing declarative syntax
  and/or languages for analysis description, interfaced with distributed
  computing backends.
  \end{quote}
\item
  \begin{quote}
  Define and improve schemes for interoperability of end-stage
  experimental software with data science/machine learning frameworks.
  \end{quote}
\end{enumerate}

Undertaking this programme of priority topics and R\&D requires, first
and foremost, investment from funding agencies to support developments.
A new generation of developers is also needed to refresh and regenerate
the long lived software projects on which the field so heavily relies.
The transition from R\&D to production ready software is usually long
and so it is urgent to undertake this investment now, with the positive
outcomes ready to become integrated in time for HL-LHC. Regenerating the
cadre of software developers in HEP also requires support for realistic
long term career prospects for those who specialise in this vital area.
Collaboration with industry can certainly be fruitful and is to be
encouraged to allow early access to technology and communication of HEP
problems and priorities.

Almost all the domain areas identify the use of compute accelerators,
particularly GPUs, as being a priority item, which matches our
expectation of how computing hardware will evolve. As a common problem
area, this is an obvious area in which expertise should be developed.

The stakeholders in HEP must be able to give their priorities and
feedback on R\&D areas. Each project has mechanisms for doing this,
while the HSF can help to provide overall input on prioritisation.

### 2020-04-30 LHCC/2020/2020-01/input_analysis.tex
Trim and correct text before first submission
Reduce verbosity, correct mistakes, add references

@@
 
-Examination of collision data is in essence the primary objective of the
-experimental collaborations, coming at the end of the data
-preparation/simulation and reconstruction chain. The stage of analysis,
+Examination of collision data is, in essence, the primary objective of the
+experiment collaborations, coming at the end of the data
+preparation, simulation and reconstruction chain. The stage of analysis,
 starting in most cases from a standard data format generically referred
@@
 broad, encompassing the production of derived data formats as well as
-end stage analysis actions such as producing histograms from those
+end stage analysis actions, such as producing histograms from those
 intermediate datasets and visualising the final results. In the
@@
 volume for both ATLAS and CMS \cite{Ref1,Ref2}. To serve precision analyses
-that require large event statistics, the HL-LHC will increase the
+that require large event statistics experiments will increase the
 recorded event rate by an order of magnitude, compared with a projected
-\textasciitilde 10\% annual growth of storage resources. The large pile-up in HL-LHC
+\textasciitilde10\% annual growth of storage resources. The large pile-up in HL-LHC
 collisions will compound the storage challenge by inflating the data
@@
 avoiding duplication of commonly selected events in multiple datasets.
-Neither of these approaches provides exhaustive coverage of analysis use
-cases on the experiments and alternative data formats are needed for a
+Neither of these approaches provides exhaustive coverage of all the analysis use
+cases for the experiments and alternative data formats are needed for the
 small but significant (10-20\%) fraction of analyses requiring custom
@@
 larger AOD to reside on tape, providing significant savings in disk
-storage. Experience on CMS shows that further reduction (down to
+storage. Experience from CMS shows that further reduction (down to
 \textasciitilde2 kB/event for the nano-AOD) could be possible while
@@
 noting the LHCb analysis model already in Run 3 has to forego the luxury
-of reprocessing Raw data. This constraint from sheer event rate is not
+of reprocessing raw data. This constraint from sheer event rate is not
 without merit as it greatly simplifies the data processing and analysis
@@
 from the LHC. Metadata may include information such as book-keeping of
-processed events, detector conditions and data quality, \emph{a
-posteriori} measured scale factors, and software versioning. This
+processed events, detector conditions and data quality, a
+posteriori measured scale factors, and software versioning. This
 information is often scattered in multiple locations such as:
-``in-file'' data, remote database resources, shared-area text or ROOT files,
+in-file metadata, remote database resources, shared-area text or ROOT files,
 TWiki pages or even in e-mail. The lack of proper metadata integration
@@
 likely not provide any understanding, for the future analyser, about
-what the analysis was effectively doing. Another use case is to re-use
+what the analysis was effectively doing. Another use case is to reuse
 the analysis code on new data, be that the same dataset with improved
@@
 \cite{Ref6}, and packages from the wider data science community. Deeper
-integration with external python tools is also particularly important
+integration with external Python tools is also particularly important
 for enabling state-of-the-art machine learning frameworks such as
@@
 with an effortless integration of HEP specific tools written in C++ with
-industry python big data toolkits.
+industry Python big data toolkits.
 
-The ROOT package, i.e. the foundation and workhorse for LHC analysis,
+The ROOT package, which is the foundation and workhorse for LHC analysis,
 has been the focus of substantial development. On the data storage
@@
 density drives compression and this can vary massively between
-experiments and analyses. This new format called RNTuple \cite{Ref9, ROOT-2020-HL-LHC}, an
+experiments and analyses. This new RNTuple format \cite{Ref9, ROOT-2020-HL-LHC}, an
 evolution of the TTree file format, shows robust and significant
-performance improvements (1-5x faster), which could potentially save
-significant storage space (10-20\%) for the HL-LHC. Another area of
+performance improvements ($1-5\times$ faster), which could potentially save
+significant storage space ($10-20\%$) for the HL-LHC. Another area of
 progress, and potential consolidation in ROOT I/O, is lossy compression,
@@
 declarative approach to analysis processing in either Python or C++
-(always executed in C++), which natively exploits multi-core
+(but always executed in C++), which natively exploits multi-core
 architectures \cite{Ref7} and could be extended to accelerators. Used from
-Python, it allows to pre-process data in C++, and export to numpy.
+Python, it allows pre-processing data in C++, and exporting to numpy.
 Flexibility in interfacing RDataFrame to non-ROOT data formats has
-allowed the ALICE collaboration to prototype its Run 3 analysis model on
+allowed the ALICE collaboration to prototype its Run-3 analysis model on
 this new technology. Growing use of Machine Learning was also
@@
 optimisations, more complex deep learning models and enhanced
-interoperability with common data science tools \cite{Ref14,Ref15}  which is of
+interoperability with common data science tools \cite{Ref14,Ref15}, which is of
 particular importance for inference.
@@
 emerging options, defining clear benchmarks for I/O and processing speed
-would be beneficial.
+is essential.
 
@@
 ``notebook'' technology such as Jupyter. This permits quasi-interactive
-exploration where the annotated history including formatted outputs and
-graphics can be saved and shared with collaborators for reproduction and
+exploration where the annotated history, including formatted outputs and
+graphics, can be saved and shared with collaborators for reproduction and
 adaptation. Although not a substitute for command line scripts in
@@
 called ``combine tools'' developed by CMS and based on RooFit for the
-Higgs discovery). Descriptions of a full analysis in terms of %% Missing reference?
+Higgs discovery. Descriptions of a full analysis in terms of %% Missing reference?
 workflows could then tie together the analysis at the top level \cite{Ref18,Ref21}. With some effort, the HEP community may be able to converge on
@@
 perhaps a few hours in order to keep people productive. As analysis
-tasks are often IO-intensive and not necessarily-CPU intensive, the
+tasks are often I/O-intensive and not necessarily-CPU intensive, the
 right trade off between CPU and high bandwidth storage should be
@@
 will make database-like interfaces seem less scary than they once were
-and may finally allow a more complete integration with analysis code. I
-have a dream.
+and may finally allow a more complete integration with analysis code. 
+% I have a dream. (We all do, but much too chatty for this document!)
 
@@
 Equally though, HEP will continue to have its own unique challenges and
-it's less likely that ultra-performant machine learning inference will
+it is less likely that ultra-performant machine learning inference will
 be top of the data scientists wishlist. Given the challenges presented

### 2020-04-30 LHCC/2020/2020-01/input_generators.tex
Trim and correct text before first submission
Reduce verbosity, correct mistakes, add references

@@
 physics output of the LHC programme. This situation will get
-significantly worse in the high-luminosity running phase of the LHC
-(HL-LHC). The fraction of the CPU resources available to the LHC
-experiments which is used for event generation today is currently
-estimated to be approximately 5-15\%. As is the case for the other big
-consumers of CPU (detector simulation and reconstruction, that have
-until now been the main focus), speedups in generator software are also
+significantly worse for 
+HL-LHC. The fraction of the CPU resources used for event generation today is approximately 5-15\%. 
+As is the case for the other big
+consumers of CPU (detector simulation and reconstruction), speedups in generator software are
 needed to address the overall resource problem expected at the HL-LHC,
-also because more accurate predictions, requiring more complex
-calculations (e.g. beyond NLO or with higher jet multiplicities) will be
-needed from generators. Many other issues, both technical and
-non-technical (e.g. funding, training, careers for those working in this
-area), exist~\cite{Alves:2017she,Gen18}.
+compounded because more accurate predictions, requiring more complex
+calculations will be
+needed (e.g. beyond NLO or with higher jet multiplicities). Many other issues, both technical and
+non-technical exist, e.g. funding, training, careers for those working in this
+area~\cite{Alves:2017she,Gen18}.
 
@@
 Computing (HPC) centres, and generators look like a natural candidate
-for this.
+for this, as they are smaller code bases without complex dependencies.
 
@@
 address them. This is finally condensed into a list of a few
-high-priority items, for the next 18 months before the LHCC review that
-is currently scheduled for fall 2021. A more detailed version of this
+high-priority items, for the next 18 months. A more detailed version of this
 contribution, including a more complete list of references, is
-uploaded on arXiv and will be submitted for publication~\cite{Gen20}.
+uploaded to arXiv and will be submitted for publication~\cite{Gen20}.
 
@@
 Various combinations of software libraries are thus possible, often
-written by different authors and some dating back many years. For a
-given process, the LHC experiments often use different software packages
+written by different authors and frequently many years old. For a
+given process the LHC experiments often use different software packages
 and settings from each other, and a single experiment can generate
@@
 cost reductions. The large number of external packages also complicates
-their long-term maintenance and integration in the experiments software
+their long-term maintenance and integration in the experiments' software
 and workflows, sometimes leading to job failures and computing
@@
 
-A very diverse combination of skills are needed in generator software,
-thus people with many distinct profiles need to contribute. Theorists
+A very diverse combination of skills and profiles are needed for generator software.
+Theorists
 (who create fundamental physics models, and design, develop and optimise
 most generator code), experimentalists working in research (who request
-types/sizes of samples) and in computing (who implement, monitor and
+different samples) and in computing (who implement, monitor and
 account execution of workflows on computing resources), software
@@
 generators, but software engineers need papers describing the software
-modules and overall data/control flow.
+modules and overall control and data flow.
 
@@
 recognition on published papers, and may not be motivated to work on
-software optimisations that are not "theoretical" enough to advance
+software optimisations that are not ``theoretical'' enough to advance
 their careers. Generator support tasks in the experiments may also not
-be enough to secure jobs/funding to experimentalists pursuing a career
+be enough to secure jobs or funding for experimentalists pursuing a career
 in research.
@@
 the way in which they are deployed and integrated by the experiments in
-their software frameworks and computing infrastructure.The goals and
+their software frameworks and computing infrastructure. The goals and
 metrics of software optimisation work may also differ. Theorists are
@@
 with n\textsubscript{jets} MEs plus parton showers. The resulting
-inefficiencies can be relatively high depending on the process, but they
+inefficiencies can be relatively high, depending on the process, but they
 are unavoidable in the algorithmic strategy used by the underlying
@@
 reconstruction and analysis codes, increasing the compute and storage
-requirements. For a fraction r of negative weight events, the number of
+requirements. For a fraction $r$ of negative weight events, the number of
 events to generate increases by a factor $1/(1-2r)^2$: for instance,
-with r=25\% (which may be regarded as a worst-case scenario in top quark
+with $r=25\%$ (which may be regarded as a worst-case scenario in top quark
 pair production~\cite{Fre20}), one needs to generate 4 times as many
 events. Negative weights can instead be almost completely avoided, by
-design, in another popular matching prescription, POWHEG, which however
+design, in another popular matching prescription, POWHEG; however, this
 is only available for a limited number of processes and describes the
@@
 generator software packages would help to further refine R\&D
-priorities:
+priorities.
 
@@
 higher than that in CMS. More detailed analysis of the different
-strategies is ongoing. A practical issue is that these figures had to be
+strategies is ongoing. These figures had to be
 harvested from logs and production system databases, which was
-particularly difficult for CMS, requiring significant person hours to
-extract the required information. It is important to establish better
+particularly difficult for CMS, requiring significant person hours.
+It is important to establish better
 mechanisms to collect this information, to allow for an easy comparison
@@
 More generally, as is the case in other areas of HEP, some R\&D on
-generator software would certainly be needed to modernise it and make it
+generator software is certainly be needed to modernise it and make it
 more efficient, or even port it to more modern computing 
@@
 kernels. Porting and optimising generators on GPUs is essential to be
-able to use modern GPU-based HPCs (such as SUMMIT, where 95\% of the
-compute capacity comes from GPUs). The work done in this direction in
+able to use modern GPU-based HPCs. The work done in this direction in
 the past on MG5\_aMC, that never reached production quality, is now
-being revamped by the WG, in collaboration with the MG5\_aMC team, and
+being reinvigorated by the WG, in collaboration with the MG5\_aMC team, and
 represents one of the main R\&D priorities of the WG. This work is
@@
 multithreading approaches. The fact that some generators are not even
-thread safe, may also be a problem, for instance to embed them in
+thread safe may also be a problem, for instance to embed them in
 multi-threaded event processing frameworks, such as that of CMS.
@@
 
-\paragraph{NNLO: status of theoretical physics research.} The first question is
+\paragraph{NNLO: status of theoretical physics research.} The first question is,
 for which processes NNLO precision will be available at the time of the
-HL-LHC. For example, when would NNLO be expected for 2-\textgreater3 or
+HL-LHC? For example, when would NNLO be expected for $2 \rightarrow 3$ or
 even higher multiplicity final states? And for lower multiplicity final
@@
 \paragraph{NNLO: experimental requirements for data analysis at HL-LHC.} The
-second question is for which final states unweighted event generation
-with NNLO precision would actually be required, and how many events
-would be needed. One should also ask if reweighting LO event samples to
+second question is, for which final states unweighted event generation
+with NNLO precision would actually be required? and how many events
+would be needed? One should also ask if reweighting LO event samples to
 NNLO would not be an acceptable cheaper alternative to address the
-experimental needs.
+experiment's needs.
 
@@
 phase space the number of unweighted events must be strictly
-proportional to the luminosity. For example, in the low pT regions of W
+proportional to the luminosity? For example, in the low $p_\text{T}$ regions of W
 boson production it is probably impossible to keep up with the data, due

### 2020-04-30 LHCC/2020/2020-01/input_introduction.tex
Trim and correct text before first submission
Reduce verbosity, correct mistakes, add references

@@
 software and GPU codes have been used in production by some experiments
-for years~\cite{calafiura2018hep, albrecht2018hepexec}.
-In other areas migration is still ongoing. In others it is
-not even yet started. Generic heterogeneous programming models have
+for years now~\cite{calafiura2018hep, albrecht2018hepexec}.
+In other areas migration is still ongoing. In yet others it has
+not even started. Generic heterogeneous programming models have
 existed for some time, and new ones are arising, but there is, as yet,
@@
 shared. Support is needed for multiple architectural targets and
-ensuring the correct runtime on heterogeneous environments.
+ensuring the correct runtime in heterogeneous environments.
 

### 2020-04-30 LHCC/2020/2020-01/input_simulation.tex
Trim and correct text before first submission
Reduce verbosity, correct mistakes, add references

@@
 transferred and analysed hundreds of billions of simulated events. This
-effort required a major fraction of the WLCG computing 
+effort required a major fraction of WLCG's computing 
 resources~\cite{ALICE-TDR-12, ATLAS-TDR-17, CMS-TDR-7, LHCb-TDR-11, WLCG-CM-Update, 
@@
 an order of magnitude more data. At the same time, the demand for
-detector simulation will grow accordingly, in order to keep the
-statistical uncertainties as low as possible. However, the expected
+detector simulation will grow accordingly so that
+statistical uncertainties are kept as low as possible. However, the expected
 computing resources will not suffice if current detector simulation is
-to be used; the availability of simulated samples of sufficient size
-will soon become a major limiting factor as far as new physics
+used; the availability of simulated samples of sufficient size
+would soon become a major limiting factor as far as new physics
 discoveries, for example through precision measurements, are concerned.
@@
 of virtuality. This fostered progress in physics by comparison of
-competing models on same energy ranges, and allowed complementary models
+competing models on the same energy ranges, and allowed complementary models
 to be combined to cover large energy ranges. The simulation applications
@@
 Improvements of physics models need to continue to take place to prevent
-systematic uncertainties from the simulation from becoming a dominant
+systematic uncertainties from simulation from becoming a dominant
 factor, however, their exact influence on the precision of physics
@@
 Taking all those elements into account, three main development paths have emerged to be
-undertaken simultaneously: these have started to be pursued, however,
+undertaken simultaneously and these have started to be pursued. However,
 effort in any one of them is far from sufficient to exploit their
@@
 
-On one hand, it is needed to continue to modernise
+Firstly, it is necessary to continue to modernise
 and refactor the simulation code by using more optimal, modern
-programming techniques. Those improvements, requiring considerable
-effort but no major paradigm changes, can reasonably bring several tens
-of percent of speed-up, by improving the throughput on modern CPUs,
-avoiding constant churn in data and instruction caches. This will
-certainly help to produce more simulated events, but it will not be
-sufficient to meet the high luminosity needs of the LHC experiments.
+programming techniques.
 
-On the other hand, different fast simulation techniques, where tracking of
+%Those improvements, requiring considerable
+%effort, but involving no major paradigm changes, can reasonably bring several tens
+%of percent of speed-up, by improving throughput on modern CPUs,
+%avoiding constant churn in data and instruction caches. This will
+%certainly help to produce more simulated events, but it will not be
+%sufficient to meet the high-luminosity needs of the LHC experiments.
+
+Secondly, different fast simulation techniques, where tracking of
 individual particles in given detectors is replaced by some sort of
-parameterisation, have already been extensively used for LHC simulation
-campaigns, where some precision is traded to achieve higher event
-throughput (thus lowering the cost of samples overall). Techniques to
-re-use part of the simulated events multiple times have also emerged.
-New technologies (like Machine Learning), should certainly be
-investigated as fast simulation approaches. All of these will become
-more and more significant and will need to be seamlessly integrated with
-the detailed simulation provided by Geant4.
+parameterisation and precision is traded to achieve higher event
+throughput.
+% (thus lowering the cost of samples overall). Techniques to
+%reuse part of the simulated events multiple times have also emerged.
+%New technologies (like machine learning), should certainly be
+%investigated as fast simulation approaches. All of these will become
+%more and more significant and will need to be seamlessly integrated with
+%the detailed simulation provided by Geant4.
 
@@
 accelerators (like GPUs or FPGAs) is the third emerging path that needs
-to be undertaken and that requires intense R\&D effort. We will discuss
-more in detail those three development axes in the following sections.
+to be undertaken and that requires intense R\&D effort.
 
-In this section we will outline the developments identified in and since the
-Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} focusing
-in particular in these three area.
+% Discussion here of the 3 approached is quite asymmetric - maybe too many details for the first two
+
+We will discuss
+more in detail these three development axes in the following sections, 
+outlining the developments identified in and since the
+Community White Paper~\cite{Albrecht2019, SimFoundation2018hep}.
 
@@
 writing more compact and more specialised code can lead in certain
-places to up to several tens percent speed-up~\cite{GeantV}. The
+places to up to several tens of percent speed-up~\cite{GeantV}. The
 technical improvements to explore here consist of avoiding too many
 virtual layers and improving the instruction and data locality, leading
-to better cache use. Going even further as far as specialisation is
+to better cache use.
+
+Going even further, as far as specialisation is
 concerned, studies suggest that implementing compact, self-contained
@@
 scoring (to simulate, for instance, the response of an EM calorimeter)
-can also bring speed-ups to the complete application~\cite{GeantV}. As
+can also bring speed-ups to the complete application~\cite{GeantV}.
+
+As
 far as architectural modifications are concerned, the design of the
@@
 certainly provide the necessary starting points for further R\&D in that
-direction. Most of these investigations have started and it is hoped to
-see some first results on a year time-scale. More effort, however, will
+direction.
+
+Most of these investigations have started and it is hoped to
+see some first results on a year long time-scale. More effort, however, will
 be required to integrate any eventual changes in the production code in
-a timely manner. The GeantV Vector prototype has allowed the assessment
+a timely manner.
+
+The GeantV Vector prototype has allowed the assessment
 of the achievable speedup using a novel, vectorised approach to particle
@@
 could not be tackled without a major additional long term effort~\cite{GeantV}.
+
 At the same time libraries developed in the context of the
@@
 All LHC experiments have dedicated simulation frameworks making use of
-the Geant4 toolkit for modelling their detectors and the physics
-processes occurring in the experimental setups. Every experiment
+the Geant4 toolkit for modelling physics processes in their detectors. Every experiment
 configures and uses what is provided by Geant4 for their specific needs.
 Effort has been spent by each experiment since the publication of the
-Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} to benchmark their 
+Community White Paper to benchmark their 
 simulation code and to maximally
@@
 investigated, neutron Russian roulette used for dedicated simulations.
-All of them can have major impacts on simulation time and sharing
+All of these can have major impacts on simulation time and sharing
 knowledge and experience between experiments can be of high benefit even
@@
 harmonising their different architectural choices with the Geant4
-implementation~\cite{CMS02, LHCb01, ATLAS01}, and have or are planning to
-use them in their production environment to gain access to a larger set
+implementation~\cite{CMS02, LHCb01, ATLAS01}, and have used, or are planning to
+use, them in their production environment to gain access to a larger set
 of distributed resources. ATLAS has been running their multi-process
@@
 simulation on available resources including HPCs~\cite{CMS03}. Both
-approaches have allowed to exploit a higher number of computing
-resources by gaining access to systems with limited memory.
+approaches have allowed the exploitation of additional computing
+resources by enabling access to lower memory systems.
 
@@
 
-The so-called `fast simulation' techniques consist of replacing the
-tracking of individual particles through (a part of) the detector,
+Fast simulation techniques consist of replacing the
+tracking of individual particles through the detector (or part thereof),
 including all the physics processes they would undergo, by a
-parameterisation where the detector response is produced directly as a
-function of the incoming particle type and energy. An example of such a
+parameterisation, where the detector response is produced directly as a
+function of the incoming particle type and energy.
+
+An example of such a
 parameterisation was implemented many years ago in the context of the H1
@@
 Geant4 toolkit under the name of the GFLASH library, became the starting
-idea for several `custom' implementations, specific for other
+idea for several custom implementations, specific for other
 calorimeters. Those were interfaced to experiments' simulation
-applications through the available `hook' in the Geant4 toolkit. The LHC
+applications through a hook in the Geant4 toolkit. The LHC
 experiments implemented, for example, dedicated parametrised response
@@
 detector geometries. This kind of `classical' fast simulation, which
-describes the particle shower shape using some complex mathematical
-functions with several parameters, will remain an important tool;
-however, it is also clear that their overall precision for the different
+describes the shower shape using some complex mathematical
+functions, with several parameters, will remain an important tool;
+however, it is also clear that their overall precision for different
 energy values, especially for highly granular and complex calorimeters
-will always be relatively low. The recent developments of Deep
-Learning-based techniques have opened an exciting possibility of
+will always be relatively low.
+
+Recent developments of deep
+learning-based techniques have opened up an exciting possibility of
 replacing those `classical' parameterisations by trained neural networks
 that would reproduce the detector response. This approach consists of
-training generative models such as Generative Adversarial Networks
+training generative models, such as Generative Adversarial Networks
 (GAN), Variational Auto-Encoders (VAE) or Autoregressive Generative
@@
 energies), still requires further effort. Detailed physics validation of
-those tools and understanding of the capability to reproduce the
-fluctuations is a prerequisite towards `production-quality' fast
+those tools and understanding their capability to reproduce
+fluctuations is a prerequisite towards production quality fast
 simulation libraries. This work is already ongoing, but will require,
-over the next few years, more effort to be invested combining the
+over the next few years, more effort to be invested, combining
 physics and Machine Learning expertise. Given that the training of the
 deep learning algorithms usually requires using sufficiently large MC
-samples produced using standard techniques, the development of the novel
+samples produced using standard techniques, the development of these novel
 tools does not necessarily remove the need to speed up Geant4.
@@
 been spent in the experiments frameworks to combine as much as possible
-fast and detailed implementation. In particular ATLAS and CMS are using
-in their baseline production shower libraries for the modelling of
-calorimeters in the forward region in combination with the detailed
+fast and detailed implementations. In particular, ATLAS and CMS are using
+shower libraries in production for the modelling of
+calorimeters in the forward region, in combination with the detailed
 simulation of the rest of the sub-detectors. In LHCb the re-use of part
@@
 care paid to ensure keeping under control bias on statistical
-uncertainties~\cite{LHCb02}, its applicability in ATLAS and CMS could be
+uncertainties~\cite{LHCb02}. Applicability of this technique in ATLAS and CMS could be
 explored for specific cases. Similarly, the re-use of simulated or real
-events to mimic the background to hard-scatter events is also exploited,
+events to mimic the background to hard-scatter events is also being exploited,
 where additional challenges exist concerning storage and I/O due to the
@@
 of reconstructed objects (tracks and clusters) for rapid evaluation of
-physic's reach is also exploited by the experiments with generic
+physics reach is also exploited by the experiments with generic
 frameworks for fast simulation of typical collider detectors being used~\cite{bib-Delphes, PGS4}. 
-The advantage of such frameworks is to allow
+The advantage of such frameworks is that they allow
 simultaneous studies for different experiments as well as their
-potential use by the theory community. Experiments specific fully
+potential use by the theory community. Experiment specific fully
 parameterised simulations providing reconstructed objects compatible
@@
 from `smeared reconstructed quantities' to parameterisation for specific
-detectors to the detailed Geant4, will need to be exploited by the
+detectors to detailed Geant4, will need to be exploited by the
 experiments for different tasks. Seamless ways of providing them in a
-transparent integrated way in the experiments' simulation frameworks
-should continue to be pursued with Geant4 supporting hooks for such
-integrations.
+transparent and integrated way in the experiments' simulation frameworks
+should continue to be pursued with Geant4 support.
 
@@
 
-The hardware landscape has been always setting direction as far as
-software evolution was concerned. The recent adaptation of the
+The hardware landscape has always set directions as far as
+software evolution is concerned. The recent adaptation of the
 simulation code to multithreading (MT) turned out to be relatively
 straightforward, but still took several years to adopt and validate.
-Geant4 can now run in the MT mode efficiently, distributing the events
-between the threads and sharing resources like the geometry description
-or physics data and thus reduce the memory footprint of a simulation
+Geant4 can now run in MT mode efficiently, distributing the events
+between threads and sharing resources, like the geometry description
+or physics data, and thus reduce the memory footprint of a simulation
 while maintaining its throughput performance. The efficient utilisation
@@
 hand turned out to be more challenging. Not only was a limited part of
-the code `vectorisable' but also, as the already mentioned GeantV R\&D
+the code vectorisable but also, as the already mentioned GeantV R\&D
 has shown, overheads related to gathering the data for vector processing
@@
 
-The most recent `revolution' in the computing hardware systems consists
-of using Graphics Processing Units (GPUs) for general computing. Several
+The most recent revolution in computing hardware is the
+use of Graphics Processing Units (GPUs) for general purpose computing. Several
 attempts have been made to port specific simulation code to GPUs. A few
-of them turned out to be successful leading to factors of several
-hundred or thousand of speedup~\cite{GPU-MPEXS-DNA, Hybrid-Gate, Opticks}, however, they were always
+of them turned out to be successful, leading to factors of several
+hundred or a thousand speedup~\cite{GPU-MPEXS-DNA, Hybrid-Gate, Opticks}; however, they were always
 limited to a very specific simulation problem, far from what is
-addressed by a general HEP simulation. The application of GPUs has been
+addressed by a general HEP simulation. This application of GPUs has been
 very successful in restricted domains like medical physics simulation,
@@
 often extremely simple compared to the LHC detectors. A natural
-`extrapolation' of those approaches to a general HEP simulation is very
+extrapolation of those approaches to a general HEP simulation is very
 difficult to imagine, because the stochasticity of the simulated events
@@
 are concerned, not to mention, simply the feasibility of porting the
-whole simulation to the GPU code for which specific expertise would be
-needed. On the other hand, running only very small pieces of the
+whole simulation to GPU code, for which specific expertise would be
+needed.
+
+On the other hand, running only very small pieces of the
 simulation application on GPUs does not seem to be efficient either, as
-gathering data and transferring it from the host to the device and back
-may strongly limit any possible performance gain, similarly to what was
+gathering data and transferring it from the host to the device and back again
+may strongly limit any possible performance gain, similar to what was
 seen with the GeantV vectorisation R\&D. The most plausible approaches
@@
 device. Such libraries could be the right compromise between the
-complexity of the algorithm that needs to be ported to the device and
-the overall time that is now spent (and could be gained) on the
-corresponding of the full simulation on the CPU. The investigation in
-those directions are starting now, but will certainly require a
-considerable effort to be further invested before being able to judge
-their real impact. The first steps that are currently being undertaken
+complexity of the algorithm that needs to be ported and
+the overall time that is now saved in
+by the full simulation on the CPU.
+
+An investigation of
+these directions is starting now, but it will certainly require
+considerable effort to be invested before being able to judge
+the real impact. The first steps that are currently being undertaken
 consist of implementing, on GPUs, prototypes based on VecGeom geometry
-and navigation capabilities. If those prototypes turn out to be
+and navigation capabilities~\cite{Wenzel_2017}. If those prototypes turn out to be
 successful, the next challenge will consist of adding tracking and
@@
 modules, stripped of some features which would be expensive to implement
-or support efficiently on GPUs. Another interesting avenue consists of
+or support efficiently on GPUs.
+
+Another interesting avenue consists of
 exploring the application of some vendor libraries (like Nvidia Optix).
 Those libraries, originally developed for ray tracing, have several
-similarities with general particle transport and if applied
-successfully, could lead to major speed-ups. All those efforts certainly
+similarities with general particle transport and, if applied
+successfully, could lead to major speed-ups. All these efforts certainly
 require a major investment and new developers, expert in those
@@
 as a basis to develop their own code. Nevertheless the large variety of
-detectors' technologies used by the experiments reduces its
-applicability. Digitisation not yet being a limiting factor in terms of
-CPU requirements also does not urge developments in this area.
+detector technologies used reduces its
+applicability. Digitisation is not yet a limiting factor, in terms of
+CPU requirements, so developments in this area have not been a priority.
 
@@
 triggers. Hardware triggers are based on very specific custom devices
-and a general approach does not seem feasible even if some
+and a general approach does not seem very feasible, even if some
 parametrisation could be generalised.
@@
 
-To conclude, it is important to realise that there is a considerable
-risk for the simulation to become a major limiting factor as far as new
+It is important to realise that there is a considerable
+risk that simulation becomes a major limiting factor as far as new
 physics discoveries are concerned if no serious effort is invested in
-R\&D. We need the concerted effort of different experts in physics,
-Machine Learning and GPUs. There are too many unknowns to focus on only
-one direction, so a lot of different prototyping is needed. We need to
-further develop Geant4 while working on fast simulation algorithms and
-doing extensive R\&D on leveraging compute accelerators. While the R\&D
+R\&D. Concerted effort of different experts in physics,
+Machine Learning and GPUs is needed. There are too many unknowns to focus on only
+one direction, so a lot of different prototyping is needed. Development of Geant4,
+while working on fast simulation algorithms and
+doing extensive R\&D on leveraging compute accelerators is required. While the R\&D
 activities are taking place, one can not forget that it is still
 necessary to support the existing code and make sure that sufficient
-funding and staffing is provided for maintenance and development of the
-physics algorithms, as well as for adapting the code to the ongoing
-hardware, operating system and compiler evolution.
+funding and staffing is provided for maintenance and development of
+physics algorithms, as well as for adapting the code to any updated CPU
+hardware, operating systems and new compilers.
 
-It is also important to stress once again the need of not only ongoing
-R\&D investigations and independent prototyping to explore novel ideas,
-but also a continuous integration activity of new runtime performance
+It is also important to stress once again the need for a continuous
+integration activity of new runtime performance
 improvements and solutions into the Geant4 code base and experiments
-applications to profit from each improvement as quickly as possible. An
+applications to profit from any improvement as quickly as possible. An
 effort needs to be invested into making it possible to evolve it to meet

### 2020-04-30 LHCC/2020/2020-01/ctcs.tex
Upload LHCC Common Tools and Community Spftware document (#131)
Upload "final" version into GitHub

N.B. This version is squashed - editing history is in lhcc-document branch in my fork

@@
 \usepackage{lineno}  % for line numbering during review
-\linenumbers
+%\linenumbers
 
@@
  & & HSF-DOC-2020-01 \\
- & & Zenodo DOI Here \\
+ & & 10.5281/zenodo.3779250 \\
  & & May 1, 2020 \\ % use \date or hardwire e.g. December 15, 2017

### 2020-04-30 LHCC/2020/2020-01/input_recoTrigger.tex
Upload LHCC Common Tools and Community Spftware document (#131)
Upload "final" version into GitHub

N.B. This version is squashed - editing history is in lhcc-document branch in my fork

@@
 
-At the HL-LHC, the central challenge for high-level triggers (e.g.,
+At the HL-LHC, the central challenge for high-level triggers (e.g.
 software triggers) and object reconstruction is to maintain excellent
 efficiency and resolution in the face of high pileup values, especially
-at low transverse momenta. Detector upgrades such as increases in
+at low transverse momenta. Detector upgrades, such as increases in
 channel density, high precision timing and improved detector layouts are
@@
 complexity at the HL-LHC also requires the development of software
-algorithms that can process events with a similar cost per event to
+algorithms that can process events with a similar cost per-event to
 Run-2 and Run-3. At the same time, algorithmic approaches need to
@@
 This section focuses on the challenges identified in the Community White
-Paper~\cite{Alves:2017she,albrecht2018hepexec,albrecht2018hep} and the development of solutions that. It also includes
+Paper~\cite{Alves:2017she,albrecht2018hepexec,albrecht2018hep} and the
+development of solutions since then. It also includes
 mentions of open source software efforts that have developed within or
@@
 analysis\footnote{LHCb \cite{Aaij:2019uij} and ALICE \cite{Buncic:2011297} will both
-  stream the full collision rate to real-time or quasi-realtime software
+  stream the full collision rate to real-time or quasi-real-time software
   trigger systems.}.
@@
 system, driven by use cases where the physics potential improves when
-analysing more events than what can be written out in full with
+analysing more events than can be written out in full with
 traditional data processing.
@@
 These approaches remove the detector data that typically makes up the
-raw data tier kept for offline reconstruction, and keep only a limited
-set of analysis objects reconstructed within the high level trigger.
-
-Real-time analysis techniques are being adopted to enable a wider range
-of physics signals to be saved by the trigger for final analysis. As
-rates increase, these techniques can become more important and
-widespread by enabling only the parts of an event associated with the
-signal candidates to be saved, reducing the required disk space. The
-experiments are focusing on the technological developments that make it
+raw data kept for offline reconstruction, and keep only a limited
+set of analysis objects reconstructed within the high level trigger or
+keepig only the parts of an event associated with the
+signal candidates, reducing the required disk space.
+The experiments are focusing on the technological developments that make it
 possible to do this with acceptable reduction of the analysis
-sensitivity and with minimal biases. The HSF trigger and reconstruction
+sensitivity and with minimal biases. The HSF Reconstruction and Software Trigger
 group has been encouraging cross-talk between LHC experiments and beyond
-on real-time analysis, as these kinds of non-standard trigger and
-reconstruction workflows increase the physics output for selected
+on real-time analysis, as these kinds of workflows increase the physics output for selected
 physics cases without adding significant overhead to the current
@@
 
-Active research topics include compression and custom data formats;
+Active research topics also include compression and custom data formats;
 toolkits for real-time detector calibration and validation which will
@@
 deploy real-time analyses without compromising data taking quality.
-Use-cases for real-time analysis techniques have expanded during the
+Use cases for real-time analysis techniques have expanded during the
 last years of Run 2 and their use appears likely to
 grow already during Run 3 data taking for all experiments owing to
-improvements in the HLT software and farms \textbf{Cite CMS and ATLAS CDRs}. Further ideas include retaining very reduced data from as much
-of the data taking rate as possible from selected regions and
+improvements in the HLT software and farms. % Missing reference to CMS and ATLAS CDRs
+Further ideas include retaining high-rate, but very reduced data, from selected regions and
 sub-detectors, even up to the 40 MHz bunch crossing rate~\cite{hannes_sakulin_2019_3598769}.
@@
 
-Algorithmic areas of particular importance at HL-LHC experiments are
+Algorithmic areas of particular importance for HL-LHC experiments are
 charged-particle trajectory reconstruction (\textit{tracking}), including
@@
 
-The CPU needed for event reconstruction in Run-2 and Run-3 has been
+The CPU needed for event reconstruction in Runs 2 and 3 is
 dominated by charged particle reconstruction (tracking). This is still a
 focus for HL-LHC triggering and event reconstruction, especially when
-the need for efficiently reconstructing low transverse momentum
+the need to efficiently reconstruct low transverse momentum
 particles is considered.
 
-The huge increase in the number of charged particles and hence the
-combinatorial load for tracking algorithms at future colliders will put
+The huge increase in the number of charged particles, and hence the
+combinatorial load for tracking algorithms, at future colliders will put
 great strain on compute resources. To minimise the memory footprint,
-tracking software needs to be thread-safe to support multi-threaded
+tracking software needs to be thread-safe and to support multi-threaded
 execution per core. At the same time, the software has to be efficient
@@
 Since the CWP, experiments have made progress towards improving software
-tracking efficiency in HL-LHC simulation, see e.g. Ref. \cite{ATL-PHYS-PUB-2019-041}. A
+tracking efficiency in HL-LHC simulation, e.g. \cite{ATL-PHYS-PUB-2019-041}. A
 number of collaboration and community initiatives have been focusing on
-tracking, targeting both offline and online, and we list them below.
+tracking, targeting both offline and online, and these are listed below.
 
-The \emph{\textbf{ACTS}} project \cite{Ai:2019kze,Gumpert_2017} is an attempt to encapsulate
+The \textbf{ACTS} project \cite{Ai:2019kze,Gumpert_2017} is an attempt to encapsulate
 the current ATLAS track reconstruction software into an
-experiment-independent and framework-independent tracking software
+experiment-independent and framework-independent tracking software,
 designed to fully exploit modern computing architectures. It builds on
@@
 reconstruction. ACTS has active collaborators from other HEP
-experiments, such as FASER, and provides support to Belle-II, EIC, CEPC,
-etc.
+experiments, such as FASER, and provides support to Belle-II, EIC and CEPC.
 
-The aim of the \textbf{mkFit} project is to speed up Kalman filter (KF)
+The aim of the \textbf{mkFit} \cite{cerati2019speeding} project is to speed up Kalman filter (KF)
 tracking algorithms using highly parallel architectures and to deliver
@@
 activities focused on delivering a production quality software setup to
-perform track building in the context of the LHC Run-3 (or Run-2). The
+perform track building in the context of LHC Runs 2 and 3. The
 implementation relies on single-precision floating point mathematics and is
-available for multicore CPUs and achieves significant gains in compute
-performance from the use of vector instructions set extending to
-AVX-512, based on Matriplex library (a part of the mkFIt project).
+available for multicore CPUs. It achieves significant gains in compute
+performance from the use of vector instructions, extending to
+AVX-512, based on Matriplex library (a part of the mkFit project).
 
@@
 reconstruction. This allows the mitigation of the effects of pile-up by
-exploiting the time-separation of collision products~\cite{Collaboration:2623663,CMS:2667167}\\
-The experiment communities have been working on timing detector
+exploiting the time-separation of collision products~\cite{Collaboration:2623663,CMS:2667167}
+
+Experiment communities have been working on timing detector
 reconstruction and object identification techniques in complex
@@
 procedures to check trigger rates as well as the performance of the low-
-and high-level physics objects in the data {[}ref ATLAS DQ and CMS
-DQ{]}.
+and high-level physics objects in the data. % Missing reference {[}ref ATLAS DQ and CMS DQ{]}.
 
@@
 current and future hardware. The goal for work in this area is to evolve
-current toolkit and algorithm implementations, and best programming
-techniques to better use SIMD capabilities of current and future
+current toolkit and algorithm implementations, and develop best programming
+techniques to better use the SIMD capabilities of current and future
 computing architectures. Since the CWP, algorithm development projects
-(e.g. \cite{LHCB-FIGURE-2019-002}
-as well as mkFit) have demonstrated success in increasing the use of vector
-units. In addition, HEP has increasingly benefited from industry
+have demonstrated success in increasing the use of vector
+units~\cite{cerati2019speeding, LHCB-FIGURE-2019-002}. In addition, HEP has increasingly benefited from industry
 developed software, including machine learning toolkits (e.g.,
 TensorFlow), that typically make excellent use of vector units.
-Challenges remain in this area. For example, full exploitation of this
+Challenges remain in this area: for example, full exploitation
 must also account for many generations of hardware that experiments must
 exploit on the grid. In addition, to realise full performance gains, a
-large fraction of the application must be improved to use SIMD processor
-capabilities (partly due to turbo-boost capabilities of processors)
+large fraction of the application must be improved to use SIMD processing
+capabilities (partly due to turbo-boost capabilities of processors).
 
@@
 
-It may be desirable or even necessary to deploy new algorithms,
-including advanced machine learning techniques, in order to face the
-increase in event complexity without increasing the per-event
+It may be desirable, or even necessary, to deploy new algorithms that
+include advanced machine learning techniques to manage the
+increase in event complexity without increasing per-event
 reconstruction resource needs.
 
-Work is already undergoing in the collaborations to evolve or rewrite
+Work is already ongoing in the collaborations to evolve or rewrite
 existing toolkits and algorithms focused on their physics and technical
@@
 
-Computing platforms are generally evolving towards having more cores in
-order to increase processing capability. In addition different
-architectural models, e.g. GPUs, FPGAs are available.
-The goal is to improve the throughput of multithreaded
+Computing platforms are generally evolving towards having more cores
+or to adopt different
+architectural models (GPUs, FPGAs) to increase processing capability~\cite{acm10.1145/3282307}.
+The goal for HEP is to improve the throughput of
 software trigger and event reconstruction applications, so current
@@
 efficiently exploit these opportunities. The first algorithms that should be
-targetted are those which are most time consuming.
+targeted are those which are most time consuming.
 
@@
 designed for, or have successfully adapted to, hardware accelerators,
-most notably to NVidia GPUs using CUDA. Two projects targeting Run-3 and
+most notably to NVIDIA GPUs using CUDA. Two projects targeting Run-3 and
 with prospects for use in Run-4 are Patatrack (from CMS) \cite{andrea_bocci_2019_3598824} and
@@
 
-\textbf{Patatrack} is a CMS initiative aimed to use heterogeneous
-computing for the charged particle reconstruction. Within the CMS
+\textbf{Patatrack} is a CMS initiative aimed at using heterogeneous
+computing for charged particle reconstruction. Within the CMS
 codebase (CMSSW), this project has demonstrated that physics
 reconstruction code can be written to leverage heterogeneous
-architectures, like GPUs, to achieve a significant speed-up while
+architectures, like GPUs, and achieve a significant speed-up, while
 reducing the overall cost and power consumption of a system. An example
@@
 \textbf{Allen} is a data processing framework for GPUs, as well as a
-specific implementation of a first-level trigger for LHCb Run 3
+specific implementation of a first-level trigger for LHCb Run-3
 data-taking. Allen is optimised for sustaining the rate required by
 real-time processing in LHCb, but can be used as a more general GPU data
-processing as it includes a scheduler and a memory manager that can be
-integrated into Gaudi. This permits the extension and implementation of
+processing system as it includes a scheduler and a memory manager that can be
+integrated into Gaudi~\cite{gaudi_paper}. This permits the extension and implementation of
 Allen within other experimental software using the same underlying
@@
 parallel algorithms \cite{attila_krasznahorkay_2019_3599103} and to enable the use of machine learning
-algorithms in object reconstruction and identification {[}\emph{waiting
-for an email from ATLAS with references}{]}.
+algorithms in object reconstruction and identification. % Missing references {[}\emph{waiting for an email from ATLAS with references}{]}.
 
 The experience acquired with these projects shows that GPUs can be used
-for increasing the throughput in cases of large-size (ALICE) \cite{david_rohr_2019_3599418} and small-size events per second (LHCb), and to speed up
-specific parts of the data processing (CMS, Patatrack).
+for increasing the throughput in cases of large-size (ALICE)
+\cite{david_rohr_2019_3599418} and small-size events per second (LHCb), and to speed up
+specific parts of the data processing (CMS Patatrack).
 
@@
 
-Heterogeneous computing is still a rapidly developing and changing
-field. The burden of writing architecture-specific code repeatedly will
-certainly exceed the available effort from the experiments. Efforts are
-ongoing in this direction, since the CWP, trying to take advantage of
-the latest development in synergy with industry and attempting to
-increase the use of centres with heterogeneous resources by experiments.
+%Heterogeneous computing is still a rapidly developing and changing
+%field. The burden of writing architecture-specific code repeatedly will
+%certainly exceed the available effort from the experiments. Efforts are
+%ongoing in this direction, since the CWP, trying to take advantage of
+%the latest development in synergy with industry and attempting to
+%increase the use of centres with heterogeneous resources by experiments.
 
@@
 
-The use of open-source elements of the LHC software stack for trigger
+The use of open-source elements of the LHC software stacks for trigger
 and reconstruction by smaller experiments, especially in case of common
@@
   adoption of software relying on similar detector descriptions.}, is
-still worthwhile This benefits experiments with a smaller developer base
-than LHC experiments, and increases the return on investment for LHC
+still worthwhile This benefits smaller experiments
+and increases the return on investment for LHC
 experiment software. For this reason, we also encourage the addition of

### 2020-04-30 LHCC/2020/2020-01/input_summary.tex
Upload LHCC Common Tools and Community Spftware document (#131)
Upload "final" version into GitHub

N.B. This version is squashed - editing history is in lhcc-document branch in my fork

@@
 above. There are clearly identified priority topics for the different
-areas in the next few years. We note that longer term R\&D, into areas
+areas in the next few years\footnote{Longer term R\&D, into high-risk, high-reward blue sky areas,
 such as quantum or neuromorphic computing, should happen at a low level,
 but are almost certainly too far from production to deliver improvements
-for Run 4 of the LHC.
+for Run 4 of the LHC.}.
 
@@
   Improve current codes through refactoring and specialised HEP models
-  to avoid bottlenecks identified,
+  to avoid bottlenecks identified.
   \end{quote}
@@
   \begin{quote}
-  Develop GPU codes for tackling very specific, time consuming, parts of
+  Undertake R\&D into GPU codes for tackling very specific, time consuming, parts of
   the simulation for HEP. Specifically calorimetry, including geometry,
@@
   libraries with a focus to generalising and mitigating their
-  dependencies on architecture-specific codes.
+  dependencies on architect\-ure specific codes.
   \end{quote}
@@
   Further promote activities in the area of data quality and software
-  monitoring, as the quality of the recorded/reconstructed data is
+  monitoring, as the quality of the recorded and reconstructed data is
   paramount for all physics analysis.
@@
   Define and improve schemes for interoperability of end-stage
-  experimental software with data science/machine learning frameworks.
+  experimental software with data science and machine learning frameworks.
   \end{quote}
@@
 problems and priorities.
+R\&D that is successful will become the next generation of production
+software, but this will require lifetime support for maintainance and
+further evolution.
 
@@
 feedback on R\&D areas. Each project has mechanisms for doing this,
-while the HSF can help to provide overall input on prioritisation.
+while the HSF can help to provide overall input on prioritisation
+and coordination of common cross-experiment activities.

### 2020-05-07 LHCC/2020/2020-01/input_simulation.tex
Add reference to GeantV arXiv

@@
 writing more compact and more specialised code can lead in certain
-places to up to several tens of percent speed-up~\cite{GeantV}. The
+places to up to several tens of percent speed-up~\cite{GeantV-2019, GeantV-2020}. The
 technical improvements to explore here consist of avoiding too many
@@
 scoring (to simulate, for instance, the response of an EM calorimeter)
-can also bring speed-ups to the complete application~\cite{GeantV}.
+can also bring speed-ups to the complete application~\cite{GeantV-2019, GeantV-2020}.
 
@@
 vectorisation where some bottleneck (for example geometry navigation)
-could not be tackled without a major additional long term effort~\cite{GeantV}.
+could not be tackled without a major additional long term effort~\cite{GeantV-2019, GeantV-2020}.
 
@@
 threads, etc.) to efficiently (development time wise and run time wise)
-profit from it~\cite{GeantV}.
+profit from it~\cite{GeantV-2019, GeantV-2020}.
 

### 2020-08-19 LHCC/2020/2020-01/a2tex.py
First version of authors
Add author processing script (taken from CWP version)
Add initial (incomplete) list of some authors

@@ -1,0 +1,324 @@
+#! /usr/bin/env python3
+#
+# Process signatory list into a LaTeX author list
+#
+# The data source is expected to be made of 3 text files:
+#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey (i)
+#      with i a footnote number. If there are multiple affiliations, separate AffiliationKeys with &.
+#    - footnotes:txt: the list of footnotes. There must be one footnote per line starting with 'i.'
+#    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
+#      AffiliationKey must not contain any spaces.
+#
+# Note: this script requires Python >= 3.5
+#
+# Initial version written by Graeme A Stewart (graeme.andrew.stewart@cern.ch) and Michel Jouvin (jouvin@lal.in2p3.fr)
+# The script is formatted with black <https://pypi.org/project/black/>
+
+import sys
+import os
+import re
+from operator import attrgetter
+import argparse
+
+# Define some constants related to application exit status
+EXIT_STATUS_SUCCESS = 0
+EXIT_STATUS_OPTION_ERROR = 3
+EXIT_STATUS_FAILURE = 5
+
+# Default input files
+AUTHOR_FILE_DEFAULT = "authors.txt"
+FOOTNOTE_FILE_DEFAULT = "footnotes.txt"
+AFFILIATION_ADDRESS_FILE_DEFAULT = "address.txt"
+LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
+ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
+
+
+class AffiliationFootnote:
+    def __init__(self, text):
+        self.text = text
+        self.mark = None
+
+    def set_mark(self, i, letter=False):
+        if letter:
+            if i > 26:
+                j = int(i / 26)
+                i = i % 26
+                j_str = chr(ord("a") + j - 1)
+            else:
+                j_str = ""
+            self.mark = "{}{}".format(j_str, chr(ord("a") + i - 1))
+        else:
+            self.mark = str(i)
+
+    def __repr__(self):
+        return repr((self.mark, self.text))
+
+
+class Author:
+    def __init__(self, author_line):
+        footnotes_search = re.search(r"\((?P<footnotes>[\d,]+)\)$", author_line)
+        if footnotes_search:
+            self.footnotes = footnotes_search.group("footnotes").split(",")
+            author_line = author_line.replace(footnotes_search.group(), "")
+        else:
[…]

### 2020-08-19 LHCC/2020/2020-01/a2tex.py
Improve author processing script
Remove JHEP option that has no use here
Catch exceptions and print when parsing files and splitting lines
Provide a summary of processing

@@ -117,7 +117,11 @@ def read_kv_file(file, separator=":", referenced_keys=None):
             kv_line = kv_line.strip()
             if kv_line.startswith("#") or kv_line == "":
                 continue
-            k, v = kv_line.split(separator, 1)
+            try:
+                k, v = kv_line.split(separator, 1)
+            except Exception:
+                print(f"In file {file} problem splitting this line:\n{kv_line}")
+                raise
             v = v.strip()
             if k in kv_map:
                 raise Exception("Duplicated key ({}) found in {}".format(k, file))
@@ -149,13 +153,6 @@ def main():
         dest="footnotes",
         default=FOOTNOTE_FILE_DEFAULT,
         help="File containining footnote list (D: {})".format(FOOTNOTE_FILE_DEFAULT),
-    )
-    parser.add_argument(
-        "--jhep",
-        dest="jhep",
-        action="store_true",
-        default=False,
-        help="Produce an author list that can be used in jheppub.sty",
     )
     parser.add_argument(
         "--output",
@@ -196,7 +193,7 @@ def main():
     for footnote_mark, affiliation in enumerate(sorted_affiliations, start=1):
         affiliation_map[affiliation].set_mark(footnote_mark)
 
-    # Parse and assign footnotes: interprete the first field as a footnote key
+    # Parse and assign footnotes: interpret the first field as a footnote key
     # matched against author footnotes rather than the real mark
     footnote_map = read_kv_file(
         options.footnotes, separator=". ", referenced_keys=referenced_foonotes
@@ -251,73 +248,42 @@ def main():
                 )
                 affiliation_list = affiliation_list + "," + footnote_str
 
-            if options.jhep:
-                formatted_name = "{} {}".format(author.forename, author.surname)
-                author_position = author_list.index(author) - len(author_list)
-                # The fiddly bit to get "a, b, c and d" as in JHEP style
-                # N.B. For position -2 doing nothing *is correct*
-                if author_position < -2:
-                    formatted_name += ","
-                elif author_position == -1:
-                    formatted_name = "and {}".format(formatted_name)
-                print(
-                    "\\author[{}]{{{}}}".format(affiliation_list, formatted_name),
-                    file=output,
-                )
+            if author == author_list[-1]:
+                eol_str = ""
             else:
-                if author == author_list[-1]:
-                    eol_str = ""
-                else:
-                    eol_str = ";"
-                print(
-                    "{}, {}$^{{{}}}${}".format(
[…]

### 2020-08-19 LHCC/2020/2020-01/a2tex.py
Lhcc author update (#139)
* First version of authors

Add author processing script (taken from CWP version)
Add initial (incomplete) list of some authors

* Improve author processing script

Remove JHEP option that has no use here
Catch exceptions and print when parsing files and splitting lines
Provide a summary of processing

* Update authors

Add all HSF coordinators and WG convenors

* Add analysis section authors

* Add generator authors

List of contributors was taken from arXiv:2004.13687

* Update Efe's affiliation

* Add authors from direct contact and simulation

@@ -1,0 +1,290 @@
+#! /usr/bin/env python3
+#
+# Process signatory list into a LaTeX author list
+#
+# The data source is expected to be made of 3 text files:
+#    - authors.txt: the list of signers in the following format: Lastname, Firstname - AffiliationKey (i)
+#      with i a footnote number. If there are multiple affiliations, separate AffiliationKeys with &.
+#    - footnotes:txt: the list of footnotes. There must be one footnote per line starting with 'i.'
+#    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
+#      AffiliationKey must not contain any spaces.
+#
+# Note: this script requires Python >= 3.5
+#
+# Initial version written by Graeme A Stewart (graeme.andrew.stewart@cern.ch) and Michel Jouvin (jouvin@lal.in2p3.fr)
+# The script is formatted with black <https://pypi.org/project/black/>
+
+import sys
+import os
+import re
+from operator import attrgetter
+import argparse
+
+# Define some constants related to application exit status
+EXIT_STATUS_SUCCESS = 0
+EXIT_STATUS_OPTION_ERROR = 3
+EXIT_STATUS_FAILURE = 5
+
+# Default input files
+AUTHOR_FILE_DEFAULT = "authors.txt"
+FOOTNOTE_FILE_DEFAULT = "footnotes.txt"
+AFFILIATION_ADDRESS_FILE_DEFAULT = "address.txt"
+LATEX_AUTHOR_FILE_DEFAULT = "authors.tex"
+ARXIV_AUTHOR_FILE_DEFAULT = "arxiv.txt"
+
+
+class AffiliationFootnote:
+    def __init__(self, text):
+        self.text = text
+        self.mark = None
+
+    def set_mark(self, i, letter=False):
+        if letter:
+            if i > 26:
+                j = int(i / 26)
+                i = i % 26
+                j_str = chr(ord("a") + j - 1)
+            else:
+                j_str = ""
+            self.mark = "{}{}".format(j_str, chr(ord("a") + i - 1))
+        else:
+            self.mark = str(i)
+
+    def __repr__(self):
+        return repr((self.mark, self.text))
+
+
+class Author:
+    def __init__(self, author_line):
+        footnotes_search = re.search(r"\((?P<footnotes>[\d,]+)\)$", author_line)
+        if footnotes_search:
+            self.footnotes = footnotes_search.group("footnotes").split(",")
+            author_line = author_line.replace(footnotes_search.group(), "")
+        else:
[…]

### 2020-08-20 LHCC/2020/2020-01/input_generators.tex
Infrastructure and fixes
Add a Makefile that regenerates the document

Add a footnote hyperlink to mail the HSF editorial secretariat
Update LaTeX footnotes to symbols to avoid clashing with authors

Remove some spurious unicode (fi ligature) from generators text

@@
 is due to the way the experiments simulate some processes, where they
-generate large inclusive event samples, which are then ﬁltered on
-ﬁnal-state criteria to decide which events are passed on to detector
+generate large inclusive event samples, which are then filtered on
+final-state criteria to decide which events are passed on to detector
 simulation and reconstruction (e.g. CMS simulations of specific
@@
 filtering has \textasciitilde10\% efficiency for V+jets). This
-inefficiency could be reduced by developing ﬁltering tools within the
+inefficiency could be reduced by developing filtering tools within the
 generators themselves, designed for compatibility with the requirements

### 2020-08-20 LHCC/2020/2020-01/a2tex.py

@@ -9,7 +9,7 @@
 #    - address.txt: the affiliation list with the following format: AffilitationKey: Affiliation text
 #      AffiliationKey must not contain any spaces.
 #
-# Note: this script requires Python >= 3.5
+# Note: this script requires Python >= 3.6
 #
 # Initial version written by Graeme A Stewart (graeme.andrew.stewart@cern.ch) and Michel Jouvin (jouvin@lal.in2p3.fr)
 # The script is formatted with black <https://pypi.org/project/black/>

### 2020-08-25 LHCC/2020/2020-01/input_generators.tex
Generation text and references, sim references
Update some of the text on event sampling and negative weights
for event generation

Add a few references to simulation and generation sections

(Suggested by Ben Nachman)

@@
 reductions in WLCG compute budgets could be achieved. Improvements in
-this area can only start from physics-motivated approaches based on the
-knowledge of phase space peaks, but they can be complemented by
-brute-force machine learning (ML) algorithmic methods~\cite{Ben17,Bot20,Gao20,Kli18}.
+this area start from physics-motivated approaches based on the
+knowledge of phase space peaks and are complemented by
+machine learning (ML) algorithmic methods~\cite{Ben17,Bot20,Gao20,Kli18}.
 
@@
 (see~\cite{Fre20} for a more in-depth discussion). Progress in this area can
-only be achieved by theorists: for instance, a new approach for
+only be achieved with physics knowledge: for instance, a new approach for
 MC@NLO-type matching with reduced negative weights has recently been
-proposed~\cite{Fre20}. It should be noted that negative weights can also
-happen at LO because of not-positive-definite parton density function
-sets and interference terms, e.g. in effective field theory
-calculations.
+proposed~\cite{Fre20} and some recent papers show how to lessen the impact of
+negative weights in a secondary step~\cite{andersen2020positive,
+nachman2020neural}. It should be noted that negative weights can also happen at
+LO because of not-positive-definite parton density function sets and
+interference terms, e.g. in effective field theory calculations.
 

### 2020-08-25 LHCC/2020/2020-01/input_simulation.tex
Generation text and references, sim references
Update some of the text on event sampling and negative weights
for event generation

Add a few references to simulation and generation sections

(Suggested by Ben Nachman)

@@
 Networks on the `images' of particle 
-showers~\cite{ML001-GAN, aishik_ghosh_2019_3599705, GAN-ATLAS}. The energy
+showers~\cite{Paganini_2018, ML001-GAN, aishik_ghosh_2019_3599705, GAN-ATLAS}. The energy
 deposition values in the calorimeter cells are considered as `pixels' of
@@
 All LHC experiments have explored the use of multi-objective regression
-and generative adversarial networks (GANs)~\cite{GAN-ATLAS, GAN-LHCb}, they are now in the process of investigating the use of
+and generative adversarial networks (GANs)~\cite{GAN-ATLAS, GAN-LHCb, aishik_ghosh_2019_3599705}, they are now in the process of investigating the use of
 fast simulations for other types of detectors than calorimeters, e.g.

### 2020-09-24 LHCC/2020/2020-01/ctcs.tex
First version of authors
Add author processing script (taken from CWP version)
Add initial (incomplete) list of some authors
Infrastructure and fixes
Add a Makefile that regenerates the document

Add a footnote hyperlink to mail the HSF editorial secretariat
Update LaTeX footnotes to symbols to avoid clashing with authors

Remove some spurious unicode (fi ligature) from generators text
Update DOI for CTCS Snowmass
Much more complete author list now

@@
  & & HSF-DOC-2020-01 \\
- & & 10.5281/zenodo.3779250 \\
- & & May 1, 2020 \\ % use \date or hardwire e.g. December 15, 2017
+ & & 10.5281/zenodo.4009114 \\
+ & & 31 August, 2020 \\ % use \date or hardwire e.g. December 15, 2017
  & & \\
@@
 
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+
 \title{HL-LHC Computing Review:\\Common Tools and Community Software}
 
-\author{High-Energy Physics Software Foundation}
+\author{HEP Software Foundation\footnote{\href{mailto:hsf-editorial-secretariat@googlegroups.com}{hsf-editorial-secretariat@googlegroups.com}}:
+  \input{authors}
+}
 

### 2022-05-17 HSF-DOC/2022-01/hsf-history-paper.tex
Add HSF History paper
Version as submitted to arXiv

% Copyright (C) 2022, CERN, Princeton and Fermilab, licence CC-BY-4.0.

\documentclass{article}
\usepackage[utf8]{inputenc}

\usepackage{jheppub}

\usepackage{graphicx}
\usepackage[titletoc]{appendix}
\usepackage{xspace}
\usepackage{ulem}

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{refs.bib}

\title{The HEP Software Foundation Community}
\author{The HEP Software Foundation\\
\vspace*{4mm}
Contact editors:\\
\vspace*{2mm}
Graeme A Stewart, CERN (\href{mailto:graeme.andrew.stewart@cern.ch}{graeme.andrew.stewart@cern.ch})\\
Peter Elmer, Princeton University (\href{mailto:Peter.Elmer@cern.ch}{Peter.Elmer@cern.ch})\\
Elizabeth Sexton-Kennedy, Fermilab (\href{mailto:sexton@fnal.gov}{sexton@fnal.gov})
}
\date{April 2022}

\begin{document}

% HSF Document Number (only for arXiv)
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-DOC-2022-01 \\
 & & May 17, 2022 \\ % use \date or hardwire e.g. December 15, 2017
 & & Copyright (C) 2022 CERN, Princeton and Fermilab, licence CC-BY-4.0 \\
 & & \\
\end{tabular*}
\vspace{2.0cm}

\abstract{The HEP Software Foundation was founded in 2014 to tackle common problems of software development and sustainability for high-energy physics. In this paper we outline the motivation for the founding of the organisation and give a brief history of its development. We describe how the organisation functions today and what challenges remain to be faced in the future.}

\maketitle

\hypertarget{history}{%
\section{History}\label{history}}

Over the past 50 years, the experimental particle, nuclear and
astroparticle physics communities have iteratively evolved a significant
amount of community structure. This was a natural result of the growing
size, scale and time duration of the experiments and the centralisation
of facilities at large laboratories. National, and now international,
collaborations are typically required to build, operate and maintain the
large detectors used in these experiments. No single university or
laboratory can provide all of the necessary expertise and required
effort. The largest collaborations have grown from 100s of collaborators
in the 1990s to 1000s at (for example) the Large Hadron Collider (LHC)
at CERN. This community has also developed a broad ecosystem of
methodologies and technologies that are used to build successive
experiments and upgrades to existing experiments. While a specific
instrument can necessarily only be used for a single experiment at any
given time, the large commonalities in methodology should permit the
development of research software which can be used widely in the
community for different projects. Despite this, much of the software
development remained somewhat siloed within individual experiments or,
at most, one or another host laboratory, with only a few exceptions.

This is not to say that in the history of HEP there was no common software. 
CERNLIB~\cite{CERNLIB} was a foundation library written in the Fortran era and used by many experiments.  Elements of it have been rewritten in C++ and constitute some of the most widely used software packages in the field.  Projects such as ROOT, Geant4, and various generators have effectively acted as common glue for many experiments.  However in the software layers above these foundation and toolkit libraries, redundant
solutions that are difficult to evolve and sustain over time
(years or decades for large experiments!) are common. 
To a large extent, software speed, performance and efficiency had been ignored previously, because the costs due to inefficient software could be ignored in the past. Software is as much an intellectual product as well as a tool, thus a new approach was needed.

First steps in the direction of collaborating on modernising, in 2011-2012, led to the formation of a
cross-experiment ``Concurrency Forum''~\cite{CERN-RD-MULTICORE} to discuss
the specific software challenges brought by changes in microprocessor
technology (multi-core, wide vectors, GPUs). Driven initially by CERN
and Fermilab, the forum demonstrated community interest in wider
software collaborations. By 2014-2015, a number of colleagues involved
in HEP software for many years were discussing a more ambitious and
broader scope for research software collaborations in HEP. This
eventually led to the formation of the \textbf{High-Energy Physics (HEP)
Software Foundation (HSF)}. The driving motivations for this initiative
were that the physics upgrades anticipated in the coming decades,
particularly the High-Luminosity LHC~\cite{HL-LHC}, would put enormous
pressure on the software used in HEP; that much of our software was
already decades old; the changes in microprocessor technology brought
new challenges to the table; and that there was an urgent need to train
new talent and to attract investment to the field, which could be better
supported when common, multi-experiment, efforts were promoted. More
generally, additional community structure which promotes research
software collaborations, not tied to single experiments or laboratories,
has greater potential to enhance the longer term sustainability of the
software.

The very first workshop~\cite{CERN-WS} attempted to build on the
community experience within the large experiments, however there was too
much discussion of ``governance'' questions. Individual experiments need
to operate a large well-integrated detector, manage pooled resources
(such as computing and storage) and at the end of the day produce
scientific publications signed by the entire collaboration. Thus
governance questions within experiments are critical. This top-down
approach was less obvious for the envisioned research software
collaborations, which can be more ``ecosystem-like''. It also made
engaging experiments of very different sizes more challenging. By the
end of the workshop most participants had concluded that a different
structure was needed. Subsequent workshops, one in North America and
one in Europe to aid inclusivity~\cite{SLAC-WS,LAL-WS}, brought together
many HEP experiments, HEP specific software projects and other non-HEP
organisations, such as the Apache Software Foundation and the Software
Sustainability Institute. Here, the principle of a ``do-ocracy'' was
enshrined, to encourage activity from the bottom-up, with developers
leading, which was a far more productive approach to community building.

From these early workshops the idea was born of preparing a Community
White Paper (CWP), laying out the roadmap for software and computing in
HEP in the 2020s. As a way to fortify these community-led efforts, the
Worldwide LHC Computing Grid (WLCG) gave a formal charge to the HSF to
prepare this paper~\cite{CWP-Charge}. Many individuals volunteered and
the US National Science Foundation was an early investor with dedicated
funding to help organise and carry out CWP workshops~\cite{SDSC-CWP,LAPP-CWP}. These workshops were focal points of an intense year of
identifying key domain areas and potential solutions for the field and
setting up working groups who would prepare topic-specific white papers.
This approach was able to greatly broaden the involvement of key
individuals and solidified the active communities around the HSF. Once
the working groups had produced their domain-specific white papers, an
editorial team took charge of synthesising a unified single version of
the white paper, which was published with extremely wide community
support, 310 signing authors from 124 institutes~\cite{Albrecht2019}.

The CWP was not the only activity happening in the HSF at this time.
Particularly where there was an identified gap between different
communities, be these inter-experiment, between projects, or even
between theory and experiment, the HSF was ideally placed to bridge gaps
and bring different people together. Workshops on analysis ecosystems~\cite{ANALYSIS-ECO} 
and on computational aspects of event generators~\cite{COMP-GEN} were notable examples. In addition the HSF was a natural
forum for considering more radical software developments and their
potential, gathering experiment and expert feedback on progress and
possibilities~\cite{GEANTV-RD}.

\hypertarget{hsf-activities}{%
\section{HSF Activities}\label{hsf-activities}}

While much of the early HSF activity was focused on software community
building through the CWP, working groups were started where common
topics were obviously identified within the community, e.g., in the
domain of packaging and distributing HEP software stacks. These groups
brought together experts and interested parties and focused around
technical discussions.

In the wake of the CWP it was clear that this model would work very well
for the areas of most concern for the future, so the model was broadened
and, over a few years, eight working groups were established with about
3 conveners in each case, appointed annually and with a nomination
system that allows both stakeholder (experiment, institution) input and
bottom-up volunteers for running these groups. In order to marshall
these activities it was necessary to have some critical amount of
binding effort, so the decision of CERN management to allow significant
(0.5 FTE) time from one person to the HSF was crucial and has had a key
multiplication effect.

Where these groups are involved in areas that are `traditional', e.g.,
detector simulation or reconstruction, there is a strong involvement
with ongoing developments in the experiments and in well established
software projects. The focus is the exchange of ideas and discussions of
common problems. In a number of cases the HSF had identified topics of
interest that were simply not covered elsewhere in the field and then
the HSF working group has had a further leading and catalysing effect.
This is particularly the case for the use of Python in HEP, led by the
PyHEP working group; and for computational aspects of physics event
generators, led by the Generators working group~\cite{Valassi2021}.

In addition to being a focus for the exchange of ideas and techniques,
when WGs identify a concrete topic where a paper can usefully be
prepared, the HSF is a natural place for organising pan-experiment input
and encouraging some standardisation (e.g., in analysis level metadata
or in detector conditions data access). This has been recognised by more
formal bodies outside the HSF, such as WLCG and the LHCC, who often ask
the HSF to marshall community inputs for updates on development
activities and reviews of development plans. This is an important point
in terms of recognising the contribution of the organisation. In turn,
this fosters recognition for the contribution of our individual members
and helps their careers to advance.

This engagement with strategic bodies in the field, where the HSF
advocates for software investment~\cite{stewart_graeme_a_2018_2413005}, leads the HSF to
work in tandem with other funded software R\&D projects in HEP, where
the HSF will support funding applications and then work with these
projects, enhancing their connection to the community and the impact of
their work. Practically projects can contribute through the working
groups closest to their R\&D areas.

The HSF also engages with other like minded bodies and collaborates on
regular series of meetings regarding accelerator programming or software
and computing for nuclear physics and also organises itself as an
umbrella organisation, with CERN, to run Google Summer of Code for HEP
software projects and experiments.

In the past the HSF organised face-to-face workshops and the intention
is to restart such activities once the pandemic passes, but in the
meantime virtual workshops have played a role. In some circumstances
these can even have a greater impact, with the PyHEP workshops in 2020
and 2021 registering more than 1000 participants~\cite{PYHEP20,PYHEP21}.
In large part this reflects a strong didactic element in the Python
area. This thread is reflected also in that the HSF, together with
IRIS-HEP, SIDIS, The Carpentries and the ROOT project~\cite{IRIS-HEP,SIDIS,CARPENTRIES,Brun1996,ROOT}, has put a strong emphasis on training
activities~\cite{Malik2021} and now runs regular training events in
fundamental software skills and in C++ programming. This is seen as a
critical activity in the field and attempts are now also being made to
have these activities as feeders to encourage trainees to be involved in
software development and training.

\hypertarget{outcomes-and-conclusions}{%
\section{Outcomes and
Conclusions}\label{outcomes-and-conclusions}}

Contrary to initial ideas, the HSF has not, by and large, run software
projects themselves - without actually having resources to disburse it
is better to allow such projects to be independent and work with the HSF
as is useful. That said, HSF events have proved to be fertile ground for
people from different backgrounds to meet (e.g., nuclear and
astroparticle physics) and even to start common software projects that
then take on a life of their own. Fostering funded projects has led to
new investment in software R\&D in HEP and a higher recognition of the
importance of promoting excellent software developers in their careers;
this is a major success that was a direct outcome of the CWP process.

The HSF has now established itself as a recognised part of the HEP
software landscape where it links strategic bodies to the community of
software developers. It remains a challenge to continue to build the
next generation of HEP software developers and make them feel involved
and part of an organisation like the HSF, but work to improve training
and engage younger colleagues through the working group process is hoped
to improve this. Looking forward to post-pandemic activities, where
face-to-face interactions can happen again, will also help to continue
to build HEP software communities.

%\hypertarget{references}{%
%\paragraph{References}\label{references}}

\sloppy
\raggedright
\clearpage
\printbibliography[title={References},heading=bibintoc]

\end{document}

### 2022-12-16 HSF-DOC/2022-02/AEII.tex
Upload Analysis Ecosystem II Report

\documentclass[a4paper,11pt]{article}
\usepackage[utf8]{inputenc}

% Highlight where bad unicode is in the references...
\DeclareUnicodeCharacter{0301}{*************************************}

\usepackage{jheppub}
\usepackage{hyperref}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[titletoc]{appendix}
\setcounter{tocdepth}{2} % Only to subsection level in ToC
\usepackage{xspace}

% Stop lists from using so much space
\usepackage{enumitem}
\setlist{noitemsep}

\usepackage[T1]{fontenc} % if needed

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{AEII.bib}

\setcounter{biburlnumpenalty}{9000}
\setcounter{biburlucpenalty}{9000}
\setcounter{biburllcpenalty}{9000}

\begin{document}

% HSF Document Number (only for arXiv)
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-DOC-2022-02 \\
 & & \href{https://doi.org/10.5281/zenodo.7418818}{10.5281/zenodo.7418818} \\
 & & Licence:
\href{https://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0} \\
& & 
\end{tabular*}
\vspace{2.0cm}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\title{\boldmath Second Analysis Ecosystem Workshop Report\\
23-25 May 2022, IJCLab Orsay, Paris}

\date{August 2022}

\author{The HEP Software Foundation and IRIS-HEP: \textrm{Mohamed Aly [3], Jackson Burzynski [12], Bryan Cardwell [17], Daniel C. Craik [9], Tal van Daalen [4], Tomas Dado [28], Ayanabha Das [23], Antonio Delgado Peris [10], Caterina Doglioni [3], Peter Elmer [2,b], Engin Eren [27], Martin B. Eriksen [15,16], Jonas Eschle [9], Giulio Eulisse [1,b], Conor Fitzpatrick [3], José Flix Molina [10,16], Alessandra Forti [3], Ben Galewsky [21], Sean Gasiorowski [39], Aman Goel [13], Loukas Gouskos [1,b], Enrico Guiraud [1], Kanhaiya Gupta [41], Stephan Hageboeck [1,b], Allison Reinsvold Hall [32,b], Lukas Heinrich [24,b], Alexander Held [31,b], José M. Hernández [10], Michel Hernández Villanueva [38], Julius Hrivnac [5], Michel Jouvin [5,b], Teng Jian Khoo [33,b], Luke Kreczko [14], Nils Krumnack [22], Thomas Kuhr [7], Baidyanath Kundu [2], Eric Lancon [8], Johannes Lange [18], Paul Laycock [8,b], Kilian Lieret [7], Nicholas J. Manganelli [29], Pere Mato Villa [1,b], Andrzej Novak [20], Antonio Perez-Calero Yzquierdo [10,16], Jim Pivarski [2,b], Mason Proffitt [4], Jonas Rembser [1,b], Eduardo Rodrigues [35,b], Grigori Rybkin [5], Jana Schaarschmidt [4,b], Henry F. Schreiner [2], Markus Schulz [1], Andrea Sciabà [1], Sezen Sekmen [36], Elizabeth Sexton-Kennedy [25,b], Oksana Shadura [34,b], Tibor Simko [1], Nathan Simpson [30,b], Jaydip Singh [26], Nicola Skidmore [3,b], Nicholas Smith [40], Michael Sokoloff [44,b], Graeme A. Stewart [1,a,b], Giles C. Strong [42], Gokhan Unel [37], Vassil Vassilev [2], Mark Waterlaat [43], Gordon Watts [4,b], Efe Yazgan [19]}
 \vspace{5mm}
 \textrm{a\,Lead Editor, b\,Editor}
 \vspace{5mm}
 \textrm{1\,CERN, 2\,Princeton University, 3\,University of Manchester, 4\,University of Washington, 5\,IJCLab, 10\,CIEMAT, 7\,LMU München, 8\,Brookhaven National Laboratory, 9\,University of Zurich, 16\,PIC, 12\,Simon Fraser University, 13\,University of Delhi, 14\,University of Bristol, 15\,IFAE, 17\,University of Virginia, 18\,Universität Hamburg, 19\,National Taiwan University, 20\,RWTH Aachen, 21\,University of Illinois, Urbana-Champaign, 22\,Iowa State University, 23\,Czech Technical University in Prague, 24\,Technische Universität München, 25\,Femilab, 26\,University of Lucknow, 27\,DESY, 28\,TU Dortmund, 29\,University of California, Riverside, 30\,Lund University, 31\,University of Wisconsin–Madison, 32\,US Naval Academy, 33\,Humbold-Universität zu Berlin, 34\,University Nebraska-Lincoln, 35\,University of Liverpool, 36\,Kyungpook National University, 37\,UC Irvine, 38\,Deutsches Elektronen–Synchrotron, 39\,SLAC National Accelerator Laboratory, 40\,FNAL, 41\,Universität Bonn, 42\,INFN, Sezione di Padova, 43\,University of Maastricht, 44\,University of Cincinnati}
}

\abstract{The second workshop on the HEP Analysis Ecosystem took place 23-25 May 2022 at IJCLab in Orsay, to look at progress and continuing challenges in scaling up HEP analysis to meet the needs of HL-LHC and DUNE, as well as the very pressing needs of LHC Run 3 analysis.

The workshop was themed around six particular topics, which were felt to capture key questions, opportunities and challenges. Each topic arranged a plenary session introduction, often with speakers summarising the state-of-the art and the next steps for analysis. This was then followed by parallel sessions, which were much more discussion focused, and where attendees could grapple with the challenges and propose solutions that could be tried. Where there was significant overlap between topics, a joint discussion between them was arranged.

In the weeks following the workshop the session conveners wrote this document, which is a summary of the main discussions, the key points raised and the conclusions and outcomes. The document was circulated amongst the participants for comments before being finalised here.}

\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The second workshop on the HEP Analysis Ecosystem~\cite{AEIIWS}
took place 23-25 May 2022 at IJCLab in Orsay, to look at progress and
continuing challenges in scaling up HEP analysis to meet the needs of
HL-LHC and DUNE, as well as the very pressing needs of LHC Run 3
analysis. Since the original
HSF Analysis Ecosystem workshop~\cite{AEIWS}, held in Amsterdam five years ago, 
the ecosystem of
software used for analysis in High Energy Physics (HEP) has evolved
considerably, which was reflected in the preparation and organisation of
the workshop.

To set the scene, looking back to the report~\cite{elmer_peter_2017_6599290} 
from that first
workshop, the central ideas put forward have been vindicated and many of
the inchoate ideas proposed have become a concrete reality. In
particular, the increasing use of Python drove, and was supported by, a
new PyROOT interface~\cite{Galli:2020boj}. Python access to the most powerful machine
learning interfaces from data science is now well supported, with
considerable development of data bridges, e.g., in ROOT~\cite{An:2021} and the widely used
Uproot package~\cite{Pivarski_Uproot_2017}. As these
external pieces have become vital to modern HEP, a suite of packages
that focus on HEP needs to access and use Python data science tools has
grown up, with the Scikit-HEP~\cite{scikit-hep,Rodrigues:2020syo}
project providing a focus for this part of the community. Using ROOT has
been made a lot easier, with a mature Conda installation option and
modularity being planned for new components. Declarative interfaces have
become capable and are being used for analysis at scale, with RDataFrame~\cite{rdataframe}
and Coffea~\cite{Smith:2020pxs,lindsey_gray_2022_7158492} being
the most notable. These interfaces are used in the development of
prototype Analysis Facilities, with backend independence, notebook
interfaces, and many attractive properties in, e.g.,
SWAN~\cite{swan, PIPARO20181071} and Coffea-Casa~\cite{refId0-coffea-casa} and the
development of Distributed RDataFrame~\cite{distrdf} (real metrics for success are
still, however, a work in progress).

The use of continuous integration, encapsulation, code analysis and
workflow management tools has improved in the community, which aids both
scale-out and data preservation.

Although users interact increasingly with Python, as expected, C++
remains wide\-spread and necessary, with no other language really
challenging it in HEP. The performance of I/O in ROOT
\cite{BRUN199781,rene_brun_2019_3895860} has been improved
hugely with RNTuple~\cite{ntuple20}, which beats all other available formats. These
critical low-level improvements link back to Analysis Facility
development and continuing R\&D on different storage interfaces, e.g.,
object stores.

In a few places technology predictions from five years ago did not come
to pass, such as memory resident analyses based on persistent memory
technology or super-high core count CPUs (e.g., Xeon Phi failing to
compete with GPUs). In other areas, such as metadata handling, there was
little general progress although the problem is still seen as relevant
and common between experiments.

Even with these considerable advances in tooling, there remain many open
questions and pressing needs that merited a second workshop. With the
improving pandemic situation, the workshop was organised as a hybrid
event, with 125 people in total registering and more than 70 attending
in-person.

The workshop was themed around six particular topics, which were felt to
capture key questions, opportunities and challenges. Each topic arranged
a plenary session introduction, often with speakers summarising the
state-of-the art and the next steps for analysis. This was then followed
by parallel sessions, which were much more discussion focused, and where
attendees could grapple with the challenges and propose solutions that
could be tried. Where there was significant overlap between topics, a
joint discussion between them was arranged.
\href{https://docs.google.com/document/d/17gPx0eKODpYfJwwler4p-9mR8KllISbDy_LlgOA9tfk/edit?usp=sharing}{{Live
notes}} were taken in all of the sessions. An additional session on
Analysis Workflow Design was held as a breakout. On the final day the
topic conveners gave a brief summary of the key points. This was then
followed by writing sessions, where the outcomes of the workshop started
to crystalise. The final result of that process and deliberation is
given in this document.

The success of this workshop can be attributed to the engagement of
experts from different experiments and projects, and their willingness
to discuss openly. The opportunities afforded by in-person participation
were significant, with many intense and interesting discussions between
participants following on from workshop sessions. The organisers would
like to warmly thank colleagues at IJCLab, particularly Michel Jouvin,
for supporting the workshop at the lab and for arranging a memorable
dinner. Sponsorship for the workshop was provided by IJCLab, CERN-HSF,
IRIS-HEP and Nvidia, to whom we are very grateful.

\hypertarget{topics}{%
\section{Topics}\label{topics}}

\hypertarget{analysis-user-experience-and-declarative-languages}{%
\subsection{Analysis User Experience and Declarative
Languages}\label{analysis-user-experience-and-declarative-languages}}

\begin{itemize}
    \item \emph{Convenors: Jonas Rembser, Alexander Held}
    \item \emph{Speakers: Axel Nauman, Jim Pivarski, Sezen Sekmen}
\end{itemize}

This session focused on the user experience of an analyser performing
the final steps in an analysis pipeline. Analysis typically starts from
centrally produced datasets, which are further reduced by filtering
events or event content. New columns are calculated and added, including
information needed for evaluating systematic uncertainties and
observables required for the analysis. analysers then commonly perform
statistical analysis by building statistical models from template
histograms or directly using data in unbinned models. This workflow was
also discussed in the
``Future
Analysis Workflow Design'' session (see section~\ref{breakout-session---analysis-workflow-design}). The user experience for
statistical analysis was not prominently discussed at the workshop, but
it is connected to the ease of systematic uncertainty bookkeeping. The
second main topic of this track was declarative or domain-specific
languages (DSLs), allowing for the decoupling of physics information
from execution details.

Workshop participants ranked the following three aspects as the
\textbf{top three difficulties} in a typical analysis workflow:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  \textbf{Systematic uncertainties:} Dealing with systematic
  uncertainties was a recurring topic throughout this workshop and not
  just confined to the dedicated track. There is no uniform way to
  handle systematic uncertainties, and analysers must find solutions
  that work for their specific use case. Difficulties arise from the
  various types of systematic uncertainties (those that alter object
  kinematics are more demanding than weight-based uncertainties) and the
  different approaches to evaluating them. The evaluation may rely on
  additional columns and metadata and may require tools that are either
  centrally provided or implemented by the analyser. See also
  section~\ref{systematics} in this report, which is focused
  on systematics.
  \end{quote}
\item
  \begin{quote}
  \textbf{Metadata:} Challenges related to metadata include finding and
  handling the relevant information, including scale factors or other
  calibration data. Another aspect is bookkeeping: tracking and
  organising all datasets required for the analysis. See also section~\ref{systematics-and-metadata} on metadata.
  \end{quote}
\item
  \begin{quote}
  \textbf{Scale-out:} The move from prototyping to running at scale on
  various sites is referred to as \emph{scale-out}. With varying
  scale-out mechanisms and different environments at each site,
  analysers cannot easily port their analysis from their local machine
  to something running at scale at various facilities.
  \end{quote}
\end{enumerate}

Another aspect deemed crucial to the user experience is
\textbf{interoperability}, particularly for interfaces between analysis
stages. Interoperability between ROOT~\cite{rene_brun_2019_3895860}
and Scikit-HEP~\cite{scikit-hep} tools is
mandatory, as well as the interoperability of key objects like
histograms (e.g., Boost::Histograms in cppyy vs.~pybind11,
boost-histogram vs. ROOT histograms, TH* vs.~ROOT7 RHist). Other
examples are Python bindings, serialised statistical models (e.g., RooFit
workspaces including their JSON version and the
pyhf~\cite{pyhf,pyhf_joss} JSON format for
HistFactory workspaces), and data interoperability at the column-level
(also in memory). Not everything needs to be interoperable, but for
novel ideas to flourish, users should not be locked in one particular
toolset after writing the first code.

A big discussion topic was \textbf{onboarding new analysers} and
first-time user experience. There was broad agreement that documentation
generally has to be improved, but the situation was better than five
years ago at the Analysis Ecosystem
Workshop I~\cite{elmer_peter_2017_6599290}. Several success stories were identified, including the
regularly-updated LHCb Starterkit lessons~\cite{lhcb-starterkit}.
Discussion forums, as well as Slack and Mattermost
channels, are also important for new users. Having a variety of channels
for problem discussion can help engage users with different needs.
Instant messaging platforms allow for quick and informal iterations.
Solutions to problems posted in large channels in instant messaging
platforms can however be difficult to discover for analysers compared to
forum posts or discussions on GitHub. It was also noted that tutorial
writing is challenging: what should be addressed in a tutorial? Feedback
from users can be essential to inform developers what to focus on when
providing documentation material. Dedicated hackathons bringing together
users and developers may be an efficient way of exchanging ideas and
creating new material.

Partially related was the discussion on \textbf{educating analysers in
programming \& software engineering techniques} and \textbf{simplifying
analysis tools}, which often appears contradictory. There was no general
consensus on which direction is more important, as both help to make
analysers more efficient. It was noted that interest in programming
training courses, such as the HSF C++ course, is very high. The
importance of having physicists with advanced knowledge of C++ in
particular was also highlighted, as the personpower for maintaining the
large experiment-specific frameworks is limited. Library developers
noted that users frequently run into relatively basic programming issues
where more formal programming experience would help. A big consideration
regarding the complexity of writing an analysis is how high- or
low-level the interfaces are that analysers are expected to use.
High-level interfaces --- including DSLs --- help users to focus on the
physics and give developers more flexibility for optimising the backends
without changing user interfaces. However, access to lower levels, via
hooks or similar, was seen as crucial to retain the possibility for
users to implement more complex analyses.

The discussion also touched on \textbf{performance}: how can a user
learn that their analysis is not suffering from an easily-avoidable
performance bottleneck? Informal conversations with other analysers were
pointed out as one solution to get a rough feeling for the expected
turnaround time of an analysis. Beyond that, experiments might dedicate
meetings to review techniques and help collect best practices. The
importance of profiling tools was also pointed out. They can not only
help the user with identifying bottlenecks, but their output may also
help experts provide targeted help. In this context, it is desirable for
tools and libraries to expose relevant metrics and performance counters
where possible. Simple tooling to study whether an analysis is
I/O-limited could be helpful. Another critical aspect related to
performance is the analysis implementation time. Time spent writing a
very efficient implementation may surpass all the time saved from the
efficiency.

Declarative or \textbf{domain-specific languages} are an emerging
approach to decoupling physics information from execution details.
General advantages of DSLs include their self-documenting nature and the
decoupling from the backend implementation, allowing updates to the
backend as new technologies become available. One strength of DSLs that
was highlighted is the readability of the analysis description and the
possibility to compare different analyses implemented in the same way.
This can simplify navigating the analysis landscape, simplifying tasks
such as finding overlap in analysis event selections. Several efforts to
develop DSLs for HEP analysis are ongoing, covering both embedded and
external domain specific languages. An embedded DSL is built on top of a
host language like Python, while an external DSL uses its own
interpreter or compiler. Embedded DSLs are more prevalent in HEP,
including FuncADL~\cite{Proffitt:2021wfh} and ROOT's
RDataFrame~\cite{rdataframe} (including abstractions built on top of it). The
Analysis Description
Language Project (ADL)~\cite{adl-project,Sekmen:2020vph} is an example of an external DSL, with
CutLang~\cite{Sekmen:2018ehb} available as an
interpreter. A disadvantage of embedded DSLs is that users might be
tempted to mangle the physics logic with execution details, negating one
main selling point of DSLs. Hence, it is important to continue research
and development on external DSLs in parallel. Furthermore, the existing
embedded DSLs often originate from one specific experiment. Care should
be taken to separate the development of DSLs and tools to abstract away
experiment specific execution details and helper routines to not stall
the development on either side. Finally, as analysis preservation is of
crucial importance to make the most of any physics analysis, it should
be integral to the design of any analysis description and workflow as a
whole. Demanding that users convert their analysis to a reproducible one
after-the-fact is not sustainable.

A \textbf{wishlist for the future} emerged when reviewing discussions at
the workshop, composed of items that we perceive as most important to
address. Related to the handling of systematics, this list includes
automatic optimisation of the analysis' computation graph to only
re-calculate quantities when needed. Some tools, like RDataFrame, are
already designed around this concept, and other tools are encouraged to
focus their R\&D in this direction to ensure sustainable analysis
efficiency. Analysis interfaces should support object facades, meaning
that columns can be grouped into object views for easier reasoning at
the level of physics objects. This can be done by a user-specified
schema or automatic aggregation based on column names. Frameworks such
as Coffea~\cite{lindsey_gray_2022_7158492} and bamboo~\cite{David:2021ohq} are
implementing this already. We recommend the development of standardised
ways to build such object facades to harmonise the access to event
content in different tools. Furthermore, small frameworks or libraries
should help with common chores related to systematics and metadata
handling where possible. The wishlist also includes more documentation
and learning material. Related to this is the demand for more available
Open Data and analyses that use this Open Data to showcase and benchmark
different analysis approaches. Tooling to help analysers with debugging,
identifying performance bottlenecks, and optimising their analysis
pipelines is another point on the list.

\hypertarget{analysis-on-reduced-formats-or-specialist-inputs}{%
\subsection{Analysis on reduced formats or specialist
inputs}\label{analysis-on-reduced-formats-or-specialist-inputs}}

\begin{itemize}
    \item \emph{Convenors: Allie Hall, Jana Schaarschmidt, Loukas Gouskos}
    \item \emph{Speakers: James Catmore, Lindsey Gray, Michel Hernandez Villanueva, Jackson Burzynski, Bryan Cardwell, Lukas Alexander Heinrich, Nick Smith}
\end{itemize}

This session included a review of the design ideas and the status of
reduced formats used in various experiments for previous runs, and
concepts for future runs. It also contained a critical discussion of
cases where specialist, non-standard inputs are needed and how reduced
formats can be adopted to fit these cases. A cross-over discussion was
held to explore questions related to storing and evaluating systematic
uncertainties using reduced formats.

During the MC production chain, or similarly the data processing chain,
several reduction tiers are used, to filter out low level information
that is not needed on the analysis level in a standard workflow.

In ATLAS and CMS, the AOD format is the output from the reconstruction,
holding detailed object information, with a size of about 500 kB/event.
In CMS, this is reduced to MiniAOD~\cite{Petrucciani_2015}, through
slimming and thinning the collections, e.g., removing some tracks and
applying preselection cuts, resulting in a size of 30 kB/event. The next
reduction stage is called NanoAOD~\cite{Peruzzi_2020},
which are flat ROOT ntuples, containing only selected and processed high
level objects stored with reduced precision for floats, which gives 1-2
kB/event. Systematic uncertainties are recomputed on the fly and are
therefore not stored. MiniAOD and NanoAOD serve 85\% of all analyses,
and were in use already during LHC Run-2. The size of NanoAOD is
strictly monitored, and official changes to the format need to undergo
approval. Further information on the CMS computing model is available in
the Phase-2 Computing Model document~\cite{Software:2815292}.

In ATLAS, during run-2, about 100 different reduced formats were used,
tailored for specific physics analyses or combined performance studies.
These formats contained slimmed object collections and entire events
were removed that did not pass some criteria (skimming). This results in
a varying file size typically, but not limited to, 30-50 kB/event, and
event skimming fractions ranging from well below 1\% to above 10\%.
Although these formats were effective for analysis, the huge event-wise
overlap between them when made from simulated events (which tend to have
higher skimming fractions), called for revisions to the model.
Consequently, for Run-3, ATLAS introduces a common format DAOD\_PHYS,
which is not skimmed, and which can serve 80\% of all physics analyses,
with a size of around 30-50 kB/event. A smaller format called
DAOD\_PHYSLITE, primarily aimed at Run-4, has a size of 10-15 kB/event.
The main physics objects in PHYSLITE are calibrated as the format is
made, removing the need to store inputs for calibrations. Currently,
inputs for evaluating systematic uncertainties still need to be stored.
These and other concepts for ATLAS computing are outlined in the HL-LHC CDR~\cite{Calafiura:2729668}.

The reduction chain in Belle II~\cite{belle2computing} is also two-staged. Starting from the RAW format (70 kB/event),
which is the detector output, then comes reconstruction, which is first
reduced to the mDST (mini Data Summary Table) format with 15 kB/event
containing a subset of objects, which is then reduced further to uDST
(user DST) with 20kB/event, which is a skimmed version of mDST but also
augmented with analysis objects. There are about 80 skims in use,
tailored for specific needs. The production of these skims is a
bottleneck in the processing chain in Belle II due to the high I/O load.
Not all analyses can use these skims, but it is important that such
analyses are supported as well, also in light of reproducibility and
long-term data preservation.

Limitations of these concepts were discussed as well. In CMS, about half
of the NanoAODs are customised, which means skimmed or the content
extended (or both), to fit specific analysis needs. There is significant
overlap between these customised formats. A possible solution would be
friend trees, including only the additional information needed for each
analysis. Another possible solution could be the so-called LegoAOD, that
uses central services like ServiceX~\cite{galewsky2020servicex}, Crab~\cite{SPIGA2008267}, 
Dask~\cite{dask} etc., to allow users to
easily add extra columns without having to copy the rest. Another
innovative approach is the use of object stores, which can avoid the
need to copy columns across processing tiers (for example some metadata
is duplicated at every stage of the reduction chain).

It was noticed that the PHYSLITE format from ATLAS is about 5-10 times
larger than CMS NanoAOD. PHYSLITE is currently at a prototype stage,
while NanoAOD are already used in physics analysis. A critical review of
the content of PHYSLITE will lead to further reduction. However, one of
the main culprits is the storage of inputs for systematic uncertainties.
Future R\&D work will therefore explore possibilities to reduce this
information, using external look-up tables, parameterisations or ML
models. Not all analyses require the same level of precision, calling
for a flexible approach for evaluating systematic uncertainties.

While the large majority of analysis can use reduced formats, it is of
crucial importance to look at the remaining cases that need special
inputs. In particular searches for BSM physics and exotic signatures
require non-standard objects, such as for example displaced muons,
disappearing tracks or unique shower shapes, or they rely on low-level
information such as energy stored in each calorimeter layer or even
individual cells. ATLAS also maintains a large list of residual formats
for combined performance work that cannot use PHYS, for example
calibration studies that need jet constituents.

One technical solution was presented, that is to add friend trees that
store additional variables, but only for a subset of events. A case
study for the ATLAS search for displaced jets in the calorimeter was
performed. Simply adding the required topocluster collection to PHYS
would increase its size by about 140\%, while adding a friend tree
holding this collection only for events that pass the trigger increases
the size by just 2\%, which is a very encouraging result.

The focus for future work on reduced formats must be on special inputs
for non-standard workflows, since these will ultimately drive the
storage needs of the experiments.

\hypertarget{machine-learning-tools-differentiable-workflows}{%
\subsection{Machine Learning Tools \& Differentiable
Workflows}\label{machine-learning-tools-differentiable-workflows}}

\begin{itemize}
    \item \emph{Convenors: Nathan Simpson, Lukas Heinrich}
    \item \emph{Speakers: Sean Gasiorowski, Vassil Vasilev, Giles Strong, Engin Eren}
\end{itemize}

Since the first Analysis Ecosystem Workshop in 2017~\cite{AEIWS}, the role of Machine
Learning (ML) in High Energy Physics (HEP) analysis has grown
significantly. Back then, our in-house solution TMVA~\cite{TMVA} was the tool under
discussion at the time, but now usage has shifted almost entirely to
industry-developed tools and automatic differentiation frameworks, which
span methods ranging from boosted decision trees (BDTs), XGBoost~\cite{XGBoost}, and
various neural network architectures.

In terms of applications, the majority of ML approaches are concerned
with \textbf{designing observables} to be used in a standard HEP
inference scheme, e.g., template-based statistical inference concerning
a physics parameter of interest. There is not yet widespread effort to
share these trained (e.g., via self-supervision) representations of
event data to be adapted for downstream tasks. This is in contrast to
fields like computer vision or natural language processing, which have
many ways to distribute pre-trained models, e.g. through
PyTorch Hub~\cite{PyTorchHub},
TensorFlow Hub~\cite{TFHub},
Hugging Face~\cite{huggingface}, and others. This
point warrants further discussion to understand potential utility.

While supervised learning is the dominant methodology,
\textbf{unsupervised learning methods} such as variational autoencoders
are gaining interest for applications such as anomaly detection. More
recently, we have also seen progress in techniques for
\textbf{likelihood-free inference}, such as likelihood-ratio estimation,
which skip the likelihood modelling step entirely in favour of directly
estimating the implicit likelihood of the underlying physics simulation.

From a community development perspective, it is important to ensure an
efficient pathway from new R\&D developments within the HEP ML community
to use within production settings within experiments. To this end, the
availability of \textbf{realistic} and \textbf{openly available datasets
and data simulators} (e.g. fast detector simulation for given
experiments) are important, and ML R\&D work should be encouraged to
assess performance in such more realistic environments.

A difficult tension arises in the context of \textbf{reduced formats}:
ML analysis pipelines may aim to incorporate increasingly lower-level
features (e.g., track or calorimeter data) in order to rely less on prior
fixed representations provided by reconstruction algorithms and optimise
task-specific performance. However, the pressure arising from storage
constraints encourages only the most high-level variables to be
retained. Here, the ideas discussed in the workshop on augmenting
reduced formats with analysis-specific columns may become important.

It was noted that to a large extent the development of ML components
within a physics analysis is \emph{regarded as a separate activity} from
the main analysis development, with its own data preprocessing pipeline,
ML training and evaluation frameworks. As most leading ML frameworks
(JAX~\cite{jax2018github}, TensorFlow~\cite{tensorflow2015-whitepaper}, PyTorch~\cite{NEURIPS2019_9015}, Scikit-Learn~\cite{scikit-learn}) focus on a Python user
interface, the ongoing efforts within the HEP community for
\textbf{python-based analysis workflows} are deemed essential to
overcome this pattern. Additionally, ROOT is maintaining integrations of
major external frameworks into TMVA and on facilities for efficient data
loading from RDataFrame. For integrating finalised ML code into larger
workflows, inference tools, such as ONNXRuntime, lwtnn or SOFIE are
important. As the developments on future \textbf{Analysis Facilities}
take shape, it is important that ML-focused workflows become
well-supported on such resources (experiment tracking, hyperparameter
optimisation, metrics analysis) and some integrations with external
services (e.g. KubeFlow, MLFlow) may be worth investigating.
Additionally, the methods developed for continuous tracking and
evaluation of ML training runs could also be transferred to a generic
continuous evaluation of HEP analyses, and is worth investigating to
test its utility.

During the workshop, there was considerable interest in the emerging
concept of
``differentiable
programming''~\cite{Baydin2020DifferentiablePI} (DP) as a generalisation of the gradient-aware
computation in the ML context. In this scheme, arbitrary numerical
programs can produce not only outputs but also efficiently and precisely
compute gradient information with respect to either program inputs or
program parameters. Gradient information can \textbf{improve ML
applications} by combining malleable ML components with more
domain-specific computation to induce stronger \textbf{inductive bias}
into the model. Similarly, loss functions can be improved by providing
differentiable physics-driven evaluations of the model output as e.g. in
INFERNO~\cite{de_Castro_2019} or
neos~\cite{NEOS}. Moreover,
gradient information of the target function can be used within as
powerful labels as demonstrated in e.g. score-based training approaches
like madminer~\cite{https://doi.org/10.48550/arxiv.1907.10621} and
madjax~\cite{madjax}. Interest in
DP also stems from use cases beyond ML, where e.g. gradient information
may be important for sensitivity analysis with respect to systematic
variations, or improve statistical inference as used for example in
fitting frameworks pyhf~\cite{pyhf,pyhf_joss} and zfit~\cite{Eschle:2019jmu}.
There is also
movement in the direction of end-to-end detector optimisation, including
a recent whitepaper on
the topic~\cite{https://doi.org/10.48550/arxiv.2203.13818}. Communities that have formed from this effort (with a
degree of cross-pollination) include
gradHEP~\cite{gradhep} and the
MODE
collaboration~\cite{MODE:2021yid}.

The current practice of analyses based on a fixed reconstruction scheme
is comparable to standard \textbf{transfer-learning} approaches in ML,
where fixed pre-trained layers are used as inputs to training a
task-specific ``tail'' (i.e.. each analysis tuning selections on fixed
outputs of reconstruction, which was optimised on its own set of
objectives). The pre-trained layers may then be fine-tuned towards the
specific task, using small learning rates, by exploiting the
differentiable nature of the model to re-optimize lower-level features
for a given task (i.e. analyses being able to adjust e.g. low-level
reconstruction parameters). This practice has recently become
commonplace in industry. Developing an \textbf{end-to-end differentiable
HEP pipeline} could enable a similar workflow, however the achievable
improvement is unknown and should be studied in the coming years.

A major hurdle towards DP is the fact that much of the existing code to
analyse HEP data is either not written in differentiable programming
languages or the computations may fundamentally not be differentiable
(e.g. hard categorical decisions, array sorting, ..). For the former,
ongoing developments such as
CLAD~\cite{Vassilev_Clad} or
Enzyme~\cite{10.1145/3458817.3476165} aim to enable AD for
general purpose language and ROOT has reported the intention to leverage
these to integrate differentiation more deeply into e.g. RooFit. On the
latter point, effort has been put into consolidating a set of
differentiable surrogates for common operations in HEP (e.g. histograms)
into a package called
\emph{relaxed}~\cite{relaxed}. These
``relaxed'' operations are smooth analogues to the equivalent hard
operation (e.g. max vs softmax) with a tunable level of approximation.
The relaxed operations can be used during optimisation to find settings
for the pipeline with no approximations, and can also be used for the
final task if desired (possibly incurring a bias, if e.g. using a soft
histogram and assuming Poisson statistics per-bin).

We assessed that more effort is needed to study the real-world benefits
of DP in different applications over other optimisation methods (e.g.
Bayesian optimisation), as well as ensuring that the community at-large
converges on common tooling and methodological efforts for the future.

\hypertarget{systematics-and-metadata}{%
\subsection{Systematics and Metadata}\label{systematics-and-metadata}}

\begin{itemize}
    \item \emph{Conveners: Paul Laycock, Teng Jian Khoo}
    \item \emph{Speakers: Thomas Kuhr, Stephan Hageboeck}
\end{itemize}

\hypertarget{metadata-and-the-data-analysis-wg-metadata-paper-review}{%
\subsubsection{\texorpdfstring{Metadata and the Data Analysis WG
Metadata Paper Review
}{Metadata and the Data Analysis WG Metadata Paper Review }}\label{metadata-and-the-data-analysis-wg-metadata-paper-review}}

In the report from
the first Analysis Ecosystem workshop~\cite{elmer_peter_2017_6599290} in 2017, the first item under
``Missing pieces'' is:

\begin{itemize}
\item
  \begin{quote}
  Easy access to bookkeeping information and other metadata. Common
  support for this across experiment boundaries.
  \end{quote}
\end{itemize}

It is fair to say that nothing happened in the intervening 5 years.

The main outcome from the Data Analysis Working Group Metadata paper
review~\cite{Khoo:2022pja}, prepared for
this workshop, was that the authors had identified a major problem for
HEP: Metadata; and the panel recommended the community follow up on this
important topic. This provoked a discussion at the workshop -- who would
follow up? Judging from the progress made in the last 5 years, the
problem is not about to be resolved. Metadata is a large topic touching
many areas, so would benefit from having HSF take responsibility for it
and systematically follow up on the diverse challenges it presents. An
\textbf{HSF Metadata workshop} is proposed, where stakeholders from the
experiments will be invited to participate and discuss the best way to
organise followup within the HSF community. The importance of leveraging
the existing connections to other HSF working groups and activities was
strongly emphasised during the AE2 workshop.

Using the HSF Metadata workshop as a catalyst, a comprehensive list of
use cases for metadata will be collected in an \textbf{HSF Metadata
Report}, delivered shortly after the conclusion of the workshop. The
report will cover the end-to-end lifecycle of metadata and the data they
pertain to, addressing issues like metadata scopes, evolution and change
of ownership during this lifecycle. Often metadata producers are not
responsible for metadata usage in analysis and these orphaned metadata
can become problematic. The HSF Metadata Report should cover all use
cases, not restricted to analysis, to be able to identify such gaps and
ensure metadata systems cover every metadata source for its entire
lifecycle. The report will detail the use cases and a derived set of
requirements, together with a more complete description of metadata
scopes. The paper could also try to look at commonality across use cases
and where common HEP solutions could be applicable given sufficient
buy-in from the experiments.

The definitions of metadata scopes should address the questions posed by
the panel:

\begin{itemize}
\item
  \begin{quote}
  Metadata is data about data. Which data?
  \end{quote}
\item
  \begin{quote}
  On what does the metadata depend?
  \end{quote}
\item
  \begin{quote}
  Is the metadata known at the time of data production?
  \end{quote}
\item
  \begin{quote}
  How is the metadata produced and by whom?
  \end{quote}
\item
  \begin{quote}
  How is the metadata used and by whom?
  \end{quote}
\item
  \begin{quote}
  Who is the owner of the metadata, and are there transfers of
  ownership?
  \end{quote}
\item
  \begin{quote}
  What relations between different levels of metadata exist and how are
  they handled?
  \end{quote}
\end{itemize}

The review panel gave more feedback useful for the HSF Metadata Report.
While they greatly valued the work of the Data Analysis Working Group
Metadata paper~\cite{Khoo:2022pja} authors and found the paper provided a good
overview of several existing approaches, the guidelines lacked
sufficient technical details to fully represent ``technical
specifications'' for future analysis metadata system implementations. A
certain technical elucidation of guidelines would be desirable, and will
also greatly facilitate the overall understanding of metadata use in
view of analysis preservation and reuse. This detailed understanding
could then help future metadata system designers find commonality while
keeping flexibility to extend metadata schemata to cover particular
specific scenarios. A concrete example of progress in this area may be
the CERN Analysis Preservation framework that uses composable JSON
schema to describe analysis steps that allow both common and particular
JSON fields. A detailed report from the panel will be made available in
the coming months.

\subsubsection{Systematics}\label{systematics}

While recognised as an essential component of experimental analysis, the
book-keeping and processing of systematic uncertainties was considered
the largest pain point in analysis software. The step from coding the
nominal analysis workflow to including all uncertainties is large: a
na\"{i}ve loop repeating all work for many variations inflates CPU and disk
footprints; while avoiding this demands a complex optimisation process
requiring tracking of all and only affected columns of data.

ATLAS is developing algorithm sequences for object corrections that
includes optimisation logic for generating collections with variations
applied and filtered for quality, but variation tracking and optimised
handling is otherwise a task for analysis frameworks, including
RDataFrame and Coffea. There is convergence in ATLAS and CMS on
standardised APIs for extracting systematic uncertainties from metadata
repositories (ATLAS `CP Tools' and CMS CorrectionLib), but these are
considered less applicable in LHCb and other experiments that are
heavily reliant on data-driven background estimates or very
analysis-specific procedures for determining uncertainties. The ATLAS
model has the benefit of an integrated representation of uncertainties,
allowing classification during event processing, and preset groupings of
relevant/recommended systematics on a per-object-collection basis. CMS
has entirely decoupled the application of systematic corrections from
the nominal object corrections, which offers a much more nimble workflow
with minimal software dependencies and a low CPU load. This decoupling
requires that uncertainties be parameterised rather than propagated via
alterations of the nominal procedure, so has implications on
experimental workflows beyond the pure technical solution.

The data dependencies for object systematics encourage the use of an
object-based APIs, but the increasing popularity of columnar analysis
methods suggests that these be satisfied through object facades, which
can be constructed e.g. in RDataFrame. Frameworks can then cleanly apply
variations and track affected columns through downstream operations.

Recommendation:

In conjunction with other ongoing innovations, a streamlined analysis
workflow incorporating systematic uncertainties can be proposed, arising
from an overall consensus at the workshop: Analysis should begin from
fully calibrated, reduced data formats (NanoAOD, DAOD\_PHYSLITE) derived
from the fully reconstructed AOD data. The nominal analysis workflow
should be complemented by a loop handling systematic variations computed
on the fly to favour CPU utilisation over heavier I/O, and with the
framework outputting analysis outputs, which might be histograms or a
small number of final discriminants for all variations.

This model offers space for collaboration on solutions to various common
problems:

\begin{itemize}
\item
  \begin{quote}
  An experiment-agnostic set of labels for systematic variations that
  can be used for tracking during analysis computations, as well as for
  downstream interpretation (visualisation/statistical analysis). This
  could facilitate inter-experiment statistical combinations, and be
  used not only for object/event uncertainties but also to represent MC
  and background uncertainties of relevance to the wider experimental
  community.
  \end{quote}
\item
  \begin{quote}
  A standard API for applying systematic variations to object
  collections, utilising the labels described above. Underlying
  implementations of access to metadata stores and calculations can
  remain experiment-specific, but a common representation also serves
  analysis description and preservation purposes well.
  \end{quote}
\item
  \begin{quote}
  Generalised analysis frameworks or other schemes handling optimisation
  logic, such as the RDF::Vary approach, propagating varied columns
  through all downstream analysis operations.
  \end{quote}
\end{itemize}

As a motivation for and test of these common libraries, an analysis
challenge is proposed, comprising a simultaneous top quark cross-section
analysis between ATLAS, CMS and ideally LHCb. This could share MC event
generation, to undergo simulation/recon\-struc\-tion in the respective
experimental frameworks, and subsequently work from the standard
analysis formats, but share code where possible. To encourage active
participation from the collaborations, the main deliverables would be
the establishment of a more coherent organisation and tracking of
uncertainties in CMS and progress with decoupling uncertainties from
nominal corrections in ATLAS. It is proposed that this challenge be
followed up and reported on in the HSF Data Analysis Working Group.

\hypertarget{real-time-onlinetrigger-level-analysis}{%
\subsection{``Real-time'' online/trigger-level
analysis}\label{real-time-onlinetrigger-level-analysis}}

\begin{itemize}
    \item \emph{Conveners: Giulio Eulisse, Mike Sokoloff}
    \item \emph{Speakers: Caterina Doglioni, Daniel Charles Craik}
\end{itemize}

``Real-time'' online and trigger-level analysis (RTTA) can encompass
anything from monitoring detector performance to reconstructing physics
objects for persistence for later data analysis to producing histograms
(perhaps multidimensional) that can be used directly in published
analyses (see \cite{RTAworkshop2019} for an introduction to the wide scope). A key challenge for the immediate future is ensuring that
trigger software and offline software produce the same
(indistinguishable) results, even if executing on significantly
different hardware. For example, LHCb's Run 3 first level software
trigger (Allen, \cite{Aaij2020}) executes on Nvidia GPUs in real time. What level of
difference is tolerable when running on different hardware, and what
tests do we need to demonstrate that we can tolerate the differences?
This is critical for generating and analysing simulated data in WLCG
facilities. In the case of LHCb, Allen provides an x86 back-end that
produces what are judged to be tolerably indistinguishable results.
While Allen compiled for x86 is around 4x slower than equivalent
hand-optimised x86 reconstruction code, this overhead is irrelevant in
the overall LHCb context. Experiments require some re-calibration of
trigger-level data prior to, or during, analysis workflow. Simulating
trigger-level hardware using WLCG resources requires the same level of
validation. An interesting question is how WLCG will provide
heterogeneous resources that reproduce those used in triggering in the
future, which is probably an inevitable trend, albeit not one primarily
motivated by this use case. These could include Nvidia GPUs (used by
LHCb), AMD GPUs (used by ALICE), and a variety of FPGAs (used at some
level by all experiments).

In the longer term, ``real-time'' triggers might be able to use elastic
resources in addition to the bespoke resources dedicated to individual
experiments. LHC experiments typically record data during 5 x
10\textsuperscript{6} seconds per year, and they do not acquire data
every year. To the extent resources sit idle (or mostly idle) for
extended periods of time, they are wasted. A possible alternative would
be moving (some) data from the experimental halls to elastic resources
for higher level triggering. This requires that the bandwidth be
available for transporting the data and that the elastic resources
(potentially CPUs, GPUs, and FPGAs) be available when needed. Such
solutions were investigated in the context of Run 3 but judged to be
unaffordable due to the cost of network infrastructure required. Careful
studies of technical, financial, and administrative (policy) issues will
therefore need to be done before such a model can be seriously
considered for the future. The potential benefits include more efficient
use of hardware (and personnel) and the possibility to dramatically
expand the computing power available to the highest level triggers on
short notice. A serious downside is the outsourcing of quality assurance
work together with associated sociological and management overheads,
given that it has to date generally proven to be difficult if not
impossible to capture and predict all requirements of HEP data processing
chains in specification documents.

\hypertarget{analysis-facilities}{%
\subsection{Analysis Facilities}\label{analysis-facilities}}

\begin{itemize}
    \item \emph{Convenors: Oksana Shadura, Nicole Skidmore}
    \item \emph{Speakers: Robert Gardner, Alessandra Forti, Lindsey Gray, Nick Smith, Enric Tejedor Saavedra}
\end{itemize}

\hypertarget{building-analysis-facility-infrastructure}{%
\subsubsection{Building Analysis Facility
Infrastructure}\label{building-analysis-facility-infrastructure}}

Analysis facilities (AF) can be broadly described as the infrastructure
and services that provide integrated data, software, and computational
resources to execute one or more elements of an analysis workflow. These
resources are shared among members of a virtual organisation and
supported by one or multiple organisations.

At the workshop, the community identified the following key areas of
analysis facilities that are of interest for further investigation:
\textbf{federated identity management, modern data management, and data
delivery techniques, resource-sharing mechanisms, and efficient methods
for sharing user environments} and many others. In this section a brief
explanation of each of the key areas and ideas discussed at the workshop are
summarised for further consideration by the community.

We expect that this document will help the recently founded the
\textbf{HSF Analysis Facilities Forum}~\cite{hsfaff} working group to provide
a generic list of requirements for analysis facility providers and
architects, providing best practice guidelines that will include the
design and key features provided by future AFs as well as a list of
possible research topics to be investigated later.

\hypertarget{interoperability}{%
\paragraph{Interoperability}\label{interoperability}}

Architects of AFs should provide interoperable solutions, meaning that
users should be able to navigate seamlessly from one Analysis Facility
to another with an extensible and modular design to accommodate future
needs without disruptions.

We would also like to acknowledge the importance of simultaneously
supporting ``leg\-acy'' analysis methods within the same computing
environment (or site, facility) to facilitate adoption from the end-user
physics community. This has the added benefit of conveniently
introducing and benchmarking new methods and providing realistic
feedback.

\hypertarget{identity-management-integration-in-analysis-facilities}{%
\paragraph{Identity management integration in Analysis
Facilities}\label{identity-management-integration-in-analysis-facilities}}

One of the most highly demanded features is the integration of federated
identity management in Analysis Facilities~\cite{broeder2012federated}, which allows using an
account from one facility to create an account and log in to a different
facility and related authorisation schemes including various projects
and account resource allocations.

The integration of Federated Identity management is not only limited to
AFs. The same problem arises when we look at HPCs and external
cloud-based resources. WLCG/OSG have to be made aware that with highly
dynamic and interactive work this problem gets more urgent. Identity
management is the main building block, but in addition, also
authorisation and accounting have to be federated.

Over the workshop, it was also highlighted multiple times that we need
to provide the "best practices" on how to integrate tokens and
federated identity into a new ecosystem. This includes a discussion of
the integration differences between the WLCG.

\hypertarget{data-organisation-management-and-access}{%
\paragraph{Data organisation, Management and
Access}\label{data-organisation-management-and-access}}

In terms of DOMA-related topics, the majority of the community is
concerned about the transparent movement of data between Analysis Facilities
and other storage systems used within HEP.

Fast access to input data is one of the most important aspects of an AF.
The access can be from local storage or can be remote. Users need to be
able to find the data as they would on any other type of experiment
resources and to be able to access these resources in a timely manner.
The output of the analysis needs also to be shareable between AFs and
with other sites. To satisfy these requirements an AF is expected to be
integrated with the experiments' Data Management systems and the
experiments' authentication and authorisation system (see Federated
identity section). The access to should be as fast as possible, so it is
expected that AFs, which have a high degree of repeatedly accessing the
same data, will have a system of local caches, in particular XRootD
based XCaches~\cite{fajardo2020creating} have the advantage of being fully integrated with ROOT,
the dominant file format, thus they are able to transfer only portions
of files, reducing the volume of transfers. The size and the QoS of the
caches should be evaluated.

New workflows may also use different data formats for performance
reasons or because they are more compatible with columnar analysis.
Evaluation of such formats is ongoing in the experiments and
transformation services such as ServiceX~\cite{galewsky2020servicex} are envisaged to give the users the capability to transform data on the fly from one format to another.
ServiceX functionality and scalability needs to be tested.

Another aspect we discussed during the workshop was the type of storage.
Experiments applications and services have been designed to use POSIX
file systems but the emergence of object stores with their scalability
and efficiency in serving data raises the possibility that new workflows
can be adapted to use distributed object stores and experiments would
need to review their requirements for POSIX. Whilst it is not for the
AFs to define this at a general level, AFs themselves might support different types of
storage, so this should be investigated.

%Leaving this out for the Latex document
%(e.g. the way Nick Smith mentioned in his \href{https://indico.cern.ch/event/1125222/contributions/4875668/attachments/2449795/4198186/ncsmith-objectstores.pdf}{{presentation}} on the AE2 workshop)

\hypertarget{resource-sharing}{%
\paragraph{Resource sharing}\label{resource-sharing}}

During the IRIS-HEP
Blueprint meeting~\cite{IRISHEPblueprint} an idea was mentioned to create a multi-site
substrate project which would federate contributions from multiple
resource providers (institutes and public cloud), offering a flexible
platform for service deployment at the scales needed to test the
viability of system designs and closely matching the concept of
"infrastructure as code", in the case of Kubernetes~\cite{k8s} - i.e. the substrate as
the medium upon which infrastructure (services) can be deployed
declaratively and flexibly and also be easily redeployed on the other
facilities. For efficient resource sharing, the community expects the
integration and co-location of analysis facilities with existing centers
and proposes the adoption of Kubernetes as a substrate to maximise
resource contributions and utilisation. This will also require
understanding how computing resources sharing will be organised,
including, importantly, the sharing of storage space among Analysis
Facilities.

Another important feature of scalable resource sharing is maintaining a
clear separation between execution engines and the resource sharing
layers, which means being able to switch between notebook provisioning
(fully interactive resources), analysis facility execution frameworks
(e.g. Dask workers~\cite{dask}), and traditional batch and SSH login environments.

An R\&D topic that should be investigated is how to implement fair
sharing in a scalable environment, especially the provision of resources
suitable to the type of analysis the user wants to perform. A possible
solution that was discussed is to design an intelligent portal,
knowledgeable of the different capabilities at various Analysis
Facilities, rather than users having to discover or keep track of this
information themselves.

\hypertarget{sharing-environment}{%
\paragraph{Sharing environment}\label{sharing-environment}}

The main requirement for users while doing collaborative physics
analyses is the ability to share code and results.

We discussed how environments can be shared efficiently, using existing
mechanisms, such as singularity containers distributed on~\cite{cvmfs}, and how
we could share instead of standardise environments by providing sharing
mechanisms for ``data analysis'', ``ML training''. Instead of trying to
replicate the same environment at different sites and force everyone
to use the same, the best approach could be to implement a sharing
mechanism, i.e. an infrastructure that can support any Data Analysis or
ML environment the users may require. No analysis can be performed with
a single environment as different parts of the workflow often have
different requirements (e.g. analysis-oriented environments such as
LHCb's lb-conda~\cite{lbconda}, powering environments to perform PID calibration
studies). Further interesting mechanisms allowing more efficient sharing
of environments and data are export mechanisms to HEPData~\cite{maguire2017hepdata}, sharing
cached (short-lived) data, and exporting metadata for analysis
preservation.

A key problem is how to preserve user environments when moving from one
facility to another. To solve this problem we discussed existing
mechanisms, such as centrally managed software stacks (e.g. LCG
releases~\cite{villanueva2019building}). Another opportunity could be to provide container images that
contain a tested combination of libraries that can be used to perform
the standard analysis workflows, or to allow users to ``bring-your-own''
image to take advantage of non-standard libraries using BinderHub~\cite{binderhub} for
Jupyter environments~\cite{jupyter} and workers. There needs to be a preservation
strategy for these container images and maintenance of their security
and authorship providence (software metadata).

\hypertarget{best-development-practices-for-analysis-facility-architects-and-developers-using-kubernetes}{%
\subsubsection{Best development practices for Analysis Facility
architects and developers using
kubernetes}\label{best-development-practices-for-analysis-facility-architects-and-developers-using-kubernetes}}

Currently, higher-level analysis systems are being developed within a
Kubernetes environment, and the role of having a flexible (programmable)
the substrate was noted. Since the time of the workshop, Kubernetes (Cloud
Native Computing Foundation~\cite{cncf}) has emerged as a promising technology for this role. Already we have demonstrated deployments of processing frameworks for columnar data
(Coffea~\cite{Smith:2020pxs}) and data delivery (ServiceX~\cite{galewsky2020servicex}), plus identity management combined
into declarative deployments providing an "analysis facility"
(Coffea-Casa~\cite{refId0-coffea-casa}). The identified requirements are flexible access to
infrastructure resources (CPU, disk, GPU and I/O), the declarative
nature of deployments for reproducibility at multiple centers and
providing the capabilities for processing frameworks and interactive
interfaces.

Similarly, we need to work on standardising analysis facility deployment
tools (today Helm~\cite{helm}, tomorrow its successors or the new tools) and
supporting the surrounding infrastructure (Helm chart catalogs,
container registries). This defines a new training area for system
administrators, to build a culture of analysis facility providers
experienced in the dependent technologies. Beyond deployment, this
includes a new set of troubleshooting skills and infrastructure
knowledge (pods, containers, controllers, etc.).

For novel analysis systems and related services, we also need to draw the
clear distinction with site-specific attributes and removal of
dependencies (e.g., a site's identity or account management service,
LDAP, etc.).

A ``best practice'' that the community recommends for future analysis
facilities is that developers should always ask themselves about
portability. For instance, could a facility be deployed at different
institutions? As well as the scalability of deployment, could an
analysis facility be used by more than one user community with the
experiment-specific dependencies removed, or the needed hooks properly
abstracted? Another question is if the deployment pattern lends itself
to open source best practices and whether it can be deployed with
confidence by a K8s cluster operator without significant supervision and
what priors are assumed if this is not the case?

From a security point of view, for analysis facilities built with one or
more internet-facing services, image security is paramount, and services
for vulnerability scans, patch updates, etc. need to be addressed.

We should also encourage analysis facility developers during the design
phase to have in mind the following capabilities: privileged interfaces
for user management, roles, and what metadata is needed. A public-facing
dashboard using open technologies with status reports about the AF such
as performance metrics and community metrics (such as analysis groups,
tasks, users, institutions) will help to measure impact.

Over the course of the workshop, we discussed the differences between
Kubernetes "distributions" (e.g. vanilla Kubernetes vs OKD~\cite{okd}) and how to
get applications to work on both without the users having to choose.
There was an agreement that there should be guidelines for K8s apps
developers that take into account the differences. The K8s application code
adjustments needed, depending on new Kubernetes releases, is another
the challenge that developers will need to keep pace with and could be
efficiently resolved through better interaction with the Cloud Native
community~\cite{cncf}.

\hypertarget{tracking-analysis-performance}{%
\subsubsection{\texorpdfstring{Tracking analysis performance
}{Tracking analysis performance }}\label{tracking-analysis-performance}}

To provide an extensive overview of how resources are used and to ensure
that an informed decision about which AF to use can be made, Analysis
Facilities should publish some key parameters related to performance
(occupancy, etc.). For this reason, we need to provide instrumentation to
understand user modalities, preferences, and analysis throughput. This
can result in a variety of metrics - session/task timings, resource
consumption, and contextual system "business" markers (the single-use
systems, many users/groups, undersubscribed and oversubscribed
resources).

Examples of metrics that could help to have an insight into how and if
efficiently used resources in the analysis facility:

\begin{itemize}
\item
  \begin{quote}
  workflow ID,
  \end{quote}
\item
  \begin{quote}
  CPU, RAM, swap,
  \end{quote}
\item
  \begin{quote}
  I/O (local storage and network),
  \end{quote}
\item
  \begin{quote}
  software stack,
  \end{quote}
\item
  \begin{quote}
  job failure rate,
  \end{quote}
\item
  \begin{quote}
  Time To Completion (TTC),
  \end{quote}
\item
  \begin{quote}
  data source - entirely local or cached from a Data Lake and formats
  used on input (PHYS, PHYSLITE, AOD, MiniAOD, NanoAOD, NTuple,etc..),
  \end{quote}
\item
  \begin{quote}
  formats written (columns), ratios.
  \end{quote}
\end{itemize}

A cost-effective or user-effective solution could require users to test
jobs on subsamples of the data prior to the final submission on the full
samples. This option is, for example, already explored by the LHCb
experiment (exploiting GitLab pipelines) for NTuple productions with the
analysis productions framework. This validation step allows one to
collect many of the metrics mentioned above.

We need to introduce a concept of telemetry for various components of
Analysis Facilities that will provide an overview of what users are
actually doing and which components of the analysis pipeline are run.
Telemetry in the software stack will likely require a dialog with
software analysis framework developers to coordinate a common approach
as well as to establish an API for tools to write their own usage logs
and metrics. Most likely a standard set of Kubernetes tools can do this,
along with some log message format standards.

Collecting more data could also allow the optimisation of resources
requested for known workflows. This means that we could, for example,
scale user jobs for heterogeneous architectures, particularly using
accelerators in a more adaptive way and to try to tune the analysis
configurations in a heuristic way through some validation phases.

The IRIS-HEP Analysis Grand Challenge (AGC)~\cite{agc} workflow defines an analysis
benchmark that could be easily re-implemented and executed on any
generic Analysis Facility and help to showcase how to use existing
analysis facilities on a scale appropriate for this analysis. During the
workshop the IRIS-HEP Analysis Grand Challenge (AGC) workflow was
suggested as one of the analysis facility performance benchmarks, giving
the possibility for AFs architects or resource managers to understand
better data analysis bottlenecks at AF. AGC also could be used to
explain to the users of the analysis facility what it could look like
for an efficient data analysis pipeline.

\hypertarget{collecting-user-requirements-and-metrics-of-success-for-analysis-facilities}{%
\subsubsection{Collecting user requirements and metrics of success for
analysis
facilities}\label{collecting-user-requirements-and-metrics-of-success-for-analysis-facilities}}

Every project, including Analysis Facilities, needs to determine
the success of a project and help project managers to evaluate a project's
status, foresee risks, and assess the quality of work. For this reason it
should be initially envisaged to perform regular user surveys and define
key parameters to capture the success of analysis facilities. As a
baseline to evaluate AF success as a project, we could propose to use
the standard key metrics such as the total number of users or a number of
analyses or the evolution of usage over time, both in terms of the number of
users and compute time spent. Other examples of metrics could include
the number of users per week versus time or the number of active users
within a given time window and by which method users are interacting
(interface, traditional batch), the total number, volume, and format of
files consumed/written and shape of workflow statistics.

Another important category of metrics is the more user-oriented
metrics, such as the number of reported issues, the time to answer
tickets, the time taken to solve the issue, feedback from the user
satisfaction surveys, etc., as well as other user engagement metrics
such as quality and user satisfaction using prepared documentation,
tutorials and quality of user-engagement training for a given facility.

We also need to continue to improve analysis facilities by collecting
additional user requirements and keeping track of improvements over time
and the uptake over time. This also could be defined as an additional
set of metrics.

\subsection{Breakout Session - Analysis Workflow
Design}\label{breakout-session---analysis-workflow-design}

During the workshop, there was a breakout session on \emph{Analysis
Workflow Design}, the summary of which is linked
\href{https://docs.google.com/document/d/116ITweumV9AKIBvMn-GcPQ4UQ--xvZLhAgtjlgGhZdc/edit?usp=sharing}{{here}}.

\hypertarget{conclusions-outcomes}{%
\section{Conclusions / Outcomes}\label{conclusions-outcomes}}

Here we summarise the main outcomes of the workshop that should be
followed up:

\begin{enumerate}
\item
  Object facades, which group columnar data into views that allow
  reasoning about physics objects, are important for a good analyser
  user experience. The HSF Data Analysis Working Group should coordinate
  a harmonisation effort to ensure consistent behaviour across different
  tools in this regard.
\item
  The interoperability of tools should be strongly encouraged and the
  HSF should encourage appropriate discussions between development
  teams.
\item
  Best practice for onboarding new analysts (in analysis models and
  programming techniques) is a crucial area in which to have continued
  commitment between the HSF Training and Data Analysis groups and the
  experiments.
\item
  Continued R\&D for reduced formats should focus on how to accommodate
  special inputs for non-standard workflows. Possible approaches include
  the use of friend trees to augment existing reduced formats or using
  object stores to access columns across processing tiers.
\item
  The curation of open datasets (or equivalently simulators) along with
  benchmarks for common problems in machine learning is of clear
  importance, and should be made a focus of the experiments that have
  sufficient data and meaningful task definitions to suit this format.
  This will bring the field more in-line with industry standards for
  replication, give a better feel for how well new methods perform on
  certain tasks, and reduce friction for non-physics experts to
  contribute in a meaningful way to particle physics research.
\item
  Differentiable programming received a clear interest from those
  present in the workshop, but lacks thorough comparisons with existing
  methods at scale, and requires concentrated effort to harmonise
  software and methodology efforts. This should be a priority going
  forward for those involved to make it clear to the field when, how,
  and if these methods warrant usage.
\item
  Metadata requires dedicated follow up and an HSF Metadata Workshop is
  proposed, tentatively for early 2023. This should be coordinated by a
  team nominated by HSF Coordination. The product of that workshop
  should be an HSF Metadata Report, covering all Metadata use cases
  through their end-to-end lifecycle. Discussions on how best to provide
  dedicated follow up, via new or existing working groups or otherwise,
  will also be an important part of the workshop.
\item
  Concerted effort is needed to tackle systematic uncertainty handling,
  identified as the largest pain point in the analysis workflow.
  Agreement on an idealised workflow for analysis of calibrated, reduced
  analysis formats provides a model for R\&D and a set of deliverables.
  A joint top-pair cross-section analysis on open MC between ATLAS/CMS
  and LHCb was proposed as a test challenge for development of common
  tools to describe and extract standard uncertainties both in analysis
  code and in downstream interpretation frameworks.
\item
  The future of computing is moving towards higher concurrency and
  increased use of accelerators. We believe that analysis usage is no
  exception to this trend. Many of the tools discussed at the workshop
  should be supported to continue the work of adapting to this future
  reality, and taking full advantage of facility resources that offer
  accelerators like GPUs and FPGAs. That is clearly an element of the
  path towards our ultimate goal of reducing time to insight, by making
  results available within interactive time scales. The HSF, through its
  various working groups, should attempt to foster best practice and
  knowledge sharing in this transition.
  \begin{enumerate}
  \item
    As a particular specific aspect, the numerical stability and
    reproducibility of multi-architecture codes should be studied, with
    an understanding of when sufficient and acceptable architecture
    independence of results has been achieved.
  \end{enumerate}
\item
  Community involved in development modern AFs identified the following
  key areas of analysis facilities that are of interest for further
  investigation and improvements: Identity management integration,
  modern data organisation solutions and heavy usage of data delivery
  techniques, investigate the object stores, the resource sharing
  mechanisms and the efficient user friendly methods for sharing user
  environments between analysis facilities and many others.
\item
  With more analysis level services deployed within the Kubernetes
  environment an essential area for improvement is standardisation of
  analysis facility deployment tools and support of the surrounding
  infrastructure. Defining a new training area for system administrators
  will give analysis facility providers sufficient experience in the
  dependent technologies.
\item
  To provide an extensive overview on how AF resources are used and to
  ensure that an informed decision about which AF to use can be made, it
  would be useful if Analysis Facilities publish key metrics related to
  performance and usage. It is initially envisaged to perform regular
  user surveys to capture the success of analysis facilities, focusing
  on user-oriented or user engagement metrics. We can also use this to
  understand user modalities, preferences and analysis throughput to
  guide AF development.
\end{enumerate}

\section{Acknowledgments}
We thank the Irène Joliot-Curie Laboratory, the HEP 
Software Foundation, IRIS-HEP, and nVidia for sponsoring
this workshop.
We especially thank the administrative staff of the
Irène Joliot-Curie Laboratory for their help.
This work was supported in part by the U.S. National Science Foundation (NSF) Cooperative Agreement OAC-1836650 (IRIS-HEP) as well as the German Research Foundation (DFG) under grant EXC-2094-390783311.
We acknowledge support of participants from their
national funding agencies, including:
CNRS/IN2P3 (France);
INFN (Italy);
DFG (Germany);
STFC (United Kingdom);
DOE and NSF (USA).

\sloppy
\raggedright
\clearpage
\printbibliography[title={References},heading=bibintoc]

\end{document}

### 2023-10-12 HSF-TN/2020-01/HSF-TN-2020-01.tex
Update of TN-2020-01
This commit has all changes, but an incorrect Zenodo DOI, that needs to be
updated once a new version is ready to be submitted

\documentclass[12pt,a4paper]{article}

% Variables that controls behaviour
\usepackage{ifthen} % for conditional statements
\newboolean{pdflatex}
\setboolean{pdflatex}{true} % False for eps figures

\newboolean{inbibliography}
\setboolean{inbibliography}{false} %True once you enter the bibliography

\textheight=230mm
\textwidth=160mm
\oddsidemargin=7mm
\evensidemargin=-10mm
\topmargin=-10mm
\headsep=20mm
\columnsep=5mm
\addtolength{\belowcaptionskip}{0.5em}

\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}

\setlength{\hoffset}{-2cm}
\setlength{\voffset}{-2cm}

% Page defaults ...
\topmargin=0.5cm
\oddsidemargin=2.5cm
\textwidth=16cm
\textheight=22cm

% Don't chase after perfection
\raggedbottom
\sloppy

\usepackage{float}
\usepackage{microtype}
\usepackage{lineno}    % Line numbering during drafting
\usepackage{xspace}    % Avoid problems with missing or double spaces after predefined symbold
\usepackage{caption}   % These three command get the figure and table captions automatically small
\renewcommand{\captionfont}{\small}
\renewcommand{\captionlabelfont}{\small}

\usepackage{booktabs}

%% Graphics
\usepackage{graphicx}  % to include figures (can also use other packages)
\usepackage{color}
\usepackage{colortbl}

%% Math
\usepackage{amsmath} % Adds a large collection of math symbols
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{upgreek} % Adds in support for greek letters in roman typeset

%% fix to allow peaceful coexistence of line numbering and
%% mathematical objects
%% http://www.latex-community.org/forum/viewtopic.php?f=5&t=163
%%
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname
end#1\endcsname
 \renewenvironment{#1}%
   {\linenomath\csname old#1\endcsname}%
   {\csname oldend#1\endcsname\endlinenomath}%
}
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}%
}
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
\patchBothAmsMathEnvironmentsForLineno{eqnarray}%
}

% Get hyperlinks to captions and in references.
% These do not work with revtex. Use "hypertext" as class option instead.
\usepackage{hyperref}    % Hyperlinks in references
\usepackage[all]{hypcap} % Internal hyperlinks to floats.

% Make this the last packages you include before the \begin{document}
\usepackage{cite} % Allows for ranges in citations
\usepackage{mciteplus}

\usepackage{longtable} % only for template

% Pandoc
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\begin{titlepage}
\pagenumbering{roman}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2020-01 v2.0 \\  % ID 
 & & 10.5281/zenodo.1469634 \\ % DOI
 & & Original Date: 4 October 2016 \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & Last Update: 12 October 2023 \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{2.0cm}

\begin{center}
  \includegraphics*[width=4cm]{hsf_logo_angled.png}
\end{center}

\vspace*{2.0cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
  Proposal for HSF Project Best Practices
\end{center}
}

\vspace*{2.0cm}

% Authors -------------------------------------------------
\begin{center}
Benedikt Hegner$^1$, Benjamin Morgan$^2$, Graeme A Stewart$^1$
\bigskip\\
{\it\footnotesize
$ ^1$CERN\\
$ ^2$University of Warwick
}
\end{center}

\vspace{\fill}

\end{titlepage}

\section{Proposal for HSF Project Best
Practices}\label{proposal-for-hsf-project-best-practices}

\subsection{Introduction}\label{introduction}

This technical note is a proposed list of best practices for HSF and
other HEP open source projects. The main motivation is to ensure
interoperability and usability of a given project by other projects and
being able to build consistent software stacks. In addition, it should
make it easier for other developers to contribute to existing projects.
In the following we discuss different practices and conventions that
ease the life of

\begin{itemize}
\tightlist
\item
  developers and new contributors,
\item
  end-users and other client projects.
\end{itemize}

Afterwards we provide a checklist of the proposals you may want to use
for your project. The proposals are mainly based on experience with the
LCG software projects and releases. You may find many of the points
discussed here trivial. However, people usually differ in what they
consider trivial. The technical recommendations in this document are
tailored towards C++-based projects, but can easily be mapped onto,
e.g., Python-based projects.

\subsection{Project Scope, Name and
Visibility}\label{project-scope-name-and-visibility}

On starting a project, make sure you have an idea of the project's scope
and goals. Try to pick a name that is suitable for that. You need to
ensure uniqueness, as the name will be used to name software artifacts,
like libraries, code namespaces, error messages, etc. In addition, have
a look around to see if it conflicts with pre-existing trademarks for
software products or services.

Though it sounds like a triviality, your project should be made known to
the community. For this, having a \emph{dedicated project website} or
another entry point for information is essential. It should concentrate
all the information useful for users and developers. If possible, it
should point at all the other information listed in this document. It is
important to find the right place to put information. Try not to repeat
yourself, as duplicated documentation can easily get out of sync. Access
to all sources of project information should be granted to search engine
spiders. Furthermore, the HSF working groups and software forums allow
you to present your project or ideas at any stage in its project
lifecycle.

There is an excellent open source
\href{https://opensource.guide/}{general guide} that covers community
interactions, getting your project known and helping find new
contributors, particularly helpful for the sociology of successful open
source projects.

\subsection{Supporting Developers and
Contributors}\label{supporting-developers-and-contributors}

The following sections discuss points mainly relevant for project
developers and potential new contributors.

\subsubsection{Code repository}\label{code-repository}

The first requirement for an open-source project is fully versioned code
in a \emph{public repository}. The code should be accessible in
anonymous read-only mode by everybody. Services like
\href{https://github.com}{GitHub} or \href{https://gitlab.com}{GitLab}
provide it for free. In addition efforts like
\href{https://www.hepforge.org/}{hepforge} or labs like CERN or DESY may
host HEP-specific packages. Services supporting a clone plus
\emph{merge-request/pull-request workflow} can be extremely helpful to
attract new contributors, as it is the current de-facto standard for
open software development. Try to make sure there are no barriers to
contributing in this way, e.g., lab hosted services may be more
difficult for users without accounts to use for merge requests.

\subsubsection{License and Copyright}\label{license-and-copyright}

The ownership and copyright of the code has to be well defined and
understood. Host labs for experiments will often hold copyright on
behalf of experiments and projects that are not legal entities and this
can simplify future life enormously. In a second step, the code and
software provided should be properly licensed in order to be able to use
code provided by others, and to allow people to re-use, update, or
improve the software you provide. The
\href{http://hepsoftwarefoundation.org/technical_notes.html}{HSF
technical note} \emph{HSF-TN-2016-01} (\emph{Software Licence Agreements
HSF Policy Guidelines}) discusses various options. This is one of the
topics that is \emph{typically ignored at the beginning of a project and
hard to fix afterwards}.

\subsubsection{Compilation and other
commands}\label{compilation-and-other-commands}

Compiling, installing and testing should each be - if possible -
single-command actions. In particular, making testing easy is important.
A good place to put the necessary information is a \emph{README} file in
the repository. Relying on community standards like
\href{https://www.cmake.org}{\emph{CMake}} make it easier for others to
use and understand the setup.

\subsubsection{Testing}\label{testing}

To improve on the quality of software, unit and integration testing are
essential. Having well-documented tests makes it as well easier for
contributors to participate. They can check whether they break old
features and can, with new tests, document what their addition is
supposed to do. Testing can also assist in the triage and fixing of
bugs. Tests can be written to reproduce reported issues and fail when
they occur, with subsequent fixes validated by the tests passing.

For \emph{unit tests} plenty of software packages exist, of which
\href{https://github.com/google/googletest}{\emph{gtest}} and
\href{https://github.com/philsquared/Catch}{\emph{catch}} are two good
choices for C++ projects. Integration tests running a software project
in a certain setup can take advantage of \emph{CTest} (supplied as part
of \emph{CMake}) and \href{http://www.cdash.org}{\emph{CDash}} or be
driven by shell scripts. Ease of use is again important here, otherwise
tests tend not to be run. For example, \emph{CMake}/\emph{CTest} add
dedicated \emph{test} targets to buildscripts so that running the tests
is a simple matter of ``building'' the target, e.g.~\texttt{make\ test}
when using Makefiles.

\paragraph{Continuous integration}\label{continuous-integration}

As well as being available from the command line, tests of the code
should run on all pull/merge requests so that reviewers can immediately
see if the proposed changes break any known functionality. There are
many options to do this, well integrated with modern code repositories.

\subsubsection{Communication and
Reporting}\label{communication-and-reporting}

A mailing list to contact developers is always useful, even as issue
trackers become also more capable of supporting discussions. It is
better to have publicly and anonymously accessible archives and to be
open for subscription and posting by the public.

\subsubsection{Issue tracking}\label{issue-tracking}

It is useful to provide an issue (bug) tracker for users and developers
to interact with, allowing a view of both open and closed tickets
anonymously by the public. Optimal solutions here are the issue tracking
capabilities the code repository itself (such as GitLab or GitHub), as
solutions which integrate directly with the code repository are much
easier to use for both users and developers. CERN's
\href{https://www.atlassian.com/software/jira}{JIRA} service is an
alternative.

\subsubsection{Reference Guide}\label{reference-guide}

For developers it is important to have a good overview of provided
interfaces, existing classes, and implementation details. For this a
reference guide is a helpful tool. The de-facto standard for creating
reference guides in C++ projects is
\href{http://www.doxygen.org/}{\emph{Doxygen}}.

\subsubsection{Conventions and
Workflows}\label{conventions-and-workflows}

Every project choses certain (coding) conventions and integration
workflows. While there is a plethora of possibilities, the concretely
chosen conventions and workflows should be documented visibly. A
\emph{How to contribute} document is good practice. This is, as well, a
nice place to add information where contributions by others would be
possible and desired.

\subsubsection{Be prepared for using external (cloud)
services}\label{be-prepared-for-using-external-cloud-services}

The project should be careful in its assumptions about the environment
available for development and testing, like access to extra storage or
connections. Make it easy to integrate your software into e.g.~a
container. Containers also help greatly in deploying a continuous
integration system, e.g., testing the build on different Linux flavours.

\subsection{End-users and client
projects}\label{end-users-and-client-projects}

\subsubsection{Documentation}\label{documentation}

In addition to the already mentioned documentation, end-user focused
documentation is important. A little checklist further below summarizes
the most important information to be given as part of the documentation.

\subsubsection{Release Information}\label{release-information}

While developers (most of the time) know the changes between various
releases, it is important to document changes between releases for
end-users. It turned out to be a good policy to have multiple categories
of releases, like production releases, development releases, bug fix
releases, etc. While each project may have different conventions here,
the chosen convention should be explained, including its meaning in
terms of changes to the project's
\href{https://en.wikipedia.org/wiki/Application_programming_interface}{\emph{API}}
and
\href{https://en.wikipedia.org/wiki/Application_binary_interface}{\emph{ABI}}.
A clear numbering scheme like ``major.minor.patch'' can support this
(also known as \href{https://semver.org/}{semantic versioning}). For
each release the \emph{supported compilers}, \emph{supported operating
systems} and \emph{required dependencies} should be listed. This helps
avoiding frustrations on the user side.

\subsubsection{Interaction with
developers}\label{interaction-with-developers}

To be able to interact with developers, both the already mentioned
\emph{mailing list} and \emph{issue tracker} are important and helpful.
The required permissions to post there should be as low as possible.
Make it easy for people to give feedback and to contribute.

\subsubsection{Relocatability and co-existence of
versions}\label{relocatability-and-co-existence-of-versions}

Often a project has to be integrated into bigger software stacks. Being
relocatable, i.e.~having no hard-coded absolute paths in any build
artifact, is often a necessity to deploy and distribute these stacks. To
enable your project to become part of such a software stack, try to make
it relocatable. In addition your software should not make too strong
assumptions about its own location.

\subsubsection{Usability and run-time
settings}\label{usability-and-run-time-settings}

It should be straight forward for a user to set up and run your project.
This can for example be ensured by providing environment setup scripts,
but the number of environment variables required should be limited as
far as possible.

\subsubsection{Be prepared for using external (cloud)
services}\label{be-prepared-for-using-external-cloud-services-1}

The project should be careful in its assumptions about the user
environment, like having access to extra storage or network connections.
Like for the development process, make it easy to integrate your
software into e.g.~a container.

\subsubsection{Publications and
References}\label{publications-and-references}

Users of your software should be able to give credit to your work. Try
to publish your work in conference proceedings or journals such as
\emph{Computing and Software for Big Science} so that it can be properly
cited.

To help users cite your software make sure that you list the recommended
citation where it can easily be found in your documentation. One highly
recommended standard is now to add a \texttt{CITATION.cff} file to your
repository. This is a human and machine readable file that tells users
exactly how to write a citation. More information is on the
\href{https://citation-file-format.github.io}{Citation File Format
website}.

\subsection{Making best practices easier - the HSF Template
Project}\label{making-best-practices-easier---the-hsf-template-project}

Many of the points mentioned are per se trivial, but need some
infrastructure to be set up. To assist new projects, an HSF project
template was created. It covers many of the technical points and
provides some canonical or example implementation for many of the
issues. It is meant as open collection point of ideas and proposals by
the community.

\subsection{Checklist}\label{checklist}

A little checklist of topics to consider is given here. Not every point
applies to every project, but it may give you a handle in improving the
quality of the software you provide.

\subsubsection{Repository and code
checklist}\label{repository-and-code-checklist}

\begin{table}[H]
  \caption{Repository and code checklist}
  \label{tab:repository-and-code-checklist}
  \begin{tabular}{lll}
  \toprule
  \textbf{Topic} & \textbf{Possible solution(s)} & \textbf{Template} \\
  \midrule
Public repository & \href{https://github.com}{github},
\href{https://gitlab.com}{gitlab} & - \\
License + file & MIT, Apache2 & MIT, Apache2 \\
README file & \href{https://en.wikipedia.org/wiki/Markdown}{Markdown},
\href{http://docutils.sourceforge.net/rst.html}{reStructuredText} &
Yes \\
Reference guide & \href{http://www.doxygen.org}{Doxygen} & Doxygen \\
build scripts & \href{https://www.cmake.org}{CMake} & CMake \\
Unit testing & \href{https://github.com/google/googletest}{gtest},
\href{https://github.com/philsquared/Catch}{catch} & catch \\
Integration testing & CTest, scripts & CTest \\
version file & headers & headers \\
Relocatability & strict policy & Yes \\
environment setup & (c)sh script & - \\
\end{tabular}
\end{table}

\subsubsection{Procedure and release checklist}\label{procedure-and-release-checklist}

The following list contains mostly ``nice-to-have'' points. Having them
well-defined and documented helps both developers as well as potential
volunteer contributors.

\begin{table}[H]
  \caption{Procedure and release checklist}
  \label{tab:procedure-and-release-checklist}
  \begin{tabular}{lll}
  \toprule
  \textbf{Topic} & \textbf{Possible solution(s)} & \textbf{Template} \\
  \midrule
Defined workflow & plenty & \\
Automatic testing & \href{https://docs.github.com/en/actions}{Github Actions},
\href{https://about.gitlab.com/solutions/continuous-integration/}{gitlab CI} & - \\
Test run+reporting & CTest,CDash & CTest \\
Static Analysis &
\href{http://clang-analyzer.llvm.org}{clang-analyzer} & - \\
  \end{tabular}
\end{table}

\subsubsection{Website and information checklist}\label{website-and-information-checklist}

\begin{table}[H]
  \caption{Website and information checklist}
  \label{tab:website-and-information-checklist}
  \begin{tabular}{ll}
    \toprule
    \textbf{Topic} & \textbf{Possible solution(s)}\\
  Website & \href{https://jekyllrb.com}{jekyll},
  \href{https://pages.github.com}{github pages}, \href{https://gohugo.io}{Hugo} \\
  How to contribute & - \\
  User manual & markdown, doxygen \\
  Reference manual & doxygen \\
  Bug tracker & github, gitlab, jira \\
  Mailing list & google groups, e-groups \\
  Link to repository & - \\
  List of releases & - \\
  List of supported OS+compilers & - \\
  List of pre-requisites & - \\
  CITATION.cff file & \href{https://citation-file-format.github.io}{CFF website} \\
  \end{tabular}
\end{table}

\subsection{Summary and Outlook}\label{summary-and-outlook}

This document described a few best practices, and potential
implementations. Updates, additional points or corrections are very
welcome.

\end{document}

### 2025-05-14 HSF-TN/2024-01/latex/accelerators.tex
Add LaTeX source

\section{Accelerator  Resources}
\label{accelerators}

The most commonly used accelerators in HEP are GPUs and this is the focus of the discussion here. It is expected that both availability and demand for GPU resources is going to increase, not least in the context of HPC facilities, and for sustainability considerations similar to those made for the introduction of ARM processors on the grid \cite{arm}. It is important that GPU resources be made available both for interactive and non-interactive workloads. Particularly in machine learning, a large fraction of time is spent in R\&D to develop a candidate model interactively, before large-scale non-interactive trainings are launched, e.g. for hyperparameter tuning.

Aside from pure availability of heterogeneous resources, it is important for AFs to provide software hooks to integrate such hardware. WLCG computing has until recently been based on the assumption of relative uniformity of hardware resources. The exclusive use of x86 architecture guaranteed that user code could run everywhere. GPUs, or accelerators in general, pose a new challenge for AFs and WLCG sites. The variety of GPU models, functionality and software needed to access them, as well as the complex memory management, make GPUs more difficult to insert both in a local context as well as at distributed facilities. The user code is still the subject of R\&D and the requirements, not yet as well defined as for CPUs, vary from GPU sharing, to partial offload of the code to a GPU,  to use of multi-GPUs for large ML models, to requests for specific versions of libraries. How to tune batch systems to support this variety of configurations and possible user requests is still being investigated by sites. For this reason GPUs so far have been used mostly on dedicated resources, such as the HLT farms for very specific workflows \cite{lhcbhltgpu}, or users have requested single (dedicated) interactive GPUs. In the few cases where multiple GPUs have been requested by a single user they tend to be allocated on a single node. Particularly interesting is the case of interactive GPUs which remain idle while the user is coding, especially when used as a notebook resource that is assigned to the session. Time sharing could be considered in these cases. Physical partitioning is also an option in newer GPU models, and permits providing multiple GPU flavors. Some AFs, like Nebraska and FNAL elastic AF, are deploying specialized software, such as NVIDIA's Triton inference servers, for their inference-as-a-service infrastructure in parallel to their batch system access. The Triton server schedules the model requests and can be used from either local HTCondor jobs or from the interactive nodes.

Users also use commercial cloud resources to access GPUs, but this type of access is usually not integrated. As demand for such resources is still very dynamic, it may be interesting to investigate models of AFs that couple with cloud computing resource providers, where such resources can be allocated on-demand. This model has been used by LHC collaborations, such as ATLAS, to investigate the technical compatibility of collaboration software, as well as validation of the physics output, for other types of non-x86 architectures like ARM and this coupling can also be used for accessing various types of accelerators, more commonly GPUs. CMS, LIGO and protoDUNE have also used commercial providers for integrated inference-as-a-service. In these cases the user interface can either be independent or mediated through the experiment grid WFMS (Workflow Management System).

### 2025-05-14 HSF-TN/2024-01/latex/doma.tex
Add LaTeX source

\section{DOMA (Data Organisation, Management, Access) }
\label{doma}

Achieving the transparent movement of data between AFs and other storage systems used within HEP is still a matter of development. Fast access to input data is one of the most important aspects of an AF. Considerable effort is going into reducing the input data volume handled by end users so that it can fit comfortably at single sites. However, not all analyses will be able to use such “reduced” formats (the current plan is to have a 70-80\% analysis using them) and whilst the site may have the storage space, it may not have the interactive computing resources for an AF (for example a T1 site). A second aspect that needs to be taken into consideration is where users’ output will be stored and if it needs to be shared. Experiment analysis models are expected to develop such that direct large-scale ntuple productions by analysis groups are very uncommon. The new reduced formats are expected to replace these,  with large productions continuing to be carried out on the grid. Support for access and distribution of data from analysis facilities needs to be guarantueed, i.e. AFs should be taken in consideration in the experiment's data model

\subsection*{Data access: single site, replicated or cached}

Users need to be able to find their input data as they would on any other type of experiment resource and be able to access them in a timely manner. There are a few possibilities to organize the data and these shape the AF infrastructure:

\begin{itemize}
    \item Put everything at CERN. This of course is possible but the interactive and batch system resources should also increase to cope as well as the user support response to the increased requests.
    \item Create other facilities that host  a large storage common to all computing resources and a large amount of interactive/batch resources, similar to those found at CERN. Typically this would be at T1s or large labs like DESY. 
    \item Remote access to all the data from interactive computing resources with the latency hidden by the caches. This already exists at some sites like UCSD and NERSC.
\end{itemize}

The last of these models can maximize the usage of any type of available computing resource and does not stop either of the previous models from being implemented.

Analysis workflows frequently access the same data repeatedly, so caching is beneficial. So far all testing has been done using XRootD~\cite{xrootd} based XCaches~\cite{xcache}, which have two advantages: the first is that the applications (including uproot~\cite{uproot} and RData Frame\cite{rdataframe}) can read only the required portions of files, reducing the volume of transfers; the second is that they are already used at HEP sites so it is just a matter of ironing out the more specific user requirements. Hosting managed copies of the data is usually painful, unless a grid storage system is available and integrated in the experiment’s system. Caches, however, require little maintenance. If the deployment of a cache is simple enough that it can be adapted to be used at Tier3 sites then these resources could continue to be used effectively without hosting managed data. This is particularly important considering the analysts will want local user formats (nanoAOD/PHYSLITE) skimmed (a typical case is  multi-lepton analysis) or augmented (i.e. custom objects that need tracks or particle flow candidates). Augmented files may point to external remote files, so the ability to copy locally the augmentation onto a cache is almost mandatory to reduce latency. There is the possibility to use https for caches, but that is not as well tested and would require further R\&D. Using caches means the HL-LHC data, in whatever format, has to be on remotely accessible storage that can be integrated with the caches, i.e., it cannot be on purely internal storage. There is also a question on the specifications of these caches. They need to be fast, e.g. on SSD and NVMe, and possibly support a large number of users; but they could also be large HDD which are not as fast but store a lot more data. For example the Southern California (SoCal) Cache is a ~2PB cache serving CMS analysis jobs from both UCSD and Caltech and is made out of retired HDDs. The scalability comes from using several data servers grouped under a redirector (similar to a regular XRootD cluster).

\subsection*{Common namespaces}

Depending on the preference of the experiment on how to track data access, access to remote data may benefit from integrating the data management tools into the user interface. Usually users query their experiment’s DDM (Distributed Data Management) system  (e.g. Rucio or Dirac~\cite{dirac}) from the command line, but considering the request of interactive resource access via IDE, it is interesting to highlight the integration of Rucio with a Jupyter notebook plug-in. This was developed by Rucio and the ESCAPE~\cite{escape} project and avoids a double step of having to query Rucio in a terminal and then upload the resulting list in the notebook. The plug-in could have configurable caches available. In this case the DDM tools provide the data common namespace. It is also possible to access a common namespace using a combination of XRootD redirectors and caches like CMS does, however here the user still has to know the list of files they want to access, requiring interaction with a file catalog.

Another example of a common namespace is when site storage underpins all its computing resources. Usually this is a POSIX-like file system. The most prominent example for this is EOS at CERN. EOS offers a POSIX-like file system, with a single namespace, that can be accessed by interactive resources, local batch and cloud resources. Grid jobs and remote non-grid resources can also read and write data via other protocols and the users can always find them easily via the POSIX-like interface. This setup is one of the reasons CERN is so popular for analysis --- everything is easily accessible with a variety of different tools.

\subsection*{POSIX vs Object stores}

Full POSIX compliance cannot be expected and is also very unlikely to be necessary  for complexity and for performance reasons. In this text, POSIX should be implicitly understood as POSIX-like.

Users favor POSIX file systems because they are more familiar and intuitive and the experiments’ applications and services are also built on POSIX. The emergence of object stores, with their scalability and efficiency in serving data, raises the possibility that new workflows can be adapted to use distributed object stores. Scalability of hierarchical POSIX file systems is a recurring worry, and further effort to understand if data access speed and scalability are limited by application logic or storage technologies is needed. One should also clearly distinguish between the use cases: accessing several TB of data from mass storage vs copying a few smaller files in a local directory. One needs to take into consideration the evolution of analysis software, like ROOT. Traditionally it is bound to local IO and relies on POSIX semantics, which may use "advanced" concepts (symlinks, hardlinks, fseek()...) for maximum performance and convenience. Some analysis software may already provide data IO interfaces for simpler protocols (such as Amazon's S3 and XRootD). However even though the tools provide those interfaces, users usually wrap their analysis in their own code that will use a local file system, because that is what they are familiar with. For example, a user may create a script to chunk a data file into smaller files relying on local Unix tools rather than learning how to do that in S3. Users may also want to explore a new dataset or download some data to their local computers in a format with which they are familiar, which is static files with POSIX semantics for access/interaction. This happens on a small scale: users are not interactively exploring or exporting terabytes to petabytes of data. Therefore, the only requirement is that an object store can export (usually small amounts of) selected data to a static file. Similarly, user output (histograms, etc.) from analysis workflows is orders of magnitude smaller and may be input to subsequent tasks that are separate from the AF and any object store semantics. Therefore, the ability to output static files is useful, but separate from storing, accessing, and caching TB-PB of data. "Skimming", keeping only selected subset of events for further analysis, is an intermediate case, and one probably wants the option to output into an object store format for this case. Taken together, all of this implies a general need for both static files and object stores to be accepted as both input and output formats, with the most efficient formats preferred for the largest-scale processing tasks.

There have been a few developments in applications looking to support object stores like the ROOT I/O system RNtuple~\cite{rntuple} or XRootD framework. However a wider shift to object stores would require a dedicated working group to assess the pros and cons of integrating them into experiment software. In the future it could be possible to access large scale object stores directly from the analysis software while continuing to support POSIX-like features for smaller more interactive tasks.

Whatever architecture an AF may eventually have (distributed, replicated at T1s, integrated with national resources) a storage system that is accessible from everywhere and can be easily replicated should be the aim. Different sites may have their own solutions, and often a different type of storage for different resources with different protocols, this poses a challenge for users to migrate to other facilities or share their data. Currently this role is filled by the grid storage which is not designed for user experience, particularly without going through the experiments’ DDM.

\subsection*{Integration of namespaces and access methods}

As expressed in the user requirements, users want the ability to move from one resource to another. This implies the ability to share the data as well as the code with colleagues. For this   a possible solution may be for facilities to offer both local storage, that computer nodes can read and write to and from efficiently, as well as storage whose contents will be registered in the wider DDM infrastructure. While DDM tools such as Rucio support user uploads, their current main use is the upload of output data from WLCG jobs, so it would be necessary to extend this to analysis facility output datasets.

For many analysis activities users currently use file systems available at well-known sites, such as EOS, to share data. Within the HSF AF Forum the idea of a global and shareable working space, has been discussed. However, it would be necessary for such infrastructure to be extendable to GB-TB scale datasets that may be produced during analysis. There is no obvious scalable and performant solution here. We cannot resort to distributed file systems like AFS for performance reasons, sharing services like CERNbox~\cite{cernbox} are not suited for large data sharing  and “common spaces” on grid storages need easier interfaces for the users.

There are all sorts of details to be ironed out to support storage systems with different namespaces and access methods. Whether talking about POSIX vs object stores, grid storage vs local storage, different access methods at the same site or integrating a WEB based service, like CERNbox, the problems are similar. It will require work to map each solution to the access of another, for example how CERN integrates SSO and Kerberos for Binder~\cite{binder}. Some solutions may work only for fuse mountable protocols (EOS/XRootD) and not for filesystems with kernel drivers. There is also the option adopted by LHCb to encourage users to use their Analysis Production Data package (apd)~\cite{apd} tools to find data also locally which ATLAS and CMS could do using Rucio.   Alternatively, moving away from POSIX has been proposed, but this clashes with the fact that POSIX is at the base of any Linux server and to avoid it we should forbid users from interacting with file systems, which is not feasible. Going forward if more cloud technologies are introduced it will be easier to integrate object stores from the AAI point of view but also from the infrastructure point of view. For example, whereas a broken POSIX/fuse mount is very intrusive to workloads, object stores are more loosely coupled and make it easier to recover from issues with the backends. Regarding integration with containers, whereas POSIX and fuse often require additional components and even additional privileges, object stores, being native cloud storage, integrate very well.

\subsection*{CERNBox data and document sharing}

File sharing services like CERNBox, heavily used for data and document sharing, can benefit from the fact that the output data of jobs run at the CERN facility can be written on a dedicated EOS space, making them automatically available to colleagues. CERNBox has other functionalities, such as integration with Zenodo that makes it really interesting as a document sharing service. Another useful workflow that CERNBox enables is to combine online work at the AF with offline work in a personal computer, and have the files synchronized via the CERNBox sync client installed on the machine. The CERNbox team is currently extending support beyond EOS to CephFS which will make the service deployable at more sites. To make it really as useful as the CERN instance though, the underlying storage should allow users to write with standard protocols data that can be shared via CERNbox. Of course other labs may have a file sharing service but again we face the access problem: an external person cannot use it without an account or a token and questions about interoperability need to be considered; furthermore such share-and-sync services usually can provide only modest storage with limited integration, making it mostly applicable to small, final outputs.

\subsection*{Data transformation services}

An important step of an analysis is to transform data from one format to another or to create a custom selection of the data. Examples include: physics groups produce ntuples and then users can further refine the event selections; users developing new ML algorithms and parallelized workflows may want to transform ROOT files to other formats developed outside of HEP for ease of use with other tools; users may need to augment existing formats with custom information. Usually these steps are carried out in isolation, but a logical goal is to simplify and reduce the number of steps required, particularly if they have to be carried out on different resources, i.e. grid first and then local resources and require to register and manage transformed copies and take more storage.  Transformation services, such as ServiceX~\cite{servicex}, have been introduced to give users the ability to transform data on the fly. Such services need to be fully integrated with performant storage and caches to serve the data in a timely manner and could be prime consumers of the object stores mentioned in the previous section. This introduces a dependency on new intermediate services that need to serve multiple users and cache intermediate data for further re-use. The efficiency of these intermediate services are being tested to understand if they are a viable solution.

\subsection*{Open Questions}

\subsubsection*{On data “locality” at facilities}

Should analysts expect the data, particularly reduced formats, to be local to any facility they wish to use (thus providing low latency access)? Often analysts produce derived datasets (with extra cuts, derived variables) and work with them, does the same apply?

\subsubsection*{On POSIX file access?}
Is POSIX required? Maybe just interacting with an object store via e..g XRootD / https / analysis software is sufficient? How much work is required to fully support object stores? Users like filesystem-like semantics, but what part of POSIX is really needed and can we decouple mass storage access from more interactive, smaller scale activities?

\subsubsection*{On user interaction with Distributed Data Management systems}
Can all Distributed Data Management queries be hidden from the user and is this desirable? Should users expect that this is managed for them?

\subsubsection*{On provenance of intermediate data products}
Which intermediate files need to be promoted from local to global storage so that users can run at different sites and how is this declared by the user?

\subsubsection*{On common file sharing services}
Do we need a common file sharing service to help users share their files? In which case can any existing services fulfill this purpose?

### 2025-05-14 HSF-TN/2024-01/latex/introduction.tex
Add LaTeX source

\section{Introduction}\label{introduction}

In the HL-LHC era (to begin in 2029) LHC analysts, depending on the type of analysis they do, may have to process up to an order of magnitude more data than currently. At the same time, there are significant changes to analysis techniques such as the introduction of columnar analysis, parallelised data access and the organization of multi-step workflows into pipelines, some of which may be better suited to run using cloud technologies and heterogeneous resources. Consequently, the analysis infrastructure must evolve. This is an active area of R\&D collectively referred to as “Analysis Facilities”.

An Analysis Facility (from here on abbreviated to AF) can be defined as infrastructure and services that provide integrated data, software and computational resources to execute one or more elements of an analysis workflow. These resources are shared among members of a Virtual Organization (VO) and supported by the organization. This R\&D is expected to be integrated with the existing infrastructure and guide any future development.

Since discussions on the evolution of analysis for the HL-LHC started, the analysis infrastructure has often been considered as something separate from the current grid infrastructure, mainly due to two aspects: large scale interactivity for fast analysis cycles and integration of cloud native technologies. The possible additional requirement to support whole VOs also sets it apart from the local Tier 3 sites currently giving access only to local users and occasionally to external collaborators. The nature of AFs has provoked long debates. CERN satisfies the requirements described above but it is the only site supporting all users and the competition for resources there is already quite intense. The question is how to define an AF that offers similar functionalities as those at CERN without hindering participation from other sites. Can they support a subset of functionalities or be open to groups of users and still be considered an AF?

We can start by stating that the most common setup for analysis (small interactive machines, a batch system and a shared file system between the two) will not be replaced and we should focus on a technology refresh. The effort is to complement and add new tools to what already exists.  A second observation is that the distinction between AFs and the Worldwide LHC Computing Grid (WLCG)~\cite{wlcg} is more blurred than currently appreciated and the integration of AFs within existing WLCG infrastructure is important for broad accessibility and a standardized user experience with smooth data access, dynamic allocation of WLCG computing as part of the AF resources and easier support from the sites. A third observation is that several analysis facility sites support multiple experiments, for instance CERN, FNAL and DESY, and the aim is to provide the same infrastructure for most if not all experiments which will not necessarily be only WLCG or even HEP. Interdisciplinarity will have to be taken in consideration when planning the AFs. In other words AFs do not exist in a vacuum and should not be considered isolated and one should not assume the resources are dedicated. The way forward is not to focus on a specific architecture, but on the building blocks that can be deployed whatever shape an AF takes. For these reasons we should talk about “analysis infrastructure” more than “analysis facilities”.

In this spirit, following the Analysis Ecosystem II workshop and the Analysis Facilities Forum presentations, a number of key areas were identified which this white paper addresses:

\begin{itemize}
    \item \textbf{Users' perspective and their use cases} (Section~\ref{userperspective}): this should drive the R\&D and naturally will be discussed first.
    \item \textbf{Compute resources access and provisioning} (Section~\ref{computeresources}): considers the evolution of computing requirements, in particular the concept of scale out resources and integration on existing infrastructures.
    \item \textbf{Data Organisation, Management, Access (DOMA)} (Section~\ref{doma}): concerns the input/output data for analysis workflows. In particular the need to integrate AFs with grid storage and methods to solve the data locality problem. 
    \item \textbf{Federated Identity Management and and Authentication, Authorization Infrastructure (AAI)} (Section~\ref{aai}): a consistent AAI across analysis infrastructure is at the core of a smooth integration particularly for data access; federated identity can also allow access to multiple facilities. 
    \item \textbf{Accelerator resources} (Section~\ref{accelerators}): considers the increased use of accelerators in analysis and the resulting need to add accelerators to AF resources.
    \item \textbf{Analysis portability and preservation}(Section~\ref{preservation}): handles the requirements on the packaging of software to enable collaborative analysis across facilities with the ability to rerun or recast analyses in the future. 
    \item \textbf{Monitoring and Metrics} (Section~\ref{monitoring}): concerns the need for appropriate monitoring and benchmarking both for users to effectively use resources and for sites to provide resources more effectively.
    \item \textbf{End user documentation and support} (Section~\ref{documentation}): expresses the requirement for  findable, quality documentation and support for users as part of the AF infrastructure is essential.

\end{itemize}

This White Paper does not aim to answer all the questions surrounding analysis facilities and infrastructures. We have identified remaining, key open questions for the community and they are highlighted at the end of the relevant sections.

### 2025-05-14 HSF-TN/2024-01/latex/preservation.tex
Add LaTeX source

\section{Analysis portability and preservation}
\label{preservation}

A core requirement of an AF then is providing the user with tools that help them realize their desired software stack and to address the user portability and preservation requirements expressed in section~\ref{userperspective}. There are few methods users can use to build the software stack.

\subsection*{CVMFS}
CVMFS is the read-only distributed file system for global distribution of HEP software, in which software packages are laid out in a directory structure with software tools that can selectively enable a certain combination of packages within a given user session (e.g., LCG views). The user can augment this global set of software through local installations, which may or may not be carried out in the user space. While this is an effective method to centrally maintain and distribute software it has disadvantages: user code built against CVMFS needs runtime access, so will not work on non-CVMFS nodes; and the user code needs to be preserved separately from the CVMFS dependencies, which is important for analysis preservation.

\subsection*{Conda}

Conda~\cite{conda} is a cross-platform and language agnostic package and environment manager, which solves portability between collaborators and is adopted particularly when python external tools are used. It is not particularly designed for preservation but it makes it easy for users to build their code in multiple environments with different sets of dependencies and switch from one to another. To be truly preserved the conda environment needs to be locked to specific versions (not the default) and, in addition, the user’s supplementary code needs to be also preserved.

\subsection*{Linux containers}

Linux containers allow users to assemble the software stack on their own, in a self-contained, preservable format. The OCI (Open Container Initiative) format, supported by Docker~\cite{docker} and others, can be read by any runtime environment and such container registries can be used to distribute images to facilities. If built with fixed dependencies a container does not change and contains all the software it needs to run the application. It should be noted that preserving only the containers may not be sufficient when older versions of OSs ( for example slc5)  do not run anymore on modern hardware --- this remains an area of R\&D.

Conda, containers and CVMFS can be combined. For example LHCb uses conda in combination with CVMFS to distribute and preserve different user setups~\cite{lbconda}. Since it is a package manager it can also be used to build containers instead of the specific container OS package manager. CVMFS is also used to distribute singularity/apptainer~\cite{singularity} images on the grid --- currently CVMFS stores and serves around 3000 custom apptainer images. This could be extended to other runtimes with containerd~\cite{containerd} images serving facilities that are built on top of cloud-computing infrastructure, such as kubernetes. Supporting different runtime images for the same software raises questions about the infrastructure as standard distribution, apptainer images and containerd images would all be duplicated software installations.

The construction of well-maintained software stacks and  images may require expertise from AF and experiment experts and an infrastructure to make it easy for the users. For example, SWAN makes a curated set of software environments available through a dedicated user interface. Projects such as repo2docker~\cite{repotodocker} and Binder abstract the image building process away from the user. Here, the user gives their desired set of requirements in a declarative format which then gets built ad-hoc during a start up phase. For reproducibility and shareability of analyses, users code should be versioned or tagged along with the Git repository and environments (conda or container images) used.

### 2025-05-14 HSF-TN/2024-01/latex/aai.tex
Add LaTeX source

\section{Federated Identity Management and AAI}
\label{aai}

Equitable access to resources in a VO-wide manner is one of the foundations on which the success of the grid is built. Users can access worldwide resources but any one workload is routed to a convenient resource given, e.g. data constraints. This is based on an agreed key piece of Federated Identity Management (FIM) infrastructure: x509 certificates. So far the majority of end user analysis has not required the use of FIM because the resources used are either private ---  provided by universities or national labs --- with their own AAI mechanisms, or are provided by CERN. If there is a significant increase in resources requested, and the need to support all users, FIM will become a key requirement for analysis infrastructure.

While the WLCG FIM model is very successful, the use and maintenance of certificates for users is difficult and incompatible with the use of cloud services and resources. WLCG is working towards an AAI based on Oauth2.0 and tokens which will replace x509 certificates. By HL-LHC the FIM on the grid will be based on the new token infrastructure, which is fully compatible with cloud and web services such as JupyterHub, CERNBox and kubernetes. This will greatly improve the user experience.  It is highly desirable that the analysis infrastructure services be fully integrated with the new AAI. Note that tokens underpin FIM for internet web services and so their integration into analysis frameworks should be easier.

This integration does not imply or necessitate that all analysis resources support all users, but it allows for a uniform access to all resources for authorized users. As expressed in the user requirements, the VOs ultimately can decide on the authorization mode, i.e. if the facilities will be usable by all users, but whatever decision is made the infrastructure should support it. This would also allow private resources to be more easily made accessible to selected colleagues, which would be a site or local group decision. Some AFs in the US are an example of the latter, currently giving access to ATLAS and CMS users depending on their affiliation.

It is important to underline that currently a Oauth2.0 based AAI is not compatible with a classical SSH/POSIX/batch system infrastructure and development work is necessary to integrate them.

### 2025-05-14 HSF-TN/2024-01/latex/computeresources.tex
Add LaTeX source

\section{Compute resources: access and provisioning}
\label{computeresources}

End-user data analysis workflows are evolving significantly, with new data formats designed by the LHC collaborations to reduce the storage footprint and be more compatible with modern data science paradigms, eventually enabling researchers to utilize frameworks that require little to no experiment-specific software. This evolution will have an impact on the infrastructure and resources dedicated to user analysis activity.

Outlined are the current R\&D activities that have been reported to the HSF AF forum, with particular focus on the evolution of access and provisioning of the compute resources to expose heterogeneous and possibly distributed resources. A set of recurrent questions and discussions are the starting point for the review targeted by this section: 
\begin{enumerate}
    \item \textbf{Compute resources access:}
    \begin{enumerate}[label=(\alph*)]
        \item Workflow management: tools leveraging batch systems in a native way are arguably the most widely adopted ones at the time of writing, nevertheless workflows are becoming larger and more complicated and the integration of cloud paradigms for a seamless transition from local “interactive” development to full scale analysis multi-step execution are emerging in many R\&D activities. That is expected to enable enhanced data exploration and reduce “time-to-insight”.
        \item UX/UI: ssh-ing into a preconfigured machine is a de-facto standard for most analysis and that is not expected to change. Instead, a possible enhancement of the UX is expected to come from WebUI interfaces largely adopted in the context of cloud/notebook based data science ecosystems.
        \item 
        AAI: the adoption of tokens for Authentication and Authorization by the HEP community is also important in the context of AFs and integration of cloud technologies as well as simplified access to data from diverse resources. Section~\ref{aai} is dedicated to AAI infrastructure.
        \item Containers: a base hypothesis of the computing models under investigation is that the payloads are containerized. Section~\ref{preservation} details the portability and reproducibility of that approach.
    \end{enumerate}
    \item \textbf{Compute resources provisioning:}
    \begin{enumerate}[label=(\alph*)]
        \item Provisioning responsiveness: point 1.a shows a significant shift in terms of the latency that is acceptable in the interactive vs. non-interactive scenarios. That opens the question of how to implement a computing model at AF level where dynamic provisioning can possibly live alongside crowded traditional batch systems.
        \item Efficiency and performance: upcoming frameworks that are enabling a whole set of new possibilities from the user perspective have to be validated against their efficiency in terms of resource utilization and overall physics throughput. 
        \item Site provisioning model: HPC centers and commercial clouds are going to have a significant impact on the future of resource provisioning. Finding a way to integrate them seamlessly is crucial, along with the understanding of which kind of analysis paradigm can benefit the most from that integration. 
        \item Resource provisioning model: the server level is the other side of the spectrum of the provisioning model. One solution could be to allocate whole nodes and reserve a whole “fat” node that is fully committed to achieve the quickest analysis turnaround time. It remains to be seen if this can be sustainable in the future and compatible with the goal of maximum overall resource efficiency.
        \item Specialized hardware: the adoption of GPUs and, potentially, FPGA acceleration, is expected to increase, and this has to be made both easy and efficient from the infrastructure perspective. Although not many case studies have been reported, we do expect this to be a strategic point to investigate. This is transversal to both access and provisioning patterns. 
    \end{enumerate}
\end{enumerate}

Many R\&Ds reported to the AF Forum on their investigations, which allows an “educated guess” on what the main traits are for future analysis infrastructures.

\subsection*{User Interfaces}

The SSH login into a pre-configured UI still holds as a requirement for many use cases, especially for non-pythonic workflows like scripting for batch job submission. For analysis of reduced data (at the moment often done in the previous step), an interactive interface can be accessed through the adoption of tools enabling a single WebUI to log in and access the resources they need. JupyterHub is a popular choice for this, because it allows multiple users to spawn JupyterLab instances on a cluster of servers. One interesting model emerged for point 2.a: providing users with a relatively small amount of “always-on” resources, ready to get the user started with an interactive workflow, while scheduling a larger fraction on lazier batch system queues, leveraging offloading mechanisms (see below). Another area of investigation is the possibility of using JupyterHub as a mechanism for ssh-UI on-demand provisioning, integrating the SSH access into a running JupyterLab instance, thus enabling users to share a common environment with access to different UIs.

\subsection*{Offloading to external resources}
In terms of supporting the offloading of “heavy lifting” workflows to external resources, and since the possibility to distribute batch jobs is still a requirement, most of the reported activities aim to integrate a mechanism for leveraging the batch systems as an offloading backend for the more “interactive”/data-science-like analysis workflows. That behavior has been demonstrated mostly via Dask cluster deployments. In fact, many of the presented solutions have their foundation in Dask distributing payloads on either HTCondor~\cite{htcondor} or SLURM~\cite{slurm}. Some sites have deployed Apache Spark~\cite{spark} managed clusters, notably CERN Analytix cluster. For the requirements of fast prototyping of ML models and even full analyses, an offloading mechanism for a whole JupyterLab instance to a remote, specialized node that is fully committed to the analysis for a short period of time could be considered. Although possible, it has yet to be proven sustainable at scale with dedicated activities. With all of these considerations in mind,  figure~\ref{fig:AFschema} attempts to schematize a high level description of an AF resource management “strawman”.

Different models for the evolution of the access and provisioning of analysis resources have been presented by several groups (SWAN~\cite{swan}, Coffea Casa~\cite{coffea}, NAF~\cite{NAF}). Most of the commonalities are in the methods of providing access to the users’ UIs. Providing SSH and/or JupyterHub access is becoming a default mode of operations in many (if not all) the presented activities. Interesting variations occur for the scale-out mechanism toward remote resources. A number of facilities have the batch resources co-located (in terms of network segregation) with the AF interactive seed, differing only in the integration with the analysis tools. On a slightly different path, there is an evaluation for integrating geo-distributed centers (e.g.,T2 grid resources) as offload candidates for tasks (INFN~\cite{INFN_distributed}); this presents the problem of having the JupyterLab instance on a different network from the worker nodes. The current solution in such a scenario is the creation of a self-contained cluster to be created at the remote site on-demand.

\begin{figure}
    \centering
    \includegraphics[height=6cm]{fig/schema.png}
    \caption{High level schema of the recurrent patterns reported at the forum by R\&D activities around compute access and provisioning models evolution.}
    \label{fig:AFschema}
\end{figure}

One final observation is that there are ongoing efforts, presented by the kubernetes~\cite{k8s} user activities, that might allow for an alternative implementation. Here, payloads and their distribution can eventually be fully delegated to kubernetes native resources with the introduction of dedicated abstractions and scaling improvements. There are preliminary implementations and R\&D efforts that have been reported at the forum but have yet to be investigated properly in order to make a clear comparison with the alternatives presented above.

To support complicated workflow pipelines, as expressed in the user requirements, workflow languages such as Snakemake~\cite{snakemake}, Yadage~\cite{yadage}, Common Workflow Language (CWL)~\cite{cwl} and others may be suitable. Also, some workload schedulers have features that allow users to express dependencies between workloads (e.g. HTCondor's tool DAGMan). Workflow services such as REANA~\cite{reana}, capable of processing such languages, should be integrated into the existing batch infrastructure at the facility in order to have access to the same data and compute resources that would be available if the pipeline were processed manually. Note that REANA also works with a kubernetes back end.

\subsection*{Open Questions}

\subsubsection*{On resource interoperability?} If we cannot guarantee uniformity of technologies how do we manage “interoperability” between resource providers?

\subsubsection*{On a common interface}

Given the large number of new technologies should users be exposed to all of these or should there be a common entry point interface such as JupyterLab? Reaching agreement on a common interface will be challenging, the introduction of Jupyter Notebooks/JupyterLab is an attempt at having a common interface but this remains an open question.

### 2025-05-14 HSF-TN/2024-01/latex/documentation.tex
Add LaTeX source

\section{End user documentation}
\label{documentation}

The ability for all users, regardless of technical capability, to use and profit from AFs is crucial. The necessary developments in analysis techniques and infrastructure must therefore be undertaken in collaboration with analysts to keep the “barrier-to-entry” on using new analysis facilities/infrastructures as low as possible. Analysts will expect comprehensive documentation and, more importantly, tutorials on how to run analyses that are similar to theirs on the facility - as an example the IRIS-HEP Analysis Grand Challenge has complete analysis workflows, including the configuration of facilities, using open-source code and CERN Open Data that can serve as valuable instructional resources. Tutorials for analysis specific tasks that illustrate how to most effectively use AFs will also be required. These tutorials should be regularly tested on the facilities (perhaps as part of a periodic testing schedule). We can target experiments’ training and onboarding initiatives such as the LHCb Starterkit and the CMS Data Analysis School, and enlist the help of the HSF training working group~\cite{hsftraining}. These initiatives/groups can provide dedicated lessons with hands-on tutorials to train early career researchers on these infrastructures from the outset --- these initiatives are documented in~\cite{Hageboeck:2023kcb}. The findability, and usability of this documentation and instructional material should be considered an important user metric. It is clear that such educational material needs to rely on a degree of uniformity between the user interfaces --- hiding, as much as possible, the differences between facilities --- to avoid filling the documentation with caveats.

End-user support can be provided both by the facility and by peers. Forums such as Slack, Mattermost and Discord channels provide vital, realtime, peer-to-peer support among countless communities and could be utilized for end-user support on a facility. The facility should endorse such a channel's existence and direct users to it as an official source of support.

### 2025-05-14 HSF-TN/2024-01/latex/monitoring.tex
Add LaTeX source

\section{Monitoring and Metrics}
\label{monitoring}

To provide an extensive overview on how resources are used in order to guide infrastructure development and to allow users to make an informed decision about which infrastructure to use, key parameters should be published for each AF.

Metrics for AFs can be categorized into four areas

\begin{enumerate}
    \item User experience metrics
    \item User trend metrics
    \item Performance metrics
    \item Facility metrics
\end{enumerate}

User experience metrics relate to the analysts’ impression of the facility. Support for users is a key component of the analyst experience. Such support metrics include time to acknowledge tickets and, as a separate metric, time to resolve tickets. Unlike time to acknowledge tickets, time to resolve tickets is very case specific, for instance, if a ticket requests or requires the implementation of new functionality it will often take considerable time (and not all requests are even reasonable!). Over an extended time period, and with some care, this can be compared between sites. Once new functionality or features are added to a facility the rate of uptake by users is a useful metric to gauge user engagement and is closely related to the findability and usability of documentation and tutorials, which are key for user on-boarding and retention (see also section~\ref{documentation}). User satisfaction surveys can gather feedback directly from users (albeit response rates can be low).

User trend metrics can be used to inform future infrastructure development and act as a “live survey” of user analysis requirements. This could include metrics on the data source being used (local or cached from a remote site or "DataLake"), input data format, resources employed per job and interactive session length which would impact parameters of the caching system. In an idealized scenario this would mean a common API across software tools used at the analysis infrastructures, but this cannot realistically be imposed.  Tools like prmon~\cite{prmon}, or in general any code profiler that can be run within the user job, are well tested and can easily collect resource usage information while the jobs run and are not tied to any specific software. Users cannot be expected to run this monitoring on their own, so it requires automation from the VOs and/or the sites. When facilities run these tools, wrapping the jobs automatically, they should also set up a visualization infrastructure for the users to track their jobs. This happens on the grid where the WFMS of the experiments have robust monitoring of every job and metrics are then aggregated to give a wide view of the resource usage by site or globally. Another example of this would be the Grafana dashboard at CERN, which displays some information on the user batch jobs. However, it does not cover other resources (cloud or interactive) or many desirable metrics. Of course tracking basic machine level or batch system resources can be done almost automatically. However for tracking each application, what it does and what data it uses there is no other option but to push the users to use an instrumented framework which collects all this extra information.

The concept of dark analysis was raised at the Analysis Ecosystem Workshop II and refers to the analysis that users run privately on, for instance, university or laboratory resources. We currently have no handle on the type and scale of this analysis and therefore a user survey was proposed to analyze these user trends. This kind of large scale survey must be treated carefully. The survey could be conducted through the HSF Data Analysis working group and would give further insight into the analysis needs of users across experiments. If we had instrumented AF the problem of dark analysis would be reduced.

Whilst all current facilities have different focuses, and this is important, there are universal aspects of an analysis pipeline that can serve as benchmarks. These can be used to assess analysis facility performance - a performance metric.  The IRIS-HEP Analysis Grand Challenge (AGC)~\cite{agc} has established complete analysis workflows using CERN Open Data that can be used to test the functionality, integration and performance of an AF. These workflows can be factorized to benchmark performance in specific areas, for instance, data delivery. They also serve as useful tutorials for analysts, showcasing complete, realistic end user analyses that exploit the tools discussed in this document as well as the scientific python ecosystem.

Publication of facility metrics is considered a necessary aspect. Facility metrics quantify the technical performance of the facility rather than the analysis performance. Resource idle time, job failure rate and downtime should all be monitored by the facility host.

Increasingly, sustainability metrics will be assessed, scrutinized and perform an important role in evaluating facilities’ viability. For instance, one could consider the carbon footprint per user/job. As reported at the 26th International Conference on Computing in High Energy \& Nuclear Physics (CHEP 2023) the environmental footprint of HEP research overall can be controlled via modernization of the facilities we use to run jobs, improvements in the software and computing models of the experiments and improvement in hardware technologies. AFs have the opportunity to influence two of these factors. This will be carefully balanced against operational cost per user/job where the weighting of each of these (potentially conflicting) optimisations will be decided by funding agencies. Facilities must provide information for both and meaningful metrics will evolve over time for all sites, not just AFs.

Infrastructure status reports would be important to track the AF efficiency, but it is a whole project in itself and would take a fair amount of operational overhead to maintain, correct, ensure the system reports fairly, etc. Facilities should certainly track metrics internally as a first step. All analysis infrastructures should define key parameters to measure success over time and a level of coordination on these parameters across infrastructures should be implemented to make fair comparisons. As a minimal set we propose the following as an evolution over time: total number of active users, total compute time spent both on local and on batch resources, storage space available and used, I/O rates, Time-To-Completion and data source (local, cached, remote).

### 2025-05-14 HSF-TN/2024-01/latex/userperspective.tex
Add LaTeX source

\section{User perspective}
\label{userperspective}

Analysis Facilities aim to facilitate the preparation and execution of HEP data analyses. During the course of the HSF Analysis Facility Forum a number of requirements have been identified in order to enable users to do their research effectively.

\subsection*{Ability to perform fast research iterations on large datasets interactively}

Research success relies crucially on the ability to test ideas during the R\&D phase of an analysis, e.g. assessing the impact of a change in event selection or determining the impact of a variation in systematics. Here a fast “time-to-insight”, i.e. minimal wall-clock time, is a priority. In this phase, the user expects to rapidly gain access to and interactively work with the resources. This requirement so far has been satisfied by classic facilities that provide a number of interactive, single “login” nodes accessible via remote terminal/ssh sessions or, more recently, by providing small single node Jupyter~\cite{jupyter} notebooks.

However dataset sizes are increasing and an interactive experience constrained to a single node with a handful of resources may no longer be sufficient. The requirement becomes to enable a distributed and interactive analysis style, where the work is coordinated by the user in an interactive session, but the user can dynamically connect to a scale-out system that provides additional resources. The expectation is that the processing on additional resources starts in parallel almost immediately, with only a short lag-time due to resource provisioning, and the resulting data is aggregated and sent back to the interactive node for the user to examine in real-time.

What constitutes “minimal wall-clock time”, “large amount of resources” and “large datasets” is analysis dependent and experiments should define acceptable/expected ranges for their analyses. The implications on how interactivity could be achieved are still under discussion.

\subsection*{Ability to convert interactive to batch-schedulable workloads}

When an analysis matures from an R\&D phase to performing the final measurement, the usage pattern often changes from a rapid iteration and frequent changes of the analysis to a more controlled pattern, in which results from various configurations must be produced. Here users may not urgently require full interactivity and, in fact, may prefer to prepare a number of analysis configurations ahead of time and asynchronously schedule them in bulk. In this mode, the total wall-clock time required for a processing is not an urgent concern to the user and facility-side optimizations towards resource use efficiency, load balancing, may be valid trade-offs. This mode is typically handled as a non-interactive batch workload, as supported by a workload scheduling system.

As more interactive distributed workloads become possible for the R\&D phase, it is important that it is easy to convert a workload from an interactive setup into a “headless” mode that is amenable to batch systems. This poses requirements on both the analysis software as well as the analysis facility. This of course is mostly important in case of Jupyter notebooks where the submission procedures to a batch system need to be developed by the AFs. A Jupyter notebook that, in interactive mode, dynamically scales to multiple nodes to execute an analysis and is driven by a user working within a notebook, should be re-configurable into a non-interactive workload that requests a fixed number of workers, when they become available, and scales down workers reliably to free resources again.

\subsection*{Ability to interact with the WLCG and scale outside of the facility on occasion}

Given the distributed nature of HEP research and the existence of the WLCG, analyzers often face the need to interact with the wider infrastructure. Common tasks may include interacting with other WLCG and other opportunistic resources, pulling or pushing data in and out of the facility. As such, AFs should not be “sealed environments”, but integrated into the wider WLCG infrastructure.  If the workload would exceed traditional analysis activities and/or the data is distributed across the WLCG and the task at hand matches well to grid-style processing the user should be able to use grid middle-ware tools to offload the work to the WLCG and connect to opportunistic resources. The latter may be particularly important for training large machine learning models for which AFs may not have sufficient GPUs. The ability to convert interactive workflows to batch-like jobs (as described above) should also help with this use case.

\subsection*{Ability to efficiently train machine learning models for HEP}

The progress over the last decade in Machine Learning (ML) techniques invariably also affects physics analysis in HEP, where it is used across the board in data-taking, simulation, reconstruction and analysis. ML methods play a dual role: they can be used to find more powerful algorithms, but also enable significant acceleration due to the high degree of parallelization possible in neural networks, which maps well to hardware such as GPUs. For AFs this implies two important requirements: 1) the need for an efficient python-based analysis environment and 2) access to heterogeneous hardware both interactively as well as in batch mode. The first is due to the dominance of python within ML research with tools like PyTorch~\cite{pytorch}, SciKit-Learn~\cite{scikit-learn}, and Tensorflow~\cite{tensorflow}, which is expected to continue in the short to medium term. ML development also requires early on data- and compute-intensive fast iteration during the development of the model, where interactive access to, e.g. Jupyter notebooks is a priority. The second requirement of accelerator availability must be met both interactively as well as non-interactively. As a user, it is expected to be able to start an interactive session with GPU access. For a positive user experience it may be more important to be able to quickly access a single GPU than to be able to keep it for a long time or scale from a single GPU to a multi-GPU environment. Access to batch GPUs is increasingly required for model selection and hyperparameter tunings but also for cutting-edge ML research, because both models and datasets are growing, and a typical training run for many models now exceeds 24 hours. Here the user should be able to use standard workload schedulers to describe the shape of the workload (e.g. GPUs, CPUs, memory).

\subsection*{Ability to reproducibly instantiate desired software stack}

Analyses are a mixture of both collaboration-wide software frameworks and more custom-tailored packages that suit the particular analysis needs of the team. As a user, it is reasonable to expect the ability to assemble this mixture. As an additional aspect, the ability to archive and possibly distribute this particular software environment is important. This may be of particular interest for both analysis preservation and reuse purposes, but also help towards the ability to reinstantiate the analysis environment at different facilities. CVMFS~\cite{cvmfs} --- the dominant solution in HEP for software distribution --- as well as Linux containers are expected to contribute here. Users may not be experts in these technologies and thus the process of listing requirements for the software environment may need to be abstracted from the underlying technology.

\subsection*{Ability to collaborate in a multi-organisational team on a single resource}

Analysis teams at HEP experiments are multi-organisational: team members may not be at the same institution, region or even country, yet the goal is to jointly work towards a physics result. Analysis is also distributed: responsibility for parts of the analysis workflow is often taken on by a subgroup and the resulting data products of one subgroup must be seamlessly passed onto another one. Such a requirement is easiest to fulfill if access to resources are equitable, where the full team can work with the same resources. The main reason for this is that maintaining copies of data, i.e. downloading datasets, ensuring all files are present, doing necessary post-processing/combining/bookkeeping on the local files before they're ready for analysis is hard work for the users, inefficient and error-prone. Having one copy means this only needs to be done once. In the current model this is ntuples, but this could also refer to skims of reduced formats in the future.

Given the data volumes of an analysis it may be desirable to work on the same resource for the duration of an analysis project, however, fine-grained control over which resource this would be may not be necessary and may be left up to the VO, i.e. the VO may assign users or physics groups to AFs rather than users flocking to a facility or another.

\subsection*{Ability to move analyses to new facilities}

While it should be possible for the full analysis team to access the same facility for jointly developing the analysis, users and/or system administrators may value the ability to move an analysis effort between facilities. This implies the requirement that the data products that are local to a single facility can be replicated to distributed storage. Note that this does not imply a requirement for a distributed, fast, automatic synchronization but rather an on-demand method of data movement.

\subsection*{Ability to efficiently access collaboration data as well as make intermediate data products available to the team}

Analysis starts after a common bulk pre-processing of the data by the collaboration on WLCG resources. It is therefore necessary to be able to interact from within the facility with distributed data management services, such as Rucio~\cite{rucio}, both from interactive and non-interactive environments. Notably, the physical presence of the data on the processing site is not a priority of the user as long as a smooth processing of the data can be ensured on-demand. This opens the possibility of inbound data-access via caching services such as XCache~\cite{xcache}, where the user only references a global identifier and the file is made available on-demand.

During user data processing new data products are being created which need to be made available within the facilities’ distributed storage. Additionally it is important that users are able to share data to other collaborators (possibly working on other facilities). Currently this is often satisfied through an ad-hoc choice of a storage solution that many collaborators have access to (e.g. EOS~\cite{eos}) as a “data exchange”. Ideally, however, it would be possible to register user data seamlessly in a global namespace. It may be sufficient that such data publishing is based on explicit intent and full POSIX semantics may not be needed. Thus an implementation of such a requirement does not imply a global, automatically synchronized filesystem (such as AFS that provides networked file storage for CERN users).

\subsection*{Ability to express interdependent distributed computations at small and large scales}

An analysis at LHC scale is often the result of a complex workflow of multiple interdependent steps that need to be executed in a certain order: for example data reduction/event selection, followed by preparation of the inputs required for statistical analysis, followed by inference and visualization steps. Here, it is often the case that due to the distributed nature of analysis teams, steps are largely independently developed by analysis subgroups and later on integrated into an overall data analysis pipeline. The individual steps may have different, and even conflicting, software environment requirements. This implies users do not just need to set up one environment, but ideally seamlessly transition between many. Nonetheless, it is desirable for analyzers to be able to express and schedule the full data analysis pipeline into an overall workflow.

Workflows in this sense should be contrasted to computational graphs that are, e.g. built during the use of a distributed data processor, such as Dask~\cite{dask}, as each node in the former may be a full analysis step (e.g., running the fit), while the latter typically considers nodes to be small computational units (e.g., compute invariant mass). Such workflows have been used both for analysis reinterpretation work, where the full analysis pipeline is re-executed on new simulation data, as well as for automation of, e.g. the production of calibration functions for analysis.

\subsection*{Open Questions}

\subsubsection*{On interactivity of analysis sessions}
What is the user expectation of an interactive session? Interactive analysis sessions on large datasets implies the ability to scale out to distributed resources and users will expect instant access to such an analysis session. What are the required disk resources, computing cores (order 100s) and network capabilities to realize this? The answers to these questions are analysis/workflow dependent, vary widely, and will need to be quantified as such. The exact definition of ``interactive analysis" remains an open question for analysis in the HL-LHC era, where datasets will be much larger (~10x) for ATLAS and CMS from Run 4 onwards and will also grow again for LHCb and ALICE from Run 5. This does not automatically mean that interactive needs jump by 10x, but an increase will happen and should be quantified.

\subsubsection*{On the Interfaces to resources provided}
Analysis work is multi-faceted and while a browser-based interface can be attractive for some use-cases such as interactive computing with notebooks, it should not be the only possible interface to the resource. Should users expect to be able to connect via their preferred text-based terminals or Integrated Development Environments (IDEs) such as Microsoft's Visual Studio Code?

### 2025-05-14 HSF-TN/2024-01/latex/whitepaper.tex
Add LaTeX source

\documentclass{article}
\usepackage[english]{babel}
\usepackage[letterpaper,top=3cm,bottom=3cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{color}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{authblk}

\usepackage{enumitem}

%%Specific defs
\usepackage{xspace} 
\newcommand{\red}{\color{red}}
\def\needref{{\textcolor{red}{[ref]}}\xspace}

\newcommand{\covid}{COVID-19\xspace}
\def\cpp{{\tt C++}\xspace}

\usepackage[sorting=none,style=numeric-comp]{biblatex}
\addbibresource{whitepaper.bib}

\begin{document}

\begin{titlepage}

\vspace*{-1.5cm}
\centerline{\large THE HEP SOFTWARE FOUNDATION (HSF)}
\vspace*{1.5cm}
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}

\\
 & & HSF-TN-2024-01 \\  % ID
 & & April 2024 \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

\vspace*{.5cm}

% Title --------------------------------------------------
{\bf\boldmath\huge
\begin{center}
  Analysis Facilities for the HL-LHC White Paper
\end{center}
}

\vspace*{1.0cm}

% Authors -------------------------------------------------
\begin{center}

%58 authors at 33 institutes!

D. Ciangottini$^{1, \rm b}$,
A. Forti$^{2, \rm b}$,
L. Heinrich$^{3, \rm b}$,
N. Skidmore$^{4, \rm b}$,

C. Alpigiani$^{5}$,
M. Aly,$^{2}$,
D. Benjamin$^{6}$,
B. Bockelman$^{7}$,
L. Bryant$^{8}$,
J. Catmore$^{9}$,
M. D'Alfonso$^{35}$,
A. Delgado Peris$^{17}$,
C. Doglioni$^{2}$,
G. Duckeck$^{10}$,
P. Elmer$^{11}$,
J. Eschle$^{12}$,
M. Feickert$^{13}$,
J. Frost$^{14}$,
R. Gardner$^{8}$,
V. Garonne$^{6}$,
M. Giffels$^{36}$,
J. Gooding$^{15}$,
E. Gramstad$^{9}$,
L. Gray$^{16}$,
B. Hegner$^{31}$,
A. Held$^{13}$,
J. Hernández$^{17}$,
B. Holzman$^{16}$,
F. Hu$^{8}$,
B. Jashal$^{18,19}$,
D. Kondratyev$^{20}$,
E. Kourlitis$^{3}$,
L. Kreczko$^{21}$,
I. Krommydas$^{22}$,
T. Kuhr$^{10}$,
E. Lancon$^{6}$,
C. Lange$^{23}$,
D. Lange$^{11}$, 
J. Lange$^{34}$, 
P. Lenzi$^{1}$,
T. Linden$^{24}$,
V. Martinez Outschoorn$^{27}$,
S. McKee$^{25}$,
J. F. Molina$^{17}$,
M. Neubauer$^{26}$,
A. Novak$^{35}$,
I. Osborne$^{11}$,
F. Ould-Saada$^{9}$,
A. P. Pages$^{28}$,
K. Pedro$^{16}$,
A. Perez-Calero Yzquierdo$^{17}$,
S. Piperov$^{20}$,
J. Pivarski$^{11}$,
E. Rodrigues$^{29}$,
N. Sahoo$^{30}$,
A. Sciaba$^{31}$,
M. Schulz$^{31}$,
L. Sexton-Kennedy$^{16}$,
O. Shadura$^{32}$,
T. \v{S}imko$^{31}$,
N. Smith$^{16}$,
D. Spiga$^{1}$,
G. Stark$^{33}$,
G. A. Stewart$^{31}$,
I. Vukotic$^{8}$,
G. Watts$^{5}$,

\bigskip
{\it\footnotesize
$^b$Editor,
$^1$INFN,
$^2$University of Manchester,
$^3$Technische Universität München,
$^4$University of Warwick
$^5$University of Washington
$^6$Brookhaven National Laboratory
$^{7}$Morgridge Institute for Research
$^{8}$University of Chicago
$^{9}$University of Oslo
$^{10}$Ludwigs-Maximilians Universität München
$^{11}$Princeton University
$^{12}$Syracuse University
$^{13}$University of Wisconsin-Madison
$^{14}$University of Oxford
$^{15}$Technische Universität Dortmund
$^{16}$FNAL
$^{17}$Centro de Investigaciones Energéticas, Medioambientales y Tecnológicas
$^{18}$Instituto de Física Corpuscular, Valencia
$^{19}$Tata Institute of Fundamental Research Mumbai
$^{20}$Purdue University
$^{21}$University of Bristol
$^{22}$Rice University
$^{23}$Paul Scherrer Institute
$^{24}$Helsinki Institute of Physics
$^{25}$University of Michigan
$^{26}$University of Illinois at Urbana-Champaign
$^{27}$University of Massachusetts Amherst
$^{28}$IFAE
$^{29}$University of Liverpool
$^{30}$University of Birmingham
$^{31}$CERN
$^{32}$University of Nebraska-Lincoln
$^{33}$SCIPP UC Santa Cruz
$^{34}$University of Hamburg
$^{35}$Massachussets Institute of Technology
$^{36}$Karlsruhe Institute of Technology
}
\end{center}

\vspace*{1.0cm}

% Abstract -----------------------------------------------
\begin{abstract}
  \noindent

This white paper presents the current status of the R\&D for Analysis Facilities (AFs) and attempts to summarize the views on the future direction of these facilities. These views have been collected through the High Energy Physics (HEP) Software Foundation’s (HSF) Analysis Facilities forum~\cite{hsfaff}, established in March 2022, the Analysis Ecosystems II workshop~\cite{ecosystemstwo}, that took place in May 2022, and the WLCG/HSF pre-CHEP workshop~\cite{prechep}, that took place in May 2023. The paper attempts to cover all the aspects of an analysis facility.

\end{abstract}

\vspace{\fill}

{\footnotesize
\centerline{\copyright~Licence \href{http://creativecommons.org/licenses/by/4.0/}{CC-BY-4.0}.}}
\vspace*{2mm}

\vskip 0.5cm
\textbf{Keywords}: High energy physics, analysis facilities, data analysis, scientific computing, data access, grid computing, federated identity management, analysis preservation, resource provisioning, HL-LHC

\end{titlepage}

\clearpage

\tableofcontents

\input{introduction}
\input{userperspective}
\input{computeresources}
\input{doma}
\input{aai}
\input{accelerators}
\input{preservation}
\input{monitoring}
\input{documentation}

\section{Summary}

Motivated by discussions at the HEP Software Foundations (HSF) Analysis Facilities forum, the Analysis Ecosystems II workshop, and the 26th International Conference on Computing in High Energy \& Nuclear Physics, this white paper documents HEP analysis infrastructure goals and the technologies that can be employed to achieve them. These goals should guide the development of such future infrastructures to enable the full scientific potential of HEP experiments in the next decades to be realized. We recommend the formation of cross-experiment working groups to follow-up these goals. In particular to collect the analysis workflows use-cases, tools for monitoring of user jobs and to discuss the integration of object stores into the experiments.

%\section*{Acknowledgements}

\sloppy
\raggedright
%\clearpage
\printbibliography[title={References},heading=bibintoc]

\clearpage

\end{document}
